### 2018-12-18 docs/manuscript.Rnw
first commit after restructuring

\documentclass[a4paper,hidelinks]{article}
\usepackage{bm,amsmath,amssymb,amsthm,amsfonts,graphics,graphicx,epsfig,rotating,caption,subcaption,natbib,appendix,titlesec,multicol,hyperref,verbatim,bbm,algorithm,algpseudocode,pgfplotstable,threeparttable, booktabs,mathtools,dsfont,parskip}

% vectors and matrices
\newcommand{\bbeta}{\bm{\beta}}
\newcommand{\btau}{\bm{\tau}}
\newcommand{\bsigma}{\bm{\sigma}}
\newcommand{\balpha}{\bm{\alpha}}
\newcommand{\bepsilon}{\bm{\epsilon}}
\newcommand{\bomega}{\bm{\omega}}
\newcommand{\bOmega}{\bm{\Omega}}
\newcommand{\bSigma}{\bm{\Sigma}}
\newcommand{\bkappa}{\bm{\kappa}}
\newcommand{\btheta}{\bm{\theta}}
\newcommand{\bmu}{\bm{\mu}}
\newcommand{\bpsi}{\bm{\psi}}
\newcommand{\blambda}{\bm{\lambda}}
\newcommand{\bLambda}{\bm{\Lambda}}
\newcommand{\bPsi}{\bm{\Psi}}
\newcommand{\bphi}{\bm{\phi}}
\newcommand{\Y}{\mathbf{Y}}
\newcommand{\y}{\mathbf{y}}
\newcommand{\X}{\mathbf{X}}
\newcommand{\x}{\mathbf{x}}
\newcommand{\I}{\mathbf{I}}
\newcommand{\bgamma}{\bm{\gamma}}
\newcommand{\T}{\mathbf{T}}
\newcommand{\Z}{\mathbf{Z}}
\newcommand{\W}{\mathbf{W}}
\renewcommand{\O}{\mathbf{O}}
\newcommand{\B}{\mathbf{B}}
\renewcommand{\H}{\mathbf{H}}
\newcommand{\U}{\mathbf{U}}
\newcommand{\Em}{\mathbf{E}}
\newcommand{\D}{\mathbf{D}}
\newcommand{\Vm}{\mathbf{V}}
\newcommand{\rv}{\mathbf{r}}
\newcommand{\A}{\mathbf{A}}
\newcommand{\Q}{\mathbf{Q}}
\newcommand{\Sm}{\mathbf{S}}
\newcommand{\0}{\bm{0}}
\newcommand{\tx}{\tilde{\mathbf{x}}}
\newcommand{\hy}{\hat{y}}
\newcommand{\tm}{\tilde{m}}
\newcommand{\tkappa}{\tilde{\kappa}}
\newcommand{\m}{\mathbf{m}}
\newcommand{\C}{\mathbf{C}}
\renewcommand{\v}{\mathbf{v}}
\newcommand{\g}{\mathbf{g}}
\newcommand{\s}{\mathbf{s}}
\renewcommand{\c}{\mathbf{c}}
\newcommand{\ones}{\mathbf{1}}
\newcommand{\R}{\mathbf{R}}
\newcommand{\rvec}{\mathbf{r}}
\newcommand{\Lm}{\mathbf{L}}

% functions and operators
\renewcommand{\L}{\mathcal{L}}
\newcommand{\G}{\mathcal{G}}
\renewcommand{\P}{\mathcal{P}}
\newcommand{\argmax}{\text{argmax} \,}
\newcommand{\expit}{\text{expit}}
\newcommand{\erfc}{\text{erfc}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\V}{\mathbb{V}}
\newcommand{\Cov}{\mathbb{C}\text{ov}}
\newcommand{\tr}{^{\text{T}}}
\newcommand{\diag}{\text{diag}}
\newcommand{\KL}{D_{\text{KL}}}
\newcommand{\trace}{\text{tr}}
\newcommand{\norm}[1]{\left\lVert #1 \right\rVert}

% distributions
\newcommand{\N}{\text{N}}
\newcommand{\TG}{\text{TG}}
\newcommand{\Bi}{\text{Binom}}
\newcommand{\PG}{\text{PG}}
\newcommand{\Mu}{\text{Multi}}
\newcommand{\GIG}{\text{GIG}}
\newcommand{\IGauss}{\text{IGauss}}
\newcommand{\Un}{\text{U}}

% syntax and readability
\renewcommand{\(}{\left(}
\renewcommand{\)}{\right)}
\renewcommand{\[}{\left[}
\renewcommand{\]}{\right]}

% settings
\graphicspath{{/Users/magnusmunch/Documents/OneDrive/PhD/project_cambridge/graphs/}}
\setlength{\evensidemargin}{.5cm} \setlength{\oddsidemargin}{.5cm}
\setlength{\textwidth}{15cm}
\title{Drug efficacy prediction in cell lines}
\date{\today}
\author{Magnus M. M\"unch$^{1,2}$\footnote{Correspondence to: \href{mailto:m.munch@vumc.nl}{m.munch@vumc.nl}}, Mark A. van de Wiel$^{1,3}$, Sylvia Richardson$^{3}$, \\ and Gwena{\"e}l G. R. Leday$^{3}$}

\begin{document}
	\maketitle

	\noindent
	1. Department of Epidemiology \& Biostatistics, Amsterdam Public Health research institute, VU University Medical Center, PO Box 7057, 1007 MB
	Amsterdam, The Netherlands\\
	2. Mathematical Institute, Leiden University, Leiden, The Netherlands \\
	3. MRC Biostatistics Unit, Cambridge Institute of Public Health, Cambridge, United Kingdom \\

	\begin{abstract}
		{...}
	\end{abstract}

	\noindent\textbf{Keywords}:

	\noindent\textbf{Software available from}:

	\section{Introduction}
	- motivation: drug response prediction from different omics data types is promising both for precision medicine and drug target discovery and relates to actual tumor response \cite[]{iorio_landscape_2016}.

	- problem: prediction of drug efficacy in cell lines is difficult (DREAM 7 challenge in \cite{costello_community_2014}).

	- example data: introduction (GDSC, CCLE, NCI-60)

	- feature information: prior experts, pathways, database summaries, previous experiments, type of feature (somatic mutations, copy number alterations, DNA methylation, and gene expression).

	- drugs information: molecular target, drug class, etc (ask Oslo).

	- current literature: review in \cite{azuaje_computational_2017}, reference Oslo, TANDEM \cite[]{aben_tandem:_2016}, DeltaNet \cite[]{noh_inferring_2016}, patient-derived tumor cell lines \cite[]{gao_high-throughput_2015}.

	- our solution: adaptive borrowing of information through empirical Bayes, guided by external feature and drug information.

	\section{Model}
	\subsection{Simultaneous equations model}\label{sec:SEM}
	- Gwen can deal with different number of features per equation, might solve missing drug-feature combination problem? Doesn't solve missing response-feature combination problem.

	- check literature for: a) conjugate version, b) similar models: "prior variance decomposition", "structured Bayesian regression"

	We have continuous efficacy measures $y_{id}$ for cell lines $i=1,\dots, n$, from tissues $t=1, \dots, T$, on drugs $d=1,\dots,D$. Throughout this paper we assume the $y_{id}$ to be centred per drug and let $\y_d = \begin{bmatrix} y_{1d} & \dots & y_{nd} \end{bmatrix} \tr$. We predict efficacy with molecular features $x_{ij}$, $j=1,\dots, p$, collected in $\x_i = \begin{bmatrix} x_{i1} & \dots & x_{ip} \end{bmatrix} \tr$. We have an \textit{a priori} partitioning of the omics features into $g=1, \dots, G$ non-overlapping groups, which might be informative for the problem at hand For convenience, we let $g(j)$ denote the group index of molecular feature $j$. Furthermore, we have characteristics of the drugs available in the form of `co-data' $\mathbf{C} = \begin{bmatrix} \mathbf{c}_1 & \dots & \mathbf{c}_D \end{bmatrix} \tr$. We code the tissue from which the cell lines were taken as a $T$ binary dummy variables 
	$$
	z_{it} = \begin{cases}
	1 & \text{if } \text{tissue}_i = t, \\
	0 & \text{otherwise},
	\end{cases}
	$$
	collected in $\mathbf{z}_i = \begin{bmatrix} z_{i1} & \dots & z_{iT} \end{bmatrix} \tr$.

	We assume that cell lines from the same tissue are related and model this relation through random effects $u_{td}$. We let the molecular feature effects $\beta_{jd}$ vary over the drugs and model drug efficacy as a linear combination of these two effects:
	\begin{subequations}\label{eq:linearmodel}
		\begin{align}
		y_{id} & = \beta_{0d} + \sum_{j=1}^p x_{ij} \beta_{jd} + \sum_{t=1}^T z_{it} u_{td} + \epsilon_{id} \\
		& = \beta_{0d} + \x_i \tr \bbeta_d + \mathbf{z}_i \tr \mathbf{u}_d + \epsilon_{id},\\
		\text{with } & \epsilon_{id} \sim \mathcal{N}(0, \sigma_d^2),
		\end{align}
	\end{subequations}
	where $p$-dimensional $\bbeta_d$ and $T$-dimensional $\bm{u}_d$ are the drug-specific omics feature and tissue effect vectors, respectively. Note that (\ref{eq:linearmodel}) gives rise to a system of $d=1, \dots, D$ linear simultaneous equations.

	\subsection{Bayesian prior model}
	We capture the uncertainty in the parameters through a Bayesian prior model. We assign a flat prior to the intercept terms $\beta_{0d}$ and integrate them out. The remaining parameters are endowed with the following priors:
	\begin{subequations}\label{eq:prior}
		\begin{align}
		\beta_{jd} & \sim \mathcal{N}_p (0, \phi^2_{g(j)} \gamma_d^2 \sigma_d^2), \\
		\mathbf{u}_{d} & \sim \mathcal{N}_T (\mathbf{0}, \bm{\Xi}_d \sigma_d^2),
		\end{align}
	\end{subequations}
	and hyper-priors:
	\begin{subequations}\label{eq:hyperprior}
		\begin{align}
			\sigma_d^{2} & \sim 1/\sigma_d^{3} \\
			\gamma_d^{2} & \sim \mathcal{IG}(\theta_d, \lambda), \\
			\bm{\Xi}_d & \sim \mathcal{W}^{-1}_T(\bm{\Omega}, \nu),
		\end{align}
	\end{subequations}
	where $\mathcal{IG}(\theta, \lambda)$ is an inverse Gaussian distribution with mean $\theta$ and shape $\lambda$, and $\mathcal{W}^{-1}_T(\bm{\Omega}, \nu)$ is an inverse Wishart distribution with $T \times T$-dimensional positive definite (PD) scale matrix $\bm{\Omega}$ and degrees of freedom $\nu$. Note that for $\nu > T + 1$, this inverse Wishart prior implies an expected prior covariance matrix $(\nu - T - 1)^{-1} \bm{\Omega}$ for the tissue effects, common to all drugs. Likewise, the molecular feature effects are related through the common hyper-prior for the $\gamma_d^2$.

	- Jeffrey's prior for variance $1/\sigma^{3/2}$ (a priori independent data mean and variance) or $1/\sigma^3$ (joint prior, but only mean and variance, not variance and betas).

	A few remarks on the choice of priors are justified here: many authors endow error variance components with vague gamma priors. \cite{gelman_prior_2006}, among others, advises against this practice. The degree of `vagueness' has a large influence on the posterior, while degree of `vagueness' is a difficult parameter to set. This influence is especially pronounced if the likelihood is relatively flat, as may be reasonably expected in the large $p$, small $n$ setting we are in. We therefore model the error variance with Jeffrey's prior \cite[]{jeffreys_invariant_1946}.

	Furthermore, we choose to model the $\gamma^2_d$ by an inverse Gaussian distribution, as has been suggested in \cite{fabrizi_specification_2016} and \cite{caron_sparse_2008}, because it allows to model the mean $\theta_d$ as a function of the drug covariates $\mathbf{c}_d$, as explained in Section \ref{sec:empiricalbayes}.

	- IG hyperprior used in finance \cite[]{barndorff-nielsen_normal_1997}, where it is called the NIG prior.
%	\begin{subequations}\label{eq:hyperprior}
%		\begin{align}
%		\sigma_d^{2} & \sim 1/\sigma_d^{3} \\
%		\log \gamma_d^{2} & \sim \mathcal{N}(\theta, \tau^2), \\
%		\bm{\Xi}_d & \sim \mathcal{W}^{-1}_T(\Vm, k),
%		\end{align}
%	\end{subequations}
%- try log-normal distribution for the $\gamma_d$'s. This is used in Example 10.4.1 in \cite{lunn_bugs_2012}, with justification same as ours: allows for modelling the variance as a function of covariates. Also used in examples 6.10 and 9.2 in \cite{abrams_bayesian_2004}.

%	\begin{subequations}\label{eq:hyperprior}
%		\begin{align}
%		\sigma_d^{-2} & \sim \Gamma(0.001, 0.001), \\
%		\gamma_d^{-2} & \sim \Gamma(k, \theta), \\
%		\bm{\Xi}_d & \sim \mathcal{W}^{-1}_T(\Vm, \tau),
%		\end{align}
%	\end{subequations}
%	where $g(j)$ denotes the group index of molecular feature $j$, $\Gamma(k, \theta)$ is a gamma distribution with shape $k$ and rate $\theta$, and $\mathcal{W}_T(\Vm, \tau)$ is a Wishart distribution with $T \times T$-dimensional scale matrix $\Vm$ and degrees of freedom $\tau$. Note that this Wishart prior implicates an expected prior covariance matrix $\tau \cdot \Vm$ for the tissue effects, common to all drugs.

	\section{Estimation}
	\subsection{Variational Bayes}\label{sec:variationalbayes}
	The posterior corresponding to the model described in (\ref{eq:linearmodel}), (\ref{eq:prior}), and (\ref{eq:hyperprior}) is not available in closed form. We therefore approximate the posterior by variational Bayes, where we force the posterior density $Q$ to factorise as:
	\begin{align}\label{eq:variationalposterior}
	p(\B,\mathbf{U},\mathbf{\sigma}^2, \bgamma^2,\bm{\Xi}_1, \dots, \bm{\Xi}_D) \approx Q(\cdot) = q(\B) \cdot q(\mathbf{U}) \cdot q(\mathbf{\sigma}^2) \cdot q(\bgamma^2) \cdot q(\bm{\Xi}_1, \dots, \bm{\Xi}_D),
	\end{align}
	where $\B = \begin{bmatrix} \bbeta_1 & \cdots & \bbeta_D \end{bmatrix}$ and $\U = \begin{bmatrix} \mathbf{u}_1 & \cdots & \mathbf{u}_D \end{bmatrix}$, are of dimensions $p \times D$ and $T \times D$, respectively. Under the factorisation in (\ref{eq:variationalposterior}), the marginal variational posteriors that minimise the Kullback-Leibler divergence of the true posterior to the variational Bayes approximation \cite[]{neal_view_1998}, are given by:
	\begin{subequations}\label{eq:variationalposterior2}
		\begin{align}
		q_{\B} (\B) & \overset{D}{=} \prod_{d=1}^{D} \mathcal{N}_p (\bmu_d, \bSigma_d), \\
		q_{\U}(\U) & \overset{D}{=} \prod_{d=1}^{D} \mathcal{N}_T (\mathbf{m}_d, \Sm_d), \\
		q_{\bgamma^{2}}(\bgamma^{2}) & \overset{D}{=} \prod_{d=1}^D \mathcal{GIG} \(-\frac{p+1}{2}, \frac{\lambda}{\theta_d^2}, \delta_{d} \), \\
		q_{\bsigma^{2}}(\bsigma^{2}) & \overset{D}{=} \prod_{d=1}^D \Gamma^{-1} \(\frac{n + p + T + 1}{2}, \zeta_{d} \), \\
		q_{\bm{\Xi}_1, \dots, \bm{\Xi}_D}(\bm{\Xi}_1, \dots, \bm{\Xi}_D) & \overset{D}{\propto} \prod_{d=1}^D \mathcal{W}_T^{-1} (\bm{\Psi}_d, \nu + 1).
		\end{align}
	\end{subequations}
	Here $\mathcal{GIG}(\cdot)$ and $\Gamma^{-1}(\cdot)$ denote the generalized inverse Gaussian and inverse gamma distributions, respectively. The full derivations may be found in Appendix \ref{app:derivations}.

	The parameters in (\ref{eq:variationalposterior2}) contain cyclic dependencies. We therefore iterate the following estimating equations until convergence to a local optimum:
	\begin{align*}
	\bSigma_d^{(h+1)} & = \frac{2\zeta_d^{(h)}}{n+p+T+1} \left\{\X \tr \X + a_d^{(h)} \cdot \diag (\phi_{g(j)}^{-2}) \right\}^{-1}, \\
	\bm{\mu}_d^{(h+1)} & = \left\{\X \tr \X + a_d^{(h)} \cdot \diag (\phi_{g(j)}^{-2}) \right\}^{-1} \X \tr ( \y_d - \Z \mathbf{m}_d^{(h)}), \\
	\Sm_d^{(h+1)} & = \frac{2\zeta_d^{(h)}}{n+p+T+1} \[ \Z \tr \Z + (\nu + 1) \cdot (\bm{\Psi}_d^{(h)})^{-1} \]^{-1}, \\
	\mathbf{m}_d^{(h+1)} & = \[ \Z \tr \Z + (\nu + 1) \cdot (\bm{\Psi}_d^{(h)})^{-1} \]^{-1} \Z \tr (\y_d - \X \bm{\mu}_d^{(h)}), \\
	\bm{\Psi}_d^{(h+1)} & = \frac{2\zeta_d^{(h)}}{n+p+T+1} \[ \Sm_d^{(h+1)} + \mathbf{m}_d^{(h+1)} (\mathbf{m}_d^{(h+1)}) \tr \] + \bm{\Omega}, \\
	\delta_d^{(h+1)} & =  \frac{n+p+T+1}{2\zeta_d^{(h)}} \sum_{j=1}^p \frac{(\mu_{jd}^{(h+1)})^2 + (\bSigma_d^{(h+1)})_{jj}}{\phi^2_{g(j)}} + \lambda, \\
	\zeta_d^{(h+1)} & = \frac{1}{2} \y_d \tr \y_d - \y_d \tr (\X \bmu_d^{(h+1)} + \Z \mathbf{m}_d^{(h+1)}) + (\bmu_d^{(h+1)}) \tr \X \tr \Z \mathbf{m}_d^{(h+1)} + \frac{1}{2} \trace (\X \tr \X \bSigma_d^{(h+1)}) \\
	& \,\,\,\,\,\,\,\,\,\, + \frac{1}{2} (\bmu_d^{(h+1)}) \tr \X \tr \X \bmu_d^{(h+1)} + \frac{1}{2} \trace (\Z \tr \Z \Sm_d^{(h+1)}) + \frac{1}{2} (\mathbf{m}_d^{(h+1)}) \tr \Z \tr \Z \mathbf{m}_d^{(h+1)} \\
	& \,\,\,\,\,\,\,\,\,\, + \frac{1}{2} a_d^{(h)} \sum_{j=1}^p \frac{(\mu_{jd}^{(h+1)})^2 + (\bSigma_d^{(h+1)})_{jj}}{\phi^2_{g(j)}} + \frac{1}{2} (\nu + 1) \cdot \trace \[ (\bm{\Psi}_d^{(h+1)})^{-1} \Sm_d^{(h+1)} \] \\
	& \,\,\,\,\,\,\,\,\,\, + \frac{1}{2} (\nu + 1) \cdot (\mathbf{m}_d^{(h+1)}) \tr (\bm{\Psi}_d^{(h+1)})^{-1} \mathbf{m}_d^{(h+1)},
	\end{align*}
	where
	$$
	a_d^{(h)} = \sqrt{\frac{\lambda}{\theta^2_d \delta_d^{(h)}}} \frac{K_{\frac{p - 1}{2}} \( \sqrt{\lambda \delta_d^{(h)}}/\theta_d \)}{K_{\frac{p+1}{2}} \( \sqrt{\lambda \delta_d^{(h)}}/\theta_d \)} + \frac{p+1}{\delta_d^{(h)}},
	$$
	and $K_{\alpha}(\cdot)$ is the modified Bessel function of the second kind.

	\subsection{Empirical Bayes}\label{sec:empiricalbayes}
	A full Bayesian model requires the specification of hyper-parameters $\bm{\eta} = \begin{bmatrix} (\bm{\phi}^2) \tr & \bm{\theta} \tr & \lambda & \text{vec}(\bm{\Omega}) \tr & \nu \end{bmatrix} \tr$. These are abstract and hard to interpret parameters for which we generally lack expert knowledge. They do, however, have a significant influence on the shape of the posterior distribution. We therefore propose to estimate the unspecified hyper parameters $\bm{\eta}$ by empirical Bayes. Simply put, empirical Bayes fits the prior model to the data and is, as such, objective (up to model specification).

	The canonical method for empirical Bayes is maximisation of the marginal likelihood with respect to the hyper parameters. In \cite{casella_empirical_2001} the marginal likelihood is maximised by an EM algorithm:
	$$
	\bm{\eta}^{(l+1)} = \underset{\bm{\eta}}{\argmax}\E_{\cdot | \Y} [\log p(\Y, \B, \U, \bm{\Xi}_1, \dots, \bm{\Xi}_D, \bgamma^2, \bsigma^2) | \bm{\eta}^{(l)}],
	$$
	where the expectation is with respect to the posterior. In our case, this posterior is not available in closed form, which renders the expectation difficult. While \cite{casella_empirical_2001} suggests to approximate the expectation by a Monte Carlo sample, we propose to use the variational Bayes approximation developed in Section \ref{sec:variationalbayes}:
	\begin{align}\label{eq:variationalbayes}
	\bm{\eta}^{(l+1)} & = \underset{\bm{\eta}}{\argmax}\E_{Q^{(l)}} [\log p(\Y, \B, \U, \bm{\Xi}_1, \dots, \bm{\Xi}_D, \bgamma^2, \bsigma^2)] \nonumber \\
	& = \underset{\bm{\eta}}{\argmax} \E_{Q^{(l)}} [\log \pi (\B | \bgamma^2, \bsigma^2)] + \E_{Q^{(l)}} [\log \pi (\bgamma^2)] + \E_{Q^{(l)}} [\log \pi (\bm{\Xi}_1, \dots, \bm{\Xi}_D)].
	\end{align}
	where now the expectation is with respect to the converged variational posterior $Q^{(l)}$. Inspection of (\ref{eq:variationalbayes}) learns us that the problem may be decomposed into three separate optimisation problems.

	\subsubsection{Group-specific variance components}
	The first sub-problem concerns the group-specific variance components of the molecular feature effects. In order to separate the overall variance of the molecular feature effects from the group-specific deviations, we impose $\prod_{g=1}^G (\phi_g^2)^{|\mathcal{G}_g|} = 1$. As a consequence, the overall level of shrinkage is determined by the $\gamma^2_d$, while the $\phi_g^2$ effectuate differential shrinkage of the groups. Because the $\phi_g^2$ are variance components, we additionally require $\phi^2_1, \dots, \phi^2_G> 0$, such that our first optimisation sub-problem becomes:
	\begin{align*}
	(\bm{\phi}^2)^{(l+1)} & = \underset{\bm{\phi}^2}{\argmax} \E_{Q^{(l)}} [\log \pi (\B | \bgamma^2, \bsigma^2)] \\
	& = \underset{\bm{\phi}^2}{\argmax} -\frac{D}{2} \sum_{g=1}^G |\G_g| \log \phi_g^2 - \frac{1}{2} \sum_{g=1}^G b_g^{(l)} \phi_g^{-2} \\
	& \text{subject to } \prod_{g=1}^G (\phi_g^2)^{|\G_g|} = 1 \text{ and } \phi^2_1, \dots, \phi^2_G> 0.
	\end{align*}
	where 
	$$
	b_g^{(l)}=\sum_{d=1}^D a_d^{(l)} \frac{n+p+T+1}{2\zeta_d^{(l)}} \sum_{j \in \mathcal{G}_g} \[ (\mu_{jd}^{(l)})^2 + (\bSigma_d^{(l)})_{jj} \].
	$$
	This is a convex optimisation problem and easily solved numerically.

	\subsubsection{Drug-specific variance components}
	The second sub-problem concerns the prior for the drug-specific variance components $\gamma_d^2$. As explained in Section \ref{sec:SEM}, we may have `covariates' $\mathbf{c}_d$ on the drugs available. We conjecture that these covariates may be informative for the omics effect size and therefore model the prior mean of the $\gamma_d^2$ as a function of these covariates: $\theta_d = (\mathbf{c}_d \tr \bm{\alpha})^{-1}$. Then, the empirical Bayes estimation of $\bm{\alpha}$ effectively boils down to an inverse Gaussian regression of the $\gamma_d^{-2}$ on the $\mathbf{c}_d$ \cite[]{fries_optimal_1986,whitmore_regression_1983}. The problem may be formulated as:
	\begin{align*}\label{eq:optproblem2}
	\lambda^{(l+1)}, \bm{\alpha}^{(l+1)} & = \underset{\lambda,\bm{\alpha}}{\argmax} \E_{Q^{(l)}} [\log \pi (\bgamma^2)] \\
	%	& = \underset{\lambda,\bm{\alpha}}{\argmax} \frac{D}{2} \log \lambda - \frac{\lambda}{2} \sum_{d=1}^D \[ \E(\gamma_d^2)(\mathbf{c} _d \tr \bm{\alpha})^2 - 2 \mathbf{c}_d \tr \bm{\alpha} + \E(\gamma_d^{-2}) \] \\
	& = \underset{\lambda,\bm{\alpha}}{\argmax} \frac{D}{2} \log \lambda - \frac{\lambda}{2} \bm{\alpha} \tr \mathbf{C} \tr \diag [\E_{Q^{(l)}}(\gamma_d^2)] \mathbf{C} \bm{\alpha} + \lambda \bm{\alpha} \tr \mathbf{C} \tr \mathbf{1} - \frac{\lambda}{2} \sum_{d=1}^D \E_{Q^{(l)}}(\gamma_d^{-2}),
	\end{align*}
	where $\mathbf{C} = \begin{bmatrix} \mathbf{c}_1 & \dots & \mathbf{c}_D \end{bmatrix} \tr$. The solutions to this problem are:
	\begin{align*}
	\bm{\alpha}^{(l+1)} & = \left\{ \mathbf{C} \tr \Vm^{(l)} \mathbf{C} \right\}^{-1} \mathbf{C} \tr \mathbf{1}, \\
	\lambda^{(l+1)} & = \[ \sum_{d=1}^D a_d^{(l)} - (\bm{\alpha}^{(l+1)}) \tr \mathbf{C} \tr \mathbf{1} \]^{-1} D,
	\end{align*}
	where 
	$$
	\Vm^{(l)} = \diag \[ \sqrt{\frac{\delta_d^{(l)}(\theta_d^{(l)})^2}{\lambda^{(l)}}} \frac{K_{\frac{p - 1}{2}} \( \sqrt{\lambda^{(l)} \delta_d^{(l)}}/\theta^{(l)}_d \)}{K_{\frac{p + 1}{2}} \( \sqrt{\lambda^{(l)} \delta_d^{(l)}}/\theta^{(l)}_d \)} \].
	$$

	- Think about parametrisation of $\mathbf{C}$. Idea: use overall mean and deviations from it in alpha, such that the starting value can be the mean and 0 for alpha. Or: use intercept and code the rest as dummy variables.

	\subsubsection{Tissue effect covariance}
	The third sub-problem pertains to the tissue effect covariance prior. To avoid overfitting when freely estimating the $\frac{T(T+1)}{2}$ elements in $\bm{\Omega}$ and $\nu$, we parametrise them as $\bm{\Omega}=  \tau [(\tau - \rho) \I_T + \rho \mathbf{1}_{T \times T}]$ and $\nu = \tau + T + 1$, respectively. The justification for this parametrisation is three-fold: (a) it implies an easy to interpret prior covariance matrix with $\tau$ on its diagonal and $\rho$ on its off-diagonals, (b) it reduces the number of parameters to  estimate to just two, and (c) simplifies the positive definite constraint on  $\bm{\Omega}$. To ensure positive definiteness and a proper covariance matrix, we impose the constraints $\tau > \max\[\rho - T \rho,\rho, 0\]$ (see Appendix \ref{app:priortissuecovariance}). The resulting third sub-problem is a bit more involved than the first two:
	\begin{align*}
	\tau^{(l + 1)}, \rho^{(l+1)} & = \underset{\tau, \rho}{\argmax} \frac{D (T^2 + T)}{2} \log \tau + \frac{DT}{2} \tau \log \tau + \frac{D(T+1)}{2} \log \[ \tau - (T - 1) \rho\] \\
	& \,\,\,\,\,\,\,\,\,\, + \frac{D}{2} \tau \log \[ \tau - (T - 1) \rho\] + \frac{D(T^2 - 1)}{2} \log(\tau - \rho) +\frac{D(T - 1)}{2} \tau  \log(\tau - \rho) \\
	& \,\,\,\,\,\,\,\,\,\, - D \sum_{t=1}^T \log \Gamma \( \frac{\tau + t + 1}{2} \) - w_1^{(l)} \tau^2 + w_2^{(l)} \tau \rho + w_3^{(l)} \tau, \\
	& \text{subject to } \tau > \max\[\rho - T \rho,\rho, 0\],
	\end{align*}
	where
	\begin{align*}
	w_1^{(l)} & =  \frac{\tau^{(l)} + T + 2}{2} \sum_{d=1}^D \trace \[ (\bm{\Psi}_d^{(l)})^{-1}\] > 0,\\
	w_2^{(l)} & = \frac{\tau^{(l)} + T + 2}{2} \sum_{d=1}^D \left\{\trace \[ (\bm{\Psi}_d^{(l)})^{-1}\] - \trace \[\mathbf{1}_{T \times T} \cdot (\bm{\Psi}_d^{(l)})^{-1}\] \right\} < w_1^{(l)},\\
	w_3^{(l)} & = \frac{1}{2} \[D \sum_{t=1}^T \psi\(\frac{\tau^{(l)} + t + 2}{2}\) - \sum_{d=1}^D \log |\bm{\Psi}_d^{(l)}| \].
	\end{align*}

	- use mathematica to check convexity

	- we use the ELBO for VB convergence and parameters for EB convergence

	- we might be lenient for VB convergence. The extreme case would be 1 VB iteration per EB iteration (corresponds to tolerance of infinity).

%	\begin{align*}
%	\tau^{(l+1)}, \rho^{(l+1)} & = \underset{\tau, \rho}{\argmax} \E_{Q^{(l)}} [\log \pi (\bm{\Xi}_1, \dots, \bm{\Xi}_D)] \\
%	& = \underset{\tau, \rho}{\argmax} w_1^{(l)} \tau - w_2^{(l)} \tau^2 - w_3^{(l)} \tau \rho - D \log \Gamma_T \(\frac{\tau}{2}\) + \frac{DT}{2} \tau \log \tau \\
%	& \,\,\,\,\,\,\,\,\,\,+ \frac{D}{2} \tau \log \[ \tau - (1-T)\rho \] + \frac{D(T-1)}{2} \tau \log (\tau - \rho), \\
%	& \text{subject to } \tau > \max\[\rho - T \rho,\rho, 0\] \text{ and } 1> |\rho| \geq 0,
%	\end{align*}
%	where 
%	\begin{align*}
%	w_1^{(l)} & = \frac{1}{2} \[D \psi_T\(\frac{\tau^{(l)} + 1}{2}\) + \sum_{d=1}^D \log |\bm{\Psi}_d^{(l)}| \], \\
%	w_2^{(l)} & = \frac{\tau^{(l)} + 1}{2} \sum_{d=1}^D \trace \(\bm{\Psi}_d^{(l)}\) > 0,\\
%	w_3^{(l)} & = \frac{\tau^{(l)} + 1}{2} \sum_{d=1}^D \trace \[(\mathbf{1}_{T \times T} - \I_T) \cdot \bm{\Psi}_d^{(l)}\],
%	\end{align*}
%	and $\Gamma_T(x)$ and $\psi_T(x)$ are the $T$-variate gamma and digamma functions, respectively. $\log \Gamma_T \(\frac{\tau}{2}\)$ is convex on the intervals $(0,1), \dots, (T-1, \infty)$

%	- we model the prior expectation of the covariance instead of the precision. Precisions are conditional on the other variables, which doesn't make sense in the tissue context.

%	The second sub problem estimates the molecular feature effect hyper prior:
%	\begin{align}\label{eq:optproblem2}
%	k^{(l+1)},\theta^{(l+1)} & = \underset{k,\theta}{\argmax} \E_{Q^{(l)}} [\log \pi (\bgamma^2)] \nonumber\\
%	& = \underset{k,\theta}{\argmax} D k \log \theta - D \log \Gamma (k) + e_1^{(l)} k - e_2^{(l)} \theta \\
%	& \text{subject to } k ,\theta > 0,
%	\end{align}
%	where
%	\begin{align*}
%	e_1^{(l)} & = D\psi\(\frac{p}{2} + k^{(l)}\) - \sum_{d=1}^D \log a_d^{(l)}, \\
%	e_2^{(l)} & = \(\frac{p}{2} + k^{(l)} \) \sum_{d=1}^D (a_d^{(l)})^{-1},
%	\end{align*}
%	and $\Gamma(x)$ and $\psi(x)$ are the gamma and digamma functions, respectively. This is again a convex problem (see Appendix \ref{app:convexity}), but more cumbersome to solve numerically due to the log-gamma function. By approximating the digamma function with $\phi(x) \approx \log x - 1/(2x)$, we get the following closed-form solutions:
%	\begin{align*}
%	k^{(l+1)} & = \frac{1}{2} \( \log e_2^{(l)} - D^{-1} e_1^{(l)} - \log D \)^{-1}, \\
%	\theta^{(l+1)} & = D k^{(l+1)} (e_2^{(l)})^{-1}.
%	\end{align*}

	\section{Simulations}

	- Hamming distance to assess variable selection

	\section{Data application}
	- check GSK open target EBI or ask Chris Wallace about it. Database that links targets to drugs and diseases. Might be useful for drugs co-data.

	- check NCI-60

	- how to transform the data and deal with missingness (ask Oslo and read papers published on GDSC and CCLE repositories)

	- what information is available on drugs (ask Oslo)

	- start with gene expression (one $g(j)$)

	\subsection{GDSC}

	\subsection{CCLE}

	\section{Discussion}
	- empirical Bayes gives extra information via estimation of the hyper-parameters, such as informativeness of the drug information (via model for the $\gamma_d$) and relatedness of the tissues (via $\bm{\Omega}$ and $\nu$)

	- regression on drug covariates very intuitive.

	- find more literature on inverse Gaussian prior for variances and modelling variances with covariates.

	\bibliographystyle{author_short3} 
	\bibliography{refs}

	\titleformat{\section}{\normalfont\Large\bfseries}{\appendixname~\thesection.}{1em}{}
	\begin{appendix}
		\section{Variational Bayes derivations}\label{app:derivations}
		In the following all expectations are with respect to the variational posterior $Q$.
		\begin{align*}
		\log q_{\B} (\B) & \propto \E_{\mathbf{U}, \bsigma^2} [\log \mathcal{L}(\Y | \B, \U, \bsigma^2 )] + \E_{\bgamma^2, \bsigma^2} [\log \pi (\B | \bgamma^2, \bsigma^2)] \\
		& \propto - \frac{1}{2} \sum_{d=1}^D \sum_{i=1}^n \E_{\mathbf{u}_d, \sigma_d^2} \[ \frac{(y_{id} - \x_i \tr \bbeta_d - \mathbf{z}_i \tr \mathbf{u}_d)^2}{\sigma_d^2} \] - \frac{1}{2} \sum_{d=1}^D \sum_{j=1}^p \phi_{g(j)}^{-2} \E_{\gamma^2_d, \sigma^2_d} \( \frac{\beta_{jd}^2}{\gamma_d^2 \sigma_d^2} \), \\
		q_{\B} (\B) & \overset{D}{=} \prod_{d=1}^{D} \mathcal{N}_p (\bmu_d, \bSigma_d), \\
		& \text{ with } \bSigma_d = \E(\sigma_d^{-2})^{-1} [\X \tr \X + \E(\gamma_d^{-2}) \cdot \diag (\phi_{g(j)}^{-2}) ]^{-1}, \\
		& \text{ and } \bmu_d = [\X \tr \X + \E(\gamma_d^{-2}) \cdot \diag (\phi_{g(j)}^{-2}) ]^{-1} \X \tr [\y_d - \Z \cdot \E(\mathbf{u}_d)].
		\end{align*}

		\begin{align*}
		\log q_{\U}(\U) & \propto \E_{\mathbf{B}, \bsigma^2} [\log \mathcal{L}(\Y | \B, \U, \bsigma^2 )] + \E_{\mathbf{\Xi}_1, \dots, \mathbf{\Xi}_D, \bsigma^2} [\log \pi (\U | \mathbf{\Xi}_1, \dots, \mathbf{\Xi}_D, \bsigma^2)] \\
		& \propto - \frac{1}{2} \sum_{d=1}^D \sum_{i=1}^n \E_{\bbeta_d, \sigma_d^2} \[ \frac{(y_{id} - \x_i \tr \bbeta_d - \mathbf{z}_i \tr \mathbf{u}_d)^2}{\sigma_d^2} \] - \frac{1}{2} \sum_{d=1}^D \E_{\bm{\Xi}_d, \sigma_d^2}\( \frac{\mathbf{u}_d \tr \bm{\Xi}_d^{-1} \mathbf{u}_d}{\sigma_d^2} \), \\
		q_{\U}(\U) & \overset{D}{=} \prod_{d=1}^{D} \mathcal{N}_T (\mathbf{m}_d, \Sm_d), \\
		& \text{ with } \Sm_d = \E(\sigma_d^{-2})^{-1} [\Z \tr \Z + \E(\bm{\Xi}_d^{-1}) ]^{-1}, \\
		& \text{ and } \mathbf{m}_d = [\Z \tr \Z + \E(\bm{\Xi}_d^{-1}) ]^{-1} \Z \tr [\y_d - \X \cdot \E(\bbeta_d)].
		\end{align*}

		\begin{align*}
		\log q_{\bgamma^{2}}(\bgamma^{2}) & \propto \E_{\B, \bsigma^2} [\log \pi (\B | \bgamma^{2}, \sigma^2)] + \log \pi (\bgamma^{2}) \\
		& \propto -\frac{p}{2} \sum_{d=1}^D \log \gamma_d^{2} - \frac{1}{2} \sum_{d=1}^D \sum_{j=1}^p \phi_{g(j)}^{-2} \E \( \frac{\beta_{jd}^2}{\sigma^{2}_d}\) \gamma_d^{-2} - \frac{3}{2} \sum_{d=1}^D \log \gamma_d^{2} \\
		& \,\,\,\,\,\,\,\,\,\, - \frac{\lambda}{2 \theta^2} \sum_{d=1}^D (\gamma_d^2 - \theta)^2 \gamma_d^{-2} \\
		& \propto \sum_{d=1}^D \left\{ \( -\frac{p + 1}{2} - 1\) \log \gamma_d^{2} - \frac{\lambda}{2\theta^2} \gamma_d^2 - \frac{1}{2}\[ \lambda + \E(\sigma_d^{-2}) \sum_{j=1}^p \frac{\E(\beta_{jd}^2)}{\phi_{g(j)}^{2}}\] \gamma_d^{-2} \right\}, \\
		q_{\bgamma^{2}}(\bgamma^{2}) & \overset{D}{=} \prod_{d=1}^D \mathcal{GIG} \(-\frac{p+1}{2}, \frac{\lambda}{\theta_d^2}, \delta_{d} \), \\
		& \text{with } \delta_{d} =\E(\sigma_d^{-2}) \sum_{j=1}^p \frac{\E(\beta_{jd})^2 + \V(\beta_{jd})}{\phi_{g(j)}^2} + \lambda.
		\end{align*}

%		\begin{align*}
%		\log q_{\bgamma^{-2}}(\bgamma^{-2}) & \propto \E_{\B, \bsigma^2} [\log \pi (\B | \bgamma^{-2}, \sigma^2)] + \log \pi (\bgamma^{-2}) \\
%		& \propto \frac{p}{2} \sum_{d=1}^D \log \gamma_d^{-2} - \frac{1}{2} \sum_{d=1}^D \sum_{j=1}^p \phi_{g(j)}^{-2} \E \( \frac{\beta_{jd}^2}{\sigma^{2}_d}\) \gamma_d^{-2} + (k - 1) \sum_{d=1}^D \log \gamma_d^{-2} - \theta \sum_{d=1}^D \gamma_d^{-2} \\
%		& = \sum_{d=1}^D \left\{ \( \frac{p}{2} + k - 1\) \log \gamma_d^{-2} - \[ \frac{\E(\sigma_d^{-2})}{2} \sum_{j=1}^p \frac{\E(\beta_{jd}^2)}{\phi_{g(j)}^{2}} + \theta \] \gamma_d^{-2} \right\}.
%		\end{align*}

%	\begin{align*}
%	q_{\bgamma^{-2}}(\bgamma^{-2}) & \overset{D}{=} \prod_{d=1}^D \Gamma \(\frac{p}{2} + k, a_{d} \), \\
%	& \text{ with } a_{d} =\frac{\E(\sigma_d^{-2})}{2} \sum_{j=1}^p \frac{\E(\beta_{jd})^2 + \V(\beta_{jd})}{\phi_{g(j)}^2} + \theta.
%	\end{align*}

		\begin{align*}
		\log q_{\bsigma^{2}}(\bsigma^{2}) & \propto \E_{\B,\mathbf{U}} [\log \mathcal{L}(\Y | \B, \U, \bsigma^2 )] + \E_{\B, \bgamma^2} [\log \pi (\B | \bgamma^2, \bsigma^2)] \\
		& \,\,\,\,\,\,\,\,\,\, + \E_{\mathbf{U}, \bm{\Xi}_1, \dots, \bm{\Xi}_D} [\log \pi (\mathbf{U} | \bm{\Xi}_1, \dots, \bm{\Xi}_D, \bsigma^2)] + \log \pi (\bsigma^{2}) \\
		& \propto -\frac{n}{2} \sum_{d=1}^D \log \sigma_d^{2} - \frac{1}{2} \sum_{d=1}^D \sum_{i=1}^n \E [(y_{id} - \x_i \tr \bbeta_d - \mathbf{z}_i \tr \mathbf{u}_d)^2] \sigma_d^{2} - \frac{p}{2} \sum_{d=1}^D \log \sigma_d^{-2} \\
		& \,\,\,\,\,\,\,\,\,\, - \frac{1}{2} \sum_{d=1}^D \sum_{j=1}^p \E( \gamma_d^{-2}) \frac{\E ( \beta_{jd}^2)}{\phi_{g(j)}^{2}} \sigma_d^{-2} - \frac{T}{2} \sum_{d=1}^D \log \sigma_d^{2} - \frac{1}{2} \sum_{d=1}^D \E(\mathbf{u}_d \tr \bm{\Xi}_d^{-1} \mathbf{u}) \sigma_d^{-2} \\
		& \,\,\,\,\,\,\,\,\,\, - \frac{3}{2} \sum_{d=1}^D \log \sigma^2_d \\
		& = -\sum_{d=1}^{D} \(\frac{n + p + T + 3}{2} \) \log \sigma_d^{2} - \sum_{d=1}^{D} \frac{1}{2}\Bigg\{ \sum_{i=1}^n \E [(y_{id} - \x_i \tr \bbeta_d - \mathbf{z}_i \tr \mathbf{u}_d)^2] \\
		& \,\,\,\,\,\,\,\,\,\, + \E(\gamma_d^{-2})\sum_{j=1}^p \frac{\E(\beta_{jd}^2)}{\phi_{g(j)}^2} + \E(\mathbf{u}_{d} \tr \bm{\Xi}_d^{-1} \mathbf{u}_d) \Bigg\} \sigma_d^{-2}, \\
		q_{\bsigma^{2}}(\bsigma^{2}) & \overset{D}{=} \prod_{d=1}^D \Gamma^{-1} \(\frac{n + p + T + 1}{2}, \zeta_{d} \), \\
		& \text{ with } \zeta_{d} = \frac{\y_d \tr \y_d}{2} - \y_d \tr [\X \E(\bbeta_d) + \Z \E(\mathbf{u}_d)] + \E(\bbeta_d \tr) \X \tr \Z \E (\mathbf{u}_d) \\
		& \,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\, + \frac{\trace [\X \tr \X \V(\bbeta_d)]}{2} + \frac{\E(\bbeta_d \tr) \X \tr \X \E(\bbeta_d)}{2} + \frac{\trace [\Z \tr \Z \V(\mathbf{u}_d)]}{2} \\
		& \,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\, + \frac{\E(\mathbf{u}_d \tr) \Z \tr \Z \E(\mathbf{u}_d)}{2} + \frac{\E(\gamma_d^{-2})}{2} \sum_{j=1}^p \frac{\E(\beta_{jd})^2 + \V(\beta_{jd})}{\phi_{g(j)}^2} \\
		& \,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,+ \frac{\E(\mathbf{u}_d \tr) \E(\bm{\Xi}_d^{-1}) \E(\mathbf{u}_d)}{2} + \frac{\trace [\E(\bm{\Xi}_d^{-1}) \V(\mathbf{u}_d)]}{2},
		\end{align*}

%		\begin{align*}
%		\log q_{\bsigma^{-2}}(\bsigma^{-2}) & \propto \E_{\B,\mathbf{U}} [\log \mathcal{L}(\Y | \B, \U, \bsigma^2 )] + \E_{\B, \bgamma^2} [\log \pi (\B | \bgamma^2, \bsigma^2)] \\
%		& \,\,\,\,\,\,\,\,\,\, + \E_{\mathbf{U}, \bm{\Xi}_1, \dots, \bm{\Xi}_D} [\log \pi (\mathbf{U} | \bm{\Xi}_1, \dots, \bm{\Xi}_D, \bsigma^2)] + \log \pi (\bsigma^{-2}) \\
%		& \propto \frac{n}{2} \sum_{d=1}^D \log \sigma_u^{-2} - \frac{1}{2} \sum_{d=1}^D \sum_{i=1}^n \E [(y_{id} - \x_i \tr \bbeta_d - \mathbf{z}_i \tr \mathbf{u}_d)^2] \sigma_d^{-2} + \frac{p}{2} \sum_{d=1}^D \log \sigma_d^{-2} \\
%		& \,\,\,\,\,\,\,\,\,\, - \frac{1}{2} \sum_{d=1}^D \sum_{j=1}^p \phi_{g(j)}^{-2} \E \( \frac{\beta_{jd}^2}{\gamma_d^{2}} \) \sigma_d^{-2} + \frac{T}{2} \sum_{d=1}^D \log \sigma_d^{-2} - \frac{1}{2} \sum_{d=1}^D \E(\mathbf{u}_d \tr \bm{\Xi}_d^{-1} \mathbf{u}) \sigma_d^{-2} \\
%		& \,\,\,\,\,\,\,\,\,\, - 0.999 \sum_{d=1}^D \log \sigma_d^{-2} - 0.001 \sum_{d=1}^D \sigma_d^{-2} \\
%		& = \sum_{d=1}^{D} \(\frac{n + p + T}{2} - 0.999 \) \log \sigma_d^{-2} - \sum_{d=1}^{D} \Bigg\{ \frac{\sum_{i=1}^n \E [(y_{id} - \x_i \tr \bbeta_d - \mathbf{z}_i \tr \mathbf{u}_d)^2]}{2} \\
%		& \,\,\,\,\,\,\,\,\,\, + \frac{\E(\gamma_d^{-2})}{2} \sum_{j=1}^p \frac{\E(\beta_{jd}^2)}{\phi_{g(j)}^2} + \frac{\E(\mathbf{u}_{d} \tr \bm{\Xi}_d^{-1} \mathbf{u}_d)}{2}  + 0.001 \Bigg\} \sigma_d^{-2}.
%		\end{align*}

%	\begin{align*}
%	q_{\bsigma^{-2}}(\bsigma^{-2}) & \overset{D}{=} \prod_{d=1}^D \Gamma \(\frac{n + p + T}{2} + 0.001, b_{d} \), \\
%	& \text{ with } b_{d} = \frac{\y_d \tr \y_d}{2} - \y_d \tr [\X \E(\bbeta_d) + \Z \E(\mathbf{u}_d)] + \E(\bbeta_d \tr) \X \tr \Z \E (\mathbf{u}_d) \\
%	& \,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\, + \frac{\trace [\X \tr \X \V(\bbeta_d)]}{2} + \frac{\E(\bbeta_d \tr) \X \tr \X \E(\bbeta_d)}{2} + \frac{\trace [\Z \tr \Z \V(\mathbf{u}_d)]}{2} \\
%	& \,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\, + \frac{\E(\mathbf{u}_d \tr) \Z \tr \Z \E(\mathbf{u}_d)}{2} + \frac{\E(\gamma_d^{-2})}{2} \sum_{j=1}^p \frac{\E(\beta_{jd})^2 + \V(\beta_{jd})}{\phi_{g(j)}^2} \\
%	& \,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,+ \frac{\E(\mathbf{u}_d \tr) \E(\bm{\Xi}_d^{-1}) \E(\mathbf{u}_d)}{2} + \frac{\trace [\E(\bm{\Xi}_d^{-1}) \V(\mathbf{u}_d)]}{2} + 0.001.
%	\end{align*}

		\begin{align*}
		\log q_{\bm{\Xi}_1, \dots, \bm{\Xi}_D}(\bm{\Xi}_1, \dots, \bm{\Xi}_D) & \propto \E_{\mathbf{U}, \bm{\sigma}^2} [\log \pi (\mathbf{U} | \bm{\Xi}_1, \dots, \bm{\Xi}_D, \bsigma^2)] + \log \pi (\bm{\Xi}_1, \dots, \bm{\Xi}_D) \\
		& \propto -\frac{1}{2} \sum_{d=1}^D \log |\bm{\Xi}_d| - \frac{1}{2} \sum_{d=1}^D \E \(\frac{\mathbf{u}_d \tr \bm{\Xi}_d^{-1} \mathbf{u}_d}{\sigma_d^2} \) \\
		& \,\,\,\,\,\,\,\,\,\, - \frac{k + T + 1}{2} \sum_{d=1}^D \log |\bm{\Xi}_d| - \frac{1}{2} \sum_{d=1}^D \trace (\bm{\Omega} \bm{\Xi}_d^{-1}) \\
		& = -\frac{\tau + T + 2}{2} \sum_{d=1}^D \log |\bm{\Xi}_d| - \frac{1}{2} \sum_{d=1}^D \trace \left\{ \[\E(\sigma_d^{-2}) \E( \mathbf{u}_d \mathbf{u}_d \tr) + \bm{\Omega}\] \bm{\Xi}_d^{-1} \right\}, \\
		q_{\bm{\Xi}_1, \dots, \bm{\Xi}_D}(\bm{\Xi}_1, \dots, \bm{\Xi}_D) & \overset{D}{\propto} \prod_{d=1}^D \mathcal{W}_T^{-1} (\bm{\Psi}_d, \nu + 1),  \\ 
		& \text{ with } \bm{\Psi}_d =\E(\sigma^{-2}_d) \[\V(\mathbf{u}_d) + \E(\mathbf{u}_d) \E( \mathbf{u}_d \tr) \] + \bm{\Omega}.
		\end{align*}

%		\section{Convexity}\label{app:convexity}
%		For (\ref{eq:optproblem2}) to be a convex problem, we need the Hessian $\mathbf{H}$ of the objective function to be positive definite, i.e., $\forall \mathbf{z} \in \mathbb{R}^2: \mathbf{z} \tr \mathbf{H} \mathbf{z} > 0$. We have
%		\begin{align*}
%		\mathbf{z} \tr \mathbf{H} \mathbf{z} = \frac{D}{\theta^2 \psi^{(1)}(k)} \( z_1 \theta \psi^{(1)}(k) - z_2 \)^2 + z_2^2 \( k \psi^{(1)}(k)  - 1 \),
%		\end{align*}
%		with $\psi^{(1)}(k)$ the trigamma function. Since the trigamma function is positive for all $k > 0$, and we have $\theta, k > 0$, we need:
%		$$
%		k \psi^{(1)}(k)> 1.
%		$$
%		Using Lemma 1 in \cite{guo_refinements_2013} we can show that for $k > 0$:
%		$$
%		k \psi^{(1)}(k) > 1 + \frac{1}{2k},
%		$$
%		such that the objective function is convex.

		\section{Prior tissue covariance constraints}\label{app:priortissuecovariance}
%		The empirical Bayes objective function of the third sub-optimisation problem in \ref{sec:empiricalbayes} is:
%		\begin{align*}
%		\E_{Q^{(l)}}[\log \pi (\mathbf{\Xi}_1, \dots, \mathbf{\Xi}_D)] & \propto \frac{D}{2} \nu \log |\bm{\Omega}| - \frac{D T \log 2}{2} \nu - D \log \Gamma_T \(\frac{\nu}{2}\) + \frac{\nu}{2}\sum_{d=1}^D \E_{Q^{(l)}} (\log | \mathbf{\Xi}_d^{-1}|) \\
%		& \,\,\,\,\,\,\,\,\,\, - \frac{1}{2} \sum_{d=1}^D \trace [ \bm{\Omega} \cdot \E_{Q^{(l)}}(\mathbf{\Xi}_d^{-1})] \\
%		& = \frac{D}{2} \nu \log |\bm{\Omega}| - D \log \Gamma_T\(\frac{\nu}{2}\) - \frac{\nu^{(l)} + 1}{2} \sum_{d=1}^D \trace [ \bm{\Omega} \cdot ( \mathbf{\Psi}_d^{(l)})^{-1}] \\
%		& \,\,\,\,\,\,\,\,\,\, + \frac{\nu}{2} \sum_{d=1}^D \[ \psi_T\(\frac{\nu^{(l)} + 1}{2}\) - \log |\mathbf{\Psi}_d^{(l)}| \],
%		\end{align*}
%		which is maximised under constraints $\nu > T + 1$ and $\bm{\Omega}$ positive definite (PD), to ensure a proper inverse Wishart distribution.

		We parametrise the prior covariance as a matrix with $\tau$ on the diagonal and $\rho$ on its off-diagonals. To that end we set $\nu = \tau + T + 1$ and the $T \times T$-dimensional matrix
		$$
		\bm{\Omega} = \tau \begin{bmatrix}
		\tau &  & \rho \\
		& \ddots &  \\
		\rho & & \tau
		\end{bmatrix},
		$$
		such that $\E_{\pi(\mathbf{\Xi}_d)} (\mathbf{\Xi}_d) = (\nu - T - 1)^{-1} \bm{\Omega}$ is of the desired form. To ensure a non-degenerate prior distribution we require $\tau > 0$ and $\bm{\Omega}$ PD. We use $\tau > 0$ to note that a PD $\bm{\Omega}$ implies positive roots $\lambda$ of the characteristic polynomial:
		\begin{align*}
		\left| \begin{bmatrix}
		\tau &  & \rho \\
		& \ddots &  \\
		\rho & & \tau
		\end{bmatrix} -\lambda \I_T \right| = (\tau - \lambda - \rho)^{T-1}[\tau - \lambda - (1-T)\rho] = 0.
		\end{align*}
		For $T > 1$, the solutions are:
		$$
		(\lambda = \tau - \rho) \lor (\lambda = T \rho - \rho + \tau),
		$$
		which are positive if the following condition holds: $\tau > \max(\rho,\rho - T \rho)$. We combine this with $\tau > 0$, to arrive at the full proper prior constraint: $\tau > \max(0,\rho,\rho - T \rho)$.

%		The resulting empirical Bayes step is now:
%		\begin{align*}
%		\tau^{(l + 1)}, \rho^{(l+1)} & = \underset{\tau, \rho}{\argmax} \frac{D (T^2 + T)}{2} \log \tau + \frac{DT}{2} \tau \log \tau + \frac{D(T+1)}{2} \log \[ \tau - (T - 1) \rho\] \\
%		& \,\,\,\,\,\,\,\,\,\, + \frac{D}{2} \tau \log \[ \tau - (T - 1) \rho\] + \frac{D(T^2 - 1)}{2} \log(\tau - \rho) +\frac{D(T - 1)}{2} \tau  \log(\tau - \rho) \\
%		& \,\,\,\,\,\,\,\,\,\, - D \sum_{t=1}^T \log \Gamma \( \frac{\tau + t + 1}{2} \) - w_1^{(l)} \tau^2 + w_2^{(l)} \tau \rho + w_3^{(l)} \tau, \\
%		& \text{subject to } \tau > \max\[\rho - T \rho,\rho, 0\],
%		\end{align*}
%		with
%		\begin{align*}
%		w_1^{(l)} & =  \frac{\tau^{(l)} + T + 2}{2} \sum_{d=1}^D \trace \[ (\bm{\Psi}_d^{(l)})^{-1}\] > 0,\\
%		w_2^{(l)} & = \frac{\tau^{(l)} + T + 2}{2} \sum_{d=1}^D \left\{\trace \[ (\bm{\Psi}_d^{(l)})^{-1}\] - \trace \[\mathbf{1}_{T \times T} \cdot (\bm{\Psi}_d^{(l)})^{-1}\] \right\} < w_1^{(l)},\\
%		w_3^{(l)} & = \frac{1}{2} \[D \sum_{t=1}^T \psi\(\frac{\tau^{(l)} + t + 2}{2}\) - \sum_{d=1}^D \log |\bm{\Psi}_d^{(l)}| \].
%		\end{align*}

		\section{Model without tissue effects}
		\subsection{Model}
		In order to investigate the inverse Gaussian prior model and its estimation, we consider the simpler model:
		\begin{align*}
		y_{id} | \bm{\beta}_d, \sigma_d^2 & \sim \mathcal{N} (\x_i \tr \bm{\beta}_d, \sigma_d^2), \\
		\beta_{jd} | \gamma_d^2, \sigma_d^2 & \sim \mathcal{N} (0, \sigma_d^2 \gamma_d^2), \\
		\gamma_d^2 & \sim \mathcal{IG}(\theta_d, \lambda), \\
		\sigma_d^2 & \sim 1/\sigma_d^3.
		\end{align*}
		So in other words, we ignore the tissue effect and fix all $\phi_g^2$ to one. 
		\subsection{Variational Bayes}
		The corresponding variational Bayes posterior is given by:
		\begin{align*}
		q(\bm{\beta}) & \overset{D}{=} \prod_{d=1}^D \mathcal{N}_p (\bm{\mu}_d, \bm{\Sigma}_d), \\
		q(\bm{\gamma}^2) & \overset{D}{=} \prod_{d=1}^D \mathcal{GIG}\(-\frac{p+1}{2}, \frac{\lambda}{\theta_d^2}, \delta_d\), \\
		q(\bm{\sigma}^2) & \overset{D}{=} \prod_{d=1}^D \Gamma^{-1} \(\frac{n + p + 1}{2}, \zeta_d\).
		\end{align*}
		We update the parameters until convergence:
		\begin{align*}
		\bm{\Sigma}_d^{(h+1)} & = \frac{2 \zeta_d^{(h)}}{n+p+1} (\X \tr \X + a_d^{(h)} \I)^{-1}, \\
		\bm{\mu}_d^{(h+1)} & = (\X \tr \X + a_d^{(h)} \I)^{-1} \X \tr \y_d, \\
		\delta_d^{(h+1)} & = \frac{n+p+1}{2 \zeta_d^{(h)}} [\trace(\bm{\Sigma}_d^{(h+1)}) + (\bm{\mu}_d^{(h+1)}) \tr \bm{\mu}_d^{(h+1)}] + \lambda, \\ \zeta_d^{(h+1)} & = \frac{a_d^{(h+1)}}{2} \[\mathbf{y}_d \tr \mathbf{y}_d -2 \mathbf{y}_d \tr \X \bm{\mu}_d^{(h+1)} + \trace [ (\X \tr \X + \I_p) \bm{\Sigma}_d^{(h+1)}] + (\bm{\mu}_d^{(h+1)}) \tr (\X \tr \X + \I_p)\bm{\mu}_d^{(h+1)}\],
		\end{align*}
		with
		$$
		a_d^{(h)} = \sqrt{\frac{\lambda}{\theta_d^2 \delta_d^{(h)}}} \frac{K_{\frac{p - 1}{2}}\(\sqrt{\lambda \delta_d^{(h)}} /\theta_d \)}{K_{\frac{p + 1}{2}}\(\sqrt{\lambda \delta_d^{(h)}} /\theta_d \)} + \frac{p+1}{\delta_d^{(h)}}.
		$$

		\subsection{Empirical Bayes}
		Again we model $\theta_d = (\mathbf{c}_d \tr \bm{\alpha})^{-1}$ to arrive at the following empirical Bayes updates:
		\begin{align*}
		\bm{\alpha}^{(l+1)} & = \left\{ \mathbf{C} \tr \Vm^{(l)} \mathbf{C} \right\}^{-1} \mathbf{C} \tr \mathbf{1}, \\
		\lambda^{(l+1)} & = \[ \sum_{d=1}^D a_d^{(l)} - (\bm{\alpha}^{(l+1)}) \tr \mathbf{C} \tr \mathbf{1} \]^{-1} D,
		\end{align*}
		where 
		$$
		\Vm^{(l)} = \diag \[ \sqrt{\frac{\delta_d^{(l)}(\theta_d^{(l)})^2}{\lambda^{(l)}}} \frac{K_{\frac{p - 1}{2}} \( \sqrt{\lambda^{(l)} \delta_d^{(l)}}/\theta^{(l)}_d \)}{K_{\frac{p + 1}{2}} \( \sqrt{\lambda^{(l)} \delta_d^{(l)}}/\theta^{(l)}_d \)} \].
		$$

		\subsection{Evidence lower bound}
		We monitor convergence through the evidence lower bound (ELBO):
		\begin{align*}
		\text{ELBO} & (Q^{(l)}) \propto \frac{1}{2} \sum_{d=1}^D \log |\bm{\Sigma}_d^{(l)}| - \frac{n + p + 1}{2} \sum_{d=1}^D \log \zeta_d^{(l)} + \sum_{d=1}^D \log K_{\frac{p+1}{2}}\(\sqrt{\lambda^{(l)} \delta_d^{(l)}}/\theta_d^{(l)}\) \\
		& - \frac{n+p+1}{4} \sum_{d=1}^D (\zeta_d^{(l)})^{-1} \[ \y_d \tr \y_d - 2\y_d \tr \X \bm{\mu}_d^{(l)} + (\bm{\mu}_d^{(l)}) \tr \X \tr \X \bm{\mu}_d^{(l)} + \trace ( \X \tr \X \bm{\Sigma}_d^{(l)})\] \\
		& + \frac{D(p+3)}{4} \log \lambda^{(l)} + \lambda^{(l)} \sum_{d=1}^D (\theta_d^{(l)})^{-1} - \frac{p+1}{2} \sum_{d=1}^D \log \theta_d^{(l)} - \frac{p+1}{4} \sum_{d=1}^D \log \delta_d^{(l)} \\
		& + \frac{1}{2} \sum_{d=1}^D \[ \delta_d^{(l)} - \lambda^{(l)} - \frac{n+p+1}{2\zeta_d^{(l)}} (\bm{\mu}_d^{(l)}) \tr \bm{\mu}_d^{(l)} - \frac{n+p+1}{2\zeta_d^{(l)}} \trace (\bm{\Sigma}_d^{(l)}) \] a_d^{(l)}. 
		\end{align*}

%		\begin{align*}
%		y_{id} | \bm{\beta}_d & \sim \mathcal{N} (\x_i \tr \bm{\beta}_d, \sigma_d^2), \\
%		\beta_{jd} | \gamma_d^2 & \sim \mathcal{N} (0, \sigma_d^2 \gamma_d^2), \\
%		\gamma_d^2 & \sim \mathcal{IG}(\theta_d, \lambda).
%		\end{align*}
%		So in other words, we ignore the tissue effect, fix the $\sigma_d^2$, and set all $\phi_g^2$ to one. The corresponding variational Bayes posterior is given by:
%		\begin{align*}
%		q(\bm{\beta}) & \overset{D}{=} \prod_{d=1}^D \mathcal{N}_p (\bm{\mu}_d, \bm{\Sigma}_d), \\
%		q(\bm{\gamma}^2) & \overset{D}{=} \prod_{d=1}^D \mathcal{GIG}\(-\frac{p+1}{2}, \frac{\lambda}{\theta_d^2}, \delta_d\).
%		\end{align*}
%		We update the parameters until convergence:
%		\begin{align*}
%		\bm{\Sigma}_d^{(h+1)} & = \sigma_d^2 (\X \tr \X + a_d^{(h)} \I)^{-1}, \\
%		& \text{with } a_d^{(h)} = \sqrt{\frac{\lambda}{\theta_d^2 \delta_d^{(h)}}} \frac{K_{\frac{p - 1}{2}}\(\sqrt{\lambda \delta_d^{(h)}} /\theta_d \)}{K_{\frac{p + 1}{2}}\(\sqrt{\lambda \delta_d^{(h)}} /\theta_d \)} + \frac{p+1}{\delta_d^{(h)}}, \\
%		\bm{\mu}_d^{(h+1)} & = (\X \tr \X + a_d^{(h)} \I)^{-1} \X \tr \y_d, \\
%		\delta_d^{(h+1)} & = \sigma_d^{-2} [\trace(\bm{\Sigma}_d^{(h+1)}) + (\bm{\mu}_d^{(h+1)}) \tr \bm{\mu}_d^{(h+1)}] + \lambda.
%		\end{align*}
%		Again we model $\theta_d = (\mathbf{c}_d \tr \bm{\alpha})^{-1}$ to arrive at the following empirical Bayes updates:
%		\begin{align*}
%		\bm{\alpha}^{(l+1)} & = \left\{ \mathbf{C} \tr \Vm^{(l)} \mathbf{C} \right\}^{-1} \mathbf{C} \tr \mathbf{1}, \\
%		\lambda^{(l+1)} & = \[ \sum_{d=1}^D a_d^{(l)} - (\bm{\alpha}^{(l+1)}) \tr \mathbf{C} \tr \mathbf{1} \]^{-1} D,
%		\end{align*}
%		where 
%		$$
%		\Vm^{(l)} = \diag \[ \sqrt{\frac{\delta_d^{(l)}(\theta_d^{(l)})^2}{\lambda^{(l)}}} \frac{K_{\frac{p - 1}{2}} \( \sqrt{\lambda^{(l)} \delta_d^{(l)}}/\theta^{(l)}_d \)}{K_{\frac{p + 1}{2}} \( \sqrt{\lambda^{(l)} \delta_d^{(l)}}/\theta^{(l)}_d \)} \].
%		$$
%		We monitor convergence through the evidence lower bound (ELBO):
%		\begin{align*}
%		\text{ELBO} & (Q^{(l)}) \propto \sum_{d=1}^D \sigma_d^{-2} \y_d \tr \X \bm{\mu}_d^{(l)} - \frac{1}{2} \sum_{d=1}^D \sigma_d^{-2} \trace ( \X \tr \X \bm{\Sigma}_d^{(l)}) - \frac{1}{2} \sum_{d=1}^D \sigma_d^{-2} (\bm{\mu}_d^{(l)}) \tr \X \tr \X \bm{\mu}_d^{(l)} \\
%		& + \frac{1}{2} \sum_{d=1} \log |\bm{\Sigma}_d^{(l)}| + \frac{D(p+3)}{4} \log \lambda^{(l)} + \lambda^{(l)} \sum_{d=1}^D (\theta_d^{(l)})^{-1} - \frac{p+1}{2} \sum_{d=1}^D \log \theta_d^{(l)} \\
%		& - \frac{p+1}{4} \sum_{d=1}^D \log \delta_d^{(l)} + \sum_{d=1}^D \log K_{\frac{p+1}{2}}\(\sqrt{\lambda^{(l)} \delta_d^{(l)}}/\theta_d^{(l)}\)\\
%		& + \frac{1}{2} \sum_{d=1}^D \left(\rule{0cm}{1.2cm}\right. \[ \delta_d^{(l)} - \lambda^{(l)} - \sigma_d^{-2} (\bm{\mu}_d^{(l)}) \tr \bm{\mu}_d^{(l)} - \sigma_d^{-2} \trace (\bm{\Sigma}_d^{(l)}) \]  \\
%		& \,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\, \cdot \[\sqrt{\frac{\lambda^{(l)}}{\delta_d^{(l)} (\theta^{(l)}_d)^2}} \frac{K_{\frac{p - 1}{2}}\(\frac{\sqrt{\lambda^{(l)} \delta_d^{(l)}}}{\theta^{(l)}_d} \)}{K_{\frac{p + 1}{2}}\(\frac{\sqrt{\lambda^{(l)} \delta_d^{(l)}}}{\theta^{(l)}_d} \)} + \frac{p+1}{\delta_d^{(l)}}\] \left. \rule{0cm}{1.2cm}\right).
%		\end{align*}

		- Different options for convergence. We may monitor parameters, ELBO, or a combination. We may also iterate a fixed number of times, especially for the VB.

		\subsection{Efficient computation}
		We need the following quantities during the EB iterations: $\zeta_d$, $a_d$, $\delta_d$, $\trace (\bm{\Sigma}_d)$, $\trace (\X \tr \X \bm{\Sigma}_d)$, $\bm{\mu}_d \tr \bm{\mu}_d$, and $\bm{\mu}_d \tr \X \tr \X \bm{\mu}_d$. In addition, we need $\log|\bm{\Sigma}|$ and $\mathbf{y}_d \tr \X \bm{\mu}_d$ to monitor the ELBO. The first three are just scalar operations. The rest is easily calculated using the SVD $\X = \mathbf{U} \mathbf{D} \mathbf{V} \tr$:
		\begin{align*}
		\trace (\bm{\Sigma}_d) & = \frac{2\zeta_d}{n + p + 1} \[\sum_{i=1}^n (v_i^2 + a_d)^{-1} + \max (p - n, 0)\cdot a_d^{-1}\], \\
		\trace (\X \tr \X \bm{\Sigma}_d) & = \frac{2\zeta_d}{n + p + 1} \sum_{i=1}^n v_i^2(v_i^2 + a_d)^{-1}, \\
		\bm{\mu}_d \tr \bm{\mu}_d & = \sum_{i=1}^n v_i^2 [(\mathbf{U} \tr \mathbf{y}_d)_i]^2/(v_i^2 + a_d)^2, \\
		\bm{\mu}_d \tr \X \tr \X \bm{\mu}_d & = \sum_{i=1}^n v_i^4 [(\mathbf{U} \tr \mathbf{y}_d)_i]^2/(v_i^2 + a_d)^2, \\
		\log|\bm{\Sigma}| & \propto p \log \zeta_d - \sum_{i=1} \log (v_i^2 + a_d) - \max(p - n, 0) \cdot \log a_d, \\
		\mathbf{y}_d \tr \X \bm{\mu}_d & = \sum_{i=1}^n v_i^2 [(\mathbf{U} \tr \mathbf{y}_d)_i]^2/(v_i^2 + a_d).
		\end{align*}
		where $v_i$ are the singular values of $\X$. The SVD and $\mathbf{U} \tr \mathbf{y}_d$ are $\mathcal{O}(pn^2)$ and $\mathcal{O}(n^2)$ once at the start of the algorithm. The rest of the calculations involve just $n$ scalar multiplications and/or additions.

		\subsection{Simulations}
		We simulate from a correctly specified model. We assume random $y_{id}$, $\x_i$, and $\bm{\beta}_d$, while we fix $\mathbf{C}$, $\gamma_d^2$, $\sigma_d^2$ and $\bm{\alpha}$. We estimate the $\bm{\mu}_d$, $\bm{\Sigma}_d$, $\delta_d$, $\bm{\alpha}$ and $\lambda$, while we fix the $\sigma_d^2$ to their true values.

		We set $n=100$, $p=200$, and $D=10$. We consider 5 evenly sized classes of drugs, such that $\mathbf{c}_{kd} = \mathbbm{1}(k = \text{class}_d)$ and $\begin{bmatrix} \mathbf{c}_1 & \cdots & \mathbf{c}_D \end{bmatrix} \tr$. From these the $\gamma_d^2 = (\mathbf{c}_d \tr \bm{\alpha})^{-1}$ are created. Next, we simulate $\bm{\beta}_{jd} \sim \mathcal{N} (0, \sigma^2_d \gamma_d^2)$. We fix the mean signal-to noise ratio $\overline{\text{SNR}}_d$, such that the predictor data variance is $s^2 = \V(\mathbf{x}_i) = \overline{\text{SNR}}_d / (\overline{\gamma^2_d} p)$. The last step consists of simulating the data by $\mathbf{x}_{ij} \sim \mathcal{N}(0,s^2)$ and $y_{id} = \mathcal{N} (\mathbf{x}_i \tr \bm{\beta}_d, \sigma_d^2)$. The remaining parameters are set as follows: $\bm{\alpha} = \begin{bmatrix} 1 & \dots & 5 \end{bmatrix} \tr$, $\sigma_d^2 = 1$ for all $d$, and $\overline{\text{SNR}}_d = 5$.

		We compare the results to a model where the $\gamma_d^2$ are drawn from a class-specific  inverse Gamma distribution; a generalisation of the models in \cite{kpogbezan_empirical_2017} and \cite{leday_gene_2017}:
		\begin{align*}
		y_{id} | \bm{\beta}_d ,\sigma_d^2 & \sim \mathcal{N} (\x_i \tr \bm{\beta}_d, \sigma_d^2), \\
		\beta_{jd} | \gamma_d^2 ,\sigma_d^2 & \sim \mathcal{N} (0, \sigma_d^2 \gamma_d^2), \\
		\gamma_d^2 & \sim \Gamma^{-1} (a_{\text{class}(d)}, b_{\text{class}(d)}), \\
		\sigma_d^2 & \sim \Gamma^{-1} (0.001, 0.001),
		\end{align*}
		where $\text{class}(c)$ denotes the class of drug $d$. This allows for more flexible empirical Bayes estimation of the $a_d$ and $b_d$, but, as a consequence, increases the risk of overfitting. Additionally, it does not allow to include continuous covariates on the drugs.

		To account for the random data and parameter generation, we repeat the simulation 100 times and present the results in Table...

		\subsection{Starting values}
%		Starting value for $\sigma_d^2$ as in \cite{chipman_bart:_2010}. They consider a scaled-Chi squared distribution for the error variance with degrees of freedom 3 and scale chosen such that the $q*100$th percentile matches the error variance in the data $s^2(\mathbf{y}_d)$. As starting value for $\sigma_d^2$ we take the mode of this distribution: $\hat{\sigma}_d^2 = 3a/5$, with $a$ the solution to:
%		$$
%		\Gamma\(\frac{3}{3}\) \[1 - q - F(3a/(2s^2(\mathbf{y}_d)); 3/2, 1) \] = 0,
%		$$
%		where $F(x; 3/2, 1)$ is the CDF of a gamma function with shape $3/2$ and scale $1$.  In \cite{moran_variance_2018}, REFERENCE to Ročková and George (2018), and \cite{chipman_bart:_2010}, $q=0.9$ is suggested. We find that $q \in ( 0.9, 0.95 )$ gives good results.

		We generate starting values as follows. Consider the $\sigma_d^2$ and $\gamma_d^2$ as fixed and estimate them by MML of the regular ridge model:
		$$
		\hat{\sigma}_d^2, \hat{\gamma}_d^2= \underset{\sigma_d^2,\gamma_d^2}{\argmax} \int_{\bm{\beta}} p(\mathbf{y}_d | \bm{\beta}, \sigma_d^2) p(\bm{\beta} | \gamma_d^2, \sigma_d^2) d\bm{\beta}.
		$$
		The regular ridge model is conjugate, so MML is relatively simple. Next we estimate a common inverse Gaussian prior for the $\gamma_d^2$ by considering the estimates to be samples from the prior:
		\begin{align*}
		\hat{\theta} & = n^{-1} \sum_{d=1}^D \hat{\gamma}_d^2 / n, \\
		\hat{\lambda} & = n \left\{\sum_{d=1}^D \[ (\hat{\gamma}_d^2)^{-1} - \hat{\theta}^{-1}\]\right\}^{-1}.
		\end{align*}
		We estimate the mode of the $\hat{m}=\hat{\gamma}_d^2$ by kernel density estimation and equate it to the theoretical mode of the generalized inverse Gaussian distribution with the estimated $\hat{\theta}$ and $\hat{\lambda}$. We solve to find a common $\delta$:
		$$
		\hat{\delta} = \frac{\hat{m}^2 \hat{\lambda}}{\hat{\theta}^2} + (p + 3) \hat{m}.
		$$
		Lastly, we consider the $\hat{\sigma}^2_d$ fixed samples from the inverse Gamma distribution to estimate a common scale:
		$$
		\hat{\zeta} = \frac{D(n + p + 1)}{2 \sum_{d=1}^D[(\hat{\sigma}_d^2)^{-1}]}.
		$$

		\subsection{Independent error variance model parameters}
		Model:
		\begin{align*}
		y_{id} | \bm{\beta}_d, \sigma_d^2 & \sim \mathcal{N} (\x_i \tr \bm{\beta}_d, \sigma_d^2), \\
		\beta_{jd} | \gamma_d^2 & \sim \mathcal{N} (0, \gamma_d^2), \\
		\gamma_d^2 & \sim \mathcal{IG}((\mathbf{c}_d \tr \bm{\alpha})^{-1}, \lambda), \\
		\sigma_d^2 & \sim 1/\sigma_d^3.
		\end{align*}
		Variational posterior:
		\begin{align*}
		q(\bm{\beta}) & \overset{D}{=} \prod_{d=1}^D \mathcal{N}_p (\bm{\mu}_d, \bm{\Sigma}_d), \\
		q(\bm{\gamma}^2) & \overset{D}{=} \prod_{d=1}^D \mathcal{GIG}\(-\frac{p+1}{2}, \lambda (\mathbf{c}_d \tr \bm{\alpha})^2, \delta_d\), \\
		q(\bm{\sigma}^2) & \overset{D}{=} \prod_{d=1}^D \Gamma^{-1} \(\frac{n + 1}{2}, \zeta_d\).
		\end{align*}
		Parameter updates:
		\begin{align*}
		\bm{\Sigma}_d^{(h+1)} & = \(\frac{n+1}{2 \zeta_d^{(h)}}\X \tr \X + a_d^{(h)} \I_p\)^{-1}, \\
		\bm{\mu}_d^{(h+1)} & = \frac{n+1}{2 \zeta_d^{(h)}} \bm{\Sigma}_d^{(h+1)} \X \tr \y_d, \\
		\delta_d^{(h+1)} & = \trace(\bm{\Sigma}_d^{(h+1)}) + (\bm{\mu}_d^{(h+1)}) \tr \bm{\mu}_d^{(h+1)} + \lambda, \\ \zeta_d^{(h+1)} & = \frac{1}{2} \[\mathbf{y}_d \tr \mathbf{y}_d -2 \mathbf{y}_d \tr \X \bm{\mu}_d^{(h+1)} + \trace (\X \tr \X \bm{\Sigma}_d^{(h+1)}) + (\bm{\mu}_d^{(h+1)}) \tr \X \tr \X \bm{\mu}_d^{(h+1)}\],
		\end{align*}
		with $a_d^{(h)}$ and the empirical Bayes updates $\bm{\alpha}^{(l+1)}$ and $\lambda^{(l+1)}$ the same as before.
		\section{Ratios of modified Bessel functions}
		Ratios of modified Bessel functions of the second kind $K_{\alpha - 1}(x)/K_{\alpha}(x)$ are prone to under- and overflow for large $\alpha$. In our case, $\alpha$ increases linearly with $p$. Since $p$ may be large, this causes numerical issues in the calculation of various quantities. We alleviate the numerical issues through the following.

		We let $n_1=p/2$ and $n_2=(p-1)/2$ and use the well-known recursive relation:
		$$
		K_{\alpha}(x) = K_{\alpha - 2}(x) + \frac{2(\alpha - 1)}{2} K_{\alpha- 1}(x),
		$$
		to rewrite the ratio:
		\begin{align*}
		\frac{K_{\frac{p - 1}{2}}(x)}{K_{\frac{p + 1}{2}}(x)} = 
		\begin{cases}
		\Big( \dots \Big( \( 1 + \frac{2 \cdot 1 - 1}{x} \)^{-1} + \frac{2 \cdot 2-1}{x}\Big)^{-1} + \dots + \frac{2 \cdot n_1 - 1}{x}\Big)^{-1}, & \text{for } p \text{ even}, \\
		\Big( \dots \Big( \( \frac{K_0(x)}{K_1(x)} + \frac{2}{x} \cdot 1 \)^{-1} + \frac{2}{x} \cdot 2 \Big)^{-1} + \dots + \frac{2}{x} \cdot n_2 \Big)^{-1}, & \text{for } p \text{ odd}.
		\end{cases}
		\end{align*} 
		The ratio $K_0(x)/K_1(x)$ is well-behaved, such that the ratio may be computed without numerical issues.

		\section{General model}
		We have drug covariates $\mathbf{c}_d$ and omics feature covariates $\mathbf{z}_j$. The model is now:
		\begin{align*}
		y_{id} | \bm{\beta}_d, \sigma_d^2 & \sim \mathcal{N} (\x_i \tr \bm{\beta}_d, \sigma_d^2), \\
		\beta_{jd} | \tau_j^2, \gamma_d^2, \sigma_d^2 & \sim \mathcal{N} (0, \sigma_d^2 \gamma_d^2 \tau_j^2), \\
		\gamma_d^2 & \sim \mathcal{IG}\((\mathbf{c}_d \tr \bm{\alpha})^{-1}, \lambda\), \\
		\tau_j^2 & \sim \mathcal{IG}\((\mathbf{z}_j \tr \bm{\theta})^{-1}, \nu\), \\
		\sigma_d^2 & \sim 1/\sigma_d^3.
		\end{align*}
		Variational posterior:
		\begin{align*}
		q(\bm{\beta}) & \overset{D}{=} \prod_{d=1}^D \mathcal{N}_p (\bm{\mu}_d, \bm{\Sigma}_d), \\
		q(\bm{\gamma}^2) & \overset{D}{=} \prod_{d=1}^D \mathcal{GIG}\(-\frac{p+1}{2}, \lambda (\mathbf{c}_d \tr \bm{\alpha})^2, \delta_d\), \\
		q(\bm{\tau}^2) & \overset{D}{=} \prod_{j=1}^p \mathcal{GIG}\(-\frac{D+1}{2}, \nu (\mathbf{z}_j \tr \bm{\theta})^2, \eta_j\), \\
		q(\bm{\sigma}^2) & \overset{D}{=} \prod_{d=1}^D \Gamma^{-1} \(\frac{n + p + 1}{2}, \zeta_d\).
		\end{align*}
		Variational parameters:
		\begin{align*}
		\bm{\Sigma}_d^{(h+1)} & = \frac{2 \zeta_d^{(h)}}{n+p+1} \[\X \tr \X + a_d^{(h)} \cdot \diag(b_j^{(h)})\]^{-1}, \\
		\bm{\mu}_d^{(h+1)} & = \[\X \tr \X + a_d^{(h)} \cdot \diag(b_j^{(h)})\]^{-1} \X \tr \y_d, \\
		\delta_d^{(h+1)} & = \frac{n+p+1}{2 \zeta_d^{(h)}} \left\{\trace \[\diag(b_j^{(h)}) \bm{\Sigma}_d^{(h+1)} \] + (\bm{\mu}_d^{(h+1)}) \tr \diag(b_j^{(h)}) \bm{\mu}_d^{(h+1)}\right\} + \lambda, \\ 
		\eta_j^{(h+1)} & = \sum_{d=1}^D \frac{a_d^{(h)}(n+p+1)}{2 \zeta_d^{(h)}} \[ (\bmu_{jd}^{(h + 1)})^2 + (\bSigma_d^{(h + 1)})_{jj} \] + \nu, \\ 
		\zeta_d^{(h+1)} & = \frac{1}{2} \Big( \mathbf{y}_d \tr \mathbf{y}_d -2 \mathbf{y}_d \tr \X \bm{\mu}_d^{(h+1)} + a_d^{(h+1)} \trace \left\{ \[\X \tr \X + \diag(b_j^{(h + 1)})\] \bm{\Sigma}_d^{(h+1)}\right\} \\
		& \,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\, + a_d^{(h+1)} (\bm{\mu}_d^{(h+1)}) \tr \[\X \tr \X + \diag(b_j^{(h + 1)})\]\bm{\mu}_d^{(h+1)}\Big),
		\end{align*}
		with
		\begin{align*}
		a_d^{(h)} & = \sqrt{\frac{\lambda (\mathbf{c}_d \tr \bm{\alpha})^2}{\delta_d^{(h)}}} \frac{K_{\frac{p-1}{2}} \( \sqrt{\lambda (\mathbf{c}_d \tr \bm{\alpha})^2 \delta_d^{(h)}} \)}{K_{\frac{p+1}{2}} \( \sqrt{\lambda (\mathbf{c}_d \tr \bm{\alpha})^2 \delta_d^{(h)}} \)} + \frac{p+1}{\delta_d^{(h)}} \text{ and} \\
		b_j^{(h)} & = \sqrt{\frac{\nu (\mathbf{z}_j \tr \bm{\theta})^2}{\eta_j^{(h)}}} \frac{K_{\frac{D-1}{2}} \( \sqrt{\nu (\mathbf{z}_j \tr \bm{\theta})^2 \eta_j^{(h)}} \)}{K_{\frac{D+1}{2}} \( \sqrt{\nu (\mathbf{z}_j \tr \bm{\theta})^2 \eta_j^{(h)}} \)} + \frac{D+1}{\eta_j^{(h)}}.
		\end{align*}
		Hyper-parameter updates:
		\begin{align*}
		\bm{\alpha}^{(l+1)} & = \left[ \mathbf{C} \tr \diag ( e_d^{(l)} ) \mathbf{C} \right]^{-1} \mathbf{C} \tr \mathbf{1}, \\
		\lambda^{(l+1)} & = \[ \sum_{d=1}^D a_d^{(l)} - (\bm{\alpha}^{(l+1)}) \tr \mathbf{C} \tr \mathbf{1} \]^{-1} D, \\
		\bm{\theta}^{(l+1)} & = \left[ \mathbf{Z} \tr \diag ( f_j^{(l)} ) \mathbf{Z} \right]^{-1} \mathbf{Z} \tr \mathbf{1}, \\
		\nu^{(l+1)} & = \[ \sum_{j=1}^p b_j^{(l)} - (\bm{\theta}^{(l+1)}) \tr \mathbf{Z} \tr \mathbf{1} \]^{-1} p,
		\end{align*}
		where 
		\begin{align*}
		e_d^{(l)} & = \sqrt{\frac{\delta_d^{(l)}}{\lambda^{(l)}(\mathbf{c}_d \tr \bm{\alpha}^{(l)})^2}} \frac{K_{\frac{p - 1}{2}} \( \sqrt{\lambda^{(l)} (\mathbf{c}_d \tr \bm{\alpha}^{(l)})^2 \delta_d^{(l)}} \)}{K_{\frac{p + 1}{2}} \( \sqrt{\lambda^{(l)} (\mathbf{c}_d \tr \bm{\alpha}^{(l)})^2 \delta_d^{(l)}} \)} \text{ and} \\
		f_j^{(l)} & = \sqrt{\frac{\eta_j^{(l)}}{\nu^{(l)}(\mathbf{z}_j \tr \bm{\theta}^{(l)})^2}} \frac{K_{\frac{D - 1}{2}} \( \sqrt{\nu^{(l)} (\mathbf{z}_j \tr \bm{\theta}^{(l)})^2 \eta_j^{(l)}} \)}{K_{\frac{D + 1}{2}} \( \sqrt{\nu^{(l)} (\mathbf{z}_j \tr \bm{\theta}^{(l)})^2 \eta_j^{(l)}} \)}.
		\end{align*}

	\end{appendix}

\end{document}

%		\section{Efficient computation}
%		Let $\X = \U \D \Vm \tr$ and $\Z=\U^* \D^* (\Vm^*) \tr$ be the singular value decomposition of $\X$ and $\Z$, respectively. Filling in all the expectations, we have the following iterative estimating equations:
%		\begin{align*}
%		\bSigma_d^{(l+1)} & = a_3 b_{3d}^{(l)} \Vm (\D^2 + a_1 b_{1d}^{(l)} \I_n)^{-1} \Vm \tr, \\
%		\Sm_d^{(l+1)} & = a_3 b_{3d}^{(l)} \Vm^* [(\D^*)^2 + a_2 b_{2d}^{(l)} \I_T]^{-1} (\Vm^*) \tr, \\
%		\bmu_d^{(l+1)} & = a_3 b_{3d}^{(l)} \Vm (\D + a_1 b_{1d}^{(l)} \D^{-1})^{-1} \U \tr (\y_d - \Z \mathbf{m}_d^{(l)}), \\
%		\mathbf{m}_d^{(l+1)} & = a_3 b_{3d}^{(l)} \Vm^* [\D^* + a_2 b_{2d}^{(l)} (\D^*)^{-1}]^{-1} (\U^*) \tr (\y_d - \X \bmu_d^{(l + 1)}), \\
%		b_{1d}^{(l+1)} & = \Bigg\{\frac{a_3 b_{3d}^{(l)}}{2} \Bigg[ (\y_d - \Z \mathbf{m}_d^{(l)}) \tr \U (\D + a_1 b_{1d}^{(l)} \D^{-1})^{-2} \U \tr (\y_d - \Z \mathbf{m}_d^{(l)}) \\
%		\,\,\,\,\,\,\,\,\,\, & + a_3 b_{3d}^{(l)} \sum_{i=1}^n (d_i^2 + a_1 b_{1d}^{(l)})^{-1} \Bigg] + \theta_1^{-1} \Bigg\}^{-1}, \\
%		b_{2d}^{(l+1)} & = \Bigg(\frac{a_3 b_{3d}^{(l)}}{2} \Bigg\{ (\y_d - \X \bmu_d^{(l+1)}) \tr \U^* [\D^* + a_2 b_{2d}^{(l)} (\D^*)^{-1}]^{-2} (\U^*) \tr (\y_d - \X \bmu_d^{(l+1)}) \\
%		\,\,\,\,\,\,\,\,\,\, & + a_3 b_{3d}^{(l)} \sum_{i=1}^n [(d^*_i)^2 + a_2 b_{2d}^{(l)}]^{-1} \Bigg\} + \theta_2^{-1} \Bigg)^{-1}, \\
%		b_{3d}^{(l+1)} & = \frac{\y_d \tr \y_d}{2}  - \y_d \tr (\X \bmu_d^{(l+1)} + \Z \mathbf{m}_d^{(l+1)}) + (\bmu_d^{(l+1)}) \tr \X \tr \Z \mathbf{m}_d^{(l+1)} + \frac{\sum_{i=1}^n \(d_i^2 - \frac{d_i^4}{d_i^2 + a_1 b_1^{(l+1)}}\)}{2 a_1 b_{1d}^{(l + 1)}} \\
%		\,\,\,\,\,\,\,\,\,\, & + \frac{(\bmu_d^{(l)}) \tr \X \tr \X \bmu_d^{(l)}}{2} + \frac{\sum_{t=1}^T \[(d^*_t)^2 - \frac{(d^*_t)^4}{(d^*_t)^2 + a_2 b_2^{(l+1)}}\]}{2 a_2 b_{2d}^{(l + 1)}} + \frac{(\mathbf{m}_d^{(l)}) \tr \Z \tr \Z \mathbf{m}_d^{(l)}}{2} \\
%		& \,\,\,\,\,\,\,\,\,\, + \frac{a_1 b_{1d}^{(l+1)} (\y_d - \Z \mathbf{m}_d^{(l)}) \tr \U (\D + a_1 b_{1d}^{(l)} \D^{-1})^{-2} \U \tr (\y_d - \Z \mathbf{m}_d^{(l)})}{2} \\
%		& \,\,\,\,\,\,\,\,\,\, + \frac{a_1 b_{1d}^{(l+1)} a_3 b_{3d}^{(l)} \sum_{i=1}^n (d_i^2 + a_1 b_{1d}^{(l)})^{-1} }{2} \\
%		& \,\,\,\,\,\,\,\,\,\, + \frac{a_2 b_{2d}^{(l+1)} (\y_d - \X \bmu_d^{(l+1)}) \tr \U^* [\D^* + a_2 b_{2d}^{(l)} (\D^*)^{-1}]^{-2} (\U^*) \tr (\y_d - \X \bmu_d^{(l+1)})}{2} \\
%		& \,\,\,\,\,\,\,\,\,\, + \frac{a_1 b_{1d}^{(l+1)} a_3 b_{3d}^{(l)} \sum_{i=1}^n [(d^*_i)^2 + a_2 b_{2d}^{(l)}]^{-1}}{2},
%		\end{align*}
%		where $(\D^2 + a_1 b_{1d}^{(l)})^{-1}$, $[(\D^*)^2 + a_2 b_{2d}^{(l)}]^{-1}$, $(\D + a_1 b_{1d}^{(l)} \D^{-1})^{-1}$, and $[\D^* + a_2 b_{2d}^{(l)} (\D^*)^{-1}]^{-1}$ are diagonal matrices with entries $(d_t^2 + a_1 b_{1d}^{(l)})^{-1}$, $i=1,\dots,n$, $[(d_t^*)^2 + a_2 b_{2d}^{(l)}]^{-1}$, $t=1,\dots,T$, $(d_i + a_1 b_{1d}^{(l)} / d_i)^{-1}$, $i=1, \dots, n$, and $(d^*_t + a_2 b_{2d}^{(l)} / d^*_t)^{-1}$, $t=1 , \dots, T$, respectively, and hence easily calculated once the singular values are known.

### 2018-12-18 code/R/functions.R

@@ -1,0 +1,482 @@
+# computes ratio besselK(x, p/2 - 1/2)/besselK(x, p/2 + 1/2) (tested)
+ratio.besselK <- function(x, p) {
+  out <- sapply(x, function(s) {
+    if((p %% 2)==0) {
+      value <- 1
+      for(i in 1:(p/2)) {
+        value <- 1/(value + 2*(i - 0.5)/s)
+      }
+    } else {
+      value <- besselK(s, 1, expon.scaled=TRUE)/besselK(s, 0, expon.scaled=TRUE)
+      for(i in 1:(p/2 + 0.5)) {
+        value <- 1/(value + 2*(i - 1)/s)
+      }
+    }
+    return(value)})
+  return(out)
+}
+
+# one VB update with sigma^2 stochastic (not tested)
+vb.update <- function(zetaold, aold, lambda, theta, x, xxt, xtx, y) {
+  p <- ncol(x)
+  D <- length(length(zetaold))
+  n <- nrow(y)
+  
+  # sort of efficient matrix inversion
+  mat <- lapply(c(1:D), function(d) {
+    mat <- xxt
+    diag(mat) <- diag(mat) + aold[d]
+    mat <- -(t(x) %*% solve(mat) %*% x)/aold[d]
+    diag(mat) <- diag(mat) + 1/aold[d]
+    return(mat)})
+  
+  # vb parameters
+  Sigma <- lapply(c(1:D), function(d) {2*zetaold[d]*mat[[d]]/(n + p + 1)})
+  mu <- sapply(c(1:D), function(d) {mat[[d]] %*% t(x) %*% y[, d]})
+  delta <- sapply(c(1:D), function(d) {
+    0.5*(n + p + 1)*(sum(diag(Sigma[[d]])) + sum(mu[, d]^2))/
+      zetaold[d] + lambda})
+  zeta <- sapply(c(1:D), function(d) {
+    0.5*aold[d]*(sum(y[, d]^2) - 2*t(y[, d]) %*% x %*% mu[, d] + 
+                   sum((xtx + diag(p))*Sigma[[d]]) + 
+                   t(mu[, d]) %*% (xtx + diag(p)) %*% mu[, d])})
+  
+  # calculate the ratio of two BesselK functions
+  ratio <- ratio_besselK_cpp(sqrt(lambda*delta/theta^2), p)
+  
+  # additional VB and EB parameters used in next iteration
[…]

### 2018-12-18 code/R/test1_gdsc.R

@@ -1,0 +1,16 @@
+path.data <- "/Users/magnusmunch/Documents/OneDrive/PhD/project_cambridge/data/"
+
+
+data <- read.table(paste(path.data, "v17.3_fitted_dose_response.csv", sep=""),
+                   header=TRUE, sep=",", stringsAsFactors=FALSE)
+data2 <- aggregate(LN_IC50 ~ CELL_LINE_NAME + DRUG_NAME, 
+                   data=data[data$DRUG_ID > 1000, ], FUN=mean, na.rm=TRUE)
+data3 <- reshape(data2, timevar="DRUG_NAME", idvar="CELL_LINE_NAME", 
+                 direction="wide")
+
+hist(colMeans(is.na(data)), xlab="Proportion missing", 
+     main="Proportion missing in features")
+hist(rowMeans(is.na(data)), xlab="Proportion missing", 
+     main="Proportion missing in observations")
+
+data4 <- na.omit(data3[, apply(data3, 2, function(x) {sum(is.na(x))}) <= 150])

### 2018-12-18 code/R/test_igaussian.R
independent beta and sigma^2 vb update implemented and tested

@@ -52,66 +52,12 @@ svd.x <- svd(x)
 sv <- svd.x$d
 uty <- t(svd.x$u) %*% y # size n x D
 yty <- colSums(y*y)
-test1 <- single.vb.update(new.vb[2, d], new.vb[3, d], new.eb$lambda, 
-                          new.eb$theta[d], sv, n, p, uty[, d], yty[d])
+test1 <- single.vb.update.ind(new.vb[2, d], new.vb[3, d], new.eb$lambda, 
+                              new.eb$theta[d], sv, n, p, uty[, d], yty[d])
 
 # naive update
-test2 <- single.naive.update(new.vb[2, d], new.vb[3, d], new.eb$lambda, 
-                             new.eb$theta[d], n, p, x, y[, d])
-
-# naive update
-all.equal(test1[-c(5, 6)], test2[-c(5, 6)])
-test2 <- c(delta=delta, zeta=zeta, a=a, v=v, elbo=NA, elbo.const=NA)
-cbind(test1, test2)
-
-elbo.seq <- numeric(0)
-alpha.seq <- numeric(0)
-lambda.seq <- numeric(0)
-theta.seq <- numeric(0)
-ebiter <- 20
-vbiter <- 2
-for(i in c(1:ebiter)) {
-  
-  for(j in c(1:vbiter)) {
-    new.vb <- sapply(c(1:D), function(d) {
-      single.vb.update(new.vb[2, d], new.vb[3, d], new.eb$lambda, 
-                       new.eb$theta[d], sv, n, p, uty[, d], yty[d])})
-  }
-  
-  new.eb <- eb.update(new.vb[4, ], new.vb[3, ], new.vb[6, ], C, D)
-  elbo.seq <- rbind(elbo.seq, new.eb$elbo)
-  alpha.seq <- rbind(alpha.seq, new.eb$alpha)
-  lambda.seq <- rbind(lambda.seq, new.eb$lambda)
-  theta.seq <- rbind(theta.seq, new.eb$theta)
-  
-}
-
-fit1.igauss1 <- est.igauss(x, y, C, 
-                           control=list(epsilon.eb=1e-3, epsilon.vb=1e-3, 
-                                        maxit.eb=20, maxit.vb=2, trace=TRUE), 
-                           init=list(alpha=c(1, rep(0, nclass - 1)), 
-                                     lambda=0.001, a=rep(0.001, D), 
[…]

### 2018-12-20 docs/manuscript.Rnw
independent beta and sigma^2 vb update implemented and tested
created supplementary material with first simulation results

@@
 	- IG hyperprior used in finance \cite[]{barndorff-nielsen_normal_1997}, where it is called the NIG prior.
-%	\begin{subequations}\label{eq:hyperprior}
-%		\begin{align}
-%		\sigma_d^{2} & \sim 1/\sigma_d^{3} \\
-%		\log \gamma_d^{2} & \sim \mathcal{N}(\theta, \tau^2), \\
-%		\bm{\Xi}_d & \sim \mathcal{W}^{-1}_T(\Vm, k),
-%		\end{align}
-%	\end{subequations}
-%- try log-normal distribution for the $\gamma_d$'s.
-This is used in Example 10.4.1 in \cite{lunn_bugs_2012}, with justification same as ours: allows for modelling the variance as a function of covariates.
-Also used in examples 6.10 and 9.2 in \cite{abrams_bayesian_2004}.
-
-%	\begin{subequations}\label{eq:hyperprior}
-%		\begin{align}
-%		\sigma_d^{-2} & \sim \Gamma(0.001, 0.001), \\
-%		\gamma_d^{-2} & \sim \Gamma(k, \theta), \\
-%		\bm{\Xi}_d & \sim \mathcal{W}^{-1}_T(\Vm, \tau),
-%		\end{align}
-%	\end{subequations}
-%	where $g(j)$ denotes the group index of molecular feature $j$, $\Gamma(k, \theta)$ is a gamma distribution with shape $k$ and rate $\theta$, and $\mathcal{W}_T(\Vm, \tau)$ is a Wishart distribution with $T \times T$-dimensional scale matrix $\Vm$ and degrees of freedom $\tau$.
-Note that this Wishart prior implicates an expected prior covariance matrix $\tau \cdot \Vm$ for the tissue effects, common to all drugs.
 
@@
 	\lambda^{(l+1)}, \bm{\alpha}^{(l+1)} & = \underset{\lambda,\bm{\alpha}}{\argmax} \E_{Q^{(l)}} [\log \pi (\bgamma^2)] \\
-	%	& = \underset{\lambda,\bm{\alpha}}{\argmax} \frac{D}{2} \log \lambda - \frac{\lambda}{2} \sum_{d=1}^D \[ \E(\gamma_d^2)(\mathbf{c} _d \tr \bm{\alpha})^2 - 2 \mathbf{c}_d \tr \bm{\alpha} + \E(\gamma_d^{-2}) \] \\
 	& = \underset{\lambda,\bm{\alpha}}{\argmax} \frac{D}{2} \log \lambda - \frac{\lambda}{2} \bm{\alpha} \tr \mathbf{C} \tr \diag [\E_{Q^{(l)}}(\gamma_d^2)] \mathbf{C} \bm{\alpha} + \lambda \bm{\alpha} \tr \mathbf{C} \tr \mathbf{1} - \frac{\lambda}{2} \sum_{d=1}^D \E_{Q^{(l)}}(\gamma_d^{-2}),
@@
 
-%	\begin{align*}
-%	\tau^{(l+1)}, \rho^{(l+1)} & = \underset{\tau, \rho}{\argmax} \E_{Q^{(l)}} [\log \pi (\bm{\Xi}_1, \dots, \bm{\Xi}_D)] \\
-%	& = \underset{\tau, \rho}{\argmax} w_1^{(l)} \tau - w_2^{(l)} \tau^2 - w_3^{(l)} \tau \rho - D \log \Gamma_T \(\frac{\tau}{2}\) + \frac{DT}{2} \tau \log \tau \\
-%	& \,\,\,\,\,\,\,\,\,\,+ \frac{D}{2} \tau \log \[ \tau - (1-T)\rho \] + \frac{D(T-1)}{2} \tau \log (\tau - \rho), \\
-%	& \text{subject to } \tau > \max\[\rho - T \rho,\rho, 0\] \text{ and } 1> |\rho| \geq 0,
-%	\end{align*}
-%	where 
-%	\begin{align*}
-%	w_1^{(l)} & = \frac{1}{2} \[D \psi_T\(\frac{\tau^{(l)} + 1}{2}\) + \sum_{d=1}^D \log |\bm{\Psi}_d^{(l)}| \], \\
-%	w_2^{(l)} & = \frac{\tau^{(l)} + 1}{2} \sum_{d=1}^D \trace \(\bm{\Psi}_d^{(l)}\) > 0,\\
-%	w_3^{(l)} & = \frac{\tau^{(l)} + 1}{2} \sum_{d=1}^D \trace \[(\mathbf{1}_{T \times T} - \I_T) \cdot \bm{\Psi}_d^{(l)}\],
-%	\end{align*}
-%	and $\Gamma_T(x)$ and $\psi_T(x)$ are the $T$-variate gamma and digamma functions, respectively. $\log \Gamma_T \(\frac{\tau}{2}\)$ is convex on the intervals $(0,1), \dots, (T-1, \infty)$
-
-%	- we model the prior expectation of the covariance instead of the precision. Precisions are conditional on the other variables, which doesn't make sense in the tissue context.
-
-%	The second sub problem estimates the molecular feature effect hyper prior:
-%	\begin{align}\label{eq:optproblem2}
-%	k^{(l+1)},\theta^{(l+1)} & = \underset{k,\theta}{\argmax} \E_{Q^{(l)}} [\log \pi (\bgamma^2)] \nonumber\\
-%	& = \underset{k,\theta}{\argmax} D k \log \theta - D \log \Gamma (k) + e_1^{(l)} k - e_2^{(l)} \theta \\
-%	& \text{subject to } k ,\theta > 0,
-%	\end{align}
-%	where
-%	\begin{align*}
-%	e_1^{(l)} & = D\psi\(\frac{p}{2} + k^{(l)}\) - \sum_{d=1}^D \log a_d^{(l)}, \\
-%	e_2^{(l)} & = \(\frac{p}{2} + k^{(l)} \) \sum_{d=1}^D (a_d^{(l)})^{-1},
-%	\end{align*}
-%	and $\Gamma(x)$ and $\psi(x)$ are the gamma and digamma functions, respectively.
-This is again a convex problem (see Appendix \ref{app:convexity}), but more cumbersome to solve numerically due to the log-gamma function.
-By approximating the digamma function with $\phi(x) \approx \log x - 1/(2x)$, we get the following closed-form solutions:
-%	\begin{align*}
-%	k^{(l+1)} & = \frac{1}{2} \( \log e_2^{(l)} - D^{-1} e_1^{(l)} - \log D \)^{-1}, \\
-%	\theta^{(l+1)} & = D k^{(l+1)} (e_2^{(l)})^{-1}.
-%	\end{align*}
-
 	\section{Simulations}
@@
 	\begin{appendix}
-		\section{Variational Bayes derivations}\label{app:derivations}
-		In the following all expectations are with respect to the variational posterior $Q$.
-		\begin{align*}
-		\log q_{\B} (\B) & \propto \E_{\mathbf{U}, \bsigma^2} [\log \mathcal{L}(\Y | \B, \U, \bsigma^2 )] + \E_{\bgamma^2, \bsigma^2} [\log \pi (\B | \bgamma^2, \bsigma^2)] \\
-		& \propto - \frac{1}{2} \sum_{d=1}^D \sum_{i=1}^n \E_{\mathbf{u}_d, \sigma_d^2} \[ \frac{(y_{id} - \x_i \tr \bbeta_d - \mathbf{z}_i \tr \mathbf{u}_d)^2}{\sigma_d^2} \] - \frac{1}{2} \sum_{d=1}^D \sum_{j=1}^p \phi_{g(j)}^{-2} \E_{\gamma^2_d, \sigma^2_d} \( \frac{\beta_{jd}^2}{\gamma_d^2 \sigma_d^2} \), \\
-		q_{\B} (\B) & \overset{D}{=} \prod_{d=1}^{D} \mathcal{N}_p (\bmu_d, \bSigma_d), \\
-		& \text{ with } \bSigma_d = \E(\sigma_d^{-2})^{-1} [\X \tr \X + \E(\gamma_d^{-2}) \cdot \diag (\phi_{g(j)}^{-2}) ]^{-1}, \\
-		& \text{ and } \bmu_d = [\X \tr \X + \E(\gamma_d^{-2}) \cdot \diag (\phi_{g(j)}^{-2}) ]^{-1} \X \tr [\y_d - \Z \cdot \E(\mathbf{u}_d)].
-		\end{align*}
-
-		\begin{align*}
-		\log q_{\U}(\U) & \propto \E_{\mathbf{B}, \bsigma^2} [\log \mathcal{L}(\Y | \B, \U, \bsigma^2 )] + \E_{\mathbf{\Xi}_1, \dots, \mathbf{\Xi}_D, \bsigma^2} [\log \pi (\U | \mathbf{\Xi}_1, \dots, \mathbf{\Xi}_D, \bsigma^2)] \\
-		& \propto - \frac{1}{2} \sum_{d=1}^D \sum_{i=1}^n \E_{\bbeta_d, \sigma_d^2} \[ \frac{(y_{id} - \x_i \tr \bbeta_d - \mathbf{z}_i \tr \mathbf{u}_d)^2}{\sigma_d^2} \] - \frac{1}{2} \sum_{d=1}^D \E_{\bm{\Xi}_d, \sigma_d^2}\( \frac{\mathbf{u}_d \tr \bm{\Xi}_d^{-1} \mathbf{u}_d}{\sigma_d^2} \), \\
-		q_{\U}(\U) & \overset{D}{=} \prod_{d=1}^{D} \mathcal{N}_T (\mathbf{m}_d, \Sm_d), \\
-		& \text{ with } \Sm_d = \E(\sigma_d^{-2})^{-1} [\Z \tr \Z + \E(\bm{\Xi}_d^{-1}) ]^{-1}, \\
-		& \text{ and } \mathbf{m}_d = [\Z \tr \Z + \E(\bm{\Xi}_d^{-1}) ]^{-1} \Z \tr [\y_d - \X \cdot \E(\bbeta_d)].
-		\end{align*}
-
-		\begin{align*}
-		\log q_{\bgamma^{2}}(\bgamma^{2}) & \propto \E_{\B, \bsigma^2} [\log \pi (\B | \bgamma^{2}, \sigma^2)] + \log \pi (\bgamma^{2}) \\
-		& \propto -\frac{p}{2} \sum_{d=1}^D \log \gamma_d^{2} - \frac{1}{2} \sum_{d=1}^D \sum_{j=1}^p \phi_{g(j)}^{-2} \E \( \frac{\beta_{jd}^2}{\sigma^{2}_d}\) \gamma_d^{-2} - \frac{3}{2} \sum_{d=1}^D \log \gamma_d^{2} \\
-		& \,\,\,\,\,\,\,\,\,\, - \frac{\lambda}{2 \theta^2} \sum_{d=1}^D (\gamma_d^2 - \theta)^2 \gamma_d^{-2} \\
-		& \propto \sum_{d=1}^D \left\{ \( -\frac{p + 1}{2} - 1\) \log \gamma_d^{2} - \frac{\lambda}{2\theta^2} \gamma_d^2 - \frac{1}{2}\[ \lambda + \E(\sigma_d^{-2}) \sum_{j=1}^p \frac{\E(\beta_{jd}^2)}{\phi_{g(j)}^{2}}\] \gamma_d^{-2} \right\}, \\
-		q_{\bgamma^{2}}(\bgamma^{2}) & \overset{D}{=} \prod_{d=1}^D \mathcal{GIG} \(-\frac{p+1}{2}, \frac{\lambda}{\theta_d^2}, \delta_{d} \), \\
-		& \text{with } \delta_{d} =\E(\sigma_d^{-2}) \sum_{j=1}^p \frac{\E(\beta_{jd})^2 + \V(\beta_{jd})}{\phi_{g(j)}^2} + \lambda.
-		\end{align*}
-
-%		\begin{align*}
-%		\log q_{\bgamma^{-2}}(\bgamma^{-2}) & \propto \E_{\B, \bsigma^2} [\log \pi (\B | \bgamma^{-2}, \sigma^2)] + \log \pi (\bgamma^{-2}) \\
-%		& \propto \frac{p}{2} \sum_{d=1}^D \log \gamma_d^{-2} - \frac{1}{2} \sum_{d=1}^D \sum_{j=1}^p \phi_{g(j)}^{-2} \E \( \frac{\beta_{jd}^2}{\sigma^{2}_d}\) \gamma_d^{-2} + (k - 1) \sum_{d=1}^D \log \gamma_d^{-2} - \theta \sum_{d=1}^D \gamma_d^{-2} \\
-%		& = \sum_{d=1}^D \left\{ \( \frac{p}{2} + k - 1\) \log \gamma_d^{-2} - \[ \frac{\E(\sigma_d^{-2})}{2} \sum_{j=1}^p \frac{\E(\beta_{jd}^2)}{\phi_{g(j)}^{2}} + \theta \] \gamma_d^{-2} \right\}.
-%		\end{align*}
-
-%	\begin{align*}
-%	q_{\bgamma^{-2}}(\bgamma^{-2}) & \overset{D}{=} \prod_{d=1}^D \Gamma \(\frac{p}{2} + k, a_{d} \), \\
-%	& \text{ with } a_{d} =\frac{\E(\sigma_d^{-2})}{2} \sum_{j=1}^p \frac{\E(\beta_{jd})^2 + \V(\beta_{jd})}{\phi_{g(j)}^2} + \theta.
-%	\end{align*}
-
-		\begin{align*}
-		\log q_{\bsigma^{2}}(\bsigma^{2}) & \propto \E_{\B,\mathbf{U}} [\log \mathcal{L}(\Y | \B, \U, \bsigma^2 )] + \E_{\B, \bgamma^2} [\log \pi (\B | \bgamma^2, \bsigma^2)] \\
-		& \,\,\,\,\,\,\,\,\,\, + \E_{\mathbf{U}, \bm{\Xi}_1, \dots, \bm{\Xi}_D} [\log \pi (\mathbf{U} | \bm{\Xi}_1, \dots, \bm{\Xi}_D, \bsigma^2)] + \log \pi (\bsigma^{2}) \\
-		& \propto -\frac{n}{2} \sum_{d=1}^D \log \sigma_d^{2} - \frac{1}{2} \sum_{d=1}^D \sum_{i=1}^n \E [(y_{id} - \x_i \tr \bbeta_d - \mathbf{z}_i \tr \mathbf{u}_d)^2] \sigma_d^{2} - \frac{p}{2} \sum_{d=1}^D \log \sigma_d^{-2} \\
-		& \,\,\,\,\,\,\,\,\,\, - \frac{1}{2} \sum_{d=1}^D \sum_{j=1}^p \E( \gamma_d^{-2}) \frac{\E ( \beta_{jd}^2)}{\phi_{g(j)}^{2}} \sigma_d^{-2} - \frac{T}{2} \sum_{d=1}^D \log \sigma_d^{2} - \frac{1}{2} \sum_{d=1}^D \E(\mathbf{u}_d \tr \bm{\Xi}_d^{-1} \mathbf{u}) \sigma_d^{-2} \\
-		& \,\,\,\,\,\,\,\,\,\, - \frac{3}{2} \sum_{d=1}^D \log \sigma^2_d \\
-		& = -\sum_{d=1}^{D} \(\frac{n + p + T + 3}{2} \) \log \sigma_d^{2} - \sum_{d=1}^{D} \frac{1}{2}\Bigg\{ \sum_{i=1}^n \E [(y_{id} - \x_i \tr \bbeta_d - \mathbf{z}_i \tr \mathbf{u}_d)^2] \\
-		& \,\,\,\,\,\,\,\,\,\, + \E(\gamma_d^{-2})\sum_{j=1}^p \frac{\E(\beta_{jd}^2)}{\phi_{g(j)}^2} + \E(\mathbf{u}_{d} \tr \bm{\Xi}_d^{-1} \mathbf{u}_d) \Bigg\} \sigma_d^{-2}, \\
-		q_{\bsigma^{2}}(\bsigma^{2}) & \overset{D}{=} \prod_{d=1}^D \Gamma^{-1} \(\frac{n + p + T + 1}{2}, \zeta_{d} \), \\
-		& \text{ with } \zeta_{d} = \frac{\y_d \tr \y_d}{2} - \y_d \tr [\X \E(\bbeta_d) + \Z \E(\mathbf{u}_d)] + \E(\bbeta_d \tr) \X \tr \Z \E (\mathbf{u}_d) \\
-		& \,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\, + \frac{\trace [\X \tr \X \V(\bbeta_d)]}{2} + \frac{\E(\bbeta_d \tr) \X \tr \X \E(\bbeta_d)}{2} + \frac{\trace [\Z \tr \Z \V(\mathbf{u}_d)]}{2} \\
-		& \,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\, + \frac{\E(\mathbf{u}_d \tr) \Z \tr \Z \E(\mathbf{u}_d)}{2} + \frac{\E(\gamma_d^{-2})}{2} \sum_{j=1}^p \frac{\E(\beta_{jd})^2 + \V(\beta_{jd})}{\phi_{g(j)}^2} \\
-		& \,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,+ \frac{\E(\mathbf{u}_d \tr) \E(\bm{\Xi}_d^{-1}) \E(\mathbf{u}_d)}{2} + \frac{\trace [\E(\bm{\Xi}_d^{-1}) \V(\mathbf{u}_d)]}{2},
-		\end{align*}
-
-%		\begin{align*}
-%		\log q_{\bsigma^{-2}}(\bsigma^{-2}) & \propto \E_{\B,\mathbf{U}} [\log \mathcal{L}(\Y | \B, \U, \bsigma^2 )] + \E_{\B, \bgamma^2} [\log \pi (\B | \bgamma^2, \bsigma^2)] \\
-%		& \,\,\,\,\,\,\,\,\,\, + \E_{\mathbf{U}, \bm{\Xi}_1, \dots, \bm{\Xi}_D} [\log \pi (\mathbf{U} | \bm{\Xi}_1, \dots, \bm{\Xi}_D, \bsigma^2)] + \log \pi (\bsigma^{-2}) \\
-%		& \propto \frac{n}{2} \sum_{d=1}^D \log \sigma_u^{-2} - \frac{1}{2} \sum_{d=1}^D \sum_{i=1}^n \E [(y_{id} - \x_i \tr \bbeta_d - \mathbf{z}_i \tr \mathbf{u}_d)^2] \sigma_d^{-2} + \frac{p}{2} \sum_{d=1}^D \log \sigma_d^{-2} \\
-%		& \,\,\,\,\,\,\,\,\,\, - \frac{1}{2} \sum_{d=1}^D \sum_{j=1}^p \phi_{g(j)}^{-2} \E \( \frac{\beta_{jd}^2}{\gamma_d^{2}} \) \sigma_d^{-2} + \frac{T}{2} \sum_{d=1}^D \log \sigma_d^{-2} - \frac{1}{2} \sum_{d=1}^D \E(\mathbf{u}_d \tr \bm{\Xi}_d^{-1} \mathbf{u}) \sigma_d^{-2} \\
-%		& \,\,\,\,\,\,\,\,\,\, - 0.999 \sum_{d=1}^D \log \sigma_d^{-2} - 0.001 \sum_{d=1}^D \sigma_d^{-2} \\
-%		& = \sum_{d=1}^{D} \(\frac{n + p + T}{2} - 0.999 \) \log \sigma_d^{-2} - \sum_{d=1}^{D} \Bigg\{ \frac{\sum_{i=1}^n \E [(y_{id} - \x_i \tr \bbeta_d - \mathbf{z}_i \tr \mathbf{u}_d)^2]}{2} \\
-%		& \,\,\,\,\,\,\,\,\,\, + \frac{\E(\gamma_d^{-2})}{2} \sum_{j=1}^p \frac{\E(\beta_{jd}^2)}{\phi_{g(j)}^2} + \frac{\E(\mathbf{u}_{d} \tr \bm{\Xi}_d^{-1} \mathbf{u}_d)}{2}  + 0.001 \Bigg\} \sigma_d^{-2}.
-%		\end{align*}
-
-%	\begin{align*}
-%	q_{\bsigma^{-2}}(\bsigma^{-2}) & \overset{D}{=} \prod_{d=1}^D \Gamma \(\frac{n + p + T}{2} + 0.001, b_{d} \), \\
-%	& \text{ with } b_{d} = \frac{\y_d \tr \y_d}{2} - \y_d \tr [\X \E(\bbeta_d) + \Z \E(\mathbf{u}_d)] + \E(\bbeta_d \tr) \X \tr \Z \E (\mathbf{u}_d) \\
-%	& \,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\, + \frac{\trace [\X \tr \X \V(\bbeta_d)]}{2} + \frac{\E(\bbeta_d \tr) \X \tr \X \E(\bbeta_d)}{2} + \frac{\trace [\Z \tr \Z \V(\mathbf{u}_d)]}{2} \\
-%	& \,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\, + \frac{\E(\mathbf{u}_d \tr) \Z \tr \Z \E(\mathbf{u}_d)}{2} + \frac{\E(\gamma_d^{-2})}{2} \sum_{j=1}^p \frac{\E(\beta_{jd})^2 + \V(\beta_{jd})}{\phi_{g(j)}^2} \\
-%	& \,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,+ \frac{\E(\mathbf{u}_d \tr) \E(\bm{\Xi}_d^{-1}) \E(\mathbf{u}_d)}{2} + \frac{\trace [\E(\bm{\Xi}_d^{-1}) \V(\mathbf{u}_d)]}{2} + 0.001.
-%	\end{align*}
-
-		\begin{align*}
-		\log q_{\bm{\Xi}_1, \dots, \bm{\Xi}_D}(\bm{\Xi}_1, \dots, \bm{\Xi}_D) & \propto \E_{\mathbf{U}, \bm{\sigma}^2} [\log \pi (\mathbf{U} | \bm{\Xi}_1, \dots, \bm{\Xi}_D, \bsigma^2)] + \log \pi (\bm{\Xi}_1, \dots, \bm{\Xi}_D) \\
-		& \propto -\frac{1}{2} \sum_{d=1}^D \log |\bm{\Xi}_d| - \frac{1}{2} \sum_{d=1}^D \E \(\frac{\mathbf{u}_d \tr \bm{\Xi}_d^{-1} \mathbf{u}_d}{\sigma_d^2} \) \\
-		& \,\,\,\,\,\,\,\,\,\, - \frac{k + T + 1}{2} \sum_{d=1}^D \log |\bm{\Xi}_d| - \frac{1}{2} \sum_{d=1}^D \trace (\bm{\Omega} \bm{\Xi}_d^{-1}) \\
-		& = -\frac{\tau + T + 2}{2} \sum_{d=1}^D \log |\bm{\Xi}_d| - \frac{1}{2} \sum_{d=1}^D \trace \left\{ \[\E(\sigma_d^{-2}) \E( \mathbf{u}_d \mathbf{u}_d \tr) + \bm{\Omega}\] \bm{\Xi}_d^{-1} \right\}, \\
-		q_{\bm{\Xi}_1, \dots, \bm{\Xi}_D}(\bm{\Xi}_1, \dots, \bm{\Xi}_D) & \overset{D}{\propto} \prod_{d=1}^D \mathcal{W}_T^{-1} (\bm{\Psi}_d, \nu + 1),  \\ 
-		& \text{ with } \bm{\Psi}_d =\E(\sigma^{-2}_d) \[\V(\mathbf{u}_d) + \E(\mathbf{u}_d) \E( \mathbf{u}_d \tr) \] + \bm{\Omega}.
-		\end{align*}
-
-%		\section{Convexity}\label{app:convexity}
-%		For (\ref{eq:optproblem2}) to be a convex problem, we need the Hessian $\mathbf{H}$ of the objective function to be positive definite, i.e., $\forall \mathbf{z} \in \mathbb{R}^2: \mathbf{z} \tr \mathbf{H} \mathbf{z} > 0$.
-We have
-%		\begin{align*}
-%		\mathbf{z} \tr \mathbf{H} \mathbf{z} = \frac{D}{\theta^2 \psi^{(1)}(k)} \( z_1 \theta \psi^{(1)}(k) - z_2 \)^2 + z_2^2 \( k \psi^{(1)}(k)  - 1 \),
-%		\end{align*}
-%		with $\psi^{(1)}(k)$ the trigamma function. Since the trigamma function is positive for all $k > 0$, and we have $\theta, k > 0$, we need:
-%		$$
-%		k \psi^{(1)}(k)> 1.
-%		$$
-%		Using Lemma 1 in \cite{guo_refinements_2013} we can show that for $k > 0$:
-%		$$
-%		k \psi^{(1)}(k) > 1 + \frac{1}{2k},
-%		$$
-%		such that the objective function is convex.
-
 		\section{Prior tissue covariance constraints}\label{app:priortissuecovariance}
-%		The empirical Bayes objective function of the third sub-optimisation problem in \ref{sec:empiricalbayes} is:
-%		\begin{align*}
-%		\E_{Q^{(l)}}[\log \pi (\mathbf{\Xi}_1, \dots, \mathbf{\Xi}_D)] & \propto \frac{D}{2} \nu \log |\bm{\Omega}| - \frac{D T \log 2}{2} \nu - D \log \Gamma_T \(\frac{\nu}{2}\) + \frac{\nu}{2}\sum_{d=1}^D \E_{Q^{(l)}} (\log | \mathbf{\Xi}_d^{-1}|) \\
-%		& \,\,\,\,\,\,\,\,\,\, - \frac{1}{2} \sum_{d=1}^D \trace [ \bm{\Omega} \cdot \E_{Q^{(l)}}(\mathbf{\Xi}_d^{-1})] \\
-%		& = \frac{D}{2} \nu \log |\bm{\Omega}| - D \log \Gamma_T\(\frac{\nu}{2}\) - \frac{\nu^{(l)} + 1}{2} \sum_{d=1}^D \trace [ \bm{\Omega} \cdot ( \mathbf{\Psi}_d^{(l)})^{-1}] \\
-%		& \,\,\,\,\,\,\,\,\,\, + \frac{\nu}{2} \sum_{d=1}^D \[ \psi_T\(\frac{\nu^{(l)} + 1}{2}\) - \log |\mathbf{\Psi}_d^{(l)}| \],
-%		\end{align*}
-%		which is maximised under constraints $\nu > T + 1$ and $\bm{\Omega}$ positive definite (PD), to ensure a proper inverse Wishart distribution.
-
 		We parametrise the prior covariance as a matrix with $\tau$ on the diagonal and $\rho$ on its off-diagonals. To that end we set $\nu = \tau + T + 1$ and the $T \times T$-dimensional matrix
@@
 
-%		The resulting empirical Bayes step is now:
-%		\begin{align*}
-%		\tau^{(l + 1)}, \rho^{(l+1)} & = \underset{\tau, \rho}{\argmax} \frac{D (T^2 + T)}{2} \log \tau + \frac{DT}{2} \tau \log \tau + \frac{D(T+1)}{2} \log \[ \tau - (T - 1) \rho\] \\
-%		& \,\,\,\,\,\,\,\,\,\, + \frac{D}{2} \tau \log \[ \tau - (T - 1) \rho\] + \frac{D(T^2 - 1)}{2} \log(\tau - \rho) +\frac{D(T - 1)}{2} \tau  \log(\tau - \rho) \\
-%		& \,\,\,\,\,\,\,\,\,\, - D \sum_{t=1}^T \log \Gamma \( \frac{\tau + t + 1}{2} \) - w_1^{(l)} \tau^2 + w_2^{(l)} \tau \rho + w_3^{(l)} \tau, \\
-%		& \text{subject to } \tau > \max\[\rho - T \rho,\rho, 0\],
-%		\end{align*}
-%		with
-%		\begin{align*}
-%		w_1^{(l)} & =  \frac{\tau^{(l)} + T + 2}{2} \sum_{d=1}^D \trace \[ (\bm{\Psi}_d^{(l)})^{-1}\] > 0,\\
-%		w_2^{(l)} & = \frac{\tau^{(l)} + T + 2}{2} \sum_{d=1}^D \left\{\trace \[ (\bm{\Psi}_d^{(l)})^{-1}\] - \trace \[\mathbf{1}_{T \times T} \cdot (\bm{\Psi}_d^{(l)})^{-1}\] \right\} < w_1^{(l)},\\
-%		w_3^{(l)} & = \frac{1}{2} \[D \sum_{t=1}^T \psi\(\frac{\tau^{(l)} + t + 2}{2}\) - \sum_{d=1}^D \log |\bm{\Psi}_d^{(l)}| \].
-%		\end{align*}
-
-		\section{Model without tissue effects}
-		\subsection{Model}
-		In order to investigate the inverse Gaussian prior model and its estimation, we consider the simpler model:
-		\begin{align*}
-		y_{id} | \bm{\beta}_d, \sigma_d^2 & \sim \mathcal{N} (\x_i \tr \bm{\beta}_d, \sigma_d^2), \\
-		\beta_{jd} | \gamma_d^2, \sigma_d^2 & \sim \mathcal{N} (0, \sigma_d^2 \gamma_d^2), \\
-		\gamma_d^2 & \sim \mathcal{IG}(\theta_d, \lambda), \\
-		\sigma_d^2 & \sim 1/\sigma_d^3.
-		\end{align*}
-		So in other words, we ignore the tissue effect and fix all $\phi_g^2$ to one. 
-		\subsection{Variational Bayes}
-		The corresponding variational Bayes posterior is given by:
-		\begin{align*}
-		q(\bm{\beta}) & \overset{D}{=} \prod_{d=1}^D \mathcal{N}_p (\bm{\mu}_d, \bm{\Sigma}_d), \\
-		q(\bm{\gamma}^2) & \overset{D}{=} \prod_{d=1}^D \mathcal{GIG}\(-\frac{p+1}{2}, \frac{\lambda}{\theta_d^2}, \delta_d\), \\
-		q(\bm{\sigma}^2) & \overset{D}{=} \prod_{d=1}^D \Gamma^{-1} \(\frac{n + p + 1}{2}, \zeta_d\).
-		\end{align*}
-		We update the parameters until convergence:
-		\begin{align*}
-		\bm{\Sigma}_d^{(h+1)} & = \frac{2 \zeta_d^{(h)}}{n+p+1} (\X \tr \X + a_d^{(h)} \I)^{-1}, \\
-		\bm{\mu}_d^{(h+1)} & = (\X \tr \X + a_d^{(h)} \I)^{-1} \X \tr \y_d, \\
-		\delta_d^{(h+1)} & = \frac{n+p+1}{2 \zeta_d^{(h)}} [\trace(\bm{\Sigma}_d^{(h+1)}) + (\bm{\mu}_d^{(h+1)}) \tr \bm{\mu}_d^{(h+1)}] + \lambda, \\ \zeta_d^{(h+1)} & = \frac{a_d^{(h+1)}}{2} \[\mathbf{y}_d \tr \mathbf{y}_d -2 \mathbf{y}_d \tr \X \bm{\mu}_d^{(h+1)} + \trace [ (\X \tr \X + \I_p) \bm{\Sigma}_d^{(h+1)}] + (\bm{\mu}_d^{(h+1)}) \tr (\X \tr \X + \I_p)\bm{\mu}_d^{(h+1)}\],
-		\end{align*}
-		with
-		$$
-		a_d^{(h)} = \sqrt{\frac{\lambda}{\theta_d^2 \delta_d^{(h)}}} \frac{K_{\frac{p - 1}{2}}\(\sqrt{\lambda \delta_d^{(h)}} /\theta_d \)}{K_{\frac{p + 1}{2}}\(\sqrt{\lambda \delta_d^{(h)}} /\theta_d \)} + \frac{p+1}{\delta_d^{(h)}}.
-		$$
-
-		\subsection{Empirical Bayes}
-		Again we model $\theta_d = (\mathbf{c}_d \tr \bm{\alpha})^{-1}$ to arrive at the following empirical Bayes updates:
-		\begin{align*}
-		\bm{\alpha}^{(l+1)} & = \left\{ \mathbf{C} \tr \Vm^{(l)} \mathbf{C} \right\}^{-1} \mathbf{C} \tr \mathbf{1}, \\
-		\lambda^{(l+1)} & = \[ \sum_{d=1}^D a_d^{(l)} - (\bm{\alpha}^{(l+1)}) \tr \mathbf{C} \tr \mathbf{1} \]^{-1} D,
-		\end{align*}
-		where 
-		$$
-		\Vm^{(l)} = \diag \[ \sqrt{\frac{\delta_d^{(l)}(\theta_d^{(l)})^2}{\lambda^{(l)}}} \frac{K_{\frac{p - 1}{2}} \( \sqrt{\lambda^{(l)} \delta_d^{(l)}}/\theta^{(l)}_d \)}{K_{\frac{p + 1}{2}} \( \sqrt{\lambda^{(l)} \delta_d^{(l)}}/\theta^{(l)}_d \)} \].
-		$$
-
-		\subsection{Evidence lower bound}
-		We monitor convergence through the evidence lower bound (ELBO):
-		\begin{align*}
-		\text{ELBO} & (Q^{(l)}) \propto \frac{1}{2} \sum_{d=1}^D \log |\bm{\Sigma}_d^{(l)}| - \frac{n + p + 1}{2} \sum_{d=1}^D \log \zeta_d^{(l)} + \sum_{d=1}^D \log K_{\frac{p+1}{2}}\(\sqrt{\lambda^{(l)} \delta_d^{(l)}}/\theta_d^{(l)}\) \\
-		& - \frac{n+p+1}{4} \sum_{d=1}^D (\zeta_d^{(l)})^{-1} \[ \y_d \tr \y_d - 2\y_d \tr \X \bm{\mu}_d^{(l)} + (\bm{\mu}_d^{(l)}) \tr \X \tr \X \bm{\mu}_d^{(l)} + \trace ( \X \tr \X \bm{\Sigma}_d^{(l)})\] \\
-		& + \frac{D(p+3)}{4} \log \lambda^{(l)} + \lambda^{(l)} \sum_{d=1}^D (\theta_d^{(l)})^{-1} - \frac{p+1}{2} \sum_{d=1}^D \log \theta_d^{(l)} - \frac{p+1}{4} \sum_{d=1}^D \log \delta_d^{(l)} \\
-		& + \frac{1}{2} \sum_{d=1}^D \[ \delta_d^{(l)} - \lambda^{(l)} - \frac{n+p+1}{2\zeta_d^{(l)}} (\bm{\mu}_d^{(l)}) \tr \bm{\mu}_d^{(l)} - \frac{n+p+1}{2\zeta_d^{(l)}} \trace (\bm{\Sigma}_d^{(l)}) \] a_d^{(l)}.
-		\end{align*}
-
-%		\begin{align*}
-%		y_{id} | \bm{\beta}_d & \sim \mathcal{N} (\x_i \tr \bm{\beta}_d, \sigma_d^2), \\
-%		\beta_{jd} | \gamma_d^2 & \sim \mathcal{N} (0, \sigma_d^2 \gamma_d^2), \\
-%		\gamma_d^2 & \sim \mathcal{IG}(\theta_d, \lambda).
-%		\end{align*}
-%		So in other words, we ignore the tissue effect, fix the $\sigma_d^2$, and set all $\phi_g^2$ to one. The corresponding variational Bayes posterior is given by:
-%		\begin{align*}
-%		q(\bm{\beta}) & \overset{D}{=} \prod_{d=1}^D \mathcal{N}_p (\bm{\mu}_d, \bm{\Sigma}_d), \\
-%		q(\bm{\gamma}^2) & \overset{D}{=} \prod_{d=1}^D \mathcal{GIG}\(-\frac{p+1}{2}, \frac{\lambda}{\theta_d^2}, \delta_d\).
-%		\end{align*}
-%		We update the parameters until convergence:
-%		\begin{align*}
-%		\bm{\Sigma}_d^{(h+1)} & = \sigma_d^2 (\X \tr \X + a_d^{(h)} \I)^{-1}, \\
-%		& \text{with } a_d^{(h)} = \sqrt{\frac{\lambda}{\theta_d^2 \delta_d^{(h)}}} \frac{K_{\frac{p - 1}{2}}\(\sqrt{\lambda \delta_d^{(h)}} /\theta_d \)}{K_{\frac{p + 1}{2}}\(\sqrt{\lambda \delta_d^{(h)}} /\theta_d \)} + \frac{p+1}{\delta_d^{(h)}}, \\
-%		\bm{\mu}_d^{(h+1)} & = (\X \tr \X + a_d^{(h)} \I)^{-1} \X \tr \y_d, \\
-%		\delta_d^{(h+1)} & = \sigma_d^{-2} [\trace(\bm{\Sigma}_d^{(h+1)}) + (\bm{\mu}_d^{(h+1)}) \tr \bm{\mu}_d^{(h+1)}] + \lambda.
-%		\end{align*}
-%		Again we model $\theta_d = (\mathbf{c}_d \tr \bm{\alpha})^{-1}$ to arrive at the following empirical Bayes updates:
-%		\begin{align*}
-%		\bm{\alpha}^{(l+1)} & = \left\{ \mathbf{C} \tr \Vm^{(l)} \mathbf{C} \right\}^{-1} \mathbf{C} \tr \mathbf{1}, \\
-%		\lambda^{(l+1)} & = \[ \sum_{d=1}^D a_d^{(l)} - (\bm{\alpha}^{(l+1)}) \tr \mathbf{C} \tr \mathbf{1} \]^{-1} D,
-%		\end{align*}
-%		where 
-%		$$
-%		\Vm^{(l)} = \diag \[ \sqrt{\frac{\delta_d^{(l)}(\theta_d^{(l)})^2}{\lambda^{(l)}}} \frac{K_{\frac{p - 1}{2}} \( \sqrt{\lambda^{(l)} \delta_d^{(l)}}/\theta^{(l)}_d \)}{K_{\frac{p + 1}{2}} \( \sqrt{\lambda^{(l)} \delta_d^{(l)}}/\theta^{(l)}_d \)} \].
-%		$$
-%		We monitor convergence through the evidence lower bound (ELBO):
-%		\begin{align*}
-%		\text{ELBO} & (Q^{(l)}) \propto \sum_{d=1}^D \sigma_d^{-2} \y_d \tr \X \bm{\mu}_d^{(l)} - \frac{1}{2} \sum_{d=1}^D \sigma_d^{-2} \trace ( \X \tr \X \bm{\Sigma}_d^{(l)}) - \frac{1}{2} \sum_{d=1}^D \sigma_d^{-2} (\bm{\mu}_d^{(l)}) \tr \X \tr \X \bm{\mu}_d^{(l)} \\
-%		& + \frac{1}{2} \sum_{d=1} \log |\bm{\Sigma}_d^{(l)}| + \frac{D(p+3)}{4} \log \lambda^{(l)} + \lambda^{(l)} \sum_{d=1}^D (\theta_d^{(l)})^{-1} - \frac{p+1}{2} \sum_{d=1}^D \log \theta_d^{(l)} \\
-%		& - \frac{p+1}{4} \sum_{d=1}^D \log \delta_d^{(l)} + \sum_{d=1}^D \log K_{\frac{p+1}{2}}\(\sqrt{\lambda^{(l)} \delta_d^{(l)}}/\theta_d^{(l)}\)\\
-%		& + \frac{1}{2} \sum_{d=1}^D \left(\rule{0cm}{1.2cm}\right.
-\[ \delta_d^{(l)} - \lambda^{(l)} - \sigma_d^{-2} (\bm{\mu}_d^{(l)}) \tr \bm{\mu}_d^{(l)} - \sigma_d^{-2} \trace (\bm{\Sigma}_d^{(l)}) \]  \\
-%		& \,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\, \cdot \[\sqrt{\frac{\lambda^{(l)}}{\delta_d^{(l)} (\theta^{(l)}_d)^2}} \frac{K_{\frac{p - 1}{2}}\(\frac{\sqrt{\lambda^{(l)} \delta_d^{(l)}}}{\theta^{(l)}_d} \)}{K_{\frac{p + 1}{2}}\(\frac{\sqrt{\lambda^{(l)} \delta_d^{(l)}}}{\theta^{(l)}_d} \)} + \frac{p+1}{\delta_d^{(l)}}\] \left.
-\rule{0cm}{1.2cm}\right).
-%		\end{align*}
-
-		- Different options for convergence. We may monitor parameters, ELBO, or a combination. We may also iterate a fixed number of times, especially for the VB.
-
-		\subsection{Efficient computation}
-		We need the following quantities during the EB iterations: $\zeta_d$, $a_d$, $\delta_d$, $\trace (\bm{\Sigma}_d)$, $\trace (\X \tr \X \bm{\Sigma}_d)$, $\bm{\mu}_d \tr \bm{\mu}_d$, and $\bm{\mu}_d \tr \X \tr \X \bm{\mu}_d$.
-In addition, we need $\log|\bm{\Sigma}|$ and $\mathbf{y}_d \tr \X \bm{\mu}_d$ to monitor the ELBO.
-The first three are just scalar operations.
-The rest is easily calculated using the SVD $\X = \mathbf{U} \mathbf{D} \mathbf{V} \tr$:
-		\begin{align*}
-		\trace (\bm{\Sigma}_d) & = \frac{2\zeta_d}{n + p + 1} \[\sum_{i=1}^n (v_i^2 + a_d)^{-1} + \max (p - n, 0)\cdot a_d^{-1}\], \\
-		\trace (\X \tr \X \bm{\Sigma}_d) & = \frac{2\zeta_d}{n + p + 1} \sum_{i=1}^n v_i^2(v_i^2 + a_d)^{-1}, \\
-		\bm{\mu}_d \tr \bm{\mu}_d & = \sum_{i=1}^n v_i^2 [(\mathbf{U} \tr \mathbf{y}_d)_i]^2/(v_i^2 + a_d)^2, \\
-		\bm{\mu}_d \tr \X \tr \X \bm{\mu}_d & = \sum_{i=1}^n v_i^4 [(\mathbf{U} \tr \mathbf{y}_d)_i]^2/(v_i^2 + a_d)^2, \\
-		\log|\bm{\Sigma}| & \propto p \log \zeta_d - \sum_{i=1} \log (v_i^2 + a_d) - \max(p - n, 0) \cdot \log a_d, \\
-		\mathbf{y}_d \tr \X \bm{\mu}_d & = \sum_{i=1}^n v_i^2 [(\mathbf{U} \tr \mathbf{y}_d)_i]^2/(v_i^2 + a_d).
-		\end{align*}
-		where $v_i$ are the singular values of $\X$.
-The SVD and $\mathbf{U} \tr \mathbf{y}_d$ are $\mathcal{O}(pn^2)$ and $\mathcal{O}(n^2)$ once at the start of the algorithm.
-The rest of the calculations involve just $n$ scalar multiplications and/or additions.
-
-		\subsection{Simulations}
-		We simulate from a correctly specified model.
-We assume random $y_{id}$, $\x_i$, and $\bm{\beta}_d$, while we fix $\mathbf{C}$, $\gamma_d^2$, $\sigma_d^2$ and $\bm{\alpha}$.
-We estimate the $\bm{\mu}_d$, $\bm{\Sigma}_d$, $\delta_d$, $\bm{\alpha}$ and $\lambda$, while we fix the $\sigma_d^2$ to their true values.
-
-		We set $n=100$, $p=200$, and $D=10$.
-We consider 5 evenly sized classes of drugs, such that $\mathbf{c}_{kd} = \mathbbm{1}(k = \text{class}_d)$ and $\begin{bmatrix} \mathbf{c}_1 & \cdots & \mathbf{c}_D \end{bmatrix} \tr$.
-From these the $\gamma_d^2 = (\mathbf{c}_d \tr \bm{\alpha})^{-1}$ are created.
-Next, we simulate $\bm{\beta}_{jd} \sim \mathcal{N} (0, \sigma^2_d \gamma_d^2)$.
-We fix the mean signal-to noise ratio $\overline{\text{SNR}}_d$, such that the predictor data variance is $s^2 = \V(\mathbf{x}_i) = \overline{\text{SNR}}_d / (\overline{\gamma^2_d} p)$.
-The last step consists of simulating the data by $\mathbf{x}_{ij} \sim \mathcal{N}(0,s^2)$ and $y_{id} = \mathcal{N} (\mathbf{x}_i \tr \bm{\beta}_d, \sigma_d^2)$.
-The remaining parameters are set as follows: $\bm{\alpha} = \begin{bmatrix} 1 & \dots & 5 \end{bmatrix} \tr$, $\sigma_d^2 = 1$ for all $d$, and $\overline{\text{SNR}}_d = 5$.
-
-		We compare the results to a model where the $\gamma_d^2$ are drawn from a class-specific  inverse Gamma distribution; a generalisation of the models in \cite{kpogbezan_empirical_2017} and \cite{leday_gene_2017}:
-		\begin{align*}
-		y_{id} | \bm{\beta}_d ,\sigma_d^2 & \sim \mathcal{N} (\x_i \tr \bm{\beta}_d, \sigma_d^2), \\
-		\beta_{jd} | \gamma_d^2 ,\sigma_d^2 & \sim \mathcal{N} (0, \sigma_d^2 \gamma_d^2), \\
-		\gamma_d^2 & \sim \Gamma^{-1} (a_{\text{class}(d)}, b_{\text{class}(d)}), \\
-		\sigma_d^2 & \sim \Gamma^{-1} (0.001, 0.001),
-		\end{align*}
-		where $\text{class}(c)$ denotes the class of drug $d$.
-This allows for more flexible empirical Bayes estimation of the $a_d$ and $b_d$, but, as a consequence, increases the risk of overfitting.
-Additionally, it does not allow to include continuous covariates on the drugs.
-
-		To account for the random data and parameter generation, we repeat the simulation 100 times and present the results in Table...
-
-		\subsection{Starting values}
-%		Starting value for $\sigma_d^2$ as in \cite{chipman_bart:_2010}.
-They consider a scaled-Chi squared distribution for the error variance with degrees of freedom 3 and scale chosen such that the $q*100$th percentile matches the error variance in the data $s^2(\mathbf{y}_d)$.
-As starting value for $\sigma_d^2$ we take the mode of this distribution: $\hat{\sigma}_d^2 = 3a/5$, with $a$ the solution to:
-%		$$
-%		\Gamma\(\frac{3}{3}\) \[1 - q - F(3a/(2s^2(\mathbf{y}_d)); 3/2, 1) \] = 0,
-%		$$
-%		where $F(x; 3/2, 1)$ is the CDF of a gamma function with shape $3/2$ and scale $1$.
-In \cite{moran_variance_2018}, REFERENCE to Ročková and George (2018), and \cite{chipman_bart:_2010}, $q=0.9$ is suggested.
-We find that $q \in ( 0.9, 0.95 )$ gives good results.
-
-		We generate starting values as follows. Consider the $\sigma_d^2$ and $\gamma_d^2$ as fixed and estimate them by MML of the regular ridge model:
-		$$
-		\hat{\sigma}_d^2, \hat{\gamma}_d^2= \underset{\sigma_d^2,\gamma_d^2}{\argmax} \int_{\bm{\beta}} p(\mathbf{y}_d | \bm{\beta}, \sigma_d^2) p(\bm{\beta} | \gamma_d^2, \sigma_d^2) d\bm{\beta}.
-		$$
-		The regular ridge model is conjugate, so MML is relatively simple. Next we estimate a common inverse Gaussian prior for the $\gamma_d^2$ by considering the estimates to be samples from the prior:
-		\begin{align*}
-		\hat{\theta} & = n^{-1} \sum_{d=1}^D \hat{\gamma}_d^2 / n, \\
-		\hat{\lambda} & = n \left\{\sum_{d=1}^D \[ (\hat{\gamma}_d^2)^{-1} - \hat{\theta}^{-1}\]\right\}^{-1}.
-		\end{align*}
-		We estimate the mode of the $\hat{m}=\hat{\gamma}_d^2$ by kernel density estimation and equate it to the theoretical mode of the generalized inverse Gaussian distribution with the estimated $\hat{\theta}$ and $\hat{\lambda}$.
-We solve to find a common $\delta$:
-		$$
-		\hat{\delta} = \frac{\hat{m}^2 \hat{\lambda}}{\hat{\theta}^2} + (p + 3) \hat{m}.
-		$$
-		Lastly, we consider the $\hat{\sigma}^2_d$ fixed samples from the inverse Gamma distribution to estimate a common scale:
-		$$
-		\hat{\zeta} = \frac{D(n + p + 1)}{2 \sum_{d=1}^D[(\hat{\sigma}_d^2)^{-1}]}.
-		$$
-
-		\subsection{Independent error variance model parameters}
-		Model:
-		\begin{align*}
-		y_{id} | \bm{\beta}_d, \sigma_d^2 & \sim \mathcal{N} (\x_i \tr \bm{\beta}_d, \sigma_d^2), \\
-		\beta_{jd} | \gamma_d^2 & \sim \mathcal{N} (0, \gamma_d^2), \\
-		\gamma_d^2 & \sim \mathcal{IG}((\mathbf{c}_d \tr \bm{\alpha})^{-1}, \lambda), \\
-		\sigma_d^2 & \sim 1/\sigma_d^3.
-		\end{align*}
-		Variational posterior:
-		\begin{align*}
-		q(\bm{\beta}) & \overset{D}{=} \prod_{d=1}^D \mathcal{N}_p (\bm{\mu}_d, \bm{\Sigma}_d), \\
-		q(\bm{\gamma}^2) & \overset{D}{=} \prod_{d=1}^D \mathcal{GIG}\(-\frac{p+1}{2}, \lambda (\mathbf{c}_d \tr \bm{\alpha})^2, \delta_d\), \\
-		q(\bm{\sigma}^2) & \overset{D}{=} \prod_{d=1}^D \Gamma^{-1} \(\frac{n + 1}{2}, \zeta_d\).
-		\end{align*}
-		Parameter updates:
-		\begin{align*}
-		\bm{\Sigma}_d^{(h+1)} & = \(\frac{n+1}{2 \zeta_d^{(h)}}\X \tr \X + a_d^{(h)} \I_p\)^{-1}, \\
-		\bm{\mu}_d^{(h+1)} & = \frac{n+1}{2 \zeta_d^{(h)}} \bm{\Sigma}_d^{(h+1)} \X \tr \y_d, \\
-		\delta_d^{(h+1)} & = \trace(\bm{\Sigma}_d^{(h+1)}) + (\bm{\mu}_d^{(h+1)}) \tr \bm{\mu}_d^{(h+1)} + \lambda, \\ \zeta_d^{(h+1)} & = \frac{1}{2} \[\mathbf{y}_d \tr \mathbf{y}_d -2 \mathbf{y}_d \tr \X \bm{\mu}_d^{(h+1)} + \trace (\X \tr \X \bm{\Sigma}_d^{(h+1)}) + (\bm{\mu}_d^{(h+1)}) \tr \X \tr \X \bm{\mu}_d^{(h+1)}\],
-		\end{align*}
-		with $a_d^{(h)}$ and the empirical Bayes updates $\bm{\alpha}^{(l+1)}$ and $\lambda^{(l+1)}$ the same as before.
-		\section{Ratios of modified Bessel functions}
-		Ratios of modified Bessel functions of the second kind $K_{\alpha - 1}(x)/K_{\alpha}(x)$ are prone to under- and overflow for large $\alpha$.
-In our case, $\alpha$ increases linearly with $p$.
-Since $p$ may be large, this causes numerical issues in the calculation of various quantities.
-We alleviate the numerical issues through the following.
-
-		We let $n_1=p/2$ and $n_2=(p-1)/2$ and use the well-known recursive relation:
-		$$
-		K_{\alpha}(x) = K_{\alpha - 2}(x) + \frac{2(\alpha - 1)}{2} K_{\alpha- 1}(x),
-		$$
-		to rewrite the ratio:
-		\begin{align*}
-		\frac{K_{\frac{p - 1}{2}}(x)}{K_{\frac{p + 1}{2}}(x)} = 
-		\begin{cases}
-		\Big( \dots \Big( \( 1 + \frac{2 \cdot 1 - 1}{x} \)^{-1} + \frac{2 \cdot 2-1}{x}\Big)^{-1} + \dots + \frac{2 \cdot n_1 - 1}{x}\Big)^{-1}, & \text{for } p \text{ even}, \\
-		\Big( \dots \Big( \( \frac{K_0(x)}{K_1(x)} + \frac{2}{x} \cdot 1 \)^{-1} + \frac{2}{x} \cdot 2 \Big)^{-1} + \dots + \frac{2}{x} \cdot n_2 \Big)^{-1}, & \text{for } p \text{ odd}.
-		\end{cases}
-		\end{align*} 
-		The ratio $K_0(x)/K_1(x)$ is well-behaved, such that the ratio may be computed without numerical issues.
-
-		\section{General model}
-		We have drug covariates $\mathbf{c}_d$ and omics feature covariates $\mathbf{z}_j$. The model is now:
-		\begin{align*}
-		y_{id} | \bm{\beta}_d, \sigma_d^2 & \sim \mathcal{N} (\x_i \tr \bm{\beta}_d, \sigma_d^2), \\
-		\beta_{jd} | \tau_j^2, \gamma_d^2, \sigma_d^2 & \sim \mathcal{N} (0, \sigma_d^2 \gamma_d^2 \tau_j^2), \\
-		\gamma_d^2 & \sim \mathcal{IG}\((\mathbf{c}_d \tr \bm{\alpha})^{-1}, \lambda\), \\
-		\tau_j^2 & \sim \mathcal{IG}\((\mathbf{z}_j \tr \bm{\theta})^{-1}, \nu\), \\
-		\sigma_d^2 & \sim 1/\sigma_d^3.
-		\end{align*}
-		Variational posterior:
-		\begin{align*}
-		q(\bm{\beta}) & \overset{D}{=} \prod_{d=1}^D \mathcal{N}_p (\bm{\mu}_d, \bm{\Sigma}_d), \\
-		q(\bm{\gamma}^2) & \overset{D}{=} \prod_{d=1}^D \mathcal{GIG}\(-\frac{p+1}{2}, \lambda (\mathbf{c}_d \tr \bm{\alpha})^2, \delta_d\), \\
-		q(\bm{\tau}^2) & \overset{D}{=} \prod_{j=1}^p \mathcal{GIG}\(-\frac{D+1}{2}, \nu (\mathbf{z}_j \tr \bm{\theta})^2, \eta_j\), \\
-		q(\bm{\sigma}^2) & \overset{D}{=} \prod_{d=1}^D \Gamma^{-1} \(\frac{n + p + 1}{2}, \zeta_d\).
-		\end{align*}
-		Variational parameters:
-		\begin{align*}
-		\bm{\Sigma}_d^{(h+1)} & = \frac{2 \zeta_d^{(h)}}{n+p+1} \[\X \tr \X + a_d^{(h)} \cdot \diag(b_j^{(h)})\]^{-1}, \\
-		\bm{\mu}_d^{(h+1)} & = \[\X \tr \X + a_d^{(h)} \cdot \diag(b_j^{(h)})\]^{-1} \X \tr \y_d, \\
-		\delta_d^{(h+1)} & = \frac{n+p+1}{2 \zeta_d^{(h)}} \left\{\trace \[\diag(b_j^{(h)}) \bm{\Sigma}_d^{(h+1)} \] + (\bm{\mu}_d^{(h+1)}) \tr \diag(b_j^{(h)}) \bm{\mu}_d^{(h+1)}\right\} + \lambda, \\ 
-		\eta_j^{(h+1)} & = \sum_{d=1}^D \frac{a_d^{(h)}(n+p+1)}{2 \zeta_d^{(h)}} \[ (\bmu_{jd}^{(h + 1)})^2 + (\bSigma_d^{(h + 1)})_{jj} \] + \nu, \\ 
-		\zeta_d^{(h+1)} & = \frac{1}{2} \Big( \mathbf{y}_d \tr \mathbf{y}_d -2 \mathbf{y}_d \tr \X \bm{\mu}_d^{(h+1)} + a_d^{(h+1)} \trace \left\{ \[\X \tr \X + \diag(b_j^{(h + 1)})\] \bm{\Sigma}_d^{(h+1)}\right\} \\
-		& \,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\, + a_d^{(h+1)} (\bm{\mu}_d^{(h+1)}) \tr \[\X \tr \X + \diag(b_j^{(h + 1)})\]\bm{\mu}_d^{(h+1)}\Big),
-		\end{align*}
-		with
-		\begin{align*}
-		a_d^{(h)} & = \sqrt{\frac{\lambda (\mathbf{c}_d \tr \bm{\alpha})^2}{\delta_d^{(h)}}} \frac{K_{\frac{p-1}{2}} \( \sqrt{\lambda (\mathbf{c}_d \tr \bm{\alpha})^2 \delta_d^{(h)}} \)}{K_{\frac{p+1}{2}} \( \sqrt{\lambda (\mathbf{c}_d \tr \bm{\alpha})^2 \delta_d^{(h)}} \)} + \frac{p+1}{\delta_d^{(h)}} \text{ and} \\
-		b_j^{(h)} & = \sqrt{\frac{\nu (\mathbf{z}_j \tr \bm{\theta})^2}{\eta_j^{(h)}}} \frac{K_{\frac{D-1}{2}} \( \sqrt{\nu (\mathbf{z}_j \tr \bm{\theta})^2 \eta_j^{(h)}} \)}{K_{\frac{D+1}{2}} \( \sqrt{\nu (\mathbf{z}_j \tr \bm{\theta})^2 \eta_j^{(h)}} \)} + \frac{D+1}{\eta_j^{(h)}}.
-		\end{align*}
-		Hyper-parameter updates:
-		\begin{align*}
-		\bm{\alpha}^{(l+1)} & = \left[ \mathbf{C} \tr \diag ( e_d^{(l)} ) \mathbf{C} \right]^{-1} \mathbf{C} \tr \mathbf{1}, \\
-		\lambda^{(l+1)} & = \[ \sum_{d=1}^D a_d^{(l)} - (\bm{\alpha}^{(l+1)}) \tr \mathbf{C} \tr \mathbf{1} \]^{-1} D, \\
-		\bm{\theta}^{(l+1)} & = \left[ \mathbf{Z} \tr \diag ( f_j^{(l)} ) \mathbf{Z} \right]^{-1} \mathbf{Z} \tr \mathbf{1}, \\
-		\nu^{(l+1)} & = \[ \sum_{j=1}^p b_j^{(l)} - (\bm{\theta}^{(l+1)}) \tr \mathbf{Z} \tr \mathbf{1} \]^{-1} p,
-		\end{align*}
-		where 
-		\begin{align*}
-		e_d^{(l)} & = \sqrt{\frac{\delta_d^{(l)}}{\lambda^{(l)}(\mathbf{c}_d \tr \bm{\alpha}^{(l)})^2}} \frac{K_{\frac{p - 1}{2}} \( \sqrt{\lambda^{(l)} (\mathbf{c}_d \tr \bm{\alpha}^{(l)})^2 \delta_d^{(l)}} \)}{K_{\frac{p + 1}{2}} \( \sqrt{\lambda^{(l)} (\mathbf{c}_d \tr \bm{\alpha}^{(l)})^2 \delta_d^{(l)}} \)} \text{ and} \\
-		f_j^{(l)} & = \sqrt{\frac{\eta_j^{(l)}}{\nu^{(l)}(\mathbf{z}_j \tr \bm{\theta}^{(l)})^2}} \frac{K_{\frac{D - 1}{2}} \( \sqrt{\nu^{(l)} (\mathbf{z}_j \tr \bm{\theta}^{(l)})^2 \eta_j^{(l)}} \)}{K_{\frac{D + 1}{2}} \( \sqrt{\nu^{(l)} (\mathbf{z}_j \tr \bm{\theta}^{(l)})^2 \eta_j^{(l)}} \)}.
-		\end{align*}
-
 	\end{appendix}
@@
 \end{document}
-
-%		\section{Efficient computation}
-%		Let $\X = \U \D \Vm \tr$ and $\Z=\U^* \D^* (\Vm^*) \tr$ be the singular value decomposition of $\X$ and $\Z$, respectively.
-Filling in all the expectations, we have the following iterative estimating equations:
-%		\begin{align*}
-%		\bSigma_d^{(l+1)} & = a_3 b_{3d}^{(l)} \Vm (\D^2 + a_1 b_{1d}^{(l)} \I_n)^{-1} \Vm \tr, \\
-%		\Sm_d^{(l+1)} & = a_3 b_{3d}^{(l)} \Vm^* [(\D^*)^2 + a_2 b_{2d}^{(l)} \I_T]^{-1} (\Vm^*) \tr, \\
-%		\bmu_d^{(l+1)} & = a_3 b_{3d}^{(l)} \Vm (\D + a_1 b_{1d}^{(l)} \D^{-1})^{-1} \U \tr (\y_d - \Z \mathbf{m}_d^{(l)}), \\
-%		\mathbf{m}_d^{(l+1)} & = a_3 b_{3d}^{(l)} \Vm^* [\D^* + a_2 b_{2d}^{(l)} (\D^*)^{-1}]^{-1} (\U^*) \tr (\y_d - \X \bmu_d^{(l + 1)}), \\
-%		b_{1d}^{(l+1)} & = \Bigg\{\frac{a_3 b_{3d}^{(l)}}{2} \Bigg[ (\y_d - \Z \mathbf{m}_d^{(l)}) \tr \U (\D + a_1 b_{1d}^{(l)} \D^{-1})^{-2} \U \tr (\y_d - \Z \mathbf{m}_d^{(l)}) \\
-%		\,\,\,\,\,\,\,\,\,\, & + a_3 b_{3d}^{(l)} \sum_{i=1}^n (d_i^2 + a_1 b_{1d}^{(l)})^{-1} \Bigg] + \theta_1^{-1} \Bigg\}^{-1}, \\
-%		b_{2d}^{(l+1)} & = \Bigg(\frac{a_3 b_{3d}^{(l)}}{2} \Bigg\{ (\y_d - \X \bmu_d^{(l+1)}) \tr \U^* [\D^* + a_2 b_{2d}^{(l)} (\D^*)^{-1}]^{-2} (\U^*) \tr (\y_d - \X \bmu_d^{(l+1)}) \\
-%		\,\,\,\,\,\,\,\,\,\, & + a_3 b_{3d}^{(l)} \sum_{i=1}^n [(d^*_i)^2 + a_2 b_{2d}^{(l)}]^{-1} \Bigg\} + \theta_2^{-1} \Bigg)^{-1}, \\
-%		b_{3d}^{(l+1)} & = \frac{\y_d \tr \y_d}{2}  - \y_d \tr (\X \bmu_d^{(l+1)} + \Z \mathbf{m}_d^{(l+1)}) + (\bmu_d^{(l+1)}) \tr \X \tr \Z \mathbf{m}_d^{(l+1)} + \frac{\sum_{i=1}^n \(d_i^2 - \frac{d_i^4}{d_i^2 + a_1 b_1^{(l+1)}}\)}{2 a_1 b_{1d}^{(l + 1)}} \\
-%		\,\,\,\,\,\,\,\,\,\, & + \frac{(\bmu_d^{(l)}) \tr \X \tr \X \bmu_d^{(l)}}{2} + \frac{\sum_{t=1}^T \[(d^*_t)^2 - \frac{(d^*_t)^4}{(d^*_t)^2 + a_2 b_2^{(l+1)}}\]}{2 a_2 b_{2d}^{(l + 1)}} + \frac{(\mathbf{m}_d^{(l)}) \tr \Z \tr \Z \mathbf{m}_d^{(l)}}{2} \\
-%		& \,\,\,\,\,\,\,\,\,\, + \frac{a_1 b_{1d}^{(l+1)} (\y_d - \Z \mathbf{m}_d^{(l)}) \tr \U (\D + a_1 b_{1d}^{(l)} \D^{-1})^{-2} \U \tr (\y_d - \Z \mathbf{m}_d^{(l)})}{2} \\
-%		& \,\,\,\,\,\,\,\,\,\, + \frac{a_1 b_{1d}^{(l+1)} a_3 b_{3d}^{(l)} \sum_{i=1}^n (d_i^2 + a_1 b_{1d}^{(l)})^{-1} }{2} \\
-%		& \,\,\,\,\,\,\,\,\,\, + \frac{a_2 b_{2d}^{(l+1)} (\y_d - \X \bmu_d^{(l+1)}) \tr \U^* [\D^* + a_2 b_{2d}^{(l)} (\D^*)^{-1}]^{-2} (\U^*) \tr (\y_d - \X \bmu_d^{(l+1)})}{2} \\
-%		& \,\,\,\,\,\,\,\,\,\, + \frac{a_1 b_{1d}^{(l+1)} a_3 b_{3d}^{(l)} \sum_{i=1}^n [(d^*_i)^2 + a_2 b_{2d}^{(l)}]^{-1}}{2},
-%		\end{align*}
-%		where $(\D^2 + a_1 b_{1d}^{(l)})^{-1}$, $[(\D^*)^2 + a_2 b_{2d}^{(l)}]^{-1}$, $(\D + a_1 b_{1d}^{(l)} \D^{-1})^{-1}$, and $[\D^* + a_2 b_{2d}^{(l)} (\D^*)^{-1}]^{-1}$ are diagonal matrices with entries $(d_t^2 + a_1 b_{1d}^{(l)})^{-1}$, $i=1,\dots,n$, $[(d_t^*)^2 + a_2 b_{2d}^{(l)}]^{-1}$, $t=1,\dots,T$, $(d_i + a_1 b_{1d}^{(l)} / d_i)^{-1}$, $i=1, \dots, n$, and $(d^*_t + a_2 b_{2d}^{(l)} / d^*_t)^{-1}$, $t=1 , \dots, T$, respectively, and hence easily calculated once the singular values are known.

### 2018-12-20 docs/supplement.Rnw
created supplementary material with first simulation results

--- 2018-12-18 docs/manuscript.Rnw
@@
 
+% R packages
+<<echo=FALSE>>=
+library(cambridge)
+@
+
 % settings
@@
 \setlength{\textwidth}{15cm}
-\title{Drug efficacy prediction in cell lines}
+\title{Supplementary material to: Drug efficacy prediction in cell lines}
 \date{\today}
@@
 
-	\begin{abstract}
-		{...}
-	\end{abstract}
+	\section{Introduction}
+	This document contains the Supplementary Material (SM) to the document `Drug efficacy prediction in cell lines'. In the following, this document is referred to as Main Document (MD).
 
-	\noindent\textbf{Keywords}:
-
-	\noindent\textbf{Software available from}:
-
-	\section{Introduction}
-	- motivation: drug response prediction from different omics data types is promising both for precision medicine and drug target discovery and relates to actual tumor response \cite[]{iorio_landscape_2016}.
-
-	- problem: prediction of drug efficacy in cell lines is difficult (DREAM 7 challenge in \cite{costello_community_2014}).
-
-	- example data: introduction (GDSC, CCLE, NCI-60)
-
-	- feature information: prior experts, pathways, database summaries, previous experiments, type of feature (somatic mutations, copy number alterations, DNA methylation, and gene expression).
-
-	- drugs information: molecular target, drug class, etc (ask Oslo).
-
-	- current literature: review in \cite{azuaje_computational_2017}, reference Oslo, TANDEM \cite[]{aben_tandem:_2016}, DeltaNet \cite[]{noh_inferring_2016}, patient-derived tumor cell lines \cite[]{gao_high-throughput_2015}.
-
-	- our solution: adaptive borrowing of information through empirical Bayes, guided by external feature and drug information.
-
-	\section{Model}
-	\subsection{Simultaneous equations model}\label{sec:SEM}
-	- Gwen can deal with different number of features per equation, might solve missing drug-feature combination problem? Doesn't solve missing response-feature combination problem.
-
-	- check literature for: a) conjugate version, b) similar models: "prior variance decomposition", "structured Bayesian regression"
-
-	We have continuous efficacy measures $y_{id}$ for cell lines $i=1,\dots, n$, from tissues $t=1, \dots, T$, on drugs $d=1,\dots,D$.
-Throughout this paper we assume the $y_{id}$ to be centred per drug and let $\y_d = \begin{bmatrix} y_{1d} & \dots & y_{nd} \end{bmatrix} \tr$.
-We predict efficacy with molecular features $x_{ij}$, $j=1,\dots, p$, collected in $\x_i = \begin{bmatrix} x_{i1} & \dots & x_{ip} \end{bmatrix} \tr$.
-We have an \textit{a priori} partitioning of the omics features into $g=1, \dots, G$ non-overlapping groups, which might be informative for the problem at hand For convenience, we let $g(j)$ denote the group index of molecular feature $j$.
-Furthermore, we have characteristics of the drugs available in the form of `co-data' $\mathbf{C} = \begin{bmatrix} \mathbf{c}_1 & \dots & \mathbf{c}_D \end{bmatrix} \tr$.
-We code the tissue from which the cell lines were taken as a $T$ binary dummy variables
-	$$
-	z_{it} = \begin{cases}
-	1 & \text{if } \text{tissue}_i = t, \\
-	0 & \text{otherwise},
-	\end{cases}
-	$$
-	collected in $\mathbf{z}_i = \begin{bmatrix} z_{i1} & \dots & z_{iT} \end{bmatrix} \tr$.
-
-	We assume that cell lines from the same tissue are related and model this relation through random effects $u_{td}$.
-We let the molecular feature effects $\beta_{jd}$ vary over the drugs and model drug efficacy as a linear combination of these two effects:
-	\begin{subequations}\label{eq:linearmodel}
-		\begin{align}
-		y_{id} & = \beta_{0d} + \sum_{j=1}^p x_{ij} \beta_{jd} + \sum_{t=1}^T z_{it} u_{td} + \epsilon_{id} \\
-		& = \beta_{0d} + \x_i \tr \bbeta_d + \mathbf{z}_i \tr \mathbf{u}_d + \epsilon_{id},\\
-		\text{with } & \epsilon_{id} \sim \mathcal{N}(0, \sigma_d^2),
-		\end{align}
-	\end{subequations}
-	where $p$-dimensional $\bbeta_d$ and $T$-dimensional $\bm{u}_d$ are the drug-specific omics feature and tissue effect vectors, respectively.
-Note that (\ref{eq:linearmodel}) gives rise to a system of $d=1, \dots, D$ linear simultaneous equations.
-
-	\subsection{Bayesian prior model}
-	We capture the uncertainty in the parameters through a Bayesian prior model.
-We assign a flat prior to the intercept terms $\beta_{0d}$ and integrate them out.
-The remaining parameters are endowed with the following priors:
-	\begin{subequations}\label{eq:prior}
-		\begin{align}
-		\beta_{jd} & \sim \mathcal{N}_p (0, \phi^2_{g(j)} \gamma_d^2 \sigma_d^2), \\
-		\mathbf{u}_{d} & \sim \mathcal{N}_T (\mathbf{0}, \bm{\Xi}_d \sigma_d^2),
-		\end{align}
-	\end{subequations}
-	and hyper-priors:
-	\begin{subequations}\label{eq:hyperprior}
-		\begin{align}
-			\sigma_d^{2} & \sim 1/\sigma_d^{3} \\
-			\gamma_d^{2} & \sim \mathcal{IG}(\theta_d, \lambda), \\
-			\bm{\Xi}_d & \sim \mathcal{W}^{-1}_T(\bm{\Omega}, \nu),
-		\end{align}
-	\end{subequations}
-	where $\mathcal{IG}(\theta, \lambda)$ is an inverse Gaussian distribution with mean $\theta$ and shape $\lambda$, and $\mathcal{W}^{-1}_T(\bm{\Omega}, \nu)$ is an inverse Wishart distribution with $T \times T$-dimensional positive definite (PD) scale matrix $\bm{\Omega}$ and degrees of freedom $\nu$.
-Note that for $\nu > T + 1$, this inverse Wishart prior implies an expected prior covariance matrix $(\nu - T - 1)^{-1} \bm{\Omega}$ for the tissue effects, common to all drugs.
-Likewise, the molecular feature effects are related through the common hyper-prior for the $\gamma_d^2$.
-
-	- Jeffrey's prior for variance $1/\sigma^{3/2}$ (a priori independent data mean and variance) or $1/\sigma^3$ (joint prior, but only mean and variance, not variance and betas).
-
-	A few remarks on the choice of priors are justified here: many authors endow error variance components with vague gamma priors.
-\cite{gelman_prior_2006}, among others, advises against this practice.
-The degree of `vagueness' has a large influence on the posterior, while degree of `vagueness' is a difficult parameter to set.
-This influence is especially pronounced if the likelihood is relatively flat, as may be reasonably expected in the large $p$, small $n$ setting we are in.
-We therefore model the error variance with Jeffrey's prior \cite[]{jeffreys_invariant_1946}.
-
-	Furthermore, we choose to model the $\gamma^2_d$ by an inverse Gaussian distribution, as has been suggested in \cite{fabrizi_specification_2016} and \cite{caron_sparse_2008}, because it allows to model the mean $\theta_d$ as a function of the drug covariates $\mathbf{c}_d$, as explained in Section \ref{sec:empiricalbayes}.
-
-	- IG hyperprior used in finance \cite[]{barndorff-nielsen_normal_1997}, where it is called the NIG prior.
-%	\begin{subequations}\label{eq:hyperprior}
-%		\begin{align}
-%		\sigma_d^{2} & \sim 1/\sigma_d^{3} \\
-%		\log \gamma_d^{2} & \sim \mathcal{N}(\theta, \tau^2), \\
-%		\bm{\Xi}_d & \sim \mathcal{W}^{-1}_T(\Vm, k),
-%		\end{align}
-%	\end{subequations}
-%- try log-normal distribution for the $\gamma_d$'s.
-This is used in Example 10.4.1 in \cite{lunn_bugs_2012}, with justification same as ours: allows for modelling the variance as a function of covariates.
-Also used in examples 6.10 and 9.2 in \cite{abrams_bayesian_2004}.
-
-%	\begin{subequations}\label{eq:hyperprior}
-%		\begin{align}
-%		\sigma_d^{-2} & \sim \Gamma(0.001, 0.001), \\
-%		\gamma_d^{-2} & \sim \Gamma(k, \theta), \\
-%		\bm{\Xi}_d & \sim \mathcal{W}^{-1}_T(\Vm, \tau),
-%		\end{align}
-%	\end{subequations}
-%	where $g(j)$ denotes the group index of molecular feature $j$, $\Gamma(k, \theta)$ is a gamma distribution with shape $k$ and rate $\theta$, and $\mathcal{W}_T(\Vm, \tau)$ is a Wishart distribution with $T \times T$-dimensional scale matrix $\Vm$ and degrees of freedom $\tau$.
-Note that this Wishart prior implicates an expected prior covariance matrix $\tau \cdot \Vm$ for the tissue effects, common to all drugs.
-
-	\section{Estimation}
-	\subsection{Variational Bayes}\label{sec:variationalbayes}
-	The posterior corresponding to the model described in (\ref{eq:linearmodel}), (\ref{eq:prior}), and (\ref{eq:hyperprior}) is not available in closed form.
-We therefore approximate the posterior by variational Bayes, where we force the posterior density $Q$ to factorise as:
-	\begin{align}\label{eq:variationalposterior}
-	p(\B,\mathbf{U},\mathbf{\sigma}^2, \bgamma^2,\bm{\Xi}_1, \dots, \bm{\Xi}_D) \approx Q(\cdot) = q(\B) \cdot q(\mathbf{U}) \cdot q(\mathbf{\sigma}^2) \cdot q(\bgamma^2) \cdot q(\bm{\Xi}_1, \dots, \bm{\Xi}_D),
-	\end{align}
-	where $\B = \begin{bmatrix} \bbeta_1 & \cdots & \bbeta_D \end{bmatrix}$ and $\U = \begin{bmatrix} \mathbf{u}_1 & \cdots & \mathbf{u}_D \end{bmatrix}$, are of dimensions $p \times D$ and $T \times D$, respectively.
-Under the factorisation in (\ref{eq:variationalposterior}), the marginal variational posteriors that minimise the Kullback-Leibler divergence of the true posterior to the variational Bayes approximation \cite[]{neal_view_1998}, are given by:
-	\begin{subequations}\label{eq:variationalposterior2}
-		\begin{align}
-		q_{\B} (\B) & \overset{D}{=} \prod_{d=1}^{D} \mathcal{N}_p (\bmu_d, \bSigma_d), \\
-		q_{\U}(\U) & \overset{D}{=} \prod_{d=1}^{D} \mathcal{N}_T (\mathbf{m}_d, \Sm_d), \\
-		q_{\bgamma^{2}}(\bgamma^{2}) & \overset{D}{=} \prod_{d=1}^D \mathcal{GIG} \(-\frac{p+1}{2}, \frac{\lambda}{\theta_d^2}, \delta_{d} \), \\
-		q_{\bsigma^{2}}(\bsigma^{2}) & \overset{D}{=} \prod_{d=1}^D \Gamma^{-1} \(\frac{n + p + T + 1}{2}, \zeta_{d} \), \\
-		q_{\bm{\Xi}_1, \dots, \bm{\Xi}_D}(\bm{\Xi}_1, \dots, \bm{\Xi}_D) & \overset{D}{\propto} \prod_{d=1}^D \mathcal{W}_T^{-1} (\bm{\Psi}_d, \nu + 1).
-		\end{align}
-	\end{subequations}
-	Here $\mathcal{GIG}(\cdot)$ and $\Gamma^{-1}(\cdot)$ denote the generalized inverse Gaussian and inverse gamma distributions, respectively.
-The full derivations may be found in Appendix \ref{app:derivations}.
-
-	The parameters in (\ref{eq:variationalposterior2}) contain cyclic dependencies. We therefore iterate the following estimating equations until convergence to a local optimum:
-	\begin{align*}
-	\bSigma_d^{(h+1)} & = \frac{2\zeta_d^{(h)}}{n+p+T+1} \left\{\X \tr \X + a_d^{(h)} \cdot \diag (\phi_{g(j)}^{-2}) \right\}^{-1}, \\
-	\bm{\mu}_d^{(h+1)} & = \left\{\X \tr \X + a_d^{(h)} \cdot \diag (\phi_{g(j)}^{-2}) \right\}^{-1} \X \tr ( \y_d - \Z \mathbf{m}_d^{(h)}), \\
-	\Sm_d^{(h+1)} & = \frac{2\zeta_d^{(h)}}{n+p+T+1} \[ \Z \tr \Z + (\nu + 1) \cdot (\bm{\Psi}_d^{(h)})^{-1} \]^{-1}, \\
-	\mathbf{m}_d^{(h+1)} & = \[ \Z \tr \Z + (\nu + 1) \cdot (\bm{\Psi}_d^{(h)})^{-1} \]^{-1} \Z \tr (\y_d - \X \bm{\mu}_d^{(h)}), \\
-	\bm{\Psi}_d^{(h+1)} & = \frac{2\zeta_d^{(h)}}{n+p+T+1} \[ \Sm_d^{(h+1)} + \mathbf{m}_d^{(h+1)} (\mathbf{m}_d^{(h+1)}) \tr \] + \bm{\Omega}, \\
-	\delta_d^{(h+1)} & =  \frac{n+p+T+1}{2\zeta_d^{(h)}} \sum_{j=1}^p \frac{(\mu_{jd}^{(h+1)})^2 + (\bSigma_d^{(h+1)})_{jj}}{\phi^2_{g(j)}} + \lambda, \\
-	\zeta_d^{(h+1)} & = \frac{1}{2} \y_d \tr \y_d - \y_d \tr (\X \bmu_d^{(h+1)} + \Z \mathbf{m}_d^{(h+1)}) + (\bmu_d^{(h+1)}) \tr \X \tr \Z \mathbf{m}_d^{(h+1)} + \frac{1}{2} \trace (\X \tr \X \bSigma_d^{(h+1)}) \\
-	& \,\,\,\,\,\,\,\,\,\, + \frac{1}{2} (\bmu_d^{(h+1)}) \tr \X \tr \X \bmu_d^{(h+1)} + \frac{1}{2} \trace (\Z \tr \Z \Sm_d^{(h+1)}) + \frac{1}{2} (\mathbf{m}_d^{(h+1)}) \tr \Z \tr \Z \mathbf{m}_d^{(h+1)} \\
-	& \,\,\,\,\,\,\,\,\,\, + \frac{1}{2} a_d^{(h)} \sum_{j=1}^p \frac{(\mu_{jd}^{(h+1)})^2 + (\bSigma_d^{(h+1)})_{jj}}{\phi^2_{g(j)}} + \frac{1}{2} (\nu + 1) \cdot \trace \[ (\bm{\Psi}_d^{(h+1)})^{-1} \Sm_d^{(h+1)} \] \\
-	& \,\,\,\,\,\,\,\,\,\, + \frac{1}{2} (\nu + 1) \cdot (\mathbf{m}_d^{(h+1)}) \tr (\bm{\Psi}_d^{(h+1)})^{-1} \mathbf{m}_d^{(h+1)},
-	\end{align*}
-	where
-	$$
-	a_d^{(h)} = \sqrt{\frac{\lambda}{\theta^2_d \delta_d^{(h)}}} \frac{K_{\frac{p - 1}{2}} \( \sqrt{\lambda \delta_d^{(h)}}/\theta_d \)}{K_{\frac{p+1}{2}} \( \sqrt{\lambda \delta_d^{(h)}}/\theta_d \)} + \frac{p+1}{\delta_d^{(h)}},
-	$$
-	and $K_{\alpha}(\cdot)$ is the modified Bessel function of the second kind.
-
-	\subsection{Empirical Bayes}\label{sec:empiricalbayes}
-	A full Bayesian model requires the specification of hyper-parameters $\bm{\eta} = \begin{bmatrix} (\bm{\phi}^2) \tr & \bm{\theta} \tr & \lambda & \text{vec}(\bm{\Omega}) \tr & \nu \end{bmatrix} \tr$.
-These are abstract and hard to interpret parameters for which we generally lack expert knowledge.
-They do, however, have a significant influence on the shape of the posterior distribution.
-We therefore propose to estimate the unspecified hyper parameters $\bm{\eta}$ by empirical Bayes.
-Simply put, empirical Bayes fits the prior model to the data and is, as such, objective (up to model specification).
-
-	The canonical method for empirical Bayes is maximisation of the marginal likelihood with respect to the hyper parameters.
-In \cite{casella_empirical_2001} the marginal likelihood is maximised by an EM algorithm:
-	$$
-	\bm{\eta}^{(l+1)} = \underset{\bm{\eta}}{\argmax}\E_{\cdot | \Y} [\log p(\Y, \B, \U, \bm{\Xi}_1, \dots, \bm{\Xi}_D, \bgamma^2, \bsigma^2) | \bm{\eta}^{(l)}],
-	$$
-	where the expectation is with respect to the posterior.
-In our case, this posterior is not available in closed form, which renders the expectation difficult.
-While \cite{casella_empirical_2001} suggests to approximate the expectation by a Monte Carlo sample, we propose to use the variational Bayes approximation developed in Section \ref{sec:variationalbayes}:
-	\begin{align}\label{eq:variationalbayes}
-	\bm{\eta}^{(l+1)} & = \underset{\bm{\eta}}{\argmax}\E_{Q^{(l)}} [\log p(\Y, \B, \U, \bm{\Xi}_1, \dots, \bm{\Xi}_D, \bgamma^2, \bsigma^2)] \nonumber \\
-	& = \underset{\bm{\eta}}{\argmax} \E_{Q^{(l)}} [\log \pi (\B | \bgamma^2, \bsigma^2)] + \E_{Q^{(l)}} [\log \pi (\bgamma^2)] + \E_{Q^{(l)}} [\log \pi (\bm{\Xi}_1, \dots, \bm{\Xi}_D)].
-	\end{align}
-	where now the expectation is with respect to the converged variational posterior $Q^{(l)}$.
-Inspection of (\ref{eq:variationalbayes}) learns us that the problem may be decomposed into three separate optimisation problems.
-
-	\subsubsection{Group-specific variance components}
-	The first sub-problem concerns the group-specific variance components of the molecular feature effects.
-In order to separate the overall variance of the molecular feature effects from the group-specific deviations, we impose $\prod_{g=1}^G (\phi_g^2)^{|\mathcal{G}_g|} = 1$.
-As a consequence, the overall level of shrinkage is determined by the $\gamma^2_d$, while the $\phi_g^2$ effectuate differential shrinkage of the groups.
-Because the $\phi_g^2$ are variance components, we additionally require $\phi^2_1, \dots, \phi^2_G> 0$, such that our first optimisation sub-problem becomes:
-	\begin{align*}
-	(\bm{\phi}^2)^{(l+1)} & = \underset{\bm{\phi}^2}{\argmax} \E_{Q^{(l)}} [\log \pi (\B | \bgamma^2, \bsigma^2)] \\
-	& = \underset{\bm{\phi}^2}{\argmax} -\frac{D}{2} \sum_{g=1}^G |\G_g| \log \phi_g^2 - \frac{1}{2} \sum_{g=1}^G b_g^{(l)} \phi_g^{-2} \\
-	& \text{subject to } \prod_{g=1}^G (\phi_g^2)^{|\G_g|} = 1 \text{ and } \phi^2_1, \dots, \phi^2_G> 0.
-	\end{align*}
-	where 
-	$$
-	b_g^{(l)}=\sum_{d=1}^D a_d^{(l)} \frac{n+p+T+1}{2\zeta_d^{(l)}} \sum_{j \in \mathcal{G}_g} \[ (\mu_{jd}^{(l)})^2 + (\bSigma_d^{(l)})_{jj} \].
-	$$
-	This is a convex optimisation problem and easily solved numerically.
-
-	\subsubsection{Drug-specific variance components}
-	The second sub-problem concerns the prior for the drug-specific variance components $\gamma_d^2$.
-As explained in Section \ref{sec:SEM}, we may have `covariates' $\mathbf{c}_d$ on the drugs available.
-We conjecture that these covariates may be informative for the omics effect size and therefore model the prior mean of the $\gamma_d^2$ as a function of these covariates: $\theta_d = (\mathbf{c}_d \tr \bm{\alpha})^{-1}$.
-Then, the empirical Bayes estimation of $\bm{\alpha}$ effectively boils down to an inverse Gaussian regression of the $\gamma_d^{-2}$ on the $\mathbf{c}_d$ \cite[]{fries_optimal_1986,whitmore_regression_1983}.
-The problem may be formulated as:
-	\begin{align*}\label{eq:optproblem2}
-	\lambda^{(l+1)}, \bm{\alpha}^{(l+1)} & = \underset{\lambda,\bm{\alpha}}{\argmax} \E_{Q^{(l)}} [\log \pi (\bgamma^2)] \\
-	%	& = \underset{\lambda,\bm{\alpha}}{\argmax} \frac{D}{2} \log \lambda - \frac{\lambda}{2} \sum_{d=1}^D \[ \E(\gamma_d^2)(\mathbf{c} _d \tr \bm{\alpha})^2 - 2 \mathbf{c}_d \tr \bm{\alpha} + \E(\gamma_d^{-2}) \] \\
-	& = \underset{\lambda,\bm{\alpha}}{\argmax} \frac{D}{2} \log \lambda - \frac{\lambda}{2} \bm{\alpha} \tr \mathbf{C} \tr \diag [\E_{Q^{(l)}}(\gamma_d^2)] \mathbf{C} \bm{\alpha} + \lambda \bm{\alpha} \tr \mathbf{C} \tr \mathbf{1} - \frac{\lambda}{2} \sum_{d=1}^D \E_{Q^{(l)}}(\gamma_d^{-2}),
-	\end{align*}
-	where $\mathbf{C} = \begin{bmatrix} \mathbf{c}_1 & \dots & \mathbf{c}_D \end{bmatrix} \tr$. The solutions to this problem are:
-	\begin{align*}
-	\bm{\alpha}^{(l+1)} & = \left\{ \mathbf{C} \tr \Vm^{(l)} \mathbf{C} \right\}^{-1} \mathbf{C} \tr \mathbf{1}, \\
-	\lambda^{(l+1)} & = \[ \sum_{d=1}^D a_d^{(l)} - (\bm{\alpha}^{(l+1)}) \tr \mathbf{C} \tr \mathbf{1} \]^{-1} D,
-	\end{align*}
-	where 
-	$$
-	\Vm^{(l)} = \diag \[ \sqrt{\frac{\delta_d^{(l)}(\theta_d^{(l)})^2}{\lambda^{(l)}}} \frac{K_{\frac{p - 1}{2}} \( \sqrt{\lambda^{(l)} \delta_d^{(l)}}/\theta^{(l)}_d \)}{K_{\frac{p + 1}{2}} \( \sqrt{\lambda^{(l)} \delta_d^{(l)}}/\theta^{(l)}_d \)} \].
-	$$
-
-	- Think about parametrisation of $\mathbf{C}$.
-Idea: use overall mean and deviations from it in alpha, such that the starting value can be the mean and 0 for alpha.
-Or: use intercept and code the rest as dummy variables.
-
-	\subsubsection{Tissue effect covariance}
-	The third sub-problem pertains to the tissue effect covariance prior.
-To avoid overfitting when freely estimating the $\frac{T(T+1)}{2}$ elements in $\bm{\Omega}$ and $\nu$, we parametrise them as $\bm{\Omega}=  \tau [(\tau - \rho) \I_T + \rho \mathbf{1}_{T \times T}]$ and $\nu = \tau + T + 1$, respectively.
-The justification for this parametrisation is three-fold: (a) it implies an easy to interpret prior covariance matrix with $\tau$ on its diagonal and $\rho$ on its off-diagonals, (b) it reduces the number of parameters to  estimate to just two, and (c) simplifies the positive definite constraint on  $\bm{\Omega}$.
-To ensure positive definiteness and a proper covariance matrix, we impose the constraints $\tau > \max\[\rho - T \rho,\rho, 0\]$ (see Appendix \ref{app:priortissuecovariance}).
-The resulting third sub-problem is a bit more involved than the first two:
-	\begin{align*}
-	\tau^{(l + 1)}, \rho^{(l+1)} & = \underset{\tau, \rho}{\argmax} \frac{D (T^2 + T)}{2} \log \tau + \frac{DT}{2} \tau \log \tau + \frac{D(T+1)}{2} \log \[ \tau - (T - 1) \rho\] \\
-	& \,\,\,\,\,\,\,\,\,\, + \frac{D}{2} \tau \log \[ \tau - (T - 1) \rho\] + \frac{D(T^2 - 1)}{2} \log(\tau - \rho) +\frac{D(T - 1)}{2} \tau  \log(\tau - \rho) \\
-	& \,\,\,\,\,\,\,\,\,\, - D \sum_{t=1}^T \log \Gamma \( \frac{\tau + t + 1}{2} \) - w_1^{(l)} \tau^2 + w_2^{(l)} \tau \rho + w_3^{(l)} \tau, \\
-	& \text{subject to } \tau > \max\[\rho - T \rho,\rho, 0\],
-	\end{align*}
-	where
-	\begin{align*}
-	w_1^{(l)} & =  \frac{\tau^{(l)} + T + 2}{2} \sum_{d=1}^D \trace \[ (\bm{\Psi}_d^{(l)})^{-1}\] > 0,\\
-	w_2^{(l)} & = \frac{\tau^{(l)} + T + 2}{2} \sum_{d=1}^D \left\{\trace \[ (\bm{\Psi}_d^{(l)})^{-1}\] - \trace \[\mathbf{1}_{T \times T} \cdot (\bm{\Psi}_d^{(l)})^{-1}\] \right\} < w_1^{(l)},\\
-	w_3^{(l)} & = \frac{1}{2} \[D \sum_{t=1}^T \psi\(\frac{\tau^{(l)} + t + 2}{2}\) - \sum_{d=1}^D \log |\bm{\Psi}_d^{(l)}| \].
-	\end{align*}
-
-	- use mathematica to check convexity
-
-	- we use the ELBO for VB convergence and parameters for EB convergence
-
-	- we might be lenient for VB convergence. The extreme case would be 1 VB iteration per EB iteration (corresponds to tolerance of infinity).
-
-%	\begin{align*}
-%	\tau^{(l+1)}, \rho^{(l+1)} & = \underset{\tau, \rho}{\argmax} \E_{Q^{(l)}} [\log \pi (\bm{\Xi}_1, \dots, \bm{\Xi}_D)] \\
-%	& = \underset{\tau, \rho}{\argmax} w_1^{(l)} \tau - w_2^{(l)} \tau^2 - w_3^{(l)} \tau \rho - D \log \Gamma_T \(\frac{\tau}{2}\) + \frac{DT}{2} \tau \log \tau \\
-%	& \,\,\,\,\,\,\,\,\,\,+ \frac{D}{2} \tau \log \[ \tau - (1-T)\rho \] + \frac{D(T-1)}{2} \tau \log (\tau - \rho), \\
-%	& \text{subject to } \tau > \max\[\rho - T \rho,\rho, 0\] \text{ and } 1> |\rho| \geq 0,
-%	\end{align*}
-%	where 
-%	\begin{align*}
-%	w_1^{(l)} & = \frac{1}{2} \[D \psi_T\(\frac{\tau^{(l)} + 1}{2}\) + \sum_{d=1}^D \log |\bm{\Psi}_d^{(l)}| \], \\
-%	w_2^{(l)} & = \frac{\tau^{(l)} + 1}{2} \sum_{d=1}^D \trace \(\bm{\Psi}_d^{(l)}\) > 0,\\
-%	w_3^{(l)} & = \frac{\tau^{(l)} + 1}{2} \sum_{d=1}^D \trace \[(\mathbf{1}_{T \times T} - \I_T) \cdot \bm{\Psi}_d^{(l)}\],
-%	\end{align*}
-%	and $\Gamma_T(x)$ and $\psi_T(x)$ are the $T$-variate gamma and digamma functions, respectively. $\log \Gamma_T \(\frac{\tau}{2}\)$ is convex on the intervals $(0,1), \dots, (T-1, \infty)$
-
-%	- we model the prior expectation of the covariance instead of the precision. Precisions are conditional on the other variables, which doesn't make sense in the tissue context.
-
-%	The second sub problem estimates the molecular feature effect hyper prior:
-%	\begin{align}\label{eq:optproblem2}
-%	k^{(l+1)},\theta^{(l+1)} & = \underset{k,\theta}{\argmax} \E_{Q^{(l)}} [\log \pi (\bgamma^2)] \nonumber\\
-%	& = \underset{k,\theta}{\argmax} D k \log \theta - D \log \Gamma (k) + e_1^{(l)} k - e_2^{(l)} \theta \\
-%	& \text{subject to } k ,\theta > 0,
-%	\end{align}
-%	where
-%	\begin{align*}
-%	e_1^{(l)} & = D\psi\(\frac{p}{2} + k^{(l)}\) - \sum_{d=1}^D \log a_d^{(l)}, \\
-%	e_2^{(l)} & = \(\frac{p}{2} + k^{(l)} \) \sum_{d=1}^D (a_d^{(l)})^{-1},
-%	\end{align*}
-%	and $\Gamma(x)$ and $\psi(x)$ are the gamma and digamma functions, respectively.
-This is again a convex problem (see Appendix \ref{app:convexity}), but more cumbersome to solve numerically due to the log-gamma function.
-By approximating the digamma function with $\phi(x) \approx \log x - 1/(2x)$, we get the following closed-form solutions:
-%	\begin{align*}
-%	k^{(l+1)} & = \frac{1}{2} \( \log e_2^{(l)} - D^{-1} e_1^{(l)} - \log D \)^{-1}, \\
-%	\theta^{(l+1)} & = D k^{(l+1)} (e_2^{(l)})^{-1}.
-%	\end{align*}
-
-	\section{Simulations}
-
-	- Hamming distance to assess variable selection
-
-	\section{Data application}
-	- check GSK open target EBI or ask Chris Wallace about it. Database that links targets to drugs and diseases. Might be useful for drugs co-data.
-
-	- check NCI-60
-
-	- how to transform the data and deal with missingness (ask Oslo and read papers published on GDSC and CCLE repositories)
-
-	- what information is available on drugs (ask Oslo)
-
-	- start with gene expression (one $g(j)$)
-
-	\subsection{GDSC}
-
-	\subsection{CCLE}
-
-	\section{Discussion}
-	- empirical Bayes gives extra information via estimation of the hyper-parameters, such as informativeness of the drug information (via model for the $\gamma_d$) and relatedness of the tissues (via $\bm{\Omega}$ and $\nu$)
-
-	- regression on drug covariates very intuitive.
-
-	- find more literature on inverse Gaussian prior for variances and modelling variances with covariates.
-
-	\bibliographystyle{author_short3} 
-	\bibliography{refs}
-
-	\titleformat{\section}{\normalfont\Large\bfseries}{\appendixname~\thesection.}{1em}{}
-	\begin{appendix}
-		\section{Variational Bayes derivations}\label{app:derivations}
+	\section{Variational Bayes derivations}\label{sec:vbderivations}
 	In the following all expectations are with respect to the variational posterior $Q$.
@@
 
-%		\begin{align*}
-%		\log q_{\bgamma^{-2}}(\bgamma^{-2}) & \propto \E_{\B, \bsigma^2} [\log \pi (\B | \bgamma^{-2}, \sigma^2)] + \log \pi (\bgamma^{-2}) \\
-%		& \propto \frac{p}{2} \sum_{d=1}^D \log \gamma_d^{-2} - \frac{1}{2} \sum_{d=1}^D \sum_{j=1}^p \phi_{g(j)}^{-2} \E \( \frac{\beta_{jd}^2}{\sigma^{2}_d}\) \gamma_d^{-2} + (k - 1) \sum_{d=1}^D \log \gamma_d^{-2} - \theta \sum_{d=1}^D \gamma_d^{-2} \\
-%		& = \sum_{d=1}^D \left\{ \( \frac{p}{2} + k - 1\) \log \gamma_d^{-2} - \[ \frac{\E(\sigma_d^{-2})}{2} \sum_{j=1}^p \frac{\E(\beta_{jd}^2)}{\phi_{g(j)}^{2}} + \theta \] \gamma_d^{-2} \right\}.
-%		\end{align*}
-
-%	\begin{align*}
-%	q_{\bgamma^{-2}}(\bgamma^{-2}) & \overset{D}{=} \prod_{d=1}^D \Gamma \(\frac{p}{2} + k, a_{d} \), \\
-%	& \text{ with } a_{d} =\frac{\E(\sigma_d^{-2})}{2} \sum_{j=1}^p \frac{\E(\beta_{jd})^2 + \V(\beta_{jd})}{\phi_{g(j)}^2} + \theta.
-%	\end{align*}
-
 	\begin{align*}
@@
 
-%		\begin{align*}
-%		\log q_{\bsigma^{-2}}(\bsigma^{-2}) & \propto \E_{\B,\mathbf{U}} [\log \mathcal{L}(\Y | \B, \U, \bsigma^2 )] + \E_{\B, \bgamma^2} [\log \pi (\B | \bgamma^2, \bsigma^2)] \\
-%		& \,\,\,\,\,\,\,\,\,\, + \E_{\mathbf{U}, \bm{\Xi}_1, \dots, \bm{\Xi}_D} [\log \pi (\mathbf{U} | \bm{\Xi}_1, \dots, \bm{\Xi}_D, \bsigma^2)] + \log \pi (\bsigma^{-2}) \\
-%		& \propto \frac{n}{2} \sum_{d=1}^D \log \sigma_u^{-2} - \frac{1}{2} \sum_{d=1}^D \sum_{i=1}^n \E [(y_{id} - \x_i \tr \bbeta_d - \mathbf{z}_i \tr \mathbf{u}_d)^2] \sigma_d^{-2} + \frac{p}{2} \sum_{d=1}^D \log \sigma_d^{-2} \\
-%		& \,\,\,\,\,\,\,\,\,\, - \frac{1}{2} \sum_{d=1}^D \sum_{j=1}^p \phi_{g(j)}^{-2} \E \( \frac{\beta_{jd}^2}{\gamma_d^{2}} \) \sigma_d^{-2} + \frac{T}{2} \sum_{d=1}^D \log \sigma_d^{-2} - \frac{1}{2} \sum_{d=1}^D \E(\mathbf{u}_d \tr \bm{\Xi}_d^{-1} \mathbf{u}) \sigma_d^{-2} \\
-%		& \,\,\,\,\,\,\,\,\,\, - 0.999 \sum_{d=1}^D \log \sigma_d^{-2} - 0.001 \sum_{d=1}^D \sigma_d^{-2} \\
-%		& = \sum_{d=1}^{D} \(\frac{n + p + T}{2} - 0.999 \) \log \sigma_d^{-2} - \sum_{d=1}^{D} \Bigg\{ \frac{\sum_{i=1}^n \E [(y_{id} - \x_i \tr \bbeta_d - \mathbf{z}_i \tr \mathbf{u}_d)^2]}{2} \\
-%		& \,\,\,\,\,\,\,\,\,\, + \frac{\E(\gamma_d^{-2})}{2} \sum_{j=1}^p \frac{\E(\beta_{jd}^2)}{\phi_{g(j)}^2} + \frac{\E(\mathbf{u}_{d} \tr \bm{\Xi}_d^{-1} \mathbf{u}_d)}{2}  + 0.001 \Bigg\} \sigma_d^{-2}.
-%		\end{align*}
-
-%	\begin{align*}
-%	q_{\bsigma^{-2}}(\bsigma^{-2}) & \overset{D}{=} \prod_{d=1}^D \Gamma \(\frac{n + p + T}{2} + 0.001, b_{d} \), \\
-%	& \text{ with } b_{d} = \frac{\y_d \tr \y_d}{2} - \y_d \tr [\X \E(\bbeta_d) + \Z \E(\mathbf{u}_d)] + \E(\bbeta_d \tr) \X \tr \Z \E (\mathbf{u}_d) \\
-%	& \,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\, + \frac{\trace [\X \tr \X \V(\bbeta_d)]}{2} + \frac{\E(\bbeta_d \tr) \X \tr \X \E(\bbeta_d)}{2} + \frac{\trace [\Z \tr \Z \V(\mathbf{u}_d)]}{2} \\
-%	& \,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\, + \frac{\E(\mathbf{u}_d \tr) \Z \tr \Z \E(\mathbf{u}_d)}{2} + \frac{\E(\gamma_d^{-2})}{2} \sum_{j=1}^p \frac{\E(\beta_{jd})^2 + \V(\beta_{jd})}{\phi_{g(j)}^2} \\
-%	& \,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,+ \frac{\E(\mathbf{u}_d \tr) \E(\bm{\Xi}_d^{-1}) \E(\mathbf{u}_d)}{2} + \frac{\trace [\E(\bm{\Xi}_d^{-1}) \V(\mathbf{u}_d)]}{2} + 0.001.
-%	\end{align*}
-
 	\begin{align*}
@@
 
-%		\section{Convexity}\label{app:convexity}
-%		For (\ref{eq:optproblem2}) to be a convex problem, we need the Hessian $\mathbf{H}$ of the objective function to be positive definite, i.e., $\forall \mathbf{z} \in \mathbb{R}^2: \mathbf{z} \tr \mathbf{H} \mathbf{z} > 0$.
-We have
-%		\begin{align*}
-%		\mathbf{z} \tr \mathbf{H} \mathbf{z} = \frac{D}{\theta^2 \psi^{(1)}(k)} \( z_1 \theta \psi^{(1)}(k) - z_2 \)^2 + z_2^2 \( k \psi^{(1)}(k)  - 1 \),
-%		\end{align*}
-%		with $\psi^{(1)}(k)$ the trigamma function. Since the trigamma function is positive for all $k > 0$, and we have $\theta, k > 0$, we need:
-%		$$
-%		k \psi^{(1)}(k)> 1.
-%		$$
-%		Using Lemma 1 in \cite{guo_refinements_2013} we can show that for $k > 0$:
-%		$$
-%		k \psi^{(1)}(k) > 1 + \frac{1}{2k},
-%		$$
-%		such that the objective function is convex.
+  \section{Model without tissue effects}
+  \subsection{Introduction}
+  In order to investigate the inverse Gaussian prior model and its estimation, we consider a simpler model. The simpler model ignores tissue effects and fixes $\forall g: \phi_g^2=1$.
 
-		\section{Prior tissue covariance constraints}\label{app:priortissuecovariance}
-%		The empirical Bayes objective function of the third sub-optimisation problem in \ref{sec:empiricalbayes} is:
-%		\begin{align*}
-%		\E_{Q^{(l)}}[\log \pi (\mathbf{\Xi}_1, \dots, \mathbf{\Xi}_D)] & \propto \frac{D}{2} \nu \log |\bm{\Omega}| - \frac{D T \log 2}{2} \nu - D \log \Gamma_T \(\frac{\nu}{2}\) + \frac{\nu}{2}\sum_{d=1}^D \E_{Q^{(l)}} (\log | \mathbf{\Xi}_d^{-1}|) \\
-%		& \,\,\,\,\,\,\,\,\,\, - \frac{1}{2} \sum_{d=1}^D \trace [ \bm{\Omega} \cdot \E_{Q^{(l)}}(\mathbf{\Xi}_d^{-1})] \\
-%		& = \frac{D}{2} \nu \log |\bm{\Omega}| - D \log \Gamma_T\(\frac{\nu}{2}\) - \frac{\nu^{(l)} + 1}{2} \sum_{d=1}^D \trace [ \bm{\Omega} \cdot ( \mathbf{\Psi}_d^{(l)})^{-1}] \\
-%		& \,\,\,\,\,\,\,\,\,\, + \frac{\nu}{2} \sum_{d=1}^D \[ \psi_T\(\frac{\nu^{(l)} + 1}{2}\) - \log |\mathbf{\Psi}_d^{(l)}| \],
-%		\end{align*}
-%		which is maximised under constraints $\nu > T + 1$ and $\bm{\Omega}$ positive definite (PD), to ensure a proper inverse Wishart distribution.
-
-		We parametrise the prior covariance as a matrix with $\tau$ on the diagonal and $\rho$ on its off-diagonals. To that end we set $\nu = \tau + T + 1$ and the $T \times T$-dimensional matrix
-		$$
-		\bm{\Omega} = \tau \begin{bmatrix}
-		\tau &  & \rho \\
-		& \ddots &  \\
-		\rho & & \tau
-		\end{bmatrix},
-		$$
-		such that $\E_{\pi(\mathbf{\Xi}_d)} (\mathbf{\Xi}_d) = (\nu - T - 1)^{-1} \bm{\Omega}$ is of the desired form.
-To ensure a non-degenerate prior distribution we require $\tau > 0$ and $\bm{\Omega}$ PD.
-We use $\tau > 0$ to note that a PD $\bm{\Omega}$ implies positive roots $\lambda$ of the characteristic polynomial:
-		\begin{align*}
-		\left| \begin{bmatrix}
-		\tau &  & \rho \\
-		& \ddots &  \\
-		\rho & & \tau
-		\end{bmatrix} -\lambda \I_T \right| = (\tau - \lambda - \rho)^{T-1}[\tau - \lambda - (1-T)\rho] = 0.
-		\end{align*}
-		For $T > 1$, the solutions are:
-		$$
-		(\lambda = \tau - \rho) \lor (\lambda = T \rho - \rho + \tau),
-		$$
-		which are positive if the following condition holds: $\tau > \max(\rho,\rho - T \rho)$.
-We combine this with $\tau > 0$, to arrive at the full proper prior constraint: $\tau > \max(0,\rho,\rho - T \rho)$.
-
-%		The resulting empirical Bayes step is now:
-%		\begin{align*}
-%		\tau^{(l + 1)}, \rho^{(l+1)} & = \underset{\tau, \rho}{\argmax} \frac{D (T^2 + T)}{2} \log \tau + \frac{DT}{2} \tau \log \tau + \frac{D(T+1)}{2} \log \[ \tau - (T - 1) \rho\] \\
-%		& \,\,\,\,\,\,\,\,\,\, + \frac{D}{2} \tau \log \[ \tau - (T - 1) \rho\] + \frac{D(T^2 - 1)}{2} \log(\tau - \rho) +\frac{D(T - 1)}{2} \tau  \log(\tau - \rho) \\
-%		& \,\,\,\,\,\,\,\,\,\, - D \sum_{t=1}^T \log \Gamma \( \frac{\tau + t + 1}{2} \) - w_1^{(l)} \tau^2 + w_2^{(l)} \tau \rho + w_3^{(l)} \tau, \\
-%		& \text{subject to } \tau > \max\[\rho - T \rho,\rho, 0\],
-%		\end{align*}
-%		with
-%		\begin{align*}
-%		w_1^{(l)} & =  \frac{\tau^{(l)} + T + 2}{2} \sum_{d=1}^D \trace \[ (\bm{\Psi}_d^{(l)})^{-1}\] > 0,\\
-%		w_2^{(l)} & = \frac{\tau^{(l)} + T + 2}{2} \sum_{d=1}^D \left\{\trace \[ (\bm{\Psi}_d^{(l)})^{-1}\] - \trace \[\mathbf{1}_{T \times T} \cdot (\bm{\Psi}_d^{(l)})^{-1}\] \right\} < w_1^{(l)},\\
-%		w_3^{(l)} & = \frac{1}{2} \[D \sum_{t=1}^T \psi\(\frac{\tau^{(l)} + t + 2}{2}\) - \sum_{d=1}^D \log |\bm{\Psi}_d^{(l)}| \].
-%		\end{align*}
-
-		\section{Model without tissue effects}
-		\subsection{Model}
-		In order to investigate the inverse Gaussian prior model and its estimation, we consider the simpler model:
+  \subsection{Model and estimation}
+  The model is given by:
   \begin{align*}
@@
   \end{align*}
-		So in other words, we ignore the tissue effect and fix all $\phi_g^2$ to one. 
-		\subsection{Variational Bayes}
   The corresponding variational Bayes posterior is given by:
@@
   \end{align*}
-		We update the parameters until convergence:
+  We update the variational parameters until convergence:
   \begin{align*}
@@
   \end{align*}
-		with
-		$$
-		a_d^{(h)} = \sqrt{\frac{\lambda}{\theta_d^2 \delta_d^{(h)}}} \frac{K_{\frac{p - 1}{2}}\(\sqrt{\lambda \delta_d^{(h)}} /\theta_d \)}{K_{\frac{p + 1}{2}}\(\sqrt{\lambda \delta_d^{(h)}} /\theta_d \)} + \frac{p+1}{\delta_d^{(h)}}.
-		$$
-
-		\subsection{Empirical Bayes}
-		Again we model $\theta_d = (\mathbf{c}_d \tr \bm{\alpha})^{-1}$ to arrive at the following empirical Bayes updates:
-		\begin{align*}
-		\bm{\alpha}^{(l+1)} & = \left\{ \mathbf{C} \tr \Vm^{(l)} \mathbf{C} \right\}^{-1} \mathbf{C} \tr \mathbf{1}, \\
-		\lambda^{(l+1)} & = \[ \sum_{d=1}^D a_d^{(l)} - (\bm{\alpha}^{(l+1)}) \tr \mathbf{C} \tr \mathbf{1} \]^{-1} D,
-		\end{align*}
-		where 
-		$$
-		\Vm^{(l)} = \diag \[ \sqrt{\frac{\delta_d^{(l)}(\theta_d^{(l)})^2}{\lambda^{(l)}}} \frac{K_{\frac{p - 1}{2}} \( \sqrt{\lambda^{(l)} \delta_d^{(l)}}/\theta^{(l)}_d \)}{K_{\frac{p + 1}{2}} \( \sqrt{\lambda^{(l)} \delta_d^{(l)}}/\theta^{(l)}_d \)} \].
-		$$
+  with $a_d^{(h)}$ as in the full model in the MD. Equivalently, the empirical Bayes updates for $\bm{\alpha}$ and $\lambda$ are the same as in the full model in the MD.
 
@@
 
-%		\begin{align*}
-%		y_{id} | \bm{\beta}_d & \sim \mathcal{N} (\x_i \tr \bm{\beta}_d, \sigma_d^2), \\
-%		\beta_{jd} | \gamma_d^2 & \sim \mathcal{N} (0, \sigma_d^2 \gamma_d^2), \\
-%		\gamma_d^2 & \sim \mathcal{IG}(\theta_d, \lambda).
-%		\end{align*}
-%		So in other words, we ignore the tissue effect, fix the $\sigma_d^2$, and set all $\phi_g^2$ to one. The corresponding variational Bayes posterior is given by:
-%		\begin{align*}
-%		q(\bm{\beta}) & \overset{D}{=} \prod_{d=1}^D \mathcal{N}_p (\bm{\mu}_d, \bm{\Sigma}_d), \\
-%		q(\bm{\gamma}^2) & \overset{D}{=} \prod_{d=1}^D \mathcal{GIG}\(-\frac{p+1}{2}, \frac{\lambda}{\theta_d^2}, \delta_d\).
-%		\end{align*}
-%		We update the parameters until convergence:
-%		\begin{align*}
-%		\bm{\Sigma}_d^{(h+1)} & = \sigma_d^2 (\X \tr \X + a_d^{(h)} \I)^{-1}, \\
-%		& \text{with } a_d^{(h)} = \sqrt{\frac{\lambda}{\theta_d^2 \delta_d^{(h)}}} \frac{K_{\frac{p - 1}{2}}\(\sqrt{\lambda \delta_d^{(h)}} /\theta_d \)}{K_{\frac{p + 1}{2}}\(\sqrt{\lambda \delta_d^{(h)}} /\theta_d \)} + \frac{p+1}{\delta_d^{(h)}}, \\
-%		\bm{\mu}_d^{(h+1)} & = (\X \tr \X + a_d^{(h)} \I)^{-1} \X \tr \y_d, \\
-%		\delta_d^{(h+1)} & = \sigma_d^{-2} [\trace(\bm{\Sigma}_d^{(h+1)}) + (\bm{\mu}_d^{(h+1)}) \tr \bm{\mu}_d^{(h+1)}] + \lambda.
-%		\end{align*}
-%		Again we model $\theta_d = (\mathbf{c}_d \tr \bm{\alpha})^{-1}$ to arrive at the following empirical Bayes updates:
-%		\begin{align*}
-%		\bm{\alpha}^{(l+1)} & = \left\{ \mathbf{C} \tr \Vm^{(l)} \mathbf{C} \right\}^{-1} \mathbf{C} \tr \mathbf{1}, \\
-%		\lambda^{(l+1)} & = \[ \sum_{d=1}^D a_d^{(l)} - (\bm{\alpha}^{(l+1)}) \tr \mathbf{C} \tr \mathbf{1} \]^{-1} D,
-%		\end{align*}
-%		where 
-%		$$
-%		\Vm^{(l)} = \diag \[ \sqrt{\frac{\delta_d^{(l)}(\theta_d^{(l)})^2}{\lambda^{(l)}}} \frac{K_{\frac{p - 1}{2}} \( \sqrt{\lambda^{(l)} \delta_d^{(l)}}/\theta^{(l)}_d \)}{K_{\frac{p + 1}{2}} \( \sqrt{\lambda^{(l)} \delta_d^{(l)}}/\theta^{(l)}_d \)} \].
-%		$$
-%		We monitor convergence through the evidence lower bound (ELBO):
-%		\begin{align*}
-%		\text{ELBO} & (Q^{(l)}) \propto \sum_{d=1}^D \sigma_d^{-2} \y_d \tr \X \bm{\mu}_d^{(l)} - \frac{1}{2} \sum_{d=1}^D \sigma_d^{-2} \trace ( \X \tr \X \bm{\Sigma}_d^{(l)}) - \frac{1}{2} \sum_{d=1}^D \sigma_d^{-2} (\bm{\mu}_d^{(l)}) \tr \X \tr \X \bm{\mu}_d^{(l)} \\
-%		& + \frac{1}{2} \sum_{d=1} \log |\bm{\Sigma}_d^{(l)}| + \frac{D(p+3)}{4} \log \lambda^{(l)} + \lambda^{(l)} \sum_{d=1}^D (\theta_d^{(l)})^{-1} - \frac{p+1}{2} \sum_{d=1}^D \log \theta_d^{(l)} \\
-%		& - \frac{p+1}{4} \sum_{d=1}^D \log \delta_d^{(l)} + \sum_{d=1}^D \log K_{\frac{p+1}{2}}\(\sqrt{\lambda^{(l)} \delta_d^{(l)}}/\theta_d^{(l)}\)\\
-%		& + \frac{1}{2} \sum_{d=1}^D \left(\rule{0cm}{1.2cm}\right.
-\[ \delta_d^{(l)} - \lambda^{(l)} - \sigma_d^{-2} (\bm{\mu}_d^{(l)}) \tr \bm{\mu}_d^{(l)} - \sigma_d^{-2} \trace (\bm{\Sigma}_d^{(l)}) \]  \\
-%		& \,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\, \cdot \[\sqrt{\frac{\lambda^{(l)}}{\delta_d^{(l)} (\theta^{(l)}_d)^2}} \frac{K_{\frac{p - 1}{2}}\(\frac{\sqrt{\lambda^{(l)} \delta_d^{(l)}}}{\theta^{(l)}_d} \)}{K_{\frac{p + 1}{2}}\(\frac{\sqrt{\lambda^{(l)} \delta_d^{(l)}}}{\theta^{(l)}_d} \)} + \frac{p+1}{\delta_d^{(l)}}\] \left.
-\rule{0cm}{1.2cm}\right).
-%		\end{align*}
-
-		- Different options for convergence. We may monitor parameters, ELBO, or a combination. We may also iterate a fixed number of times, especially for the VB.
+  - Different options for convergence.
+We may monitor parameters, ELBO, or a combination.
+We may also iterate a fixed number of times, especially for the VB (currently doing this, using 2 VB iterations and 20 EB iterations).
 
   \subsection{Efficient computation}
-		We need the following quantities during the EB iterations: $\zeta_d$, $a_d$, $\delta_d$, $\trace (\bm{\Sigma}_d)$, $\trace (\X \tr \X \bm{\Sigma}_d)$, $\bm{\mu}_d \tr \bm{\mu}_d$, and $\bm{\mu}_d \tr \X \tr \X \bm{\mu}_d$.
+  The empirical Bayes updates require the following quantities: $\zeta_d$, $a_d$, $\delta_d$, $\trace (\bm{\Sigma}_d)$, $\trace (\X \tr \X \bm{\Sigma}_d)$, $\bm{\mu}_d \tr \bm{\mu}_d$, and $\bm{\mu}_d \tr \X \tr \X \bm{\mu}_d$.
 In addition, we need $\log|\bm{\Sigma}|$ and $\mathbf{y}_d \tr \X \bm{\mu}_d$ to monitor the ELBO.
-The first three are just scalar operations.
+The first three quantities are obtained by scalar operations of $\mathcal{O}(n)$.
 The rest is easily calculated using the SVD $\X = \mathbf{U} \mathbf{D} \mathbf{V} \tr$:
   \begin{align*}
-		\trace (\bm{\Sigma}_d) & = \frac{2\zeta_d}{n + p + 1} \[\sum_{i=1}^n (v_i^2 + a_d)^{-1} + \max (p - n, 0)\cdot a_d^{-1}\], \\
-		\trace (\X \tr \X \bm{\Sigma}_d) & = \frac{2\zeta_d}{n + p + 1} \sum_{i=1}^n v_i^2(v_i^2 + a_d)^{-1}, \\
-		\bm{\mu}_d \tr \bm{\mu}_d & = \sum_{i=1}^n v_i^2 [(\mathbf{U} \tr \mathbf{y}_d)_i]^2/(v_i^2 + a_d)^2, \\
-		\bm{\mu}_d \tr \X \tr \X \bm{\mu}_d & = \sum_{i=1}^n v_i^4 [(\mathbf{U} \tr \mathbf{y}_d)_i]^2/(v_i^2 + a_d)^2, \\
-		\log|\bm{\Sigma}| & \propto p \log \zeta_d - \sum_{i=1} \log (v_i^2 + a_d) - \max(p - n, 0) \cdot \log a_d, \\
-		\mathbf{y}_d \tr \X \bm{\mu}_d & = \sum_{i=1}^n v_i^2 [(\mathbf{U} \tr \mathbf{y}_d)_i]^2/(v_i^2 + a_d).
+  \trace (\bm{\Sigma}_d^{(h+1)}) & = \frac{2\zeta_d^{(h)}}{n + p + 1} \[\sum_{i=1}^n (v_i^2 + a_d^{(h)})^{-1} + \max (p - n, 0)\cdot (a_d^{(h)})^{-1}\], \\
+  \trace (\X \tr \X \bm{\Sigma}_d^{(h+1)}) & = \frac{2\zeta_d^{(h)}}{n + p + 1} \sum_{i=1}^n v_i^2(v_i^2 + a_d^{(h)})^{-1}, \\
+  (\bm{\mu}_d^{(h+1)}) \tr \bm{\mu}_d^{(h+1)} & = \sum_{i=1}^n v_i^2 [(\mathbf{U} \tr \mathbf{y}_d)_i]^2/(v_i^2 + a_d^{(h)})^2, \\
+  (\bm{\mu}_d^{(h+1)}) \tr \X \tr \X \bm{\mu}_d^{(h+1)} & = \sum_{i=1}^n v_i^4 [(\mathbf{U} \tr \mathbf{y}_d)_i]^2/(v_i^2 + a_d^{(h)})^2, \\
+  \log|\bm{\Sigma}^{(h+1)}| & \propto p \log \zeta_d^{(h)} - \sum_{i=1} \log (v_i^2 + a_d^{(h)}) - \max(p - n, 0) \cdot \log a_d^{(h)}, \\
+  \mathbf{y}_d \tr \X \bm{\mu}_d^{(h+1)} & = \sum_{i=1}^n v_i^2 [(\mathbf{U} \tr \mathbf{y}_d)_i]^2/(v_i^2 + a_d^{(h)}).
   \end{align*}
   where $v_i$ are the singular values of $\X$.
-The SVD and $\mathbf{U} \tr \mathbf{y}_d$ are $\mathcal{O}(pn^2)$ and $\mathcal{O}(n^2)$ once at the start of the algorithm.
+The SVD and $\mathbf{U} \tr \mathbf{y}_d$ are $\mathcal{O}(pn^2)$ and $\mathcal{O}(n^2)$ operations, respectively, but have to be calculated only once at the start of the algorithm.
 The rest of the calculations involve just $n$ scalar multiplications and/or additions.
+
+  \subsection{Independent error variance model parameters}
+  \cite{moran_variance_2018} argue that in regression models, $\bm{\beta}_d$ and $\sigma_d^2$ should be independent \textit{a priori}.
+However, in our case that is debatable.
+The $D$ regressions need not be on the same scale, so some form of calibration of the $\bm{\beta}_d$ posteriors to this scale is required.
+A simple way of doing this is to include the scales $\sigma_d^2$ in the priors for the $\bm{\beta}_d$.
+For a sense of completeness we still compare to the independent prior model in our simulations.
+The independent prior model is given by:
+  \begin{align*}
+  y_{id} | \bm{\beta}_d, \sigma_d^2 & \sim \mathcal{N} (\x_i \tr \bm{\beta}_d, \sigma_d^2), \\
+  \beta_{jd} | \gamma_d^2 & \sim \mathcal{N} (0, \gamma_d^2), \\
+  \gamma_d^2 & \sim \mathcal{IG}((\mathbf{c}_d \tr \bm{\alpha})^{-1}, \lambda), \\
+  \sigma_d^2 & \sim 1/\sigma_d^3.
+  \end{align*}
+  Its variational posterior is given by:
+  \begin{align*}
+  q(\bm{\beta}) & \overset{D}{=} \prod_{d=1}^D \mathcal{N}_p (\bm{\mu}_d, \bm{\Sigma}_d), \\
+  q(\bm{\gamma}^2) & \overset{D}{=} \prod_{d=1}^D \mathcal{GIG}\(-\frac{p+1}{2}, \lambda (\mathbf{c}_d \tr \bm{\alpha})^2, \delta_d\), \\
+  q(\bm{\sigma}^2) & \overset{D}{=} \prod_{d=1}^D \Gamma^{-1} \(\frac{n + 1}{2}, \zeta_d\),
+  \end{align*}
+  with variational parameter updates:
+  \begin{align*}
+  \bm{\Sigma}_d^{(h+1)} & = \(\frac{n+1}{2 \zeta_d^{(h)}}\X \tr \X + a_d^{(h)} \I_p\)^{-1}, \\
+  \bm{\mu}_d^{(h+1)} & = \frac{n+1}{2 \zeta_d^{(h)}} \bm{\Sigma}_d^{(h+1)} \X \tr \y_d, \\
+  \delta_d^{(h+1)} & = \trace(\bm{\Sigma}_d^{(h+1)}) + (\bm{\mu}_d^{(h+1)}) \tr \bm{\mu}_d^{(h+1)} + \lambda, \\ \zeta_d^{(h+1)} & = \frac{1}{2} \[\mathbf{y}_d \tr \mathbf{y}_d -2 \mathbf{y}_d \tr \X \bm{\mu}_d^{(h+1)} + \trace (\X \tr \X \bm{\Sigma}_d^{(h+1)}) + (\bm{\mu}_d^{(h+1)}) \tr \X \tr \X \bm{\mu}_d^{(h+1)}\],
+  \end{align*}
+  with $a_d^{(h)}$ and the empirical Bayes updates $\bm{\alpha}^{(l+1)}$ and $\lambda^{(l+1)}$ as in the full model in the MD.
+A singular value decomposition $\X = \mathbf{U} \mathbf{D} \mathbf{V} \tr$ leads to the following efficient computations:
+  \begin{align*}
+  \trace (\bm{\Sigma}^{(h+1)}_d) & = \frac{2\zeta^{(h)}_d}{n + 1} \sum_{i=1}^n \(v_i^2 + \frac{2\zeta^{(h)}_d a^{(h)}_d}{n+1}\)^{-1} + \max (p - n, 0)\cdot (a^{(h)}_d)^{-1}, \\
+  \trace (\X \tr \X \bm{\Sigma}^{(h + 1)}_d) & = \frac{2\zeta^{(h)}_d}{n + 1} \sum_{i=1}^n v_i^2 \(v_i^2 + \frac{2\zeta^{(h)}_d a^{(h)}_d}{n+1}\)^{-1}, \\
+  (\bm{\mu}_d^{(h + 1)}) \tr \bm{\mu}_d^{(h + 1)} & = \sum_{i=1}^n [(\mathbf{U} \tr \mathbf{y}_d)_i]^2 v_i^2 \(v_i^2 + \frac{2 \zeta_d^{(h)} a_d^{(h)}}{n + 1}\)^{-2}, \\
+  (\bm{\mu}_d^{(h + 1)}) \tr \X \tr \X \bm{\mu}_d^{(h + 1)} & = \sum_{i=1}^n [(\mathbf{U} \tr \mathbf{y}_d)_i]^2v_i^4 \(v_i^2 + \frac{2 \zeta_d^{(h)} a_d^{(h)}}{n + 1}\)^{-2}, \\
+  \log|\bm{\Sigma}^{(h + 1)}| & \propto p \log \zeta_d^{(h)} - \sum_{i=1} \log \(v_i^2 + \frac{2 \zeta_d^{(h)} a_d^{(h)}}{n + 1}\) \\
+  & \,\,\,\,\,\,\,\,\,\,\, - \max(p - n, 0) \cdot (\log a_d^{(h)} + \log \zeta_d^{(h)}), \\
+  \mathbf{y}_d \tr \X \bm{\mu}_d^{(h + 1)} & = \sum_{i=1}^n [(\mathbf{U} \tr \mathbf{y}_d)_i]^2 v_i^2 \(v_i^2 + \frac{2 \zeta_d^{(h)} a_d^{(h)}}{n + 1}\),
+  \end{align*}
+  where $v_i$ are the singular values of $\X$.
+As before, this renders $\mathcal{O}(pn^2)$ and $\mathcal{O}(n^2)$ calculations once at the start of the algorithm and iterative updates of $\mathcal{O}(n)$.
 
@@
 Next, we simulate $\bm{\beta}_{jd} \sim \mathcal{N} (0, \sigma^2_d \gamma_d^2)$.
-We fix the mean signal-to noise ratio $\overline{\text{SNR}}_d$, such that the predictor data variance is $s^2 = \V(\mathbf{x}_i) = \overline{\text{SNR}}_d / (\overline{\gamma^2_d} p)$.
+We fix the mean signal-to noise ratio $\overline{\text{SNR}}_d$, such that the predictor data variance is $s^2 = \V(\mathbf{x}_i) = \overline{\text{SNR}}_d / (\overline{\gamma^2_d} p)$ (NOT CORRECT).
 The last step consists of simulating the data by $\mathbf{x}_{ij} \sim \mathcal{N}(0,s^2)$ and $y_{id} = \mathcal{N} (\mathbf{x}_i \tr \bm{\beta}_d, \sigma_d^2)$.
-The remaining parameters are set as follows: $\bm{\alpha} = \begin{bmatrix} 1 & \dots & 5 \end{bmatrix} \tr$, $\sigma_d^2 = 1$ for all $d$, and $\overline{\text{SNR}}_d = 5$.
+The remaining parameters are set as follows: $\bm{\alpha} = \begin{bmatrix} 1 & \dots & 5 \end{bmatrix} \tr$, $\sigma_d^2 = 1$ for all $d$, and $\overline{\text{SNR}}_d = 10$.
+To account for the random data and parameter generation, we repeat the simulation 100 times.
 
@@
   where $\text{class}(c)$ denotes the class of drug $d$.
-This allows for more flexible empirical Bayes estimation of the $a_d$ and $b_d$, but, as a consequence, increases the risk of overfitting.
+This allows for more flexible empirical Bayes estimation of the $a_d$ and $b_d$, and thus prior means of the different drug classes.
+As a consequence however, the risk of overfitting is increased.
 Additionally, it does not allow to include continuous covariates on the drugs.
+We present the results in Figures \ref{fig:boxplot_igaussian_res1_sigma2}-\ref{fig:boxplot_igaussian_res1_alpha}.
 
-		To account for the random data and parameter generation, we repeat the simulation 100 times and present the results in Table...
+<<"boxplot_igaussian_res1_sigma2", fig.cap="MSEs for posterior means of $\\sigma_d^2$", out.width="50%", fig.align="center", echo=FALSE>>=
+load("../results/simulations_igaussian_res1.RData")
+temp <- sapply(res1$zeta, function(z) {
+  colMeans((2*t(z)/(set1$n + (z==1)*set1$p - 1) - set1$sigma^2)^2)})
+par(cex=1.3)
+boxplot(temp, outline=FALSE, names=names(res1$mse.mu), ylab="MSE")
+legend("topright", title=expression(bold("median MSE")), bty="n", 
+       paste(names(res1$mse.mu), round(apply(temp, 2, median), 2), sep="="))
+par(cex=1)
+@
 
-		\subsection{Starting values}
-%		Starting value for $\sigma_d^2$ as in \cite{chipman_bart:_2010}.
+<<"boxplot_igaussian_res1_gamma2", fig.cap="a) MSE and b) correlation for posterior means of $\\gamma_d^2$", fig.align="center", echo=FALSE>>=
+load("../results/simulations_igaussian_res1.RData")
+temp1 <- sapply(1:3, function(m) {
+  colMeans((t(sqrt(res1$delta[[m]]*res1$theta[[m]]^2/res1$lambda[[m]])*
+                ratio_besselK_cpp(sqrt(res1$delta[[m]]*res1$lambda[[m]])/
+                                    res1$theta[[m]], set1$p)) - set1$gamma^2)^2)})
+temp2 <- sapply(1:3, function(m) {
+  apply(sqrt(res1$delta[[m]]*res1$theta[[m]]^2/res1$lambda[[m]])*
+          ratio_besselK_cpp(sqrt(res1$delta[[m]]*res1$lambda[[m]])/
+                              res1$theta[[m]], set1$p), 1, function(r) {
+                                cor(r, set1$gamma^2)})})
+par(cex=1.3)
+layout(matrix(c(rep(1, 4), rep(2, 4)), 2, 4), widths=1, heights=1, respect=TRUE)
+boxplot(temp1, names=names(res1$mse.mu), ylab="MSE", main="a)")
+legend("topright", title=expression(bold("median MSE")), bty="n", 
+       paste(names(res1$mse.mu), round(apply(temp1, 2, median), 2), sep="="))
+boxplot(temp2, names=names(res1$mse.mu), ylab="Correlation", main="b)")
+legend("topright", title=expression(bold("median correlation")), bty="n", 
+       paste(names(res1$mse.mu), round(apply(temp2, 2, median), 2), sep="="))
+par(cex=1)
+@
+
+<<"boxplot_igaussian_res1_mu", fig.cap="a) MSE and b) correlation for posterior means of $\\beta_d$", fig.align="center", echo=FALSE>>=
+load("../results/simulations_igaussian_res1.RData")
+par(cex=1.3)
+layout(matrix(c(rep(1, 4), rep(2, 4)), 2, 4), widths=1, heights=1, respect=TRUE)
+boxplot(res1$mse.mu, names=names(res1$mse.mu), ylab="MSE", main="a)")
+legend("topright", title=expression(bold("median MSE")), bty="n", 
+       paste(names(res1$mse.mu), round(sapply(res1$mse.mu, median), 2), 
+             sep="="))
+boxplot(res1$cor.mu, names=names(res1$mse.mu), ylab="Correlation", main="b)")
+legend("topright", title=expression(bold("median correlation")), bty="n", 
+       paste(names(res1$mse.mu), round(sapply(res1$cor.mu, median), 2), 
+             sep="="))
+par(cex=1)
+@
+
+<<"boxplot_igaussian_res1_theta", fig.cap="a) MSE and b) correlation for prior means of $\\gamma_d^2$ ($\\theta_d$)", fig.align="center", echo=FALSE>>=
+load("../results/simulations_igaussian_res1.RData")
+temp1 <- sapply(res1$theta, function(th) {colMeans((t(th) - set1$theta)^2)})
+temp2 <- sapply(res1$theta, function(th) {apply(th, 1, function(r) {
+  cor(r, set1$theta)})})
+par(cex=1.3)
+layout(matrix(c(rep(1, 4), rep(2, 4)), 2, 4), widths=1, heights=1, respect=TRUE)
+boxplot(temp1, names=names(res1$mse.mu), ylab="MSE", main="a)")
+legend("topright", title=expression(bold("median MSE")), bty="n", 
+       paste(names(res1$mse.mu), round(apply(temp1, 2, median), 2), sep="="))
+boxplot(temp2, names=names(res1$mse.mu), ylab="Correlation", main="b)")
+legend("topright", title=expression(bold("median correlation")), bty="n", 
+       paste(names(res1$mse.mu), round(apply(temp2, 2, median), 2), sep="="))
+par(cex=1)
+@
+
+<<"boxplot_igaussian_res1_alpha", fig.cap="a) MSE and b) correlation for $\\alpha$", fig.align="center", echo=FALSE>>=
+load("../results/simulations_igaussian_res1.RData")
+temp1 <- sapply(res1$alpha, function(a) {colMeans((t(a) - set1$alpha)^2)})
+temp2 <- sapply(res1$alpha, function(a) {apply(a, 1, function(r) {
+  cor(r, set1$alpha)})})
+par(cex=1.3)
+layout(matrix(c(rep(1, 4), rep(2, 4)), 2, 4), widths=1, heights=1, respect=TRUE)
+boxplot(temp1, names=names(res1$mse.mu), ylab="MSE", main="a)")
+legend("topright", title=expression(bold("median MSE")), bty="n", 
+       paste(names(res1$mse.mu), round(apply(temp1, 2, median), 2), sep="="))
+boxplot(temp2, names=names(res1$mse.mu), ylab="Correlation", main="b)")
+legend("topright", title=expression(bold("median correlation")), bty="n", 
+       paste(names(res1$mse.mu), round(apply(temp2, 2, median), 2), sep="="))
+par(cex=1)
+@
+
+  \section{Starting values}
+  \subsection{Method 1}
+  Starting value for $\sigma_d^2$ as in \cite{chipman_bart:_2010}.
 They consider a scaled-Chi squared distribution for the error variance with degrees of freedom 3 and scale chosen such that the $q*100$th percentile matches the error variance in the data $s^2(\mathbf{y}_d)$.
 As starting value for $\sigma_d^2$ we take the mode of this distribution: $\hat{\sigma}_d^2 = 3a/5$, with $a$ the solution to:
-%		$$
-%		\Gamma\(\frac{3}{3}\) \[1 - q - F(3a/(2s^2(\mathbf{y}_d)); 3/2, 1) \] = 0,
-%		$$
-%		where $F(x; 3/2, 1)$ is the CDF of a gamma function with shape $3/2$ and scale $1$.
+  $$
+  \Gamma\(\frac{3}{3}\) \[1 - q - F(3a/(2s^2(\mathbf{y}_d)); 3/2, 1) \] = 0,
+  $$
+  where $F(x; 3/2, 1)$ is the CDF of a gamma function with shape $3/2$ and scale $1$.
 In \cite{moran_variance_2018}, REFERENCE to Ročková and George (2018), and \cite{chipman_bart:_2010}, $q=0.9$ is suggested.
@@
 
+  \subsection{Method 2}
   We generate starting values as follows. Consider the $\sigma_d^2$ and $\gamma_d^2$ as fixed and estimate them by MML of the regular ridge model:
@@
 
-		\subsection{Independent error variance model parameters}
-		Model:
-		\begin{align*}
-		y_{id} | \bm{\beta}_d, \sigma_d^2 & \sim \mathcal{N} (\x_i \tr \bm{\beta}_d, \sigma_d^2), \\
-		\beta_{jd} | \gamma_d^2 & \sim \mathcal{N} (0, \gamma_d^2), \\
-		\gamma_d^2 & \sim \mathcal{IG}((\mathbf{c}_d \tr \bm{\alpha})^{-1}, \lambda), \\
-		\sigma_d^2 & \sim 1/\sigma_d^3.
-		\end{align*}
-		Variational posterior:
-		\begin{align*}
-		q(\bm{\beta}) & \overset{D}{=} \prod_{d=1}^D \mathcal{N}_p (\bm{\mu}_d, \bm{\Sigma}_d), \\
-		q(\bm{\gamma}^2) & \overset{D}{=} \prod_{d=1}^D \mathcal{GIG}\(-\frac{p+1}{2}, \lambda (\mathbf{c}_d \tr \bm{\alpha})^2, \delta_d\), \\
-		q(\bm{\sigma}^2) & \overset{D}{=} \prod_{d=1}^D \Gamma^{-1} \(\frac{n + 1}{2}, \zeta_d\).
-		\end{align*}
-		Parameter updates:
-		\begin{align*}
-		\bm{\Sigma}_d^{(h+1)} & = \(\frac{n+1}{2 \zeta_d^{(h)}}\X \tr \X + a_d^{(h)} \I_p\)^{-1}, \\
-		\bm{\mu}_d^{(h+1)} & = \frac{n+1}{2 \zeta_d^{(h)}} \bm{\Sigma}_d^{(h+1)} \X \tr \y_d, \\
-		\delta_d^{(h+1)} & = \trace(\bm{\Sigma}_d^{(h+1)}) + (\bm{\mu}_d^{(h+1)}) \tr \bm{\mu}_d^{(h+1)} + \lambda, \\ \zeta_d^{(h+1)} & = \frac{1}{2} \[\mathbf{y}_d \tr \mathbf{y}_d -2 \mathbf{y}_d \tr \X \bm{\mu}_d^{(h+1)} + \trace (\X \tr \X \bm{\Sigma}_d^{(h+1)}) + (\bm{\mu}_d^{(h+1)}) \tr \X \tr \X \bm{\mu}_d^{(h+1)}\],
-		\end{align*}
-		with $a_d^{(h)}$ and the empirical Bayes updates $\bm{\alpha}^{(l+1)}$ and $\lambda^{(l+1)}$ the same as before.
   \section{Ratios of modified Bessel functions}
@@
 
-	\end{appendix}
+  \bibliographystyle{author_short3} 
+	\bibliography{refs}
 
 \end{document}
-
-%		\section{Efficient computation}
-%		Let $\X = \U \D \Vm \tr$ and $\Z=\U^* \D^* (\Vm^*) \tr$ be the singular value decomposition of $\X$ and $\Z$, respectively.
-Filling in all the expectations, we have the following iterative estimating equations:
-%		\begin{align*}
-%		\bSigma_d^{(l+1)} & = a_3 b_{3d}^{(l)} \Vm (\D^2 + a_1 b_{1d}^{(l)} \I_n)^{-1} \Vm \tr, \\
-%		\Sm_d^{(l+1)} & = a_3 b_{3d}^{(l)} \Vm^* [(\D^*)^2 + a_2 b_{2d}^{(l)} \I_T]^{-1} (\Vm^*) \tr, \\
-%		\bmu_d^{(l+1)} & = a_3 b_{3d}^{(l)} \Vm (\D + a_1 b_{1d}^{(l)} \D^{-1})^{-1} \U \tr (\y_d - \Z \mathbf{m}_d^{(l)}), \\
-%		\mathbf{m}_d^{(l+1)} & = a_3 b_{3d}^{(l)} \Vm^* [\D^* + a_2 b_{2d}^{(l)} (\D^*)^{-1}]^{-1} (\U^*) \tr (\y_d - \X \bmu_d^{(l + 1)}), \\
-%		b_{1d}^{(l+1)} & = \Bigg\{\frac{a_3 b_{3d}^{(l)}}{2} \Bigg[ (\y_d - \Z \mathbf{m}_d^{(l)}) \tr \U (\D + a_1 b_{1d}^{(l)} \D^{-1})^{-2} \U \tr (\y_d - \Z \mathbf{m}_d^{(l)}) \\
-%		\,\,\,\,\,\,\,\,\,\, & + a_3 b_{3d}^{(l)} \sum_{i=1}^n (d_i^2 + a_1 b_{1d}^{(l)})^{-1} \Bigg] + \theta_1^{-1} \Bigg\}^{-1}, \\
-%		b_{2d}^{(l+1)} & = \Bigg(\frac{a_3 b_{3d}^{(l)}}{2} \Bigg\{ (\y_d - \X \bmu_d^{(l+1)}) \tr \U^* [\D^* + a_2 b_{2d}^{(l)} (\D^*)^{-1}]^{-2} (\U^*) \tr (\y_d - \X \bmu_d^{(l+1)}) \\
-%		\,\,\,\,\,\,\,\,\,\, & + a_3 b_{3d}^{(l)} \sum_{i=1}^n [(d^*_i)^2 + a_2 b_{2d}^{(l)}]^{-1} \Bigg\} + \theta_2^{-1} \Bigg)^{-1}, \\
-%		b_{3d}^{(l+1)} & = \frac{\y_d \tr \y_d}{2}  - \y_d \tr (\X \bmu_d^{(l+1)} + \Z \mathbf{m}_d^{(l+1)}) + (\bmu_d^{(l+1)}) \tr \X \tr \Z \mathbf{m}_d^{(l+1)} + \frac{\sum_{i=1}^n \(d_i^2 - \frac{d_i^4}{d_i^2 + a_1 b_1^{(l+1)}}\)}{2 a_1 b_{1d}^{(l + 1)}} \\
-%		\,\,\,\,\,\,\,\,\,\, & + \frac{(\bmu_d^{(l)}) \tr \X \tr \X \bmu_d^{(l)}}{2} + \frac{\sum_{t=1}^T \[(d^*_t)^2 - \frac{(d^*_t)^4}{(d^*_t)^2 + a_2 b_2^{(l+1)}}\]}{2 a_2 b_{2d}^{(l + 1)}} + \frac{(\mathbf{m}_d^{(l)}) \tr \Z \tr \Z \mathbf{m}_d^{(l)}}{2} \\
-%		& \,\,\,\,\,\,\,\,\,\, + \frac{a_1 b_{1d}^{(l+1)} (\y_d - \Z \mathbf{m}_d^{(l)}) \tr \U (\D + a_1 b_{1d}^{(l)} \D^{-1})^{-2} \U \tr (\y_d - \Z \mathbf{m}_d^{(l)})}{2} \\
-%		& \,\,\,\,\,\,\,\,\,\, + \frac{a_1 b_{1d}^{(l+1)} a_3 b_{3d}^{(l)} \sum_{i=1}^n (d_i^2 + a_1 b_{1d}^{(l)})^{-1} }{2} \\
-%		& \,\,\,\,\,\,\,\,\,\, + \frac{a_2 b_{2d}^{(l+1)} (\y_d - \X \bmu_d^{(l+1)}) \tr \U^* [\D^* + a_2 b_{2d}^{(l)} (\D^*)^{-1}]^{-2} (\U^*) \tr (\y_d - \X \bmu_d^{(l+1)})}{2} \\
-%		& \,\,\,\,\,\,\,\,\,\, + \frac{a_1 b_{1d}^{(l+1)} a_3 b_{3d}^{(l)} \sum_{i=1}^n [(d^*_i)^2 + a_2 b_{2d}^{(l)}]^{-1}}{2},
-%		\end{align*}
-%		where $(\D^2 + a_1 b_{1d}^{(l)})^{-1}$, $[(\D^*)^2 + a_2 b_{2d}^{(l)}]^{-1}$, $(\D + a_1 b_{1d}^{(l)} \D^{-1})^{-1}$, and $[\D^* + a_2 b_{2d}^{(l)} (\D^*)^{-1}]^{-1}$ are diagonal matrices with entries $(d_t^2 + a_1 b_{1d}^{(l)})^{-1}$, $i=1,\dots,n$, $[(d_t^*)^2 + a_2 b_{2d}^{(l)}]^{-1}$, $t=1,\dots,T$, $(d_i + a_1 b_{1d}^{(l)} / d_i)^{-1}$, $i=1, \dots, n$, and $(d^*_t + a_2 b_{2d}^{(l)} / d^*_t)^{-1}$, $t=1 , \dots, T$, respectively, and hence easily calculated once the singular values are known.

### 2018-12-28 docs/supplement.Rnw
fixing travis ci button

@@
 <<"boxplot_igaussian_res1_mu", fig.cap="a) MSE and b) correlation for posterior means of $\\beta_d$", fig.align="center", echo=FALSE>>=
-load("../results/simulations_igaussian_res1.RData")
+res1 <- read.table("../results/simulations_igaussian_res1.csv")
+set1 <- read.table("../results/simulations_igaussian_set1.csv")
+names1 <- c("inv Gauss", "ind inv Gauss", "inv Gamma")
+temp1 <- cbind(res1$mse.mu.inv.Gauss, res1$mse.mu.ind.inv.Gauss, 
+               res1$mse.mu.inv.Gamma)
+temp2 <- cbind(res1$cor.mu.inv.Gauss, res1$cor.mu.ind.inv.Gauss, 
+               res1$cor.mu.inv.Gamma)
 par(cex=1.3)
 layout(matrix(c(rep(1, 4), rep(2, 4)), 2, 4), widths=1, heights=1, respect=TRUE)
-boxplot(res1$mse.mu, names=names(res1$mse.mu), ylab="MSE", main="a)")
-legend("topright", title=expression(bold("median MSE")), bty="n", 
-       paste(names(res1$mse.mu), round(sapply(res1$mse.mu, median), 2), 
-             sep="="))
-boxplot(res1$cor.mu, names=names(res1$mse.mu), ylab="Correlation", main="b)")
-legend("topright", title=expression(bold("median correlation")), bty="n", 
-       paste(names(res1$mse.mu), round(sapply(res1$cor.mu, median), 2), 
+boxplot(temp1, outline=FALSE, names=names1, ylab="MSE", main="a)")
+legend("topleft", title=expression(bold("median MSE")), bty="n", 
+       paste(names1, round(apply(temp1, 2, median), 2), sep="="))
+boxplot(temp2, names=names1, ylab="Correlation", main="b)")
+legend("bottomleft", title=expression(bold("median correlation")), bty="n", 
+       paste(names1, round(apply(temp2, 2, median), 2), 
              sep="="))
 par(cex=1)
+layout(matrix(1, 1, 1), widths=1, heights=1, respect=TRUE)
 @

### 2018-12-28 code/R/simulations_igaussian.R

@@ -1,7 +1,5 @@
 #!/usr/bin/env Rscript
 
-getwd()
-
 ### change these to user-specific values
 auth_token <- "da10f2b37c513e3383c5b2e0aa1300288329c636"
 
@@ -153,13 +151,9 @@ for(r in 1:nreps) {
 
 warnings()
 
-# test <- read.table(paste(path.res, "simulations_igaussian_res1.csv", sep=""))
-# 
-# 
-# colnames(test)
-# 
-# 
-# 
+
+
+
 # load(paste(path.res, "simulations_igaussian_res1.RData", sep=""))
 # ### convergence
 # # elbo convergence

### 2018-12-28 docs/supplement.Rnw
changed plots creation to .csv version, updated repo README.md file

--- 2018-12-20 docs/supplement.Rnw
@@
 <<"boxplot_igaussian_res1_sigma2", fig.cap="MSEs for posterior means of $\\sigma_d^2$", out.width="50%", fig.align="center", echo=FALSE>>=
-load("../results/simulations_igaussian_res1.RData")
-temp <- sapply(res1$zeta, function(z) {
-  colMeans((2*t(z)/(set1$n + (z==1)*set1$p - 1) - set1$sigma^2)^2)})
+res1 <- read.table("../results/simulations_igaussian_res1.csv")
+set1 <- read.table("../results/simulations_igaussian_set1.csv")
+names1 <- c("inv Gauss", "ind inv Gauss", "inv Gamma")
+temp1 <- list(res1[, substr(colnames(res1), 1, 14)=="zeta.inv.Gauss"],
+              res1[, substr(colnames(res1), 1, 18)=="zeta.ind.inv.Gauss"],
+              res1[, substr(colnames(res1), 1, 5)=="cpost"],
+              res1[, substr(colnames(res1), 1, 5)=="dpost"])
+temp2 <- cbind(colMeans((2*temp1[[1]]/(set1[rownames(set1)=="n", ] + set1[
+  rownames(set1)=="p", ] - 1) - set1[substr(rownames(set1), 1, 5)=="sigma", ]^2)^2),
+  colMeans((2*temp1[[2]]/(set1[rownames(set1)=="n", ] - 1) - 
+              set1[substr(rownames(set1), 1, 5)=="sigma", ]^2)^2),
+               colMeans((temp1[[4]]/(temp1[[3]] - 1) - 
+                           set1[substr(rownames(set1), 1, 5)=="sigma", ]^2)^2))
 par(cex=1.3)
-boxplot(temp, outline=FALSE, names=names(res1$mse.mu), ylab="MSE")
+boxplot(temp2, outline=FALSE, names=names1, ylab="MSE")
 legend("topright", title=expression(bold("median MSE")), bty="n", 
-       paste(names(res1$mse.mu), round(apply(temp, 2, median), 2), sep="="))
+       paste(names1, round(apply(temp2, 2, median), 2), sep="="))
 par(cex=1)
@@
 <<"boxplot_igaussian_res1_gamma2", fig.cap="a) MSE and b) correlation for posterior means of $\\gamma_d^2$", fig.align="center", echo=FALSE>>=
-load("../results/simulations_igaussian_res1.RData")
-temp1 <- sapply(1:3, function(m) {
-  colMeans((t(sqrt(res1$delta[[m]]*res1$theta[[m]]^2/res1$lambda[[m]])*
-                ratio_besselK_cpp(sqrt(res1$delta[[m]]*res1$lambda[[m]])/
-                                    res1$theta[[m]], set1$p)) - set1$gamma^2)^2)})
-temp2 <- sapply(1:3, function(m) {
-  apply(sqrt(res1$delta[[m]]*res1$theta[[m]]^2/res1$lambda[[m]])*
-          ratio_besselK_cpp(sqrt(res1$delta[[m]]*res1$lambda[[m]])/
-                              res1$theta[[m]], set1$p), 1, function(r) {
-                                cor(r, set1$gamma^2)})})
+res1 <- read.table("../results/simulations_igaussian_res1.csv")
+set1 <- read.table("../results/simulations_igaussian_set1.csv")
+names1 <- c("inv Gauss", "ind inv Gauss", "inv Gamma")
+temp1 <- list(res1[, substr(colnames(res1), 1, 15)=="delta.inv.Gauss"],
+              res1[, substr(colnames(res1), 1, 15)=="theta.inv.Gauss"],
+              res1[, substr(colnames(res1), 1, 16)=="lambda.inv.Gauss"],
+              res1[, substr(colnames(res1), 1, 19)=="delta.ind.inv.Gauss"],
+              res1[, substr(colnames(res1), 1, 19)=="theta.ind.inv.Gauss"],
+              res1[, substr(colnames(res1), 1, 20)=="lambda.ind.inv.Gauss"],
+              res1[, substr(colnames(res1), 1, 15)=="apost.inv.Gamma"],
+              res1[, substr(colnames(res1), 1, 15)=="bpost.inv.Gamma"])
+temp2 <- list(as.matrix(sqrt(temp1[[1]]*temp1[[2]]^2/temp1[[3]]))*
+                ratio_besselK_cpp(as.matrix(sqrt(
+                  temp1[[1]]*temp1[[3]])/temp1[[2]]), set1[rownames(set1)=="p", ]),
+              as.matrix(sqrt(temp1[[4]]*temp1[[5]]^2/temp1[[6]]))*
+                ratio_besselK_cpp(as.matrix(sqrt(
+                  temp1[[4]]*temp1[[6]])/temp1[[5]]), set1[rownames(set1)=="p", ]),
+              as.matrix(temp1[[8]]/(temp1[[7]] - 1)))
+temp3 <- sapply(temp2, function(m) {
+  colMeans((t(m) - set1[substr(rownames(set1), 1, 5)=="gamma", ]^2)^2)})
+temp4 <- sapply(temp2, function(m) {apply(t(m), 2, function(r) {
+  cor(r, set1[substr(rownames(set1), 1, 5)=="gamma", ]^2)})})
 par(cex=1.3)
 layout(matrix(c(rep(1, 4), rep(2, 4)), 2, 4), widths=1, heights=1, respect=TRUE)
-boxplot(temp1, names=names(res1$mse.mu), ylab="MSE", main="a)")
-legend("topright", title=expression(bold("median MSE")), bty="n", 
-       paste(names(res1$mse.mu), round(apply(temp1, 2, median), 2), sep="="))
-boxplot(temp2, names=names(res1$mse.mu), ylab="Correlation", main="b)")
-legend("topright", title=expression(bold("median correlation")), bty="n", 
-       paste(names(res1$mse.mu), round(apply(temp2, 2, median), 2), sep="="))
+boxplot(temp3, names=names1, ylab="MSE", main="a)")
+legend("right", title=expression(bold("median MSE")), bty="n", 
+       paste(names1, round(apply(temp3, 2, median), 2), sep="="))
+boxplot(temp4, names=names1, ylab="Correlation", main="b)")
+legend("bottomleft", title=expression(bold("median correlation")), bty="n", 
+       paste(names1, round(apply(temp4, 2, median), 2), sep="="))
 par(cex=1)
+layout(matrix(1, 1, 1), widths=1, heights=1, respect=TRUE)
 @
@@
 <<"boxplot_igaussian_res1_mu", fig.cap="a) MSE and b) correlation for posterior means of $\\beta_d$", fig.align="center", echo=FALSE>>=
-load("../results/simulations_igaussian_res1.RData")
+res1 <- read.table("../results/simulations_igaussian_res1.csv")
+names1 <- c("inv Gauss", "ind inv Gauss", "inv Gamma")
+temp1 <- cbind(res1$mse.mu.inv.Gauss, res1$mse.mu.ind.inv.Gauss, 
+               res1$mse.mu.inv.Gamma)
+temp2 <- cbind(res1$cor.mu.inv.Gauss, res1$cor.mu.ind.inv.Gauss, 
+               res1$cor.mu.inv.Gamma)
 par(cex=1.3)
 layout(matrix(c(rep(1, 4), rep(2, 4)), 2, 4), widths=1, heights=1, respect=TRUE)
-boxplot(res1$mse.mu, names=names(res1$mse.mu), ylab="MSE", main="a)")
-legend("topright", title=expression(bold("median MSE")), bty="n", 
-       paste(names(res1$mse.mu), round(sapply(res1$mse.mu, median), 2), 
-             sep="="))
-boxplot(res1$cor.mu, names=names(res1$mse.mu), ylab="Correlation", main="b)")
-legend("topright", title=expression(bold("median correlation")), bty="n", 
-       paste(names(res1$mse.mu), round(sapply(res1$cor.mu, median), 2), 
+boxplot(temp1, outline=FALSE, names=names1, ylab="MSE", main="a)")
+legend("topleft", title=expression(bold("median MSE")), bty="n", 
+       paste(names1, round(apply(temp1, 2, median), 2), sep="="))
+boxplot(temp2, names=names1, ylab="Correlation", main="b)")
+legend("bottomleft", title=expression(bold("median correlation")), bty="n", 
+       paste(names1, round(apply(temp2, 2, median), 2), 
              sep="="))
 par(cex=1)
+layout(matrix(1, 1, 1), widths=1, heights=1, respect=TRUE)
 @
@@
 <<"boxplot_igaussian_res1_theta", fig.cap="a) MSE and b) correlation for prior means of $\\gamma_d^2$ ($\\theta_d$)", fig.align="center", echo=FALSE>>=
-load("../results/simulations_igaussian_res1.RData")
-temp1 <- sapply(res1$theta, function(th) {colMeans((t(th) - set1$theta)^2)})
-temp2 <- sapply(res1$theta, function(th) {apply(th, 1, function(r) {
-  cor(r, set1$theta)})})
+res1 <- read.table("../results/simulations_igaussian_res1.csv")
+set1 <- read.table("../results/simulations_igaussian_set1.csv")
+names1 <- c("inv Gauss", "ind inv Gauss", "inv Gamma")
+temp1 <- list(res1[, substr(colnames(res1), 1, 15)=="theta.inv.Gauss"],
+              res1[, substr(colnames(res1), 1, 19)=="theta.ind.inv.Gauss"],
+              res1[, substr(colnames(res1), 1, 16)=="aprior.inv.Gamma"],
+              res1[, substr(colnames(res1), 1, 16)=="bprior.inv.Gamma"])
+temp2 <- list(colMeans((t(temp1[[1]]) - 
+                          set1[substr(rownames(set1), 1, 5)=="theta", ])^2),
+              colMeans((t(temp1[[2]]) - 
+                          set1[substr(rownames(set1), 1, 5)=="theta", ])^2),
+              colMeans((t(temp1[[4]]/(temp1[[3]] - 1)) - 
+                          unique(set1[substr(rownames(set1), 1, 5)=="theta", ]))^2))
+temp3 <- list(apply(t(temp1[[1]]), 2, function(r) {
+  cor(r, set1[substr(rownames(set1), 1, 5)=="theta", ])}),
+              apply(t(temp1[[2]]), 2, function(r) {
+                cor(r, set1[substr(rownames(set1), 1, 5)=="theta", ])}),
+              apply(t(temp1[[4]]/(temp1[[3]] - 1)), 2, function(r) {
+                cor(r, unique(set1[substr(rownames(set1), 1, 5)=="theta", ]))}))
 par(cex=1.3)
 layout(matrix(c(rep(1, 4), rep(2, 4)), 2, 4), widths=1, heights=1, respect=TRUE)
-boxplot(temp1, names=names(res1$mse.mu), ylab="MSE", main="a)")
-legend("topright", title=expression(bold("median MSE")), bty="n", 
-       paste(names(res1$mse.mu), round(apply(temp1, 2, median), 2), sep="="))
-boxplot(temp2, names=names(res1$mse.mu), ylab="Correlation", main="b)")
-legend("topright", title=expression(bold("median correlation")), bty="n", 
-       paste(names(res1$mse.mu), round(apply(temp2, 2, median), 2), sep="="))
+boxplot(temp2, names=names1, ylab="MSE", main="a)")
+legend("bottomright", title=expression(bold("median MSE")), bty="n",
+       paste(names1, round(sapply(temp2, median), 2), sep="="))
+boxplot(temp3, names=names1, ylab="Correlation", main="b)")
+legend("bottomleft", title=expression(bold("median correlation")), bty="n",
+       paste(names1, round(sapply(temp3, median), 2), sep="="))
 par(cex=1)
+layout(matrix(1, 1, 1), widths=1, heights=1, respect=TRUE)
 @
@@
 <<"boxplot_igaussian_res1_alpha", fig.cap="a) MSE and b) correlation for $\\alpha$", fig.align="center", echo=FALSE>>=
-load("../results/simulations_igaussian_res1.RData")
-temp1 <- sapply(res1$alpha, function(a) {colMeans((t(a) - set1$alpha)^2)})
-temp2 <- sapply(res1$alpha, function(a) {apply(a, 1, function(r) {
-  cor(r, set1$alpha)})})
+res1 <- read.table("../results/simulations_igaussian_res1.csv")
+set1 <- read.table("../results/simulations_igaussian_set1.csv")
+set1[substr(rownames(set1), 1, 5)=="alpha", ]
+names1 <- c("inv Gauss", "ind inv Gauss")
+temp1 <- list(res1[, substr(colnames(res1), 1, 15)=="alpha.inv.Gauss"],
+              res1[, substr(colnames(res1), 1, 19)=="alpha.ind.inv.Gauss"])
+temp2 <- list(colMeans((t(temp1[[1]]) - set1[substr(rownames(set1), 1, 5)=="alpha", ])^2),
+              colMeans((t(temp1[[2]]) - set1[substr(rownames(set1), 1, 5)=="alpha", ])^2))
+temp3 <- list(apply(t(temp1[[1]]), 2, function(r) {cor(r, set1[substr(rownames(set1), 1, 5)=="alpha", ])}),
+              apply(t(temp1[[2]]), 2, function(r) {cor(r, set1[substr(rownames(set1), 1, 5)=="alpha", ])}))
 par(cex=1.3)
 layout(matrix(c(rep(1, 4), rep(2, 4)), 2, 4), widths=1, heights=1, respect=TRUE)
-boxplot(temp1, names=names(res1$mse.mu), ylab="MSE", main="a)")
+boxplot(temp2, names=names1, ylab="MSE", main="a)")
 legend("topright", title=expression(bold("median MSE")), bty="n",
-       paste(names(res1$mse.mu), round(apply(temp1, 2, median), 2), sep="="))
-boxplot(temp2, names=names(res1$mse.mu), ylab="Correlation", main="b)")
-legend("topright", title=expression(bold("median correlation")), bty="n", 
-       paste(names(res1$mse.mu), round(apply(temp2, 2, median), 2), sep="="))
+       paste(names1, round(sapply(temp2, median), 2), sep="="))
+boxplot(temp3, names=names1, ylab="Correlation", main="b)")
+legend("bottomright", title=expression(bold("median correlation")), bty="n",
+       paste(names1, round(sapply(temp3, median), 2), sep="="))
 par(cex=1)
+layout(matrix(1, 1, 1), widths=1, heights=1, respect=TRUE)
 @

### 2019-01-02 docs/supplement.Rnw
added convergence plots for inverse Gaussian simulations

--- 2018-12-20 docs/supplement.Rnw
@@
 <<"boxplot_igaussian_res1_sigma2", fig.cap="MSEs for posterior means of $\\sigma_d^2$", out.width="50%", fig.align="center", echo=FALSE>>=
-load("../results/simulations_igaussian_res1.RData")
-temp <- sapply(res1$zeta, function(z) {
-  colMeans((2*t(z)/(set1$n + (z==1)*set1$p - 1) - set1$sigma^2)^2)})
+res1 <- read.table("../results/simulations_igaussian_res1.csv")
+set1 <- read.table("../results/simulations_igaussian_set1.csv")
+names1 <- c("inv Gauss", "ind inv Gauss", "inv Gamma")
+temp1 <- list(res1[, substr(colnames(res1), 1, 14)=="zeta.inv.Gauss"],
+              res1[, substr(colnames(res1), 1, 18)=="zeta.ind.inv.Gauss"],
+              res1[, substr(colnames(res1), 1, 5)=="cpost"],
+              res1[, substr(colnames(res1), 1, 5)=="dpost"])
+temp2 <- cbind(colMeans((2*temp1[[1]]/(set1$n + set1$p - 1) - as.numeric(set1[, substr(colnames(set1), 1, 5)=="sigma"])^2)^2),
+  colMeans((2*temp1[[2]]/(set1$n - 1) - 
+              as.numeric(set1[, substr(colnames(set1), 1, 5)=="sigma"])^2)^2),
+               colMeans((temp1[[4]]/(temp1[[3]] - 1) - 
+                           as.numeric(set1[, substr(colnames(set1), 1, 5)=="sigma"])^2)^2))
 par(cex=1.3)
-boxplot(temp, outline=FALSE, names=names(res1$mse.mu), ylab="MSE")
+boxplot(temp2, outline=FALSE, names=names1, ylab="MSE")
 legend("topright", title=expression(bold("median MSE")), bty="n", 
-       paste(names(res1$mse.mu), round(apply(temp, 2, median), 2), sep="="))
+       paste(names1, round(apply(temp2, 2, median), 2), sep="="))
 par(cex=1)
@@
 <<"boxplot_igaussian_res1_gamma2", fig.cap="a) MSE and b) correlation for posterior means of $\\gamma_d^2$", fig.align="center", echo=FALSE>>=
-load("../results/simulations_igaussian_res1.RData")
-temp1 <- sapply(1:3, function(m) {
-  colMeans((t(sqrt(res1$delta[[m]]*res1$theta[[m]]^2/res1$lambda[[m]])*
-                ratio_besselK_cpp(sqrt(res1$delta[[m]]*res1$lambda[[m]])/
-                                    res1$theta[[m]], set1$p)) - set1$gamma^2)^2)})
-temp2 <- sapply(1:3, function(m) {
-  apply(sqrt(res1$delta[[m]]*res1$theta[[m]]^2/res1$lambda[[m]])*
-          ratio_besselK_cpp(sqrt(res1$delta[[m]]*res1$lambda[[m]])/
-                              res1$theta[[m]], set1$p), 1, function(r) {
-                                cor(r, set1$gamma^2)})})
+res1 <- read.table("../results/simulations_igaussian_res1.csv")
+set1 <- read.table("../results/simulations_igaussian_set1.csv")
+names1 <- c("inv Gauss", "ind inv Gauss", "inv Gamma")
+temp1 <- list(res1[, substr(colnames(res1), 1, 15)=="delta.inv.Gauss"],
+              res1[, substr(colnames(res1), 1, 15)=="theta.inv.Gauss"],
+              res1[, substr(colnames(res1), 1, 16)=="lambda.inv.Gauss"],
+              res1[, substr(colnames(res1), 1, 19)=="delta.ind.inv.Gauss"],
+              res1[, substr(colnames(res1), 1, 19)=="theta.ind.inv.Gauss"],
+              res1[, substr(colnames(res1), 1, 20)=="lambda.ind.inv.Gauss"],
+              res1[, substr(colnames(res1), 1, 15)=="apost.inv.Gamma"],
+              res1[, substr(colnames(res1), 1, 15)=="bpost.inv.Gamma"])
+temp2 <- list(as.matrix(sqrt(temp1[[1]]*temp1[[2]]^2/temp1[[3]]))*
+                ratio_besselK_cpp(as.matrix(sqrt(
+                  temp1[[1]]*temp1[[3]])/temp1[[2]]), set1$p),
+              as.matrix(sqrt(temp1[[4]]*temp1[[5]]^2/temp1[[6]]))*
+                ratio_besselK_cpp(as.matrix(sqrt(
+                  temp1[[4]]*temp1[[6]])/temp1[[5]]), set1$p),
+              as.matrix(temp1[[8]]/(temp1[[7]] - 1)))
+temp3 <- sapply(temp2, function(m) {
+  colMeans((t(m) - as.numeric(set1[, substr(colnames(set1), 1, 5)=="gamma"])^2)^2)})
+temp4 <- sapply(temp2, function(m) {apply(t(m), 2, function(r) {
+  cor(r, as.numeric(set1[, substr(colnames(set1), 1, 5)=="gamma"])^2)})})
 par(cex=1.3)
 layout(matrix(c(rep(1, 4), rep(2, 4)), 2, 4), widths=1, heights=1, respect=TRUE)
-boxplot(temp1, names=names(res1$mse.mu), ylab="MSE", main="a)")
-legend("topright", title=expression(bold("median MSE")), bty="n", 
-       paste(names(res1$mse.mu), round(apply(temp1, 2, median), 2), sep="="))
-boxplot(temp2, names=names(res1$mse.mu), ylab="Correlation", main="b)")
-legend("topright", title=expression(bold("median correlation")), bty="n", 
-       paste(names(res1$mse.mu), round(apply(temp2, 2, median), 2), sep="="))
+boxplot(temp3, names=names1, ylab="MSE", main="a)")
+legend("right", title=expression(bold("median MSE")), bty="n", 
+       paste(names1, round(apply(temp3, 2, median), 2), sep="="))
+boxplot(temp4, names=names1, ylab="Correlation", main="b)")
+legend("bottomleft", title=expression(bold("median correlation")), bty="n", 
+       paste(names1, round(apply(temp4, 2, median), 2), sep="="))
 par(cex=1)
+layout(matrix(1, 1, 1), widths=1, heights=1, respect=TRUE)
 @
@@
 <<"boxplot_igaussian_res1_theta", fig.cap="a) MSE and b) correlation for prior means of $\\gamma_d^2$ ($\\theta_d$)", fig.align="center", echo=FALSE>>=
-load("../results/simulations_igaussian_res1.RData")
-temp1 <- sapply(res1$theta, function(th) {colMeans((t(th) - set1$theta)^2)})
-temp2 <- sapply(res1$theta, function(th) {apply(th, 1, function(r) {
-  cor(r, set1$theta)})})
+res1 <- read.table("../results/simulations_igaussian_res1.csv")
+set1 <- read.table("../results/simulations_igaussian_set1.csv")
+names1 <- c("inv Gauss", "ind inv Gauss", "inv Gamma")
+temp1 <- list(res1[, substr(colnames(res1), 1, 15)=="theta.inv.Gauss"],
+              res1[, substr(colnames(res1), 1, 19)=="theta.ind.inv.Gauss"],
+              res1[, substr(colnames(res1), 1, 16)=="aprior.inv.Gamma"],
+              res1[, substr(colnames(res1), 1, 16)=="bprior.inv.Gamma"])
+temp2 <- list(colMeans((t(temp1[[1]]) - 
+                          as.numeric(set1[, substr(colnames(set1), 1, 5)=="theta"]))^2),
+              colMeans((t(temp1[[2]]) - 
+                          as.numeric(set1[, substr(colnames(set1), 1, 5)=="theta"]))^2),
+              colMeans((t(temp1[[4]]/(temp1[[3]] - 1)) - 
+                          unique(as.numeric(set1[, substr(colnames(set1), 1, 5)=="theta"])))^2))
+temp3 <- list(apply(t(temp1[[1]]), 2, function(r) {
+  cor(r, as.numeric(set1[, substr(colnames(set1), 1, 5)=="theta"]))}),
+              apply(t(temp1[[2]]), 2, function(r) {
+                cor(r, as.numeric(set1[, substr(colnames(set1), 1, 5)=="theta"]))}),
+              apply(t(temp1[[4]]/(temp1[[3]] - 1)), 2, function(r) {
+                cor(r, unique(as.numeric(set1[, substr(colnames(set1), 1, 5)=="theta"])))}))
 par(cex=1.3)
 layout(matrix(c(rep(1, 4), rep(2, 4)), 2, 4), widths=1, heights=1, respect=TRUE)
-boxplot(temp1, names=names(res1$mse.mu), ylab="MSE", main="a)")
-legend("topright", title=expression(bold("median MSE")), bty="n", 
-       paste(names(res1$mse.mu), round(apply(temp1, 2, median), 2), sep="="))
-boxplot(temp2, names=names(res1$mse.mu), ylab="Correlation", main="b)")
-legend("topright", title=expression(bold("median correlation")), bty="n", 
-       paste(names(res1$mse.mu), round(apply(temp2, 2, median), 2), sep="="))
+boxplot(temp2, names=names1, ylab="MSE", main="a)")
+legend("bottomright", title=expression(bold("median MSE")), bty="n",
+       paste(names1, round(sapply(temp2, median), 2), sep="="))
+boxplot(temp3, names=names1, ylab="Correlation", main="b)")
+legend("bottomleft", title=expression(bold("median correlation")), bty="n",
+       paste(names1, round(sapply(temp3, median), 2), sep="="))
 par(cex=1)
+layout(matrix(1, 1, 1), widths=1, heights=1, respect=TRUE)
 @
@@
 <<"boxplot_igaussian_res1_alpha", fig.cap="a) MSE and b) correlation for $\\alpha$", fig.align="center", echo=FALSE>>=
-load("../results/simulations_igaussian_res1.RData")
-temp1 <- sapply(res1$alpha, function(a) {colMeans((t(a) - set1$alpha)^2)})
-temp2 <- sapply(res1$alpha, function(a) {apply(a, 1, function(r) {
-  cor(r, set1$alpha)})})
+res1 <- read.table("../results/simulations_igaussian_res1.csv")
+set1 <- read.table("../results/simulations_igaussian_set1.csv")
+
+names1 <- c("inv Gauss", "ind inv Gauss")
+temp1 <- list(res1[, substr(colnames(res1), 1, 15)=="alpha.inv.Gauss"],
+              res1[, substr(colnames(res1), 1, 19)=="alpha.ind.inv.Gauss"])
+temp2 <- list(colMeans((t(temp1[[1]]) - as.numeric(set1[, substr(colnames(set1), 1, 5)=="alpha"]))^2),
+              colMeans((t(temp1[[2]]) - as.numeric(set1[, substr(colnames(set1), 1, 5)=="alpha"]))^2))
+temp3 <- list(apply(t(temp1[[1]]), 2, function(r) {cor(r, as.numeric(set1[, substr(colnames(set1), 1, 5)=="alpha"]))}),
+              apply(t(temp1[[2]]), 2, function(r) {cor(r, as.numeric(set1[, substr(colnames(set1), 1, 5)=="alpha"]))}))
 par(cex=1.3)
 layout(matrix(c(rep(1, 4), rep(2, 4)), 2, 4), widths=1, heights=1, respect=TRUE)
-boxplot(temp1, names=names(res1$mse.mu), ylab="MSE", main="a)")
+boxplot(temp2, names=names1, ylab="MSE", main="a)")
 legend("topright", title=expression(bold("median MSE")), bty="n",
-       paste(names(res1$mse.mu), round(apply(temp1, 2, median), 2), sep="="))
-boxplot(temp2, names=names(res1$mse.mu), ylab="Correlation", main="b)")
-legend("topright", title=expression(bold("median correlation")), bty="n", 
-       paste(names(res1$mse.mu), round(apply(temp2, 2, median), 2), sep="="))
+       paste(names1, round(sapply(temp2, median), 2), sep="="))
+boxplot(temp3, names=names1, ylab="Correlation", main="b)")
+legend("bottomright", title=expression(bold("median correlation")), bty="n",
+       paste(names1, round(sapply(temp3, median), 2), sep="="))
 par(cex=1)
+layout(matrix(1, 1, 1), widths=1, heights=1, respect=TRUE)
+@
+
+<<"lines_igaussian_res1_elbo", fig.cap="Convergence of ELBO for a) inverse Gaussian, b) independent inverse Gaussian, and c) inverse Gamma models", fig.align="center", echo=FALSE>>=
+fit1 <- read.table("../results/simulations_igaussian_fit1.csv")
+par(cex=1.3)
+layout(matrix(c(rep(c(1, 1, 2, 2), times=2), rep(c(0, 3, 3, 0), times=2)), 
+              4, 4, byrow=TRUE), widths=1, heights=1, respect=TRUE)
+plot(fit1[, substr(colnames(fit1), 1, 14)=="inv.Gauss.elbo"][, 1], type="l",
+     ylim=range(fit1[, substr(colnames(fit1), 1, 14)=="inv.Gauss.elbo"]),
+     main="a)", xlab="Iteration", ylab="ELBO")
+for(i in 2:100) {
+  lines(fit1[, substr(colnames(fit1), 1, 14)=="inv.Gauss.elbo"][, i], col=i)
+}
+plot(fit1[, substr(colnames(fit1), 1, 18)=="ind.inv.Gauss.elbo"][, 1], type="l",
+     ylim=range(fit1[, substr(colnames(fit1), 1, 18)=="ind.inv.Gauss.elbo"]),
+     main="b)", xlab="Iteration", ylab="ELBO")
+for(i in 2:100) {
+  lines(fit1[, substr(colnames(fit1), 1, 18)=="ind.inv.Gauss.elbo"][, i], col=i)
+}
+plot(fit1[, substr(colnames(fit1), 1, 14)=="inv.Gamma.elbo"][, 1], type="l",
+     ylim=range(fit1[, substr(colnames(fit1), 1, 14)=="inv.Gamma.elbo"]),
+     main="c)", xlab="Iteration", ylab="ELBO")
+for(i in 2:100) {
+  lines(fit1[, substr(colnames(fit1), 1, 14)=="inv.Gamma.elbo"][, i], col=i)
+}
+par(cex=1)
+layout(matrix(1, 1, 1), widths=1, heights=1, respect=TRUE)
+@
+
+<<"lines_igaussian_res1_theta", fig.cap="Convergence of prior means for a) inverse Gaussian, b) independent inverse Gaussian, and c) inverse Gamma models", fig.align="center", echo=FALSE>>=
+fit1 <- read.table("../results/simulations_igaussian_fit1.csv")
+par(cex=1.3)
+layout(matrix(c(rep(c(1, 1, 2, 2), times=2), rep(c(0, 3, 3, 0), times=2)), 
+              4, 4, byrow=TRUE), widths=1, heights=1, respect=TRUE)
+plot(fit1[, substr(colnames(fit1), 1, 15)=="inv.Gauss.theta"][, 1], type="l",
+     ylim=range(fit1[, substr(colnames(fit1), 1, 15)=="inv.Gauss.theta"]),
+     main="a)", xlab="Iteration", ylab=expression(theta[d]))
+for(i in 2:5) {
+  lines(fit1[, substr(colnames(fit1), 1, 15)=="inv.Gauss.theta"][
+    , (i - 1)*(set1$D/set1$nclass) + 1], col=i)
+}
+plot(fit1[, substr(colnames(fit1), 1, 19)=="ind.inv.Gauss.theta"][, 1], 
+     type="l", 
+     ylim=range(fit1[, substr(colnames(fit1), 1, 19)=="ind.inv.Gauss.theta"]),
+     main="b)", xlab="Iteration", ylab=expression(theta[d]))
+for(i in 2:5) {
+  lines(fit1[, substr(colnames(fit1), 1, 19)=="ind.inv.Gauss.theta"][
+    , (i - 1)*(set1$D/set1$nclass) + 1], col=i)
+}
+plot(sort(fit1[, substr(colnames(fit1), 1, 16)=="inv.Gamma.bprior"][, 1]/
+            (fit1[, substr(colnames(fit1), 1, 16)=="inv.Gamma.aprior"][
+              , 1] - 1)), 
+     type="l", 
+     ylim=range(fit1[, substr(colnames(fit1), 1, 16)=="inv.Gamma.bprior"]/
+                  (fit1[, substr(colnames(fit1), 1, 16)=="inv.Gamma.aprior"] - 
+                     1)),
+     main="c)", xlab="Iteration", ylab=expression(beta/(alpha - 1)))
+for(i in 2:5) {
+  lines(sort((fit1[, substr(colnames(fit1), 1, 16)=="inv.Gamma.bprior"][, ]/
+               (fit1[, substr(colnames(fit1), 1, 16)=="inv.Gamma.aprior"][
+                 , 1] - 1))[, i]), col=i)
+}
+par(cex=1)
+layout(matrix(1, 1, 1), widths=1, heights=1, respect=TRUE)
 @

### 2019-01-02 code/R/simulations_igaussian.R

@@ -8,7 +8,7 @@
 #   library(devtools)
 #   install_github("magnusmunch/cambridge/code", local=FALSE, 
 #                  auth_token=Sys.getenv("GITHUB_PAT"))
-# }getwd
+# }
 library(devtools)
 install_github("magnusmunch/cambridge/code", local=FALSE, 
                auth_token=Sys.getenv("GITHUB_PAT"))
@@ -33,10 +33,9 @@ sigma <- rep(1, D)
 SNR <- 10
 
 # store settings in an object
-set1 <- c(nreps=nreps, n=n, p=p, D=D, nclass=nclass, alpha=alpha, C=C, 
+temp <- c(nreps=nreps, n=n, p=p, D=D, nclass=nclass, alpha=alpha, C=C, 
           theta=theta, gamma=gamma, sigma=sigma, SNR=SNR)
-set1 <- data.frame(nreps=nreps, n=n, p=p, D=D, nclass=nclass, alpha=alpha, C=C, 
-                   theta=theta, gamma=gamma, sigma=sigma, SNR=SNR)
+set1 <- t(matrix(temp)); colnames(set1) <- names(temp)
 
 # objects to store simulation results
 list1 <- sapply(c("inv Gauss", "ind inv Gauss"), function(m) {
@@ -137,11 +136,11 @@ for(r in 1:nreps) {
   temp1 <- temp1[rep(seq_len(nrow(temp1)), times=fit1.igauss1$iter$vb + 1), ]
   temp2 <- data.frame(ind.inv.Gauss=fit1.igauss2$seq.eb)
   temp2 <- temp2[rep(seq_len(nrow(temp2)), times=fit1.igauss2$iter$vb + 1), ]
-  fit1 <- data.frame(inv.Gauss.elbo=fit1.igauss1$seq.elbo, 
-                     inv.Gauss=temp1, ind.inv.Gauss.elbo=fit1.igauss2$seq.elbo,
-                     ind.inv.Gauss=temp2, inv.Gamma.elbo=fit1.gwen$seq.elbo,
+  fit1 <- data.frame(inv.Gauss.elbo=fit1.igauss1$seq.elbo, temp1, 
+                     ind.inv.Gauss.elbo=fit1.igauss2$seq.elbo, temp2, 
+                     inv.Gamma.elbo=fit1.gwen$seq.elbo,
[…]

### 2019-01-02 docs/supplement.Rnw
second simulation run finished

@@
 
+<<"boxplot_igaussian_res2_theta", fig.cap="a) MSE and b) correlation for prior means of $\\gamma_d^2$ ($\\theta_d$) in simulation 2", fig.align="center", echo=FALSE>>=
+res2 <- read.table("../results/simulations_igaussian_res2.csv")
+set2 <- read.table("../results/simulations_igaussian_set2.csv")
+names1 <- c("inv Gauss", "ind inv Gauss", "inv Gamma")
+temp1 <- list(res2[, substr(colnames(res2), 1, 15)=="theta.inv.Gauss"],
+              res2[, substr(colnames(res2), 1, 19)=="theta.ind.inv.Gauss"],
+              res2[, substr(colnames(res2), 1, 16)=="aprior.inv.Gamma"],
+              res2[, substr(colnames(res2), 1, 16)=="bprior.inv.Gamma"])
+temp2 <- list(colMeans((t(temp1[[1]]) - 
+                          as.numeric(set2[, substr(colnames(set2), 1, 5)=="theta"]))^2),
+              colMeans((t(temp1[[2]]) - 
+                          as.numeric(set2[, substr(colnames(set2), 1, 5)=="theta"]))^2),
+              colMeans((t(temp1[[4]]/(temp1[[3]] - 1)) - 
+                          unique(as.numeric(set2[, substr(colnames(set2), 1, 5)=="theta"])))^2))
+temp3 <- list(apply(t(temp1[[1]]), 2, function(r) {
+  cor(r, as.numeric(set2[, substr(colnames(set2), 1, 5)=="theta"]))}),
+              apply(t(temp1[[2]]), 2, function(r) {
+                cor(r, as.numeric(set2[, substr(colnames(set2), 1, 5)=="theta"]))}),
+              apply(t(temp1[[4]]/(temp1[[3]] - 1)), 2, function(r) {
+                cor(r, unique(as.numeric(set2[, substr(colnames(set2), 1, 5)=="theta"])))}))
+par(cex=1.3)
+layout(matrix(c(rep(1, 4), rep(2, 4)), 2, 4), widths=1, heights=1, respect=TRUE)
+boxplot(temp2, names=names1, ylab="MSE", main="a)")
+legend("bottomright", title=expression(bold("median MSE")), bty="n",
+       paste(names1, round(sapply(temp2, median), 2), sep="="))
+boxplot(temp3, names=names1, ylab="Correlation", main="b)")
+legend("bottomleft", title=expression(bold("median correlation")), bty="n",
+       paste(names1, round(sapply(temp3, median), 2), sep="="))
+par(cex=1)
+layout(matrix(1, 1, 1), widths=1, heights=1, respect=TRUE)
+@
+
+<<"boxplot_igaussian_res1_alpha", fig.cap="a) MSE and b) correlation for $\\alpha$", fig.align="center", echo=FALSE>>=
+res2 <- read.table("../results/simulations_igaussian_res2.csv")
+set2 <- read.table("../results/simulations_igaussian_set2.csv")
+
+names1 <- c("inv Gauss", "ind inv Gauss")
+temp1 <- list(res2[, substr(colnames(res2), 1, 15)=="alpha.inv.Gauss"],
+              res2[, substr(colnames(res2), 1, 19)=="alpha.ind.inv.Gauss"])
+temp2 <- list(colMeans((t(temp1[[1]]) - as.numeric(set2[, substr(colnames(set2), 1, 5)=="alpha"]))^2),
+              colMeans((t(temp1[[2]]) - as.numeric(set2[, substr(colnames(set2), 1, 5)=="alpha"]))^2))
+temp3 <- list(apply(t(temp1[[1]]), 2, function(r) {cor(r, as.numeric(set2[, substr(colnames(set2), 1, 5)=="alpha"]))}),
+              apply(t(temp1[[2]]), 2, function(r) {cor(r, as.numeric(set2[, substr(colnames(set2), 1, 5)=="alpha"]))}))
+par(cex=1.3)
+layout(matrix(c(rep(1, 4), rep(2, 4)), 2, 4), widths=1, heights=1, respect=TRUE)
+boxplot(temp2, names=names1, ylab="MSE", main="a)")
+legend("topright", title=expression(bold("median MSE")), bty="n",
+       paste(names1, round(sapply(temp2, median), 2), sep="="))
+boxplot(temp3, names=names1, ylab="Correlation", main="b)")
+legend("bottomright", title=expression(bold("median correlation")), bty="n",
+       paste(names1, round(sapply(temp3, median), 2), sep="="))
+par(cex=1)
+layout(matrix(1, 1, 1), widths=1, heights=1, respect=TRUE)
+@
+
+<<>>=
+fit2 <- read.table("../results/simulations_igaussian_fit2.csv")
+plot(fit2[, substr(colnames(fit2), 1, 15)=="inv.Gauss.alpha"][, 1], type="l",
+     ylim=range(fit2[, substr(colnames(fit2), 1, 15)=="inv.Gauss.alpha"]))
+for(i in 2:5) {
+  lines(fit2[, substr(colnames(fit2), 1, 15)=="inv.Gauss.alpha"][, 2], col=i)  
+}
+@
+
   \section{Starting values}

### 2019-01-02 code/R/ccle.R

@@ -33,4 +33,3 @@ cell.lines <- unlist(strsplit("5637 	22Rv1 	786-O 	A-204 	A-253 	A2780 	A3/KAW 	
 cell.lines2 <- data2[data2$Primary.Cell.Line.Name %in% cell.lines & data2$Compound=="Nutlin-3", c(2, 11)]
 cell.lines3 <- cell.lines2[order(cell.lines2$Primary.Cell.Line.Name), ]
 cell.lines3
-

### 2019-01-02 code/R/gdsc.R

@@ -1,3 +1,27 @@
+#!/usr/bin/env Rscript
+
+### installation of package
+# if(!("cambridge" %in% installed.packages())) {
+#   if(!("devtools" %in% installed.packages())) {
+#     install.packages("devtools")
+#   }
+#   library(devtools)
+#   install_github("magnusmunch/cambridge/code", local=FALSE, 
+#                  auth_token=Sys.getenv("GITHUB_PAT"))
+# }
+library(devtools)
+install_github("magnusmunch/cambridge/code", local=FALSE, 
+               auth_token=Sys.getenv("GITHUB_PAT"))
+install_github("cancerrxgene/gdscIC50", build_vignettes=TRUE)
+
+### libraries
+library(cambridge)
+library(gdscIC50)
+vignette("gdscIC50")
+
+test <- gdsc_example
+
+
 path.data <- "/Users/magnusmunch/Documents/OneDrive/PhD/project_cambridge/data/"
 
 
@@ -14,3 +38,11 @@ hist(rowMeans(is.na(data)), xlab="Proportion missing",
      main="Proportion missing in observations")
 
 data4 <- na.omit(data3[, apply(data3, 2, function(x) {sum(is.na(x))}) <= 150])
+
+# 1.1) Deleting missing drug response data: we sequentially deleted missing drugs and missing cell lines from bigger proportions to small proportions, to be more possible to get as big nonmissing matrix as possible. But we still cannot guarantee that is the biggest nonmissing matrix.
+# 
+# 1.2) Preselecting GEX features: we selected gene expression features explaining 50% variations over all cell lines, since this results in not too features and probably remains those important features. 
+# 
+# 1.3) Preselecting CNV: we used the same strategy as Garnett et al. (2012) where the GDSC data are from.
+# 
+# 1.4) Preselecting MUT: we deleted mutated genes with less than two cell lines.

### 2019-01-02 code/R/simulations_igaussian.R

@@ -16,7 +16,7 @@ install_github("magnusmunch/cambridge/code", local=FALSE,
 ### libraries
 library(cambridge)
 
-### simulations
+################################# simulation 1 #################################
 # settings
 nreps <- 100
 n <- 100
@@ -33,9 +33,143 @@ sigma <- rep(1, D)
 SNR <- 10
 
 # store settings in an object
+temp <- c(nreps=nreps, n=n, p=p, D=D, nclass=nclass, alpha=alpha, C=C,
+          theta=theta, gamma=gamma, sigma=sigma, SNR=SNR)
+set1 <- t(matrix(temp))
+colnames(set1) <- names(temp)
+
+# objects to store simulation results
+list1 <- sapply(c("inv Gauss", "ind inv Gauss"), function(m) {
+  matrix(NA, ncol=D, nrow=nreps,
+         dimnames=list(NULL, paste("drug", c(1:D), sep="")))}, simplify=FALSE)
+list2 <- sapply(c("inv Gamma"), function(m) {
+  matrix(NA, ncol=D, nrow=nreps,
+         dimnames=list(NULL, paste("drug", c(1:D), sep="")))}, simplify=FALSE)
+list3 <- sapply(c("inv Gauss", "ind inv Gauss"), function(m) {
+  matrix(NA, ncol=nclass, nrow=nreps,
+         dimnames=list(NULL, paste("class", c(1:nclass), sep="")))},
+  simplify=FALSE)
+list4 <- sapply(c("inv Gamma"), function(m) {
+  matrix(NA, ncol=nclass, nrow=nreps,
+         dimnames=list(NULL, paste("class", c(1:nclass), sep="")))},
+  simplify=FALSE)
+list5 <- sapply(c("inv Gauss", "ind inv Gauss"), function(m) {
+  rep(NA, nreps)}, simplify=FALSE)
+list6 <- sapply(c("inv Gauss", "ind inv Gauss", "inv Gamma"), function(m) {
+  rep(NA, nreps)}, simplify=FALSE)
+res1 <- data.frame(theta=list1, delta=list1, zeta=list1, apost=list2,
+                   bpost=list2, cpost=list2, dpost=list2, alpha=list3,
[…]

### 2019-01-02 docs/manuscript.Rnw
start with Oslo pipeline for GDSC and CCLE

@@
 	\section{Data application}
+	We predict IC50 of the drugs.
+Other choices of efficacy measures are possible, but a discussion on the pros and cons of these different choices is not the aim of this paper.
+In any case, the presented methods may be used with any continuous efficay measure.
+
 	- check GSK open target EBI or ask Chris Wallace about it. Database that links targets to drugs and diseases. Might be useful for drugs co-data.
@@
 	\subsection{GDSC}
+	After iteratively removing the cell line or drug with the highest proportion of missings until there are no more missing values, we end up with  IC50 estimates for 495 cell lines on 167 drugs.
 

### 2019-01-02 docs/supplement.Rnw
start with Oslo pipeline for GDSC and CCLE

--- 2018-12-28 docs/supplement.Rnw
@@
 
-<<"boxplot_igaussian_res1_sigma2", fig.cap="MSEs for posterior means of $\\sigma_d^2$", out.width="50%", fig.align="center", echo=FALSE>>=
+<<"boxplot_igaussian_res1_sigma2", fig.cap="MSEs for posterior means of $\\sigma_d^2$ in simulation 1", out.width="50%", fig.align="center", echo=FALSE>>=
 res1 <- read.table("../results/simulations_igaussian_res1.csv")
@@
               res1[, substr(colnames(res1), 1, 5)=="dpost"])
-temp2 <- cbind(colMeans((2*temp1[[1]]/(set1[rownames(set1)=="n", ] + set1[
-  rownames(set1)=="p", ] - 1) - set1[substr(rownames(set1), 1, 5)=="sigma", ]^2)^2),
-  colMeans((2*temp1[[2]]/(set1[rownames(set1)=="n", ] - 1) - 
-              set1[substr(rownames(set1), 1, 5)=="sigma", ]^2)^2),
+temp2 <- cbind(colMeans((2*temp1[[1]]/(set1$n + set1$p - 1) - as.numeric(set1[, substr(colnames(set1), 1, 5)=="sigma"])^2)^2),
+  colMeans((2*temp1[[2]]/(set1$n - 1) - 
+              as.numeric(set1[, substr(colnames(set1), 1, 5)=="sigma"])^2)^2),
                colMeans((temp1[[4]]/(temp1[[3]] - 1) - 
-                           set1[substr(rownames(set1), 1, 5)=="sigma", ]^2)^2))
+                           as.numeric(set1[, substr(colnames(set1), 1, 5)=="sigma"])^2)^2))
 par(cex=1.3)
@@
 
-<<"boxplot_igaussian_res1_gamma2", fig.cap="a) MSE and b) correlation for posterior means of $\\gamma_d^2$", fig.align="center", echo=FALSE>>=
+<<"boxplot_igaussian_res1_gamma2", fig.cap="a) MSE and b) correlation for posterior means of $\\gamma_d^2$ in simulation 1", fig.align="center", echo=FALSE>>=
 res1 <- read.table("../results/simulations_igaussian_res1.csv")
@@
                 ratio_besselK_cpp(as.matrix(sqrt(
-                  temp1[[1]]*temp1[[3]])/temp1[[2]]), set1[rownames(set1)=="p", ]),
+                  temp1[[1]]*temp1[[3]])/temp1[[2]]), set1$p),
               as.matrix(sqrt(temp1[[4]]*temp1[[5]]^2/temp1[[6]]))*
                 ratio_besselK_cpp(as.matrix(sqrt(
-                  temp1[[4]]*temp1[[6]])/temp1[[5]]), set1[rownames(set1)=="p", ]),
+                  temp1[[4]]*temp1[[6]])/temp1[[5]]), set1$p),
               as.matrix(temp1[[8]]/(temp1[[7]] - 1)))
 temp3 <- sapply(temp2, function(m) {
-  colMeans((t(m) - set1[substr(rownames(set1), 1, 5)=="gamma", ]^2)^2)})
+  colMeans((t(m) - as.numeric(set1[, substr(colnames(set1), 1, 5)=="gamma"])^2)^2)})
 temp4 <- sapply(temp2, function(m) {apply(t(m), 2, function(r) {
-  cor(r, set1[substr(rownames(set1), 1, 5)=="gamma", ]^2)})})
+  cor(r, as.numeric(set1[, substr(colnames(set1), 1, 5)=="gamma"])^2)})})
 par(cex=1.3)
@@
 
-<<"boxplot_igaussian_res1_mu", fig.cap="a) MSE and b) correlation for posterior means of $\\beta_d$", fig.align="center", echo=FALSE>>=
+<<"boxplot_igaussian_res1_mu", fig.cap="a) MSE and b) correlation for posterior means of $\\beta_d$ in simulation 1", fig.align="center", echo=FALSE>>=
 res1 <- read.table("../results/simulations_igaussian_res1.csv")
@@
 
-<<"boxplot_igaussian_res1_theta", fig.cap="a) MSE and b) correlation for prior means of $\\gamma_d^2$ ($\\theta_d$)", fig.align="center", echo=FALSE>>=
+<<"boxplot_igaussian_res1_theta", fig.cap="a) MSE and b) correlation for prior means of $\\gamma_d^2$ ($\\theta_d$) in simulation 1", fig.align="center", echo=FALSE>>=
 res1 <- read.table("../results/simulations_igaussian_res1.csv")
@@
 temp2 <- list(colMeans((t(temp1[[1]]) - 
-                          set1[substr(rownames(set1), 1, 5)=="theta", ])^2),
+                          as.numeric(set1[, substr(colnames(set1), 1, 5)=="theta"]))^2),
               colMeans((t(temp1[[2]]) - 
-                          set1[substr(rownames(set1), 1, 5)=="theta", ])^2),
+                          as.numeric(set1[, substr(colnames(set1), 1, 5)=="theta"]))^2),
               colMeans((t(temp1[[4]]/(temp1[[3]] - 1)) - 
-                          unique(set1[substr(rownames(set1), 1, 5)=="theta", ]))^2))
+                          unique(as.numeric(set1[, substr(colnames(set1), 1, 5)=="theta"])))^2))
 temp3 <- list(apply(t(temp1[[1]]), 2, function(r) {
-  cor(r, set1[substr(rownames(set1), 1, 5)=="theta", ])}),
+  cor(r, as.numeric(set1[, substr(colnames(set1), 1, 5)=="theta"]))}),
               apply(t(temp1[[2]]), 2, function(r) {
-                cor(r, set1[substr(rownames(set1), 1, 5)=="theta", ])}),
+                cor(r, as.numeric(set1[, substr(colnames(set1), 1, 5)=="theta"]))}),
               apply(t(temp1[[4]]/(temp1[[3]] - 1)), 2, function(r) {
-                cor(r, unique(set1[substr(rownames(set1), 1, 5)=="theta", ]))}))
+                cor(r, unique(as.numeric(set1[, substr(colnames(set1), 1, 5)=="theta"])))}))
 par(cex=1.3)
@@
 
-<<"boxplot_igaussian_res1_alpha", fig.cap="a) MSE and b) correlation for $\\alpha$", fig.align="center", echo=FALSE>>=
+<<"boxplot_igaussian_res1_alpha", fig.cap="a) MSE and b) correlation for $\\alpha$ in simulation 1", fig.align="center", echo=FALSE>>=
 res1 <- read.table("../results/simulations_igaussian_res1.csv")
 set1 <- read.table("../results/simulations_igaussian_set1.csv")
-set1[substr(rownames(set1), 1, 5)=="alpha", ]
+
 names1 <- c("inv Gauss", "ind inv Gauss")
@@
               res1[, substr(colnames(res1), 1, 19)=="alpha.ind.inv.Gauss"])
-temp2 <- list(colMeans((t(temp1[[1]]) - set1[substr(rownames(set1), 1, 5)=="alpha", ])^2),
-              colMeans((t(temp1[[2]]) - set1[substr(rownames(set1), 1, 5)=="alpha", ])^2))
-temp3 <- list(apply(t(temp1[[1]]), 2, function(r) {cor(r, set1[substr(rownames(set1), 1, 5)=="alpha", ])}),
-              apply(t(temp1[[2]]), 2, function(r) {cor(r, set1[substr(rownames(set1), 1, 5)=="alpha", ])}))
+temp2 <- list(colMeans((t(temp1[[1]]) - as.numeric(set1[, substr(colnames(set1), 1, 5)=="alpha"]))^2),
+              colMeans((t(temp1[[2]]) - as.numeric(set1[, substr(colnames(set1), 1, 5)=="alpha"]))^2))
+temp3 <- list(apply(t(temp1[[1]]), 2, function(r) {cor(r, as.numeric(set1[, substr(colnames(set1), 1, 5)=="alpha"]))}),
+              apply(t(temp1[[2]]), 2, function(r) {cor(r, as.numeric(set1[, substr(colnames(set1), 1, 5)=="alpha"]))}))
 par(cex=1.3)
@@
        paste(names1, round(sapply(temp3, median), 2), sep="="))
+par(cex=1)
+layout(matrix(1, 1, 1), widths=1, heights=1, respect=TRUE)
+@
+
+<<"lines_igaussian_res1_elbo", fig.cap="Convergence of ELBO for a) inverse Gaussian, b) independent inverse Gaussian, and c) inverse Gamma models in simulation 1", fig.align="center", echo=FALSE>>=
+fit1 <- read.table("../results/simulations_igaussian_fit1.csv")
+par(cex=1.3)
+layout(matrix(c(rep(c(1, 1, 2, 2), times=2), rep(c(0, 3, 3, 0), times=2)), 
+              4, 4, byrow=TRUE), widths=1, heights=1, respect=TRUE)
+plot(fit1[, substr(colnames(fit1), 1, 14)=="inv.Gauss.elbo"][, 1], type="l",
+     ylim=range(fit1[, substr(colnames(fit1), 1, 14)=="inv.Gauss.elbo"]),
+     main="a)", xlab="Iteration", ylab="ELBO")
+for(i in 2:100) {
+  lines(fit1[, substr(colnames(fit1), 1, 14)=="inv.Gauss.elbo"][, i], col=i)
+}
+plot(fit1[, substr(colnames(fit1), 1, 18)=="ind.inv.Gauss.elbo"][, 1], type="l",
+     ylim=range(fit1[, substr(colnames(fit1), 1, 18)=="ind.inv.Gauss.elbo"]),
+     main="b)", xlab="Iteration", ylab="ELBO")
+for(i in 2:100) {
+  lines(fit1[, substr(colnames(fit1), 1, 18)=="ind.inv.Gauss.elbo"][, i], col=i)
+}
+plot(fit1[, substr(colnames(fit1), 1, 14)=="inv.Gamma.elbo"][, 1], type="l",
+     ylim=range(fit1[, substr(colnames(fit1), 1, 14)=="inv.Gamma.elbo"]),
+     main="c)", xlab="Iteration", ylab="ELBO")
+for(i in 2:100) {
+  lines(fit1[, substr(colnames(fit1), 1, 14)=="inv.Gamma.elbo"][, i], col=i)
+}
+par(cex=1)
+layout(matrix(1, 1, 1), widths=1, heights=1, respect=TRUE)
+@
+
+<<"lines_igaussian_res1_theta", fig.cap="Convergence of prior means for a) inverse Gaussian, b) independent inverse Gaussian, and c) inverse Gamma models in simulation 1", fig.align="center", echo=FALSE>>=
+fit1 <- read.table("../results/simulations_igaussian_fit1.csv")
+set1 <- read.table("../results/simulations_igaussian_set1.csv")
+
+par(cex=1.3)
+layout(matrix(c(rep(c(1, 1, 2, 2), times=2), rep(c(0, 3, 3, 0), times=2)), 
+              4, 4, byrow=TRUE), widths=1, heights=1, respect=TRUE)
+plot(fit1[, substr(colnames(fit1), 1, 15)=="inv.Gauss.theta"][, 1], type="l",
+     ylim=range(fit1[, substr(colnames(fit1), 1, 15)=="inv.Gauss.theta"]),
+     main="a)", xlab="Iteration", ylab=expression(theta[d]))
+for(i in 2:5) {
+  lines(fit1[, substr(colnames(fit1), 1, 15)=="inv.Gauss.theta"][
+    , (i - 1)*(set1$D/set1$nclass) + 1], col=i)
+}
+plot(fit1[, substr(colnames(fit1), 1, 19)=="ind.inv.Gauss.theta"][, 1], 
+     type="l", 
+     ylim=range(fit1[, substr(colnames(fit1), 1, 19)=="ind.inv.Gauss.theta"]),
+     main="b)", xlab="Iteration", ylab=expression(theta[d]))
+for(i in 2:5) {
+  lines(fit1[, substr(colnames(fit1), 1, 19)=="ind.inv.Gauss.theta"][
+    , (i - 1)*(set1$D/set1$nclass) + 1], col=i)
+}
+plot(sort(fit1[, substr(colnames(fit1), 1, 16)=="inv.Gamma.bprior"][, 1]/
+            (fit1[, substr(colnames(fit1), 1, 16)=="inv.Gamma.aprior"][
+              , 1] - 1)), 
+     type="l", 
+     ylim=range(fit1[, substr(colnames(fit1), 1, 16)=="inv.Gamma.bprior"]/
+                  (fit1[, substr(colnames(fit1), 1, 16)=="inv.Gamma.aprior"] - 
+                     1)),
+     main="c)", xlab="Iteration", ylab=expression(beta/(alpha - 1)))
+for(i in 2:5) {
+  lines(sort((fit1[, substr(colnames(fit1), 1, 16)=="inv.Gamma.bprior"][, ]/
+               (fit1[, substr(colnames(fit1), 1, 16)=="inv.Gamma.aprior"][
+                 , 1] - 1))[, i]), col=i)
+}
+par(cex=1)
+layout(matrix(1, 1, 1), widths=1, heights=1, respect=TRUE)
+@
+
+Simulations with no difference between classes:
+
+<<"boxplot_igaussian_res2_theta", fig.cap="a) MSE and b) correlation for prior means of $\\gamma_d^2$ ($\\theta_d$) in simulation 2", out.width="50%", fig.align="center", echo=FALSE>>=
+res2 <- read.table("../results/simulations_igaussian_res2.csv")
+set2 <- read.table("../results/simulations_igaussian_set2.csv")
+names1 <- c("inv Gauss", "ind inv Gauss", "inv Gamma")
+temp1 <- list(res2[, substr(colnames(res2), 1, 15)=="theta.inv.Gauss"],
+              res2[, substr(colnames(res2), 1, 19)=="theta.ind.inv.Gauss"],
+              res2[, substr(colnames(res2), 1, 16)=="aprior.inv.Gamma"],
+              res2[, substr(colnames(res2), 1, 16)=="bprior.inv.Gamma"])
+temp2 <- list(colMeans((t(temp1[[1]]) - 
+                          as.numeric(set2[, substr(colnames(set2), 1, 5)=="theta"]))^2),
+              colMeans((t(temp1[[2]]) - 
+                          as.numeric(set2[, substr(colnames(set2), 1, 5)=="theta"]))^2),
+              colMeans((t(temp1[[4]]/(temp1[[3]] - 1)) - 
+                          unique(as.numeric(set2[, substr(colnames(set2), 1, 5)=="theta"])))^2))
+par(cex=1.3)
+boxplot(temp2, names=names1, ylab="MSE", main="a)")
+legend("bottomright", title=expression(bold("median MSE")), bty="n",
+       paste(names1, round(sapply(temp2, median), 2), sep="="))
+par(cex=1)
+@
+
+<<"lines_igaussian_res2_alpha", fig.cap="Convergence of $\\alpha$ for a) inverse Gaussian and b) independent inverse Gaussian in simulation 2", fig.align="center", echo=FALSE>>=
+fit2 <- read.table("../results/simulations_igaussian_fit2.csv")
+set2 <- read.table("../results/simulations_igaussian_set2.csv")
+
+par(cex=1.3)
+layout(matrix(rep(c(1, 1, 2, 2), times=2), 2, 4, byrow=TRUE), widths=1, 
+       heights=1, respect=TRUE)
+plot(fit2[, substr(colnames(fit2), 1, 15)=="inv.Gauss.alpha"][, 1], type="l",
+     ylim=range(fit2[, substr(colnames(fit2), 1, 15)=="inv.Gauss.alpha"]),
+     main="a)", xlab="Iteration", ylab=expression(theta[d]))
+for(i in 2:5) {
+  lines(fit2[, substr(colnames(fit2), 1, 15)=="inv.Gauss.alpha"][, i], col=i)
+}
+plot(fit2[, substr(colnames(fit2), 1, 19)=="ind.inv.Gauss.alpha"][, 1], 
+     type="l", 
+     ylim=range(fit2[, substr(colnames(fit2), 1, 19)=="ind.inv.Gauss.alpha"]),
+     main="b)", xlab="Iteration", ylab=expression(theta[d]))
+for(i in 2:5) {
+  lines(fit2[, substr(colnames(fit2), 1, 19)=="ind.inv.Gauss.alpha"][, i], 
+        col=i)
+}
 par(cex=1)

### 2019-01-03 code/R/gdsc.R
gdsc.R now reads in drug response and expression

@@ -1,5 +1,10 @@
 #!/usr/bin/env Rscript
 
+# set working directory to source (this) file (only works with Rstudio)
+if(any(grepl("rstudio", search()))) {
+  setwd(gsub("(.*)/.*", "\\1", rstudioapi::getActiveDocumentContext()$path))  
+}
+
 ### installation of package
 # if(!("cambridge" %in% installed.packages())) {
 #   if(!("devtools" %in% installed.packages())) {
@@ -15,24 +20,73 @@ install_github("magnusmunch/cambridge/code", local=FALSE,
 
 ### libraries
 library(cambridge)
+library(gdata)
 
 ################################## analysis 1 ##################################
 # loading the data either from a local file or from the GDSC website
-# data <- read.table("ftp://ftp.sanger.ac.uk/pub/project/cancerrxgene/releases/current_release/v17.3_fitted_dose_response.xlsx",
-#                    header=TRUE, sep=",", stringsAsFactors=FALSE)
-resp <- read.table("../../data/v17.3_fitted_dose_response.csv", header=TRUE, 
-                   sep=",", stringsAsFactors=FALSE)
+if(file.exists("../../data/v17.3_fitted_dose_response.csv")) {
+  resp1 <- read.table("../../data/v17.3_fitted_dose_response.csv", header=TRUE, 
+                      sep=",", stringsAsFactors=FALSE)
+} else {
+  resp1 <- read.table("ftp://ftp.sanger.ac.uk/pub/project/cancerrxgene/releases/current_release/v17.3_fitted_dose_response.xlsx",
+                      header=TRUE, sep=",", stringsAsFactors=FALSE)
+}
+if(file.exists("../../data/sanger1018_brainarray_ensemblgene_rma.txt.gz")) {
+  expr1 <- read.table("../../data/sanger1018_brainarray_ensemblgene_rma.txt.gz", 
+                      header=TRUE, stringsAsFactors=FALSE)
+} else {
+  expr1 <- read.table(gcon(url("ftp://ftp.sanger.ac.uk/pub/project/cancerrxgene/releases/current_release/sanger1018_brainarray_ensemblgene_rma.txt.gz"), 
+                           text=TRUE), header=TRUE, stringsAsFactors=FALSE)
+}
[…]

### 2019-01-04 code/R/gdsc.R
gdsc.R includes parsing of pubchem data

@@ -21,6 +21,7 @@ install_github("magnusmunch/cambridge/code", local=FALSE,
 ### libraries
 library(cambridge)
 library(gdata)
+library(rcdk)
 
 ################################## analysis 1 ##################################
 # loading the data either from a local file or from the GDSC website
@@ -115,23 +116,81 @@ expr <- expr2[rownames(expr2) %in% rownames(resp5), ]
 # remove all in-between steps
 rm(list=c("resp1", "resp2", "resp3", "resp4", "resp5", "expr1", "expr2"))
 
+# scaling the response and expressions to mean 0 and variance 1
 expr.scale <- scale(expr)
 resp.scale <- scale(resp)
 
+# creating some drug covariates
 target <- drug1$TARGET[match(colnames(resp.scale), drug1$DRUG_NAME)]
 pathway <- drug1$TARGET_PATHWAY[match(colnames(resp.scale), drug1$DRUG_NAME)]
 stage <- drug2$Clinical.Stage
 action <- drug2$Action
 cluster <- drug4$Cluster.identifier
 
-paste("https://pubchem.ncbi.nlm.nih.gov/rest/pug/compound/name/",
-      paste(drug1$DRUG_NAME, collapse=","), "/property/MolecularFormula/CSV", 
-      sep="")
+# retrieve CID and SMILES from PUG REST
+smiles.list <- vector("list", nrow(drug1))
+for(i in 1:nrow(drug1)) {
+  cat("\r", "drug", i)
+  
+  # create a vector of all drug name synonyms
+  syn <- c(drug1$DRUG_NAME[i], trimws(unlist(strsplit(drug1$SYNONYMS[i], ","))))
+  
+  # loop over the synonyms to find SMILES and CID
+  sfound <- FALSE
+  snum <- 0
+  start <- -1
+  while(!sfound & (snum < length(syn))) {
+    snum <- snum + 1
+    
+    # see PUG REST website for url explanation
+    url <- paste("https://pubchem.ncbi.nlm.nih.gov/rest/pug/compound/name/",
+                 paste(syn[snum], collapse=","), 
+                 "/property/CanonicalSMILES/CSV", sep="")
+    
+    # PUG REST asks for not more than 5 requests per second
+    elapsed <- proc.time()[3] - start
+    Sys.sleep((0.2 - elapsed)*(elapsed < 0.2))
+    start <- proc.time()[3]
+    
[…]

### 2019-01-05 docs/supplement.Rnw
figure creation is external code called from .rnw file

--- 2019-01-02 docs/supplement.Rnw
@@
 
+<<figures, cache=FALSE, echo=FALSE>>=
+read_chunk("../code/R/figures.R")
+@
+
 % settings
+\pgfplotsset{compat=1.13}
 \graphicspath{{/Users/magnusmunch/Documents/OneDrive/PhD/project_cambridge/graphs/}}
@@
 
-<<"boxplot_igaussian_res1_sigma2", fig.cap="MSEs for posterior means of $\\sigma_d^2$", out.width="50%", fig.align="center", echo=FALSE>>=
-res1 <- read.table("../results/simulations_igaussian_res1.csv")
-set1 <- read.table("../results/simulations_igaussian_set1.csv")
-names1 <- c("inv Gauss", "ind inv Gauss", "inv Gamma")
-temp1 <- list(res1[, substr(colnames(res1), 1, 14)=="zeta.inv.Gauss"],
-              res1[, substr(colnames(res1), 1, 18)=="zeta.ind.inv.Gauss"],
-              res1[, substr(colnames(res1), 1, 5)=="cpost"],
-              res1[, substr(colnames(res1), 1, 5)=="dpost"])
-temp2 <- cbind(colMeans((2*temp1[[1]]/(set1$n + set1$p - 1) - as.numeric(set1[, substr(colnames(set1), 1, 5)=="sigma"])^2)^2),
-  colMeans((2*temp1[[2]]/(set1$n - 1) - 
-              as.numeric(set1[, substr(colnames(set1), 1, 5)=="sigma"])^2)^2),
-               colMeans((temp1[[4]]/(temp1[[3]] - 1) - 
-                           as.numeric(set1[, substr(colnames(set1), 1, 5)=="sigma"])^2)^2))
-par(cex=1.3)
-boxplot(temp2, outline=FALSE, names=names1, ylab="MSE")
-legend("topright", title=expression(bold("median MSE")), bty="n", 
-       paste(names1, round(apply(temp2, 2, median), 2), sep="="))
-par(cex=1)
+<<boxplot_igaussian_res1_sigma2, cache=FALSE, echo=FALSE, fig.cap="MSEs for posterior means of $\\sigma_d^2$ in simulation 1", out.width="50%", fig.align="center">>=
 @
 
-<<"boxplot_igaussian_res1_gamma2", fig.cap="a) MSE and b) correlation for posterior means of $\\gamma_d^2$", fig.align="center", echo=FALSE>>=
-res1 <- read.table("../results/simulations_igaussian_res1.csv")
-set1 <- read.table("../results/simulations_igaussian_set1.csv")
-names1 <- c("inv Gauss", "ind inv Gauss", "inv Gamma")
-temp1 <- list(res1[, substr(colnames(res1), 1, 15)=="delta.inv.Gauss"],
-              res1[, substr(colnames(res1), 1, 15)=="theta.inv.Gauss"],
-              res1[, substr(colnames(res1), 1, 16)=="lambda.inv.Gauss"],
-              res1[, substr(colnames(res1), 1, 19)=="delta.ind.inv.Gauss"],
-              res1[, substr(colnames(res1), 1, 19)=="theta.ind.inv.Gauss"],
-              res1[, substr(colnames(res1), 1, 20)=="lambda.ind.inv.Gauss"],
-              res1[, substr(colnames(res1), 1, 15)=="apost.inv.Gamma"],
-              res1[, substr(colnames(res1), 1, 15)=="bpost.inv.Gamma"])
-temp2 <- list(as.matrix(sqrt(temp1[[1]]*temp1[[2]]^2/temp1[[3]]))*
-                ratio_besselK_cpp(as.matrix(sqrt(
-                  temp1[[1]]*temp1[[3]])/temp1[[2]]), set1$p),
-              as.matrix(sqrt(temp1[[4]]*temp1[[5]]^2/temp1[[6]]))*
-                ratio_besselK_cpp(as.matrix(sqrt(
-                  temp1[[4]]*temp1[[6]])/temp1[[5]]), set1$p),
-              as.matrix(temp1[[8]]/(temp1[[7]] - 1)))
-temp3 <- sapply(temp2, function(m) {
-  colMeans((t(m) - as.numeric(set1[, substr(colnames(set1), 1, 5)=="gamma"])^2)^2)})
-temp4 <- sapply(temp2, function(m) {apply(t(m), 2, function(r) {
-  cor(r, as.numeric(set1[, substr(colnames(set1), 1, 5)=="gamma"])^2)})})
-par(cex=1.3)
-layout(matrix(c(rep(1, 4), rep(2, 4)), 2, 4), widths=1, heights=1, respect=TRUE)
-boxplot(temp3, names=names1, ylab="MSE", main="a)")
-legend("right", title=expression(bold("median MSE")), bty="n", 
-       paste(names1, round(apply(temp3, 2, median), 2), sep="="))
-boxplot(temp4, names=names1, ylab="Correlation", main="b)")
-legend("bottomleft", title=expression(bold("median correlation")), bty="n", 
-       paste(names1, round(apply(temp4, 2, median), 2), sep="="))
-par(cex=1)
-layout(matrix(1, 1, 1), widths=1, heights=1, respect=TRUE)
+<<boxplot_igaussian_res1_gamma2, cache=FALSE, echo=FALSE, fig.cap="a) MSE and b) correlation for posterior means of $\\gamma_d^2$ in simulation 1", fig.align="center">>=
 @
 
-<<"boxplot_igaussian_res1_mu", fig.cap="a) MSE and b) correlation for posterior means of $\\beta_d$", fig.align="center", echo=FALSE>>=
-res1 <- read.table("../results/simulations_igaussian_res1.csv")
-names1 <- c("inv Gauss", "ind inv Gauss", "inv Gamma")
-temp1 <- cbind(res1$mse.mu.inv.Gauss, res1$mse.mu.ind.inv.Gauss, 
-               res1$mse.mu.inv.Gamma)
-temp2 <- cbind(res1$cor.mu.inv.Gauss, res1$cor.mu.ind.inv.Gauss, 
-               res1$cor.mu.inv.Gamma)
-par(cex=1.3)
-layout(matrix(c(rep(1, 4), rep(2, 4)), 2, 4), widths=1, heights=1, respect=TRUE)
-boxplot(temp1, outline=FALSE, names=names1, ylab="MSE", main="a)")
-legend("topleft", title=expression(bold("median MSE")), bty="n", 
-       paste(names1, round(apply(temp1, 2, median), 2), sep="="))
-boxplot(temp2, names=names1, ylab="Correlation", main="b)")
-legend("bottomleft", title=expression(bold("median correlation")), bty="n", 
-       paste(names1, round(apply(temp2, 2, median), 2), 
-             sep="="))
-par(cex=1)
-layout(matrix(1, 1, 1), widths=1, heights=1, respect=TRUE)
+<<boxplot_igaussian_res1_mu, cache=FALSE, echo=FALSE, fig.cap="a) MSE and b) correlation for posterior means of $\\beta_d$ in simulation 1", fig.align="center">>=
 @
 
-<<"boxplot_igaussian_res1_theta", fig.cap="a) MSE and b) correlation for prior means of $\\gamma_d^2$ ($\\theta_d$)", fig.align="center", echo=FALSE>>=
-res1 <- read.table("../results/simulations_igaussian_res1.csv")
-set1 <- read.table("../results/simulations_igaussian_set1.csv")
-names1 <- c("inv Gauss", "ind inv Gauss", "inv Gamma")
-temp1 <- list(res1[, substr(colnames(res1), 1, 15)=="theta.inv.Gauss"],
-              res1[, substr(colnames(res1), 1, 19)=="theta.ind.inv.Gauss"],
-              res1[, substr(colnames(res1), 1, 16)=="aprior.inv.Gamma"],
-              res1[, substr(colnames(res1), 1, 16)=="bprior.inv.Gamma"])
-temp2 <- list(colMeans((t(temp1[[1]]) - 
-                          as.numeric(set1[, substr(colnames(set1), 1, 5)=="theta"]))^2),
-              colMeans((t(temp1[[2]]) - 
-                          as.numeric(set1[, substr(colnames(set1), 1, 5)=="theta"]))^2),
-              colMeans((t(temp1[[4]]/(temp1[[3]] - 1)) - 
-                          unique(as.numeric(set1[, substr(colnames(set1), 1, 5)=="theta"])))^2))
-temp3 <- list(apply(t(temp1[[1]]), 2, function(r) {
-  cor(r, as.numeric(set1[, substr(colnames(set1), 1, 5)=="theta"]))}),
-              apply(t(temp1[[2]]), 2, function(r) {
-                cor(r, as.numeric(set1[, substr(colnames(set1), 1, 5)=="theta"]))}),
-              apply(t(temp1[[4]]/(temp1[[3]] - 1)), 2, function(r) {
-                cor(r, unique(as.numeric(set1[, substr(colnames(set1), 1, 5)=="theta"])))}))
-par(cex=1.3)
-layout(matrix(c(rep(1, 4), rep(2, 4)), 2, 4), widths=1, heights=1, respect=TRUE)
-boxplot(temp2, names=names1, ylab="MSE", main="a)")
-legend("bottomright", title=expression(bold("median MSE")), bty="n",
-       paste(names1, round(sapply(temp2, median), 2), sep="="))
-boxplot(temp3, names=names1, ylab="Correlation", main="b)")
-legend("bottomleft", title=expression(bold("median correlation")), bty="n",
-       paste(names1, round(sapply(temp3, median), 2), sep="="))
-par(cex=1)
-layout(matrix(1, 1, 1), widths=1, heights=1, respect=TRUE)
+<<boxplot_igaussian_res1_theta, cache=FALSE, echo=FALSE, fig.cap="a) MSE and b) correlation for prior means of $\\gamma_d^2$ ($\\theta_d$) in simulation 1", fig.align="center">>=
 @
 
-<<"boxplot_igaussian_res1_alpha", fig.cap="a) MSE and b) correlation for $\\alpha$", fig.align="center", echo=FALSE>>=
-res1 <- read.table("../results/simulations_igaussian_res1.csv")
-set1 <- read.table("../results/simulations_igaussian_set1.csv")
-
-names1 <- c("inv Gauss", "ind inv Gauss")
-temp1 <- list(res1[, substr(colnames(res1), 1, 15)=="alpha.inv.Gauss"],
-              res1[, substr(colnames(res1), 1, 19)=="alpha.ind.inv.Gauss"])
-temp2 <- list(colMeans((t(temp1[[1]]) - as.numeric(set1[, substr(colnames(set1), 1, 5)=="alpha"]))^2),
-              colMeans((t(temp1[[2]]) - as.numeric(set1[, substr(colnames(set1), 1, 5)=="alpha"]))^2))
-temp3 <- list(apply(t(temp1[[1]]), 2, function(r) {cor(r, as.numeric(set1[, substr(colnames(set1), 1, 5)=="alpha"]))}),
-              apply(t(temp1[[2]]), 2, function(r) {cor(r, as.numeric(set1[, substr(colnames(set1), 1, 5)=="alpha"]))}))
-par(cex=1.3)
-layout(matrix(c(rep(1, 4), rep(2, 4)), 2, 4), widths=1, heights=1, respect=TRUE)
-boxplot(temp2, names=names1, ylab="MSE", main="a)")
-legend("topright", title=expression(bold("median MSE")), bty="n",
-       paste(names1, round(sapply(temp2, median), 2), sep="="))
-boxplot(temp3, names=names1, ylab="Correlation", main="b)")
-legend("bottomright", title=expression(bold("median correlation")), bty="n",
-       paste(names1, round(sapply(temp3, median), 2), sep="="))
-par(cex=1)
-layout(matrix(1, 1, 1), widths=1, heights=1, respect=TRUE)
+<<boxplot_igaussian_res1_alpha, cache=FALSE, echo=FALSE, fig.cap="a) MSE and b) correlation for $\\alpha$ in simulation 1", fig.align="center">>=
 @
 
-<<"lines_igaussian_res1_elbo", fig.cap="Convergence of ELBO for a) inverse Gaussian, b) independent inverse Gaussian, and c) inverse Gamma models", fig.align="center", echo=FALSE>>=
-fit1 <- read.table("../results/simulations_igaussian_fit1.csv")
-par(cex=1.3)
-layout(matrix(c(rep(c(1, 1, 2, 2), times=2), rep(c(0, 3, 3, 0), times=2)), 
-              4, 4, byrow=TRUE), widths=1, heights=1, respect=TRUE)
-plot(fit1[, substr(colnames(fit1), 1, 14)=="inv.Gauss.elbo"][, 1], type="l",
-     ylim=range(fit1[, substr(colnames(fit1), 1, 14)=="inv.Gauss.elbo"]),
-     main="a)", xlab="Iteration", ylab="ELBO")
-for(i in 2:100) {
-  lines(fit1[, substr(colnames(fit1), 1, 14)=="inv.Gauss.elbo"][, i], col=i)
-}
-plot(fit1[, substr(colnames(fit1), 1, 18)=="ind.inv.Gauss.elbo"][, 1], type="l",
-     ylim=range(fit1[, substr(colnames(fit1), 1, 18)=="ind.inv.Gauss.elbo"]),
-     main="b)", xlab="Iteration", ylab="ELBO")
-for(i in 2:100) {
-  lines(fit1[, substr(colnames(fit1), 1, 18)=="ind.inv.Gauss.elbo"][, i], col=i)
-}
-plot(fit1[, substr(colnames(fit1), 1, 14)=="inv.Gamma.elbo"][, 1], type="l",
-     ylim=range(fit1[, substr(colnames(fit1), 1, 14)=="inv.Gamma.elbo"]),
-     main="c)", xlab="Iteration", ylab="ELBO")
-for(i in 2:100) {
-  lines(fit1[, substr(colnames(fit1), 1, 14)=="inv.Gamma.elbo"][, i], col=i)
-}
-par(cex=1)
-layout(matrix(1, 1, 1), widths=1, heights=1, respect=TRUE)
+<<lines_igaussian_res1_elbo, cache=FALSE, echo=FALSE, fig.cap="Convergence of ELBO for a) inverse Gaussian, b) independent inverse Gaussian, and c) inverse Gamma models in simulation 1", fig.align="center">>=
 @
 
-<<"lines_igaussian_res1_theta", fig.cap="Convergence of prior means for a) inverse Gaussian, b) independent inverse Gaussian, and c) inverse Gamma models", fig.align="center", echo=FALSE>>=
-fit1 <- read.table("../results/simulations_igaussian_fit1.csv")
-par(cex=1.3)
-layout(matrix(c(rep(c(1, 1, 2, 2), times=2), rep(c(0, 3, 3, 0), times=2)), 
-              4, 4, byrow=TRUE), widths=1, heights=1, respect=TRUE)
-plot(fit1[, substr(colnames(fit1), 1, 15)=="inv.Gauss.theta"][, 1], type="l",
-     ylim=range(fit1[, substr(colnames(fit1), 1, 15)=="inv.Gauss.theta"]),
-     main="a)", xlab="Iteration", ylab=expression(theta[d]))
-for(i in 2:5) {
-  lines(fit1[, substr(colnames(fit1), 1, 15)=="inv.Gauss.theta"][
-    , (i - 1)*(set1$D/set1$nclass) + 1], col=i)
-}
-plot(fit1[, substr(colnames(fit1), 1, 19)=="ind.inv.Gauss.theta"][, 1], 
-     type="l", 
-     ylim=range(fit1[, substr(colnames(fit1), 1, 19)=="ind.inv.Gauss.theta"]),
-     main="b)", xlab="Iteration", ylab=expression(theta[d]))
-for(i in 2:5) {
-  lines(fit1[, substr(colnames(fit1), 1, 19)=="ind.inv.Gauss.theta"][
-    , (i - 1)*(set1$D/set1$nclass) + 1], col=i)
-}
-plot(sort(fit1[, substr(colnames(fit1), 1, 16)=="inv.Gamma.bprior"][, 1]/
-            (fit1[, substr(colnames(fit1), 1, 16)=="inv.Gamma.aprior"][
-              , 1] - 1)), 
-     type="l", 
-     ylim=range(fit1[, substr(colnames(fit1), 1, 16)=="inv.Gamma.bprior"]/
-                  (fit1[, substr(colnames(fit1), 1, 16)=="inv.Gamma.aprior"] - 
-                     1)),
-     main="c)", xlab="Iteration", ylab=expression(beta/(alpha - 1)))
-for(i in 2:5) {
-  lines(sort((fit1[, substr(colnames(fit1), 1, 16)=="inv.Gamma.bprior"][, ]/
-               (fit1[, substr(colnames(fit1), 1, 16)=="inv.Gamma.aprior"][
-                 , 1] - 1))[, i]), col=i)
-}
-par(cex=1)
-layout(matrix(1, 1, 1), widths=1, heights=1, respect=TRUE)
+<<lines_igaussian_res1_theta, cache=FALSE, echo=FALSE, fig.cap="Convergence of prior means for a) inverse Gaussian, b) independent inverse Gaussian, and c) inverse Gamma models in simulation 1", fig.align="center">>=
+@
+
+Simulations with no difference between classes:
+
+<<boxplot_igaussian_res2_theta, cache=FALSE, echo=FALSE, fig.cap="a) MSE and b) correlation for prior means of $\\gamma_d^2$ ($\\theta_d$) in simulation 2", out.width="50%", fig.align="center">>=
+@
+
+<<lines_igaussian_res2_alpha, cache=FALSE, echo=FALSE, fig.cap="Convergence of $\\alpha$ for a) inverse Gaussian and b) independent inverse Gaussian in simulation 2", fig.align="center">>=
 @

### 2019-01-07 docs/supplement.Rnw
changed file organisation, added Makefile

@@
 <<figures, cache=FALSE, echo=FALSE>>=
-read_chunk("../code/R/figures.R")
+read_chunk("../code/figures.R")
 @
@@
 \pgfplotsset{compat=1.13}
-\graphicspath{{/Users/magnusmunch/Documents/OneDrive/PhD/project_cambridge/graphs/}}
 \setlength{\evensidemargin}{.5cm} \setlength{\oddsidemargin}{.5cm}
@@
 
-  \section{Model without tissue effects}
-  \subsection{Introduction}
-  In order to investigate the inverse Gaussian prior model and its estimation, we consider a simpler model. The simpler model ignores tissue effects and fixes $\forall g: \phi_g^2=1$.
-
-  \subsection{Model and estimation}
+  \section{Model without tissue effects}\label{sec:notissuemodel}
+  In order to investigate the inverse Gaussian prior model and its estimation, we consider a simpler model.
+The simpler model ignores tissue effects and fixes $\forall g: \phi_g^2=1$.
 The model is given by:
@@
   \bm{\mu}_d^{(h+1)} & = (\X \tr \X + a_d^{(h)} \I)^{-1} \X \tr \y_d, \\
-  \delta_d^{(h+1)} & = \frac{n+p+1}{2 \zeta_d^{(h)}} [\trace(\bm{\Sigma}_d^{(h+1)}) + (\bm{\mu}_d^{(h+1)}) \tr \bm{\mu}_d^{(h+1)}] + \lambda, \\ \zeta_d^{(h+1)} & = \frac{a_d^{(h+1)}}{2} \[\mathbf{y}_d \tr \mathbf{y}_d -2 \mathbf{y}_d \tr \X \bm{\mu}_d^{(h+1)} + \trace [ (\X \tr \X + \I_p) \bm{\Sigma}_d^{(h+1)}] + (\bm{\mu}_d^{(h+1)}) \tr (\X \tr \X + \I_p)\bm{\mu}_d^{(h+1)}\],
+  \delta_d^{(h+1)} & = \frac{n+p+1}{2 \zeta_d^{(h)}} [\trace(\bm{\Sigma}_d^{(h+1)}) + (\bm{\mu}_d^{(h+1)}) \tr \bm{\mu}_d^{(h+1)}] + \lambda, \\ \zeta_d^{(h+1)} & = \frac{1}{2} \bigg[\mathbf{y}_d \tr \mathbf{y}_d -2 \mathbf{y}_d \tr \X \bm{\mu}_d^{(h+1)} + \trace ( \X \tr \X \bm{\Sigma}_d^{(h+1)}) + (\bm{\mu}_d^{(h+1)}) \tr \X \tr \X \bm{\mu}_d^{(h+1)} + \\
+  & \,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\, a_d^{(h+1)} \trace ( \bm{\Sigma}_d^{(h+1)}) + a_d^{(h+1)} (\bm{\mu}_d^{(h+1)}) \tr \bm{\mu}_d^{(h+1)}\bigg],
   \end{align*}
@@
 
-  \subsection{Evidence lower bound}
   We monitor convergence through the evidence lower bound (ELBO):
@@
 
-  \subsection{Efficient computation}
   The empirical Bayes updates require the following quantities: $\zeta_d$, $a_d$, $\delta_d$, $\trace (\bm{\Sigma}_d)$, $\trace (\X \tr \X \bm{\Sigma}_d)$, $\bm{\mu}_d \tr \bm{\mu}_d$, and $\bm{\mu}_d \tr \X \tr \X \bm{\mu}_d$.
@@
 
-  \subsection{Independent error variance model parameters}
+  \section{Independent error variance model}\label{sec:indvarmodel}
   \cite{moran_variance_2018} argue that in regression models, $\bm{\beta}_d$ and $\sigma_d^2$ should be independent \textit{a priori}.
@@
   \end{align*}
+
   Its variational posterior is given by:
@@
   with $a_d^{(h)}$ and the empirical Bayes updates $\bm{\alpha}^{(l+1)}$ and $\lambda^{(l+1)}$ as in the full model in the MD.
+
   A singular value decomposition $\X = \mathbf{U} \mathbf{D} \mathbf{V} \tr$ leads to the following efficient computations:
@@
 
-  \subsection{Simulations}
-  We simulate from a correctly specified model.
+  \section{Inverse Gamma model}\label{sec:invgammamodel}
+  An alternative model is one where the $\gamma_d^2$ are drawn from a class-specific inverse Gamma distribution; a generalisation of the models in \cite{kpogbezan_empirical_2017} and \cite{leday_gene_2017}:
+  \begin{align*}
+  y_{id} | \bm{\beta}_d ,\sigma_d^2 & \sim \mathcal{N} (\x_i \tr \bm{\beta}_d, \sigma_d^2), \\
+  \beta_{jd} | \gamma_d^2 ,\sigma_d^2 & \sim \mathcal{N} (0, \sigma_d^2 \gamma_d^2), \\
+  \gamma_d^2 & \sim \Gamma^{-1} (a_{\text{class}(d)}, b_{\text{class}(d)}), \\
+  \sigma_d^2 & \sim \Gamma^{-1} (0.001, 0.001),
+  \end{align*}
+  where $\text{class}(c)$ denotes the class of drug $d$.
+This allows for more flexible empirical Bayes estimation of the $a_d$ and $b_d$, and thus prior means of the different drug classes.
+As a consequence however, the risk of overfitting is increased.
+Additionally, it does not allow to include continuous covariates on the drugs.
+
+  Its variational posterior is given by:
+  \begin{align*}
+  q(\bm{\beta}) & \overset{D}{=} \prod_{d=1}^D \mathcal{N}_p (\bm{\mu}_d, \bm{\Sigma}_d), \\
+  q(\bm{\gamma}^2) & \overset{D}{=} \prod_{d=1}^D \Gamma^{-1} \( \frac{p}{2} + a_{\text{class}(d)}, b^*_{\text{class}(d)}\), \\
+  q(\bm{\sigma}^2) & \overset{D}{=} \prod_{d=1}^D \Gamma^{-1} \( \frac{n+p}{2} + 0.001, d^*_{\text{class}(d)} \),
+  \end{align*}
+  with variational parameter updates:
+  \begin{align*}
+  \bm{\Sigma}_d^{(h+1)} & = \frac{n + p - 1.998}{2(d^*_{\text{class}(d)})^{(h)}}\(\X \tr \X + \frac{p + 2a_{\text{class}(d)} - 2}{2(b^*_{\text{class}(d)})^{(h)}} \I_p\)^{-1}, \\
+  \bm{\mu}_d^{(h+1)} & = \(\X \tr \X + \frac{p + 2a_{\text{class}(d)} - 2}{2(b^*_{\text{class}(d)})^{(h)}} \I_p\)^{-1} \X \tr \y_d, \\
+  b^*_{\text{class}(d)} & = b_{\text{class}(d)} + \frac{(d^*_{\text{class}(d)})^{(h)}}{n + p -1.998} [\trace(\bm{\Sigma}_d^{(h+1)}) + (\bm{\mu}_d^{(h+1)}) \tr \bm{\mu}_d^{(h+1)}],\\
+  d^*_{\text{class}(d)} & = \frac{1}{2} \[\mathbf{y}_d \tr \mathbf{y}_d -2 \mathbf{y}_d \tr \X \bm{\mu}_d^{(h+1)} + \trace (\X \tr \X \bm{\Sigma}_d^{(h+1)}) + (\bm{\mu}_d^{(h+1)}) \tr \X \tr \X \bm{\mu}_d^{(h+1)}\].
+  \end{align*}
+
+  \section{Simulations}
+  We simulate from a correctly specified model given in \ref{sec:notissuemodel}.
 We assume random $y_{id}$, $\x_i$, and $\bm{\beta}_d$, while we fix $\mathbf{C}$, $\gamma_d^2$, $\sigma_d^2$ and $\bm{\alpha}$.
@@
 To account for the random data and parameter generation, we repeat the simulation 100 times.
-
-  We compare the results to a model where the $\gamma_d^2$ are drawn from a class-specific inverse Gamma distribution; a generalisation of the models in \cite{kpogbezan_empirical_2017} and \cite{leday_gene_2017}:
-  \begin{align*}
-  y_{id} | \bm{\beta}_d ,\sigma_d^2 & \sim \mathcal{N} (\x_i \tr \bm{\beta}_d, \sigma_d^2), \\
-  \beta_{jd} | \gamma_d^2 ,\sigma_d^2 & \sim \mathcal{N} (0, \sigma_d^2 \gamma_d^2), \\
-  \gamma_d^2 & \sim \Gamma^{-1} (a_{\text{class}(d)}, b_{\text{class}(d)}), \\
-  \sigma_d^2 & \sim \Gamma^{-1} (0.001, 0.001),
-  \end{align*}
-  where $\text{class}(c)$ denotes the class of drug $d$.
-This allows for more flexible empirical Bayes estimation of the $a_d$ and $b_d$, and thus prior means of the different drug classes.
-As a consequence however, the risk of overfitting is increased.
-Additionally, it does not allow to include continuous covariates on the drugs.
-We present the results in Figures \ref{fig:boxplot_igaussian_res1_sigma2}-\ref{fig:boxplot_igaussian_res1_alpha}.
+We compare the results to the models in \ref{sec:indvarmodel} and \ref{sec:invgammamodel}, and present the results in Figures \ref{fig:boxplot_igaussian_res1_sigma2}-\ref{fig:boxplot_igaussian_res1_alpha}.
 

### 2019-01-07 rpackage/R/functions.R

@@ -14,7 +14,7 @@ single.vb.update <- function(zetaold, aold, lambda, theta, sv, n, p, uty, yty) {
   delta <- 0.5*(n + p + 1)*(trSigma + mutmu)/zetaold + lambda
   ratio <- ratio_besselK_cpp(sqrt(lambda*delta/theta^2), p)
   a <- sqrt(lambda/(theta^2*delta))*ratio + (p + 1)/delta
-  zeta <- 0.5*a*(yty + trSigma + mutmu + trXtXSigma + mutXtXmu - 2*ytXmu)
+  zeta <- 0.5*(yty + a*trSigma + a*mutmu + trXtXSigma + mutXtXmu - 2*ytXmu)
   v <- sqrt(delta*theta^2/lambda)*ratio
   
   # calculate the elbo part that is constant after next eb update

### 2019-01-07 rpackage/tests/testthat/test-igaussian.R

@@ -13,9 +13,10 @@ test_that("dependent beta and sigma^2 caluclations", {
   delta <- 0.5*(n + p + 1)*(sum(diag(Sigma)) + sum(mu^2))/zetaold + lambda
   a <- sqrt(lambda/(theta^2*delta))*
     ratio_besselK_cpp(sqrt(lambda*delta)/theta, p) + (p + 1)/delta
-  zeta <- as.numeric(0.5*a*(sum(y^2) - 2*t(y) %*% x %*% mu +
-                              sum(diag((t(x) %*% x + diag(p)) %*% Sigma)) +
-                              t(mu) %*% (t(x) %*% x + diag(p)) %*% mu))
+  zeta <- as.numeric(0.5*(sum(y^2) - 2*t(y) %*% x %*% mu +
+                            sum(diag(t(x) %*% x %*% Sigma)) +
+                            t(mu) %*% t(x) %*% x  %*% mu + 
+                            a*sum(diag(Sigma)) + a*t(mu) %*% mu))
   v <- sqrt(delta*theta^2/lambda)*
     ratio_besselK_cpp(sqrt(lambda*delta)/theta, p)
   

### 2019-01-08 docs/supplement.Rnw
rewrote simulation models in supplement.Rnw

@@
 
-% R packages
-<<echo=FALSE>>=
-library(cambridge)
-@
-
-<<figures, cache=FALSE, echo=FALSE>>=
-read_chunk("../code/figures.R")
-@
-
 % settings
@@
 \begin{document}
+
+% load packages and set options
+<<include=FALSE>>=
+library(cambridge)
+opts_knit$set(base.dir="../figs")
+opts_chunk$set(fig.align='center', echo=FALSE, fig.path="../figs/")
+@
+
+% read figure code
+<<figures, include=FALSE>>=
+read_chunk("../code/figures.R")
+@
+
 	\maketitle
@@
 
+	\section{Simulations}
+	\subsection{Models}
+	In order to investigate the inverse Gaussian prior model and its estimation, we consider a simpler model.
+The simpler model ignores tissue effects and fixes $\forall g: \phi_g^2=1$.
+The model is given by:
+	\begin{align*}
+  y_{id} | \bm{\beta}_d, \sigma_d^2 & \sim \mathcal{N} (\x_i \tr \bm{\beta}_d, \sigma_d^2), \\
+  \beta_{jd} | \gamma_d^2, \sigma_d^2 & \sim p(\beta_{jd}), \\
+  \gamma_d^2 & \sim p(\gamma_d^2), \\
+  \sigma_d^2 & \sim 1/\sigma_d^3.
+  \end{align*}
+  We use a conjugate and non-conjugate model for $\beta_{jd}$:
+	\begin{align*}
+  \beta_{jd} | \gamma_d^2, \sigma_d^2 & \sim \mathcal{N} (0, \sigma_d^2 \gamma_d^2), \\
+  \beta_{jd} | \gamma_d^2, \sigma_d^2 \sim \beta_{jd} | \gamma_d^2 & \sim \mathcal{N} (0, \gamma_d^2),
+  \end{align*}
+	respectively. For the prior variances we use an inverse Gaussian and inverse Gamma model
+	\begin{align*}
+  \gamma_d^2 \sim \mathcal{IG}(\theta_d, \lambda_d) = \mathcal{GIG}(-1/2, \lambda_d \theta_d^{-2}, \lambda_d), \\
+  \gamma_d^2 \sim \Gamma^{-1}(\alpha_d/2, \lambda_d/2) = \mathcal{GIG}(-\alpha_d/2, 0, \lambda_d),
+  \end{align*}
+	respectively. We have rescaled the inverse Gamma parameters to make them more comparable in the following. The two distributions coincide if $\theta_d \to \infty$ and $\alpha_d = 1$.
+
+	\subsection{Estimation}
+	The variational distributions are:
+	\begin{align*}
+  q(\bm{\beta}) & \overset{D}{=} \prod_{d=1}^D \mathcal{N}_p (\bm{\mu}_d, \bm{\Sigma}_d), \\
+  q(\bm{\gamma}^2) & \overset{D}{=} \prod_{d=1}^D \mathcal{GIG}\(-\frac{p+\alpha_d}{2}, \lambda_d \theta_d^{-2}, \delta_d\), \\
+  q(\bm{\sigma}^2) & \overset{D}{=} \prod_{d=1}^D \Gamma^{-1} \(\frac{n + p + 1}{2}, \zeta_d\),
+  \end{align*}
+	where $\alpha_d=1$ if we use the inverse Gaussian prior and $\theta_d \to \infty$ if we use the inverse Gamma prior.
+The parameters contain cyclic dependencies and are, in the conjugate setting updated by:
+	\begin{align*}
+  \bm{\Sigma}_d^{(h+1)} & = (a_d^{(h)})^{-1} (\X \tr \X + b_d^{(h)} \I)^{-1}, \\
+  \bm{\mu}_d^{(h+1)} & = (\X \tr \X + b_d^{(h)} \I)^{-1} \X \tr \y_d, \\
+  \delta_d^{(h+1)} & = a_d^{(h)} \bmu_d \tr \bmu_d + \lambda_d, \\ 
+  \zeta_d^{(h+1)} & = \frac{1}{2} \bigg[\mathbf{y}_d \tr \mathbf{y}_d -2 \mathbf{y}_d \tr \X \bm{\mu}_d^{(h+1)} + \trace ( \X \tr \X \bm{\Sigma}_d^{(h+1)}) + (\bm{\mu}_d^{(h+1)}) \tr \X \tr \X \bm{\mu}_d^{(h+1)} \\
+  & \,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\, + b_d^{(h+1)} \trace ( \bm{\Sigma}_d^{(h+1)}) + b_d^{(h+1)} (\bm{\mu}_d^{(h+1)}) \tr \bm{\mu}_d^{(h+1)}\bigg].
+  \end{align*}
+	Likewise in the non-conjugate setting:
+	\begin{align*}
+  \bm{\Sigma}_d^{(h+1)} & = (a_d^{(h)})^{-1} (\X \tr \X + b_d^{(h)} (a_d^{(h)})^{-1} \I)^{-1}, \\
+  \bm{\mu}_d^{(h+1)} & = (\X \tr \X + b_d^{(h)} (a_d^{(h)})^{-1} \I)^{-1} \X \tr \y_d, \\
+  \delta_d^{(h+1)} & = \bmu_d \tr \bmu_d + \lambda_d, \\ 
+  \zeta_d^{(h+1)} & = \frac{1}{2} \[\mathbf{y}_d \tr \mathbf{y}_d -2 \mathbf{y}_d \tr \X \bm{\mu}_d^{(h+1)} + \trace ( \X \tr \X \bm{\Sigma}_d^{(h+1)}) + (\bm{\mu}_d^{(h+1)}) \tr \X \tr \X \bm{\mu}_d^{(h+1)} \].
+  \end{align*}
+  We have $a_d^{(h)}=\E_{Q^{(h)}}(\sigma_d^{-2})=(n+p+1)/(2 \zeta_d^{(h)})$ and
+  $$
+  b_d^{(h)} = \E_{Q^{(h)}}(\gamma_d^{-2}) = \sqrt{\frac{\lambda}{\theta_d^2 \delta_d^{(h)}}} \frac{K_{\frac{p + \alpha_d -2}{2}} \( \sqrt{\lambda \delta_d^{(h)}}/\theta_d \)}{K_{\frac{p+\alpha_d}{2}} \( \sqrt{\lambda \delta_d^{(h)}}/\theta_d \)} + \frac{p+\alpha_d^{(h)}}{\delta_d^{(h)}}.\\
+  $$
+  This simplifies to $b_d^{(h)}=(p+\alpha_d)/\delta_d^{(h)}$ in the inverse Gamma model.
+
+  \subsection{Efficient computation}
+  Let the SVD of $\X$ be $\X = \mathbf{U} \mathbf{D} \mathbf{V} \tr$, $v_i$ be the singular values, and $c_d^{(h)}=b_d^{(h)}$ in the conjugate setting and $c_d^{(h)}=b_d^{(h)}/a_d^{(h)}$ in the non-conjugate setting.
+Then:
+  \begin{align*}
+  \trace (\bm{\Sigma}_d^{(h+1)}) & = (a_d^{(h)})^{-1} \[\sum_{i=1}^n (v_i^2 + c_d^{(h)})^{-1} + \max (p - n, 0)\cdot (c_d^{(h)})^{-1}\], \\
+  \trace (\X \tr \X \bm{\Sigma}_d^{(h+1)}) & = (a_d^{(h)})^{-1} \sum_{i=1}^n v_i^2(v_i^2 + c_d^{(h)})^{-1}, \\
+  (\bm{\mu}_d^{(h+1)}) \tr \bm{\mu}_d^{(h+1)} & = \sum_{i=1}^n v_i^2 [(\mathbf{U} \tr \mathbf{y}_d)_i]^2/(v_i^2 + c_d^{(h)})^2, \\
+  (\bm{\mu}_d^{(h+1)}) \tr \X \tr \X \bm{\mu}_d^{(h+1)} & = \sum_{i=1}^n v_i^4 [(\mathbf{U} \tr \mathbf{y}_d)_i]^2/(v_i^2 + c_d^{(h)})^2, \\
+  \log|\bm{\Sigma}^{(h+1)}| & \propto -p \log a_d^{(h)} - \sum_{i=1} \log (v_i^2 + c_d^{(h)}) - \max(p - n, 0) \cdot \log c_d^{(h)}, \\
+  \mathbf{y}_d \tr \X \bm{\mu}_d^{(h+1)} & = \sum_{i=1}^n v_i^2 [(\mathbf{U} \tr \mathbf{y}_d)_i]^2/(v_i^2 + c_d^{(h)}).
+  \end{align*}
+  The SVD and $\mathbf{U} \tr \mathbf{y}_d$ are $\mathcal{O}(pn^2)$ and $\mathcal{O}(n^2)$ operations, respectively, but have to be calculated only once at the start of the algorithm.
+The rest of the calculations involve just $n$ scalar multiplications and/or additions.
+
   \section{Model without tissue effects}\label{sec:notissuemodel}
@@
   \bm{\mu}_d^{(h+1)} & = (\X \tr \X + a_d^{(h)} \I)^{-1} \X \tr \y_d, \\
-  \delta_d^{(h+1)} & = \frac{n+p+1}{2 \zeta_d^{(h)}} [\trace(\bm{\Sigma}_d^{(h+1)}) + (\bm{\mu}_d^{(h+1)}) \tr \bm{\mu}_d^{(h+1)}] + \lambda, \\ \zeta_d^{(h+1)} & = \frac{1}{2} \bigg[\mathbf{y}_d \tr \mathbf{y}_d -2 \mathbf{y}_d \tr \X \bm{\mu}_d^{(h+1)} + \trace ( \X \tr \X \bm{\Sigma}_d^{(h+1)}) + (\bm{\mu}_d^{(h+1)}) \tr \X \tr \X \bm{\mu}_d^{(h+1)} + \\
-  & \,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\, a_d^{(h+1)} \trace ( \bm{\Sigma}_d^{(h+1)}) + a_d^{(h+1)} (\bm{\mu}_d^{(h+1)}) \tr \bm{\mu}_d^{(h+1)}\bigg],
+  \delta_d^{(h+1)} & = \frac{n+p+1}{2 \zeta_d^{(h)}} [\trace(\bm{\Sigma}_d^{(h+1)}) + (\bm{\mu}_d^{(h+1)}) \tr \bm{\mu}_d^{(h+1)}] + \lambda, \\ 
+  \zeta_d^{(h+1)} & = \frac{1}{2} \bigg[\mathbf{y}_d \tr \mathbf{y}_d -2 \mathbf{y}_d \tr \X \bm{\mu}_d^{(h+1)} + \trace ( \X \tr \X \bm{\Sigma}_d^{(h+1)}) + (\bm{\mu}_d^{(h+1)}) \tr \X \tr \X \bm{\mu}_d^{(h+1)} \\
+  & \,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\, + a_d^{(h+1)} \trace ( \bm{\Sigma}_d^{(h+1)}) + a_d^{(h+1)} (\bm{\mu}_d^{(h+1)}) \tr \bm{\mu}_d^{(h+1)}\bigg],
   \end{align*}
@@
   q(\bm{\beta}) & \overset{D}{=} \prod_{d=1}^D \mathcal{N}_p (\bm{\mu}_d, \bm{\Sigma}_d), \\
-  q(\bm{\gamma}^2) & \overset{D}{=} \prod_{d=1}^D \Gamma^{-1} \( \frac{p}{2} + a_{\text{class}(d)}, b^*_{\text{class}(d)}\), \\
-  q(\bm{\sigma}^2) & \overset{D}{=} \prod_{d=1}^D \Gamma^{-1} \( \frac{n+p}{2} + 0.001, d^*_{\text{class}(d)} \),
+  q(\bm{\gamma}^2) & \overset{D}{=} \prod_{d=1}^D \Gamma^{-1} \( \frac{p}{2} + a_{\text{class}(d)}, b^*_d\), \\
+  q(\bm{\sigma}^2) & \overset{D}{=} \prod_{d=1}^D \Gamma^{-1} \( \frac{n+p}{2} + 0.001, d^*_d \),
   \end{align*}
@@
   \begin{align*}
-  \bm{\Sigma}_d^{(h+1)} & = \frac{n + p - 1.998}{2(d^*_{\text{class}(d)})^{(h)}}\(\X \tr \X + \frac{p + 2a_{\text{class}(d)} - 2}{2(b^*_{\text{class}(d)})^{(h)}} \I_p\)^{-1}, \\
-  \bm{\mu}_d^{(h+1)} & = \(\X \tr \X + \frac{p + 2a_{\text{class}(d)} - 2}{2(b^*_{\text{class}(d)})^{(h)}} \I_p\)^{-1} \X \tr \y_d, \\
-  b^*_{\text{class}(d)} & = b_{\text{class}(d)} + \frac{(d^*_{\text{class}(d)})^{(h)}}{n + p -1.998} [\trace(\bm{\Sigma}_d^{(h+1)}) + (\bm{\mu}_d^{(h+1)}) \tr \bm{\mu}_d^{(h+1)}],\\
-  d^*_{\text{class}(d)} & = \frac{1}{2} \[\mathbf{y}_d \tr \mathbf{y}_d -2 \mathbf{y}_d \tr \X \bm{\mu}_d^{(h+1)} + \trace (\X \tr \X \bm{\Sigma}_d^{(h+1)}) + (\bm{\mu}_d^{(h+1)}) \tr \X \tr \X \bm{\mu}_d^{(h+1)}\].
+  \bm{\Sigma}_d^{(h+1)} & = (c_d^{(h)})^{-1} \(\X \tr \X + (e_d^{(h)})^{-1} \I_p\)^{-1}, \\
+  \bm{\mu}_d^{(h+1)} & = \(\X \tr \X + (e_d^{(h)})^{-1} \I_p\)^{-1} \X \tr \y_d, \\
+  (b^*_d)^{(h+1)} & = b_{\text{class}(d)} + \frac{c_d^{(h)}}{2} [\trace(\bm{\Sigma}_d^{(h+1)}) + (\bm{\mu}_d^{(h+1)}) \tr \bm{\mu}_d^{(h+1)}],\\
+  (d^*_d)^{(h+1)} & = \frac{1}{2} \[\mathbf{y}_d \tr \mathbf{y}_d -2 \mathbf{y}_d \tr \X \bm{\mu}_d^{(h+1)} + \trace (\X \tr \X \bm{\Sigma}_d^{(h+1)}) + (\bm{\mu}_d^{(h+1)}) \tr \X \tr \X \bm{\mu}_d^{(h+1)}\] + \\
+  & \frac{e_d^{(h)}}{2} \trace ( \bm{\Sigma}_d^{(h+1)}) + \frac{e_d^{(h)}}{2} (\bm{\mu}_d^{(h+1)}) \tr \bm{\mu}_d^{(h+1)},
+  \end{align*}
+  where
+  $$
+  c_d^{(h)} = \frac{2(d_d^*)^{(h)}}{n + p -1.998} \text{ and } e_d^{(h)} = \frac{2(b_d^*)^{(h)}}{p + 2a_{\text{class}(d)} - 2}.
+  $$
+
+  \begin{align*}
+  \text{ELBO}_d & (Q^{(l)}) \propto -\(\frac{n+p}{2} + 0.001 \) \log \[ (d^*_d)^{(l)}\] + \frac{1}{2} \log |\bm{\Sigma}_d^{(l)}| - \[ (\bm{\mu}_d^{(l)}) \tr \bm{\mu}_d^{(l)} + \trace (\bm{\Sigma}_d^{(l)}) \] \frac{c_d^{(l)} e_d^{(l)}}{2} \\
+  & - \[ \bm{y}_d \tr \bm{y}_d - \bm{y}_d \tr \X \bm{\mu}_d^{(l)} + (\bm{\mu}_d^{(l)}) \tr \X \tr \X \bm{\mu}_d^{(l)} + \trace (\X \tr \X \bm{\Sigma}_d^{(l)})\] \frac{c_d^{(l)}}{2} \\
+  & + a_{\text{class}(d)}^{(l)} + \log \[\frac{\Gamma (a_{\text{class}(d)}^{(l)} + p/2)}{\Gamma (a_{\text{class}(d)}^{(l)})}\] - \( a_{\text{class}(d)}^{(l)} +\frac{p}{2} \) \log \[(b_d^*)^{(l)}\]  \\
+  & + a_{\text{class}(d)}^{(l)} \log b_{\text{class}(d)}^{(l)} - b_{\text{class}(d)}^{(l)} \frac{a_{\text{class}(d)}^{(l)} + p/2}{(b_d^*)^{(h)}}.
   \end{align*}

### 2019-01-10 docs/supplement.Rnw
simulation models now same framework, partly implemented in R

--- 2019-01-05 docs/supplement.Rnw
@@
 
-% R packages
-<<echo=FALSE>>=
-library(cambridge)
-@
-
-<<figures, cache=FALSE, echo=FALSE>>=
-read_chunk("../code/R/figures.R")
-@
-
 % settings
 \pgfplotsset{compat=1.13}
-\graphicspath{{/Users/magnusmunch/Documents/OneDrive/PhD/project_cambridge/graphs/}}
 \setlength{\evensidemargin}{.5cm} \setlength{\oddsidemargin}{.5cm}
@@
 
-	\section{Introduction}
+	\section{Content overview}
 	This document contains the Supplementary Material (SM) to the document `Drug efficacy prediction in cell lines'. In the following, this document is referred to as Main Document (MD).
@@
 
-  \section{Model without tissue effects}
-  \subsection{Introduction}
-  In order to investigate the inverse Gaussian prior model and its estimation, we consider a simpler model. The simpler model ignores tissue effects and fixes $\forall g: \phi_g^2=1$.
-
-  \subsection{Model and estimation}
-  The model is given by:
+	\section{Simulations}
+	\subsection{Models}
+	In order to investigate the inverse Gaussian prior model and its estimation, we consider a simpler model. The simpler model ignores tissue effects and fixes $\forall g: \phi_g^2=1$:
 	\begin{align*}
   y_{id} | \bm{\beta}_d, \sigma_d^2 & \sim \mathcal{N} (\x_i \tr \bm{\beta}_d, \sigma_d^2), \\
-  \beta_{jd} | \gamma_d^2, \sigma_d^2 & \sim \mathcal{N} (0, \sigma_d^2 \gamma_d^2), \\
-  \gamma_d^2 & \sim \mathcal{IG}(\theta_d, \lambda), \\
+  \beta_{jd} | \gamma_d^2, \sigma_d^2 & \sim p(\beta_{jd}), \\
+  \gamma_d^2 & \sim p(\gamma_d^2), \\
   \sigma_d^2 & \sim 1/\sigma_d^3.
   \end{align*}
-  The corresponding variational Bayes posterior is given by:
+
+  \cite{moran_variance_2018} argue that in regression models, $\bm{\beta}_d$ and $\sigma_d^2$ should be independent \textit{a priori}.
+However, in our case that is debatable.
+The $D$ regressions need not be on the same scale, so some form of calibration of the $\bm{\beta}_d$ posteriors to this scale is required.
+A simple way of doing this is to include the scales $\sigma_d^2$ in the priors for the $\bm{\beta}_d$.
+We call the two versions of the model conjugate (with $\bm{\beta}_d$ and $\sigma_d^2$ dependent):
+  $$
+  \beta_{jd} | \gamma_d^2, \sigma_d^2 \sim \mathcal{N} (0, \sigma_d^2 \gamma_d^2)
+  $$
+  and non-conjugate (with $\bm{\beta}_d$ and $\sigma_d^2$ independent):
+  $$
+  \beta_{jd} | \gamma_d^2, \sigma_d^2 \sim \beta_{jd} | \gamma_d^2 \sim \mathcal{N} (0, \gamma_d^2),
+  $$
+  and compare them in our simulations.
+
+	In addition to the inverse Gaussian model for the prior variances $\gamma_d^2$ (introduced in the MD):
+	$$
+  \gamma_d^2 \sim \mathcal{IG}(\theta_d, \lambda_d),
+  $$
+	we investigate a generalisation of the models in \cite{kpogbezan_empirical_2017} and \cite{leday_gene_2017}, where we draw the $\gamma_d^2$ from an inverse Gamma distribution; :
+	$$
+	\gamma_d^2 \sim \Gamma^{-1}(\alpha_d/2, \lambda_d/2).
+	$$
+	For the sake of comparability, we have rescaled the inverse Gamma parameters and reparametrised both distributions as generalized inverse Gaussians:
+	\begin{align*}
+  \gamma_d^2 & \sim \mathcal{GIG}(-1/2, \lambda_d \theta_d^{-2}, \lambda_d), \\
+  \gamma_d^2 & \sim \mathcal{GIG}(-\alpha_d/2, 0, \lambda_d).
+  \end{align*}
+	The two models coincide if $\theta_d \to \infty$ and $\alpha_d = 1$.
+
+	\subsection{Estimation}
+	The variational distributions are as follows:
 	\begin{align*}
   q(\bm{\beta}) & \overset{D}{=} \prod_{d=1}^D \mathcal{N}_p (\bm{\mu}_d, \bm{\Sigma}_d), \\
-  q(\bm{\gamma}^2) & \overset{D}{=} \prod_{d=1}^D \mathcal{GIG}\(-\frac{p+1}{2}, \frac{\lambda}{\theta_d^2}, \delta_d\), \\
-  q(\bm{\sigma}^2) & \overset{D}{=} \prod_{d=1}^D \Gamma^{-1} \(\frac{n + p + 1}{2}, \zeta_d\).
+  q(\bm{\gamma}^2) & \overset{D}{=} \prod_{d=1}^D \mathcal{GIG}\(-\frac{p+\alpha_d}{2}, \lambda_d \theta_d^{-2}, \delta_d\), \\
+  q(\bm{\sigma}^2) & \overset{D}{=} \prod_{d=1}^D \Gamma^{-1} \(\frac{df + 1}{2}, \zeta_d\),
   \end{align*}
-  We update the variational parameters until convergence:
+	where $\alpha_d=1$ and $\theta_d \to \infty$ in the inverse Gaussian and inverse Gamma models, respectively.
+Furthermore, $df=n + p$ in the conjugate model and $df=n$ in the non-conjugate model.
+The variational parameters contain cyclic dependencies and are, in the conjugate setting, iteratively updated by:
 	\begin{align*}
-  \bm{\Sigma}_d^{(h+1)} & = \frac{2 \zeta_d^{(h)}}{n+p+1} (\X \tr \X + a_d^{(h)} \I)^{-1}, \\
-  \bm{\mu}_d^{(h+1)} & = (\X \tr \X + a_d^{(h)} \I)^{-1} \X \tr \y_d, \\
-  \delta_d^{(h+1)} & = \frac{n+p+1}{2 \zeta_d^{(h)}} [\trace(\bm{\Sigma}_d^{(h+1)}) + (\bm{\mu}_d^{(h+1)}) \tr \bm{\mu}_d^{(h+1)}] + \lambda, \\ \zeta_d^{(h+1)} & = \frac{a_d^{(h+1)}}{2} \[\mathbf{y}_d \tr \mathbf{y}_d -2 \mathbf{y}_d \tr \X \bm{\mu}_d^{(h+1)} + \trace [ (\X \tr \X + \I_p) \bm{\Sigma}_d^{(h+1)}] + (\bm{\mu}_d^{(h+1)}) \tr (\X \tr \X + \I_p)\bm{\mu}_d^{(h+1)}\],
+  \bm{\Sigma}_d^{(h+1)} & = (a_d^{(h)})^{-1} (\X \tr \X + b_d^{(h)} \I)^{-1}, \\
+  \bm{\mu}_d^{(h+1)} & = (\X \tr \X + b_d^{(h)} \I)^{-1} \X \tr \y_d, \\
+  \delta_d^{(h+1)} & = a_d^{(h)} \bmu_d \tr \bmu_d + \lambda_d, \\ 
+  \zeta_d^{(h+1)} & = \frac{1}{2} \bigg[\mathbf{y}_d \tr \mathbf{y}_d -2 \mathbf{y}_d \tr \X \bm{\mu}_d^{(h+1)} + \trace ( \X \tr \X \bm{\Sigma}_d^{(h+1)}) + (\bm{\mu}_d^{(h+1)}) \tr \X \tr \X \bm{\mu}_d^{(h+1)} \\
+  & \,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\, + b_d^{(h+1)} \trace ( \bm{\Sigma}_d^{(h+1)}) + b_d^{(h+1)} (\bm{\mu}_d^{(h+1)}) \tr \bm{\mu}_d^{(h+1)}\bigg].
   \end{align*}
-  with $a_d^{(h)}$ as in the full model in the MD. Equivalently, the empirical Bayes updates for $\bm{\alpha}$ and $\lambda$ are the same as in the full model in the MD.
+	while in the non-conjugate setting, we use the following updating equations:
+	\begin{align*}
+  \bm{\Sigma}_d^{(h+1)} & = (a_d^{(h)})^{-1} (\X \tr \X + b_d^{(h)} (a_d^{(h)})^{-1} \I)^{-1}, \\
+  \bm{\mu}_d^{(h+1)} & = (\X \tr \X + b_d^{(h)} (a_d^{(h)})^{-1} \I)^{-1} \X \tr \y_d, \\
+  \delta_d^{(h+1)} & = \bmu_d \tr \bmu_d + \lambda_d, \\ 
+  \zeta_d^{(h+1)} & = \frac{1}{2} \[\mathbf{y}_d \tr \mathbf{y}_d -2 \mathbf{y}_d \tr \X \bm{\mu}_d^{(h+1)} + \trace ( \X \tr \X \bm{\Sigma}_d^{(h+1)}) + (\bm{\mu}_d^{(h+1)}) \tr \X \tr \X \bm{\mu}_d^{(h+1)} \].
+  \end{align*}
+  Here, we set
+  \begin{align}
+  a_d^{(h)} & =\E_{Q^{(h)}}(\sigma_d^{-2})=(df+1)/(2 \zeta_d^{(h)}), \nonumber \\
+  b_d^{(h)} & = \E_{Q^{(h)}}(\gamma_d^{-2}) = \sqrt{\frac{\lambda_d}{\theta_d^2 \delta_d^{(h)}}} \frac{K_{\frac{p + \alpha_d -2}{2}} \( \sqrt{\lambda_d \delta_d^{(h)}}/\theta_d \)}{K_{\frac{p+\alpha_d}{2}} \( \sqrt{\lambda_d \delta_d^{(h)}}/\theta_d \)} + \frac{p+\alpha_d^{(h)}}{\delta_d^{(h)}}.
+\label{eq:auxvar1}
+  \end{align}
+  Note that in the inverse Gamma model, the computation of (\ref{eq:auxvar1}) simplifies to $b_d^{(h)}=(p+\alpha_d)/\delta_d^{(h)}$.
+
+  \subsection{Empirical Bayes}
+  To avoid subjectivity in our choice of prior, we estimate the hyper-parameters with empirical Bayes.
+The empirical Bayes procedure differs for the two prior models.
+In the inverse Gaussian case, we parametrise $\theta_d = (\mathbf{c}_d \tr \bm{\alpha})^{-1}$ to allow for the inclusion of drug covariates in the empirical Bayes estimation.
+The estimating equations are:
+  \begin{align*}
+  \bm{\alpha} & = \[ \mathbf{C} \tr \diag (\lambda_d e_d^{(l)}) \mathbf{C} \]^{-1} \diag (\lambda_d) \mathbf{1}_{D \times 1}, \\
+  \lambda_d & = \[ b_d^{(l)} + e_d^{(l)} (\mathbf{c}_d \tr \bm{\alpha})^2 - 2\mathbf{c}_d \tr \bm{\alpha} \]^{-1},
+  \end{align*}
+  where $e_d^{(l)} = (b_d^{(l)} + \delta_d^{(l)}/p) \cdot \delta_d^{(l)} (\theta_d^{(h)})^2/\lambda_d^{(l)}$.
+Solving the equations is done by iteratively reweighted least squares of responses $(e_d^{(l)})^{-1}$ on predictors $\mathbf{c}_d$ with weights $\lambda_d e_d^{(l)}$.
+If we use one common $\lambda=\lambda_d$ then the solution is in closed form:
+  \begin{align*}
+	\bm{\alpha}^{(l+1)} & = \[ \mathbf{C} \tr \diag(e_d^{(l)}) \mathbf{C} \]^{-1} \mathbf{C} \tr \mathbf{1}_{D \times 1}, \\
+	\lambda^{(l+1)} & = \[ \sum_{d=1}^D b_d^{(l)} - (\bm{\alpha}^{(l+1)}) \tr \mathbf{C} \tr \mathbf{1} \]^{-1} D.
+	\end{align*}
+	This corresponds to the model in the MD.
+
+  In the inverse Gamma model we use the following empirical Bayes procedure: Suppose we have a partitioning of our drugs into drug classes and we let prior variance components $\gamma_d^2$ and $\gamma^2_{d'}$ be drawn from the same distribution if drug $d$ and drug $d'$ are from the same class.
+Or simply put: $\gamma^2_{d}, \gamma^2_{d'} \sim \Gamma^{-1}(\alpha_c/2, \lambda_c/2)$ if $d, d' \in \text{class}_c$.
+Empirical bayes estimation of the $\alpha_c$ and $\lambda_c$ by maximisation of the marginal likelihood amounts to iteratively solving the estimating equations:
+  \begin{subequations}
+    \begin{align}
+      \log \(\frac{\lambda_c}{2}\) - \psi\(\frac{\alpha_c}{2}\) - \frac{\sum_{d \in \text{class}_c} e^{(l)}_d}{|\text{class}_c|} & = 0, \label{eq:estequation1}\\
+      \frac{\alpha_c}{\lambda_c} - \frac{\sum_{d \in \text{class}_c} b_d^{(l)}}{|\text{class}_c|} & = 0, \label{eq:estequation2}
+    \end{align}
+  \end{subequations}
+  where $e^{(l)}_d = \log \lambda_d^{(l)} - \psi ( \alpha_d^{(l)}/2) - \log 2$.
+We solve (\ref{eq:estequation2}) for $\lambda_c$, plug it into (\ref{eq:estequation1}) and bound the result using \cite{alzer_inequalities_1997} to obtain an interval $( a_c^*, 2a_c^* )$ containing the solution to (\ref{eq:estequation1}), where:
+  $$
+  a_c^* = \left\{ \frac{\sum_{d \in \text{class}_c} e^{(l)}_d}{|\text{class}_c|} + \log \[ \frac{\sum_{d \in \text{class}_c} b_d^{(l)}}{|\text{class}_c|} \]  \right\}^{-1}.
+  $$
+  With this interval it is straightforward to find $\alpha^{(l+1)}_c$ by any root-finding algorithm and plugging the solution into (\ref{eq:estequation2}) to find $\lambda^{(l+1)}_c$.
+
+	The inverse Gamma model with independent drug classes allows for more flexible empirical Bayes estimation of the prior means for the different drug classes.
+As a consequence however, the risk of overfitting is increased.
+Additionally, it does not allow to include continuous covariates on the drugs.
 
 	\subsection{Evidence lower bound}
-  We monitor convergence through the evidence lower bound (ELBO):
+	We monitor convergence through the evidence lower bound (ELBO).
+To compute the evidence lower bound after updating the EB parameters from $\alpha_d^{(l)}, \theta_d^{(l)}, \lambda_d^{(l)}$ to $\alpha_d^{(l + 1)}, \theta_d^{(l + 1)}, \lambda_d^{(l + 1)}$, we compute $\text{ELBO}^{(l + 1)} = \sum_{d=1}^D \text{ELBO}_d^{(l + 1)}$, with, in the conjugate case:
 	\begin{align*}
-  \text{ELBO} & (Q^{(l)}) \propto \frac{1}{2} \sum_{d=1}^D \log |\bm{\Sigma}_d^{(l)}| - \frac{n + p + 1}{2} \sum_{d=1}^D \log \zeta_d^{(l)} + \sum_{d=1}^D \log K_{\frac{p+1}{2}}\(\sqrt{\lambda^{(l)} \delta_d^{(l)}}/\theta_d^{(l)}\) \\
-  & - \frac{n+p+1}{4} \sum_{d=1}^D (\zeta_d^{(l)})^{-1} \[ \y_d \tr \y_d - 2\y_d \tr \X \bm{\mu}_d^{(l)} + (\bm{\mu}_d^{(l)}) \tr \X \tr \X \bm{\mu}_d^{(l)} + \trace ( \X \tr \X \bm{\Sigma}_d^{(l)})\] \\
-  & + \frac{D(p+3)}{4} \log \lambda^{(l)} + \lambda^{(l)} \sum_{d=1}^D (\theta_d^{(l)})^{-1} - \frac{p+1}{2} \sum_{d=1}^D \log \theta_d^{(l)} - \frac{p+1}{4} \sum_{d=1}^D \log \delta_d^{(l)} \\
-  & + \frac{1}{2} \sum_{d=1}^D \[ \delta_d^{(l)} - \lambda^{(l)} - \frac{n+p+1}{2\zeta_d^{(l)}} (\bm{\mu}_d^{(l)}) \tr \bm{\mu}_d^{(l)} - \frac{n+p+1}{2\zeta_d^{(l)}} \trace (\bm{\Sigma}_d^{(l)}) \] a_d^{(l)}.
+	\text{ELBO}_d^{(l + 1)} & = \frac{1}{2} \log |\bm{\Sigma}_d^{(l)}| - \frac{a_d^{(l)}}{2} \[ \y_d \tr \y_d - 2\y_d \tr \X \bm{\mu}_d^{(l)} + (\bm{\mu}_d^{(l)}) \tr \X \tr \X \bm{\mu}_d^{(l)} + \trace ( \X \tr \X \bm{\Sigma}_d^{(l)})\] \\
+	& - \frac{df + 1}{2} \log \zeta_d^{(l)} + \frac{p + \alpha_d^{(l)}}{4} \log \lambda_d^{(l)} - \frac{p + \alpha_d^{(l)}}{2} \log \theta_d^{(l)} - \frac{p+\alpha_d^{(l)}}{4} \log \delta_d^{(l)} \\
+	& + \log K_{\frac{p+\alpha_d^{(l)}}{2}}\(\sqrt{\lambda_d^{(l)} \delta_d^{(l)}}/\theta_d^{(l)}\) + \frac{b_d^{(l)}}{2} \[ \delta_d^{(l)} - a_d^{(l)} (\bm{\mu}_d^{(l)}) \tr \bm{\mu}_d^{(l)} - a_d^{(l)} \trace (\bm{\Sigma}_d^{(l)}) \] \\
+	& + \frac{\lambda_d^{(l)} e_d^{(l)}}{2(\theta_d^{(l)})^2} + \frac{(\alpha_d^{(l)} - \alpha_d^{(l + 1)})h_d^{(l)}}{2} - \log K_{\frac{\alpha_d^{(l + 1)}}{2}}\(\lambda_d^{(l + 1)}/\theta_d^{(l + 1)}\) + \frac{\alpha_d^{(l + 1)}}{2} \log \theta_d^{(l)} \\
+	& - \frac{\lambda_d^{(l+1)} e_d^{(l)}}{2 (\theta_d^{(l + 1)})^2} - \frac{\lambda_d^{(l + 1)} b_d^{(l)}}{2},
 	\end{align*}
+	where $h_d^{(l)} = \log \delta_d^{(l)} - \psi [(p + \alpha_d^{(l)})/2] - \log 2$ in the inverse Gamma model and
+	$$
+	h_d^{(l)} = \frac{1}{2}\log \delta_d^{(l)} - \frac{1}{2}\log \lambda_d^{(l)} + \log \theta_d^{(l)} + \frac{\partial}{\partial \nu} \log K_{\nu} (\sqrt{\lambda_d^{(l)} \delta_d^{(l)}}/\theta_d^{(l)}) \bigg|_{\nu = \frac{p + \alpha_d^{(l)}}{2}}
+	$$
+	in the inverse Gaussian case.
+Note that in the inverse Gaussian case, the computationally expensive $h_d^{(l)}$ terms never have to be calculated, since we have $\alpha_d^{(l)} - \alpha_d^{(l + 1)}=0$.
+The non-conjugate ELBO simply exchanges the $a_d^{(l)} (\bm{\mu}_d^{(l)}) \tr \bm{\mu}_d^{(l)}$ and $a_d^{(l)} \trace (\bm{\Sigma}_d^{(l)})$ terms with $(\bm{\mu}_d^{(l)}) \tr \bm{\mu}_d^{(l)}$ and $\trace (\bm{\Sigma}_d^{(l)})$, respectively.
+To compute the ELBO after a VB update we simply let $\alpha_d^{(l)}, \theta_d^{(l)}, \lambda_d^{(l)}=\alpha_d^{(l + 1)}, \theta_d^{(l + 1)}, \lambda_d^{(l + 1)}$.
 
-  - Different options for convergence.
-We may monitor parameters, ELBO, or a combination.
-We may also iterate a fixed number of times, especially for the VB (currently doing this, using 2 VB iterations and 20 EB iterations).
+  In practice, there are different options to assess convergence.
+We may monitor the parameters themselves, the ELBO, or a combination of the two.
+We may also iterate for an \textit{a priori} fixed number of times.
+We may also choose to use a different convergence criterium for the EB and VB iterations.
+Currently we fix the number of EB iterations to 20, while we do 2 VB iterations per EB iteration.
 
@@
 The first three quantities are obtained by scalar operations of $\mathcal{O}(n)$.
-The rest is easily calculated using the SVD $\X = \mathbf{U} \mathbf{D} \mathbf{V} \tr$:
+Let $c_d^{(h)}=b_d^{(h)}$ in the conjugate setting and $c_d^{(h)}=b_d^{(h)}/a_d^{(h)}$ in the non-conjugate setting.
+Then, beforementioned quantities are easily calculated using the SVD $\X = \mathbf{U} \mathbf{D} \mathbf{V} \tr$:
   \begin{align*}
-  \trace (\bm{\Sigma}_d^{(h+1)}) & = \frac{2\zeta_d^{(h)}}{n + p + 1} \[\sum_{i=1}^n (v_i^2 + a_d^{(h)})^{-1} + \max (p - n, 0)\cdot (a_d^{(h)})^{-1}\], \\
-  \trace (\X \tr \X \bm{\Sigma}_d^{(h+1)}) & = \frac{2\zeta_d^{(h)}}{n + p + 1} \sum_{i=1}^n v_i^2(v_i^2 + a_d^{(h)})^{-1}, \\
-  (\bm{\mu}_d^{(h+1)}) \tr \bm{\mu}_d^{(h+1)} & = \sum_{i=1}^n v_i^2 [(\mathbf{U} \tr \mathbf{y}_d)_i]^2/(v_i^2 + a_d^{(h)})^2, \\
-  (\bm{\mu}_d^{(h+1)}) \tr \X \tr \X \bm{\mu}_d^{(h+1)} & = \sum_{i=1}^n v_i^4 [(\mathbf{U} \tr \mathbf{y}_d)_i]^2/(v_i^2 + a_d^{(h)})^2, \\
-  \log|\bm{\Sigma}^{(h+1)}| & \propto p \log \zeta_d^{(h)} - \sum_{i=1} \log (v_i^2 + a_d^{(h)}) - \max(p - n, 0) \cdot \log a_d^{(h)}, \\
-  \mathbf{y}_d \tr \X \bm{\mu}_d^{(h+1)} & = \sum_{i=1}^n v_i^2 [(\mathbf{U} \tr \mathbf{y}_d)_i]^2/(v_i^2 + a_d^{(h)}).
+  \trace (\bm{\Sigma}_d^{(h+1)}) & = (a_d^{(h)})^{-1} \[\sum_{i=1}^n (v_i^2 + c_d^{(h)})^{-1} + \max (p - n, 0)\cdot (c_d^{(h)})^{-1}\], \\
+  \trace (\X \tr \X \bm{\Sigma}_d^{(h+1)}) & = (a_d^{(h)})^{-1} \sum_{i=1}^n v_i^2(v_i^2 + c_d^{(h)})^{-1}, \\
+  (\bm{\mu}_d^{(h+1)}) \tr \bm{\mu}_d^{(h+1)} & = \sum_{i=1}^n v_i^2 [(\mathbf{U} \tr \mathbf{y}_d)_i]^2/(v_i^2 + c_d^{(h)})^2, \\
+  (\bm{\mu}_d^{(h+1)}) \tr \X \tr \X \bm{\mu}_d^{(h+1)} & = \sum_{i=1}^n v_i^4 [(\mathbf{U} \tr \mathbf{y}_d)_i]^2/(v_i^2 + c_d^{(h)})^2, \\
+  \log|\bm{\Sigma}^{(h+1)}| & \propto -p \log a_d^{(h)} - \sum_{i=1} \log (v_i^2 + c_d^{(h)}) - \max(p - n, 0) \cdot \log c_d^{(h)}, \\
+  \mathbf{y}_d \tr \X \bm{\mu}_d^{(h+1)} & = \sum_{i=1}^n v_i^2 [(\mathbf{U} \tr \mathbf{y}_d)_i]^2/(v_i^2 + c_d^{(h)}).
   \end{align*}
@@
 
-  \subsection{Independent error variance model parameters}
-  \cite{moran_variance_2018} argue that in regression models, $\bm{\beta}_d$ and $\sigma_d^2$ should be independent \textit{a priori}.
-However, in our case that is debatable.
-The $D$ regressions need not be on the same scale, so some form of calibration of the $\bm{\beta}_d$ posteriors to this scale is required.
-A simple way of doing this is to include the scales $\sigma_d^2$ in the priors for the $\bm{\beta}_d$.
-For a sense of completeness we still compare to the independent prior model in our simulations.
-The independent prior model is given by:
-  \begin{align*}
-  y_{id} | \bm{\beta}_d, \sigma_d^2 & \sim \mathcal{N} (\x_i \tr \bm{\beta}_d, \sigma_d^2), \\
-  \beta_{jd} | \gamma_d^2 & \sim \mathcal{N} (0, \gamma_d^2), \\
-  \gamma_d^2 & \sim \mathcal{IG}((\mathbf{c}_d \tr \bm{\alpha})^{-1}, \lambda), \\
-  \sigma_d^2 & \sim 1/\sigma_d^3.
-  \end{align*}
-  Its variational posterior is given by:
-  \begin{align*}
-  q(\bm{\beta}) & \overset{D}{=} \prod_{d=1}^D \mathcal{N}_p (\bm{\mu}_d, \bm{\Sigma}_d), \\
-  q(\bm{\gamma}^2) & \overset{D}{=} \prod_{d=1}^D \mathcal{GIG}\(-\frac{p+1}{2}, \lambda (\mathbf{c}_d \tr \bm{\alpha})^2, \delta_d\), \\
-  q(\bm{\sigma}^2) & \overset{D}{=} \prod_{d=1}^D \Gamma^{-1} \(\frac{n + 1}{2}, \zeta_d\),
-  \end{align*}
-  with variational parameter updates:
-  \begin{align*}
-  \bm{\Sigma}_d^{(h+1)} & = \(\frac{n+1}{2 \zeta_d^{(h)}}\X \tr \X + a_d^{(h)} \I_p\)^{-1}, \\
-  \bm{\mu}_d^{(h+1)} & = \frac{n+1}{2 \zeta_d^{(h)}} \bm{\Sigma}_d^{(h+1)} \X \tr \y_d, \\
-  \delta_d^{(h+1)} & = \trace(\bm{\Sigma}_d^{(h+1)}) + (\bm{\mu}_d^{(h+1)}) \tr \bm{\mu}_d^{(h+1)} + \lambda, \\ \zeta_d^{(h+1)} & = \frac{1}{2} \[\mathbf{y}_d \tr \mathbf{y}_d -2 \mathbf{y}_d \tr \X \bm{\mu}_d^{(h+1)} + \trace (\X \tr \X \bm{\Sigma}_d^{(h+1)}) + (\bm{\mu}_d^{(h+1)}) \tr \X \tr \X \bm{\mu}_d^{(h+1)}\],
-  \end{align*}
-  with $a_d^{(h)}$ and the empirical Bayes updates $\bm{\alpha}^{(l+1)}$ and $\lambda^{(l+1)}$ as in the full model in the MD.
-A singular value decomposition $\X = \mathbf{U} \mathbf{D} \mathbf{V} \tr$ leads to the following efficient computations:
-  \begin{align*}
-  \trace (\bm{\Sigma}^{(h+1)}_d) & = \frac{2\zeta^{(h)}_d}{n + 1} \sum_{i=1}^n \(v_i^2 + \frac{2\zeta^{(h)}_d a^{(h)}_d}{n+1}\)^{-1} + \max (p - n, 0)\cdot (a^{(h)}_d)^{-1}, \\
-  \trace (\X \tr \X \bm{\Sigma}^{(h + 1)}_d) & = \frac{2\zeta^{(h)}_d}{n + 1} \sum_{i=1}^n v_i^2 \(v_i^2 + \frac{2\zeta^{(h)}_d a^{(h)}_d}{n+1}\)^{-1}, \\
-  (\bm{\mu}_d^{(h + 1)}) \tr \bm{\mu}_d^{(h + 1)} & = \sum_{i=1}^n [(\mathbf{U} \tr \mathbf{y}_d)_i]^2 v_i^2 \(v_i^2 + \frac{2 \zeta_d^{(h)} a_d^{(h)}}{n + 1}\)^{-2}, \\
-  (\bm{\mu}_d^{(h + 1)}) \tr \X \tr \X \bm{\mu}_d^{(h + 1)} & = \sum_{i=1}^n [(\mathbf{U} \tr \mathbf{y}_d)_i]^2v_i^4 \(v_i^2 + \frac{2 \zeta_d^{(h)} a_d^{(h)}}{n + 1}\)^{-2}, \\
-  \log|\bm{\Sigma}^{(h + 1)}| & \propto p \log \zeta_d^{(h)} - \sum_{i=1} \log \(v_i^2 + \frac{2 \zeta_d^{(h)} a_d^{(h)}}{n + 1}\) \\
-  & \,\,\,\,\,\,\,\,\,\,\, - \max(p - n, 0) \cdot (\log a_d^{(h)} + \log \zeta_d^{(h)}), \\
-  \mathbf{y}_d \tr \X \bm{\mu}_d^{(h + 1)} & = \sum_{i=1}^n [(\mathbf{U} \tr \mathbf{y}_d)_i]^2 v_i^2 \(v_i^2 + \frac{2 \zeta_d^{(h)} a_d^{(h)}}{n + 1}\),
-  \end{align*}
-  where $v_i$ are the singular values of $\X$.
-As before, this renders $\mathcal{O}(pn^2)$ and $\mathcal{O}(n^2)$ calculations once at the start of the algorithm and iterative updates of $\mathcal{O}(n)$.
-
-  \subsection{Simulations}
-  We simulate from a correctly specified model.
+  \section{Simulation setting}
+  We simulate from a correctly specified model given in \ref{sec:notissuemodel}.
 We assume random $y_{id}$, $\x_i$, and $\bm{\beta}_d$, while we fix $\mathbf{C}$, $\gamma_d^2$, $\sigma_d^2$ and $\bm{\alpha}$.

### 2019-01-14 docs/manuscript.Rnw
gdsc.R now reads in drug response and expression
gdsc.R includes parsing of pubchem data
new suggestions in manuscript.rnw

@@
 
+	- check CTRP study
+
+	- check data in PharamcoGx R-package
+
 	- how to transform the data and deal with missingness (ask Oslo and read papers published on GDSC and CCLE repositories)
 
-	- what information is available on drugs (ask Oslo)
+	- create drug covariates from SMILES formulas with R-package rcdk.
+
+	- try non-negative matrix factorisation/independent component analysis on binary fingerprints to get principal components that characterise the different drugs.
+
+	- check whether compounds are PAINS \cite[]{baell_new_2010} or not. PAINS indicates whether compound is frequent false hitter.
 
@@
 	\section{Discussion}
+	- use copula's and marginal posteriors to find multivariate posteriors for contrast testing.
+
 	- empirical Bayes gives extra information via estimation of the hyper-parameters, such as informativeness of the drug information (via model for the $\gamma_d$) and relatedness of the tissues (via $\bm{\Omega}$ and $\nu$)

### 2019-01-14 docs/supplement.Rnw
new suggestions in manuscript.rnw

@@
   \section{Simulation setting}
+  - check model with Stan.
+
+  - make one group of drugs and compare methods
+
+  - use drug-specific lambda's
+
+  - try smaller $p$ to see whether $\sigma_d^2$ become more identifiable.
+
+  - scale-invariant instead of conjugate.
+
   We simulate from a correctly specified model given in \ref{sec:notissuemodel}.

### 2019-01-15 docs/supplement.Rnw
VB update now generic

@@
 	We monitor convergence through the evidence lower bound (ELBO).
-To compute the evidence lower bound after updating the EB parameters from $\alpha_d^{(l)}, \theta_d^{(l)}, \lambda_d^{(l)}$ to $\alpha_d^{(l + 1)}, \theta_d^{(l + 1)}, \lambda_d^{(l + 1)}$, we compute $\text{ELBO}^{(l + 1)} = \sum_{d=1}^D \text{ELBO}_d^{(l + 1)}$, with, in the conjugate case:
+To compute the evidence lower bound after updating the EB parameters from $\alpha_d^{(l)}, \theta_d^{(l)}, \lambda_d^{(l)}$ to $\alpha_d^{(l + 1)}, \theta_d^{(l + 1)}, \lambda_d^{(l + 1)}$, we compute $\text{ELBO}^{(l + 1)} = \sum_{d=1}^D \text{ELBO}_d^{(l + 1)}$, with, in the inverse Gaussian conjugate case:
 	\begin{align*}
 	\text{ELBO}_d^{(l + 1)} & = \frac{1}{2} \log |\bm{\Sigma}_d^{(l)}| - \frac{a_d^{(l)}}{2} \[ \y_d \tr \y_d - 2\y_d \tr \X \bm{\mu}_d^{(l)} + (\bm{\mu}_d^{(l)}) \tr \X \tr \X \bm{\mu}_d^{(l)} + \trace ( \X \tr \X \bm{\Sigma}_d^{(l)})\] \\
-	& - \frac{df + 1}{2} \log \zeta_d^{(l)} + \frac{p + \alpha_d^{(l)}}{4} \log \lambda_d^{(l)} - \frac{p + \alpha_d^{(l)}}{2} \log \theta_d^{(l)} - \frac{p+\alpha_d^{(l)}}{4} \log \delta_d^{(l)} \\
-	& + \log K_{\frac{p+\alpha_d^{(l)}}{2}}\(\sqrt{\lambda_d^{(l)} \delta_d^{(l)}}/\theta_d^{(l)}\) + \frac{b_d^{(l)}}{2} \[ \delta_d^{(l)} - a_d^{(l)} (\bm{\mu}_d^{(l)}) \tr \bm{\mu}_d^{(l)} - a_d^{(l)} \trace (\bm{\Sigma}_d^{(l)}) \] \\
-	& + \frac{\lambda_d^{(l)} e_d^{(l)}}{2(\theta_d^{(l)})^2} + \frac{(\alpha_d^{(l)} - \alpha_d^{(l + 1)})h_d^{(l)}}{2} - \log K_{\frac{\alpha_d^{(l + 1)}}{2}}\(\lambda_d^{(l + 1)}/\theta_d^{(l + 1)}\) + \frac{\alpha_d^{(l + 1)}}{2} \log \theta_d^{(l)} \\
-	& - \frac{\lambda_d^{(l+1)} e_d^{(l)}}{2 (\theta_d^{(l + 1)})^2} - \frac{\lambda_d^{(l + 1)} b_d^{(l)}}{2},
+	& - \frac{df + 1}{2} \log \zeta_d^{(l)} + \frac{b_d^{(l)}}{2} \[ \delta_d^{(l)} - a_d^{(l)} (\bm{\mu}_d^{(l)}) \tr \bm{\mu}_d^{(l)} - a_d^{(l)} \trace (\bm{\Sigma}_d^{(l)}) \] + \frac{p + 1}{4} \log \lambda_d^{(l)} \\
+	& - \frac{p + 1}{2} \log \theta_d^{(l)} - \frac{p+1}{4} \log \delta_d^{(l)} + \log K_{\frac{p+1}{2}}\(\sqrt{\lambda_d^{(l)} \delta_d^{(l)}}/\theta_d^{(l)}\) \\
+	& + \frac{\lambda_d^{(l)} e_d^{(l)}}{2(\theta_d^{(l)})^2} +\frac{\lambda_d^{(l+1)}}{\theta_d^{(l+1)}} + \frac{1}{2}\log \lambda_d^{(l+1)} - \frac{\lambda_d^{(l+1)} e_d^{(l)}}{2 (\theta_d^{(l + 1)})^2} - \frac{\lambda_d^{(l + 1)} b_d^{(l)}}{2}.
 	\end{align*}
-	where $h_d^{(l)} = \log \delta_d^{(l)} - \psi [(p + \alpha_d^{(l)})/2] - \log 2$ in the inverse Gamma model and
-	$$
-	h_d^{(l)} = \frac{1}{2}\log \delta_d^{(l)} - \frac{1}{2}\log \lambda_d^{(l)} + \log \theta_d^{(l)} + \frac{\partial}{\partial \nu} \log K_{\nu} (\sqrt{\lambda_d^{(l)} \delta_d^{(l)}}/\theta_d^{(l)}) \bigg|_{\nu = \frac{p + \alpha_d^{(l)}}{2}}
-	$$
-	in the inverse Gaussian case.
-Note that in the inverse Gaussian case, the computationally expensive $h_d^{(l)}$ terms never have to be calculated, since we have $\alpha_d^{(l)} - \alpha_d^{(l + 1)}=0$.
-The non-conjugate ELBO simply exchanges the $a_d^{(l)} (\bm{\mu}_d^{(l)}) \tr \bm{\mu}_d^{(l)}$ and $a_d^{(l)} \trace (\bm{\Sigma}_d^{(l)})$ terms with $(\bm{\mu}_d^{(l)}) \tr \bm{\mu}_d^{(l)}$ and $\trace (\bm{\Sigma}_d^{(l)})$, respectively.
+
+	In the conjugate inverse Gamma setting, we have:
+	\begin{align*}
+	\text{ELBO}_d^{(l + 1)} & = \frac{1}{2} \log |\bm{\Sigma}_d^{(l)}| - \frac{a_d^{(l)}}{2} \[ \y_d \tr \y_d - 2\y_d \tr \X \bm{\mu}_d^{(l)} + (\bm{\mu}_d^{(l)}) \tr \X \tr \X \bm{\mu}_d^{(l)} + \trace ( \X \tr \X \bm{\Sigma}_d^{(l)})\] \\
+	& - \frac{df + 1}{2} \log \zeta_d^{(l)} + \frac{b_d^{(l)}}{2} \[ \delta_d^{(l)} - a_d^{(l)} (\bm{\mu}_d^{(l)}) \tr \bm{\mu}_d^{(l)} - a_d^{(l)} \trace (\bm{\Sigma}_d^{(l)}) \] -\frac{\alpha_d^{(l)}}{2} \psi \(\frac{p + \alpha_d^{(l)}}{2}\) \\
+	& + \log \Gamma \( \frac{p + \alpha_d^{(l)}}{2} \) - \frac{p+\alpha_d^{(l + 1)}}{2} \log \delta_d^{(l)} - \log \Gamma \( \frac{\alpha_d^{(l + 1)}}{2} \) + \frac{\alpha_d^{(l + 1)}}{2} \psi \(\frac{p + \alpha_d^{(l)}}{2}\) \\
+	& + \frac{\alpha_d^{(l + 1)}}{2} \log \lambda_d^{(l+1)} - \frac{\lambda_d^{(l + 1)} b_d^{(l)}}{2}.
+	\end{align*}
+
+	The non-conjugate ELBO versions of both the inverse Gaussian and inverse Gamma models simply exchanges the $a_d^{(l)} (\bm{\mu}_d^{(l)}) \tr \bm{\mu}_d^{(l)}$ and $a_d^{(l)} \trace (\bm{\Sigma}_d^{(l)})$ terms with $(\bm{\mu}_d^{(l)}) \tr \bm{\mu}_d^{(l)}$ and $\trace (\bm{\Sigma}_d^{(l)})$, respectively.
 To compute the ELBO after a VB update we simply let $\alpha_d^{(l)}, \theta_d^{(l)}, \lambda_d^{(l)}=\alpha_d^{(l + 1)}, \theta_d^{(l + 1)}, \lambda_d^{(l + 1)}$.

### 2019-01-21 docs/supplement.Rnw
implemented VB and EB steps for conjugate and non-conjugate inv. Gaussian and inv. Gamma model

--- 2019-01-14 docs/supplement.Rnw
@@
 	$$
-	\gamma_d^2 \sim \Gamma^{-1}(\alpha_d/2, \lambda_d/2).
+	\gamma_d^2 \sim \Gamma^{-1}(\eta_d/2, \lambda_d/2).
 	$$
@@
   \gamma_d^2 & \sim \mathcal{GIG}(-1/2, \lambda_d \theta_d^{-2}, \lambda_d), \\
-  \gamma_d^2 & \sim \mathcal{GIG}(-\alpha_d/2, 0, \lambda_d).
+  \gamma_d^2 & \sim \mathcal{GIG}(-\eta_d/2, 0, \lambda_d).
   \end{align*}
-	The two models coincide if $\theta_d \to \infty$ and $\alpha_d = 1$.
+	The two models coincide if $\theta_d \to \infty$ and $\eta_d = 1$.
 
@@
   q(\bm{\beta}) & \overset{D}{=} \prod_{d=1}^D \mathcal{N}_p (\bm{\mu}_d, \bm{\Sigma}_d), \\
-  q(\bm{\gamma}^2) & \overset{D}{=} \prod_{d=1}^D \mathcal{GIG}\(-\frac{p+\alpha_d}{2}, \lambda_d \theta_d^{-2}, \delta_d\), \\
+  q(\bm{\gamma}^2) & \overset{D}{=} \prod_{d=1}^D \mathcal{GIG}\(-\frac{p+\eta_d}{2}, \lambda_d \theta_d^{-2}, \delta_d\), \\
   q(\bm{\sigma}^2) & \overset{D}{=} \prod_{d=1}^D \Gamma^{-1} \(\frac{df + 1}{2}, \zeta_d\),
   \end{align*}
-	where $\alpha_d=1$ and $\theta_d \to \infty$ in the inverse Gaussian and inverse Gamma models, respectively.
+	where $\eta_d=1$ and $\theta_d \to \infty$ in the inverse Gaussian and inverse Gamma models, respectively.
 Furthermore, $df=n + p$ in the conjugate model and $df=n$ in the non-conjugate model.
@@
   \begin{align*}
-  \bm{\Sigma}_d^{(h+1)} & = (a_d^{(h)})^{-1} (\X \tr \X + b_d^{(h)} \I)^{-1}, \\
-  \bm{\mu}_d^{(h+1)} & = (\X \tr \X + b_d^{(h)} \I)^{-1} \X \tr \y_d, \\
-  \delta_d^{(h+1)} & = a_d^{(h)} \bmu_d \tr \bmu_d + \lambda_d, \\ 
+  \bm{\Sigma}_d^{(h+1)} & = (a_d^{(h)})^{-1} (\X \tr \X + c_d^{(h)} \I)^{-1}, \\
+  \bm{\mu}_d^{(h+1)} & = (\X \tr \X + c_d^{(h)} \I)^{-1} \X \tr \y_d, \\
+  \delta_d^{(h+1)} & = b_d^{(h)} (c_d^{(h)})^{-1} \bmu_d \tr \bmu_d + \lambda_d, \\ 
   \zeta_d^{(h+1)} & = \frac{1}{2} \bigg[\mathbf{y}_d \tr \mathbf{y}_d -2 \mathbf{y}_d \tr \X \bm{\mu}_d^{(h+1)} + \trace ( \X \tr \X \bm{\Sigma}_d^{(h+1)}) + (\bm{\mu}_d^{(h+1)}) \tr \X \tr \X \bm{\mu}_d^{(h+1)} \\
-  & \,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\, + b_d^{(h+1)} \trace ( \bm{\Sigma}_d^{(h+1)}) + b_d^{(h+1)} (\bm{\mu}_d^{(h+1)}) \tr \bm{\mu}_d^{(h+1)}\bigg].
-  \end{align*}
-	while in the non-conjugate setting, we use the following updating equations:
-	\begin{align*}
-  \bm{\Sigma}_d^{(h+1)} & = (a_d^{(h)})^{-1} (\X \tr \X + b_d^{(h)} (a_d^{(h)})^{-1} \I)^{-1}, \\
-  \bm{\mu}_d^{(h+1)} & = (\X \tr \X + b_d^{(h)} (a_d^{(h)})^{-1} \I)^{-1} \X \tr \y_d, \\
-  \delta_d^{(h+1)} & = \bmu_d \tr \bmu_d + \lambda_d, \\ 
-  \zeta_d^{(h+1)} & = \frac{1}{2} \[\mathbf{y}_d \tr \mathbf{y}_d -2 \mathbf{y}_d \tr \X \bm{\mu}_d^{(h+1)} + \trace ( \X \tr \X \bm{\Sigma}_d^{(h+1)}) + (\bm{\mu}_d^{(h+1)}) \tr \X \tr \X \bm{\mu}_d^{(h+1)} \].
+  & \,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\, + \frac{b_d^{(h+1)} - c_d^{(h+1)} a_d^{(h+1)}}{1 - a_d^{(h+1)}} \trace ( \bm{\Sigma}_d^{(h+1)}) + \frac{b_d^{(h+1)} - c_d^{(h+1)} a_d^{(h+1)}}{1 - a_d^{(h+1)}} (\bm{\mu}_d^{(h+1)}) \tr \bm{\mu}_d^{(h+1)}\bigg].
   \end{align*}
@@
   a_d^{(h)} & =\E_{Q^{(h)}}(\sigma_d^{-2})=(df+1)/(2 \zeta_d^{(h)}), \nonumber \\
-  b_d^{(h)} & = \E_{Q^{(h)}}(\gamma_d^{-2}) = \sqrt{\frac{\lambda_d}{\theta_d^2 \delta_d^{(h)}}} \frac{K_{\frac{p + \alpha_d -2}{2}} \( \sqrt{\lambda_d \delta_d^{(h)}}/\theta_d \)}{K_{\frac{p+\alpha_d}{2}} \( \sqrt{\lambda_d \delta_d^{(h)}}/\theta_d \)} + \frac{p+\alpha_d^{(h)}}{\delta_d^{(h)}}.
-\label{eq:auxvar1}
+  b_d^{(h)} & = \E_{Q^{(h)}}(\gamma_d^{-2}) = \sqrt{\frac{\lambda_d}{\theta_d^2 \delta_d^{(h)}}} \frac{K_{\frac{p + \eta_d -2}{2}} \( \sqrt{\lambda_d \delta_d^{(h)}}/\theta_d \)}{K_{\frac{p+\eta_d}{2}} \( \sqrt{\lambda_d \delta_d^{(h)}}/\theta_d \)} + \frac{p+\eta_d^{(h)}}{\delta_d^{(h)}}, \label{eq:auxvar1} \\
+  c_d^{(h)} & = \begin{cases*}
+  b_d^{(h)} & in the conjugate setting \\
+  b_d^{(h)}/a_d^{(h)} & in the non-conjugate setting .
+  \end{cases*} \nonumber
   \end{align}
-  Note that in the inverse Gamma model, the computation of (\ref{eq:auxvar1}) simplifies to $b_d^{(h)}=(p+\alpha_d)/\delta_d^{(h)}$.
+  Note that in the inverse Gamma model, the computation of (\ref{eq:auxvar1}) simplifies to $b_d^{(h)}=(p+\eta_d)/\delta_d^{(h)}$.
 
@@
   \begin{align*}
-  \bm{\alpha} & = \[ \mathbf{C} \tr \diag (\lambda_d e_d^{(l)}) \mathbf{C} \]^{-1} \diag (\lambda_d) \mathbf{1}_{D \times 1}, \\
+  \bm{\alpha} & = \[ \mathbf{C} \tr \diag (\lambda_d e_d^{(l)}) \mathbf{C} \]^{-1} \mathbf{C} \tr \diag (\lambda_d) \mathbf{1}_{D \times 1}, \\
   \lambda_d & = \[ b_d^{(l)} + e_d^{(l)} (\mathbf{c}_d \tr \bm{\alpha})^2 - 2\mathbf{c}_d \tr \bm{\alpha} \]^{-1},
   \end{align*}
-  where $e_d^{(l)} = (b_d^{(l)} + \delta_d^{(l)}/p) \cdot \delta_d^{(l)} (\theta_d^{(h)})^2/\lambda_d^{(l)}$.
+  where $e_d^{(l)} = \E_{Q^{(l)}}(\gamma_d^2) = [b_d^{(l)} - (p + \eta_d^{(h)})/\delta_d^{(l)})] \cdot \delta_d^{(l)} (\theta_d^{(h)})^2/\lambda_d^{(l)}$.
 Solving the equations is done by iteratively reweighted least squares of responses $(e_d^{(l)})^{-1}$ on predictors $\mathbf{c}_d$ with weights $\lambda_d e_d^{(l)}$.
@@
   In the inverse Gamma model we use the following empirical Bayes procedure: Suppose we have a partitioning of our drugs into drug classes and we let prior variance components $\gamma_d^2$ and $\gamma^2_{d'}$ be drawn from the same distribution if drug $d$ and drug $d'$ are from the same class.
-Or simply put: $\gamma^2_{d}, \gamma^2_{d'} \sim \Gamma^{-1}(\alpha_c/2, \lambda_c/2)$ if $d, d' \in \text{class}_c$.
-Empirical bayes estimation of the $\alpha_c$ and $\lambda_c$ by maximisation of the marginal likelihood amounts to iteratively solving the estimating equations:
+Or simply put: $\gamma^2_{d}, \gamma^2_{d'} \sim \Gamma^{-1}(\eta_c/2, \lambda_c/2)$ if $d, d' \in \text{class}_c$.
+Empirical bayes estimation of the $\eta_c$ and $\lambda_c$ by maximisation of the (approximated) marginal likelihood amounts to iteratively solving the estimating equations:
   \begin{subequations}
     \begin{align}
-      \log \(\frac{\lambda_c}{2}\) - \psi\(\frac{\alpha_c}{2}\) - \frac{\sum_{d \in \text{class}_c} e^{(l)}_d}{|\text{class}_c|} & = 0, \label{eq:estequation1}\\
-      \frac{\alpha_c}{\lambda_c} - \frac{\sum_{d \in \text{class}_c} b_d^{(l)}}{|\text{class}_c|} & = 0, \label{eq:estequation2}
+      \psi\(\frac{\eta_c}{2}\) & = \log \lambda_c - \frac{\sum_{d \in \text{class}_c} e^{(l)}_d}{|\text{class}_c|} - \log 2, \label{eq:estequation1}\\
+      \lambda_c & = \eta_c \cdot |\text{class}_c| \cdot \(\sum_{d \in \text{class}_c} b_d^{(l)}\)^{-1}, \label{eq:estequation2}
     \end{align}
   \end{subequations}
-  where $e^{(l)}_d = \log \lambda_d^{(l)} - \psi ( \alpha_d^{(l)}/2) - \log 2$.
-We solve (\ref{eq:estequation2}) for $\lambda_c$, plug it into (\ref{eq:estequation1}) and bound the result using \cite{alzer_inequalities_1997} to obtain an interval $( a_c^*, 2a_c^* )$ containing the solution to (\ref{eq:estequation1}), where:
+  until convergence, where $e^{(l)}_d = \E_{Q^{(l)}}(\log \gamma_d^2) = \log \delta_d^{(l)} - \psi [ (p + \eta_d^{(l)})/2] - \log 2$. 
+  We propose to do that by first solving (\ref{eq:estequation2}) for $\lambda_c$, plugging it into (\ref{eq:estequation1}) and bounding the result using \cite{alzer_inequalities_1997} to obtain an interval $( \eta_c^*, 2\eta_c^* )$ containing the solution to (\ref{eq:estequation1}), where:
   $$
-  a_c^* = \left\{ \frac{\sum_{d \in \text{class}_c} e^{(l)}_d}{|\text{class}_c|} + \log \[ \frac{\sum_{d \in \text{class}_c} b_d^{(l)}}{|\text{class}_c|} \]  \right\}^{-1}.
+  \eta_c^* = \left\{ \frac{\sum_{d \in \text{class}_c} e^{(l)}_d}{|\text{class}_c|} + \log \[ \frac{\sum_{d \in \text{class}_c} b_d^{(l)}}{|\text{class}_c|} \]  \right\}^{-1}.
   $$
   With this interval it is straightforward to find $\alpha^{(l+1)}_c$ by any root-finding algorithm and plugging the solution into (\ref{eq:estequation2}) to find $\lambda^{(l+1)}_c$.
+We may use $\eta_c^*$ as a starting value.
 
@@
 	We monitor convergence through the evidence lower bound (ELBO).
-To compute the evidence lower bound after updating the EB parameters from $\alpha_d^{(l)}, \theta_d^{(l)}, \lambda_d^{(l)}$ to $\alpha_d^{(l + 1)}, \theta_d^{(l + 1)}, \lambda_d^{(l + 1)}$, we compute $\text{ELBO}^{(l + 1)} = \sum_{d=1}^D \text{ELBO}_d^{(l + 1)}$, with, in the conjugate case:
+To compute the evidence lower bound after updating the EB parameters from $\eta_d^{(l)}, \theta_d^{(l)}, \lambda_d^{(l)}$ to $\eta_d^{(l + 1)}, \theta_d^{(l + 1)}, \lambda_d^{(l + 1)}$, we compute $\text{ELBO}^{(l + 1)} = \sum_{d=1}^D \text{ELBO}_d^{(l + 1)}$, with, in the inverse Gaussian case:
 	\begin{align*}
 	\text{ELBO}_d^{(l + 1)} & = \frac{1}{2} \log |\bm{\Sigma}_d^{(l)}| - \frac{a_d^{(l)}}{2} \[ \y_d \tr \y_d - 2\y_d \tr \X \bm{\mu}_d^{(l)} + (\bm{\mu}_d^{(l)}) \tr \X \tr \X \bm{\mu}_d^{(l)} + \trace ( \X \tr \X \bm{\Sigma}_d^{(l)})\] \\
-	& - \frac{df + 1}{2} \log \zeta_d^{(l)} + \frac{p + \alpha_d^{(l)}}{4} \log \lambda_d^{(l)} - \frac{p + \alpha_d^{(l)}}{2} \log \theta_d^{(l)} - \frac{p+\alpha_d^{(l)}}{4} \log \delta_d^{(l)} \\
-	& + \log K_{\frac{p+\alpha_d^{(l)}}{2}}\(\sqrt{\lambda_d^{(l)} \delta_d^{(l)}}/\theta_d^{(l)}\) + \frac{b_d^{(l)}}{2} \[ \delta_d^{(l)} - a_d^{(l)} (\bm{\mu}_d^{(l)}) \tr \bm{\mu}_d^{(l)} - a_d^{(l)} \trace (\bm{\Sigma}_d^{(l)}) \] \\
-	& + \frac{\lambda_d^{(l)} e_d^{(l)}}{2(\theta_d^{(l)})^2} + \frac{(\alpha_d^{(l)} - \alpha_d^{(l + 1)})h_d^{(l)}}{2} - \log K_{\frac{\alpha_d^{(l + 1)}}{2}}\(\lambda_d^{(l + 1)}/\theta_d^{(l + 1)}\) + \frac{\alpha_d^{(l + 1)}}{2} \log \theta_d^{(l)} \\
-	& - \frac{\lambda_d^{(l+1)} e_d^{(l)}}{2 (\theta_d^{(l + 1)})^2} - \frac{\lambda_d^{(l + 1)} b_d^{(l)}}{2},
+	& + \frac{b_d^{(l)}}{2} \[ \delta_d^{(l)} - (a_d^{(l)} + 1 - b_d^{(l)}/c_d^{(l)})(\bm{\mu}_d^{(l)}) \tr \bm{\mu}_d^{(l)} - (a_d^{(l)} + 1 - b_d^{(l)}/c_d^{(l)}) \trace (\bm{\Sigma}_d^{(l)}) \] \\
+	& - \frac{df + 1}{2} \log \zeta_d^{(l)} + \frac{p + 1}{4} \log \lambda_d^{(l)} - \frac{p + 1}{2} \log \theta_d^{(l)} - \frac{p+1}{4} \log \delta_d^{(l)} \\
+	& + \log K_{\frac{p+1}{2}}\(\sqrt{\lambda_d^{(l)} \delta_d^{(l)}}/\theta_d^{(l)}\) + \frac{\lambda_d^{(l)} e_d^{(l)}}{2(\theta_d^{(l)})^2} \\
+	& + \frac{\lambda_d^{(l+1)}}{\theta_d^{(l+1)}} + \frac{1}{2}\log \lambda_d^{(l+1)} - \frac{\lambda_d^{(l+1)} e_d^{(l)}}{2 (\theta_d^{(l + 1)})^2} - \frac{\lambda_d^{(l + 1)} b_d^{(l)}}{2}.
 	\end{align*}
-	where $h_d^{(l)} = \log \delta_d^{(l)} - \psi [(p + \alpha_d^{(l)})/2] - \log 2$ in the inverse Gamma model and
-	$$
-	h_d^{(l)} = \frac{1}{2}\log \delta_d^{(l)} - \frac{1}{2}\log \lambda_d^{(l)} + \log \theta_d^{(l)} + \frac{\partial}{\partial \nu} \log K_{\nu} (\sqrt{\lambda_d^{(l)} \delta_d^{(l)}}/\theta_d^{(l)}) \bigg|_{\nu = \frac{p + \alpha_d^{(l)}}{2}}
-	$$
-	in the inverse Gaussian case.
-Note that in the inverse Gaussian case, the computationally expensive $h_d^{(l)}$ terms never have to be calculated, since we have $\alpha_d^{(l)} - \alpha_d^{(l + 1)}=0$.
-The non-conjugate ELBO simply exchanges the $a_d^{(l)} (\bm{\mu}_d^{(l)}) \tr \bm{\mu}_d^{(l)}$ and $a_d^{(l)} \trace (\bm{\Sigma}_d^{(l)})$ terms with $(\bm{\mu}_d^{(l)}) \tr \bm{\mu}_d^{(l)}$ and $\trace (\bm{\Sigma}_d^{(l)})$, respectively.
-To compute the ELBO after a VB update we simply let $\alpha_d^{(l)}, \theta_d^{(l)}, \lambda_d^{(l)}=\alpha_d^{(l + 1)}, \theta_d^{(l + 1)}, \lambda_d^{(l + 1)}$.
+
+	In the inverse Gamma setting, we have:
+	\begin{align*}
+	\text{ELBO}_d^{(l + 1)} & = \frac{1}{2} \log |\bm{\Sigma}_d^{(l)}| - \frac{a_d^{(l)}}{2} \[ \y_d \tr \y_d - 2\y_d \tr \X \bm{\mu}_d^{(l)} + (\bm{\mu}_d^{(l)}) \tr \X \tr \X \bm{\mu}_d^{(l)} + \trace ( \X \tr \X \bm{\Sigma}_d^{(l)})\] \\
+	& + \frac{b_d^{(l)}}{2} \[ \delta_d^{(l)} - (a_d^{(l)} + 1 - b_d^{(l)}/c_d^{(l)})(\bm{\mu}_d^{(l)}) \tr \bm{\mu}_d^{(l)} - (a_d^{(l)} + 1 - b_d^{(l)}/c_d^{(l)}) \trace (\bm{\Sigma}_d^{(l)}) \]  \\
+	& - \frac{df + 1}{2} \log \zeta_d^{(l)} - \frac{p + \eta_d^{(l)}}{2} \log \delta_d^{(l)} + \frac{\eta_d^{(l)}}{2} e_d^{(l)} + \frac{\eta_d^{(l)}}{2} \log 2 + \log \Gamma \( \frac{p + \eta_d^{(l)}}{2} \) \\
+	& - \log \Gamma \( \frac{\eta_d^{(l + 1)}}{2} \) - \frac{\eta_d^{(l+1)}}{2} e_d^{(l)} - \frac{\eta_d^{(l+1)}}{2} \log 2 + \frac{\eta_d^{(l + 1)}}{2} \log \lambda_d^{(l+1)} - \frac{\lambda_d^{(l + 1)} b_d^{(l)}}{2}.
+	\end{align*}
+  Note that the computation of $e_d^{(l)}$ differs between the inverse Gaussian and Gamma models.
+Furthermore, to compute the ELBO after a VB update we simply let $\eta_d^{(l)}, \theta_d^{(l)}, \lambda_d^{(l)}=\eta_d^{(l + 1)}, \theta_d^{(l + 1)}, \lambda_d^{(l + 1)}$.
 

### 2019-01-21 rpackage/R/RcppExports.R

@@ -1,8 +1,8 @@
 # Generated by using Rcpp::compileAttributes() -> do not edit by hand
 # Generator token: 10BE3573-1514-4C36-9D1C-5A225CD40393
 
-ratio_besselK_cpp <- function(x, p) {
-    .Call(`_cambridge_ratio_besselK_cpp`, x, p)
+ratio_besselK_cpp <- function(x, nu) {
+    .Call(`_cambridge_ratio_besselK_cpp`, x, nu)
 }
 
 vb_update_cpp <- function(zetaold, aold, lambda, theta, x, xxt, xtx, y) {

### 2019-01-21 rpackage/R/functions.R

@@ -1,10 +1,5 @@
-# single VB update conjugate, non-conjugate, inv. Gauss and Gamma (not tested)
-single.vb.update <- function(aold, bold, alpha, theta, lambda, sv, n, p, uty, 
-                             yty, conjugate, inv.gauss) {
-  
-  # variables that change with conjugacy
-  cold <- bold/aold^conjugate
-  df <- n + conjugate*p
+# calculates the auxiliary variables in the VB step (tested)
+.aux.var <- function(cold, aold, sv, uty, n, p) {
   
   # auxiliary variables involving mu and Sigma
   trSigma <- (sum(1/(sv^2 + cold)) + max(p - n, 0)/cold)/aold
@@ -15,186 +10,190 @@ single.vb.update <- function(aold, bold, alpha, theta, lambda, sv, n, p, uty,
     max(p - n, 0)*log(cold)
   ytXmu <- sum(sv^2*uty^2/(sv^2 + cold))
   
-  # vb parameters
-  delta <- aold^conjugate*mutmu + lambda
-  b <- ifelse(inv.gauss, 
-              sqrt(lambda/(theta^2*delta))*ratio_besselK_cpp(sqrt(
-                lambda*delta/theta^2), p)+ (p + 1)/delta,
-              (p + alpha)/delta)
-  zeta <- 0.5*(yty - 2*ytXmu + trXtXSigma + mutXtXmu + 
-                 conjugate*b*(trSigma + mutmu))
-  a <- (df + 1)/(2*zeta)
+  out <- list(trSigma=trSigma, trXtXSigma=trXtXSigma, mutmu=mutmu,
+              mutXtXmu=mutXtXmu, logdetSigma=logdetSigma,
+              ytXmu=ytXmu)
+  return(out)
+}
+
+# calculates the ELBO in the inverse Gaussian model (tested)
+.elbo.inv.gauss <- function(aux, zeta, delta, a, b, c, e, lambda, theta, df, p, 
+                            yty) {
   
-  # variable needed in eb steps
-  e <- ifelse(inv.gauss, (b + delta/p)*delta*theta^2/lambda, 
-              log(lambda) - digamma(alpha/2) - log(2))
[…]

### 2019-01-21 rpackage/src/RcppExports.cpp

@@ -7,14 +7,14 @@
 using namespace Rcpp;
 
 // ratio_besselK_cpp
-NumericVector ratio_besselK_cpp(arma::vec x, int p);
-RcppExport SEXP _cambridge_ratio_besselK_cpp(SEXP xSEXP, SEXP pSEXP) {
+NumericVector ratio_besselK_cpp(arma::vec x, double nu);
+RcppExport SEXP _cambridge_ratio_besselK_cpp(SEXP xSEXP, SEXP nuSEXP) {
 BEGIN_RCPP
     Rcpp::RObject rcpp_result_gen;
     Rcpp::RNGScope rcpp_rngScope_gen;
     Rcpp::traits::input_parameter< arma::vec >::type x(xSEXP);
-    Rcpp::traits::input_parameter< int >::type p(pSEXP);
-    rcpp_result_gen = Rcpp::wrap(ratio_besselK_cpp(x, p));
+    Rcpp::traits::input_parameter< double >::type nu(nuSEXP);
+    rcpp_result_gen = Rcpp::wrap(ratio_besselK_cpp(x, nu));
     return rcpp_result_gen;
 END_RCPP
 }

### 2019-01-28 docs/presentation.Rnw
first version of presentation added

\documentclass[10pt,letterpaper]{beamer} %
\usetheme{CambridgeUS}
\usefonttheme{professionalfonts}

%\usepackage{footnote}
%\makesavenoteenv{figure}

\usepackage{amsmath,verbatim,bm,caption,soul,graphicx}
\usepackage[plain]{algorithm}
\usepackage{algorithmic}
\usepackage[export]{adjustbox}
\usepackage[american]{babel}
\usepackage[backend=bibtex,style=verbose]{biblatex}

\addtobeamertemplate{block begin}{\setlength\abovedisplayskip{0pt}}

\setbeamertemplate{sidebar right}{}% or get rid of navigation entries there somehow
\addtobeamertemplate{footline}{\hfill\usebeamertemplate***{navigation symbols}%
    \hspace*{0.1cm}\par\vskip 2pt}{}

\setbeamerfont{caption}{size=\scriptsize}

\makeatletter
\setbeamertemplate{headline}{%
	\begin{beamercolorbox}[ht=2.25ex,dp=1ex]{section in head/foot}
		\insertnavigation{\paperwidth}
	\end{beamercolorbox}%
}%
\makeatother

% vectors and matrices
\newcommand{\bbeta}{\bm{\beta}}
\newcommand{\btau}{\bm{\tau}}
\newcommand{\bsigma}{\bm{\sigma}}
\newcommand{\balpha}{\bm{\alpha}}
\newcommand{\bepsilon}{\bm{\epsilon}}
\newcommand{\bomega}{\bm{\omega}}
\newcommand{\bOmega}{\bm{\Omega}}
\newcommand{\bSigma}{\bm{\Sigma}}
\newcommand{\bkappa}{\bm{\kappa}}
\newcommand{\btheta}{\bm{\theta}}
\newcommand{\bmu}{\bm{\mu}}
\newcommand{\bpsi}{\bm{\psi}}
\newcommand{\blambda}{\bm{\lambda}}
\newcommand{\bLambda}{\bm{\Lambda}}
\newcommand{\bPsi}{\bm{\Psi}}
\newcommand{\bphi}{\bm{\phi}}
\newcommand{\Y}{\mathbf{Y}}
\newcommand{\y}{\mathbf{y}}
\newcommand{\X}{\mathbf{X}}
\newcommand{\x}{\mathbf{x}}
\newcommand{\I}{\mathbf{I}}
\newcommand{\bgamma}{\bm{\gamma}}
\newcommand{\T}{\mathbf{T}}
\newcommand{\Z}{\mathbf{Z}}
\newcommand{\W}{\mathbf{W}}
\renewcommand{\O}{\mathbf{O}}
\newcommand{\B}{\mathbf{B}}
\renewcommand{\H}{\mathbf{H}}
\newcommand{\U}{\mathbf{U}}
\newcommand{\Em}{\mathbf{E}}
\newcommand{\D}{\mathbf{D}}
\newcommand{\Vm}{\mathbf{V}}
\newcommand{\rv}{\mathbf{r}}
\newcommand{\A}{\mathbf{A}}
\newcommand{\Q}{\mathbf{Q}}
\newcommand{\Sm}{\mathbf{S}}
\newcommand{\0}{\bm{0}}
\newcommand{\tx}{\tilde{\mathbf{x}}}
\newcommand{\hy}{\hat{y}}
\newcommand{\tm}{\tilde{m}}
\newcommand{\tkappa}{\tilde{\kappa}}
\newcommand{\m}{\mathbf{m}}
\newcommand{\C}{\mathbf{C}}
\renewcommand{\v}{\mathbf{v}}
\newcommand{\g}{\mathbf{g}}
\newcommand{\s}{\mathbf{s}}
\renewcommand{\c}{\mathbf{c}}
\newcommand{\ones}{\mathbf{1}}
\newcommand{\R}{\mathbf{R}}
\newcommand{\rvec}{\mathbf{r}}
\newcommand{\Lm}{\mathbf{L}}

% functions and operators
\renewcommand{\L}{\mathcal{L}}
\newcommand{\G}{\mathcal{G}}
\renewcommand{\P}{\mathcal{P}}
\newcommand{\argmax}{\text{argmax} \,}
\newcommand{\expit}{\text{expit}}
\newcommand{\erfc}{\text{erfc}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\V}{\mathbb{V}}
\newcommand{\Cov}{\mathbb{C}\text{ov}}
\newcommand{\tr}{^{\text{T}}}
\newcommand{\diag}{\text{diag}}
\newcommand{\KL}{D_{\text{KL}}}
\newcommand{\trace}{\text{tr}}
\newcommand{\norm}[1]{\left\lVert #1 \right\rVert}

% distributions
\newcommand{\N}{\text{N}}
\newcommand{\TG}{\text{TG}}
\newcommand{\Bi}{\text{Binom}}
\newcommand{\PG}{\text{PG}}
\newcommand{\Mu}{\text{Multi}}
\newcommand{\GIG}{\text{GIG}}
\newcommand{\IGauss}{\text{IGauss}}
\newcommand{\Un}{\text{U}}

% syntax and readability
\renewcommand{\(}{\left(}
\renewcommand{\)}{\right)}
\renewcommand{\[}{\left[}
\renewcommand{\]}{\right]}
\renewcommand*{\thefootnote}{\fnsymbol{footnote}}

\author[Magnus M\"unch]{Magnus M\"unch \\[3mm] \footnotesize{Joint work with Mark A. van de Wiel, Sylvia Richardson, and Gwena{\"e}l G. R. Leday}}
\title[EB for drug efficacy prediction]{Emprical Bayes for drug efficacy prediction in cell lines}
\institute[LU \& AUMC]{Leiden University and Amsterdam UMC}
\date{\today} % Date, can be changed to a custom date
\bibliography{refs}

\begin{document}

% load packages and set options
<<include=FALSE>>=
library(cambridge)
library(statmod)
opts_knit$set(root.dir="..", base.dir="../figs")
opts_chunk$set(fig.align='center', echo=FALSE, fig.path="../figs/")
@

% read figure code
<<figures, include=FALSE>>=
read_chunk("code/figures.R")
@

  \begin{frame}
    \titlepage % Print the title page as the first slide
    \begin{figure}
      \raisebox{-0.5\height}{\includegraphics[width=0.3\linewidth]{../figs/logo-universiteitleiden-cmyk.pdf}}
		  \raisebox{-0.5\height}{\includegraphics[width=0.3\linewidth]{../figs/Amsterdam_UMC_UK_Logo_CMYK_U_150_dpi.jpg}}
	  \end{figure}
  \end{frame}

  \section{The challenge}
  \begin{frame}
    \frametitle{Motivation}
    \begin{figure}[h]
      \centering
			\includegraphics[width=1\linewidth]{../figs/{F2.large}.jpg}
		\end{figure}
  \end{frame}

  \begin{frame}
    \frametitle{Challenges}
    \begin{itemize}
      \item \textbf<2>{number of treatments \only<1>{$D$} \only<2>{$\bm{D}$} large}
			\item \only<1>{$p > n$} \only<2>{$\bm{p > n}$}
			\item translation from cell lines to patient
			\item quantification of drug efficacy
			\item ...
		\end{itemize}
  \end{frame}

  \begin{frame}
    \frametitle{Opportunities}
    \begin{columns}
	    \begin{column}{0.45\linewidth}
		    \begin{itemize}
          \item Highly structured setting
			    \item Correlated treatments
			    \item Extra information on drugs
			    \item ...
		    \end{itemize}
	    \end{column}
	    \begin{column}{0.45\linewidth}
	      \only<1>{\begin{figure}[h]
          \centering
			    \includegraphics[width=0.5\linewidth]{../figs/pubchem.png}
		    \end{figure}
		    \begin{columns}
		      \begin{column}{0.5\linewidth}
<<drug_effect, include=TRUE, echo=FALSE, fig.align="center", out.width="100%">>=
curve(dnorm(x), -4, 4, main="", yaxt='n', xaxt="n", bty="n", ann=FALSE, cex=10, asp=10,
      lwd=4)
axis(1, line=-6, cex=100, lwd=4)
title(xlab="Drug effect", line=-3.3, cex=10, cex.lab=1.5)
@
		      \end{column}
		      \begin{column}{0.5\linewidth}
		        \begin{figure}[h]
              \centering
			        \includegraphics[width=1\linewidth]{../figs/300px-SMILES.png}
		        \end{figure}
		      \end{column}
		    \end{columns}
		    }
		    \begin{enumerate}[$\Rightarrow$]
          \item<2-3> Simple models possible
			    \item<2-3> \textbf<3>{Learning across treatments}
			    \item<2-3> \textbf<3>{Incorporate to aid estimation}
			    \item<2-3> ...
		    \end{enumerate}
	    \end{column}
		\end{columns}
		\bigskip
		\only<3> {\centering \textbf{Suggest Bayesian model}}
  \end{frame}

  \section{Our solution}
  \begin{frame}
    \frametitle{Model}
    \textbf{Likelihood} for cell line $i$'s response $y_i$ to drug $d$, using $p$ omics features $\mathbf{x}_{ij}$.
    \begin{align*}
		  \only<1>{y_{id} & = \beta_{0d} + \x_i \tr \bbeta_d + \mathbf{z}_i \tr \mathbf{u}_d + \epsilon_{id},} 
		  \only<2>{y_{id} & = \beta_{0d} + \x_i \tr \bbeta_d \phantom{ + \mathbf{z}_i \tr \mathbf{u}_d} + \epsilon_{id},} \\
		  \epsilon_{id} & \sim \mathcal{N}(0, \sigma_d^2),
		\end{align*}
		\textbf{Model parameters}
		\begin{align*}
		  \beta_{jd} & \sim \mathcal{N}_p (0, \gamma_d^2 \sigma_d^2), \\
		  \only<1>{\mathbf{u}_{d} & \sim \mathcal{N}_T (\mathbf{0}, \bm{\Xi}_d \sigma_d^2),}
		\end{align*}
	  \textbf{Error and prior variance}
		\begin{align*}
			\sigma_d^{2} & \sim 1/\sigma_d^{3}, \\
			\gamma_d^{2} & \sim \mathcal{IG}(\theta_d, \lambda_d)\only<1>{,}\only<2>{.} \\
			\only<1>{\bm{\Xi}_d & \sim \mathcal{W}^{-1}_T(\bm{\Omega}, \nu).}
		\end{align*}
  \end{frame}

  \begin{frame}
    \frametitle{Hyperparameters}
    \textbf{How to choose $\theta_d$ and $\lambda_d$?} We want to somehow:
    \begin{columns}
      \begin{column}{0.5\linewidth}
        \begin{itemize}
          \item Learn across drugs
          \item Be objective
          \item Incorporate extra drug information $\mathbf{c}_d$
        \end{itemize}
      \end{column}
      \begin{column}{0.5\linewidth}
        \only<2>{\centering $\Rightarrow$ \textbf{Empirical Bayes} (EB) \\
        Let $\theta_d = (\mathbf{c}_d \tr \bm{\alpha})^{-1}$ \\
        $\Downarrow$ \\
        estimate $\bm{\alpha}$ and $\lambda_d$
        }
      \end{column}
    \end{columns}
  \end{frame}

  \begin{frame}
    \frametitle{Empirical Bayes}
    \textbf{Cannonical EB}: Maximise the marginal likelihood (MML)
    \begin{itemize}
      \item analytical MML
      \item Laplace approximation
      \item using EM
      \begin{itemize}
        \item MCMC
        \item \textbf<2>{Variational Bayes} (VB)
      \end{itemize}
    \end{itemize}
    \only<2>{
    \begin{block}{Empirical Bayes EM}
      $$
      \bm{\alpha}^{(l+1)}, \bm{\lambda}^{(l+1)} = \underset{\bm{\alpha}, \bm{\lambda}}{\argmax} \, \E_{\bbeta | \y} \[ \log \L_{\bm{\alpha}, \bm{\lambda}}(\y, \bbeta) | \bm{\alpha}^{(l)}, \bm{\lambda}^{(l)} \]
      $$
    \end{block}
    }
  \end{frame}

  \begin{frame}
    \frametitle{Variational Bayes}
    $\E_{\bbeta | \y}$ hard to compute, so we approximate:
    $$
    p(\bbeta , \bm{\gamma}^2, \bm{\sigma}^2 | \y) \approx Q(\cdot) = q(\bbeta)q(\bm{\gamma}^2)q(\bm{\sigma}^2)
    $$
    Variational calculus tells us $D_{KL}(Q || P)$ is minimised by:
    \begin{align*}
    q(\bbeta) & \overset{D}{=} \prod_{d=1}^D \mathcal{N}_p(\bm{\mu}_d,\bm{\Sigma}_d), \\
    q(\bm{\gamma}^2) & \overset{D}{=} \prod_{d=1}^D \mathcal{GIG}(-1/2,\lambda_d/\theta^2_d,\delta_d), \text{ with } \theta_d = (\mathbf{c}_d \tr \bm{\alpha})^{-1} \\
    q(\bm{\sigma}^2) & \overset{D}{=} \prod_{d=1}^D \Gamma^{-1} ((n + p + 1)/2, \zeta_d).
    \end{align*}
    Requires iteratively updating $\bm{\mu}_d$, $\bm{\Sigma}_d$, $\delta_d$, and $\zeta_d$
  \end{frame}

  \begin{frame}
    \frametitle{Full algorithm}
    \begin{columns}
      \begin{column}{0.5\linewidth}
        \begin{algorithm}[H]
        \caption{Empirical-variational Bayes EM}
          \begin{algorithmic}[1]
            \WHILE{EB not converged}
              \WHILE{VB not converged}
                \FOR{$d=1,\dots,D$}
                  \STATE {update $\bm{\mu}_d$, $\bm{\Sigma}_d$, $\delta_d$, and $\zeta_d$}
                \ENDFOR
              \ENDWHILE
              \STATE {update $\bm{\alpha}$, and $\bm{\lambda}$}
            \ENDWHILE
          \end{algorithmic}
        \end{algorithm}
      \end{column}
      \begin{column}{0.5\linewidth}
      \smallskip
      \bigskip \\
      $\Bigg\}$ E-step \\
      \bigskip
      \medskip
      M-step
      \end{column}
    \end{columns}
  \end{frame}

  \begin{frame}
    \frametitle{Output}
    \textbf{Hyperparameters} $\hat{\lambda}_d$ (scale) and $\hat{\theta}_d=(\mathbf{c}_d \tr \hat{\bm{\alpha}})^{-1}$ (location)
    \begin{itemize}
      \item adaptively estimated (non-informative if bad prior info)
      \item indication of important drugs
      \item indication of important drug `covariates'
    \end{itemize}
    \begin{columns}
      \begin{column}{0.45\linewidth}
<<eb_effect1, include=TRUE, echo=FALSE, fig.align="center", out.width="80%", fig.asp=2/3>>=
colors <- sp::bpy.colors(4)[-c(1, 4)]
curve(dinvgauss(x, 1, 1), 0, 5, col=colors[1], lwd=4, main="", yaxt="n", xaxt="n", bty="n", ann=FALSE)
curve(dinvgauss(x, 5, 3), 0, 5, add=TRUE, col=colors[2], lwd=4, main="", yaxt="n", xaxt="n", bty="n", ann=FALSE)
title(xlab=expression(gamma[d]^2), cex=10, cex.lab=3)
legend("topright", legend=c(expression(lambda[d]~","~theta[d]~"small"), expression(lambda[d]~","~theta[d]~"large")), lty=1, col=colors, lwd=4, cex=3, box.col=NA)
@
      \end{column}
      \begin{column}{0.1\linewidth}
        $$\Rightarrow$$
      \end{column}
      \begin{column}{0.45\linewidth}
<<eb_effect2, include=TRUE, echo=FALSE, fig.align="center", out.width="80%", fig.asp=2/3>>=
colors <- sp::bpy.colors(4)[-c(1, 4)]
curve(dnorm(x, 0, 1), -7, 7, col=colors[1], lwd=4, main="", yaxt="n", xaxt="n", bty="n", ann=FALSE)
curve(dnorm(x, 0, 3), -7, 7, col=colors[2], lwd=4, add=TRUE, main="", yaxt="n", xaxt="n", bty="n", ann=FALSE)
title(xlab=expression(beta[jd]^2~"|"~gamma[d]^2), cex=10, cex.lab=3)
@
      \end{column}
    \end{columns}
    \textbf{Posterior} parameters $\bm{\mu}_d$ and $\bm{\Sigma}_d$
    \begin{itemize}
      \item predict future cell line data
      \item feature selection (with caution)
    \end{itemize}
  \end{frame}

  \section{Simulations}
  \begin{frame}
    \frametitle{Motivation}
    \textbf{How well does the EB estimation work?} \\ 
    We consider a simple setting with $K$ classes of drugs:
    $$
	  c_{dk} = \begin{cases}
	    1 & \text{if } \text{class}_d = k, \\
	    0 & \text{otherwise}.
	  \end{cases}
	  $$
	  and compare to
	  \begin{itemize}
	    \item `default' inverse Gamma prior (scaled to match inverse Gaussian):
        $$
        \gamma_d^{2} \sim \Gamma^{-1} (\eta_d/2, \lambda_d/2) = \Gamma^{-1} (\eta_{\text{class}_d}/2, \lambda_{\text{class}_d}/2).
        $$
      \item non-conjugate models $p(\beta_{jd}|\sigma_d^2,\gamma_d^2)=(\beta_{jd}|\gamma_d^2)$
    \end{itemize}
  \end{frame}

  \section{Simulations}
  \begin{frame}
    \frametitle{Setup}
    We use $n=100$, $p=100$, $D=100$, $\text{class}=4$, $\forall d: \lambda_d,\sigma_d^2=1$, $x_{ij} \overset{ind}{\sim} \mathcal{N}(0,1)$
    and simulate from
    \begin{enumerate}
      \item inv. Gaussian model (correct) with $\bm{\alpha} =\begin{bmatrix} 1 & 2 & 3 & 4 \end{bmatrix} \tr$ and $\theta_d =(\mathbf{c}_d \tr \bm{\alpha})^{-1}$
      \item inv. Gamma model (incorrect) with $\bm{\alpha} = \begin{bmatrix} 3 & 4 & 5 & 6 \end{bmatrix} \tr$ and $\eta_d =\mathbf{c}_d \tr \bm{\alpha}$
    \end{enumerate}
  \end{frame}

  \begin{frame}
    \frametitle{Prior means 1}
<<boxplot_igaussian_res1_prior_mean, cache=FALSE, echo=FALSE, fig.cap="Prior mean estimates", out.width="100%", fig.align="center", fig.asp=2/3>>=
@
  \end{frame}

  \begin{frame}
    \frametitle{Model parameters 1}
<<boxplot_igaussian_res1_mu_mse, cache=FALSE, echo=FALSE, fig.cap="MSE of model parameters", out.width="100%", fig.align="center", fig.asp=2/3>>=
@
  \end{frame}

  \begin{frame}
    \frametitle{Prior means 2}
<<boxplot_igaussian_res2_prior_mean, cache=FALSE, echo=FALSE, fig.cap="Prior mean estimates", out.width="100%", fig.align="center", fig.asp=2/3>>=
@
  \end{frame}

  \begin{frame}
    \frametitle{Model parameters 2}
<<boxplot_igaussian_res2_mu_mse, cache=FALSE, echo=FALSE, fig.cap="MSE of model parameters", out.width="100%", fig.align="center", fig.asp=2/3>>=
@
  \end{frame}

  \begin{frame}
    \frametitle{Convergence issues}
<<lines_igaussian_res2_igauss_conj_convergence, cache=FALSE, echo=FALSE, fig.cap="a) lambda, b) prior mean, and c) prior variance estimates", out.width="50%", fig.align="center">>=
@
  \end{frame}

  \section{Future}
  \begin{frame}
    \frametitle{Future work}
    Lots to do
    \begin{itemize}
      \item implement tissue effects
      \item solve convergence issues
      \item more simulations
      \item investigate real data (GDSC and CCLE)
      \item write award-winning nature/science paper
    \end{itemize}
  \end{frame}

  \begin{frame}
    \frametitle{Thank you}
    \centering {\usebeamercolor[fg]{structure} \textbf{Material}} \\
    \textit{Paper:} Working progress... \\
    \textit{R-package:} Working progress on https://github.com/magnusmunch/cambridge
    \bigskip

    \centering {\usebeamercolor[fg]{structure} \textbf{Website}} \\
    \textit{Group:} https://www.bigstatistics.nl/ \\
    \textit{Personal:} http://pub.math.leidenuniv.nl/~munchmm/
    \bigskip

    \centering {\usebeamercolor[fg]{structure} \textbf{Acknowledgements}} \\
    This research has received funding from the European Research Council under ERC Grant Agreement 320637
  \end{frame}

\end{document}

### 2019-01-28 docs/supplement.Rnw
first version of presentation added

@@
 % load packages and set options
-<<include=FALSE>>=
+<<settings, include=FALSE, echo=FALSE>>=
 library(cambridge)
@@
 % read figure code
-<<figures, include=FALSE>>=
+<<figures, include=FALSE, echo=FALSE>>=
 read_chunk("../code/figures.R")
@@
 To account for the random data and parameter generation, we repeat the simulation 100 times.
-We compare the results to the models in \ref{sec:indvarmodel} and \ref{sec:invgammamodel}, and present the results in Figures \ref{fig:boxplot_igaussian_res1_sigma2}-\ref{fig:boxplot_igaussian_res1_alpha}.
+We compare the results to the models in \ref{sec:indvarmodel} and \ref{sec:invgammamodel}, and present the results in Figures \ref{fig:boxplot_igaussian_res1_prior_mean}-\ref{fig:boxplot_igaussian_res2_prior_mean}.
 
-<<boxplot_igaussian_res1_sigma2, cache=FALSE, echo=FALSE, fig.cap="MSEs for posterior means of $\\sigma_d^2$ in simulation 1", out.width="50%", fig.align="center">>=
+<<boxplot_igaussian_res1_prior_mean, cache=FALSE, echo=FALSE, fig.cap="Prior mean estimates", out.width="80%", fig.align="center", fig.asp=2/3>>=
 @
 
-<<boxplot_igaussian_res1_gamma2, cache=FALSE, echo=FALSE, fig.cap="a) MSE and b) correlation for posterior means of $\\gamma_d^2$ in simulation 1", fig.align="center">>=
-@
-
-<<boxplot_igaussian_res1_mu, cache=FALSE, echo=FALSE, fig.cap="a) MSE and b) correlation for posterior means of $\\beta_d$ in simulation 1", fig.align="center">>=
-@
-
-<<boxplot_igaussian_res1_theta, cache=FALSE, echo=FALSE, fig.cap="a) MSE and b) correlation for prior means of $\\gamma_d^2$ ($\\theta_d$) in simulation 1", fig.align="center">>=
-@
-
-<<boxplot_igaussian_res1_alpha, cache=FALSE, echo=FALSE, fig.cap="a) MSE and b) correlation for $\\alpha$ in simulation 1", fig.align="center">>=
-@
-
-<<lines_igaussian_res1_elbo, cache=FALSE, echo=FALSE, fig.cap="Convergence of ELBO for a) inverse Gaussian, b) independent inverse Gaussian, and c) inverse Gamma models in simulation 1", fig.align="center">>=
-@
-
-<<lines_igaussian_res1_theta, cache=FALSE, echo=FALSE, fig.cap="Convergence of prior means for a) inverse Gaussian, b) independent inverse Gaussian, and c) inverse Gamma models in simulation 1", fig.align="center">>=
-@
-
-Simulations with no difference between classes:
-
-<<boxplot_igaussian_res2_theta, cache=FALSE, echo=FALSE, fig.cap="a) MSE and b) correlation for prior means of $\\gamma_d^2$ ($\\theta_d$) in simulation 2", out.width="50%", fig.align="center">>=
-@
-
-<<lines_igaussian_res2_alpha, cache=FALSE, echo=FALSE, fig.cap="Convergence of $\\alpha$ for a) inverse Gaussian and b) independent inverse Gaussian in simulation 2", fig.align="center">>=
+<<boxplot_igaussian_res2_prior_mean, cache=FALSE, echo=FALSE, fig.cap="Prior mean estimates", out.width="80%", fig.align="center", fig.asp=2/3>>=
 @

### 2019-01-28 code/figures.R

@@ -1,233 +1,215 @@
+################################## supplement ##################################
 # ---- figures ----
-# ---- boxplot_igaussian_res1_sigma2 ----  
-res1 <- read.table("../results/simulations_igaussian_res1.csv")
-set1 <- read.table("../results/simulations_igaussian_set1.csv")
-names1 <- c("inv Gauss", "ind inv Gauss", "inv Gamma")
-temp1 <- list(res1[, substr(colnames(res1), 1, 14)=="zeta.inv.Gauss"],
-              res1[, substr(colnames(res1), 1, 18)=="zeta.ind.inv.Gauss"],
-              res1[, substr(colnames(res1), 1, 5)=="cpost"],
-              res1[, substr(colnames(res1), 1, 5)=="dpost"])
-temp2 <- cbind(colMeans((2*temp1[[1]]/(set1$n + set1$p - 1) - as.numeric(set1[, substr(colnames(set1), 1, 5)=="sigma"])^2)^2),
-               colMeans((2*temp1[[2]]/(set1$n - 1) - 
-                           as.numeric(set1[, substr(colnames(set1), 1, 5)=="sigma"])^2)^2),
-               colMeans((temp1[[4]]/(temp1[[3]] - 1) - 
-                           as.numeric(set1[, substr(colnames(set1), 1, 5)=="sigma"])^2)^2))
-par(cex=1.3)
-boxplot(temp2, outline=FALSE, names=names1, ylab="MSE")
-legend("topright", title=expression(bold("median MSE")), bty="n", 
-       paste(names1, round(apply(temp2, 2, median), 2), sep="="))
-par(cex=1)
+# ---- boxplot_igaussian_res1_prior_mean ----  
+res1 <- as.matrix(read.table("results/simulations_igaussian_res1.csv"))
+set1 <- read.table("results/simulations_igaussian_set1.csv")
+labels1 <- c("c. inv. \n Gaussian", "nc. inv. \n Gaussian",
+              "c. inv. \n Gamma", "nc. inv. \n Gamma")
+colors1 <- sp::bpy.colors(6)[-c(1, 6)]
 
-# ---- boxplot_igaussian_res1_gamma2 ----
-res1 <- read.table("../results/simulations_igaussian_res1.csv")
-set1 <- read.table("../results/simulations_igaussian_set1.csv")
-names1 <- c("inv Gauss", "ind inv Gauss", "inv Gamma")
[…]

### 2019-01-28 code/simulations_igaussian.R

@@ -1,282 +1,224 @@
 #!/usr/bin/env Rscript
 
-### installation of package
+### installation of packages
 # if(!("devtools" %in% installed.packages())) {
 #   install.packages("devtools")
 # }
 # library(devtools)
 # install_github("magnusmunch/cambridge/rpackage", local=FALSE,
 #                auth_token=Sys.getenv("GITHUB_PAT"))
+if(!("statmod" %in% installed.packages())) {
+  install.packages("statmod")
+}
 
 ### libraries
 library(cambridge)
+library(statmod)
 
 ################################# simulation 1 #################################
 # settings
 nreps <- 100
 n <- 100
-p <- 200
-D <- 300
-nclass <- 5
+p <- 100
+D <- 100
+nclass <- 4
 
 # create fixed parameters
 alpha <- c(1:nclass)
-C <- model.matrix(~ 1 + factor(rep(c(1:nclass), each=D/nclass)))
-theta <- as.numeric(1/(C %*% alpha))
-gamma <- sqrt(theta)
+C.inv.gauss <- model.matrix(~ -1 + factor(rep(c(1:nclass), each=D/nclass)))
+C.inv.gamma <- as.numeric(C.inv.gauss %*% c(1:nclass))
+theta <- as.numeric(1/(C.inv.gauss %*% alpha))
+lambda <- rep(1, D)
 sigma <- rep(1, D)
-SNR <- 10
+SNR <- 50
 
 # store settings in an object
-temp <- c(nreps=nreps, n=n, p=p, D=D, nclass=nclass, alpha=alpha, C=C,
-          theta=theta, gamma=gamma, sigma=sigma, SNR=SNR)
-set1 <- t(matrix(temp))
-colnames(set1) <- names(temp)
+temp1 <- c(nreps=nreps, n=n, p=p, D=D, nclass=nclass, alpha=alpha, 
+           C=C.inv.gauss, theta=theta, lambda=lambda, sigma=sigma, SNR=SNR)
+set1 <- t(matrix(temp1))
+colnames(set1) <- names(temp1)
 
 # objects to store simulation results
-list1 <- sapply(c("inv Gauss", "ind inv Gauss"), function(m) {
-  matrix(NA, ncol=D, nrow=nreps,
-         dimnames=list(NULL, paste("drug", c(1:D), sep="")))}, simplify=FALSE)
-list2 <- sapply(c("inv Gamma"), function(m) {
[…]

### 2019-01-28 rpackage/R/functions.R

@@ -59,7 +59,7 @@
 
 # single VB update, drugs covariate model only (tested)
 .single.vb.update <- function(aold, bold, eta, theta, lambda, sv, n, p, uty, 
-                              yty, conjugate, inv.gauss) {
+                              yty, conjugate, hyperprior) {
   
   # variables that change with conjugacy
   cold <- bold/(aold^(1 - conjugate))
@@ -69,25 +69,25 @@
   aux <- .aux.var(cold, aold, sv, uty, n, p)
   
   # vb parameters
-  delta <- (bold/cold)*aux$mutmu + lambda
-  if(inv.gauss) {
-    b <- sqrt(lambda/(theta^2*delta))*
-      ratio_besselK_cpp(sqrt(lambda*delta/theta^2), 0.5*(p + eta)) + 
-      (p + eta)/delta
+  delta <- unname((bold/cold)*aux$mutmu + lambda)
+  if(hyperprior=="inv. Gaussian") {
+    b <- unname(sqrt(lambda/(theta^2*delta))*
+                  ratio_besselK(sqrt(lambda*delta/theta^2), 
+                                     0.5*(p + eta)) + (p + eta)/delta)
     
     # needed in EB step
     e <- (b - (p + eta)/delta)*delta*theta^2/lambda
   } else {
-    b <- (p + eta)/delta
+    b <- unname((p + eta)/delta)
     e <- log(delta) - digamma(0.5*(p + eta)) - log(2)
   }
-  zeta <- 0.5*(yty - 2*aux$ytXmu + aux$trXtXSigma + aux$mutXtXmu + 
-                 ifelse(conjugate, b, 0)*(aux$trSigma + aux$mutmu))
-  a <- (df + 1)/(2*zeta)
+  zeta <- unname(0.5*(yty - 2*aux$ytXmu + aux$trXtXSigma + aux$mutXtXmu + 
+                        ifelse(conjugate, b, 0)*(aux$trSigma + aux$mutmu)))
+  a <- unname((df + 1)/(2*zeta))
   c <- b/(a^(1 - conjugate))
   
   # calculate the elbo separately for the inverse Gaussian and Gamma models
-  if(inv.gauss) {
+  if(hyperprior=="inv. Gaussian") {
     elbo <- .elbo.inv.gauss(aux, zeta, delta, a, b, c, e, lambda, theta, df, p, 
                             yty)
   } else {
@@ -96,8 +96,8 @@
   }
   
   
[…]

### 2019-01-28 rpackage/src/RcppExports.cpp

@@ -6,98 +6,21 @@
 
 using namespace Rcpp;
 
-// ratio_besselK_cpp
-NumericVector ratio_besselK_cpp(arma::vec x, double nu);
-RcppExport SEXP _cambridge_ratio_besselK_cpp(SEXP xSEXP, SEXP nuSEXP) {
+// ratio_besselK
+NumericVector ratio_besselK(arma::vec x, arma::vec nu);
+RcppExport SEXP _cambridge_ratio_besselK(SEXP xSEXP, SEXP nuSEXP) {
 BEGIN_RCPP
     Rcpp::RObject rcpp_result_gen;
     Rcpp::RNGScope rcpp_rngScope_gen;
     Rcpp::traits::input_parameter< arma::vec >::type x(xSEXP);
-    Rcpp::traits::input_parameter< double >::type nu(nuSEXP);
-    rcpp_result_gen = Rcpp::wrap(ratio_besselK_cpp(x, nu));
-    return rcpp_result_gen;
-END_RCPP
-}
-// vb_update_cpp
-List vb_update_cpp(arma::vec zetaold, arma::vec aold, double lambda, arma::vec theta, arma::mat x, arma::mat xxt, arma::mat xtx, arma::mat y);
-RcppExport SEXP _cambridge_vb_update_cpp(SEXP zetaoldSEXP, SEXP aoldSEXP, SEXP lambdaSEXP, SEXP thetaSEXP, SEXP xSEXP, SEXP xxtSEXP, SEXP xtxSEXP, SEXP ySEXP) {
-BEGIN_RCPP
-    Rcpp::RObject rcpp_result_gen;
-    Rcpp::RNGScope rcpp_rngScope_gen;
-    Rcpp::traits::input_parameter< arma::vec >::type zetaold(zetaoldSEXP);
-    Rcpp::traits::input_parameter< arma::vec >::type aold(aoldSEXP);
-    Rcpp::traits::input_parameter< double >::type lambda(lambdaSEXP);
-    Rcpp::traits::input_parameter< arma::vec >::type theta(thetaSEXP);
-    Rcpp::traits::input_parameter< arma::mat >::type x(xSEXP);
-    Rcpp::traits::input_parameter< arma::mat >::type xxt(xxtSEXP);
-    Rcpp::traits::input_parameter< arma::mat >::type xtx(xtxSEXP);
-    Rcpp::traits::input_parameter< arma::mat >::type y(ySEXP);
[…]

### 2019-01-28 rpackage/tests/testthat/test-igaussian.R

@@ -180,17 +180,18 @@ test_that("low dim. conjugate inverse Gaussian VB update", {
   cold <- bold; x <- matrix(rnorm(p*n), nrow=n, ncol=p); y <- matrix(rnorm(n))
   svd.x <- svd(x); sv <- svd.x$d; uty <- t(svd.x$u) %*% y
   yty <- as.numeric(t(y) %*% y); eta <- rchisq(1, 2); lambda <- rchisq(1, 2)
-  theta <- rchisq(1, 1); conjugate <- TRUE; inv.gauss <- TRUE; df <- n + p
+  theta <- rchisq(1, 1); conjugate <- TRUE; hyperprior <- "inv. Gaussian"
+  df <- n + p
 
   new.vb <- .single.vb.update(aold, bold, eta, theta, lambda, sv, n, p, uty,
-                              yty, conjugate, inv.gauss)
+                              yty, conjugate, hyperprior)
 
   # the actual calculation are this (see manuscript):
   Sigma <- solve(t(x) %*% x + cold*diag(p))/aold
   mu <- solve(t(x) %*% x + cold*diag(p)) %*% t(x) %*% y
   delta <- c(delta=(bold/cold)*as.numeric(t(mu) %*% mu) + lambda)
   b <- sqrt(lambda/(theta^2*delta))*
-    ratio_besselK_cpp(sqrt(lambda*delta)/theta, 0.5*(p + eta)) + (p + eta)/delta
+    ratio_besselK(sqrt(lambda*delta)/theta, 0.5*(p + eta)) + (p + eta)/delta
   zeta <- 0.5*(yty - 2*t(y) %*% x %*% mu + sum(diag(t(x) %*% x %*% Sigma)) +
                  t(mu) %*% t(x) %*% x %*% mu + 
                  b*(sum(diag(Sigma)) + t(mu) %*% mu))
@@ -211,17 +212,18 @@ test_that("high dim. conjugate inverse Gaussian VB update", {
   cold <- bold; x <- matrix(rnorm(p*n), nrow=n, ncol=p); y <- matrix(rnorm(n))
   svd.x <- svd(x); sv <- svd.x$d; uty <- t(svd.x$u) %*% y
   yty <- as.numeric(t(y) %*% y); eta <- rchisq(1, 2); lambda <- rchisq(1, 2)
-  theta <- rchisq(1, 1); conjugate <- TRUE; inv.gauss <- TRUE; df <- n + p
[…]

### 2019-01-30 docs/presentation.Rnw
BSU together 2019 talk version

@@
+% document settings
 \documentclass[10pt,letterpaper]{beamer} %
@@
 
-%\usepackage{footnote}
-%\makesavenoteenv{figure}
-
-\usepackage{amsmath,verbatim,bm,caption,soul,graphicx}
+% loading packages
+\usepackage{amsmath,verbatim,bm,caption,soul,graphicx,url}
 \usepackage[plain]{algorithm}
@@
 \usepackage[american]{babel}
-\usepackage[backend=bibtex,style=verbose]{biblatex}
+\usepackage[backend=bibtex,citestyle=authoryear]{biblatex}
+\usepackage[outdir=./]{epstopdf}
+% \usepackage[outdir=../figs/]{epstopdf}
 
+% beamer settings
 \addtobeamertemplate{block begin}{\setlength\abovedisplayskip{0pt}}
@@
     \hspace*{0.1cm}\par\vskip 2pt}{}
-
 \setbeamerfont{caption}{size=\scriptsize}
-
 \makeatletter
@@
 
-\author[Magnus M\"unch]{Magnus M\"unch \\[3mm] \footnotesize{Joint work with Mark A. van de Wiel, Sylvia Richardson, and Gwena{\"e}l G. R. Leday}}
-\title[EB for drug efficacy prediction]{Emprical Bayes for drug efficacy prediction in cell lines}
-\institute[LU \& AUMC]{Leiden University and Amsterdam UMC}
-\date{\today} % Date, can be changed to a custom date
-\bibliography{refs}
+% content settings
+\author[Magnus M\"unch]{Magnus M. M\"unch \\[3mm] \footnotesize{Joint work with Mark A. van de Wiel, Sylvia Richardson, and Gwena{\"e}l G. R. Leday}}
+\title[EB for drug efficacy prediction]{Empirical Bayes for drug efficacy prediction in cell lines}
+\institute[LU, AUMC, \& MRC BSU]{Leiden University, Amsterdam UMC, and MRC Biostatistics Unit}
+\date{\today}
+\bibliography{refs} % unfortunately custom bibliography doesn't work with biblatex
 
@@
 		  \raisebox{-0.5\height}{\includegraphics[width=0.3\linewidth]{../figs/Amsterdam_UMC_UK_Logo_CMYK_U_150_dpi.jpg}}
+		  \raisebox{-0.5\height}{\includegraphics[width=0.4\linewidth,trim={0 0.5cm 14cm 0},clip]{../figs/MRC_BSU_Cambridge_colour_profprint_A4.eps}}
 	  \end{figure}
@@
     \begin{figure}[h]
-      \centering
-			\includegraphics[width=1\linewidth]{../figs/{F2.large}.jpg}
+      {\centering \includegraphics[width=1\linewidth]{../figs/{F2.large}.jpg}}
+			{\scriptsize{Source:~\cite{goodspeed_tumor-derived_2016}}}
 		\end{figure}
@@
 @
+		        \begin{figure}[h]
+              \centering
+			        \includegraphics[width=0.9\linewidth]{../figs/FDA_approval.png}
+		        \end{figure}
 		      \end{column}
@@
 		        \begin{figure}[h]
-              \centering
-			        \includegraphics[width=1\linewidth]{../figs/300px-SMILES.png}
+              {\centering \includegraphics[width=1\linewidth]{../figs/300px-SMILES.png}}
+			        {\scriptsize{Source:~\cite{noauthor_simplified_2019}}}
 		        \end{figure}
@@
 		\begin{align*}
-		  \beta_{jd} & \sim \mathcal{N}_p (0, \gamma_d^2 \sigma_d^2), \\
+		  \beta_{jd} & \sim \mathcal{N} (0, \gamma_d^2 \sigma_d^2), \\
 		  \only<1>{\mathbf{u}_{d} & \sim \mathcal{N}_T (\mathbf{0}, \bm{\Xi}_d \sigma_d^2),}
@@
           \item Be objective
-          \item Incorporate extra drug information $\mathbf{c}_d$
+          \item Incorporate drug information $\mathbf{c}_d$
         \end{itemize}
@@
       \begin{column}{0.5\linewidth}
-        \only<2>{\centering $\Rightarrow$ \textbf{Empirical Bayes} (EB) \\
-        Let $\theta_d = (\mathbf{c}_d \tr \bm{\alpha})^{-1}$ \\
-        $\Downarrow$ \\
-        estimate $\bm{\alpha}$ and $\lambda_d$
-        }
+
       \end{column}
     \end{columns}
+    \bigskip
+    \only<2>{
+      \textbf{Empirical Bayes} (EB) \\
+      Let $\theta_d = \E_{\gamma_d^2}(\gamma_d^2) = (\mathbf{c}_d \tr \bm{\alpha})^{-1}$ $\Rightarrow$ estimate $\bm{\alpha}$ and $\lambda_d$ \\
+      \bigskip
+      \textbf{Note:} smaller dimension of $\bm{\alpha}$ simplifies estimation of $\theta_d$
+    }
   \end{frame}
@@
     \frametitle{Empirical Bayes}
-    \textbf{Cannonical EB}: Maximise the marginal likelihood (MML)
+    \textbf{Canonical EB}: Maximise the marginal likelihood (MML)
     \begin{itemize}
@@
     \end{itemize}
+    \only<1>{
+      \bigskip
+      $$
+      \text{ML}(\bm{\alpha})=\int_{\bm{\beta}} \mathcal{L} (\y | \bm{\beta}) \pi_{\bm{\alpha}}(\bm{\beta}) d \bm{\beta}
+      $$
+    }
     \only<2>{
@@
 title(xlab=expression(gamma[d]^2), cex=10, cex.lab=3)
-legend("topright", legend=c(expression(lambda[d]~","~theta[d]~"small"), expression(lambda[d]~","~theta[d]~"large")), lty=1, col=colors, lwd=4, cex=3, box.col=NA)
+legend("topright", legend=c(expression(hat(lambda)[d]~","~hat(theta)[d]~"small"), expression(hat(lambda)[d]~","~hat(theta)[d]~"large")), lty=1, col=colors, lwd=4, cex=3, box.col=NA)
 @
@@
     \end{columns}
-    \textbf{Posterior} parameters $\bm{\mu}_d$ and $\bm{\Sigma}_d$
+    \textbf{Posterior} $p(\bbeta_d | \y)$ parameters $\bm{\mu}_d$ and $\bm{\Sigma}_d$
     \begin{itemize}
@@
         $$
-      \item non-conjugate models $p(\beta_{jd}|\sigma_d^2,\gamma_d^2)=(\beta_{jd}|\gamma_d^2)$
+      \item non-conjugate models $p(\beta_{jd}|\sigma_d^2,\gamma_d^2)=p(\beta_{jd}|\gamma_d^2)$ \\
+        (conjugate model: \textit{a priori} dependence of $\beta_{jd}$ on $\sigma_d^2$)    
     \end{itemize}
@@
 
-  \section{Simulations}
   \begin{frame}
@@
     We use $n=100$, $p=100$, $D=100$, $\text{class}=4$, $\forall d: \lambda_d,\sigma_d^2=1$, $x_{ij} \overset{ind}{\sim} \mathcal{N}(0,1)$
-    and simulate from
+    and simulate $\gamma_d^2$ from
     \begin{enumerate}
@@
     \end{enumerate}
+    \bigskip
+    \textbf{Note:} results in same $\theta_d$
   \end{frame}
@@
 
-  \section{Future}
+  \section{Outlook}
   \begin{frame}
@@
     \begin{itemize}
-      \item implement tissue effects
       \item solve convergence issues
       \item more simulations
+      \item implement tissue effects
       \item investigate real data (GDSC and CCLE)
@@
     \textit{Paper:} Working progress... \\
-    \textit{R-package:} Working progress on https://github.com/magnusmunch/cambridge
+    \textit{R-package (devel):} \url{https://github.com/magnusmunch/cambridge}
     \bigskip
@@
     \centering {\usebeamercolor[fg]{structure} \textbf{Website}} \\
-    \textit{Group:} https://www.bigstatistics.nl/ \\
-    \textit{Personal:} http://pub.math.leidenuniv.nl/~munchmm/
+    \textit{Group:} \url{https://www.bigstatistics.nl/} \\
+    \textit{Personal:} \url{http://pub.math.leidenuniv.nl/~munchmm/}
     \bigskip
@@
 
+  \begin{frame}{Bibliography}
+    \frametitle{References}
+	  \printbibliography
+	\end{frame}
+
 \end{document}

### 2019-01-30 docs/supplement.Rnw
BSU together 2019 talk version

@@
 opts_knit$set(base.dir="../figs")
-opts_chunk$set(fig.align='center', echo=FALSE, fig.path="../figs/")
+opts_chunk$set(fig.align='center', echo=FALSE, fig.path="../figs/", cache=FALSE)
 @
@@
 % read figure code
-<<figures, include=FALSE, echo=FALSE>>=
+<<figures, include=FALSE>>=
 read_chunk("../code/figures.R")
@@
 
-<<boxplot_igaussian_res1_prior_mean, cache=FALSE, echo=FALSE, fig.cap="Prior mean estimates", out.width="80%", fig.align="center", fig.asp=2/3>>=
+<<boxplot_igaussian_res1_prior_mean, fig.cap="Prior mean estimates", out.width="80%", fig.asp=2/3>>=
 @
 
-<<boxplot_igaussian_res2_prior_mean, cache=FALSE, echo=FALSE, fig.cap="Prior mean estimates", out.width="80%", fig.align="center", fig.asp=2/3>>=
+<<boxplot_igaussian_res2_prior_mean, fig.cap="Prior mean estimates", out.width="80%", fig.asp=2/3>>=
 @
@@
   where $F(x; 3/2, 1)$ is the CDF of a gamma function with shape $3/2$ and scale $1$.
-In \cite{moran_variance_2018}, REFERENCE to Ročková and George (2018), and \cite{chipman_bart:_2010}, $q=0.9$ is suggested.
+In \cite{moran_variance_2018}, REFERENCE to Rockova and George (2018), and \cite{chipman_bart:_2010}, $q=0.9$ is suggested.
 We find that $q \in ( 0.9, 0.95 )$ gives good results.

### 2019-01-30 docs/manuscript.Rnw
VB update now generic
investigating convergence issues

@@
 		\begin{align}
-		\beta_{jd} & \sim \mathcal{N}_p (0, \phi^2_{g(j)} \gamma_d^2 \sigma_d^2), \\
+		\beta_{jd} & \sim \mathcal{N}_p (0, \phi^2_{g(j)} \gamma_d^2 \sigma_d^2), \label{eq:betaprior}\\
 		\mathbf{u}_{d} & \sim \mathcal{N}_T (\mathbf{0}, \bm{\Xi}_d \sigma_d^2),
@@
 
+	Recently, \cite{moran_variance_2018} have argued that in high-dimensional regression problems, \text{a priori} dependence between the regression parameters and error variance may induce biased estimation.
+In (\ref{eq:betaprior}), we assume such an \textit{a priori} dependence.
+However, we contend that independent regression parameters and error variance are not desirable in our setting for two reasons: (a) the empirical Bayes estimation requires some notion of scale to combine the separate regressions, which may be done by inclusion of the error variance in the regression parameter prior model.
+(b) The variational Bayes approximation introduced in Section \ref{sec:variationalbayes} gives a posterior that is similar to the one we would obtain with a dependent regression parameters and error variance prior model.
+This renders a better approximate expectation in the empirical Bayes iterations.
+
 	\bibliographystyle{author_short3} 

### 2019-01-30 code/simulations_igaussian.R

@@ -132,7 +132,7 @@ C.inv.gamma <- as.numeric(C.inv.gauss %*% c(1:nclass))
 eta <- as.numeric(C.inv.gauss %*% alpha)
 lambda <- rep(1, D)
 sigma <- rep(1, D)
-SNR <- 50
+SNR <- 3
 
 # store settings in an object
 temp1 <- c(nreps=nreps, n=n, p=p, D=D, nclass=nclass, alpha=alpha, 
@@ -157,7 +157,7 @@ res2 <- as.data.frame(
 # set initial values and control parameters
 control <- list(epsilon.eb=-1, epsilon.vb=-1, 
                 epsilon.opt=sqrt(sqrt(.Machine$double.eps)), 
-                maxit.eb=20, maxit.vb=2, maxit.opt=100, trace=FALSE)
+                maxit.eb=30, maxit.vb=4, maxit.opt=100, trace=FALSE)
 init.inv.gauss <- list(a=rep(mean(1/sigma^2), D), 
                        b=rep(mean(lambda/(eta - 2)), D), 
                        theta=rep(mean(lambda/(eta - 2)), D), lambda=rep(1, D))
@@ -222,3 +222,51 @@ for(r in 1:nreps) {
 }
 
 warnings()
+
+
+opar <- par(no.readonly=TRUE)
+layout(matrix(c(rep(rep(c(1:3), each=2), 2), rep(c(0, 4, 4, 5, 5, 0), 2)), 
+              byrow=TRUE, nrow=4, ncol=6), widths=1, heights=1, respect=TRUE)
+par(cex=1.3, mar=opar$mar*c(1/2, 1.1, 1/2, 1/2))
+plot(fit2.igauss.conj$seq.eb$alpha[, 1], type="l", xlab="iteration", main="a)",
+     ylab=expression(hat(alpha)[c]), ylim=range(fit2.igauss.conj$seq.eb$alpha))
+for(d in 2:ncol(fit2.igauss.conj$seq.eb$alpha)) {
+  lines(fit2.igauss.conj$seq.eb$alpha[, d], col=d)
+}
+plot(fit2.igauss.conj$seq.eb$theta[, 1], type="l", xlab="iteration", main="b)",
+     ylab=expression(hat(theta)[d]), ylim=range(fit2.igauss.conj$seq.eb$theta))
[…]

### 2019-02-19 docs/manuscript.Rnw
changed R code chuck options manuscript.Rnw and supplement.Rnw

@@
+% load packages
+<<packages, include=FALSE, echo=FALSE>>=
+library(cambridge)
+@
+
+% set options
+<<settings, include=FALSE, echo=FALSE>>=
+opts_knit$set(root.dir="..", base.dir="../figs")
+opts_chunk$set(fig.align='center', fig.path="../figs/", 
+               echo=FALSE, cache=FALSE, message=FALSE)
+@
+
+% read figure code
+<<figures, include=FALSE>>=
+read_chunk("code/figures.R")
+@
+
 \documentclass[a4paper,hidelinks]{article}
@@
 \begin{document}
+
 	\maketitle
@@
 
-	- drugs information: molecular target, drug class, etc (ask Oslo).
+	- drugs information: molecular target, drug class, etc (ask Oslo), pubchem, omnipath DB, pharmaco DB
 

### 2019-02-19 docs/supplement.Rnw
changed R code chuck options manuscript.Rnw and supplement.Rnw

@@
+% load packages
+<<packages, include=FALSE, echo=FALSE>>=
+library(cambridge)
+@
+
+% set options
+<<settings, include=FALSE, echo=FALSE>>=
+opts_knit$set(root.dir="..", base.dir="../figs")
+opts_chunk$set(fig.align='center', fig.path="../figs/", 
+               echo=FALSE, cache=FALSE, message=FALSE)
+@
+
+% read figure code
+<<figures, include=FALSE>>=
+read_chunk("code/figures.R")
+@
+
 \documentclass[a4paper,hidelinks]{article}
-\usepackage{bm,amsmath,amssymb,amsthm,amsfonts,graphics,graphicx,epsfig,rotating,caption,subcaption,natbib,appendix,titlesec,multicol,hyperref,verbatim,bbm,algorithm,algpseudocode,pgfplotstable,threeparttable, booktabs,mathtools,dsfont,parskip}
+\usepackage{bm,amsmath,amssymb,amsthm,amsfonts,graphics,graphicx,epsfig,
+rotating,caption,subcaption,natbib,appendix,titlesec,multicol,hyperref,verbatim,
+bbm,algorithm,algpseudocode,pgfplotstable,threeparttable,booktabs,mathtools,
+dsfont,parskip}
 
@@
 \begin{document}
-
-% load packages and set options
-<<settings, include=FALSE, echo=FALSE>>=
-library(cambridge)
-opts_knit$set(base.dir="../figs")
-opts_chunk$set(fig.align='center', echo=FALSE, fig.path="../figs/", cache=FALSE)
-@
-
-% read figure code
-<<figures, include=FALSE>>=
-read_chunk("../code/figures.R")
-@
 

### 2019-04-17 docs/manuscript1.Rnw
added mcmc version of inverse Gaussian model in stan

% set options
<<settings, include=FALSE, echo=FALSE>>=
opts_knit$set(root.dir="..", base.dir="../figs")
opts_chunk$set(fig.align='center', fig.path="../figs/", 
               echo=FALSE, cache=FALSE, message=FALSE, fig.pos='!ht')
knit_hooks$set(document=function(x) {
  sub('\\usepackage[]{color}', '\\usepackage{xcolor}', x, fixed=TRUE)})
@

% read figure code
<<figures, include=FALSE>>=
read_chunk("code/figures.R")
@

\documentclass[a4paper,hidelinks]{article}
\usepackage{bm,amsmath,amssymb,amsthm,amsfonts,graphics,graphicx,epsfig,
rotating,caption,subcaption,natbib,appendix,titlesec,multicol,hyperref,verbatim,
bbm,algorithm,algpseudocode,pgfplotstable,threeparttable,booktabs,mathtools,
dsfont,parskip,xr}
\externaldocument[sm-]{supplement1}

% vectors and matrices
\newcommand{\bbeta}{\bm{\beta}}
\newcommand{\btau}{\bm{\tau}}
\newcommand{\bsigma}{\bm{\sigma}}
\newcommand{\balpha}{\bm{\alpha}}
\newcommand{\bepsilon}{\bm{\epsilon}}
\newcommand{\bomega}{\bm{\omega}}
\newcommand{\bOmega}{\bm{\Omega}}
\newcommand{\bSigma}{\bm{\Sigma}}
\newcommand{\bkappa}{\bm{\kappa}}
\newcommand{\btheta}{\bm{\theta}}
\newcommand{\bmu}{\bm{\mu}}
\newcommand{\bpsi}{\bm{\psi}}
\newcommand{\blambda}{\bm{\lambda}}
\newcommand{\bLambda}{\bm{\Lambda}}
\newcommand{\bPsi}{\bm{\Psi}}
\newcommand{\bphi}{\bm{\phi}}
\newcommand{\Y}{\mathbf{Y}}
\newcommand{\y}{\mathbf{y}}
\newcommand{\X}{\mathbf{X}}
\newcommand{\x}{\mathbf{x}}
\newcommand{\I}{\mathbf{I}}
\newcommand{\bgamma}{\bm{\gamma}}
\newcommand{\T}{\mathbf{T}}
\newcommand{\Z}{\mathbf{Z}}
\newcommand{\W}{\mathbf{W}}
\renewcommand{\O}{\mathbf{O}}
\newcommand{\B}{\mathbf{B}}
\renewcommand{\H}{\mathbf{H}}
\newcommand{\U}{\mathbf{U}}
\newcommand{\Em}{\mathbf{E}}
\newcommand{\D}{\mathbf{D}}
\newcommand{\Vm}{\mathbf{V}}
\newcommand{\rv}{\mathbf{r}}
\newcommand{\A}{\mathbf{A}}
\newcommand{\Q}{\mathbf{Q}}
\newcommand{\Sm}{\mathbf{S}}
\newcommand{\0}{\bm{0}}
\newcommand{\tx}{\tilde{\mathbf{x}}}
\newcommand{\hy}{\hat{y}}
\newcommand{\tm}{\tilde{m}}
\newcommand{\tkappa}{\tilde{\kappa}}
\newcommand{\m}{\mathbf{m}}
\newcommand{\C}{\mathbf{C}}
\renewcommand{\v}{\mathbf{v}}
\newcommand{\g}{\mathbf{g}}
\newcommand{\s}{\mathbf{s}}
\renewcommand{\c}{\mathbf{c}}
\newcommand{\ones}{\mathbf{1}}
\newcommand{\R}{\mathbf{R}}
\newcommand{\rvec}{\mathbf{r}}
\newcommand{\Lm}{\mathbf{L}}

% functions and operators
\renewcommand{\L}{\mathcal{L}}
\newcommand{\G}{\mathcal{G}}
\renewcommand{\P}{\mathcal{P}}
\newcommand{\argmax}{\text{argmax} \,}
\newcommand{\expit}{\text{expit}}
\newcommand{\erfc}{\text{erfc}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\V}{\mathbb{V}}
\newcommand{\Cov}{\mathbb{C}\text{ov}}
\newcommand{\tr}{^{\text{T}}}
\newcommand{\diag}{\text{diag}}
\newcommand{\KL}{D_{\text{KL}}}
\newcommand{\trace}{\text{tr}}
\newcommand{\norm}[1]{\left\lVert #1 \right\rVert}

% distributions
\newcommand{\N}{\text{N}}
\newcommand{\TG}{\text{TG}}
\newcommand{\Bi}{\text{Binom}}
\newcommand{\PG}{\text{PG}}
\newcommand{\Mu}{\text{Multi}}
\newcommand{\GIG}{\text{GIG}}
\newcommand{\IGauss}{\text{IGauss}}
\newcommand{\Un}{\text{U}}

% syntax and readability
\renewcommand{\(}{\left(}
\renewcommand{\)}{\right)}
\renewcommand{\[}{\left[}
\renewcommand{\]}{\right]}

% settings
\graphicspath{{../figs/}}
\pgfplotsset{compat=1.16}
\setlength{\evensidemargin}{.5cm} \setlength{\oddsidemargin}{.5cm}
\setlength{\textwidth}{15cm}
\title{Empirical Bayes estimation of simultaneous-equation models}
\date{\today}
\author{Magnus M. M\"unch$^{1,2}$\footnote{Correspondence to: 
\href{mailto:m.munch@vumc.nl}{m.munch@vumc.nl}}, Mark A. van de Wiel$^{1,3}$, 
Sylvia Richardson$^{3}$, \\ and Gwena{\"e}l G. R. Leday$^{3}$}

\begin{document}

	\maketitle

	\noindent
	1. Department of Epidemiology \& Biostatistics, Amsterdam Public Health
	research institute, Amsterdam University medical centers, PO Box 7057, 1007 MB
	Amsterdam, The Netherlands \\*
	2. Mathematical Institute, Leiden University, Leiden, The Netherlands \\*
	3. MRC Biostatistics Unit, Cambridge Institute of Public Health, Cambridge, 
	United Kingdom

	\begin{abstract}
		{...}
	\end{abstract}

	\noindent\textbf{Keywords}: Empirical Bayes; Simultaneous-equations model; 
	Variational Bayes; Multivariate data

	\section{Introduction}
	- Motivation

	- Regression information

	- current literature

	- our solution: adaptive borrowing of information through empirical Bayes, 
	guided by external regression information.

	\section{Model}
	\subsection{Simultaneous equations model}\label{sec:SEM}
	We have continuous measurements $y_{id}$ for observation $i=1,\dots, n$ on 
	outcome $d=1,\dots,D$. Throughout this paper we assume the $y_{id}$ to be 
	centred per outcome and let $\y_d = \begin{bmatrix} y_{1d} & \dots & y_{nd} 
	\end{bmatrix} \tr$. We predict the $y_{id}$ with features $x_{ij}$, 
	$j=1,\dots, p$, collected in $\x_i = \begin{bmatrix} x_{i1} & \dots & x_{ip} 
	\end{bmatrix} \tr$. We have characteristics of the outcomes available in the 
	form of `co-data' $\mathbf{C} = \begin{bmatrix} \mathbf{c}_1 & \dots & 
	\mathbf{c}_D \end{bmatrix} \tr$.

	We model the outcomes as a linear function of the features:
	\begin{subequations}\label{eq:linearmodel}
		\begin{align}
		y_{id} & = \beta_{0d} + \x_i \tr \bbeta_d + \epsilon_{id},\\
		\text{with } & \epsilon_{id} \sim \mathcal{N}(0, \sigma_d^2),
		\end{align}
	\end{subequations}
	where the $p$-dimensional $\bbeta_d$ are the outcome-specific feature effects.
	Note that (\ref{eq:linearmodel}) gives rise to a system of 
	$D$ linear simultaneous equations.

	\subsection{Bayesian prior model}
	We capture the uncertainty in the parameters through a Bayesian prior model. 
	We assign a flat prior to the intercept terms $\beta_{0d}$ and integrate 
	them out. The remaining parameters are endowed with the following priors:
	\begin{subequations}\label{eq:priormodel}
	  \begin{align}
      y_{id} | \bm{\beta}_d, \sigma_d^2 & \sim \mathcal{N} 
      (\x_i \tr \bm{\beta}_d, \sigma_d^2), \\
      \beta_{jd} | \gamma_d^2, \sigma_d^2 & \sim 
      \mathcal{N} (0, \sigma_d^2 \gamma_d^2), \label{eq:betaprior} \\
      \gamma_d^2 & \sim \mathcal{IG}(\theta_d, \lambda_d), \\
      \sigma_d^2 & \sim 1/\sigma_d^3,
    \end{align}
  \end{subequations}
  where $\mathcal{IG}(\theta, \lambda)$ denotes the inverse Gaussian 
  distribution with mean $\theta$ and shape $\lambda$.
  \cite{moran_variance_2018} argue that in regression models, $\bm{\beta}_d$ 
  and $\sigma_d^2$ should be independent \textit{a priori}, i.e.
  $\beta_{jd} | \gamma_d^2, \sigma_d^2 \sim \beta_{jd} | 
  \gamma_d^2$. However, in our case
  that is debatable. The $D$ regressions need not be on the same scale, so in
  order to jointly consider the $\bm{\beta}_d$ some 
  form of calibration is required. A straightforward way 
  of doing this is to include the scale the $\bm{\beta}_d$ prior with 
  $\sigma_d^2$ as in (\ref{eq:betaprior}).

  The normal inverse Gaussian (NIG) prior model for $\beta_{jd}$ and 
  $\gamma^2_d$ in (\ref{eq:priormodel}) was introduced in 
  \cite{barndorff-nielsen_hyperbolic_1978} and is, since 
  \cite{barndorff-nielsen_normal_1997}, routinely applied in mathematical 
  finance \cite[]{kalemanova_normal_2007}. In Figure 
  \ref{fig:dens_igaussian_marginalbeta}, the NIG 
  prior distribution is depicted together with two common prior choices, ridge 
  and lasso. From this Figure we see that, depending on the choice
  of hyperparameters, the NIG prior can vary from heavier tails than the lasso, 
  to more Gaussian tails like the ridge.
<<dens_igaussian_marginalbeta, fig.cap="Marginal $\\beta_j$ prior for several choices of $\\gamma^2_d$ hyperpior, scaled to $\\V(\\beta_j)=1$.", out.width="60%", fig.asp=1>>=
library(GeneralizedHyperbolic)
# all scaled such that variance is one
sigma <- 1
theta <- 1/sigma^2
lambda <- c(0.1, 2)
lambda1 <- sqrt(2)
col <- c(1:4)
lty <- c(1:4)
labels <- as.expression(c(bquote("IG, "~lambda==.(lambda[1])~", "~
                                   theta==.(theta)),
                          bquote("IG, "~lambda==.(lambda[2])~", "~
                                   theta==.(theta)),
                          "Point mass (ridge)", "Exponential (lasso)"))
dprior <- function(x, lambda, theta, sigma) {
  dnig(x, 0, sigma*sqrt(lambda), sqrt(lambda/(theta*sigma)), 0)
}
dlasso <- function(x, lambda1) {
  0.5*lambda1*exp(-lambda1*abs(x))
}

curve(dprior(x, lambda[1], theta, sigma), -3, 3, ylab=expression(p(beta[j])), 
      xlab=expression(beta[j]), n=1000, 
      col=col[1], lty=lty[1])
curve(dprior(x, lambda[2], theta, sigma), add=TRUE, n=1000, 
      col=col[2], lty=lty[2])
curve(dnorm(x, 0, 1), add=TRUE, n=1000, col=col[3], lty=lty[3])
curve(dlasso(x, lambda1), add=TRUE, n=1000, col=col[4], lty=lty[4])
legend("topright", legend=labels, col=col, lty=lty, 
       title="Hyperprior")
@

  Our argumentation to model the $\gamma^2_d$ by an inverse Gaussian 
  distribution, as has been suggested in \cite{fabrizi_specification_2016} and 
  \cite{caron_sparse_2008}, is two-fold: (i) the NIG model is more flexible than
  the standard ridge and (ii) allows to model the mean $\theta_d$ as a 
  function of the co-data $\mathbf{c}_d$, as explained in Section 
  \ref{sec:empiricalbayes}.

  A few remarks on the choice of error variance prior are justified here: 
  many authors endow error variance components with vague gamma priors. 
  \cite{gelman_prior_2006}, 
  among others, advises against this practice. The degree of `vagueness' has a 
  large influence on the posterior, while degree of `vagueness' is a difficult 
  parameter to set. This influence is especially pronounced if the likelihood is
  relatively flat, as may be reasonably expected in the large $p$, small $n$ 
  setting we are in. We therefore model the error variance with Jeffrey's prior 
  \cite[]{jeffreys_invariant_1946}. In the derivation of our Jeffrey's prior
  for the error variance, we jointly consider an unknown mean and variance.

  - Jeffrey's prior for variance $1/\sigma^{3/2}$ (a priori independent data 
	mean and variance) or $1/\sigma^3$ (joint prior, but only mean and variance, 
	not variance and betas).

	\section{Estimation}
	\subsection{Variational Bayes}\label{sec:variationalbayes}
	The posterior corresponding to the model described in (\ref{eq:linearmodel})
	and (\ref{eq:priormodel}) is not available in closed form. We therefore 
	approximate the posterior by variational Bayes, where we force the posterior 
	density $Q_d$ to factorise as: $p(\bbeta_d, \sigma_d^2, \gamma_d^2) \approx 
	Q_d(\cdot) = q(\bbeta_d) \cdot q(\sigma_d^2) \cdot q(\gamma_d^2)$.
	Under such a factorisation, the marginal variational posteriors that 
	minimise the Kullback-Leibler divergence of the true posterior to the 
	variational Bayes approximation \cite[]{neal_view_1998}, are given by:
	\begin{align*}
    q(\bm{\beta}_d) & \overset{D}{=} \mathcal{N}_p 
    (\bm{\mu}_d, \bm{\Sigma}_d), \\
    q(\gamma^2_d) & \overset{D}{=} \mathcal{GIG}
    \(-\frac{p+1}{2}, \lambda_d \theta_d^{-2}, \delta_d\), \\
    q(\sigma^2_d) & \overset{D}{=} \Gamma^{-1} \(\frac{n + p + 1}{2}, \zeta_d\).
  \end{align*}
	The variational parameters $\bm{\mu}_d$, $\bm{\Sigma}_d$, $\delta_d$, and
	$\zeta_d$ contain cyclic dependencies and are iteratively 
	updated by:
  \begin{align*}
    \bm{\Sigma}_d^{(h+1)} & = (a_d^{(h)})^{-1} 
    (\X \tr \X + b_d^{(h)} \I)^{-1}, \\
    \bm{\mu}_d^{(h+1)} & = (\X \tr \X + b_d^{(h)} \I)^{-1} \X \tr \y_d, \\
    \delta_d^{(h+1)} & = \bmu_d \tr \bmu_d + \lambda_d, \\ 
    \zeta_d^{(h+1)} & = \frac{1}{2} \bigg[\mathbf{y}_d \tr \mathbf{y}_d -
    2 \mathbf{y}_d \tr \X \bm{\mu}_d^{(h+1)} + 
    \trace ( \X \tr \X \bm{\Sigma}_d^{(h+1)}) + 
    (\bm{\mu}_d^{(h+1)}) \tr \X \tr \X \bm{\mu}_d^{(h+1)} \\
    & \,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\, + 
    b_d^{(h+1)} \trace ( \bm{\Sigma}_d^{(h+1)}) + 
    b_d^{(h+1)} (\bm{\mu}_d^{(h+1)}) \tr \bm{\mu}_d^{(h+1)}\bigg],
  \end{align*}
  until convergence. Here, we set
  \begin{align*}
    a_d^{(h)} & =\E_{Q^{(h)}}(\sigma_d^{-2})=(n + p + 1)/(2 \zeta_d^{(h)}), \\
    b_d^{(h)} & = \E_{Q^{(h)}}(\gamma_d^{-2}) = \sqrt{\frac{\lambda_d}
    {\theta_d^2 \delta_d^{(h)}}} \frac{K_{\frac{p - 1}{2}} 
    \( \sqrt{\lambda_d \delta_d^{(h)}}/\theta_d \)}
    {K_{\frac{p+1}{2}} \( \sqrt{\lambda_d \delta_d^{(h)}}/\theta_d \)} + 
    \frac{p+1}{\delta_d^{(h)}}.
  \end{align*}

  \subsection{Empirical Bayes}\label{sec:empiricalbayes}
  A full Bayesian model requires the specification of hyper parameters 
  $\bm{\theta}=\begin{bmatrix} \theta_1 & \cdots & \theta_D \end{bmatrix} \tr$ 
  and $\begin{bmatrix} \lambda_1 & \cdots & \lambda_D \end{bmatrix} \tr$. 
  These are abstract and hard to interpret parameters 
  for which we generally lack expert knowledge. They do, however, have a 
  significant influence on the shape of the posterior distribution. We 
  therefore propose to estimate these hyper parameters by empirical Bayes. 
  Simply put, empirical Bayes fits the prior model to the data and is, as such, 
  objective (up to model specification).

	The canonical method for empirical Bayes is maximisation of the marginal 
	likelihood with respect to the hyper parameters. In 
	\cite{casella_empirical_2001} the marginal likelihood is maximised by an EM 
	algorithm:
	$$
	\bm{\theta}^{(l+1)},\bm{\lambda}^{(l+1)} = \underset{\bm{\theta},\bm{\lambda}}
	{\argmax}\E_{\cdot | \Y} 
	[\log p(\Y, \B, \bgamma^2, \bsigma^2) | \bm{\theta}^{(l)},\bm{\lambda}^{(l)}],
	$$
	where $\Y = \begin{bmatrix} \y_1 & \cdots & \y_D \end{bmatrix}$,
	$\B = \begin{bmatrix} \bbeta_1 & \cdots & \bbeta_D \end{bmatrix}$, 
	$\bgamma^2 = \begin{bmatrix} \gamma^2_1 & \cdots & \gamma^2_D 
	\end{bmatrix}\tr$, $\bsigma^2 = \begin{bmatrix} \sigma^2_1 & \cdots & 
	\sigma^2_D \end{bmatrix} \tr$,
	and the expectation is with respect to the posterior. In our case, this 
	posterior is not available in closed form, which renders the expectation 
	difficult. While \cite{casella_empirical_2001} suggests to approximate the 
	expectation by a Monte Carlo sample, we propose to use the variational Bayes 
	approximation developed in Section \ref{sec:variationalbayes}:
	\begin{align*}
	  \bm{\theta}^{(l+1)},\bm{\lambda}^{(l+1)} & = 
	  \underset{\bm{\theta},\bm{\lambda}}{\argmax}\E_{Q^{(l)}} 
	  [\log p(\Y, \B, \bgamma^2, \bsigma^2)] \\
	  & = \underset{\bm{\theta},\bm{\lambda}}{\argmax} \E_{Q^{(l)}} 
	  [\log \pi (\B | \bgamma^2, \bsigma^2)] + 
	  \E_{Q^{(l)}} [\log \pi (\bgamma^2)].
	\end{align*}
	where now the expectation is with respect to the converged variational 
	posterior $Q^{(l)}=\prod_{d=1}^D Q_d^{(l)}$.

	To make use of the outcome-specific codata, we parametrise 
	$\theta_d = (\mathbf{c}_d \tr \bm{\alpha})^{-1}$ and estimate $\blambda$ and
	$\bm{\alpha}$. This allows us to include continuous co-data on the outcomes
	in an objective, data-driven manner. The estimating equations are:
  \begin{align*}
    \bm{\alpha} & = \[ \mathbf{C} \tr \diag 
    (\lambda_d e_d^{(l)}) \mathbf{C} \]^{-1} \mathbf{C} \tr \diag 
    (\lambda_d) \mathbf{1}_{D \times 1}, \\
    \lambda_d & = \[ b_d^{(l)} + e_d^{(l)} (\mathbf{c}_d \tr \bm{\alpha})^2 - 
    2\mathbf{c}_d \tr \bm{\alpha} \]^{-1},
  \end{align*}
  where $e_d^{(l)} = \E_{Q^{(l)}}(\gamma_d^2) = [b_d^{(l)} - 
  (p + 1)/\delta_d^{(l)})] \cdot \delta_d^{(l)} 
  (\theta_d^{(h)})^2/\lambda_d^{(l)}$. Solving the equations is done by 
  iteratively reweighted least squares of responses $(e_d^{(l)})^{-1}$ on 
  predictors $\mathbf{c}_d$ with weights $\lambda_d e_d^{(l)}$.

  \section{Comparison of NIG model to Normal inverse Gamma}
  - check model with Stan.

  \section{Comparison of conjugate to non-conjugate}

  \section{Variational Bayes versus MCMC}

  \section{Inverse Gaussian Empirical Bayes versus inverse Gamma}

  \section{Simulations for VB}
	\subsection{Setup}

  We simulate from a correctly specified model given in \ref{eq:priormodel}. We assume random $y_{id}$, $\x_i$, and $\bm{\beta}_d$, while we fix $\mathbf{C}$, $\gamma_d^2$, $\sigma_d^2$ and $\bm{\alpha}$. We estimate the $\bm{\mu}_d$, $\bm{\Sigma}_d$, $\delta_d$, $\bm{\alpha}$ and $\lambda$, while we fix the $\sigma_d^2$ to their true values.

  We set $n=100$, $p=200$, and $D=10$. We consider 5 evenly sized classes of drugs, such that $\mathbf{c}_{kd} = \mathbbm{1}(k = \text{class}_d)$ and $\begin{bmatrix} \mathbf{c}_1 & \cdots & \mathbf{c}_D \end{bmatrix} \tr$. From these the $\gamma_d^2 = (\mathbf{c}_d \tr \bm{\alpha})^{-1}$ are created. Next, we simulate $\bm{\beta}_{jd} \sim \mathcal{N} (0, \sigma^2_d \gamma_d^2)$. We fix the mean signal-to noise ratio $\overline{\text{SNR}}_d$, such that the predictor data variance is $s^2 = \V(\mathbf{x}_i) = \overline{\text{SNR}}_d / (\overline{\gamma^2_d} p)$ (NOT CORRECT). The last step consists of simulating the data by $\mathbf{x}_{ij} \sim \mathcal{N}(0,s^2)$ and $y_{id} = \mathcal{N} (\mathbf{x}_i \tr \bm{\beta}_d, \sigma_d^2)$. The remaining parameters are set as follows: $\bm{\alpha} = \begin{bmatrix} 1 & \dots & 5 \end{bmatrix} \tr$, $\sigma_d^2 = 1$ for all $d$, and $\overline{\text{SNR}}_d = 10$. To account for the random data and parameter generation, we repeat the simulation 100 times. We compare the results to the inverse Gamma and non-conjugate model, and present the results in Figures \ref{fig:boxplot_igaussian_res1_prior_mean}-\ref{fig:boxplot_igaussian_res2_prior_mean}.

<<boxplot_igaussian_res1_prior_mean, fig.cap="Prior mean estimates", out.width="80%", fig.asp=2/3>>=
@

<<boxplot_igaussian_res2_prior_mean, fig.cap="Prior mean estimates", out.width="80%", fig.asp=2/3>>=
@

  \section{Application}

  \section{Discussion}

  \section*{Software}
	A (developmental) \texttt{R} package is available from 
	\url{https://github.com/magnusmunch/cambridge}.

	\section*{Supplementary Material}
	Supplementary Material is available online from 
	\url{https://github.com/magnusmunch/cambridge}.

	\section*{Reproducible Research}
	All results and documents may be recreated from 
	\url{https://github.com/magnusmunch/cambridge}.

	\section*{Acknowledgements}
	\textit{Conflict of Interest}: None declared.

	\bibliographystyle{author_short3.bst}
	\bibliography{refs}

	\section*{Session info}

<<r session-info, eval=TRUE, echo=TRUE, tidy=TRUE>>=
devtools::session_info()
@

\end{document}

### 2019-04-17 docs/presentation.Rnw
added mcmc version of inverse Gaussian model in stan

@@
   \begin{frame}
+    \frametitle{Inverse Gaussian distribution}
+<<dens_igaussian_igamma, fig.cap="Several densities of the inverse Gaussian and inverse Gamma families", out.width="85%", fig.asp=2/3>>=
+digauss <- function(x, theta, lambda, eta) {
+  sqrt(lambda/(x^3*2*pi))*exp(-lambda*(x - theta)^2/(2*theta^2*x))
+}
+
+digamma <- function(x, theta, lambda, eta) {
+  alpha <- eta/2
+  beta <- lambda/2
+  beta^alpha/gamma(alpha)*x^(-alpha - 1)*exp(-beta/x)
+}
+
+eta <- c(1, 2, 1, 1, 5)
+lambda <- c(1, 1, 2, 1/2, 2)
+theta <- c(1000, 1000, 1, 1, 1)
+
+labels <- c("inv. Gaussian", "inv. Gamma")
+col <- c(1, 2)
+
+opar <- par(no.readonly=TRUE)
+par(mar=opar$mar*c(1, 1.3, 1, 1), cex=1.2, lwd=2)
+layout(matrix(c(rep(c(1, 1, 2, 2, 3, 3), 2), rep(c(0, 4, 4, 5, 5, 0), 2)),
+                 nrow=4, ncol=6, byrow=TRUE))
+for(i in 1:length(eta)) {
+  modes <- c(theta[i]*(sqrt(1 + 9*theta[i]^2/
+                            (4*lambda[i]^2)) - 3*theta[i]/(2*lambda[i])),
+             lambda[i]/(eta[i] + 2))
+  ylim <- c(0, max(digauss(modes[1], theta[i], lambda[i], eta[i]),
+                   digamma(modes[2], theta[i], lambda[i], eta[i])))
+  curve(digauss(x, theta[i], lambda[i], eta[i]), 0.001, 5, n=1000,
+        ylim=ylim, ylab="Density", "x", col=col[1], 
+        main=bquote(eta==.(eta[i])*","~lambda==.(lambda[i])*","~theta==.(
+          theta[i])))
+  curve(digamma(x, theta[i], lambda[i], eta[i]), add=TRUE, 
+        col=col[2], n=1000)
+}
+legend("topright", legend=labels, lty=1, col=col, seg.len=1)
+par(opar)
+@
+  \end{frame}
+
+  \begin{frame}
+    \frametitle{Normal inverse Gaussian model}
+<<dens_igaussian_marginalbeta, fig.cap="Marginal $\\beta_j$ prior for several choices of $\\gamma^2_d$ hyperpior, scaled to $\\V(\\beta_j)=1$.", out.width="80%", fig.asp=2/3>>=
+library(GeneralizedHyperbolic)
+# all scaled such that variance is one
+sigma <- 1
+theta <- 1/sigma^2
+lambda <- c(0.1, 2)
+lambda1 <- sqrt(2)
+col <- c(1:4)
+lty <- c(1:4)
+labels <- as.expression(c(bquote("IG, "~lambda==.(lambda[1])~", "~
+                                   theta==.(theta)),
+                          bquote("IG, "~lambda==.(lambda[2])~", "~
+                                   theta==.(theta)),
+                          "Point mass (ridge)", "Exponential (lasso)"))
+dprior <- function(x, lambda, theta, sigma) {
+  dnig(x, 0, sigma*sqrt(lambda), sqrt(lambda/(theta*sigma)), 0)
+}
+dlasso <- function(x, lambda1) {
+  0.5*lambda1*exp(-lambda1*abs(x))
+}
+
+opar <- par(no.readonly=TRUE)
+par(mar=opar$mar*c(1, 1.3, 1/1.3, 1), cex.lab=1.5, lwd=3)
+curve(dprior(x, lambda[1], theta, sigma), -3, 3, ylab=expression(p(beta[j])), 
+      xlab=expression(beta[j]), n=1000, 
+      col=col[1], lty=lty[1])
+curve(dprior(x, lambda[2], theta, sigma), add=TRUE, n=1000, 
+      col=col[2], lty=lty[2])
+curve(dnorm(x, 0, 1), add=TRUE, n=1000, col=col[3], lty=lty[3])
+curve(dlasso(x, lambda1), add=TRUE, n=1000, col=col[4], lty=lty[4])
+legend("topright", legend=labels, col=col, lty=lty, 
+       title="Hyperprior")
+par(opar)
+@
+  \end{frame}
+
+  \begin{frame}
     \frametitle{Hyperparameters}

### 2019-04-17 docs/supplement1.Rnw
added mcmc version of inverse Gaussian model in stan

--- 2019-02-19 docs/supplement.Rnw
@@
 bbm,algorithm,algpseudocode,pgfplotstable,threeparttable,booktabs,mathtools,
-dsfont,parskip}
+dsfont,parskip,xr}
+\externaldocument[md-]{manuscript1}
 
@@
 % settings
-\pgfplotsset{compat=1.13}
+\graphicspath{{../figs/}}
+\pgfplotsset{compat=1.16}
 \setlength{\evensidemargin}{.5cm} \setlength{\oddsidemargin}{.5cm}
@@
 \date{\today}
-\author{Magnus M.
-M\"unch$^{1,2}$\footnote{Correspondence to: \href{mailto:m.munch@vumc.nl}{m.munch@vumc.nl}}, Mark A.
-van de Wiel$^{1,3}$, Sylvia Richardson$^{3}$, \\ and Gwena{\"e}l G.
-R.
-Leday$^{3}$}
+\author{Magnus M. M\"unch$^{1,2}$\footnote{Correspondence to: 
+\href{mailto:m.munch@vumc.nl}{m.munch@vumc.nl}}, Mark A. van de Wiel$^{1,3}$, 
+Sylvia Richardson$^{3}$, \\ and Gwena{\"e}l G. R. Leday$^{3}$}
 
@@
 	\noindent
-	1. Department of Epidemiology \& Biostatistics, Amsterdam Public Health research institute, VU University Medical Center, PO Box 7057, 1007 MB
-	Amsterdam, The Netherlands\\
-	2. Mathematical Institute, Leiden University, Leiden, The Netherlands \\
-	3. MRC Biostatistics Unit, Cambridge Institute of Public Health, Cambridge, United Kingdom \\
+	1. Department of Epidemiology \& Biostatistics, Amsterdam Public Health
+	research institute, Amsterdam University medical centers, PO Box 7057, 1007 MB
+	Amsterdam, The Netherlands \\*
+	2. Mathematical Institute, Leiden University, Leiden, The Netherlands \\*
+	3. MRC Biostatistics Unit, Cambridge Institute of Public Health, Cambridge, 
+	United Kingdom
 
 	\section{Content overview}
-	This document contains the Supplementary Material (SM) to the document `Drug efficacy prediction in cell lines'. In the following, this document is referred to as Main Document (MD).
+	This document contains the Supplementary Material (SM) to the document `Empirical Bayes estimation of simultaneous-equation models'.
+In the following, this document is referred to as Main Document (MD).
 
@@
 	In the following all expectations are with respect to the variational posterior $Q$.
+	\subsection{Evidence lower bound}
+	We monitor convergence through the evidence lower bound (ELBO).
+To compute the evidence lower bound after updating the EB parameters from $\eta_d^{(l)}, \theta_d^{(l)}, \lambda_d^{(l)}$ to $\eta_d^{(l + 1)}, \theta_d^{(l + 1)}, \lambda_d^{(l + 1)}$, we compute $\text{ELBO}^{(l + 1)} = \sum_{d=1}^D \text{ELBO}_d^{(l + 1)}$, with, in the inverse Gaussian case:
 	\begin{align*}
-	\log q_{\B} (\B) & \propto \E_{\mathbf{U}, \bsigma^2} [\log \mathcal{L}(\Y | \B, \U, \bsigma^2 )] + \E_{\bgamma^2, \bsigma^2} [\log \pi (\B | \bgamma^2, \bsigma^2)] \\
-	& \propto - \frac{1}{2} \sum_{d=1}^D \sum_{i=1}^n \E_{\mathbf{u}_d, \sigma_d^2} \[ \frac{(y_{id} - \x_i \tr \bbeta_d - \mathbf{z}_i \tr \mathbf{u}_d)^2}{\sigma_d^2} \] - \frac{1}{2} \sum_{d=1}^D \sum_{j=1}^p \phi_{g(j)}^{-2} \E_{\gamma^2_d, \sigma^2_d} \( \frac{\beta_{jd}^2}{\gamma_d^2 \sigma_d^2} \), \\
-	q_{\B} (\B) & \overset{D}{=} \prod_{d=1}^{D} \mathcal{N}_p (\bmu_d, \bSigma_d), \\
-	& \text{ with } \bSigma_d = \E(\sigma_d^{-2})^{-1} [\X \tr \X + \E(\gamma_d^{-2}) \cdot \diag (\phi_{g(j)}^{-2}) ]^{-1}, \\
-	& \text{ and } \bmu_d = [\X \tr \X + \E(\gamma_d^{-2}) \cdot \diag (\phi_{g(j)}^{-2}) ]^{-1} \X \tr [\y_d - \Z \cdot \E(\mathbf{u}_d)].
+	\text{ELBO}_d^{(l + 1)} & = \frac{1}{2} \log |\bm{\Sigma}_d^{(l)}| - \frac{a_d^{(l)}}{2} \[ \y_d \tr \y_d - 2\y_d \tr \X \bm{\mu}_d^{(l)} + (\bm{\mu}_d^{(l)}) \tr \X \tr \X \bm{\mu}_d^{(l)} + \trace ( \X \tr \X \bm{\Sigma}_d^{(l)})\] \\
+	& + \frac{b_d^{(l)}}{2} \[ \delta_d^{(l)} - (a_d^{(l)} + 1 - b_d^{(l)}/c_d^{(l)})(\bm{\mu}_d^{(l)}) \tr \bm{\mu}_d^{(l)} - (a_d^{(l)} + 1 - b_d^{(l)}/c_d^{(l)}) \trace (\bm{\Sigma}_d^{(l)}) \] \\
+	& - \frac{df + 1}{2} \log \zeta_d^{(l)} + \frac{p + 1}{4} \log \lambda_d^{(l)} - \frac{p + 1}{2} \log \theta_d^{(l)} - \frac{p+1}{4} \log \delta_d^{(l)} \\
+	& + \log K_{\frac{p+1}{2}}\(\sqrt{\lambda_d^{(l)} \delta_d^{(l)}}/\theta_d^{(l)}\) + \frac{\lambda_d^{(l)} e_d^{(l)}}{2(\theta_d^{(l)})^2} \\
+	& + \frac{\lambda_d^{(l+1)}}{\theta_d^{(l+1)}} + \frac{1}{2}\log \lambda_d^{(l+1)} - \frac{\lambda_d^{(l+1)} e_d^{(l)}}{2 (\theta_d^{(l + 1)})^2} - \frac{\lambda_d^{(l + 1)} b_d^{(l)}}{2}.
 	\end{align*}
 
+	In the inverse Gamma setting, we have:
 	\begin{align*}
-	\log q_{\U}(\U) & \propto \E_{\mathbf{B}, \bsigma^2} [\log \mathcal{L}(\Y | \B, \U, \bsigma^2 )] + \E_{\mathbf{\Xi}_1, \dots, \mathbf{\Xi}_D, \bsigma^2} [\log \pi (\U | \mathbf{\Xi}_1, \dots, \mathbf{\Xi}_D, \bsigma^2)] \\
-	& \propto - \frac{1}{2} \sum_{d=1}^D \sum_{i=1}^n \E_{\bbeta_d, \sigma_d^2} \[ \frac{(y_{id} - \x_i \tr \bbeta_d - \mathbf{z}_i \tr \mathbf{u}_d)^2}{\sigma_d^2} \] - \frac{1}{2} \sum_{d=1}^D \E_{\bm{\Xi}_d, \sigma_d^2}\( \frac{\mathbf{u}_d \tr \bm{\Xi}_d^{-1} \mathbf{u}_d}{\sigma_d^2} \), \\
-	q_{\U}(\U) & \overset{D}{=} \prod_{d=1}^{D} \mathcal{N}_T (\mathbf{m}_d, \Sm_d), \\
-	& \text{ with } \Sm_d = \E(\sigma_d^{-2})^{-1} [\Z \tr \Z + \E(\bm{\Xi}_d^{-1}) ]^{-1}, \\
-	& \text{ and } \mathbf{m}_d = [\Z \tr \Z + \E(\bm{\Xi}_d^{-1}) ]^{-1} \Z \tr [\y_d - \X \cdot \E(\bbeta_d)].
+	\text{ELBO}_d^{(l + 1)} & = \frac{1}{2} \log |\bm{\Sigma}_d^{(l)}| - \frac{a_d^{(l)}}{2} \[ \y_d \tr \y_d - 2\y_d \tr \X \bm{\mu}_d^{(l)} + (\bm{\mu}_d^{(l)}) \tr \X \tr \X \bm{\mu}_d^{(l)} + \trace ( \X \tr \X \bm{\Sigma}_d^{(l)})\] \\
+	& + \frac{b_d^{(l)}}{2} \[ \delta_d^{(l)} - (a_d^{(l)} + 1 - b_d^{(l)}/c_d^{(l)})(\bm{\mu}_d^{(l)}) \tr \bm{\mu}_d^{(l)} - (a_d^{(l)} + 1 - b_d^{(l)}/c_d^{(l)}) \trace (\bm{\Sigma}_d^{(l)}) \]  \\
+	& - \frac{df + 1}{2} \log \zeta_d^{(l)} - \frac{p + \eta_d^{(l)}}{2} \log \delta_d^{(l)} + \frac{\eta_d^{(l)}}{2} e_d^{(l)} + \frac{\eta_d^{(l)}}{2} \log 2 + \log \Gamma \( \frac{p + \eta_d^{(l)}}{2} \) \\
+	& - \log \Gamma \( \frac{\eta_d^{(l + 1)}}{2} \) - \frac{\eta_d^{(l+1)}}{2} e_d^{(l)} - \frac{\eta_d^{(l+1)}}{2} \log 2 + \frac{\eta_d^{(l + 1)}}{2} \log \lambda_d^{(l+1)} - \frac{\lambda_d^{(l + 1)} b_d^{(l)}}{2}.
 	\end{align*}
+  Note that the computation of $e_d^{(l)}$ differs between the inverse Gaussian and Gamma models.
+Furthermore, to compute the ELBO after a VB update we simply let $\eta_d^{(l)}, \theta_d^{(l)}, \lambda_d^{(l)}=\eta_d^{(l + 1)}, \theta_d^{(l + 1)}, \lambda_d^{(l + 1)}$.
 
-	\begin{align*}
-	\log q_{\bgamma^{2}}(\bgamma^{2}) & \propto \E_{\B, \bsigma^2} [\log \pi (\B | \bgamma^{2}, \sigma^2)] + \log \pi (\bgamma^{2}) \\
-	& \propto -\frac{p}{2} \sum_{d=1}^D \log \gamma_d^{2} - \frac{1}{2} \sum_{d=1}^D \sum_{j=1}^p \phi_{g(j)}^{-2} \E \( \frac{\beta_{jd}^2}{\sigma^{2}_d}\) \gamma_d^{-2} - \frac{3}{2} \sum_{d=1}^D \log \gamma_d^{2} \\
-	& \,\,\,\,\,\,\,\,\,\, - \frac{\lambda}{2 \theta^2} \sum_{d=1}^D (\gamma_d^2 - \theta)^2 \gamma_d^{-2} \\
-	& \propto \sum_{d=1}^D \left\{ \( -\frac{p + 1}{2} - 1\) \log \gamma_d^{2} - \frac{\lambda}{2\theta^2} \gamma_d^2 - \frac{1}{2}\[ \lambda + \E(\sigma_d^{-2}) \sum_{j=1}^p \frac{\E(\beta_{jd}^2)}{\phi_{g(j)}^{2}}\] \gamma_d^{-2} \right\}, \\
-	q_{\bgamma^{2}}(\bgamma^{2}) & \overset{D}{=} \prod_{d=1}^D \mathcal{GIG} \(-\frac{p+1}{2}, \frac{\lambda}{\theta_d^2}, \delta_{d} \), \\
-	& \text{with } \delta_{d} =\E(\sigma_d^{-2}) \sum_{j=1}^p \frac{\E(\beta_{jd})^2 + \V(\beta_{jd})}{\phi_{g(j)}^2} + \lambda.
-	\end{align*}
+  In practice, there are different options to assess convergence.
+We may monitor the parameters themselves, the ELBO, or a combination of the two.
+We may also iterate for an \textit{a priori} fixed number of times.
+We may also choose to use a different convergence criterium for the EB and VB iterations.
+Currently we fix the number of EB iterations to 20, while we do 2 VB iterations per EB iteration.
 
-	\begin{align*}
-	\log q_{\bsigma^{2}}(\bsigma^{2}) & \propto \E_{\B,\mathbf{U}} [\log \mathcal{L}(\Y | \B, \U, \bsigma^2 )] + \E_{\B, \bgamma^2} [\log \pi (\B | \bgamma^2, \bsigma^2)] \\
-	& \,\,\,\,\,\,\,\,\,\, + \E_{\mathbf{U}, \bm{\Xi}_1, \dots, \bm{\Xi}_D} [\log \pi (\mathbf{U} | \bm{\Xi}_1, \dots, \bm{\Xi}_D, \bsigma^2)] + \log \pi (\bsigma^{2}) \\
-	& \propto -\frac{n}{2} \sum_{d=1}^D \log \sigma_d^{2} - \frac{1}{2} \sum_{d=1}^D \sum_{i=1}^n \E [(y_{id} - \x_i \tr \bbeta_d - \mathbf{z}_i \tr \mathbf{u}_d)^2] \sigma_d^{2} - \frac{p}{2} \sum_{d=1}^D \log \sigma_d^{-2} \\
-	& \,\,\,\,\,\,\,\,\,\, - \frac{1}{2} \sum_{d=1}^D \sum_{j=1}^p \E( \gamma_d^{-2}) \frac{\E ( \beta_{jd}^2)}{\phi_{g(j)}^{2}} \sigma_d^{-2} - \frac{T}{2} \sum_{d=1}^D \log \sigma_d^{2} - \frac{1}{2} \sum_{d=1}^D \E(\mathbf{u}_d \tr \bm{\Xi}_d^{-1} \mathbf{u}) \sigma_d^{-2} \\
-	& \,\,\,\,\,\,\,\,\,\, - \frac{3}{2} \sum_{d=1}^D \log \sigma^2_d \\
-	& = -\sum_{d=1}^{D} \(\frac{n + p + T + 3}{2} \) \log \sigma_d^{2} - \sum_{d=1}^{D} \frac{1}{2}\Bigg\{ \sum_{i=1}^n \E [(y_{id} - \x_i \tr \bbeta_d - \mathbf{z}_i \tr \mathbf{u}_d)^2] \\
-	& \,\,\,\,\,\,\,\,\,\, + \E(\gamma_d^{-2})\sum_{j=1}^p \frac{\E(\beta_{jd}^2)}{\phi_{g(j)}^2} + \E(\mathbf{u}_{d} \tr \bm{\Xi}_d^{-1} \mathbf{u}_d) \Bigg\} \sigma_d^{-2}, \\
-	q_{\bsigma^{2}}(\bsigma^{2}) & \overset{D}{=} \prod_{d=1}^D \Gamma^{-1} \(\frac{n + p + T + 1}{2}, \zeta_{d} \), \\
-	& \text{ with } \zeta_{d} = \frac{\y_d \tr \y_d}{2} - \y_d \tr [\X \E(\bbeta_d) + \Z \E(\mathbf{u}_d)] + \E(\bbeta_d \tr) \X \tr \Z \E (\mathbf{u}_d) \\
-	& \,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\, + \frac{\trace [\X \tr \X \V(\bbeta_d)]}{2} + \frac{\E(\bbeta_d \tr) \X \tr \X \E(\bbeta_d)}{2} + \frac{\trace [\Z \tr \Z \V(\mathbf{u}_d)]}{2} \\
-	& \,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\, + \frac{\E(\mathbf{u}_d \tr) \Z \tr \Z \E(\mathbf{u}_d)}{2} + \frac{\E(\gamma_d^{-2})}{2} \sum_{j=1}^p \frac{\E(\beta_{jd})^2 + \V(\beta_{jd})}{\phi_{g(j)}^2} \\
-	& \,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,+ \frac{\E(\mathbf{u}_d \tr) \E(\bm{\Xi}_d^{-1}) \E(\mathbf{u}_d)}{2} + \frac{\trace [\E(\bm{\Xi}_d^{-1}) \V(\mathbf{u}_d)]}{2},
-	\end{align*}
-
-	\begin{align*}
-	\log q_{\bm{\Xi}_1, \dots, \bm{\Xi}_D}(\bm{\Xi}_1, \dots, \bm{\Xi}_D) & \propto \E_{\mathbf{U}, \bm{\sigma}^2} [\log \pi (\mathbf{U} | \bm{\Xi}_1, \dots, \bm{\Xi}_D, \bsigma^2)] + \log \pi (\bm{\Xi}_1, \dots, \bm{\Xi}_D) \\
-	& \propto -\frac{1}{2} \sum_{d=1}^D \log |\bm{\Xi}_d| - \frac{1}{2} \sum_{d=1}^D \E \(\frac{\mathbf{u}_d \tr \bm{\Xi}_d^{-1} \mathbf{u}_d}{\sigma_d^2} \) \\
-	& \,\,\,\,\,\,\,\,\,\, - \frac{k + T + 1}{2} \sum_{d=1}^D \log |\bm{\Xi}_d| - \frac{1}{2} \sum_{d=1}^D \trace (\bm{\Omega} \bm{\Xi}_d^{-1}) \\
-	& = -\frac{\tau + T + 2}{2} \sum_{d=1}^D \log |\bm{\Xi}_d| - \frac{1}{2} \sum_{d=1}^D \trace \left\{ \[\E(\sigma_d^{-2}) \E( \mathbf{u}_d \mathbf{u}_d \tr) + \bm{\Omega}\] \bm{\Xi}_d^{-1} \right\}, \\
-	q_{\bm{\Xi}_1, \dots, \bm{\Xi}_D}(\bm{\Xi}_1, \dots, \bm{\Xi}_D) & \overset{D}{\propto} \prod_{d=1}^D \mathcal{W}_T^{-1} (\bm{\Psi}_d, \nu + 1),  \\ 
-	& \text{ with } \bm{\Psi}_d =\E(\sigma^{-2}_d) \[\V(\mathbf{u}_d) + \E(\mathbf{u}_d) \E( \mathbf{u}_d \tr) \] + \bm{\Omega}.
-	\end{align*}
-
-	\section{Simulations}
-	\subsection{Models}
-	In order to investigate the inverse Gaussian prior model and its estimation, we consider a simpler model. The simpler model ignores tissue effects and fixes $\forall g: \phi_g^2=1$:
-	\begin{align*}
-  y_{id} | \bm{\beta}_d, \sigma_d^2 & \sim \mathcal{N} (\x_i \tr \bm{\beta}_d, \sigma_d^2), \\
-  \beta_{jd} | \gamma_d^2, \sigma_d^2 & \sim p(\beta_{jd}), \\
-  \gamma_d^2 & \sim p(\gamma_d^2), \\
-  \sigma_d^2 & \sim 1/\sigma_d^3.
-  \end{align*}
-
-  \cite{moran_variance_2018} argue that in regression models, $\bm{\beta}_d$ and $\sigma_d^2$ should be independent \textit{a priori}.
-However, in our case that is debatable.
-The $D$ regressions need not be on the same scale, so some form of calibration of the $\bm{\beta}_d$ posteriors to this scale is required.
-A simple way of doing this is to include the scales $\sigma_d^2$ in the priors for the $\bm{\beta}_d$.
-We call the two versions of the model conjugate (with $\bm{\beta}_d$ and $\sigma_d^2$ dependent):
-  $$
-  \beta_{jd} | \gamma_d^2, \sigma_d^2 \sim \mathcal{N} (0, \sigma_d^2 \gamma_d^2)
-  $$
-  and non-conjugate (with $\bm{\beta}_d$ and $\sigma_d^2$ independent):
-  $$
-  \beta_{jd} | \gamma_d^2, \sigma_d^2 \sim \beta_{jd} | \gamma_d^2 \sim \mathcal{N} (0, \gamma_d^2),
-  $$
-  and compare them in our simulations.
-
-	In addition to the inverse Gaussian model for the prior variances $\gamma_d^2$ (introduced in the MD):
-	$$
-  \gamma_d^2 \sim \mathcal{IG}(\theta_d, \lambda_d),
-  $$
-	we investigate a generalisation of the models in \cite{kpogbezan_empirical_2017} and \cite{leday_gene_2017}, where we draw the $\gamma_d^2$ from an inverse Gamma distribution; :
-	$$
-	\gamma_d^2 \sim \Gamma^{-1}(\eta_d/2, \lambda_d/2).
-	$$
-	For the sake of comparability, we have rescaled the inverse Gamma parameters and reparametrised both distributions as generalized inverse Gaussians:
-	\begin{align*}
-  \gamma_d^2 & \sim \mathcal{GIG}(-1/2, \lambda_d \theta_d^{-2}, \lambda_d), \\
-  \gamma_d^2 & \sim \mathcal{GIG}(-\eta_d/2, 0, \lambda_d).
-  \end{align*}
-	The two models coincide if $\theta_d \to \infty$ and $\eta_d = 1$.
-
-	\subsection{Estimation}
-	The variational distributions are as follows:
-	\begin{align*}
-  q(\bm{\beta}) & \overset{D}{=} \prod_{d=1}^D \mathcal{N}_p (\bm{\mu}_d, \bm{\Sigma}_d), \\
-  q(\bm{\gamma}^2) & \overset{D}{=} \prod_{d=1}^D \mathcal{GIG}\(-\frac{p+\eta_d}{2}, \lambda_d \theta_d^{-2}, \delta_d\), \\
-  q(\bm{\sigma}^2) & \overset{D}{=} \prod_{d=1}^D \Gamma^{-1} \(\frac{df + 1}{2}, \zeta_d\),
-  \end{align*}
-	where $\eta_d=1$ and $\theta_d \to \infty$ in the inverse Gaussian and inverse Gamma models, respectively.
-Furthermore, $df=n + p$ in the conjugate model and $df=n$ in the non-conjugate model.
-The variational parameters contain cyclic dependencies and are, in the conjugate setting, iteratively updated by:
-  \begin{align*}
-  \bm{\Sigma}_d^{(h+1)} & = (a_d^{(h)})^{-1} (\X \tr \X + c_d^{(h)} \I)^{-1}, \\
-  \bm{\mu}_d^{(h+1)} & = (\X \tr \X + c_d^{(h)} \I)^{-1} \X \tr \y_d, \\
-  \delta_d^{(h+1)} & = b_d^{(h)} (c_d^{(h)})^{-1} \bmu_d \tr \bmu_d + \lambda_d, \\ 
-  \zeta_d^{(h+1)} & = \frac{1}{2} \bigg[\mathbf{y}_d \tr \mathbf{y}_d -2 \mathbf{y}_d \tr \X \bm{\mu}_d^{(h+1)} + \trace ( \X \tr \X \bm{\Sigma}_d^{(h+1)}) + (\bm{\mu}_d^{(h+1)}) \tr \X \tr \X \bm{\mu}_d^{(h+1)} \\
-  & \,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\, + \frac{b_d^{(h+1)} - c_d^{(h+1)} a_d^{(h+1)}}{1 - a_d^{(h+1)}} \trace ( \bm{\Sigma}_d^{(h+1)}) + \frac{b_d^{(h+1)} - c_d^{(h+1)} a_d^{(h+1)}}{1 - a_d^{(h+1)}} (\bm{\mu}_d^{(h+1)}) \tr \bm{\mu}_d^{(h+1)}\bigg].
-  \end{align*}
-  Here, we set
-  \begin{align}
-  a_d^{(h)} & =\E_{Q^{(h)}}(\sigma_d^{-2})=(df+1)/(2 \zeta_d^{(h)}), \nonumber \\
-  b_d^{(h)} & = \E_{Q^{(h)}}(\gamma_d^{-2}) = \sqrt{\frac{\lambda_d}{\theta_d^2 \delta_d^{(h)}}} \frac{K_{\frac{p + \eta_d -2}{2}} \( \sqrt{\lambda_d \delta_d^{(h)}}/\theta_d \)}{K_{\frac{p+\eta_d}{2}} \( \sqrt{\lambda_d \delta_d^{(h)}}/\theta_d \)} + \frac{p+\eta_d^{(h)}}{\delta_d^{(h)}}, \label{eq:auxvar1} \\
-  c_d^{(h)} & = \begin{cases*}
-  b_d^{(h)} & in the conjugate setting \\
-  b_d^{(h)}/a_d^{(h)} & in the non-conjugate setting .
-  \end{cases*} \nonumber
-  \end{align}
-  Note that in the inverse Gamma model, the computation of (\ref{eq:auxvar1}) simplifies to $b_d^{(h)}=(p+\eta_d)/\delta_d^{(h)}$.
-
-  \subsection{Empirical Bayes}
-  To avoid subjectivity in our choice of prior, we estimate the hyper-parameters with empirical Bayes.
-The empirical Bayes procedure differs for the two prior models.
-In the inverse Gaussian case, we parametrise $\theta_d = (\mathbf{c}_d \tr \bm{\alpha})^{-1}$ to allow for the inclusion of drug covariates in the empirical Bayes estimation.
-The estimating equations are:
-  \begin{align*}
-  \bm{\alpha} & = \[ \mathbf{C} \tr \diag (\lambda_d e_d^{(l)}) \mathbf{C} \]^{-1} \mathbf{C} \tr \diag (\lambda_d) \mathbf{1}_{D \times 1}, \\
-  \lambda_d & = \[ b_d^{(l)} + e_d^{(l)} (\mathbf{c}_d \tr \bm{\alpha})^2 - 2\mathbf{c}_d \tr \bm{\alpha} \]^{-1},
-  \end{align*}
-  where $e_d^{(l)} = \E_{Q^{(l)}}(\gamma_d^2) = [b_d^{(l)} - (p + \eta_d^{(h)})/\delta_d^{(l)})] \cdot \delta_d^{(l)} (\theta_d^{(h)})^2/\lambda_d^{(l)}$.
-Solving the equations is done by iteratively reweighted least squares of responses $(e_d^{(l)})^{-1}$ on predictors $\mathbf{c}_d$ with weights $\lambda_d e_d^{(l)}$.
-If we use one common $\lambda=\lambda_d$ then the solution is in closed form:
-  \begin{align*}
-	\bm{\alpha}^{(l+1)} & = \[ \mathbf{C} \tr \diag(e_d^{(l)}) \mathbf{C} \]^{-1} \mathbf{C} \tr \mathbf{1}_{D \times 1}, \\
-	\lambda^{(l+1)} & = \[ \sum_{d=1}^D b_d^{(l)} - (\bm{\alpha}^{(l+1)}) \tr \mathbf{C} \tr \mathbf{1} \]^{-1} D.
-	\end{align*}
-	This corresponds to the model in the MD.
-
-  In the inverse Gamma model we use the following empirical Bayes procedure: Suppose we have a partitioning of our drugs into drug classes and we let prior variance components $\gamma_d^2$ and $\gamma^2_{d'}$ be drawn from the same distribution if drug $d$ and drug $d'$ are from the same class.
-Or simply put: $\gamma^2_{d}, \gamma^2_{d'} \sim \Gamma^{-1}(\eta_c/2, \lambda_c/2)$ if $d, d' \in \text{class}_c$.
-Empirical bayes estimation of the $\eta_c$ and $\lambda_c$ by maximisation of the (approximated) marginal likelihood amounts to iteratively solving the estimating equations:
-  \begin{subequations}
-    \begin{align}
-      \psi\(\frac{\eta_c}{2}\) & = \log \lambda_c - \frac{\sum_{d \in \text{class}_c} e^{(l)}_d}{|\text{class}_c|} - \log 2, \label{eq:estequation1}\\
-      \lambda_c & = \eta_c \cdot |\text{class}_c| \cdot \(\sum_{d \in \text{class}_c} b_d^{(l)}\)^{-1}, \label{eq:estequation2}
-    \end{align}
-  \end{subequations}
-  until convergence, where $e^{(l)}_d = \E_{Q^{(l)}}(\log \gamma_d^2) = \log \delta_d^{(l)} - \psi [ (p + \eta_d^{(l)})/2] - \log 2$. 
-  We propose to do that by first solving (\ref{eq:estequation2}) for $\lambda_c$, plugging it into (\ref{eq:estequation1}) and bounding the result using \cite{alzer_inequalities_1997} to obtain an interval $( \eta_c^*, 2\eta_c^* )$ containing the solution to (\ref{eq:estequation1}), where:
-  $$
-  \eta_c^* = \left\{ \frac{\sum_{d \in \text{class}_c} e^{(l)}_d}{|\text{class}_c|} + \log \[ \frac{\sum_{d \in \text{class}_c} b_d^{(l)}}{|\text{class}_c|} \]  \right\}^{-1}.
-  $$
-  With this interval it is straightforward to find $\alpha^{(l+1)}_c$ by any root-finding algorithm and plugging the solution into (\ref{eq:estequation2}) to find $\lambda^{(l+1)}_c$.
-We may use $\eta_c^*$ as a starting value.
-
-	The inverse Gamma model with independent drug classes allows for more flexible empirical Bayes estimation of the prior means for the different drug classes.
-As a consequence however, the risk of overfitting is increased.
-Additionally, it does not allow to include continuous covariates on the drugs.
-
-	\subsection{Evidence lower bound}
+	\section{Evidence lower bound}
 	We monitor convergence through the evidence lower bound (ELBO).
@@
 
-  \section{Simulation setting}
-  - check model with Stan.
+  \section{Inverse Gamma model}
+  In addition to the inverse Gaussian model, we investigate a generalisation of 
+  the models in \cite{kpogbezan_empirical_2017} and \cite{leday_gene_2017}, 
+  where we draw the $\gamma_d^2$ from an inverse Gamma distribution:
+	$$
+	\gamma_d^2 \sim \Gamma^{-1}(\eta_d/2, \lambda_d/2),
+	$$
+	with shape $\eta_d/2$ and scale $\lambda/2$. For the sake of comparability, we
+	have rescaled the inverse Gamma parameters and reparametrised both 
+	distributions as generalized inverse Gaussians:
+	\begin{align*}
+  \gamma_d^2 & \sim \mathcal{GIG}(-1/2, \lambda_d \theta_d^{-2}, \lambda_d), \\
+  \gamma_d^2 & \sim \mathcal{GIG}(-\eta_d/2, 0, \lambda_d).
+  \end{align*}
+	The two models coincide if $\theta_d \to \infty$ and $\eta_d = 1$.
 
-  - make one group of drugs and compare methods
+	\begin{align*}
+  q(\bm{\beta}_d) & \overset{D}{=} \prod_{d=1}^D \mathcal{N}_p (\bm{\mu}_d, \bm{\Sigma}_d), \\
+  q(\bm{\gamma}^2) & \overset{D}{=} \prod_{d=1}^D \mathcal{GIG}\(-\frac{p+\eta_d}{2}, \lambda_d \theta_d^{-2}, \delta_d\), \\
+  q(\bm{\sigma}^2) & \overset{D}{=} \prod_{d=1}^D \Gamma^{-1} \(\frac{df + 1}{2}, \zeta_d\),
+  \end{align*}
+	where $\eta_d=1$ and $\theta_d \to \infty$ in the inverse Gaussian and inverse Gamma models, respectively.
+Furthermore, $df=n + p$ in the conjugate model and $df=n$ in the non-conjugate model.
+The variational parameters contain cyclic dependencies and are, in the conjugate setting, iteratively updated by:
+  \begin{align*}
+  \bm{\Sigma}_d^{(h+1)} & = (a_d^{(h)})^{-1} (\X \tr \X + c_d^{(h)} \I)^{-1}, \\
+  \bm{\mu}_d^{(h+1)} & = (\X \tr \X + c_d^{(h)} \I)^{-1} \X \tr \y_d, \\
+  \delta_d^{(h+1)} & = b_d^{(h)} (c_d^{(h)})^{-1} \bmu_d \tr \bmu_d + \lambda_d, \\ 
+  \zeta_d^{(h+1)} & = \frac{1}{2} \bigg[\mathbf{y}_d \tr \mathbf{y}_d -2 \mathbf{y}_d \tr \X \bm{\mu}_d^{(h+1)} + \trace ( \X \tr \X \bm{\Sigma}_d^{(h+1)}) + (\bm{\mu}_d^{(h+1)}) \tr \X \tr \X \bm{\mu}_d^{(h+1)} \\
+  & \,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\, + \frac{b_d^{(h+1)} - c_d^{(h+1)} a_d^{(h+1)}}{1 - a_d^{(h+1)}} \trace ( \bm{\Sigma}_d^{(h+1)}) + \frac{b_d^{(h+1)} - c_d^{(h+1)} a_d^{(h+1)}}{1 - a_d^{(h+1)}} (\bm{\mu}_d^{(h+1)}) \tr \bm{\mu}_d^{(h+1)}\bigg].
+  \end{align*}
+  Here, we set
+  \begin{align}
+  a_d^{(h)} & =\E_{Q^{(h)}}(\sigma_d^{-2})=(df+1)/(2 \zeta_d^{(h)}), \nonumber \\
+  b_d^{(h)} & = \E_{Q^{(h)}}(\gamma_d^{-2}) = \sqrt{\frac{\lambda_d}{\theta_d^2 \delta_d^{(h)}}} \frac{K_{\frac{p + \eta_d -2}{2}} \( \sqrt{\lambda_d \delta_d^{(h)}}/\theta_d \)}{K_{\frac{p+\eta_d}{2}} \( \sqrt{\lambda_d \delta_d^{(h)}}/\theta_d \)} + \frac{p+\eta_d^{(h)}}{\delta_d^{(h)}}, \label{eq:auxvar1} \\
+  c_d^{(h)} & = \begin{cases*}
+  b_d^{(h)} & in the conjugate setting \\
+  b_d^{(h)}/a_d^{(h)} & in the non-conjugate setting .
+  \end{cases*} \nonumber
+  \end{align}
+  Note that in the inverse Gamma model, the computation of (\ref{eq:auxvar1}) simplifies to $b_d^{(h)}=(p+\eta_d)/\delta_d^{(h)}$.
 
-  - use drug-specific lambda's
+	<<dens_igaussian_igamma, fig.cap="Several densities of the inverse Gaussian and inverse Gamma families", out.width="100%", fig.asp=2/3>>=
+digauss <- function(x, theta, lambda, eta) {
+  sqrt(lambda/(x^3*2*pi))*exp(-lambda*(x - theta)^2/(2*theta^2*x))
+}
 
-  - try smaller $p$ to see whether $\sigma_d^2$ become more identifiable.
+digamma <- function(x, theta, lambda, eta) {
+  alpha <- eta/2
+  beta <- lambda/2
+  beta^alpha/gamma(alpha)*x^(-alpha - 1)*exp(-beta/x)
+}
 
-  - scale-invariant instead of conjugate.
+eta <- c(1, 2, 1, 1, 5)
+lambda <- c(1, 1, 2, 1/2, 2)
+theta <- c(1000, 1000, 1, 1, 1)
 
-  We simulate from a correctly specified model given in \ref{sec:notissuemodel}.
-We assume random $y_{id}$, $\x_i$, and $\bm{\beta}_d$, while we fix $\mathbf{C}$, $\gamma_d^2$, $\sigma_d^2$ and $\bm{\alpha}$.
-We estimate the $\bm{\mu}_d$, $\bm{\Sigma}_d$, $\delta_d$, $\bm{\alpha}$ and $\lambda$, while we fix the $\sigma_d^2$ to their true values.
+labels <- c("inv. Gaussian", "inv. Gamma")
+col <- c(1, 2)
 
-  We set $n=100$, $p=200$, and $D=10$.
-We consider 5 evenly sized classes of drugs, such that $\mathbf{c}_{kd} = \mathbbm{1}(k = \text{class}_d)$ and $\begin{bmatrix} \mathbf{c}_1 & \cdots & \mathbf{c}_D \end{bmatrix} \tr$.
-From these the $\gamma_d^2 = (\mathbf{c}_d \tr \bm{\alpha})^{-1}$ are created.
-Next, we simulate $\bm{\beta}_{jd} \sim \mathcal{N} (0, \sigma^2_d \gamma_d^2)$.
-We fix the mean signal-to noise ratio $\overline{\text{SNR}}_d$, such that the predictor data variance is $s^2 = \V(\mathbf{x}_i) = \overline{\text{SNR}}_d / (\overline{\gamma^2_d} p)$ (NOT CORRECT).
-The last step consists of simulating the data by $\mathbf{x}_{ij} \sim \mathcal{N}(0,s^2)$ and $y_{id} = \mathcal{N} (\mathbf{x}_i \tr \bm{\beta}_d, \sigma_d^2)$.
-The remaining parameters are set as follows: $\bm{\alpha} = \begin{bmatrix} 1 & \dots & 5 \end{bmatrix} \tr$, $\sigma_d^2 = 1$ for all $d$, and $\overline{\text{SNR}}_d = 10$.
-To account for the random data and parameter generation, we repeat the simulation 100 times.
-We compare the results to the models in \ref{sec:indvarmodel} and \ref{sec:invgammamodel}, and present the results in Figures \ref{fig:boxplot_igaussian_res1_prior_mean}-\ref{fig:boxplot_igaussian_res2_prior_mean}.
-
-<<boxplot_igaussian_res1_prior_mean, fig.cap="Prior mean estimates", out.width="80%", fig.asp=2/3>>=
+opar <- par(no.readonly=TRUE)
+par(mar=opar$mar*c(1, 1.3, 1, 1))
+layout(matrix(c(rep(c(1, 1, 2, 2, 3, 3), 2), rep(c(0, 4, 4, 5, 5, 0), 2)),
+                 nrow=4, ncol=6, byrow=TRUE))
+for(i in 1:length(eta)) {
+  modes <- c(theta[i]*(sqrt(1 + 9*theta[i]^2/
+                            (4*lambda[i]^2)) - 3*theta[i]/(2*lambda[i])),
+             lambda[i]/(eta[i] + 2))
+  ylim <- c(0, max(digauss(modes[1], theta[i], lambda[i], eta[i]),
+                   digamma(modes[2], theta[i], lambda[i], eta[i])))
+  curve(digauss(x, theta[i], lambda[i], eta[i]), 0.001, 5, n=1000,
+        ylim=ylim, ylab="Density", "x", col=col[1], 
+        main=bquote(eta==.(eta[i])*","~lambda==.(lambda[i])*","~theta==.(
+          theta[i])))
+  curve(digamma(x, theta[i], lambda[i], eta[i]), add=TRUE, 
+        col=col[2], n=1000)
+}
+legend("topright", legend=labels, lty=1, col=col, seg.len=1)
+par(opar)
 @
 
-<<boxplot_igaussian_res2_prior_mean, fig.cap="Prior mean estimates", out.width="80%", fig.asp=2/3>>=
-@
+	In the inverse Gamma model we use the following empirical Bayes procedure: Suppose we have a partitioning of our drugs into drug classes and we let prior variance components $\gamma_d^2$ and $\gamma^2_{d'}$ be drawn from the same distribution if drug $d$ and drug $d'$ are from the same class.
+Or simply put: $\gamma^2_{d}, \gamma^2_{d'} \sim \Gamma^{-1}(\eta_c/2, \lambda_c/2)$ if $d, d' \in \text{class}_c$.
+Empirical bayes estimation of the $\eta_c$ and $\lambda_c$ by maximisation of the (approximated) marginal likelihood amounts to iteratively solving the estimating equations:
+  \begin{subequations}
+    \begin{align}
+      \psi\(\frac{\eta_c}{2}\) & = \log \lambda_c - \frac{\sum_{d \in \text{class}_c} e^{(l)}_d}{|\text{class}_c|} - \log 2, \label{eq:estequation1}\\
+      \lambda_c & = \eta_c \cdot |\text{class}_c| \cdot \(\sum_{d \in \text{class}_c} b_d^{(l)}\)^{-1}, \label{eq:estequation2}
+    \end{align}
+  \end{subequations}
+  until convergence, where $e^{(l)}_d = \E_{Q^{(l)}}(\log \gamma_d^2) = \log \delta_d^{(l)} - \psi [ (p + \eta_d^{(l)})/2] - \log 2$. 
+  We propose to do that by first solving (\ref{eq:estequation2}) for $\lambda_c$, plugging it into (\ref{eq:estequation1}) and bounding the result using \cite{alzer_inequalities_1997} to obtain an interval $( \eta_c^*, 2\eta_c^* )$ containing the solution to (\ref{eq:estequation1}), where:
+  $$
+  \eta_c^* = \left\{ \frac{\sum_{d \in \text{class}_c} e^{(l)}_d}{|\text{class}_c|} + \log \[ \frac{\sum_{d \in \text{class}_c} b_d^{(l)}}{|\text{class}_c|} \]  \right\}^{-1}.
+  $$
+  With this interval it is straightforward to find $\alpha^{(l+1)}_c$ by any root-finding algorithm and plugging the solution into (\ref{eq:estequation2}) to find $\lambda^{(l+1)}_c$.
+We may use $\eta_c^*$ as a starting value.
+
+	The inverse Gamma model with independent drug classes allows for more flexible empirical Bayes estimation of the prior means for the different drug classes.
+As a consequence however, the risk of overfitting is increased.
+Additionally, it does not allow to include continuous covariates on the drugs.
+
+	\section{Non-conjugate model}
+	We call the two versions of
+  the model conjugate (with the $\beta_{jd}$ and $\sigma_d^2$ dependent 
+  \textit{a priori}):
+  $$
+  \beta_{jd} | \gamma_d^2, \sigma_d^2 \sim \mathcal{N} 
+  (0, \sigma_d^2 \gamma_d^2)
+  $$
+  and non-conjugate (with $\beta_{jd}$ and $\sigma_d^2$ independent 
+  \textit{a priori}):
+  $$
+  \beta_{jd} | \gamma_d^2, \sigma_d^2 \sim \beta_{jd} | 
+  \gamma_d^2 \sim \mathcal{N} (0, \gamma_d^2).
+  $$
+
+	The variational distributions are as follows:
+	\begin{align*}
+  q(\bm{\beta}) & \overset{D}{=} \prod_{d=1}^D \mathcal{N}_p (\bm{\mu}_d, \bm{\Sigma}_d), \\
+  q(\bm{\gamma}^2) & \overset{D}{=} \prod_{d=1}^D \mathcal{GIG}\(-\frac{p+\eta_d}{2}, \lambda_d \theta_d^{-2}, \delta_d\), \\
+  q(\bm{\sigma}^2) & \overset{D}{=} \prod_{d=1}^D \Gamma^{-1} \(\frac{df + 1}{2}, \zeta_d\),
+  \end{align*}
+	where $\eta_d=1$ and $\theta_d \to \infty$ in the inverse Gaussian and inverse Gamma models, respectively.
+Furthermore, $df=n + p$ in the conjugate model and $df=n$ in the non-conjugate model.
+The variational parameters contain cyclic dependencies and are, in the conjugate setting, iteratively updated by:
+  \begin{align*}
+  \bm{\Sigma}_d^{(h+1)} & = (a_d^{(h)})^{-1} (\X \tr \X + c_d^{(h)} \I)^{-1}, \\
+  \bm{\mu}_d^{(h+1)} & = (\X \tr \X + c_d^{(h)} \I)^{-1} \X \tr \y_d, \\
+  \delta_d^{(h+1)} & = b_d^{(h)} (c_d^{(h)})^{-1} \bmu_d \tr \bmu_d + \lambda_d, \\ 
+  \zeta_d^{(h+1)} & = \frac{1}{2} \bigg[\mathbf{y}_d \tr \mathbf{y}_d -2 \mathbf{y}_d \tr \X \bm{\mu}_d^{(h+1)} + \trace ( \X \tr \X \bm{\Sigma}_d^{(h+1)}) + (\bm{\mu}_d^{(h+1)}) \tr \X \tr \X \bm{\mu}_d^{(h+1)} \\
+  & \,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\, + \frac{b_d^{(h+1)} - c_d^{(h+1)} a_d^{(h+1)}}{1 - a_d^{(h+1)}} \trace ( \bm{\Sigma}_d^{(h+1)}) + \frac{b_d^{(h+1)} - c_d^{(h+1)} a_d^{(h+1)}}{1 - a_d^{(h+1)}} (\bm{\mu}_d^{(h+1)}) \tr \bm{\mu}_d^{(h+1)}\bigg].
+  \end{align*}
+  Here, we set
+  \begin{align}
+  a_d^{(h)} & =\E_{Q^{(h)}}(\sigma_d^{-2})=(df+1)/(2 \zeta_d^{(h)}), \nonumber \\
+  b_d^{(h)} & = \E_{Q^{(h)}}(\gamma_d^{-2}) = \sqrt{\frac{\lambda_d}{\theta_d^2 \delta_d^{(h)}}} \frac{K_{\frac{p + \eta_d -2}{2}} \( \sqrt{\lambda_d \delta_d^{(h)}}/\theta_d \)}{K_{\frac{p+\eta_d}{2}} \( \sqrt{\lambda_d \delta_d^{(h)}}/\theta_d \)} + \frac{p+\eta_d^{(h)}}{\delta_d^{(h)}}, \label{eq:auxvar1} \\
+  c_d^{(h)} & = \begin{cases*}
+  b_d^{(h)} & in the conjugate setting \\
+  b_d^{(h)}/a_d^{(h)} & in the non-conjugate setting .
+  \end{cases*} \nonumber
+  \end{align}
+  Note that in the inverse Gamma model, the computation of (\ref{eq:auxvar1}) simplifies to $b_d^{(h)}=(p+\eta_d)/\delta_d^{(h)}$.
 
@@
 
-  \section{General model}
-  We have drug covariates $\mathbf{c}_d$ and omics feature covariates $\mathbf{z}_j$. The model is now:
-  \begin{align*}
-  y_{id} | \bm{\beta}_d, \sigma_d^2 & \sim \mathcal{N} (\x_i \tr \bm{\beta}_d, \sigma_d^2), \\
-  \beta_{jd} | \tau_j^2, \gamma_d^2, \sigma_d^2 & \sim \mathcal{N} (0, \sigma_d^2 \gamma_d^2 \tau_j^2), \\
-  \gamma_d^2 & \sim \mathcal{IG}\((\mathbf{c}_d \tr \bm{\alpha})^{-1}, \lambda\), \\
-  \tau_j^2 & \sim \mathcal{IG}\((\mathbf{z}_j \tr \bm{\theta})^{-1}, \nu\), \\
-  \sigma_d^2 & \sim 1/\sigma_d^3.
-  \end{align*}
-  Variational posterior:
-  \begin{align*}
-  q(\bm{\beta}) & \overset{D}{=} \prod_{d=1}^D \mathcal{N}_p (\bm{\mu}_d, \bm{\Sigma}_d), \\
-  q(\bm{\gamma}^2) & \overset{D}{=} \prod_{d=1}^D \mathcal{GIG}\(-\frac{p+1}{2}, \lambda (\mathbf{c}_d \tr \bm{\alpha})^2, \delta_d\), \\
-  q(\bm{\tau}^2) & \overset{D}{=} \prod_{j=1}^p \mathcal{GIG}\(-\frac{D+1}{2}, \nu (\mathbf{z}_j \tr \bm{\theta})^2, \eta_j\), \\
-  q(\bm{\sigma}^2) & \overset{D}{=} \prod_{d=1}^D \Gamma^{-1} \(\frac{n + p + 1}{2}, \zeta_d\).
-  \end{align*}
-  Variational parameters:
-  \begin{align*}
-  \bm{\Sigma}_d^{(h+1)} & = \frac{2 \zeta_d^{(h)}}{n+p+1} \[\X \tr \X + a_d^{(h)} \cdot \diag(b_j^{(h)})\]^{-1}, \\
-  \bm{\mu}_d^{(h+1)} & = \[\X \tr \X + a_d^{(h)} \cdot \diag(b_j^{(h)})\]^{-1} \X \tr \y_d, \\
-  \delta_d^{(h+1)} & = \frac{n+p+1}{2 \zeta_d^{(h)}} \left\{\trace \[\diag(b_j^{(h)}) \bm{\Sigma}_d^{(h+1)} \] + (\bm{\mu}_d^{(h+1)}) \tr \diag(b_j^{(h)}) \bm{\mu}_d^{(h+1)}\right\} + \lambda, \\ 
-  \eta_j^{(h+1)} & = \sum_{d=1}^D \frac{a_d^{(h)}(n+p+1)}{2 \zeta_d^{(h)}} \[ (\bmu_{jd}^{(h + 1)})^2 + (\bSigma_d^{(h + 1)})_{jj} \] + \nu, \\ 
-  \zeta_d^{(h+1)} & = \frac{1}{2} \Big( \mathbf{y}_d \tr \mathbf{y}_d -2 \mathbf{y}_d \tr \X \bm{\mu}_d^{(h+1)} + a_d^{(h+1)} \trace \left\{ \[\X \tr \X + \diag(b_j^{(h + 1)})\] \bm{\Sigma}_d^{(h+1)}\right\} \\
-  & \,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\, + a_d^{(h+1)} (\bm{\mu}_d^{(h+1)}) \tr \[\X \tr \X + \diag(b_j^{(h + 1)})\]\bm{\mu}_d^{(h+1)}\Big),
-  \end{align*}
-  with
-  \begin{align*}
-  a_d^{(h)} & = \sqrt{\frac{\lambda (\mathbf{c}_d \tr \bm{\alpha})^2}{\delta_d^{(h)}}} \frac{K_{\frac{p-1}{2}} \( \sqrt{\lambda (\mathbf{c}_d \tr \bm{\alpha})^2 \delta_d^{(h)}} \)}{K_{\frac{p+1}{2}} \( \sqrt{\lambda (\mathbf{c}_d \tr \bm{\alpha})^2 \delta_d^{(h)}} \)} + \frac{p+1}{\delta_d^{(h)}} \text{ and} \\
-  b_j^{(h)} & = \sqrt{\frac{\nu (\mathbf{z}_j \tr \bm{\theta})^2}{\eta_j^{(h)}}} \frac{K_{\frac{D-1}{2}} \( \sqrt{\nu (\mathbf{z}_j \tr \bm{\theta})^2 \eta_j^{(h)}} \)}{K_{\frac{D+1}{2}} \( \sqrt{\nu (\mathbf{z}_j \tr \bm{\theta})^2 \eta_j^{(h)}} \)} + \frac{D+1}{\eta_j^{(h)}}.
-  \end{align*}
-  Hyper-parameter updates:
-  \begin{align*}
-  \bm{\alpha}^{(l+1)} & = \left[ \mathbf{C} \tr \diag ( e_d^{(l)} ) \mathbf{C} \right]^{-1} \mathbf{C} \tr \mathbf{1}, \\
-  \lambda^{(l+1)} & = \[ \sum_{d=1}^D a_d^{(l)} - (\bm{\alpha}^{(l+1)}) \tr \mathbf{C} \tr \mathbf{1} \]^{-1} D, \\
-  \bm{\theta}^{(l+1)} & = \left[ \mathbf{Z} \tr \diag ( f_j^{(l)} ) \mathbf{Z} \right]^{-1} \mathbf{Z} \tr \mathbf{1}, \\
-  \nu^{(l+1)} & = \[ \sum_{j=1}^p b_j^{(l)} - (\bm{\theta}^{(l+1)}) \tr \mathbf{Z} \tr \mathbf{1} \]^{-1} p,
-  \end{align*}
-  where 
-  \begin{align*}
-  e_d^{(l)} & = \sqrt{\frac{\delta_d^{(l)}}{\lambda^{(l)}(\mathbf{c}_d \tr \bm{\alpha}^{(l)})^2}} \frac{K_{\frac{p - 1}{2}} \( \sqrt{\lambda^{(l)} (\mathbf{c}_d \tr \bm{\alpha}^{(l)})^2 \delta_d^{(l)}} \)}{K_{\frac{p + 1}{2}} \( \sqrt{\lambda^{(l)} (\mathbf{c}_d \tr \bm{\alpha}^{(l)})^2 \delta_d^{(l)}} \)} \text{ and} \\
-  f_j^{(l)} & = \sqrt{\frac{\eta_j^{(l)}}{\nu^{(l)}(\mathbf{z}_j \tr \bm{\theta}^{(l)})^2}} \frac{K_{\frac{D - 1}{2}} \( \sqrt{\nu^{(l)} (\mathbf{z}_j \tr \bm{\theta}^{(l)})^2 \eta_j^{(l)}} \)}{K_{\frac{D + 1}{2}} \( \sqrt{\nu^{(l)} (\mathbf{z}_j \tr \bm{\theta}^{(l)})^2 \eta_j^{(l)}} \)}.
-  \end{align*}
-
   \bibliographystyle{author_short3}  

### 2019-04-24 docs/manuscript1.Rnw
first two simulations ran, intercept added, split functions.R

@@
 \setlength{\textwidth}{15cm}
-\title{Empirical Bayes estimation of simultaneous-equation models}
+\title{Empirical Bayes estimation of the normal inverse Gaussian prior in 
+simultaneous-equation models}
 \date{\today}
@@
 
-	\noindent\textbf{Keywords}: Empirical Bayes; Simultaneous-equations model; 
-	Variational Bayes; Multivariate data
+	\noindent\textbf{Keywords}: ...
 
@@
 
-	- Regression information
+	- Co-data introduction
 
@@
 	- our solution: adaptive borrowing of information through empirical Bayes, 
-	guided by external regression information.
+	guided by co-data.
 
@@
   form of calibration is required. A straightforward way 
-  of doing this is to include the scale the $\bm{\beta}_d$ prior with 
-  $\sigma_d^2$ as in (\ref{eq:betaprior}).
+  of doing this is to scale the $\bm{\beta}_d$ prior with 
+  $\sigma_d^2$ as in (\ref{eq:betaprior}), resulting in a 
+  scale-free prior with respect to the response.
 
@@
   \cite{barndorff-nielsen_normal_1997}, routinely applied in mathematical 
-  finance \cite[]{kalemanova_normal_2007}. In Figure 
+  finance (see, e.g., \cite{kalemanova_normal_2007}). In Figure 
   \ref{fig:dens_igaussian_marginalbeta}, the NIG 
-  prior distribution is depicted together with two common prior choices, ridge 
-  and lasso. From this Figure we see that, depending on the choice
+  prior distribution marginalised over $\gamma_d^2$ is depicted together with 
+  three other common prior choices,
+  Student's $t$ (see Section \ref{sec:inversegammamodel} for more details on 
+  this prior), ridge and lasso, all scaled to variance one. From this Figure 
+  we see that, depending on the choice
   of hyperparameters, the NIG prior can vary from heavier tails than the lasso, 
-  to more Gaussian tails like the ridge.
-<<dens_igaussian_marginalbeta, fig.cap="Marginal $\\beta_j$ prior for several choices of $\\gamma^2_d$ hyperpior, scaled to $\\V(\\beta_j)=1$.", out.width="60%", fig.asp=1>>=
+  to more Gaussian tails like the ridge, but behaves similarly to the 
+  Student's $t$ prior. Our argumentation to model the 
+  $\gamma^2_d$ by an inverse Gaussian 
+  distribution, as has been suggested in \cite{fabrizi_specification_2016} and 
+  \cite{caron_sparse_2008}, is two-fold: (i) the NIG model is more flexible than
+  the standard ridge and lasso priors (as seen from Figure 
+  \ref{fig:dens_igaussian_marginalbeta}), and (ii) the NIG prior allows to model
+  the mean $\theta_d$ as a 
+  function of the co-data $\mathbf{c}_d$ more conveniently than the Student's 
+  $t$ prior, as explained in Section 
+  \ref{sec:empiricalbayes}. 
+<<dens_igaussian_marginalbeta, fig.cap="Several choices of marginal $\\beta_j$ prior, scaled to $\\V(\\beta_j)=1$.", out.width="80%", fig.asp=2/3>>=
 library(GeneralizedHyperbolic)
+library(metRology)
 # all scaled such that variance is one
@@
 lambda1 <- sqrt(2)
-col <- c(1:4)
-lty <- c(1:4)
-labels <- as.expression(c(bquote("IG, "~lambda==.(lambda[1])~", "~
+eta <- lambda/theta + 2
+col <- c(1:6)
+lty <- c(1:6)
+labels <- as.expression(c(bquote("NIG, "~lambda==.(lambda[1])~", "~
                                    theta==.(theta)),
-                          bquote("IG, "~lambda==.(lambda[2])~", "~
+                          bquote("NIG, "~lambda==.(lambda[2])~", "~
                                    theta==.(theta)),
-                          "Point mass (ridge)", "Exponential (lasso)"))
+                          bquote("Student's t, "~lambda==.(lambda[1])~", "~
+                                   eta==.(eta[1])),
+                          bquote("Student's t, "~lambda==.(lambda[2])~", "~
+                                   eta==.(eta[2])),
+                          "ridge", "lasso"))
 dprior <- function(x, lambda, theta, sigma) {
@@
 
+ylim <- c(0, dt.scaled(0, eta[1]/2, 0, sqrt(lambda[1]*sigma^2/eta[1])))
 curve(dprior(x, lambda[1], theta, sigma), -3, 3, ylab=expression(p(beta[j])), 
-      xlab=expression(beta[j]), n=1000, 
+      xlab=expression(beta[j]), n=1000, ylim=ylim,
       col=col[1], lty=lty[1])
@@
 curve(dlasso(x, lambda1), add=TRUE, n=1000, col=col[4], lty=lty[4])
+curve(dt.scaled(x, eta[1]/2, 0, sqrt(lambda[1]*sigma^2/eta[1])), add=TRUE, 
+      n=1000, col=col[5], lty=lty[5])
+curve(dt.scaled(x, eta[2]/2, 0, sqrt(lambda[2]*sigma^2/eta[2])), add=TRUE, 
+      n=1000, col=col[6], lty=lty[6])
 legend("topright", legend=labels, col=col, lty=lty, 
-       title="Hyperprior")
+       title="Prior", seg.len=1)
 @
-
-  Our argumentation to model the $\gamma^2_d$ by an inverse Gaussian 
-  distribution, as has been suggested in \cite{fabrizi_specification_2016} and 
-  \cite{caron_sparse_2008}, is two-fold: (i) the NIG model is more flexible than
-  the standard ridge and (ii) allows to model the mean $\theta_d$ as a 
-  function of the co-data $\mathbf{c}_d$, as explained in Section 
-  \ref{sec:empiricalbayes}.
 
@@
   relatively flat, as may be reasonably expected in the large $p$, small $n$ 
-  setting we are in. We therefore model the error variance with Jeffrey's prior 
+  setting we are in. We therefore model the error variance with Jeffrey's 
+  objective prior 
   \cite[]{jeffreys_invariant_1946}. In the derivation of our Jeffrey's prior
@@
 	and (\ref{eq:priormodel}) is not available in closed form. We therefore 
-	approximate the posterior by variational Bayes, where we force the posterior 
-	density $Q_d$ to factorise as: $p(\bbeta_d, \sigma_d^2, \gamma_d^2) \approx 
-	Q_d(\cdot) = q(\bbeta_d) \cdot q(\sigma_d^2) \cdot q(\gamma_d^2)$.
+	approximate the posterior by variational Bayes, where we force the approximate
+	posterior 
+	density to factorise as: $p(\bbeta_d, \sigma_d^2, \gamma_d^2) \approx 
+	Q_d(\cdot) = q(\bbeta_d) \cdot q(\sigma_d^2) \cdot q(\gamma_d^2)$. For
+	notational convenience, we slightly abuse notation and let $q(\cdot)$ denote
+	different densities for different inputs.
 	Under such a factorisation, the marginal variational posteriors that 
 	minimise the Kullback-Leibler divergence of the true posterior to the 
-	variational Bayes approximation \cite[]{neal_view_1998}, are given by:
+	variational Bayes approximation \cite[]{neal_view_1998} are given by:
 	\begin{align*}
@@
   \end{align*}
-	The variational parameters $\bm{\mu}_d$, $\bm{\Sigma}_d$, $\delta_d$, and
+	See SM Section \ref{sm-sec:vbderivations} for the derivations. The variational
+	parameters $\bm{\mu}_d$, $\bm{\Sigma}_d$, $\delta_d$, and
 	$\zeta_d$ contain cyclic dependencies and are iteratively 
@@
     \bm{\mu}_d^{(h+1)} & = (\X \tr \X + b_d^{(h)} \I)^{-1} \X \tr \y_d, \\
-    \delta_d^{(h+1)} & = \bmu_d \tr \bmu_d + \lambda_d, \\ 
+    \delta_d^{(h+1)} & = a_d^{(h)} \[(\bmu^{(h+1)}_d) \tr \bmu^{(h+1)}_d +
+    \trace (\bSigma_d^{(h+1)})\] + \lambda_d, \\ 
     \zeta_d^{(h+1)} & = \frac{1}{2} \bigg[\mathbf{y}_d \tr \mathbf{y}_d -
@@
   until convergence. Here, we set
-  \begin{align*}
-    a_d^{(h)} & =\E_{Q^{(h)}}(\sigma_d^{-2})=(n + p + 1)/(2 \zeta_d^{(h)}), \\
+  \begin{align}
+    a_d^{(h)} & =\E_{Q^{(h)}}(\sigma_d^{-2})=(n + p + 1)/(2 \zeta_d^{(h)}), 
+    \nonumber\\
     b_d^{(h)} & = \E_{Q^{(h)}}(\gamma_d^{-2}) = \sqrt{\frac{\lambda_d}
@@
     {K_{\frac{p+1}{2}} \( \sqrt{\lambda_d \delta_d^{(h)}}/\theta_d \)} + 
-    \frac{p+1}{\delta_d^{(h)}}.
-  \end{align*}
+    \frac{p+1}{\delta_d^{(h)}}, \label{eq:ratiomodifiedbessel}
+  \end{align}
+  where $K_{\nu}(x)$ denotes the modified Bessel function of the second kind.
+  A method for fast and numerically stable calculation of ratios of such 
+  functions as in (\ref{eq:ratiomodifiedbessel}) is given in SM Section 
+  \ref{sm-sec:ratiosmodifiedbessels}.
 
@@
 	$\bm{\alpha}$. This allows us to include continuous co-data on the outcomes
-	in an objective, data-driven manner. The estimating equations are:
+	in an objective, data-driven manner. Additionally, it reduces the number of
+	hyperparameters even further. The estimating equations are:
   \begin{align*}
@@
 
-  \section{Comparison of NIG model to Normal inverse Gamma}
-  - check model with Stan.
+  Variaional Bayes approximations are known to underestimate posterior 
+  variances \cite[]{rue_approximate_2009,consonni_mean-field_2007,
+  bishop_pattern_2006,wang_inadequacy_2005}. If posterior variances are needed, 
+  we suggest generating samples from the posterior after the empirical Bayes 
+  iterations have converged. A Gibbs sampler to do so is described in 
+  the SM Section \ref{sm-sec:gibbssampler}. Alternatively, the \texttt{R} 
+  package \texttt{rstan} \cite[]{guo_rstan:_2018} allows the user to sample
+  from a generic posterior. An implementation of model
+  (\ref{eq:priormodel}) in \texttt{stan} code may be found on 
+  \url{https://github.com/magnusmunch/cambridge}.
 
-  \section{Comparison of conjugate to non-conjugate}
+  \section{Simulations}
+  \subsection{Comparison of NIG to normal inverse Gamma model}
+  \label{sec:inversegammamodel}
+  A common default prior for latent variance components is the inverse Gamma
+  prior:
+  $$
+  \gamma_d^2 \sim \Gamma^{-1}(\eta_d/2,\lambda_d/2),
+  $$
+  with shape $\eta_d/2$ and scale $\lambda_d/2$, both rescaled to match our 
+  inverse Gaussian prior formulation. The resulting marginal prior for the model
+  parameters
+  (conditional on the error variance $\sigma_d^2$) is then a scaled Student's 
+  $t$-distribution:
+  $$
+  \beta_j | \sigma_d^2 \sim \text{Student's } 
+  t\(\eta_d,\sigma_d \sqrt{\lambda_d/\eta_d}\),
+  $$
+  with degrees of freedom $\eta_d$ and scale $\sigma_d \sqrt{\lambda_d/\eta_d}$.
+  See SM Section \ref{sm-sec:inversegamma} for more details on this prior. 
+  We compare the 
+  posteriors of the proposed NIG prior to the scaled Student's $t$-distribution
+  prior in a simulated setting. We generate draws from these posteriors using 
+  an MCMC sampler implemented in \texttt{rstan}. The aim of this simulation 
+  study is 
+  to investigate the performance of the NIG prior model, so we set $D=1$ and 
+  drop the $d$ indices in the following.
 
-  \section{Variational Bayes versus MCMC}
+  \subsection{Variational Bayes approximation}
+  To investigate the accuracy of our variational Bayes approximation we compare 
+  the resulting posterior to an MCMC sample from the posterior obtained with 
+  \texttt{rstan}. Since the MCMC
+  sample represents the exact posterior (up to Monte Carlo error), we use it as 
+  a proxy
+  for the true posterior. For completeness, we include the comparison to the 
+  automatic differentiation variational inference (ADVI) 
+  algorithm in \texttt{rstan} \cite[]{kucukelbir_automatic_2015}. ADVI is a 
+  variational Bayes algorithm that assumes a full factorization of the posterior
+  and is fit through gradient based optimization of the variational objective.
+  \texttt{rstan} offers an ADVI algorithm that assumes a fullrank posterior 
+  covariance, but this algorithm failed often resulted in errors and
+  convergence problems, so we do not include it in the comparisons.
 
-  \section{Inverse Gaussian Empirical Bayes versus inverse Gamma}
+  To assess the performance of the variational approximations we evaluate 
+  several measures. Divergence between the MCMC and variational posterior
+  is quantified through a multivariate extension of the Cram{\'e}r-von Mises
+  criterion \cite[]{baringhaus_new_2004}, where smaller values of the criterion
+  indicate more similar distributions. In addition, we calculate the correlation
+  and mean squared distance (MSD) between the posterior variances of
+  the MCMC posterior and variational posterior. \texttt{rstan} allows the user 
+  to maximise the posterior to obtain the maximum \textit{a posteriori} (MAP)
+  estimates of the model parameters. We include these MAP estimates together
+  with the ridge estimates obtained from the \texttt{glmnet} \texttt{R} package
+  \cite[]{friedman_regularization_2010} in the comparison of the posterior mean
+  estimates.
 
-  \section{Simulations for VB}
-	\subsection{Setup}
+  The interest of this simulation lies in the variational Bayes approximation, 
+  so we restrict ourselves to the univariate setting (i.e., $D=1$) and drop any
+  $d$ indices from notation in the following. We simulated
+  $n=100$ features $\x_i$ from a $p=200$ dimensional Gaussian distribution with
+  zero mean vector, variances equal to one, and all covariances set to 
+  $\rho=0.5$. The outcomes are generated through $y_i = \x_i\tr \bbeta + 
+  \epsilon_i$, with the $\epsilon_i$ from the standard normal distribution. 
+  Likewise, we generate the model parameters $\beta_j$ from a standard normal
+  distribution. For estimation, we set our hyperparameters $\lambda=0.1$ and 
+  $\theta=2$. We choose the penalty parameter in \texttt{glmnet} by assuming a 
+  known error variance and matching the marginal prior variance of the NIG and
+  ridge models, which results in the penalty parameter 
+  $1/(n\sigma^2\theta)=1/200$.To mitigate the effect of random variation, we
+  repeat the simulation 100 times and report the results in Figures 
+  \ref{fig:boxplots_igaussian_res3_vb}-
+  \ref{fig:boxplots_igaussian_res3_best}.
 
-  We simulate from a correctly specified model given in \ref{eq:priormodel}.
-We assume random $y_{id}$, $\x_i$, and $\bm{\beta}_d$, while we fix $\mathbf{C}$, $\gamma_d^2$, $\sigma_d^2$ and $\bm{\alpha}$.
-We estimate the $\bm{\mu}_d$, $\bm{\Sigma}_d$, $\delta_d$, $\bm{\alpha}$ and $\lambda$, while we fix the $\sigma_d^2$ to their true values.
+<<boxplots_igaussian_res3_vb, fig.cap="(a) Cramér-von Mises criterion, (b) correlation, and (c) MSD between the variances of the MCMC and variational bayes posteriors.", out.width="80%", fig.asp=1/2>>=
+@
 
-  We set $n=100$, $p=200$, and $D=10$.
-We consider 5 evenly sized classes of drugs, such that $\mathbf{c}_{kd} = \mathbbm{1}(k = \text{class}_d)$ and $\begin{bmatrix} \mathbf{c}_1 & \cdots & \mathbf{c}_D \end{bmatrix} \tr$.
-From these the $\gamma_d^2 = (\mathbf{c}_d \tr \bm{\alpha})^{-1}$ are created.
-Next, we simulate $\bm{\beta}_{jd} \sim \mathcal{N} (0, \sigma^2_d \gamma_d^2)$.
-We fix the mean signal-to noise ratio $\overline{\text{SNR}}_d$, such that the predictor data variance is $s^2 = \V(\mathbf{x}_i) = \overline{\text{SNR}}_d / (\overline{\gamma^2_d} p)$ (NOT CORRECT).
-The last step consists of simulating the data by $\mathbf{x}_{ij} \sim \mathcal{N}(0,s^2)$ and $y_{id} = \mathcal{N} (\mathbf{x}_i \tr \bm{\beta}_d, \sigma_d^2)$.
-The remaining parameters are set as follows: $\bm{\alpha} = \begin{bmatrix} 1 & \dots & 5 \end{bmatrix} \tr$, $\sigma_d^2 = 1$ for all $d$, and $\overline{\text{SNR}}_d = 10$.
-To account for the random data and parameter generation, we repeat the simulation 100 times.
-We compare the results to the inverse Gamma and non-conjugate model, and present the results in Figures \ref{fig:boxplot_igaussian_res1_prior_mean}-\ref{fig:boxplot_igaussian_res2_prior_mean}.
+<<boxplots_igaussian_res3_best, fig.cap="(a) Correlation and (b) MSD between the $\\beta_j$ point estimates of MCMC and ADVI, VB, glmnet, and MAP.", out.width="80%", fig.asp=3/5>>=
+@
+
+  \section{Empirical Bayes estimation}
+  We simulate from a correctly specified model given in \ref{eq:priormodel}. We 
+  assume random $y_{id}$, $\x_i$, and $\bm{\beta}_d$, while we fix $\mathbf{C}$,
+  $\gamma_d^2$, $\sigma_d^2$ and $\bm{\alpha}$. We estimate the $\bm{\mu}_d$,
+  $\bm{\Sigma}_d$, $\delta_d$, $\bm{\alpha}$ and $\lambda$, while we fix the
+  $\sigma_d^2$ to their true values.
+
+  We set $n=100$, $p=200$, and $D=10$. We consider 5 evenly sized classes of
+  drugs, such that $\mathbf{c}_{kd} = \mathbbm{1}(k = \text{class}_d)$ and
+  $\begin{bmatrix} \mathbf{c}_1 & \cdots & \mathbf{c}_D \end{bmatrix} \tr$. From
+  these the $\gamma_d^2 = (\mathbf{c}_d \tr \bm{\alpha})^{-1}$ are created. 
+  Next, we simulate $\bm{\beta}_{jd} \sim \mathcal{N} 
+  (0, \sigma^2_d \gamma_d^2)$. We fix the mean signal-to noise ratio 
+  $\overline{\text{SNR}}_d$, such that the predictor data variance is 
+  $s^2 = \V(\mathbf{x}_i) = \overline{\text{SNR}}_d / 
+  (\overline{\gamma^2_d} p)$ (NOT CORRECT). The last step consists of 
+  simulating the data by $\mathbf{x}_{ij} \sim \mathcal{N}(0,s^2)$ and 
+  $y_{id} = \mathcal{N} (\mathbf{x}_i \tr \bm{\beta}_d, \sigma_d^2)$. The
+  remaining parameters are set as follows: $\bm{\alpha} = \begin{bmatrix} 1 &
+  \dots & 5 \end{bmatrix} \tr$, $\sigma_d^2 = 1$ for all $d$, and 
+  $\overline{\text{SNR}}_d = 10$. To account for the random data and parameter
+  generation, we repeat the simulation 100 times. We compare the results to the
+  inverse Gamma and non-conjugate model, and present the results in Figures
+  \ref{fig:boxplot_igaussian_res1_prior_mean}-
+  \ref{fig:boxplot_igaussian_res2_prior_mean}.
 
@@
 
+  - investigate convergence: (i) look at both error and prior variance 
+  convergence and (ii) look at actual prior distribution convergence (
+  plot true distribution and estimated prior every 5 iterations or so)
+
+  - try to fix convergence: (i) track only expectation and variance or ELBO, 
+  (ii) reparametrise, and (iii) fix $\forall d: \lambda_d=\lambda$.
+
   \section{Application}
@@
   \section{Discussion}
+  - inverse Gaussian more convenient than inverse Gamma, because EB is closed-
+  form
+
+  - use copula's and marginal posteriors to find multivariate posteriors for
+  contrast testing.
+
+  - apart from the scaling issue, the conjugate prior model is preferred
+  over the non-conjugate model in \cite{moran_variance_2018}.
+  The assumed factorization in the variational Bayes approximation 
+  $q(\bbeta_d) \cdot q(\sigma_d)$ is actually true for this model. The 
+  approximate expectation used in the empirical Bayes step is therefore more 
+  accurate
 

### 2019-04-24 docs/supplement1.Rnw
first two simulations ran, intercept added, split functions.R

@@
 \setlength{\textwidth}{15cm}
-\title{Supplementary material to: Drug efficacy prediction in cell lines}
+\title{Supplementary material to: Empirical Bayes estimation of the normal
+inverse Gaussian prior in 
+simultaneous-equation models}
 \date{\today}
@@
 	In the following all expectations are with respect to the variational posterior $Q$.
-	\subsection{Evidence lower bound}
+	\begin{align*}
+	\log q_{\B} (\B) & \propto \E_{\mathbf{U}, \bsigma^2} [\log \mathcal{L}(\Y | \B, \U, \bsigma^2 )] + \E_{\bgamma^2, \bsigma^2} [\log \pi (\B | \bgamma^2, \bsigma^2)] \\
+	& \propto - \frac{1}{2} \sum_{d=1}^D \sum_{i=1}^n \E_{\mathbf{u}_d, \sigma_d^2} \[ \frac{(y_{id} - \x_i \tr \bbeta_d - \mathbf{z}_i \tr \mathbf{u}_d)^2}{\sigma_d^2} \] - \frac{1}{2} \sum_{d=1}^D \sum_{j=1}^p \phi_{g(j)}^{-2} \E_{\gamma^2_d, \sigma^2_d} \( \frac{\beta_{jd}^2}{\gamma_d^2 \sigma_d^2} \), \\
+	q_{\B} (\B) & \overset{D}{=} \prod_{d=1}^{D} \mathcal{N}_p (\bmu_d, \bSigma_d), \\
+	& \text{ with } \bSigma_d = \E(\sigma_d^{-2})^{-1} [\X \tr \X + \E(\gamma_d^{-2}) \cdot \diag (\phi_{g(j)}^{-2}) ]^{-1}, \\
+	& \text{ and } \bmu_d = [\X \tr \X + \E(\gamma_d^{-2}) \cdot \diag (\phi_{g(j)}^{-2}) ]^{-1} \X \tr [\y_d - \Z \cdot \E(\mathbf{u}_d)].
+	\end{align*}
+
+	\begin{align*}
+	\log q_{\U}(\U) & \propto \E_{\mathbf{B}, \bsigma^2} [\log \mathcal{L}(\Y | \B, \U, \bsigma^2 )] + \E_{\mathbf{\Xi}_1, \dots, \mathbf{\Xi}_D, \bsigma^2} [\log \pi (\U | \mathbf{\Xi}_1, \dots, \mathbf{\Xi}_D, \bsigma^2)] \\
+	& \propto - \frac{1}{2} \sum_{d=1}^D \sum_{i=1}^n \E_{\bbeta_d, \sigma_d^2} \[ \frac{(y_{id} - \x_i \tr \bbeta_d - \mathbf{z}_i \tr \mathbf{u}_d)^2}{\sigma_d^2} \] - \frac{1}{2} \sum_{d=1}^D \E_{\bm{\Xi}_d, \sigma_d^2}\( \frac{\mathbf{u}_d \tr \bm{\Xi}_d^{-1} \mathbf{u}_d}{\sigma_d^2} \), \\
+	q_{\U}(\U) & \overset{D}{=} \prod_{d=1}^{D} \mathcal{N}_T (\mathbf{m}_d, \Sm_d), \\
+	& \text{ with } \Sm_d = \E(\sigma_d^{-2})^{-1} [\Z \tr \Z + \E(\bm{\Xi}_d^{-1}) ]^{-1}, \\
+	& \text{ and } \mathbf{m}_d = [\Z \tr \Z + \E(\bm{\Xi}_d^{-1}) ]^{-1} \Z \tr [\y_d - \X \cdot \E(\bbeta_d)].
+	\end{align*}
+
+	\begin{align*}
+	\log q_{\bgamma^{2}}(\bgamma^{2}) & \propto \E_{\B, \bsigma^2} [\log \pi (\B | \bgamma^{2}, \sigma^2)] + \log \pi (\bgamma^{2}) \\
+	& \propto -\frac{p}{2} \sum_{d=1}^D \log \gamma_d^{2} - \frac{1}{2} \sum_{d=1}^D \sum_{j=1}^p \phi_{g(j)}^{-2} \E \( \frac{\beta_{jd}^2}{\sigma^{2}_d}\) \gamma_d^{-2} - \frac{3}{2} \sum_{d=1}^D \log \gamma_d^{2} \\
+	& \,\,\,\,\,\,\,\,\,\, - \frac{\lambda}{2 \theta^2} \sum_{d=1}^D (\gamma_d^2 - \theta)^2 \gamma_d^{-2} \\
+	& \propto \sum_{d=1}^D \left\{ \( -\frac{p + 1}{2} - 1\) \log \gamma_d^{2} - \frac{\lambda}{2\theta^2} \gamma_d^2 - \frac{1}{2}\[ \lambda + \E(\sigma_d^{-2}) \sum_{j=1}^p \frac{\E(\beta_{jd}^2)}{\phi_{g(j)}^{2}}\] \gamma_d^{-2} \right\}, \\
+	q_{\bgamma^{2}}(\bgamma^{2}) & \overset{D}{=} \prod_{d=1}^D \mathcal{GIG} \(-\frac{p+1}{2}, \frac{\lambda}{\theta_d^2}, \delta_{d} \), \\
+	& \text{with } \delta_{d} =\E(\sigma_d^{-2}) \sum_{j=1}^p \frac{\E(\beta_{jd})^2 + \V(\beta_{jd})}{\phi_{g(j)}^2} + \lambda.
+	\end{align*}
+
+	\begin{align*}
+	\log q_{\bsigma^{2}}(\bsigma^{2}) & \propto \E_{\B,\mathbf{U}} [\log \mathcal{L}(\Y | \B, \U, \bsigma^2 )] + \E_{\B, \bgamma^2} [\log \pi (\B | \bgamma^2, \bsigma^2)] \\
+	& \,\,\,\,\,\,\,\,\,\, + \E_{\mathbf{U}, \bm{\Xi}_1, \dots, \bm{\Xi}_D} [\log \pi (\mathbf{U} | \bm{\Xi}_1, \dots, \bm{\Xi}_D, \bsigma^2)] + \log \pi (\bsigma^{2}) \\
+	& \propto -\frac{n}{2} \sum_{d=1}^D \log \sigma_d^{2} - \frac{1}{2} \sum_{d=1}^D \sum_{i=1}^n \E [(y_{id} - \x_i \tr \bbeta_d - \mathbf{z}_i \tr \mathbf{u}_d)^2] \sigma_d^{2} - \frac{p}{2} \sum_{d=1}^D \log \sigma_d^{-2} \\
+	& \,\,\,\,\,\,\,\,\,\, - \frac{1}{2} \sum_{d=1}^D \sum_{j=1}^p \E( \gamma_d^{-2}) \frac{\E ( \beta_{jd}^2)}{\phi_{g(j)}^{2}} \sigma_d^{-2} - \frac{T}{2} \sum_{d=1}^D \log \sigma_d^{2} - \frac{1}{2} \sum_{d=1}^D \E(\mathbf{u}_d \tr \bm{\Xi}_d^{-1} \mathbf{u}) \sigma_d^{-2} \\
+	& \,\,\,\,\,\,\,\,\,\, - \frac{3}{2} \sum_{d=1}^D \log \sigma^2_d \\
+	& = -\sum_{d=1}^{D} \(\frac{n + p + T + 3}{2} \) \log \sigma_d^{2} - \sum_{d=1}^{D} \frac{1}{2}\Bigg\{ \sum_{i=1}^n \E [(y_{id} - \x_i \tr \bbeta_d - \mathbf{z}_i \tr \mathbf{u}_d)^2] \\
+	& \,\,\,\,\,\,\,\,\,\, + \E(\gamma_d^{-2})\sum_{j=1}^p \frac{\E(\beta_{jd}^2)}{\phi_{g(j)}^2} + \E(\mathbf{u}_{d} \tr \bm{\Xi}_d^{-1} \mathbf{u}_d) \Bigg\} \sigma_d^{-2}, \\
+	q_{\bsigma^{2}}(\bsigma^{2}) & \overset{D}{=} \prod_{d=1}^D \Gamma^{-1} \(\frac{n + p + T + 1}{2}, \zeta_{d} \), \\
+	& \text{ with } \zeta_{d} = \frac{\y_d \tr \y_d}{2} - \y_d \tr [\X \E(\bbeta_d) + \Z \E(\mathbf{u}_d)] + \E(\bbeta_d \tr) \X \tr \Z \E (\mathbf{u}_d) \\
+	& \,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\, + \frac{\trace [\X \tr \X \V(\bbeta_d)]}{2} + \frac{\E(\bbeta_d \tr) \X \tr \X \E(\bbeta_d)}{2} + \frac{\trace [\Z \tr \Z \V(\mathbf{u}_d)]}{2} \\
+	& \,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\, + \frac{\E(\mathbf{u}_d \tr) \Z \tr \Z \E(\mathbf{u}_d)}{2} + \frac{\E(\gamma_d^{-2})}{2} \sum_{j=1}^p \frac{\E(\beta_{jd})^2 + \V(\beta_{jd})}{\phi_{g(j)}^2} \\
+	& \,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,+ \frac{\E(\mathbf{u}_d \tr) \E(\bm{\Xi}_d^{-1}) \E(\mathbf{u}_d)}{2} + \frac{\trace [\E(\bm{\Xi}_d^{-1}) \V(\mathbf{u}_d)]}{2},
+	\end{align*}
+
+	\begin{align*}
+	\log q_{\bm{\Xi}_1, \dots, \bm{\Xi}_D}(\bm{\Xi}_1, \dots, \bm{\Xi}_D) & \propto \E_{\mathbf{U}, \bm{\sigma}^2} [\log \pi (\mathbf{U} | \bm{\Xi}_1, \dots, \bm{\Xi}_D, \bsigma^2)] + \log \pi (\bm{\Xi}_1, \dots, \bm{\Xi}_D) \\
+	& \propto -\frac{1}{2} \sum_{d=1}^D \log |\bm{\Xi}_d| - \frac{1}{2} \sum_{d=1}^D \E \(\frac{\mathbf{u}_d \tr \bm{\Xi}_d^{-1} \mathbf{u}_d}{\sigma_d^2} \) \\
+	& \,\,\,\,\,\,\,\,\,\, - \frac{k + T + 1}{2} \sum_{d=1}^D \log |\bm{\Xi}_d| - \frac{1}{2} \sum_{d=1}^D \trace (\bm{\Omega} \bm{\Xi}_d^{-1}) \\
+	& = -\frac{\tau + T + 2}{2} \sum_{d=1}^D \log |\bm{\Xi}_d| - \frac{1}{2} \sum_{d=1}^D \trace \left\{ \[\E(\sigma_d^{-2}) \E( \mathbf{u}_d \mathbf{u}_d \tr) + \bm{\Omega}\] \bm{\Xi}_d^{-1} \right\}, \\
+	q_{\bm{\Xi}_1, \dots, \bm{\Xi}_D}(\bm{\Xi}_1, \dots, \bm{\Xi}_D) & \overset{D}{\propto} \prod_{d=1}^D \mathcal{W}_T^{-1} (\bm{\Psi}_d, \nu + 1),  \\ 
+	& \text{ with } \bm{\Psi}_d =\E(\sigma^{-2}_d) \[\V(\mathbf{u}_d) + \E(\mathbf{u}_d) \E( \mathbf{u}_d \tr) \] + \bm{\Omega}.
+	\end{align*}
+
+	\section{Evidence lower bound}\label{sec:evidencelowerbound}
 	We monitor convergence through the evidence lower bound (ELBO).
@@
 
-	\section{Evidence lower bound}
 	We monitor convergence through the evidence lower bound (ELBO).
@@
 
-  \subsection{Efficient computation}
+  \section{Efficient computation}
   The empirical Bayes updates require the following quantities: $\zeta_d$, $a_d$, $\delta_d$, $\trace (\bm{\Sigma}_d)$, $\trace (\X \tr \X \bm{\Sigma}_d)$, $\bm{\mu}_d \tr \bm{\mu}_d$, and $\bm{\mu}_d \tr \X \tr \X \bm{\mu}_d$.
@@
 
-  \section{Inverse Gamma model}
+  \section{Ratios of modified Bessel functions}\label{sec:ratiosmodifiedbessels}
+  Ratios of modified Bessel functions of the second kind 
+  $K_{\alpha - 1}(x)/K_{\alpha}(x)$ are prone to under- and overflow for large
+  $\alpha$. In our case, $\alpha$ increases linearly with $p$. Since $p$ may be
+  large, this causes numerical issues in the calculation of various quantities.
+  We alleviate the numerical issues through the following.
+
+  We let $n_1=p/2$ and $n_2=(p-1)/2$ and use the well-known recursive relation:
+  $$
+  K_{\alpha}(x) = K_{\alpha - 2}(x) + \frac{2(\alpha - 1)}{2} K_{\alpha- 1}(x),
+  $$
+  to rewrite the ratio:
+  \begin{align*}
+    \frac{K_{\frac{p - 1}{2}}(x)}{K_{\frac{p + 1}{2}}(x)} = 
+    \begin{cases}
+      \Big( \dots \Big( \( 1 + \frac{2 \cdot 1 - 1}{x} \)^{-1} + 
+      \frac{2 \cdot 2-1}{x}\Big)^{-1} + \dots + 
+      \frac{2 \cdot n_1 - 1}{x}\Big)^{-1}, & \text{for } p \text{ even}, \\
+      \Big( \dots \Big( \( \frac{K_0(x)}{K_1(x)} + \frac{2}{x} \cdot 1 \)^{-1} +
+      \frac{2}{x} \cdot 2 \Big)^{-1} + \dots + \frac{2}{x} \cdot n_2 \Big)^{-1},
+      & \text{for } p \text{ odd}.
+    \end{cases}
+  \end{align*} 
+  The ratio $K_0(x)/K_1(x)$ is well-behaved, such that the ratio may be computed
+  without numerical issues.
+
+  \section{Gibbs sampler}\label{sec:gibbssampler}
+  MCMC samples from the posterior corresponding to model 
+  (\ref{md-eq:priormodel}) may be generated for each equation independently. 
+  For clarity, we therefore omit the index $d$ in the following. MCMC samples 
+  may be generated by iteratively sampling the following conditional
+  distributions:
+  \begin{align}
+    \bm{\beta} | \gamma^2, \sigma^2, \y & \sim \mathcal{N}_p 
+    (\bm{\mu}, \bm{\Sigma}), \label{eq:betasample} \\
+    \gamma^2 | \bm{\beta}, \sigma^2, \y & \sim \mathcal{GIG}
+    \(-\frac{p+1}{2}, \lambda \theta^{-2}, \delta\), \nonumber \\
+    \sigma^2 | \bm{\beta}, \gamma^2, \y & \sim \Gamma^{-1} 
+    \(\frac{n + p + 1}{2}, \zeta\), \nonumber
+  \end{align}
+  with
+  \begin{subequations}\label{eq:gibbsparameters}
+    \begin{align}
+      \bm{\Sigma} & = \sigma^2 (\X \tr \X + \gamma^{-2} \I)^{-1}, \\
+      \bm{\mu} & = (\X \tr \X + \gamma^{-2} \I)^{-1} \X \tr \y, \\
+      \delta & = \sigma^{-2} \bm{\beta} \tr \bm{\beta} + \lambda, \nonumber \\ 
+      \zeta & = \frac{1}{2} \[\mathbf{y} \tr \mathbf{y} -
+      2 \mathbf{y} \tr \X \bm{\mu} + \trace ( \X \tr \X \bm{\Sigma}) + 
+      (\bm{\mu}) \tr \X \tr \X \bm{\mu}  + 
+      \gamma^{-2} \trace ( \bm{\Sigma}) + 
+      \gamma^{-2} (\bm{\mu}) \tr \bm{\mu}\]. \nonumber
+    \end{align}
+  \end{subequations}
+  In high dimensional space, the $p times p$ matrix inversions in 
+  (\ref{eq:gibbsparameters}) are   
+  significant computational bottlenecks. \cite{bhattacharya_fast_2016} describe a
+  method to sample from (\ref{eq:betasample}) without explicit calculation
+  of this inverse, thereby offering a siginifcant speed-up compared to naive
+  sampling.
+
+  \section{Inverse Gamma model}\label{sec:inversegamma}
   In addition to the inverse Gaussian model, we investigate a generalisation of 
@@
   where we draw the $\gamma_d^2$ from an inverse Gamma distribution:
-	$$
-	\gamma_d^2 \sim \Gamma^{-1}(\eta_d/2, \lambda_d/2),
-	$$
+	\begin{equation}\label{eq:invgammaprior}
+	  \gamma_d^2 \sim \Gamma^{-1}(\eta_d/2, \lambda_d/2)=
+	  \mathcal{GIG}(-\eta_d/2, 0, \lambda_d),
+	\end{equation}
 	with shape $\eta_d/2$ and scale $\lambda/2$. For the sake of comparability, we
-	have rescaled the inverse Gamma parameters and reparametrised both 
-	distributions as generalized inverse Gaussians:
-	\begin{align*}
-  \gamma_d^2 & \sim \mathcal{GIG}(-1/2, \lambda_d \theta_d^{-2}, \lambda_d), \\
-  \gamma_d^2 & \sim \mathcal{GIG}(-\eta_d/2, 0, \lambda_d).
-  \end{align*}
-	The two models coincide if $\theta_d \to \infty$ and $\eta_d = 1$.
+	have rescaled the inverse Gamma parameters and reparametrised as a generalized
+	inverse Gaussian. This allows for an easier comparison to the inverse 
+	Gaussian hyperprior:
+	\begin{equation}\label{eq:invgaussianprior}
+	  \mathcal{IG}(\theta_d, \lambda_d) =
+	  \mathcal{GIG}(-1/2, \lambda_d \theta_d^{-2}, \lambda_d).
+  \end{equation}
+	From (\ref{eq:invgammaprior}) and (\ref{eq:invgaussianprior}) it is 
+	that the two models coincide if 
+	$\theta_d \to \infty$ and $\eta_d = 1$.
 
@@
   \end{align*}
-  Here, we set
-  \begin{align}
-  a_d^{(h)} & =\E_{Q^{(h)}}(\sigma_d^{-2})=(df+1)/(2 \zeta_d^{(h)}), \nonumber \\
-  b_d^{(h)} & = \E_{Q^{(h)}}(\gamma_d^{-2}) = \sqrt{\frac{\lambda_d}{\theta_d^2 \delta_d^{(h)}}} \frac{K_{\frac{p + \eta_d -2}{2}} \( \sqrt{\lambda_d \delta_d^{(h)}}/\theta_d \)}{K_{\frac{p+\eta_d}{2}} \( \sqrt{\lambda_d \delta_d^{(h)}}/\theta_d \)} + \frac{p+\eta_d^{(h)}}{\delta_d^{(h)}}, \label{eq:auxvar1} \\
-  c_d^{(h)} & = \begin{cases*}
-  b_d^{(h)} & in the conjugate setting \\
-  b_d^{(h)}/a_d^{(h)} & in the non-conjugate setting .
-  \end{cases*} \nonumber
-  \end{align}
-  Note that in the inverse Gamma model, the computation of (\ref{eq:auxvar1}) simplifies to $b_d^{(h)}=(p+\eta_d)/\delta_d^{(h)}$.
+  Here, we set $a_d^{(h)}$, $b_d^{(h)}$, and $c_d^{(h)}$ as in 
+  (\ref{eq:auxvar1})
+  Note that in the inverse Gamma model, the computation of (\ref{eq:auxvar1})
+  simplifies to $b_d^{(h)}=(p+\eta_d)/\delta_d^{(h)}$.
 
@@
 
-  \section{Ratios of modified Bessel functions}
-  Ratios of modified Bessel functions of the second kind $K_{\alpha - 1}(x)/K_{\alpha}(x)$ are prone to under- and overflow for large $\alpha$.
-In our case, $\alpha$ increases linearly with $p$.
-Since $p$ may be large, this causes numerical issues in the calculation of various quantities.
-We alleviate the numerical issues through the following.
-
-  We let $n_1=p/2$ and $n_2=(p-1)/2$ and use the well-known recursive relation:
-  $$
-  K_{\alpha}(x) = K_{\alpha - 2}(x) + \frac{2(\alpha - 1)}{2} K_{\alpha- 1}(x),
-  $$
-  to rewrite the ratio:
-  \begin{align*}
-  \frac{K_{\frac{p - 1}{2}}(x)}{K_{\frac{p + 1}{2}}(x)} = 
-  \begin{cases}
-  \Big( \dots \Big( \( 1 + \frac{2 \cdot 1 - 1}{x} \)^{-1} + \frac{2 \cdot 2-1}{x}\Big)^{-1} + \dots + \frac{2 \cdot n_1 - 1}{x}\Big)^{-1}, & \text{for } p \text{ even}, \\
-  \Big( \dots \Big( \( \frac{K_0(x)}{K_1(x)} + \frac{2}{x} \cdot 1 \)^{-1} + \frac{2}{x} \cdot 2 \Big)^{-1} + \dots + \frac{2}{x} \cdot n_2 \Big)^{-1}, & \text{for } p \text{ odd}.
-  \end{cases}
-  \end{align*} 
-  The ratio $K_0(x)/K_1(x)$ is well-behaved, such that the ratio may be computed without numerical issues.
-
   \bibliographystyle{author_short3}  

### 2019-04-24 rpackage/R/aux_functions.R

@@ -1,0 +1,165 @@
+# calculates the auxiliary variables in the VB step (tested)
+.aux.var <- function(cold, aold, sv, uty, n, p) {
+  
+  # auxiliary variables involving mu and Sigma
+  trSigma <- (sum(1/(sv^2 + cold)) + max(p - n, 0)/cold)/aold
+  trXtXSigma <- sum(sv^2/(sv^2 + cold))/aold
+  mutmu <- sum(sv^2*uty^2/(sv^2 + cold)^2)
+  mutXtXmu <- sum(sv^4*uty^2/(sv^2 + cold)^2)
+  logdetSigma <- -p*log(aold) - sum(log(sv^2 + cold)) - 
+    max(p - n, 0)*log(cold)
+  ytXmu <- sum(sv^2*uty^2/(sv^2 + cold))
+  
+  out <- list(trSigma=trSigma, trXtXSigma=trXtXSigma, mutmu=mutmu,
+              mutXtXmu=mutXtXmu, logdetSigma=logdetSigma,
+              ytXmu=ytXmu)
+  return(out)
+}
+
+# calculates the auxiliary variables in the VB step with intercept (not tested)
+.aux.var.int <- function(cold, aold, sv, uty, sumu, sumy, n, p) {
+  
+  s <- 1/(n - sum(sumu^2*sv^2/(sv^2 + cold)))
+  
+  aux1 <- sum(sv^2*sumu^2/(sv^2 + cold))
+  aux2 <- sum(sv^2*uty^2/(sv^2 + cold))
+  aux3 <- sum(sv^2*sumu*uty/(sv^2 + cold))
+  aux4 <- sum(sv^4*sumu^2/(sv^2 + cold)^2)
+  aux5 <- sum(sv^4*uty^2/(sv^2 + cold)^2)
+  aux6 <- sum(sv^4*sumu*uty/(sv^2 + cold)^2)
+  
+  # auxiliary variables involving mu and Sigma
+  trSigma <- (sum(1/(sv^2 + cold)) + s*sum(sv^2*sumu^2/(sv^2 + cold)^2) + 
+                max(p - n, 0)/cold)/aold 
+  trXtXSigma <- (sum(sv^2/(sv^2 + cold)) + s*(n - 2*aux1 + aux4))/aold
+  mutmu <- s^2*(sumy^2 - 2*sumy*aux3 + aux3^2)*
[…]

### 2019-04-24 code/simulations_igaussian.R
first two simulations ran, intercept added, split functions.R

@@ -7,9 +7,9 @@
 # library(devtools)
 # install_github("magnusmunch/cambridge/rpackage", local=FALSE,
 #                auth_token=Sys.getenv("GITHUB_PAT"))
-if(!("statmod" %in% installed.packages())) {
-  install.packages("statmod")
-}
+
+### parallelisation
+parallel <- FALSE
 
 ### libraries
 library(cambridge)
@@ -19,6 +19,8 @@ library(rstan)
 library(glmnet)
 library(mvtnorm)
 library(cramer)
+library(foreach)
+library(doParallel)
 
 ################################# simulation 1 #################################
 # settings
@@ -233,20 +235,22 @@ warnings()
 # settings
 nreps <- 100
 n <- 100
-p <- 30
+p <- 200
 D <- 1
 
 # create fixed parameters
+rho <- 0.5
+Sigma <- matrix(rho, ncol=p, nrow=p); diag(Sigma) <- 1
 gamma <- rep(1, D)
 sigma <- rep(1, D)
 lambda <- 0.1
 theta <- 2
 
 # store settings in an object
-temp1 <- c(nreps=nreps, n=n, p=p, D=D, gamma=gamma, sigma=sigma, lambda=lambda,
-           theta=theta)
-set1 <- t(matrix(temp1))
-colnames(set1) <- names(temp1)
+set <- data.frame(nreps=nreps, n=n, p=p, D=D, gamma=t(as.matrix(gamma)), 
+                  sigma=t(as.matrix(sigma)), 
+                  lambda=lambda, theta=theta)
+rownames(set) <- paste0("set", 1:nrow(set))
 
 # objects to store simulation results
 methods <- c("MCMC", "ADVI", "VB", "MAP", "glmnet")
@@ -254,155 +258,216 @@ params <- c("beta0", "beta", "sigma", "gamma")
 
 # set initial values and control parameters
 stan.model <- stan_model("code/igauss.stan")
+nsamples <- 1000
+nwarmup <- 1000
 
-
-
-# simulation
-set.seed(2019)
-beta <- sapply(1:D, function(d) {rnorm(p, 0, sigma[d]*gamma[d])})
-x <- matrix(rnorm(n*p, 0, 1), nrow=n, ncol=p)
-y <- sapply(1:D, function(d) {rnorm(n, x %*% beta[, d], sigma[d])})
-
-
-
-
-fit.mcmc <- sapply(1:D, function(d) {
[…]

### 2019-04-24 rpackage/R/functions.R

@@ -1,736 +1,0 @@
-################################# VB functions #################################
-# calculates the auxiliary variables in the VB step (tested)
-.aux.var <- function(cold, aold, sv, uty, n, p) {
-  
-  # auxiliary variables involving mu and Sigma
-  trSigma <- (sum(1/(sv^2 + cold)) + max(p - n, 0)/cold)/aold
-  trXtXSigma <- sum(sv^2/(sv^2 + cold))/aold
-  mutmu <- sum(sv^2*uty^2/(sv^2 + cold)^2)
-  mutXtXmu <- sum(sv^4*uty^2/(sv^2 + cold)^2)
-  logdetSigma <- -p*log(aold) - sum(log(sv^2 + cold)) - 
-    max(p - n, 0)*log(cold)
-  ytXmu <- sum(sv^2*uty^2/(sv^2 + cold))
-  
-  out <- list(trSigma=trSigma, trXtXSigma=trXtXSigma, mutmu=mutmu,
-              mutXtXmu=mutXtXmu, logdetSigma=logdetSigma,
-              ytXmu=ytXmu)
-  return(out)
-}
-
-# calculates the auxiliary variables in the VB step with intercept (not tested)
-.aux.var.int <- function(cold, aold, sv, uty, sumu, sumy, n, p) {
-  
-  s <- 1/(n - sum(sumu^2*sv^2/(sv^2 + cold)))
-  
-  aux1 <- sum(sv^2*sumu^2/(sv^2 + c))
-  aux2 <- sum(sv^2*uty^2/(sv^2 + c))
-  aux3 <- sum(sv^2*sumu*uty/(sv^2 + c))
-  aux4 <- sum(sv^4*sumu^2/(sv^2 + c)^2)
-  aux5 <- sum(sv^4*uty^2/(sv^2 + c)^2)
-  aux6 <- sum(sv^4*sumu*uty/(sv^2 + c)^2)
-  
-  # auxiliary variables involving mu and Sigma
-  trSigma <- (sum(1/(sv^2 + cold)) + s*sum(sv^2*sumu^2/(sv^2 + c)^2) + 
-                max(p - n, 0)/cold)/aold 
-  trXtXSigma <- (sum(sv^2/(sv^2 + cold)) + s*(n - 2*aux1 + aux4))/aold
-  mutmu <- s^2*(sumy^2 - 2*sumy*aux3 + aux3^2)*
[…]

### 2019-05-13 docs/supplement1.Rnw
changed some nomenclature and added nigig model

--- 2019-04-17 docs/supplement1.Rnw
@@
-% load packages
-<<packages, include=FALSE, echo=FALSE>>=
-library(cambridge)
-@
-
 % set options
@@
 opts_chunk$set(fig.align='center', fig.path="../figs/", 
-               echo=FALSE, cache=FALSE, message=FALSE)
+               echo=FALSE, cache=FALSE, message=FALSE, fig.pos='!ht')
+knit_hooks$set(document=function(x) {
+  sub('\\usepackage[]{color}', '\\usepackage{xcolor}', x, fixed=TRUE)})
 @
@@
 	\section{Content overview}
-	This document contains the Supplementary Material (SM) to the document `Empirical Bayes estimation of simultaneous-equation models'.
-In the following, this document is referred to as Main Document (MD).
+	This document contains the Supplementary Material (SM) to the document 
+	`Empirical Bayes estimation of simultaneous-equation models'. In the 
+	following, this document is referred to as Main Document (MD).
+
+	\section{NIG prior}\label{sec:nigprior}
+	For a fixed error variance $\sigma_d^2$, we have a marginal prior variance 
+	and kurtosis:
+	\begin{align*}
+	  \V(\beta_{jd} | \sigma_d^2) & = \sigma_d^2 \theta_d, \\
+	  \mathcal{K}(\beta_{jd} | \sigma_d^2) & = \frac{3\theta_d}{\lambda_d} + 3.
+	\end{align*}
+
+	\section{NIGIG prior}\label{sec:nigigprior}
+	For a fixed error variance $\sigma_d^2$, we have a marginal prior variance 
+	and kurtosis:
+	\begin{align*}
+	  \V(\beta_{jd} | \sigma_d^2) & = \sigma_d^2 \theta_d^c \theta_j^z  
+    \(r_d + \frac{2\theta_d^c}{\lambda_d^c}\), \\
+	  \mathcal{K}(\beta_{jd} | \sigma_d^2) & = \(\frac{3\theta_j^z}{\lambda_j^z} +
+	  3\) \frac{4 \theta_d^c (r_d + 2\theta_d^c/\lambda_d^c) + 1}{
+	  \lambda_d^c(r_d + 2\theta_d^c/\lambda_d^c)^2},
+	\end{align*}
+	with $r_d = K_0(\lambda_d^c/\theta_d^c)/K_1(\lambda_d^c/\theta_d^c)$.
+
+	\section{Student's \texorpdfstring{$t$}{t} prior}\label{sec:studentstprior}
+	For a fixed error variance $\sigma_d^2$, we have a marginal prior variance 
+	and kurtosis:
+	\begin{align*}
+	  \V(\beta_{jd} | \sigma_d^2) & = \frac{\sigma_d^2 \lambda_d}{\eta_d - 2}, \\
+	  \mathcal{K}(\beta_{jd} | \sigma_d^2) & = \frac{6}{\eta_d - 4} + 3.
+	\end{align*}
 
 	\section{Variational Bayes derivations}\label{sec:vbderivations}
-	In the following all expectations are with respect to the variational posterior $Q$.
-	\subsection{Evidence lower bound}
-	We monitor convergence through the evidence lower bound (ELBO).
-To compute the evidence lower bound after updating the EB parameters from $\eta_d^{(l)}, \theta_d^{(l)}, \lambda_d^{(l)}$ to $\eta_d^{(l + 1)}, \theta_d^{(l + 1)}, \lambda_d^{(l + 1)}$, we compute $\text{ELBO}^{(l + 1)} = \sum_{d=1}^D \text{ELBO}_d^{(l + 1)}$, with, in the inverse Gaussian case:
+	For clarity we have indexed the variational 
+	density functions with their respective parameters, which we omitted in the 
+	MD. In the following all expectations are with respect to the 
+	variational posterior $Q_d$. The variational posterior for $\bbeta_d$ is 
+	found as follows:
 	\begin{align*}
-	\text{ELBO}_d^{(l + 1)} & = \frac{1}{2} \log |\bm{\Sigma}_d^{(l)}| - \frac{a_d^{(l)}}{2} \[ \y_d \tr \y_d - 2\y_d \tr \X \bm{\mu}_d^{(l)} + (\bm{\mu}_d^{(l)}) \tr \X \tr \X \bm{\mu}_d^{(l)} + \trace ( \X \tr \X \bm{\Sigma}_d^{(l)})\] \\
-	& + \frac{b_d^{(l)}}{2} \[ \delta_d^{(l)} - (a_d^{(l)} + 1 - b_d^{(l)}/c_d^{(l)})(\bm{\mu}_d^{(l)}) \tr \bm{\mu}_d^{(l)} - (a_d^{(l)} + 1 - b_d^{(l)}/c_d^{(l)}) \trace (\bm{\Sigma}_d^{(l)}) \] \\
-	& - \frac{df + 1}{2} \log \zeta_d^{(l)} + \frac{p + 1}{4} \log \lambda_d^{(l)} - \frac{p + 1}{2} \log \theta_d^{(l)} - \frac{p+1}{4} \log \delta_d^{(l)} \\
-	& + \log K_{\frac{p+1}{2}}\(\sqrt{\lambda_d^{(l)} \delta_d^{(l)}}/\theta_d^{(l)}\) + \frac{\lambda_d^{(l)} e_d^{(l)}}{2(\theta_d^{(l)})^2} \\
-	& + \frac{\lambda_d^{(l+1)}}{\theta_d^{(l+1)}} + \frac{1}{2}\log \lambda_d^{(l+1)} - \frac{\lambda_d^{(l+1)} e_d^{(l)}}{2 (\theta_d^{(l + 1)})^2} - \frac{\lambda_d^{(l + 1)} b_d^{(l)}}{2}.
+	  \log q_{\bbeta_d} (\bbeta_d) & \propto \E [\log \mathcal{L}
+	  (\y_d | \bbeta_d, \sigma_d^2 )] + \E [\log \pi (\bbeta_d | \gamma_d^2, 
+	  \sigma_d^2)] \\
+	  & \propto - \frac{1}{2} \sum_{i=1}^n 
+	  \E \[ \frac{(y_{id} - \x_i \tr \bbeta_d)^2}{\sigma_d^2} \] - 
+	  \frac{1}{2} \sum_{j=1}^p 
+	  \E \( \frac{\beta_{jd}^2}{\gamma_d^2 \sigma_d^2} \), \\
+	  q_{\bbeta_d} (\bbeta_d) & \overset{D}{=} 
+	  \mathcal{N}_p (\bmu_d, \bSigma_d), \\
+	  & \text{with } \bSigma_d = \E(\sigma_d^{-2})^{-1} 
+	  [\X \tr \X + \E(\gamma_d^{-2}) \cdot \I ]^{-1}, \\
+	  & \text{and } \bmu_d = [\X \tr \X + \E(\gamma_d^{-2}) \cdot \I ]^{-1} 
+	  \X \tr \y_d.
+	\end{align*}
+	The variational posterior for $\gamma_d^2$ is given by:
+	\begin{align*}
+	  \log q_{\gamma_d^{2}}(\gamma_d^{2}) & \propto \E 
+	  [\log \pi (\bbeta_d | \gamma_d^{2}, \sigma_d^2)] + \log \pi(\gamma_d^{2}) \\
+	  & \propto -\frac{p}{2} \log \gamma_d^{2} - \frac{1}{2} \sum_{j=1}^p 
+	  \E \( \frac{\beta_{jd}^2}{\sigma^{2}_d}\) \gamma_d^{-2} - 
+	  \frac{3}{2} \log \gamma_d^{2} - 
+	  \frac{\lambda_d (\gamma_d^2 - \theta_d)^2}{2 \theta_d^2} \gamma_d^{-2} \\
+	  & \propto \( -\frac{p + 1}{2} - 1\) \log \gamma_d^{2} - 
+	  \frac{\lambda_d}{2\theta_d^2} \gamma_d^2 - 
+	  \frac{1}{2}\[ \lambda_d + \E(\sigma_d^{-2}) \sum_{j=1}^p \E(\beta_{jd}^2)\] 
+	  \gamma_d^{-2} , \\
+	  q_{\gamma_d^{2}}(\gamma_d^{2}) & \overset{D}{=} \mathcal{GIG} 
+	  \(-\frac{p+1}{2}, \frac{\lambda_d}{\theta_d^2}, \delta_{d} \), \\
+	  & \text{with } \delta_{d} =\E(\sigma_d^{-2})
+	  \left\{\E(\bbeta_{d} \tr) \E(\bbeta_{d}) + \trace[\V(\bbeta_{d})]\right\} + 
+	  \lambda_d.
+	\end{align*}
+	Lastly, we have the variational posterior for $\sigma_d^2$:
+	\begin{align*}
+	  \log q_{\sigma_d^{2}}(\sigma_d^{2}) & \propto \E 
+	  [\log \mathcal{L}(\y_d | \bbeta)d, \sigma_d^2 )] + 
+	  \E [\log \pi (\bbeta_d | \gamma_d^2, \sigma_d^2)] + 
+	  \log \pi (\sigma_d^{2}) \\
+	  & \propto -\frac{n}{2} \log \sigma_d^{2} - \frac{1}{2} \sum_{i=1}^n 
+	  \frac{\E [(y_{id} - \x_i \tr \bbeta_d)^2]}{\sigma_d^{2}} - 
+	  \frac{p}{2} \log \sigma_d^{2} - \frac{1}{2} \sum_{j=1}^p 
+	  \E \( \frac{\beta_{jd}^2)}{\gamma_d^{-2}}\) \sigma_d^{-2} - 
+	  \frac{3}{2} \log \sigma^2_d \\
+	  & = \(-\frac{n + p + 1}{2} - 1 \) \log \sigma_d^{2} - 
+	  \frac{1}{2}\left\{ \sum_{i=1}^n \E [(y_{id} - \x_i \tr \bbeta_d)^2] + 
+	  \E(\gamma_d^{-2})\sum_{j=1}^p \E(\beta_{jd}^2) \right\} \sigma_d^{-2}, \\
+	  q_{\sigma_d^{2}}(\sigma_d^{2}) & \overset{D}{=} 
+	  \Gamma^{-1} \(\frac{n + p + 1}{2}, \zeta_{d} \), \\
+	  & \text{with } \zeta_{d} = \frac{1}{2} \bigg\{\mathbf{y}_d \tr \mathbf{y}_d -
+    2 \mathbf{y}_d \tr \X \E(\bbeta_d) + 
+    \trace [ \X \tr \X \V(\bbeta_d)] + 
+    \E(\bbeta_d \tr)  \X \tr \X \E(\bbeta_d ) \\
+    & \,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\, + 
+    \E(\gamma_d^{-2}) \trace [ \V(\bbeta_d)] + 
+    \E(\gamma_d^{-2}) \E(\bbeta_d \tr) \E(\bbeta_d ) \bigg\}.
 	\end{align*}
 
-	In the inverse Gamma setting, we have:
+	\section{Evidence lower bound}\label{sec:evidencelowerbound}
+	The variational Bayes posterior is found through maximisation of an evidence
+	lower bound (ELBO). Additionally, we may monitor convergence through this 
+	ELBO. To compute the ELBO after updating the EB parameters from 
+	$\theta_d^{(l)}, \lambda_d^{(l)}$ to 
+	$\theta_d^{(l + 1)}, \lambda_d^{(l + 1)}$, we compute 
+	$\text{ELBO}^{(l + 1)} = \sum_{d=1}^D \text{ELBO}_d^{(l + 1)}$, with:
 	\begin{align*}
-	\text{ELBO}_d^{(l + 1)} & = \frac{1}{2} \log |\bm{\Sigma}_d^{(l)}| - \frac{a_d^{(l)}}{2} \[ \y_d \tr \y_d - 2\y_d \tr \X \bm{\mu}_d^{(l)} + (\bm{\mu}_d^{(l)}) \tr \X \tr \X \bm{\mu}_d^{(l)} + \trace ( \X \tr \X \bm{\Sigma}_d^{(l)})\] \\
-	& + \frac{b_d^{(l)}}{2} \[ \delta_d^{(l)} - (a_d^{(l)} + 1 - b_d^{(l)}/c_d^{(l)})(\bm{\mu}_d^{(l)}) \tr \bm{\mu}_d^{(l)} - (a_d^{(l)} + 1 - b_d^{(l)}/c_d^{(l)}) \trace (\bm{\Sigma}_d^{(l)}) \]  \\
-	& - \frac{df + 1}{2} \log \zeta_d^{(l)} - \frac{p + \eta_d^{(l)}}{2} \log \delta_d^{(l)} + \frac{\eta_d^{(l)}}{2} e_d^{(l)} + \frac{\eta_d^{(l)}}{2} \log 2 + \log \Gamma \( \frac{p + \eta_d^{(l)}}{2} \) \\
-	& - \log \Gamma \( \frac{\eta_d^{(l + 1)}}{2} \) - \frac{\eta_d^{(l+1)}}{2} e_d^{(l)} - \frac{\eta_d^{(l+1)}}{2} \log 2 + \frac{\eta_d^{(l + 1)}}{2} \log \lambda_d^{(l+1)} - \frac{\lambda_d^{(l + 1)} b_d^{(l)}}{2}.
+	  \text{ELBO}_d^{(l + 1)} = & \frac{1}{2} \log |\bm{\Sigma}_d^{(l)}| - 
+	  \frac{a_d^{(l)}}{2} \[ \y_d \tr \y_d - 2\y_d \tr \X \bm{\mu}_d^{(l)} + 
+	  (\bm{\mu}_d^{(l)}) \tr \X \tr \X \bm{\mu}_d^{(l)} + 
+	  \trace ( \X \tr \X \bm{\Sigma}_d^{(l)})\] \\
+	  & + \frac{b_d^{(l)}}{2} \[ \delta_d^{(l)} - 
+	  a_d^{(l)}(\bm{\mu}_d^{(l)}) \tr \bm{\mu}_d^{(l)} - 
+	  a_d^{(l)} \trace (\bm{\Sigma}_d^{(l)}) \] \\
+	  & - \frac{p + n + 1}{2} \log \zeta_d^{(l)} + 
+	  \frac{p + 1}{4} \log \lambda_d^{(l)} - 
+	  \frac{p + 1}{2} \log \theta_d^{(l)} - \frac{p+1}{4} \log \delta_d^{(l)} \\
+	  & + \log K_{\frac{p+1}{2}}\(\sqrt{\lambda_d^{(l)} 
+	  \delta_d^{(l)}}/\theta_d^{(l)}\) + 
+	  \frac{\lambda_d^{(l)} e_d^{(l)}}{2(\theta_d^{(l)})^2} \\
+	  & + \frac{\lambda_d^{(l+1)}}{\theta_d^{(l+1)}} + 
+	  \frac{1}{2}\log \lambda_d^{(l+1)} - \frac{\lambda_d^{(l+1)} 
+	  e_d^{(l)}}{2 (\theta_d^{(l + 1)})^2} - 
+	  \frac{\lambda_d^{(l + 1)} b_d^{(l)}}{2},
 	\end{align*}
-  Note that the computation of $e_d^{(l)}$ differs between the inverse Gaussian and Gamma models.
-Furthermore, to compute the ELBO after a VB update we simply let $\eta_d^{(l)}, \theta_d^{(l)}, \lambda_d^{(l)}=\eta_d^{(l + 1)}, \theta_d^{(l + 1)}, \lambda_d^{(l + 1)}$.
+	where $a^{(l)}$, $b^{(l)}$, and $e^{(l)}$ are as in the MD. To compute the 
+	ELBO after a VB update we simply let 
+  $\theta_d^{(l)}, \lambda_d^{(l)}=\theta_d^{(l + 1)}, \lambda_d^{(l + 1)}$.
 
-  In practice, there are different options to assess convergence.
-We may monitor the parameters themselves, the ELBO, or a combination of the two.
-We may also iterate for an \textit{a priori} fixed number of times.
-We may also choose to use a different convergence criterium for the EB and VB iterations.
-Currently we fix the number of EB iterations to 20, while we do 2 VB iterations per EB iteration.
+  In practice, there are different options to assess convergence. We may monitor
+  the parameters themselves, the ELBO, or a combination of the two. We may also 
+  iterate for an \textit{a priori} fixed number of times. We may also choose to 
+  use a different convergence criterium for the EB and VB iterations. Currently 
+  we fix the number of EB iterations to 20, while we do 2 VB iterations per EB 
+  iteration.
 
-	\section{Evidence lower bound}
-	We monitor convergence through the evidence lower bound (ELBO).
-To compute the evidence lower bound after updating the EB parameters from $\eta_d^{(l)}, \theta_d^{(l)}, \lambda_d^{(l)}$ to $\eta_d^{(l + 1)}, \theta_d^{(l + 1)}, \lambda_d^{(l + 1)}$, we compute $\text{ELBO}^{(l + 1)} = \sum_{d=1}^D \text{ELBO}_d^{(l + 1)}$, with, in the inverse Gaussian case:
-	\begin{align*}
-	\text{ELBO}_d^{(l + 1)} & = \frac{1}{2} \log |\bm{\Sigma}_d^{(l)}| - \frac{a_d^{(l)}}{2} \[ \y_d \tr \y_d - 2\y_d \tr \X \bm{\mu}_d^{(l)} + (\bm{\mu}_d^{(l)}) \tr \X \tr \X \bm{\mu}_d^{(l)} + \trace ( \X \tr \X \bm{\Sigma}_d^{(l)})\] \\
-	& + \frac{b_d^{(l)}}{2} \[ \delta_d^{(l)} - (a_d^{(l)} + 1 - b_d^{(l)}/c_d^{(l)})(\bm{\mu}_d^{(l)}) \tr \bm{\mu}_d^{(l)} - (a_d^{(l)} + 1 - b_d^{(l)}/c_d^{(l)}) \trace (\bm{\Sigma}_d^{(l)}) \] \\
-	& - \frac{df + 1}{2} \log \zeta_d^{(l)} + \frac{p + 1}{4} \log \lambda_d^{(l)} - \frac{p + 1}{2} \log \theta_d^{(l)} - \frac{p+1}{4} \log \delta_d^{(l)} \\
-	& + \log K_{\frac{p+1}{2}}\(\sqrt{\lambda_d^{(l)} \delta_d^{(l)}}/\theta_d^{(l)}\) + \frac{\lambda_d^{(l)} e_d^{(l)}}{2(\theta_d^{(l)})^2} \\
-	& + \frac{\lambda_d^{(l+1)}}{\theta_d^{(l+1)}} + \frac{1}{2}\log \lambda_d^{(l+1)} - \frac{\lambda_d^{(l+1)} e_d^{(l)}}{2 (\theta_d^{(l + 1)})^2} - \frac{\lambda_d^{(l + 1)} b_d^{(l)}}{2}.
-	\end{align*}
-
-	In the inverse Gamma setting, we have:
-	\begin{align*}
-	\text{ELBO}_d^{(l + 1)} & = \frac{1}{2} \log |\bm{\Sigma}_d^{(l)}| - \frac{a_d^{(l)}}{2} \[ \y_d \tr \y_d - 2\y_d \tr \X \bm{\mu}_d^{(l)} + (\bm{\mu}_d^{(l)}) \tr \X \tr \X \bm{\mu}_d^{(l)} + \trace ( \X \tr \X \bm{\Sigma}_d^{(l)})\] \\
-	& + \frac{b_d^{(l)}}{2} \[ \delta_d^{(l)} - (a_d^{(l)} + 1 - b_d^{(l)}/c_d^{(l)})(\bm{\mu}_d^{(l)}) \tr \bm{\mu}_d^{(l)} - (a_d^{(l)} + 1 - b_d^{(l)}/c_d^{(l)}) \trace (\bm{\Sigma}_d^{(l)}) \]  \\
-	& - \frac{df + 1}{2} \log \zeta_d^{(l)} - \frac{p + \eta_d^{(l)}}{2} \log \delta_d^{(l)} + \frac{\eta_d^{(l)}}{2} e_d^{(l)} + \frac{\eta_d^{(l)}}{2} \log 2 + \log \Gamma \( \frac{p + \eta_d^{(l)}}{2} \) \\
-	& - \log \Gamma \( \frac{\eta_d^{(l + 1)}}{2} \) - \frac{\eta_d^{(l+1)}}{2} e_d^{(l)} - \frac{\eta_d^{(l+1)}}{2} \log 2 + \frac{\eta_d^{(l + 1)}}{2} \log \lambda_d^{(l+1)} - \frac{\lambda_d^{(l + 1)} b_d^{(l)}}{2}.
-	\end{align*}
-  Note that the computation of $e_d^{(l)}$ differs between the inverse Gaussian and Gamma models.
-Furthermore, to compute the ELBO after a VB update we simply let $\eta_d^{(l)}, \theta_d^{(l)}, \lambda_d^{(l)}=\eta_d^{(l + 1)}, \theta_d^{(l + 1)}, \lambda_d^{(l + 1)}$.
-
-  In practice, there are different options to assess convergence.
-We may monitor the parameters themselves, the ELBO, or a combination of the two.
-We may also iterate for an \textit{a priori} fixed number of times.
-We may also choose to use a different convergence criterium for the EB and VB iterations.
-Currently we fix the number of EB iterations to 20, while we do 2 VB iterations per EB iteration.
-
-  \subsection{Efficient computation}
+  \section{Efficient computation}
   The empirical Bayes updates require the following quantities: $\zeta_d$, $a_d$, $\delta_d$, $\trace (\bm{\Sigma}_d)$, $\trace (\X \tr \X \bm{\Sigma}_d)$, $\bm{\mu}_d \tr \bm{\mu}_d$, and $\bm{\mu}_d \tr \X \tr \X \bm{\mu}_d$.
@@
 
-  \section{Inverse Gamma model}
-  In addition to the inverse Gaussian model, we investigate a generalisation of 
-  the models in \cite{kpogbezan_empirical_2017} and \cite{leday_gene_2017}, 
-  where we draw the $\gamma_d^2$ from an inverse Gamma distribution:
+  \section{Ratios of modified Bessel functions}\label{sec:ratiosmodifiedbessels}
+  Ratios of modified Bessel functions of the second kind 
+  $K_{\alpha - 1}(x)/K_{\alpha}(x)$ are prone to under- and overflow for large
+  $\alpha$. In our case, $\alpha$ increases linearly with $p$. Since $p$ may be
+  large, this causes numerical issues in the calculation of various quantities.
+  We alleviate the numerical issues through the following.
+
+  We let $n_1=p/2$ and $n_2=(p-1)/2$ and use the well-known recursive relation:
   $$
-	\gamma_d^2 \sim \Gamma^{-1}(\eta_d/2, \lambda_d/2),
+  K_{\alpha}(x) = K_{\alpha - 2}(x) + \frac{2(\alpha - 1)}{2} K_{\alpha- 1}(x),
   $$
+  to rewrite the ratio:
+  \begin{align*}
+    \frac{K_{\frac{p - 1}{2}}(x)}{K_{\frac{p + 1}{2}}(x)} = 
+    \begin{cases}
+      \Big( \dots \Big( \( 1 + \frac{2 \cdot 1 - 1}{x} \)^{-1} + 
+      \frac{2 \cdot 2-1}{x}\Big)^{-1} + \dots + 
+      \frac{2 \cdot n_1 - 1}{x}\Big)^{-1}, & \text{for } p \text{ even}, \\
+      \Big( \dots \Big( \( \frac{K_0(x)}{K_1(x)} + \frac{2}{x} \cdot 1 \)^{-1} +
+      \frac{2}{x} \cdot 2 \Big)^{-1} + \dots + \frac{2}{x} \cdot n_2 \Big)^{-1},
+      & \text{for } p \text{ odd}.
+    \end{cases}
+  \end{align*} 
+  The ratio $K_0(x)/K_1(x)$ is well-behaved, such that the ratio may be computed
+  without numerical issues.
+
+  \section{Gibbs sampler}\label{sec:gibbssampler}
+  MCMC samples from the posterior corresponding to model 
+  (\ref{md-eq:priormodel}) may be generated for each equation independently. 
+  MCMC samples may be generated by iteratively sampling the following 
+  conditional distributions:
+  \begin{subequations}
+    \begin{align}
+      \bm{\beta}_d | \gamma_d^2, \sigma_d^2, \y_d & \sim \mathcal{N}_p 
+      (\bm{\mu}_d, \bm{\Sigma}_d), \label{eq:betasample} \\
+      \text{with } & \bm{\Sigma}_d = \sigma_d^2 (\X \tr \X + 
+      \gamma_d^{-2} \I)^{-1} \label{eq:sigmasample} \\
+      \text{and } & \bm{\mu}_d = (\X \tr \X + 
+      \gamma_d^{-2} \I)^{-1} \X \tr \y_d \label{eq:musample} \\
+      \gamma_d^2 | \bm{\beta}_d, \sigma_d^2, \y & \sim \mathcal{GIG}
+      \(-\frac{p+1}{2}, \lambda_d \theta_d^{-2}, \delta_d\), \nonumber \\
+      \text{with } & \delta_d = \sigma_d^{-2} \bm{\beta}_d \tr \bm{\beta}_d + 
+      \lambda_d, \nonumber \\
+      \sigma_d^2 | \bm{\beta}_d, \gamma_d^2, \y & \sim \Gamma^{-1} 
+      \(\frac{n + p + 1}{2}, \zeta_d\), \nonumber \\
+      \text{with } & \zeta = \frac{1}{2} \[\mathbf{y} \tr \mathbf{y} -
+      2 \mathbf{y} \tr \X \bm{\mu} + \trace ( \X \tr \X \bm{\Sigma}) + 
+      (\bm{\mu}) \tr \X \tr \X \bm{\mu}  + 
+      \gamma^{-2} \trace ( \bm{\Sigma}) + 
+      \gamma^{-2} (\bm{\mu}) \tr \bm{\mu}\]. \nonumber
+    \end{align}
+  \end{subequations}
+  In high dimensional space, the $p \times p$ matrix inversions in 
+  (\ref{eq:sigmasample}) and (\ref{eq:musample}) are   
+  significant computational bottlenecks. \cite{bhattacharya_fast_2016} describe 
+  a method to sample from (\ref{eq:betasample}) without explicit calculation
+  of this inverse, thereby offering a siginifcant speed-up compared to naive
+  sampling.
+
+  \section{Inverse Gamma model}\label{sec:inversegamma}
+  The model introduced in MD Section \ref{md-sec:inversegammamodel} draws the 
+  $\gamma_d^2$ from an inverse Gamma distribution:
+	\begin{equation}\label{eq:invgammaprior}
+	  \gamma_d^2 \sim \Gamma^{-1}(\eta_d/2, \lambda_d/2)=
+	  \mathcal{GIG}(-\eta_d/2, 0, \lambda_d),
+	\end{equation}
 	with shape $\eta_d/2$ and scale $\lambda/2$. For the sake of comparability, we
-	have rescaled the inverse Gamma parameters and reparametrised both 
-	distributions as generalized inverse Gaussians:
-	\begin{align*}
-  \gamma_d^2 & \sim \mathcal{GIG}(-1/2, \lambda_d \theta_d^{-2}, \lambda_d), \\
-  \gamma_d^2 & \sim \mathcal{GIG}(-\eta_d/2, 0, \lambda_d).
-  \end{align*}
-	The two models coincide if $\theta_d \to \infty$ and $\eta_d = 1$.
-
-	\begin{align*}
-  q(\bm{\beta}_d) & \overset{D}{=} \prod_{d=1}^D \mathcal{N}_p (\bm{\mu}_d, \bm{\Sigma}_d), \\
-  q(\bm{\gamma}^2) & \overset{D}{=} \prod_{d=1}^D \mathcal{GIG}\(-\frac{p+\eta_d}{2}, \lambda_d \theta_d^{-2}, \delta_d\), \\
-  q(\bm{\sigma}^2) & \overset{D}{=} \prod_{d=1}^D \Gamma^{-1} \(\frac{df + 1}{2}, \zeta_d\),
-  \end{align*}
-	where $\eta_d=1$ and $\theta_d \to \infty$ in the inverse Gaussian and inverse Gamma models, respectively.
-Furthermore, $df=n + p$ in the conjugate model and $df=n$ in the non-conjugate model.
-The variational parameters contain cyclic dependencies and are, in the conjugate setting, iteratively updated by:
-  \begin{align*}
-  \bm{\Sigma}_d^{(h+1)} & = (a_d^{(h)})^{-1} (\X \tr \X + c_d^{(h)} \I)^{-1}, \\
-  \bm{\mu}_d^{(h+1)} & = (\X \tr \X + c_d^{(h)} \I)^{-1} \X \tr \y_d, \\
-  \delta_d^{(h+1)} & = b_d^{(h)} (c_d^{(h)})^{-1} \bmu_d \tr \bmu_d + \lambda_d, \\ 
-  \zeta_d^{(h+1)} & = \frac{1}{2} \bigg[\mathbf{y}_d \tr \mathbf{y}_d -2 \mathbf{y}_d \tr \X \bm{\mu}_d^{(h+1)} + \trace ( \X \tr \X \bm{\Sigma}_d^{(h+1)}) + (\bm{\mu}_d^{(h+1)}) \tr \X \tr \X \bm{\mu}_d^{(h+1)} \\
-  & \,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\, + \frac{b_d^{(h+1)} - c_d^{(h+1)} a_d^{(h+1)}}{1 - a_d^{(h+1)}} \trace ( \bm{\Sigma}_d^{(h+1)}) + \frac{b_d^{(h+1)} - c_d^{(h+1)} a_d^{(h+1)}}{1 - a_d^{(h+1)}} (\bm{\mu}_d^{(h+1)}) \tr \bm{\mu}_d^{(h+1)}\bigg].
-  \end{align*}
-  Here, we set
-  \begin{align}
-  a_d^{(h)} & =\E_{Q^{(h)}}(\sigma_d^{-2})=(df+1)/(2 \zeta_d^{(h)}), \nonumber \\
-  b_d^{(h)} & = \E_{Q^{(h)}}(\gamma_d^{-2}) = \sqrt{\frac{\lambda_d}{\theta_d^2 \delta_d^{(h)}}} \frac{K_{\frac{p + \eta_d -2}{2}} \( \sqrt{\lambda_d \delta_d^{(h)}}/\theta_d \)}{K_{\frac{p+\eta_d}{2}} \( \sqrt{\lambda_d \delta_d^{(h)}}/\theta_d \)} + \frac{p+\eta_d^{(h)}}{\delta_d^{(h)}}, \label{eq:auxvar1} \\
-  c_d^{(h)} & = \begin{cases*}
-  b_d^{(h)} & in the conjugate setting \\
-  b_d^{(h)}/a_d^{(h)} & in the non-conjugate setting .
-  \end{cases*} \nonumber
-  \end{align}
-  Note that in the inverse Gamma model, the computation of (\ref{eq:auxvar1}) simplifies to $b_d^{(h)}=(p+\eta_d)/\delta_d^{(h)}$.
-
+	have rescaled the inverse Gamma parameters and reparametrised as a generalized
+	inverse Gaussian. This allows for an easier comparison to the inverse 
+	Gaussian hyperprior:
+	\begin{equation}\label{eq:invgaussianprior}
+	  \mathcal{IG}(\theta_d, \lambda_d) =
+	  \mathcal{GIG}(-1/2, \lambda_d \theta_d^{-2}, \lambda_d).
+  \end{equation}
+	From (\ref{eq:invgammaprior}) and (\ref{eq:invgaussianprior}) it is obvious 
+	that the two models coincide for $\theta_d \to \infty$ and $\eta_d = 1$. In
+	Figure \ref{fig:dens_igaussian_igamma} the two distributions are compared for 
+	a range of parameter values.
 <<dens_igaussian_igamma, fig.cap="Several densities of the inverse Gaussian and inverse Gamma families", out.width="100%", fig.asp=2/3>>=
@@
 
-	In the inverse Gamma model we use the following empirical Bayes procedure: Suppose we have a partitioning of our drugs into drug classes and we let prior variance components $\gamma_d^2$ and $\gamma^2_{d'}$ be drawn from the same distribution if drug $d$ and drug $d'$ are from the same class.
-Or simply put: $\gamma^2_{d}, \gamma^2_{d'} \sim \Gamma^{-1}(\eta_c/2, \lambda_c/2)$ if $d, d' \in \text{class}_c$.
-Empirical bayes estimation of the $\eta_c$ and $\lambda_c$ by maximisation of the (approximated) marginal likelihood amounts to iteratively solving the estimating equations:
+	The variational posterior is given by:
+	\begin{align*}
+	  q(\bm{\beta}_d) & \overset{D}{=} \mathcal{N}_p 
+	  (\bm{\mu}_d, \bm{\Sigma}_d), \\
+	  q(\gamma_d^2) & \overset{D}{=} \mathcal{GIG}\(-\frac{p+\eta_d}{2}, 
+	  \lambda_d \theta_d^{-2}, \delta_d\), \\
+	  q(\sigma_d^2) & \overset{D}{=} \Gamma^{-1} \(\frac{n + p + 1}{2}, \zeta_d\).
+  \end{align*}
+	The variational parameters are the same as in the NIG model, except that now:
+  $b_d^{(h)} = (p + \eta_d)/\delta_d^{(h)}$.
+
+	The inclusion of continuous co-data is not so straightforward in the inverse
+	Gamma hyperprior model. We therefore focus on categorical co-data and use the 
+	following empirical Bayes procedure. Suppose we have a partitioning of our 
+	equations into classes and we let prior variance components $\gamma_d^2$ and
+	$\gamma^2_{d'}$ be drawn from the same distribution if equations $d$ and 
+	$d'$ are from the same class. Or simply put: $\gamma^2_{d}, \gamma^2_{d'} 
+	\sim \Gamma^{-1}(\eta_c/2, \lambda_c/2)$ if $d, d' \in \text{class}_c$. 
+	Empirical bayes estimation of the $\eta_c$ and $\lambda_c$ by maximisation of
+	the (approximated) marginal likelihood amounts to iteratively solving the
+	estimating equations:
   \begin{subequations}
     \begin{align}
-      \psi\(\frac{\eta_c}{2}\) & = \log \lambda_c - \frac{\sum_{d \in \text{class}_c} e^{(l)}_d}{|\text{class}_c|} - \log 2, \label{eq:estequation1}\\
-      \lambda_c & = \eta_c \cdot |\text{class}_c| \cdot \(\sum_{d \in \text{class}_c} b_d^{(l)}\)^{-1}, \label{eq:estequation2}
+      \psi\(\frac{\eta_c}{2}\) & = \log \lambda_c - 
+      \frac{\sum_{d \in \text{class}_c} e^{(l)}_d}{|\text{class}_c|} - \log 2, 
+      \label{eq:estequation1}\\
+      \lambda_c & = \eta_c \cdot |\text{class}_c| \cdot 
+      \(\sum_{d \in \text{class}_c} b_d^{(l)}\)^{-1}, \label{eq:estequation2}
     \end{align}
   \end{subequations}
-  until convergence, where $e^{(l)}_d = \E_{Q^{(l)}}(\log \gamma_d^2) = \log \delta_d^{(l)} - \psi [ (p + \eta_d^{(l)})/2] - \log 2$. 
-  We propose to do that by first solving (\ref{eq:estequation2}) for $\lambda_c$, plugging it into (\ref{eq:estequation1}) and bounding the result using \cite{alzer_inequalities_1997} to obtain an interval $( \eta_c^*, 2\eta_c^* )$ containing the solution to (\ref{eq:estequation1}), where:
+  until convergence, where $e^{(l)}_d = \E_{Q^{(l)}}(\log \gamma_d^2) = 
+  \log \delta_d^{(l)} - \psi [ (p + \eta_d^{(l)})/2] - \log 2$. 
+  We propose to do so by first solving (\ref{eq:estequation2}) for $\lambda_c$, 
+  plugging it into (\ref{eq:estequation1}) and bounding the result using
+  \cite{alzer_inequalities_1997} to obtain an interval $( \eta_c^*, 2\eta_c^* )$
+  that contains the solution to (\ref{eq:estequation1}), where:
   $$
-  \eta_c^* = \left\{ \frac{\sum_{d \in \text{class}_c} e^{(l)}_d}{|\text{class}_c|} + \log \[ \frac{\sum_{d \in \text{class}_c} b_d^{(l)}}{|\text{class}_c|} \]  \right\}^{-1}.
+  \eta_c^* = \left\{ \frac{\sum_{d \in \text{class}_c} 
+  e^{(l)}_d}{|\text{class}_c|} + \log \[ \frac{\sum_{d \in \text{class}_c} 
+  b_d^{(l)}}{|\text{class}_c|} \]  \right\}^{-1}.
   $$
-  With this interval it is straightforward to find $\alpha^{(l+1)}_c$ by any root-finding algorithm and plugging the solution into (\ref{eq:estequation2}) to find $\lambda^{(l+1)}_c$.
-We may use $\eta_c^*$ as a starting value.
+  With this interval it is straightforward to find $\eta^{(l+1)}_c$ by any 
+  root-finding algorithm and plugging the solution into (\ref{eq:estequation2})
+  to find $\lambda^{(l+1)}_c$. We may use $\eta_c^*$ as a starting value.
 
-	The inverse Gamma model with independent drug classes allows for more flexible empirical Bayes estimation of the prior means for the different drug classes.
-As a consequence however, the risk of overfitting is increased.
-Additionally, it does not allow to include continuous covariates on the drugs.
+	In the inverse Gamma setting, $\text{ELBO}^{(l + 1)} = 
+	\sum_{d=1}^D \text{ELBO}_d^{(l + 1)}$ after a EB update is given by:
+	\begin{align*}
+	  \text{ELBO}_d^{(l + 1)}  = & \frac{1}{2} \log |\bm{\Sigma}_d^{(l)}| -
+	  \frac{a_d^{(l)}}{2} \[ \y_d \tr \y_d - 2\y_d \tr \X \bm{\mu}_d^{(l)} +
+	  (\bm{\mu}_d^{(l)}) \tr \X \tr \X \bm{\mu}_d^{(l)} + \trace ( \X \tr \X 
+	  \bm{\Sigma}_d^{(l)})\] \\
+	  & + \frac{b_d^{(l)}}{2} \[ \delta_d^{(l)} - 
+	  a_d^{(l)}(\bm{\mu}_d^{(l)}) \tr \bm{\mu}_d^{(l)} - 
+	  a_d^{(l)} \trace (\bm{\Sigma}_d^{(l)}) \] - 
+	  \frac{n + p + 1}{2} \log \zeta_d^{(l)} \\
+	  & - \frac{p + \eta_d^{(l)}}{2} \log \delta_d^{(l)} + 
+	  \frac{\eta_d^{(l)}}{2} e_d^{(l)} + \frac{\eta_d^{(l)}}{2} \log 2 + 
+	  \log \Gamma \( \frac{p + \eta_d^{(l)}}{2} \) \\
+	  & - \log \Gamma \( \frac{\eta_d^{(l + 1)}}{2} \) - 
+	  \frac{\eta_d^{(l+1)}}{2} e_d^{(l)} - \frac{\eta_d^{(l+1)}}{2} \log 2 +
+	  \frac{\eta_d^{(l + 1)}}{2} \log \lambda_d^{(l+1)} - 
+	  \frac{\lambda_d^{(l + 1)} b_d^{(l)}}{2}.
+	\end{align*}
+  To compute the ELBO after a VB update we simply let 
+  $\eta_d^{(l)}, \lambda_d^{(l)}=\eta_d^{(l + 1)}, \lambda_d^{(l + 1)}$.
 
 	\section{Non-conjugate model}
-	We call the two versions of
-  the model conjugate (with the $\beta_{jd}$ and $\sigma_d^2$ dependent 
-  \textit{a priori}):
-  $$
-  \beta_{jd} | \gamma_d^2, \sigma_d^2 \sim \mathcal{N} 
-  (0, \sigma_d^2 \gamma_d^2)
-  $$
-  and non-conjugate (with $\beta_{jd}$ and $\sigma_d^2$ independent 
-  \textit{a priori}):
+	The non-conjugate model assumes $\beta_{jd}$ and $\sigma_d^2$ 
+	independent \textit{a priori}:
   $$
@@
 	\begin{align*}
-  q(\bm{\beta}) & \overset{D}{=} \prod_{d=1}^D \mathcal{N}_p (\bm{\mu}_d, \bm{\Sigma}_d), \\
-  q(\bm{\gamma}^2) & \overset{D}{=} \prod_{d=1}^D \mathcal{GIG}\(-\frac{p+\eta_d}{2}, \lambda_d \theta_d^{-2}, \delta_d\), \\
-  q(\bm{\sigma}^2) & \overset{D}{=} \prod_{d=1}^D \Gamma^{-1} \(\frac{df + 1}{2}, \zeta_d\),
+	  q(\bm{\beta}_d) & \overset{D}{=} \mathcal{N}_p 
+	  (\bm{\mu}_d, \bm{\Sigma}_d), \\
+	  q(\gamma_d^2) & \overset{D}{=} \mathcal{GIG}
+	  \(-\frac{p+1}{2}, \lambda_d \theta_d^{-2}, \delta_d\), \\
+	  q(\sigma_d^2) & \overset{D}{=} \Gamma^{-1} 
+	  \(\frac{n + 1}{2}, \zeta_d\),
   \end{align*}
-	where $\eta_d=1$ and $\theta_d \to \infty$ in the inverse Gaussian and inverse Gamma models, respectively.
-Furthermore, $df=n + p$ in the conjugate model and $df=n$ in the non-conjugate model.
-The variational parameters contain cyclic dependencies and are, in the conjugate setting, iteratively updated by:
+	The variational parameters are iteratively updated by:
   \begin{align*}
@@
   \end{align*}
+  Here, we set $a_d^{(h)}$, $b_d^{(h)}$, and $c_d^{(h)}$ as in 
+  (\ref{eq:auxvar1})
+  Note that in the inverse Gamma model, the computation of (\ref{eq:auxvar1})
+  simplifies to $b_d^{(h)}=(p+\eta_d)/\delta_d^{(h)}$.
+
   Here, we set

### 2019-05-13 docs/manuscript1.Rnw
changed some nomenclature and added nigig model

@@
 	\section{Introduction}
-	- Motivation
+	- Motivation:
+	simultaneous regressions in: network reconstruction, eQTL mapping 
+	\cite[]{kpogbezan_incorporating_2019}, drug target discovery 
+	\cite[]{noh_inferring_2016}.
+
+	- Introduce data example
 
@@
   \cite{barndorff-nielsen_normal_1997}, routinely applied in mathematical 
-  finance (see, e.g., \cite{kalemanova_normal_2007}). In Figure 
-  \ref{fig:dens_igaussian_marginalbeta}, the NIG 
+  finance (see, e.g., \cite{kalemanova_normal_2007}). See SM Section 
+  \ref{sm-sec:nigprior} for more details on this prior.
+  In Figure \ref{fig:dens_igaussian_marginalbeta}, the NIG 
   prior distribution marginalised over $\gamma_d^2$ is depicted together with 
@@
 
+  \section{General model}
+  In addition to the outcome co-data $\mathbf{c}_d$, we may have feature co-data
+  $\mathbf{z}_j$ available. Examples of this in omics settings are pathway 
+  memberships or previously obtained $p$-values for the features. In
+  that case, we extend model (\ref{eq:priormodel}) to:
+  \begin{subequations}\label{eq:generalmodel}
+    \begin{align}
+      y_{id} | \bm{\beta}_d, \sigma_d^2 & \sim \mathcal{N} 
+      (\x_i \tr \bm{\beta}_d, \sigma_d^2), \nonumber \\
+      \beta_{jd} | \tau_j^2, \gamma_d^2, \sigma_d^2 & \sim \mathcal{N} 
+      (0, \sigma_d^2 \gamma_d^2 \tau_j^2), \\
+      \gamma_d^2 & \sim \mathcal{IG}\(\theta_d^c, \lambda^c_d\), \\
+      \tau_j^2 & \sim \mathcal{IG}\(\theta_j^z, \lambda^z_j\), \\
+      \sigma_d^2 & \sim 1/\sigma_d^3, \nonumber
+    \end{align}
+  \end{subequations}
+  where we now have an extra variance component $\tau_j$ that is 
+  \textit{a priori} inverse Gaussian distributed with mean dependent on the 
+  feature co-data $\mathbf{z}_j$. Equivalently, we may set 
+  $\eta_{jd}^2=\tau_j^2 \gamma_d^2/(\theta_j^z \theta_d^c)$ and write 
+  (\ref{eq:generalmodel}) as:
+  \begin{align*}
+    \beta_{jd} | \eta_{jd}^2, \sigma_d^2 & \sim \mathcal{N} 
+    \(0, \frac{\lambda_{jd}}{\chi_{jd}}\eta_{jd}^2\sigma_d^2\), \\
+    \eta_{jd}^2 & \overset{D}{=}\pi^{-1}\sqrt{\chi_{jd}}\exp(\psi_{jd}) 
+    \eta_{jd}^{-3/2} \cdot K_{0} \(\sqrt{\psi_{jd}^2 + \chi_{jd} (\eta_{jd}^2 
+    + \eta_{jd}^{-2} - 2)} \),
+  \end{align*}
+  where $\lambda_{jd}=\lambda_j^z\lambda_d^c$, 
+  $\chi_{jd}=\lambda_j^z\lambda_d^c/(\theta_j^z \theta_d^c)$, and
+  $\psi_{jd}=\lambda_j^z/\theta_j^z + \lambda_d^c/ \theta_d^c$. A third way
+  of writing the model is:
+  \begin{subequations}\label{eq:generalprior}
+    \begin{align}
+      \beta_{jd} | \eta_{jd}^2, \sigma_d^2 & \sim \mathcal{N} 
+      \(0, \eta_{jd}^2 \sigma_d^2\), \\
+      \eta_{jd}^2 | \xi_{d}^2 & \sim \mathcal{IG} 
+      (\theta_j^z \xi_d^2, \lambda_j^z \xi_d^2 ), \\
+      \xi_d^2 & \sim \mathcal{GIG} (1, \lambda_d^c/(\theta_d^c)^2, \lambda_d^c).
+    \end{align}
+  \end{subequations}
+  From (\ref{eq:generalprior}) we find the marginal variance and kurtosis for 
+  $\beta_{jd}$ (conditional on $\sigma_d^2$):
+  \begin{align}
+    \V(\beta_{jd} | \sigma_d^2) & =\sigma_d^2 \theta_j^z \theta_d^c 
+    \[\frac{K_0(\lambda_d^c/\theta_d^c)}{K_1(\lambda_d^c/\theta_d^c)} + 
+    \frac{2\theta_d^c}{\lambda_d^c}\],\nonumber \\
+    \mathcal{K}(\beta_{jd} | \sigma_d^2) & = 3 \(\frac{\theta_j^z}{\lambda_j^z}
+    + 1\) \theta_d^c \[r_d - r_d^{-1} + 8\(\frac{\theta_d^c}{\lambda_d^c}\)^2
+    K_0 (\lambda_d^c/\theta_d^c) + \frac{4 \theta_d^c}{\lambda_d^c}r_d 
+    K_0 (\lambda_d^c/\theta_d^c) + 1/\theta_d^c\], \label{eq:kurtnigig}
+  \end{align}
+  with $r_d = K_0(\lambda_d^c/\theta_d^c)/K_1(\lambda_d^c/\theta_d^c)$. If 
+  we fix the prior expectation $\E(\gamma_d^2)=\theta_d^c$ and let the prior 
+  variance 
+  $\V(\gamma_d^2)$ go to zero, we have $\V(\beta_{jd} | \sigma_d^2) \rightarrow
+  \sigma_d^2 \theta_j^z \theta_d^c$ and $\mathcal{K}(\beta_{jd} | \sigma_d^2)
+  \rightarrow 3(\theta_j^z/\lambda_j^z + 1)$. From (\ref{eq:kurtnigig}) we see
+  that we can keep the variance constant by fixing $\theta_j^z$, $\theta_d^c$,
+  and $\lambda_d^c$, while we increasing the kurtosis by decrearing 
+  $\lambda_j^z$. This makes the model (\ref{eq:generalmodel}), which we 
+  refer to by the NIGIG model hereafter, a flexible model.
+
+  The formulation as in (\ref{eq:generalmodel}) results in an additional 
+  variational distribution for the $\tau_j$:
+  $$
+  q(\tau_j^2) \overset{D}{=} \mathcal{GIG}\(-\frac{D+1}{2}, \lambda^z_j 
+  (\theta_j^z)^{-2}, \delta_j^z\).
+  $$
+  The variational parameters are now updated by:
+  \begin{align*}
+    \bm{\Sigma}_d^{(h+1)} & = (a_d^{(h)})^{-1} 
+    \[\X \tr \X + b_d^{(h)} \diag (c_j^{(h)})\]^{-1}, \\
+    \bm{\mu}_d^{(h+1)} & = \[\X \tr \X + b_d^{(h)} \diag(c_j^{(h)})\]^{-1} 
+    \X \tr \y_d, \\
+    (\delta^c_d)^{(h+1)} & = a_d^{(h)} \left\{(\bmu^{(h+1)}_d) \tr 
+    \diag (c_j^{(h)}) \bmu^{(h+1)}_d + \trace \[ \diag (c_j^{(h)}) 
+    \bSigma_d^{(h+1)}\] \right\} + \lambda^c_d, \\ 
+    (\delta^z_j)^{(h+1)} & = \sum_{d=1}^D a_d^{(h)} b_d^{(h)} 
+    \[(\bmu^{(h+1)}_d) \tr \bmu^{(h+1)}_d + \trace (\bSigma_d^{(h+1)}) \] + 
+    \lambda^z_j, \\ 
+    \zeta_d^{(h+1)} & = \frac{1}{2} \bigg[\mathbf{y}_d \tr \mathbf{y}_d -
+    2 \mathbf{y}_d \tr \X \bm{\mu}_d^{(h+1)} + 
+    \trace ( \X \tr \X \bm{\Sigma}_d^{(h+1)}) + 
+    (\bm{\mu}_d^{(h+1)}) \tr \X \tr \X \bm{\mu}_d^{(h+1)} \\
+    & \,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\, + 
+    b_d^{(h+1)} \trace \[ \diag (c_j^{(h)}) \bm{\Sigma}_d^{(h+1)}\] + 
+    b_d^{(h+1)} (\bm{\mu}_d^{(h+1)}) \tr \diag (c_j^{(h)}) 
+    \bm{\mu}_d^{(h+1)}\bigg],
+  \end{align*}
+  until convergence, where
+  $$
+  c_j^{(h)} = \E_{Q^{(h)}}(\tau_j^{-2}) = \sqrt{\frac{\lambda^z_j}
+  {(\theta_j^z)^2 (\delta^z_j)^{(h)}}} 
+  \frac{K_{\frac{D - 1}{2}} 
+  \( \sqrt{\lambda^z_j (\delta^z_j)^{(h)}}/\theta_j^z \)}
+  {K_{\frac{D+1}{2}} \( \sqrt{\lambda^z_j (\delta_j^z)^{(h)}}/\theta_j^z \)} +
+  \frac{D+1}{(\delta_j^z)^{(h)}}
+  $$
+  If we set $\theta_d^c=(\mathbf{c}_d \tr \bm{\alpha}^c)^{-1}$ and
+  $\theta_j^z=(\mathbf{z}_j \tr \bm{\alpha}^z)^{-1}$, we have an additional
+  empirical Bayes step that consists of an iteratively 
+  reweighted least squares regression of responses $(f_j^{(l)})^{-1}$ on
+  predictors $\mathbf{z}_j$ with weights $\lambda_j^z f_j^{(l)}$, where 
+  \begin{align*}
+  f_j^{(l)} & = \E_{Q^{(h)}}(\tau_j) = [c_j^{(l)} - 
+  (D + 1)/(\delta_j^z)^{(l)}] \cdot (\delta_j^z)^{(l)} [\mathbf{z}_j \tr 
+  (\bm{\alpha}^z)^{(l)} ]^{-2} / (\lambda_j^z)^{(l)}, \\
+  \lambda_j^z & = \[ c_j^{(l)} + f_j^{(l)} 
+  (\mathbf{z}_j \tr \bm{\alpha}^z)^2 - 
+    2\mathbf{z}_j \tr \bm{\alpha}^z \]^{-1}.
+  \end{align*}
+
   \section{Simulations}
-  \subsection{Comparison of NIG to normal inverse Gamma model}
+  \subsection{Comparison of NIG and Student's \texorpdfstring{$t$}{t} priors}
   \label{sec:inversegammamodel}
@@
   We compare the 
-  posteriors of the proposed NIG prior to the scaled Student's $t$-distribution
-  prior in a simulated setting. We generate draws from these posteriors using 
-  an MCMC sampler implemented in \texttt{rstan}. The aim of this simulation 
-  study is 
-  to investigate the performance of the NIG prior model, so we set $D=1$ and 
-  drop the $d$ indices in the following.
+  posteriors of the proposed NIG prior and the scaled Student's $t$-distribution
+  prior in a simulated setting.
+
+  We compare the two posterior distributions to the true posterior for a 
+  number of measures. Divergence between the distributions
+  is quantified through a multivariate extension of the Cram{\'e}r-von Mises
+  criterion \cite[]{baringhaus_new_2004}, where smaller values of the criterion
+  indicate more similar distributions. We also assess 95\% coverage of the 
+  posterior credible intervals. The accuracy of the 
+  point estimates of $\beta_j$ in the form of posterior means is assessed by
+  the mean squared distance (MSD) and correlation between the point estimates 
+  and the true $\beta_j$.
+
+  The aim of this simulation study is to investigate the performance of the NIG 
+  prior model, so we set $D=1$ and drop the $d$ indices in the following. 
+  We studied two settings: one dense and one sparse. In the dense setting, 
+  we simulated $n=100$ features $\x_i$ from a $p=200$ dimensional Gaussian 
+  distribution with zero mean vector, variances equal to one, and all 
+  covariances set to $\rho=0.5$. The outcomes were generated through $y_i = 
+  \x_i\tr \bbeta + \epsilon_i$, with the $\epsilon_i$ from the standard normal 
+  distribution. 
+  Likewise, we generated the model parameters $\beta_j$ from a standard normal
+  distribution. For estimation, we set our hyperparameters $\lambda=0.5$, 
+  $\theta=1$, and $\eta=2.5$. such that the marginal $\beta_j$ variances
+  of the NIG and Student's $t$ priors match. For both priors, we generated 
+  1000 draws (after 1000 burnin) from the posteriors using an MCMC sampler 
+  implemented in \texttt{rstan}.
+
+  In the sparse setting, the observed data generating process is the same. The
+  model parameters are are simulated from a spike-and-slab prior with a 
+  Laplace slab:
+  \begin{align*}
+    w & \sim \text{Beta}(\nu, \kappa), \\
+    z_j | w & \sim \text{Binom}(w), \\
+    \beta_j | z_j & \sim z_j\text{Laplace}
+    \(\sqrt{\frac{\nu + \kappa}{2\nu}} \sigma \gamma\) + (1 - z_j)\delta_0,
+  \end{align*}
+  where $\delta_0$ denotes the Dirac mass at zero. The Laplace scale is chosen
+  such that the we have marginal prior variance $\V(\beta_j)=1$. The sparsity 
+  level is determined by the proportion of zero $\beta_j$, 
+  $s=1 - \nu/(\nu + \kappa)$. Note that the resulting marginal prior for
+  $\beta_j$ is:
+  $$
+  \beta_j \sim (s - 1)\text{Laplace}\(\sqrt{\frac{\nu + \kappa}{2\nu}} 
+  \sigma \gamma\) + s\delta_0.
+  $$
+  We use two settings, $s=0.5$ and $s=0.9$. The 
+  hyperparameters for estimation are set the same as in the dense setting.
+  To mitigate the effect of random variation, we
+  repeated the simulations 100 times and report the results in Figures 
+  \ref{fig:boxplots_igaussian_res4_post} and [! ref to figure here]
+<<boxplots_igaussian_res4_post, fig.cap="Comparison of the NIG and Student's t priors through (a) Cramér-von Mises criterion between the model and true posteriors, (b) correlation and (c) MSD between the posterior means and true parameters, and (d) 95\\% credible interval coverage of the true values.", out.width="100%", fig.asp=2/5>>=
+@
+
+  [! describe results here]
 
@@
   To assess the performance of the variational approximations we evaluate 
-  several measures. Divergence between the MCMC and variational posterior
-  is quantified through a multivariate extension of the Cram{\'e}r-von Mises
-  criterion \cite[]{baringhaus_new_2004}, where smaller values of the criterion
-  indicate more similar distributions. In addition, we calculate the correlation
-  and mean squared distance (MSD) between the posterior variances of
-  the MCMC posterior and variational posterior. \texttt{rstan} allows the user 
+  several measures. We measure divergence between the MCMC and variational 
+  posterior with the Cram{\'e}r-von Mises criterion. We compute the 
+  correlation and MSD between the MCMC and variational posterior variances. 
+  \texttt{rstan} allows the user 
   to maximise the posterior to obtain the maximum \textit{a posteriori} (MAP)
@@
   The interest of this simulation lies in the variational Bayes approximation, 
-  so we restrict ourselves to the univariate setting (i.e., $D=1$) and drop any
-  $d$ indices from notation in the following. We simulated
-  $n=100$ features $\x_i$ from a $p=200$ dimensional Gaussian distribution with
-  zero mean vector, variances equal to one, and all covariances set to 
-  $\rho=0.5$. The outcomes are generated through $y_i = \x_i\tr \bbeta + 
-  \epsilon_i$, with the $\epsilon_i$ from the standard normal distribution. 
-  Likewise, we generate the model parameters $\beta_j$ from a standard normal
-  distribution. For estimation, we set our hyperparameters $\lambda=0.1$ and 
+  so we restrict ourselves to the univariate setting (i.e., $D=1$). The
+  simulation setup is the same as in Section \ref{sec:inversegammamodel}.
+  The hyperparameters are set as follows: $\lambda=0.1$ and 
   $\theta=2$. We choose the penalty parameter in \texttt{glmnet} by assuming a 
@@
   ridge models, which results in the penalty parameter 
-  $1/(n\sigma^2\theta)=1/200$.To mitigate the effect of random variation, we
-  repeat the simulation 100 times and report the results in Figures 
+  $1/(n\sigma^2\theta)=1/200$. We
+  repeated the simulation 100 times and report the results in Figures 
   \ref{fig:boxplots_igaussian_res3_vb}-
@@
 @
+
+  [! describe results here]
 
@@
   \section{Application}
+  - Possible: network reconstruction, medical surveys, drug response prediction
 

### 2019-05-21 docs/supplement1.Rnw
added sparse simulations 1

@@
 	\section{NIGIG prior}\label{sec:nigigprior}
-	For a fixed error variance $\sigma_d^2$, we have a marginal prior variance 
-	and kurtosis:
+	Equivalently, we may set 
+  $\eta_{jd}^2=\tau_j^2 \gamma_d^2/(\theta_j^z \theta_d^c)$ and write 
+  MD equation (\ref{md-eq:generalmodel}) as:
   \begin{align*}
+    \beta_{jd} | \eta_{jd}^2, \sigma_d^2 & \sim \mathcal{N} 
+    \(0, \frac{\lambda_{jd}}{\chi_{jd}}\eta_{jd}^2\sigma_d^2\), \\
+    \eta_{jd}^2 & \overset{D}{=}\pi^{-1}\sqrt{\chi_{jd}}\exp(\psi_{jd}) 
+    \eta_{jd}^{-3/2} \cdot K_{0} \(\sqrt{\psi_{jd}^2 + \chi_{jd} (\eta_{jd}^2 
+    + \eta_{jd}^{-2} - 2)} \),
+  \end{align*}
+  where $\lambda_{jd}=\lambda_j^z\lambda_d^c$, 
+  $\chi_{jd}=\lambda_j^z\lambda_d^c/(\theta_j^z \theta_d^c)$, and
+  $\psi_{jd}=\lambda_j^z/\theta_j^z + \lambda_d^c/ \theta_d^c$. A third way
+  of writing the model is:
+  \begin{subequations}\label{eq:generalprior}
+    \begin{align}
+      \beta_{jd} | \eta_{jd}^2, \sigma_d^2 & \sim \mathcal{N} 
+      \(0, \eta_{jd}^2 \sigma_d^2\), \\
+      \eta_{jd}^2 | \xi_{d}^2 & \sim \mathcal{IG} 
+      (\theta_j^z \xi_d^2, \lambda_j^z \xi_d^2 ), \\
+      \xi_d^2 & \sim \mathcal{GIG} (1, \lambda_d^c/(\theta_d^c)^2, \lambda_d^c).
+    \end{align}
+  \end{subequations}
+	From (\ref{eq:generalprior}), for a fixed error variance
+	$\sigma_d^2$, we find the marginal prior variance and kurtosis:
+	\begin{subequations}\label{eq:varkurtnigig}
+	  \begin{align}
 	    \V(\beta_{jd} | \sigma_d^2) & = \sigma_d^2 \theta_d^c \theta_j^z  
       \(r_d + \frac{2\theta_d^c}{\lambda_d^c}\), \\
-	  \mathcal{K}(\beta_{jd} | \sigma_d^2) & = \(\frac{3\theta_j^z}{\lambda_j^z} +
-	  3\) \frac{4 \theta_d^c (r_d + 2\theta_d^c/\lambda_d^c) + 1}{
-	  \lambda_d^c(r_d + 2\theta_d^c/\lambda_d^c)^2},
-	\end{align*}
+	    \mathcal{K}(\beta_{jd} | \sigma_d^2) & = \(\frac{3\theta_j^z}
+	    {\lambda_j^z} + 3\) \frac{4 \theta_d^c (r_d + 2\theta_d^c/\lambda_d^c) +
+	    1}{\lambda_d^c(r_d + 2\theta_d^c/\lambda_d^c)^2},
+	  \end{align}
+	\end{subequations}
 	with $r_d = K_0(\lambda_d^c/\theta_d^c)/K_1(\lambda_d^c/\theta_d^c)$.
+	If 
+  we fix the prior expectation $\E(\gamma_d^2)=\theta_d^c$ and let the prior 
+  variance 
+  $\V(\gamma_d^2)$ go to zero, we have $\V(\beta_{jd} | \sigma_d^2) \rightarrow
+  \sigma_d^2 \theta_j^z \theta_d^c$ and $\mathcal{K}(\beta_{jd} | \sigma_d^2)
+  \rightarrow 3(\theta_j^z/\lambda_j^z + 1)$. From (\ref{eq:varkurtnigig}) we see
+  that we can keep the variance constant by fixing $\theta_j^z$, $\theta_d^c$,
+  and $\lambda_d^c$, while we increasing the kurtosis by decrearing 
+  $\lambda_j^z$. This makes the model MD equations (\ref{md-eq:generalmodel}) a
+  flexible model.
 

### 2019-05-21 docs/manuscript2.Rnw

--- 2019-02-19 docs/manuscript.Rnw
@@
 opts_chunk$set(fig.align='center', fig.path="../figs/", 
-               echo=FALSE, cache=FALSE, message=FALSE)
+               echo=FALSE, cache=FALSE, message=FALSE, fig.pos='!ht')
 @
@@
 \documentclass[a4paper,hidelinks]{article}
-\usepackage{bm,amsmath,amssymb,amsthm,amsfonts,graphics,graphicx,epsfig,rotating,caption,subcaption,natbib,appendix,titlesec,multicol,hyperref,verbatim,bbm,algorithm,algpseudocode,pgfplotstable,threeparttable, booktabs,mathtools,dsfont,parskip}
+\usepackage{url,bm,algorithm,algpseudocode,bbm,amsmath,amssymb,amsthm,
+amsfonts,graphics,graphicx,epsfig,rotating,caption,subcaption,natbib,appendix,
+titlesec,multicol,verbatim,threeparttable,booktabs,dsfont,parskip,
+xr-hyper,hyperref,enumerate}
+\externaldocument[sm-]{supplement}
 
@@
 
+	\section*{Session info}
+
+<<r session-info, eval=TRUE, echo=TRUE, tidy=TRUE>>=
+devtools::session_info()
+@
+
 	\titleformat{\section}{\normalfont\Large\bfseries}{\appendixname~\thesection.}{1em}{}

### 2019-05-21 docs/manuscript1.Rnw

@@
   \textit{a priori} inverse Gaussian distributed with mean dependent on the 
-  feature co-data $\mathbf{z}_j$. Equivalently, we may set 
-  $\eta_{jd}^2=\tau_j^2 \gamma_d^2/(\theta_j^z \theta_d^c)$ and write 
-  (\ref{eq:generalmodel}) as:
-  \begin{align*}
-    \beta_{jd} | \eta_{jd}^2, \sigma_d^2 & \sim \mathcal{N} 
-    \(0, \frac{\lambda_{jd}}{\chi_{jd}}\eta_{jd}^2\sigma_d^2\), \\
-    \eta_{jd}^2 & \overset{D}{=}\pi^{-1}\sqrt{\chi_{jd}}\exp(\psi_{jd}) 
-    \eta_{jd}^{-3/2} \cdot K_{0} \(\sqrt{\psi_{jd}^2 + \chi_{jd} (\eta_{jd}^2 
-    + \eta_{jd}^{-2} - 2)} \),
-  \end{align*}
-  where $\lambda_{jd}=\lambda_j^z\lambda_d^c$, 
-  $\chi_{jd}=\lambda_j^z\lambda_d^c/(\theta_j^z \theta_d^c)$, and
-  $\psi_{jd}=\lambda_j^z/\theta_j^z + \lambda_d^c/ \theta_d^c$. A third way
-  of writing the model is:
-  \begin{subequations}\label{eq:generalprior}
-    \begin{align}
-      \beta_{jd} | \eta_{jd}^2, \sigma_d^2 & \sim \mathcal{N} 
-      \(0, \eta_{jd}^2 \sigma_d^2\), \\
-      \eta_{jd}^2 | \xi_{d}^2 & \sim \mathcal{IG} 
-      (\theta_j^z \xi_d^2, \lambda_j^z \xi_d^2 ), \\
-      \xi_d^2 & \sim \mathcal{GIG} (1, \lambda_d^c/(\theta_d^c)^2, \lambda_d^c).
-    \end{align}
-  \end{subequations}
-  From (\ref{eq:generalprior}) we find the marginal variance and kurtosis for 
-  $\beta_{jd}$ (conditional on $\sigma_d^2$):
-  \begin{align}
-    \V(\beta_{jd} | \sigma_d^2) & =\sigma_d^2 \theta_j^z \theta_d^c 
-    \[\frac{K_0(\lambda_d^c/\theta_d^c)}{K_1(\lambda_d^c/\theta_d^c)} + 
-    \frac{2\theta_d^c}{\lambda_d^c}\],\nonumber \\
-    \mathcal{K}(\beta_{jd} | \sigma_d^2) & = 3 \(\frac{\theta_j^z}{\lambda_j^z}
-    + 1\) \theta_d^c \[r_d - r_d^{-1} + 8\(\frac{\theta_d^c}{\lambda_d^c}\)^2
-    K_0 (\lambda_d^c/\theta_d^c) + \frac{4 \theta_d^c}{\lambda_d^c}r_d 
-    K_0 (\lambda_d^c/\theta_d^c) + 1/\theta_d^c\], \label{eq:kurtnigig}
-  \end{align}
-  with $r_d = K_0(\lambda_d^c/\theta_d^c)/K_1(\lambda_d^c/\theta_d^c)$. If 
-  we fix the prior expectation $\E(\gamma_d^2)=\theta_d^c$ and let the prior 
-  variance 
-  $\V(\gamma_d^2)$ go to zero, we have $\V(\beta_{jd} | \sigma_d^2) \rightarrow
-  \sigma_d^2 \theta_j^z \theta_d^c$ and $\mathcal{K}(\beta_{jd} | \sigma_d^2)
-  \rightarrow 3(\theta_j^z/\lambda_j^z + 1)$. From (\ref{eq:kurtnigig}) we see
-  that we can keep the variance constant by fixing $\theta_j^z$, $\theta_d^c$,
-  and $\lambda_d^c$, while we increasing the kurtosis by decrearing 
-  $\lambda_j^z$. This makes the model (\ref{eq:generalmodel}), which we 
-  refer to by the NIGIG model hereafter, a flexible model.
+  feature co-data $\mathbf{z}_j$. We will refer to this model as the normal
+  inverse Gaussian-inverse Gaussian (NIGIG) model. See SM Section 
+  \ref{sm-sec:nigigprior} for more details on this prior.
 
@@
   \section{Simulations}
-  \subsection{Comparison of NIG and Student's \texorpdfstring{$t$}{t} priors}
+  \subsection{Comparison of NIG, NIGIG and Student's \texorpdfstring{$t$}{t} 
+  priors}
   \label{sec:inversegammamodel}
@@
   We compare the 
-  posteriors of the proposed NIG prior and the scaled Student's $t$-distribution
-  prior in a simulated setting.
+  posteriors of the proposed NIG and NIGIG priors to the scaled Student's 
+  $t$-distribution prior in a simulated setting.
 
-  We compare the two posterior distributions to the true posterior for a 
+  We compare the three posterior distributions to the true posterior for a 
   number of measures. Divergence between the distributions
@@
   Likewise, we generated the model parameters $\beta_j$ from a standard normal
-  distribution. For estimation, we set our hyperparameters $\lambda=0.5$, 
-  $\theta=1$, and $\eta=2.5$. such that the marginal $\beta_j$ variances
-  of the NIG and Student's $t$ priors match. For both priors, we generated 
-  1000 draws (after 1000 burnin) from the posteriors using an MCMC sampler 
-  implemented in \texttt{rstan}.
+  distribution.
+
+  We choose the hyperparameters such that we get all combinations of marginal
+  prior variances and kurtoses $\V(\beta_{jd} | \sigma_d^2) \in 
+  \{ 1/2, 1, 2 \}$ and $\mathcal{K}(\beta_{jd} | \sigma_d^2) \in 
+  \{ 4, 10 \}$. This leads to the settings in Table \ref{tab:hyper_set1}. For 
+  all priors, we generated 1000 draws (after 1000 burnin) from the 
+  posteriors using an MCMC sampler implemented in \texttt{rstan}. We repeated
+  the simulations 100 times and report 
+  results in Figures \ref{fig:boxplots_igaussian_res4_post1} and 
+  \ref{fig:boxplots_igaussian_res4_post2}. 
+<<hyper_set1>>= 
+set <- read.table("results/simulations_igaussian_set4.csv")
+kableExtra::kable_styling(
+  knitr::kable(set[, c("lambdac", "thetac", "lambdaz", "thetaz",
+                       "eta", "lambda", "vars", "kurts")], 
+               col.names=c("$\\lambda^c$", "$\\theta^c$", "$\\lambda^z$",
+                           "$\\theta^z$", "$\\eta$", "$\\lambda$", 
+                           "$\\V(\\beta_{jd} | \\sigma_d^2)$",
+                           "$\\mathcal{K}(\\beta_{jd} | \\sigma_d^2)$"), 
+               caption="Hyperparameter settings", 
+               "latex", booktabs=TRUE, escape=FALSE, digits=2),
+  latex_options=c("HOLD_position"))
+@
+
+  <<boxplots_igaussian_res4_post1, fig.cap="Comparison of the NIG, NIGIG, and Student's t priors in (a)-(d) simulation setting 1, (e)-(h) setting 2, and (i)-(l) setting 3, through Cramér-von Mises criterion between the model and true posteriors, correlation and MSD between the posterior means and true parameters, and 95\\% credible interval coverage of the true values.", out.width="100%", fig.asp=1>>=
+@
+
+<<boxplots_igaussian_res4_post2, fig.cap="Comparison of the NIG, NIGIG, and Student's t priors in (a)-(d) simulation setting 4, (e)-(h) setting 5, and (i)-(l) setting 6, through Cramér-von Mises criterion between the model and true posteriors, correlation and MSD between the posterior means and true parameters, and 95\\% credible interval coverage of the true values.", out.width="100%", fig.asp=1>>=
+@
+
+  [! describe results here]
 
@@
   model parameters are are simulated from a spike-and-slab prior with a 
-  Laplace slab:
+  Gaussian slab:
   \begin{align*}
@@
     z_j | w & \sim \text{Binom}(w), \\
-    \beta_j | z_j & \sim z_j\text{Laplace}
-    \(\sqrt{\frac{\nu + \kappa}{2\nu}} \sigma \gamma\) + (1 - z_j)\delta_0,
+    \beta_j | z_j & \sim z_j \mathcal{N} \( 0, 1\) +
+    (1 - z_j)\delta_0,
   \end{align*}
-  where $\delta_0$ denotes the Dirac mass at zero. The Laplace scale is chosen
+  where $\delta_0$ denotes the Dirac mass at zero. The slab is chosen
   such that the we have marginal prior variance $\V(\beta_j)=1$. The sparsity 
@@
   $$
-  \beta_j \sim (s - 1)\text{Laplace}\(\sqrt{\frac{\nu + \kappa}{2\nu}} 
-  \sigma \gamma\) + s\delta_0.
+  \beta_j \sim (s - 1)\mathcal{N} \( 0, 1\) + s\delta_0.
   $$
   We use two settings, $s=0.5$ and $s=0.9$. The 
-  hyperparameters for estimation are set the same as in the dense setting.
+  hyperparameters for estimation are set the same as in the dense setting, 
+  resulting in a total of 12 different simulation settings.
   To mitigate the effect of random variation, we
   repeated the simulations 100 times and report the results in Figures 
-  \ref{fig:boxplots_igaussian_res4_post} and [! ref to figure here]
-<<boxplots_igaussian_res4_post, fig.cap="Comparison of the NIG and Student's t priors through (a) Cramér-von Mises criterion between the model and true posteriors, (b) correlation and (c) MSD between the posterior means and true parameters, and (d) 95\\% credible interval coverage of the true values.", out.width="100%", fig.asp=2/5>>=
-@
-
-  [! describe results here]
+  [! ref to figure here]
 
@@
 
+  - metabolomics data: oxidative stress als uitkomst van de andere metabolieten
+  min organic acids omdat ze downstream zijn. oxidateve stress onderverdelen in
+  high and low als uitkomst co-data en de andere metabolieten onderverdelen
+  als predictor co-data.
   \section{Discussion}

### 2019-05-21 docs/supplement2.Rnw

--- 2019-02-19 docs/supplement.Rnw
@@
 A simple way of doing this is to include the scales $\sigma_d^2$ in the priors for the $\bm{\beta}_d$.
-We call the two versions of the model conjugate (with $\bm{\beta}_d$ and $\sigma_d^2$ dependent):
+We call the two versions of the model conjugate (with $\bm{\beta}_d$ and $\sigma_d^2$ dependent \textit{a priori}):
   $$
@@
   $$
-  and non-conjugate (with $\bm{\beta}_d$ and $\sigma_d^2$ independent):
+  and non-conjugate (with $\bm{\beta}_d$ and $\sigma_d^2$ independent \textit{a priori}):
   $$
@@
   $$
-	we investigate a generalisation of the models in \cite{kpogbezan_empirical_2017} and \cite{leday_gene_2017}, where we draw the $\gamma_d^2$ from an inverse Gamma distribution; :
+	we investigate a generalisation of the models in \cite{kpogbezan_empirical_2017} and \cite{leday_gene_2017}, where we draw the $\gamma_d^2$ from an inverse Gamma distribution:
 	$$
-	\gamma_d^2 \sim \Gamma^{-1}(\eta_d/2, \lambda_d/2).
+	\gamma_d^2 \sim \Gamma^{-1}(\eta_d/2, \lambda_d/2),
 	$$
-	For the sake of comparability, we have rescaled the inverse Gamma parameters and reparametrised both distributions as generalized inverse Gaussians:
+	with shape $\eta_d/2$ and scale $\lambda/2$. For the sake of comparability, we
+	have rescaled the inverse Gamma parameters and reparametrised both 
+	distributions as generalized inverse Gaussians:
 	\begin{align*}
@@
 	The two models coincide if $\theta_d \to \infty$ and $\eta_d = 1$.
+
+<<dens_igaussian_igamma, fig.cap="Several densities of the inverse Gaussian and inverse Gamma families", out.width="100%", fig.asp=2/3>>=
+digauss <- function(x, theta, lambda, eta) {
+  sqrt(lambda/(x^3*2*pi))*exp(-lambda*(x - theta)^2/(2*theta^2*x))
+}
+
+digamma <- function(x, theta, lambda, eta) {
+  alpha <- eta/2
+  beta <- lambda/2
+  beta^alpha/gamma(alpha)*x^(-alpha - 1)*exp(-beta/x)
+}
+
+eta <- c(1, 2, 1, 1, 5)
+lambda <- c(1, 1, 2, 1/2, 2)
+theta <- c(1000, 1000, 1, 1, 1)
+
+labels <- c("inv. Gaussian", "inv. Gamma")
+col <- c(1, 2)
+
+opar <- par(no.readonly=TRUE)
+par(mar=opar$mar*c(1, 1.3, 1, 1))
+layout(matrix(c(rep(c(1, 1, 2, 2, 3, 3), 2), rep(c(0, 4, 4, 5, 5, 0), 2)),
+                 nrow=4, ncol=6, byrow=TRUE))
+for(i in 1:length(eta)) {
+  modes <- c(theta[i]*(sqrt(1 + 9*theta[i]^2/
+                            (4*lambda[i]^2)) - 3*theta[i]/(2*lambda[i])),
+             lambda[i]/(eta[i] + 2))
+  ylim <- c(0, max(digauss(modes[1], theta[i], lambda[i], eta[i]),
+                   digamma(modes[2], theta[i], lambda[i], eta[i])))
+  curve(digauss(x, theta[i], lambda[i], eta[i]), 0.001, 5, n=1000,
+        ylim=ylim, ylab="Density", "x", col=col[1], 
+        main=bquote(eta==.(eta[i])*","~lambda==.(lambda[i])*","~theta==.(
+          theta[i])))
+  curve(digamma(x, theta[i], lambda[i], eta[i]), add=TRUE, 
+        col=col[2], n=1000)
+}
+legend("topright", legend=labels, lty=1, col=col, seg.len=1)
+par(opar)
+@
+
+<<dens_igaussian_marginalbeta, fig.cap="Marginal $p(\\beta_j)$ prior for several choices of $p(\\gamma^2_j)$ scaled to $\\V(\\beta_j)=1$.", out.width="60%", fig.asp=1>>=
+library(GeneralizedHyperbolic)
+# all scaled such that variance is one
+sigma <- 1
+theta <- 1/sigma^2
+lambda <- c(0.1, 2)
+lambda1 <- sqrt(2)
+col <- c(1:4)
+lty <- c(1:4)
+labels <- as.expression(c(bquote("IG, "~lambda==.(lambda[1])~", "~
+                                   theta==.(theta)),
+                          bquote("IG, "~lambda==.(lambda[2])~", "~
+                                   theta==.(theta)),
+                          "Point mass (ridge)", "Exponential (lasso)"))
+dprior <- function(x, lambda, theta, sigma) {
+  dnig(x, 0, sigma*sqrt(lambda), sqrt(lambda/(theta*sigma)), 0)
+}
+dlasso <- function(x, lambda1) {
+  0.5*lambda1*exp(-lambda1*abs(x))
+}
+
+curve(dprior(x, lambda[1], theta, sigma), -3, 3, ylab="Density", n=1000, 
+      col=col[1], lty=lty[1])
+curve(dprior(x, lambda[2], theta, sigma), add=TRUE, n=1000, 
+      col=col[2], lty=lty[2])
+curve(dnorm(x, 0, 1), add=TRUE, n=1000, col=col[3], lty=lty[3])
+curve(dlasso(x, lambda1), add=TRUE, n=1000, col=col[4], lty=lty[4])
+legend("topright", legend=labels, col=col, lty=lty)
+@
 

### 2019-05-22 code/simulations_igaussian.R
commit before push

@@ -778,9 +778,6 @@ methods <- c("nig", "studentst", "nigig")
 params <- c("beta0", "beta", "sigma", "gamma", "tau")
 
 # set initial values and control parameters
-stan.nig <- stan_model("code/igauss.stan")
-stan.nigig <- stan_model("code/nigig.stan")
-stan.studentst <- stan_model("code/igamma.stan")
 nsamples <- 1000
 nwarmup <- 1000
 
@@ -791,6 +788,10 @@ ncores <- 50
 res <- lapply(1:nrow(set), function(s) {
   mclapply(1:nreps, function(k) {
     print(paste0("set", s, ", rep", k))
+    
+    stan.nig <- stan_model("code/igauss.stan")
+    stan.nigig <- stan_model("code/nigig.stan")
+    stan.studentst <- stan_model("code/igamma.stan")
     
     set.seed(2019 + k)
     beta <- sapply(1:D, function(d) {

### 2019-05-22 docs/manuscript1.Rnw
commit before push

@@
 
+  \section{General model 2}
+  We have co-data $\mathbf{c}_{jd}$ available that describes the relationship
+  between the covariate $j$ and outcome $d$. An example from eQTL is the 
+  distance between SNP and gene. Model (\ref{eq:priormodel}) is then rewritten
+  as:
+  \begin{subequations}\label{eq:generalmodel2}
+    \begin{align*}
+      y_{id} | \bm{\beta}_d, \sigma_d^2 & \sim \mathcal{N} 
+      \(\x_i \tr \bm{\beta}_d, \sigma_d^2\), \nonumber \\
+      \beta_{jd} | \gamma_{jd}^2, \sigma_d^2 & \sim \mathcal{N} 
+      \(0, \sigma_d^2 \gamma_{jd}^2 \), \\
+      \gamma_{jd}^2 & \sim \mathcal{IG}
+      \( (\mathbf{c}_{jd} \tr \bm{\alpha})^{-1}, \lambda_{jd} \), \\
+      \sigma_d^2 & \sim 1/\sigma_d^3, \nonumber
+    \end{align*}
+  \end{subequations}
+  where now the variance components $\gamma_{jd}^2$ are both outcome and 
+  covariate specific and modelled as a function of the co-data. 
+  The formulation as in (\ref{eq:generalmodel2}) results in the following
+  variational distribution for the $\gamma^2_{jd}$:
+  $$
+  q(\gamma_{jd}^2) \overset{D}{=} \mathcal{GIG}
+  \(-1, \lambda_{jd}(\mathbf{c}_{jd} \tr \bm{\alpha})^2, \delta_{jd}\).
+  $$
+  The variational parameters are now updated by:
+  \begin{align*}
+    \bm{\Sigma}_d^{(h+1)} & = (a_d^{(h)})^{-1} 
+    \[\X \tr \X + \diag(b_{jd}^{(h)})\]^{-1}, \\
+    \bm{\mu}_d^{(h+1)} & = \[\X \tr \X + \diag(b_{jd}^{(h)})\]^{-1} 
+    \X \tr \y_d, \\
+    \delta_{jd}^{(h+1)} & = a_d^{(h)} \[(\bmu^{(h+1)}_{jd})^2 + 
+    (\bSigma_d)_{jj}\] + \lambda_{jd}, \\ 
+    \zeta_d^{(h+1)} & = \frac{1}{2} \bigg[\mathbf{y}_d \tr \mathbf{y}_d -
+    2 \mathbf{y}_d \tr \X \bm{\mu}_d^{(h+1)} + 
+    \trace ( \X \tr \X \bm{\Sigma}_d^{(h+1)}) + 
+    (\bm{\mu}_d^{(h+1)}) \tr \X \tr \X \bm{\mu}_d^{(h+1)} \\
+    & \,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\, + 
+    \trace \[ \diag (b_{jd}^{(h)}) \bm{\Sigma}_d^{(h+1)}\] + 
+    (\bm{\mu}_d^{(h+1)}) \tr \diag (b_{jd}^{(h)}) 
+    \bm{\mu}_d^{(h+1)}\bigg],
+  \end{align*}
+  until convergence. The empirical Bayes step that consists of an iteratively 
+  reweighted least squares regression of responses $(f_j^{(l)})^{-1}$ on
+  predictors $\mathbf{z}_j$ with weights $\lambda_j^z f_j^{(l)}$, where 
+  \begin{align*}
+  f_j^{(l)} & = \E_{Q^{(h)}}(\tau_j) = [c_j^{(l)} - 
+  (D + 1)/(\delta_j^z)^{(l)}] \cdot (\delta_j^z)^{(l)} [\mathbf{z}_j \tr 
+  (\bm{\alpha}^z)^{(l)} ]^{-2} / (\lambda_j^z)^{(l)}, \\
+  \lambda_j^z & = \[ c_j^{(l)} + f_j^{(l)} 
+  (\mathbf{z}_j \tr \bm{\alpha}^z)^2 - 
+    2\mathbf{z}_j \tr \bm{\alpha}^z \]^{-1}.
+  \end{align*}
+
+  Note that so far, the design matrix $\X$ is assumed to be shared across the
+  equations. This is not a requirement, so that we may have $\X^d$, for 
+  $d=1, \dots, D$ different design matrices. As a consequence, the index $j$
+  may run over differently sized sequences.
+
   \section{Simulations}

### 2019-05-22 docs/manuscript1.Rnw
commit before push

@@
   \end{align*}
-  until convergence. The empirical Bayes step that consists of an iteratively 
-  reweighted least squares regression of responses $(f_j^{(l)})^{-1}$ on
-  predictors $\mathbf{z}_j$ with weights $\lambda_j^z f_j^{(l)}$, where 
+  until convergence, where $a_d^{(h)}$ is as before and 
+  $$
+  b_d^{(h)}=\sqrt{\frac{\delta_{jd}}
+  {\lambda_{jd}(\mathbf{c}_{jd} \bm{\alpha})^2}}
+  \frac{K_0(\sqrt{\delta_{jd}\lambda_{jd}(\mathbf{c}_{jd} \bm{\alpha})^2})}
+  {K_1(\sqrt{\delta_{jd}\lambda_{jd}(\mathbf{c}_{jd} \bm{\alpha})^2})} + 
+  \frac{2}{\delta_{jd}}.
+  $$ 
+  The empirical Bayes step that consists of an iteratively 
+  reweighted least squares regression with steps:
   \begin{align*}
-  f_j^{(l)} & = \E_{Q^{(h)}}(\tau_j) = [c_j^{(l)} - 
-  (D + 1)/(\delta_j^z)^{(l)}] \cdot (\delta_j^z)^{(l)} [\mathbf{z}_j \tr 
-  (\bm{\alpha}^z)^{(l)} ]^{-2} / (\lambda_j^z)^{(l)}, \\
-  \lambda_j^z & = \[ c_j^{(l)} + f_j^{(l)} 
-  (\mathbf{z}_j \tr \bm{\alpha}^z)^2 - 
-    2\mathbf{z}_j \tr \bm{\alpha}^z \]^{-1}.
+    \bm{\alpha} & = \[ \mathbf{C} \tr \diag 
+    (\lambda_{jd} e_{jd}^{(l)}) \mathbf{C} \]^{-1} \mathbf{C} \tr \diag 
+    (\lambda_{jd}) \mathbf{1}_{p \cdot D \times 1}, \\
+    \lambda_{jd} & = \[ b_{jd}^{(l)} + e_{jd}^{(l)} 
+    (\mathbf{c}_{jd} \tr \bm{\alpha})^2 - 
+    2\mathbf{c}_{jd} \tr \bm{\alpha} \]^{-1},
   \end{align*}
+  where $e_{jd}^{(l)} = \E_{Q^{(l)}}(\gamma_d^2) = [b_{jd}^{(l)} - 
+  2/\delta_{jd}^{(l)})] \cdot \delta_{jd}^{(l)} 
+  /[\lambda_d^{(l)} (\mathbf{c}_{jd} \tr \bm{\alpha})^2]$.
 

### 2019-06-04 docs/manuscript1.Rnw
commit before push
first ENIG simulations

--- 2019-05-21 docs/manuscript1.Rnw
@@
 	\subsection{Simultaneous equations model}\label{sec:SEM}
+	- change the name to seemingly unrelated regressions?
+
 	We have continuous measurements $y_{id}$ for observation $i=1,\dots, n$ on 
@@
       \mathcal{N} (0, \sigma_d^2 \gamma_d^2), \label{eq:betaprior} \\
-      \gamma_d^2 & \sim \mathcal{IG}(\theta_d, \lambda_d), \\
+      \gamma_d^2 & \sim \mathcal{IG} \( (\mathbf{c}_d \tr \bm{\alpha})^{-1}, 
+      \lambda_d \), \\
       \sigma_d^2 & \sim 1/\sigma_d^3,
@@
   \end{subequations}
-  where $\mathcal{IG}(\theta, \lambda)$ denotes the inverse Gaussian 
-  distribution with mean $\theta$ and shape $\lambda$.
+  where $\mathcal{IG}((\mathbf{c}_d \tr \bm{\alpha})^{-1}, \lambda_d)$ 
+  denotes the inverse Gaussian distribution with mean 
+  $(\mathbf{c}_d \tr \bm{\alpha})^{-1}$ and shape $\lambda_d$.
+
+  The normal inverse Gaussian (NIG) prior model for $\beta_{jd}$ and 
+  $\gamma^2_d$ in (\ref{eq:priormodel}) was introduced in 
+  \cite{barndorff-nielsen_hyperbolic_1978} and is, since 
+  \cite{barndorff-nielsen_normal_1997}, routinely applied in mathematical 
+  finance (see, e.g., \cite{kalemanova_normal_2007}). See SM Section 
+  \ref{sm-sec:nigprior} for more details on this prior.
+  In Figure \ref{fig:dens_igaussian_marginalbeta}, the NIG 
+  prior distribution marginalised over $\gamma_d^2$ is depicted together with 
+  three other common prior choices,
+  Student's $t$ (see Section \ref{sec:inversegammamodel} for more details on 
+  this prior), normal (ridge) and Laplace (lasso), all scaled to variance one. 
+  From this Figure we see that, depending on the choice
+  of hyperparameters, the NIG prior can vary from heavier tails than the lasso, 
+  to more Gaussian tails like the ridge, but behaves similarly to the 
+  Student's $t$ prior. Our argumentation to model the 
+  $\gamma^2_d$ by an inverse Gaussian 
+  distribution, as has been suggested in \cite{fabrizi_specification_2016} and 
+  \cite{caron_sparse_2008}, is two-fold: (i) the NIG model is more flexible than
+  the standard ridge and lasso priors (as seen from Figure 
+  \ref{fig:dens_igaussian_marginalbeta}), and (ii) the NIG prior allows to model
+  the mean as a 
+  function of the co-data $\mathbf{c}_d$ more conveniently than the Student's 
+  $t$ prior, as explained in Section 
+  \ref{sec:empiricalbayes}. 
+<<dens_igaussian_marginalbeta, fig.cap="Several choices of marginal $\\beta_j$ prior, scaled to $\\V(\\beta_j)=1$.", out.width="80%", fig.asp=2/3>>=
+library(GeneralizedHyperbolic)
+library(metRology)
+# all scaled such that variance is one
+sigma <- 1
+mgammasq <- 1/sigma^2
+lambda <- c(0.1, 2)
+lambda1 <- sqrt(2)
+eta <- lambda/mgammasq + 2
+col <- c(1:6)
+lty <- c(1:6)
+labels <- as.expression(c(bquote("NIG, "~lambda==.(lambda[1])~", "~
+                                   (bold(c)^"T"~bold(alpha))^-1==.(mgammasq)),
+                          bquote("NIG, "~lambda==.(lambda[2])~", "~
+                                   (bold(c)^"T"~bold(alpha))^-1==.(mgammasq)),
+                          bquote("Student's t, "~lambda==.(lambda[1])~", "~
+                                   eta==.(eta[1])),
+                          bquote("Student's t, "~lambda==.(lambda[2])~", "~
+                                   eta==.(eta[2])),
+                          "ridge", "lasso"))
+dprior <- function(x, lambda, mgammasq, sigma) {
+  dnig(x, 0, sigma*sqrt(lambda), sqrt(lambda/(mgammasq*sigma)), 0)
+}
+dlasso <- function(x, lambda1) {
+  0.5*lambda1*exp(-lambda1*abs(x))
+}
+
+ylim <- c(0, dt.scaled(0, eta[1]/2, 0, sqrt(lambda[1]*sigma^2/eta[1])))
+opar <- par(no.readonly=TRUE)
+par(mar=opar$mar*c(1, 1.1, 1, 1))
+curve(dprior(x, lambda[1], mgammasq, sigma), -3, 3, 
+      ylab=expression(p(beta[j])), 
+      xlab=expression(beta[j]), n=1000, ylim=ylim,
+      col=col[1], lty=lty[1])
+curve(dprior(x, lambda[2], mgammasq, sigma), add=TRUE, n=1000, 
+      col=col[2], lty=lty[2])
+curve(dnorm(x, 0, 1), add=TRUE, n=1000, col=col[3], lty=lty[3])
+curve(dlasso(x, lambda1), add=TRUE, n=1000, col=col[4], lty=lty[4])
+curve(dt.scaled(x, eta[1]/2, 0, sqrt(lambda[1]*sigma^2/eta[1])), add=TRUE, 
+      n=1000, col=col[5], lty=lty[5])
+curve(dt.scaled(x, eta[2]/2, 0, sqrt(lambda[2]*sigma^2/eta[2])), add=TRUE, 
+      n=1000, col=col[6], lty=lty[6])
+legend("topright", legend=labels, col=col, lty=lty, 
+       title="Prior", seg.len=1)
+par(opar)
+@
+
   \cite{moran_variance_2018} argue that in regression models, $\bm{\beta}_d$ 
@@
 
-  The normal inverse Gaussian (NIG) prior model for $\beta_{jd}$ and 
-  $\gamma^2_d$ in (\ref{eq:priormodel}) was introduced in 
-  \cite{barndorff-nielsen_hyperbolic_1978} and is, since 
-  \cite{barndorff-nielsen_normal_1997}, routinely applied in mathematical 
-  finance (see, e.g., \cite{kalemanova_normal_2007}). See SM Section 
-  \ref{sm-sec:nigprior} for more details on this prior.
-  In Figure \ref{fig:dens_igaussian_marginalbeta}, the NIG 
-  prior distribution marginalised over $\gamma_d^2$ is depicted together with 
-  three other common prior choices,
-  Student's $t$ (see Section \ref{sec:inversegammamodel} for more details on 
-  this prior), ridge and lasso, all scaled to variance one. From this Figure 
-  we see that, depending on the choice
-  of hyperparameters, the NIG prior can vary from heavier tails than the lasso, 
-  to more Gaussian tails like the ridge, but behaves similarly to the 
-  Student's $t$ prior. Our argumentation to model the 
-  $\gamma^2_d$ by an inverse Gaussian 
-  distribution, as has been suggested in \cite{fabrizi_specification_2016} and 
-  \cite{caron_sparse_2008}, is two-fold: (i) the NIG model is more flexible than
-  the standard ridge and lasso priors (as seen from Figure 
-  \ref{fig:dens_igaussian_marginalbeta}), and (ii) the NIG prior allows to model
-  the mean $\theta_d$ as a 
-  function of the co-data $\mathbf{c}_d$ more conveniently than the Student's 
-  $t$ prior, as explained in Section 
-  \ref{sec:empiricalbayes}. 
-<<dens_igaussian_marginalbeta, fig.cap="Several choices of marginal $\\beta_j$ prior, scaled to $\\V(\\beta_j)=1$.", out.width="80%", fig.asp=2/3>>=
-library(GeneralizedHyperbolic)
-library(metRology)
-# all scaled such that variance is one
-sigma <- 1
-theta <- 1/sigma^2
-lambda <- c(0.1, 2)
-lambda1 <- sqrt(2)
-eta <- lambda/theta + 2
-col <- c(1:6)
-lty <- c(1:6)
-labels <- as.expression(c(bquote("NIG, "~lambda==.(lambda[1])~", "~
-                                   theta==.(theta)),
-                          bquote("NIG, "~lambda==.(lambda[2])~", "~
-                                   theta==.(theta)),
-                          bquote("Student's t, "~lambda==.(lambda[1])~", "~
-                                   eta==.(eta[1])),
-                          bquote("Student's t, "~lambda==.(lambda[2])~", "~
-                                   eta==.(eta[2])),
-                          "ridge", "lasso"))
-dprior <- function(x, lambda, theta, sigma) {
-  dnig(x, 0, sigma*sqrt(lambda), sqrt(lambda/(theta*sigma)), 0)
-}
-dlasso <- function(x, lambda1) {
-  0.5*lambda1*exp(-lambda1*abs(x))
-}
-
-ylim <- c(0, dt.scaled(0, eta[1]/2, 0, sqrt(lambda[1]*sigma^2/eta[1])))
-curve(dprior(x, lambda[1], theta, sigma), -3, 3, ylab=expression(p(beta[j])), 
-      xlab=expression(beta[j]), n=1000, ylim=ylim,
-      col=col[1], lty=lty[1])
-curve(dprior(x, lambda[2], theta, sigma), add=TRUE, n=1000, 
-      col=col[2], lty=lty[2])
-curve(dnorm(x, 0, 1), add=TRUE, n=1000, col=col[3], lty=lty[3])
-curve(dlasso(x, lambda1), add=TRUE, n=1000, col=col[4], lty=lty[4])
-curve(dt.scaled(x, eta[1]/2, 0, sqrt(lambda[1]*sigma^2/eta[1])), add=TRUE, 
-      n=1000, col=col[5], lty=lty[5])
-curve(dt.scaled(x, eta[2]/2, 0, sqrt(lambda[2]*sigma^2/eta[2])), add=TRUE, 
-      n=1000, col=col[6], lty=lty[6])
-legend("topright", legend=labels, col=col, lty=lty, 
-       title="Prior", seg.len=1)
-@
-
   A few remarks on the choice of error variance prior are justified here: 
@@
   \cite[]{jeffreys_invariant_1946}. In the derivation of our Jeffrey's prior
-  for the error variance, we jointly consider an unknown mean and variance.
-
-  - Jeffrey's prior for variance $1/\sigma^{3/2}$ (a priori independent data 
-	mean and variance) or $1/\sigma^3$ (joint prior, but only mean and variance, 
-	not variance and betas).
+  for the error variance, we jointly consider an unknown mean and variance. This
+  joint consideration results in the somewhat unorthodox $1/\sigma^3$ Jeffrey's
+  prior.
 
@@
     q(\gamma^2_d) & \overset{D}{=} \mathcal{GIG}
-    \(-\frac{p+1}{2}, \lambda_d \theta_d^{-2}, \delta_d\), \\
+    \(-\frac{p+1}{2}, \lambda_d (\mathbf{c}_d \tr \bm{\alpha})^2, \delta_d\), \\
     q(\sigma^2_d) & \overset{D}{=} \Gamma^{-1} \(\frac{n + p + 1}{2}, \zeta_d\).
@@
     \nonumber\\
-    b_d^{(h)} & = \E_{Q^{(h)}}(\gamma_d^{-2}) = \sqrt{\frac{\lambda_d}
-    {\theta_d^2 \delta_d^{(h)}}} \frac{K_{\frac{p - 1}{2}} 
-    \( \sqrt{\lambda_d \delta_d^{(h)}}/\theta_d \)}
-    {K_{\frac{p+1}{2}} \( \sqrt{\lambda_d \delta_d^{(h)}}/\theta_d \)} + 
+    b_d^{(h)} & = \E_{Q^{(h)}}(\gamma_d^{-2}) = 
+    \sqrt{\frac{\lambda_d(\mathbf{c}_d \tr \bm{\alpha})^{2}}
+    {\delta_d^{(h)}}} \frac{K_{\frac{p - 1}{2}} 
+    \( \sqrt{\lambda_d \delta_d^{(h)}}(\mathbf{c}_d \tr \bm{\alpha}) \)}
+    {K_{\frac{p+1}{2}} \( \sqrt{\lambda_d \delta_d^{(h)}}
+    (\mathbf{c}_d \tr \bm{\alpha}) \)} + 
     \frac{p+1}{\delta_d^{(h)}}, \label{eq:ratiomodifiedbessel}
@@
   \subsection{Empirical Bayes}\label{sec:empiricalbayes}
-  A full Bayesian model requires the specification of hyper parameters 
-  $\bm{\theta}=\begin{bmatrix} \theta_1 & \cdots & \theta_D \end{bmatrix} \tr$ 
-  and $\begin{bmatrix} \lambda_1 & \cdots & \lambda_D \end{bmatrix} \tr$. 
-  These are abstract and hard to interpret parameters 
+  To make use of the outcome-specific codata, we parametrised the prior 
+  $\gamma_d^2$ mean as $(\mathbf{c}_d \tr \bm{\alpha})^{-1}$. This allows us to
+  include continuous co-data on the outcomes into the model. Additionally, it 
+  reduces the number of hyperparameters. A full Bayesian model then requires the
+  specification of these hyper parameters 
+  $\bm{\alpha}$ and $\begin{bmatrix} \lambda_1 & \cdots & \lambda_D 
+  \end{bmatrix} \tr$. These are abstract and hard to interpret parameters 
   for which we generally lack expert knowledge. They do, however, have a 
@@
   Simply put, empirical Bayes fits the prior model to the data and is, as such, 
-  objective (up to model specification).
+  objective (up to model specification). In our case, this results in an   
+  objective and data-driven inclusion of the co-data.
 
@@
 	$$
-	\bm{\theta}^{(l+1)},\bm{\lambda}^{(l+1)} = \underset{\bm{\theta},\bm{\lambda}}
+	\bm{\alpha}^{(l+1)},\bm{\lambda}^{(l+1)} = \underset{\bm{\alpha},\bm{\lambda}}
 	{\argmax}\E_{\cdot | \Y} 
-	[\log p(\Y, \B, \bgamma^2, \bsigma^2) | \bm{\theta}^{(l)},\bm{\lambda}^{(l)}],
+	[\log p(\Y, \B, \bgamma^2, \bsigma^2) | \bm{\alpha}^{(l)},\bm{\lambda}^{(l)}],
 	$$
@@
 	\begin{align*}
-	  \bm{\theta}^{(l+1)},\bm{\lambda}^{(l+1)} & = 
-	  \underset{\bm{\theta},\bm{\lambda}}{\argmax}\E_{Q^{(l)}} 
+	  \bm{\alpha}^{(l+1)},\bm{\lambda}^{(l+1)} & = 
+	  \underset{\bm{\alpha},\bm{\lambda}}{\argmax}\E_{Q^{(l)}} 
 	  [\log p(\Y, \B, \bgamma^2, \bsigma^2)] \\
-	  & = \underset{\bm{\theta},\bm{\lambda}}{\argmax} \E_{Q^{(l)}} 
-	  [\log \pi (\B | \bgamma^2, \bsigma^2)] + 
+	  & = \underset{\bm{\alpha},\bm{\lambda}}{\argmax} 
 	  \E_{Q^{(l)}} [\log \pi (\bgamma^2)].
@@
 
-	To make use of the outcome-specific codata, we parametrise 
-	$\theta_d = (\mathbf{c}_d \tr \bm{\alpha})^{-1}$ and estimate $\blambda$ and
-	$\bm{\alpha}$. This allows us to include continuous co-data on the outcomes
-	in an objective, data-driven manner. Additionally, it reduces the number of
-	hyperparameters even further. The estimating equations are:
+	The estimating equations are:
   \begin{align*}
@@
   (p + 1)/\delta_d^{(l)})] \cdot \delta_d^{(l)}
-  (\theta_d^{(h)})^2/\lambda_d^{(l)}$. Solving the equations is done by 
+  /[\lambda_d^{(l)}(\mathbf{c}_d \tr \alpha^{(h)})^2]$. 
+  Solving the equations is done by 
   iteratively reweighted least squares of responses $(e_d^{(l)})^{-1}$ on 
@@
 
-  \section{General model}
+  \section{Extended model I}
   In addition to the outcome co-data $\mathbf{c}_d$, we may have feature co-data
@@
   that case, we extend model (\ref{eq:priormodel}) to:
-  \begin{subequations}\label{eq:generalmodel}
+  \begin{subequations}\label{eq:extendedmodel1}
     \begin{align}
@@
       (0, \sigma_d^2 \gamma_d^2 \tau_j^2), \\
-      \gamma_d^2 & \sim \mathcal{IG}\(\theta_d^c, \lambda^c_d\), \\
-      \tau_j^2 & \sim \mathcal{IG}\(\theta_j^z, \lambda^z_j\), \\
+      \gamma_d^2 & \sim \mathcal{IG}
+      \((\mathbf{c}_d \tr \bm{\alpha})^{-1}, \lambda_d\), \\
+      \tau_j^2 & \sim \mathcal{IG}
+      \((\mathbf{z}_j \tr \bm{\theta})^{-1}, \kappa_j\), \\
       \sigma_d^2 & \sim 1/\sigma_d^3, \nonumber
@@
   \textit{a priori} inverse Gaussian distributed with mean dependent on the 
-  feature co-data $\mathbf{z}_j$. We will refer to this model as the normal
-  inverse Gaussian-inverse Gaussian (NIGIG) model. See SM Section 
-  \ref{sm-sec:nigigprior} for more details on this prior.
+  feature co-data $\mathbf{z}_j$ and hyper parameter $\bm{\theta}$. We will
+  refer to this model as the double normal
+  inverse Gaussian (DNIG) model. See SM Section 
+  \ref{sm-sec:dnigprior} for more details on this prior.
 
-  The formulation as in (\ref{eq:generalmodel}) results in an additional 
+  The formulation as in (\ref{eq:extendedmodel1}) results in an additional 
   variational distribution for the $\tau_j$:
   $$
-  q(\tau_j^2) \overset{D}{=} \mathcal{GIG}\(-\frac{D+1}{2}, \lambda^z_j 
-  (\theta_j^z)^{-2}, \delta_j^z\).
+  q(\tau_j^2) \overset{D}{=} \mathcal{GIG}\(-\frac{D+1}{2}, \kappa_j 
+  (\mathbf{z}_j \tr \bm{\theta})^{2}, \nu_j\).
   $$
@@
     \X \tr \y_d, \\
-    (\delta^c_d)^{(h+1)} & = a_d^{(h)} \left\{(\bmu^{(h+1)}_d) \tr 
+    \delta_d^{(h+1)} & = a_d^{(h)} \left\{(\bmu^{(h+1)}_d) \tr 
     \diag (c_j^{(h)}) \bmu^{(h+1)}_d + \trace \[ \diag (c_j^{(h)}) 
-    \bSigma_d^{(h+1)}\] \right\} + \lambda^c_d, \\ 
-    (\delta^z_j)^{(h+1)} & = \sum_{d=1}^D a_d^{(h)} b_d^{(h)} 
+    \bSigma_d^{(h+1)}\] \right\} + \lambda_d, \\ 
+    \nu_j^{(h+1)} & = \sum_{d=1}^D a_d^{(h)} b_d^{(h)} 
     \[(\bmu^{(h+1)}_d) \tr \bmu^{(h+1)}_d + \trace (\bSigma_d^{(h+1)}) \] + 
-    \lambda^z_j, \\ 
+    \kappa_j, \\ 
     \zeta_d^{(h+1)} & = \frac{1}{2} \bigg[\mathbf{y}_d \tr \mathbf{y}_d -
@@
     & \,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\, + 
-    b_d^{(h+1)} \trace \[ \diag (c_j^{(h)}) \bm{\Sigma}_d^{(h+1)}\] + 
-    b_d^{(h+1)} (\bm{\mu}_d^{(h+1)}) \tr \diag (c_j^{(h)}) 
+    b_d^{(h+1)} \trace \[ \diag (c_j^{(h+1)}) \bm{\Sigma}_d^{(h+1)}\] + 
+    b_d^{(h+1)} (\bm{\mu}_d^{(h+1)}) \tr \diag (c_j^{(h+1)}) 
     \bm{\mu}_d^{(h+1)}\bigg],
@@
   $$
-  c_j^{(h)} = \E_{Q^{(h)}}(\tau_j^{-2}) = \sqrt{\frac{\lambda^z_j}
-  {(\theta_j^z)^2 (\delta^z_j)^{(h)}}} 
-  \frac{K_{\frac{D - 1}{2}} 
-  \( \sqrt{\lambda^z_j (\delta^z_j)^{(h)}}/\theta_j^z \)}
-  {K_{\frac{D+1}{2}} \( \sqrt{\lambda^z_j (\delta_j^z)^{(h)}}/\theta_j^z \)} +
-  \frac{D+1}{(\delta_j^z)^{(h)}}
+  c_j^{(h)} = \E_{Q^{(h)}}(\tau_j^{-2}) = \sqrt{\frac{\kappa_j 
+  (\mathbf{z}_j \tr \bm{\theta})^2}{\nu_j^{(h)}}} 
+  \frac{K_{\frac{D - 1}{2}} \( \sqrt{\kappa_j \nu_j^{(h)}} 
+  (\mathbf{z}_j \tr \bm{\theta}) \)}{K_{\frac{D+1}{2}} 
+  \( \sqrt{\kappa_j \nu_j^{(h)}} (\mathbf{z}_j \tr \bm{\theta}) \)} +
+  \frac{D+1}{\nu_j^{(h)}}
   $$
-  If we set $\theta_d^c=(\mathbf{c}_d \tr \bm{\alpha}^c)^{-1}$ and
-  $\theta_j^z=(\mathbf{z}_j \tr \bm{\alpha}^z)^{-1}$, we have an additional
-  empirical Bayes step that consists of an iteratively 
+  We have an additional empirical Bayes step that consists of an iteratively 
   reweighted least squares regression of responses $(f_j^{(l)})^{-1}$ on
-  predictors $\mathbf{z}_j$ with weights $\lambda_j^z f_j^{(l)}$, where 
+  predictors $\mathbf{z}_j$ with weights $\kappa_j f_j^{(l)}$:
   \begin{align*}
-  f_j^{(l)} & = \E_{Q^{(h)}}(\tau_j) = [c_j^{(l)} - 
-  (D + 1)/(\delta_j^z)^{(l)}] \cdot (\delta_j^z)^{(l)} [\mathbf{z}_j \tr 
-  (\bm{\alpha}^z)^{(l)} ]^{-2} / (\lambda_j^z)^{(l)}, \\
-  \lambda_j^z & = \[ c_j^{(l)} + f_j^{(l)} 
-  (\mathbf{z}_j \tr \bm{\alpha}^z)^2 - 
-    2\mathbf{z}_j \tr \bm{\alpha}^z \]^{-1}.
+    \bm{\theta} & = \[ \mathbf{Z} \tr \diag 
+    (\kappa_j f_j^{(l)}) \mathbf{Z} \]^{-1} \mathbf{Z} \tr \diag 
+    (\kappa_j) \mathbf{1}_{p \times 1}, \\
+    \kappa_j & = \[ c_j^{(l)} + f_j^{(l)} (\mathbf{z}_j \tr \bm{\theta})^2 - 
+    2\mathbf{z}_j \tr \bm{\theta}^z \]^{-1},
+  \end{align*}
+  where $f_j^{(l)} = \E_{Q^{(h)}}(\tau_j) = [c_j^{(l)} - 
+  (D + 1)/\nu_j^{(l)}] \cdot \nu_j^{(l)}/ [\kappa^{(l)}_j (\mathbf{z}_j \tr 
+  \bm{\theta}^{(l)})^2 ]$ and $\mathbf{Z} = \begin{bmatrix} \mathbf{z}_1 & 
+  \dots & \mathbf{z}_p \end{bmatrix} \tr$.
+
+  \section{Extended model II}
+  We propose a second extension of model (\ref{eq:priormodel}) 
+  that incorporates co-data $\mathbf{c}_{d}^j$ that describes the relationship
+  between covariate $j$ and outcome $d$, collected in
+  $\mathbf{C} = \begin{bmatrix} \mathbf{c}_{1}^1 & \cdots & \mathbf{c}_{1}^p &
+  \cdots & \mathbf{c}_{D}^1 & \cdots & \mathbf{c}_{D}^p
+  \end{bmatrix}_{1 \times pD} \tr$. An example from eQTL is the 
+  distance between SNP and gene. Model (\ref{eq:priormodel}) is then rewritten
+  as:
+  \begin{subequations}\label{eq:extendedmodel2}
+    \begin{align*}
+      y_{id} | \bm{\beta}_d, \sigma_d^2 & \sim \mathcal{N} 
+      \(\x_i \tr \bm{\beta}_d, \sigma_d^2\), \nonumber \\
+      \beta_{jd} | \gamma_{jd}^2, \sigma_d^2 & \sim \mathcal{N} 
+      \(0, \sigma_d^2 \gamma_{jd}^2 \), \\
+      \gamma_{jd}^2 & \sim \mathcal{IG}
+      \( [(\mathbf{c}_{d}^j) \tr \bm{\alpha}]^{-1}, \lambda_{d} \), \\
+      \sigma_d^2 & \sim 1/\sigma_d^3, \nonumber
+    \end{align*}
+  \end{subequations}
+  where now the variance components $\gamma_{jd}^2$ are both outcome and 
+  covariate specific and modelled as a function of the co-data 
+  $\mathbf{c}_{d}^j$. We will
+  refer to this model as the extended normal
+  inverse Gaussian (ENIG) model
+  The formulation as in (\ref{eq:extendedmodel2}) results in the following
+  variational distribution for the $\gamma^2_{jd}$:
+  $$
+  q(\gamma_{jd}^2) \overset{D}{=} \mathcal{GIG}
+  \(-1, \lambda_{d}[(\mathbf{c}_{d}^j) \tr \bm{\alpha}]^2, \delta_{jd}\).
+  $$
+  The variational parameters are iteratively updated by:
+  \begin{align*}
+    \bm{\Sigma}_d^{(h+1)} & = (a_d^{(h)})^{-1} 
+    \[\X \tr \X + \diag(b_{jd}^{(h)})\]^{-1}, \\
+    \bm{\mu}_d^{(h+1)} & = \[\X \tr \X + \diag(b_{jd}^{(h)})\]^{-1} 
+    \X \tr \y_d, \\
+    \delta_{jd}^{(h+1)} & = a_d^{(h)} \[(\bmu^{(h+1)}_{jd})^2 + 
+    (\bSigma_d)_{jj}\] + \lambda_{d}, \\ 
+    \zeta_d^{(h+1)} & = \frac{1}{2} \bigg[\mathbf{y}_d \tr \mathbf{y}_d -
+    2 \mathbf{y}_d \tr \X \bm{\mu}_d^{(h+1)} + 
+    \trace ( \X \tr \X \bm{\Sigma}_d^{(h+1)}) + 
+    (\bm{\mu}_d^{(h+1)}) \tr \X \tr \X \bm{\mu}_d^{(h+1)} \\
+    & \,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\, + 
+    \trace \[ \diag (b_{jd}^{(h+1)}) \bm{\Sigma}_d^{(h+1)}\] + 
+    (\bm{\mu}_d^{(h+1)}) \tr \diag (b_{jd}^{(h+1)}) 
+    \bm{\mu}_d^{(h+1)}\bigg],
+  \end{align*}
+  until convergence, where
+  $$
+  b_{jd}^{(h)}=\sqrt{\frac{\delta_{jd}^{(h)}}
+  {\lambda_{d}[(\mathbf{c}_{d}^j) \tr \bm{\alpha}]^2}}
+  \frac{K_0(\sqrt{\delta_{jd}^{(h)}\lambda_{d}}
+  [(\mathbf{c}_{d}^j) \tr \bm{\alpha}])}
+  {K_1(\sqrt{\delta_{jd}^{(h)}\lambda_{d}}
+  [(\mathbf{c}_{d}^j) \tr \bm{\alpha}])} + 
+  \frac{2}{\delta_{jd}^{(h)}}.
+  $$ 
+  The empirical Bayes step consists of an iteratively 
+  reweighted least squares regression with iterations:
+  \begin{align*}
+    \bm{\alpha} & = \[ \mathbf{C} \tr \bLambda \diag 
+    (e_{jd}^{(l)}) \mathbf{C} \]^{-1} \mathbf{C} \tr \bLambda
+    \mathbf{1}_{pD \times 1}, \\
+    \lambda_{d} & = p \left\{ \sum_{j=1}^p b_{jd}^{(l)} + 
+    \sum_{j=1}^p e_{jd}^{(l)} [(\mathbf{c}_{d}^j) \tr \bm{\alpha}]^2 - 
+    2 \sum_{j=1}^p (\mathbf{c}_{d}^j) \tr \bm{\alpha} \right\}^{-1},
+  \end{align*}
+  where $\bLambda=\I_p \otimes \diag(\lambda_d)$ and 
+  $e_{jd}^{(l)} = (b_{jd}^{(l)} - 
+  2/\delta_{jd}^{(l)}) \cdot \delta_{jd}^{(l)}/[\lambda_{d}^{(l)}
+  \left\{(\mathbf{c}_{d}^j) \tr \bm{\alpha}]^2\right\}$.
+
+  Another option is to consider a constant $\lambda$ over the equations, i.e.,
+  $\forall d: \lambda_d = \lambda$. In that case the EB step reduces to an
+  ordinary least squares fit with:
+  \begin{align*}
+    \hat{\bm{\alpha}} & = \[ \mathbf{C} \tr \diag (e_{jd}^{(l)}) 
+    \mathbf{C} \]^{-1} \mathbf{C} \tr \mathbf{1}_{p D \times 1}, \\
+    \hat{\lambda} & = pD \[ \sum_{d=1}^D\sum_{j=1}^p b_{jd}^{(l)} + 
+    \hat{\bm{\alpha}} \tr \mathbf{C} \tr \diag (e_{jd}^{(l)}) \mathbf{C}
+    \hat{\bm{\alpha}} - 2 \hat{\bm{\alpha}} \tr \mathbf{C} \tr 
+    \mathbf{1}_{pD \times 1} \]^{-1}.
   \end{align*}
 
+  Note that so far, the design matrix $\X$ is assumed to be shared across the
+  equations. This is not a requirement. That is, we may have D different design 
+  matrices $\X^d$. These design matrices need not have the same number of 
+  columns. This becomes useful in, e.g. eQTL, where we regress genes on a
+  selection of SNPs, often based on distance from the gene. As a consequence of
+  different design matrices the number of features is now indexed by the 
+  number of equations, i.e., we have $p_d$ instead of $p$.
+
   \section{Simulations}
-  \subsection{Comparison of NIG, NIGIG and Student's \texorpdfstring{$t$}{t} 
-  priors}
+  \subsection{Comparison of priors}
   \label{sec:inversegammamodel}
@@
   We compare the 
-  posteriors of the proposed NIG and NIGIG priors to the scaled Student's 
+  posteriors of the proposed NIG, DNIG, and EGIG priors to the scaled Student's 
   $t$-distribution prior in a simulated setting.

### 2019-06-04 docs/supplement1.Rnw
first ENIG simulations

@@
 
-	\section{NIGIG prior}\label{sec:nigigprior}
+	\section{NIGIG prior}\label{sec:dnigprior}
 	Equivalently, we may set 
   $\eta_{jd}^2=\tau_j^2 \gamma_d^2/(\theta_j^z \theta_d^c)$ and write 
-  MD equation (\ref{md-eq:generalmodel}) as:
+  MD equation (\ref{md-eq:extendedmodel1}) as:
   \begin{align*}
@@
   of writing the model is:
-  \begin{subequations}\label{eq:generalprior}
+  \begin{subequations}\label{eq:extendedmodel1.2}
     \begin{align}
@@
   \end{subequations}
-	From (\ref{eq:generalprior}), for a fixed error variance
+	From (\ref{eq:extendedmodel1.2}), for a fixed error variance
 	$\sigma_d^2$, we find the marginal prior variance and kurtosis:
-	\begin{subequations}\label{eq:varkurtnigig}
+	\begin{subequations}\label{eq:varkurtdgig}
 	  \begin{align}
@@
   \sigma_d^2 \theta_j^z \theta_d^c$ and $\mathcal{K}(\beta_{jd} | \sigma_d^2)
-  \rightarrow 3(\theta_j^z/\lambda_j^z + 1)$. From (\ref{eq:varkurtnigig}) we see
+  \rightarrow 3(\theta_j^z/\lambda_j^z + 1)$. From (\ref{eq:varkurtdgig}) we see
   that we can keep the variance constant by fixing $\theta_j^z$, $\theta_d^c$,
   and $\lambda_d^c$, while we increasing the kurtosis by decrearing 
-  $\lambda_j^z$. This makes the model MD equations (\ref{md-eq:generalmodel}) a
-  flexible model.
+  $\lambda_j^z$. This makes the model MD equations (\ref{md-eq:extendedmodel1})
+  a flexible model.
 
@@
   \section{Efficient computation}
+  \subsection{NIG model}
   The empirical Bayes updates require the following quantities: $\zeta_d$, $a_d$, $\delta_d$, $\trace (\bm{\Sigma}_d)$, $\trace (\X \tr \X \bm{\Sigma}_d)$, $\bm{\mu}_d \tr \bm{\mu}_d$, and $\bm{\mu}_d \tr \X \tr \X \bm{\mu}_d$.
@@
 The rest of the calculations involve just $n$ scalar multiplications and/or additions.
+
+  \subsection{DNIG and ENIG models}
+  If we have the following forms:
+  \begin{align*}
+    \bm{\Sigma} = a^{-1} \(\tilde{\X} \tr \tilde{\X} + 
+    \tilde{\mathbf{H}}\)^{-1}, \\
+    \bm{\mu} = \(\tilde{\X} \tr \tilde{\X} + 
+    \tilde{\mathbf{H}}\)^{-1} \tilde{\X} \tr \y,
+  \end{align*}
+  with $\tilde{\X} = \begin{bmatrix} \mathbf{1} & \X \end{bmatrix}$ and
+  $$
+  \tilde{\mathbf{H}} = 
+  \begin{bmatrix} 
+    0 & \mathbf{0} \\
+    \mathbf{0} & \mathbf{H}
+  \end{bmatrix},
+  $$
+  then 
+  \begin{align*}
+    \diag (\bSigma) = a^{-1} 
+    \begin{bmatrix}
+      q \\
+      \mathbf{r}
+    \end{bmatrix},
+  \end{align*}
+  with
+  \begin{align*}
+    \mathbf{D} & = (\I_n + \X \H^{-1} \X \tr), \\
+    q & = \[ n - \mathbf{1} \X \H^{-1} \X \tr \mathbf{1} + 
+    \mathbf{1} \X \H^{-1} \X \tr \mathbf{D}^{-1} 
+    \X \H^{-1} \X \tr \mathbf{1}\]^{-1} \\
+    \mathbf{r} & = \diag( \H^{-1}) - \left\{ \[\H^{-1} \X \tr 
+    \mathbf{D}^{-1}\] \circ (\H^{-1} \X \tr) \right\} \mathbf{1} +
+    q (\H^{-1} \X \tr \mathbf{1}) \circ (\H^{-1} \X \tr \mathbf{1}) \\
+    & \,\,\,\,\, - 2q \[\H^{-1} \X \tr \mathbf{D}^{-1} 
+    \X \H^{-1} \X \tr \mathbf{1}\] \circ (\H^{-1} \X \tr \mathbf{1}) \\
+    & \,\,\,\,\, + q \[\H^{-1} \X \tr \mathbf{D}^{-1} 
+    \X \H^{-1} \X \tr \mathbf{1}\] \circ \[\H^{-1} \X \tr 
+    \mathbf{D}^{-1} \X \H^{-1} \X \tr \mathbf{1}\].
+  \end{align*}
+  Likewise we find 
+  \begin{align*}
+    \bm{\mu} =  
+    \begin{bmatrix}
+      q \sum_{i=1}^n y_i - q t + qs \\
+      \mathbf{u}
+    \end{bmatrix},
+  \end{align*}
+  with
+  \begin{align*}
+    s & = 1 \X \H^{-1} \X \tr \mathbf{D}^{-1} \X \tr \H^{-1}
+    \X \tr \y \\
+    t & = \mathbf{1} \X \H^{-1} \X \tr \y \\
+    \mathbf{u} & = q \(t - \sum_{i=1}^n y_i - 2s \) \H^{-1} \X \tr \mathbf{1} + 
+    + \H^{-1} \X \tr \y + q \( \sum_{i=1}^n y_i + s\) 
+    \H^{-1} \X \tr \mathbf{D}^{-1} \X \H^{-1} \X \tr \mathbf{1} \\
+    & \,\,\,\,\, - \H^{-1} \X \tr \mathbf{D}^{-1} \X \H^{-1} \X \tr \y.
+  \end{align*}
+  We have
+  \begin{align*}
+    \X \bmu & = q \( \sum_{i=1}^n y_i + s - t \) \mathbf{1} + 
+    q \(t - \sum_{i=1}^n y_i - 2s \) \X \H^{-1} \X \tr \mathbf{1} +
+    \X \H^{-1} \X \tr \y \\
+    & \,\,\,\,\, + q \(s + \sum_{i=1}^n \) \X \H^{-1} \X \tr \mathbf{D}^{-1} 
+    \X \H^{-1} \X \tr \mathbf{1} - \X \H^{-1} \X \tr \mathbf{D}^{-1} 
+    \X \H^{-1} \X \tr \y.
+  \end{align*}
+  Furthermore
+  \begin{align*}
+    \bSigma \X \tr & = q \( \sum_{i=1}^n y_i + s - t \) \mathbf{1} + 
+    q \(t - \sum_{i=1}^n y_i - 2s \) \X \H^{-1} \X \tr \mathbf{1} +
+    \X \H^{-1} \X \tr \y \\
+    & \,\,\,\,\, + q \(s + \sum_{i=1}^n \) \X \H^{-1} \X \tr \mathbf{D}^{-1} 
+    \X \H^{-1} \X \tr \mathbf{1} - \X \H^{-1} \X \tr \mathbf{D}^{-1} 
+    \X \H^{-1} \X \tr \y.
+  \end{align*}
 

### 2019-06-04 rpackage/R/eb_functions.R

@@ -1,3 +1,65 @@
+# update EB parameters in ENIG model (not tested)
+.eb.update.enig <- function(e, b, old.lambda, Cmat, p, D, mult.lambda, fixed.eb,
+                            epsilon, maxit) {
+    
+  # eb parameters
+  alpha <- rowSums(solve(t(Cmat*Reduce("c", e)) %*% Cmat) %*% t(Cmat))
+  if(fixed.eb=="lambda") {
+    lambda <- old.lambda
+  } else {
+    lambda <- sum(p)/(sum(Reduce("c", b)) + 
+                        sum(colSums(t(Cmat*sqrt(Reduce("c", e)))*alpha)^2) - 
+                        2*sum(t(Cmat)*alpha))  
+  }
+  
+  # IRLS
+  if(mult.lambda & fixed.eb=="none") {
+    new.eb <-  list(alpha=alpha, lambda=rep(lambda, D), conv=FALSE)
+    iter <- 0
+    while(!(new.eb$conv | iter >= maxit)) {
+      
+      # update iteration number and EB parameters
+      iter <- iter + 1
+      new.eb <- .eb.update.mult.enig(new.eb$lambda, new.eb$alpha, e, b, Cmat, 
+                                     p, D, epsilon)
+    }
+    alpha <- new.eb$alpha
+    lambda <- new.eb$lambda
+    conv <- new.eb$conv
+  } else {
+    conv <- TRUE
+    iter <- 1
+  }
+  
+  # compute inner product, mean and variance
+  Calpha <- unname(split(as.numeric(Cmat %*% alpha), rep(1:D, times=p)))
+  mprior <- lapply(1:D, function(d) {1/Calpha[[d]]})
+  vprior <- lapply(1:D, function(d) {
+    1/(Calpha[[d]]^3*ifelse(mult.lambda, lambda[d], lambda))})
+  
+  out <- list(alpha=alpha, lambda=lambda, mprior=mprior, vprior=vprior, 
+              Calpha=Calpha, conv=conv, iter=iter)
+  return(out)
+}
+
+# one ENIG eb IRLS iteration (not tested)
+.eb.update.mult.enig <- function(old.lambda, old.alpha, e, b, Cmat, p, D, 
+                                 epsilon) {
+  
+  # create new IRLS updates
+  alpha <- rowSums(solve(t(Cmat*Reduce("c", e)*rep(old.lambda, p)) %*% Cmat) %*%
[…]

### 2019-06-04 code/eQTL.R
first ENIG simulations

@@ -1,0 +1,91 @@
+#!/usr/bin/env Rscript
+
+### installation of packages
+if(!("devtools" %in% installed.packages())) {
+  install.packages("devtools")
+}
+library(devtools)
+install_github("magnusmunch/cambridge/rpackage", local=FALSE,
+               auth_token=Sys.getenv("GITHUB_PAT"))
+
+### libraries
+library(cambridge)
+library(glmnet)
+
+### load data
+load("data/P38MAPKpathwayKb100K.RData")
+
+### data preparation
+y <- sapply(Gn, "[[", 1)
+x <- lapply(sapply(Gn, "[[", 2), scale)
+G <- length(Gn)
+p <- sapply(x, ncol)
+n <- nrow(y)
+
+### creating co-data
+inorout <- lapply(Gn, "[[", 4)
+distance <- lapply(Gn, function(g) {  
+  sapply(g[[5]], function(s) {
+    (sign(prod(s - g[[3]][c(1:2)]))!=1)*
+      min(abs(s - g[[3]][c(1:2)]))})})
+genelength <- lapply(1:G, function(g) {
+  rep(diff(as.numeric(Gn[[g]][[3]][c(1, 2)])), p[g])})
+minoraf <- lapply(x, function(g) {f <- apply(g, 2, function(s) {
+  f <- mean(s); unname(min(f, 1 - f))})})
+# library(biomaRt)
+# ensembl <- useEnsembl(biomart="ensembl", dataset="hsapiens_gene_ensembl")
+# # View(listFilters(ensembl))
+# gb <- getBM(attributes=c('ensembl_exon_id', "exon_chrom_start", 
+#                          "exon_chrom_end"), filters=c("hgnc_symbol"), 
+#             values=Gn[[1]][[7]]$gene_name, mart=ensembl)
+C <- sapply(1:G, function(g) {
+  cbind(inorout=inorout[[g]], distance=distance[[g]], 
+        genelength=genelength[[g]], minoraf=minoraf[[g]])})
+
+### fitting models
+# ENIG model
+fit.enig <- enig(x, y, C, mult.lambda=FALSE, fixed.eb="none", init=NULL, 
+                 control=list(epsilon.eb=1e-3, epsilon.vb=1e-3, 
+                              epsilon.opt=sqrt(.Machine$double.eps), 
[…]

### 2019-06-04 rpackage/R/vb_functions.R

@@ -1,23 +1,55 @@
-# single VB update in general model (not tested)
-.single.vb.gen.update <- function(aold, bold, cold, thetac, thetaz, lambdac,
-                                  lambdaz, x, ytx, p, D) {
+# single VB update in ENIG model (not tested)
+.single.vb.update.enig <- function(aold, bold, Calpha, lambda, y, x, ytx, yty, 
+                                   n, p, D) {
   
   # auxiliary variables involving mu and Sigma
-  aux <- sapply(1:D, function(d) {
-    .aux.var.gen(aold[d], bold[d], cold, x, ytx[d, ])})
-    
+  aux <- .aux.var.enig(aold, bold, y, x, ytx)
+  
   # vb parameters
-  deltac <- sapply(1:D, function(d) {
-    aold[d]*(aux[[d]]$mutdiagcmu + aux[[d]]$trdiagcSigma) + lambdac[d]})
-  dz.nolambda <- sum(sapply(1:D, function(d) {
-    aold[d]*bold[d]*(aux[[d]]$mutmu + aux[[d]]$trSigma)}) + lambda)
-        
-  b <- sqrt(lambdac/deltac)/thetac^2*
-    ratio_besselK(sqrt(lambdac*deltac)/thetac, 0.5*(p + 1)) + (p + 1)/deltac
-        
-  out <- c(deltac=deltac, deltaz=deltaz, zeta=zeta, a=a, b=b, c=c, e=e, f=f)
+  delta <- aold*(aux$mu^2 + aux$dSigma) + lambda
+  b <- sqrt(lambda*Calpha^2/delta)*
+    ratio_besselK(sqrt(lambda*delta)*Calpha, 1) + 2/delta
+  zeta <- (yty - 2*aux$ytXmu + aux$trXtXSigma + aux$mutXtXmu + aux$trHSigma +
+             aux$mutHmu)/2
+  a <- (n + p + 1)/(2*zeta)
+  e <- (b - 2/delta)*delta/(lambda*Calpha^2)
+    
+  out <- list(delta=delta, zeta=zeta, a=a, b=b, e=e)
   return(out)
 }
+
+# VB update in DNIG model for all equations (not tested)
+# .vb.update.dnig <- function(aold, bold, cold, Calpha, lambda, Ztheta, kappa, 
[…]

### 2019-07-29 code/remote.Rmd
testing enig on gdsc data

---
title: "Geuvadis"
author: "Armin Rauschenberger"
date: "`r format(Sys.time(),'%d %B %Y')`"
output:
    html_document:
        theme: default
        highlight: textmate
        code_folding: hide
---

```{r knitr,include=FALSE}
knitr::opts_chunk$set(eval=FALSE)
```

********************************************************************************
rm(list=ls());gc()

# __1. Initialise__

Install missing R packages.

```{r ini_install}
inst <- rownames(utils::installed.packages())
cran <- c("R.utils","refGenome","glmnet")
if(!all(cran %in% inst)){
    for(i in seq_along(cran)){
        if(!cran[i] %in% inst){
            install.packages(cran[i])
        }
    }
}

bioc <- c("snpStats","globaltest","rtracklayer")
if(!all(bioc %in% inst)){
    source("http://bioconductor.org/biocLite.R")
    for(i in seq_along(bioc)){
        if(!bioc[i] %in% inst){
            biocLite(bioc[i])
        }
    }
}
```

Verify PLINK installation.

```{r ini_plink}
setwd("/Users/gino/Documents/project3/data_plink")
if(system('plink -version') != 0){
    utils::browseURL("https://www.cog-genomics.org/plink2")
    stop("PLINK not found!")
}
```

Define the working directory.

```{r ini_path}
path <- "/Users/gino/Documents/project3/data_plink"
```

Choose the chromosome.

```{r ini_chr}
# chr <- 1
```

********************************************************************************

# __2. Download__

Download expression data:

```{r get_expression}
site <- "https://www.ebi.ac.uk/arrayexpress/files/E-GEUV-1"
file <- "GD462.GeneQuantRPKM.50FN.samplename.resk10.txt.gz"
url <- file.path(site,file)
destfile <- file.path(path,file)
utils::download.file(url=url,destfile=destfile,method="auto")
R.utils::gunzip(filename=destfile,remove=FALSE,overwrite=TRUE)
```

Download sample annotation:

```{r get_samples}
site <- "http://www.ebi.ac.uk/arrayexpress/files/E-GEUV-1"
file <- "E-GEUV-1.sdrf.txt"
url  <- file.path(site,file)
destfile <- file.path(path,file)
utils::download.file(url=url,destfile=destfile,method="libcurl")
```

```{r get_genes}
## Gene annotation: Unknown release of reference genome hg19!
# site <- paste0("ftp://ftp.ensembl.org/pub/release-75/gtf/homo_sapiens/")
# file <- paste0("Homo_sapiens.GRCh37.75.gtf.gz")
# url  <- file.path(site,file)
# destfile <- file.path(path,file)
# utils::download.file(url=url,destfile=destfile,method="auto")
# R.utils::gunzip(filename=destfile,remove=FALSE,overwrite=TRUE)
###############################################################################
mart <- biomaRt::useMart(biomart="ensembl",dataset="hsapiens_gene_ensembl")
name <- substr(Y$map$TargetID,start=1,stop=15)
genesAnn <- biomaRt::getBM(attributes=c("ensembl_gene_id","hgnc_symbol","chromosome_name","start_position","end_position"), filters ="ensembl_gene_id",values =name,mart=mart)
martSnp <- biomaRt::useEnsembl(biomart="ENSEMBL_MART_SNP",dataset="hsapiens_snp")
name = head(X$map$snp.name)
###############################################################################

gtf_dir <- "gencode.v12.annotation.gtf"
gtf <- import(gtf_dir)
head(gtf)

### Keep protein coding genes only
index <- mcols(gtf)$gene_type == "protein_coding" & mcols(gtf)$type == "gene"
gtf2  <- gtf[index]
P38MAPK_path <- c("ATF1", "ATF2", "ATF6", "CEBPB", "CREB1", "CSNK2A1", "CSNK2A2", "CSNK2B", "DDIT3", "EIF4E", "EIF4EBP1", "ELK4", "ESR1", "GDI1", "HBP1", "HSPB1", "JUN", "KRT19", "KRT8", "MAPK11", "MAPK14", "MAPKAPK2", "MAPKAPK3", "MAPKAPK5", "MEF2A", "MEF2C", "MITF", "MKNK1", "NOS2", "PLA2G4A", "PPARGC1A", "PTGS2", "RAB5A", "RPS6KA4", "RPS6KA5", "SLC9A1", "TP53", "USF1", "BLK", "CCM2", "CDC42", "DUSP1", "DUSP10", "DUSP16", "DUSP8", "FGR", "FYN", "HCK", "LCK", "LYN", "MAP2K3", "MAP2K4", "MAP2K6", "MAP3K12", "MAP3K3", "PAK1", "PAK2", "PAK3", "PRKG1", "RAC1", "RALA", "RALB", "RIPK1", "SRC", "TAB1", "TRAF6", "YES1", "CCND1", "EEF2K", "MAPK12", "MAPK13", "MAPT", "PKN1", "SNTA1", "STMN1", "AC013461.1", "CDC25B", "ETV1", "LSP1", "RAF1", "SFN", "SRF", "TCF3", "TH", "TSC2", "YWHAB", "YWHAE", "YWHAG", "YWHAH", "YWHAQ", "YWHAZ", "ATM", "CAMK2B", "GADD45A", "GADD45B", "GADD45G", "MAP3K1", "MAP3K10", "MAP3K4", "MAP3K5", "MAP3K6", "MAP3K7", "TAB2", "TAOK1", "TAOK2", "TAOK3", "TRAF2", "TXN")

P38MAPK <- c("ATF1", "ATF2", "ATF6", "CEBPB", "CREB1", "CSNK2A1", "CSNK2A2", "CSNK2B", "DDIT3", "EIF4E", "EIF4EBP1", "ELK4", "ESR1", "GDI1", "HBP1", "HSPB1", "JUN", "KRT8", "MAPK11", "MAPK14", "MAPKAPK2", "MAPKAPK3", "MAPKAPK5", "MEF2A", "MEF2C", "MITF", "MKNK1", "NOS2", "PLA2G4A", "PPARGC1A", "PTGS2", "RAB5A", "RPS6KA4", "RPS6KA5", "SLC9A1", "TP53", "USF1", "BLK", "CCM2", "CDC42", "DUSP1", "DUSP10", "DUSP16", "DUSP8", "FGR", "FYN", "HCK", "LCK", "LYN", "MAP2K3", "MAP2K4", "MAP2K6", "MAP3K12", "MAP3K3", "PAK1", "PAK2", "PRKG1", "RAC1", "RALA", "RALB", "RIPK1", "SRC", "TAB1", "TRAF6", "YES1", "CCND1", "EEF2K", "MAPK12", "MAPK13", "MAPT", "PKN1", "SNTA1", "STMN1", "AC013461.1", "CDC25B", "LSP1", "RAF1", "SFN", "SRF", "TCF3", "TSC2", "YWHAB", "YWHAE", "YWHAG", "YWHAH", "YWHAQ", "YWHAZ", "ATM", "CAMK2B", "GADD45A", "GADD45B", "GADD45G", "MAP3K1", "MAP3K10", "MAP3K4", "MAP3K5", "MAP3K6", "MAP3K7", "TAB2", "TAOK1", "TAOK2", "TAOK3", "TRAF2", "TXN")

sum(mcols(gtf2)$gene_name %in% P38MAPK)
sum(P38MAPK %in% mcols(gtf2)$gene_name)
# P38MAPK_path[which(!P38MAPK_path %in% mcols(gtf2)$gene_name)]
# which(substr(mcols(gtf2)$gene_id, start=1,stop=15)=="ENSG00000091436")
# which(substr(mcols(gtf2)$gene_id, start=1,stop=15)=="ENSG00000224330")
# which(mcols(gtf2)$gene_name=="AC005019.2")
# which(mcols(gtf2)$gene_name=="AC013461.1")
# mcols(gtf2)[2863,]
# gtf2[2863,]
# AC013461.1 = ZAK
gtf_P38MAPK <- gtf2[mcols(gtf2)$gene_name %in% P38MAPK]
P38MAPK_id  <- substr(mcols(gtf_P38MAPK)$gene_id, start=1,stop=15)
head(mcols(gtf_P38MAPK))
head(mcols(gtf_P38MAPK@ranges))
df_iranges <- data.frame(iranges = gtf_P38MAPK@ranges)
sum(df_iranges$iranges.start %in% Y$map$Coord)
sum(Y$map$Coord %in% df_iranges$iranges.start)
#####
P38MAPK_id %in% substr(as.character(Y$data$TargetID), start=1, stop=15)
index2      <- match(substr(as.character(Y$data$TargetID), start=1, stop=15), P38MAPK_id)
P38MAPK_id  <- P38MAPK_id[index2]

df_gtf_P38MAPK <- mcols(gtf_P38MAPK)[index2,]
df_iranges  <- df_iranges[index2,]
```

Load and order expression data.

```{r pro_expression}
Y <- list()
Y$data <- utils::read.table(file=file.path(path,"GD462.GeneQuantRPKM.50FN.samplename.resk10.txt"),header=TRUE,nrows=-1)
Y$data <- Y$data[order(Y$data$Coord),]
Y$data <- Y$data[order(Y$data$Chr),]
Y$data <- Y$data[substr(as.character(Y$data$TargetID), start=1, stop=15) %in% P38MAPK_id,]
Y$map  <- Y$data[,c(1:4)]
Y$map_gtf  <- df_gtf_P38MAPK
Y$exData <- t(as.matrix(Y$data[,-c(1:4)]))
Y$map$Chr
substr(as.character(Y$map$TargetID), start=1, stop=15) == P38MAPK_id  
intersect(df_iranges$iranges.start, Y$map$Coord)
# path_geneId[which(!path_geneId %in% substr(as.character(Y$data$TargetID), start=1, stop=15))]
# gtf_pathAnno[which(!path_geneId %in% substr(as.character(Y$data$TargetID), start=1, stop=15))]

# which(substr(as.character(Y$data$TargetID), start=1, stop=15)=="ENSG00000203879")
ENSG00000203879
which(Y$map$Chr=="X")
# Y$map[23606,]
# sum(path_geneId %in% substr(as.character(Y$data$TargetID), start=1, stop=15))
# Y$data <- Y$data[Y$data$Chr==chr,] # test this???
for(i in 1:22){
  indices <- which(Y$map$Chr==i)
  if(length(indices)>=1){
    for(j in indices){
      chr <- i
      ## Download genotype data for chromosome i
      site <- "http://www.ebi.ac.uk/arrayexpress/files/E-GEUV-1/genotypes"
      file <- paste0("GEUVADIS.chr",chr,".PH1PH2_465.IMPFRQFILT_BIALLELIC_PH.annotv2.genotypes.vcf.gz")
      url <- file.path(site,file)
      destfile <- file.path(path,file)
      utils::download.file(url=url,destfile=destfile,method="libcurl")

      ## Exclude SNPs with $\text{maf}<5\%$ or $\text{geno}>0\%$
      setwd(path)
      system(paste0("plink --vcf GEUVADIS.chr",chr,".PH1PH2_465.IMPFRQFILT_BIALLELIC_PH.annotv2.genotypes.vcf.gz        --maf 0.05 --geno 0 --make-bed --out snps",chr),invisible=FALSE)

      chr <- 22
      ## Load genotype data
      library(snpStats)
      bed <- file.path(path,paste0("snps",chr,".bed"))
      bim <- file.path(path,paste0("snps",chr,".bim"))
      fam <- file.path(path,paste0("snps",chr,".fam"))
      X   <- snpStats::read.plink(bed=bed,bim=bim,fam=fam)
      X$fam <- NULL
      names(X)[1] <- "data"
      all(diff(X$map$position) > 0)

      ##Exclude samples without expression or genotype data
      intersect  <- intersect(rownames(Y$eData),rownames(X$data))
      Y$eData   <- Y$eData[intersect,]
      X$data     <- X$data[intersect,]

      ## Exclude samples from population YRI
      samples <- utils::read.delim(file=file.path(path,"E-GEUV-1.sdrf.txt"))
      match   <- match(rownames(Y$eData),samples$Source.Name)
      cond    <- samples$Factor.Value.population[match]!="YRI"
      Y$eData  <- Y$eData[cond,]
      X$data  <- X$data[cond,]

      ## Exclude SNPs with $\text{maf}<5\%$
      maf <- snpStats::col.summary(X$data)$MAF
      cond <- maf >= 0.05
      X$data <- X$data[,cond]
      X$map <- X$map[cond,]

      ## Exclude genes with zero-variance
      cond <- apply(Y$eData,2,function(x) sd(x)>0)
      Y$eData <- Y$eData[,cond]
      Y$map <- Y$map[cond,]
    }
  } 
}    
```

Download genotype data:

```{r get_genotype}
site <- "http://www.ebi.ac.uk/arrayexpress/files/E-GEUV-1/genotypes"
file <- paste0("GEUVADIS.chr",chr,".PH1PH2_465.IMPFRQFILT_BIALLELIC_PH.annotv2.genotypes.vcf.gz")
url <- file.path(site,file)
destfile <- file.path(path,file)
utils::download.file(url=url,destfile=destfile,method="libcurl")
```

Exclude SNPs with $\text{maf}<5\%$ or $\text{geno}>0\%$.

```{r get_selection}
setwd(path)
system(paste0("plink --vcf GEUVADIS.chr",chr,".PH1PH2_465.IMPFRQFILT_BIALLELIC_PH.annotv2.genotypes.vcf.gz --maf 0.05 --geno 0 --make-bed --out snps",chr),invisible=FALSE)
```

********************************************************************************

## __3. Process__

Load genotype data:

```{r pro_genotype}
library(snpStats)
bed <- file.path(path,paste0("snps",chr,".bed"))
bim <- file.path(path,paste0("snps",chr,".bim"))
fam <- file.path(path,paste0("snps",chr,".fam"))
X   <- snpStats::read.plink(bed=bed,bim=bim,fam=fam)
X$fam <- NULL
names(X)[1] <- "data"
all(diff(X$map$position) > 0)
```

Exclude samples without expression or genotype data.

```{r pro_samples}
intersect <- intersect(rownames(Y$data),rownames(X$data))
Y$data    <- Y$data[intersect,]
X$data    <- X$data[intersect,]
```

Exclude samples from population YRI.

```{r pro_population}
samples <- utils::read.delim(file=file.path(path,"E-GEUV-1.sdrf.txt"))
match   <- match(rownames(Y$data),samples$Source.Name)
cond    <- samples$Factor.Value.population[match]!="YRI"
Y$data  <- Y$data[cond,]
X$data  <- X$data[cond,]
```

Exclude SNPs with $\text{maf}<5\%$.

```{r pro_frequency}
maf <- snpStats::col.summary(X$data)$MAF
cond <- maf >= 0.05
X$data <- X$data[,cond]
X$map <- X$map[cond,]
```

Exclude genes with zero-variance.

```{r pro_variance}
cond <- apply(Y$data,2,function(x) sd(x)>0)
Y$data <- Y$data[,cond]
Y$map <- Y$map[cond,]
```

********************************************************************************

## __4. Match__

```{r mat_annot}
## Identify all protein-coding genes:
# map <- list()
# setwd(path)
# object <- refGenome::ensemblGenome()
# refGenome::basedir(object) <- file.path(path)
# filename <- paste0("Homo_sapiens.GRCh37.75.gtf")
# refGenome::read.gtf(object,filename=filename)
# map <- refGenome::getGenePositions(object=object,by="gene_id")
# map <- map[map$seqid==chr & map$gene_biotype=="protein_coding",
#            c("gene_id","seqid","start","end")]
## Unknown release of reference genome hg19!
```

Identify SNPs within one million base pairs around gene start.

```{r mat_compress}
p <- ncol(Y$data)
match <- data.frame(from=integer(length=p),to=integer(length=p))

pb <- utils::txtProgressBar(min=0,max=p,style=3)
for(i in seq_len(p)){ # decrease p to sum(Y$map$Chr==1)
    utils::setTxtProgressBar(pb=pb,value=i)

    # conditions: same chromosome, within window
    cond1 <- X$map$chromosome == Y$map$Chr[i]
    if(!any(cond1)){
        stop("Wrong chromosome!")
    }
    cond2 <- X$map$position >= Y$map$Coord[i] - 1*10^6 # change!
    cond3 <- X$map$position <= Y$map$Coord[i] + 1*10^6 # change!
    which <- as.integer(which(cond1 & cond2 & cond3))

    # compression: first and last entry
    if(length(which)==0){next}
    match$from[i] <- min(which)
    match$to[i] <- max(which)
    if(length(which)==1){next}
    if(!all(diff(which)==1)){
        stop("Wrong order!")
    }
}
```

********************************************************************************

## __5. Predict__

Regress gene expression on SNPs (example).

```{r pre_lasso}
i <- sample(which(Y$map$Chr==chr),size=1) # index
if(Y$map$Chr[i]!=chr){
    stop("Wrong chromosome!")
}
y <- Y$data[,i]
if((match$to[i]-match$from[i]<2)){
    warning("Too few covariates!")
} 
x <- X$data[,seq(from=match$from[i],to=match$to[i],by=1)]
x <- abs(as(x,"numeric")-2) # major/minor allele
net <- glmnet::cv.glmnet(y=y,x=x)
```

### 2019-07-29 docs/manuscript1.Rnw
testing enig on gdsc data

@@
   \cdots & \mathbf{c}_{D}^1 & \cdots & \mathbf{c}_{D}^p
-  \end{bmatrix}_{1 \times pD} \tr$. An example from eQTL is the 
+  \end{bmatrix}_{|\bm{\alpha}| \times pD} \tr$. An example from eQTL is the 
   distance between SNP and gene. Model (\ref{eq:priormodel}) is then rewritten
@@
   \section{Simulations}
+  Our simulation study consists of three parts investigating the model and its
+  estimation:
+  \begin{enumerate}
+    \item Prior performance
+    \item Variational Bayes approximation
+    \item Empirical Bayes estimation
+  \end{enumerate}
+
   \subsection{Comparison of priors}
@@
   $$
-  \gamma_d^2 \sim \Gamma^{-1}(\eta_d/2,\lambda_d/2),
+  \gamma_d^2 \sim \Gamma^{-1}(\eta_d/2,\xi_d/2),
   $$
-  with shape $\eta_d/2$ and scale $\lambda_d/2$, both rescaled to match our 
+  with shape $\eta_d/2$ and scale $\xi_d/2$, both rescaled to match our 
   inverse Gaussian prior formulation. The resulting marginal prior for the model
@@
   \beta_j | \sigma_d^2 \sim \text{Student's } 
-  t\(\eta_d,\sigma_d \sqrt{\lambda_d/\eta_d}\),
+  t\(\eta_d,\sigma_d \sqrt{\xi_d/\eta_d}\),
   $$
-  with degrees of freedom $\eta_d$ and scale $\sigma_d \sqrt{\lambda_d/\eta_d}$.
+  with degrees of freedom $\eta_d$ and scale $\sigma_d \sqrt{\xi_d/\eta_d}$.
   See SM Section \ref{sm-sec:inversegamma} for more details on this prior. 
@@
   In the sparse setting, the observed data generating process is the same. The
-  model parameters are are simulated from a spike-and-slab prior with a 
-  Gaussian slab:
+  model parameters are are simulated from a spike-and-slab prior with both the
+  spike and slab Gaussian:
   \begin{align*}
-    w & \sim \text{Beta}(\nu, \kappa), \\
-    z_j | w & \sim \text{Binom}(w), \\
-    \beta_j | z_j & \sim z_j \mathcal{N} \( 0, 1\) +
-    (1 - z_j)\delta_0,
+    w & \sim \text{Beta}(\phi, \chi), \\
+    z_j | w & \sim \text{Bernoulli}(w), \\
+    \beta_j | z_j & \sim z_j \mathcal{N} \( 0, 1 + 0.99 \frac{\chi}{\phi}\) +
+    (1 - z_j)\mathcal{N} \( 0, 0.01\).
   \end{align*}
-  where $\delta_0$ denotes the Dirac mass at zero. The slab is chosen
+  The slab is chosen
   such that the we have marginal prior variance $\V(\beta_j)=1$. The sparsity 
   level is determined by the proportion of zero $\beta_j$, 
-  $s=1 - \nu/(\nu + \kappa)$. Note that the resulting marginal prior for
+  $s=1 - \phi/(\phi + \chi)$. Note that the resulting marginal prior for
   $\beta_j$ is:
   $$
-  \beta_j \sim (s - 1)\mathcal{N} \( 0, 1\) + s\delta_0.
+  \beta_j \sim (1 - s)\mathcal{N} \( 0, \frac{1 - 0.01 s}{1 - s}\) + 
+  s \mathcal{N} \( 0, 0.01\).
   $$
@@
   hyperparameters for estimation are set the same as in the dense setting, 
-  resulting in a total of 12 different simulation settings.
+  resulting in a total of 12 different simulation settings 
+  (6 per sparsity setting).
   To mitigate the effect of random variation, we
@@
 
-  \section{Empirical Bayes estimation}
+  \subsection{Empirical Bayes estimation}
   We simulate from a correctly specified model given in \ref{eq:priormodel}. We 
@@
 
-  \section{Application}
-  - Possible: network reconstruction, medical surveys, drug response prediction
+  \section{Application in eQTL detection}
+  - Genes from p38MAPK pathway. Possible co-data are:
+  \begin{itemize}
+    \item gene $\times$ SNP: distance of SNP to gene (continuous)
+    \item gene: length of the gene (continuous) 
+      (and possibly chromosome number (categorical))
+    \item SNP: within or outside of gene (binary), intron or exon (binary),
+      and minor allele frequency (continuous)
+  \end{itemize}
+  Within or outside of gene and distance of SNP to gene are probably highly
+  collinear.
 
-  - metabolomics data: oxidative stress als uitkomst van de andere metabolieten
+  From Gino: 108 Geuvadis genes, 9 removed because they contained just one 
+  subject. Annotation is from GRCh37 reference genome: 
+  \url{https://www.ncbi.nlm.nih.gov/assembly/GCF_000001405.13/}. 
+  Annotation file that Gino used: 
+  \url{https://www.gencodegenes.org/human/release_12.html}.
+  Ensembl version of this genome:
+  \url{https://grch37.ensembl.org/Homo_sapiens/Info/Index}.
+  Possibly use GeuvadisTranscriptExpr R package.
+
+  - compare to intercept only co-data model, i.e., 
+  $\mathbf{C} = \mathbf{1}_{pD \times 1}$.
+
+  - show influence of co-data with standardized $\bm{\alpha}$'s.
+
+  - comparison in terms of explained heritability by plotting explained 
+  heritability versus number of SNPs used to calculate explained heritability,
+  where explained heritability is fraction of heritability explained with top 
+  SNPs and heritability explained with all SNPs. We may calculate the AUC of
+  this curve to quantify performance in one number per regression.
+
+  The posterior mean of the heritability is approximated by 
+  \begin{align*}
+    \E(h_d^2 | \y_d) & = \E \( \frac{\sigma^2_d \sum_{j=1}^{p_d} 
+    \gamma_{jd}^2}{\sigma^2_d \sum_{j=1}^{p_d} \gamma_{jd}^2 + 
+    \sigma_d^2} \bigg| \y_d \) 
+    \approx \E_Q \( \frac{\sum_{j=1}^{p_d} \gamma_{jd}^2}
+    {\sum_{j=1}^{p_d} \gamma_{jd}^2 + 1} \) \\
+    & \approx \frac{\sum_{j=1}^{p_d} \E_Q (\gamma_{jd}^2)}
+    {\sum_{j=1}^{p_d} \E_Q (\gamma_{jd}^2) + 1} - 
+    \frac{\sum_{j=1}^{p_d} \V_Q (\gamma_{jd}^2)}
+    {\[\sum_{j=1}^{p_d} \E_Q (\gamma_{jd}^2) + 1\]^3},
+  \end{align*}
+  where the last step is the result of a second order Taylor approximation.
+
+  Also compare by log pseudo marginal likelihood 
+  $LPML_d = n^{-1}\sum_{i=1}^n \log (CPO_{id})$ \cite[]{gelfand_model_1992}. Let 
+  $\theta_{id}=\x_i \tr \bbeta_d$. Then the conditional 
+  predictive ordinates (CPO) are given by:
+  \begin{align*}
+    CPO_{id} & = p(\y_{id} | \y_{-id}) = \int_{-\infty}^{\infty} \int_0^{\infty}
+    p(\y_{id} | \bbeta_d, \sigma_d^2 ) \pi (\bbeta | \y_{-id}, \sigma_d^2) 
+    \pi (\sigma_d^2 | \y_{-id}) d\sigma_d^2 d\bbeta_d \\
+    & = \int_{-\infty}^{\infty} \int_0^{\infty}
+    p(\y_{id} | \theta_{id}, \sigma_d^2 ) 
+    \pi (\theta_{id} | \y_{-id}, \sigma_d^2) 
+    \pi (\sigma_d^2 | \y_{-id}) d\sigma_d^2 d\theta_{id},
+  \end{align*}
+  where
+  \begin{align*}
+    \y_i | \theta_{id}, \sigma_d^2 & \sim \mathcal{N} (\theta_{id}, 
+    \sigma_d^2), \\
+    \theta_{id} | \y_{-id} & \sim \mathcal{N}(\x_i \tr \bmu_{-id}, 
+    \x_i \tr \bSigma_{-id} \x_i), \\
+    \sigma_d^2 | \y_{-id} & \sim \Gamma^{-1} \( \frac{n - n_{id} + p + 1}{2},
+    \zeta_{-id} \),
+  \end{align*}
+  $n_{id}$ denotes the number of elements in $\y_{id}$ and $\bmu_{-id}$, 
+  $\bSigma_{-id}$, and $\zeta_{-id}$ are the corresponding parameters estimated
+  from $\y_{-id}$. Plugging in the densities we see that we have:
+  \begin{align*}
+    CPO_{id} = & (2 \pi)^{-1} \frac{\Gamma\(\frac{n - n_{id} + p + 2}{2}\)}
+    {\Gamma\(\frac{n - n_{id} + p + 1}{2}\)}
+    (\zeta_{-id} \x_i \tr \bSigma_{-id} \x_i)^{-1/2} \\
+    & \cdot \int_{-\infty}^{\infty}
+    \( \frac{\theta^2}{2\zeta_{-id}} + 1 \)^{-\frac{n - n_{id} + p + 2}{2}}
+    \exp \[ - \frac{(\theta - \x_i \tr \bmu_{-id} + y_i)^2}
+    {2 \x_i \tr \bSigma_{-id} \x_i} \] d\theta,
+  \end{align*}
+  a one-dimensional (well-behaved) integral that is easily solved with some 
+  numerical integration routine
+
+  - Also possible: network reconstruction, medical surveys, drug response 
+  prediction, metabolomics data: oxidative stress als uitkomst van de andere 
+  metabolieten
   min organic acids omdat ze downstream zijn. oxidateve stress onderverdelen in
@@
   als predictor co-data.
+
   \section{Discussion}
@@
   accurate
+
+  - issues: co-data needs to be positive
 

### 2019-07-29 docs/manuscript2.Rnw
testing enig on gdsc data

@@
 
+	From Oslo:
+	1.2) Preselecting GEX features: we selected gene expression features
+explaining 50\% variations over all cell lines, since this results in not too
+features and probably remains those important features.
+
+1.3) Preselecting CNV: we used the same strategy as Garnett et al. (2012) where the GDSC data are from.
+
+1.4) Preselecting MUT: we deleted mutated genes with less than two cell lines.
+
 	\subsection{CCLE}

### 2019-07-29 docs/supplement1.Rnw
testing enig on gdsc data

@@
 
-	\section{NIG prior}\label{sec:nigprior}
+	\section{NIG and ENIG prior}\label{sec:nigprior}
 	For a fixed error variance $\sigma_d^2$, we have a marginal prior variance 
@@
 
-	\section{NIGIG prior}\label{sec:dnigprior}
+	\section{DGIG prior}\label{sec:dnigprior}
 	Equivalently, we may set 

### 2019-07-29 code/eQTL.R

@@ -10,7 +10,9 @@ install_github("magnusmunch/cambridge/rpackage", local=FALSE,
 
 ### libraries
 library(cambridge)
+library(biomaRt)
 library(glmnet)
+library(MASS)
 
 ### load data
 load("data/P38MAPKpathwayKb100K.RData")
@@ -22,6 +24,14 @@ G <- length(Gn)
 p <- sapply(x, ncol)
 n <- nrow(y)
 
+set.seed(2019)
+id.train <- sample(1:n, floor(n/3))
+ytrain <- y[id.train, ]
+ytest <- y[-id.train, ]
+
+xtrain <- sapply(x, function(s) {s[id.train, ]})
+xtest <- sapply(x, function(s) {s[-id.train, ]})
+
 ### creating co-data
 inorout <- lapply(Gn, "[[", 4)
 distance <- lapply(Gn, function(g) {  
@@ -30,62 +40,412 @@ distance <- lapply(Gn, function(g) {
       min(abs(s - g[[3]][c(1:2)]))})})
 genelength <- lapply(1:G, function(g) {
   rep(diff(as.numeric(Gn[[g]][[3]][c(1, 2)])), p[g])})
-minoraf <- lapply(x, function(g) {f <- apply(g, 2, function(s) {
-  f <- mean(s); unname(min(f, 1 - f))})})
-# library(biomaRt)
-# ensembl <- useEnsembl(biomart="ensembl", dataset="hsapiens_gene_ensembl")
-# # View(listFilters(ensembl))
-# gb <- getBM(attributes=c('ensembl_exon_id', "exon_chrom_start", 
-#                          "exon_chrom_end"), filters=c("hgnc_symbol"), 
-#             values=Gn[[1]][[7]]$gene_name, mart=ensembl)
+minoraf <- lapply(sapply(Gn, "[[", 2), function(g) {
+  f <- apply(g, 2, function(s) {f <- mean(s); unname(min(f, 1 - f))})})
+
+### retrieve exon status of SNPs
+# View(listEnsembl())
+geneid <- sapply(Gn, function(g) {
+  strsplit(g[[7]]$gene_id, "\\.")[[1]][1]})
+ensembl <- useEnsembl(biomart="ensembl", dataset="hsapiens_gene_ensembl",
+                      GRCh=37)
+# View(listFilters(ensembl))
[…]

### 2019-08-13 docs/manuscript1.Rnw
updated simulations and manuscript

--- 2019-05-22 docs/manuscript1.Rnw
@@
 	\section{Introduction}
-	- Motivation:
-	simultaneous regressions in: network reconstruction, eQTL mapping 
-	\cite[]{kpogbezan_incorporating_2019}, drug target discovery 
-	\cite[]{noh_inferring_2016}.
+	Recently, the dimensions of data sets have increased in three directions:
+	(i) samples, (ii) variables, and (iii) outcomes. Here, we will 
+	concern ourselves with directions (ii) and (iii). More spefically, we will 
+	focus on the regression setting
+	where we have multiple outcomes and a medium to high dimensional feature 
+	space, i.e., $p \gtrsim n$. Standard methods
+	underperform or even break down in such medium to high dimensional settings, 
+	so new methodology is required.
 
-	- Introduce data example
+	As a motivating example, we present an expression quantitative loci (eQTLs)
+	detection problem (see Section \ref{sec:application}). In eQTLs detection, 
+	the aim is to find a set of single nucleotide loci (SNPs) that explains a
+	(part of a) gene's expression. The resulting heritability of the gene 
+	expression traits is thought to be indicative of the relationship between 
+	genotype and phenotype. eQTLs detection is often done for a set of genes 
+	simultaneously and as such comprises a simultaneous-equations model (SEM) with
+	multiple outcomes and a high dimensional feature set (the SNPs). Apart from
+	eQTLs, SEM models are also applied in network reconstruction 
+	\cite[]{kpogbezan_incorporating_2019, leday_gene_2017} 
+	and drug target discovery \cite[]{noh_inferring_2016}, [! ref to second 
+	manuscript].
 
-	- Co-data introduction
+	In many multi-outcome regression settings, extra information on both the 
+	outcomes and 
+	features is available. In our motivating eQTLs detection example, extra 
+	information may come in the form of (i) gene annotation (e.g., length of the
+	gene), (ii) SNP annotation (e.g., whether the SNP is in the gene or not), 
+	(iii) SNP-gene annotation (e.g., distance of SNP to gene). The presented 
+	method allows for the incorporation of such information to aid in the 
+	estimation, by pooling information both across equations (genes) and 
+	features (SNPs).
 
-	- current literature
+	We are not the first to propose to learn from external data 
+	across equations and features. \cite{kpogbezan_incorporating_2019} uses 
+	the presence or absence of edges in a normal tissue apoptosis pathway, to aid 
+	in the estimation of a cancer tissue apoptosis pathway. Furhtermore, in an 
+	eQTLs detection problem they include the location of the SNPs (inside
+	the gene or not) to aid in the estimation. \cite{oman_minimax_2002} uses 
+	empirical Bayes to learn from external data concerning both the outcomes and 
+	the features in a low dimensional chemometrics dataset. Our proposed method,
+	termed [! acronym here],
+	extends \cite{kpogbezan_incorporating_2019,kpogbezan_empirical_2017} by 
+	allowing for the inclusion of continuous external information and several 
+	sources of information.
 
-	- our solution: adaptive borrowing of information through empirical Bayes, 
-	guided by co-data.
+	- think of acronym
 
-	\section{Model}
+	The rest of the paper is structured as follows. In Section 
+	\ref{sec:model} we introduce our model, the estimation of which is detailed
+	in Section \ref{sec:estimation}. In Sections \ref{sec:simulations} and
+	\ref{sec:application} we demonstrate the method in simulations and an 
+  eQTLs detection problem, respectively. We end with a discussion in Section 
+  \ref{sec:discussion}.
+
+	\section{Model}\label{sec:model}
 	\subsection{Simultaneous equations model}\label{sec:SEM}
@@
 	$j=1,\dots, p$, collected in $\x_i = \begin{bmatrix} x_{i1} & \dots & x_{ip} 
-	\end{bmatrix} \tr$. We have characteristics of the outcomes available in the 
-	form of `co-data' $\mathbf{C} = \begin{bmatrix} \mathbf{c}_1 & \dots & 
-	\mathbf{c}_D \end{bmatrix} \tr$.
+	\end{bmatrix} \tr$ and centered per feature. We have $|\bm{\alpha}|$ 
+	external variables available that describes the outcome $d$, feature $j$, or
+	relationship between feature $j$ and outcome $d$, collected in 
+  $\mathbf{C} = \begin{bmatrix} \mathbf{c}_{1}^1 & \cdots &
+	\mathbf{c}_{1}^p & \cdots & \mathbf{c}_{D}^1 & \cdots & \mathbf{c}_{D}^p
+  \end{bmatrix}_{|\bm{\alpha}| \times pD} \tr$. An example of such an external
+  variable from eQTL detection is the distance between SNP and gene.
 
-	We model the outcomes as a linear function of the features:
+	We model the outcomes $y_{id}$ as a linear function of the features $x_{ij}$:
 	\begin{subequations}\label{eq:linearmodel}
@@
 	We assign a flat prior to the intercept terms $\beta_{0d}$ and integrate 
-	them out. The remaining parameters are endowed with the following priors:
+	them out. The remaining parameters are endowed with priors such that we end
+	up with the following model:
 	\begin{subequations}\label{eq:priormodel}
@@
       y_{id} | \bm{\beta}_d, \sigma_d^2 & \sim \mathcal{N} 
-      (\x_i \tr \bm{\beta}_d, \sigma_d^2), \\
-      \beta_{jd} | \gamma_d^2, \sigma_d^2 & \sim 
-      \mathcal{N} (0, \sigma_d^2 \gamma_d^2), \label{eq:betaprior} \\
-      \gamma_d^2 & \sim \mathcal{IG}(\theta_d, \lambda_d), \\
+      \(\x_i \tr \bm{\beta}_d, \sigma_d^2\), \\
+      \beta_{jd} | \gamma_{jd}^2, \sigma_d^2 & \sim \mathcal{N} 
+      \(0, \sigma_d^2 \gamma_{jd}^2 \), \label{eq:betaprior} \\
+      \gamma_{jd}^2 & \sim \mathcal{IG}
+      \( [(\mathbf{c}_{d}^j) \tr \bm{\alpha}]^{-1}, \lambda_{d} \), \\
       \sigma_d^2 & \sim 1/\sigma_d^3,
@@
   \end{subequations}
-  where $\mathcal{IG}(\theta, \lambda)$ denotes the inverse Gaussian 
-  distribution with mean $\theta$ and shape $\lambda$.
+  where $\mathcal{IG} ([(\mathbf{c}_{d}^j) \tr \bm{\alpha}]^{-1}, \lambda_{d})$ 
+  denotes the inverse Gaussian distribution with mean 
+  $[(\mathbf{c}_{d}^j) \tr \bm{\alpha}]^{-1}$ and shape $\lambda_d$.
+
+  The normal inverse Gaussian (NIG) prior model for $\beta_{jd}$ and 
+  $\gamma^2_{jd}$ in (\ref{eq:priormodel}) was introduced in 
+  \cite{barndorff-nielsen_hyperbolic_1978} and is, since 
+  \cite{barndorff-nielsen_normal_1997}, routinely applied in mathematical 
+  finance (see, e.g., \cite{kalemanova_normal_2007}). Supplementary Material
+  (SM) Section \ref{sm-sec:nigprior} contains more details on this prior.
+  In Figure \ref{fig:dens_igaussian_marginalbeta}, the NIG 
+  prior distribution, marginalised over $\gamma_{jd}^2$, is depicted together 
+  with three other common choices,
+  the Student's $t$ (see Section \ref{sec:inversegammamodel} for more details on 
+  this prior), normal (ridge) and Laplace (lasso) priors, 
+  all scaled to variance one. 
+  From this Figure we see that, depending on the choice
+  of hyperparameters, the NIG prior can vary from heavier tails than the lasso, 
+  to more Gaussian tails like the ridge, but behaves similarly to the 
+  Student's $t$ prior. Our argumentation to model the 
+  $\gamma^2_{jd}$ by an inverse Gaussian 
+  distribution, as has been suggested in \cite{fabrizi_specification_2016} and 
+  \cite{caron_sparse_2008}, is two-fold: (i) the NIG model is more flexible than
+  the standard ridge and lasso priors (as seen from Figure 
+  \ref{fig:dens_igaussian_marginalbeta}), and (ii) the NIG prior allows to model
+  the mean as a 
+  function of the external variables $\mathbf{c}_{d}^j$ more conveniently than 
+  the Student's $t$ prior, as explained in Section 
+  \ref{sec:empiricalbayes}. 
+<<dens_igaussian_marginalbeta, fig.cap="Several choices of marginal $\\beta_j$ prior, scaled to $\\V(\\beta_j)=1$.", out.width="60%", fig.asp=3/4>>=
+@
+
   \cite{moran_variance_2018} argue that in regression models, $\bm{\beta}_d$ 
   and $\sigma_d^2$ should be independent \textit{a priori}, i.e.
-  $\beta_{jd} | \gamma_d^2, \sigma_d^2 \sim \beta_{jd} | 
-  \gamma_d^2$. However, in our case
+  $\beta_{jd} | \gamma_{jd}^2, \sigma_d^2 \sim \beta_{jd} | 
+  \gamma_{jd}^2$. However, in our case
   that is debatable. The $D$ regressions need not be on the same scale, so in
@@
   $\sigma_d^2$ as in (\ref{eq:betaprior}), resulting in a 
-  scale-free prior with respect to the response.
-
-  The normal inverse Gaussian (NIG) prior model for $\beta_{jd}$ and 
-  $\gamma^2_d$ in (\ref{eq:priormodel}) was introduced in 
-  \cite{barndorff-nielsen_hyperbolic_1978} and is, since 
-  \cite{barndorff-nielsen_normal_1997}, routinely applied in mathematical 
-  finance (see, e.g., \cite{kalemanova_normal_2007}). See SM Section 
-  \ref{sm-sec:nigprior} for more details on this prior.
-  In Figure \ref{fig:dens_igaussian_marginalbeta}, the NIG 
-  prior distribution marginalised over $\gamma_d^2$ is depicted together with 
-  three other common prior choices,
-  Student's $t$ (see Section \ref{sec:inversegammamodel} for more details on 
-  this prior), ridge and lasso, all scaled to variance one. From this Figure 
-  we see that, depending on the choice
-  of hyperparameters, the NIG prior can vary from heavier tails than the lasso, 
-  to more Gaussian tails like the ridge, but behaves similarly to the 
-  Student's $t$ prior. Our argumentation to model the 
-  $\gamma^2_d$ by an inverse Gaussian 
-  distribution, as has been suggested in \cite{fabrizi_specification_2016} and 
-  \cite{caron_sparse_2008}, is two-fold: (i) the NIG model is more flexible than
-  the standard ridge and lasso priors (as seen from Figure 
-  \ref{fig:dens_igaussian_marginalbeta}), and (ii) the NIG prior allows to model
-  the mean $\theta_d$ as a 
-  function of the co-data $\mathbf{c}_d$ more conveniently than the Student's 
-  $t$ prior, as explained in Section 
-  \ref{sec:empiricalbayes}. 
-<<dens_igaussian_marginalbeta, fig.cap="Several choices of marginal $\\beta_j$ prior, scaled to $\\V(\\beta_j)=1$.", out.width="80%", fig.asp=2/3>>=
-library(GeneralizedHyperbolic)
-library(metRology)
-# all scaled such that variance is one
-sigma <- 1
-theta <- 1/sigma^2
-lambda <- c(0.1, 2)
-lambda1 <- sqrt(2)
-eta <- lambda/theta + 2
-col <- c(1:6)
-lty <- c(1:6)
-labels <- as.expression(c(bquote("NIG, "~lambda==.(lambda[1])~", "~
-                                   theta==.(theta)),
-                          bquote("NIG, "~lambda==.(lambda[2])~", "~
-                                   theta==.(theta)),
-                          bquote("Student's t, "~lambda==.(lambda[1])~", "~
-                                   eta==.(eta[1])),
-                          bquote("Student's t, "~lambda==.(lambda[2])~", "~
-                                   eta==.(eta[2])),
-                          "ridge", "lasso"))
-dprior <- function(x, lambda, theta, sigma) {
-  dnig(x, 0, sigma*sqrt(lambda), sqrt(lambda/(theta*sigma)), 0)
-}
-dlasso <- function(x, lambda1) {
-  0.5*lambda1*exp(-lambda1*abs(x))
-}
-
-ylim <- c(0, dt.scaled(0, eta[1]/2, 0, sqrt(lambda[1]*sigma^2/eta[1])))
-curve(dprior(x, lambda[1], theta, sigma), -3, 3, ylab=expression(p(beta[j])), 
-      xlab=expression(beta[j]), n=1000, ylim=ylim,
-      col=col[1], lty=lty[1])
-curve(dprior(x, lambda[2], theta, sigma), add=TRUE, n=1000, 
-      col=col[2], lty=lty[2])
-curve(dnorm(x, 0, 1), add=TRUE, n=1000, col=col[3], lty=lty[3])
-curve(dlasso(x, lambda1), add=TRUE, n=1000, col=col[4], lty=lty[4])
-curve(dt.scaled(x, eta[1]/2, 0, sqrt(lambda[1]*sigma^2/eta[1])), add=TRUE, 
-      n=1000, col=col[5], lty=lty[5])
-curve(dt.scaled(x, eta[2]/2, 0, sqrt(lambda[2]*sigma^2/eta[2])), add=TRUE, 
-      n=1000, col=col[6], lty=lty[6])
-legend("topright", legend=labels, col=col, lty=lty, 
-       title="Prior", seg.len=1)
-@
+  scale-free prior with respect to the response. Apart from the scaling,
+  in this setting, the conjugate prior model is preferred
+  over the non-conjugate model, because the variational Bayes approximation 
+  (see Section \ref{sec:estimation}) assumes independent $\sigma_d^2$ and 
+  $\bm{\beta}_d$ in the posterior, which is actually true for 
+  the conjugate model. Therefore, the approximate expectation used in the 
+  empirical Bayes step (see Section \ref{sec:estimation}) is more accurate.
 
@@
   \cite[]{jeffreys_invariant_1946}. In the derivation of our Jeffrey's prior
-  for the error variance, we jointly consider an unknown mean and variance.
+  for the error variance, we jointly consider an unknown mean and variance. This
+  joint consideration results in the somewhat unorthodox $1/\sigma^3$ Jeffrey's
+  prior.
 
-  - Jeffrey's prior for variance $1/\sigma^{3/2}$ (a priori independent data 
-	mean and variance) or $1/\sigma^3$ (joint prior, but only mean and variance, 
-	not variance and betas).
-
-	\section{Estimation}
+	\section{Estimation}\label{sec:estimation}
 	\subsection{Variational Bayes}\label{sec:variationalbayes}
@@
 	posterior 
-	density to factorise as: $p(\bbeta_d, \sigma_d^2, \gamma_d^2) \approx 
-	Q_d(\cdot) = q(\bbeta_d) \cdot q(\sigma_d^2) \cdot q(\gamma_d^2)$. For
+	density to factorise as: $p(\bbeta_d, \sigma_d^2, \gamma_{1d}^2, \dots, 
+	\gamma_{pd}^2) \approx 
+	Q_d(\cdot) = q(\bbeta_d) \cdot q(\sigma_d^2) \cdot \prod_{j=1}^p 
+	q(\gamma_{jd}^2)$. For
 	notational convenience, we slightly abuse notation and let $q(\cdot)$ denote
@@
     (\bm{\mu}_d, \bm{\Sigma}_d), \\
-    q(\gamma^2_d) & \overset{D}{=} \mathcal{GIG}
-    \(-\frac{p+1}{2}, \lambda_d \theta_d^{-2}, \delta_d\), \\
+    q(\gamma_{jd}^2) & \overset{D}{=} \mathcal{GIG}
+    \(-1, \lambda_{d}[(\mathbf{c}_{d}^j) \tr \bm{\alpha}]^2, \delta_{jd}\), \\
     q(\sigma^2_d) & \overset{D}{=} \Gamma^{-1} \(\frac{n + p + 1}{2}, \zeta_d\).
@@
     \bm{\Sigma}_d^{(h+1)} & = (a_d^{(h)})^{-1} 
-    (\X \tr \X + b_d^{(h)} \I)^{-1}, \\
-    \bm{\mu}_d^{(h+1)} & = (\X \tr \X + b_d^{(h)} \I)^{-1} \X \tr \y_d, \\
-    \delta_d^{(h+1)} & = a_d^{(h)} \[(\bmu^{(h+1)}_d) \tr \bmu^{(h+1)}_d +
-    \trace (\bSigma_d^{(h+1)})\] + \lambda_d, \\ 
+    \[\X \tr \X + \diag(b_{jd}^{(h)})\]^{-1}, \\
+    \bm{\mu}_d^{(h+1)} & = \[\X \tr \X + \diag(b_{jd}^{(h)})\]^{-1} 
+    \X \tr \y_d, \\
+    \delta_{jd}^{(h+1)} & = a_d^{(h)} \[(\bmu^{(h+1)}_{jd})^2 + 
+    (\bSigma_d)_{jj}\] + \lambda_{d}, \\ 
     \zeta_d^{(h+1)} & = \frac{1}{2} \bigg[\mathbf{y}_d \tr \mathbf{y}_d -
@@
     & \,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\, + 
-    b_d^{(h+1)} \trace ( \bm{\Sigma}_d^{(h+1)}) + 
-    b_d^{(h+1)} (\bm{\mu}_d^{(h+1)}) \tr \bm{\mu}_d^{(h+1)}\bigg],
+    \trace \[ \diag (b_{jd}^{(h+1)}) \bm{\Sigma}_d^{(h+1)}\] + 
+    (\bm{\mu}_d^{(h+1)}) \tr \diag (b_{jd}^{(h+1)}) 
+    \bm{\mu}_d^{(h+1)}\bigg],
   \end{align*}
@@
     \nonumber\\
-    b_d^{(h)} & = \E_{Q^{(h)}}(\gamma_d^{-2}) = \sqrt{\frac{\lambda_d}
-    {\theta_d^2 \delta_d^{(h)}}} \frac{K_{\frac{p - 1}{2}} 
-    \( \sqrt{\lambda_d \delta_d^{(h)}}/\theta_d \)}
-    {K_{\frac{p+1}{2}} \( \sqrt{\lambda_d \delta_d^{(h)}}/\theta_d \)} + 
-    \frac{p+1}{\delta_d^{(h)}}, \label{eq:ratiomodifiedbessel}
+    b_{jd}^{(h)} & = \E_{Q^{(h)}}(\gamma_{jd}^{-2}) = 
+    \sqrt{\frac{\delta_{jd}^{(h)}}
+    {\lambda_{d}[(\mathbf{c}_{d}^j) \tr \bm{\alpha}]^2}}
+    \frac{K_0(\sqrt{\delta_{jd}^{(h)}\lambda_{d}}
+    [(\mathbf{c}_{d}^j) \tr \bm{\alpha}])}
+    {K_1(\sqrt{\delta_{jd}^{(h)}\lambda_{d}}
+    [(\mathbf{c}_{d}^j) \tr \bm{\alpha}])} + 
+    \frac{2}{\delta_{jd}^{(h)}}, \label{eq:ratiomodifiedbessel}
   \end{align}
   where $K_{\nu}(x)$ denotes the modified Bessel function of the second kind.
-  A method for fast and numerically stable calculation of ratios of such 
-  functions as in (\ref{eq:ratiomodifiedbessel}) is given in SM Section 
+  A method for fast and numerically stable calculation of ratios of
+  modified Bessel functions of the second kind, as in 
+  (\ref{eq:ratiomodifiedbessel}), is given in SM Section 
   \ref{sm-sec:ratiosmodifiedbessels}.
@@
   \subsection{Empirical Bayes}\label{sec:empiricalbayes}
-  A full Bayesian model requires the specification of hyper parameters 
-  $\bm{\theta}=\begin{bmatrix} \theta_1 & \cdots & \theta_D \end{bmatrix} \tr$ 
-  and $\begin{bmatrix} \lambda_1 & \cdots & \lambda_D \end{bmatrix} \tr$. 
-  These are abstract and hard to interpret parameters 
+  To make use of the external data, we parametrise the prior mean of the
+  $\gamma_{jd}^2$ as $[(\mathbf{c}_{d}^j) \tr \bm{\alpha}]^{-1}$. 
+  This allows us to
+  include continuous external variables $\mathbf{c}_{d}^j$ into the model. 
+  Additionally, it reduces the number of hyperparameters from $pD$ to 
+  $|\bm{\alpha}| + D$. A full Bayesian model
+  then requires the specification of the hyperparameters 
+  $\bm{\alpha}$ and $\blambda = \begin{bmatrix} \lambda_1 & \cdots & \lambda_D 
+  \end{bmatrix} \tr$. These are abstract and hard to interpret parameters 
   for which we generally lack expert knowledge. They do, however, have a 
   significant influence on the shape of the posterior distribution. We 
-  therefore propose to estimate these hyper parameters by empirical Bayes. 
+  therefore propose to estimate these hyperparameters by empirical Bayes. 
   Simply put, empirical Bayes fits the prior model to the data and is, as such, 
-  objective (up to model specification).
+  objective Bayes (up to model specification). In our case, this results in an   
+  objective and data-driven inclusion of the external data.
 
@@
 	algorithm:
-	$$
-	\bm{\theta}^{(l+1)},\bm{\lambda}^{(l+1)} = \underset{\bm{\theta},\bm{\lambda}}
-	{\argmax}\E_{\cdot | \Y} 
-	[\log p(\Y, \B, \bgamma^2, \bsigma^2) | \bm{\theta}^{(l)},\bm{\lambda}^{(l)}],
-	$$
+	\begin{align*}
+	  \bm{\alpha}^{(l+1)},\bm{\lambda}^{(l+1)} & = \underset{\bm{\alpha},
+	  \bm{\lambda}}{\argmax}\E_{\cdot | \Y} 
+	  [\log p(\Y, \B, \bm{\Gamma}^2, \bsigma^2) | \bm{\alpha}^{(l)},
+	  \bm{\lambda}^{(l)}] \\
+	  & = \underset{\bm{\alpha},\bm{\lambda}}{\argmax} 
+	  \E_{\cdot | \Y} [\log \pi (\bm{\Gamma}^2) | \bm{\alpha}^{(l)},
+	  \bm{\lambda}^{(l)}],
+	\end{align*}
 	where $\Y = \begin{bmatrix} \y_1 & \cdots & \y_D \end{bmatrix}$,
 	$\B = \begin{bmatrix} \bbeta_1 & \cdots & \bbeta_D \end{bmatrix}$, 
-	$\bgamma^2 = \begin{bmatrix} \gamma^2_1 & \cdots & \gamma^2_D 
-	\end{bmatrix}\tr$, $\bsigma^2 = \begin{bmatrix} \sigma^2_1 & \cdots & 
+	$\bsigma^2 = \begin{bmatrix} \sigma^2_1 & \cdots & 
 	\sigma^2_D \end{bmatrix} \tr$,
-	and the expectation is with respect to the posterior. In our case, this 
+	and $\bm{\Gamma}^2 = \begin{bmatrix} \bgamma^2_1 & \cdots & \bgamma^2_D 
+	\end{bmatrix}\tr$, with 
+	$\bgamma^2_d = \begin{bmatrix} \gamma^2_{1d} & \cdots & \gamma^2_{pd} 
+	\end{bmatrix}\tr$,
+	and the expectation is with respect to the joint posterior. In our case, this 
 	posterior is not available in closed form, which renders the expectation 
@@
 	approximation developed in Section \ref{sec:variationalbayes}:
+	$$
+	\bm{\alpha}^{(l+1)},\bm{\lambda}^{(l+1)} = 
+	\underset{\bm{\alpha},\bm{\lambda}}{\argmax}\E_{Q^{(l)}} 
+	[\log \pi(\bm{\Gamma}^2)],
+	$$
+	where now the expectation is with respect to the converged variational 
+	posterior $Q^{(l)}=\prod_{d=1}^D Q_d^{(l)}$. The resulting estimating 
+	equations are:
   \begin{align*}
-	  \bm{\theta}^{(l+1)},\bm{\lambda}^{(l+1)} & = 
-	  \underset{\bm{\theta},\bm{\lambda}}{\argmax}\E_{Q^{(l)}} 
-	  [\log p(\Y, \B, \bgamma^2, \bsigma^2)] \\
-	  & = \underset{\bm{\theta},\bm{\lambda}}{\argmax} \E_{Q^{(l)}} 
-	  [\log \pi (\B | \bgamma^2, \bsigma^2)] + 
-	  \E_{Q^{(l)}} [\log \pi (\bgamma^2)].
+    \bm{\alpha} & = \[ \mathbf{C} \tr \bLambda \diag 
+    (e_{jd}^{(l)}) \mathbf{C} \]^{-1} \mathbf{C} \tr \bLambda
+    \mathbf{1}_{pD \times 1}, \\
+    \lambda_{d} & = p \left\{ \sum_{j=1}^p b_{jd}^{(l)} + 
+    \sum_{j=1}^p e_{jd}^{(l)} [(\mathbf{c}_{d}^j) \tr \bm{\alpha}]^2 - 
+    2 \sum_{j=1}^p (\mathbf{c}_{d}^j) \tr \bm{\alpha} \right\}^{-1},
   \end{align*}
-	where now the expectation is with respect to the converged variational 
-	posterior $Q^{(l)}=\prod_{d=1}^D Q_d^{(l)}$.
-
-	To make use of the outcome-specific codata, we parametrise 
-	$\theta_d = (\mathbf{c}_d \tr \bm{\alpha})^{-1}$ and estimate $\blambda$ and
-	$\bm{\alpha}$. This allows us to include continuous co-data on the outcomes
-	in an objective, data-driven manner. Additionally, it reduces the number of
-	hyperparameters even further. The estimating equations are:
-  \begin{align*}
-    \bm{\alpha} & = \[ \mathbf{C} \tr \diag 
-    (\lambda_d e_d^{(l)}) \mathbf{C} \]^{-1} \mathbf{C} \tr \diag 
-    (\lambda_d) \mathbf{1}_{D \times 1}, \\
-    \lambda_d & = \[ b_d^{(l)} + e_d^{(l)} (\mathbf{c}_d \tr \bm{\alpha})^2 - 
-    2\mathbf{c}_d \tr \bm{\alpha} \]^{-1},
-  \end{align*}
-  where $e_d^{(l)} = \E_{Q^{(l)}}(\gamma_d^2) = [b_d^{(l)} - 
-  (p + 1)/\delta_d^{(l)})] \cdot \delta_d^{(l)} 
-  (\theta_d^{(h)})^2/\lambda_d^{(l)}$. Solving the equations is done by 
-  iteratively reweighted least squares of responses $(e_d^{(l)})^{-1}$ on 
-  predictors $\mathbf{c}_d$ with weights $\lambda_d e_d^{(l)}$.
+  where $\bLambda=\diag(\lambda_d) \otimes \I_p$
+  and $e_{jd}^{(l)} = \E_{Q^{(l)}}(\gamma_{jd}^2) = (b_{jd}^{(l)} - 
+  2/\delta_{jd}^{(l)}) \cdot \delta_{jd}^{(l)}/[\lambda_{d}^{(l)}
+  \left\{(\mathbf{c}_{d}^j) \tr \bm{\alpha}^{(l)}]^2\right\}$. 
+  Solving the equations is done by 
+  iteratively reweighted least squares.
 
@@
   variances \cite[]{rue_approximate_2009,consonni_mean-field_2007,
-  bishop_pattern_2006,wang_inadequacy_2005}. If posterior variances are needed, 
+  bishop_pattern_2006,wang_inadequacy_2005}. If posterior variances are 
+  required, 
   we suggest generating samples from the posterior after the empirical Bayes 
@@
 
-  \section{General model}
-  In addition to the outcome co-data $\mathbf{c}_d$, we may have feature co-data
-  $\mathbf{z}_j$ available. Examples of this in omics settings are pathway 
-  memberships or previously obtained $p$-values for the features. In
-  that case, we extend model (\ref{eq:priormodel}) to:
-  \begin{subequations}\label{eq:generalmodel}
-    \begin{align}
-      y_{id} | \bm{\beta}_d, \sigma_d^2 & \sim \mathcal{N} 
-      (\x_i \tr \bm{\beta}_d, \sigma_d^2), \nonumber \\
-      \beta_{jd} | \tau_j^2, \gamma_d^2, \sigma_d^2 & \sim \mathcal{N} 
-      (0, \sigma_d^2 \gamma_d^2 \tau_j^2), \\
-      \gamma_d^2 & \sim \mathcal{IG}\(\theta_d^c, \lambda^c_d\), \\
-      \tau_j^2 & \sim \mathcal{IG}\(\theta_j^z, \lambda^z_j\), \\
-      \sigma_d^2 & \sim 1/\sigma_d^3, \nonumber
-    \end{align}
-  \end{subequations}
-  where we now have an extra variance component $\tau_j$ that is 
-  \textit{a priori} inverse Gaussian distributed with mean dependent on the 
-  feature co-data $\mathbf{z}_j$. We will refer to this model as the normal
-  inverse Gaussian-inverse Gaussian (NIGIG) model. See SM Section 
-  \ref{sm-sec:nigigprior} for more details on this prior.
+  Note that so far, the design matrix $\X$ is assumed to be shared across the
+  equations. This is not a requirement. That is, we may have D different design 
+  matrices $\X^d$ that need not have the same number of 
+  columns, i.e., we have $p_d$ instead of $p$. This becomes useful in, e.g. 
+  eQTLs detection, where we regress genes on a gene-specific selection
+  of SNPs, often based on SNP-gene distance.
 
-  The formulation as in (\ref{eq:generalmodel}) results in an additional 
-  variational distribution for the $\tau_j$:
-  $$
-  q(\tau_j^2) \overset{D}{=} \mathcal{GIG}\(-\frac{D+1}{2}, \lambda^z_j 
-  (\theta_j^z)^{-2}, \delta_j^z\).
-  $$
-  The variational parameters are now updated by:
-  \begin{align*}
-    \bm{\Sigma}_d^{(h+1)} & = (a_d^{(h)})^{-1} 
-    \[\X \tr \X + b_d^{(h)} \diag (c_j^{(h)})\]^{-1}, \\
-    \bm{\mu}_d^{(h+1)} & = \[\X \tr \X + b_d^{(h)} \diag(c_j^{(h)})\]^{-1} 
-    \X \tr \y_d, \\
-    (\delta^c_d)^{(h+1)} & = a_d^{(h)} \left\{(\bmu^{(h+1)}_d) \tr 
-    \diag (c_j^{(h)}) \bmu^{(h+1)}_d + \trace \[ \diag (c_j^{(h)}) 
-    \bSigma_d^{(h+1)}\] \right\} + \lambda^c_d, \\ 
-    (\delta^z_j)^{(h+1)} & = \sum_{d=1}^D a_d^{(h)} b_d^{(h)} 
-    \[(\bmu^{(h+1)}_d) \tr \bmu^{(h+1)}_d + \trace (\bSigma_d^{(h+1)}) \] + 
-    \lambda^z_j, \\ 
-    \zeta_d^{(h+1)} & = \frac{1}{2} \bigg[\mathbf{y}_d \tr \mathbf{y}_d -
-    2 \mathbf{y}_d \tr \X \bm{\mu}_d^{(h+1)} + 
-    \trace ( \X \tr \X \bm{\Sigma}_d^{(h+1)}) + 
-    (\bm{\mu}_d^{(h+1)}) \tr \X \tr \X \bm{\mu}_d^{(h+1)} \\
-    & \,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\, + 
-    b_d^{(h+1)} \trace \[ \diag (c_j^{(h)}) \bm{\Sigma}_d^{(h+1)}\] + 
-    b_d^{(h+1)} (\bm{\mu}_d^{(h+1)}) \tr \diag (c_j^{(h)}) 
-    \bm{\mu}_d^{(h+1)}\bigg],
-  \end{align*}
-  until convergence, where
-  $$
-  c_j^{(h)} = \E_{Q^{(h)}}(\tau_j^{-2}) = \sqrt{\frac{\lambda^z_j}
-  {(\theta_j^z)^2 (\delta^z_j)^{(h)}}} 
-  \frac{K_{\frac{D - 1}{2}} 
-  \( \sqrt{\lambda^z_j (\delta^z_j)^{(h)}}/\theta_j^z \)}
-  {K_{\frac{D+1}{2}} \( \sqrt{\lambda^z_j (\delta_j^z)^{(h)}}/\theta_j^z \)} +
-  \frac{D+1}{(\delta_j^z)^{(h)}}
-  $$
-  If we set $\theta_d^c=(\mathbf{c}_d \tr \bm{\alpha}^c)^{-1}$ and
-  $\theta_j^z=(\mathbf{z}_j \tr \bm{\alpha}^z)^{-1}$, we have an additional
-  empirical Bayes step that consists of an iteratively 
-  reweighted least squares regression of responses $(f_j^{(l)})^{-1}$ on
-  predictors $\mathbf{z}_j$ with weights $\lambda_j^z f_j^{(l)}$, where 
-  \begin{align*}
-  f_j^{(l)} & = \E_{Q^{(h)}}(\tau_j) = [c_j^{(l)} - 
-  (D + 1)/(\delta_j^z)^{(l)}] \cdot (\delta_j^z)^{(l)} [\mathbf{z}_j \tr 
-  (\bm{\alpha}^z)^{(l)} ]^{-2} / (\lambda_j^z)^{(l)}, \\
-  \lambda_j^z & = \[ c_j^{(l)} + f_j^{(l)} 
-  (\mathbf{z}_j \tr \bm{\alpha}^z)^2 - 
-    2\mathbf{z}_j \tr \bm{\alpha}^z \]^{-1}.
-  \end{align*}
+  \section{Simulations}\label{sec:simulations}
+  We conducted a simulation study consisting of three parts investigating (i) 
+  the NIG prior performance, (ii) the variational Bayes approximation, and (iii)
+  the empirical Bayes estimation.
 
-  \section{General model 2}
-  We have co-data $\mathbf{c}_{jd}$ available that describes the relationship
-  between the covariate $j$ and outcome $d$. An example from eQTL is the 
-  distance between SNP and gene. Model (\ref{eq:priormodel}) is then rewritten
-  as:
-  \begin{subequations}\label{eq:generalmodel2}
-    \begin{align*}
-      y_{id} | \bm{\beta}_d, \sigma_d^2 & \sim \mathcal{N} 
-      \(\x_i \tr \bm{\beta}_d, \sigma_d^2\), \nonumber \\
-      \beta_{jd} | \gamma_{jd}^2, \sigma_d^2 & \sim \mathcal{N} 
-      \(0, \sigma_d^2 \gamma_{jd}^2 \), \\
-      \gamma_{jd}^2 & \sim \mathcal{IG}
-      \( (\mathbf{c}_{jd} \tr \bm{\alpha})^{-1}, \lambda_{jd} \), \\
-      \sigma_d^2 & \sim 1/\sigma_d^3, \nonumber
-    \end{align*}
-  \end{subequations}
-  where now the variance components $\gamma_{jd}^2$ are both outcome and 
-  covariate specific and modelled as a function of the co-data. 
-  The formulation as in (\ref{eq:generalmodel2}) results in the following
-  variational distribution for the $\gamma^2_{jd}$:
-  $$
-  q(\gamma_{jd}^2) \overset{D}{=} \mathcal{GIG}
-  \(-1, \lambda_{jd}(\mathbf{c}_{jd} \tr \bm{\alpha})^2, \delta_{jd}\).
-  $$
-  The variational parameters are now updated by:
-  \begin{align*}
-    \bm{\Sigma}_d^{(h+1)} & = (a_d^{(h)})^{-1} 
-    \[\X \tr \X + \diag(b_{jd}^{(h)})\]^{-1}, \\
-    \bm{\mu}_d^{(h+1)} & = \[\X \tr \X + \diag(b_{jd}^{(h)})\]^{-1} 
-    \X \tr \y_d, \\
-    \delta_{jd}^{(h+1)} & = a_d^{(h)} \[(\bmu^{(h+1)}_{jd})^2 + 
-    (\bSigma_d)_{jj}\] + \lambda_{jd}, \\ 
-    \zeta_d^{(h+1)} & = \frac{1}{2} \bigg[\mathbf{y}_d \tr \mathbf{y}_d -
-    2 \mathbf{y}_d \tr \X \bm{\mu}_d^{(h+1)} + 
-    \trace ( \X \tr \X \bm{\Sigma}_d^{(h+1)}) + 
-    (\bm{\mu}_d^{(h+1)}) \tr \X \tr \X \bm{\mu}_d^{(h+1)} \\
-    & \,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\, + 
-    \trace \[ \diag (b_{jd}^{(h)}) \bm{\Sigma}_d^{(h+1)}\] + 
-    (\bm{\mu}_d^{(h+1)}) \tr \diag (b_{jd}^{(h)}) 
-    \bm{\mu}_d^{(h+1)}\bigg],
-  \end{align*}
-  until convergence, where $a_d^{(h)}$ is as before and 
-  $$
-  b_d^{(h)}=\sqrt{\frac{\delta_{jd}}
-  {\lambda_{jd}(\mathbf{c}_{jd} \bm{\alpha})^2}}
-  \frac{K_0(\sqrt{\delta_{jd}\lambda_{jd}(\mathbf{c}_{jd} \bm{\alpha})^2})}
-  {K_1(\sqrt{\delta_{jd}\lambda_{jd}(\mathbf{c}_{jd} \bm{\alpha})^2})} + 
-  \frac{2}{\delta_{jd}}.
-  $$ 
-  The empirical Bayes step that consists of an iteratively 
-  reweighted least squares regression with steps:
-  \begin{align*}
-    \bm{\alpha} & = \[ \mathbf{C} \tr \diag 
-    (\lambda_{jd} e_{jd}^{(l)}) \mathbf{C} \]^{-1} \mathbf{C} \tr \diag 
-    (\lambda_{jd}) \mathbf{1}_{p \cdot D \times 1}, \\
-    \lambda_{jd} & = \[ b_{jd}^{(l)} + e_{jd}^{(l)} 
-    (\mathbf{c}_{jd} \tr \bm{\alpha})^2 - 
-    2\mathbf{c}_{jd} \tr \bm{\alpha} \]^{-1},
-  \end{align*}
-  where $e_{jd}^{(l)} = \E_{Q^{(l)}}(\gamma_d^2) = [b_{jd}^{(l)} - 
-  2/\delta_{jd}^{(l)})] \cdot \delta_{jd}^{(l)} 
-  /[\lambda_d^{(l)} (\mathbf{c}_{jd} \tr \bm{\alpha})^2]$.
-
-  Note that so far, the design matrix $\X$ is assumed to be shared across the
-  equations. This is not a requirement, so that we may have $\X^d$, for 
-  $d=1, \dots, D$ different design matrices. As a consequence, the index $j$
-  may run over differently sized sequences.
-
-  \section{Simulations}
-  \subsection{Comparison of NIG, NIGIG and Student's \texorpdfstring{$t$}{t} 
-  priors}
+  \subsection{NIG prior performance}
   \label{sec:inversegammamodel}
+  The aim of this part is to investigate the 
+  performance of the NIG prior model. Since we may do so in the univariate 
+  setting, so we drop all $d$ indices in the following. 
   A common default prior for latent variance components is the inverse Gamma 
@@
   $$
-  \gamma_d^2 \sim \Gamma^{-1}(\eta_d/2,\lambda_d/2),
+  \gamma_j^2 \sim \Gamma^{-1}(\eta/2,\xi/2),
   $$
-  with shape $\eta_d/2$ and scale $\lambda_d/2$, both rescaled to match our 
+  with shape $\eta/2$ and scale $\xi/2$, both rescaled to match our 
   inverse Gaussian prior formulation. The resulting marginal prior for the model
   parameters
-  (conditional on the error variance $\sigma_d^2$) is then a scaled Student's 
+  (conditional on the error variance $\sigma^2$) is then a scaled Student's 
   $t$-distribution:
   $$
-  \beta_j | \sigma_d^2 \sim \text{Student's } 
-  t\(\eta_d,\sigma_d \sqrt{\lambda_d/\eta_d}\),
+  \beta_j | \sigma^2 \sim \text{Student's } 
+  t\(\eta,\sigma \sqrt{\xi/\eta}\),
   $$
-  with degrees of freedom $\eta_d$ and scale $\sigma_d \sqrt{\lambda_d/\eta_d}$.
+  with degrees of freedom $\eta$ and scale $\sigma \sqrt{\xi/\eta}$.
   See SM Section \ref{sm-sec:inversegamma} for more details on this prior. 
   We compare the 
-  posteriors of the proposed NIG and NIGIG priors to the scaled Student's 
+  posteriors of the proposed NIG to the scaled Student's 
   $t$-distribution prior in a simulated setting.
 
-  We compare the three posterior distributions to the true posterior for a 
+  We compare the posterior distributions to the true posterior for a 
   number of measures. Divergence between the distributions
@@
   criterion \cite[]{baringhaus_new_2004}, where smaller values of the criterion
-  indicate more similar distributions. We also assess 95\% coverage of the 
+  indicate a better fit. We also assess 95\% coverage of the 
   posterior credible intervals. The accuracy of the 
   point estimates of $\beta_j$ in the form of posterior means is assessed by
-  the mean squared distance (MSD) and correlation between the point estimates 
+  the mean squared error (MSE) and correlation between the point estimates 
   and the true $\beta_j$.
 
-  The aim of this simulation study is to investigate the performance of the NIG 
-  prior model, so we set $D=1$ and drop the $d$ indices in the following. 
   We studied two settings: one dense and one sparse. In the dense setting, 
@@
   We choose the hyperparameters such that we get all combinations of marginal
-  prior variances and kurtoses $\V(\beta_{jd} | \sigma_d^2) \in 
-  \{ 1/2, 1, 2 \}$ and $\mathcal{K}(\beta_{jd} | \sigma_d^2) \in 
+  prior variances and kurtoses $\V(\beta_{j} | \sigma^2) \in 
+  \{ 1/2, 1, 2 \}$ and $\mathcal{K}(\beta_{j} | \sigma^2) \in 
   \{ 4, 10 \}$. This leads to the settings in Table \ref{tab:hyper_set1}. For 
@@
   posteriors using an MCMC sampler implemented in \texttt{rstan}. We repeated
-  the simulations 100 times and report 
-  results in Figures \ref{fig:boxplots_igaussian_res4_post1} and 
-  \ref{fig:boxplots_igaussian_res4_post2}. 
+  the simulation 100 times and report 
+  results in Figures \ref{fig:boxplots_igaussian_res4.1_post1} and 
+  \ref{fig:boxplots_igaussian_res4.1_post2}. 
 <<hyper_set1>>= 
 set <- read.table("results/simulations_igaussian_set4.csv")
-kableExtra::kable_styling(
-  knitr::kable(set[, c("lambdac", "thetac", "lambdaz", "thetaz",
-                       "eta", "lambda", "vars", "kurts")], 
-               col.names=c("$\\lambda^c$", "$\\theta^c$", "$\\lambda^z$",
-                           "$\\theta^z$", "$\\eta$", "$\\lambda$", 
-                           "$\\V(\\beta_{jd} | \\sigma_d^2)$",
-                           "$\\mathcal{K}(\\beta_{jd} | \\sigma_d^2)$"), 
+kableExtra::add_header_above(kableExtra::kable_styling(
+  knitr::kable(set[, c("ctalphainv", "lambda", "eta", "xi", "vars", "kurts")],
+               col.names=c("$(\\mathbf{c} \\tr \\bm{\\alpha})^{-1}$",
+                           "$\\lambda$", "$\\eta$", "$\\xi$",
+                           "$\\V(\\beta_{j} | \\sigma^2)$",
+                           "$\\mathcal{K}(\\beta_{j} | \\sigma^2)$"),
                caption="Hyperparameter settings",
                "latex", booktabs=TRUE, escape=FALSE, digits=2),
-  latex_options=c("HOLD_position"))
+  latex_options=c("HOLD_position")),
+  header=c(" "=1, "NIG"=2, "Student's t"=2, "Prior"=2))
 @
 
-  <<boxplots_igaussian_res4_post1, fig.cap="Comparison of the NIG, NIGIG, and Student's t priors in (a)-(d) simulation setting 1, (e)-(h) setting 2, and (i)-(l) setting 3, through Cramér-von Mises criterion between the model and true posteriors, correlation and MSD between the posterior means and true parameters, and 95\\% credible interval coverage of the true values.", out.width="100%", fig.asp=1>>=
+<<boxplots_igaussian_res4.1_post1, fig.cap="Comparison of the NIG and Student's t priors in the dense setting in (a)-(d) simulation setting 1, (e)-(h) setting 2, and (i)-(l) setting 3, through Cramér-von Mises criterion between the model and true posteriors, correlation and MSD between the posterior means and true parameters, and 95\\% credible interval coverage of the true values.", out.width="100%", fig.asp=1>>=
 @
 
-<<boxplots_igaussian_res4_post2, fig.cap="Comparison of the NIG, NIGIG, and Student's t priors in (a)-(d) simulation setting 4, (e)-(h) setting 5, and (i)-(l) setting 6, through Cramér-von Mises criterion between the model and true posteriors, correlation and MSD between the posterior means and true parameters, and 95\\% credible interval coverage of the true values.", out.width="100%", fig.asp=1>>=
+<<boxplots_igaussian_res4.1_post2, fig.cap="Comparison of the NIG and Student's t priors in the dense setting in (a)-(d) simulation setting 4, (e)-(h) setting 5, and (i)-(l) setting 6, through Cramér-von Mises criterion between the model and true posteriors, correlation and MSD between the posterior means and true parameters, and 95\\% credible interval coverage of the true values.", out.width="100%", fig.asp=1>>=
 @
@@
 
-  In the sparse setting, the observed data generating process is the same. The
-  model parameters are are simulated from a spike-and-slab prior with a 
-  Gaussian slab:
+  In the sparse setting, the observed data generating process is the same as in
+  the dense setting. The
+  model parameters are simulated from an (absolutely continuous) spike-and-slab 
+  prior with both Gaussian spike and slab:
   \begin{align*}
-    w & \sim \text{Beta}(\nu, \kappa), \\
-    z_j | w & \sim \text{Binom}(w), \\
-    \beta_j | z_j & \sim z_j \mathcal{N} \( 0, 1\) +
-    (1 - z_j)\delta_0,
+    w & \sim \text{Beta}(\phi, \chi), \\
+    z_j | w & \sim \text{Bernoulli}(w), \\
+    \beta_j | z_j & \sim z_j \mathcal{N} \( 0, 1 + 0.99 \frac{\chi}{\phi}\) +
+    (1 - z_j)\mathcal{N} \( 0, 0.01\).
   \end{align*}
-  where $\delta_0$ denotes the Dirac mass at zero. The slab is chosen
+  The slab is chosen
   such that the we have marginal prior variance $\V(\beta_j)=1$. The sparsity 
-  level is determined by the proportion of zero $\beta_j$, 
-  $s=1 - \nu/(\nu + \kappa)$. Note that the resulting marginal prior for
+  level is determined by the proportion of zero $\beta_j$: 
+  $s=1 - \phi/(\phi + \chi)$. Note that the resulting marginal prior for
   $\beta_j$ is:
   $$
-  \beta_j \sim (s - 1)\mathcal{N} \( 0, 1\) + s\delta_0.
+  \beta_j \sim (1 - s)\mathcal{N} \( 0, \frac{1 - 0.01 s}{1 - s}\) + 
+  s \mathcal{N} \( 0, 0.01\).
   $$
@@
   hyperparameters for estimation are set the same as in the dense setting, 
-  resulting in a total of 12 different simulation settings.
+  resulting in a total of 12 different simulation settings 
+  (6 per sparsity setting).
   To mitigate the effect of random variation, we
   repeated the simulations 100 times and report the results in Figures 
-  [! ref to figure here]
+  \ref{fig:boxplots_igaussian_res5_post1} and 
+  \ref{fig:boxplots_igaussian_res5_post2}.
+
+<<boxplots_igaussian_res5_post1, fig.cap="Comparison of the NIG and Student's t priors in the sparse setting in (a)-(d) simulation setting 1, (e)-(h) setting 2, and (i)-(l) setting 3, through Cramér-von Mises criterion between the model and true posteriors, correlation and MSD between the posterior means and true parameters, and 95\\% credible interval coverage of the true values.", out.width="100%", fig.asp=1>>=
+@
+
+<<boxplots_igaussian_res5_post2, fig.cap="Comparison of the NIG and Student's t priors in the sparse setting in (a)-(d) simulation setting 4, (e)-(h) setting 5, and (i)-(l) setting 6, through Cramér-von Mises criterion between the model and true posteriors, correlation and MSD between the posterior means and true parameters, and 95\\% credible interval coverage of the true values.", out.width="100%", fig.asp=1>>=
+@
+
+  [! describe results here]
 
   \subsection{Variational Bayes approximation}
-  To investigate the accuracy of our variational Bayes approximation we compare 
-  the resulting posterior to an MCMC sample from the posterior obtained with 
-  \texttt{rstan}. Since the MCMC
-  sample represents the exact posterior (up to Monte Carlo error), we use it as 
-  a proxy
-  for the true posterior. For completeness, we include the comparison to the 
+  To investigate the accuracy of the variational Bayes approximation we compare 
+  the variational Bayes posterior to 1000 (after 1000 burn-in) MCMC samples from
+  the posterior obtained with \texttt{rstan}. Since the MCMC
+  samples represent the exact posterior (up to Monte Carlo error), we use it as 
+  a proxy for the true posterior. We include the comparison of our variational
+  Bayes approximation to the approximate posterior from the
   automatic differentiation variational inference (ADVI)
@@
   \texttt{rstan} offers an ADVI algorithm that assumes a fullrank posterior
-  covariance, but this algorithm failed often resulted in errors and
-  convergence problems, so we do not include it in the comparisons.
+  covariance, but this algorithm often failed to converge or resulted in errors, 
+  so we do not include it in the comparisons. We measure divergence between the 
+  MCMC and variational posteriors with the Cram{\'e}r-von Mises criterion.
 
-  To assess the performance of the variational approximations we evaluate 
-  several measures. We measure divergence between the MCMC and variational 
-  posterior with the Cram{\'e}r-von Mises criterion. We compute the 
-  correlation and MSD between the MCMC and variational posterior variances. 
-  \texttt{rstan} allows the user 
+  In addition to ADVI, \texttt{rstan} allows the user
   to maximise the posterior to obtain the maximum \textit{a posteriori} (MAP)
-  estimates of the model parameters. We include these MAP estimates together
+  estimates of the model parameters. We compare these MAP estimates together
   with the ridge estimates obtained from the \texttt{glmnet} \texttt{R} package
-  \cite[]{friedman_regularization_2010} in the comparison of the posterior mean
-  estimates.
+  \cite[]{friedman_regularization_2010}, and the ADVI posterior means,
+  to the MAP estimates obtained from our model through the correlation with
+  and the mean squared distance (MSD) from the MCMC posterior means.
 
   The interest of this simulation lies in the variational Bayes approximation, 
-  so we restrict ourselves to the univariate setting (i.e., $D=1$). The
-  simulation setup is the same as in Section \ref{sec:inversegammamodel}.
-  The hyperparameters are set as follows: $\lambda=0.1$ and 
-  $\theta=2$. We choose the penalty parameter in \texttt{glmnet} by assuming a 
-  known error variance and matching the marginal prior variance of the NIG and
-  ridge models, which results in the penalty parameter 
-  $1/(n\sigma^2\theta)=1/200$. We
-  repeated the simulation 100 times and report the results in Figures 
-  \ref{fig:boxplots_igaussian_res3_vb}-
-  \ref{fig:boxplots_igaussian_res3_best}.
+  so we restrict ourselves to the univariate setting again (i.e., $D=1$). The
+  simulation setup is the same as in the dense setting 
+  in Section \ref{sec:inversegammamodel}. Similarly, we use the same 
+  hyperparameters settings (Table \ref{tab:hyper_set1}. For \texttt{glmnet}
+  we either cross-validate the penalty parameter $\lambda$ or match it to the
+  true prior variance. We
+  repeated the simulations 100 times and report the results in Figure 
+  \ref{fig:boxplots_igaussian_res4.2_post1}.
 
-<<boxplots_igaussian_res3_vb, fig.cap="(a) Cramér-von Mises criterion, (b) correlation, and (c) MSD between the variances of the MCMC and variational bayes posteriors.", out.width="80%", fig.asp=1/2>>=
-@
-
-<<boxplots_igaussian_res3_best, fig.cap="(a) Correlation and (b) MSD between the $\\beta_j$ point estimates of MCMC and ADVI, VB, glmnet, and MAP.", out.width="80%", fig.asp=3/5>>=
+<<boxplots_igaussian_res4.2_post1, fig.cap="(a) \\& (d) Cramér-von Mises criterion, (b) \\& (e) correlation, and (c) \\& (f) MSD between the $\\beta_j$ point estimates of the MCMC and variational bayes posteriors.", out.width="80%", fig.asp=2/3>>=
 @
@@
 
-  \section{Empirical Bayes estimation}
-  We simulate from a correctly specified model given in \ref{eq:priormodel}. We 
-  assume random $y_{id}$, $\x_i$, and $\bm{\beta}_d$, while we fix $\mathbf{C}$,
-  $\gamma_d^2$, $\sigma_d^2$ and $\bm{\alpha}$. We estimate the $\bm{\mu}_d$,
-  $\bm{\Sigma}_d$, $\delta_d$, $\bm{\alpha}$ and $\lambda$, while we fix the
-  $\sigma_d^2$ to their true values.
+  \subsection{Empirical Bayes estimation}
+  To assess the empirical Bayes estimation of the inverse Gaussian prior, 
+  we conduct a small simulation study where we simulate from
+  model (\ref{eq:priormodel}), while keeping $\forall d: \sigma^2_d=1$ fixed. 
+  More specifically, we simulate 
+  $\gamma_{jd}^2 \sim \mathcal{IG}([(\mathbf{c}_d^j) \tr \bm{\alpha}]^{-1},
+  \lambda_d)$, for $j=1,\dots,200$, $d=1,\dots,10$, where we set 
+  $\bm{\alpha}= \begin{bmatrix} 1 & 1 & 2 & 3 & 4 \end{bmatrix}$. We use 
+  balanced dummy coding for the $\mathbf{c}_d^j$, such that we end up with
+  $\E(\gamma_{jd}) \in \{ 1, 0.5, 0.33, 0.25, 0.2 \}$, and fix 
+  $\forall d: \lambda_d=1$.  
+  Next, we simulate $\beta_{jd} \sim \mathcal{N} 
+  (0, \gamma_{jd}^2)$.
+  The observed data is generated from $x_{ij} \sim \mathcal{N} (0, 1)$ and 
+  $y_{id} \sim \mathcal{N} (\mathbf{x}_i \tr \bm{\beta}_d, 1)$, for $i=100$. 
+  We repeat the simulation 100 times and, for each simulation, estimate the 
+  hyperparameters by the method described in Section \ref{sec:empiricalbayes} 
+  and present the results in Figure \ref{fig:boxplots_igaussian_res6_eb}.
 
-  We set $n=100$, $p=200$, and $D=10$. We consider 5 evenly sized classes of
-  drugs, such that $\mathbf{c}_{kd} = \mathbbm{1}(k = \text{class}_d)$ and
-  $\begin{bmatrix} \mathbf{c}_1 & \cdots & \mathbf{c}_D \end{bmatrix} \tr$. From
-  these the $\gamma_d^2 = (\mathbf{c}_d \tr \bm{\alpha})^{-1}$ are created. 
-  Next, we simulate $\bm{\beta}_{jd} \sim \mathcal{N} 
-  (0, \sigma^2_d \gamma_d^2)$. We fix the mean signal-to noise ratio 
-  $\overline{\text{SNR}}_d$, such that the predictor data variance is 
-  $s^2 = \V(\mathbf{x}_i) = \overline{\text{SNR}}_d / 
-  (\overline{\gamma^2_d} p)$ (NOT CORRECT). The last step consists of 
-  simulating the data by $\mathbf{x}_{ij} \sim \mathcal{N}(0,s^2)$ and 
-  $y_{id} = \mathcal{N} (\mathbf{x}_i \tr \bm{\beta}_d, \sigma_d^2)$. The
-  remaining parameters are set as follows: $\bm{\alpha} = \begin{bmatrix} 1 &
-  \dots & 5 \end{bmatrix} \tr$, $\sigma_d^2 = 1$ for all $d$, and 
-  $\overline{\text{SNR}}_d = 10$. To account for the random data and parameter
-  generation, we repeat the simulation 100 times. We compare the results to the
-  inverse Gamma and non-conjugate model, and present the results in Figures
-  \ref{fig:boxplot_igaussian_res1_prior_mean}-
-  \ref{fig:boxplot_igaussian_res2_prior_mean}.
-
-<<boxplot_igaussian_res1_prior_mean, fig.cap="Prior mean estimates", out.width="80%", fig.asp=2/3>>=
+<<boxplots_igaussian_res6_eb, fig.cap="Empirical Bayes $\\hat{\\bm{\\alpha}}$ estimates", out.width="80%", fig.asp=2/3>>=
 @
 
-<<boxplot_igaussian_res2_prior_mean, fig.cap="Prior mean estimates", out.width="80%", fig.asp=2/3>>=
+  [! describe results here]
+
+  - seems to be an overall scaling issue, relative to each other the estimates
+  are not so bad.
+
+  - investigate convergence: (i) look at both error and prior variance 
+  convergence and (ii) look at actual prior distribution convergence 
+  (plot true distribution and estimated prior every 5 iterations or so)
+
+  \section{Application in eQTLs mapping}\label{sec:application}
+  This application concerns eQTLs detection from the GEUVADIS data
+  \cite[]{lappalainen_transcriptome_2013}. After the removal of 9 genes due to
+  missing SNP information, we have 99 genes (outcomes) from the p38MAPK pathway. 
+  After removal of the subjects without any expression or SNP data, 373 subjects
+  remained. For each gene, we selected SNPs with minor alles frequency (MAF) 
+  larger than 0.05 and within 105 bases up- or downstream from the gene. This 
+  resulted in 99 sets of SNPs (one for each gene) from the 373 subjects, where 
+  the number of SNPs per gene ranges from 56 to 1169.
+
+  We used Ensembl \cite[]{hunt_ensembl_2018} to create several external 
+  variables: (i) SNP location 1 (binary, in- or outside of the gene),
+  (ii) SNP location 2 (binary, in an exon or an intron), (iii) 
+  genelength (continuous), (iv) minor allele frequency of the SNP
+  (continuous), and (v) squared MAF. We expect that SNPs that are in the gene 
+  and/or in an exon are more important in eQTLs detection. Since a larger gene
+  might depend on more SNPs, we expect that longer genes yield lower 
+  regression parameter estimates. We predict a positive correlation between
+  the regression parameters and the MAF.
+
+  We fit five different models to a subset of the subjects and 
+  estimated performance measures on the rest of the subjects. The five models we
+  fit differed only in the included external variables. The intercept only model
+  is considered a reference model. The other four models all included the 
+  genelengths and MAFs (variables (iii) and (iv)). 
+  Models 1 and 2 included variable 
+  (i), while models 1, 3, and 4 included variable (ii). In addition, model 4
+  included the squared MAFs (v).
+
+  For each model, we ran the algorithm for 100 EB iterations and obtained 
+  the $\hat{\bm{\alpha}}$ hyperparameter estimates in Table 
+  \ref{tab:hyperparameters_eQTL}. In addition, we calculated posterior means
+  for the $\bm{\beta}_d$ from MCMC samples from the posterior with the estimated
+  hyperparameters. We estimated mean squared error (MSE) for these posterior 
+  means. A common measure 
+  of Bayesian model fit is the log pseudo marginal likelihood
+  (LPML) \cite[]{gelfand_model_1992}, the Bayesian version of the leave-one-out 
+  cross-validated log likelihood (see Section \ref{sm-sec:lpml} for more 
+  details). We estimated LPML for the different models and added them,
+  together with the MSEs to Table \ref{tab:hyperparameters_eQTL}.
+<<hyperparameters_eQTL>>= 
+load("results/eQTL_fit1.Rdata")
+seq.mse <- c(2.0699, 2.0629, 2.0649, 2.0628, 2.0638)
+seq.lpml <- round(c(mean(lpml.ienig$lpml), mean(lpml1.enig$lpml),
+                    mean(lpml2.enig$lpml), mean(lpml3.enig$lpml), 
+                    mean(lpml4.enig$lpml)), 5)
+tab <- cbind(round(rbind(c(fit.ienig$eb$alpha, rep(NA, 5)),
+                         c(fit1.enig$eb$alpha, NA),
+                         c(fit2.enig$eb$alpha[c(1:2)], NA, 
+                           fit2.enig$eb$alpha[-c(1:2)], NA),
+                         c(fit3.enig$eb$alpha[1], NA, 
+                           fit3.enig$eb$alpha[-1], NA),
+                         c(fit4.enig$eb$alpha[1], NA, 
+                           fit4.enig$eb$alpha[-1])), 2),
+             seq.mse, seq.lpml)
+tab[which.min(tab[, 7]), 7] <- kableExtra::cell_spec(
+  tab[which.min(tab[, 7]), 7], format="latex", bold=TRUE)
+tab[which.max(tab[, 8]), 8] <- kableExtra::cell_spec(
+  tab[which.min(tab[, 8]), 8], format="latex", bold=TRUE)
+rownames(tab) <- c("intercept only", paste("model", 1:4))
+options(knitr.kable.NA = '')
+kableExtra::kable_styling(kableExtra::column_spec(
+  knitr::kable(tab, align=rep("r", ncol(tab)),
+               col.names=c("intercept", "in gene", "in exon", "length", 
+                           "MAF", "MAF$^2$", "MSE", "LPML"),
+               caption="Estimated $\\hat{\\bm{\\alpha}}$ hyperparameters, 
+               MSE (lowest in bold), and LPML (highest in bold).", 
+               "latex", booktabs=TRUE, escape=FALSE, 
+               digits=c(rep(2, 6), 4, 4)), 8, border_left=TRUE),
+  latex_options=c("HOLD_position"))
 @
 
-  - investigate convergence: (i) look at both error and prior variance 
-  convergence and (ii) look at actual prior distribution convergence (
-  plot true distribution and estimated prior every 5 iterations or so)
+  - show influence of co-data with standardized $\bm{\alpha}$'s.
 
-  - try to fix convergence: (i) track only expectation and variance or ELBO, 
-  (ii) reparametrise, and (iii) fix $\forall d: \lambda_d=\lambda$.
+  A quantity that is often of interest in eQTLs detection studies, is the 
+  gene expression heritability. It is the proportion of a gene's expression that
+  is explained with the SNPs and a measure of the relationship between trait 
+  phenotypic variation and genetic variation. Heritability (per gene) is 
+  given by $h_d^2 = (\sigma^2_d \sum_{j=1}^{p_d} \gamma_{jd}^2)/
+  (\sigma^2_d \sum_{j=1}^{p_d} \gamma_{jd}^2 + \sigma_d^2)$. We assess 
+  mean heritability of the five models for a rank-based selection of SNPs.
+  We present MCMC-based posterior means and one standard deviation error bars in
+  Figure \ref{fig:lines_eqtl_hsq}.
+<<lines_eqtl_hsq, fig.cap="Mean heritability in the eQTL detection data.", out.width="80%", fig.asp=2/3>>=
+@
 
-  \section{Application}
-  - Possible: network reconstruction, medical surveys, drug response prediction
+  [! describe results here]
 
-  - metabolomics data: oxidative stress als uitkomst van de andere metabolieten
-  min organic acids omdat ze downstream zijn. oxidateve stress onderverdelen in
-  high and low als uitkomst co-data en de andere metabolieten onderverdelen
-  als predictor co-data.
-  \section{Discussion}
-  - inverse Gaussian more convenient than inverse Gamma, because EB is closed-
-  form
+  - maybe use explained heritability (fraction of full model heritability)
 
-  - use copula's and marginal posteriors to find multivariate posteriors for
-  contrast testing.
+  - AUC of curve as overal measure
 
-  - apart from the scaling issue, the conjugate prior model is preferred
-  over the non-conjugate model in \cite{moran_variance_2018}.
-  The assumed factorization in the variational Bayes approximation 
-  $q(\bbeta_d) \cdot q(\sigma_d)$ is actually true for this model. The 
-  approximate expectation used in the empirical Bayes step is therefore more 
-  accurate
+  % Annotation is from GRCh37 reference genome: 
+  % \url{https://www.ncbi.nlm.nih.gov/assembly/GCF_000001405.13/}. 
+  % Annotation file that Gino used: 
+  % \url{https://www.gencodegenes.org/human/release_12.html}.
+  % Ensembl version of this genome:
+  % \url{https://grch37.ensembl.org/Homo_sapiens/Info/Index}.
+  % Possibly use GeuvadisTranscriptExpr R package.
+
+  \section{Discussion}\label{sec:discussion}
+  In this paper we have introduced a normal inverse Gaussian prior for 
+  multivariate high dimensional linear regression models that is estimated by
+  empirical-variational Bayes. The estimation is adaptive to external data
+  that describes the relation between the outcomes and features in a data-driven
+  manner. In simulations we have shown that the inverse Gaussian model 
+  does not perform worse, and often bettern, than a common alternative, 
+  the Student's $t$ prior. in addition, the Student's $t$ prior does not admit 
+  relatively straightforward estimation, compared to the inverse Gaussian 
+  prior, as explained in Section \ref{sec:estimation}. The 
+  simulations also show that the model is reasonably well approximated and 
+  estimated using empirical-variational Bayes. An eQTL detection application 
+  shows that external data may indeed benefit prior estimation and prediction
+  performance.
+
+  Apart from eQTL detection, the model may be used for network reconstruction as 
+  in \cite{leday_gene_2017}, drug response prediction (manuscript in 
+  preparation by the current authors), and other applications with multiple
+  outcomes, large feature space, and available external information.
+
+  A drawback of the method is the fact that the external data is required to
+  be positive. That may not always be the case in practice. We note, however,
+  that the $\mathbf{c}_d^j$ may be translated to be positive by adding a 
+  constant. In the usual predictive feature sense this doesn't work, since one
+  might still encounter negative future values. Here, the external data is
+  only used for estimation and not for prediction, so this issue is 
+  non-existent.
 
@@
 	\section*{Acknowledgements}
+	We thank Gino Kpogbezan for providing the pre-processed data in Section 
+	\ref{sec:application}.
 	\textit{Conflict of Interest}: None declared.

### 2019-08-13 docs/manuscript2.Rnw
updated simulations and manuscript

--- 2019-05-21 docs/manuscript2.Rnw
@@
 \usepackage{url,bm,algorithm,algpseudocode,bbm,amsmath,amssymb,amsthm,
-amsfonts,graphics,graphicx,epsfig,rotating,caption,subcaption,natbib,appendix,
+amsfonts,graphics,graphicx,epsfig,rotating,caption,subcaption,natbib,
 titlesec,multicol,verbatim,threeparttable,booktabs,dsfont,parskip,
@@
 	\section{Introduction}
-	- motivation: drug response prediction from different omics data types is promising both for precision medicine and drug target discovery and relates to actual tumor response \cite[]{iorio_landscape_2016}.
+	- motivation: drug response prediction from different omics data types is 
+	promising both for precision medicine and drug target discovery and relates to
+	actual tumor response \cite[]{iorio_landscape_2016}.
 
-	- problem: prediction of drug efficacy in cell lines is difficult (DREAM 7 challenge in \cite{costello_community_2014}).
+	- problem: prediction of drug efficacy in cell lines is difficult (DREAM 7 
+	challenge in \cite{costello_community_2014}).
 
@@
 
-	- feature information: prior experts, pathways, database summaries, previous experiments, type of feature (somatic mutations, copy number alterations, DNA methylation, and gene expression).
+	- feature information: prior experts, pathways, database summaries, previous 
+	experiments, type of feature (somatic mutations, copy number alterations, DNA 
+	methylation, and gene expression).
 
-	- drugs information: molecular target, drug class, etc (ask Oslo), pubchem, omnipath DB, pharmaco DB
+	- drugs information: molecular target, drug class, pubchem, 
+	omnipath DB, pharmaco DB
 
-	- current literature: review in \cite{azuaje_computational_2017}, reference Oslo, TANDEM \cite[]{aben_tandem:_2016}, DeltaNet \cite[]{noh_inferring_2016}, patient-derived tumor cell lines \cite[]{gao_high-throughput_2015}.
+	- drugs x feature information
 
-	- our solution: adaptive borrowing of information through empirical Bayes, guided by external feature and drug information.
+	- current literature: reviews in \cite{azuaje_computational_2017} and
+	\cite{ali_machine_2019}, 
+	full multivariate approach by Oslo people in \cite{mai_composite_2019}, 
+	TANDEM \cite[]{aben_tandem:_2016}, DeltaNet 
+	by \cite{noh_inferring_2016}, patient-derived tumor cell lines 
+	in \cite{gao_high-throughput_2015}, 
+	multiple kernel learning approach in \cite{costello_community_2014},
+	\cite{ammad-ud-din_drug_2016} combine a 
+	multiple kernel learning approach with matrix factorization, using the model 
+	to uncover the latent 
+	relationships between drug targets and intracellular pathways, and
+	structured penalized regression in \cite{zhao_structured_2019}
 
-	\section{Model}
+	- our solution: adaptive borrowing of information through empirical Bayes, 
+	guided by external feature and drug information.
+
+	\section{Methods}
 	\subsection{Simultaneous equations model}\label{sec:SEM}
-	- Gwen can deal with different number of features per equation, might solve missing drug-feature combination problem? Doesn't solve missing response-feature combination problem.
+	Assume we have continuous efficacy measures $y_{id}$ for cell lines 
+	$i=1,\dots, n$, centred per drug $d$. We predict efficacy from molecular 
+	features 
+	$x_{ij}$, $j=1,\dots, p$, collected in $\x_i = \begin{bmatrix} x_{i1} & 
+	\dots & x_{ip} \end{bmatrix} \tr$. We have an \textit{a priori} (possibly 
+	informative) partitioning 
+	of the omics features into $g=1, \dots, G$ non-overlapping groups.
+	For convenience, we let 
+	$g(j)$ denote the group index of molecular feature $j$. Furthermore, we have 
+	characteristics of the drugs available in the form of external variables 
+	$\mathbf{C} = \begin{bmatrix} \mathbf{c}_1 & \dots & \mathbf{c}_D 
+	\end{bmatrix} \tr$.
 
-	- check literature for: a) conjugate version, b) similar models: "prior variance decomposition", "structured Bayesian regression"
-
-	We have continuous efficacy measures $y_{id}$ for cell lines $i=1,\dots, n$, from tissues $t=1, \dots, T$, on drugs $d=1,\dots,D$.
-Throughout this paper we assume the $y_{id}$ to be centred per drug and let $\y_d = \begin{bmatrix} y_{1d} & \dots & y_{nd} \end{bmatrix} \tr$.
-We predict efficacy with molecular features $x_{ij}$, $j=1,\dots, p$, collected in $\x_i = \begin{bmatrix} x_{i1} & \dots & x_{ip} \end{bmatrix} \tr$.
-We have an \textit{a priori} partitioning of the omics features into $g=1, \dots, G$ non-overlapping groups, which might be informative for the problem at hand For convenience, we let $g(j)$ denote the group index of molecular feature $j$.
-Furthermore, we have characteristics of the drugs available in the form of `co-data' $\mathbf{C} = \begin{bmatrix} \mathbf{c}_1 & \dots & \mathbf{c}_D \end{bmatrix} \tr$.
-We code the tissue from which the cell lines were taken as a $T$ binary dummy variables
-	$$
-	z_{it} = \begin{cases}
-	1 & \text{if } \text{tissue}_i = t, \\
-	0 & \text{otherwise},
-	\end{cases}
-	$$
-	collected in $\mathbf{z}_i = \begin{bmatrix} z_{i1} & \dots & z_{iT} \end{bmatrix} \tr$.
-
-	We assume that cell lines from the same tissue are related and model this relation through random effects $u_{td}$.
-We let the molecular feature effects $\beta_{jd}$ vary over the drugs and model drug efficacy as a linear combination of these two effects:
-	\begin{subequations}\label{eq:linearmodel}
+	We regress the drug efficacies on the molecular feature effects $\beta_{jd}$:
+	\begin{subequations}
 		\begin{align}
-		y_{id} & = \beta_{0d} + \sum_{j=1}^p x_{ij} \beta_{jd} + \sum_{t=1}^T z_{it} u_{td} + \epsilon_{id} \\
-		& = \beta_{0d} + \x_i \tr \bbeta_d + \mathbf{z}_i \tr \mathbf{u}_d + \epsilon_{id},\\
+		  y_{id} & = \x_i \tr \bbeta_d + \epsilon_{id}, \label{eq:linearmodel}\\
 		  \text{with } & \epsilon_{id} \sim \mathcal{N}(0, \sigma_d^2),
@@
 	\end{subequations}
-	where $p$-dimensional $\bbeta_d$ and $T$-dimensional $\bm{u}_d$ are the drug-specific omics feature and tissue effect vectors, respectively.
-Note that (\ref{eq:linearmodel}) gives rise to a system of $d=1, \dots, D$ linear simultaneous equations.
+	where $p$-dimensional $\bbeta_d$ are the drug-specific omics feature effects. 
+	Note that (\ref{eq:linearmodel}) gives rise to a system of $d=1, \dots, D$ 
+	linear simultaneous equations.
 
 	\subsection{Bayesian prior model}
-	We capture the uncertainty in the parameters through a Bayesian prior model.
-We assign a flat prior to the intercept terms $\beta_{0d}$ and integrate them out.
-The remaining parameters are endowed with the following priors:
+	We capture the uncertainty in the parameters through a Bayesian prior model
+	and endow the parameters with the following prior:
 	\begin{subequations}\label{eq:prior}
 		\begin{align}
-		\beta_{jd} & \sim \mathcal{N}_p (0, \phi^2_{g(j)} \gamma_d^2 \sigma_d^2), \label{eq:betaprior}\\
-		\mathbf{u}_{d} & \sim \mathcal{N}_T (\mathbf{0}, \bm{\Xi}_d \sigma_d^2),
+		  \beta_{jd} & \sim \mathcal{N}_p (0, \phi^2_{g(j)} \gamma_{jd}^2 
+		  \sigma_d^2), 
+		  \label{eq:betaprior}\\
+		  \gamma_{jd}^{2} & \sim \mathcal{IG}( 
+		  [(\mathbf{c}_d^j) \tr \bm{\alpha}]^{-1}, \lambda_d), \\
+		  \sigma_d^{2} & \sim 1/\sigma_d^{3},
 		\end{align}
 	\end{subequations}
-	and hyper-priors:
-	\begin{subequations}\label{eq:hyperprior}
-		\begin{align}
-			\sigma_d^{2} & \sim 1/\sigma_d^{3} \\
-			\gamma_d^{2} & \sim \mathcal{IG}(\theta_d, \lambda), \\
-			\bm{\Xi}_d & \sim \mathcal{W}^{-1}_T(\bm{\Omega}, \nu),
-		\end{align}
-	\end{subequations}
-	where $\mathcal{IG}(\theta, \lambda)$ is an inverse Gaussian distribution with mean $\theta$ and shape $\lambda$, and $\mathcal{W}^{-1}_T(\bm{\Omega}, \nu)$ is an inverse Wishart distribution with $T \times T$-dimensional positive definite (PD) scale matrix $\bm{\Omega}$ and degrees of freedom $\nu$.
-Note that for $\nu > T + 1$, this inverse Wishart prior implies an expected prior covariance matrix $(\nu - T - 1)^{-1} \bm{\Omega}$ for the tissue effects, common to all drugs.
-Likewise, the molecular feature effects are related through the common hyper-prior for the $\gamma_d^2$.
+	where $\mathcal{IG}(\theta, \lambda)$ denotes an inverse Gaussian distribution 
+	with mean $\theta$ and shape $\lambda$. The different molecular feature types
+	are given their own weights through the $\phi^2_{g}$, while the external 
+	outcome-feature data comes in through 
+	our mean model for the $\gamma_{jd}^2$ hyperprior model: 
+	$[(\mathbf{c}_d^j) \tr \bm{\alpha}]^{-1}$. Both the $\phi^2_{g}$ and
+	$\bm{\alpha}$ are estimated in a data-driven manner. More details on the model
+	itself, as well as its estimation, may be found in 
+	the SM and in [! ref to first paper here].
 
-	- Jeffrey's prior for variance $1/\sigma^{3/2}$ (a priori independent data mean and variance) or $1/\sigma^3$ (joint prior, but only mean and variance, not variance and betas).
-
-	A few remarks on the choice of priors are justified here: many authors endow error variance components with vague gamma priors.
-\cite{gelman_prior_2006}, among others, advises against this practice.
-The degree of `vagueness' has a large influence on the posterior, while degree of `vagueness' is a difficult parameter to set.
-This influence is especially pronounced if the likelihood is relatively flat, as may be reasonably expected in the large $p$, small $n$ setting we are in.
-We therefore model the error variance with Jeffrey's prior \cite[]{jeffreys_invariant_1946}.
-
-	Furthermore, we choose to model the $\gamma^2_d$ by an inverse Gaussian distribution, as has been suggested in \cite{fabrizi_specification_2016} and \cite{caron_sparse_2008}, because it allows to model the mean $\theta_d$ as a function of the drug covariates $\mathbf{c}_d$, as explained in Section \ref{sec:empiricalbayes}.
-
-	- IG hyperprior used in finance \cite[]{barndorff-nielsen_normal_1997}, where it is called the NIG prior.
-
-	\section{Estimation}
-	\subsection{Variational Bayes}\label{sec:variationalbayes}
-	The posterior corresponding to the model described in (\ref{eq:linearmodel}), (\ref{eq:prior}), and (\ref{eq:hyperprior}) is not available in closed form.
-We therefore approximate the posterior by variational Bayes, where we force the posterior density $Q$ to factorise as:
-	\begin{align}\label{eq:variationalposterior}
-	p(\B,\mathbf{U},\mathbf{\sigma}^2, \bgamma^2,\bm{\Xi}_1, \dots, \bm{\Xi}_D) \approx Q(\cdot) = q(\B) \cdot q(\mathbf{U}) \cdot q(\mathbf{\sigma}^2) \cdot q(\bgamma^2) \cdot q(\bm{\Xi}_1, \dots, \bm{\Xi}_D),
-	\end{align}
-	where $\B = \begin{bmatrix} \bbeta_1 & \cdots & \bbeta_D \end{bmatrix}$ and $\U = \begin{bmatrix} \mathbf{u}_1 & \cdots & \mathbf{u}_D \end{bmatrix}$, are of dimensions $p \times D$ and $T \times D$, respectively.
-Under the factorisation in (\ref{eq:variationalposterior}), the marginal variational posteriors that minimise the Kullback-Leibler divergence of the true posterior to the variational Bayes approximation \cite[]{neal_view_1998}, are given by:
-	\begin{subequations}\label{eq:variationalposterior2}
-		\begin{align}
-		q_{\B} (\B) & \overset{D}{=} \prod_{d=1}^{D} \mathcal{N}_p (\bmu_d, \bSigma_d), \\
-		q_{\U}(\U) & \overset{D}{=} \prod_{d=1}^{D} \mathcal{N}_T (\mathbf{m}_d, \Sm_d), \\
-		q_{\bgamma^{2}}(\bgamma^{2}) & \overset{D}{=} \prod_{d=1}^D \mathcal{GIG} \(-\frac{p+1}{2}, \frac{\lambda}{\theta_d^2}, \delta_{d} \), \\
-		q_{\bsigma^{2}}(\bsigma^{2}) & \overset{D}{=} \prod_{d=1}^D \Gamma^{-1} \(\frac{n + p + T + 1}{2}, \zeta_{d} \), \\
-		q_{\bm{\Xi}_1, \dots, \bm{\Xi}_D}(\bm{\Xi}_1, \dots, \bm{\Xi}_D) & \overset{D}{\propto} \prod_{d=1}^D \mathcal{W}_T^{-1} (\bm{\Psi}_d, \nu + 1).
-		\end{align}
-	\end{subequations}
-	Here $\mathcal{GIG}(\cdot)$ and $\Gamma^{-1}(\cdot)$ denote the generalized inverse Gaussian and inverse gamma distributions, respectively.
-The full derivations may be found in Appendix \ref{app:derivations}.
-
-	The parameters in (\ref{eq:variationalposterior2}) contain cyclic dependencies. We therefore iterate the following estimating equations until convergence to a local optimum:
-	\begin{align*}
-	\bSigma_d^{(h+1)} & = \frac{2\zeta_d^{(h)}}{n+p+T+1} \left\{\X \tr \X + a_d^{(h)} \cdot \diag (\phi_{g(j)}^{-2}) \right\}^{-1}, \\
-	\bm{\mu}_d^{(h+1)} & = \left\{\X \tr \X + a_d^{(h)} \cdot \diag (\phi_{g(j)}^{-2}) \right\}^{-1} \X \tr ( \y_d - \Z \mathbf{m}_d^{(h)}), \\
-	\Sm_d^{(h+1)} & = \frac{2\zeta_d^{(h)}}{n+p+T+1} \[ \Z \tr \Z + (\nu + 1) \cdot (\bm{\Psi}_d^{(h)})^{-1} \]^{-1}, \\
-	\mathbf{m}_d^{(h+1)} & = \[ \Z \tr \Z + (\nu + 1) \cdot (\bm{\Psi}_d^{(h)})^{-1} \]^{-1} \Z \tr (\y_d - \X \bm{\mu}_d^{(h)}), \\
-	\bm{\Psi}_d^{(h+1)} & = \frac{2\zeta_d^{(h)}}{n+p+T+1} \[ \Sm_d^{(h+1)} + \mathbf{m}_d^{(h+1)} (\mathbf{m}_d^{(h+1)}) \tr \] + \bm{\Omega}, \\
-	\delta_d^{(h+1)} & =  \frac{n+p+T+1}{2\zeta_d^{(h)}} \sum_{j=1}^p \frac{(\mu_{jd}^{(h+1)})^2 + (\bSigma_d^{(h+1)})_{jj}}{\phi^2_{g(j)}} + \lambda, \\
-	\zeta_d^{(h+1)} & = \frac{1}{2} \y_d \tr \y_d - \y_d \tr (\X \bmu_d^{(h+1)} + \Z \mathbf{m}_d^{(h+1)}) + (\bmu_d^{(h+1)}) \tr \X \tr \Z \mathbf{m}_d^{(h+1)} + \frac{1}{2} \trace (\X \tr \X \bSigma_d^{(h+1)}) \\
-	& \,\,\,\,\,\,\,\,\,\, + \frac{1}{2} (\bmu_d^{(h+1)}) \tr \X \tr \X \bmu_d^{(h+1)} + \frac{1}{2} \trace (\Z \tr \Z \Sm_d^{(h+1)}) + \frac{1}{2} (\mathbf{m}_d^{(h+1)}) \tr \Z \tr \Z \mathbf{m}_d^{(h+1)} \\
-	& \,\,\,\,\,\,\,\,\,\, + \frac{1}{2} a_d^{(h)} \sum_{j=1}^p \frac{(\mu_{jd}^{(h+1)})^2 + (\bSigma_d^{(h+1)})_{jj}}{\phi^2_{g(j)}} + \frac{1}{2} (\nu + 1) \cdot \trace \[ (\bm{\Psi}_d^{(h+1)})^{-1} \Sm_d^{(h+1)} \] \\
-	& \,\,\,\,\,\,\,\,\,\, + \frac{1}{2} (\nu + 1) \cdot (\mathbf{m}_d^{(h+1)}) \tr (\bm{\Psi}_d^{(h+1)})^{-1} \mathbf{m}_d^{(h+1)},
-	\end{align*}
-	where
-	$$
-	a_d^{(h)} = \sqrt{\frac{\lambda}{\theta^2_d \delta_d^{(h)}}} \frac{K_{\frac{p - 1}{2}} \( \sqrt{\lambda \delta_d^{(h)}}/\theta_d \)}{K_{\frac{p+1}{2}} \( \sqrt{\lambda \delta_d^{(h)}}/\theta_d \)} + \frac{p+1}{\delta_d^{(h)}},
-	$$
-	and $K_{\alpha}(\cdot)$ is the modified Bessel function of the second kind.
-
-	\subsection{Empirical Bayes}\label{sec:empiricalbayes}
-	A full Bayesian model requires the specification of hyper-parameters $\bm{\eta} = \begin{bmatrix} (\bm{\phi}^2) \tr & \bm{\theta} \tr & \lambda & \text{vec}(\bm{\Omega}) \tr & \nu \end{bmatrix} \tr$.
-These are abstract and hard to interpret parameters for which we generally lack expert knowledge.
-They do, however, have a significant influence on the shape of the posterior distribution.
-We therefore propose to estimate the unspecified hyper parameters $\bm{\eta}$ by empirical Bayes.
-Simply put, empirical Bayes fits the prior model to the data and is, as such, objective (up to model specification).
-
-	The canonical method for empirical Bayes is maximisation of the marginal likelihood with respect to the hyper parameters.
-In \cite{casella_empirical_2001} the marginal likelihood is maximised by an EM algorithm:
-	$$
-	\bm{\eta}^{(l+1)} = \underset{\bm{\eta}}{\argmax}\E_{\cdot | \Y} [\log p(\Y, \B, \U, \bm{\Xi}_1, \dots, \bm{\Xi}_D, \bgamma^2, \bsigma^2) | \bm{\eta}^{(l)}],
-	$$
-	where the expectation is with respect to the posterior.
-In our case, this posterior is not available in closed form, which renders the expectation difficult.
-While \cite{casella_empirical_2001} suggests to approximate the expectation by a Monte Carlo sample, we propose to use the variational Bayes approximation developed in Section \ref{sec:variationalbayes}:
-	\begin{align}\label{eq:variationalbayes}
-	\bm{\eta}^{(l+1)} & = \underset{\bm{\eta}}{\argmax}\E_{Q^{(l)}} [\log p(\Y, \B, \U, \bm{\Xi}_1, \dots, \bm{\Xi}_D, \bgamma^2, \bsigma^2)] \nonumber \\
-	& = \underset{\bm{\eta}}{\argmax} \E_{Q^{(l)}} [\log \pi (\B | \bgamma^2, \bsigma^2)] + \E_{Q^{(l)}} [\log \pi (\bgamma^2)] + \E_{Q^{(l)}} [\log \pi (\bm{\Xi}_1, \dots, \bm{\Xi}_D)].
-	\end{align}
-	where now the expectation is with respect to the converged variational posterior $Q^{(l)}$.
-Inspection of (\ref{eq:variationalbayes}) learns us that the problem may be decomposed into three separate optimisation problems.
-
-	\subsubsection{Group-specific variance components}
-	The first sub-problem concerns the group-specific variance components of the molecular feature effects.
-In order to separate the overall variance of the molecular feature effects from the group-specific deviations, we impose $\prod_{g=1}^G (\phi_g^2)^{|\mathcal{G}_g|} = 1$.
-As a consequence, the overall level of shrinkage is determined by the $\gamma^2_d$, while the $\phi_g^2$ effectuate differential shrinkage of the groups.
-Because the $\phi_g^2$ are variance components, we additionally require $\phi^2_1, \dots, \phi^2_G> 0$, such that our first optimisation sub-problem becomes:
-	\begin{align*}
-	(\bm{\phi}^2)^{(l+1)} & = \underset{\bm{\phi}^2}{\argmax} \E_{Q^{(l)}} [\log \pi (\B | \bgamma^2, \bsigma^2)] \\
-	& = \underset{\bm{\phi}^2}{\argmax} -\frac{D}{2} \sum_{g=1}^G |\G_g| \log \phi_g^2 - \frac{1}{2} \sum_{g=1}^G b_g^{(l)} \phi_g^{-2} \\
-	& \text{subject to } \prod_{g=1}^G (\phi_g^2)^{|\G_g|} = 1 \text{ and } \phi^2_1, \dots, \phi^2_G> 0.
-	\end{align*}
-	where 
-	$$
-	b_g^{(l)}=\sum_{d=1}^D a_d^{(l)} \frac{n+p+T+1}{2\zeta_d^{(l)}} \sum_{j \in \mathcal{G}_g} \[ (\mu_{jd}^{(l)})^2 + (\bSigma_d^{(l)})_{jj} \].
-	$$
-	This is a convex optimisation problem and easily solved numerically.
-
-	\subsubsection{Drug-specific variance components}
-	The second sub-problem concerns the prior for the drug-specific variance components $\gamma_d^2$.
-As explained in Section \ref{sec:SEM}, we may have `covariates' $\mathbf{c}_d$ on the drugs available.
-We conjecture that these covariates may be informative for the omics effect size and therefore model the prior mean of the $\gamma_d^2$ as a function of these covariates: $\theta_d = (\mathbf{c}_d \tr \bm{\alpha})^{-1}$.
-Then, the empirical Bayes estimation of $\bm{\alpha}$ effectively boils down to an inverse Gaussian regression of the $\gamma_d^{-2}$ on the $\mathbf{c}_d$ \cite[]{fries_optimal_1986,whitmore_regression_1983}.
-The problem may be formulated as:
-	\begin{align*}\label{eq:optproblem2}
-	\lambda^{(l+1)}, \bm{\alpha}^{(l+1)} & = \underset{\lambda,\bm{\alpha}}{\argmax} \E_{Q^{(l)}} [\log \pi (\bgamma^2)] \\
-	& = \underset{\lambda,\bm{\alpha}}{\argmax} \frac{D}{2} \log \lambda - \frac{\lambda}{2} \bm{\alpha} \tr \mathbf{C} \tr \diag [\E_{Q^{(l)}}(\gamma_d^2)] \mathbf{C} \bm{\alpha} + \lambda \bm{\alpha} \tr \mathbf{C} \tr \mathbf{1} - \frac{\lambda}{2} \sum_{d=1}^D \E_{Q^{(l)}}(\gamma_d^{-2}),
-	\end{align*}
-	where $\mathbf{C} = \begin{bmatrix} \mathbf{c}_1 & \dots & \mathbf{c}_D \end{bmatrix} \tr$. The solutions to this problem are:
-	\begin{align*}
-	\bm{\alpha}^{(l+1)} & = \left\{ \mathbf{C} \tr \Vm^{(l)} \mathbf{C} \right\}^{-1} \mathbf{C} \tr \mathbf{1}, \\
-	\lambda^{(l+1)} & = \[ \sum_{d=1}^D a_d^{(l)} - (\bm{\alpha}^{(l+1)}) \tr \mathbf{C} \tr \mathbf{1} \]^{-1} D,
-	\end{align*}
-	where 
-	$$
-	\Vm^{(l)} = \diag \[ \sqrt{\frac{\delta_d^{(l)}(\theta_d^{(l)})^2}{\lambda^{(l)}}} \frac{K_{\frac{p - 1}{2}} \( \sqrt{\lambda^{(l)} \delta_d^{(l)}}/\theta^{(l)}_d \)}{K_{\frac{p + 1}{2}} \( \sqrt{\lambda^{(l)} \delta_d^{(l)}}/\theta^{(l)}_d \)} \].
-	$$
-
-	- Think about parametrisation of $\mathbf{C}$.
-Idea: use overall mean and deviations from it in alpha, such that the starting value can be the mean and 0 for alpha.
-Or: use intercept and code the rest as dummy variables.
-
-	\subsubsection{Tissue effect covariance}
-	The third sub-problem pertains to the tissue effect covariance prior.
-To avoid overfitting when freely estimating the $\frac{T(T+1)}{2}$ elements in $\bm{\Omega}$ and $\nu$, we parametrise them as $\bm{\Omega}=  \tau [(\tau - \rho) \I_T + \rho \mathbf{1}_{T \times T}]$ and $\nu = \tau + T + 1$, respectively.
-The justification for this parametrisation is three-fold: (a) it implies an easy to interpret prior covariance matrix with $\tau$ on its diagonal and $\rho$ on its off-diagonals, (b) it reduces the number of parameters to  estimate to just two, and (c) simplifies the positive definite constraint on  $\bm{\Omega}$.
-To ensure positive definiteness and a proper covariance matrix, we impose the constraints $\tau > \max\[\rho - T \rho,\rho, 0\]$ (see Appendix \ref{app:priortissuecovariance}).
-The resulting third sub-problem is a bit more involved than the first two:
-	\begin{align*}
-	\tau^{(l + 1)}, \rho^{(l+1)} & = \underset{\tau, \rho}{\argmax} \frac{D (T^2 + T)}{2} \log \tau + \frac{DT}{2} \tau \log \tau + \frac{D(T+1)}{2} \log \[ \tau - (T - 1) \rho\] \\
-	& \,\,\,\,\,\,\,\,\,\, + \frac{D}{2} \tau \log \[ \tau - (T - 1) \rho\] + \frac{D(T^2 - 1)}{2} \log(\tau - \rho) +\frac{D(T - 1)}{2} \tau  \log(\tau - \rho) \\
-	& \,\,\,\,\,\,\,\,\,\, - D \sum_{t=1}^T \log \Gamma \( \frac{\tau + t + 1}{2} \) - w_1^{(l)} \tau^2 + w_2^{(l)} \tau \rho + w_3^{(l)} \tau, \\
-	& \text{subject to } \tau > \max\[\rho - T \rho,\rho, 0\],
-	\end{align*}
-	where
-	\begin{align*}
-	w_1^{(l)} & =  \frac{\tau^{(l)} + T + 2}{2} \sum_{d=1}^D \trace \[ (\bm{\Psi}_d^{(l)})^{-1}\] > 0,\\
-	w_2^{(l)} & = \frac{\tau^{(l)} + T + 2}{2} \sum_{d=1}^D \left\{\trace \[ (\bm{\Psi}_d^{(l)})^{-1}\] - \trace \[\mathbf{1}_{T \times T} \cdot (\bm{\Psi}_d^{(l)})^{-1}\] \right\} < w_1^{(l)},\\
-	w_3^{(l)} & = \frac{1}{2} \[D \sum_{t=1}^T \psi\(\frac{\tau^{(l)} + t + 2}{2}\) - \sum_{d=1}^D \log |\bm{\Psi}_d^{(l)}| \].
-	\end{align*}
-
-	- use mathematica to check convexity
-
-	- we use the ELBO for VB convergence and parameters for EB convergence
-
-	- we might be lenient for VB convergence. The extreme case would be 1 VB iteration per EB iteration (corresponds to tolerance of infinity).
-
-	\section{Simulations}
-
-	- Hamming distance to assess variable selection
-
-	\section{Data application}
-	We predict IC50 of the drugs.
-Other choices of efficacy measures are possible, but a discussion on the pros and cons of these different choices is not the aim of this paper.
-In any case, the presented methods may be used with any continuous efficay measure.
-
-	- check GSK open target EBI or ask Chris Wallace about it. Database that links targets to drugs and diseases. Might be useful for drugs co-data.
-
-	- check NCI-60
-
-	- check CTRP study
-
-	- check data in PharamcoGx R-package
-
-	- how to transform the data and deal with missingness (ask Oslo and read papers published on GDSC and CCLE repositories)
-
-	- create drug covariates from SMILES formulas with R-package rcdk.
-
-	- try non-negative matrix factorisation/independent component analysis on binary fingerprints to get principal components that characterise the different drugs.
-
-	- check whether compounds are PAINS \cite[]{baell_new_2010} or not. PAINS indicates whether compound is frequent false hitter.
-
-	- start with gene expression (one $g(j)$)
+	\section{Data}
 
 	\subsection{GDSC}
-	After iteratively removing the cell line or drug with the highest proportion of missings until there are no more missing values, we end up with  IC50 estimates for 495 cell lines on 167 drugs.
+	The GDSC project's \cite[]{yang_genomics_2013} aim is "to improve cancer 
+	treatments by discovering therapeutic biomarkers that can be used to identify 
+	patients most likely to respond to anticancer drug". Part of the project is to
+	screen $>1000$ human cancer cell lines with drugs. The cell lines have been
+	genetically characterised and several drug sensitivity measures are recorded.
+	The data is freely available from \cite{garnett_systematic_2012} and consist 
+	of: (i) the sensitivity measures 
+	(mainly half maximal inhibitory concentration, i.e., IC50) of the cell lines 
+	to the drugs, (ii) annotation of the 
+	screened compounds, and (iii) the cell lines' genomic profile (mutations, 
+	copy numbers, methylation profiles, and expression). We will attempt to 
+	predict drug sensitivity of the cell lines, as quantified by IC50, using the
+	genomic data. Other choices of efficacy measures than IC50 are possible, 
+	but a discussion on the pros and cons of these different choices is not the 
+	aim of this paper.
+
+	\subsubsection{Primary data}
+	We took the logarithm of the IC50 values and took the average if there were
+	multiple measurements per cell line-drug combination. In the following, if we 
+	refer to IC50, it denotes the log-transformed values. After iteratively 
+	removing the cell line or drug with the highest proportion 
+	of missings until there are no more missing values, we end up with IC50 
+	estimates for 482 cell lines on 168 drugs. For the gene expression 
+	data, we fit a lasso with IC50 as response to the drugs separately and 
+	adjusted the penalty parameter to select approximately 300 genes per drug.
+
+	- From Oslo:
+	1.2) Preselecting GEX features: we selected gene expression features
+	explaining 50\% variations over all cell lines, since this results in not too
+	features and probably remains those important features.
+	1.3) Preselecting CNV: we used the same strategy as Garnett et al. (2012)
+	where the GDSC data are from.
+	1.4) Preselecting MUT: we deleted mutated genes with less than two cell lines.
+
+	- check out selection method in \cite{zhao_structured_2019}.
+
+	- how to transform the data and deal with missingness? Read GDSC papers.
+
+	\subsubsection{External data}
+	Several external drug variables are available: (i) the developmental stage
+	(experimental, in clinical development, or clinically approved) and (ii)
+	the action (cytotoxic or targeted). We also have a external variable 
+	available that relates the drugs to the features: a binary indicator whether 
+	a gene is in the target pathway of the drug or not.
+
+  - check whether compounds are PAINS \cite[]{baell_new_2010} or not. 
+  PAINS indicates whether compound is frequent false hitter.
+
+  - clinical trials database CTRP might contain info
+
+  - extra data on GDSC might be in PharamcoGx R-package. Also contains bunch 
+  of functions.
+
+  - check GSK open target EBI or ask Chris Wallace about it. Database that links
+	targets to drugs and diseases. Might be useful for drugs co-data.
+
+	- try non-negative matrix factorisation/independent component analysis on 
+	binary fingerprints to get principal components that characterise the 
+	different drugs.
 
 	\subsection{CCLE}
+	- extra data on CCLE might be in PharamcoGx R-package
+
+	\subsection{NCI-50}
+
+	\subsection{CMap}
+	- Cmap \cite[]{subramanian_next_2017,lamb_connectivity_2006} contains 
+	pharma-genomic data (similar to CCLE and GDSC).
+
+	- Data available at: 
+	https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE92742.
+
+	- Cmap website: https://clue.io/cmap.
+
+	\section{Results}
+	\subsection{GDSC}
 
 	\section{Discussion}
-	- use copula's and marginal posteriors to find multivariate posteriors for contrast testing.
+	- use copula's and marginal posteriors to find multivariate posteriors for 
+	contrast testing.
 
-	- empirical Bayes gives extra information via estimation of the hyper-parameters, such as informativeness of the drug information (via model for the $\gamma_d$) and relatedness of the tissues (via $\bm{\Omega}$ and $\nu$)
+	- empirical Bayes gives extra information via estimation of the 
+	hyperparameters, such as informativeness of the drug information (via model 
+	for the $\gamma_d$) and relatedness of the tissues (via $\bm{\Omega}$ and 
+	$\nu$)
 
@@
 
-	- find more literature on inverse Gaussian prior for variances and modelling variances with covariates.
+	\section*{Supplementary Material}
+	Supplementary Material is available online from 
+	\url{https://github.com/magnusmunch/cambridge}.
 
-	Recently, \cite{moran_variance_2018} have argued that in high-dimensional regression problems, \text{a priori} dependence between the regression parameters and error variance may induce biased estimation.
-In (\ref{eq:betaprior}), we assume such an \textit{a priori} dependence.
-However, we contend that independent regression parameters and error variance are not desirable in our setting for two reasons: (a) the empirical Bayes estimation requires some notion of scale to combine the separate regressions, which may be done by inclusion of the error variance in the regression parameter prior model.
-(b) The variational Bayes approximation introduced in Section \ref{sec:variationalbayes} gives a posterior that is similar to the one we would obtain with a dependent regression parameters and error variance prior model.
-This renders a better approximate expectation in the empirical Bayes iterations.
+	\section*{Reproducible Research}
+	All results and documents may be recreated from 
+	\url{https://github.com/magnusmunch/cambridge}.
+
+	\section*{Acknowledgements}
+	We thank Zhi Zhao and Leiv Tore Salte R{\o}nneberg from the Oslo Centre for 
+	Biostatistics and Epidemiology for help with the pre-processing of the data. 
+	\textit{Conflict of Interest}: None declared.
 
@@
 
-	\titleformat{\section}{\normalfont\Large\bfseries}{\appendixname~\thesection.}{1em}{}
-	\begin{appendix}
-		\section{Prior tissue covariance constraints}\label{app:priortissuecovariance}
-		We parametrise the prior covariance as a matrix with $\tau$ on the diagonal and $\rho$ on its off-diagonals. To that end we set $\nu = \tau + T + 1$ and the $T \times T$-dimensional matrix
-		$$
-		\bm{\Omega} = \tau \begin{bmatrix}
-		\tau &  & \rho \\
-		& \ddots &  \\
-		\rho & & \tau
-		\end{bmatrix},
-		$$
-		such that $\E_{\pi(\mathbf{\Xi}_d)} (\mathbf{\Xi}_d) = (\nu - T - 1)^{-1} \bm{\Omega}$ is of the desired form.
-To ensure a non-degenerate prior distribution we require $\tau > 0$ and $\bm{\Omega}$ PD.
-We use $\tau > 0$ to note that a PD $\bm{\Omega}$ implies positive roots $\lambda$ of the characteristic polynomial:
-		\begin{align*}
-		\left| \begin{bmatrix}
-		\tau &  & \rho \\
-		& \ddots &  \\
-		\rho & & \tau
-		\end{bmatrix} -\lambda \I_T \right| = (\tau - \lambda - \rho)^{T-1}[\tau - \lambda - (1-T)\rho] = 0.
-		\end{align*}
-		For $T > 1$, the solutions are:
-		$$
-		(\lambda = \tau - \rho) \lor (\lambda = T \rho - \rho + \tau),
-		$$
-		which are positive if the following condition holds: $\tau > \max(\rho,\rho - T \rho)$.
-We combine this with $\tau > 0$, to arrive at the full proper prior constraint: $\tau > \max(0,\rho,\rho - T \rho)$.
-
-	\end{appendix}
-
 \end{document}

### 2019-08-13 docs/supplement1.Rnw
updated simulations and manuscript

--- 2019-05-21 docs/supplement1.Rnw
@@
 
-	\section{NIG prior}\label{sec:nigprior}
-	For a fixed error variance $\sigma_d^2$, we have a marginal prior variance 
+	\section{NIG and Student's \texorpdfstring{$t$}{t} Priors}\label{sec:priors}
+	\subsection{Normal inverse Gaussian}\label{sec:nigprior}
+	Let $\theta = (\mathbf{c} \tr \bm{\alpha})^{-1}$. Then, 
+	for a fixed error variance $\sigma^2$, we have marginal prior variance 
 	and kurtosis:
 	\begin{align*}
-	  \V(\beta_{jd} | \sigma_d^2) & = \sigma_d^2 \theta_d, \\
-	  \mathcal{K}(\beta_{jd} | \sigma_d^2) & = \frac{3\theta_d}{\lambda_d} + 3.
+	  \V(\beta | \sigma^2) & = \sigma^2 \theta, \\
+	  \mathcal{K}(\beta | \sigma^2) & = \frac{3\theta}{\lambda} + 3.
 	\end{align*}
 
-	\section{NIGIG prior}\label{sec:nigigprior}
-	Equivalently, we may set 
-  $\eta_{jd}^2=\tau_j^2 \gamma_d^2/(\theta_j^z \theta_d^c)$ and write 
-  MD equation (\ref{md-eq:generalmodel}) as:
-  \begin{align*}
-    \beta_{jd} | \eta_{jd}^2, \sigma_d^2 & \sim \mathcal{N} 
-    \(0, \frac{\lambda_{jd}}{\chi_{jd}}\eta_{jd}^2\sigma_d^2\), \\
-    \eta_{jd}^2 & \overset{D}{=}\pi^{-1}\sqrt{\chi_{jd}}\exp(\psi_{jd}) 
-    \eta_{jd}^{-3/2} \cdot K_{0} \(\sqrt{\psi_{jd}^2 + \chi_{jd} (\eta_{jd}^2 
-    + \eta_{jd}^{-2} - 2)} \),
-  \end{align*}
-  where $\lambda_{jd}=\lambda_j^z\lambda_d^c$, 
-  $\chi_{jd}=\lambda_j^z\lambda_d^c/(\theta_j^z \theta_d^c)$, and
-  $\psi_{jd}=\lambda_j^z/\theta_j^z + \lambda_d^c/ \theta_d^c$. A third way
-  of writing the model is:
-  \begin{subequations}\label{eq:generalprior}
-    \begin{align}
-      \beta_{jd} | \eta_{jd}^2, \sigma_d^2 & \sim \mathcal{N} 
-      \(0, \eta_{jd}^2 \sigma_d^2\), \\
-      \eta_{jd}^2 | \xi_{d}^2 & \sim \mathcal{IG} 
-      (\theta_j^z \xi_d^2, \lambda_j^z \xi_d^2 ), \\
-      \xi_d^2 & \sim \mathcal{GIG} (1, \lambda_d^c/(\theta_d^c)^2, \lambda_d^c).
-    \end{align}
-  \end{subequations}
-	From (\ref{eq:generalprior}), for a fixed error variance
-	$\sigma_d^2$, we find the marginal prior variance and kurtosis:
-	\begin{subequations}\label{eq:varkurtnigig}
-	  \begin{align}
-	    \V(\beta_{jd} | \sigma_d^2) & = \sigma_d^2 \theta_d^c \theta_j^z  
-      \(r_d + \frac{2\theta_d^c}{\lambda_d^c}\), \\
-	    \mathcal{K}(\beta_{jd} | \sigma_d^2) & = \(\frac{3\theta_j^z}
-	    {\lambda_j^z} + 3\) \frac{4 \theta_d^c (r_d + 2\theta_d^c/\lambda_d^c) +
-	    1}{\lambda_d^c(r_d + 2\theta_d^c/\lambda_d^c)^2},
-	  \end{align}
-	\end{subequations}
-	with $r_d = K_0(\lambda_d^c/\theta_d^c)/K_1(\lambda_d^c/\theta_d^c)$.
-	If 
-  we fix the prior expectation $\E(\gamma_d^2)=\theta_d^c$ and let the prior 
-  variance 
-  $\V(\gamma_d^2)$ go to zero, we have $\V(\beta_{jd} | \sigma_d^2) \rightarrow
-  \sigma_d^2 \theta_j^z \theta_d^c$ and $\mathcal{K}(\beta_{jd} | \sigma_d^2)
-  \rightarrow 3(\theta_j^z/\lambda_j^z + 1)$. From (\ref{eq:varkurtnigig}) we see
-  that we can keep the variance constant by fixing $\theta_j^z$, $\theta_d^c$,
-  and $\lambda_d^c$, while we increasing the kurtosis by decrearing 
-  $\lambda_j^z$. This makes the model MD equations (\ref{md-eq:generalmodel}) a
-  flexible model.
-
-	\section{Student's \texorpdfstring{$t$}{t} prior}\label{sec:studentstprior}
-	For a fixed error variance $\sigma_d^2$, we have a marginal prior variance 
+	\subsection{Student's \texorpdfstring{$t$}{t} prior}\label{sec:studentstprior}
+	For a fixed error variance $\sigma^2$, we have marginal prior variance 
 	and kurtosis:
 	\begin{align*}
-	  \V(\beta_{jd} | \sigma_d^2) & = \frac{\sigma_d^2 \lambda_d}{\eta_d - 2}, \\
-	  \mathcal{K}(\beta_{jd} | \sigma_d^2) & = \frac{6}{\eta_d - 4} + 3.
+	  \V(\beta | \sigma^2) & = \frac{\sigma^2 \xi}{\eta - 2}, \\
+	  \mathcal{K}(\beta | \sigma^2) & = \frac{6}{\eta - 4} + 3.
 	\end{align*}
@@
 	  \log q_{\bbeta_d} (\bbeta_d) & \propto \E [\log \mathcal{L}
-	  (\y_d | \bbeta_d, \sigma_d^2 )] + \E [\log \pi (\bbeta_d | \gamma_d^2, 
+	  (\y_d | \bbeta_d, \sigma_d^2 )] + \E [\log \pi (\bbeta_d | \bgamma_{d}^2, 
 	  \sigma_d^2)] \\
@@
 	  \frac{1}{2} \sum_{j=1}^p 
-	  \E \( \frac{\beta_{jd}^2}{\gamma_d^2 \sigma_d^2} \), \\
+	  \E \( \frac{\beta_{jd}^2}{\gamma_{jd}^2 \sigma_d^2} \), \\
 	  q_{\bbeta_d} (\bbeta_d) & \overset{D}{=} 
@@
 	  & \text{with } \bSigma_d = \E(\sigma_d^{-2})^{-1} 
-	  [\X \tr \X + \E(\gamma_d^{-2}) \cdot \I ]^{-1}, \\
-	  & \text{and } \bmu_d = [\X \tr \X + \E(\gamma_d^{-2}) \cdot \I ]^{-1} 
+	  \{\X \tr \X + \diag [\E(\gamma_{jd}^{-2})] \}^{-1}, \\
+	  & \text{and } \bmu_d = \{\X \tr \X + \diag [\E(\gamma_{jd}^{-2})] \}^{-1} 
 	  \X \tr \y_d.
 	\end{align*}
-	The variational posterior for $\gamma_d^2$ is given by:
+	The variational posterior for $\gamma_{jd}^2$ is given by:
 	\begin{align*}
-	  \log q_{\gamma_d^{2}}(\gamma_d^{2}) & \propto \E 
-	  [\log \pi (\bbeta_d | \gamma_d^{2}, \sigma_d^2)] + \log \pi(\gamma_d^{2}) \\
-	  & \propto -\frac{p}{2} \log \gamma_d^{2} - \frac{1}{2} \sum_{j=1}^p 
-	  \E \( \frac{\beta_{jd}^2}{\sigma^{2}_d}\) \gamma_d^{-2} - 
-	  \frac{3}{2} \log \gamma_d^{2} - 
-	  \frac{\lambda_d (\gamma_d^2 - \theta_d)^2}{2 \theta_d^2} \gamma_d^{-2} \\
-	  & \propto \( -\frac{p + 1}{2} - 1\) \log \gamma_d^{2} - 
-	  \frac{\lambda_d}{2\theta_d^2} \gamma_d^2 - 
-	  \frac{1}{2}\[ \lambda_d + \E(\sigma_d^{-2}) \sum_{j=1}^p \E(\beta_{jd}^2)\] 
-	  \gamma_d^{-2} , \\
-	  q_{\gamma_d^{2}}(\gamma_d^{2}) & \overset{D}{=} \mathcal{GIG} 
-	  \(-\frac{p+1}{2}, \frac{\lambda_d}{\theta_d^2}, \delta_{d} \), \\
-	  & \text{with } \delta_{d} =\E(\sigma_d^{-2})
-	  \left\{\E(\bbeta_{d} \tr) \E(\bbeta_{d}) + \trace[\V(\bbeta_{d})]\right\} + 
-	  \lambda_d.
+	  \log q_{\gamma_{jd}^{2}}(\gamma_{jd}^{2}) & \propto \E 
+	  [\log \pi (\beta_{jd} | \gamma_{jd}^{2}, \sigma_d^2)] + 
+	  \log \pi(\gamma_{jd}^{2}) \\
+	  & \propto -\frac{1}{2} \log \gamma_{jd}^{2} - \frac{1}{2} 
+	  \E \( \frac{\beta_{jd}^2}{\sigma^{2}_d}\) \gamma_{jd}^{-2} - 
+	  \frac{3}{2} \log \gamma_{jd}^{2} - 
+	  \frac{\lambda_d \{\gamma_{jd}^2 - 
+	  [(\mathbf{c}_{d}^j) \tr \bm{\alpha}]^{-1}\}^2}
+	  {2[(\mathbf{c}_{d}^j) \tr \bm{\alpha}]^{-2}} \gamma_{jd}^{-2} \\
+	  & \propto \( -1 - 1\) \log \gamma_{jd}^{2} - 
+	  \frac{\lambda_d[(\mathbf{c}_{d}^j) \tr \bm{\alpha}]^{2}}{2} \gamma_{jd}^2 - 
+	  \frac{1}{2}\[ \lambda_d + \E(\sigma_d^{-2}) \E(\beta_{jd}^2)\] 
+	  \gamma_{jd}^{-2} , \\
+	  q_{\gamma_{jd}^{2}}(\gamma_{jd}^{2}) & \overset{D}{=} \mathcal{GIG} 
+	  \(-1, \lambda_d[(\mathbf{c}_{d}^j) \tr \bm{\alpha}]^{2}, \delta_{jd} \), \\
+	  & \text{with } \delta_{jd} =\E(\sigma_d^{-2})
+	  \[\E(\beta_{jd})^2 + \V(\beta_{jd})\] + \lambda_d.
 	\end{align*}
@@
 	  [\log \mathcal{L}(\y_d | \bbeta)d, \sigma_d^2 )] + 
-	  \E [\log \pi (\bbeta_d | \gamma_d^2, \sigma_d^2)] + 
+	  \E [\log \pi (\bbeta_d | \bgamma_d^2, \sigma_d^2)] + 
 	  \log \pi (\sigma_d^{2}) \\
@@
 	  \frac{p}{2} \log \sigma_d^{2} - \frac{1}{2} \sum_{j=1}^p 
-	  \E \( \frac{\beta_{jd}^2)}{\gamma_d^{-2}}\) \sigma_d^{-2} - 
+	  \E \( \frac{\beta_{jd}^2)}{\gamma_{jd}^{2}}\) \sigma_d^{-2} - 
 	  \frac{3}{2} \log \sigma^2_d \\
@@
 	  \frac{1}{2}\left\{ \sum_{i=1}^n \E [(y_{id} - \x_i \tr \bbeta_d)^2] + 
-	  \E(\gamma_d^{-2})\sum_{j=1}^p \E(\beta_{jd}^2) \right\} \sigma_d^{-2}, \\
+	  \sum_{j=1}^p \E(\gamma_{jd}^{-2})\E(\beta_{jd}^2) \right\} \sigma_d^{-2}, \\
 	  q_{\sigma_d^{2}}(\sigma_d^{2}) & \overset{D}{=} 
 	  \Gamma^{-1} \(\frac{n + p + 1}{2}, \zeta_{d} \), \\
-	  & \text{with } \zeta_{d} = \frac{1}{2} \bigg\{\mathbf{y}_d \tr \mathbf{y}_d -
-    2 \mathbf{y}_d \tr \X \E(\bbeta_d) + 
+	  & \text{with } \zeta_{d} = \frac{1}{2} \bigg\{\mathbf{y}_d \tr 
+	  \mathbf{y}_d - 2 \mathbf{y}_d \tr \X \E(\bbeta_d) + 
     \trace [ \X \tr \X \V(\bbeta_d)] + 
@@
     & \,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\, + 
-    \E(\gamma_d^{-2}) \trace [ \V(\bbeta_d)] + 
-    \E(\gamma_d^{-2}) \E(\bbeta_d \tr) \E(\bbeta_d ) \bigg\}.
+    \trace \{ \diag[\E(\gamma_{jd}^{-2})] \V(\bbeta_d)\} + 
+     \E(\bbeta_d \tr) \diag[\E(\gamma_{jd}^{-2})] \E(\bbeta_d ) \bigg\}.
 	\end{align*}
 
-	\section{Evidence lower bound}\label{sec:evidencelowerbound}
-	The variational Bayes posterior is found through maximisation of an evidence
-	lower bound (ELBO). Additionally, we may monitor convergence through this 
-	ELBO. To compute the ELBO after updating the EB parameters from 
-	$\theta_d^{(l)}, \lambda_d^{(l)}$ to 
-	$\theta_d^{(l + 1)}, \lambda_d^{(l + 1)}$, we compute 
-	$\text{ELBO}^{(l + 1)} = \sum_{d=1}^D \text{ELBO}_d^{(l + 1)}$, with:
-	\begin{align*}
-	  \text{ELBO}_d^{(l + 1)} = & \frac{1}{2} \log |\bm{\Sigma}_d^{(l)}| - 
-	  \frac{a_d^{(l)}}{2} \[ \y_d \tr \y_d - 2\y_d \tr \X \bm{\mu}_d^{(l)} + 
-	  (\bm{\mu}_d^{(l)}) \tr \X \tr \X \bm{\mu}_d^{(l)} + 
-	  \trace ( \X \tr \X \bm{\Sigma}_d^{(l)})\] \\
-	  & + \frac{b_d^{(l)}}{2} \[ \delta_d^{(l)} - 
-	  a_d^{(l)}(\bm{\mu}_d^{(l)}) \tr \bm{\mu}_d^{(l)} - 
-	  a_d^{(l)} \trace (\bm{\Sigma}_d^{(l)}) \] \\
-	  & - \frac{p + n + 1}{2} \log \zeta_d^{(l)} + 
-	  \frac{p + 1}{4} \log \lambda_d^{(l)} - 
-	  \frac{p + 1}{2} \log \theta_d^{(l)} - \frac{p+1}{4} \log \delta_d^{(l)} \\
-	  & + \log K_{\frac{p+1}{2}}\(\sqrt{\lambda_d^{(l)} 
-	  \delta_d^{(l)}}/\theta_d^{(l)}\) + 
-	  \frac{\lambda_d^{(l)} e_d^{(l)}}{2(\theta_d^{(l)})^2} \\
-	  & + \frac{\lambda_d^{(l+1)}}{\theta_d^{(l+1)}} + 
-	  \frac{1}{2}\log \lambda_d^{(l+1)} - \frac{\lambda_d^{(l+1)} 
-	  e_d^{(l)}}{2 (\theta_d^{(l + 1)})^2} - 
-	  \frac{\lambda_d^{(l + 1)} b_d^{(l)}}{2},
-	\end{align*}
-	where $a^{(l)}$, $b^{(l)}$, and $e^{(l)}$ are as in the MD. To compute the 
-	ELBO after a VB update we simply let 
-  $\theta_d^{(l)}, \lambda_d^{(l)}=\theta_d^{(l + 1)}, \lambda_d^{(l + 1)}$.
-
-  In practice, there are different options to assess convergence. We may monitor
-  the parameters themselves, the ELBO, or a combination of the two. We may also 
-  iterate for an \textit{a priori} fixed number of times. We may also choose to 
-  use a different convergence criterium for the EB and VB iterations. Currently 
-  we fix the number of EB iterations to 20, while we do 2 VB iterations per EB 
-  iteration.
+% 	\section{Evidence lower bound}\label{sec:evidencelowerbound}
+% 	The variational Bayes posterior is found through maximisation of an evidence
+% 	lower bound (ELBO). Additionally, we may monitor convergence through this 
+% 	ELBO. To compute the ELBO after updating the EB parameters from 
+% 	$\theta_d^{(l)}, \lambda_d^{(l)}$ to 
+% 	$\theta_d^{(l + 1)}, \lambda_d^{(l + 1)}$, we compute 
+% 	$\text{ELBO}^{(l + 1)} = \sum_{d=1}^D \text{ELBO}_d^{(l + 1)}$, with:
+% 	\begin{align*}
+% 	  \text{ELBO}_d^{(l + 1)} = & \frac{1}{2} \log |\bm{\Sigma}_d^{(l)}| - 
+% 	  \frac{a_d^{(l)}}{2} \[ \y_d \tr \y_d - 2\y_d \tr \X \bm{\mu}_d^{(l)} + 
+% 	  (\bm{\mu}_d^{(l)}) \tr \X \tr \X \bm{\mu}_d^{(l)} + 
+% 	  \trace ( \X \tr \X \bm{\Sigma}_d^{(l)})\] \\
+% 	  & + \frac{b_d^{(l)}}{2} \[ \delta_d^{(l)} - 
+% 	  a_d^{(l)}(\bm{\mu}_d^{(l)}) \tr \bm{\mu}_d^{(l)} - 
+% 	  a_d^{(l)} \trace (\bm{\Sigma}_d^{(l)}) \] \\
+% 	  & - \frac{p + n + 1}{2} \log \zeta_d^{(l)} + 
+% 	  \frac{p + 1}{4} \log \lambda_d^{(l)} - 
+% 	  \frac{p + 1}{2} \log \theta_d^{(l)} - \frac{p+1}{4} \log \delta_d^{(l)} \\
+% 	  & + \log K_{\frac{p+1}{2}}\(\sqrt{\lambda_d^{(l)} 
+% 	  \delta_d^{(l)}}/\theta_d^{(l)}\) + 
+% 	  \frac{\lambda_d^{(l)} e_d^{(l)}}{2(\theta_d^{(l)})^2} \\
+% 	  & + \frac{\lambda_d^{(l+1)}}{\theta_d^{(l+1)}} + 
+% 	  \frac{1}{2}\log \lambda_d^{(l+1)} - \frac{\lambda_d^{(l+1)} 
+% 	  e_d^{(l)}}{2 (\theta_d^{(l + 1)})^2} - 
+% 	  \frac{\lambda_d^{(l + 1)} b_d^{(l)}}{2},
+% 	\end{align*}
+% 	where $a^{(l)}$, $b^{(l)}$, and $e^{(l)}$ are as in the MD. To compute the 
+% 	ELBO after a VB update we simply let 
+%   $\theta_d^{(l)}, \lambda_d^{(l)}=\theta_d^{(l + 1)}, \lambda_d^{(l + 1)}$.
+% 	 
+%   In practice, there are different options to assess convergence. We may monitor
+%   the parameters themselves, the ELBO, or a combination of the two. We may also 
+%   iterate for an \textit{a priori} fixed number of times. We may also choose to 
+%   use a different convergence criterium for the EB and VB iterations. Currently 
+%   we fix the number of EB iterations to 20, while we do 2 VB iterations per EB 
+%   iteration.
 
   \section{Efficient computation}
-  The empirical Bayes updates require the following quantities: $\zeta_d$, $a_d$, $\delta_d$, $\trace (\bm{\Sigma}_d)$, $\trace (\X \tr \X \bm{\Sigma}_d)$, $\bm{\mu}_d \tr \bm{\mu}_d$, and $\bm{\mu}_d \tr \X \tr \X \bm{\mu}_d$.
-In addition, we need $\log|\bm{\Sigma}|$ and $\mathbf{y}_d \tr \X \bm{\mu}_d$ to monitor the ELBO.
-The first three quantities are obtained by scalar operations of $\mathcal{O}(n)$.
-Let $c_d^{(h)}=b_d^{(h)}$ in the conjugate setting and $c_d^{(h)}=b_d^{(h)}/a_d^{(h)}$ in the non-conjugate setting.
-Then, beforementioned quantities are easily calculated using the SVD $\X = \mathbf{U} \mathbf{D} \mathbf{V} \tr$:
-  \begin{align*}
-  \trace (\bm{\Sigma}_d^{(h+1)}) & = (a_d^{(h)})^{-1} \[\sum_{i=1}^n (v_i^2 + c_d^{(h)})^{-1} + \max (p - n, 0)\cdot (c_d^{(h)})^{-1}\], \\
-  \trace (\X \tr \X \bm{\Sigma}_d^{(h+1)}) & = (a_d^{(h)})^{-1} \sum_{i=1}^n v_i^2(v_i^2 + c_d^{(h)})^{-1}, \\
-  (\bm{\mu}_d^{(h+1)}) \tr \bm{\mu}_d^{(h+1)} & = \sum_{i=1}^n v_i^2 [(\mathbf{U} \tr \mathbf{y}_d)_i]^2/(v_i^2 + c_d^{(h)})^2, \\
-  (\bm{\mu}_d^{(h+1)}) \tr \X \tr \X \bm{\mu}_d^{(h+1)} & = \sum_{i=1}^n v_i^4 [(\mathbf{U} \tr \mathbf{y}_d)_i]^2/(v_i^2 + c_d^{(h)})^2, \\
-  \log|\bm{\Sigma}^{(h+1)}| & \propto -p \log a_d^{(h)} - \sum_{i=1} \log (v_i^2 + c_d^{(h)}) - \max(p - n, 0) \cdot \log c_d^{(h)}, \\
-  \mathbf{y}_d \tr \X \bm{\mu}_d^{(h+1)} & = \sum_{i=1}^n v_i^2 [(\mathbf{U} \tr \mathbf{y}_d)_i]^2/(v_i^2 + c_d^{(h)}).
-  \end{align*}
-  where $v_i$ are the singular values of $\X$.
-The SVD and $\mathbf{U} \tr \mathbf{y}_d$ are $\mathcal{O}(pn^2)$ and $\mathcal{O}(n^2)$ operations, respectively, but have to be calculated only once at the start of the algorithm.
-The rest of the calculations involve just $n$ scalar multiplications and/or additions.
+  The empirical Bayes updates require the quantities
+  $e_{jd}$ and $b_{jd}$, which in turn depend on the quantities 
+  $\diag(\bSigma_d)$, $\bmu_{d}$, $\y_d \tr \X \bmu_d$, 
+  $\trace(\X \tr \X \bSigma_d)$, $\bmu_d \tr \X \tr \X \bmu_d$,
+  $\trace[\diag(b_{jd})\bSigma_d]$, and $\bmu_d \tr \diag(b_{jd}) \bmu_d$.
+  Let $\mathbf{H}_d=\diag[(b_{jd}^{(h)})^{-1}]$. Then, the first two quantities 
+  are more efficiently calculated as:
+  \begin{subequations}\label{eq:complexity1}
+    \begin{align}
+      \diag (\bm{\Sigma}_d^{(h+1)}) & = (a_d^{(h)})^{-1} 
+      \H_d^{(h)} - (a_d^{(h)})^{-1} \{ [ \H_d^{(h)} \X \tr 
+      ( \I_n + \X \H_d^{(h)} \X \tr )^{-1} ] \circ
+      (\H_d^{(h)} \X \tr ) \} \mathbf{1}_{n \times 1}, \\
+      \bmu_d^{(h+1)}) & = \H_d^{(h)} \X \tr \y_d - \H_d^{(h)} \X \tr 
+      ( \I_n + \X \H_d^{(h)} \X \tr )^{-1} \X \H_d^{(h)} \X \tr \y_d.
+    \end{align}
+  \end{subequations}
+  The quantities $\y_d \tr \X \bmu_d$, $\bmu_d \tr \X \tr \X \bmu_d$,
+  $\trace[\diag(b_{jd})\bSigma_d]$, and $\bmu_d \tr \diag(b_{jd}) \bmu_d$ are
+  easily calculated from the first two. For the remaining one we have:
+  \begin{align}\label{eq:complexity2}
+    \trace(\X \tr \X \bSigma_d^{(h+1)})  = 
+    & (a_d^{(h)})^{-1} \mathbf{1}_{1 \times p} \H_d^{(h)} \X \tr 
+    \mathbf{1}_{n \times 1} \nonumber \\  
+    & - (a_d^{(h)})^{-1} \mathbf{1}_{1 \times p} \{ [  \H_d^{(h)} \X \tr 
+    ( \I_n + \X \H_d^{(h)} \X \tr )^{-1} \X \H_d^{(h)} \X \tr ] \circ 
+    \X \tr \} \mathbf{1}_{n \times 1}.
+  \end{align}
+  Both (\ref{eq:complexity1}) and (\ref{eq:complexity2}) are operations of
+  complexity $\mathcal{O}(pn^2)$.
 
@@
   $\alpha$. In our case, $\alpha$ increases linearly with $p$. Since $p$ may be
-  large, this causes numerical issues in the calculation of various quantities.
-  We alleviate the numerical issues through the following.
+  large, this may cause numerical issues in the calculation of various 
+  quantities. We alleviate the numerical issues through the following.
 
@@
 
-	\section{Non-conjugate model}
-	The non-conjugate model assumes $\beta_{jd}$ and $\sigma_d^2$ 
-	independent \textit{a priori}:
-  $$
-  \beta_{jd} | \gamma_d^2, \sigma_d^2 \sim \beta_{jd} | 
-  \gamma_d^2 \sim \mathcal{N} (0, \gamma_d^2).
-  $$
+	\section{Log pseudo marginal likelihood}\label{sec:lpml}
+	The log pseudo marginal likelihood is calculated as:
+	$\text{LPML}_d = n^{-1}\sum_{i=1}^n \log (\text{CPO}_{id})$ 
+  \cite[]{gelfand_model_1992}. Let 
+  $\theta_{id}=\x_i \tr \bbeta_d$ and $\y_{-id}$ denote $\y_{d}$ with 
+  observation $y_{id}$ excluded. Then the conditional 
+  predictive ordinates (CPO) are given by:
+  \begin{align*}
+    \text{CPO}_{id} & = p(y_{id} | \y_{-id}) = 
+    \int_{-\infty}^{\infty} \int_0^{\infty}
+    p(y_{id} | \bbeta_d, \sigma_d^2 ) \pi (\bbeta_d | \y_{-id}, \sigma_d^2) 
+    \pi (\sigma_d^2 | \y_{-id}) d\sigma_d^2 d\bbeta_d \\
+    & = \int_{-\infty}^{\infty} \int_0^{\infty}
+    p(y_{id} | \theta_{id}, \sigma_d^2 ) 
+    \pi (\theta_{id} | \y_{-id}, \sigma_d^2) 
+    \pi (\sigma_d^2 | \y_{-id}) d\sigma_d^2 d\theta_{id},
+  \end{align*}
+  where
+  \begin{align}
+    y_{id} | \theta_{id}, \sigma_d^2 & \sim \mathcal{N} (\theta_{id}, 
+    \sigma_d^2), \nonumber \\
+    \theta_{id} | \y_{-id} & \sim \mathcal{N}(\x_i \tr \bmu_{-id}, 
+    \x_i \tr \bSigma_{-id} \x_i), \nonumber \\
+    \sigma_d^2 | \y_{-id} & \sim \Gamma^{-1} \( \frac{n + p}{2},
+    \zeta_{-id} \), \label{eq:gammavb}
+  \end{align}
+  where $\bmu_{-id}$, $\bSigma_{-id}$, and $\zeta_{-id}$ are the variational
+  parameters estimated from $\y_{-id}$. Plugging in the densities we see that 
+  we have:
+  \begin{subequations}
+    \begin{align}
+      \text{CPO}_{id} = & (2 \pi)^{-1} \frac{\Gamma\(\frac{n + p + 1}
+      {2}\)}{\Gamma\(\frac{n + p}{2}\)}
+      (\zeta_{-id} \x_i \tr \bSigma_{-id} \x_i)^{-1/2} \label{eq:cpo1} \\
+      & \cdot \int_{-\infty}^{\infty}
+      \( \frac{\theta^2}{2\zeta_{-id}} + 1 \)^{-\frac{n + p + 1}{2}}
+      \exp \[ - \frac{(\theta - \x_i \tr \bmu_{-id} + y_i)^2}
+      {2 \x_i \tr \bSigma_{-id} \x_i} \] d\theta, \label{eq:cpo2}
+    \end{align}
+  \end{subequations}
+  a one-dimensional (well-behaved) integral that is easily solved with some 
+  numerical integration routine.
 
-	The variational distributions are as follows:
-	\begin{align*}
-	  q(\bm{\beta}_d) & \overset{D}{=} \mathcal{N}_p 
-	  (\bm{\mu}_d, \bm{\Sigma}_d), \\
-	  q(\gamma_d^2) & \overset{D}{=} \mathcal{GIG}
-	  \(-\frac{p+1}{2}, \lambda_d \theta_d^{-2}, \delta_d\), \\
-	  q(\sigma_d^2) & \overset{D}{=} \Gamma^{-1} 
-	  \(\frac{n + 1}{2}, \zeta_d\),
-  \end{align*}
-	The variational parameters are iteratively updated by:
-  \begin{align*}
-  \bm{\Sigma}_d^{(h+1)} & = (a_d^{(h)})^{-1} (\X \tr \X + c_d^{(h)} \I)^{-1}, \\
-  \bm{\mu}_d^{(h+1)} & = (\X \tr \X + c_d^{(h)} \I)^{-1} \X \tr \y_d, \\
-  \delta_d^{(h+1)} & = b_d^{(h)} (c_d^{(h)})^{-1} \bmu_d \tr \bmu_d + \lambda_d, \\ 
-  \zeta_d^{(h+1)} & = \frac{1}{2} \bigg[\mathbf{y}_d \tr \mathbf{y}_d -2 \mathbf{y}_d \tr \X \bm{\mu}_d^{(h+1)} + \trace ( \X \tr \X \bm{\Sigma}_d^{(h+1)}) + (\bm{\mu}_d^{(h+1)}) \tr \X \tr \X \bm{\mu}_d^{(h+1)} \\
-  & \,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\, + \frac{b_d^{(h+1)} - c_d^{(h+1)} a_d^{(h+1)}}{1 - a_d^{(h+1)}} \trace ( \bm{\Sigma}_d^{(h+1)}) + \frac{b_d^{(h+1)} - c_d^{(h+1)} a_d^{(h+1)}}{1 - a_d^{(h+1)}} (\bm{\mu}_d^{(h+1)}) \tr \bm{\mu}_d^{(h+1)}\bigg].
-  \end{align*}
-  Here, we set $a_d^{(h)}$, $b_d^{(h)}$, and $c_d^{(h)}$ as in 
-  (\ref{eq:auxvar1})
-  Note that in the inverse Gamma model, the computation of (\ref{eq:auxvar1})
-  simplifies to $b_d^{(h)}=(p+\eta_d)/\delta_d^{(h)}$.
+  In practice we would like to avoid re-estimating the model for every 
+  observation $i$. Instead we divide the observations into $K$ parts and 
+  estimate $K$ models, so that now $\y_{-id}$ denotes $\y$ with all observations
+  in part $k$ removed. Additionally, we add $1 - n_{id}$ to the numerator of the
+  Gamma distribution in (\ref{eq:gammavb}), the numerators in the Gamma 
+  functions in (\ref{eq:cpo1}), and the numerator of the power in
+  (\ref{eq:cpo2}), where $n_{id}$ denotes the number of elements in the removed 
+  part.
 
-  Here, we set
-  \begin{align}
-  a_d^{(h)} & =\E_{Q^{(h)}}(\sigma_d^{-2})=(df+1)/(2 \zeta_d^{(h)}), \nonumber \\
-  b_d^{(h)} & = \E_{Q^{(h)}}(\gamma_d^{-2}) = \sqrt{\frac{\lambda_d}{\theta_d^2 \delta_d^{(h)}}} \frac{K_{\frac{p + \eta_d -2}{2}} \( \sqrt{\lambda_d \delta_d^{(h)}}/\theta_d \)}{K_{\frac{p+\eta_d}{2}} \( \sqrt{\lambda_d \delta_d^{(h)}}/\theta_d \)} + \frac{p+\eta_d^{(h)}}{\delta_d^{(h)}}, \label{eq:auxvar1} \\
-  c_d^{(h)} & = \begin{cases*}
-  b_d^{(h)} & in the conjugate setting \\
-  b_d^{(h)}/a_d^{(h)} & in the non-conjugate setting .
-  \end{cases*} \nonumber
-  \end{align}
-  Note that in the inverse Gamma model, the computation of (\ref{eq:auxvar1}) simplifies to $b_d^{(h)}=(p+\eta_d)/\delta_d^{(h)}$.
-
-  \section{Starting values}
-  \subsection{Method 1}
-  Starting value for $\sigma_d^2$ as in \cite{chipman_bart:_2010}.
+% 	\section{Non-conjugate model}
+% 	The non-conjugate model assumes $\beta_{jd}$ and $\sigma_d^2$ 
+% 	independent \textit{a priori}:
+%   $$
+%   \beta_{jd} | \gamma_d^2, \sigma_d^2 \sim \beta_{jd} | 
+%   \gamma_d^2 \sim \mathcal{N} (0, \gamma_d^2).
+%   $$
+%   
+% 	The variational distributions are as follows:
+% 	\begin{align*}
+% 	  q(\bm{\beta}_d) & \overset{D}{=} \mathcal{N}_p 
+% 	  (\bm{\mu}_d, \bm{\Sigma}_d), \\
+% 	  q(\gamma_d^2) & \overset{D}{=} \mathcal{GIG}
+% 	  \(-\frac{p+1}{2}, \lambda_d \theta_d^{-2}, \delta_d\), \\
+% 	  q(\sigma_d^2) & \overset{D}{=} \Gamma^{-1} 
+% 	  \(\frac{n + 1}{2}, \zeta_d\),
+%   \end{align*}
+% 	The variational parameters are iteratively updated by:
+%   \begin{align*}
+%   \bm{\Sigma}_d^{(h+1)} & = (a_d^{(h)})^{-1} (\X \tr \X + c_d^{(h)} \I)^{-1}, \\
+%   \bm{\mu}_d^{(h+1)} & = (\X \tr \X + c_d^{(h)} \I)^{-1} \X \tr \y_d, \\
+%   \delta_d^{(h+1)} & = b_d^{(h)} (c_d^{(h)})^{-1} \bmu_d \tr \bmu_d + \lambda_d, \\ 
+%   \zeta_d^{(h+1)} & = \frac{1}{2} \bigg[\mathbf{y}_d \tr \mathbf{y}_d -2 \mathbf{y}_d \tr \X \bm{\mu}_d^{(h+1)} + \trace ( \X \tr \X \bm{\Sigma}_d^{(h+1)}) + (\bm{\mu}_d^{(h+1)}) \tr \X \tr \X \bm{\mu}_d^{(h+1)} \\
+%   & \,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\, + \frac{b_d^{(h+1)} - c_d^{(h+1)} a_d^{(h+1)}}{1 - a_d^{(h+1)}} \trace ( \bm{\Sigma}_d^{(h+1)}) + \frac{b_d^{(h+1)} - c_d^{(h+1)} a_d^{(h+1)}}{1 - a_d^{(h+1)}} (\bm{\mu}_d^{(h+1)}) \tr \bm{\mu}_d^{(h+1)}\bigg].
+%   \end{align*}
+%   Here, we set $a_d^{(h)}$, $b_d^{(h)}$, and $c_d^{(h)}$ as in 
+%   (\ref{eq:auxvar1})
+%   Note that in the inverse Gamma model, the computation of (\ref{eq:auxvar1})
+%   simplifies to $b_d^{(h)}=(p+\eta_d)/\delta_d^{(h)}$.
+%   
+%   Here, we set
+%   \begin{align}
+%   a_d^{(h)} & =\E_{Q^{(h)}}(\sigma_d^{-2})=(df+1)/(2 \zeta_d^{(h)}), \nonumber \\
+%   b_d^{(h)} & = \E_{Q^{(h)}}(\gamma_d^{-2}) = \sqrt{\frac{\lambda_d}{\theta_d^2 \delta_d^{(h)}}} \frac{K_{\frac{p + \eta_d -2}{2}} \( \sqrt{\lambda_d \delta_d^{(h)}}/\theta_d \)}{K_{\frac{p+\eta_d}{2}} \( \sqrt{\lambda_d \delta_d^{(h)}}/\theta_d \)} + \frac{p+\eta_d^{(h)}}{\delta_d^{(h)}}, \label{eq:auxvar1} \\
+%   c_d^{(h)} & = \begin{cases*}
+%   b_d^{(h)} & in the conjugate setting \\
+%   b_d^{(h)}/a_d^{(h)} & in the non-conjugate setting .
+%   \end{cases*} \nonumber
+%   \end{align}
+%   Note that in the inverse Gamma model, the computation of (\ref{eq:auxvar1}) simplifies to $b_d^{(h)}=(p+\eta_d)/\delta_d^{(h)}$.
+%   
+%   \section{Starting values}
+%   \subsection{Method 1}
+%   Starting value for $\sigma_d^2$ as in \cite{chipman_bart:_2010}.
 They consider a scaled-Chi squared distribution for the error variance with degrees of freedom 3 and scale chosen such that the $q*100$th percentile matches the error variance in the data $s^2(\mathbf{y}_d)$.
 As starting value for $\sigma_d^2$ we take the mode of this distribution: $\hat{\sigma}_d^2 = 3a/5$, with $a$ the solution to:
-  $$
-  \Gamma\(\frac{3}{3}\) \[1 - q - F(3a/(2s^2(\mathbf{y}_d)); 3/2, 1) \] = 0,
-  $$
-  where $F(x; 3/2, 1)$ is the CDF of a gamma function with shape $3/2$ and scale $1$.
+%   $$
+%   \Gamma\(\frac{3}{3}\) \[1 - q - F(3a/(2s^2(\mathbf{y}_d)); 3/2, 1) \] = 0,
+%   $$
+%   where $F(x; 3/2, 1)$ is the CDF of a gamma function with shape $3/2$ and scale $1$.
 In \cite{moran_variance_2018}, REFERENCE to Rockova and George (2018), and \cite{chipman_bart:_2010}, $q=0.9$ is suggested.
 We find that $q \in ( 0.9, 0.95 )$ gives good results.
-
-  \subsection{Method 2}
-  We generate starting values as follows. Consider the $\sigma_d^2$ and $\gamma_d^2$ as fixed and estimate them by MML of the regular ridge model:
-  $$
-  \hat{\sigma}_d^2, \hat{\gamma}_d^2= \underset{\sigma_d^2,\gamma_d^2}{\argmax} \int_{\bm{\beta}} p(\mathbf{y}_d | \bm{\beta}, \sigma_d^2) p(\bm{\beta} | \gamma_d^2, \sigma_d^2) d\bm{\beta}.
-  $$
-  The regular ridge model is conjugate, so MML is relatively simple. Next we estimate a common inverse Gaussian prior for the $\gamma_d^2$ by considering the estimates to be samples from the prior:
-  \begin{align*}
-  \hat{\theta} & = n^{-1} \sum_{d=1}^D \hat{\gamma}_d^2 / n, \\
-  \hat{\lambda} & = n \left\{\sum_{d=1}^D \[ (\hat{\gamma}_d^2)^{-1} - \hat{\theta}^{-1}\]\right\}^{-1}.
-  \end{align*}
-  We estimate the mode of the $\hat{m}=\hat{\gamma}_d^2$ by kernel density estimation and equate it to the theoretical mode of the generalized inverse Gaussian distribution with the estimated $\hat{\theta}$ and $\hat{\lambda}$.
+%   
+%   \subsection{Method 2}
+%   We generate starting values as follows. Consider the $\sigma_d^2$ and $\gamma_d^2$ as fixed and estimate them by MML of the regular ridge model:
+%   $$
+%   \hat{\sigma}_d^2, \hat{\gamma}_d^2= \underset{\sigma_d^2,\gamma_d^2}{\argmax} \int_{\bm{\beta}} p(\mathbf{y}_d | \bm{\beta}, \sigma_d^2) p(\bm{\beta} | \gamma_d^2, \sigma_d^2) d\bm{\beta}.
+%   $$
+%   The regular ridge model is conjugate, so MML is relatively simple. Next we estimate a common inverse Gaussian prior for the $\gamma_d^2$ by considering the estimates to be samples from the prior:
+%   \begin{align*}
+%   \hat{\theta} & = n^{-1} \sum_{d=1}^D \hat{\gamma}_d^2 / n, \\
+%   \hat{\lambda} & = n \left\{\sum_{d=1}^D \[ (\hat{\gamma}_d^2)^{-1} - \hat{\theta}^{-1}\]\right\}^{-1}.
+%   \end{align*}
+%   We estimate the mode of the $\hat{m}=\hat{\gamma}_d^2$ by kernel density estimation and equate it to the theoretical mode of the generalized inverse Gaussian distribution with the estimated $\hat{\theta}$ and $\hat{\lambda}$.
 We solve to find a common $\delta$:
-  $$
-  \hat{\delta} = \frac{\hat{m}^2 \hat{\lambda}}{\hat{\theta}^2} + (p + 3) \hat{m}.
-  $$
-  Lastly, we consider the $\hat{\sigma}^2_d$ fixed samples from the inverse Gamma distribution to estimate a common scale:
-  $$
-  \hat{\zeta} = \frac{D(n + p + 1)}{2 \sum_{d=1}^D[(\hat{\sigma}_d^2)^{-1}]}.
-  $$
+%   $$
+%   \hat{\delta} = \frac{\hat{m}^2 \hat{\lambda}}{\hat{\theta}^2} + (p + 3) \hat{m}.
+%   $$
+%   Lastly, we consider the $\hat{\sigma}^2_d$ fixed samples from the inverse Gamma distribution to estimate a common scale:
+%   $$
+%   \hat{\zeta} = \frac{D(n + p + 1)}{2 \sum_{d=1}^D[(\hat{\sigma}_d^2)^{-1}]}.
+%   $$
 

### 2019-08-13 docs/supplement2.Rnw
updated simulations and manuscript

@@
 \usepackage{bm,amsmath,amssymb,amsthm,amsfonts,graphics,graphicx,epsfig,
-rotating,caption,subcaption,natbib,appendix,titlesec,multicol,hyperref,verbatim,
+rotating,caption,subcaption,natbib,appendix,titlesec,multicol,verbatim,
 bbm,algorithm,algpseudocode,pgfplotstable,threeparttable,booktabs,mathtools,
-dsfont,parskip}
+dsfont,parskip,xr-hyper,hyperref}
+\externaldocument[md-]{manuscript2}
 
@@
 	This document contains the Supplementary Material (SM) to the document `Drug efficacy prediction in cell lines'. In the following, this document is referred to as Main Document (MD).
+
+	\section{Model}
+	\subsection{Simultaneous equations model}\label{sec:SEM}
+	We have continuous efficacy measures $y_{id}$ for cell lines $i=1,\dots, n$, 
+	from tissues $t=1, \dots, T$, on drugs $d=1,\dots,D$. Throughout this paper 
+	we assume the $y_{id}$ to be centred per drug and let $\y_d = \begin{bmatrix} 
+	y_{1d} & \dots & y_{nd} \end{bmatrix} \tr$. We predict efficacy with molecular
+	features $x_{ij}$, $j=1,\dots, p$, collected in 
+	$\x_i = \begin{bmatrix} x_{i1} & \dots & x_{ip} \end{bmatrix} \tr$. We have an
+	\textit{a priori} partitioning of the omics features into $g=1, \dots, G$ 
+	non-overlapping groups, which might be informative for the problem at hand For convenience, we let $g(j)$ denote the group index of molecular feature $j$.
+Furthermore, we have characteristics of the drugs available in the form of `co-data' $\mathbf{C} = \begin{bmatrix} \mathbf{c}_1 & \dots & \mathbf{c}_D \end{bmatrix} \tr$.
+We code the tissue from which the cell lines were taken as a $T$ binary dummy variables
+	$$
+	z_{it} = \begin{cases}
+	1 & \text{if } \text{tissue}_i = t, \\
+	0 & \text{otherwise},
+	\end{cases}
+	$$
+	collected in $\mathbf{z}_i = \begin{bmatrix} z_{i1} & \dots & z_{iT} \end{bmatrix} \tr$.
+
+	We assume that cell lines from the same tissue are related and model this relation through random effects $u_{td}$.
+We let the molecular feature effects $\beta_{jd}$ vary over the drugs and model drug efficacy as a linear combination of these two effects:
+	\begin{subequations}\label{eq:linearmodel}
+		\begin{align}
+		y_{id} & = \beta_{0d} + \sum_{j=1}^p x_{ij} \beta_{jd} + \sum_{t=1}^T z_{it} u_{td} + \epsilon_{id} \\
+		& = \beta_{0d} + \x_i \tr \bbeta_d + \mathbf{z}_i \tr \mathbf{u}_d + \epsilon_{id},\\
+		\text{with } & \epsilon_{id} \sim \mathcal{N}(0, \sigma_d^2),
+		\end{align}
+	\end{subequations}
+	where $p$-dimensional $\bbeta_d$ and $T$-dimensional $\bm{u}_d$ are the drug-specific omics feature and tissue effect vectors, respectively.
+Note that (\ref{eq:linearmodel}) gives rise to a system of $d=1, \dots, D$ linear simultaneous equations.
+
+	\subsection{Bayesian prior model}
+	We capture the uncertainty in the parameters through a Bayesian prior model.
+We assign a flat prior to the intercept terms $\beta_{0d}$ and integrate them out.
+The remaining parameters are endowed with the following priors:
+	\begin{subequations}\label{eq:prior}
+		\begin{align}
+		\beta_{jd} & \sim \mathcal{N}_p (0, \phi^2_{g(j)} \gamma_d^2 \sigma_d^2), \label{eq:betaprior}\\
+		\mathbf{u}_{d} & \sim \mathcal{N}_T (\mathbf{0}, \bm{\Xi}_d \sigma_d^2),
+		\end{align}
+	\end{subequations}
+	and hyper-priors:
+	\begin{subequations}\label{eq:hyperprior}
+		\begin{align}
+			\sigma_d^{2} & \sim 1/\sigma_d^{3} \\
+			\gamma_d^{2} & \sim \mathcal{IG}(\theta_d, \lambda), \\
+			\bm{\Xi}_d & \sim \mathcal{W}^{-1}_T(\bm{\Omega}, \nu),
+		\end{align}
+	\end{subequations}
+	where $\mathcal{IG}(\theta, \lambda)$ is an inverse Gaussian distribution with mean $\theta$ and shape $\lambda$, and $\mathcal{W}^{-1}_T(\bm{\Omega}, \nu)$ is an inverse Wishart distribution with $T \times T$-dimensional positive definite (PD) scale matrix $\bm{\Omega}$ and degrees of freedom $\nu$.
+Note that for $\nu > T + 1$, this inverse Wishart prior implies an expected prior covariance matrix $(\nu - T - 1)^{-1} \bm{\Omega}$ for the tissue effects, common to all drugs.
+Likewise, the molecular feature effects are related through the common hyper-prior for the $\gamma_d^2$.
+
+	- Jeffrey's prior for variance $1/\sigma^{3/2}$ (a priori independent data mean and variance) or $1/\sigma^3$ (joint prior, but only mean and variance, not variance and betas).
+
+	A few remarks on the choice of priors are justified here: many authors endow error variance components with vague gamma priors.
+\cite{gelman_prior_2006}, among others, advises against this practice.
+The degree of `vagueness' has a large influence on the posterior, while degree of `vagueness' is a difficult parameter to set.
+This influence is especially pronounced if the likelihood is relatively flat, as may be reasonably expected in the large $p$, small $n$ setting we are in.
+We therefore model the error variance with Jeffrey's prior \cite[]{jeffreys_invariant_1946}.
+
+	Furthermore, we choose to model the $\gamma^2_d$ by an inverse Gaussian distribution, as has been suggested in \cite{fabrizi_specification_2016} and \cite{caron_sparse_2008}, because it allows to model the mean $\theta_d$ as a function of the drug covariates $\mathbf{c}_d$, as explained in Section \ref{sec:empiricalbayes}.
+
+	- IG hyperprior used in finance \cite[]{barndorff-nielsen_normal_1997}, where it is called the NIG prior.
+
+	\section{Estimation}
+	\subsection{Variational Bayes}\label{sec:variationalbayes}
+	The posterior corresponding to the model described in (\ref{eq:linearmodel}), (\ref{eq:prior}), and (\ref{eq:hyperprior}) is not available in closed form.
+We therefore approximate the posterior by variational Bayes, where we force the posterior density $Q$ to factorise as:
+	\begin{align}\label{eq:variationalposterior}
+	p(\B,\mathbf{U},\mathbf{\sigma}^2, \bgamma^2,\bm{\Xi}_1, \dots, \bm{\Xi}_D) \approx Q(\cdot) = q(\B) \cdot q(\mathbf{U}) \cdot q(\mathbf{\sigma}^2) \cdot q(\bgamma^2) \cdot q(\bm{\Xi}_1, \dots, \bm{\Xi}_D),
+	\end{align}
+	where $\B = \begin{bmatrix} \bbeta_1 & \cdots & \bbeta_D \end{bmatrix}$ and $\U = \begin{bmatrix} \mathbf{u}_1 & \cdots & \mathbf{u}_D \end{bmatrix}$, are of dimensions $p \times D$ and $T \times D$, respectively.
+Under the factorisation in (\ref{eq:variationalposterior}), the marginal variational posteriors that minimise the Kullback-Leibler divergence of the true posterior to the variational Bayes approximation \cite[]{neal_view_1998}, are given by:
+	\begin{subequations}\label{eq:variationalposterior2}
+		\begin{align}
+		q_{\B} (\B) & \overset{D}{=} \prod_{d=1}^{D} \mathcal{N}_p (\bmu_d, \bSigma_d), \\
+		q_{\U}(\U) & \overset{D}{=} \prod_{d=1}^{D} \mathcal{N}_T (\mathbf{m}_d, \Sm_d), \\
+		q_{\bgamma^{2}}(\bgamma^{2}) & \overset{D}{=} \prod_{d=1}^D \mathcal{GIG} \(-\frac{p+1}{2}, \frac{\lambda}{\theta_d^2}, \delta_{d} \), \\
+		q_{\bsigma^{2}}(\bsigma^{2}) & \overset{D}{=} \prod_{d=1}^D \Gamma^{-1} \(\frac{n + p + T + 1}{2}, \zeta_{d} \), \\
+		q_{\bm{\Xi}_1, \dots, \bm{\Xi}_D}(\bm{\Xi}_1, \dots, \bm{\Xi}_D) & \overset{D}{\propto} \prod_{d=1}^D \mathcal{W}_T^{-1} (\bm{\Psi}_d, \nu + 1).
+		\end{align}
+	\end{subequations}
+	Here $\mathcal{GIG}(\cdot)$ and $\Gamma^{-1}(\cdot)$ denote the generalized inverse Gaussian and inverse gamma distributions, respectively.
+
+	The parameters in (\ref{eq:variationalposterior2}) contain cyclic dependencies. We therefore iterate the following estimating equations until convergence to a local optimum:
+	\begin{align*}
+	\bSigma_d^{(h+1)} & = \frac{2\zeta_d^{(h)}}{n+p+T+1} \left\{\X \tr \X + a_d^{(h)} \cdot \diag (\phi_{g(j)}^{-2}) \right\}^{-1}, \\
+	\bm{\mu}_d^{(h+1)} & = \left\{\X \tr \X + a_d^{(h)} \cdot \diag (\phi_{g(j)}^{-2}) \right\}^{-1} \X \tr ( \y_d - \Z \mathbf{m}_d^{(h)}), \\
+	\Sm_d^{(h+1)} & = \frac{2\zeta_d^{(h)}}{n+p+T+1} \[ \Z \tr \Z + (\nu + 1) \cdot (\bm{\Psi}_d^{(h)})^{-1} \]^{-1}, \\
+	\mathbf{m}_d^{(h+1)} & = \[ \Z \tr \Z + (\nu + 1) \cdot (\bm{\Psi}_d^{(h)})^{-1} \]^{-1} \Z \tr (\y_d - \X \bm{\mu}_d^{(h)}), \\
+	\bm{\Psi}_d^{(h+1)} & = \frac{2\zeta_d^{(h)}}{n+p+T+1} \[ \Sm_d^{(h+1)} + \mathbf{m}_d^{(h+1)} (\mathbf{m}_d^{(h+1)}) \tr \] + \bm{\Omega}, \\
+	\delta_d^{(h+1)} & =  \frac{n+p+T+1}{2\zeta_d^{(h)}} \sum_{j=1}^p \frac{(\mu_{jd}^{(h+1)})^2 + (\bSigma_d^{(h+1)})_{jj}}{\phi^2_{g(j)}} + \lambda, \\
+	\zeta_d^{(h+1)} & = \frac{1}{2} \y_d \tr \y_d - \y_d \tr (\X \bmu_d^{(h+1)} + \Z \mathbf{m}_d^{(h+1)}) + (\bmu_d^{(h+1)}) \tr \X \tr \Z \mathbf{m}_d^{(h+1)} + \frac{1}{2} \trace (\X \tr \X \bSigma_d^{(h+1)}) \\
+	& \,\,\,\,\,\,\,\,\,\, + \frac{1}{2} (\bmu_d^{(h+1)}) \tr \X \tr \X \bmu_d^{(h+1)} + \frac{1}{2} \trace (\Z \tr \Z \Sm_d^{(h+1)}) + \frac{1}{2} (\mathbf{m}_d^{(h+1)}) \tr \Z \tr \Z \mathbf{m}_d^{(h+1)} \\
+	& \,\,\,\,\,\,\,\,\,\, + \frac{1}{2} a_d^{(h)} \sum_{j=1}^p \frac{(\mu_{jd}^{(h+1)})^2 + (\bSigma_d^{(h+1)})_{jj}}{\phi^2_{g(j)}} + \frac{1}{2} (\nu + 1) \cdot \trace \[ (\bm{\Psi}_d^{(h+1)})^{-1} \Sm_d^{(h+1)} \] \\
+	& \,\,\,\,\,\,\,\,\,\, + \frac{1}{2} (\nu + 1) \cdot (\mathbf{m}_d^{(h+1)}) \tr (\bm{\Psi}_d^{(h+1)})^{-1} \mathbf{m}_d^{(h+1)},
+	\end{align*}
+	where
+	$$
+	a_d^{(h)} = \sqrt{\frac{\lambda}{\theta^2_d \delta_d^{(h)}}} \frac{K_{\frac{p - 1}{2}} \( \sqrt{\lambda \delta_d^{(h)}}/\theta_d \)}{K_{\frac{p+1}{2}} \( \sqrt{\lambda \delta_d^{(h)}}/\theta_d \)} + \frac{p+1}{\delta_d^{(h)}},
+	$$
+	and $K_{\alpha}(\cdot)$ is the modified Bessel function of the second kind.
+
+	\subsection{Empirical Bayes}\label{sec:empiricalbayes}
+	A full Bayesian model requires the specification of hyper-parameters $\bm{\eta} = \begin{bmatrix} (\bm{\phi}^2) \tr & \bm{\theta} \tr & \lambda & \text{vec}(\bm{\Omega}) \tr & \nu \end{bmatrix} \tr$.
+These are abstract and hard to interpret parameters for which we generally lack expert knowledge.
+They do, however, have a significant influence on the shape of the posterior distribution.
+We therefore propose to estimate the unspecified hyper parameters $\bm{\eta}$ by empirical Bayes.
+Simply put, empirical Bayes fits the prior model to the data and is, as such, objective (up to model specification).
+
+	The canonical method for empirical Bayes is maximisation of the marginal likelihood with respect to the hyper parameters.
+In \cite{casella_empirical_2001} the marginal likelihood is maximised by an EM algorithm:
+	$$
+	\bm{\eta}^{(l+1)} = \underset{\bm{\eta}}{\argmax}\E_{\cdot | \Y} [\log p(\Y, \B, \U, \bm{\Xi}_1, \dots, \bm{\Xi}_D, \bgamma^2, \bsigma^2) | \bm{\eta}^{(l)}],
+	$$
+	where the expectation is with respect to the posterior.
+In our case, this posterior is not available in closed form, which renders the expectation difficult.
+While \cite{casella_empirical_2001} suggests to approximate the expectation by a Monte Carlo sample, we propose to use the variational Bayes approximation developed in Section \ref{sec:variationalbayes}:
+	\begin{align}\label{eq:variationalbayes}
+	\bm{\eta}^{(l+1)} & = \underset{\bm{\eta}}{\argmax}\E_{Q^{(l)}} [\log p(\Y, \B, \U, \bm{\Xi}_1, \dots, \bm{\Xi}_D, \bgamma^2, \bsigma^2)] \nonumber \\
+	& = \underset{\bm{\eta}}{\argmax} \E_{Q^{(l)}} [\log \pi (\B | \bgamma^2, \bsigma^2)] + \E_{Q^{(l)}} [\log \pi (\bgamma^2)] + \E_{Q^{(l)}} [\log \pi (\bm{\Xi}_1, \dots, \bm{\Xi}_D)].
+	\end{align}
+	where now the expectation is with respect to the converged variational posterior $Q^{(l)}$.
+Inspection of (\ref{eq:variationalbayes}) learns us that the problem may be decomposed into three separate optimisation problems.
+
+	\subsubsection{Group-specific variance components}
+	The first sub-problem concerns the group-specific variance components of the molecular feature effects.
+In order to separate the overall variance of the molecular feature effects from the group-specific deviations, we impose $\prod_{g=1}^G (\phi_g^2)^{|\mathcal{G}_g|} = 1$.
+As a consequence, the overall level of shrinkage is determined by the $\gamma^2_d$, while the $\phi_g^2$ effectuate differential shrinkage of the groups.
+Because the $\phi_g^2$ are variance components, we additionally require $\phi^2_1, \dots, \phi^2_G> 0$, such that our first optimisation sub-problem becomes:
+	\begin{align*}
+	(\bm{\phi}^2)^{(l+1)} & = \underset{\bm{\phi}^2}{\argmax} \E_{Q^{(l)}} [\log \pi (\B | \bgamma^2, \bsigma^2)] \\
+	& = \underset{\bm{\phi}^2}{\argmax} -\frac{D}{2} \sum_{g=1}^G |\G_g| \log \phi_g^2 - \frac{1}{2} \sum_{g=1}^G b_g^{(l)} \phi_g^{-2} \\
+	& \text{subject to } \prod_{g=1}^G (\phi_g^2)^{|\G_g|} = 1 \text{ and } \phi^2_1, \dots, \phi^2_G> 0.
+	\end{align*}
+	where 
+	$$
+	b_g^{(l)}=\sum_{d=1}^D a_d^{(l)} \frac{n+p+T+1}{2\zeta_d^{(l)}} \sum_{j \in \mathcal{G}_g} \[ (\mu_{jd}^{(l)})^2 + (\bSigma_d^{(l)})_{jj} \].
+	$$
+	This is a convex optimisation problem and easily solved numerically.
+
+	\subsubsection{Drug-specific variance components}
+	The second sub-problem concerns the prior for the drug-specific variance components $\gamma_d^2$.
+As explained in Section \ref{sec:SEM}, we may have `covariates' $\mathbf{c}_d$ on the drugs available.
+We conjecture that these covariates may be informative for the omics effect size and therefore model the prior mean of the $\gamma_d^2$ as a function of these covariates: $\theta_d = (\mathbf{c}_d \tr \bm{\alpha})^{-1}$.
+Then, the empirical Bayes estimation of $\bm{\alpha}$ effectively boils down to an inverse Gaussian regression of the $\gamma_d^{-2}$ on the $\mathbf{c}_d$ \cite[]{fries_optimal_1986,whitmore_regression_1983}.
+The problem may be formulated as:
+	% \begin{align*}\label{eq:optproblem2}
+	\begin{align*}
+	\lambda^{(l+1)}, \bm{\alpha}^{(l+1)} & = \underset{\lambda,\bm{\alpha}}{\argmax} \E_{Q^{(l)}} [\log \pi (\bgamma^2)] \\
+	& = \underset{\lambda,\bm{\alpha}}{\argmax} \frac{D}{2} \log \lambda - \frac{\lambda}{2} \bm{\alpha} \tr \mathbf{C} \tr \diag [\E_{Q^{(l)}}(\gamma_d^2)] \mathbf{C} \bm{\alpha} + \lambda \bm{\alpha} \tr \mathbf{C} \tr \mathbf{1} - \frac{\lambda}{2} \sum_{d=1}^D \E_{Q^{(l)}}(\gamma_d^{-2}),
+	\end{align*}
+	where $\mathbf{C} = \begin{bmatrix} \mathbf{c}_1 & \dots & \mathbf{c}_D \end{bmatrix} \tr$. The solutions to this problem are:
+	\begin{align*}
+	\bm{\alpha}^{(l+1)} & = \left\{ \mathbf{C} \tr \Vm^{(l)} \mathbf{C} \right\}^{-1} \mathbf{C} \tr \mathbf{1}, \\
+	\lambda^{(l+1)} & = \[ \sum_{d=1}^D a_d^{(l)} - (\bm{\alpha}^{(l+1)}) \tr \mathbf{C} \tr \mathbf{1} \]^{-1} D,
+	\end{align*}
+	where 
+	$$
+	\Vm^{(l)} = \diag \[ \sqrt{\frac{\delta_d^{(l)}(\theta_d^{(l)})^2}{\lambda^{(l)}}} \frac{K_{\frac{p - 1}{2}} \( \sqrt{\lambda^{(l)} \delta_d^{(l)}}/\theta^{(l)}_d \)}{K_{\frac{p + 1}{2}} \( \sqrt{\lambda^{(l)} \delta_d^{(l)}}/\theta^{(l)}_d \)} \].
+	$$
+
+	- Think about parametrisation of $\mathbf{C}$.
+Idea: use overall mean and deviations from it in alpha, such that the starting value can be the mean and 0 for alpha.
+Or: use intercept and code the rest as dummy variables.
+
+	\subsubsection{Tissue effect covariance}
+	The third sub-problem pertains to the tissue effect covariance prior.
+To avoid overfitting when freely estimating the $\frac{T(T+1)}{2}$ elements in $\bm{\Omega}$ and $\nu$, we parametrise them as $\bm{\Omega}=  \tau [(\tau - \rho) \I_T + \rho \mathbf{1}_{T \times T}]$ and $\nu = \tau + T + 1$, respectively.
+The justification for this parametrisation is three-fold: (a) it implies an easy to interpret prior covariance matrix with $\tau$ on its diagonal and $\rho$ on its off-diagonals, (b) it reduces the number of parameters to  estimate to just two, and (c) simplifies the positive definite constraint on  $\bm{\Omega}$.
+To ensure positive definiteness and a proper covariance matrix, we impose the constraints $\tau > \max\[\rho - T \rho,\rho, 0\]$ (see Appendix \ref{app:priortissuecovariance}).
+The resulting third sub-problem is a bit more involved than the first two:
+	\begin{align*}
+	\tau^{(l + 1)}, \rho^{(l+1)} & = \underset{\tau, \rho}{\argmax} \frac{D (T^2 + T)}{2} \log \tau + \frac{DT}{2} \tau \log \tau + \frac{D(T+1)}{2} \log \[ \tau - (T - 1) \rho\] \\
+	& \,\,\,\,\,\,\,\,\,\, + \frac{D}{2} \tau \log \[ \tau - (T - 1) \rho\] + \frac{D(T^2 - 1)}{2} \log(\tau - \rho) +\frac{D(T - 1)}{2} \tau  \log(\tau - \rho) \\
+	& \,\,\,\,\,\,\,\,\,\, - D \sum_{t=1}^T \log \Gamma \( \frac{\tau + t + 1}{2} \) - w_1^{(l)} \tau^2 + w_2^{(l)} \tau \rho + w_3^{(l)} \tau, \\
+	& \text{subject to } \tau > \max\[\rho - T \rho,\rho, 0\],
+	\end{align*}
+	where
+	\begin{align*}
+	w_1^{(l)} & =  \frac{\tau^{(l)} + T + 2}{2} \sum_{d=1}^D \trace \[ (\bm{\Psi}_d^{(l)})^{-1}\] > 0,\\
+	w_2^{(l)} & = \frac{\tau^{(l)} + T + 2}{2} \sum_{d=1}^D \left\{\trace \[ (\bm{\Psi}_d^{(l)})^{-1}\] - \trace \[\mathbf{1}_{T \times T} \cdot (\bm{\Psi}_d^{(l)})^{-1}\] \right\} < w_1^{(l)},\\
+	w_3^{(l)} & = \frac{1}{2} \[D \sum_{t=1}^T \psi\(\frac{\tau^{(l)} + t + 2}{2}\) - \sum_{d=1}^D \log |\bm{\Psi}_d^{(l)}| \].
+	\end{align*}
+
+	- use mathematica to check convexity
+
+	- we use the ELBO for VB convergence and parameters for EB convergence
+
+	- we might be lenient for VB convergence. The extreme case would be 1 VB 
+	iteration per EB iteration (corresponds to tolerance of infinity).
+
+	\section{Prior tissue covariance constraints}\label{app:priortissuecovariance}
+		We parametrise the prior covariance as a matrix with $\tau$ on the diagonal and $\rho$ on its off-diagonals. To that end we set $\nu = \tau + T + 1$ and the $T \times T$-dimensional matrix
+		$$
+		\bm{\Omega} = \tau \begin{bmatrix}
+		\tau &  & \rho \\
+		& \ddots &  \\
+		\rho & & \tau
+		\end{bmatrix},
+		$$
+		such that $\E_{\pi(\mathbf{\Xi}_d)} (\mathbf{\Xi}_d) = (\nu - T - 1)^{-1} \bm{\Omega}$ is of the desired form.
+To ensure a non-degenerate prior distribution we require $\tau > 0$ and $\bm{\Omega}$ PD.
+We use $\tau > 0$ to note that a PD $\bm{\Omega}$ implies positive roots $\lambda$ of the characteristic polynomial:
+		\begin{align*}
+		\left| \begin{bmatrix}
+		\tau &  & \rho \\
+		& \ddots &  \\
+		\rho & & \tau
+		\end{bmatrix} -\lambda \I_T \right| = (\tau - \lambda - \rho)^{T-1}[\tau - \lambda - (1-T)\rho] = 0.
+		\end{align*}
+		For $T > 1$, the solutions are:
+		$$
+		(\lambda = \tau - \rho) \lor (\lambda = T \rho - \rho + \tau),
+		$$
+		which are positive if the following condition holds: $\tau > \max(\rho,\rho - T \rho)$.
+We combine this with $\tau > 0$, to arrive at the full proper prior constraint: $\tau > \max(0,\rho,\rho - T \rho)$.
 

### 2019-08-13 code/figures.R

@@ -1,19 +1,74 @@
 ################################# main document ################################
 # ---- figures ----
+# ---- dens_igaussian_marginalbeta ----
+library(GeneralizedHyperbolic)
+library(metRology)
+library(sp)
+sigma <- 1
+mgammasq <- 1/sigma^2
+lambda <- c(0.1, 2)
+lambda1 <- sqrt(2)
+eta <- lambda/mgammasq + 2
+labels <- as.expression(c(bquote("NIG, "~lambda==.(lambda[1])*","~
+                                   (bold(c)^"T"*bold(alpha))^-1==.(mgammasq)),
+                          bquote("NIG, "~lambda==.(lambda[2])*","~
+                                   (bold(c)^"T"*bold(alpha))^-1==.(mgammasq)),
+                          bquote("Student's t, "~xi==.(lambda[1])*","~
+                                   eta==.(eta[1])),
+                          bquote("Student's t, "~xi==.(lambda[2])*","~
+                                   eta==.(eta[2])),
+                          "ridge", "lasso"))
+dprior <- function(x, lambda, mgammasq, sigma) {
+  dnig(x, 0, sigma*sqrt(lambda), sqrt(lambda/(mgammasq*sigma)), 0)
+}
+dlasso <- function(x, lambda1) {
+  0.5*lambda1*exp(-lambda1*abs(x))
+}
+
+col <- bpy.colors(length(labels), cutoff.tail=0.3)
+lty <- c(1:length(labels))
+
+ylim <- c(0, dt.scaled(0, eta[1]/2, 0, sqrt(lambda[1]*sigma^2/eta[1])))
+opar <- par(no.readonly=TRUE)
+par(mar=opar$mar*c(1, 1.1, 1, 1))
+curve(dprior(x, lambda[1], mgammasq, sigma), -3, 3, 
+      ylab=expression(p(beta)), 
+      xlab=expression(beta), n=1000, ylim=ylim,
+      col=col[1], lty=lty[1])
+curve(dprior(x, lambda[2], mgammasq, sigma), add=TRUE, n=1000, 
+      col=col[2], lty=lty[2])
+curve(dt.scaled(x, eta[1]/2, 0, sqrt(lambda[1]*sigma^2/eta[1])), add=TRUE, 
+      n=1000, col=col[3], lty=lty[3])
[…]

### 2019-08-13 rpackage/R/aux_functions.R

@@ -300,3 +300,78 @@ ridge.mll <- function(par, sv, uty, yty, n, p) {
               lambda=lambda, zeta=rep(zeta, D), a=rep(a, D))
   return(out)
 }
+
+### CPO
+.f.int.cpo <- function(val, y, zeta, n, p, xtmu, xtSigmax) {
+  (val^2/(2*zeta) + 1)^(-(n - length(xtmu) + p + 2)/2)*
+    exp(-(val - xtmu + y)^2/(2*xtSigmax))
+}
+
+.cpoi <- function(model, x, y, n, p, D) {
+  out <- lapply(1:D, function(d) {
+    zeta <- model$vb$zeta[d]
+    xtmu <- as.numeric(x[[d]] %*% model$vb$mu[[d]])
+    xtSigmax <- rowSums((x[[d]] %*% model$vb$Sigma[[d]])*x[[d]])
+    int <- sapply(1:nrow(matrix(y, ncol=D)), function(i) {
+      unlist(integrate(.f.int.cpo, -Inf, Inf, y=matrix(y, ncol=D)[i, d], 
+                       zeta=zeta, n=n, p=p[d], xtmu=xtmu[i], 
+                       xtSigmax=xtSigmax[i])[c(1:2)])})
+    
+    
+    list((2*pi)^(-1)*exp(lgamma((n- nrow(matrix(y, ncol=D)) + p[d] + 2)/2) - 
+                           lgamma((n - nrow(matrix(y, ncol=D)) + p[d] + 1)/2))*
+           (zeta*xtSigmax)^(-1/2)*int[1, ], int[2, ])})
+  
+  return(list(value=unname(sapply(out, "[[", 1)), 
+              abs.error=unname(sapply(out, "[[", 2))))
+}
+
+lpml <- function(x, y, C, mult.lambda=FALSE, intercept.eb=TRUE,
+                 fixed.eb=c("none", "lambda", "both"), full.post=FALSE, 
+                 init=NULL,
+                 control=list(conv.post=TRUE, trace=TRUE,
+                              epsilon.eb=1e-3, epsilon.vb=1e-3, 
+                              epsilon.opt=sqrt(.Machine$double.eps),
+                              maxit.eb=10, maxit.vb=2, maxit.opt=100,
+                              maxit.post=100),
+                 nfolds, foldid=NULL) {
+  
+  # create the folds
+  p <- sapply(x, ncol)
+  D <- ncol(y)
+  n <- nrow(y)
[…]

### 2019-08-15 docs/manuscript1.Rnw
added unpenalized covariates and intercept (not tested)

@@
 	the features in a low dimensional chemometrics dataset. Our proposed method,
-	termed [! acronym here],
+	termed \texttt{semnig},
 	extends \cite{kpogbezan_incorporating_2019,kpogbezan_empirical_2017} by 
@@
 	sources of information.
-
-	- think of acronym
 

### 2019-08-15 code/data_gdsc.R

@@ -36,13 +36,13 @@ if(file.exists("data/sanger1018_brainarray_ensemblgene_rma.txt.gz")) {
   expr <- read.table(gcon(url("ftp://ftp.sanger.ac.uk/pub/project/cancerrxgene/releases/current_release/sanger1018_brainarray_ensemblgene_rma.txt.gz"),
                           text=TRUE), header=TRUE, stringsAsFactors=FALSE)
 }
-# if(file.exists("data/Cell_Lines_Details.xlsx")) {
-#   cell1 <- read.xls("data/Cell_Lines_Details.xlsx", 
-#                     stringsAsFactors=FALSE)
-# } else {
-#   cell1 <- read.xls("ftp://ftp.sanger.ac.uk/pub/project/cancerrxgene/releases/current_release/Cell_Lines_Details.xlsx", 
-#                     stringsAsFactors=FALSE)
-# }
+if(file.exists("data/Cell_Lines_Details.xlsx")) {
+  cell <- read.xls("data/Cell_Lines_Details.xlsx",
+                   stringsAsFactors=FALSE)
+} else {
+  cell <- read.xls("ftp://ftp.sanger.ac.uk/pub/project/cancerrxgene/releases/current_release/Cell_Lines_Details.xlsx",
+                   stringsAsFactors=FALSE)
+}
 
 # reading in the preformatted list of compounds
 if(file.exists("data/Screened_Compounds.xlsx")) {

### 2019-08-15 code/tryout.R

@@ -1,0 +1,22 @@
+r <- 200
+u <- 5
+n <- 100
+xu <- matrix(rnorm(u*n), nrow=n, ncol=u)
+xr <- matrix(rnorm(r*n), nrow=n, ncol=r)
+h <- rep(1, r)
+
+mat1 <- solve(cbind(rbind(t(xu) %*% xu, t(xr) %*% xu), 
+                    rbind(t(xu) %*% xr, t(xr) %*% xr + diag(h))))
+
+Om <- solve(xr %*% diag(1/h) %*% t(xr) + diag(n))
+mat2.11 <- solve(t(xu) %*% Om %*% xu)
+mat2.12 <- - mat2.11 %*% t(xu) %*% xr %*% diag(1/h) +
+  mat2.11 %*% t(xu) %*% xr %*% diag(1/h) %*% t(xr) %*% Om %*% xr %*% diag(1/h)
+mat2.21 <- t(mat2.12)
+mat2.22 <- diag(1/h) - diag(1/h) %*% t(xr) %*% Om %*% xr %*% diag(1/h) -
+  diag(1/h) %*% t(xr) %*% xu %*% mat2.12 + 
+  diag(1/h) %*% t(xr) %*% Om %*% xr %*% diag(1/h) %*% t(xr) %*% xu %*% mat2.12
+mat2 <- cbind(rbind(mat2.11, mat2.21), rbind(mat2.12, mat2.22))
+mat1[1:5, 1:5]
+mat2[1:5, 1:5]
+

### 2019-08-15 rpackage/R/main_functions.R

@@ -11,9 +11,10 @@
 #              maxit.post=100)
 
 # estimate ENIG model (not tested)
-enig <- function(x, y, C, mult.lambda=FALSE, intercept.eb=TRUE,
-                 fixed.eb=c("none", "lambda", "both"), full.post=FALSE, 
-                 init=NULL,
+enig <- function(x, y, C, unpenalized=NULL, standardize=TRUE, intercept=TRUE, 
+                 intercept.eb=TRUE, mult.lambda=FALSE, 
+                 fixed.eb=c("none", "lambda", "both"), 
+                 full.post=FALSE, init=NULL,
                  control=list(conv.post=TRUE, trace=TRUE,
                               epsilon.eb=1e-3, epsilon.vb=1e-3, 
                               epsilon.opt=sqrt(.Machine$double.eps),
@@ -23,11 +24,26 @@ enig <- function(x, y, C, mult.lambda=FALSE, intercept.eb=TRUE,
   # save the arguments
   cl <- match.call()
   
+  # check for unpenalized covariates
+  unp <- intercept | !is.null(unpenalized)
+  
   # fixed parameters
-  n <- nrow(y)
-  p <- sapply(x, ncol)
   D <- ncol(y)
+  n <- nrow(y)
+  u <- sapply(unpenalized, function(s) {
+    ifelse(is.null(s), 0, ncol(s)) + intercept})
+  r <- sapply(x, ncol)
+  p <- sapply(1:D, function(d) {u[d] + p[d]})
   ncd <- ifelse(is.null(C), 0, ncol(C[[1]]))
+  
+  # standardize data and add intercept if need be
+  xr <- x
+  if(standardize) {
+    xr <- scale(xr)
+  }
+  if(intercept) {
+    xu <- cbind(1, xu)
+  }
   
   # add intercept and transform co-data to matrix
   if(intercept.eb & is.null(C)) {
@@ -40,12 +56,14 @@ enig <- function(x, y, C, mult.lambda=FALSE, intercept.eb=TRUE,
   
   # fixed data functions
   yty <- colSums(y^2)
-  ytx <- lapply(1:D, function(d) {as.numeric(y[, d] %*% x[[d]])})
+  if(!unp) {
+    ytx <- lapply(1:D, function(d) {as.numeric(y[, d] %*% x[[d]])})  
+  }
   
   # set initial values if not given
   if(is.null(init)) {
     init$aold <- rep(1, D)
[…]

### 2019-08-19 docs/manuscript1.Rnw
unpenalized covariates added to rpackage, slight reorganization

@@
 
+  Several extensions of the model are described in [! ref to second paper here].
+
 	\section{Estimation}\label{sec:estimation}
@@
     b_{jd}^{(h)} & = \E_{Q^{(h)}}(\gamma_{jd}^{-2}) = 
-    \sqrt{\frac{\delta_{jd}^{(h)}}
-    {\lambda_{d}[(\mathbf{c}_{d}^j) \tr \bm{\alpha}]^2}}
+    \sqrt{\frac{\lambda_{d}[(\mathbf{c}_{d}^j) \tr \bm{\alpha}]^2}
+    {\delta_{jd}^{(h)}}}
     \frac{K_0(\sqrt{\delta_{jd}^{(h)}\lambda_{d}}
@@
   to maximise the posterior to obtain the maximum \textit{a posteriori} (MAP)
-  estimates of the model parameters. We compare these MAP estimates together
-  with the ridge estimates obtained from the \texttt{glmnet} \texttt{R} package
-  \cite[]{friedman_regularization_2010}, and the ADVI posterior means,
+  estimates of the model parameters. We compare these MAP estimates, and the 
+  ADVI posterior means,
   to the MAP estimates obtained from our model through the correlation with
@@
   true prior variance. We
-  repeated the simulations 100 times and report the results in Figure 
-  \ref{fig:boxplots_igaussian_res4.2_post1}.
+  repeated the simulations 100 times and report the results in Figures 
+  \ref{fig:boxplots_igaussian_res4.2_post1}-
+  \ref{fig:boxplots_igaussian_res4.2_post2}.
 
-<<boxplots_igaussian_res4.2_post1, fig.cap="(a) \\& (d) Cramér-von Mises criterion, (b) \\& (e) correlation, and (c) \\& (f) MSD between the $\\beta_j$ point estimates of the MCMC and variational bayes posteriors.", out.width="80%", fig.asp=2/3>>=
+  In addition to the performance measures, we have plotted posterior samples 
+  histograms of 8 randomly selected model parameters in setting 1, 
+  along with the esimated
+  marginal variational Bayes posteriors, and the MAP estimates in Figure 
+  \ref{fig:hist_igaussian_fit4}.
+
+<<boxplots_igaussian_res4.2_post1, fig.cap="Comparison of different approximations in (a)-(c) simulation setting 1, (d)-(f) setting 2, and (g)-(i) setting 3, through Cramér-von Mises criterion between the approximate and MCMC posteriors, correlation and MSD between the approximate and MCMC posterior means.", out.width="80%", fig.asp=1>>=
+@
+
+<<boxplots_igaussian_res4.2_post2, fig.cap="Comparison of different approximations in (a)-(c) simulation setting 4, (d)-(f) setting 5, and (g)-(i) setting 6, through Cramér-von Mises criterion between the approximate and MCMC posteriors, and correlation and MSD between the approximate and MCMC posterior means.", out.width="80%", fig.asp=1>>=
+@
+
+<<hist_igaussian_fit4, fig.cap="Histogram of posterior MCMC samples of 8 randomly selected model parameters in setting 1, along with the estimated variational posteriors and MAP estimates.", out.width="80%", fig.asp=1>>=
 @

### 2019-08-19 docs/manuscript2.Rnw
unpenalized covariates added to rpackage, slight reorganization

@@
 	$x_{ij}$, $j=1,\dots, p$, collected in $\x_i = \begin{bmatrix} x_{i1} & 
-	\dots & x_{ip} \end{bmatrix} \tr$. We have an \textit{a priori} (possibly 
-	informative) partitioning 
-	of the omics features into $g=1, \dots, G$ non-overlapping groups.
-	For convenience, we let 
-	$g(j)$ denote the group index of molecular feature $j$. Furthermore, we have 
+	\dots & x_{ip} \end{bmatrix} \tr$. Furthermore, we have 
 	characteristics of the drugs available in the form of external variables 
@@
 	We regress the drug efficacies on the molecular feature effects $\beta_{jd}$:
-	\begin{subequations}
+	\begin{subequations}\label{eq:likelihood}
 		\begin{align}
@@
 		\begin{align}
-		  \beta_{jd} & \sim \mathcal{N}_p (0, \phi^2_{g(j)} \gamma_{jd}^2 
-		  \sigma_d^2), 
+		  \beta_{jd} | \gamma^2_{jd}, \sigma^2_d & \sim \mathcal{N}_p 
+		  (0, \gamma_{jd}^2 \sigma_d^2), 
 		  \label{eq:betaprior}\\
@@
 	where $\mathcal{IG}(\theta, \lambda)$ denotes an inverse Gaussian distribution 
-	with mean $\theta$ and shape $\lambda$. The different molecular feature types
-	are given their own weights through the $\phi^2_{g}$, while the external 
+	with mean $\theta$ and shape $\lambda$. The external 
 	outcome-feature data comes in through 
 	our mean model for the $\gamma_{jd}^2$ hyperprior model: 
-	$[(\mathbf{c}_d^j) \tr \bm{\alpha}]^{-1}$. Both the $\phi^2_{g}$ and
-	$\bm{\alpha}$ are estimated in a data-driven manner. More details on the model
+	$[(\mathbf{c}_d^j) \tr \bm{\alpha}]^{-1}$. The
+	$\bm{\alpha}$ and $\lambda_d$ are estimated in a data-driven manner. 
+	More details on the model
 	itself, as well as its estimation, may be found in 
 	the SM and in [! ref to first paper here].
+
+	\subsection{Tissue effect}
+	The cell lines used in drug response models are often taken from different
+	tissues. In addition, other clinical covariates are often available. 
+	For unbiased omics feature effects, one may wish to account for these, 
+	without shrinkage of their effects. We do that by adding such unpenalized 
+	covariates 
+	$\mathbf{z}_{i} = \begin{bmatrix} z_{i1} & \cdots & z_{iT} \end{bmatrix} \tr$ 
+	to model (\ref{eq:likelihood}):
+  \begin{align*}
+    y_{id} & = \x_i \tr \bbeta_d + \mathbf{z}_i \tr \bm{\psi} + 
+    \epsilon_{id}, \\
+    \text{with } & \epsilon_{id} \sim \mathcal{N}(0, \sigma_d^2),
+	\end{align*}
+	with $\bm{\psi} = \begin{bmatrix} \psi_1 & \cdots & \psi_T \end{bmatrix} \tr$ 
+	the unpenalized covariates effects. We assign $\bm{\psi}$ a flat prior, which 
+	alters the estimation as described in SM Section [! ref to section here].
 

### 2019-08-19 docs/supplement2.Rnw
unpenalized covariates added to rpackage, slight reorganization

--- 2019-02-19 docs/supplement.Rnw
@@
 % settings
-\pgfplotsset{compat=1.13}
+\graphicspath{{../figs/}}
+\pgfplotsset{compat=1.16}
 \setlength{\evensidemargin}{.5cm} \setlength{\oddsidemargin}{.5cm}
 \setlength{\textwidth}{15cm}
-\title{Supplementary material to: Drug efficacy prediction in cell lines}
+\title{Supplementary material to: Empirical Bayes estimation of the normal
+inverse Gaussian prior in 
+simultaneous-equation models}
 \date{\today}
-\author{Magnus M.
-M\"unch$^{1,2}$\footnote{Correspondence to: \href{mailto:m.munch@vumc.nl}{m.munch@vumc.nl}}, Mark A.
-van de Wiel$^{1,3}$, Sylvia Richardson$^{3}$, \\ and Gwena{\"e}l G.
-R.
-Leday$^{3}$}
+\author{Magnus M. M\"unch$^{1,2}$\footnote{Correspondence to: 
+\href{mailto:m.munch@vumc.nl}{m.munch@vumc.nl}}, Mark A. van de Wiel$^{1,3}$, 
+Sylvia Richardson$^{3}$, \\ and Gwena{\"e}l G. R. Leday$^{3}$}
 
@@
 	\noindent
-	1. Department of Epidemiology \& Biostatistics, Amsterdam Public Health research institute, VU University Medical Center, PO Box 7057, 1007 MB
-	Amsterdam, The Netherlands\\
-	2. Mathematical Institute, Leiden University, Leiden, The Netherlands \\
-	3. MRC Biostatistics Unit, Cambridge Institute of Public Health, Cambridge, United Kingdom \\
+	1. Department of Epidemiology \& Biostatistics, Amsterdam Public Health
+	research institute, Amsterdam University medical centers, PO Box 7057, 1007 MB
+	Amsterdam, The Netherlands \\*
+	2. Mathematical Institute, Leiden University, Leiden, The Netherlands \\*
+	3. MRC Biostatistics Unit, Cambridge Institute of Public Health, Cambridge, 
+	United Kingdom
 
 	\section{Content overview}
-	This document contains the Supplementary Material (SM) to the document `Drug efficacy prediction in cell lines'. In the following, this document is referred to as Main Document (MD).
+	This document contains the Supplementary Material (SM) to the document `Drug
+	efficacy prediction in cell lines'. In the following, this document is
+	referred to as Main Document (MD).
 
-	\section{Variational Bayes derivations}\label{sec:vbderivations}
-	In the following all expectations are with respect to the variational posterior $Q$.
-	\begin{align*}
-	\log q_{\B} (\B) & \propto \E_{\mathbf{U}, \bsigma^2} [\log \mathcal{L}(\Y | \B, \U, \bsigma^2 )] + \E_{\bgamma^2, \bsigma^2} [\log \pi (\B | \bgamma^2, \bsigma^2)] \\
-	& \propto - \frac{1}{2} \sum_{d=1}^D \sum_{i=1}^n \E_{\mathbf{u}_d, \sigma_d^2} \[ \frac{(y_{id} - \x_i \tr \bbeta_d - \mathbf{z}_i \tr \mathbf{u}_d)^2}{\sigma_d^2} \] - \frac{1}{2} \sum_{d=1}^D \sum_{j=1}^p \phi_{g(j)}^{-2} \E_{\gamma^2_d, \sigma^2_d} \( \frac{\beta_{jd}^2}{\gamma_d^2 \sigma_d^2} \), \\
-	q_{\B} (\B) & \overset{D}{=} \prod_{d=1}^{D} \mathcal{N}_p (\bmu_d, \bSigma_d), \\
-	& \text{ with } \bSigma_d = \E(\sigma_d^{-2})^{-1} [\X \tr \X + \E(\gamma_d^{-2}) \cdot \diag (\phi_{g(j)}^{-2}) ]^{-1}, \\
-	& \text{ and } \bmu_d = [\X \tr \X + \E(\gamma_d^{-2}) \cdot \diag (\phi_{g(j)}^{-2}) ]^{-1} \X \tr [\y_d - \Z \cdot \E(\mathbf{u}_d)].
-	\end{align*}
-
-	\begin{align*}
-	\log q_{\U}(\U) & \propto \E_{\mathbf{B}, \bsigma^2} [\log \mathcal{L}(\Y | \B, \U, \bsigma^2 )] + \E_{\mathbf{\Xi}_1, \dots, \mathbf{\Xi}_D, \bsigma^2} [\log \pi (\U | \mathbf{\Xi}_1, \dots, \mathbf{\Xi}_D, \bsigma^2)] \\
-	& \propto - \frac{1}{2} \sum_{d=1}^D \sum_{i=1}^n \E_{\bbeta_d, \sigma_d^2} \[ \frac{(y_{id} - \x_i \tr \bbeta_d - \mathbf{z}_i \tr \mathbf{u}_d)^2}{\sigma_d^2} \] - \frac{1}{2} \sum_{d=1}^D \E_{\bm{\Xi}_d, \sigma_d^2}\( \frac{\mathbf{u}_d \tr \bm{\Xi}_d^{-1} \mathbf{u}_d}{\sigma_d^2} \), \\
-	q_{\U}(\U) & \overset{D}{=} \prod_{d=1}^{D} \mathcal{N}_T (\mathbf{m}_d, \Sm_d), \\
-	& \text{ with } \Sm_d = \E(\sigma_d^{-2})^{-1} [\Z \tr \Z + \E(\bm{\Xi}_d^{-1}) ]^{-1}, \\
-	& \text{ and } \mathbf{m}_d = [\Z \tr \Z + \E(\bm{\Xi}_d^{-1}) ]^{-1} \Z \tr [\y_d - \X \cdot \E(\bbeta_d)].
-	\end{align*}
-
-	\begin{align*}
-	\log q_{\bgamma^{2}}(\bgamma^{2}) & \propto \E_{\B, \bsigma^2} [\log \pi (\B | \bgamma^{2}, \sigma^2)] + \log \pi (\bgamma^{2}) \\
-	& \propto -\frac{p}{2} \sum_{d=1}^D \log \gamma_d^{2} - \frac{1}{2} \sum_{d=1}^D \sum_{j=1}^p \phi_{g(j)}^{-2} \E \( \frac{\beta_{jd}^2}{\sigma^{2}_d}\) \gamma_d^{-2} - \frac{3}{2} \sum_{d=1}^D \log \gamma_d^{2} \\
-	& \,\,\,\,\,\,\,\,\,\, - \frac{\lambda}{2 \theta^2} \sum_{d=1}^D (\gamma_d^2 - \theta)^2 \gamma_d^{-2} \\
-	& \propto \sum_{d=1}^D \left\{ \( -\frac{p + 1}{2} - 1\) \log \gamma_d^{2} - \frac{\lambda}{2\theta^2} \gamma_d^2 - \frac{1}{2}\[ \lambda + \E(\sigma_d^{-2}) \sum_{j=1}^p \frac{\E(\beta_{jd}^2)}{\phi_{g(j)}^{2}}\] \gamma_d^{-2} \right\}, \\
-	q_{\bgamma^{2}}(\bgamma^{2}) & \overset{D}{=} \prod_{d=1}^D \mathcal{GIG} \(-\frac{p+1}{2}, \frac{\lambda}{\theta_d^2}, \delta_{d} \), \\
-	& \text{with } \delta_{d} =\E(\sigma_d^{-2}) \sum_{j=1}^p \frac{\E(\beta_{jd})^2 + \V(\beta_{jd})}{\phi_{g(j)}^2} + \lambda.
-	\end{align*}
-
-	\begin{align*}
-	\log q_{\bsigma^{2}}(\bsigma^{2}) & \propto \E_{\B,\mathbf{U}} [\log \mathcal{L}(\Y | \B, \U, \bsigma^2 )] + \E_{\B, \bgamma^2} [\log \pi (\B | \bgamma^2, \bsigma^2)] \\
-	& \,\,\,\,\,\,\,\,\,\, + \E_{\mathbf{U}, \bm{\Xi}_1, \dots, \bm{\Xi}_D} [\log \pi (\mathbf{U} | \bm{\Xi}_1, \dots, \bm{\Xi}_D, \bsigma^2)] + \log \pi (\bsigma^{2}) \\
-	& \propto -\frac{n}{2} \sum_{d=1}^D \log \sigma_d^{2} - \frac{1}{2} \sum_{d=1}^D \sum_{i=1}^n \E [(y_{id} - \x_i \tr \bbeta_d - \mathbf{z}_i \tr \mathbf{u}_d)^2] \sigma_d^{2} - \frac{p}{2} \sum_{d=1}^D \log \sigma_d^{-2} \\
-	& \,\,\,\,\,\,\,\,\,\, - \frac{1}{2} \sum_{d=1}^D \sum_{j=1}^p \E( \gamma_d^{-2}) \frac{\E ( \beta_{jd}^2)}{\phi_{g(j)}^{2}} \sigma_d^{-2} - \frac{T}{2} \sum_{d=1}^D \log \sigma_d^{2} - \frac{1}{2} \sum_{d=1}^D \E(\mathbf{u}_d \tr \bm{\Xi}_d^{-1} \mathbf{u}) \sigma_d^{-2} \\
-	& \,\,\,\,\,\,\,\,\,\, - \frac{3}{2} \sum_{d=1}^D \log \sigma^2_d \\
-	& = -\sum_{d=1}^{D} \(\frac{n + p + T + 3}{2} \) \log \sigma_d^{2} - \sum_{d=1}^{D} \frac{1}{2}\Bigg\{ \sum_{i=1}^n \E [(y_{id} - \x_i \tr \bbeta_d - \mathbf{z}_i \tr \mathbf{u}_d)^2] \\
-	& \,\,\,\,\,\,\,\,\,\, + \E(\gamma_d^{-2})\sum_{j=1}^p \frac{\E(\beta_{jd}^2)}{\phi_{g(j)}^2} + \E(\mathbf{u}_{d} \tr \bm{\Xi}_d^{-1} \mathbf{u}_d) \Bigg\} \sigma_d^{-2}, \\
-	q_{\bsigma^{2}}(\bsigma^{2}) & \overset{D}{=} \prod_{d=1}^D \Gamma^{-1} \(\frac{n + p + T + 1}{2}, \zeta_{d} \), \\
-	& \text{ with } \zeta_{d} = \frac{\y_d \tr \y_d}{2} - \y_d \tr [\X \E(\bbeta_d) + \Z \E(\mathbf{u}_d)] + \E(\bbeta_d \tr) \X \tr \Z \E (\mathbf{u}_d) \\
-	& \,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\, + \frac{\trace [\X \tr \X \V(\bbeta_d)]}{2} + \frac{\E(\bbeta_d \tr) \X \tr \X \E(\bbeta_d)}{2} + \frac{\trace [\Z \tr \Z \V(\mathbf{u}_d)]}{2} \\
-	& \,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\, + \frac{\E(\mathbf{u}_d \tr) \Z \tr \Z \E(\mathbf{u}_d)}{2} + \frac{\E(\gamma_d^{-2})}{2} \sum_{j=1}^p \frac{\E(\beta_{jd})^2 + \V(\beta_{jd})}{\phi_{g(j)}^2} \\
-	& \,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,+ \frac{\E(\mathbf{u}_d \tr) \E(\bm{\Xi}_d^{-1}) \E(\mathbf{u}_d)}{2} + \frac{\trace [\E(\bm{\Xi}_d^{-1}) \V(\mathbf{u}_d)]}{2},
-	\end{align*}
-
-	\begin{align*}
-	\log q_{\bm{\Xi}_1, \dots, \bm{\Xi}_D}(\bm{\Xi}_1, \dots, \bm{\Xi}_D) & \propto \E_{\mathbf{U}, \bm{\sigma}^2} [\log \pi (\mathbf{U} | \bm{\Xi}_1, \dots, \bm{\Xi}_D, \bsigma^2)] + \log \pi (\bm{\Xi}_1, \dots, \bm{\Xi}_D) \\
-	& \propto -\frac{1}{2} \sum_{d=1}^D \log |\bm{\Xi}_d| - \frac{1}{2} \sum_{d=1}^D \E \(\frac{\mathbf{u}_d \tr \bm{\Xi}_d^{-1} \mathbf{u}_d}{\sigma_d^2} \) \\
-	& \,\,\,\,\,\,\,\,\,\, - \frac{k + T + 1}{2} \sum_{d=1}^D \log |\bm{\Xi}_d| - \frac{1}{2} \sum_{d=1}^D \trace (\bm{\Omega} \bm{\Xi}_d^{-1}) \\
-	& = -\frac{\tau + T + 2}{2} \sum_{d=1}^D \log |\bm{\Xi}_d| - \frac{1}{2} \sum_{d=1}^D \trace \left\{ \[\E(\sigma_d^{-2}) \E( \mathbf{u}_d \mathbf{u}_d \tr) + \bm{\Omega}\] \bm{\Xi}_d^{-1} \right\}, \\
-	q_{\bm{\Xi}_1, \dots, \bm{\Xi}_D}(\bm{\Xi}_1, \dots, \bm{\Xi}_D) & \overset{D}{\propto} \prod_{d=1}^D \mathcal{W}_T^{-1} (\bm{\Psi}_d, \nu + 1),  \\ 
-	& \text{ with } \bm{\Psi}_d =\E(\sigma^{-2}_d) \[\V(\mathbf{u}_d) + \E(\mathbf{u}_d) \E( \mathbf{u}_d \tr) \] + \bm{\Omega}.
-	\end{align*}
-
-	\section{Simulations}
-	\subsection{Models}
-	In order to investigate the inverse Gaussian prior model and its estimation, we consider a simpler model. The simpler model ignores tissue effects and fixes $\forall g: \phi_g^2=1$:
-	\begin{align*}
-  y_{id} | \bm{\beta}_d, \sigma_d^2 & \sim \mathcal{N} (\x_i \tr \bm{\beta}_d, \sigma_d^2), \\
-  \beta_{jd} | \gamma_d^2, \sigma_d^2 & \sim p(\beta_{jd}), \\
-  \gamma_d^2 & \sim p(\gamma_d^2), \\
-  \sigma_d^2 & \sim 1/\sigma_d^3.
-  \end{align*}
-
-  \cite{moran_variance_2018} argue that in regression models, $\bm{\beta}_d$ and $\sigma_d^2$ should be independent \textit{a priori}.
-However, in our case that is debatable.
-The $D$ regressions need not be on the same scale, so some form of calibration of the $\bm{\beta}_d$ posteriors to this scale is required.
-A simple way of doing this is to include the scales $\sigma_d^2$ in the priors for the $\bm{\beta}_d$.
-We call the two versions of the model conjugate (with $\bm{\beta}_d$ and $\sigma_d^2$ dependent):
+	\section{Unpenalized covariates}
+  Inclusion of an intercept $\beta_0$ and other unpenalized covariates is 
+  achieved by letting $\X = \begin{bmatrix}\X_u & 
+  \X_r\end{bmatrix}$, where $\X_u$ are the $u$
+  unpenalized variables (possibly with a column of ones if an intercept is
+  to be included) and $\X_r$ are the $r$ penalized variables. 
+  Additionally, we let 
   $$
-  \beta_{jd} | \gamma_d^2, \sigma_d^2 \sim \mathcal{N} (0, \sigma_d^2 \gamma_d^2)
+  \mathbf{H} = 
+  \begin{bmatrix}
+  \mathbf{0}_{u \times u} & \mathbf{0}_{u \times r} \\
+  \mathbf{0}_{r \times u} & \diag[\E(\gamma_{jd}^{-2})]. \\
+  \end{bmatrix}
   $$
-  and non-conjugate (with $\bm{\beta}_d$ and $\sigma_d^2$ independent):
-  $$
-  \beta_{jd} | \gamma_d^2, \sigma_d^2 \sim \beta_{jd} | \gamma_d^2 \sim \mathcal{N} (0, \gamma_d^2),
-  $$
-  and compare them in our simulations.
-
-	In addition to the inverse Gaussian model for the prior variances $\gamma_d^2$ (introduced in the MD):
-	$$
-  \gamma_d^2 \sim \mathcal{IG}(\theta_d, \lambda_d),
-  $$
-	we investigate a generalisation of the models in \cite{kpogbezan_empirical_2017} and \cite{leday_gene_2017}, where we draw the $\gamma_d^2$ from an inverse Gamma distribution; :
-	$$
-	\gamma_d^2 \sim \Gamma^{-1}(\eta_d/2, \lambda_d/2).
-	$$
-	For the sake of comparability, we have rescaled the inverse Gamma parameters and reparametrised both distributions as generalized inverse Gaussians:
-	\begin{align*}
-  \gamma_d^2 & \sim \mathcal{GIG}(-1/2, \lambda_d \theta_d^{-2}, \lambda_d), \\
-  \gamma_d^2 & \sim \mathcal{GIG}(-\eta_d/2, 0, \lambda_d).
-  \end{align*}
-	The two models coincide if $\theta_d \to \infty$ and $\eta_d = 1$.
-
-	\subsection{Estimation}
-	The variational distributions are as follows:
-	\begin{align*}
-  q(\bm{\beta}) & \overset{D}{=} \prod_{d=1}^D \mathcal{N}_p (\bm{\mu}_d, \bm{\Sigma}_d), \\
-  q(\bm{\gamma}^2) & \overset{D}{=} \prod_{d=1}^D \mathcal{GIG}\(-\frac{p+\eta_d}{2}, \lambda_d \theta_d^{-2}, \delta_d\), \\
-  q(\bm{\sigma}^2) & \overset{D}{=} \prod_{d=1}^D \Gamma^{-1} \(\frac{df + 1}{2}, \zeta_d\),
-  \end{align*}
-	where $\eta_d=1$ and $\theta_d \to \infty$ in the inverse Gaussian and inverse Gamma models, respectively.
-Furthermore, $df=n + p$ in the conjugate model and $df=n$ in the non-conjugate model.
-The variational parameters contain cyclic dependencies and are, in the conjugate setting, iteratively updated by:
-  \begin{align*}
-  \bm{\Sigma}_d^{(h+1)} & = (a_d^{(h)})^{-1} (\X \tr \X + c_d^{(h)} \I)^{-1}, \\
-  \bm{\mu}_d^{(h+1)} & = (\X \tr \X + c_d^{(h)} \I)^{-1} \X \tr \y_d, \\
-  \delta_d^{(h+1)} & = b_d^{(h)} (c_d^{(h)})^{-1} \bmu_d \tr \bmu_d + \lambda_d, \\ 
-  \zeta_d^{(h+1)} & = \frac{1}{2} \bigg[\mathbf{y}_d \tr \mathbf{y}_d -2 \mathbf{y}_d \tr \X \bm{\mu}_d^{(h+1)} + \trace ( \X \tr \X \bm{\Sigma}_d^{(h+1)}) + (\bm{\mu}_d^{(h+1)}) \tr \X \tr \X \bm{\mu}_d^{(h+1)} \\
-  & \,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\, + \frac{b_d^{(h+1)} - c_d^{(h+1)} a_d^{(h+1)}}{1 - a_d^{(h+1)}} \trace ( \bm{\Sigma}_d^{(h+1)}) + \frac{b_d^{(h+1)} - c_d^{(h+1)} a_d^{(h+1)}}{1 - a_d^{(h+1)}} (\bm{\mu}_d^{(h+1)}) \tr \bm{\mu}_d^{(h+1)}\bigg].
-  \end{align*}
-  Here, we set
-  \begin{align}
-  a_d^{(h)} & =\E_{Q^{(h)}}(\sigma_d^{-2})=(df+1)/(2 \zeta_d^{(h)}), \nonumber \\
-  b_d^{(h)} & = \E_{Q^{(h)}}(\gamma_d^{-2}) = \sqrt{\frac{\lambda_d}{\theta_d^2 \delta_d^{(h)}}} \frac{K_{\frac{p + \eta_d -2}{2}} \( \sqrt{\lambda_d \delta_d^{(h)}}/\theta_d \)}{K_{\frac{p+\eta_d}{2}} \( \sqrt{\lambda_d \delta_d^{(h)}}/\theta_d \)} + \frac{p+\eta_d^{(h)}}{\delta_d^{(h)}}, \label{eq:auxvar1} \\
-  c_d^{(h)} & = \begin{cases*}
-  b_d^{(h)} & in the conjugate setting \\
-  b_d^{(h)}/a_d^{(h)} & in the non-conjugate setting .
-  \end{cases*} \nonumber
+  Then,
+  \begin{align}\label{eq:unpencov}
+    \bSigma &= \E(\sigma_d^{-2})^{-1}\(\X \X + \mathbf{H} \)^{-1} \nonumber \\
+    & = 
+    \E(\sigma_d^{-2})^{-1}\begin{bmatrix} 
+      \X_u \tr \X_u & \X_u \tr \X_r \\
+      \X_r \tr \X_u & \X_r \tr \X_r + \diag[\E(\gamma_{jd}^{-2})]
+    \end{bmatrix}^{-1}.
   \end{align}
-  Note that in the inverse Gamma model, the computation of (\ref{eq:auxvar1}) simplifies to $b_d^{(h)}=(p+\eta_d)/\delta_d^{(h)}$.
-
-  \subsection{Empirical Bayes}
-  To avoid subjectivity in our choice of prior, we estimate the hyper-parameters with empirical Bayes.
-The empirical Bayes procedure differs for the two prior models.
-In the inverse Gaussian case, we parametrise $\theta_d = (\mathbf{c}_d \tr \bm{\alpha})^{-1}$ to allow for the inclusion of drug covariates in the empirical Bayes estimation.
-The estimating equations are:
-  \begin{align*}
-  \bm{\alpha} & = \[ \mathbf{C} \tr \diag (\lambda_d e_d^{(l)}) \mathbf{C} \]^{-1} \mathbf{C} \tr \diag (\lambda_d) \mathbf{1}_{D \times 1}, \\
-  \lambda_d & = \[ b_d^{(l)} + e_d^{(l)} (\mathbf{c}_d \tr \bm{\alpha})^2 - 2\mathbf{c}_d \tr \bm{\alpha} \]^{-1},
-  \end{align*}
-  where $e_d^{(l)} = \E_{Q^{(l)}}(\gamma_d^2) = [b_d^{(l)} - (p + \eta_d^{(h)})/\delta_d^{(l)})] \cdot \delta_d^{(l)} (\theta_d^{(h)})^2/\lambda_d^{(l)}$.
-Solving the equations is done by iteratively reweighted least squares of responses $(e_d^{(l)})^{-1}$ on predictors $\mathbf{c}_d$ with weights $\lambda_d e_d^{(l)}$.
-If we use one common $\lambda=\lambda_d$ then the solution is in closed form:
-  \begin{align*}
-	\bm{\alpha}^{(l+1)} & = \[ \mathbf{C} \tr \diag(e_d^{(l)}) \mathbf{C} \]^{-1} \mathbf{C} \tr \mathbf{1}_{D \times 1}, \\
-	\lambda^{(l+1)} & = \[ \sum_{d=1}^D b_d^{(l)} - (\bm{\alpha}^{(l+1)}) \tr \mathbf{C} \tr \mathbf{1} \]^{-1} D.
-	\end{align*}
-	This corresponds to the model in the MD.
-
-  In the inverse Gamma model we use the following empirical Bayes procedure: Suppose we have a partitioning of our drugs into drug classes and we let prior variance components $\gamma_d^2$ and $\gamma^2_{d'}$ be drawn from the same distribution if drug $d$ and drug $d'$ are from the same class.
-Or simply put: $\gamma^2_{d}, \gamma^2_{d'} \sim \Gamma^{-1}(\eta_c/2, \lambda_c/2)$ if $d, d' \in \text{class}_c$.
-Empirical bayes estimation of the $\eta_c$ and $\lambda_c$ by maximisation of the (approximated) marginal likelihood amounts to iteratively solving the estimating equations:
-  \begin{subequations}
-    \begin{align}
-      \psi\(\frac{\eta_c}{2}\) & = \log \lambda_c - \frac{\sum_{d \in \text{class}_c} e^{(l)}_d}{|\text{class}_c|} - \log 2, \label{eq:estequation1}\\
-      \lambda_c & = \eta_c \cdot |\text{class}_c| \cdot \(\sum_{d \in \text{class}_c} b_d^{(l)}\)^{-1}, \label{eq:estequation2}
-    \end{align}
-  \end{subequations}
-  until convergence, where $e^{(l)}_d = \E_{Q^{(l)}}(\log \gamma_d^2) = \log \delta_d^{(l)} - \psi [ (p + \eta_d^{(l)})/2] - \log 2$. 
-  We propose to do that by first solving (\ref{eq:estequation2}) for $\lambda_c$, plugging it into (\ref{eq:estequation1}) and bounding the result using \cite{alzer_inequalities_1997} to obtain an interval $( \eta_c^*, 2\eta_c^* )$ containing the solution to (\ref{eq:estequation1}), where:
-  $$
-  \eta_c^* = \left\{ \frac{\sum_{d \in \text{class}_c} e^{(l)}_d}{|\text{class}_c|} + \log \[ \frac{\sum_{d \in \text{class}_c} b_d^{(l)}}{|\text{class}_c|} \]  \right\}^{-1}.
-  $$
-  With this interval it is straightforward to find $\alpha^{(l+1)}_c$ by any root-finding algorithm and plugging the solution into (\ref{eq:estequation2}) to find $\lambda^{(l+1)}_c$.
-We may use $\eta_c^*$ as a starting value.
-
-	The inverse Gamma model with independent drug classes allows for more flexible empirical Bayes estimation of the prior means for the different drug classes.
-As a consequence however, the risk of overfitting is increased.
-Additionally, it does not allow to include continuous covariates on the drugs.
-
-	\subsection{Evidence lower bound}
-	We monitor convergence through the evidence lower bound (ELBO).
-To compute the evidence lower bound after updating the EB parameters from $\eta_d^{(l)}, \theta_d^{(l)}, \lambda_d^{(l)}$ to $\eta_d^{(l + 1)}, \theta_d^{(l + 1)}, \lambda_d^{(l + 1)}$, we compute $\text{ELBO}^{(l + 1)} = \sum_{d=1}^D \text{ELBO}_d^{(l + 1)}$, with, in the inverse Gaussian case:
-	\begin{align*}
-	\text{ELBO}_d^{(l + 1)} & = \frac{1}{2} \log |\bm{\Sigma}_d^{(l)}| - \frac{a_d^{(l)}}{2} \[ \y_d \tr \y_d - 2\y_d \tr \X \bm{\mu}_d^{(l)} + (\bm{\mu}_d^{(l)}) \tr \X \tr \X \bm{\mu}_d^{(l)} + \trace ( \X \tr \X \bm{\Sigma}_d^{(l)})\] \\
-	& + \frac{b_d^{(l)}}{2} \[ \delta_d^{(l)} - (a_d^{(l)} + 1 - b_d^{(l)}/c_d^{(l)})(\bm{\mu}_d^{(l)}) \tr \bm{\mu}_d^{(l)} - (a_d^{(l)} + 1 - b_d^{(l)}/c_d^{(l)}) \trace (\bm{\Sigma}_d^{(l)}) \] \\
-	& - \frac{df + 1}{2} \log \zeta_d^{(l)} + \frac{p + 1}{4} \log \lambda_d^{(l)} - \frac{p + 1}{2} \log \theta_d^{(l)} - \frac{p+1}{4} \log \delta_d^{(l)} \\
-	& + \log K_{\frac{p+1}{2}}\(\sqrt{\lambda_d^{(l)} \delta_d^{(l)}}/\theta_d^{(l)}\) + \frac{\lambda_d^{(l)} e_d^{(l)}}{2(\theta_d^{(l)})^2} \\
-	& + \frac{\lambda_d^{(l+1)}}{\theta_d^{(l+1)}} + \frac{1}{2}\log \lambda_d^{(l+1)} - \frac{\lambda_d^{(l+1)} e_d^{(l)}}{2 (\theta_d^{(l + 1)})^2} - \frac{\lambda_d^{(l + 1)} b_d^{(l)}}{2}.
-	\end{align*}
-
-	In the inverse Gamma setting, we have:
-	\begin{align*}
-	\text{ELBO}_d^{(l + 1)} & = \frac{1}{2} \log |\bm{\Sigma}_d^{(l)}| - \frac{a_d^{(l)}}{2} \[ \y_d \tr \y_d - 2\y_d \tr \X \bm{\mu}_d^{(l)} + (\bm{\mu}_d^{(l)}) \tr \X \tr \X \bm{\mu}_d^{(l)} + \trace ( \X \tr \X \bm{\Sigma}_d^{(l)})\] \\
-	& + \frac{b_d^{(l)}}{2} \[ \delta_d^{(l)} - (a_d^{(l)} + 1 - b_d^{(l)}/c_d^{(l)})(\bm{\mu}_d^{(l)}) \tr \bm{\mu}_d^{(l)} - (a_d^{(l)} + 1 - b_d^{(l)}/c_d^{(l)}) \trace (\bm{\Sigma}_d^{(l)}) \]  \\
-	& - \frac{df + 1}{2} \log \zeta_d^{(l)} - \frac{p + \eta_d^{(l)}}{2} \log \delta_d^{(l)} + \frac{\eta_d^{(l)}}{2} e_d^{(l)} + \frac{\eta_d^{(l)}}{2} \log 2 + \log \Gamma \( \frac{p + \eta_d^{(l)}}{2} \) \\
-	& - \log \Gamma \( \frac{\eta_d^{(l + 1)}}{2} \) - \frac{\eta_d^{(l+1)}}{2} e_d^{(l)} - \frac{\eta_d^{(l+1)}}{2} \log 2 + \frac{\eta_d^{(l + 1)}}{2} \log \lambda_d^{(l+1)} - \frac{\lambda_d^{(l + 1)} b_d^{(l)}}{2}.
-	\end{align*}
-  Note that the computation of $e_d^{(l)}$ differs between the inverse Gaussian and Gamma models.
-Furthermore, to compute the ELBO after a VB update we simply let $\eta_d^{(l)}, \theta_d^{(l)}, \lambda_d^{(l)}=\eta_d^{(l + 1)}, \theta_d^{(l + 1)}, \lambda_d^{(l + 1)}$.
-
-  In practice, there are different options to assess convergence.
-We may monitor the parameters themselves, the ELBO, or a combination of the two.
-We may also iterate for an \textit{a priori} fixed number of times.
-We may also choose to use a different convergence criterium for the EB and VB iterations.
-Currently we fix the number of EB iterations to 20, while we do 2 VB iterations per EB iteration.
-
-  \subsection{Efficient computation}
-  The empirical Bayes updates require the following quantities: $\zeta_d$, $a_d$, $\delta_d$, $\trace (\bm{\Sigma}_d)$, $\trace (\X \tr \X \bm{\Sigma}_d)$, $\bm{\mu}_d \tr \bm{\mu}_d$, and $\bm{\mu}_d \tr \X \tr \X \bm{\mu}_d$.
-In addition, we need $\log|\bm{\Sigma}|$ and $\mathbf{y}_d \tr \X \bm{\mu}_d$ to monitor the ELBO.
-The first three quantities are obtained by scalar operations of $\mathcal{O}(n)$.
-Let $c_d^{(h)}=b_d^{(h)}$ in the conjugate setting and $c_d^{(h)}=b_d^{(h)}/a_d^{(h)}$ in the non-conjugate setting.
-Then, beforementioned quantities are easily calculated using the SVD $\X = \mathbf{U} \mathbf{D} \mathbf{V} \tr$:
-  \begin{align*}
-  \trace (\bm{\Sigma}_d^{(h+1)}) & = (a_d^{(h)})^{-1} \[\sum_{i=1}^n (v_i^2 + c_d^{(h)})^{-1} + \max (p - n, 0)\cdot (c_d^{(h)})^{-1}\], \\
-  \trace (\X \tr \X \bm{\Sigma}_d^{(h+1)}) & = (a_d^{(h)})^{-1} \sum_{i=1}^n v_i^2(v_i^2 + c_d^{(h)})^{-1}, \\
-  (\bm{\mu}_d^{(h+1)}) \tr \bm{\mu}_d^{(h+1)} & = \sum_{i=1}^n v_i^2 [(\mathbf{U} \tr \mathbf{y}_d)_i]^2/(v_i^2 + c_d^{(h)})^2, \\
-  (\bm{\mu}_d^{(h+1)}) \tr \X \tr \X \bm{\mu}_d^{(h+1)} & = \sum_{i=1}^n v_i^4 [(\mathbf{U} \tr \mathbf{y}_d)_i]^2/(v_i^2 + c_d^{(h)})^2, \\
-  \log|\bm{\Sigma}^{(h+1)}| & \propto -p \log a_d^{(h)} - \sum_{i=1} \log (v_i^2 + c_d^{(h)}) - \max(p - n, 0) \cdot \log c_d^{(h)}, \\
-  \mathbf{y}_d \tr \X \bm{\mu}_d^{(h+1)} & = \sum_{i=1}^n v_i^2 [(\mathbf{U} \tr \mathbf{y}_d)_i]^2/(v_i^2 + c_d^{(h)}).
-  \end{align*}
-  where $v_i$ are the singular values of $\X$.
-The SVD and $\mathbf{U} \tr \mathbf{y}_d$ are $\mathcal{O}(pn^2)$ and $\mathcal{O}(n^2)$ operations, respectively, but have to be calculated only once at the start of the algorithm.
-The rest of the calculations involve just $n$ scalar multiplications and/or additions.
-
-  \section{Simulation setting}
-  - check model with Stan.
-
-  - make one group of drugs and compare methods
-
-  - use drug-specific lambda's
-
-  - try smaller $p$ to see whether $\sigma_d^2$ become more identifiable.
-
-  - scale-invariant instead of conjugate.
-
-  We simulate from a correctly specified model given in \ref{sec:notissuemodel}.
-We assume random $y_{id}$, $\x_i$, and $\bm{\beta}_d$, while we fix $\mathbf{C}$, $\gamma_d^2$, $\sigma_d^2$ and $\bm{\alpha}$.
-We estimate the $\bm{\mu}_d$, $\bm{\Sigma}_d$, $\delta_d$, $\bm{\alpha}$ and $\lambda$, while we fix the $\sigma_d^2$ to their true values.
-
-  We set $n=100$, $p=200$, and $D=10$.
-We consider 5 evenly sized classes of drugs, such that $\mathbf{c}_{kd} = \mathbbm{1}(k = \text{class}_d)$ and $\begin{bmatrix} \mathbf{c}_1 & \cdots & \mathbf{c}_D \end{bmatrix} \tr$.
-From these the $\gamma_d^2 = (\mathbf{c}_d \tr \bm{\alpha})^{-1}$ are created.
-Next, we simulate $\bm{\beta}_{jd} \sim \mathcal{N} (0, \sigma^2_d \gamma_d^2)$.
-We fix the mean signal-to noise ratio $\overline{\text{SNR}}_d$, such that the predictor data variance is $s^2 = \V(\mathbf{x}_i) = \overline{\text{SNR}}_d / (\overline{\gamma^2_d} p)$ (NOT CORRECT).
-The last step consists of simulating the data by $\mathbf{x}_{ij} \sim \mathcal{N}(0,s^2)$ and $y_{id} = \mathcal{N} (\mathbf{x}_i \tr \bm{\beta}_d, \sigma_d^2)$.
-The remaining parameters are set as follows: $\bm{\alpha} = \begin{bmatrix} 1 & \dots & 5 \end{bmatrix} \tr$, $\sigma_d^2 = 1$ for all $d$, and $\overline{\text{SNR}}_d = 10$.
-To account for the random data and parameter generation, we repeat the simulation 100 times.
-We compare the results to the models in \ref{sec:indvarmodel} and \ref{sec:invgammamodel}, and present the results in Figures \ref{fig:boxplot_igaussian_res1_prior_mean}-\ref{fig:boxplot_igaussian_res2_prior_mean}.
-
-<<boxplot_igaussian_res1_prior_mean, fig.cap="Prior mean estimates", out.width="80%", fig.asp=2/3>>=
-@
-
-<<boxplot_igaussian_res2_prior_mean, fig.cap="Prior mean estimates", out.width="80%", fig.asp=2/3>>=
-@
+  With the choice of blocks as in (\ref{eq:unpencov}), blockwise inversion
+  renders the largest required matrix inverse $(\X_r \tr \bOmega \X_r +
+  \diag[\E(\gamma_{jd}^{-2})])^{-1}$, of dimension 
+  $r \times r$. Inversion of this matrix is done efficiently by applying the
+  Woodbury matrix identity.
 
@@
 
-  \section{Ratios of modified Bessel functions}
-  Ratios of modified Bessel functions of the second kind $K_{\alpha - 1}(x)/K_{\alpha}(x)$ are prone to under- and overflow for large $\alpha$.
-In our case, $\alpha$ increases linearly with $p$.
-Since $p$ may be large, this causes numerical issues in the calculation of various quantities.
-We alleviate the numerical issues through the following.
-
-  We let $n_1=p/2$ and $n_2=(p-1)/2$ and use the well-known recursive relation:
-  $$
-  K_{\alpha}(x) = K_{\alpha - 2}(x) + \frac{2(\alpha - 1)}{2} K_{\alpha- 1}(x),
-  $$
-  to rewrite the ratio:
-  \begin{align*}
-  \frac{K_{\frac{p - 1}{2}}(x)}{K_{\frac{p + 1}{2}}(x)} = 
-  \begin{cases}
-  \Big( \dots \Big( \( 1 + \frac{2 \cdot 1 - 1}{x} \)^{-1} + \frac{2 \cdot 2-1}{x}\Big)^{-1} + \dots + \frac{2 \cdot n_1 - 1}{x}\Big)^{-1}, & \text{for } p \text{ even}, \\
-  \Big( \dots \Big( \( \frac{K_0(x)}{K_1(x)} + \frac{2}{x} \cdot 1 \)^{-1} + \frac{2}{x} \cdot 2 \Big)^{-1} + \dots + \frac{2}{x} \cdot n_2 \Big)^{-1}, & \text{for } p \text{ odd}.
-  \end{cases}
-  \end{align*} 
-  The ratio $K_0(x)/K_1(x)$ is well-behaved, such that the ratio may be computed without numerical issues.
-
-  \section{General model}
-  We have drug covariates $\mathbf{c}_d$ and omics feature covariates $\mathbf{z}_j$. The model is now:
-  \begin{align*}
-  y_{id} | \bm{\beta}_d, \sigma_d^2 & \sim \mathcal{N} (\x_i \tr \bm{\beta}_d, \sigma_d^2), \\
-  \beta_{jd} | \tau_j^2, \gamma_d^2, \sigma_d^2 & \sim \mathcal{N} (0, \sigma_d^2 \gamma_d^2 \tau_j^2), \\
-  \gamma_d^2 & \sim \mathcal{IG}\((\mathbf{c}_d \tr \bm{\alpha})^{-1}, \lambda\), \\
-  \tau_j^2 & \sim \mathcal{IG}\((\mathbf{z}_j \tr \bm{\theta})^{-1}, \nu\), \\
-  \sigma_d^2 & \sim 1/\sigma_d^3.
-  \end{align*}
-  Variational posterior:
-  \begin{align*}
-  q(\bm{\beta}) & \overset{D}{=} \prod_{d=1}^D \mathcal{N}_p (\bm{\mu}_d, \bm{\Sigma}_d), \\
-  q(\bm{\gamma}^2) & \overset{D}{=} \prod_{d=1}^D \mathcal{GIG}\(-\frac{p+1}{2}, \lambda (\mathbf{c}_d \tr \bm{\alpha})^2, \delta_d\), \\
-  q(\bm{\tau}^2) & \overset{D}{=} \prod_{j=1}^p \mathcal{GIG}\(-\frac{D+1}{2}, \nu (\mathbf{z}_j \tr \bm{\theta})^2, \eta_j\), \\
-  q(\bm{\sigma}^2) & \overset{D}{=} \prod_{d=1}^D \Gamma^{-1} \(\frac{n + p + 1}{2}, \zeta_d\).
-  \end{align*}
-  Variational parameters:
-  \begin{align*}
-  \bm{\Sigma}_d^{(h+1)} & = \frac{2 \zeta_d^{(h)}}{n+p+1} \[\X \tr \X + a_d^{(h)} \cdot \diag(b_j^{(h)})\]^{-1}, \\
-  \bm{\mu}_d^{(h+1)} & = \[\X \tr \X + a_d^{(h)} \cdot \diag(b_j^{(h)})\]^{-1} \X \tr \y_d, \\
-  \delta_d^{(h+1)} & = \frac{n+p+1}{2 \zeta_d^{(h)}} \left\{\trace \[\diag(b_j^{(h)}) \bm{\Sigma}_d^{(h+1)} \] + (\bm{\mu}_d^{(h+1)}) \tr \diag(b_j^{(h)}) \bm{\mu}_d^{(h+1)}\right\} + \lambda, \\ 
-  \eta_j^{(h+1)} & = \sum_{d=1}^D \frac{a_d^{(h)}(n+p+1)}{2 \zeta_d^{(h)}} \[ (\bmu_{jd}^{(h + 1)})^2 + (\bSigma_d^{(h + 1)})_{jj} \] + \nu, \\ 
-  \zeta_d^{(h+1)} & = \frac{1}{2} \Big( \mathbf{y}_d \tr \mathbf{y}_d -2 \mathbf{y}_d \tr \X \bm{\mu}_d^{(h+1)} + a_d^{(h+1)} \trace \left\{ \[\X \tr \X + \diag(b_j^{(h + 1)})\] \bm{\Sigma}_d^{(h+1)}\right\} \\
-  & \,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\, + a_d^{(h+1)} (\bm{\mu}_d^{(h+1)}) \tr \[\X \tr \X + \diag(b_j^{(h + 1)})\]\bm{\mu}_d^{(h+1)}\Big),
-  \end{align*}
-  with
-  \begin{align*}
-  a_d^{(h)} & = \sqrt{\frac{\lambda (\mathbf{c}_d \tr \bm{\alpha})^2}{\delta_d^{(h)}}} \frac{K_{\frac{p-1}{2}} \( \sqrt{\lambda (\mathbf{c}_d \tr \bm{\alpha})^2 \delta_d^{(h)}} \)}{K_{\frac{p+1}{2}} \( \sqrt{\lambda (\mathbf{c}_d \tr \bm{\alpha})^2 \delta_d^{(h)}} \)} + \frac{p+1}{\delta_d^{(h)}} \text{ and} \\
-  b_j^{(h)} & = \sqrt{\frac{\nu (\mathbf{z}_j \tr \bm{\theta})^2}{\eta_j^{(h)}}} \frac{K_{\frac{D-1}{2}} \( \sqrt{\nu (\mathbf{z}_j \tr \bm{\theta})^2 \eta_j^{(h)}} \)}{K_{\frac{D+1}{2}} \( \sqrt{\nu (\mathbf{z}_j \tr \bm{\theta})^2 \eta_j^{(h)}} \)} + \frac{D+1}{\eta_j^{(h)}}.
-  \end{align*}
-  Hyper-parameter updates:
-  \begin{align*}
-  \bm{\alpha}^{(l+1)} & = \left[ \mathbf{C} \tr \diag ( e_d^{(l)} ) \mathbf{C} \right]^{-1} \mathbf{C} \tr \mathbf{1}, \\
-  \lambda^{(l+1)} & = \[ \sum_{d=1}^D a_d^{(l)} - (\bm{\alpha}^{(l+1)}) \tr \mathbf{C} \tr \mathbf{1} \]^{-1} D, \\
-  \bm{\theta}^{(l+1)} & = \left[ \mathbf{Z} \tr \diag ( f_j^{(l)} ) \mathbf{Z} \right]^{-1} \mathbf{Z} \tr \mathbf{1}, \\
-  \nu^{(l+1)} & = \[ \sum_{j=1}^p b_j^{(l)} - (\bm{\theta}^{(l+1)}) \tr \mathbf{Z} \tr \mathbf{1} \]^{-1} p,
-  \end{align*}
-  where 
-  \begin{align*}
-  e_d^{(l)} & = \sqrt{\frac{\delta_d^{(l)}}{\lambda^{(l)}(\mathbf{c}_d \tr \bm{\alpha}^{(l)})^2}} \frac{K_{\frac{p - 1}{2}} \( \sqrt{\lambda^{(l)} (\mathbf{c}_d \tr \bm{\alpha}^{(l)})^2 \delta_d^{(l)}} \)}{K_{\frac{p + 1}{2}} \( \sqrt{\lambda^{(l)} (\mathbf{c}_d \tr \bm{\alpha}^{(l)})^2 \delta_d^{(l)}} \)} \text{ and} \\
-  f_j^{(l)} & = \sqrt{\frac{\eta_j^{(l)}}{\nu^{(l)}(\mathbf{z}_j \tr \bm{\theta}^{(l)})^2}} \frac{K_{\frac{D - 1}{2}} \( \sqrt{\nu^{(l)} (\mathbf{z}_j \tr \bm{\theta}^{(l)})^2 \eta_j^{(l)}} \)}{K_{\frac{D + 1}{2}} \( \sqrt{\nu^{(l)} (\mathbf{z}_j \tr \bm{\theta}^{(l)})^2 \eta_j^{(l)}} \)}.
-  \end{align*}
-
   \bibliographystyle{author_short3}  

### 2019-08-19 code/figures.R

@@ -212,65 +212,128 @@ library(sp)
 res <- read.table("results/simulations_igaussian_res4.2.csv")
 set <- read.table("results/simulations_igaussian_set4.csv")
 
-plot.data1 <- list(cram=res[, substr(colnames(res), 4, 4) %in% 
-                              c("1", "2", "3") &
-                              substr(colnames(res), nchar(colnames(res)) - 7 
-                                     + 1, nchar(colnames(res)))=="enig.vb" &
-                              substr(colnames(res), 6, 9)=="cram"],
-                   corbest=res[, substr(colnames(res), 4, 4) %in% 
-                                 c("1", "2", "3") &
-                                 substr(colnames(res), nchar(colnames(res)) - 7 
-                                        + 1, nchar(colnames(res)))=="enig.vb" &
-                                 substr(colnames(res), 6, 12)=="corbest"],
-                   msebest=res[, substr(colnames(res), 4, 4) %in% 
-                                 c("1", "2", "3") &
-                                 substr(colnames(res), nchar(colnames(res)) - 7 
-                                        + 1, nchar(colnames(res)))=="enig.vb" &
-                                 substr(colnames(res), 6, 12)=="msebest"])
-plot.data2 <- list(cram=res[, substr(colnames(res), 4, 4) %in% 
-                              c("4", "5", "6") &
-                              substr(colnames(res), nchar(colnames(res)) - 7 
-                                     + 1, nchar(colnames(res)))=="enig.vb" &
-                              substr(colnames(res), 6, 9)=="cram"],
-                   corbest=res[, substr(colnames(res), 4, 4) %in% 
-                                 c("4", "5", "6") &
-                                 substr(colnames(res), nchar(colnames(res)) - 7 
-                                        + 1, nchar(colnames(res)))=="enig.vb" &
-                                 substr(colnames(res), 6, 12)=="corbest"],
[…]

### 2019-08-19 code/simulations.R

@@ -422,17 +422,17 @@ for(s in 1:nrow(set)) {
                         data=list(n=n, p=p, y=y, x=x, 
                                   ctalphainv=rep(ctalphainv, p),
                                   lambda=lambda))
-    fit.enig.vb <- enig(list(x, x), cbind(y, y), NULL, mult.lambda=FALSE, 
-                        intercept.eb=TRUE, fixed.eb=c("both"), 
-                        full.post=TRUE, 
-                        init=list(aold=rep(1, 2), bold=list(p, p), 
-                                  lambda=lambda, alpha=1, 
-                                  Calpha=rep(1/ctalphainv, 2)),
-                        control=list(conv.post=TRUE, trace=TRUE,
-                                     epsilon.eb=1e-3, epsilon.vb=1e-3, 
-                                     epsilon.opt=sqrt(.Machine$double.eps),
-                                     maxit.eb=10, maxit.vb=2, maxit.opt=100,
-                                     maxit.post=100))
+    fit.enig.vb <- semnig(list(x, x), cbind(y, y), C=NULL, unpenalized=NULL, 
+                          standardize=FALSE, intercept=FALSE, intercept.eb=TRUE, 
+                          mult.lambda=FALSE, fixed.eb=c("both"), full.post=TRUE, 
+                          init=list(aold=rep(1, 2), bold=list(p, p), 
+                                    lambda=lambda, alpha=1, 
+                                    Calpha=rep(1/ctalphainv, 2)),
+                          control=list(conv.post=TRUE, trace=TRUE,
+                                       epsilon.eb=1e-3, epsilon.vb=1e-3, 
+                                       epsilon.opt=sqrt(.Machine$double.eps),
+                                       maxit.eb=10, maxit.vb=2, maxit.opt=100,
+                                       maxit.post=100))
               
     fit.enig.map <- optimizing(stan.enig, 
                                data=list(n=n, p=p, y=y, x=x, 
@@ -502,6 +502,12 @@ for(s in 1:nrow(set)) {
     res2[k, substr(colnames(res2), 1, 4)==paste0("set", s)] <- 
       c(cram2, corbest2, msebest2)
   
+    if(s==1 & k==1) {
+      save(fit.enig, fit.studentst, fit.enig.advi, fit.enig.vb, fit.enig.map, 
+           fit.glmnet, fit.cv.glmnet, 
+           file="results/simulations_igaussian_fit4.Rdata")
+    }
+    
[…]

### 2019-08-19 rpackage/R/aux_functions.R

@@ -9,7 +9,7 @@ ratio_besselK <- function(x, nu) {
   return(val)
 }
 
-# calculate full covariance with unpenalized covariates
+# calculate full covariance with unpenalized covariates (not tested)
 .Sigma.unp <- function(aold, bold, xu, xr, u, r) {
   
   hinv <- 1/bold
@@ -34,7 +34,7 @@ ratio_besselK <- function(x, nu) {
   return(Sigma)
 }
 
-# calculate full covariance
+# calculate full covariance (not tested)
 .Sigma <- function(aold, bold, x) {
   hinv <- 1/bold
   HinvXt <- t(x)*hinv
@@ -55,14 +55,14 @@ ratio_besselK <- function(x, nu) {
   diag(Om) <- diag(Om) + 1
   Om <- solve(Om)
   HinvXrtOm <- HinvXrt %*% Om
-  M11 <- solve(t(xu) %*% Om %*% xu)
+  M11 <- solve(t(xu) %*% Om %*% xu) ###
   M12 <- - M11 %*% t(xu) %*% t(HinvXrt) + M11 %*% t(xu) %*% t(XrHinvXrt) %*% 
     t(HinvXrtOm)
   
   # auxiliaries
   dSigma <- c(diag(M11), hinv - rowSums(HinvXrt*HinvXrtOm) - 
                 rowSums((HinvXrt %*% xu)*t(M12)) + 
-                rowSums(HinvXrtOm*(t(M12) %*% t(xu) %*% XrHinvXrt)))
+                rowSums(HinvXrtOm*(t(M12) %*% t(xu) %*% XrHinvXrt)))/aold
   mu <- rbind(M11 %*% t(xu) + M12 %*% t(xr),
               HinvXrt - HinvXrtOm %*% XrHinvXrt -
                 HinvXrt %*% xu %*% M12 %*% t(xr) + HinvXrtOm %*% XrHinvXrt %*%
@@ -74,11 +74,9 @@ ratio_besselK <- function(x, nu) {
   ytXmu <- cbind(xu, xr) %*% mu
   mutXtXmu <- as.numeric(t(ytXmu) %*% ytXmu)
   ytXmu <- sum(as.numeric(ytXmu)*y)
-  trHSigma <- sum(dSigma[-c(1:u)]*bold)
-  mutHmu <- sum(mu[-c(1:u)]^2*bold)
   
[…]

### 2019-08-19 rpackage/tests/testthat/test-aux_functions.R

@@ -1,5 +1,4 @@
 context("miscellaneous functions")
-
 test_that("ratio of modified Bessel functions", {
   
   # setting parameters
@@ -19,74 +18,206 @@ test_that("ratio of modified Bessel functions", {
   
 })
 
-test_that("low dim. aux. variables in general model without intercept", {
-
-  n <- 20
-  p <- 10
-  y <- rnorm(n)
-  x <- matrix(rnorm(n*p), ncol=p, nrow=n)
-  aold <- rchisq(1, 1)
-  bold <- rchisq(1, 1)
-  cold <- rchisq(p, 1)
-  ytx <- as.numeric(t(y) %*% x)
+context("aux variables without unpenalized covariates")
+test_that("low dim. auxiliary variables calculation", {
   
-  # fast calculation
-  aux <- .aux.var.gen(aold, bold, cold, x, ytx)
-
-  # actual calculation
-  Sigma <- solve(t(x) %*% x + bold*diag(cold))/aold
-  mu <- solve(t(x) %*% x + bold*diag(cold)) %*% t(x) %*% y
-  mutdiagcmu <- as.numeric(t(mu) %*% diag(cold) %*% mu)
-  trdiagcSigma <- sum(diag(diag(cold) %*% Sigma))
-  mutmu <- as.numeric(t(mu) %*% mu)
-  trSigma <- sum(diag(Sigma))
-  ytXmu <- as.numeric(t(y) %*% x %*% mu)
-  trXtXSigma <- sum(diag(t(x) %*% x %*% Sigma))
-  mutXtXmu <- as.numeric(t(mu) %*% t(x) %*% x %*% mu)
-
+  # setting parameters
+  n <- 100
+  p <- 10
+  x <- matrix(rnorm(n*p), nrow=n, ncol=p)
+  y <- rnorm(n)
+  ytx <- as.numeric(t(y) %*% x)
+  aold <- rchisq(1, 1) + 1
+  bold <- rchisq(p, 1) + 1
+  
+  # implemented function
+  test.aux <- .aux.var(aold, bold, y, x, ytx)
+  
+  # actual calculation (see supplement)
+  H <- diag(bold)
+  Sigma <- solve(t(x) %*% x + H)/aold
+  mu <- as.numeric(solve(t(x) %*% x + H) %*% t(x) %*% y)
[…]

### 2019-08-19 rpackage/tests/testthat/test-vb_functions.R

@@ -1,0 +1,155 @@
+context("vb update without unpenalized covariates")
+test_that("low dim. vb update", {
+  
+  # setting parameters
+  n <- 100
+  p <- 10
+  x <- matrix(rnorm(p*n), nrow=n, ncol=p)
+  y <- rnorm(n)
+  aold <- rchisq(1, 1) + 1
+  bold <- rchisq(p, 1) + 1
+  Calpha <- rchisq(p, 1) + 1
+  lambda <- rchisq(1, 1) + 1
+  ytx <- as.numeric(t(y) %*% x)
+  yty <- as.numeric(t(y) %*% y)
+  
+  # implemented function
+  test.vb <- .single.vb.update(aold, bold, Calpha, lambda, y, x, ytx, yty, n, p)
+  
+  # actual calculation (see supplement)
+  Sigma <- solve(t(x) %*% x + diag(bold))/aold
+  mu <- as.numeric(solve(t(x) %*% x + diag(bold)) %*% t(x) %*% y)
+  delta <- aold*(mu^2 + diag(Sigma)) + lambda
+  b <- sqrt(lambda*Calpha^2/delta)*
+    ratio_besselK(sqrt(lambda*delta)*Calpha, 1) + 2/delta
+  zeta <- 0.5*as.numeric(
+    t(y) %*% y  - 2*t(y) %*% x %*% mu + sum(diag(t(x) %*% x %*% Sigma)) +
+      t(mu) %*% t(x) %*% x %*% mu + sum(diag(diag(b) %*% Sigma)) + 
+      t(mu) %*% diag(b) %*% mu)
+  a <- (n + p + 1)/(2*zeta)
+  e <- (b - 2/delta)*delta/(lambda*Calpha^2)
+  vb <- list(delta=delta, zeta=zeta, a=a, b=b, e=e)
+  
+  # check
+  expect_equal(vb, test.vb)
+  
+})
+
+test_that("high dim. vb update", {
+  
+  # setting parameters
+  n <- 10
+  p <- 100
+  x <- matrix(rnorm(p*n), nrow=n, ncol=p)
+  y <- rnorm(n)
+  aold <- rchisq(1, 1) + 1
+  bold <- rchisq(p, 1) + 1
+  Calpha <- rchisq(p, 1) + 1
+  lambda <- rchisq(1, 1) + 1
+  ytx <- as.numeric(t(y) %*% x)
+  yty <- as.numeric(t(y) %*% y)
+  
[…]

### 2019-11-19 docs/biometrical_main.Rnw
redoing simulations

% set options
<<settings, include=FALSE, echo=FALSE>>=
opts_knit$set(root.dir="..", base.dir="../figs")
opts_chunk$set(fig.align='center', fig.path="../figs/", 
               echo=FALSE, cache=FALSE, message=FALSE, fig.pos='!ht')
knit_hooks$set(document=function(x) {
  sub('\\usepackage[]{color}', '\\usepackage{xcolor}', x, fixed=TRUE)})
@

% read figure code
<<figures, include=FALSE>>=
read_chunk("code/figures.R")
@

\documentclass[bimj,fleqn]{w-art}
\usepackage{bm,amsmath,amssymb,amsthm,amsfonts,graphics,epsfig,
rotating,appendix,titlesec,multicol,hyperref,verbatim,
bbm,algorithm,algpseudocode,pgfplotstable,threeparttable,booktabs,mathtools,
dsfont,parskip,xr,tikz,times,w-thm}
\usepackage[]{graphicx}
\usepackage[authoryear]{natbib}
\externaldocument[sm-]{supplement1}
\usetikzlibrary{bayesnet}

\setlength{\bibsep}{2pt}
\setlength{\bibhang}{2em}
\newcommand{\J}{J\"{o}reskog}
\newcommand{\So}{S\"{o}rbom}
\newcommand{\bcx}{{\bf X}}
\newcommand{\bcy}{{\bf Y}}
\newcommand{\bcz}{{\bf Z}}
\newcommand{\bcu}{{\bf U}}
\newcommand{\bcv}{{\bf V}}
\newcommand{\bcw}{{\bf W}}
\newcommand{\bci}{{\bf I}}
\newcommand{\bch}{{\bf H}}
\newcommand{\bcb}{{\bf B}}
\newcommand{\bcr}{{\bf R}}
\newcommand{\bcm}{{\bf M}}
\newcommand{\bcf}{{\bf F}}
\newcommand{\bcg}{{\bf G}}
\newcommand{\bcs}{{\bf S}}
\newcommand{\bca}{{\bf A}}
\newcommand{\bcd}{{\bf D}}
\newcommand{\bcc}{{\bf C}}
\newcommand{\bce}{{\bf E}}
\newcommand{\ba}{{\bf a}}
\newcommand{\bb}{{\bf b}}
\newcommand{\bc}{{\bf c}}
\newcommand{\bd}{{\bf d}}
\newcommand{\bx}{{\bf x}}
\newcommand{\by}{{\bf y}}
\newcommand{\bz}{{\bf z}}
\newcommand{\bu}{{\bf u}}
\newcommand{\bv}{{\bf v}}
\newcommand{\bh}{{\bf h}}
\newcommand{\bl}{{\bf l}}
\newcommand{\be}{{\bf e}}
\newcommand{\br}{{\bf r}}
\newcommand{\bw}{{\bf w}}
\newcommand{\de}{\stackrel{D}{=}}
\newcommand{\bt}{\bigtriangleup}
\newcommand{\bfequiv}{\mbox{\boldmath $\equiv$}}
\newcommand{\bmu}{\mbox{\boldmath $\mu$}}
\newcommand{\bnu}{\mbox{\boldmath $\nu$}}
\newcommand{\bxi}{\mbox{\boldmath $\xi$}}
\newcommand{\btau}{\mbox{\boldmath $\tau$}}
\newcommand{\bgamma}{\mbox{\boldmath $\Gamma$}}
\newcommand{\bphi}{\mbox{\boldmath $\Phi$}}
\newcommand{\bfphi}{\mbox{\boldmath $\varphi$}}
\newcommand{\bfeta}{\mbox{\boldmath $\eta$}}
\newcommand{\bpi}{\mbox{\boldmath $\Pi$}}
\newcommand{\bequiv}{\mbox{\boldmath $\equiv$}}
\newcommand{\bvarepsilon}{\mbox{\boldmath $\varepsilon$}}
\newcommand{\btriangle}{\mbox{\boldmath $\triangle$}}
\newcommand{\bdelta}{\mbox{\boldmath $\Delta$}}
\newcommand{\beps}{\mbox{\boldmath $\epsilon$}}
\newcommand{\btheta}{\mbox{\boldmath $\theta$}}
\newcommand{\balpha}{\mbox{\boldmath $\alpha$}}
\newcommand{\bsphi}{\mbox{\boldmath $\varphi$}}
\newcommand{\bsig}{\mbox{\boldmath $\sigma$}}
\newcommand{\bfpsi}{\mbox{\boldmath $\psi$}}
\newcommand{\bfdelta}{\mbox{\boldmath $\delta$}}
\newcommand{\bsigma}{{\bf \Sigma}}
\newcommand{\bzero}{{\bf 0}}
\newcommand{\bpsi}{\mbox{\boldmath $\Psi$}}
\newcommand{\bep}{\mbox{\boldmath $\epsilon$}}
\newcommand{\bomega}{\mbox{\boldmath $\Omega$}}
\newcommand{\bfomega}{\mbox{\boldmath $\omega$}}
\newcommand{\blambda}{\mbox{\boldmath $\Lambda$}}
\newcommand{\bflambda}{\mbox{\boldmath $\lambda$}}
\newcommand{\bfsigma}{\mbox{\boldmath $\sigma$}}
\newcommand{\bfpi}{{\mbox{\boldmath $\pi$}}}
\newcommand{\bupsilon}{\mbox{\boldmath $\upsilon$}}
\newcommand{\obs}{{\rm obs}}
\newcommand{\mis}{{\rm mis}}
\theoremstyle{plain}
\newtheorem{criterion}{Criterion}
\theoremstyle{definition}
\newtheorem{condition}[theorem]{Condition}

\pgfplotsset{compat=1.16}
\chardef\bslash=`\\ % p. 424, TeXbook
\newcommand{\ntt}{\normalfont\ttfamily}
\newcommand{\cn}[1]{{\protect\ntt\bslash#1}}
\newcommand{\pkg}[1]{{\protect\ntt#1}}
\let\fn\pkg
\let\env\pkg
\let\opt\pkg
\hfuzz1pc % Don't bother to report overfull boxes if overage is < 1pc
\newcommand{\envert}[1]{\left\lvert#1\right\rvert}
\let\abs=\envert

% vectors and matrices
\newcommand{\bbeta}{\bm{\beta}}

\newcommand{\bepsilon}{\bm{\epsilon}}
\newcommand{\bOmega}{\bm{\Omega}}
\newcommand{\bSigma}{\bm{\Sigma}}
\newcommand{\bkappa}{\bm{\kappa}}
\newcommand{\bLambda}{\bm{\Lambda}}
\newcommand{\bPsi}{\bm{\Psi}}
\newcommand{\Y}{\mathbf{Y}}
\newcommand{\y}{\mathbf{y}}
\newcommand{\X}{\mathbf{X}}
\newcommand{\x}{\mathbf{x}}
\newcommand{\I}{\mathbf{I}}
\newcommand{\T}{\mathbf{T}}
\newcommand{\Z}{\mathbf{Z}}
\newcommand{\W}{\mathbf{W}}
\renewcommand{\O}{\mathbf{O}}
\newcommand{\B}{\mathbf{B}}
\renewcommand{\H}{\mathbf{H}}
\newcommand{\U}{\mathbf{U}}
\newcommand{\Em}{\mathbf{E}}
\newcommand{\D}{\mathbf{D}}
\newcommand{\Vm}{\mathbf{V}}
\newcommand{\rv}{\mathbf{r}}
\newcommand{\A}{\mathbf{A}}
\newcommand{\Q}{\mathbf{Q}}
\newcommand{\Sm}{\mathbf{S}}
\newcommand{\0}{\bm{0}}
\newcommand{\tx}{\tilde{\mathbf{x}}}
\newcommand{\hy}{\hat{y}}
\newcommand{\tm}{\tilde{m}}
\newcommand{\tkappa}{\tilde{\kappa}}
\newcommand{\m}{\mathbf{m}}
\newcommand{\C}{\mathbf{C}}
\renewcommand{\v}{\mathbf{v}}
\newcommand{\g}{\mathbf{g}}
\newcommand{\s}{\mathbf{s}}
\renewcommand{\c}{\mathbf{c}}
\newcommand{\ones}{\mathbf{1}}
\newcommand{\R}{\mathbf{R}}
\newcommand{\rvec}{\mathbf{r}}
\newcommand{\Lm}{\mathbf{L}}

% functions and operators
\renewcommand{\L}{\mathcal{L}}
\newcommand{\G}{\mathcal{G}}
\renewcommand{\P}{\mathcal{P}}
\newcommand{\argmax}{\text{argmax} \,}
\newcommand{\expit}{\text{expit}}
\newcommand{\erfc}{\text{erfc}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\V}{\mathbb{V}}
\newcommand{\Cov}{\mathbb{C}\text{ov}}
\newcommand{\tr}{^{\text{T}}}
\newcommand{\diag}{\text{diag}}
\newcommand{\KL}{D_{\text{KL}}}
\newcommand{\trace}{\text{tr}}
\newcommand{\norm}[1]{\left\lVert #1 \right\rVert}

% distributions
\newcommand{\N}{\text{N}}
\newcommand{\TG}{\text{TG}}
\newcommand{\Bi}{\text{Binom}}
\newcommand{\PG}{\text{PG}}
\newcommand{\Mu}{\text{Multi}}
\newcommand{\GIG}{\text{GIG}}
\newcommand{\IGauss}{\text{IGauss}}
\newcommand{\Un}{\text{U}}

% syntax and readability
\renewcommand{\(}{\left(}
\renewcommand{\)}{\right)}
\renewcommand{\[}{\left[}
\renewcommand{\]}{\right]}

\begin{document}

  \DOIsuffix{bimj.200100000}\Volume{XX}\Issue{XX}\Year{XXX}
  \pagespan{1}{}
  \keywords{Drug sensitivity; Empirical Bayes; 
  Genomics of Drug Sensitivity in Cancer (GDSC); Variational Bayes\\[1pc]
  \noindent\hspace*{-4.2pc} Supporting Information for this article is available
  from the author or on the WWW under\break \hspace*{-4pc} 
  \underline{http://dx.doi.org/10.1022/bimj.XXXXXXX} (please delete if not
  applicable)}  %%% semicolon and fullpoint added here for keyword style

  \title[Drug sensitivity prediction with NIG prior]
  {Drug sensitivity prediction with the normal inverse Gaussian prior 
  informed by external data}

  %% Institutes
  \address[\inst{1}]{Department of Epidemiology \& Biostatistics, 
  Amsterdam UMC, VU University, PO Box 7057, 1007 MB Amsterdam, The Netherlands}
  \address[\inst{2}]{Mathematical Institute, Leiden University, Leiden, 
  The Netherlands}
  \address[\inst{3}]{MRC Biostatistics Unit, 
  Cambridge Institute of Public Health, Cambridge, United Kingdom}

  %% Authors
  \author[M\"unch]{Magnus M. M\"unch
  \footnote{Corresponding author: {\sf{e-mail: m.munch@amsterdamumc.nl}}, 
  Phone: +31 71 527 4844}\inst{,1,2}} 
  \author[van de Wiel]{Mark A. van de Wiel\inst{1,3}}
  \author[Richardson]{Sylvia Richardson\inst{3}}
  \author[Leday]{Gwena{\"e}l G. R. Leday\inst{3}}

  %%%%    \dedicatory{This is a dedicatory.}
  \Receiveddate{zzz} \Reviseddate{zzz} \Accepteddate{zzz}

	\begin{abstract}
		{In precision medicine, a common problem is drug sensitivity prediction from
		cancer tissue cell lines. These types of problems entail modelling 
		multivariate drug responses on high dimensional molecular feature sets in 
		typically $>1000$ cell lines. The dimensions of the problem require 
		specialised models and estimation methods. In addition, external information
		on both the drugs and the features is often available. We propose to 
		model the drug responses through a linear regression with the normal
		inverse Gaussian prior. We let the prior depend on the external information,
		and estimate the model and external information dependence in an 
		empirical-variational Bayes framework. We demonstrate the usefullness of
		this model in both a simulated setting and in the publicly available
		Genomics of Drug Sensitivity in Cancer data.}
	\end{abstract}

	\maketitle

	\section{Introduction}
	Recently, promising results in precision medicine 
	have sparked an interest in cancer drug sensitivity prediction models
	\cite[]{iorio_landscape_2016}. Typically, these models 
	predict the drug sensitivity for new patients from a set of molecular 
	features.
	Development of such models is often done in well-characterised human cancer
	tissue cell lines. The current paper presents a novel drug sensitivity
	prediction model and an application to a real drug
	sensitivity data set.

	Development of such models from cell lines has proven to be difficult
	(see e.g., the DREAM 7 challenge in \cite{costello_community_2014}). 
	Difficulties arise, among others, from the dimensions of the problem. 
	Typically, the data contains
	hundreds of drugs, thousands of cell lines, and thousands of molecular 
	features. An example of a large database of drug responses and molecular
	features is the Genomics of Drug Sensitivity in Cancer (GDSC) data 
	\cite[]{yang_genomics_2013}, which we will further 
	investigate in Section \ref{sec:gdsc}. Other examples of such databases 
	include
	the Cancer Cell Line Encyclopedia (CCLE) \cite[]{li_landscape_2019} and 
	the US National Cancer Institute 60 human tumour cell line anticancer drug
	screen (NCI60) \cite[]{shoemaker_NCI60_2006}. The dimensions of these data
	prohibit the estimation of standard regression models and
	typically require some form of regularisation.

	The GDSC database contains additional information on both the drugs and 
	molecular features, such as the target 
	pathways and developmental stages of the drugs. Additional online
	repositories might provide extra information such as the molecular weight of
	the compounds or the publication signatures of the molecular features. In some
	cases, prior knowledge on the drug efficacies might be available,
	from previous experiments. We propose to include these possibly beneficial 
	information sources in the estimation of the sensitivity prediction models in 
	a data-driven manner. More specifically, we estimate a normal 
	inverse Gaussian model, where the amount of regularisation is estimated by an 
	adaptive empirical Bayes procedure, guided by the external information.

	We are not the first to work on drug sensitivity prediction models. Reviews on 
	the topic are \cite{azuaje_computational_2017} and \cite{ali_machine_2019}.
	\cite{zhao_structured_2019} and \cite{mai_composite_2019} consider a
	structured penalized multivariate regression approach. 
	\cite[]{aben_tandem:_2016} introduces a two-stage penalized regression 
	model that includes two different types of molecular features.
	\cite{ammad-ud-din_drug_2016} and \cite{costello_community_2014} tackle the
	problem through a multiple kernel learning approach.

  Our solution allows for the adaptive incorporation of the external 
  information on drugs and features. This is done by
  pooling, both across drugs and features. Estimation of the model
  is through computational feasible variational bayes approximations, while 
  empirical Bayes estimation of tuning parameters pools information across drugs
  and features in a data-driven manner.

	The rest of the paper is structured as follows. In Section 
	\ref{sec:model} we introduce our model, the estimation of which is detailed
	in Section \ref{sec:estimation}. In Section \ref{sec:gdsc} we analyse the
	GDSC data, and we end with a discussion in Section \ref{sec:discussion}.

	\section{Model}\label{sec:model}
	\subsection{Simultaneous equations model}\label{sec:SEM}
	Let $y_{id}$ be the continuous sensitivity measures 
	for cell lines $i=1,\dots, n$, and drug $d=1, \dots, D$. 
	We predict sensitivity from molecular features 
	$x_{ij}$, $j=1,\dots, p$, collected in 
	$\x_i = \begin{bmatrix} x_{i1} & \dots & x_{ip} \end{bmatrix} \tr$. 
	We assume that both covariates and responses have been centred per drug
	and regress the drug sensitivities on the molecular features:
	\begin{equation}\label{eq:likelihood}
	  y_{id} = \x_i \tr \bbeta_d + \epsilon_{id}, 
	  \text{ with } \epsilon_{id} \sim \mathcal{N}(0, \sigma_d^2),
	\end{equation}
	where the $p$-dimensional 
	$\bbeta_d = \begin{bmatrix} \beta_{1d} & \cdots & \beta_{pd} \end{bmatrix} 
	\tr$ are the drug-specific omics feature 
	effects. Note that (\ref{eq:likelihood}) gives rise to a system of 
	D linear regression equations.

	The cell lines used in drug response models are often taken from different
	tissues. In addition, other clinical covariates might be available. 
	To obtain unbiased feature effects, one may wish to account for these. 
	We do so by introducing unpenalized covariates, the $\beta_{jd}$ coefficients 
	of which are endowed with a flat prior. For the sake of clarity, in the
	following, such unpenalized covariates are omitted. However, the available
	software allows for their inclusion.

	\subsection{Bayesian prior model}
	We carry out inference by endowing the parameters with the following priors:
	\begin{subequations}\label{eq:prior}
		\begin{align}
		  \beta_{jd} | \gamma^2_{jd}, \tau_d^2, \sigma^2_d & \sim \mathcal{N}_{p} 
		  (0, \gamma_{jd}^2 \tau_d^2 \sigma_d^2), 
		  \label{eq:betaprior}\\
		  \gamma_{jd}^{2} & \sim \mathcal{IG}(\phi_{jd}, \lambda_{\text{feat}}), \\
		  \tau_d^{2} & \sim \mathcal{IG}(\chi_d, \lambda_{\text{drug}}), 
		  \label{eq:tauprior}\\
		  \sigma_d^{2} & \sim 1/\sigma_d^{3},
		\end{align}
	\end{subequations}
	where $\mathcal{IG}(\phi, \lambda)$ denotes an inverse Gaussian distribution 
	with mean $\phi$ and shape $\lambda$. Prior distributions of the form
	(\ref{eq:prior}) are often referred to as global-local shrinkage rules 
	\cite[]{bernardo_shrink_2011}, due to the multiplicative separation of the 
	prior 
	variance into a local components $\gamma_{jd}^2$ and a global component
	$\tau_d^2$. The local components are supposed to capture local variation in
	the model parameters $\beta_{jd}$, while the global component is supposed to
	capture the general trend in $\beta_{jd}$. Addionally, for appropriate local
	shrinkage in global-local 
	shrinkage models it is important  to account for 
  different noise levels $\sigma_d^2$ by scaling the $\beta_{jd}$ variances 
  accordingly.

	The normal inverse Gaussian (NIG) prior model was introduced in 
  \cite{barndorff-nielsen_hyperbolic_1978} and since 
  \cite{barndorff-nielsen_normal_1997} it is routinely applied in mathematical 
  finance (see, e.g., \cite{kalemanova_normal_2007}). Here we extend it with an
  additional global variance component $\tau_d^2$. Supplementary Material
  (SM) Section \ref{sm-sec:prior} contains more details on the NIG prior.
  To illustrate the effect of the NIG prior on the posterior mean, we consider 
  the prior reparametrised as in 
  \cite{carvalho_handling_2009}, i.e., in terms of shrinkage weights 
  $\kappa_{jd}=1/(1 + \gamma_{jd}^2) \in (0, 1)$.
  Under the (simplified) normal means model, i.e., 
  $\X= \begin{bmatrix} \x_1 & \cdots \x_n \end{bmatrix} \tr =\I_p$, with fixed 
  $\tau_d^2=\sigma_d^2=1$, the resulting
  posterior mean for the $\beta_{jd}$ is
  $\E(\beta_{jd} | y_{jd}, \kappa_{jd}) = (1 - \kappa_{jd}) y_{jd}$. 
  Thus, $\kappa_{jd}=0$ implies no shrinkage of $\beta_{jd}$ and $\kappa_{jd}=1$
  implies full shrinkage towards zero. Figure 
  \ref{fig:dens_kappa} depicts the prior on $\kappa_{jd}$ implied by 
  several choices of $\beta_{jd}$ prior.
<<dens_kappa, fig.cap="Implied prior densities $\\pi(\\kappa_{jd})$ for the (a) NIG, (b) Student's $t$, and (c) lasso priors. Different line types correspond to different hyperparameter settings. The hyperparameter settings were chosen to show some possible, distinct shapes that each of the priors can take.", out.width="90%", fig.asp=1/3>>=
@

  Figure \ref{fig:dens_kappa} shows that, depending on the choice
  of hyperparameters, the NIG prior can behave similarly to the Student's $t$ 
  prior (decreasing form zero, with substantial mass close to 0), 
  but also rather differently. Our argumentation to model the 
  $\gamma^2_{jd}$ by an inverse Gaussian 
  distribution, as has been suggested in \cite{fabrizi_specification_2016} and 
  \cite{caron_sparse_2008}, is three-fold: (i) the NIG model is more flexible 
  than the lasso prior (as seen from Figure 
  \ref{fig:dens_kappa}), (ii) the NIG prior allows to model
  the means of the $\gamma_{jd}^2$ ($\phi_{jd}$) and $\tau_{jd}^2$ 
  ($\chi_{d}$) as a function of external data 
  more conveniently than the Student's $t$ prior, as explained in Section 
  \ref{sec:empiricalbayes}, and (iii) like the horseshoe, the NIG shrinkage
  weights prior can put mass both near 0 and 1, a desirable property of 
  shrinkage priors.

  A few remarks on the choice of error variance prior are justified here: 
  many authors endow error variance components with vague gamma priors. 
  \cite{gelman_prior_2006}, 
  among others, advises against this practice. The degree of `vagueness' has a 
  large influence on the posterior, while degree of `vagueness' is a difficult 
  parameter to set. This influence is especially pronounced if the likelihood is
  relatively flat, as may be reasonably expected in the large $p$, 
  small $n$ setting. We therefore model the error variance with Jeffrey's 
  objective prior 
  \cite[]{jeffreys_invariant_1946} that does not depend on any 
  subjective specification of hyperparameters. In the derivation of our 
  Jeffrey's prior
  for the error variance, we jointly consider an unknown mean and variance 
  \cite[]{kass_selection_1996}. This
  joint consideration results in the somewhat unorthodox $1/\sigma^3$ Jeffrey's
  prior.

	\subsection{External information}\label{sec:external}
	In drug sensitivity prediction models external information on both the drugs
	and features is often available. Here, we assume this information to be 
	available as
	external feature `covariates' $\mathbf{c}_{jdg}$, for $g=1, \dots, G$, and 
	drug `covariates' $\mathbf{z}_{dh}$, for $h=1,\dots, H$. An example of a 
	(binary) features covariate is target pathway presence, with $c_{jdg}=0$ if 
	gene $j$ is present in the 
	target pathway of drug $d$ and $c_{jdg}=1$ if it is not. An example of a 
	(ternary) drug
	covariate is developmental phase, with levels experimental phase, 
	clinical development, and approved by a governing agency.

	The external covariates come in through our mean models for 
	the $\gamma_{jd}^2$ and $\tau_d^2$ hyperpriors:
	$\phi_{jd} = (\mathbf{c}_{jd} \tr \bm{\alpha}_{\text{feat}})^{-1}$ and
	$\chi_d = (\mathbf{z}_{d} \tr \bm{\alpha}_{\text{drug}})^{-1}$, with 
	$\mathbf{c}_{jd} = \begin{bmatrix} c_{jd1} & \cdots & c_{jdG} \end{bmatrix}$
	and $\mathbf{z}_{d} = \begin{bmatrix} z_{d1} & \cdots & z_{dH} \end{bmatrix}$,
	where categorical external covariates are dummy coded.
	The model now requires hyperparameters $\bm{\alpha}_{\text{feat}}$, 
	$\lambda_{\text{feat}}$, $\bm{\alpha}_{\text{drug}}$, and $
	\lambda_{\text{drug}}$, which we estimate in a data-driven manner (see Section
	\ref{sec:empiricalbayes}).

  A representation of our model as a Bayesian DAG is given in Figure
  \ref{fig:dag}. We note that in many settings, the set of features might be
  different for different drugs. In that case the covariates are indexed by
  the drug $d$: $\X^d$, a trivial extension of model 
  (\ref{eq:likelihood}) and (\ref{eq:prior}). This extension is included in the
  available software, but for clarity it is omitted in the following. 
  Furthermore, several additional extensions of the model are 
  described in [! ref to second paper here].

  \begin{figure}
    \centering
    \tikz{ %
      % Y
      \node[obs] (y) {$y_{id}$}; %
      \factor[right=of y, yshift=1.5cm, xshift=-0.1cm] {y-f} 
      {above:$\mathcal{N}$} {} {} ; %

      % X
      \node[det, above=of y, yshift=1cm] (x) {$x_{ij}$}; %

      % sigma2
      \node[latent, right=of y] (sigma2) {$\sigma_d^2$}; %

      % beta
      \node[latent, right=of x] (beta) {$\beta_{jd}$}; %
      \factor[right=of beta, yshift=-1.4cm, xshift=-0.1cm] {beta-f} 
      {above:$\mathcal{N}$} {} {} ; %

      % gamma2
      \node[latent, right=of beta] (gamma2) {$\gamma_{jd}^2$}; %
      \node[const, right=of gamma2, yshift=0.5cm] (phi) {$\phi_{jd}$}; %
      \node[const, right=of gamma2, yshift=-0.5cm] (lambdaf) 
      {$\lambda_{\text{feat}}$}; %
      \factor[right=of gamma2] {gamma2-f} {above:$\mathcal{IG}$} {phi,lambdaf} 
      {gamma2} ; %
      \node[det, right=of gamma2, xshift=2cm, yshift=0.5cm] (c) {$c_{jdg}$}; %

      % tau2
      \node[latent, right=of sigma2] (tau2) {$\tau_d^2$}; %
      \node[const, right=of tau2, yshift=0.5cm] (chi) {$\chi_{d}$}; %
      \node[const, right=of tau2, yshift=-0.5cm] (lambdad) 
      {$\lambda_{\text{drug}}$}; %
      \factor[right=of tau2] {tau2-f} {above:$\mathcal{IG}$} {chi,lambdad} 
      {tau2} ; %
      \node[det, right=of tau2, xshift=2cm, yshift=0.5cm] (z) {$z_{dh}$}; %

      % edges
      \factoredge {x,beta,sigma2} {y-f} {y} ; %
      \factoredge {sigma2,gamma2,tau2} {beta-f} {beta} ; %
      \edge {c} {phi}; %
      \edge {z} {chi}; %

      % plates
      \plate{plateg} {(c)} {$g=1,\dots,G$}
      \plate{plateh} {(z)} {$h=1,\dots,H$}
      \plate{platei} {(y) (x)}{$i=1,\dots,n$}; %
      \plate{platej} {(x) (beta) (gamma2) (phi) (lambdaf) (c) (plateg)}
      {$j=1,\dots,p$}; %
      \plate{plated} 
      {(y) (x) (sigma2) (beta) (gamma2) (tau2) (phi) (lambdaf)
      (chi) (lambdad) (c) (z) (platei) (platej) (plateg) (plateh)}
      {$d=1,\dots,D$}; %
    }
    \caption{Hierarchical representation of the drug sensitivity prediction 
    model. Grey circles 
    represent observed variables, white circles represent unobserved variables,
    squares represent fixed data, and unenclosed letters are parameters to be
    estimated. Cell lines are indexed by $i$, features
    by $j$, drugs by $d$, drug covariates by $h$, and feature covariates by 
    $g$. The $y_{id}$ are the drug sensitivities, $x_{ij}$ the 
    molecular features, 
    $c_{jdg}$ the external feature covariates, $z_{dh}$ the external drug
    covariates, $\beta_{jd}$ the regression coefficients, $\sigma_d^2$ the 
    error variances, $\tau_d^2$ and $\gamma_{jd}^2$ the drug and feature
    specific variance components, respectively, and $\phi_{jd}$, 
    $\lambda_{\text{feat}}$, $\chi_{d}$, and $\lambda_{\text{drug}}$ the 
    hyperparameters.}
    \label{fig:dag}
  \end{figure}

	\section{Estimation}\label{sec:estimation}
	\subsection{Variational Bayes}\label{sec:variationalbayes}
	The posterior corresponding to the model described in (\ref{eq:likelihood})
	and (\ref{eq:prior}) is not available in closed form. 
	To avoid computationally intensive markov chain Monte Carlo (MCMC) 
	algorithms, we approximate the joint posterior by variational Bayes (see 
	\cite{blei_variational_2017} for a review), where the approximate
	posterior density factorises as: $p(\bbeta_d, \bm{\gamma}^2_d, \tau_{d}^2, 
	\sigma_d^2 | \y_d) \approx 
	Q_d(\cdot) = q(\bbeta_d) \cdot q(\bm{\gamma}_{d}^2) \cdot q(\tau_{d}^2) 
	\cdot q(\sigma_d^2)$, where $\bm{\gamma}_{d}^2 = 
	\begin{bmatrix} \gamma_{1d}^2 & \cdots & \gamma_{pd}^2 \end{bmatrix} \tr$. For
	notational convenience, we slightly abuse notation and let $q(\cdot)$ denote
	different densities for different inputs.
	Under such a factorisation, the marginal variational posteriors that 
	minimise the Kullback-Leibler divergence of the true posterior to the 
	variational Bayes approximation \cite[]{neal_view_1998} are given by:
	\begin{align*}
    q(\bm{\beta}_d) & \overset{D}{=} \mathcal{N}_{p} 
    (\bm{\mu}_d, \bm{\Sigma}_d), \\
    q(\bm{\gamma}_{d}^2) & \overset{D}{=} \prod_{j=1}^p \mathcal{GIG}
    \(-1, \lambda_{\text{feat}}/\phi_{jd}^2, \delta_{jd}\), \\
    q(\tau^2_d) & \overset{D}{=} \mathcal{GIG} \(-\frac{p + 1}{2}, 
    \lambda_{\text{drug}}/\chi_d^2, \eta_d\), \\
    q(\sigma^2_d) & \overset{D}{=} \Gamma^{-1} \(\frac{n + p + 1}{2}, \zeta_d\).
  \end{align*}
	See SM Section \ref{sm-sec:vbderivations} for the derivations. The variational
	parameters $\bm{\mu}_d$, $\bm{\Sigma}_d$, $\delta_{jd}$, $\eta_d$, and
	$\zeta_d$ contain cyclic dependencies and are iteratively 
	updated by:
	\begin{subequations}\label{eq:vbequations}
    \begin{align}
      \bm{\Sigma}_d^{(h+1)} & = (a_d^{(h)})^{-1} 
      \[\X \tr \X + g_{d}^{(h)} \diag(b_{jd}^{(h)})\]^{-1}, \\
      \bm{\mu}_d^{(h+1)} & = \[\X \tr \X + g_{d}^{(h)} 
      \diag(b_{jd}^{(h)})\]^{-1} 
      \X \tr \y_d, \\
      \delta_{jd}^{(h+1)} & = a_d^{(h)} g_{d}^{(h)} \[(\bmu^{(h+1)}_{jd})^2 + 
      (\bSigma^{(h + 1)}_d)_{jj}\] + \lambda_{\text{feat}}, \\ 
      \eta_d^{(h+1)} & = a_d^{(h)} \sum_{j=1}^{p} b_{jd}^{(h + 1)}
      \[ (\bmu^{(h+1)}_{jd})^2 + (\bSigma^{(h + 1)}_d)_{jj} \] + 
      \lambda_{\text{drug}}, \\
      \zeta_d^{(h+1)} & = \frac{1}{2} \bigg[\mathbf{y}_d \tr \mathbf{y}_d -
      2 \mathbf{y}_d \tr \X \bm{\mu}_d^{(h+1)} + 
      \trace ( \X \tr \X \bm{\Sigma}_d^{(h+1)}) + 
      (\bm{\mu}_d^{(h+1)}) \tr \X \tr \X \bm{\mu}_d^{(h+1)} \\
      & \,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\, + 
      g_{d}^{(h + 1)} \trace \[ \diag (b_{jd}^{(h+1)}) \bm{\Sigma}_d^{(h+1)}\] + 
      g_{d}^{(h + 1)} (\bm{\mu}_d^{(h+1)}) \tr \diag (b_{jd}^{(h+1)}) 
      \bm{\mu}_d^{(h+1)}\bigg],
    \end{align}
  \end{subequations}
  until convergence, where $\y_d = \begin{bmatrix} y_{1d} & \cdots & y_{nd}
  \end{bmatrix} \tr$. Here, we set
  \begin{align}
    a_d^{(h)} & =\E_{Q^{(h)}}(\sigma_d^{-2})=(n + p + 1)/(2 \zeta_d^{(h)}), 
    \nonumber\\
    b_{jd}^{(h)} & = \E_{Q^{(h)}}(\gamma_{jd}^{-2}) = 
    \sqrt{\frac{\lambda_{\text{feat}}}{\phi_{jd}^2 \delta_{jd}^{(h)}}}
    \frac{K_0\(\sqrt{\delta_{jd}^{(h)}\lambda_{\text{feat}}/\phi_{jd}^2}\)}
    {K_1\(\sqrt{\delta_{jd}^{(h)}\lambda_{\text{feat}}/\phi_{jd}^2}\)} + 
    \frac{2}{\delta_{jd}^{(h)}}, \label{eq:ratiomodifiedbessel} \\
    g_d^{(h)} & =\E_{Q^{(h)}}(\tau_d^{-2})=
    \sqrt{\frac{\lambda_{\text{drug}}}{\chi_{d}^2 \eta_{d}^{(h)}}}
    \frac{K_{(p-1)/2}\(\sqrt{\eta_{d}^{(h)}\lambda_{\text{drug}}/\chi_{d}^2}\)}
    {K_{(p+1)/2}\(\sqrt{\eta_{d}^{(h)}\lambda_{\text{drug}}/\chi_{d}^2}\)} + 
    \frac{p + 1}{\eta_{d}^{(h)}}.
    \nonumber
  \end{align}
  where $K_{\nu}(x)$ denotes the modified Bessel function of the second kind.
  A method for fast and numerically stable calculation of ratios of
  modified Bessel functions of the second kind, as in 
  (\ref{eq:ratiomodifiedbessel}), is given in SM Section 
  \ref{sm-sec:ratiosmodifiedbessels}.

  \subsection{Empirical Bayes}\label{sec:empiricalbayes}
  We parametrised the prior 
  mean of the $\gamma_{jd}^2$ as $\phi_{jd}=(\mathbf{c}_{jd} \tr 
  \bm{\alpha}_{\text{feat}})^{-1}$ and the prior mean of $\tau_d^2$ as
  $\chi_{d}=(\mathbf{z}_{d} \tr \bm{\alpha}_{\text{drug}})^{-1}$.
  This parametrisation allows us to include feature and drug covariates,
  both continuous and discrete, into the 
  model. Additionally, it reduces the number of hyperparameters from 
  $pD$ to $|\bm{\alpha}_{\text{feat}}| + |\bm{\alpha}_{\text{drug}}| + 2$. The
  Bayesian model
  then requires the specification of the hyperparameters 
  $\bm{\alpha}=\begin{bmatrix} \bm{\alpha}_{\text{feat}} \tr & 
  \bm{\alpha}_{\text{drug}} \tr 
  \end{bmatrix} \tr$ and $\bm{\lambda} = \begin{bmatrix} \lambda_{\text{feat}} &
  \lambda_{\text{drug}}
  \end{bmatrix} \tr$. These are
  abstract and hard to interpret parameters 
  for which we generally lack expert knowledge. They do, however, have a 
  significant influence on the shape of the posterior distribution. We 
  therefore propose to estimate these hyperparameters by empirical Bayes.
  In our case, this results in an   
  objective and data-driven inclusion of the feature and drug covariates.

	The canonical method for empirical Bayes is to maximise the marginal 
	likelihood with respect to the hyperparameters. In 
	\cite{casella_empirical_2001} the marginal likelihood is maximised by an EM 
	algorithm:
	\begin{align*}
	  \bm{\alpha}^{(l+1)},\bm{\lambda}^{(l+1)} & = \underset{\bm{\alpha},
	  \bm{\lambda}}{\argmax}\E_{\cdot | \Y} 
	  [\log p(\Y, \B, \bm{\Gamma}^2, \bm{\tau}^2, \bsigma^2) | \bm{\alpha}^{(l)},
	  \bm{\lambda}^{(l)}] \\
	  & = \underset{\bm{\alpha},\bm{\lambda}}{\argmax} 
	  \E_{\cdot | \Y} [\log \pi (\bm{\Gamma}^2) | \bm{\alpha}_{\text{feat}}^{(l)},
	  \bm{\lambda}_{\text{feat}}^{(l)}] + 
	  \E_{\cdot | \Y} [\log \pi (\bm{\tau}^2) | 
	  \bm{\alpha}_{\text{drug}}^{(l)}, \bm{\lambda}_{\text{drug}}^{(l)}],
	\end{align*}
	where $\Y = \begin{bmatrix} \y_1 & \cdots & \y_D \end{bmatrix}$,
	$\B = \begin{bmatrix} \bbeta_1, \dots, \bbeta_D \end{bmatrix}$, 
	$\bm{\tau}^2 = \begin{bmatrix} \tau^2_1 & \cdots & 
	\tau^2_D \end{bmatrix} \tr$,
	$\bsigma^2 = \begin{bmatrix} \sigma^2_1 & \cdots & 
	\sigma^2_D \end{bmatrix} \tr$,
	and $\bm{\Gamma}^2 = \begin{bmatrix} \bgamma^2_1 & \cdots & \bgamma^2_D 
	\end{bmatrix}$,
	and the expectation is with respect to the joint posterior. In our case, this 
	posterior is not available in closed form, which renders the expectation 
	difficult. While \cite{casella_empirical_2001} suggests to approximate the 
	expectation by a Monte Carlo sample, we propose to use the variational Bayes 
	approximation developed in Section \ref{sec:variationalbayes}:
	$$
	\bm{\alpha}^{(l+1)},\bm{\lambda}^{(l+1)} = 
	\underset{\bm{\alpha},\bm{\lambda}}{\argmax}\E_{Q^{(l)}} 
	[\log \pi(\bm{\Gamma}^2)| \bm{\alpha}_{\text{feat}}^{(l)}, 
	\bm{\lambda}_{\text{feat}}^{(l)}] + 
	\E_{Q^{(l)}} [\log \pi(\bm{\tau}^2)| 
	\bm{\alpha}_{\text{drug}}^{(l)}, \bm{\lambda}_{\text{drug}}^{(l)}],
	$$
	where now the expectation is with respect to the converged variational 
	posterior $Q^{(l)}=\prod_{d=1}^D Q_d^{(l)}$.

	If we stack the drug and feature covariates:
	\begin{align*}
	  \mathbf{C} = 
	  \begin{bmatrix} \mathbf{c}_{11} \tr \\ 
	    \vdots \\
	    \mathbf{c}_{p1} \tr \\
	    \vdots \\
	    \mathbf{c}_{1D} \tr \\
	    \vdots \\
	    \mathbf{c}_{pD} \tr
	  \end{bmatrix} \text{ and }
	  \mathbf{Z} =
	  \begin{bmatrix} \mathbf{z}_{1} \tr \\ 
	    \vdots \\
	    \mathbf{z}_D \tr
	  \end{bmatrix},
	\end{align*}
	the empirical Bayes updates are given by:
  \begin{align*}
    \bm{\alpha}^{(l+1)}_{\text{feat}} & = \[ \mathbf{C} \tr \diag 
    (e_{jd}^{(l)}) \mathbf{C} \]^{-1} \mathbf{C} \tr 
    \mathbf{1}_{pD \times 1}, \\
    \lambda^{(l+1)}_{\text{feat}} & = pD \[ \sum_{d=1}^D \sum_{j=1}^{p} 
    b_{jd}^{(l)} + (\bm{\alpha}^{(l+1)}_{\text{feat}}) \tr \mathbf{C} \tr 
    \diag (e_{jd}^{(l)}) \mathbf{C} \bm{\alpha}^{(l+1)}_{\text{feat}} - 
    2 (\bm{\alpha}^{(l+1)}_{\text{feat}}) \tr \mathbf{C} \tr 
    \mathbf{1}_{pD \times 1} \]^{-1},\\
    \bm{\alpha}^{(l+1)}_{\text{drug}} & = \[ \mathbf{Z} \tr \diag 
    (f_{d}^{(l)}) \mathbf{Z} \]^{-1} \mathbf{Z} \tr 
    \mathbf{1}_{D \times 1}, \\
    \lambda^{(l+1)}_{\text{drug}} & = D \[ \sum_{d=1}^D g_{d}^{(l)} + 
    (\bm{\alpha}^{(l+1)}_{\text{drug}}) \tr \mathbf{Z} \tr \diag (f_{d}^{(l)})
    \mathbf{Z} \bm{\alpha}^{(l+1)}_{\text{drug}} - 
    2 (\bm{\alpha}^{(l+1)}_{\text{drug}}) \tr \mathbf{Z} \tr 
    \mathbf{1}_{D \times 1} \]^{-1},
  \end{align*}
  where 
  \begin{align*}
    e_{jd}^{(l)} & = \E_{Q^{(l)}}(\gamma_{jd}^2| 
    \bm{\alpha}_{\text{feat}}^{(l)}, 
    \lambda_{\text{feat}}^{(l)}) = (b_{jd}^{(l)} - 
    2/\delta_{jd}^{(l)}) \cdot \delta_{jd}^{(l)} (\phi_{jd}^{(l)})^2/
    \lambda_{\text{feat}}^{(l)}, \\
    f_{d}^{(l)} & = \E_{Q^{(l)}}(\tau_{d}^2| \bm{\alpha}_{\text{drug}}^{(l)}, 
    \lambda_{\text{drug}}^{(l)}) = (g_{d}^{(l)} - 
    (p + 1)/\eta_{d}^{(l)}) \cdot \eta_{d}^{(l)} (\chi_{d}^{(l)})^2/
    \lambda_{\text{drug}}^{(l)}.
  \end{align*}

  To ensure proper and unbiased shrinkage, intercepts are included in 
  $\bm{\alpha}_{\text{feat}}$ and $\bm{\alpha}_{\text{drug}}$. This is achieved 
  by appending both
  $\mathbf{C}$ and $\mathbf{Z}$ with a column of 
  ones. These intercepts are roughly interpreted as the
  expected prior precisions $\E(\gamma_{jd}^{-2})$ and $\E(\tau_d^{-2})$
  if the feature and drug covariates are all zero. 
  Likewise, an $\alpha$ corresponding to an external covariate may be
  interpreted as an additive effect of the external covariate on the prior
  expected precision. So an $\alpha=1$ translates to an increase in expected 
  prior precision of
  1 for every increase in the external covariate of 1,
	keeping all the other external covariates fixed.

  Variational Bayes approximations are known to underestimate posterior 
  variances \cite[]{rue_approximate_2009,consonni_mean-field_2007,
  bishop_pattern_2006,wang_inadequacy_2005}. If posterior variances are 
  required, 
  we suggest generating samples from the posterior with fixed hyperparameters
  estimates (after the procedure described in Section \ref{sec:empiricalbayes}
  has converged). A Gibbs sampler is described in 
  SM Section \ref{sm-sec:gibbssampler}. Alternatively, we provide an
  implementation of the proposed model in stan using the R package rstan 
  \cite[]{guo_rstan:_2018} at 
  \url{https://github.com/magnusmunch/cambridge}.

	\section{Simulations}
	\subsection{Fixed drug component}
	This section investigates the empirical 
	Bayes estimation properties of the model in a simulated setting; its main aim 
	is to assess hyperparameter recovery. It is a data based simulation,
	wherein the responses are simulated from a synthetic model, but the
	features are taken from the real GDSC data introducted in Section 
	\ref{sec:gdsc}. 
	The real GDSC features contain strong collinearities. Such strong
	collinearities in the design matrix impede correct parameter estimation. 
	We therefore replace the ambition of correctly estimating the $\beta_{jd}$
	with the more modest aim of approximately correct estimation of the 
	hyperparameters.

	A pre-processing step, as explained in Section \ref{sec:gdsc}, selects around 
	300 features per drug, while the 168 drug sensitivities for 482 cell lines are 
	simulated from model (\ref{eq:likelihood}) and 
	(\ref{eq:prior}), with two adjustments: we fix $\forall d: \tau_d^2=1$ 
	and draw the error variances as $\forall d: \sigma_d^2 \sim \Gamma^{-1}(3,2)$,
	such that the prior $\sigma_d^2$ mean and variance are both one.
	We create four external dummy feature covariates that code for four 
	randomly assigned groups of features. We set $\bm{\alpha}_{\text{feat}}$ such
	that the $\gamma_{jd}^2$ of the four groups of features have prior means
	$\phi_{jd} \in \{ 1, 1/2, 1/4, 1/8 \}$ 
	($\bm{\alpha}_{\text{feat}}=\begin{bmatrix} 1 & 1 & 3 & 7 \end{bmatrix} \tr$). 
	The prior scale parameter is set to $\lambda_{\text{feat}}=1$. We estimate 
	four models: (i) the NIG model as in Section \ref{sec:estimation}, (ii) an 
	NIG model where we fix $\lambda_{\text{feat}}$ to its true value, and two NIG
	models without the inclusion of the external covariates, again with 
	$\lambda_{\text{feat}}$ either (iii) estimated or (iv) fixed to its true 
	value. Exclusion of the external covariates as in models (iii) and (iv) 
	amounts to direct estimation of one common expected 
	prior mean $\phi$, instead of regression estimates for the $\phi_{jd}$ as
	in models (i) and (ii). All 
	models are fit on 241 randomly selected cell lines, while performance measures 
	are estimated from the remaining 241 cell lines.

<<include=FALSE>>=
load("results/simulations_gdsc_set1.Rdata")
mapmse <- apply(res[, c(7:12)], 2, median)
mapmset <- apply(res[, c(13:18)], 2, median)
melbo <- apply(elbo, 2, median)
phi <- 1/(set$alphaf %*% t(unname(model.matrix(~ factor(1:4)))))
mphi <- apply(1/est[, c(11, 13)], 2, median)
mlambdaf <- apply(est[, c(5, 12)], 2, median)
@	
	Figure \ref{fig:simulations_gdsc_est1} shows the estimated
	$\bm{\alpha}_{\text{feat}}$ together with its true value for models (i) and 
	(ii). The figure shows that $\bm{\alpha}_{\text{feat}}$ is consistently
	underestimated 
	in the full NIG model (both with estimated and fixed $\lambda_{\text{feat}}$),
	while $\lambda_{\text{feat}}$ is overestimated as 
	\Sexpr{round(mlambdaf[1], 2)} in model (i) (the true value is 1). 
	This translates into
	underestimated prior means $\phi_{jd}$ as presented in Figure 
	\ref{fig:simulations_gdsc_phi1}, which in turn implies underestimated 
	$\beta_{jd}$ variances. The overestimated $\lambda_{\text{feat}}$ translates 
	into underestimated $\gamma_{jd}^2$ variances. Underestimation of variances is 
	a common phenomenon in variational Bayes approximations. In addition, 
	$\bm{\alpha}_{\text{feat}}$ and $\lambda_{\text{feat}}$ are higher level 
	parameters, which are, in general, hard to estimate. We note 
	that even though $\bm{\alpha}_{\text{feat}}$ and thus $\phi_{jd}$ are 
	estimated with bias, the orders of their elements are correct.
<<simulations_gdsc_est1, fig.cap="Estimated hyperparameters $\\bm{\\alpha}_{\\text{feat}}$ from simulations with (a) estimated and (b) fixed $\\lambda_{\\text{feat}}$ (models (i) and (ii), respectively).", out.width="100%", fig.asp=1/2>>= 
@

<<simulations_gdsc_phi1, fig.cap="Estimated prior means $\\bm{\\phi}_{jd}$ for the four groups of features from simulations with (a) estimated and (b) fixed $\\lambda_{\\text{feat}}$ (models (i) and (ii), respectively).", out.width="100%", fig.asp=1/2>>= 
@

	The models (iii) and (iv) give median $\phi$ estimates of
	\Sexpr{round(mphi[1], 2)} and \Sexpr{round(mphi[2], 2)}, 
	respectively. The $\phi$ estimates are
	underestimated, compared to true $\phi$ mean \Sexpr{round(mean(phi), 2)}. 
	Model (iii) slightly underestimates the median
	$\lambda_{\text{feat}}$ at \Sexpr{round(mlambdaf, 2)[2]} 
	(the true value is one) and thus overestimates the 
	prior $\gamma_{jd}^2$ variance. This is likely to compensate for the common
	prior mean 
	$\phi$, that is not able to capture part of the variation in the true feature
	specific $\phi_{jd}$. The median evidence 
	lower bounds (ELBO), which approximate the marginal likelihoods of models
	(i)-(iv), are estimated as \Sexpr{round(melbo, 2)[1:3]}, and 
	\Sexpr{round(melbo, 2)[4]}, respectively. This implies that model (i),
	with estimated $\bm{\alpha}_{\text{feat}}$ and $\lambda_{\text{feat}}$ fits
	the data best (the log Bayes factors comparing models (ii)-(iv) to model (i) 
	are \Sexpr{round(melbo[-c(1, 4)] - melbo[1], 2)}, and 
	\Sexpr{round(melbo[4] - melbo[1], 2)}, respectively, and thus highly favouring
	model (i)). The prediction MSEs for the four models on the test 
	data and gaps between training and test prediction MSE are displayed in Table
	\ref{tab:pmse1}. Table \ref{tab:pmse1} confirms that the models that
	include the external data (models (i) and (ii)) learn the underlying structure 
	in the data better than models (iii) and (iv) that do not include the external 
	data.
<<pmse1>>= 
library(knitr, quietly=TRUE, verbose=FALSE)
suppressWarnings(library(kableExtra))
load("results/simulations_gdsc_set1.Rdata")
mapmse <- apply(res[, c(7:12)], 2, median)
mapmset <- apply(res[, c(13:18)], 2, median)
tab <- rbind(mapmse, mapmse - mapmset)[, -c(5, 6)]
rownames(tab) <- c("$\\text{MSE}_{\\text{test}}$",
                  "$\\text{MSE}_{\\text{test}} - \\text{MSE}_{\\text{train}}$")
options(knitr.kable.NA="", digits=3)
kable(tab,
      col.names=c("(i)", "(ii)", "(iii)", "(iv)"),
      caption="Median prediction MSEs for the models (i)-(iv) in the fixed 
      $\\tau_d^2$ simulation setting.", 
      format="latex", booktabs=TRUE, escape=FALSE) %>%
  kable_styling(latex_options=c("HOLD_position"))
@

  \subsection{Random drug component}
	To investigate the interplay of the feature and drug variance components, we 
	repeat the whole simulation study with random $\tau_d^2$. Similarly as with 
	the $\gamma_{jd}^2$, we randomly create four groups of drugs and corresponding
	external drug dummy covariates. We simulate the $\tau_d^2$ as in
	(\ref{eq:tauprior}), where we set $\bm{\alpha}_{\text{drug}} = 
	\begin{bmatrix} 1 & 1 & 3 & 7 \end{bmatrix} \tr$, such
	that the $\tau_{d}^2$ of the four groups of features have prior means
	$\chi_{d} \in \{ 1, 1/2, 1/4, 1/8 \}$. The scale $\lambda_{\text{drug}}$ is 
	set to one. We estimate the same four models as in the fixed $\tau_d^2$ 
	setting, but in addition estimate the $\tau_d^2$ priors as well. Again, we
	use two different versions: one where we estimate 
	$\bm{\lambda}$ and one where we fix it to its true value.

<<include=FALSE>>=
load("results/simulations_gdsc_set2.Rdata")
mapmse <- apply(res[, c(7:12)], 2, median)
mapmset <- apply(res[, c(13:18)], 2, median)
melbo <- apply(elbo, 2, median)
phi <- 1/(set$alphaf %*% t(unname(model.matrix(~ factor(1:4)))))
chi <- 1/(set$alphad %*% t(unname(model.matrix(~ factor(1:4)))))
mphi <- apply(1/est[, c(21, 25)], 2, median)
mchi <- apply(1/est[, c(22, 26)], 2, median)
mlambdaf <- apply(est[, c(9, 23)], 2, median)
mlambdad <- apply(est[, c(10, 24)], 2, median)
@	
	Figures \ref{fig:simulations_gdsc_est2} and 
	\ref{fig:simulations_gdsc_phi2} display the estimated $\bm{\alpha}$, 
	$\phi_{jd}$, and $\chi_d$ for models (i) and (ii). The scale parameters were
	estimated as $\lambda_{\text{feat}}=\Sexpr{round(mlambdaf[1], 2)}$ and
	$\lambda_{\text{drug}}=\Sexpr{round(mlambdad[1], 2)}$ in model (i). From 
	Figure \ref{fig:simulations_gdsc_est2} we see that estimation of
	$\bm{\alpha}_{\text{feat}}$ is again biased. However, from 
	Figure \ref{fig:simulations_gdsc_phi2} we see that the scale and ordering of
	the $\phi_{jd}$ are correclty estimated. We conjecture that the drop in 
	estimation accuracy compared to the previous setting stems from the
	difficult estimation of the $\tau_d^2$ prior. Estimation of this prior 
	depends on only 168 observed drugs, while the $\gamma_{jd}^2$ prior is 
	is estimated from approximately $300 \cdot 168=50400$ features. 
	\cite{wiel_learning_2019} show in a simple regression setting that if the 
	number of samples and the number of parameters on which the EB is based are of
	the same order, empirical Bayes performs worse than when one of the two is
	large compared to the other. 
	Here, the number of samples $n=241$ and $D=168$ are of the same order, leading
	to a suboptimal EB estimation performance.

<<simulations_gdsc_est2, fig.cap="Estimated hyperparameters $\\bm{\\alpha}_{\\text{feat}}$ and $\\bm{\\alpha}_{\\text{drug}}$ from simulations with (a) \\& (c) estimated and (b) \\& (d) fixed $\\lambda_{\\text{feat}}$ and $\\lambda_{\\text{drug}}$ (models (i) and (ii), respectively).", out.width="100%", fig.asp=1>>= 
@

<<simulations_gdsc_phi2, fig.cap="Estimated prior means $\\bm{\\phi}_{jd}$ and $\\bm{\\chi}_{d}$ for the four groups of features and drugs from simulations with (a) \\& (c) estimated and (b) \\& (d) fixed $\\lambda_{\\text{feat}}$ and $\\lambda_{\\text{drug}}$ (models (i) and (ii), respectively).", out.width="100%", fig.asp=1>>= 
@

  We argue however, that even though the prior parameters are not estimated
  correctly, there is still some gain in predictive performance.
  Prediction MSEs on the test data the training-test gaps in prediction MSE are 
  shown in Table \ref{tab:pmse2}. The comparisons of models (i) and
	(ii) vesus models (iii) and (iv) confirm that even though EB estimation is
	suboptimal, the external data do add something in predictive performance. 
<<pmse2>>= 
library(knitr, quietly=TRUE, verbose=FALSE)
suppressWarnings(library(kableExtra))
load("results/simulations_gdsc_set2.Rdata")
mapmse <- apply(res[, c(7:12)], 2, median)
mapmset <- apply(res[, c(13:18)], 2, median)
tab <- rbind(mapmse, mapmse - mapmset)[, -c(5, 6)]
rownames(tab) <- c("$\\text{MSE}_{\\text{test}}$",
                  "$\\text{MSE}_{\\text{test}} - \\text{MSE}_{\\text{train}}$")
options(knitr.kable.NA="", digits=3)
kable(tab,
      col.names=c("(i)", "(ii)", "(iii)", "(iv)"),
      caption="Median prediction MSEs for the models (i)-(iv) in the random 
      $\\tau_d^2$ simulation setting.", 
      format="latex", booktabs=TRUE, escape=FALSE) %>%
  kable_styling(latex_options=c("HOLD_position"))
@

	\section{GDSC data}\label{sec:gdsc}
	\subsection{Primary data}
	The GDSC project's \cite[]{yang_genomics_2013} aim is ``to improve cancer 
	treatments by discovering therapeutic biomarkers that can be used to identify 
	patients most likely to respond to anticancer drug". Part of the project is to
	screen $>1000$ human cancer cell lines for drug sensitivities. The cell lines
	have been genetically characterised and several drug sensitivity measures are
	recorded.
	The data is freely available from \cite{garnett_systematic_2012} and consist 
	of: (i) the sensitivity measures of the cell lines 
	to the drugs, (ii) annotation of the 
	screened compounds, and (iii) the cell lines' genomic profile (mutations, 
	copy numbers, methylation profiles, and gene expression). We will attempt to 
	predict drug sensitivities of the cell lines, as quantified by half maximal
	inhibitory concentration (IC50), using the
	gene expression data. Other choices of sensitivity measures than IC50 are 
	possible, but a discussion on the pros and cons of different sensitivity
	measures is beyond the aim of this paper. A comprehensive penalized regression
	approach such as in 
	{\cite{mai_composite_2019} uses all the available molecular markers 
	instead of just gene expression (68 mutations, 426 copy numbers, 
	and 2602 gene expression levels for 498 cell lines and 97 drugs). The 
	current version of our software is not able to handle such problem sizes,
	due to the repeated calculation of posterior parameters in the empirical
	Bayes iterations. Currently, problems of up to 1000 outcomes, 1000 features,
	and 1000 samples are computationally feasible.

	We average repeated measures over cell line-drug combinations and model the
	logarithm of the IC50 values. In the following, IC50 refers 
	to the log-transformed values. 
	After iteratively removing the cell line or drug with the highest proportion 
	of missings until no more missing values remain, we end up with IC50 
	estimates for 482 cell lines on 168 drugs. As a preliminary step, we fit
	separate elastic net models for each drug with the \texttt{glmnet} package
	\cite[]{friedman_regularization_2010}, with $\alpha=0.01$ and $\lambda$
	chosen such that we select approximately 300 genes per drug.

	\subsection{External data}
	Two ternary drug covariates are available: the developmental stage
	(experimental, in clinical development, or clinically approved) of the
	drugs and the action (unkown, cytotoxic, or targeted) of the drugs. 
	Furthermore, we have a binary feature covariate available that indicates
	whether a gene belongs to the drug target pathway. The drug covariates are 
	taken
	directly from the GDSC database's annotation file. The feature covariate
	was created by comparing the target pathways in the GDSC annotation to
	the KEGG \cite[]{kanehisa_kegg:_2000} and reactome 
	\cite[]{fabregat_reactome_2018} repositories. The external covariates are 
	dummy-coded with
	reference categories clinically approved drugs, cytotoxic drugs, and 
	features that are not in the target pathway.

	We expect that drugs that have been clinically approved are easiest to 
	predict and hence yield the largest prior $\beta_{jd}$ variances, followed by
	the drugs in clinical development, and the experimental drugs. Likewise we 
	expect the targeted drugs to yield the largest prior $\beta_{jd}$ variances, 
	followed by the cytotoxic drugs, and the unkown target drugs. For the
	feature covariate, we expect that genes that are in the pathway of the drug
	are more predictive than genes that are not, i.e., 
	they have larger prior $\beta_{jd}$ variances than drugs that are not in 
	the pathway.

%   - check whether compounds are PAINS \cite[]{baell_new_2010} or not. 
%   PAINS indicates whether compound is frequent false hitter.
%   
%   - clinical trials database CTRP might contain info
% 
%   - extra data on GDSC might be in PharamcoGx R-package. Also contains bunch 
%   of functions.
%   
%   - check GSK open target EBI or ask Chris Wallace about it. Database that links
% 	targets to drugs and diseases. Might be useful for drugs co-data.
% 	
% 	- try non-negative matrix factorisation/independent component analysis on 
% 	binary fingerprints to get principal components that characterise the 
% 	different drugs.

	\subsection{Results}
	To compute performance measures, we randomly split the data into training and
	test sets, containing 
	241 cell lines each, estimate the models on the training data and compute 
	performance measures on the test data. We repeated this procedure 50 times
	and average performance measures over the splits.
	The estimated 
	models are: two models like in (\ref{eq:prior}) but with different external 
	variables, named (i) NIG1 and (ii) NIG2 in the following, (iii) a model 
	as described in \cite{kpogbezan_empirical_2017}, named bSEM, (iv) a regular
	ridge model, and (v) a regular lasso model. The NIG1 incorporates no external
	variables and estimates the common prior $\gamma_{jd}^2$ and $\tau_d^2$ 
	means $\phi$ and $\chi$ directly, while the NIG2 model includes the
	developmental phase and action drug covariates, and the pathway
	feature covariate, and estimates $\bm{\alpha}$. The bSEM model is similar to 
	the NIG model in that it 
	draws the $\beta_{jd}$ from a conditionally normal distribution, but 
	fixes $\tau_d^2=1$ and draws 
	$\gamma_{jd}^2 \sim \Gamma^{-1} (\phi_{jd}, \lambda_{jd})$, instead of 
	from the inverse Gaussian distribution. Additionally, 
	the bSEM model can 
	only incorporate one binary external covariate that divides the features into 
	two groups, i.e., $\bm{\alpha} = \begin{bmatrix} \alpha_0 & \alpha_1 
	\end{bmatrix} \tr$ and $\bm{c}_{jd} = \begin{bmatrix} 1 & 0 \end{bmatrix} 
	\tr$ if feature $jd$ belongs to group one and $\bm{c}_{jd} = \begin{bmatrix} 
	1 & 1 \end{bmatrix} \tr$ if feature $jd$ belongs to group two. Both 
	$\bm{\alpha}_{\text{feat}}$ and $\lambda_{\text{feat}}$ are
	estimated in a similar fashion as in Section \ref{sec:estimation}.

  From the $\bm{\alpha}$ estimates in Tables \ref{tab:results_drug} and
  \ref{tab:results_feat} we see that there
  seems to be an effect of the
  external variables. If we look at the NIG1 model, that includes all external 
  covariates, we see that approved drugs are easier to predict, due to small
  prior precision, followed by drugs that are in clinical development,
  as we expected. The cytotoxic drugs are the hardest to predict, 
  with the largest prior precision (smallest prior variance), followed by the 
  targeted drugs and the unknown drugs. This is again according to expectation.
  For the unkown drugs, we are unable to confirm the small precision. Genes 
  that are in the target pathway have 
  smaller prior precisions, according to expectation, which confirms that these
  genes are more important for prediction of drug sensitivity. The estimates 
  from
  the bSEM model are also according to expectation: genes that are in the target
  pathway have a much smaller precision than genes that are not in the target
  pathway. The 
  effects of the pathway indicator seem to be much larger in the bSEM model. 
<<results_drug>>= 
library(knitr, quietly=TRUE, verbose=FALSE)
suppressWarnings(library(kableExtra))
load("results/data_gdsc_res1.Rdata")
tab1 <- cbind(tab[, 1], c(NA, 0, NA), tab[, c(3, 2)], c(NA, 0, NA), 
              tab[, c(4, 5)])
options(knitr.kable.NA="", digits=3)
kable(tab1,
      col.names=c("intercept", "approved", "development", 
                  "experimental", "targeted", "unkown", "cytotoxic"),
      caption="$\\bm{\\alpha}_{\\text{drug}}$ estimates from GDSC data.", 
      format="latex", booktabs=TRUE, escape=FALSE) %>%
  kable_styling(latex_options=c("HOLD_position"))
@

<<results_feat>>= 
library(knitr, quietly=TRUE, verbose=FALSE)
suppressWarnings(library(kableExtra))
load("results/data_gdsc_res1.Rdata")
tab2 <- cbind(tab[, 6], c(NA, 0, 0), tab[, 7])
options(knitr.kable.NA="", digits=3)
kable(tab2,
      col.names=c("intercept", "not in pathway", "in pathway"),
      caption="$\\bm{\\alpha}_{\\text{feat}}$ estimates from GDSC data.", 
      format="latex", booktabs=TRUE, escape=FALSE) %>%
  kable_styling(latex_options=c("HOLD_position"))
@

  The median prediction MSEs and the median ELBOs, calculated on the test data
  are displayed in Table \ref{tab:results_data}. In terms of 
  is lowest in the NIG models 
  (\Sexpr{round(mapmse[2], 2)} and \Sexpr{round(mapmse[3], 2)}, in the
  NIG1 and NIG2 models, respectively). The bSEM model is not able to 
  improve on the null model with an median prediction MSE of 
  \Sexpr{round(mapmse[4], 2)} compared to the null model's MSE of 
  \Sexpr{round(mapmse[1], 2)}.
<<results_data>>= 
load("results/data_gdsc_res1.Rdata")
mapmse <- apply(apmse, 2, median)
melbo <- apply(elbo, 2, median)
tab <- rbind(mapmse, c(melbo, NA, NA))
rownames(tab) <- c("MSE", "ELBO")
options(knitr.kable.NA="", digits=3)
kable(tab,
      col.names=c("NIG1", "NIG2", "bSEM", "lasso", "ridge"),
      caption="Median prediction MSE and median ELBO estimated on the test data", 
      format="latex", booktabs=TRUE, escape=FALSE) %>%
  kable_styling(latex_options=c("HOLD_position"))
@

  In addition to the empirical Bayes estimates and the prediction MSE, we 
  investigated the stability of the estimates by ranking the posterior means
  $\E_{Q}(\bbeta_d)$ and calculating the Spearman's rank
  correlation over the 50 data splits. They are depicted in Figure 
  \ref{fig:data_gdsc_res1_brank_hist}. In both NIG models the correlations
  between splits are higher than in the bSEM model; an indication that 
  estimation in the NIG model is more stable than in the bSEM model.
<<data_gdsc_res1_brank_hist, fig.cap="Spearman's rank correlation between the estimated posterior means $\\E_{Q}(\\bbeta_d)$ in the 50 data splits.", out.width="70%", fig.asp=2/3>>= 
@

	% 
	% \subsection{CCLE}
	% - extra data on CCLE might be in PharamcoGx R-package
	% 
	% \subsection{NCI-50}
	% 
	% \subsection{CMap}
	% - Cmap \cite[]{subramanian_next_2017,lamb_connectivity_2006} contains 
	% pharma-genomic data (similar to CCLE and GDSC). 
	% 
	% - Data available at: 
	% https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE92742. 
	% 
	% - Cmap website: https://clue.io/cmap.

	\section{Discussion}\label{sec:discussion}
	The preceding presents a novel model for drug sensitivity
	prediction from a set of high dimensional molecular features. The model allows
	for the inclusion of discrete and continuous external covariates on both the 
	drugs and features. Inclusion of the external information is through 
	data-driven and adaptive empirical Bayes estimation of the hyperparameters in 
	the normal inverse Gaussian prior model (\ref{eq:prior}). Estimation is 
	efficient and scales well with the number of features and samples.

	The model is put into practice on the GDSC data. The results in Section
	\ref{sec:gdsc} show that the inclusion of the external covariates is 
	beneficial for drug sensitivity prediction. Although the prediction MSE is
	not lower if the external covariates are included, 
	the resulting hyperparameter estimates are still informative
	and may guide future research efforts.

	The comparison to the bSEM model that allows for the inclusion of 
	one binary external feature covariate favours the proposed NIG model. 
	Prediction MSEs are almost uniformly lower over the drugs for the NIG model 
	compared to bSEM. Consequently, the average prediction MSE is significantly
	lower for the NIG model.

	Although the empirical Bayes estimates of the external covariate effects 
	$\bm{\alpha}$ are informative, they are small compared to 
	bSEM's estimates. Consequently, the posteriors of the models with and without
	the external covariates (NIG2 and NIG1 in Section \ref{sec:gdsc}) do not 
	differ by much. This results similar prediction MSEs with
	the inclusion of the external covariates. This is a topic that deserves 
	further attention. One proposed solution is to include an extra 
	prior variance component that promotes a more heavy-tailed $\beta_{jd}$ prior
	distribution. An example is a variance component with a half-cauchy prior 
	distribution, i.e., the horseshoe \cite[]{carvalho_handling_2009}. 
	This solution is similar to the pInc model from 
	\cite{kpogbezan_incorporating_2019}. 
	Like the bSEM model, the pInc model gives significantly larger external
	covariate effects, 
	so we conjecture that the inclusion of a similar heavy-tailed prior variance 
	component will have a similar effect in the NIG model.

	\section*{Supplementary Material}
	Supplementary Material is available online from 
	\url{https://github.com/magnusmunch/cambridge}.

	\section*{Conflict of interest}
	The authors have declared no conflict of interest.

	\bibliography{refs}

\end{document}

### 2019-11-19 docs/manuscript2.Rnw
redoing simulations

--- 2019-08-13 docs/manuscript2.Rnw
@@
                echo=FALSE, cache=FALSE, message=FALSE, fig.pos='!ht')
+knit_hooks$set(document=function(x) {
+  sub('\\usepackage[]{color}', '\\usepackage{xcolor}', x, fixed=TRUE)})
 @
@@
 \documentclass[a4paper,hidelinks]{article}
-\usepackage{url,bm,algorithm,algpseudocode,bbm,amsmath,amssymb,amsthm,
-amsfonts,graphics,graphicx,epsfig,rotating,caption,subcaption,natbib,
-titlesec,multicol,verbatim,threeparttable,booktabs,dsfont,parskip,
-xr-hyper,hyperref,enumerate}
-\externaldocument[sm-]{supplement}
+\usepackage{bm,amsmath,amssymb,amsthm,amsfonts,graphics,graphicx,epsfig,
+rotating,caption,subcaption,natbib,appendix,titlesec,multicol,hyperref,verbatim,
+bbm,algorithm,algpseudocode,pgfplotstable,threeparttable,booktabs,mathtools,
+dsfont,parskip,xr}
+\externaldocument[sm-]{supplement2}
 
@@
 % settings
-\graphicspath{{/Users/magnusmunch/Documents/OneDrive/PhD/project_cambridge/graphs/}}
+\graphicspath{{../figs/}}
+\pgfplotsset{compat=1.16}
 \setlength{\evensidemargin}{.5cm} \setlength{\oddsidemargin}{.5cm}
 \setlength{\textwidth}{15cm}
-\title{Drug efficacy prediction in cell lines}
+\title{Empirical Bayes estimation of the normal inverse Gaussian prior in 
+simultaneous-equation models}
 \date{\today}
-\author{Magnus M.
-M\"unch$^{1,2}$\footnote{Correspondence to: \href{mailto:m.munch@vumc.nl}{m.munch@vumc.nl}}, Mark A.
-van de Wiel$^{1,3}$, Sylvia Richardson$^{3}$, \\ and Gwena{\"e}l G.
-R.
-Leday$^{3}$}
+\author{Magnus M. M\"unch$^{1,2}$\footnote{Correspondence to: 
+\href{mailto:m.munch@vumc.nl}{m.munch@vumc.nl}}, Mark A. van de Wiel$^{1,3}$, 
+Sylvia Richardson$^{3}$, \\ and Gwena{\"e}l G. R. Leday$^{3}$}
 
@@
 	\noindent
-	1. Department of Epidemiology \& Biostatistics, Amsterdam Public Health research institute, VU University Medical Center, PO Box 7057, 1007 MB
-	Amsterdam, The Netherlands\\
-	2. Mathematical Institute, Leiden University, Leiden, The Netherlands \\
-	3. MRC Biostatistics Unit, Cambridge Institute of Public Health, Cambridge, United Kingdom \\
+	1. Department of Epidemiology \& Biostatistics, Amsterdam Public Health
+	research institute, Amsterdam University medical centers, PO Box 7057, 1007 MB
+	Amsterdam, The Netherlands \\*
+	2. Mathematical Institute, Leiden University, Leiden, The Netherlands \\*
+	3. MRC Biostatistics Unit, Cambridge Institute of Public Health, Cambridge, 
+	United Kingdom
 
@@
 
-	\noindent\textbf{Keywords}:
-
-	\noindent\textbf{Software available from}:
+	\noindent\textbf{Keywords}: ...
 
 	\section{Introduction}
-	- motivation: drug response prediction from different omics data types is 
-	promising both for precision medicine and drug target discovery and relates to
-	actual tumor response \cite[]{iorio_landscape_2016}.
+	Recently, the dimensions of data sets have increased in three directions:
+	(i) samples, (ii) variables, and (iii) outcomes. Here, we will 
+	concern ourselves with directions (ii) and (iii). More spefically, we will 
+	focus on the regression setting
+	where we have multiple outcomes and a medium to high dimensional feature 
+	space, i.e., $p \gtrsim n$. Standard methods
+	underperform or even break down in such medium to high dimensional settings, 
+	so new methodology is required.
 
-	- problem: prediction of drug efficacy in cell lines is difficult (DREAM 7 
-	challenge in \cite{costello_community_2014}).
+	As a motivating example, we present an expression quantitative loci (eQTLs)
+	detection problem (see Section \ref{sec:application}). In eQTLs detection, 
+	the aim is to find a set of single nucleotide loci (SNPs) that explains a
+	(part of a) gene's expression. The resulting heritability of the gene 
+	expression traits is thought to be indicative of the relationship between 
+	genotype and phenotype. eQTLs detection is often done for a set of genes 
+	simultaneously and as such comprises a simultaneous-equations model (SEM) with
+	multiple outcomes and a high dimensional feature set (the SNPs). Apart from
+	eQTLs, SEM models are also applied in network reconstruction 
+	\cite[]{kpogbezan_incorporating_2019, leday_gene_2017} 
+	and drug target discovery \cite[]{noh_inferring_2016}, [! ref to second 
+	manuscript].
 
-	- example data: introduction (GDSC, CCLE, NCI-60)
+	In many multi-outcome regression settings, extra information on both the 
+	outcomes and 
+	features is available. In our motivating eQTLs detection example, extra 
+	information may come in the form of (i) gene annotation (e.g., length of the
+	gene), (ii) SNP annotation (e.g., whether the SNP is in the gene or not), 
+	(iii) SNP-gene annotation (e.g., distance of SNP to gene). The presented 
+	method allows for the incorporation of such information to aid in the 
+	estimation, by pooling information both across equations (genes) and 
+	features (SNPs).
 
-	- feature information: prior experts, pathways, database summaries, previous 
-	experiments, type of feature (somatic mutations, copy number alterations, DNA 
-	methylation, and gene expression).
+	We are not the first to propose to learn from external data 
+	across equations and features. \cite{kpogbezan_incorporating_2019} uses 
+	the presence or absence of edges in a normal tissue apoptosis pathway, to aid 
+	in the estimation of a cancer tissue apoptosis pathway. Furhtermore, in an 
+	eQTLs detection problem they include the location of the SNPs (inside
+	the gene or not) to aid in the estimation. \cite{oman_minimax_2002} uses 
+	empirical Bayes to learn from external data concerning both the outcomes and 
+	the features in a low dimensional chemometrics dataset. Our proposed method,
+	termed \texttt{semnig},
+	extends \cite{kpogbezan_incorporating_2019,kpogbezan_empirical_2017} by 
+	allowing for the inclusion of continuous external information and several 
+	sources of information.
 
-	- drugs information: molecular target, drug class, pubchem, 
-	omnipath DB, pharmaco DB
+	The rest of the paper is structured as follows. In Section 
+	\ref{sec:model} we introduce our model, the estimation of which is detailed
+	in Section \ref{sec:estimation}. In Sections \ref{sec:simulations} and
+	\ref{sec:application} we demonstrate the method in simulations and an 
+  eQTLs detection problem, respectively. We end with a discussion in Section 
+  \ref{sec:discussion}.
 
-	- drugs x feature information
+	\section{Model}\label{sec:model}
+	\subsection{Simultaneous equations model}\label{sec:SEM}
+	We have continuous measurements $y_{id}$ for observation $i=1,\dots, n$ on 
+	outcome $d=1,\dots,D$. Throughout this paper we assume the $y_{id}$ to be 
+	centred per outcome and let $\y_d = \begin{bmatrix} y_{1d} & \dots & y_{nd} 
+	\end{bmatrix} \tr$. We predict the $y_{id}$ with features $x_{ij}$, 
+	$j=1,\dots, p$, collected in $\x_i = \begin{bmatrix} x_{i1} & \dots & x_{ip} 
+	\end{bmatrix} \tr$ and centered per feature. We have $|\bm{\alpha}|$ 
+	external variables available that describes the outcome $d$, feature $j$, or
+	relationship between feature $j$ and outcome $d$, collected in 
+  $\mathbf{C} = \begin{bmatrix} \mathbf{c}_{1}^1 & \cdots &
+	\mathbf{c}_{1}^p & \cdots & \mathbf{c}_{D}^1 & \cdots & \mathbf{c}_{D}^p
+  \end{bmatrix}_{|\bm{\alpha}| \times pD} \tr$. An example of such an external
+  variable from eQTL detection is the distance between SNP and gene.
 
-	- current literature: reviews in \cite{azuaje_computational_2017} and
-	\cite{ali_machine_2019}, 
-	full multivariate approach by Oslo people in \cite{mai_composite_2019}, 
-	TANDEM \cite[]{aben_tandem:_2016}, DeltaNet 
-	by \cite{noh_inferring_2016}, patient-derived tumor cell lines 
-	in \cite{gao_high-throughput_2015}, 
-	multiple kernel learning approach in \cite{costello_community_2014},
-	\cite{ammad-ud-din_drug_2016} combine a 
-	multiple kernel learning approach with matrix factorization, using the model 
-	to uncover the latent 
-	relationships between drug targets and intracellular pathways, and
-	structured penalized regression in \cite{zhao_structured_2019}
-
-	- our solution: adaptive borrowing of information through empirical Bayes, 
-	guided by external feature and drug information.
-
-	\section{Methods}
-	\subsection{Simultaneous equations model}\label{sec:SEM}
-	Assume we have continuous efficacy measures $y_{id}$ for cell lines 
-	$i=1,\dots, n$, centred per drug $d$. We predict efficacy from molecular 
-	features 
-	$x_{ij}$, $j=1,\dots, p$, collected in $\x_i = \begin{bmatrix} x_{i1} & 
-	\dots & x_{ip} \end{bmatrix} \tr$. We have an \textit{a priori} (possibly 
-	informative) partitioning 
-	of the omics features into $g=1, \dots, G$ non-overlapping groups.
-	For convenience, we let 
-	$g(j)$ denote the group index of molecular feature $j$. Furthermore, we have 
-	characteristics of the drugs available in the form of external variables 
-	$\mathbf{C} = \begin{bmatrix} \mathbf{c}_1 & \dots & \mathbf{c}_D 
-	\end{bmatrix} \tr$.
-
-	We regress the drug efficacies on the molecular feature effects $\beta_{jd}$:
-	\begin{subequations}
+	We model the outcomes $y_{id}$ as a linear function of the features $x_{ij}$:
+	\begin{subequations}\label{eq:linearmodel}
 		\begin{align}
-		  y_{id} & = \x_i \tr \bbeta_d + \epsilon_{id}, \label{eq:linearmodel}\\
+		y_{id} & = \beta_{0d} + \x_i \tr \bbeta_d + \epsilon_{id},\\
 		\text{with } & \epsilon_{id} \sim \mathcal{N}(0, \sigma_d^2),
@@
 	\end{subequations}
-	where $p$-dimensional $\bbeta_d$ are the drug-specific omics feature effects. 
-	Note that (\ref{eq:linearmodel}) gives rise to a system of $d=1, \dots, D$ 
-	linear simultaneous equations.
+	where the $p$-dimensional $\bbeta_d$ are the outcome-specific feature effects.
+	Note that (\ref{eq:linearmodel}) gives rise to a system of 
+	$D$ linear simultaneous equations.
 
 	\subsection{Bayesian prior model}
-	We capture the uncertainty in the parameters through a Bayesian prior model
-	and endow the parameters with the following prior:
-	\begin{subequations}\label{eq:prior}
+	We capture the uncertainty in the parameters through a Bayesian prior model. 
+	We assign a flat prior to the intercept terms $\beta_{0d}$ and integrate 
+	them out. The remaining parameters are endowed with priors such that we end
+	up with the following model:
+	\begin{subequations}\label{eq:priormodel}
     \begin{align}
-		  \beta_{jd} & \sim \mathcal{N}_p (0, \phi^2_{g(j)} \gamma_{jd}^2 
-		  \sigma_d^2), 
-		  \label{eq:betaprior}\\
-		  \gamma_{jd}^{2} & \sim \mathcal{IG}( 
-		  [(\mathbf{c}_d^j) \tr \bm{\alpha}]^{-1}, \lambda_d), \\
-		  \sigma_d^{2} & \sim 1/\sigma_d^{3},
+      y_{id} | \bm{\beta}_d, \sigma_d^2 & \sim \mathcal{N} 
+      \(\x_i \tr \bm{\beta}_d, \sigma_d^2\), \\
+      \beta_{jd} | \gamma_{jd}^2, \sigma_d^2 & \sim \mathcal{N} 
+      \(0, \sigma_d^2 \gamma_{jd}^2 \), \label{eq:betaprior} \\
+      \gamma_{jd}^2 & \sim \mathcal{IG}
+      \( [(\mathbf{c}_{d}^j) \tr \bm{\alpha}]^{-1}, \lambda_{d} \), \\
+      \sigma_d^2 & \sim 1/\sigma_d^3,
     \end{align}
   \end{subequations}
-	where $\mathcal{IG}(\theta, \lambda)$ denotes an inverse Gaussian distribution 
-	with mean $\theta$ and shape $\lambda$. The different molecular feature types
-	are given their own weights through the $\phi^2_{g}$, while the external 
-	outcome-feature data comes in through 
-	our mean model for the $\gamma_{jd}^2$ hyperprior model: 
-	$[(\mathbf{c}_d^j) \tr \bm{\alpha}]^{-1}$. Both the $\phi^2_{g}$ and
-	$\bm{\alpha}$ are estimated in a data-driven manner. More details on the model
-	itself, as well as its estimation, may be found in 
-	the SM and in [! ref to first paper here].
+  where $\mathcal{IG} ([(\mathbf{c}_{d}^j) \tr \bm{\alpha}]^{-1}, \lambda_{d})$ 
+  denotes the inverse Gaussian distribution with mean 
+  $[(\mathbf{c}_{d}^j) \tr \bm{\alpha}]^{-1}$ and shape $\lambda_d$.
 
-	\section{Data}
+  The normal inverse Gaussian (NIG) prior model for $\beta_{jd}$ and 
+  $\gamma^2_{jd}$ in (\ref{eq:priormodel}) was introduced in 
+  \cite{barndorff-nielsen_hyperbolic_1978} and is, since 
+  \cite{barndorff-nielsen_normal_1997}, routinely applied in mathematical 
+  finance (see, e.g., \cite{kalemanova_normal_2007}). Supplementary Material
+  (SM) Section \ref{sm-sec:nigprior} contains more details on this prior.
+  In Figure \ref{fig:dens_igaussian_marginalbeta}, the NIG 
+  prior distribution, marginalised over $\gamma_{jd}^2$, is depicted together 
+  with three other common choices,
+  the Student's $t$ (see Section \ref{sec:inversegammamodel} for more details on 
+  this prior), normal (ridge) and Laplace (lasso) priors, 
+  all scaled to variance one. 
+  From this Figure we see that, depending on the choice
+  of hyperparameters, the NIG prior can vary from heavier tails than the lasso, 
+  to more Gaussian tails like the ridge, but behaves similarly to the 
+  Student's $t$ prior. Our argumentation to model the 
+  $\gamma^2_{jd}$ by an inverse Gaussian 
+  distribution, as has been suggested in \cite{fabrizi_specification_2016} and 
+  \cite{caron_sparse_2008}, is two-fold: (i) the NIG model is more flexible than
+  the standard ridge and lasso priors (as seen from Figure 
+  \ref{fig:dens_igaussian_marginalbeta}), and (ii) the NIG prior allows to model
+  the mean as a 
+  function of the external variables $\mathbf{c}_{d}^j$ more conveniently than 
+  the Student's $t$ prior, as explained in Section 
+  \ref{sec:empiricalbayes}. 
+<<dens_igaussian_marginalbeta, fig.cap="Several choices of marginal $\\beta_j$ prior, scaled to $\\V(\\beta_j)=1$.", out.width="60%", fig.asp=3/4>>=
+@
 
-	\subsection{GDSC}
-	The GDSC project's \cite[]{yang_genomics_2013} aim is "to improve cancer 
-	treatments by discovering therapeutic biomarkers that can be used to identify 
-	patients most likely to respond to anticancer drug". Part of the project is to
-	screen $>1000$ human cancer cell lines with drugs. The cell lines have been
-	genetically characterised and several drug sensitivity measures are recorded.
-	The data is freely available from \cite{garnett_systematic_2012} and consist 
-	of: (i) the sensitivity measures 
-	(mainly half maximal inhibitory concentration, i.e., IC50) of the cell lines 
-	to the drugs, (ii) annotation of the 
-	screened compounds, and (iii) the cell lines' genomic profile (mutations, 
-	copy numbers, methylation profiles, and expression). We will attempt to 
-	predict drug sensitivity of the cell lines, as quantified by IC50, using the
-	genomic data. Other choices of efficacy measures than IC50 are possible, 
-	but a discussion on the pros and cons of these different choices is not the 
-	aim of this paper.
+  \cite{moran_variance_2018} argue that in regression models, $\bm{\beta}_d$ 
+  and $\sigma_d^2$ should be independent \textit{a priori}, i.e.
+  $\beta_{jd} | \gamma_{jd}^2, \sigma_d^2 \sim \beta_{jd} | 
+  \gamma_{jd}^2$. However, in our case
+  that is debatable. The $D$ regressions need not be on the same scale, so in
+  order to jointly consider the $\bm{\beta}_d$ some 
+  form of calibration is required. A straightforward way 
+  of doing this is to scale the $\bm{\beta}_d$ prior with 
+  $\sigma_d^2$ as in (\ref{eq:betaprior}), resulting in a 
+  scale-free prior with respect to the response. Apart from the scaling,
+  in this setting, the conjugate prior model is preferred
+  over the non-conjugate model, because the variational Bayes approximation 
+  (see Section \ref{sec:estimation}) assumes independent $\sigma_d^2$ and 
+  $\bm{\beta}_d$ in the posterior, which is actually true for 
+  the conjugate model. Therefore, the approximate expectation used in the 
+  empirical Bayes step (see Section \ref{sec:estimation}) is more accurate.
 
-	\subsubsection{Primary data}
-	We took the logarithm of the IC50 values and took the average if there were
-	multiple measurements per cell line-drug combination. In the following, if we 
-	refer to IC50, it denotes the log-transformed values. After iteratively 
-	removing the cell line or drug with the highest proportion 
-	of missings until there are no more missing values, we end up with IC50 
-	estimates for 482 cell lines on 168 drugs. For the gene expression 
-	data, we fit a lasso with IC50 as response to the drugs separately and 
-	adjusted the penalty parameter to select approximately 300 genes per drug.
+  A few remarks on the choice of error variance prior are justified here: 
+  many authors endow error variance components with vague gamma priors. 
+  \cite{gelman_prior_2006}, 
+  among others, advises against this practice. The degree of `vagueness' has a 
+  large influence on the posterior, while degree of `vagueness' is a difficult 
+  parameter to set. This influence is especially pronounced if the likelihood is
+  relatively flat, as may be reasonably expected in the large $p$, small $n$ 
+  setting we are in. We therefore model the error variance with Jeffrey's 
+  objective prior 
+  \cite[]{jeffreys_invariant_1946}. In the derivation of our Jeffrey's prior
+  for the error variance, we jointly consider an unknown mean and variance. This
+  joint consideration results in the somewhat unorthodox $1/\sigma^3$ Jeffrey's
+  prior.
 
-	- From Oslo:
-	1.2) Preselecting GEX features: we selected gene expression features
-	explaining 50\% variations over all cell lines, since this results in not too
-	features and probably remains those important features.
-	1.3) Preselecting CNV: we used the same strategy as Garnett et al. (2012)
-	where the GDSC data are from.
-	1.4) Preselecting MUT: we deleted mutated genes with less than two cell lines.
+  Several extensions of the model are described in [! ref to second paper here].
 
-	- check out selection method in \cite{zhao_structured_2019}.
+	\section{Estimation}\label{sec:estimation}
+	\subsection{Variational Bayes}\label{sec:variationalbayes}
+	The posterior corresponding to the model described in (\ref{eq:linearmodel})
+	and (\ref{eq:priormodel}) is not available in closed form. We therefore 
+	approximate the posterior by variational Bayes, where we force the approximate
+	posterior 
+	density to factorise as: $p(\bbeta_d, \sigma_d^2, \gamma_{1d}^2, \dots, 
+	\gamma_{pd}^2) \approx 
+	Q_d(\cdot) = q(\bbeta_d) \cdot q(\sigma_d^2) \cdot \prod_{j=1}^p 
+	q(\gamma_{jd}^2)$. For
+	notational convenience, we slightly abuse notation and let $q(\cdot)$ denote
+	different densities for different inputs.
+	Under such a factorisation, the marginal variational posteriors that 
+	minimise the Kullback-Leibler divergence of the true posterior to the 
+	variational Bayes approximation \cite[]{neal_view_1998} are given by:
+	\begin{align*}
+    q(\bm{\beta}_d) & \overset{D}{=} \mathcal{N}_p 
+    (\bm{\mu}_d, \bm{\Sigma}_d), \\
+    q(\gamma_{jd}^2) & \overset{D}{=} \mathcal{GIG}
+    \(-1, \lambda_{d}[(\mathbf{c}_{d}^j) \tr \bm{\alpha}]^2, \delta_{jd}\), \\
+    q(\sigma^2_d) & \overset{D}{=} \Gamma^{-1} \(\frac{n + p + 1}{2}, \zeta_d\).
+  \end{align*}
+	See SM Section \ref{sm-sec:vbderivations} for the derivations. The variational
+	parameters $\bm{\mu}_d$, $\bm{\Sigma}_d$, $\delta_d$, and
+	$\zeta_d$ contain cyclic dependencies and are iteratively 
+	updated by:
+  \begin{align*}
+    \bm{\Sigma}_d^{(h+1)} & = (a_d^{(h)})^{-1} 
+    \[\X \tr \X + \diag(b_{jd}^{(h)})\]^{-1}, \\
+    \bm{\mu}_d^{(h+1)} & = \[\X \tr \X + \diag(b_{jd}^{(h)})\]^{-1} 
+    \X \tr \y_d, \\
+    \delta_{jd}^{(h+1)} & = a_d^{(h)} \[(\bmu^{(h+1)}_{jd})^2 + 
+    (\bSigma_d)_{jj}\] + \lambda_{d}, \\ 
+    \zeta_d^{(h+1)} & = \frac{1}{2} \bigg[\mathbf{y}_d \tr \mathbf{y}_d -
+    2 \mathbf{y}_d \tr \X \bm{\mu}_d^{(h+1)} + 
+    \trace ( \X \tr \X \bm{\Sigma}_d^{(h+1)}) + 
+    (\bm{\mu}_d^{(h+1)}) \tr \X \tr \X \bm{\mu}_d^{(h+1)} \\
+    & \,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\, + 
+    \trace \[ \diag (b_{jd}^{(h+1)}) \bm{\Sigma}_d^{(h+1)}\] + 
+    (\bm{\mu}_d^{(h+1)}) \tr \diag (b_{jd}^{(h+1)}) 
+    \bm{\mu}_d^{(h+1)}\bigg],
+  \end{align*}
+  until convergence. Here, we set
+  \begin{align}
+    a_d^{(h)} & =\E_{Q^{(h)}}(\sigma_d^{-2})=(n + p + 1)/(2 \zeta_d^{(h)}), 
+    \nonumber\\
+    b_{jd}^{(h)} & = \E_{Q^{(h)}}(\gamma_{jd}^{-2}) = 
+    \sqrt{\frac{\lambda_{d}[(\mathbf{c}_{d}^j) \tr \bm{\alpha}]^2}
+    {\delta_{jd}^{(h)}}}
+    \frac{K_0(\sqrt{\delta_{jd}^{(h)}\lambda_{d}}
+    [(\mathbf{c}_{d}^j) \tr \bm{\alpha}])}
+    {K_1(\sqrt{\delta_{jd}^{(h)}\lambda_{d}}
+    [(\mathbf{c}_{d}^j) \tr \bm{\alpha}])} + 
+    \frac{2}{\delta_{jd}^{(h)}}, \label{eq:ratiomodifiedbessel}
+  \end{align}
+  where $K_{\nu}(x)$ denotes the modified Bessel function of the second kind.
+  A method for fast and numerically stable calculation of ratios of
+  modified Bessel functions of the second kind, as in 
+  (\ref{eq:ratiomodifiedbessel}), is given in SM Section 
+  \ref{sm-sec:ratiosmodifiedbessels}.
 
-	- how to transform the data and deal with missingness? Read GDSC papers.
+  \subsection{Empirical Bayes}\label{sec:empiricalbayes}
+  To make use of the external data, we parametrise the prior mean of the
+  $\gamma_{jd}^2$ as $[(\mathbf{c}_{d}^j) \tr \bm{\alpha}]^{-1}$. 
+  This allows us to
+  include continuous external variables $\mathbf{c}_{d}^j$ into the model. 
+  Additionally, it reduces the number of hyperparameters from $pD$ to 
+  $|\bm{\alpha}| + D$. A full Bayesian model
+  then requires the specification of the hyperparameters 
+  $\bm{\alpha}$ and $\blambda = \begin{bmatrix} \lambda_1 & \cdots & \lambda_D 
+  \end{bmatrix} \tr$. These are abstract and hard to interpret parameters 
+  for which we generally lack expert knowledge. They do, however, have a 
+  significant influence on the shape of the posterior distribution. We 
+  therefore propose to estimate these hyperparameters by empirical Bayes. 
+  Simply put, empirical Bayes fits the prior model to the data and is, as such, 
+  objective Bayes (up to model specification). In our case, this results in an   
+  objective and data-driven inclusion of the external data.
 
-	\subsubsection{External data}
-	Several external drug variables are available: (i) the developmental stage
-	(experimental, in clinical development, or clinically approved) and (ii)
-	the action (cytotoxic or targeted). We also have a external variable 
-	available that relates the drugs to the features: a binary indicator whether 
-	a gene is in the target pathway of the drug or not.
+	The canonical method for empirical Bayes is maximisation of the marginal 
+	likelihood with respect to the hyper parameters. In 
+	\cite{casella_empirical_2001} the marginal likelihood is maximised by an EM 
+	algorithm:
+	\begin{align*}
+	  \bm{\alpha}^{(l+1)},\bm{\lambda}^{(l+1)} & = \underset{\bm{\alpha},
+	  \bm{\lambda}}{\argmax}\E_{\cdot | \Y} 
+	  [\log p(\Y, \B, \bm{\Gamma}^2, \bsigma^2) | \bm{\alpha}^{(l)},
+	  \bm{\lambda}^{(l)}] \\
+	  & = \underset{\bm{\alpha},\bm{\lambda}}{\argmax} 
+	  \E_{\cdot | \Y} [\log \pi (\bm{\Gamma}^2) | \bm{\alpha}^{(l)},
+	  \bm{\lambda}^{(l)}],
+	\end{align*}
+	where $\Y = \begin{bmatrix} \y_1 & \cdots & \y_D \end{bmatrix}$,
+	$\B = \begin{bmatrix} \bbeta_1 & \cdots & \bbeta_D \end{bmatrix}$, 
+	$\bsigma^2 = \begin{bmatrix} \sigma^2_1 & \cdots & 
+	\sigma^2_D \end{bmatrix} \tr$,
+	and $\bm{\Gamma}^2 = \begin{bmatrix} \bgamma^2_1 & \cdots & \bgamma^2_D 
+	\end{bmatrix}\tr$, with 
+	$\bgamma^2_d = \begin{bmatrix} \gamma^2_{1d} & \cdots & \gamma^2_{pd} 
+	\end{bmatrix}\tr$,
+	and the expectation is with respect to the joint posterior. In our case, this 
+	posterior is not available in closed form, which renders the expectation 
+	difficult. While \cite{casella_empirical_2001} suggests to approximate the 
+	expectation by a Monte Carlo sample, we propose to use the variational Bayes 
+	approximation developed in Section \ref{sec:variationalbayes}:
+	$$
+	\bm{\alpha}^{(l+1)},\bm{\lambda}^{(l+1)} = 
+	\underset{\bm{\alpha},\bm{\lambda}}{\argmax}\E_{Q^{(l)}} 
+	[\log \pi(\bm{\Gamma}^2)],
+	$$
+	where now the expectation is with respect to the converged variational 
+	posterior $Q^{(l)}=\prod_{d=1}^D Q_d^{(l)}$. The resulting estimating 
+	equations are:
+  \begin{align*}
+    \bm{\alpha} & = \[ \mathbf{C} \tr \bLambda \diag 
+    (e_{jd}^{(l)}) \mathbf{C} \]^{-1} \mathbf{C} \tr \bLambda
+    \mathbf{1}_{pD \times 1}, \\
+    \lambda_{d} & = p \left\{ \sum_{j=1}^p b_{jd}^{(l)} + 
+    \sum_{j=1}^p e_{jd}^{(l)} [(\mathbf{c}_{d}^j) \tr \bm{\alpha}]^2 - 
+    2 \sum_{j=1}^p (\mathbf{c}_{d}^j) \tr \bm{\alpha} \right\}^{-1},
+  \end{align*}
+  where $\bLambda=\diag(\lambda_d) \otimes \I_p$
+  and $e_{jd}^{(l)} = \E_{Q^{(l)}}(\gamma_{jd}^2) = (b_{jd}^{(l)} - 
+  2/\delta_{jd}^{(l)}) \cdot \delta_{jd}^{(l)}/[\lambda_{d}^{(l)}
+  \left\{(\mathbf{c}_{d}^j) \tr \bm{\alpha}^{(l)}]^2\right\}$. 
+  Solving the equations is done by 
+  iteratively reweighted least squares.
 
-  - check whether compounds are PAINS \cite[]{baell_new_2010} or not. 
-  PAINS indicates whether compound is frequent false hitter.
+  Variaional Bayes approximations are known to underestimate posterior 
+  variances \cite[]{rue_approximate_2009,consonni_mean-field_2007,
+  bishop_pattern_2006,wang_inadequacy_2005}. If posterior variances are 
+  required, 
+  we suggest generating samples from the posterior after the empirical Bayes 
+  iterations have converged. A Gibbs sampler to do so is described in 
+  the SM Section \ref{sm-sec:gibbssampler}. Alternatively, the \texttt{R} 
+  package \texttt{rstan} \cite[]{guo_rstan:_2018} allows the user to sample
+  from a generic posterior. An implementation of model
+  (\ref{eq:priormodel}) in \texttt{stan} code may be found on 
+  \url{https://github.com/magnusmunch/cambridge}.
 
-  - clinical trials database CTRP might contain info
+  Note that so far, the design matrix $\X$ is assumed to be shared across the
+  equations. This is not a requirement. That is, we may have D different design 
+  matrices $\X^d$ that need not have the same number of 
+  columns, i.e., we have $p_d$ instead of $p$. This becomes useful in, e.g. 
+  eQTLs detection, where we regress genes on a gene-specific selection
+  of SNPs, often based on SNP-gene distance.
 
-  - extra data on GDSC might be in PharamcoGx R-package. Also contains bunch 
-  of functions.
+  \section{Simulations}\label{sec:simulations}
+  We conducted a simulation study consisting of three parts investigating (i) 
+  the NIG prior performance, (ii) the variational Bayes approximation, and (iii)
+  the empirical Bayes estimation.
 
-  - check GSK open target EBI or ask Chris Wallace about it. Database that links
-	targets to drugs and diseases. Might be useful for drugs co-data.
+  \subsection{NIG prior performance}
+  \label{sec:inversegammamodel}
+  The aim of this part is to investigate the 
+  performance of the NIG prior model. Since we may do so in the univariate 
+  setting, so we drop all $d$ indices in the following. 
+  A common default prior for latent variance components is the inverse Gamma 
+  prior:
+  $$
+  \gamma_j^2 \sim \Gamma^{-1}(\eta/2,\xi/2),
+  $$
+  with shape $\eta/2$ and scale $\xi/2$, both rescaled to match our 
+  inverse Gaussian prior formulation. The resulting marginal prior for the model
+  parameters
+  (conditional on the error variance $\sigma^2$) is then a scaled Student's 
+  $t$-distribution:
+  $$
+  \beta_j | \sigma^2 \sim \text{Student's } 
+  t\(\eta,\sigma \sqrt{\xi/\eta}\),
+  $$
+  with degrees of freedom $\eta$ and scale $\sigma \sqrt{\xi/\eta}$.
+  See SM Section \ref{sm-sec:inversegamma} for more details on this prior. 
+  We compare the 
+  posteriors of the proposed NIG to the scaled Student's 
+  $t$-distribution prior in a simulated setting.
 
-	- try non-negative matrix factorisation/independent component analysis on 
-	binary fingerprints to get principal components that characterise the 
-	different drugs.
+  We compare the posterior distributions to the true posterior for a 
+  number of measures. Divergence between the distributions
+  is quantified through a multivariate extension of the Cram{\'e}r-von Mises
+  criterion \cite[]{baringhaus_new_2004}, where smaller values of the criterion
+  indicate a better fit. We also assess 95\% coverage of the 
+  posterior credible intervals. The accuracy of the 
+  point estimates of $\beta_j$ in the form of posterior means is assessed by
+  the mean squared error (MSE) and correlation between the point estimates 
+  and the true $\beta_j$.
 
-	\subsection{CCLE}
-	- extra data on CCLE might be in PharamcoGx R-package
+  We studied two settings: one dense and one sparse. In the dense setting, 
+  we simulated $n=100$ features $\x_i$ from a $p=200$ dimensional Gaussian 
+  distribution with zero mean vector, variances equal to one, and all 
+  covariances set to $\rho=0.5$. The outcomes were generated through $y_i = 
+  \x_i\tr \bbeta + \epsilon_i$, with the $\epsilon_i$ from the standard normal 
+  distribution. 
+  Likewise, we generated the model parameters $\beta_j$ from a standard normal
+  distribution.
 
-	\subsection{NCI-50}
+  We choose the hyperparameters such that we get all combinations of marginal
+  prior variances and kurtoses $\V(\beta_{j} | \sigma^2) \in 
+  \{ 1/2, 1, 2 \}$ and $\mathcal{K}(\beta_{j} | \sigma^2) \in 
+  \{ 4, 10 \}$. This leads to the settings in Table \ref{tab:hyper_set1}. For 
+  all priors, we generated 1000 draws (after 1000 burnin) from the 
+  posteriors using an MCMC sampler implemented in \texttt{rstan}. We repeated
+  the simulation 100 times and report 
+  results in Figures \ref{fig:boxplots_igaussian_res4.1_post1} and 
+  \ref{fig:boxplots_igaussian_res4.1_post2}. 
+<<hyper_set1>>= 
+set <- read.table("results/simulations_igaussian_set4.csv")
+kableExtra::add_header_above(kableExtra::kable_styling(
+  knitr::kable(set[, c("ctalphainv", "lambda", "eta", "xi", "vars", "kurts")],
+               col.names=c("$(\\mathbf{c} \\tr \\bm{\\alpha})^{-1}$",
+                           "$\\lambda$", "$\\eta$", "$\\xi$",
+                           "$\\V(\\beta_{j} | \\sigma^2)$",
+                           "$\\mathcal{K}(\\beta_{j} | \\sigma^2)$"),
+               caption="Hyperparameter settings",
+               "latex", booktabs=TRUE, escape=FALSE, digits=2),
+  latex_options=c("HOLD_position")),
+  header=c(" "=1, "NIG"=2, "Student's t"=2, "Prior"=2))
+@
 
-	\subsection{CMap}
-	- Cmap \cite[]{subramanian_next_2017,lamb_connectivity_2006} contains 
-	pharma-genomic data (similar to CCLE and GDSC).
+<<boxplots_igaussian_res4.1_post1, fig.cap="Comparison of the NIG and Student's t priors in the dense setting in (a)-(d) simulation setting 1, (e)-(h) setting 2, and (i)-(l) setting 3, through Cramér-von Mises criterion between the model and true posteriors, correlation and MSD between the posterior means and true parameters, and 95\\% credible interval coverage of the true values.", out.width="100%", fig.asp=1>>=
+@
 
-	- Data available at: 
-	https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE92742.
+<<boxplots_igaussian_res4.1_post2, fig.cap="Comparison of the NIG and Student's t priors in the dense setting in (a)-(d) simulation setting 4, (e)-(h) setting 5, and (i)-(l) setting 6, through Cramér-von Mises criterion between the model and true posteriors, correlation and MSD between the posterior means and true parameters, and 95\\% credible interval coverage of the true values.", out.width="100%", fig.asp=1>>=
+@
 
-	- Cmap website: https://clue.io/cmap.
+  [! describe results here]
 
-	\section{Results}
-	\subsection{GDSC}
+  In the sparse setting, the observed data generating process is the same as in
+  the dense setting. The
+  model parameters are simulated from an (absolutely continuous) spike-and-slab 
+  prior with both Gaussian spike and slab:
+  \begin{align*}
+    w & \sim \text{Beta}(\phi, \chi), \\
+    z_j | w & \sim \text{Bernoulli}(w), \\
+    \beta_j | z_j & \sim z_j \mathcal{N} \( 0, 1 + 0.99 \frac{\chi}{\phi}\) +
+    (1 - z_j)\mathcal{N} \( 0, 0.01\).
+  \end{align*}
+  The slab is chosen
+  such that the we have marginal prior variance $\V(\beta_j)=1$. The sparsity 
+  level is determined by the proportion of zero $\beta_j$: 
+  $s=1 - \phi/(\phi + \chi)$. Note that the resulting marginal prior for
+  $\beta_j$ is:
+  $$
+  \beta_j \sim (1 - s)\mathcal{N} \( 0, \frac{1 - 0.01 s}{1 - s}\) + 
+  s \mathcal{N} \( 0, 0.01\).
+  $$
+  We use two settings, $s=0.5$ and $s=0.9$. The 
+  hyperparameters for estimation are set the same as in the dense setting, 
+  resulting in a total of 12 different simulation settings 
+  (6 per sparsity setting).
+  To mitigate the effect of random variation, we
+  repeated the simulations 100 times and report the results in Figures 
+  \ref{fig:boxplots_igaussian_res5_post1} and 
+  \ref{fig:boxplots_igaussian_res5_post2}.
 
-	\section{Discussion}
-	- use copula's and marginal posteriors to find multivariate posteriors for 
-	contrast testing.
+<<boxplots_igaussian_res5_post1, fig.cap="Comparison of the NIG and Student's t priors in the sparse setting in (a)-(d) simulation setting 1, (e)-(h) setting 2, and (i)-(l) setting 3, through Cramér-von Mises criterion between the model and true posteriors, correlation and MSD between the posterior means and true parameters, and 95\\% credible interval coverage of the true values.", out.width="100%", fig.asp=1>>=
+@
 
-	- empirical Bayes gives extra information via estimation of the 
-	hyperparameters, such as informativeness of the drug information (via model 
-	for the $\gamma_d$) and relatedness of the tissues (via $\bm{\Omega}$ and 
-	$\nu$)
+<<boxplots_igaussian_res5_post2, fig.cap="Comparison of the NIG and Student's t priors in the sparse setting in (a)-(d) simulation setting 4, (e)-(h) setting 5, and (i)-(l) setting 6, through Cramér-von Mises criterion between the model and true posteriors, correlation and MSD between the posterior means and true parameters, and 95\\% credible interval coverage of the true values.", out.width="100%", fig.asp=1>>=
+@
 
-	- regression on drug covariates very intuitive.
+  [! describe results here]
+
+  \subsection{Variational Bayes approximation}
+  To investigate the accuracy of the variational Bayes approximation we compare 
+  the variational Bayes posterior to 1000 (after 1000 burn-in) MCMC samples from
+  the posterior obtained with \texttt{rstan}. Since the MCMC
+  samples represent the exact posterior (up to Monte Carlo error), we use it as 
+  a proxy for the true posterior. We include the comparison of our variational
+  Bayes approximation to the approximate posterior from the
+  automatic differentiation variational inference (ADVI)
+  algorithm in \texttt{rstan} \cite[]{kucukelbir_automatic_2015}. ADVI is a
+  variational Bayes algorithm that assumes a full factorization of the posterior
+  and is fit through gradient based optimization of the variational objective.
+  \texttt{rstan} offers an ADVI algorithm that assumes a fullrank posterior
+  covariance, but this algorithm often failed to converge or resulted in errors, 
+  so we do not include it in the comparisons. We measure divergence between the 
+  MCMC and variational posteriors with the Cram{\'e}r-von Mises criterion.
+
+  In addition to ADVI, \texttt{rstan} allows the user
+  to maximise the posterior to obtain the maximum \textit{a posteriori} (MAP)
+  estimates of the model parameters. We compare these MAP estimates, and the 
+  ADVI posterior means,
+  to the MAP estimates obtained from our model through the correlation with
+  and the mean squared distance (MSD) from the MCMC posterior means.
+
+  The interest of this simulation lies in the variational Bayes approximation, 
+  so we restrict ourselves to the univariate setting again (i.e., $D=1$). The
+  simulation setup is the same as in the dense setting 
+  in Section \ref{sec:inversegammamodel}. Similarly, we use the same 
+  hyperparameters settings (Table \ref{tab:hyper_set1}. For \texttt{glmnet}
+  we either cross-validate the penalty parameter $\lambda$ or match it to the
+  true prior variance. We
+  repeated the simulations 100 times and report the results in Figures 
+  \ref{fig:boxplots_igaussian_res4.2_post1}-
+  \ref{fig:boxplots_igaussian_res4.2_post2}.
+
+  In addition to the performance measures, we have plotted posterior samples 
+  histograms of 8 randomly selected model parameters in setting 1, 
+  along with the esimated
+  marginal variational Bayes posteriors, and the MAP estimates in Figure 
+  \ref{fig:hist_igaussian_fit4}.
+
+<<boxplots_igaussian_res4.2_post1, fig.cap="Comparison of different approximations in (a)-(c) simulation setting 1, (d)-(f) setting 2, and (g)-(i) setting 3, through Cramér-von Mises criterion between the approximate and MCMC posteriors, correlation and MSD between the approximate and MCMC posterior means.", out.width="80%", fig.asp=1>>=
+@
+
+<<boxplots_igaussian_res4.2_post2, fig.cap="Comparison of different approximations in (a)-(c) simulation setting 4, (d)-(f) setting 5, and (g)-(i) setting 6, through Cramér-von Mises criterion between the approximate and MCMC posteriors, and correlation and MSD between the approximate and MCMC posterior means.", out.width="80%", fig.asp=1>>=
+@
+
+<<hist_igaussian_fit4, fig.cap="Histogram of posterior MCMC samples of 8 randomly selected model parameters in setting 1, along with the estimated variational posteriors and MAP estimates.", out.width="80%", fig.asp=1>>=
+@
+
+  [! describe results here]
+
+  \subsection{Empirical Bayes estimation}
+  To assess the empirical Bayes estimation of the inverse Gaussian prior, 
+  we conduct a small simulation study where we simulate from
+  model (\ref{eq:priormodel}), while keeping $\forall d: \sigma^2_d=1$ fixed. 
+  More specifically, we simulate 
+  $\gamma_{jd}^2 \sim \mathcal{IG}([(\mathbf{c}_d^j) \tr \bm{\alpha}]^{-1},
+  \lambda_d)$, for $j=1,\dots,200$, $d=1,\dots,10$, where we set 
+  $\bm{\alpha}= \begin{bmatrix} 1 & 1 & 2 & 3 & 4 \end{bmatrix}$. We use 
+  balanced dummy coding for the $\mathbf{c}_d^j$, such that we end up with
+  $\E(\gamma_{jd}) \in \{ 1, 0.5, 0.33, 0.25, 0.2 \}$, and fix 
+  $\forall d: \lambda_d=1$.  
+  Next, we simulate $\beta_{jd} \sim \mathcal{N} 
+  (0, \gamma_{jd}^2)$.
+  The observed data is generated from $x_{ij} \sim \mathcal{N} (0, 1)$ and 
+  $y_{id} \sim \mathcal{N} (\mathbf{x}_i \tr \bm{\beta}_d, 1)$, for $i=100$. 
+  We repeat the simulation 100 times and, for each simulation, estimate the 
+  hyperparameters by the method described in Section \ref{sec:empiricalbayes} 
+  and present the results in Figure \ref{fig:boxplots_igaussian_res6_eb}.
+
+<<boxplots_igaussian_res6_eb, fig.cap="Empirical Bayes $\\hat{\\bm{\\alpha}}$ estimates", out.width="80%", fig.asp=2/3>>=
+@
+
+  [! describe results here]
+
+  - seems to be an overall scaling issue, relative to each other the estimates
+  are not so bad.
+
+  - investigate convergence: (i) look at both error and prior variance 
+  convergence and (ii) look at actual prior distribution convergence 
+  (plot true distribution and estimated prior every 5 iterations or so)
+
+  \section{Application in eQTLs mapping}\label{sec:application}
+  This application concerns eQTLs detection from the GEUVADIS data
+  \cite[]{lappalainen_transcriptome_2013}. After the removal of 9 genes due to
+  missing SNP information, we have 99 genes (outcomes) from the p38MAPK pathway. 
+  After removal of the subjects without any expression or SNP data, 373 subjects
+  remained. For each gene, we selected SNPs with minor alles frequency (MAF) 
+  larger than 0.05 and within 105 bases up- or downstream from the gene. This 
+  resulted in 99 sets of SNPs (one for each gene) from the 373 subjects, where 
+  the number of SNPs per gene ranges from 56 to 1169.
+
+  We used Ensembl \cite[]{hunt_ensembl_2018} to create several external 
+  variables: (i) SNP location 1 (binary, in- or outside of the gene),
+  (ii) SNP location 2 (binary, in an exon or an intron), (iii) 
+  genelength (continuous), (iv) minor allele frequency of the SNP
+  (continuous), and (v) squared MAF. We expect that SNPs that are in the gene 
+  and/or in an exon are more important in eQTLs detection. Since a larger gene
+  might depend on more SNPs, we expect that longer genes yield lower 
+  regression parameter estimates. We predict a positive correlation between
+  the regression parameters and the MAF.
+
+  We fit five different models to a subset of the subjects and 
+  estimated performance measures on the rest of the subjects. The five models we
+  fit differed only in the included external variables. The intercept only model
+  is considered a reference model. The other four models all included the 
+  genelengths and MAFs (variables (iii) and (iv)). 
+  Models 1 and 2 included variable 
+  (i), while models 1, 3, and 4 included variable (ii). In addition, model 4
+  included the squared MAFs (v).
+
+  For each model, we ran the algorithm for 100 EB iterations and obtained 
+  the $\hat{\bm{\alpha}}$ hyperparameter estimates in Table 
+  \ref{tab:hyperparameters_eQTL}. In addition, we calculated posterior means
+  for the $\bm{\beta}_d$ from MCMC samples from the posterior with the estimated
+  hyperparameters. We estimated mean squared error (MSE) for these posterior 
+  means. A common measure 
+  of Bayesian model fit is the log pseudo marginal likelihood
+  (LPML) \cite[]{gelfand_model_1992}, the Bayesian version of the leave-one-out 
+  cross-validated log likelihood (see Section \ref{sm-sec:lpml} for more 
+  details). We estimated LPML for the different models and added them,
+  together with the MSEs to Table \ref{tab:hyperparameters_eQTL}.
+<<hyperparameters_eQTL>>= 
+load("results/eQTL_fit1.Rdata")
+seq.mse <- c(2.0699, 2.0629, 2.0649, 2.0628, 2.0638)
+seq.lpml <- round(c(mean(lpml.ienig$lpml), mean(lpml1.enig$lpml),
+                    mean(lpml2.enig$lpml), mean(lpml3.enig$lpml), 
+                    mean(lpml4.enig$lpml)), 5)
+tab <- cbind(round(rbind(c(fit.ienig$eb$alpha, rep(NA, 5)),
+                         c(fit1.enig$eb$alpha, NA),
+                         c(fit2.enig$eb$alpha[c(1:2)], NA, 
+                           fit2.enig$eb$alpha[-c(1:2)], NA),
+                         c(fit3.enig$eb$alpha[1], NA, 
+                           fit3.enig$eb$alpha[-1], NA),
+                         c(fit4.enig$eb$alpha[1], NA, 
+                           fit4.enig$eb$alpha[-1])), 2),
+             seq.mse, seq.lpml)
+tab[which.min(tab[, 7]), 7] <- kableExtra::cell_spec(
+  tab[which.min(tab[, 7]), 7], format="latex", bold=TRUE)
+tab[which.max(tab[, 8]), 8] <- kableExtra::cell_spec(
+  tab[which.min(tab[, 8]), 8], format="latex", bold=TRUE)
+rownames(tab) <- c("intercept only", paste("model", 1:4))
+options(knitr.kable.NA = '')
+kableExtra::kable_styling(kableExtra::column_spec(
+  knitr::kable(tab, align=rep("r", ncol(tab)),
+               col.names=c("intercept", "in gene", "in exon", "length", 
+                           "MAF", "MAF$^2$", "MSE", "LPML"),
+               caption="Estimated $\\hat{\\bm{\\alpha}}$ hyperparameters, 
+               MSE (lowest in bold), and LPML (highest in bold).", 
+               "latex", booktabs=TRUE, escape=FALSE, 
+               digits=c(rep(2, 6), 4, 4)), 8, border_left=TRUE),
+  latex_options=c("HOLD_position"))
+@
+
+  - show influence of co-data with standardized $\bm{\alpha}$'s.
+
+  A quantity that is often of interest in eQTLs detection studies, is the 
+  gene expression heritability. It is the proportion of a gene's expression that
+  is explained with the SNPs and a measure of the relationship between trait 
+  phenotypic variation and genetic variation. Heritability (per gene) is 
+  given by $h_d^2 = (\sigma^2_d \sum_{j=1}^{p_d} \gamma_{jd}^2)/
+  (\sigma^2_d \sum_{j=1}^{p_d} \gamma_{jd}^2 + \sigma_d^2)$. We assess 
+  mean heritability of the five models for a rank-based selection of SNPs.
+  We present MCMC-based posterior means and one standard deviation error bars in
+  Figure \ref{fig:lines_eqtl_hsq}.
+<<lines_eqtl_hsq, fig.cap="Mean heritability in the eQTL detection data.", out.width="80%", fig.asp=2/3>>=
+@
+
+  [! describe results here]
+
+  - maybe use explained heritability (fraction of full model heritability)
+
+  - AUC of curve as overal measure
+
+  % Annotation is from GRCh37 reference genome: 
+  % \url{https://www.ncbi.nlm.nih.gov/assembly/GCF_000001405.13/}. 
+  % Annotation file that Gino used: 
+  % \url{https://www.gencodegenes.org/human/release_12.html}.
+  % Ensembl version of this genome:
+  % \url{https://grch37.ensembl.org/Homo_sapiens/Info/Index}.
+  % Possibly use GeuvadisTranscriptExpr R package.
+
+  \section{Discussion}\label{sec:discussion}
+  In this paper we have introduced a normal inverse Gaussian prior for 
+  multivariate high dimensional linear regression models that is estimated by
+  empirical-variational Bayes. The estimation is adaptive to external data
+  that describes the relation between the outcomes and features in a data-driven
+  manner. In simulations we have shown that the inverse Gaussian model 
+  does not perform worse, and often bettern, than a common alternative, 
+  the Student's $t$ prior. in addition, the Student's $t$ prior does not admit 
+  relatively straightforward estimation, compared to the inverse Gaussian 
+  prior, as explained in Section \ref{sec:estimation}. The 
+  simulations also show that the model is reasonably well approximated and 
+  estimated using empirical-variational Bayes. An eQTL detection application 
+  shows that external data may indeed benefit prior estimation and prediction
+  performance.
+
+  Apart from eQTL detection, the model may be used for network reconstruction as 
+  in \cite{leday_gene_2017}, drug response prediction (manuscript in 
+  preparation by the current authors), and other applications with multiple
+  outcomes, large feature space, and available external information.
+
+  A drawback of the method is the fact that the external data is required to
+  be positive. That may not always be the case in practice. We note, however,
+  that the $\mathbf{c}_d^j$ may be translated to be positive by adding a 
+  constant. In the usual predictive feature sense this doesn't work, since one
+  might still encounter negative future values. Here, the external data is
+  only used for estimation and not for prediction, so this issue is 
+  non-existent.
+
+  \section*{Software}
+	A (developmental) \texttt{R} package is available from 
+	\url{https://github.com/magnusmunch/cambridge}.
 
@@
 	\section*{Acknowledgements}
-	We thank Zhi Zhao and Leiv Tore Salte R{\o}nneberg from the Oslo Centre for 
-	Biostatistics and Epidemiology for help with the pre-processing of the data. 
+	We thank Gino Kpogbezan for providing the pre-processed data in Section 
+	\ref{sec:application}.
 	\textit{Conflict of Interest}: None declared.
 
-	\bibliographystyle{author_short3} 
+	\bibliographystyle{author_short3.bst}
 	\bibliography{refs}

### 2019-11-19 docs/supplement1.Rnw
redoing simulations

@@
 \setlength{\textwidth}{15cm}
-\title{Supplementary material to: Empirical Bayes estimation of the normal
-inverse Gaussian prior in 
-simultaneous-equation models}
+\title{Supplementary material to: Drug sensitivity prediction with the normal 
+inverse Gaussian prior informed by external data}
 \date{\today}
 \author{Magnus M. M\"unch$^{1,2}$\footnote{Correspondence to: 
-\href{mailto:m.munch@vumc.nl}{m.munch@vumc.nl}}, Mark A. van de Wiel$^{1,3}$, 
-Sylvia Richardson$^{3}$, \\ and Gwena{\"e}l G. R. Leday$^{3}$}
+\href{mailto:m.munch@amsterdamumc.nl}{m.munch@amsterdamumc.nl}}, Mark A. van de 
+Wiel$^{1,3}$, Sylvia Richardson$^{3}$, \\ and Gwena{\"e}l G. R. Leday$^{3}$}
 
@@
 	\noindent
-	1. Department of Epidemiology \& Biostatistics, Amsterdam Public Health
-	research institute, Amsterdam University medical centers, PO Box 7057, 1007 MB
-	Amsterdam, The Netherlands \\*
-	2. Mathematical Institute, Leiden University, Leiden, The Netherlands \\*
+	1. Department of Epidemiology \& Biostatistics, Amsterdam UMC, VU University, 
+	PO Box 7057, 1007 MB Amsterdam, The Netherlands \\
+	2. Mathematical Institute, Leiden University, Leiden, The Netherlands \\
 	3. MRC Biostatistics Unit, Cambridge Institute of Public Health, Cambridge,
@@
 	This document contains the Supplementary Material (SM) to the document 
-	`Empirical Bayes estimation of simultaneous-equation models'. In the 
+	`Drug sensitivity prediction with the normal inverse Gaussian prior 
+	informed by external data'. In the 
 	following, this document is referred to as Main Document (MD).
 
-	\section{NIG and Student's \texorpdfstring{$t$}{t} Priors}\label{sec:priors}
-	\subsection{Normal inverse Gaussian}\label{sec:nigprior}
-	Let $\theta = (\mathbf{c} \tr \bm{\alpha})^{-1}$. Then, 
-	for a fixed error variance $\sigma^2$, we have marginal prior variance 
-	and kurtosis:
+	Section \ref{sec:prior} describes the moments of the normal
+	inverse Gaussian (NIG) prior, Section \ref{sec:vbderivations} gives
+	the variational Bayes (BV) derivations that lead to the estimating equations 
+	in the MD, Section \ref{sec:elbo} gives the evidence lower bound (ELBO) that 
+	is maximised during the VB iterations. In Section \ref{sec:computation} 
+	efficient VB calculations are given, while Section 
+	\ref{sec:ratiosmodifiedbessels} outlines a computationally stable method
+	to calculate ratios of modified Bessel functions. Section \ref{sec:eb} shows
+	that the empirical Bayes (EB) scale estimates $\bm{\lambda}$ are non-negative.
+	The last Section \ref{sec:gibbssampler} introduces a Gibbs sampling scheme
+	to sample from the posterior corresponding to the NIG model.
+
+	\section{Normal inverse Gaussian prior moments}\label{sec:prior}
+	Let $\beta \sim \mathcal{N}(0,\sigma^2 \tau^2 \gamma^2)$, where 
+	$\sigma^2$ and $\tau^2$ are fixed and $\gamma^2 \sim 
+	\mathcal{IG}(\phi,\lambda)$, then
 	\begin{align*}
-	  \V(\beta | \sigma^2) & = \sigma^2 \theta, \\
-	  \mathcal{K}(\beta | \sigma^2) & = \frac{3\theta}{\lambda} + 3.
+	  \E(\beta | \sigma^2, \tau^2) & = 0, \\
+	  \V(\beta | \sigma^2, \tau^2) & = \sigma^2 \tau^2 \phi, \\
+	  \mathcal{K}(\beta | \sigma^2, \tau^2) & = 
+	  \frac{3\phi}{\lambda} + 3,
 	\end{align*}
-
-	\subsection{Student's \texorpdfstring{$t$}{t} prior}\label{sec:studentstprior}
-	For a fixed error variance $\sigma^2$, we have marginal prior variance 
-	and kurtosis:
-	\begin{align*}
-	  \V(\beta | \sigma^2) & = \frac{\sigma^2 \xi}{\eta - 2}, \\
-	  \mathcal{K}(\beta | \sigma^2) & = \frac{6}{\eta - 4} + 3.
-	\end{align*}
+	where $\mathcal{K}(\cdot)$ denotes the kurtosis.
 
@@
 	  (\y_d | \bbeta_d, \sigma_d^2 )] + \E [\log \pi (\bbeta_d | \bgamma_{d}^2, 
-	  \sigma_d^2)] \\
+	  \sigma_d^2, \tau_d^2)] \\
 	  & \propto - \frac{1}{2} \sum_{i=1}^n 
@@
 	  \frac{1}{2} \sum_{j=1}^p 
-	  \E \( \frac{\beta_{jd}^2}{\gamma_{jd}^2 \sigma_d^2} \), \\
+	  \E \( \frac{\beta_{jd}^2}{\gamma_{jd}^2 \sigma_d^2 \tau_d^2} \), \\
 	  q_{\bbeta_d} (\bbeta_d) & \overset{D}{=} 
@@
 	  & \text{with } \bSigma_d = \E(\sigma_d^{-2})^{-1} 
-	  \{\X \tr \X + \diag [\E(\gamma_{jd}^{-2})] \}^{-1}, \\
-	  & \text{and } \bmu_d = \{\X \tr \X + \diag [\E(\gamma_{jd}^{-2})] \}^{-1} 
+	  \{\X \tr \X + \E(\tau_d^{-2})\diag [\E(\gamma_{jd}^{-2})] \}^{-1}, \\
+	  & \text{and } \bmu_d = \{\X \tr \X + \E(\tau_d^{-2})
+	  \diag [\E(\gamma_{jd}^{-2})] \}^{-1} 
 	  \X \tr \y_d.
@@
 	  \log q_{\gamma_{jd}^{2}}(\gamma_{jd}^{2}) & \propto \E 
-	  [\log \pi (\beta_{jd} | \gamma_{jd}^{2}, \sigma_d^2)] + 
+	  [\log \pi (\beta_{jd} | \gamma_{jd}^{2}, \sigma_d^2, \tau_d^2)] + 
 	  \log \pi(\gamma_{jd}^{2}) \\
 	  & \propto -\frac{1}{2} \log \gamma_{jd}^{2} - \frac{1}{2} 
-	  \E \( \frac{\beta_{jd}^2}{\sigma^{2}_d}\) \gamma_{jd}^{-2} - 
+	  \E \( \frac{\beta_{jd}^2}{\sigma^{2}_d \tau_d^2}\) \gamma_{jd}^{-2} - 
 	  \frac{3}{2} \log \gamma_{jd}^{2} - 
-	  \frac{\lambda_d \{\gamma_{jd}^2 - 
-	  [(\mathbf{c}_{d}^j) \tr \bm{\alpha}]^{-1}\}^2}
-	  {2[(\mathbf{c}_{d}^j) \tr \bm{\alpha}]^{-2}} \gamma_{jd}^{-2} \\
+	  \frac{\lambda_{\text{feat}} \{\gamma_{jd}^2 - 
+	  \phi_{jd}\}^2}
+	  {2\phi_{jd}^{2}} \gamma_{jd}^{-2} \\
 	  & \propto \( -1 - 1\) \log \gamma_{jd}^{2} - 
-	  \frac{\lambda_d[(\mathbf{c}_{d}^j) \tr \bm{\alpha}]^{2}}{2} \gamma_{jd}^2 - 
-	  \frac{1}{2}\[ \lambda_d + \E(\sigma_d^{-2}) \E(\beta_{jd}^2)\] 
-	  \gamma_{jd}^{-2} , \\
+	  \frac{\lambda_{\text{feat}}}{2\phi_{jd}^2} \gamma_{jd}^2 - 
+	  \frac{1}{2}\[ \lambda_{\text{feat}} + \E(\tau_d^{-2})\E(\sigma_d^{-2}) 
+	  \E(\beta_{jd}^2)\] \gamma_{jd}^{-2} , \\
 	  q_{\gamma_{jd}^{2}}(\gamma_{jd}^{2}) & \overset{D}{=} \mathcal{GIG} 
-	  \(-1, \lambda_d[(\mathbf{c}_{d}^j) \tr \bm{\alpha}]^{2}, \delta_{jd} \), \\
-	  & \text{with } \delta_{jd} =\E(\sigma_d^{-2})
-	  \[\E(\beta_{jd})^2 + \V(\beta_{jd})\] + \lambda_d.
+	  \(-1, \lambda_{\text{feat}}/\phi_{jd}^{2}, \delta_{jd} \), \\
+	  & \text{with } \delta_{jd} =\E(\tau_d^{-2})\E(\sigma_d^{-2})
+	  \[\E(\beta_{jd})^2 + \V(\beta_{jd})\] + \lambda_{\text{feat}}.
+	\end{align*}
+	Similarly, we derive the variational posterior for $\tau_d^2$ as:
+	\begin{align*}
+	  \log q_{\tau_{d}^{2}}(\tau_{d}^{2}) & \propto \E 
+	  [\log \pi (\bbeta_{d} | \bgamma_{d}^{2}, \sigma_d^2, \tau_d^2)] + 
+	  \log \pi(\tau_{d}^{2}) \\
+	  & \propto -\frac{p}{2} \log \tau_{d}^{2} - \frac{1}{2} \sum_{j=1}^p
+	  \E \( \frac{\beta_{jd}^2}{\sigma^{2}_d \gamma_{jd}^{2}}\)\tau_d^{-2}  - 
+	  \frac{3}{2} \log \tau_{d}^{2} - 
+	  \frac{\lambda_{\text{drug}} \{\tau_{d}^2 - \chi_{d}\}^2}
+	  {2\chi_{d}^{2}} \tau_{d}^{-2} \\
+	  & \propto \( -\frac{p + 1}{2} - 1\) \log \tau_{d}^{2} - 
+	  \frac{\lambda_{\text{drug}}}{2\chi_{d}^2} \tau_{d}^2 - 
+	  \frac{1}{2}\[ \lambda_{\text{drug}} + \E(\sigma_d^{-2}) \sum_{j=1}^p 
+	  \E(\gamma_{jd}^{-2}) \E(\beta_{jd}^2)\] \tau_{d}^{-2} , \\
+	  q_{\tau_{d}^{2}}(\tau_{d}^{2}) & \overset{D}{=} \mathcal{GIG} 
+	  \(-\frac{p + 1}{2}, \lambda_{\text{drug}}/\chi_{d}^{2}, \eta_{d} \), \\
+	  & \text{with } \eta_{d} =\E(\sigma_d^{-2}) \sum_{j=1}^p
+	  \E(\gamma_{jd}^{-2})\[\E(\beta_{jd})^2 + \V(\beta_{jd})\] + 
+	  \lambda_{\text{drug}}.
 	\end{align*}
@@
 	  \log q_{\sigma_d^{2}}(\sigma_d^{2}) & \propto \E 
-	  [\log \mathcal{L}(\y_d | \bbeta)d, \sigma_d^2 )] + 
-	  \E [\log \pi (\bbeta_d | \bgamma_d^2, \sigma_d^2)] + 
+	  [\log \mathcal{L}(\y_d | \bbeta_d, \sigma_d^2 )] + 
+	  \E [\log \pi (\bbeta_d | \bgamma_d^2, \sigma_d^2,\tau_d^2)] + 
 	  \log \pi (\sigma_d^{2}) \\
@@
 	  \frac{p}{2} \log \sigma_d^{2} - \frac{1}{2} \sum_{j=1}^p 
-	  \E \( \frac{\beta_{jd}^2)}{\gamma_{jd}^{2}}\) \sigma_d^{-2} - 
+	  \E \( \frac{\beta_{jd}^2}{\gamma_{jd}^{2}\tau_d^2}\) \sigma_d^{-2} - 
 	  \frac{3}{2} \log \sigma^2_d \\
-	  & = \(-\frac{n + p + 1}{2} - 1 \) \log \sigma_d^{2} - 
+	  & = \(-\frac{n + p + 1}{2} - 1 \) \log \sigma_d^{2} \\
+	  & \,\,\,\,\,\,\,\,\,\, - 
 	  \frac{1}{2}\left\{ \sum_{i=1}^n \E [(y_{id} - \x_i \tr \bbeta_d)^2] + 
-	  \sum_{j=1}^p \E(\gamma_{jd}^{-2})\E(\beta_{jd}^2) \right\} \sigma_d^{-2}, \\
+	  \E(\tau_d^{-2}) \sum_{j=1}^p \E(\gamma_{jd}^{-2})
+	  \E(\beta_{jd}^2) \right\} \sigma_d^{-2}, \\
 	  q_{\sigma_d^{2}}(\sigma_d^{2}) & \overset{D}{=} 
@@
     & \,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\, + 
-    \trace \{ \diag[\E(\gamma_{jd}^{-2})] \V(\bbeta_d)\} + 
-     \E(\bbeta_d \tr) \diag[\E(\gamma_{jd}^{-2})] \E(\bbeta_d ) \bigg\}.
+    \E(\tau_d^{-2})\trace \{ \diag[\E(\gamma_{jd}^{-2})] \V(\bbeta_d)\} + 
+    \E(\tau_d^{-2})\E(\bbeta_d \tr) \diag[\E(\gamma_{jd}^{-2})] \E(\bbeta_d )
+    \bigg\}.
+  \end{align*}
+  If we fill in the expectations and variances, we arrive at the estimating
+  equations in MD equations (\ref{md-eq:vbequations}).
+
+  \section{Evidence lower bound}\label{sec:elbo}
+  The evidence lower bound that is maximised during the iterations is: 
+  $\text{ELBO}^{(l)}=\text{ELBO}_d^{(l)}$, with:
+  \begin{align*}
+    \text{ELBO}&^{(l)}_d \\
+    & = -\frac{n+p+1}{2} \log \pi + \frac{p+1-n}{2}\log 2 + \frac{n+1}{2} + p
+    + \log \Gamma \(\frac{n + p + 1}{2} \) \\
+    & - \frac{n + p + 1}{2} \log \zeta_d^{(l)} + 
+    \frac{1}{2} \log |\bSigma_d^{(l)}| - 
+    \frac{n+p+1}{4\zeta_d^{(l)}}  \bigg[\mathbf{y}_d \tr \mathbf{y}_d -
+    2 \mathbf{y}_d \tr \X \bm{\mu}_d^{(l)} + 
+    \trace ( \X \tr \X \bm{\Sigma}_d^{(l)}) \\ 
+    & + (\bm{\mu}_d^{(l)}) \tr \X \tr \X \bm{\mu}_d^{(l)} +
+    g_{d}^{(l)} \trace \[ \diag (b_{jd}^{(l)}) \bm{\Sigma}_d^{(l)}\] + 
+    g_{d}^{(l)} (\bm{\mu}_d^{(l)}) \tr \diag (b_{jd}^{(l)}) 
+    \bm{\mu}_d^{(l)}\bigg] + p \log \lambda_{\text{feat}}^{(l)} \\
+    & + \frac{p + 3}{4} \log \lambda_{\text{drug}}^{(l)} + 
+    \frac{\eta_d^{(l)} - \lambda_{\text{drug}}^{(l)}}{2} g_d^{(l)} + 
+    \sum_{j=1}^p \frac{\delta_{jd}^{(l)} - \lambda_{\text{feat}}^{(l)}}{2} 
+    b_{jd}^{(l)} + \frac{\lambda_{\text{drug}}^{(l)}}{\chi_d^{(l)}} +
+    \sum_{j=1}^p \frac{\lambda_{\text{feat}}^{(l)}}{\phi_{jd}^{(l)}} + \\
+    & - \frac{p + 1}{2} \log \chi_d^{(l)} - \sum_{j=1}^p \log \phi_{jd}^{(l)}
+    -\frac{p+1}{4} \log \eta_d^{(l)} - 
+    \frac{1}{2} \sum_{j=1}^p \log \delta_{jd}^{(l)} \\
+    & + \log K_{\frac{p+1}{2}}\(\sqrt{\eta_{d}^{(l)}
+    \lambda_{\text{drug}}^{(l)}}/
+    \chi_{d}^{(l)}\) + \sum_{j=1}^p \log K_{1}\(\sqrt{\delta_{jd}^{(l)}
+    \lambda_{\text{feat}}^{(l)}}/\phi_{jd}^{(l)}\).
   \end{align*}
 
-% 	\section{Evidence lower bound}\label{sec:evidencelowerbound}
-% 	The variational Bayes posterior is found through maximisation of an evidence
-% 	lower bound (ELBO). Additionally, we may monitor convergence through this 
-% 	ELBO. To compute the ELBO after updating the EB parameters from 
-% 	$\theta_d^{(l)}, \lambda_d^{(l)}$ to 
-% 	$\theta_d^{(l + 1)}, \lambda_d^{(l + 1)}$, we compute 
-% 	$\text{ELBO}^{(l + 1)} = \sum_{d=1}^D \text{ELBO}_d^{(l + 1)}$, with:
-% 	\begin{align*}
-% 	  \text{ELBO}_d^{(l + 1)} = & \frac{1}{2} \log |\bm{\Sigma}_d^{(l)}| - 
-% 	  \frac{a_d^{(l)}}{2} \[ \y_d \tr \y_d - 2\y_d \tr \X \bm{\mu}_d^{(l)} + 
-% 	  (\bm{\mu}_d^{(l)}) \tr \X \tr \X \bm{\mu}_d^{(l)} + 
-% 	  \trace ( \X \tr \X \bm{\Sigma}_d^{(l)})\] \\
-% 	  & + \frac{b_d^{(l)}}{2} \[ \delta_d^{(l)} - 
-% 	  a_d^{(l)}(\bm{\mu}_d^{(l)}) \tr \bm{\mu}_d^{(l)} - 
-% 	  a_d^{(l)} \trace (\bm{\Sigma}_d^{(l)}) \] \\
-% 	  & - \frac{p + n + 1}{2} \log \zeta_d^{(l)} + 
-% 	  \frac{p + 1}{4} \log \lambda_d^{(l)} - 
-% 	  \frac{p + 1}{2} \log \theta_d^{(l)} - \frac{p+1}{4} \log \delta_d^{(l)} \\
-% 	  & + \log K_{\frac{p+1}{2}}\(\sqrt{\lambda_d^{(l)} 
-% 	  \delta_d^{(l)}}/\theta_d^{(l)}\) + 
-% 	  \frac{\lambda_d^{(l)} e_d^{(l)}}{2(\theta_d^{(l)})^2} \\
-% 	  & + \frac{\lambda_d^{(l+1)}}{\theta_d^{(l+1)}} + 
-% 	  \frac{1}{2}\log \lambda_d^{(l+1)} - \frac{\lambda_d^{(l+1)} 
-% 	  e_d^{(l)}}{2 (\theta_d^{(l + 1)})^2} - 
-% 	  \frac{\lambda_d^{(l + 1)} b_d^{(l)}}{2},
-% 	\end{align*}
-% 	where $a^{(l)}$, $b^{(l)}$, and $e^{(l)}$ are as in the MD. To compute the 
-% 	ELBO after a VB update we simply let 
-%   $\theta_d^{(l)}, \lambda_d^{(l)}=\theta_d^{(l + 1)}, \lambda_d^{(l + 1)}$.
-% 	 
-%   In practice, there are different options to assess convergence. We may monitor
-%   the parameters themselves, the ELBO, or a combination of the two. We may also 
-%   iterate for an \textit{a priori} fixed number of times. We may also choose to 
-%   use a different convergence criterium for the EB and VB iterations. Currently 
-%   we fix the number of EB iterations to 20, while we do 2 VB iterations per EB 
-%   iteration.
-
-  \section{Efficient computation}
+  \section{Efficient computation}\label{sec:computation}
   The empirical Bayes updates require the quantities
-  $e_{jd}$ and $b_{jd}$, which in turn depend on the quantities 
+  $e_{jd}$, $b_{jd}$, $g_d$, and $f_d$, which in turn depend on the quantities 
   $\diag(\bSigma_d)$, $\bmu_{d}$, $\y_d \tr \X \bmu_d$, 
   $\trace(\X \tr \X \bSigma_d)$, $\bmu_d \tr \X \tr \X \bmu_d$,
-  $\trace[\diag(b_{jd})\bSigma_d]$, and $\bmu_d \tr \diag(b_{jd}) \bmu_d$.
-  Let $\mathbf{H}_d=\diag[(b_{jd}^{(h)})^{-1}]$. Then, the first two quantities 
-  are more efficiently calculated as:
+  $g_d\trace[\diag(b_{jd})\bSigma_d]$, and $g_d\bmu_d \tr \diag(b_{jd}) \bmu_d$.
+  Let $\mathbf{H}_d=g_d \diag[(b_{jd}^{(h)})^{-1}]$. Then, the first two
+  quantities are efficiently calculated as:
   \begin{subequations}\label{eq:complexity1}
@@
   The quantities $\y_d \tr \X \bmu_d$, $\bmu_d \tr \X \tr \X \bmu_d$,
-  $\trace[\diag(b_{jd})\bSigma_d]$, and $\bmu_d \tr \diag(b_{jd}) \bmu_d$ are
-  easily calculated from the first two. For the remaining one we have:
+  $g_d\trace[\diag(b_{jd})\bSigma_d]$, and $g_d\bmu_d \tr \diag(b_{jd}) \bmu_d$
+  are easily calculated from the first two. For the remaining one we have:
   \begin{align}\label{eq:complexity2}
@@
   \end{align*} 
-  The ratio $K_0(x)/K_1(x)$ is well-behaved, such that the ratio may be computed
+  The ratio $K_0(x)/K_1(x)$ is well-behaved, so that it may be computed
   without numerical issues.
+
+  \section{Empirical Bayes}\label{sec:eb}
+  For the feature scale parameter $\lambda_{\text{feat}}$ updates to be 
+  non-negative, we have to show that:
+  \begin{align}\label{eq:inequality}
+    &\sum_{d=1}^D \sum_{j=1}^{p} 
+    b_{jd} + \bm{\alpha}_{\text{feat}} \tr \mathbf{C} \tr 
+    \diag (e_{jd}) \mathbf{C} \bm{\alpha}_{\text{feat}} - 
+    2 \bm{\alpha}_{\text{feat}} \tr \mathbf{C} \tr 
+    \mathbf{1}_{pD \times 1}  = \sum_{d=1}^D \sum_{j=1}^{p} \(b_{jd} + 
+    \phi_{jd}^{-2} e_{jd} - 2 \phi_{jd}^{-1} \) \geq 0,
+  \end{align}
+  which holds if all summands in the left-hand side of (\ref{eq:inequality})
+  are non-negative.
+
+  Next, we note that through Jensen's inequality 
+  $e_{jd}^{-1}=\E(\gamma_{jd}^{2})^{-1} \leq \E(\gamma_{jd}^{-2})=b_{jd}$,
+  we may lowerbound the summands in left-hand side of (\ref{eq:inequality}):
+  $$
+  b_{jd} + \phi_{jd}^{-2} e_{jd} - 2 \phi_{jd}^{-1} \geq
+  e_{jd}^{-1} + \phi_{jd}^{-2} e_{jd} - 2 \phi_{jd}^{-1}.
+  $$
+  which we use to show that:
+  \begin{align*}
+    e_{jd}^{-1} + \phi_{jd}^{-2} e_{jd} - 2 \phi_{jd}^{-1} & \geq 0 \\
+    1 + \phi_{jd}^{-2} e_{jd}^2 - 2 \phi_{jd}^{-1} e_{jd} & \geq 0 \\
+    \( 1 - \phi_{jd}^{-1} e_{jd} \)^2 & \geq 0.
+  \end{align*}
+  Similar reasoning holds for the drug scale parameter $\lambda_{\text{drug}}$.
 
@@
   MCMC samples from the posterior corresponding to model 
-  (\ref{md-eq:priormodel}) may be generated for each equation independently. 
+  (\ref{md-eq:prior}) may be generated for each equation independently. 
   MCMC samples may be generated by iteratively sampling the following 
@@
     \begin{align}
-      \bm{\beta}_d | \gamma_d^2, \sigma_d^2, \y_d & \sim \mathcal{N}_p 
-      (\bm{\mu}_d, \bm{\Sigma}_d), \label{eq:betasample} \\
-      \text{with } & \bm{\Sigma}_d = \sigma_d^2 (\X \tr \X + 
-      \gamma_d^{-2} \I)^{-1} \label{eq:sigmasample} \\
-      \text{and } & \bm{\mu}_d = (\X \tr \X + 
-      \gamma_d^{-2} \I)^{-1} \X \tr \y_d \label{eq:musample} \\
-      \gamma_d^2 | \bm{\beta}_d, \sigma_d^2, \y & \sim \mathcal{GIG}
-      \(-\frac{p+1}{2}, \lambda_d \theta_d^{-2}, \delta_d\), \nonumber \\
-      \text{with } & \delta_d = \sigma_d^{-2} \bm{\beta}_d \tr \bm{\beta}_d + 
-      \lambda_d, \nonumber \\
-      \sigma_d^2 | \bm{\beta}_d, \gamma_d^2, \y & \sim \Gamma^{-1} 
+      \bm{\beta}_d | \bgamma_d^2, \sigma_d^2, \tau_d^2, \y_d & \sim 
+      \mathcal{N}_p (\bm{\mu}_d, \bm{\Sigma}_d), \label{eq:betasample} \\
+      \text{with } & \bm{\Sigma}_d = \sigma_d^2 [\X \tr \X + 
+      \tau_d^{-2} \diag (\gamma_{jd}^{-2})]^{-1}, \label{eq:sigmasample} \\
+      \text{and } & \bm{\mu}_d = [\X \tr \X + 
+      \tau_d^{-2} \diag (\gamma_{jd}^{-2})]^{-1} \X \tr 
+      \y_d \label{eq:musample}, \\
+      \gamma_{jd}^2 | \beta_{jd}, \sigma_d^2, \tau_d^2, \y &
+      \sim \mathcal{GIG}
+      \(-1, \lambda_{\text{feat}}/ \phi_{jd}^{2}, \delta_{jd}\), \nonumber \\
+      \text{with } & \delta_{jd} = \sigma_d^{-2} \tau_d^{-2} \beta_{jd}^2 + 
+      \lambda_{\text{feat}}, \nonumber \\
+      \tau_d^2 | \bm{\beta}_d, \sigma_d^2, \bgamma_{d}^2, \y & \sim
+      \mathcal{GIG}
+      \(-\frac{p+1}{2}, \lambda_{\text{drug}}/ \chi_{d}^{2}, \eta_d\), 
+      \nonumber \\
+      \text{with } & \eta_d = \sigma_d^{-2} \sum_{j=1}^p \gamma_{jd}^{-2}
+      \beta_{jd}^2 + \lambda_{\text{drug}}, \nonumber \\
+      \sigma_d^2 | \bm{\beta}_d, \bgamma_d^2, \tau_d^2, \y & \sim \Gamma^{-1} 
       \(\frac{n + p + 1}{2}, \zeta_d\), \nonumber \\
-      \text{with } & \zeta = \frac{1}{2} \[\mathbf{y} \tr \mathbf{y} -
+      \text{with } & \zeta_d = \frac{1}{2} \bigg[\mathbf{y} \tr \mathbf{y} -
       2 \mathbf{y} \tr \X \bm{\mu} + \trace ( \X \tr \X \bm{\Sigma}) + 
       (\bm{\mu}) \tr \X \tr \X \bm{\mu} + 
-      \gamma^{-2} \trace ( \bm{\Sigma}) + 
-      \gamma^{-2} (\bm{\mu}) \tr \bm{\mu}\]. \nonumber
+      \tau^{-2}_d \trace (\diag(\gamma_{jd}^{-2}) \bm{\Sigma}) \nonumber \\
+      & \,\,\,\,\,\,\,\,\,\, + \tau_d^{-2} (\bm{\mu}) \tr 
+      \diag(\gamma_{jd}^{-2}) \bm{\mu}\bigg]. \nonumber
     \end{align}
@@
 
-  \section{Inverse Gamma model}\label{sec:inversegamma}
-  The model introduced in MD Section \ref{md-sec:inversegammamodel} draws the 
-  $\gamma_d^2$ from an inverse Gamma distribution:
-	\begin{equation}\label{eq:invgammaprior}
-	  \gamma_d^2 \sim \Gamma^{-1}(\eta_d/2, \lambda_d/2)=
-	  \mathcal{GIG}(-\eta_d/2, 0, \lambda_d),
-	\end{equation}
-	with shape $\eta_d/2$ and scale $\lambda/2$. For the sake of comparability, we
-	have rescaled the inverse Gamma parameters and reparametrised as a generalized
-	inverse Gaussian. This allows for an easier comparison to the inverse 
-	Gaussian hyperprior:
-	\begin{equation}\label{eq:invgaussianprior}
-	  \mathcal{IG}(\theta_d, \lambda_d) =
-	  \mathcal{GIG}(-1/2, \lambda_d \theta_d^{-2}, \lambda_d).
-  \end{equation}
-	From (\ref{eq:invgammaprior}) and (\ref{eq:invgaussianprior}) it is obvious 
-	that the two models coincide for $\theta_d \to \infty$ and $\eta_d = 1$. In
-	Figure \ref{fig:dens_igaussian_igamma} the two distributions are compared for 
-	a range of parameter values.
-<<dens_igaussian_igamma, fig.cap="Several densities of the inverse Gaussian and inverse Gamma families", out.width="100%", fig.asp=2/3>>=
-digauss <- function(x, theta, lambda, eta) {
-  sqrt(lambda/(x^3*2*pi))*exp(-lambda*(x - theta)^2/(2*theta^2*x))
-}
-
-digamma <- function(x, theta, lambda, eta) {
-  alpha <- eta/2
-  beta <- lambda/2
-  beta^alpha/gamma(alpha)*x^(-alpha - 1)*exp(-beta/x)
-}
-
-eta <- c(1, 2, 1, 1, 5)
-lambda <- c(1, 1, 2, 1/2, 2)
-theta <- c(1000, 1000, 1, 1, 1)
-
-labels <- c("inv. Gaussian", "inv. Gamma")
-col <- c(1, 2)
-
-opar <- par(no.readonly=TRUE)
-par(mar=opar$mar*c(1, 1.3, 1, 1))
-layout(matrix(c(rep(c(1, 1, 2, 2, 3, 3), 2), rep(c(0, 4, 4, 5, 5, 0), 2)),
-                 nrow=4, ncol=6, byrow=TRUE))
-for(i in 1:length(eta)) {
-  modes <- c(theta[i]*(sqrt(1 + 9*theta[i]^2/
-                            (4*lambda[i]^2)) - 3*theta[i]/(2*lambda[i])),
-             lambda[i]/(eta[i] + 2))
-  ylim <- c(0, max(digauss(modes[1], theta[i], lambda[i], eta[i]),
-                   digamma(modes[2], theta[i], lambda[i], eta[i])))
-  curve(digauss(x, theta[i], lambda[i], eta[i]), 0.001, 5, n=1000,
-        ylim=ylim, ylab="Density", "x", col=col[1], 
-        main=bquote(eta==.(eta[i])*","~lambda==.(lambda[i])*","~theta==.(
-          theta[i])))
-  curve(digamma(x, theta[i], lambda[i], eta[i]), add=TRUE, 
-        col=col[2], n=1000)
-}
-legend("topright", legend=labels, lty=1, col=col, seg.len=1)
-par(opar)
-@
-
-	The variational posterior is given by:
-	\begin{align*}
-	  q(\bm{\beta}_d) & \overset{D}{=} \mathcal{N}_p 
-	  (\bm{\mu}_d, \bm{\Sigma}_d), \\
-	  q(\gamma_d^2) & \overset{D}{=} \mathcal{GIG}\(-\frac{p+\eta_d}{2}, 
-	  \lambda_d \theta_d^{-2}, \delta_d\), \\
-	  q(\sigma_d^2) & \overset{D}{=} \Gamma^{-1} \(\frac{n + p + 1}{2}, \zeta_d\).
-  \end{align*}
-	The variational parameters are the same as in the NIG model, except that now:
-  $b_d^{(h)} = (p + \eta_d)/\delta_d^{(h)}$.
-
-	The inclusion of continuous co-data is not so straightforward in the inverse
-	Gamma hyperprior model. We therefore focus on categorical co-data and use the 
-	following empirical Bayes procedure. Suppose we have a partitioning of our 
-	equations into classes and we let prior variance components $\gamma_d^2$ and
-	$\gamma^2_{d'}$ be drawn from the same distribution if equations $d$ and 
-	$d'$ are from the same class. Or simply put: $\gamma^2_{d}, \gamma^2_{d'} 
-	\sim \Gamma^{-1}(\eta_c/2, \lambda_c/2)$ if $d, d' \in \text{class}_c$. 
-	Empirical bayes estimation of the $\eta_c$ and $\lambda_c$ by maximisation of
-	the (approximated) marginal likelihood amounts to iteratively solving the
-	estimating equations:
-  \begin{subequations}
-    \begin{align}
-      \psi\(\frac{\eta_c}{2}\) & = \log \lambda_c - 
-      \frac{\sum_{d \in \text{class}_c} e^{(l)}_d}{|\text{class}_c|} - \log 2, 
-      \label{eq:estequation1}\\
-      \lambda_c & = \eta_c \cdot |\text{class}_c| \cdot 
-      \(\sum_{d \in \text{class}_c} b_d^{(l)}\)^{-1}, \label{eq:estequation2}
-    \end{align}
-  \end{subequations}
-  until convergence, where $e^{(l)}_d = \E_{Q^{(l)}}(\log \gamma_d^2) = 
-  \log \delta_d^{(l)} - \psi [ (p + \eta_d^{(l)})/2] - \log 2$. 
-  We propose to do so by first solving (\ref{eq:estequation2}) for $\lambda_c$, 
-  plugging it into (\ref{eq:estequation1}) and bounding the result using
-  \cite{alzer_inequalities_1997} to obtain an interval $( \eta_c^*, 2\eta_c^* )$
-  that contains the solution to (\ref{eq:estequation1}), where:
-  $$
-  \eta_c^* = \left\{ \frac{\sum_{d \in \text{class}_c} 
-  e^{(l)}_d}{|\text{class}_c|} + \log \[ \frac{\sum_{d \in \text{class}_c} 
-  b_d^{(l)}}{|\text{class}_c|} \]  \right\}^{-1}.
-  $$
-  With this interval it is straightforward to find $\eta^{(l+1)}_c$ by any 
-  root-finding algorithm and plugging the solution into (\ref{eq:estequation2})
-  to find $\lambda^{(l+1)}_c$. We may use $\eta_c^*$ as a starting value.
-
-	In the inverse Gamma setting, $\text{ELBO}^{(l + 1)} = 
-	\sum_{d=1}^D \text{ELBO}_d^{(l + 1)}$ after a EB update is given by:
-	\begin{align*}
-	  \text{ELBO}_d^{(l + 1)}  = & \frac{1}{2} \log |\bm{\Sigma}_d^{(l)}| -
-	  \frac{a_d^{(l)}}{2} \[ \y_d \tr \y_d - 2\y_d \tr \X \bm{\mu}_d^{(l)} +
-	  (\bm{\mu}_d^{(l)}) \tr \X \tr \X \bm{\mu}_d^{(l)} + \trace ( \X \tr \X 
-	  \bm{\Sigma}_d^{(l)})\] \\
-	  & + \frac{b_d^{(l)}}{2} \[ \delta_d^{(l)} - 
-	  a_d^{(l)}(\bm{\mu}_d^{(l)}) \tr \bm{\mu}_d^{(l)} - 
-	  a_d^{(l)} \trace (\bm{\Sigma}_d^{(l)}) \] - 
-	  \frac{n + p + 1}{2} \log \zeta_d^{(l)} \\
-	  & - \frac{p + \eta_d^{(l)}}{2} \log \delta_d^{(l)} + 
-	  \frac{\eta_d^{(l)}}{2} e_d^{(l)} + \frac{\eta_d^{(l)}}{2} \log 2 + 
-	  \log \Gamma \( \frac{p + \eta_d^{(l)}}{2} \) \\
-	  & - \log \Gamma \( \frac{\eta_d^{(l + 1)}}{2} \) - 
-	  \frac{\eta_d^{(l+1)}}{2} e_d^{(l)} - \frac{\eta_d^{(l+1)}}{2} \log 2 +
-	  \frac{\eta_d^{(l + 1)}}{2} \log \lambda_d^{(l+1)} - 
-	  \frac{\lambda_d^{(l + 1)} b_d^{(l)}}{2}.
-	\end{align*}
-  To compute the ELBO after a VB update we simply let 
-  $\eta_d^{(l)}, \lambda_d^{(l)}=\eta_d^{(l + 1)}, \lambda_d^{(l + 1)}$.
-
-	\section{Log pseudo marginal likelihood}\label{sec:lpml}
-	The log pseudo marginal likelihood is calculated as:
-	$\text{LPML}_d = n^{-1}\sum_{i=1}^n \log (\text{CPO}_{id})$ 
-  \cite[]{gelfand_model_1992}. Let 
-  $\theta_{id}=\x_i \tr \bbeta_d$ and $\y_{-id}$ denote $\y_{d}$ with 
-  observation $y_{id}$ excluded. Then the conditional 
-  predictive ordinates (CPO) are given by:
-  \begin{align*}
-    \text{CPO}_{id} & = p(y_{id} | \y_{-id}) = 
-    \int_{-\infty}^{\infty} \int_0^{\infty}
-    p(y_{id} | \bbeta_d, \sigma_d^2 ) \pi (\bbeta_d | \y_{-id}, \sigma_d^2) 
-    \pi (\sigma_d^2 | \y_{-id}) d\sigma_d^2 d\bbeta_d \\
-    & = \int_{-\infty}^{\infty} \int_0^{\infty}
-    p(y_{id} | \theta_{id}, \sigma_d^2 ) 
-    \pi (\theta_{id} | \y_{-id}, \sigma_d^2) 
-    \pi (\sigma_d^2 | \y_{-id}) d\sigma_d^2 d\theta_{id},
-  \end{align*}
-  where
-  \begin{align}
-    y_{id} | \theta_{id}, \sigma_d^2 & \sim \mathcal{N} (\theta_{id}, 
-    \sigma_d^2), \nonumber \\
-    \theta_{id} | \y_{-id} & \sim \mathcal{N}(\x_i \tr \bmu_{-id}, 
-    \x_i \tr \bSigma_{-id} \x_i), \nonumber \\
-    \sigma_d^2 | \y_{-id} & \sim \Gamma^{-1} \( \frac{n + p}{2},
-    \zeta_{-id} \), \label{eq:gammavb}
-  \end{align}
-  where $\bmu_{-id}$, $\bSigma_{-id}$, and $\zeta_{-id}$ are the variational
-  parameters estimated from $\y_{-id}$. Plugging in the densities we see that 
-  we have:
-  \begin{subequations}
-    \begin{align}
-      \text{CPO}_{id} = & (2 \pi)^{-1} \frac{\Gamma\(\frac{n + p + 1}
-      {2}\)}{\Gamma\(\frac{n + p}{2}\)}
-      (\zeta_{-id} \x_i \tr \bSigma_{-id} \x_i)^{-1/2} \label{eq:cpo1} \\
-      & \cdot \int_{-\infty}^{\infty}
-      \( \frac{\theta^2}{2\zeta_{-id}} + 1 \)^{-\frac{n + p + 1}{2}}
-      \exp \[ - \frac{(\theta - \x_i \tr \bmu_{-id} + y_i)^2}
-      {2 \x_i \tr \bSigma_{-id} \x_i} \] d\theta, \label{eq:cpo2}
-    \end{align}
-  \end{subequations}
-  a one-dimensional (well-behaved) integral that is easily solved with some 
-  numerical integration routine.
-
-  In practice we would like to avoid re-estimating the model for every 
-  observation $i$. Instead we divide the observations into $K$ parts and 
-  estimate $K$ models, so that now $\y_{-id}$ denotes $\y$ with all observations
-  in part $k$ removed. Additionally, we add $1 - n_{id}$ to the numerator of the
-  Gamma distribution in (\ref{eq:gammavb}), the numerators in the Gamma 
-  functions in (\ref{eq:cpo1}), and the numerator of the power in
-  (\ref{eq:cpo2}), where $n_{id}$ denotes the number of elements in the removed 
-  part.
-
-% 	\section{Non-conjugate model}
-% 	The non-conjugate model assumes $\beta_{jd}$ and $\sigma_d^2$ 
-% 	independent \textit{a priori}:
-%   $$
-%   \beta_{jd} | \gamma_d^2, \sigma_d^2 \sim \beta_{jd} | 
-%   \gamma_d^2 \sim \mathcal{N} (0, \gamma_d^2).
-%   $$
-%   
-% 	The variational distributions are as follows:
-% 	\begin{align*}
-% 	  q(\bm{\beta}_d) & \overset{D}{=} \mathcal{N}_p 
-% 	  (\bm{\mu}_d, \bm{\Sigma}_d), \\
-% 	  q(\gamma_d^2) & \overset{D}{=} \mathcal{GIG}
-% 	  \(-\frac{p+1}{2}, \lambda_d \theta_d^{-2}, \delta_d\), \\
-% 	  q(\sigma_d^2) & \overset{D}{=} \Gamma^{-1} 
-% 	  \(\frac{n + 1}{2}, \zeta_d\),
-%   \end{align*}
-% 	The variational parameters are iteratively updated by:
-%   \begin{align*}
-%   \bm{\Sigma}_d^{(h+1)} & = (a_d^{(h)})^{-1} (\X \tr \X + c_d^{(h)} \I)^{-1}, \\
-%   \bm{\mu}_d^{(h+1)} & = (\X \tr \X + c_d^{(h)} \I)^{-1} \X \tr \y_d, \\
-%   \delta_d^{(h+1)} & = b_d^{(h)} (c_d^{(h)})^{-1} \bmu_d \tr \bmu_d + \lambda_d, \\ 
-%   \zeta_d^{(h+1)} & = \frac{1}{2} \bigg[\mathbf{y}_d \tr \mathbf{y}_d -2 \mathbf{y}_d \tr \X \bm{\mu}_d^{(h+1)} + \trace ( \X \tr \X \bm{\Sigma}_d^{(h+1)}) + (\bm{\mu}_d^{(h+1)}) \tr \X \tr \X \bm{\mu}_d^{(h+1)} \\
-%   & \,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\, + \frac{b_d^{(h+1)} - c_d^{(h+1)} a_d^{(h+1)}}{1 - a_d^{(h+1)}} \trace ( \bm{\Sigma}_d^{(h+1)}) + \frac{b_d^{(h+1)} - c_d^{(h+1)} a_d^{(h+1)}}{1 - a_d^{(h+1)}} (\bm{\mu}_d^{(h+1)}) \tr \bm{\mu}_d^{(h+1)}\bigg].
-%   \end{align*}
-%   Here, we set $a_d^{(h)}$, $b_d^{(h)}$, and $c_d^{(h)}$ as in 
-%   (\ref{eq:auxvar1})
-%   Note that in the inverse Gamma model, the computation of (\ref{eq:auxvar1})
-%   simplifies to $b_d^{(h)}=(p+\eta_d)/\delta_d^{(h)}$.
-%   
-%   Here, we set
-%   \begin{align}
-%   a_d^{(h)} & =\E_{Q^{(h)}}(\sigma_d^{-2})=(df+1)/(2 \zeta_d^{(h)}), \nonumber \\
-%   b_d^{(h)} & = \E_{Q^{(h)}}(\gamma_d^{-2}) = \sqrt{\frac{\lambda_d}{\theta_d^2 \delta_d^{(h)}}} \frac{K_{\frac{p + \eta_d -2}{2}} \( \sqrt{\lambda_d \delta_d^{(h)}}/\theta_d \)}{K_{\frac{p+\eta_d}{2}} \( \sqrt{\lambda_d \delta_d^{(h)}}/\theta_d \)} + \frac{p+\eta_d^{(h)}}{\delta_d^{(h)}}, \label{eq:auxvar1} \\
-%   c_d^{(h)} & = \begin{cases*}
-%   b_d^{(h)} & in the conjugate setting \\
-%   b_d^{(h)}/a_d^{(h)} & in the non-conjugate setting .
-%   \end{cases*} \nonumber
-%   \end{align}
-%   Note that in the inverse Gamma model, the computation of (\ref{eq:auxvar1}) simplifies to $b_d^{(h)}=(p+\eta_d)/\delta_d^{(h)}$.
-%   
-%   \section{Starting values}
-%   \subsection{Method 1}
-%   Starting value for $\sigma_d^2$ as in \cite{chipman_bart:_2010}.
-They consider a scaled-Chi squared distribution for the error variance with degrees of freedom 3 and scale chosen such that the $q*100$th percentile matches the error variance in the data $s^2(\mathbf{y}_d)$.
-As starting value for $\sigma_d^2$ we take the mode of this distribution: $\hat{\sigma}_d^2 = 3a/5$, with $a$ the solution to:
-%   $$
-%   \Gamma\(\frac{3}{3}\) \[1 - q - F(3a/(2s^2(\mathbf{y}_d)); 3/2, 1) \] = 0,
-%   $$
-%   where $F(x; 3/2, 1)$ is the CDF of a gamma function with shape $3/2$ and scale $1$.
-In \cite{moran_variance_2018}, REFERENCE to Rockova and George (2018), and \cite{chipman_bart:_2010}, $q=0.9$ is suggested.
-We find that $q \in ( 0.9, 0.95 )$ gives good results.
-%   
-%   \subsection{Method 2}
-%   We generate starting values as follows. Consider the $\sigma_d^2$ and $\gamma_d^2$ as fixed and estimate them by MML of the regular ridge model:
-%   $$
-%   \hat{\sigma}_d^2, \hat{\gamma}_d^2= \underset{\sigma_d^2,\gamma_d^2}{\argmax} \int_{\bm{\beta}} p(\mathbf{y}_d | \bm{\beta}, \sigma_d^2) p(\bm{\beta} | \gamma_d^2, \sigma_d^2) d\bm{\beta}.
-%   $$
-%   The regular ridge model is conjugate, so MML is relatively simple. Next we estimate a common inverse Gaussian prior for the $\gamma_d^2$ by considering the estimates to be samples from the prior:
-%   \begin{align*}
-%   \hat{\theta} & = n^{-1} \sum_{d=1}^D \hat{\gamma}_d^2 / n, \\
-%   \hat{\lambda} & = n \left\{\sum_{d=1}^D \[ (\hat{\gamma}_d^2)^{-1} - \hat{\theta}^{-1}\]\right\}^{-1}.
-%   \end{align*}
-%   We estimate the mode of the $\hat{m}=\hat{\gamma}_d^2$ by kernel density estimation and equate it to the theoretical mode of the generalized inverse Gaussian distribution with the estimated $\hat{\theta}$ and $\hat{\lambda}$.
-We solve to find a common $\delta$:
-%   $$
-%   \hat{\delta} = \frac{\hat{m}^2 \hat{\lambda}}{\hat{\theta}^2} + (p + 3) \hat{m}.
-%   $$
-%   Lastly, we consider the $\hat{\sigma}^2_d$ fixed samples from the inverse Gamma distribution to estimate a common scale:
-%   $$
-%   \hat{\zeta} = \frac{D(n + p + 1)}{2 \sum_{d=1}^D[(\hat{\sigma}_d^2)^{-1}]}.
-%   $$
-
   \bibliographystyle{author_short3}  
@@
 
+	\section*{Session info}
+
+<<r session-info, eval=TRUE, echo=TRUE, tidy=TRUE>>=
+devtools::session_info()
+@
+
 \end{document}

### 2019-11-19 docs/supplement2.Rnw
redoing simulations

@@
 \usepackage{bm,amsmath,amssymb,amsthm,amsfonts,graphics,graphicx,epsfig,
-rotating,caption,subcaption,natbib,appendix,titlesec,multicol,verbatim,
+rotating,caption,subcaption,natbib,appendix,titlesec,multicol,hyperref,verbatim,
 bbm,algorithm,algpseudocode,pgfplotstable,threeparttable,booktabs,mathtools,
-dsfont,parskip,xr-hyper,hyperref}
+dsfont,parskip,xr}
 \externaldocument[md-]{manuscript2}
@@
 	\section{Content overview}
-	This document contains the Supplementary Material (SM) to the document `Drug
-	efficacy prediction in cell lines'. In the following, this document is
-	referred to as Main Document (MD).
+	This document contains the Supplementary Material (SM) to the document 
+	`Empirical Bayes estimation of simultaneous-equation models'. In the 
+	following, this document is referred to as Main Document (MD).
 
-	\section{Unpenalized covariates}
-  Inclusion of an intercept $\beta_0$ and other unpenalized covariates is 
-  achieved by letting $\X = \begin{bmatrix}\X_u & 
-  \X_r\end{bmatrix}$, where $\X_u$ are the $u$
-  unpenalized variables (possibly with a column of ones if an intercept is
-  to be included) and $\X_r$ are the $r$ penalized variables. 
-  Additionally, we let 
+	\section{NIG and Student's \texorpdfstring{$t$}{t} Priors}\label{sec:priors}
+	\subsection{Normal inverse Gaussian}\label{sec:nigprior}
+	Let $\theta = (\mathbf{c} \tr \bm{\alpha})^{-1}$. Then, 
+	for a fixed error variance $\sigma^2$, we have marginal prior variance 
+	and kurtosis:
+	\begin{align*}
+	  \V(\beta | \sigma^2) & = \sigma^2 \theta, \\
+	  \mathcal{K}(\beta | \sigma^2) & = \frac{3\theta}{\lambda} + 3.
+	\end{align*}
+
+	\subsection{Student's \texorpdfstring{$t$}{t} prior}\label{sec:studentstprior}
+	For a fixed error variance $\sigma^2$, we have marginal prior variance 
+	and kurtosis:
+	\begin{align*}
+	  \V(\beta | \sigma^2) & = \frac{\sigma^2 \xi}{\eta - 2}, \\
+	  \mathcal{K}(\beta | \sigma^2) & = \frac{6}{\eta - 4} + 3.
+	\end{align*}
+
+	\section{Variational Bayes derivations}\label{sec:vbderivations}
+	For clarity we have indexed the variational 
+	density functions with their respective parameters, which we omitted in the 
+	MD. In the following all expectations are with respect to the 
+	variational posterior $Q_d$. The variational posterior for $\bbeta_d$ is 
+	found as follows:
+	\begin{align*}
+	  \log q_{\bbeta_d} (\bbeta_d) & \propto \E [\log \mathcal{L}
+	  (\y_d | \bbeta_d, \sigma_d^2 )] + \E [\log \pi (\bbeta_d | \bgamma_{d}^2, 
+	  \sigma_d^2)] \\
+	  & \propto - \frac{1}{2} \sum_{i=1}^n 
+	  \E \[ \frac{(y_{id} - \x_i \tr \bbeta_d)^2}{\sigma_d^2} \] - 
+	  \frac{1}{2} \sum_{j=1}^p 
+	  \E \( \frac{\beta_{jd}^2}{\gamma_{jd}^2 \sigma_d^2} \), \\
+	  q_{\bbeta_d} (\bbeta_d) & \overset{D}{=} 
+	  \mathcal{N}_p (\bmu_d, \bSigma_d), \\
+	  & \text{with } \bSigma_d = \E(\sigma_d^{-2})^{-1} 
+	  \{\X \tr \X + \diag [\E(\gamma_{jd}^{-2})] \}^{-1}, \\
+	  & \text{and } \bmu_d = \{\X \tr \X + \diag [\E(\gamma_{jd}^{-2})] \}^{-1} 
+	  \X \tr \y_d.
+	\end{align*}
+	The variational posterior for $\gamma_{jd}^2$ is given by:
+	\begin{align*}
+	  \log q_{\gamma_{jd}^{2}}(\gamma_{jd}^{2}) & \propto \E 
+	  [\log \pi (\beta_{jd} | \gamma_{jd}^{2}, \sigma_d^2)] + 
+	  \log \pi(\gamma_{jd}^{2}) \\
+	  & \propto -\frac{1}{2} \log \gamma_{jd}^{2} - \frac{1}{2} 
+	  \E \( \frac{\beta_{jd}^2}{\sigma^{2}_d}\) \gamma_{jd}^{-2} - 
+	  \frac{3}{2} \log \gamma_{jd}^{2} - 
+	  \frac{\lambda_d \{\gamma_{jd}^2 - 
+	  [(\mathbf{c}_{d}^j) \tr \bm{\alpha}]^{-1}\}^2}
+	  {2[(\mathbf{c}_{d}^j) \tr \bm{\alpha}]^{-2}} \gamma_{jd}^{-2} \\
+	  & \propto \( -1 - 1\) \log \gamma_{jd}^{2} - 
+	  \frac{\lambda_d[(\mathbf{c}_{d}^j) \tr \bm{\alpha}]^{2}}{2} \gamma_{jd}^2 - 
+	  \frac{1}{2}\[ \lambda_d + \E(\sigma_d^{-2}) \E(\beta_{jd}^2)\] 
+	  \gamma_{jd}^{-2} , \\
+	  q_{\gamma_{jd}^{2}}(\gamma_{jd}^{2}) & \overset{D}{=} \mathcal{GIG} 
+	  \(-1, \lambda_d[(\mathbf{c}_{d}^j) \tr \bm{\alpha}]^{2}, \delta_{jd} \), \\
+	  & \text{with } \delta_{jd} =\E(\sigma_d^{-2})
+	  \[\E(\beta_{jd})^2 + \V(\beta_{jd})\] + \lambda_d.
+	\end{align*}
+	Lastly, we have the variational posterior for $\sigma_d^2$:
+	\begin{align*}
+	  \log q_{\sigma_d^{2}}(\sigma_d^{2}) & \propto \E 
+	  [\log \mathcal{L}(\y_d | \bbeta)d, \sigma_d^2 )] + 
+	  \E [\log \pi (\bbeta_d | \bgamma_d^2, \sigma_d^2)] + 
+	  \log \pi (\sigma_d^{2}) \\
+	  & \propto -\frac{n}{2} \log \sigma_d^{2} - \frac{1}{2} \sum_{i=1}^n 
+	  \frac{\E [(y_{id} - \x_i \tr \bbeta_d)^2]}{\sigma_d^{2}} - 
+	  \frac{p}{2} \log \sigma_d^{2} - \frac{1}{2} \sum_{j=1}^p 
+	  \E \( \frac{\beta_{jd}^2)}{\gamma_{jd}^{2}}\) \sigma_d^{-2} - 
+	  \frac{3}{2} \log \sigma^2_d \\
+	  & = \(-\frac{n + p + 1}{2} - 1 \) \log \sigma_d^{2} - 
+	  \frac{1}{2}\left\{ \sum_{i=1}^n \E [(y_{id} - \x_i \tr \bbeta_d)^2] + 
+	  \sum_{j=1}^p \E(\gamma_{jd}^{-2})\E(\beta_{jd}^2) \right\} \sigma_d^{-2}, \\
+	  q_{\sigma_d^{2}}(\sigma_d^{2}) & \overset{D}{=} 
+	  \Gamma^{-1} \(\frac{n + p + 1}{2}, \zeta_{d} \), \\
+	  & \text{with } \zeta_{d} = \frac{1}{2} \bigg\{\mathbf{y}_d \tr 
+	  \mathbf{y}_d - 2 \mathbf{y}_d \tr \X \E(\bbeta_d) + 
+    \trace [ \X \tr \X \V(\bbeta_d)] + 
+    \E(\bbeta_d \tr)  \X \tr \X \E(\bbeta_d ) \\
+    & \,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\, + 
+    \trace \{ \diag[\E(\gamma_{jd}^{-2})] \V(\bbeta_d)\} + 
+     \E(\bbeta_d \tr) \diag[\E(\gamma_{jd}^{-2})] \E(\bbeta_d ) \bigg\}.
+	\end{align*}
+
+% 	\section{Evidence lower bound}\label{sec:evidencelowerbound}
+% 	The variational Bayes posterior is found through maximisation of an evidence
+% 	lower bound (ELBO). Additionally, we may monitor convergence through this 
+% 	ELBO. To compute the ELBO after updating the EB parameters from 
+% 	$\theta_d^{(l)}, \lambda_d^{(l)}$ to 
+% 	$\theta_d^{(l + 1)}, \lambda_d^{(l + 1)}$, we compute 
+% 	$\text{ELBO}^{(l + 1)} = \sum_{d=1}^D \text{ELBO}_d^{(l + 1)}$, with:
+% 	\begin{align*}
+% 	  \text{ELBO}_d^{(l + 1)} = & \frac{1}{2} \log |\bm{\Sigma}_d^{(l)}| - 
+% 	  \frac{a_d^{(l)}}{2} \[ \y_d \tr \y_d - 2\y_d \tr \X \bm{\mu}_d^{(l)} + 
+% 	  (\bm{\mu}_d^{(l)}) \tr \X \tr \X \bm{\mu}_d^{(l)} + 
+% 	  \trace ( \X \tr \X \bm{\Sigma}_d^{(l)})\] \\
+% 	  & + \frac{b_d^{(l)}}{2} \[ \delta_d^{(l)} - 
+% 	  a_d^{(l)}(\bm{\mu}_d^{(l)}) \tr \bm{\mu}_d^{(l)} - 
+% 	  a_d^{(l)} \trace (\bm{\Sigma}_d^{(l)}) \] \\
+% 	  & - \frac{p + n + 1}{2} \log \zeta_d^{(l)} + 
+% 	  \frac{p + 1}{4} \log \lambda_d^{(l)} - 
+% 	  \frac{p + 1}{2} \log \theta_d^{(l)} - \frac{p+1}{4} \log \delta_d^{(l)} \\
+% 	  & + \log K_{\frac{p+1}{2}}\(\sqrt{\lambda_d^{(l)} 
+% 	  \delta_d^{(l)}}/\theta_d^{(l)}\) + 
+% 	  \frac{\lambda_d^{(l)} e_d^{(l)}}{2(\theta_d^{(l)})^2} \\
+% 	  & + \frac{\lambda_d^{(l+1)}}{\theta_d^{(l+1)}} + 
+% 	  \frac{1}{2}\log \lambda_d^{(l+1)} - \frac{\lambda_d^{(l+1)} 
+% 	  e_d^{(l)}}{2 (\theta_d^{(l + 1)})^2} - 
+% 	  \frac{\lambda_d^{(l + 1)} b_d^{(l)}}{2},
+% 	\end{align*}
+% 	where $a^{(l)}$, $b^{(l)}$, and $e^{(l)}$ are as in the MD. To compute the 
+% 	ELBO after a VB update we simply let 
+%   $\theta_d^{(l)}, \lambda_d^{(l)}=\theta_d^{(l + 1)}, \lambda_d^{(l + 1)}$.
+% 	 
+%   In practice, there are different options to assess convergence. We may monitor
+%   the parameters themselves, the ELBO, or a combination of the two. We may also 
+%   iterate for an \textit{a priori} fixed number of times. We may also choose to 
+%   use a different convergence criterium for the EB and VB iterations. Currently 
+%   we fix the number of EB iterations to 20, while we do 2 VB iterations per EB 
+%   iteration.
+
+  \section{Efficient computation}
+  The empirical Bayes updates require the quantities
+  $e_{jd}$ and $b_{jd}$, which in turn depend on the quantities 
+  $\diag(\bSigma_d)$, $\bmu_{d}$, $\y_d \tr \X \bmu_d$, 
+  $\trace(\X \tr \X \bSigma_d)$, $\bmu_d \tr \X \tr \X \bmu_d$,
+  $\trace[\diag(b_{jd})\bSigma_d]$, and $\bmu_d \tr \diag(b_{jd}) \bmu_d$.
+  Let $\mathbf{H}_d=\diag[(b_{jd}^{(h)})^{-1}]$. Then, the first two quantities 
+  are more efficiently calculated as:
+  \begin{subequations}\label{eq:complexity1}
+    \begin{align}
+      \diag (\bm{\Sigma}_d^{(h+1)}) & = (a_d^{(h)})^{-1} 
+      \H_d^{(h)} - (a_d^{(h)})^{-1} \{ [ \H_d^{(h)} \X \tr 
+      ( \I_n + \X \H_d^{(h)} \X \tr )^{-1} ] \circ
+      (\H_d^{(h)} \X \tr ) \} \mathbf{1}_{n \times 1}, \\
+      \bmu_d^{(h+1)}) & = \H_d^{(h)} \X \tr \y_d - \H_d^{(h)} \X \tr 
+      ( \I_n + \X \H_d^{(h)} \X \tr )^{-1} \X \H_d^{(h)} \X \tr \y_d.
+    \end{align}
+  \end{subequations}
+  The quantities $\y_d \tr \X \bmu_d$, $\bmu_d \tr \X \tr \X \bmu_d$,
+  $\trace[\diag(b_{jd})\bSigma_d]$, and $\bmu_d \tr \diag(b_{jd}) \bmu_d$ are
+  easily calculated from the first two. For the remaining one we have:
+  \begin{align}\label{eq:complexity2}
+    \trace(\X \tr \X \bSigma_d^{(h+1)})  = 
+    & (a_d^{(h)})^{-1} \mathbf{1}_{1 \times p} \H_d^{(h)} \X \tr 
+    \mathbf{1}_{n \times 1} \nonumber \\  
+    & - (a_d^{(h)})^{-1} \mathbf{1}_{1 \times p} \{ [  \H_d^{(h)} \X \tr 
+    ( \I_n + \X \H_d^{(h)} \X \tr )^{-1} \X \H_d^{(h)} \X \tr ] \circ 
+    \X \tr \} \mathbf{1}_{n \times 1}.
+  \end{align}
+  Both (\ref{eq:complexity1}) and (\ref{eq:complexity2}) are operations of
+  complexity $\mathcal{O}(pn^2)$.
+
+  \section{Ratios of modified Bessel functions}\label{sec:ratiosmodifiedbessels}
+  Ratios of modified Bessel functions of the second kind 
+  $K_{\alpha - 1}(x)/K_{\alpha}(x)$ are prone to under- and overflow for large
+  $\alpha$. In our case, $\alpha$ increases linearly with $p$. Since $p$ may be
+  large, this may cause numerical issues in the calculation of various 
+  quantities. We alleviate the numerical issues through the following.
+
+  We let $n_1=p/2$ and $n_2=(p-1)/2$ and use the well-known recursive relation:
   $$
-  \mathbf{H} = 
-  \begin{bmatrix}
-  \mathbf{0}_{u \times u} & \mathbf{0}_{u \times r} \\
-  \mathbf{0}_{r \times u} & \diag[\E(\gamma_{jd}^{-2})]. \\
-  \end{bmatrix}
+  K_{\alpha}(x) = K_{\alpha - 2}(x) + \frac{2(\alpha - 1)}{2} K_{\alpha- 1}(x),
   $$
-  Then,
-  \begin{align}\label{eq:unpencov}
-    \bSigma &= \E(\sigma_d^{-2})^{-1}\(\X \X + \mathbf{H} \)^{-1} \nonumber \\
-    & = 
-    \E(\sigma_d^{-2})^{-1}\begin{bmatrix} 
-      \X_u \tr \X_u & \X_u \tr \X_r \\
-      \X_r \tr \X_u & \X_r \tr \X_r + \diag[\E(\gamma_{jd}^{-2})]
-    \end{bmatrix}^{-1}.
+  to rewrite the ratio:
+  \begin{align*}
+    \frac{K_{\frac{p - 1}{2}}(x)}{K_{\frac{p + 1}{2}}(x)} = 
+    \begin{cases}
+      \Big( \dots \Big( \( 1 + \frac{2 \cdot 1 - 1}{x} \)^{-1} + 
+      \frac{2 \cdot 2-1}{x}\Big)^{-1} + \dots + 
+      \frac{2 \cdot n_1 - 1}{x}\Big)^{-1}, & \text{for } p \text{ even}, \\
+      \Big( \dots \Big( \( \frac{K_0(x)}{K_1(x)} + \frac{2}{x} \cdot 1 \)^{-1} +
+      \frac{2}{x} \cdot 2 \Big)^{-1} + \dots + \frac{2}{x} \cdot n_2 \Big)^{-1},
+      & \text{for } p \text{ odd}.
+    \end{cases}
+  \end{align*} 
+  The ratio $K_0(x)/K_1(x)$ is well-behaved, such that the ratio may be computed
+  without numerical issues.
+
+  \section{EB updates}
+  For the scale parameter $\lambda_d$ updates to be non-negative, we have to 
+  show that $\forall j$:
+  \begin{equation}\label{eq:inequality}
+    b_{jd}^{(l)} + e_{jd} [(\mathbf{c}_d^j) \tr \bm{\alpha}]^2 - 
+    2[(\mathbf{c}_d^j) \tr \bm{\alpha}] = 
+    \E_{Q^{(l)}}(\gamma_{jd}^{-2}) + \E_{Q^{(l)}}(\gamma_{jd}^{2}) 
+    [(\mathbf{c}_d^j) \tr \bm{\alpha}]^2 - 
+    2[(\mathbf{c}_d^j) \tr \bm{\alpha}] \geq 0.
+  \end{equation}
+  We first note that through Jensen's inequality 
+  $\E_{Q^{(l)}}(\gamma_{jd}^{2})^{-1} \leq \E_{Q^{(l)}}(\gamma_{jd}^{-2})$,
+  we may lowerbound the left-hand side of (\ref{eq:inequality}):
+  $$
+  \E_{Q^{(l)}}(\gamma_{jd}^{-2}) + \E_{Q^{(l)}}(\gamma_{jd}^{2}) 
+  [(\mathbf{c}_d^j) \tr \bm{\alpha}]^2 - 
+  2[(\mathbf{c}_d^j) \tr \bm{\alpha}] \geq 
+  \E_{Q^{(l)}}(\gamma_{jd}^{2})^{-1} + \E_{Q^{(l)}}(\gamma_{jd}^{2}) 
+  [(\mathbf{c}_d^j) \tr \bm{\alpha}]^2 - 
+  2[(\mathbf{c}_d^j) \tr \bm{\alpha}], 
+  $$
+  which we use to show that:
+  \begin{align*}
+  \E_{Q^{(l)}}(\gamma_{jd}^{2})^{-1} + \E_{Q^{(l)}}(\gamma_{jd}^{2}) 
+  [(\mathbf{c}_d^j) \tr \bm{\alpha}]^2 - 
+  2[(\mathbf{c}_d^j) \tr \bm{\alpha}] & \geq 0 \\
+  1 + \E_{Q^{(l)}}(\gamma_{jd}^{2})^2 
+  [(\mathbf{c}_d^j) \tr \bm{\alpha}]^2 - 
+  2 \E_{Q^{(l)}}(\gamma_{jd}^{2}) [(\mathbf{c}_d^j) \tr \bm{\alpha}] & \geq 0 \\
+  \left\{ 1 - \E_{Q^{(l)}}(\gamma_{jd}^{2}) [(\mathbf{c}_d^j) \tr \bm{\alpha}]
+  \right\}^2 \geq 0.
+  \end{align*}
+
+  \section{Gibbs sampler}\label{sec:gibbssampler}
+  MCMC samples from the posterior corresponding to model 
+  (\ref{md-eq:priormodel}) may be generated for each equation independently. 
+  MCMC samples may be generated by iteratively sampling the following 
+  conditional distributions:
+  \begin{subequations}
+    \begin{align}
+      \bm{\beta}_d | \gamma_d^2, \sigma_d^2, \y_d & \sim \mathcal{N}_p 
+      (\bm{\mu}_d, \bm{\Sigma}_d), \label{eq:betasample} \\
+      \text{with } & \bm{\Sigma}_d = \sigma_d^2 (\X \tr \X + 
+      \gamma_d^{-2} \I)^{-1} \label{eq:sigmasample} \\
+      \text{and } & \bm{\mu}_d = (\X \tr \X + 
+      \gamma_d^{-2} \I)^{-1} \X \tr \y_d \label{eq:musample} \\
+      \gamma_d^2 | \bm{\beta}_d, \sigma_d^2, \y & \sim \mathcal{GIG}
+      \(-\frac{p+1}{2}, \lambda_d \theta_d^{-2}, \delta_d\), \nonumber \\
+      \text{with } & \delta_d = \sigma_d^{-2} \bm{\beta}_d \tr \bm{\beta}_d + 
+      \lambda_d, \nonumber \\
+      \sigma_d^2 | \bm{\beta}_d, \gamma_d^2, \y & \sim \Gamma^{-1} 
+      \(\frac{n + p + 1}{2}, \zeta_d\), \nonumber \\
+      \text{with } & \zeta = \frac{1}{2} \[\mathbf{y} \tr \mathbf{y} -
+      2 \mathbf{y} \tr \X \bm{\mu} + \trace ( \X \tr \X \bm{\Sigma}) + 
+      (\bm{\mu}) \tr \X \tr \X \bm{\mu}  + 
+      \gamma^{-2} \trace ( \bm{\Sigma}) + 
+      \gamma^{-2} (\bm{\mu}) \tr \bm{\mu}\]. \nonumber
     \end{align}
-  With the choice of blocks as in (\ref{eq:unpencov}), blockwise inversion
-  renders the largest required matrix inverse $(\X_r \tr \bOmega \X_r +
-  \diag[\E(\gamma_{jd}^{-2})])^{-1}$, of dimension 
-  $r \times r$. Inversion of this matrix is done efficiently by applying the
-  Woodbury matrix identity.
+  \end{subequations}
+  In high dimensional space, the $p \times p$ matrix inversions in 
+  (\ref{eq:sigmasample}) and (\ref{eq:musample}) are   
+  significant computational bottlenecks. \cite{bhattacharya_fast_2016} describe 
+  a method to sample from (\ref{eq:betasample}) without explicit calculation
+  of this inverse, thereby offering a siginifcant speed-up compared to naive
+  sampling.
 
-  \section{Starting values}
-  \subsection{Method 1}
-  Starting value for $\sigma_d^2$ as in \cite{chipman_bart:_2010}.
+  \section{Inverse Gamma model}\label{sec:inversegamma}
+  The model introduced in MD Section \ref{md-sec:inversegammamodel} draws the 
+  $\gamma_d^2$ from an inverse Gamma distribution:
+	\begin{equation}\label{eq:invgammaprior}
+	  \gamma_d^2 \sim \Gamma^{-1}(\eta_d/2, \lambda_d/2)=
+	  \mathcal{GIG}(-\eta_d/2, 0, \lambda_d),
+	\end{equation}
+	with shape $\eta_d/2$ and scale $\lambda/2$. For the sake of comparability, we
+	have rescaled the inverse Gamma parameters and reparametrised as a generalized
+	inverse Gaussian. This allows for an easier comparison to the inverse 
+	Gaussian hyperprior:
+	\begin{equation}\label{eq:invgaussianprior}
+	  \mathcal{IG}(\theta_d, \lambda_d) =
+	  \mathcal{GIG}(-1/2, \lambda_d \theta_d^{-2}, \lambda_d).
+  \end{equation}
+	From (\ref{eq:invgammaprior}) and (\ref{eq:invgaussianprior}) it is obvious 
+	that the two models coincide for $\theta_d \to \infty$ and $\eta_d = 1$. In
+	Figure \ref{fig:dens_igaussian_igamma} the two distributions are compared for 
+	a range of parameter values.
+<<dens_igaussian_igamma, fig.cap="Several densities of the inverse Gaussian and inverse Gamma families", out.width="100%", fig.asp=2/3>>=
+digauss <- function(x, theta, lambda, eta) {
+  sqrt(lambda/(x^3*2*pi))*exp(-lambda*(x - theta)^2/(2*theta^2*x))
+}
+
+digamma <- function(x, theta, lambda, eta) {
+  alpha <- eta/2
+  beta <- lambda/2
+  beta^alpha/gamma(alpha)*x^(-alpha - 1)*exp(-beta/x)
+}
+
+eta <- c(1, 2, 1, 1, 5)
+lambda <- c(1, 1, 2, 1/2, 2)
+theta <- c(1000, 1000, 1, 1, 1)
+
+labels <- c("inv. Gaussian", "inv. Gamma")
+col <- c(1, 2)
+
+opar <- par(no.readonly=TRUE)
+par(mar=opar$mar*c(1, 1.3, 1, 1))
+layout(matrix(c(rep(c(1, 1, 2, 2, 3, 3), 2), rep(c(0, 4, 4, 5, 5, 0), 2)),
+                 nrow=4, ncol=6, byrow=TRUE))
+for(i in 1:length(eta)) {
+  modes <- c(theta[i]*(sqrt(1 + 9*theta[i]^2/
+                            (4*lambda[i]^2)) - 3*theta[i]/(2*lambda[i])),
+             lambda[i]/(eta[i] + 2))
+  ylim <- c(0, max(digauss(modes[1], theta[i], lambda[i], eta[i]),
+                   digamma(modes[2], theta[i], lambda[i], eta[i])))
+  curve(digauss(x, theta[i], lambda[i], eta[i]), 0.001, 5, n=1000,
+        ylim=ylim, ylab="Density", "x", col=col[1], 
+        main=bquote(eta==.(eta[i])*","~lambda==.(lambda[i])*","~theta==.(
+          theta[i])))
+  curve(digamma(x, theta[i], lambda[i], eta[i]), add=TRUE, 
+        col=col[2], n=1000)
+}
+legend("topright", legend=labels, lty=1, col=col, seg.len=1)
+par(opar)
+@
+
+	The variational posterior is given by:
+	\begin{align*}
+	  q(\bm{\beta}_d) & \overset{D}{=} \mathcal{N}_p 
+	  (\bm{\mu}_d, \bm{\Sigma}_d), \\
+	  q(\gamma_d^2) & \overset{D}{=} \mathcal{GIG}\(-\frac{p+\eta_d}{2}, 
+	  \lambda_d \theta_d^{-2}, \delta_d\), \\
+	  q(\sigma_d^2) & \overset{D}{=} \Gamma^{-1} \(\frac{n + p + 1}{2}, \zeta_d\).
+  \end{align*}
+	The variational parameters are the same as in the NIG model, except that now:
+  $b_d^{(h)} = (p + \eta_d)/\delta_d^{(h)}$.
+
+	The inclusion of continuous co-data is not so straightforward in the inverse
+	Gamma hyperprior model. We therefore focus on categorical co-data and use the 
+	following empirical Bayes procedure. Suppose we have a partitioning of our 
+	equations into classes and we let prior variance components $\gamma_d^2$ and
+	$\gamma^2_{d'}$ be drawn from the same distribution if equations $d$ and 
+	$d'$ are from the same class. Or simply put: $\gamma^2_{d}, \gamma^2_{d'} 
+	\sim \Gamma^{-1}(\eta_c/2, \lambda_c/2)$ if $d, d' \in \text{class}_c$. 
+	Empirical bayes estimation of the $\eta_c$ and $\lambda_c$ by maximisation of
+	the (approximated) marginal likelihood amounts to iteratively solving the
+	estimating equations:
+  \begin{subequations}
+    \begin{align}
+      \psi\(\frac{\eta_c}{2}\) & = \log \lambda_c - 
+      \frac{\sum_{d \in \text{class}_c} e^{(l)}_d}{|\text{class}_c|} - \log 2, 
+      \label{eq:estequation1}\\
+      \lambda_c & = \eta_c \cdot |\text{class}_c| \cdot 
+      \(\sum_{d \in \text{class}_c} b_d^{(l)}\)^{-1}, \label{eq:estequation2}
+    \end{align}
+  \end{subequations}
+  until convergence, where $e^{(l)}_d = \E_{Q^{(l)}}(\log \gamma_d^2) = 
+  \log \delta_d^{(l)} - \psi [ (p + \eta_d^{(l)})/2] - \log 2$. 
+  We propose to do so by first solving (\ref{eq:estequation2}) for $\lambda_c$, 
+  plugging it into (\ref{eq:estequation1}) and bounding the result using
+  \cite{alzer_inequalities_1997} to obtain an interval $( \eta_c^*, 2\eta_c^* )$
+  that contains the solution to (\ref{eq:estequation1}), where:
+  $$
+  \eta_c^* = \left\{ \frac{\sum_{d \in \text{class}_c} 
+  e^{(l)}_d}{|\text{class}_c|} + \log \[ \frac{\sum_{d \in \text{class}_c} 
+  b_d^{(l)}}{|\text{class}_c|} \]  \right\}^{-1}.
+  $$
+  With this interval it is straightforward to find $\eta^{(l+1)}_c$ by any 
+  root-finding algorithm and plugging the solution into (\ref{eq:estequation2})
+  to find $\lambda^{(l+1)}_c$. We may use $\eta_c^*$ as a starting value.
+
+	In the inverse Gamma setting, $\text{ELBO}^{(l + 1)} = 
+	\sum_{d=1}^D \text{ELBO}_d^{(l + 1)}$ after a EB update is given by:
+	\begin{align*}
+	  \text{ELBO}_d^{(l + 1)}  = & \frac{1}{2} \log |\bm{\Sigma}_d^{(l)}| -
+	  \frac{a_d^{(l)}}{2} \[ \y_d \tr \y_d - 2\y_d \tr \X \bm{\mu}_d^{(l)} +
+	  (\bm{\mu}_d^{(l)}) \tr \X \tr \X \bm{\mu}_d^{(l)} + \trace ( \X \tr \X 
+	  \bm{\Sigma}_d^{(l)})\] \\
+	  & + \frac{b_d^{(l)}}{2} \[ \delta_d^{(l)} - 
+	  a_d^{(l)}(\bm{\mu}_d^{(l)}) \tr \bm{\mu}_d^{(l)} - 
+	  a_d^{(l)} \trace (\bm{\Sigma}_d^{(l)}) \] - 
+	  \frac{n + p + 1}{2} \log \zeta_d^{(l)} \\
+	  & - \frac{p + \eta_d^{(l)}}{2} \log \delta_d^{(l)} + 
+	  \frac{\eta_d^{(l)}}{2} e_d^{(l)} + \frac{\eta_d^{(l)}}{2} \log 2 + 
+	  \log \Gamma \( \frac{p + \eta_d^{(l)}}{2} \) \\
+	  & - \log \Gamma \( \frac{\eta_d^{(l + 1)}}{2} \) - 
+	  \frac{\eta_d^{(l+1)}}{2} e_d^{(l)} - \frac{\eta_d^{(l+1)}}{2} \log 2 +
+	  \frac{\eta_d^{(l + 1)}}{2} \log \lambda_d^{(l+1)} - 
+	  \frac{\lambda_d^{(l + 1)} b_d^{(l)}}{2}.
+	\end{align*}
+  To compute the ELBO after a VB update we simply let 
+  $\eta_d^{(l)}, \lambda_d^{(l)}=\eta_d^{(l + 1)}, \lambda_d^{(l + 1)}$.
+
+	\section{Log pseudo marginal likelihood}\label{sec:lpml}
+	The log pseudo marginal likelihood is calculated as:
+	$\text{LPML}_d = n^{-1}\sum_{i=1}^n \log (\text{CPO}_{id})$ 
+  \cite[]{gelfand_model_1992}. Let 
+  $\theta_{id}=\x_i \tr \bbeta_d$ and $\y_{-id}$ denote $\y_{d}$ with 
+  observation $y_{id}$ excluded. Then the conditional 
+  predictive ordinates (CPO) are given by:
+  \begin{align*}
+    \text{CPO}_{id} & = p(y_{id} | \y_{-id}) = 
+    \int_{-\infty}^{\infty} \int_0^{\infty}
+    p(y_{id} | \bbeta_d, \sigma_d^2 ) \pi (\bbeta_d | \y_{-id}, \sigma_d^2) 
+    \pi (\sigma_d^2 | \y_{-id}) d\sigma_d^2 d\bbeta_d \\
+    & = \int_{-\infty}^{\infty} \int_0^{\infty}
+    p(y_{id} | \theta_{id}, \sigma_d^2 ) 
+    \pi (\theta_{id} | \y_{-id}, \sigma_d^2) 
+    \pi (\sigma_d^2 | \y_{-id}) d\sigma_d^2 d\theta_{id},
+  \end{align*}
+  where
+  \begin{align}
+    y_{id} | \theta_{id}, \sigma_d^2 & \sim \mathcal{N} (\theta_{id}, 
+    \sigma_d^2), \nonumber \\
+    \theta_{id} | \y_{-id} & \sim \mathcal{N}(\x_i \tr \bmu_{-id}, 
+    \x_i \tr \bSigma_{-id} \x_i), \nonumber \\
+    \sigma_d^2 | \y_{-id} & \sim \Gamma^{-1} \( \frac{n + p}{2},
+    \zeta_{-id} \), \label{eq:gammavb}
+  \end{align}
+  where $\bmu_{-id}$, $\bSigma_{-id}$, and $\zeta_{-id}$ are the variational
+  parameters estimated from $\y_{-id}$. Plugging in the densities we see that 
+  we have:
+  \begin{subequations}
+    \begin{align}
+      \text{CPO}_{id} = & (2 \pi)^{-1} \frac{\Gamma\(\frac{n + p + 1}
+      {2}\)}{\Gamma\(\frac{n + p}{2}\)}
+      (\zeta_{-id} \x_i \tr \bSigma_{-id} \x_i)^{-1/2} \label{eq:cpo1} \\
+      & \cdot \int_{-\infty}^{\infty}
+      \( \frac{\theta^2}{2\zeta_{-id}} + 1 \)^{-\frac{n + p + 1}{2}}
+      \exp \[ - \frac{(\theta - \x_i \tr \bmu_{-id} + y_i)^2}
+      {2 \x_i \tr \bSigma_{-id} \x_i} \] d\theta, \label{eq:cpo2}
+    \end{align}
+  \end{subequations}
+  a one-dimensional (well-behaved) integral that is easily solved with some 
+  numerical integration routine.
+
+  In practice we would like to avoid re-estimating the model for every 
+  observation $i$. Instead we divide the observations into $K$ parts and 
+  estimate $K$ models, so that now $\y_{-id}$ denotes $\y$ with all observations
+  in part $k$ removed. Additionally, we add $1 - n_{id}$ to the numerator of the
+  Gamma distribution in (\ref{eq:gammavb}), the numerators in the Gamma 
+  functions in (\ref{eq:cpo1}), and the numerator of the power in
+  (\ref{eq:cpo2}), where $n_{id}$ denotes the number of elements in the removed 
+  part.
+
+% 	\section{Non-conjugate model}
+% 	The non-conjugate model assumes $\beta_{jd}$ and $\sigma_d^2$ 
+% 	independent \textit{a priori}:
+%   $$
+%   \beta_{jd} | \gamma_d^2, \sigma_d^2 \sim \beta_{jd} | 
+%   \gamma_d^2 \sim \mathcal{N} (0, \gamma_d^2).
+%   $$
+%   
+% 	The variational distributions are as follows:
+% 	\begin{align*}
+% 	  q(\bm{\beta}_d) & \overset{D}{=} \mathcal{N}_p 
+% 	  (\bm{\mu}_d, \bm{\Sigma}_d), \\
+% 	  q(\gamma_d^2) & \overset{D}{=} \mathcal{GIG}
+% 	  \(-\frac{p+1}{2}, \lambda_d \theta_d^{-2}, \delta_d\), \\
+% 	  q(\sigma_d^2) & \overset{D}{=} \Gamma^{-1} 
+% 	  \(\frac{n + 1}{2}, \zeta_d\),
+%   \end{align*}
+% 	The variational parameters are iteratively updated by:
+%   \begin{align*}
+%   \bm{\Sigma}_d^{(h+1)} & = (a_d^{(h)})^{-1} (\X \tr \X + c_d^{(h)} \I)^{-1}, \\
+%   \bm{\mu}_d^{(h+1)} & = (\X \tr \X + c_d^{(h)} \I)^{-1} \X \tr \y_d, \\
+%   \delta_d^{(h+1)} & = b_d^{(h)} (c_d^{(h)})^{-1} \bmu_d \tr \bmu_d + \lambda_d, \\ 
+%   \zeta_d^{(h+1)} & = \frac{1}{2} \bigg[\mathbf{y}_d \tr \mathbf{y}_d -2 \mathbf{y}_d \tr \X \bm{\mu}_d^{(h+1)} + \trace ( \X \tr \X \bm{\Sigma}_d^{(h+1)}) + (\bm{\mu}_d^{(h+1)}) \tr \X \tr \X \bm{\mu}_d^{(h+1)} \\
+%   & \,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\, + \frac{b_d^{(h+1)} - c_d^{(h+1)} a_d^{(h+1)}}{1 - a_d^{(h+1)}} \trace ( \bm{\Sigma}_d^{(h+1)}) + \frac{b_d^{(h+1)} - c_d^{(h+1)} a_d^{(h+1)}}{1 - a_d^{(h+1)}} (\bm{\mu}_d^{(h+1)}) \tr \bm{\mu}_d^{(h+1)}\bigg].
+%   \end{align*}
+%   Here, we set $a_d^{(h)}$, $b_d^{(h)}$, and $c_d^{(h)}$ as in 
+%   (\ref{eq:auxvar1})
+%   Note that in the inverse Gamma model, the computation of (\ref{eq:auxvar1})
+%   simplifies to $b_d^{(h)}=(p+\eta_d)/\delta_d^{(h)}$.
+%   
+%   Here, we set
+%   \begin{align}
+%   a_d^{(h)} & =\E_{Q^{(h)}}(\sigma_d^{-2})=(df+1)/(2 \zeta_d^{(h)}), \nonumber \\
+%   b_d^{(h)} & = \E_{Q^{(h)}}(\gamma_d^{-2}) = \sqrt{\frac{\lambda_d}{\theta_d^2 \delta_d^{(h)}}} \frac{K_{\frac{p + \eta_d -2}{2}} \( \sqrt{\lambda_d \delta_d^{(h)}}/\theta_d \)}{K_{\frac{p+\eta_d}{2}} \( \sqrt{\lambda_d \delta_d^{(h)}}/\theta_d \)} + \frac{p+\eta_d^{(h)}}{\delta_d^{(h)}}, \label{eq:auxvar1} \\
+%   c_d^{(h)} & = \begin{cases*}
+%   b_d^{(h)} & in the conjugate setting \\
+%   b_d^{(h)}/a_d^{(h)} & in the non-conjugate setting .
+%   \end{cases*} \nonumber
+%   \end{align}
+%   Note that in the inverse Gamma model, the computation of (\ref{eq:auxvar1}) simplifies to $b_d^{(h)}=(p+\eta_d)/\delta_d^{(h)}$.
+%   
+%   \section{Starting values}
+%   \subsection{Method 1}
+%   Starting value for $\sigma_d^2$ as in \cite{chipman_bart:_2010}.
 They consider a scaled-Chi squared distribution for the error variance with degrees of freedom 3 and scale chosen such that the $q*100$th percentile matches the error variance in the data $s^2(\mathbf{y}_d)$.
 As starting value for $\sigma_d^2$ we take the mode of this distribution: $\hat{\sigma}_d^2 = 3a/5$, with $a$ the solution to:
-  $$
-  \Gamma\(\frac{3}{3}\) \[1 - q - F(3a/(2s^2(\mathbf{y}_d)); 3/2, 1) \] = 0,
-  $$
-  where $F(x; 3/2, 1)$ is the CDF of a gamma function with shape $3/2$ and scale $1$.
+%   $$
+%   \Gamma\(\frac{3}{3}\) \[1 - q - F(3a/(2s^2(\mathbf{y}_d)); 3/2, 1) \] = 0,
+%   $$
+%   where $F(x; 3/2, 1)$ is the CDF of a gamma function with shape $3/2$ and scale $1$.
 In \cite{moran_variance_2018}, REFERENCE to Rockova and George (2018), and \cite{chipman_bart:_2010}, $q=0.9$ is suggested.
 We find that $q \in ( 0.9, 0.95 )$ gives good results.
-
-  \subsection{Method 2}
-  We generate starting values as follows. Consider the $\sigma_d^2$ and $\gamma_d^2$ as fixed and estimate them by MML of the regular ridge model:
-  $$
-  \hat{\sigma}_d^2, \hat{\gamma}_d^2= \underset{\sigma_d^2,\gamma_d^2}{\argmax} \int_{\bm{\beta}} p(\mathbf{y}_d | \bm{\beta}, \sigma_d^2) p(\bm{\beta} | \gamma_d^2, \sigma_d^2) d\bm{\beta}.
-  $$
-  The regular ridge model is conjugate, so MML is relatively simple. Next we estimate a common inverse Gaussian prior for the $\gamma_d^2$ by considering the estimates to be samples from the prior:
-  \begin{align*}
-  \hat{\theta} & = n^{-1} \sum_{d=1}^D \hat{\gamma}_d^2 / n, \\
-  \hat{\lambda} & = n \left\{\sum_{d=1}^D \[ (\hat{\gamma}_d^2)^{-1} - \hat{\theta}^{-1}\]\right\}^{-1}.
-  \end{align*}
-  We estimate the mode of the $\hat{m}=\hat{\gamma}_d^2$ by kernel density estimation and equate it to the theoretical mode of the generalized inverse Gaussian distribution with the estimated $\hat{\theta}$ and $\hat{\lambda}$.
+%   
+%   \subsection{Method 2}
+%   We generate starting values as follows. Consider the $\sigma_d^2$ and $\gamma_d^2$ as fixed and estimate them by MML of the regular ridge model:
+%   $$
+%   \hat{\sigma}_d^2, \hat{\gamma}_d^2= \underset{\sigma_d^2,\gamma_d^2}{\argmax} \int_{\bm{\beta}} p(\mathbf{y}_d | \bm{\beta}, \sigma_d^2) p(\bm{\beta} | \gamma_d^2, \sigma_d^2) d\bm{\beta}.
+%   $$
+%   The regular ridge model is conjugate, so MML is relatively simple. Next we estimate a common inverse Gaussian prior for the $\gamma_d^2$ by considering the estimates to be samples from the prior:
+%   \begin{align*}
+%   \hat{\theta} & = n^{-1} \sum_{d=1}^D \hat{\gamma}_d^2 / n, \\
+%   \hat{\lambda} & = n \left\{\sum_{d=1}^D \[ (\hat{\gamma}_d^2)^{-1} - \hat{\theta}^{-1}\]\right\}^{-1}.
+%   \end{align*}
+%   We estimate the mode of the $\hat{m}=\hat{\gamma}_d^2$ by kernel density estimation and equate it to the theoretical mode of the generalized inverse Gaussian distribution with the estimated $\hat{\theta}$ and $\hat{\lambda}$.
 We solve to find a common $\delta$:
-  $$
-  \hat{\delta} = \frac{\hat{m}^2 \hat{\lambda}}{\hat{\theta}^2} + (p + 3) \hat{m}.
-  $$
-  Lastly, we consider the $\hat{\sigma}^2_d$ fixed samples from the inverse Gamma distribution to estimate a common scale:
-  $$
-  \hat{\zeta} = \frac{D(n + p + 1)}{2 \sum_{d=1}^D[(\hat{\sigma}_d^2)^{-1}]}.
-  $$
+%   $$
+%   \hat{\delta} = \frac{\hat{m}^2 \hat{\lambda}}{\hat{\theta}^2} + (p + 3) \hat{m}.
+%   $$
+%   Lastly, we consider the $\hat{\sigma}^2_d$ fixed samples from the inverse Gamma distribution to estimate a common scale:
+%   $$
+%   \hat{\zeta} = \frac{D(n + p + 1)}{2 \sum_{d=1}^D[(\hat{\sigma}_d^2)^{-1}]}.
+%   $$
 

### 2019-11-19 code/data_eqtl.R

@@ -1,0 +1,361 @@
+#!/usr/bin/env Rscript
+
+### installation of packages
+if(!("devtools" %in% installed.packages())) {
+  install.packages("devtools")
+}
+library(devtools)
+install_github("magnusmunch/cambridge/rpackage", local=FALSE,
+               auth_token=Sys.getenv("GITHUB_PAT"))
+
+### libraries
+library(cambridge)
+library(biomaRt)
+library(glmnet)
+library(rstan)
+library(MASS)
+
+### load data
+load("data/P38MAPKpathwayKb100K.RData")
+rm(G1, G2)
+
+### data preparation
+y <- sapply(Gn, "[[", 1)
+x <- sapply(Gn, "[[", 2)
+G <- length(Gn)
+p <- sapply(x, ncol)
+n <- nrow(y)
+
+### creating co-data
+inorout <- lapply(Gn, "[[", 4)
+distance <- lapply(Gn, function(g) {  
+  sapply(g[[5]], function(s) {
+    (sign(prod(s - g[[3]][c(1:2)]))!=1)*
+      min(abs(s - g[[3]][c(1:2)]))})})
+genelength <- lapply(1:G, function(g) {
+  rep(diff(as.numeric(Gn[[g]][[3]][c(1, 2)])), p[g])})
+minoraf <- lapply(sapply(Gn, "[[", 2), function(g) {
+  f <- apply(g, 2, function(s) {f <- mean(s); unname(min(f, 1 - f))})})
+
+### retrieve exon status of SNPs
+# View(listEnsembl())
+geneid <- sapply(Gn, function(g) {
+  strsplit(g[[7]]$gene_id, "\\.")[[1]][1]})
+ensembl <- useEnsembl(biomart="ensembl", dataset="hsapiens_gene_ensembl",
+                      GRCh=37)
+# View(listFilters(ensembl))
+gb <- getBM(attributes=c("ensembl_gene_id", "ensembl_exon_id", 
+                         "exon_chrom_start", "exon_chrom_end",
+                         "chromosome_name"), 
+            filters=c("ensembl_gene_id"), 
+            values=geneid, 
+            mart=ensembl)
+exon <- sapply(1:G, function(g) {as.numeric(sapply(Gn[[g]][[5]], function(s) {
+  any(apply(gb[gb$ensembl_gene_id==geneid[[g]], c(3, 4)], 1, function(e) {
[…]

### 2019-11-19 code/data_gdsc.R

@@ -11,14 +11,16 @@
 # }
 
 ### libraries
+detach("package:cambridge", unload=TRUE)
 library(cambridge)
 library(gdata)
-library(rcdk)
-library(fingerprint)
-library(glmnet)
+# library(rcdk)
+# library(fingerprint)
 library(MASS)
 library(biomaRt)
-library(rstan)
+detach("package:pInc", unload=TRUE)
+library(pInc)
+library(glmnet)
 
 ################################## analysis 1 ##################################
 # loading the data either from a local file or from the GDSC website
@@ -46,17 +48,17 @@ if(file.exists("data/Cell_Lines_Details.xlsx")) {
 
 # reading in the preformatted list of compounds
 if(file.exists("data/Screened_Compounds.xlsx")) {
-  drug1 <- read.xls("data/Screened_Compounds.xlsx", 
+  drug1 <- read.xls("data/Screened_Compounds.xlsx",
                     stringsAsFactors=FALSE)
 } else {
-  drug1 <- read.xls("ftp://ftp.sanger.ac.uk/pub/project/cancerrxgene/releases/current_release/Screened_Compounds.xlsx", 
+  drug1 <- read.xls("ftp://ftp.sanger.ac.uk/pub/project/cancerrxgene/releases/current_release/Screened_Compounds.xlsx",
                     stringsAsFactors=FALSE)
 }
 
 # reading in the curent list of compounds
 if(length(list.files(path="data", patt="Drug_list")) > 0) {
-  drug2 <- read.table(paste0("data/", list.files(path="data", 
-                                                 patt="Drug_list")[1]), 
+  drug2 <- read.table(paste0("data/", list.files(path="data",
+                                                 patt="Drug_list")[1]),
                       header=TRUE, sep=",", quote='"', stringsAsFactors=FALSE)
 } else {
[…]

### 2019-11-19 code/simulations_eqtl.R

@@ -1,0 +1,314 @@
+#!/usr/bin/env Rscript
+
+### installation of packages
+if(!("devtools" %in% installed.packages())) {
+  install.packages("devtools")
+}
+library(devtools)
+install_github("magnusmunch/cambridge/rpackage", local=FALSE,
+               auth_token=Sys.getenv("GITHUB_PAT"))
+
+### libraries
+library(cambridge)
+library(statmod)
+library(glmnet)
+
+### load data
+load("data/P38MAPKpathwayKb100K.RData")
+rm(G1, G2)
+
+### data preparation
+y <- sapply(Gn, "[[", 1)
+x <- lapply(sapply(Gn, "[[", 2), scale)
+G <- length(Gn)
+p <- sapply(x, ncol)
+n <- nrow(y)
+
+### creating co-data
+inorout <- lapply(Gn, "[[", 4)
+
+### simulation settings
+msigma.sq <- 1
+vsigma.sq <- 0
+mgamma.sq <- c(1, 2)
+vgamma.sq <- rep(0, G)
+
+### simulate model parameters
+set.seed(2019)
+if(vsigma.sq==0) {
+  sigma.sq <- rep(msigma.sq, G)
+} else {
+  sigma.sq <- 1/rgamma(G, msigma.sq^2/vsigma.sq + 2, 
+                       msigma.sq^3/vsigma.sq + msigma.sq)  
+}
+gamma.sq <- sapply(1:G, function(g) {
+  if(vgamma.sq[g]==0) {
+    mgamma.sq[inorout[[g]] + 1]
+  } else {
+    rinvgauss(p[g], mgamma.sq[inorout[[g]] + 1], 
+              mgamma.sq[inorout[[g]] + 1]^3/vgamma.sq[g])
+  }})
+
+
+beta <- sapply(1:G, function(g) {
+  rnorm(p[g], 0, sigma.sq[g]*gamma.sq[[g]])})
+
+### simulate data
+set.seed(2019)
+ntrain <- floor(n/2)
+ntest <- n - ntrain
+id.train <- sample(1:n, ntrain)
+xtrain <- sapply(x, function(s) {scale(s[id.train, ])})
+xtest <- sapply(x, function(s) {scale(s[-id.train, ])})
+ytrain <- scale(sapply(1:G, function(g) {
[…]

### 2019-11-19 code/tryout.R

@@ -1,51 +1,20 @@
-D <- 10
-r <- 200
-u <- 5
 n <- 100
-xu <- matrix(rnorm(u*n), nrow=n, ncol=u)
-xr <- matrix(rnorm(r*n), nrow=n, ncol=r)
-y <- matrix(rnorm(D*n), nrow=n, ncol=D)
-C <- lapply(1:D, function(d) {as.matrix(rbinom(r, 1, 0.5))})
-
-library(cambridge)
-test <- semnig(x=rep(list(xr), D), y=y, C=C, unpenalized=rep(list(xu), D), 
-               standardize=TRUE, intercept=TRUE, intercept.eb=TRUE, 
-               mult.lambda=FALSE, fixed.eb="none", full.post=TRUE, init=NULL,
-               control=list(conv.post=TRUE, trace=TRUE,
-                            epsilon.eb=1e-3, epsilon.vb=1e-3, 
-                            epsilon.opt=sqrt(.Machine$double.eps),
-                            maxit.eb=100, maxit.vb=2, maxit.opt=100,
-                            maxit.post=100))
-plot(test$seq.eb$alpha[, 1], type="l", ylim=range(test$seq.eb$alpha))
-lines(test$seq.eb$alpha[, 2], col=2)
-
-test$eb$mprior
-
-str(test)
-x=rep(list(xr), D); y=y; C=NULL; unpenalized=rep(list(xu), D); 
-standardize=TRUE; intercept=TRUE; intercept.eb=TRUE; 
-mult.lambda=FALSE; fixed.eb="both"; full.post=TRUE; init=NULL;
+p <- 150
+D <- 100
+x <- replicate(D, matrix(rnorm(n*p), nrow=n, ncol=p), simplify=FALSE)
+beta <- replicate(D, rnorm(p), simplify=FALSE)
+y <- sapply(1:D, function(d) {x[[d]] %*% beta[[d]]})
+Z <- NULL
+# Z <- cbind(1, rbinom(D, 1, 0.5))
+C <- lapply(1:D, function(d) {cbind(1, rbinom(p, 1, 0.5))})
+unpenalized=NULL
+standardize=FALSE
+intercept=FALSE
+fixed.eb="none"
+full.post=FALSE
+init=NULL
[…]

### 2019-11-19 rpackage/R/aux_functions.R

@@ -1,3 +1,38 @@
+# ELBO (not tested)
+.single.elbo <- function(p, n, zeta, yty, aux, g, b, delta, eta, lambdaf, 
+                         lambdad, Zalphad, Calphaf) {
+  
+  # elbo <- -0.5*(n + p + 1)*log(pi) + 0.5*(p + 1 - n)*log(2) + 0.5*(n + 1) + 
+  #   p + lgamma(0.5*(n + p + 1)) - 0.5*(n + p + 1)*log(zeta) + 
+  #   0.5*aux$ldetSigma - 0.25*(n + p + 1)/zeta*
+  #   (yty - 2*aux$ytXmu + aux$trXtXSigma + aux$mutXtXmu + 
+  #      g*sum(b*aux$dSigma) + g*sum(aux$mu^2*b)) + p*log(lambdaf) + 
+  #   0.25*(p + 3)/4*log(lambdad) + 0.5*(eta - lambdad)*g + 
+  #   sum(0.5*(delta - lambdaf)*b) + lambdad/chi + sum(lambdaf/phi) - 
+  #   0.5*(p + 1)*log(chi) - sum(log(phi)) - 0.25*(p + 1)*log(eta) - 
+  #   0.5*sum(log(delta)) + 
+  #   gsl::bessel_lnKnu(0.5*(p + 1), sqrt(lambdad*eta)/chi) +
+  #   sum(gsl::bessel_lnKnu(1, sqrt(lambdaf*delta)/phi))
+  chi <- switch(is.null(Zalphad) + 1, 1/Zalphad, NULL)
+  phi <- switch(is.null(Calphaf) + 1, 1/Calphaf, NULL)
+  elbo <- -0.5*(n + p + 1)*log(pi) + 0.5*(p + 1 - n)*log(2) + 0.5*(n + 1) + 
+    p + lgamma(0.5*(n + p + 1)) - 0.5*(n + p + 1)*log(zeta) + 
+    0.5*aux$ldetSigma - 0.25*(n + p + 1)/zeta*
+    (yty - 2*aux$ytXmu + aux$trXtXSigma + aux$mutXtXmu + 
+       g*sum(b*aux$dSigma) + g*sum(aux$mu^2*b)) + 
+    switch(is.null(Zalphad) + 1, 0.25*(p + 3)/4*log(lambdad) + 
+             0.5*(eta - lambdad)*g + lambdad/chi - 0.5*(p + 1)*log(chi) - 
[…]

### 2019-11-19 rpackage/R/main_functions.R

@@ -1,27 +1,19 @@
-# x <- list(x, x); y <- cbind(y, y); C=NULL; unpenalized=NULL;
-# standardize=FALSE; intercept=FALSE; intercept.eb=TRUE;
-# mult.lambda=FALSE; fixed.eb=c("both"); full.post=TRUE;
-# init=list(aold=rep(1, 2), bold=list(p, p),
-#           lambda=lambda, alpha=1,
-#           Calpha=rep(1/ctalphainv, 2));
-# control=list(conv.post=TRUE, trace=TRUE,
-#              epsilon.eb=1e-3, epsilon.vb=1e-3,
-#              epsilon.opt=sqrt(.Machine$double.eps),
-#              maxit.eb=10, maxit.vb=2, maxit.opt=100,
-#              maxit.post=100)
+# x=xtrain; y=ytrain; C=C; Z=Z; unpenalized=NULL;
+# standardize=FALSE; intercept=FALSE; fixed.eb="none";
+# full.post=FALSE; init=NULL; control=control
+# 
+# library(Rcpp)
+# sourceCpp("rpackage/src/aux_functions.cpp")
 # detach("package:cambridge", unload=TRUE)
 # library(cambridge)
 
-# estimate SEMNIG model
-semnig <- function(x, y, C, unpenalized=NULL, standardize=TRUE, intercept=TRUE, 
-                   intercept.eb=TRUE, mult.lambda=FALSE, 
-                   fixed.eb=c("none", "lambda", "both"), 
+# estimate NIG model (not tested)
+semnig <- function(x, y, C, Z, unpenalized=NULL, standardize=FALSE, 
+                   intercept=FALSE, fixed.eb="none", 
                    full.post=FALSE, init=NULL,
                    control=list(conv.post=TRUE, trace=TRUE,
-                                epsilon.eb=1e-3, epsilon.vb=1e-3, 
-                                epsilon.opt=sqrt(.Machine$double.eps),
-                                maxit.eb=10, maxit.vb=2, maxit.opt=100,
-                                maxit.post=100)) {
+                                epsilon.eb=1e-3, epsilon.vb=1e-3,
+                                maxit.eb=10, maxit.vb=2, maxit.post=100)) {
   
   # save the arguments
   cl <- match.call()
[…]

### 2019-11-19 rpackage/tests/testthat/test-aux_functions.R

@@ -39,7 +39,8 @@ test_that("low dim. auxiliary variables calculation", {
   mu <- as.numeric(solve(t(x) %*% x + H) %*% t(x) %*% y)
   aux <- list(mu=mu, dSigma=diag(Sigma), ytXmu=as.numeric(t(y) %*% x %*% mu),
               trXtXSigma=sum(diag(t(x) %*% x %*% Sigma)), 
-              mutXtXmu=as.numeric(t(mu) %*% t(x) %*% x %*% mu))
+              mutXtXmu=as.numeric(t(mu) %*% t(x) %*% x %*% mu),
+              ldetSigma=as.numeric(determinant(Sigma)$modulus))
   
   # check
   expect_equal(aux, test.aux)
@@ -66,7 +67,8 @@ test_that("high dim. auxiliary variables calculation", {
   mu <- as.numeric(solve(t(x) %*% x + H) %*% t(x) %*% y)
   aux <- list(mu=mu, dSigma=diag(Sigma), ytXmu=as.numeric(t(y) %*% x %*% mu),
               trXtXSigma=sum(diag(t(x) %*% x %*% Sigma)), 
-              mutXtXmu=as.numeric(t(mu) %*% t(x) %*% x %*% mu))
+              mutXtXmu=as.numeric(t(mu) %*% t(x) %*% x %*% mu),
+              ldetSigma=as.numeric(determinant(Sigma)$modulus))
   
   # check
   expect_equal(aux, test.aux)
@@ -96,7 +98,8 @@ test_that("low dim. auxiliary variables calculation", {
   mu <- as.numeric(solve(t(x) %*% x + H) %*% t(x) %*% y)
   aux <- list(mu=mu, dSigma=diag(Sigma), ytXmu=as.numeric(t(y) %*% x %*% mu),
               trXtXSigma=sum(diag(t(x) %*% x %*% Sigma)), 
-              mutXtXmu=as.numeric(t(mu) %*% t(x) %*% x %*% mu))
+              mutXtXmu=as.numeric(t(mu) %*% t(x) %*% x %*% mu),
+              ldetSigma=as.numeric(determinant(Sigma)$modulus))
   
   # check
   expect_equal(aux, test.aux)
@@ -125,7 +128,8 @@ test_that("high dim. auxiliary variables calculation", {
   mu <- as.numeric(solve(t(x) %*% x + H) %*% t(x) %*% y)
[…]

### 2019-11-19 rpackage/tests/testthat/test-vb_functions.R

@@ -8,27 +8,37 @@ test_that("low dim. vb update", {
   y <- rnorm(n)
   aold <- rchisq(1, 1) + 1
   bold <- rchisq(p, 1) + 1
-  Calpha <- rchisq(p, 1) + 1
-  lambda <- rchisq(1, 1) + 1
+  gold <- rchisq(1, 1) + 1
+  Calphaf <- rchisq(p, 1) + 1
+  Zalphad <- rchisq(1, 1) + 1
+  lambdaf <- rchisq(1, 1) + 1
+  lambdad <- rchisq(1, 1) + 1
   ytx <- as.numeric(t(y) %*% x)
   yty <- as.numeric(t(y) %*% y)
   
   # implemented function
-  test.vb <- .single.vb.update(aold, bold, Calpha, lambda, y, x, ytx, yty, n, p)
+  test.vb <- .single.vb.update(aold, bold, gold, Calphaf, Zalphad, lambdaf, 
+                               lambdad, y, x, ytx, yty, n, p)
   
   # actual calculation (see supplement)
-  Sigma <- solve(t(x) %*% x + diag(bold))/aold
-  mu <- as.numeric(solve(t(x) %*% x + diag(bold)) %*% t(x) %*% y)
-  delta <- aold*(mu^2 + diag(Sigma)) + lambda
-  b <- sqrt(lambda*Calpha^2/delta)*
-    ratio_besselK(sqrt(lambda*delta)*Calpha, 1) + 2/delta
+  Sigma <- solve(t(x) %*% x + gold*diag(bold))/aold
+  mu <- as.numeric(solve(t(x) %*% x + gold*diag(bold)) %*% t(x) %*% y)
+  delta <- aold*gold*(mu^2 + diag(Sigma)) + lambdaf
+  b <- sqrt(lambdaf*Calphaf^2/delta)*
+    ratio_besselK(sqrt(lambdaf*delta)*Calphaf, 1) + 2/delta
+  eta <- aold*sum(bold*(mu^2 + diag(Sigma))) + lambdad 
+  g <- sqrt(lambdad*Zalphad^2/eta)*
+    ratio_besselK(sqrt(lambdad*eta)*Zalphad, (p + 1)/2) + (p + 1)/eta
   zeta <- 0.5*as.numeric(
     t(y) %*% y  - 2*t(y) %*% x %*% mu + sum(diag(t(x) %*% x %*% Sigma)) +
[…]

### 2019-11-19 docs/manuscript1.Rnw

--- 2019-08-15 docs/manuscript1.Rnw
@@
 bbm,algorithm,algpseudocode,pgfplotstable,threeparttable,booktabs,mathtools,
-dsfont,parskip,xr}
+dsfont,parskip,xr,tikz}
 \externaldocument[sm-]{supplement1}
+\usetikzlibrary{bayesnet}
 
@@
 % settings
+\pgfplotsset{compat=1.16}
 \graphicspath{{../figs/}}
-\pgfplotsset{compat=1.16}
 \setlength{\evensidemargin}{.5cm} \setlength{\oddsidemargin}{.5cm}
 \setlength{\textwidth}{15cm}
-\title{Empirical Bayes estimation of the normal inverse Gaussian prior in 
-simultaneous-equation models}
+\title{Drug sensitivity prediction with the normal inverse Gaussian prior 
+informed by external data}
 \date{\today}
 \author{Magnus M. M\"unch$^{1,2}$\footnote{Correspondence to: 
-\href{mailto:m.munch@vumc.nl}{m.munch@vumc.nl}}, Mark A. van de Wiel$^{1,3}$, 
-Sylvia Richardson$^{3}$, \\ and Gwena{\"e}l G. R. Leday$^{3}$}
+\href{mailto:m.munch@amsterdamumc.nl}{m.munch@amsterdamumc.nl}}, Mark A. van de 
+Wiel$^{1,3}$, Sylvia Richardson$^{3}$, \\ and Gwena{\"e}l G. R. Leday$^{3}$}
 
@@
 	\begin{abstract}
-		{...}
+		{In precision medicine, a common problem is drug sensitivity prediction from
+		cancer tissue cell lines. These types of problems entail modelling 
+		multivariate drug responses on high dimensional molecular feature sets in 
+		typically $>1000$ cell lines. The dimensions of the problem require 
+		specialised models and estimation methods. In addition, external information
+		on both the drugs and the features is often available. We propose to 
+		model the drug responses through a linear regression with the normal
+		inverse Gaussian prior. We let the prior depend on the external information,
+		and estimate the model and external information dependence in an 
+		empirical-variational Bayes framework. We demonstrate the usefullness of
+		this model in both a simulated setting and in the publicly available
+		Genomics of Drug Sensitivity in Cancer data.}
 	\end{abstract}
 
-	\noindent\textbf{Keywords}: ...
+	\noindent\textbf{Keywords}: Drug sensitivity; Empirical Bayes; 
+	Genomics of Drug Sensitivity in Cancer (GDSC); Variational Bayes
+
+	\noindent\textbf{Software available from}: 
+	\url{https://github.com/magnusmunch/cambridge}
 
 	\section{Introduction}
-	Recently, the dimensions of data sets have increased in three directions:
-	(i) samples, (ii) variables, and (iii) outcomes. Here, we will 
-	concern ourselves with directions (ii) and (iii). More spefically, we will 
-	focus on the regression setting
-	where we have multiple outcomes and a medium to high dimensional feature 
-	space, i.e., $p \gtrsim n$. Standard methods
-	underperform or even break down in such medium to high dimensional settings, 
-	so new methodology is required.
+	Recently, promising results in precision medicine 
+	have sparked an interest in cancer drug sensitivity prediction models
+	\cite[]{iorio_landscape_2016}. Typically, these models 
+	predict the drug sensitivity for new patients from a set of molecular 
+	features.
+	Development of such models is often done in well-characterised human cancer
+	tissue cell lines. The current paper presents a novel drug sensitivity
+	prediction model and an application to a real drug
+	sensitivity data set.
 
-	As a motivating example, we present an expression quantitative loci (eQTLs)
-	detection problem (see Section \ref{sec:application}). In eQTLs detection, 
-	the aim is to find a set of single nucleotide loci (SNPs) that explains a
-	(part of a) gene's expression. The resulting heritability of the gene 
-	expression traits is thought to be indicative of the relationship between 
-	genotype and phenotype. eQTLs detection is often done for a set of genes 
-	simultaneously and as such comprises a simultaneous-equations model (SEM) with
-	multiple outcomes and a high dimensional feature set (the SNPs). Apart from
-	eQTLs, SEM models are also applied in network reconstruction 
-	\cite[]{kpogbezan_incorporating_2019, leday_gene_2017} 
-	and drug target discovery \cite[]{noh_inferring_2016}, [! ref to second 
-	manuscript].
+	Development of such models from cell lines has proven to be difficult
+	(see e.g., the DREAM 7 challenge in \cite{costello_community_2014}). 
+	Difficulties arise, among others, from the dimensions of the problem. 
+	Typically, the data contains
+	hundreds of drugs, thousands of cell lines, and thousands of molecular 
+	features. An example of a large database of drug responses and molecular
+	features is the Genomics of Drug Sensitivity in Cancer (GDSC) data 
+	\cite[]{yang_genomics_2013}, which we will further 
+	investigate in Section \ref{sec:gdsc}. Other examples of such databases 
+	include
+	the Cancer Cell Line Encyclopedia (CCLE) \cite[]{li_landscape_2019} and 
+	the US National Cancer Institute 60 human tumour cell line anticancer drug
+	screen (NCI60) \cite[]{shoemaker_NCI60_2006}. The dimensions of these data
+	prohibit the estimation of standard regression models and
+	typically require some form of regularisation.
 
-	In many multi-outcome regression settings, extra information on both the 
-	outcomes and 
-	features is available. In our motivating eQTLs detection example, extra 
-	information may come in the form of (i) gene annotation (e.g., length of the
-	gene), (ii) SNP annotation (e.g., whether the SNP is in the gene or not), 
-	(iii) SNP-gene annotation (e.g., distance of SNP to gene). The presented 
-	method allows for the incorporation of such information to aid in the 
-	estimation, by pooling information both across equations (genes) and 
-	features (SNPs).
+	The GDSC database contains additional information on both the drugs and 
+	molecular features, such as the target 
+	pathways and developmental stages of the drugs. Additional online
+	repositories might provide extra information such as the molecular weight of
+	the compounds or the publication signatures of the molecular features. In some
+	cases, prior knowledge on the drug efficacies might be available,
+	from previous experiments. We propose to include these possibly beneficial 
+	information sources in the estimation of the sensitivity prediction models in 
+	a data-driven manner. More specifically, we estimate a normal 
+	inverse Gaussian model, where the amount of regularisation is estimated by an 
+	adaptive empirical Bayes procedure, guided by the external information.
 
-	We are not the first to propose to learn from external data 
-	across equations and features. \cite{kpogbezan_incorporating_2019} uses 
-	the presence or absence of edges in a normal tissue apoptosis pathway, to aid 
-	in the estimation of a cancer tissue apoptosis pathway. Furhtermore, in an 
-	eQTLs detection problem they include the location of the SNPs (inside
-	the gene or not) to aid in the estimation. \cite{oman_minimax_2002} uses 
-	empirical Bayes to learn from external data concerning both the outcomes and 
-	the features in a low dimensional chemometrics dataset. Our proposed method,
-	termed \texttt{semnig},
-	extends \cite{kpogbezan_incorporating_2019,kpogbezan_empirical_2017} by 
-	allowing for the inclusion of continuous external information and several 
-	sources of information.
+	We are not the first to work on drug sensitivity prediction models. Reviews on 
+	the topic are \cite{azuaje_computational_2017} and \cite{ali_machine_2019}.
+	\cite{zhao_structured_2019} and \cite{mai_composite_2019} consider a
+	structured penalized multivariate regression approach. 
+	\cite[]{aben_tandem:_2016} introduces a two-stage penalized regression 
+	model that includes two different types of molecular features.
+	\cite{ammad-ud-din_drug_2016} and \cite{costello_community_2014} tackle the
+	problem through a multiple kernel learning approach.
+
+  Our solution allows for the adaptive incorporation of the external 
+  information on drugs and features. This is done by
+  pooling, both across drugs and features. Estimation of the model
+  is through computational feasible variational bayes approximations, while 
+  empirical Bayes estimation of tuning parameters pools information across drugs
+  and features in a data-driven manner.
 
@@
 	\ref{sec:model} we introduce our model, the estimation of which is detailed
-	in Section \ref{sec:estimation}. In Sections \ref{sec:simulations} and
-	\ref{sec:application} we demonstrate the method in simulations and an 
-  eQTLs detection problem, respectively. We end with a discussion in Section 
-  \ref{sec:discussion}.
+	in Section \ref{sec:estimation}. In Section \ref{sec:gdsc} we analyse the
+	GDSC data, and we end with a discussion in Section \ref{sec:discussion}.
 
@@
 	\subsection{Simultaneous equations model}\label{sec:SEM}
-	We have continuous measurements $y_{id}$ for observation $i=1,\dots, n$ on 
-	outcome $d=1,\dots,D$. Throughout this paper we assume the $y_{id}$ to be 
-	centred per outcome and let $\y_d = \begin{bmatrix} y_{1d} & \dots & y_{nd} 
-	\end{bmatrix} \tr$. We predict the $y_{id}$ with features $x_{ij}$, 
-	$j=1,\dots, p$, collected in $\x_i = \begin{bmatrix} x_{i1} & \dots & x_{ip} 
-	\end{bmatrix} \tr$ and centered per feature. We have $|\bm{\alpha}|$ 
-	external variables available that describes the outcome $d$, feature $j$, or
-	relationship between feature $j$ and outcome $d$, collected in 
-  $\mathbf{C} = \begin{bmatrix} \mathbf{c}_{1}^1 & \cdots &
-	\mathbf{c}_{1}^p & \cdots & \mathbf{c}_{D}^1 & \cdots & \mathbf{c}_{D}^p
-  \end{bmatrix}_{|\bm{\alpha}| \times pD} \tr$. An example of such an external
-  variable from eQTL detection is the distance between SNP and gene.
+	Let $y_{id}$ be the continuous sensitivity measures 
+	for cell lines $i=1,\dots, n$, and drug $d=1, \dots, D$. 
+	We predict sensitivity from molecular features 
+	$x_{ij}$, $j=1,\dots, p$, collected in 
+	$\x_i = \begin{bmatrix} x_{i1} & \dots & x_{ip} \end{bmatrix} \tr$. 
+	We assume that both covariates and responses have been centred per drug
+	and regress the drug sensitivities on the molecular features:
+	\begin{equation}\label{eq:likelihood}
+	  y_{id} = \x_i \tr \bbeta_d + \epsilon_{id}, 
+	  \text{ with } \epsilon_{id} \sim \mathcal{N}(0, \sigma_d^2),
+	\end{equation}
+	where the $p$-dimensional 
+	$\bbeta_d = \begin{bmatrix} \beta_{1d} & \cdots & \beta_{pd} \end{bmatrix} 
+	\tr$ are the drug-specific omics feature 
+	effects. Note that (\ref{eq:likelihood}) gives rise to a system of 
+	D linear regression equations.
 
-	We model the outcomes $y_{id}$ as a linear function of the features $x_{ij}$:
-	\begin{subequations}\label{eq:linearmodel}
+	The cell lines used in drug response models are often taken from different
+	tissues. In addition, other clinical covariates might be available. 
+	To obtain unbiased feature effects, one may wish to account for these. 
+	We do so by introducing unpenalized covariates, the $\beta_{jd}$ coefficients 
+	of which are endowed with a flat prior. For the sake of clarity, in the
+	following, such unpenalized covariates are omitted. However, the available
+	software allows for their inclusion.
+
+	\subsection{Bayesian prior model}
+	We carry out inference by endowing the parameters with the following priors:
+	\begin{subequations}\label{eq:prior}
 		\begin{align}
-		y_{id} & = \beta_{0d} + \x_i \tr \bbeta_d + \epsilon_{id},\\
-		\text{with } & \epsilon_{id} \sim \mathcal{N}(0, \sigma_d^2),
+		  \beta_{jd} | \gamma^2_{jd}, \tau_d^2, \sigma^2_d & \sim \mathcal{N}_{p} 
+		  (0, \gamma_{jd}^2 \tau_d^2 \sigma_d^2), 
+		  \label{eq:betaprior}\\
+		  \gamma_{jd}^{2} & \sim \mathcal{IG}(\phi_{jd}, \lambda_{\text{feat}}), \\
+		  \tau_d^{2} & \sim \mathcal{IG}(\chi_d, \lambda_{\text{drug}}), 
+		  \label{eq:tauprior}\\
+		  \sigma_d^{2} & \sim 1/\sigma_d^{3},
 		\end{align}
 	\end{subequations}
-	where the $p$-dimensional $\bbeta_d$ are the outcome-specific feature effects.
-	Note that (\ref{eq:linearmodel}) gives rise to a system of 
-	$D$ linear simultaneous equations.
+	where $\mathcal{IG}(\phi, \lambda)$ denotes an inverse Gaussian distribution 
+	with mean $\phi$ and shape $\lambda$. Prior distributions of the form
+	(\ref{eq:prior}) are often referred to as global-local shrinkage rules 
+	\cite[]{bernardo_shrink_2011}, due to the multiplicative separation of the 
+	prior 
+	variance into a local components $\gamma_{jd}^2$ and a global component
+	$\tau_d^2$. The local components are supposed to capture local variation in
+	the model parameters $\beta_{jd}$, while the global component is supposed to
+	capture the general trend in $\beta_{jd}$. Addionally, for appropriate local
+	shrinkage in global-local 
+	shrinkage models it is important  to account for 
+  different noise levels $\sigma_d^2$ by scaling the $\beta_{jd}$ variances 
+  accordingly.
 
-	\subsection{Bayesian prior model}
-	We capture the uncertainty in the parameters through a Bayesian prior model. 
-	We assign a flat prior to the intercept terms $\beta_{0d}$ and integrate 
-	them out. The remaining parameters are endowed with priors such that we end
-	up with the following model:
-	\begin{subequations}\label{eq:priormodel}
-    \begin{align}
-      y_{id} | \bm{\beta}_d, \sigma_d^2 & \sim \mathcal{N} 
-      \(\x_i \tr \bm{\beta}_d, \sigma_d^2\), \\
-      \beta_{jd} | \gamma_{jd}^2, \sigma_d^2 & \sim \mathcal{N} 
-      \(0, \sigma_d^2 \gamma_{jd}^2 \), \label{eq:betaprior} \\
-      \gamma_{jd}^2 & \sim \mathcal{IG}
-      \( [(\mathbf{c}_{d}^j) \tr \bm{\alpha}]^{-1}, \lambda_{d} \), \\
-      \sigma_d^2 & \sim 1/\sigma_d^3,
-    \end{align}
-  \end{subequations}
-  where $\mathcal{IG} ([(\mathbf{c}_{d}^j) \tr \bm{\alpha}]^{-1}, \lambda_{d})$ 
-  denotes the inverse Gaussian distribution with mean 
-  $[(\mathbf{c}_{d}^j) \tr \bm{\alpha}]^{-1}$ and shape $\lambda_d$.
+	The normal inverse Gaussian (NIG) prior model was introduced in 
+  \cite{barndorff-nielsen_hyperbolic_1978} and since 
+  \cite{barndorff-nielsen_normal_1997} it is routinely applied in mathematical 
+  finance (see, e.g., \cite{kalemanova_normal_2007}). Here we extend it with an
+  additional global variance component $\tau_d^2$. Supplementary Material
+  (SM) Section \ref{sm-sec:prior} contains more details on the NIG prior.
+  To illustrate the effect of the NIG prior on the posterior mean, we consider 
+  the prior reparametrised as in 
+  \cite{carvalho_handling_2009}, i.e., in terms of shrinkage weights 
+  $\kappa_{jd}=1/(1 + \gamma_{jd}^2) \in (0, 1)$.
+  Under the (simplified) normal means model, i.e., 
+  $\X= \begin{bmatrix} \x_1 & \cdots \x_n \end{bmatrix} \tr =\I_p$, with fixed 
+  $\tau_d^2=\sigma_d^2=1$, the resulting
+  posterior mean for the $\beta_{jd}$ is
+  $\E(\beta_{jd} | y_{jd}, \kappa_{jd}) = (1 - \kappa_{jd}) y_{jd}$. 
+  Thus, $\kappa_{jd}=0$ implies no shrinkage of $\beta_{jd}$ and $\kappa_{jd}=1$
+  implies full shrinkage towards zero. Figure 
+  \ref{fig:dens_kappa} depicts the prior on $\kappa_{jd}$ implied by 
+  several choices of $\beta_{jd}$ prior.
+<<dens_kappa, fig.cap="Implied prior densities $\\pi(\\kappa_{jd})$ for the (a) NIG, (b) Student's $t$, and (c) lasso priors.
+Different line types correspond to different hyperparameter settings.
+The hyperparameter settings were chosen to show some possible, distinct shapes that each of the priors can take.", out.width="100%", fig.asp=1/3>>=
+@
 
-  The normal inverse Gaussian (NIG) prior model for $\beta_{jd}$ and 
-  $\gamma^2_{jd}$ in (\ref{eq:priormodel}) was introduced in 
-  \cite{barndorff-nielsen_hyperbolic_1978} and is, since 
-  \cite{barndorff-nielsen_normal_1997}, routinely applied in mathematical 
-  finance (see, e.g., \cite{kalemanova_normal_2007}). Supplementary Material
-  (SM) Section \ref{sm-sec:nigprior} contains more details on this prior.
-  In Figure \ref{fig:dens_igaussian_marginalbeta}, the NIG 
-  prior distribution, marginalised over $\gamma_{jd}^2$, is depicted together 
-  with three other common choices,
-  the Student's $t$ (see Section \ref{sec:inversegammamodel} for more details on 
-  this prior), normal (ridge) and Laplace (lasso) priors, 
-  all scaled to variance one. 
-  From this Figure we see that, depending on the choice
-  of hyperparameters, the NIG prior can vary from heavier tails than the lasso, 
-  to more Gaussian tails like the ridge, but behaves similarly to the 
-  Student's $t$ prior. Our argumentation to model the 
+  Figure \ref{fig:dens_kappa} shows that, depending on the choice
+  of hyperparameters, the NIG prior can behave similarly to the Student's $t$ 
+  prior (decreasing form zero, with substantial mass close to 0), 
+  but also rather differently. Our argumentation to model the 
   $\gamma^2_{jd}$ by an inverse Gaussian 
   distribution, as has been suggested in \cite{fabrizi_specification_2016} and 
-  \cite{caron_sparse_2008}, is two-fold: (i) the NIG model is more flexible than
-  the standard ridge and lasso priors (as seen from Figure 
-  \ref{fig:dens_igaussian_marginalbeta}), and (ii) the NIG prior allows to model
-  the mean as a 
-  function of the external variables $\mathbf{c}_{d}^j$ more conveniently than 
-  the Student's $t$ prior, as explained in Section 
-  \ref{sec:empiricalbayes}. 
-<<dens_igaussian_marginalbeta, fig.cap="Several choices of marginal $\\beta_j$ prior, scaled to $\\V(\\beta_j)=1$.", out.width="60%", fig.asp=3/4>>=
-@
-
-  \cite{moran_variance_2018} argue that in regression models, $\bm{\beta}_d$ 
-  and $\sigma_d^2$ should be independent \textit{a priori}, i.e.
-  $\beta_{jd} | \gamma_{jd}^2, \sigma_d^2 \sim \beta_{jd} | 
-  \gamma_{jd}^2$. However, in our case
-  that is debatable. The $D$ regressions need not be on the same scale, so in
-  order to jointly consider the $\bm{\beta}_d$ some 
-  form of calibration is required. A straightforward way 
-  of doing this is to scale the $\bm{\beta}_d$ prior with 
-  $\sigma_d^2$ as in (\ref{eq:betaprior}), resulting in a 
-  scale-free prior with respect to the response. Apart from the scaling,
-  in this setting, the conjugate prior model is preferred
-  over the non-conjugate model, because the variational Bayes approximation 
-  (see Section \ref{sec:estimation}) assumes independent $\sigma_d^2$ and 
-  $\bm{\beta}_d$ in the posterior, which is actually true for 
-  the conjugate model. Therefore, the approximate expectation used in the 
-  empirical Bayes step (see Section \ref{sec:estimation}) is more accurate.
+  \cite{caron_sparse_2008}, is three-fold: (i) the NIG model is more flexible 
+  than the lasso prior (as seen from Figure 
+  \ref{fig:dens_kappa}), (ii) the NIG prior allows to model
+  the means of the $\gamma_{jd}^2$ ($\phi_{jd}$) and $\tau_{jd}^2$ 
+  ($\chi_{d}$) as a function of external data 
+  more conveniently than the Student's $t$ prior, as explained in Section 
+  \ref{sec:empiricalbayes}, and (iii) like the horseshoe, the NIG shrinkage
+  weights prior can put mass both near 0 and 1, a desirable property of 
+  shrinkage priors.
 
@@
   parameter to set. This influence is especially pronounced if the likelihood is
-  relatively flat, as may be reasonably expected in the large $p$, small $n$ 
-  setting we are in. We therefore model the error variance with Jeffrey's 
+  relatively flat, as may be reasonably expected in the large $p$, 
+  small $n$ setting. We therefore model the error variance with Jeffrey's 
   objective prior 
-  \cite[]{jeffreys_invariant_1946}. In the derivation of our Jeffrey's prior
-  for the error variance, we jointly consider an unknown mean and variance. This
+  \cite[]{jeffreys_invariant_1946} that does not depend on any 
+  subjective specification of hyperparameters. In the derivation of our 
+  Jeffrey's prior
+  for the error variance, we jointly consider an unknown mean and variance 
+  \cite[]{kass_selection_1996}. This
   joint consideration results in the somewhat unorthodox $1/\sigma^3$ Jeffrey's
@@
 
+	\subsection{External information}\label{sec:external}
+	In drug sensitivity prediction models external information on both the drugs
+	and features is often available. Here, we assume this information to be 
+	available as
+	external feature `covariates' $\mathbf{c}_{jdg}$, for $g=1, \dots, G$, and 
+	drug `covariates' $\mathbf{z}_{dh}$, for $h=1,\dots, H$. An example of a 
+	(binary) features covariate is target pathway presence, with $c_{jdg}=0$ if 
+	gene $j$ is present in the 
+	target pathway of drug $d$ and $c_{jdg}=1$ if it is not. An example of a 
+	(ternary) drug
+	covariate is developmental phase, with levels experimental phase, 
+	clinical development, and approved by a governing agency.
+
+	The external covariates come in through our mean models for 
+	the $\gamma_{jd}^2$ and $\tau_d^2$ hyperpriors:
+	$\phi_{jd} = (\mathbf{c}_{jd} \tr \bm{\alpha}_{\text{feat}})^{-1}$ and
+	$\chi_d = (\mathbf{z}_{d} \tr \bm{\alpha}_{\text{drug}})^{-1}$, with 
+	$\mathbf{c}_{jd} = \begin{bmatrix} c_{jd1} & \cdots & c_{jdG} \end{bmatrix}$
+	and $\mathbf{z}_{d} = \begin{bmatrix} z_{d1} & \cdots & z_{dH} \end{bmatrix}$,
+	where categorical external covariates are dummy coded.
+	The model now requires hyperparameters $\bm{\alpha}_{\text{feat}}$, 
+	$\lambda_{\text{feat}}$, $\bm{\alpha}_{\text{drug}}$, and $
+	\lambda_{\text{drug}}$, which we estimate in a data-driven manner (see Section
+	\ref{sec:empiricalbayes}).
+
+  A representation of our model as a Bayesian DAG is given in Figure
+  \ref{fig:dag}. We note that in many settings, the set of features might be
+  different for different drugs. In that case the covariates are indexed by
+  the drug $d$: $\X^d$, a trivial extension of model 
+  (\ref{eq:likelihood}) and (\ref{eq:prior}). This extension is included in the
+  available software, but for clarity it is omitted in the following. 
+  Furthermore, several additional extensions of the model are 
+  described in [! ref to second paper here].
+
+  \begin{figure}
+    \centering
+    \tikz{ %
+      % Y
+      \node[obs] (y) {$y_{id}$}; %
+      \factor[right=of y, yshift=1.5cm, xshift=-0.1cm] {y-f} 
+      {above:$\mathcal{N}$} {} {} ; %
+
+      % X
+      \node[det, above=of y, yshift=1cm] (x) {$x_{ij}$}; %
+
+      % sigma2
+      \node[latent, right=of y] (sigma2) {$\sigma_d^2$}; %
+
+      % beta
+      \node[latent, right=of x] (beta) {$\beta_{jd}$}; %
+      \factor[right=of beta, yshift=-1.4cm, xshift=-0.1cm] {beta-f} 
+      {above:$\mathcal{N}$} {} {} ; %
+
+      % gamma2
+      \node[latent, right=of beta] (gamma2) {$\gamma_{jd}^2$}; %
+      \node[const, right=of gamma2, yshift=0.5cm] (phi) {$\phi_{jd}$}; %
+      \node[const, right=of gamma2, yshift=-0.5cm] (lambdaf) 
+      {$\lambda_{\text{feat}}$}; %
+      \factor[right=of gamma2] {gamma2-f} {above:$\mathcal{IG}$} {phi,lambdaf} 
+      {gamma2} ; %
+      \node[det, right=of gamma2, xshift=2cm, yshift=0.5cm] (c) {$c_{jdg}$}; %
+
+      % tau2
+      \node[latent, right=of sigma2] (tau2) {$\tau_d^2$}; %
+      \node[const, right=of tau2, yshift=0.5cm] (chi) {$\chi_{d}$}; %
+      \node[const, right=of tau2, yshift=-0.5cm] (lambdad) 
+      {$\lambda_{\text{drug}}$}; %
+      \factor[right=of tau2] {tau2-f} {above:$\mathcal{IG}$} {chi,lambdad} 
+      {tau2} ; %
+      \node[det, right=of tau2, xshift=2cm, yshift=0.5cm] (z) {$z_{dh}$}; %
+
+      % edges
+      \factoredge {x,beta,sigma2} {y-f} {y} ; %
+      \factoredge {sigma2,gamma2,tau2} {beta-f} {beta} ; %
+      \edge {c} {phi}; %
+      \edge {z} {chi}; %
+
+      % plates
+      \plate{plateg} {(c)} {$g=1,\dots,G$}
+      \plate{plateh} {(z)} {$h=1,\dots,H$}
+      \plate{platei} {(y) (x)}{$i=1,\dots,n$}; %
+      \plate{platej} {(x) (beta) (gamma2) (phi) (lambdaf) (c) (plateg)}
+      {$j=1,\dots,p$}; %
+      \plate{plated} 
+      {(y) (x) (sigma2) (beta) (gamma2) (tau2) (phi) (lambdaf)
+      (chi) (lambdad) (c) (z) (platei) (platej) (plateg) (plateh)}
+      {$d=1,\dots,D$}; %
+    }
+    \caption{Hierarchical representation of the drug sensitivity prediction 
+    model. Grey circles 
+    represent observed variables, white circles represent unobserved variables,
+    squares represent fixed data, and unenclosed letters are parameters to be
+    estimated. Cell lines are indexed by $i$, features
+    by $j$, drugs by $d$, drug covariates by $h$, and feature covariates by 
+    $g$. The $y_{id}$ are the drug sensitivities, $x_{ij}$ the 
+    molecular features, 
+    $c_{jdg}$ the external feature covariates, $z_{dh}$ the external drug
+    covariates, $\beta_{jd}$ the regression coefficients, $\sigma_d^2$ the 
+    error variances, $\tau_d^2$ and $\gamma_{jd}^2$ the drug and feature
+    specific variance components, respectively, and $\phi_{jd}$, 
+    $\lambda_{\text{feat}}$, $\chi_{d}$, and $\lambda_{\text{drug}}$ the 
+    hyperparameters.}
+    \label{fig:dag}
+  \end{figure}
+
 	\section{Estimation}\label{sec:estimation}
 	\subsection{Variational Bayes}\label{sec:variationalbayes}
-	The posterior corresponding to the model described in (\ref{eq:linearmodel})
-	and (\ref{eq:priormodel}) is not available in closed form. We therefore 
-	approximate the posterior by variational Bayes, where we force the approximate
-	posterior 
-	density to factorise as: $p(\bbeta_d, \sigma_d^2, \gamma_{1d}^2, \dots, 
-	\gamma_{pd}^2) \approx 
-	Q_d(\cdot) = q(\bbeta_d) \cdot q(\sigma_d^2) \cdot \prod_{j=1}^p 
-	q(\gamma_{jd}^2)$. For
+	The posterior corresponding to the model described in (\ref{eq:likelihood})
+	and (\ref{eq:prior}) is not available in closed form. 
+	To avoid computationally intensive markov chain Monte Carlo (MCMC) 
+	algorithms, we approximate the joint posterior by variational Bayes (see 
+	\cite{blei_variational_2017} for a review), where the approximate
+	posterior density factorises as: $p(\bbeta_d, \bm{\gamma}^2_d, \tau_{d}^2, 
+	\sigma_d^2 | \y_d) \approx 
+	Q_d(\cdot) = q(\bbeta_d) \cdot q(\bm{\gamma}_{d}^2) \cdot q(\tau_{d}^2) 
+	\cdot q(\sigma_d^2)$, where $\bm{\gamma}_{d}^2 = 
+	\begin{bmatrix} \gamma_{1d}^2 & \cdots & \gamma_{pd}^2 \end{bmatrix} \tr$. For
   notational convenience, we slightly abuse notation and let $q(\cdot)$ denote
@@
 	\begin{align*}
-    q(\bm{\beta}_d) & \overset{D}{=} \mathcal{N}_p 
+    q(\bm{\beta}_d) & \overset{D}{=} \mathcal{N}_{p} 
     (\bm{\mu}_d, \bm{\Sigma}_d), \\
-    q(\gamma_{jd}^2) & \overset{D}{=} \mathcal{GIG}
-    \(-1, \lambda_{d}[(\mathbf{c}_{d}^j) \tr \bm{\alpha}]^2, \delta_{jd}\), \\
+    q(\bm{\gamma}_{d}^2) & \overset{D}{=} \prod_{j=1}^p \mathcal{GIG}
+    \(-1, \lambda_{\text{feat}}/\phi_{jd}^2, \delta_{jd}\), \\
+    q(\tau^2_d) & \overset{D}{=} \mathcal{GIG} \(-\frac{p + 1}{2}, 
+    \lambda_{\text{drug}}/\chi_d^2, \eta_d\), \\
     q(\sigma^2_d) & \overset{D}{=} \Gamma^{-1} \(\frac{n + p + 1}{2}, \zeta_d\).
@@
 	See SM Section \ref{sm-sec:vbderivations} for the derivations. The variational
-	parameters $\bm{\mu}_d$, $\bm{\Sigma}_d$, $\delta_d$, and
+	parameters $\bm{\mu}_d$, $\bm{\Sigma}_d$, $\delta_{jd}$, $\eta_d$, and
 	$\zeta_d$ contain cyclic dependencies and are iteratively 
 	updated by:
-  \begin{align*}
+	\begin{subequations}\label{eq:vbequations}
+    \begin{align}
       \bm{\Sigma}_d^{(h+1)} & = (a_d^{(h)})^{-1} 
-    \[\X \tr \X + \diag(b_{jd}^{(h)})\]^{-1}, \\
-    \bm{\mu}_d^{(h+1)} & = \[\X \tr \X + \diag(b_{jd}^{(h)})\]^{-1} 
+      \[\X \tr \X + g_{d}^{(h)} \diag(b_{jd}^{(h)})\]^{-1}, \\
+      \bm{\mu}_d^{(h+1)} & = \[\X \tr \X + g_{d}^{(h)} 
+      \diag(b_{jd}^{(h)})\]^{-1} 
       \X \tr \y_d, \\
-    \delta_{jd}^{(h+1)} & = a_d^{(h)} \[(\bmu^{(h+1)}_{jd})^2 + 
-    (\bSigma_d)_{jj}\] + \lambda_{d}, \\ 
+      \delta_{jd}^{(h+1)} & = a_d^{(h)} g_{d}^{(h)} \[(\bmu^{(h+1)}_{jd})^2 + 
+      (\bSigma^{(h + 1)}_d)_{jj}\] + \lambda_{\text{feat}}, \\ 
+      \eta_d^{(h+1)} & = a_d^{(h)} \sum_{j=1}^{p} b_{jd}^{(h + 1)}
+      \[ (\bmu^{(h+1)}_{jd})^2 + (\bSigma^{(h + 1)}_d)_{jj} \] + 
+      \lambda_{\text{drug}}, \\
       \zeta_d^{(h+1)} & = \frac{1}{2} \bigg[\mathbf{y}_d \tr \mathbf{y}_d -
@@
       & \,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\, + 
-    \trace \[ \diag (b_{jd}^{(h+1)}) \bm{\Sigma}_d^{(h+1)}\] + 
-    (\bm{\mu}_d^{(h+1)}) \tr \diag (b_{jd}^{(h+1)}) 
+      g_{d}^{(h + 1)} \trace \[ \diag (b_{jd}^{(h+1)}) \bm{\Sigma}_d^{(h+1)}\] + 
+      g_{d}^{(h + 1)} (\bm{\mu}_d^{(h+1)}) \tr \diag (b_{jd}^{(h+1)}) 
       \bm{\mu}_d^{(h+1)}\bigg],
-  \end{align*}
-  until convergence. Here, we set
+    \end{align}
+  \end{subequations}
+  until convergence, where $\y_d = \begin{bmatrix} y_{1d} & \cdots & y_{nd}
+  \end{bmatrix} \tr$. Here, we set
   \begin{align}
@@
     b_{jd}^{(h)} & = \E_{Q^{(h)}}(\gamma_{jd}^{-2}) = 
-    \sqrt{\frac{\delta_{jd}^{(h)}}
-    {\lambda_{d}[(\mathbf{c}_{d}^j) \tr \bm{\alpha}]^2}}
-    \frac{K_0(\sqrt{\delta_{jd}^{(h)}\lambda_{d}}
-    [(\mathbf{c}_{d}^j) \tr \bm{\alpha}])}
-    {K_1(\sqrt{\delta_{jd}^{(h)}\lambda_{d}}
-    [(\mathbf{c}_{d}^j) \tr \bm{\alpha}])} + 
-    \frac{2}{\delta_{jd}^{(h)}}, \label{eq:ratiomodifiedbessel}
+    \sqrt{\frac{\lambda_{\text{feat}}}{\phi_{jd}^2 \delta_{jd}^{(h)}}}
+    \frac{K_0\(\sqrt{\delta_{jd}^{(h)}\lambda_{\text{feat}}/\phi_{jd}^2}\)}
+    {K_1\(\sqrt{\delta_{jd}^{(h)}\lambda_{\text{feat}}/\phi_{jd}^2}\)} + 
+    \frac{2}{\delta_{jd}^{(h)}}, \label{eq:ratiomodifiedbessel} \\
+    g_d^{(h)} & =\E_{Q^{(h)}}(\tau_d^{-2})=
+    \sqrt{\frac{\lambda_{\text{drug}}}{\chi_{d}^2 \eta_{d}^{(h)}}}
+    \frac{K_{(p-1)/2}\(\sqrt{\eta_{d}^{(h)}\lambda_{\text{drug}}/\chi_{d}^2}\)}
+    {K_{(p+1)/2}\(\sqrt{\eta_{d}^{(h)}\lambda_{\text{drug}}/\chi_{d}^2}\)} + 
+    \frac{p + 1}{\eta_{d}^{(h)}}.
+    \nonumber
   \end{align}
@@
   \subsection{Empirical Bayes}\label{sec:empiricalbayes}
-  To make use of the external data, we parametrise the prior mean of the
-  $\gamma_{jd}^2$ as $[(\mathbf{c}_{d}^j) \tr \bm{\alpha}]^{-1}$. 
-  This allows us to
-  include continuous external variables $\mathbf{c}_{d}^j$ into the model. 
-  Additionally, it reduces the number of hyperparameters from $pD$ to 
-  $|\bm{\alpha}| + D$. A full Bayesian model
+  We parametrised the prior 
+  mean of the $\gamma_{jd}^2$ as $\phi_{jd}=(\mathbf{c}_{jd} \tr 
+  \bm{\alpha}_{\text{feat}})^{-1}$ and the prior mean of $\tau_d^2$ as
+  $\chi_{d}=(\mathbf{z}_{d} \tr \bm{\alpha}_{\text{drug}})^{-1}$.
+  This parametrisation allows us to include feature and drug covariates,
+  both continuous and discrete, into the 
+  model. Additionally, it reduces the number of hyperparameters from 
+  $pD$ to $|\bm{\alpha}_{\text{feat}}| + |\bm{\alpha}_{\text{drug}}| + 2$. The
+  Bayesian model
   then requires the specification of the hyperparameters 
-  $\bm{\alpha}$ and $\blambda = \begin{bmatrix} \lambda_1 & \cdots & \lambda_D 
-  \end{bmatrix} \tr$. These are abstract and hard to interpret parameters 
+  $\bm{\alpha}=\begin{bmatrix} \bm{\alpha}_{\text{feat}} \tr & 
+  \bm{\alpha}_{\text{drug}} \tr 
+  \end{bmatrix} \tr$ and $\bm{\lambda} = \begin{bmatrix} \lambda_{\text{feat}} &
+  \lambda_{\text{drug}}
+  \end{bmatrix} \tr$. These are
+  abstract and hard to interpret parameters 
   for which we generally lack expert knowledge. They do, however, have a 
@@
   therefore propose to estimate these hyperparameters by empirical Bayes.
-  Simply put, empirical Bayes fits the prior model to the data and is, as such, 
-  objective Bayes (up to model specification). In our case, this results in an   
-  objective and data-driven inclusion of the external data.
-
-	The canonical method for empirical Bayes is maximisation of the marginal 
-	likelihood with respect to the hyper parameters. In 
+  In our case, this results in an   
+  objective and data-driven inclusion of the feature and drug covariates.
+	The canonical method for empirical Bayes is to maximise the marginal 
+	likelihood with respect to the hyperparameters. In 
 	\cite{casella_empirical_2001} the marginal likelihood is maximised by an EM 
@@
 	  \bm{\lambda}}{\argmax}\E_{\cdot | \Y} 
-	  [\log p(\Y, \B, \bm{\Gamma}^2, \bsigma^2) | \bm{\alpha}^{(l)},
+	  [\log p(\Y, \B, \bm{\Gamma}^2, \bm{\tau}^2, \bsigma^2) | \bm{\alpha}^{(l)},
 	  \bm{\lambda}^{(l)}] \\
 	  & = \underset{\bm{\alpha},\bm{\lambda}}{\argmax} 
-	  \E_{\cdot | \Y} [\log \pi (\bm{\Gamma}^2) | \bm{\alpha}^{(l)},
-	  \bm{\lambda}^{(l)}],
+	  \E_{\cdot | \Y} [\log \pi (\bm{\Gamma}^2) | \bm{\alpha}_{\text{feat}}^{(l)},
+	  \bm{\lambda}_{\text{feat}}^{(l)}] + 
+	  \E_{\cdot | \Y} [\log \pi (\bm{\tau}^2) | 
+	  \bm{\alpha}_{\text{drug}}^{(l)}, \bm{\lambda}_{\text{drug}}^{(l)}],
 	\end{align*}
 	where $\Y = \begin{bmatrix} \y_1 & \cdots & \y_D \end{bmatrix}$,
-	$\B = \begin{bmatrix} \bbeta_1 & \cdots & \bbeta_D \end{bmatrix}$, 
+	$\B = \begin{bmatrix} \bbeta_1, \dots, \bbeta_D \end{bmatrix}$, 
+	$\bm{\tau}^2 = \begin{bmatrix} \tau^2_1 & \cdots & 
+	\tau^2_D \end{bmatrix} \tr$,
 	$\bsigma^2 = \begin{bmatrix} \sigma^2_1 & \cdots & 
@@
 	and $\bm{\Gamma}^2 = \begin{bmatrix} \bgamma^2_1 & \cdots & \bgamma^2_D 
-	\end{bmatrix}\tr$, with 
-	$\bgamma^2_d = \begin{bmatrix} \gamma^2_{1d} & \cdots & \gamma^2_{pd} 
-	\end{bmatrix}\tr$,
+	\end{bmatrix}$,
 	and the expectation is with respect to the joint posterior. In our case, this 
@@
 	\underset{\bm{\alpha},\bm{\lambda}}{\argmax}\E_{Q^{(l)}} 
-	[\log \pi(\bm{\Gamma}^2)],
+	[\log \pi(\bm{\Gamma}^2)| \bm{\alpha}_{\text{feat}}^{(l)}, 
+	\bm{\lambda}_{\text{feat}}^{(l)}] + 
+	\E_{Q^{(l)}} [\log \pi(\bm{\tau}^2)| 
+	\bm{\alpha}_{\text{drug}}^{(l)}, \bm{\lambda}_{\text{drug}}^{(l)}],
 	$$
 	where now the expectation is with respect to the converged variational 
-	posterior $Q^{(l)}=\prod_{d=1}^D Q_d^{(l)}$. The resulting estimating 
-	equations are:
+	posterior $Q^{(l)}=\prod_{d=1}^D Q_d^{(l)}$.
+
+	If we stack the drug and feature covariates:
 	\begin{align*}
-    \bm{\alpha} & = \[ \mathbf{C} \tr \bLambda \diag 
-    (e_{jd}^{(l)}) \mathbf{C} \]^{-1} \mathbf{C} \tr \bLambda
+	  \mathbf{C} = 
+	  \begin{bmatrix} \mathbf{c}_{11} \tr \\ 
+	    \vdots \\
+	    \mathbf{c}_{p1} \tr \\
+	    \vdots \\
+	    \mathbf{c}_{1D} \tr \\
+	    \vdots \\
+	    \mathbf{c}_{pD} \tr
+	  \end{bmatrix} \text{ and }
+	  \mathbf{Z} =
+	  \begin{bmatrix} \mathbf{z}_{1} \tr \\ 
+	    \vdots \\
+	    \mathbf{z}_D \tr
+	  \end{bmatrix},
+	\end{align*}
+	the empirical Bayes updates are given by:
+  \begin{align*}
+    \bm{\alpha}^{(l+1)}_{\text{feat}} & = \[ \mathbf{C} \tr \diag 
+    (e_{jd}^{(l)}) \mathbf{C} \]^{-1} \mathbf{C} \tr 
     \mathbf{1}_{pD \times 1}, \\
-    \lambda_{d} & = p \left\{ \sum_{j=1}^p b_{jd}^{(l)} + 
-    \sum_{j=1}^p e_{jd}^{(l)} [(\mathbf{c}_{d}^j) \tr \bm{\alpha}]^2 - 
-    2 \sum_{j=1}^p (\mathbf{c}_{d}^j) \tr \bm{\alpha} \right\}^{-1},
+    \lambda^{(l+1)}_{\text{feat}} & = pD \[ \sum_{d=1}^D \sum_{j=1}^{p} 
+    b_{jd}^{(l)} + (\bm{\alpha}^{(l+1)}_{\text{feat}}) \tr \mathbf{C} \tr 
+    \diag (e_{jd}^{(l)}) \mathbf{C} \bm{\alpha}^{(l+1)}_{\text{feat}} - 
+    2 (\bm{\alpha}^{(l+1)}_{\text{feat}}) \tr \mathbf{C} \tr 
+    \mathbf{1}_{pD \times 1} \]^{-1},\\
+    \bm{\alpha}^{(l+1)}_{\text{drug}} & = \[ \mathbf{Z} \tr \diag 
+    (f_{d}^{(l)}) \mathbf{Z} \]^{-1} \mathbf{Z} \tr 
+    \mathbf{1}_{D \times 1}, \\
+    \lambda^{(l+1)}_{\text{drug}} & = D \[ \sum_{d=1}^D g_{d}^{(l)} + 
+    (\bm{\alpha}^{(l+1)}_{\text{drug}}) \tr \mathbf{Z} \tr \diag (f_{d}^{(l)})
+    \mathbf{Z} \bm{\alpha}^{(l+1)}_{\text{drug}} - 
+    2 (\bm{\alpha}^{(l+1)}_{\text{drug}}) \tr \mathbf{Z} \tr 
+    \mathbf{1}_{D \times 1} \]^{-1},
   \end{align*}
-  where $\bLambda=\diag(\lambda_d) \otimes \I_p$
-  and $e_{jd}^{(l)} = \E_{Q^{(l)}}(\gamma_{jd}^2) = (b_{jd}^{(l)} - 
-  2/\delta_{jd}^{(l)}) \cdot \delta_{jd}^{(l)}/[\lambda_{d}^{(l)}
-  \left\{(\mathbf{c}_{d}^j) \tr \bm{\alpha}^{(l)}]^2\right\}$. 
-  Solving the equations is done by 
-  iteratively reweighted least squares.
+  where 
+  \begin{align*}
+    e_{jd}^{(l)} & = \E_{Q^{(l)}}(\gamma_{jd}^2| 
+    \bm{\alpha}_{\text{feat}}^{(l)}, 
+    \lambda_{\text{feat}}^{(l)}) = (b_{jd}^{(l)} - 
+    2/\delta_{jd}^{(l)}) \cdot \delta_{jd}^{(l)} (\phi_{jd}^{(l)})^2/
+    \lambda_{\text{feat}}^{(l)}, \\
+    f_{d}^{(l)} & = \E_{Q^{(l)}}(\tau_{d}^2| \bm{\alpha}_{\text{drug}}^{(l)}, 
+    \lambda_{\text{drug}}^{(l)}) = (g_{d}^{(l)} - 
+    (p + 1)/\eta_{d}^{(l)}) \cdot \eta_{d}^{(l)} (\chi_{d}^{(l)})^2/
+    \lambda_{\text{drug}}^{(l)}.
+  \end{align*}
 
-  Variaional Bayes approximations are known to underestimate posterior 
+  To ensure proper and unbiased shrinkage, intercepts are included in 
+  $\bm{\alpha}_{\text{feat}}$ and $\bm{\alpha}_{\text{drug}}$. This is achieved 
+  by appending both
+  $\mathbf{C}$ and $\mathbf{Z}$ with a column of 
+  ones. These intercepts are roughly interpreted as the
+  expected prior precisions $\E(\gamma_{jd}^{-2})$ and $\E(\tau_d^{-2})$
+  if the feature and drug covariates are all zero. 
+  Likewise, an $\alpha$ corresponding to an external covariate may be
+  interpreted as an additive effect of the external covariate on the prior
+  expected precision. So an $\alpha=1$ translates to an increase in expected 
+  prior precision of
+  1 for every increase in the external covariate of 1,
+	keeping all the other external covariates fixed.
+
+  Variational Bayes approximations are known to underestimate posterior 
   variances \cite[]{rue_approximate_2009,consonni_mean-field_2007,
@@
   required, 
-  we suggest generating samples from the posterior after the empirical Bayes 
-  iterations have converged. A Gibbs sampler to do so is described in 
-  the SM Section \ref{sm-sec:gibbssampler}. Alternatively, the \texttt{R} 
-  package \texttt{rstan} \cite[]{guo_rstan:_2018} allows the user to sample
-  from a generic posterior. An implementation of model
-  (\ref{eq:priormodel}) in \texttt{stan} code may be found on 
+  we suggest generating samples from the posterior with fixed hyperparameters
+  estimates (after the procedure described in Section \ref{sec:empiricalbayes}
+  has converged). A Gibbs sampler is described in 
+  SM Section \ref{sm-sec:gibbssampler}. Alternatively, we provide an
+  implementation of the proposed model in stan using the R package rstan 
+  \cite[]{guo_rstan:_2018} at 
   \url{https://github.com/magnusmunch/cambridge}.
 
-  Note that so far, the design matrix $\X$ is assumed to be shared across the
-  equations. This is not a requirement. That is, we may have D different design 
-  matrices $\X^d$ that need not have the same number of 
-  columns, i.e., we have $p_d$ instead of $p$. This becomes useful in, e.g. 
-  eQTLs detection, where we regress genes on a gene-specific selection
-  of SNPs, often based on SNP-gene distance.
+	\section{Simulations}
+	\subsection{Fixed drug component}
+	This section investigates the empirical 
+	Bayes estimation properties of the model in a simulated setting; its main aim 
+	is to assess hyperparameter recovery. It is a data based simulation,
+	wherein the responses are simulated from a synthetic model, but the
+	features are taken from the real GDSC data introducted in Section 
+	\ref{sec:gdsc}. 
+	The real GDSC features contain strong collinearities. Such strong
+	collinearities in the design matrix impede correct parameter estimation. 
+	We therefore replace the ambition of correctly estimating the $\beta_{jd}$
+	with the more modest aim of approximately correct estimation of the 
+	hyperparameters.
 
-  \section{Simulations}\label{sec:simulations}
-  We conducted a simulation study consisting of three parts investigating (i) 
-  the NIG prior performance, (ii) the variational Bayes approximation, and (iii)
-  the empirical Bayes estimation.
+	A pre-processing step, as explained in Section \ref{sec:gdsc}, selects around 
+	300 features per drug, while the 168 drug sensitivities for 482 cell lines are 
+	simulated from model (\ref{eq:likelihood}) and 
+	(\ref{eq:prior}), with two adjustments: we fix $\forall d: \tau_d^2=1$ 
+	and draw the error variances as $\forall d: \sigma_d^2 \sim \Gamma^{-1}(3,2)$,
+	such that the prior $\sigma_d^2$ mean and variance are both one.
+	We create four external dummy feature covariates that code for four 
+	randomly assigned groups of features. We set $\bm{\alpha}_{\text{feat}}$ such
+	that the $\gamma_{jd}^2$ of the four groups of features have prior means
+	$\phi_{jd} \in \{ 1, 1/2, 1/4, 1/8 \}$ 
+	($\bm{\alpha}_{\text{feat}}=\begin{bmatrix} 1 & 1 & 3 & 7 \end{bmatrix} \tr$). 
+	The prior scale parameter is set to $\lambda_{\text{feat}}=1$. We estimate 
+	four models: (i) the NIG model as in Section \ref{sec:estimation}, (ii) an 
+	NIG model where we fix $\lambda_{\text{feat}}$ to its true value, and two NIG
+	models without the inclusion of the external covariates, again with 
+	$\lambda_{\text{feat}}$ either (iii) estimated or (iv) fixed to its true 
+	value. Exclusion of the external covariates as in models (iii) and (iv) 
+	amounts to direct estimation of one common expected 
+	prior mean $\phi$, instead of regression estimates for the $\phi_{jd}$ as
+	in models (i) and (ii). All 
+	models are fit on 241 randomly selected cell lines, while performance measures 
+	are estimated from the remaining 241 cell lines.
 
-  \subsection{NIG prior performance}
-  \label{sec:inversegammamodel}
-  The aim of this part is to investigate the 
-  performance of the NIG prior model. Since we may do so in the univariate 
-  setting, so we drop all $d$ indices in the following. 
-  A common default prior for latent variance components is the inverse Gamma 
-  prior:
-  $$
-  \gamma_j^2 \sim \Gamma^{-1}(\eta/2,\xi/2),
-  $$
-  with shape $\eta/2$ and scale $\xi/2$, both rescaled to match our 
-  inverse Gaussian prior formulation. The resulting marginal prior for the model
-  parameters
-  (conditional on the error variance $\sigma^2$) is then a scaled Student's 
-  $t$-distribution:
-  $$
-  \beta_j | \sigma^2 \sim \text{Student's } 
-  t\(\eta,\sigma \sqrt{\xi/\eta}\),
-  $$
-  with degrees of freedom $\eta$ and scale $\sigma \sqrt{\xi/\eta}$.
-  See SM Section \ref{sm-sec:inversegamma} for more details on this prior. 
-  We compare the 
-  posteriors of the proposed NIG to the scaled Student's 
-  $t$-distribution prior in a simulated setting.
-
-  We compare the posterior distributions to the true posterior for a 
-  number of measures. Divergence between the distributions
-  is quantified through a multivariate extension of the Cram{\'e}r-von Mises
-  criterion \cite[]{baringhaus_new_2004}, where smaller values of the criterion
-  indicate a better fit. We also assess 95\% coverage of the 
-  posterior credible intervals. The accuracy of the 
-  point estimates of $\beta_j$ in the form of posterior means is assessed by
-  the mean squared error (MSE) and correlation between the point estimates 
-  and the true $\beta_j$.
-
-  We studied two settings: one dense and one sparse. In the dense setting, 
-  we simulated $n=100$ features $\x_i$ from a $p=200$ dimensional Gaussian 
-  distribution with zero mean vector, variances equal to one, and all 
-  covariances set to $\rho=0.5$. The outcomes were generated through $y_i = 
-  \x_i\tr \bbeta + \epsilon_i$, with the $\epsilon_i$ from the standard normal 
-  distribution. 
-  Likewise, we generated the model parameters $\beta_j$ from a standard normal
-  distribution.
-
-  We choose the hyperparameters such that we get all combinations of marginal
-  prior variances and kurtoses $\V(\beta_{j} | \sigma^2) \in 
-  \{ 1/2, 1, 2 \}$ and $\mathcal{K}(\beta_{j} | \sigma^2) \in 
-  \{ 4, 10 \}$. This leads to the settings in Table \ref{tab:hyper_set1}. For 
-  all priors, we generated 1000 draws (after 1000 burnin) from the 
-  posteriors using an MCMC sampler implemented in \texttt{rstan}. We repeated
-  the simulation 100 times and report 
-  results in Figures \ref{fig:boxplots_igaussian_res4.1_post1} and 
-  \ref{fig:boxplots_igaussian_res4.1_post2}. 
-<<hyper_set1>>= 
-set <- read.table("results/simulations_igaussian_set4.csv")
-kableExtra::add_header_above(kableExtra::kable_styling(
-  knitr::kable(set[, c("ctalphainv", "lambda", "eta", "xi", "vars", "kurts")],
-               col.names=c("$(\\mathbf{c} \\tr \\bm{\\alpha})^{-1}$",
-                           "$\\lambda$", "$\\eta$", "$\\xi$",
-                           "$\\V(\\beta_{j} | \\sigma^2)$",
-                           "$\\mathcal{K}(\\beta_{j} | \\sigma^2)$"),
-               caption="Hyperparameter settings",
-               "latex", booktabs=TRUE, escape=FALSE, digits=2),
-  latex_options=c("HOLD_position")),
-  header=c(" "=1, "NIG"=2, "Student's t"=2, "Prior"=2))
+<<include=FALSE>>=
+load("results/simulations_gdsc_set1.Rdata")
+mapmse <- apply(res[, c(7:12)], 2, median)
+mapmset <- apply(res[, c(13:18)], 2, median)
+melbo <- apply(elbo, 2, median)
+phi <- 1/(set$alphaf %*% t(unname(model.matrix(~ factor(1:4)))))
+mphi <- apply(1/est[, c(11, 13)], 2, median)
+mlambdaf <- apply(est[, c(5, 12)], 2, median)
+@	
+	Figure \ref{fig:simulations_gdsc_est1} shows the estimated
+	$\bm{\alpha}_{\text{feat}}$ together with its true value for models (i) and 
+	(ii). The figure shows that $\bm{\alpha}_{\text{feat}}$ is consistently
+	underestimated 
+	in the full NIG model (both with estimated and fixed $\lambda_{\text{feat}}$),
+	while $\lambda_{\text{feat}}$ is overestimated as 
+	\Sexpr{round(mlambdaf[1], 2)} in model (i) (the true value is 1). 
+	This translates into
+	underestimated prior means $\phi_{jd}$ as presented in Figure 
+	\ref{fig:simulations_gdsc_phi1}, which in turn implies underestimated 
+	$\beta_{jd}$ variances. The overestimated $\lambda_{\text{feat}}$ translates 
+	into underestimated $\gamma_{jd}^2$ variances. Underestimation of variances is 
+	a common phenomenon in variational Bayes approximations. In addition, 
+	$\bm{\alpha}_{\text{feat}}$ and $\lambda_{\text{feat}}$ are higher level 
+	parameters, which are, in general, hard to estimate. We note 
+	that even though $\bm{\alpha}_{\text{feat}}$ and thus $\phi_{jd}$ are 
+	estimated with bias, the orders of their elements are correct.
+<<simulations_gdsc_est1, fig.cap="Estimated hyperparameters $\\bm{\\alpha}_{\\text{feat}}$ from simulations with (a) estimated and (b) fixed $\\lambda_{\\text{feat}}$ (models (i) and (ii), respectively).", out.width="100%", fig.asp=1/2>>=
 @
 
-<<boxplots_igaussian_res4.1_post1, fig.cap="Comparison of the NIG and Student's t priors in the dense setting in (a)-(d) simulation setting 1, (e)-(h) setting 2, and (i)-(l) setting 3, through Cramér-von Mises criterion between the model and true posteriors, correlation and MSD between the posterior means and true parameters, and 95\\% credible interval coverage of the true values.", out.width="100%", fig.asp=1>>=
+<<simulations_gdsc_phi1, fig.cap="Estimated prior means $\\bm{\\phi}_{jd}$ for the four groups of features from simulations with (a) estimated and (b) fixed $\\lambda_{\\text{feat}}$ (models (i) and (ii), respectively).", out.width="100%", fig.asp=1/2>>=
 @
 
-<<boxplots_igaussian_res4.1_post2, fig.cap="Comparison of the NIG and Student's t priors in the dense setting in (a)-(d) simulation setting 4, (e)-(h) setting 5, and (i)-(l) setting 6, through Cramér-von Mises criterion between the model and true posteriors, correlation and MSD between the posterior means and true parameters, and 95\\% credible interval coverage of the true values.", out.width="100%", fig.asp=1>>=
+	The models (iii) and (iv) give median $\phi$ estimates of
+	\Sexpr{round(mphi[1], 2)} and \Sexpr{round(mphi[2], 2)}, 
+	respectively. The $\phi$ estimates are
+	underestimated, compared to true $\phi$ mean \Sexpr{round(mean(phi), 2)}. 
+	Model (iii) slightly underestimates the median
+	$\lambda_{\text{feat}}$ at \Sexpr{round(mlambdaf, 2)[2]} 
+	(the true value is one) and thus overestimates the 
+	prior $\gamma_{jd}^2$ variance. This is likely to compensate for the common
+	prior mean 
+	$\phi$, that is not able to capture part of the variation in the true feature
+	specific $\phi_{jd}$. The median evidence 
+	lower bounds (ELBO), which approximate the marginal likelihoods of models
+	(i)-(iv), are estimated as \Sexpr{round(melbo, 2)[1:3]}, and 
+	\Sexpr{round(melbo, 2)[4]}, respectively. This implies that model (i),
+	with estimated $\bm{\alpha}_{\text{feat}}$ and $\lambda_{\text{feat}}$ fits
+	the data best (the log Bayes factors comparing models (ii)-(iv) to model (i) 
+	are \Sexpr{round(melbo[-c(1, 4)] - melbo[1], 2)}, and 
+	\Sexpr{round(melbo[4] - melbo[1], 2)}, respectively, and thus highly favouring
+	model (i)). The prediction MSEs for the four models on the test 
+	data and gaps between training and test prediction MSE are displayed in Table
+	\ref{tab:pmse1}. Table \ref{tab:pmse1} confirms that the models that
+	include the external data (models (i) and (ii)) learn the underlying structure 
+	in the data better than models (iii) and (iv) that do not include the external 
+	data.
+<<pmse1>>= 
+library(knitr, quietly=TRUE, verbose=FALSE)
+suppressWarnings(library(kableExtra))
+load("results/simulations_gdsc_set1.Rdata")
+mapmse <- apply(res[, c(7:12)], 2, median)
+mapmset <- apply(res[, c(13:18)], 2, median)
+tab <- rbind(mapmse, mapmse - mapmset)[, -c(5, 6)]
+rownames(tab) <- c("$\\text{MSE}_{\\text{test}}$",
+                  "$\\text{MSE}_{\\text{test}} - \\text{MSE}_{\\text{train}}$")
+options(knitr.kable.NA="", digits=3)
+kable(tab,
+      col.names=c("(i)", "(ii)", "(iii)", "(iv)"),
+      caption="Median prediction MSEs for the models (i)-(iv) in the fixed 
+      $\\tau_d^2$ simulation setting.", 
+      format="latex", booktabs=TRUE, escape=FALSE) %>%
+  kable_styling(latex_options=c("HOLD_position"))
 @
 
-  [! describe results here]
+  \subsection{Random drug component}
+	To investigate the interplay of the feature and drug variance components, we 
+	repeat the whole simulation study with random $\tau_d^2$. Similarly as with 
+	the $\gamma_{jd}^2$, we randomly create four groups of drugs and corresponding
+	external drug dummy covariates. We simulate the $\tau_d^2$ as in
+	(\ref{eq:tauprior}), where we set $\bm{\alpha}_{\text{drug}} = 
+	\begin{bmatrix} 1 & 1 & 3 & 7 \end{bmatrix} \tr$, such
+	that the $\tau_{d}^2$ of the four groups of features have prior means
+	$\chi_{d} \in \{ 1, 1/2, 1/4, 1/8 \}$. The scale $\lambda_{\text{drug}}$ is 
+	set to one. We estimate the same four models as in the fixed $\tau_d^2$ 
+	setting, but in addition estimate the $\tau_d^2$ priors as well. Again, we
+	use two different versions: one where we estimate 
+	$\bm{\lambda}$ and one where we fix it to its true value.
 
-  In the sparse setting, the observed data generating process is the same as in
-  the dense setting. The
-  model parameters are simulated from an (absolutely continuous) spike-and-slab 
-  prior with both Gaussian spike and slab:
-  \begin{align*}
-    w & \sim \text{Beta}(\phi, \chi), \\
-    z_j | w & \sim \text{Bernoulli}(w), \\
-    \beta_j | z_j & \sim z_j \mathcal{N} \( 0, 1 + 0.99 \frac{\chi}{\phi}\) +
-    (1 - z_j)\mathcal{N} \( 0, 0.01\).
-  \end{align*}
-  The slab is chosen
-  such that the we have marginal prior variance $\V(\beta_j)=1$. The sparsity 
-  level is determined by the proportion of zero $\beta_j$: 
-  $s=1 - \phi/(\phi + \chi)$. Note that the resulting marginal prior for
-  $\beta_j$ is:
-  $$
-  \beta_j \sim (1 - s)\mathcal{N} \( 0, \frac{1 - 0.01 s}{1 - s}\) + 
-  s \mathcal{N} \( 0, 0.01\).
-  $$
-  We use two settings, $s=0.5$ and $s=0.9$. The 
-  hyperparameters for estimation are set the same as in the dense setting, 
-  resulting in a total of 12 different simulation settings 
-  (6 per sparsity setting).
-  To mitigate the effect of random variation, we
-  repeated the simulations 100 times and report the results in Figures 
-  \ref{fig:boxplots_igaussian_res5_post1} and 
-  \ref{fig:boxplots_igaussian_res5_post2}.
+<<include=FALSE>>=
+load("results/simulations_gdsc_set2.Rdata")
+mapmse <- apply(res[, c(7:12)], 2, median)
+mapmset <- apply(res[, c(13:18)], 2, median)
+melbo <- apply(elbo, 2, median)
+phi <- 1/(set$alphaf %*% t(unname(model.matrix(~ factor(1:4)))))
+chi <- 1/(set$alphad %*% t(unname(model.matrix(~ factor(1:4)))))
+mphi <- apply(1/est[, c(21, 25)], 2, median)
+mchi <- apply(1/est[, c(22, 26)], 2, median)
+mlambdaf <- apply(est[, c(9, 23)], 2, median)
+mlambdad <- apply(est[, c(10, 24)], 2, median)
+@	
+	Figures \ref{fig:simulations_gdsc_est2} and 
+	\ref{fig:simulations_gdsc_phi2} display the estimated $\bm{\alpha}$, 
+	$\phi_{jd}$, and $\chi_d$ for models (i) and (ii). The scale parameters were
+	estimated as $\lambda_{\text{feat}}=\Sexpr{round(mlambdaf[1], 2)}$ and
+	$\lambda_{\text{drug}}=\Sexpr{round(mlambdad[1], 2)}$ in model (i). From 
+	Figure \ref{fig:simulations_gdsc_est2} we see that estimation of
+	$\bm{\alpha}_{\text{feat}}$ is again biased. However, from 
+	Figure \ref{fig:simulations_gdsc_phi2} we see that the scale and ordering of
+	the $\phi_{jd}$ are correclty estimated. We conjecture that the drop in 
+	estimation accuracy compared to the previous setting stems from the
+	difficult estimation of the $\tau_d^2$ prior. Estimation of this prior 
+	depends on only 168 observed drugs, while the $\gamma_{jd}^2$ prior is 
+	is estimated from approximately $300 \cdot 168=50400$ features. 
+	\cite{wiel_learning_2019} show in a simple regression setting that if the 
+	number of samples and the number of parameters on which the EB is based are of
+	the same order, empirical Bayes performs worse than when one of the two is
+	large compared to the other. 
+	Here, the number of samples $n=241$ and $D=168$ are of the same order, leading
+	to a suboptimal EB estimation performance.
 
-<<boxplots_igaussian_res5_post1, fig.cap="Comparison of the NIG and Student's t priors in the sparse setting in (a)-(d) simulation setting 1, (e)-(h) setting 2, and (i)-(l) setting 3, through Cramér-von Mises criterion between the model and true posteriors, correlation and MSD between the posterior means and true parameters, and 95\\% credible interval coverage of the true values.", out.width="100%", fig.asp=1>>=
+<<simulations_gdsc_est2, fig.cap="Estimated hyperparameters $\\bm{\\alpha}_{\\text{feat}}$ and $\\bm{\\alpha}_{\\text{drug}}$ from simulations with (a) \\& (c) estimated and (b) \\& (d) fixed $\\lambda_{\\text{feat}}$ and $\\lambda_{\\text{drug}}$ (models (i) and (ii), respectively).", out.width="100%", fig.asp=1>>=
 @
 
-<<boxplots_igaussian_res5_post2, fig.cap="Comparison of the NIG and Student's t priors in the sparse setting in (a)-(d) simulation setting 4, (e)-(h) setting 5, and (i)-(l) setting 6, through Cramér-von Mises criterion between the model and true posteriors, correlation and MSD between the posterior means and true parameters, and 95\\% credible interval coverage of the true values.", out.width="100%", fig.asp=1>>=
+<<simulations_gdsc_phi2, fig.cap="Estimated prior means $\\bm{\\phi}_{jd}$ and $\\bm{\\chi}_{d}$ for the four groups of features and drugs from simulations with (a) \\& (c) estimated and (b) \\& (d) fixed $\\lambda_{\\text{feat}}$ and $\\lambda_{\\text{drug}}$ (models (i) and (ii), respectively).", out.width="100%", fig.asp=1>>=
 @
 
-  [! describe results here]
-
-  \subsection{Variational Bayes approximation}
-  To investigate the accuracy of the variational Bayes approximation we compare 
-  the variational Bayes posterior to 1000 (after 1000 burn-in) MCMC samples from
-  the posterior obtained with \texttt{rstan}. Since the MCMC
-  samples represent the exact posterior (up to Monte Carlo error), we use it as 
-  a proxy for the true posterior. We include the comparison of our variational
-  Bayes approximation to the approximate posterior from the
-  automatic differentiation variational inference (ADVI)
-  algorithm in \texttt{rstan} \cite[]{kucukelbir_automatic_2015}. ADVI is a
-  variational Bayes algorithm that assumes a full factorization of the posterior
-  and is fit through gradient based optimization of the variational objective.
-  \texttt{rstan} offers an ADVI algorithm that assumes a fullrank posterior
-  covariance, but this algorithm often failed to converge or resulted in errors, 
-  so we do not include it in the comparisons. We measure divergence between the 
-  MCMC and variational posteriors with the Cram{\'e}r-von Mises criterion.
-
-  In addition to ADVI, \texttt{rstan} allows the user
-  to maximise the posterior to obtain the maximum \textit{a posteriori} (MAP)
-  estimates of the model parameters. We compare these MAP estimates together
-  with the ridge estimates obtained from the \texttt{glmnet} \texttt{R} package
-  \cite[]{friedman_regularization_2010}, and the ADVI posterior means,
-  to the MAP estimates obtained from our model through the correlation with
-  and the mean squared distance (MSD) from the MCMC posterior means.
-
-  The interest of this simulation lies in the variational Bayes approximation, 
-  so we restrict ourselves to the univariate setting again (i.e., $D=1$). The
-  simulation setup is the same as in the dense setting 
-  in Section \ref{sec:inversegammamodel}. Similarly, we use the same 
-  hyperparameters settings (Table \ref{tab:hyper_set1}. For \texttt{glmnet}
-  we either cross-validate the penalty parameter $\lambda$ or match it to the
-  true prior variance. We
-  repeated the simulations 100 times and report the results in Figure 
-  \ref{fig:boxplots_igaussian_res4.2_post1}.
-
-<<boxplots_igaussian_res4.2_post1, fig.cap="(a) \\& (d) Cramér-von Mises criterion, (b) \\& (e) correlation, and (c) \\& (f) MSD between the $\\beta_j$ point estimates of the MCMC and variational bayes posteriors.", out.width="80%", fig.asp=2/3>>=
+  We argue however, that even though the prior parameters are not estimated
+  correctly, there is still some gain in predictive performance.
+  Prediction MSEs on the test data the training-test gaps in prediction MSE are 
+  shown in Table \ref{tab:pmse2}. The comparisons of models (i) and
+	(ii) vesus models (iii) and (iv) confirm that even though EB estimation is
+	suboptimal, the external data do add something in predictive performance. 
+<<pmse2>>= 
+library(knitr, quietly=TRUE, verbose=FALSE)
+suppressWarnings(library(kableExtra))
+load("results/simulations_gdsc_set2.Rdata")
+mapmse <- apply(res[, c(7:12)], 2, median)
+mapmset <- apply(res[, c(13:18)], 2, median)
+tab <- rbind(mapmse, mapmse - mapmset)[, -c(5, 6)]
+rownames(tab) <- c("$\\text{MSE}_{\\text{test}}$",
+                  "$\\text{MSE}_{\\text{test}} - \\text{MSE}_{\\text{train}}$")
+options(knitr.kable.NA="", digits=3)
+kable(tab,
+      col.names=c("(i)", "(ii)", "(iii)", "(iv)"),
+      caption="Median prediction MSEs for the models (i)-(iv) in the random 
+      $\\tau_d^2$ simulation setting.", 
+      format="latex", booktabs=TRUE, escape=FALSE) %>%
+  kable_styling(latex_options=c("HOLD_position"))
 @
 
-  [! describe results here]
+	\section{GDSC data}\label{sec:gdsc}
+	\subsection{Primary data}
+	The GDSC project's \cite[]{yang_genomics_2013} aim is ``to improve cancer 
+	treatments by discovering therapeutic biomarkers that can be used to identify 
+	patients most likely to respond to anticancer drug". Part of the project is to
+	screen $>1000$ human cancer cell lines for drug sensitivities. The cell lines
+	have been genetically characterised and several drug sensitivity measures are
+	recorded.
+	The data is freely available from \cite{garnett_systematic_2012} and consist 
+	of: (i) the sensitivity measures of the cell lines 
+	to the drugs, (ii) annotation of the 
+	screened compounds, and (iii) the cell lines' genomic profile (mutations, 
+	copy numbers, methylation profiles, and gene expression). We will attempt to 
+	predict drug sensitivities of the cell lines, as quantified by half maximal
+	inhibitory concentration (IC50), using the
+	gene expression data. Other choices of sensitivity measures than IC50 are 
+	possible, but a discussion on the pros and cons of different sensitivity
+	measures is beyond the aim of this paper. A comprehensive penalized regression
+	approach such as in 
+	{\cite{mai_composite_2019} uses all the available molecular markers 
+	instead of just gene expression (68 mutations, 426 copy numbers, 
+	and 2602 gene expression levels for 498 cell lines and 97 drugs). The 
+	current version of our software is not able to handle such problem sizes,
+	due to the repeated calculation of posterior parameters in the empirical
+	Bayes iterations. Currently, problems of up to 1000 outcomes, 1000 features,
+	and 1000 samples are computationally feasible.
 
-  \subsection{Empirical Bayes estimation}
-  To assess the empirical Bayes estimation of the inverse Gaussian prior, 
-  we conduct a small simulation study where we simulate from
-  model (\ref{eq:priormodel}), while keeping $\forall d: \sigma^2_d=1$ fixed. 
-  More specifically, we simulate 
-  $\gamma_{jd}^2 \sim \mathcal{IG}([(\mathbf{c}_d^j) \tr \bm{\alpha}]^{-1},
-  \lambda_d)$, for $j=1,\dots,200$, $d=1,\dots,10$, where we set 
-  $\bm{\alpha}= \begin{bmatrix} 1 & 1 & 2 & 3 & 4 \end{bmatrix}$. We use 
-  balanced dummy coding for the $\mathbf{c}_d^j$, such that we end up with
-  $\E(\gamma_{jd}) \in \{ 1, 0.5, 0.33, 0.25, 0.2 \}$, and fix 
-  $\forall d: \lambda_d=1$.  
-  Next, we simulate $\beta_{jd} \sim \mathcal{N} 
-  (0, \gamma_{jd}^2)$.
-  The observed data is generated from $x_{ij} \sim \mathcal{N} (0, 1)$ and 
-  $y_{id} \sim \mathcal{N} (\mathbf{x}_i \tr \bm{\beta}_d, 1)$, for $i=100$. 
-  We repeat the simulation 100 times and, for each simulation, estimate the 
-  hyperparameters by the method described in Section \ref{sec:empiricalbayes} 
-  and present the results in Figure \ref{fig:boxplots_igaussian_res6_eb}.
+	We average repeated measures over cell line-drug combinations and model the
+	logarithm of the IC50 values. In the following, IC50 refers 
+	to the log-transformed values. 
+	After iteratively removing the cell line or drug with the highest proportion 
+	of missings until no more missing values remain, we end up with IC50 
+	estimates for 482 cell lines on 168 drugs. As a preliminary step, we fit
+	separate elastic net models for each drug with the \texttt{glmnet} package
+	\cite[]{friedman_regularization_2010}, with $\alpha=0.01$ and $\lambda$
+	chosen such that we select approximately 300 genes per drug.
 
-<<boxplots_igaussian_res6_eb, fig.cap="Empirical Bayes $\\hat{\\bm{\\alpha}}$ estimates", out.width="80%", fig.asp=2/3>>=
+	\subsection{External data}
+	Two ternary drug covariates are available: the developmental stage
+	(experimental, in clinical development, or clinically approved) of the
+	drugs and the action (unkown, cytotoxic, or targeted) of the drugs. 
+	Furthermore, we have a binary feature covariate available that indicates
+	whether a gene belongs to the drug target pathway. The drug covariates are 
+	taken
+	directly from the GDSC database's annotation file. The feature covariate
+	was created by comparing the target pathways in the GDSC annotation to
+	the KEGG \cite[]{kanehisa_kegg:_2000} and reactome 
+	\cite[]{fabregat_reactome_2018} repositories. The external covariates are 
+	dummy-coded with
+	reference categories clinically approved drugs, cytotoxic drugs, and 
+	features that are not in the target pathway.
+
+	We expect that drugs that have been clinically approved are easiest to 
+	predict and hence yield the largest prior $\beta_{jd}$ variances, followed by
+	the drugs in clinical development, and the experimental drugs. Likewise we 
+	expect the targeted drugs to yield the largest prior $\beta_{jd}$ variances, 
+	followed by the cytotoxic drugs, and the unkown target drugs. For the
+	feature covariate, we expect that genes that are in the pathway of the drug
+	are more predictive than genes that are not, i.e., 
+	they have larger prior $\beta_{jd}$ variances than drugs that are not in 
+	the pathway.
+
+	\subsection{Results}
+	To compute performance measures, we randomly split the data into training and
+	test sets, containing 
+	241 cell lines each, estimate the models on the training data and compute 
+	performance measures on the test data. We repeated this procedure 50 times
+	and average performance measures over the splits.
+	The estimated 
+	models are: two models like in (\ref{eq:prior}) but with different external 
+	variables, named (i) NIG1 and (ii) NIG2 in the following, (iii) a model 
+	as described in \cite{kpogbezan_empirical_2017}, named bSEM, (iv) a regular
+	ridge model, and (v) a regular lasso model. The NIG1 incorporates no external
+	variables and estimates the common prior $\gamma_{jd}^2$ and $\tau_d^2$ 
+	means $\phi$ and $\chi$ directly, while the NIG2 model includes the
+	developmental phase and action drug covariates, and the pathway
+	feature covariate, and estimates $\bm{\alpha}$. The bSEM model is similar to 
+	the NIG model in that it 
+	draws the $\beta_{jd}$ from a conditionally normal distribution, but 
+	fixes $\tau_d^2=1$ and draws 
+	$\gamma_{jd}^2 \sim \Gamma^{-1} (\phi_{jd}, \lambda_{jd})$, instead of 
+	from the inverse Gaussian distribution. Additionally, 
+	the bSEM model can 
+	only incorporate one binary external covariate that divides the features into 
+	two groups, i.e., $\bm{\alpha} = \begin{bmatrix} \alpha_0 & \alpha_1 
+	\end{bmatrix} \tr$ and $\bm{c}_{jd} = \begin{bmatrix} 1 & 0 \end{bmatrix} 
+	\tr$ if feature $jd$ belongs to group one and $\bm{c}_{jd} = \begin{bmatrix} 
+	1 & 1 \end{bmatrix} \tr$ if feature $jd$ belongs to group two. Both 
+	$\bm{\alpha}_{\text{feat}}$ and $\lambda_{\text{feat}}$ are
+	estimated in a similar fashion as in Section \ref{sec:estimation}.
+
+  From the $\bm{\alpha}$ estimates in Tables \ref{tab:results_drug} and
+  \ref{tab:results_feat} we see that there
+  seems to be an effect of the
+  external variables. If we look at the NIG1 model, that includes all external 
+  covariates, we see that approved drugs are easier to predict, due to small
+  prior precision, followed by drugs that are in clinical development,
+  as we expected. The cytotoxic drugs are the hardest to predict, 
+  with the largest prior precision (smallest prior variance), followed by the 
+  targeted drugs and the unknown drugs. This is again according to expectation.
+  For the unkown drugs, we are unable to confirm the small precision. Genes 
+  that are in the target pathway have 
+  smaller prior precisions, according to expectation, which confirms that these
+  genes are more important for prediction of drug sensitivity. The estimates 
+  from
+  the bSEM model are also according to expectation: genes that are in the target
+  pathway have a much smaller precision than genes that are not in the target
+  pathway. The 
+  effects of the pathway indicator seem to be much larger in the bSEM model. 
+<<results_drug>>= 
+library(knitr, quietly=TRUE, verbose=FALSE)
+suppressWarnings(library(kableExtra))
+load("results/data_gdsc_res1.Rdata")
+tab1 <- cbind(tab[, 1], c(NA, 0, NA), tab[, c(3, 2)], c(NA, 0, NA), 
+              tab[, c(4, 5)])
+options(knitr.kable.NA="", digits=3)
+kable(tab1,
+      col.names=c("intercept", "approved", "development", 
+                  "experimental", "targeted", "unkown", "cytotoxic"),
+      caption="$\\bm{\\alpha}_{\\text{drug}}$ estimates from GDSC data.", 
+      format="latex", booktabs=TRUE, escape=FALSE) %>%
+  kable_styling(latex_options=c("HOLD_position"))
 @
 
-  [! describe results here]
-
-  - seems to be an overall scaling issue, relative to each other the estimates
-  are not so bad.
-
-  - investigate convergence: (i) look at both error and prior variance 
-  convergence and (ii) look at actual prior distribution convergence 
-  (plot true distribution and estimated prior every 5 iterations or so)
-
-  \section{Application in eQTLs mapping}\label{sec:application}
-  This application concerns eQTLs detection from the GEUVADIS data
-  \cite[]{lappalainen_transcriptome_2013}. After the removal of 9 genes due to
-  missing SNP information, we have 99 genes (outcomes) from the p38MAPK pathway. 
-  After removal of the subjects without any expression or SNP data, 373 subjects
-  remained. For each gene, we selected SNPs with minor alles frequency (MAF) 
-  larger than 0.05 and within 105 bases up- or downstream from the gene. This 
-  resulted in 99 sets of SNPs (one for each gene) from the 373 subjects, where 
-  the number of SNPs per gene ranges from 56 to 1169.
-
-  We used Ensembl \cite[]{hunt_ensembl_2018} to create several external 
-  variables: (i) SNP location 1 (binary, in- or outside of the gene),
-  (ii) SNP location 2 (binary, in an exon or an intron), (iii) 
-  genelength (continuous), (iv) minor allele frequency of the SNP
-  (continuous), and (v) squared MAF. We expect that SNPs that are in the gene 
-  and/or in an exon are more important in eQTLs detection. Since a larger gene
-  might depend on more SNPs, we expect that longer genes yield lower 
-  regression parameter estimates. We predict a positive correlation between
-  the regression parameters and the MAF.
-
-  We fit five different models to a subset of the subjects and 
-  estimated performance measures on the rest of the subjects. The five models we
-  fit differed only in the included external variables. The intercept only model
-  is considered a reference model. The other four models all included the 
-  genelengths and MAFs (variables (iii) and (iv)). 
-  Models 1 and 2 included variable 
-  (i), while models 1, 3, and 4 included variable (ii). In addition, model 4
-  included the squared MAFs (v).
-
-  For each model, we ran the algorithm for 100 EB iterations and obtained 
-  the $\hat{\bm{\alpha}}$ hyperparameter estimates in Table 
-  \ref{tab:hyperparameters_eQTL}. In addition, we calculated posterior means
-  for the $\bm{\beta}_d$ from MCMC samples from the posterior with the estimated
-  hyperparameters. We estimated mean squared error (MSE) for these posterior 
-  means. A common measure 
-  of Bayesian model fit is the log pseudo marginal likelihood
-  (LPML) \cite[]{gelfand_model_1992}, the Bayesian version of the leave-one-out 
-  cross-validated log likelihood (see Section \ref{sm-sec:lpml} for more 
-  details). We estimated LPML for the different models and added them,
-  together with the MSEs to Table \ref{tab:hyperparameters_eQTL}.
-<<hyperparameters_eQTL>>= 
-load("results/eQTL_fit1.Rdata")
-seq.mse <- c(2.0699, 2.0629, 2.0649, 2.0628, 2.0638)
-seq.lpml <- round(c(mean(lpml.ienig$lpml), mean(lpml1.enig$lpml),
-                    mean(lpml2.enig$lpml), mean(lpml3.enig$lpml), 
-                    mean(lpml4.enig$lpml)), 5)
-tab <- cbind(round(rbind(c(fit.ienig$eb$alpha, rep(NA, 5)),
-                         c(fit1.enig$eb$alpha, NA),
-                         c(fit2.enig$eb$alpha[c(1:2)], NA, 
-                           fit2.enig$eb$alpha[-c(1:2)], NA),
-                         c(fit3.enig$eb$alpha[1], NA, 
-                           fit3.enig$eb$alpha[-1], NA),
-                         c(fit4.enig$eb$alpha[1], NA, 
-                           fit4.enig$eb$alpha[-1])), 2),
-             seq.mse, seq.lpml)
-tab[which.min(tab[, 7]), 7] <- kableExtra::cell_spec(
-  tab[which.min(tab[, 7]), 7], format="latex", bold=TRUE)
-tab[which.max(tab[, 8]), 8] <- kableExtra::cell_spec(
-  tab[which.min(tab[, 8]), 8], format="latex", bold=TRUE)
-rownames(tab) <- c("intercept only", paste("model", 1:4))
-options(knitr.kable.NA = '')
-kableExtra::kable_styling(kableExtra::column_spec(
-  knitr::kable(tab, align=rep("r", ncol(tab)),
-               col.names=c("intercept", "in gene", "in exon", "length", 
-                           "MAF", "MAF$^2$", "MSE", "LPML"),
-               caption="Estimated $\\hat{\\bm{\\alpha}}$ hyperparameters, 
-               MSE (lowest in bold), and LPML (highest in bold).", 
-               "latex", booktabs=TRUE, escape=FALSE, 
-               digits=c(rep(2, 6), 4, 4)), 8, border_left=TRUE),
-  latex_options=c("HOLD_position"))
+<<results_feat>>= 
+library(knitr, quietly=TRUE, verbose=FALSE)
+suppressWarnings(library(kableExtra))
+load("results/data_gdsc_res1.Rdata")
+tab2 <- cbind(tab[, 6], c(NA, 0, 0), tab[, 7])
+options(knitr.kable.NA="", digits=3)
+kable(tab2,
+      col.names=c("intercept", "not in pathway", "in pathway"),
+      caption="$\\bm{\\alpha}_{\\text{feat}}$ estimates from GDSC data.", 
+      format="latex", booktabs=TRUE, escape=FALSE) %>%
+  kable_styling(latex_options=c("HOLD_position"))
 @
 
-  - show influence of co-data with standardized $\bm{\alpha}$'s.
-
-  A quantity that is often of interest in eQTLs detection studies, is the 
-  gene expression heritability. It is the proportion of a gene's expression that
-  is explained with the SNPs and a measure of the relationship between trait 
-  phenotypic variation and genetic variation. Heritability (per gene) is 
-  given by $h_d^2 = (\sigma^2_d \sum_{j=1}^{p_d} \gamma_{jd}^2)/
-  (\sigma^2_d \sum_{j=1}^{p_d} \gamma_{jd}^2 + \sigma_d^2)$. We assess 
-  mean heritability of the five models for a rank-based selection of SNPs.
-  We present MCMC-based posterior means and one standard deviation error bars in
-  Figure \ref{fig:lines_eqtl_hsq}.
-<<lines_eqtl_hsq, fig.cap="Mean heritability in the eQTL detection data.", out.width="80%", fig.asp=2/3>>=
+  The median prediction MSEs and the median ELBOs, calculated on the test data
+  are displayed in Table \ref{tab:results_data}. In terms of 
+  is lowest in the NIG models 
+  (\Sexpr{round(mapmse[2], 2)} and \Sexpr{round(mapmse[3], 2)}, in the
+  NIG1 and NIG2 models, respectively). The bSEM model is not able to 
+  improve on the null model with an median prediction MSE of 
+  \Sexpr{round(mapmse[4], 2)} compared to the null model's MSE of 
+  \Sexpr{round(mapmse[1], 2)}.
+<<results_data>>= 
+load("results/data_gdsc_res1.Rdata")
+mapmse <- apply(apmse, 2, median)
+melbo <- apply(elbo, 2, median)
+tab <- rbind(mapmse, c(melbo, NA, NA))
+rownames(tab) <- c("MSE", "ELBO")
+options(knitr.kable.NA="", digits=3)
+kable(tab,
+      col.names=c("NIG1", "NIG2", "bSEM", "lasso", "ridge"),
+      caption="Median prediction MSE and median ELBO estimated on the test data", 
+      format="latex", booktabs=TRUE, escape=FALSE) %>%
+  kable_styling(latex_options=c("HOLD_position"))
 @
 
-  [! describe results here]
-
-  - maybe use explained heritability (fraction of full model heritability)
-
-  - AUC of curve as overal measure
-
-  % Annotation is from GRCh37 reference genome: 
-  % \url{https://www.ncbi.nlm.nih.gov/assembly/GCF_000001405.13/}. 
-  % Annotation file that Gino used: 
-  % \url{https://www.gencodegenes.org/human/release_12.html}.
-  % Ensembl version of this genome:
-  % \url{https://grch37.ensembl.org/Homo_sapiens/Info/Index}.
-  % Possibly use GeuvadisTranscriptExpr R package.
+  In addition to the empirical Bayes estimates and the prediction MSE, we 
+  investigated the stability of the estimates by ranking the posterior means
+  $\E_{Q}(\bbeta_d)$ and calculating the Spearman's rank
+  correlation over the 50 data splits. They are depicted in Figure 
+  \ref{fig:data_gdsc_res1_brank_hist}. In both NIG models the correlations
+  between splits are higher than in the bSEM model; an indication that 
+  estimation in the NIG model is more stable than in the bSEM model.
+<<data_gdsc_res1_brank_hist, fig.cap="Spearman's rank correlation between the estimated posterior means $\\E_{Q}(\\bbeta_d)$ in the 50 data splits.", out.width="70%", fig.asp=2/3>>= 
+@
 
 	\section{Discussion}\label{sec:discussion}
-  In this paper we have introduced a normal inverse Gaussian prior for 
-  multivariate high dimensional linear regression models that is estimated by
-  empirical-variational Bayes. The estimation is adaptive to external data
-  that describes the relation between the outcomes and features in a data-driven
-  manner. In simulations we have shown that the inverse Gaussian model 
-  does not perform worse, and often bettern, than a common alternative, 
-  the Student's $t$ prior. in addition, the Student's $t$ prior does not admit 
-  relatively straightforward estimation, compared to the inverse Gaussian 
-  prior, as explained in Section \ref{sec:estimation}. The 
-  simulations also show that the model is reasonably well approximated and 
-  estimated using empirical-variational Bayes. An eQTL detection application 
-  shows that external data may indeed benefit prior estimation and prediction
-  performance.
+	The preceding presents a novel model for drug sensitivity
+	prediction from a set of high dimensional molecular features. The model allows
+	for the inclusion of discrete and continuous external covariates on both the 
+	drugs and features. Inclusion of the external information is through 
+	data-driven and adaptive empirical Bayes estimation of the hyperparameters in 
+	the normal inverse Gaussian prior model (\ref{eq:prior}). Estimation is 
+	efficient and scales well with the number of features and samples.
 
-  Apart from eQTL detection, the model may be used for network reconstruction as 
-  in \cite{leday_gene_2017}, drug response prediction (manuscript in 
-  preparation by the current authors), and other applications with multiple
-  outcomes, large feature space, and available external information.
+	The model is put into practice on the GDSC data. The results in Section
+	\ref{sec:gdsc} show that the inclusion of the external covariates is 
+	beneficial for drug sensitivity prediction. Although the prediction MSE is
+	not lower if the external covariates are included, 
+	the resulting hyperparameter estimates are still informative
+	and may guide future research efforts.
 
-  A drawback of the method is the fact that the external data is required to
-  be positive. That may not always be the case in practice. We note, however,
-  that the $\mathbf{c}_d^j$ may be translated to be positive by adding a 
-  constant. In the usual predictive feature sense this doesn't work, since one
-  might still encounter negative future values. Here, the external data is
-  only used for estimation and not for prediction, so this issue is 
-  non-existent.
+	The comparison to the bSEM model that allows for the inclusion of 
+	one binary external feature covariate favours the proposed NIG model. 
+	Prediction MSEs are almost uniformly lower over the drugs for the NIG model 
+	compared to bSEM. Consequently, the average prediction MSE is significantly
+	lower for the NIG model.
 
-  \section*{Software}
-	A (developmental) \texttt{R} package is available from 
-	\url{https://github.com/magnusmunch/cambridge}.
+	Although the empirical Bayes estimates of the external covariate effects 
+	$\bm{\alpha}$ are informative, they are small compared to 
+	bSEM's estimates. Consequently, the posteriors of the models with and without
+	the external covariates (NIG2 and NIG1 in Section \ref{sec:gdsc}) do not 
+	differ by much. This results similar prediction MSEs with
+	the inclusion of the external covariates. This is a topic that deserves 
+	further attention. One proposed solution is to include an extra 
+	prior variance component that promotes a more heavy-tailed $\beta_{jd}$ prior
+	distribution. An example is a variance component with a half-cauchy prior 
+	distribution, i.e., the horseshoe \cite[]{carvalho_handling_2009}. 
+	This solution is similar to the pInc model from 
+	\cite{kpogbezan_incorporating_2019}. 
+	Like the bSEM model, the pInc model gives significantly larger external
+	covariate effects, 
+	so we conjecture that the inclusion of a similar heavy-tailed prior variance 
+	component will have a similar effect in the NIG model.
 
@@
 	\section*{Acknowledgements}
-	We thank Gino Kpogbezan for providing the pre-processed data in Section 
-	\ref{sec:application}.
-	\textit{Conflict of Interest}: None declared.
+	We thank Zhi Zhao and Leiv Tore Salte R{\o}nneberg from the Oslo Centre for 
+	Biostatistics and Epidemiology for their suggestions on the pre-processing of 
+	the GDSC data. \textit{Conflict of Interest}: None declared.
 
-	\bibliographystyle{author_short3.bst}
+	\bibliographystyle{author_short3} 
 	\bibliography{refs}

### 2019-11-19 docs/manuscript2.Rnw

--- 2019-08-13 docs/manuscript2.Rnw
@@
 \usepackage{url,bm,algorithm,algpseudocode,bbm,amsmath,amssymb,amsthm,
-amsfonts,graphics,graphicx,epsfig,rotating,caption,subcaption,natbib,
+amsfonts,graphics,graphicx,epsfig,rotating,caption,subcaption,natbib,appendix,
 titlesec,multicol,verbatim,threeparttable,booktabs,dsfont,parskip,
@@
 	\section{Introduction}
-	- motivation: drug response prediction from different omics data types is 
-	promising both for precision medicine and drug target discovery and relates to
-	actual tumor response \cite[]{iorio_landscape_2016}.
+	- motivation: drug response prediction from different omics data types is promising both for precision medicine and drug target discovery and relates to actual tumor response \cite[]{iorio_landscape_2016}.
 
-	- problem: prediction of drug efficacy in cell lines is difficult (DREAM 7 
-	challenge in \cite{costello_community_2014}).
+	- problem: prediction of drug efficacy in cell lines is difficult (DREAM 7 challenge in \cite{costello_community_2014}).
 
@@
 
-	- feature information: prior experts, pathways, database summaries, previous 
-	experiments, type of feature (somatic mutations, copy number alterations, DNA 
-	methylation, and gene expression).
+	- feature information: prior experts, pathways, database summaries, previous experiments, type of feature (somatic mutations, copy number alterations, DNA methylation, and gene expression).
 
-	- drugs information: molecular target, drug class, pubchem, 
-	omnipath DB, pharmaco DB
+	- drugs information: molecular target, drug class, etc (ask Oslo), pubchem, omnipath DB, pharmaco DB
 
-	- drugs x feature information
-
-	- current literature: reviews in \cite{azuaje_computational_2017} and
-	\cite{ali_machine_2019}, 
-	full multivariate approach by Oslo people in \cite{mai_composite_2019}, 
-	TANDEM \cite[]{aben_tandem:_2016}, DeltaNet 
-	by \cite{noh_inferring_2016}, patient-derived tumor cell lines 
-	in \cite{gao_high-throughput_2015}, 
-	multiple kernel learning approach in \cite{costello_community_2014},
-	\cite{ammad-ud-din_drug_2016} combine a 
-	multiple kernel learning approach with matrix factorization, using the model 
-	to uncover the latent 
-	relationships between drug targets and intracellular pathways, and
-	structured penalized regression in \cite{zhao_structured_2019}
-
-	- our solution: adaptive borrowing of information through empirical Bayes, 
-	guided by external feature and drug information.
-
-	\section{Methods}
-	\subsection{Simultaneous equations model}\label{sec:SEM}
-	Assume we have continuous efficacy measures $y_{id}$ for cell lines 
-	$i=1,\dots, n$, centred per drug $d$. We predict efficacy from molecular 
-	features 
-	$x_{ij}$, $j=1,\dots, p$, collected in $\x_i = \begin{bmatrix} x_{i1} & 
-	\dots & x_{ip} \end{bmatrix} \tr$. We have an \textit{a priori} (possibly 
-	informative) partitioning 
-	of the omics features into $g=1, \dots, G$ non-overlapping groups.
-	For convenience, we let 
-	$g(j)$ denote the group index of molecular feature $j$. Furthermore, we have 
-	characteristics of the drugs available in the form of external variables 
-	$\mathbf{C} = \begin{bmatrix} \mathbf{c}_1 & \dots & \mathbf{c}_D 
-	\end{bmatrix} \tr$.
-
-	We regress the drug efficacies on the molecular feature effects $\beta_{jd}$:
-	\begin{subequations}
-		\begin{align}
-		  y_{id} & = \x_i \tr \bbeta_d + \epsilon_{id}, \label{eq:linearmodel}\\
-		  \text{with } & \epsilon_{id} \sim \mathcal{N}(0, \sigma_d^2),
-		\end{align}
-	\end{subequations}
-	where $p$-dimensional $\bbeta_d$ are the drug-specific omics feature effects. 
-	Note that (\ref{eq:linearmodel}) gives rise to a system of $d=1, \dots, D$ 
-	linear simultaneous equations.
-
-	\subsection{Bayesian prior model}
-	We capture the uncertainty in the parameters through a Bayesian prior model
-	and endow the parameters with the following prior:
-	\begin{subequations}\label{eq:prior}
-		\begin{align}
-		  \beta_{jd} & \sim \mathcal{N}_p (0, \phi^2_{g(j)} \gamma_{jd}^2 
-		  \sigma_d^2), 
-		  \label{eq:betaprior}\\
-		  \gamma_{jd}^{2} & \sim \mathcal{IG}( 
-		  [(\mathbf{c}_d^j) \tr \bm{\alpha}]^{-1}, \lambda_d), \\
-		  \sigma_d^{2} & \sim 1/\sigma_d^{3},
-		\end{align}
-	\end{subequations}
-	where $\mathcal{IG}(\theta, \lambda)$ denotes an inverse Gaussian distribution 
-	with mean $\theta$ and shape $\lambda$. The different molecular feature types
-	are given their own weights through the $\phi^2_{g}$, while the external 
-	outcome-feature data comes in through 
-	our mean model for the $\gamma_{jd}^2$ hyperprior model: 
-	$[(\mathbf{c}_d^j) \tr \bm{\alpha}]^{-1}$. Both the $\phi^2_{g}$ and
-	$\bm{\alpha}$ are estimated in a data-driven manner. More details on the model
-	itself, as well as its estimation, may be found in 
-	the SM and in [! ref to first paper here].
-
-	\section{Data}
-
-	\subsection{GDSC}
-	The GDSC project's \cite[]{yang_genomics_2013} aim is "to improve cancer 
-	treatments by discovering therapeutic biomarkers that can be used to identify 
-	patients most likely to respond to anticancer drug". Part of the project is to
-	screen $>1000$ human cancer cell lines with drugs. The cell lines have been
-	genetically characterised and several drug sensitivity measures are recorded.
-	The data is freely available from \cite{garnett_systematic_2012} and consist 
-	of: (i) the sensitivity measures 
-	(mainly half maximal inhibitory concentration, i.e., IC50) of the cell lines 
-	to the drugs, (ii) annotation of the 
-	screened compounds, and (iii) the cell lines' genomic profile (mutations, 
-	copy numbers, methylation profiles, and expression). We will attempt to 
-	predict drug sensitivity of the cell lines, as quantified by IC50, using the
-	genomic data. Other choices of efficacy measures than IC50 are possible, 
-	but a discussion on the pros and cons of these different choices is not the 
-	aim of this paper.
-
-	\subsubsection{Primary data}
-	We took the logarithm of the IC50 values and took the average if there were
-	multiple measurements per cell line-drug combination. In the following, if we 
-	refer to IC50, it denotes the log-transformed values. After iteratively 
-	removing the cell line or drug with the highest proportion 
-	of missings until there are no more missing values, we end up with IC50 
-	estimates for 482 cell lines on 168 drugs. For the gene expression 
-	data, we fit a lasso with IC50 as response to the drugs separately and 
-	adjusted the penalty parameter to select approximately 300 genes per drug.
-
-	- From Oslo:
-	1.2) Preselecting GEX features: we selected gene expression features
-	explaining 50\% variations over all cell lines, since this results in not too
-	features and probably remains those important features.
-	1.3) Preselecting CNV: we used the same strategy as Garnett et al. (2012)
-	where the GDSC data are from.
-	1.4) Preselecting MUT: we deleted mutated genes with less than two cell lines.
-
-	- check out selection method in \cite{zhao_structured_2019}.
-
-	- how to transform the data and deal with missingness? Read GDSC papers.
-
-	\subsubsection{External data}
-	Several external drug variables are available: (i) the developmental stage
-	(experimental, in clinical development, or clinically approved) and (ii)
-	the action (cytotoxic or targeted). We also have a external variable 
-	available that relates the drugs to the features: a binary indicator whether 
-	a gene is in the target pathway of the drug or not.
+	- current literature: review in \cite{azuaje_computational_2017}, reference Oslo, TANDEM \cite[]{aben_tandem:_2016}, DeltaNet \cite[]{noh_inferring_2016}, patient-derived tumor cell lines \cite[]{gao_high-throughput_2015}.
 
@@
 
-	\section{Results}
+	- our solution: adaptive borrowing of information through empirical Bayes, guided by external feature and drug information.
+
+	\section{Model}
+	\subsection{Simultaneous equations model}\label{sec:SEM}
+	- Gwen can deal with different number of features per equation, might solve missing drug-feature combination problem? Doesn't solve missing response-feature combination problem.
+
+	- check literature for: a) conjugate version, b) similar models: "prior variance decomposition", "structured Bayesian regression"
+
+	We have continuous efficacy measures $y_{id}$ for cell lines $i=1,\dots, n$, from tissues $t=1, \dots, T$, on drugs $d=1,\dots,D$.
+Throughout this paper we assume the $y_{id}$ to be centred per drug and let $\y_d = \begin{bmatrix} y_{1d} & \dots & y_{nd} \end{bmatrix} \tr$.
+We predict efficacy with molecular features $x_{ij}$, $j=1,\dots, p$, collected in $\x_i = \begin{bmatrix} x_{i1} & \dots & x_{ip} \end{bmatrix} \tr$.
+We have an \textit{a priori} partitioning of the omics features into $g=1, \dots, G$ non-overlapping groups, which might be informative for the problem at hand For convenience, we let $g(j)$ denote the group index of molecular feature $j$.
+Furthermore, we have characteristics of the drugs available in the form of `co-data' $\mathbf{C} = \begin{bmatrix} \mathbf{c}_1 & \dots & \mathbf{c}_D \end{bmatrix} \tr$.
+We code the tissue from which the cell lines were taken as a $T$ binary dummy variables
+	$$
+	z_{it} = \begin{cases}
+	1 & \text{if } \text{tissue}_i = t, \\
+	0 & \text{otherwise},
+	\end{cases}
+	$$
+	collected in $\mathbf{z}_i = \begin{bmatrix} z_{i1} & \dots & z_{iT} \end{bmatrix} \tr$.
+
+	We assume that cell lines from the same tissue are related and model this relation through random effects $u_{td}$.
+We let the molecular feature effects $\beta_{jd}$ vary over the drugs and model drug efficacy as a linear combination of these two effects:
+	\begin{subequations}\label{eq:linearmodel}
+		\begin{align}
+		y_{id} & = \beta_{0d} + \sum_{j=1}^p x_{ij} \beta_{jd} + \sum_{t=1}^T z_{it} u_{td} + \epsilon_{id} \\
+		& = \beta_{0d} + \x_i \tr \bbeta_d + \mathbf{z}_i \tr \mathbf{u}_d + \epsilon_{id},\\
+		\text{with } & \epsilon_{id} \sim \mathcal{N}(0, \sigma_d^2),
+		\end{align}
+	\end{subequations}
+	where $p$-dimensional $\bbeta_d$ and $T$-dimensional $\bm{u}_d$ are the drug-specific omics feature and tissue effect vectors, respectively.
+Note that (\ref{eq:linearmodel}) gives rise to a system of $d=1, \dots, D$ linear simultaneous equations.
+
+	\subsection{Bayesian prior model}
+	We capture the uncertainty in the parameters through a Bayesian prior model.
+We assign a flat prior to the intercept terms $\beta_{0d}$ and integrate them out.
+The remaining parameters are endowed with the following priors:
+	\begin{subequations}\label{eq:prior}
+		\begin{align}
+		\beta_{jd} & \sim \mathcal{N}_p (0, \phi^2_{g(j)} \gamma_d^2 \sigma_d^2), \label{eq:betaprior}\\
+		\mathbf{u}_{d} & \sim \mathcal{N}_T (\mathbf{0}, \bm{\Xi}_d \sigma_d^2),
+		\end{align}
+	\end{subequations}
+	and hyper-priors:
+	\begin{subequations}\label{eq:hyperprior}
+		\begin{align}
+			\sigma_d^{2} & \sim 1/\sigma_d^{3} \\
+			\gamma_d^{2} & \sim \mathcal{IG}(\theta_d, \lambda), \\
+			\bm{\Xi}_d & \sim \mathcal{W}^{-1}_T(\bm{\Omega}, \nu),
+		\end{align}
+	\end{subequations}
+	where $\mathcal{IG}(\theta, \lambda)$ is an inverse Gaussian distribution with mean $\theta$ and shape $\lambda$, and $\mathcal{W}^{-1}_T(\bm{\Omega}, \nu)$ is an inverse Wishart distribution with $T \times T$-dimensional positive definite (PD) scale matrix $\bm{\Omega}$ and degrees of freedom $\nu$.
+Note that for $\nu > T + 1$, this inverse Wishart prior implies an expected prior covariance matrix $(\nu - T - 1)^{-1} \bm{\Omega}$ for the tissue effects, common to all drugs.
+Likewise, the molecular feature effects are related through the common hyper-prior for the $\gamma_d^2$.
+
+	- Jeffrey's prior for variance $1/\sigma^{3/2}$ (a priori independent data mean and variance) or $1/\sigma^3$ (joint prior, but only mean and variance, not variance and betas).
+
+	A few remarks on the choice of priors are justified here: many authors endow error variance components with vague gamma priors.
+\cite{gelman_prior_2006}, among others, advises against this practice.
+The degree of `vagueness' has a large influence on the posterior, while degree of `vagueness' is a difficult parameter to set.
+This influence is especially pronounced if the likelihood is relatively flat, as may be reasonably expected in the large $p$, small $n$ setting we are in.
+We therefore model the error variance with Jeffrey's prior \cite[]{jeffreys_invariant_1946}.
+
+	Furthermore, we choose to model the $\gamma^2_d$ by an inverse Gaussian distribution, as has been suggested in \cite{fabrizi_specification_2016} and \cite{caron_sparse_2008}, because it allows to model the mean $\theta_d$ as a function of the drug covariates $\mathbf{c}_d$, as explained in Section \ref{sec:empiricalbayes}.
+
+	- IG hyperprior used in finance \cite[]{barndorff-nielsen_normal_1997}, where it is called the NIG prior.
+
+	\section{Estimation}
+	\subsection{Variational Bayes}\label{sec:variationalbayes}
+	The posterior corresponding to the model described in (\ref{eq:linearmodel}), (\ref{eq:prior}), and (\ref{eq:hyperprior}) is not available in closed form.
+We therefore approximate the posterior by variational Bayes, where we force the posterior density $Q$ to factorise as:
+	\begin{align}\label{eq:variationalposterior}
+	p(\B,\mathbf{U},\mathbf{\sigma}^2, \bgamma^2,\bm{\Xi}_1, \dots, \bm{\Xi}_D) \approx Q(\cdot) = q(\B) \cdot q(\mathbf{U}) \cdot q(\mathbf{\sigma}^2) \cdot q(\bgamma^2) \cdot q(\bm{\Xi}_1, \dots, \bm{\Xi}_D),
+	\end{align}
+	where $\B = \begin{bmatrix} \bbeta_1 & \cdots & \bbeta_D \end{bmatrix}$ and $\U = \begin{bmatrix} \mathbf{u}_1 & \cdots & \mathbf{u}_D \end{bmatrix}$, are of dimensions $p \times D$ and $T \times D$, respectively.
+Under the factorisation in (\ref{eq:variationalposterior}), the marginal variational posteriors that minimise the Kullback-Leibler divergence of the true posterior to the variational Bayes approximation \cite[]{neal_view_1998}, are given by:
+	\begin{subequations}\label{eq:variationalposterior2}
+		\begin{align}
+		q_{\B} (\B) & \overset{D}{=} \prod_{d=1}^{D} \mathcal{N}_p (\bmu_d, \bSigma_d), \\
+		q_{\U}(\U) & \overset{D}{=} \prod_{d=1}^{D} \mathcal{N}_T (\mathbf{m}_d, \Sm_d), \\
+		q_{\bgamma^{2}}(\bgamma^{2}) & \overset{D}{=} \prod_{d=1}^D \mathcal{GIG} \(-\frac{p+1}{2}, \frac{\lambda}{\theta_d^2}, \delta_{d} \), \\
+		q_{\bsigma^{2}}(\bsigma^{2}) & \overset{D}{=} \prod_{d=1}^D \Gamma^{-1} \(\frac{n + p + T + 1}{2}, \zeta_{d} \), \\
+		q_{\bm{\Xi}_1, \dots, \bm{\Xi}_D}(\bm{\Xi}_1, \dots, \bm{\Xi}_D) & \overset{D}{\propto} \prod_{d=1}^D \mathcal{W}_T^{-1} (\bm{\Psi}_d, \nu + 1).
+		\end{align}
+	\end{subequations}
+	Here $\mathcal{GIG}(\cdot)$ and $\Gamma^{-1}(\cdot)$ denote the generalized inverse Gaussian and inverse gamma distributions, respectively.
+The full derivations may be found in Appendix \ref{app:derivations}.
+
+	The parameters in (\ref{eq:variationalposterior2}) contain cyclic dependencies. We therefore iterate the following estimating equations until convergence to a local optimum:
+	\begin{align*}
+	\bSigma_d^{(h+1)} & = \frac{2\zeta_d^{(h)}}{n+p+T+1} \left\{\X \tr \X + a_d^{(h)} \cdot \diag (\phi_{g(j)}^{-2}) \right\}^{-1}, \\
+	\bm{\mu}_d^{(h+1)} & = \left\{\X \tr \X + a_d^{(h)} \cdot \diag (\phi_{g(j)}^{-2}) \right\}^{-1} \X \tr ( \y_d - \Z \mathbf{m}_d^{(h)}), \\
+	\Sm_d^{(h+1)} & = \frac{2\zeta_d^{(h)}}{n+p+T+1} \[ \Z \tr \Z + (\nu + 1) \cdot (\bm{\Psi}_d^{(h)})^{-1} \]^{-1}, \\
+	\mathbf{m}_d^{(h+1)} & = \[ \Z \tr \Z + (\nu + 1) \cdot (\bm{\Psi}_d^{(h)})^{-1} \]^{-1} \Z \tr (\y_d - \X \bm{\mu}_d^{(h)}), \\
+	\bm{\Psi}_d^{(h+1)} & = \frac{2\zeta_d^{(h)}}{n+p+T+1} \[ \Sm_d^{(h+1)} + \mathbf{m}_d^{(h+1)} (\mathbf{m}_d^{(h+1)}) \tr \] + \bm{\Omega}, \\
+	\delta_d^{(h+1)} & =  \frac{n+p+T+1}{2\zeta_d^{(h)}} \sum_{j=1}^p \frac{(\mu_{jd}^{(h+1)})^2 + (\bSigma_d^{(h+1)})_{jj}}{\phi^2_{g(j)}} + \lambda, \\
+	\zeta_d^{(h+1)} & = \frac{1}{2} \y_d \tr \y_d - \y_d \tr (\X \bmu_d^{(h+1)} + \Z \mathbf{m}_d^{(h+1)}) + (\bmu_d^{(h+1)}) \tr \X \tr \Z \mathbf{m}_d^{(h+1)} + \frac{1}{2} \trace (\X \tr \X \bSigma_d^{(h+1)}) \\
+	& \,\,\,\,\,\,\,\,\,\, + \frac{1}{2} (\bmu_d^{(h+1)}) \tr \X \tr \X \bmu_d^{(h+1)} + \frac{1}{2} \trace (\Z \tr \Z \Sm_d^{(h+1)}) + \frac{1}{2} (\mathbf{m}_d^{(h+1)}) \tr \Z \tr \Z \mathbf{m}_d^{(h+1)} \\
+	& \,\,\,\,\,\,\,\,\,\, + \frac{1}{2} a_d^{(h)} \sum_{j=1}^p \frac{(\mu_{jd}^{(h+1)})^2 + (\bSigma_d^{(h+1)})_{jj}}{\phi^2_{g(j)}} + \frac{1}{2} (\nu + 1) \cdot \trace \[ (\bm{\Psi}_d^{(h+1)})^{-1} \Sm_d^{(h+1)} \] \\
+	& \,\,\,\,\,\,\,\,\,\, + \frac{1}{2} (\nu + 1) \cdot (\mathbf{m}_d^{(h+1)}) \tr (\bm{\Psi}_d^{(h+1)})^{-1} \mathbf{m}_d^{(h+1)},
+	\end{align*}
+	where
+	$$
+	a_d^{(h)} = \sqrt{\frac{\lambda}{\theta^2_d \delta_d^{(h)}}} \frac{K_{\frac{p - 1}{2}} \( \sqrt{\lambda \delta_d^{(h)}}/\theta_d \)}{K_{\frac{p+1}{2}} \( \sqrt{\lambda \delta_d^{(h)}}/\theta_d \)} + \frac{p+1}{\delta_d^{(h)}},
+	$$
+	and $K_{\alpha}(\cdot)$ is the modified Bessel function of the second kind.
+
+	\subsection{Empirical Bayes}\label{sec:empiricalbayes}
+	A full Bayesian model requires the specification of hyper-parameters $\bm{\eta} = \begin{bmatrix} (\bm{\phi}^2) \tr & \bm{\theta} \tr & \lambda & \text{vec}(\bm{\Omega}) \tr & \nu \end{bmatrix} \tr$.
+These are abstract and hard to interpret parameters for which we generally lack expert knowledge.
+They do, however, have a significant influence on the shape of the posterior distribution.
+We therefore propose to estimate the unspecified hyper parameters $\bm{\eta}$ by empirical Bayes.
+Simply put, empirical Bayes fits the prior model to the data and is, as such, objective (up to model specification).
+
+	The canonical method for empirical Bayes is maximisation of the marginal likelihood with respect to the hyper parameters.
+In \cite{casella_empirical_2001} the marginal likelihood is maximised by an EM algorithm:
+	$$
+	\bm{\eta}^{(l+1)} = \underset{\bm{\eta}}{\argmax}\E_{\cdot | \Y} [\log p(\Y, \B, \U, \bm{\Xi}_1, \dots, \bm{\Xi}_D, \bgamma^2, \bsigma^2) | \bm{\eta}^{(l)}],
+	$$
+	where the expectation is with respect to the posterior.
+In our case, this posterior is not available in closed form, which renders the expectation difficult.
+While \cite{casella_empirical_2001} suggests to approximate the expectation by a Monte Carlo sample, we propose to use the variational Bayes approximation developed in Section \ref{sec:variationalbayes}:
+	\begin{align}\label{eq:variationalbayes}
+	\bm{\eta}^{(l+1)} & = \underset{\bm{\eta}}{\argmax}\E_{Q^{(l)}} [\log p(\Y, \B, \U, \bm{\Xi}_1, \dots, \bm{\Xi}_D, \bgamma^2, \bsigma^2)] \nonumber \\
+	& = \underset{\bm{\eta}}{\argmax} \E_{Q^{(l)}} [\log \pi (\B | \bgamma^2, \bsigma^2)] + \E_{Q^{(l)}} [\log \pi (\bgamma^2)] + \E_{Q^{(l)}} [\log \pi (\bm{\Xi}_1, \dots, \bm{\Xi}_D)].
+	\end{align}
+	where now the expectation is with respect to the converged variational posterior $Q^{(l)}$.
+Inspection of (\ref{eq:variationalbayes}) learns us that the problem may be decomposed into three separate optimisation problems.
+
+	\subsubsection{Group-specific variance components}
+	The first sub-problem concerns the group-specific variance components of the molecular feature effects.
+In order to separate the overall variance of the molecular feature effects from the group-specific deviations, we impose $\prod_{g=1}^G (\phi_g^2)^{|\mathcal{G}_g|} = 1$.
+As a consequence, the overall level of shrinkage is determined by the $\gamma^2_d$, while the $\phi_g^2$ effectuate differential shrinkage of the groups.
+Because the $\phi_g^2$ are variance components, we additionally require $\phi^2_1, \dots, \phi^2_G> 0$, such that our first optimisation sub-problem becomes:
+	\begin{align*}
+	(\bm{\phi}^2)^{(l+1)} & = \underset{\bm{\phi}^2}{\argmax} \E_{Q^{(l)}} [\log \pi (\B | \bgamma^2, \bsigma^2)] \\
+	& = \underset{\bm{\phi}^2}{\argmax} -\frac{D}{2} \sum_{g=1}^G |\G_g| \log \phi_g^2 - \frac{1}{2} \sum_{g=1}^G b_g^{(l)} \phi_g^{-2} \\
+	& \text{subject to } \prod_{g=1}^G (\phi_g^2)^{|\G_g|} = 1 \text{ and } \phi^2_1, \dots, \phi^2_G> 0.
+	\end{align*}
+	where 
+	$$
+	b_g^{(l)}=\sum_{d=1}^D a_d^{(l)} \frac{n+p+T+1}{2\zeta_d^{(l)}} \sum_{j \in \mathcal{G}_g} \[ (\mu_{jd}^{(l)})^2 + (\bSigma_d^{(l)})_{jj} \].
+	$$
+	This is a convex optimisation problem and easily solved numerically.
+
+	\subsubsection{Drug-specific variance components}
+	The second sub-problem concerns the prior for the drug-specific variance components $\gamma_d^2$.
+As explained in Section \ref{sec:SEM}, we may have `covariates' $\mathbf{c}_d$ on the drugs available.
+We conjecture that these covariates may be informative for the omics effect size and therefore model the prior mean of the $\gamma_d^2$ as a function of these covariates: $\theta_d = (\mathbf{c}_d \tr \bm{\alpha})^{-1}$.
+Then, the empirical Bayes estimation of $\bm{\alpha}$ effectively boils down to an inverse Gaussian regression of the $\gamma_d^{-2}$ on the $\mathbf{c}_d$ \cite[]{fries_optimal_1986,whitmore_regression_1983}.
+The problem may be formulated as:
+	\begin{align*}\label{eq:optproblem2}
+	\lambda^{(l+1)}, \bm{\alpha}^{(l+1)} & = \underset{\lambda,\bm{\alpha}}{\argmax} \E_{Q^{(l)}} [\log \pi (\bgamma^2)] \\
+	& = \underset{\lambda,\bm{\alpha}}{\argmax} \frac{D}{2} \log \lambda - \frac{\lambda}{2} \bm{\alpha} \tr \mathbf{C} \tr \diag [\E_{Q^{(l)}}(\gamma_d^2)] \mathbf{C} \bm{\alpha} + \lambda \bm{\alpha} \tr \mathbf{C} \tr \mathbf{1} - \frac{\lambda}{2} \sum_{d=1}^D \E_{Q^{(l)}}(\gamma_d^{-2}),
+	\end{align*}
+	where $\mathbf{C} = \begin{bmatrix} \mathbf{c}_1 & \dots & \mathbf{c}_D \end{bmatrix} \tr$. The solutions to this problem are:
+	\begin{align*}
+	\bm{\alpha}^{(l+1)} & = \left\{ \mathbf{C} \tr \Vm^{(l)} \mathbf{C} \right\}^{-1} \mathbf{C} \tr \mathbf{1}, \\
+	\lambda^{(l+1)} & = \[ \sum_{d=1}^D a_d^{(l)} - (\bm{\alpha}^{(l+1)}) \tr \mathbf{C} \tr \mathbf{1} \]^{-1} D,
+	\end{align*}
+	where 
+	$$
+	\Vm^{(l)} = \diag \[ \sqrt{\frac{\delta_d^{(l)}(\theta_d^{(l)})^2}{\lambda^{(l)}}} \frac{K_{\frac{p - 1}{2}} \( \sqrt{\lambda^{(l)} \delta_d^{(l)}}/\theta^{(l)}_d \)}{K_{\frac{p + 1}{2}} \( \sqrt{\lambda^{(l)} \delta_d^{(l)}}/\theta^{(l)}_d \)} \].
+	$$
+
+	- Think about parametrisation of $\mathbf{C}$.
+Idea: use overall mean and deviations from it in alpha, such that the starting value can be the mean and 0 for alpha.
+Or: use intercept and code the rest as dummy variables.
+
+	\subsubsection{Tissue effect covariance}
+	The third sub-problem pertains to the tissue effect covariance prior.
+To avoid overfitting when freely estimating the $\frac{T(T+1)}{2}$ elements in $\bm{\Omega}$ and $\nu$, we parametrise them as $\bm{\Omega}=  \tau [(\tau - \rho) \I_T + \rho \mathbf{1}_{T \times T}]$ and $\nu = \tau + T + 1$, respectively.
+The justification for this parametrisation is three-fold: (a) it implies an easy to interpret prior covariance matrix with $\tau$ on its diagonal and $\rho$ on its off-diagonals, (b) it reduces the number of parameters to  estimate to just two, and (c) simplifies the positive definite constraint on  $\bm{\Omega}$.
+To ensure positive definiteness and a proper covariance matrix, we impose the constraints $\tau > \max\[\rho - T \rho,\rho, 0\]$ (see Appendix \ref{app:priortissuecovariance}).
+The resulting third sub-problem is a bit more involved than the first two:
+	\begin{align*}
+	\tau^{(l + 1)}, \rho^{(l+1)} & = \underset{\tau, \rho}{\argmax} \frac{D (T^2 + T)}{2} \log \tau + \frac{DT}{2} \tau \log \tau + \frac{D(T+1)}{2} \log \[ \tau - (T - 1) \rho\] \\
+	& \,\,\,\,\,\,\,\,\,\, + \frac{D}{2} \tau \log \[ \tau - (T - 1) \rho\] + \frac{D(T^2 - 1)}{2} \log(\tau - \rho) +\frac{D(T - 1)}{2} \tau  \log(\tau - \rho) \\
+	& \,\,\,\,\,\,\,\,\,\, - D \sum_{t=1}^T \log \Gamma \( \frac{\tau + t + 1}{2} \) - w_1^{(l)} \tau^2 + w_2^{(l)} \tau \rho + w_3^{(l)} \tau, \\
+	& \text{subject to } \tau > \max\[\rho - T \rho,\rho, 0\],
+	\end{align*}
+	where
+	\begin{align*}
+	w_1^{(l)} & =  \frac{\tau^{(l)} + T + 2}{2} \sum_{d=1}^D \trace \[ (\bm{\Psi}_d^{(l)})^{-1}\] > 0,\\
+	w_2^{(l)} & = \frac{\tau^{(l)} + T + 2}{2} \sum_{d=1}^D \left\{\trace \[ (\bm{\Psi}_d^{(l)})^{-1}\] - \trace \[\mathbf{1}_{T \times T} \cdot (\bm{\Psi}_d^{(l)})^{-1}\] \right\} < w_1^{(l)},\\
+	w_3^{(l)} & = \frac{1}{2} \[D \sum_{t=1}^T \psi\(\frac{\tau^{(l)} + t + 2}{2}\) - \sum_{d=1}^D \log |\bm{\Psi}_d^{(l)}| \].
+	\end{align*}
+
+	- use mathematica to check convexity
+
+	- we use the ELBO for VB convergence and parameters for EB convergence
+
+	- we might be lenient for VB convergence. The extreme case would be 1 VB iteration per EB iteration (corresponds to tolerance of infinity).
+
+	\section{Simulations}
+
+	- Hamming distance to assess variable selection
+
+	\section{Data application}
+	We predict IC50 of the drugs.
+Other choices of efficacy measures are possible, but a discussion on the pros and cons of these different choices is not the aim of this paper.
+In any case, the presented methods may be used with any continuous efficay measure.
+
+	- check GSK open target EBI or ask Chris Wallace about it. Database that links targets to drugs and diseases. Might be useful for drugs co-data.
+
+	- check NCI-60
+
+	- check CTRP study
+
+	- check data in PharamcoGx R-package
+
+	- how to transform the data and deal with missingness (ask Oslo and read papers published on GDSC and CCLE repositories)
+
+	- create drug covariates from SMILES formulas with R-package rcdk.
+
+	- try non-negative matrix factorisation/independent component analysis on binary fingerprints to get principal components that characterise the different drugs.
+
+	- check whether compounds are PAINS \cite[]{baell_new_2010} or not. PAINS indicates whether compound is frequent false hitter.
+
+	- start with gene expression (one $g(j)$)
+
 	\subsection{GDSC}
+	After iteratively removing the cell line or drug with the highest proportion of missings until there are no more missing values, we end up with  IC50 estimates for 495 cell lines on 167 drugs.
+
+	\subsection{CCLE}
 
 	\section{Discussion}
-	- use copula's and marginal posteriors to find multivariate posteriors for 
-	contrast testing.
+	- use copula's and marginal posteriors to find multivariate posteriors for contrast testing.
 
-	- empirical Bayes gives extra information via estimation of the 
-	hyperparameters, such as informativeness of the drug information (via model 
-	for the $\gamma_d$) and relatedness of the tissues (via $\bm{\Omega}$ and 
-	$\nu$)
+	- empirical Bayes gives extra information via estimation of the hyper-parameters, such as informativeness of the drug information (via model for the $\gamma_d$) and relatedness of the tissues (via $\bm{\Omega}$ and $\nu$)
 
@@
 
-	\section*{Supplementary Material}
-	Supplementary Material is available online from 
-	\url{https://github.com/magnusmunch/cambridge}.
+	- find more literature on inverse Gaussian prior for variances and modelling variances with covariates.
 
-	\section*{Reproducible Research}
-	All results and documents may be recreated from 
-	\url{https://github.com/magnusmunch/cambridge}.
-
-	\section*{Acknowledgements}
-	We thank Zhi Zhao and Leiv Tore Salte R{\o}nneberg from the Oslo Centre for 
-	Biostatistics and Epidemiology for help with the pre-processing of the data. 
-	\textit{Conflict of Interest}: None declared.
+	Recently, \cite{moran_variance_2018} have argued that in high-dimensional regression problems, \text{a priori} dependence between the regression parameters and error variance may induce biased estimation.
+In (\ref{eq:betaprior}), we assume such an \textit{a priori} dependence.
+However, we contend that independent regression parameters and error variance are not desirable in our setting for two reasons: (a) the empirical Bayes estimation requires some notion of scale to combine the separate regressions, which may be done by inclusion of the error variance in the regression parameter prior model.
+(b) The variational Bayes approximation introduced in Section \ref{sec:variationalbayes} gives a posterior that is similar to the one we would obtain with a dependent regression parameters and error variance prior model.
+This renders a better approximate expectation in the empirical Bayes iterations.
 
@@
 
+	\titleformat{\section}{\normalfont\Large\bfseries}{\appendixname~\thesection.}{1em}{}
+	\begin{appendix}
+		\section{Prior tissue covariance constraints}\label{app:priortissuecovariance}
+		We parametrise the prior covariance as a matrix with $\tau$ on the diagonal and $\rho$ on its off-diagonals. To that end we set $\nu = \tau + T + 1$ and the $T \times T$-dimensional matrix
+		$$
+		\bm{\Omega} = \tau \begin{bmatrix}
+		\tau &  & \rho \\
+		& \ddots &  \\
+		\rho & & \tau
+		\end{bmatrix},
+		$$
+		such that $\E_{\pi(\mathbf{\Xi}_d)} (\mathbf{\Xi}_d) = (\nu - T - 1)^{-1} \bm{\Omega}$ is of the desired form.
+To ensure a non-degenerate prior distribution we require $\tau > 0$ and $\bm{\Omega}$ PD.
+We use $\tau > 0$ to note that a PD $\bm{\Omega}$ implies positive roots $\lambda$ of the characteristic polynomial:
+		\begin{align*}
+		\left| \begin{bmatrix}
+		\tau &  & \rho \\
+		& \ddots &  \\
+		\rho & & \tau
+		\end{bmatrix} -\lambda \I_T \right| = (\tau - \lambda - \rho)^{T-1}[\tau - \lambda - (1-T)\rho] = 0.
+		\end{align*}
+		For $T > 1$, the solutions are:
+		$$
+		(\lambda = \tau - \rho) \lor (\lambda = T \rho - \rho + \tau),
+		$$
+		which are positive if the following condition holds: $\tau > \max(\rho,\rho - T \rho)$.
+We combine this with $\tau > 0$, to arrive at the full proper prior constraint: $\tau > \max(0,\rho,\rho - T \rho)$.
+
+	\end{appendix}
+
 \end{document}

### 2019-11-20 docs/manuscript1.Rnw
created different selection strategies in analysis_gdsc.R

@@
 	pathways and developmental stages of the drugs. Additional online
-	repositories might provide extra information such as the molecular weight of
+	repositories may provide extra information such as the molecular weight of
 	the compounds or the publication signatures of the molecular features. In some
-	cases, prior knowledge on the drug efficacies might be available,
+	cases, prior knowledge on the drug efficacies may be available,
 	from previous experiments. We propose to include these possibly beneficial 
@@
 	a data-driven manner. More specifically, we estimate a normal 
-	inverse Gaussian model, where the amount of regularisation is estimated by an 
+	inverse Gaussian model, where the extent of regularisation is estimated by an 
 	adaptive empirical Bayes procedure, guided by the external information.
@@
 	in Section \ref{sec:estimation}. In Section \ref{sec:gdsc} we analyse the
-	GDSC data, and we end with a discussion in Section \ref{sec:discussion}.
+	GDSC data, and we end with a discussion in Section \ref{sec:discussion} on the 
+	pros and cons of the proposed method.
 
@@
 	prior 
-	variance into a local components $\gamma_{jd}^2$ and a global component
+	variance into a local component $\gamma_{jd}^2$ and a global component
 	$\tau_d^2$. The local components are supposed to capture local variation in
 	the model parameters $\beta_{jd}$, while the global component is supposed to
-	capture the general trend in $\beta_{jd}$. Addionally, for appropriate local
-	shrinkage in global-local 
+	capture the general trend in $\bm{\beta}_{d}$. For appropriate 
+	local shrinkage in global-local 
 	shrinkage models it is important  to account for 
@@
   $\X= \begin{bmatrix} \x_1 & \cdots \x_n \end{bmatrix} \tr =\I_p$, with fixed 
-  $\tau_d^2=\sigma_d^2=1$, the resulting
+  $\tau_d^2=\sigma_d^2=1$, the resulting conditional
   posterior mean for the $\beta_{jd}$ is
@@
 	\subsection{External information}\label{sec:external}
-	In drug sensitivity prediction models external information on both the drugs
+	In drug sensitivity prediction models, external information on both the drugs
 	and features is often available. Here, we assume this information to be 
@@
 	drug `covariates' $\mathbf{z}_{dh}$, for $h=1,\dots, H$. An example of a 
-	(binary) features covariate is target pathway presence, with $c_{jdg}=0$ if 
+	(binary) feature covariate is target pathway presence, with $c_{jdg}=0$ if 
 	gene $j$ is present in the 

### 2019-11-20 code/data_gdsc.R

@@ -15,6 +15,9 @@ if(file.exists("data/v17.3_fitted_dose_response.csv")) {
 if(file.exists("data/sanger1018_brainarray_ensemblgene_rma.txt.gz")) {
   expr <- read.table("data/sanger1018_brainarray_ensemblgene_rma.txt.gz",
                      header=TRUE, stringsAsFactors=FALSE)
+  # same data, different file
+  # expr <- read.csv("data/Cell_line_RMA_proc_basalExp.txt", 
+  #                  header=TRUE, sep="\t")
   
 } else {
   expr <- read.table("ftp://ftp.sanger.ac.uk/pub/project/cancerrxgene/releases/current_release/sanger1018_brainarray_ensemblgene_rma.txt.gz", 

### 2019-11-20 rpackage/R/main_functions.R

@@ -292,7 +292,7 @@ semnig <- function(x, y, C, Z, unpenalized=NULL, standardize=FALSE,
   }
   
   # create posterior mean and variance objects
-  mpost <- list(beta=mu, gamma.sq=sapply(1:D, function(d) {
+  mpost <- list(beta=mu, gamma.sq=lapply(1:D, function(d) {
     old.eb$mpriorf[[d]]*sqrt(old.vb$delta[[d]]/old.eb$lambdaf)*
       ratio_besselK(sqrt(old.vb$delta[[d]]*old.eb$lambdaf)/
                       old.eb$mpriorf[[d]], 1)}),
@@ -301,7 +301,7 @@ semnig <- function(x, y, C, Z, unpenalized=NULL, standardize=FALSE,
       old.eb$mpriord[d]*sqrt(old.vb$eta[[d]]/old.eb$lambdad)*
         ratio_besselK(sqrt(old.vb$eta[[d]]*old.eb$lambdad)/
                         old.eb$mpriord[d], (r[d] + 1)/2)}))
-  vpost <- list(beta=Sigma, gamma.sq=sapply(1:D, function(d) {
+  vpost <- list(beta=Sigma, gamma.sq=lapply(1:D, function(d) {
     old.eb$mpriorf[[d]]^2*old.vb$delta[[d]]/old.eb$lambdaf*
       (1 - ratio_besselK(sqrt(old.vb$delta[[d]]*old.eb$lambdaf)/
                            old.eb$mpriorf[[d]], 1)^2)}),
@@ -325,4 +325,32 @@ semnig <- function(x, y, C, Z, unpenalized=NULL, standardize=FALSE,
   class(out) <- c("semnig")
   return(out)
   
+}
+
+# estimate ELBO on new data
+new.elbo <- function(object, newx, newy) {
+  D <- ncol(newy)
+  p <- sapply(newx, ncol)
+  n <- nrow(newy)
+  d <- 1
+  out <- sapply(1:D, function(d) {
+    aux <- list(ldetSigma=determinant(object$vb$Sigma[[d]])$modulus, 
+                ytXmu=as.numeric(t(newy[, d]) %*% newx[[d]] %*% 
+                                  object$vb$mu[[d]]), 
+                trXtXSigma=sum(diag(t(newx[[d]]) %*% newx[[d]] %*% 
+                                      object$vb$Sigma[[d]])), 
[…]

### 2019-11-22 docs/manuscript1.Rnw
fixed error in simulation_gdsc.R simulation 3

@@
   several choices of $\beta_{jd}$ prior.
+
 <<dens_kappa, fig.cap="Implied prior densities $\\pi(\\kappa_{jd})$ for the (a) NIG, (b) Student's $t$, and (c) lasso priors.
@@
 	\section{Simulations}
-	\subsection{Fixed drug component}
+	\subsection{Setup}
 	This section investigates the empirical 
@@
 
-	A pre-processing step, as explained in Section \ref{sec:gdsc}, selects around 
-	300 features per drug, while the 168 drug sensitivities for 482 cell lines are 
-	simulated from model (\ref{eq:likelihood}) and 
-	(\ref{eq:prior}), with two adjustments: we fix $\forall d: \tau_d^2=1$ 
-	and draw the error variances as $\forall d: \sigma_d^2 \sim \Gamma^{-1}(3,2)$,
+	A pre-processing step selects 100 features with largest variance, while the 
+	168 drug sensitivities for 482 cell lines are 
+	simulated from three versions of model (\ref{eq:likelihood}) and 
+	(\ref{eq:prior}). All versions draw the error variances as 
+	$\forall d: \sigma_d^2 \sim \Gamma^{-1}(3,2)$,
 	such that the prior $\sigma_d^2$ mean and variance are both one. 
-	We create four external dummy feature covariates that code for four 
-	randomly assigned groups of features. We set $\bm{\alpha}_{\text{feat}}$ such
+	Version 1 fixes $\forall d: \tau_d^2=1$, version 2 fixes
+	$\forall j,d: \gamma_{jd}^2=1$, while version 3 draws both the $\tau_d^2$ and
+	$\gamma_{jd}^2$ according to (\ref{eq:prior}). In version 1, we
+	create four external dummy feature covariates that code for four 
+	randomly assigned approximately equally sized groups of features. We set
+	$\bm{\alpha}_{\text{feat}}$ such
 	that the $\gamma_{jd}^2$ of the four groups of features have prior means
@@
 	($\bm{\alpha}_{\text{feat}}=\begin{bmatrix} 1 & 1 & 3 & 7 \end{bmatrix} \tr$). 
-	The prior scale parameter is set to $\lambda_{\text{feat}}=1$. We estimate 
-	four models: (i) the NIG model as in Section \ref{sec:estimation}, (ii) an 
-	NIG model where we fix $\lambda_{\text{feat}}$ to its true value, and two NIG
-	models without the inclusion of the external covariates, again with 
-	$\lambda_{\text{feat}}$ either (iii) estimated or (iv) fixed to its true 
-	value. Exclusion of the external covariates as in models (iii) and (iv) 
+	The prior scale parameter is set to $\lambda_{\text{feat}}=1$. Version 2 
+	follows a similar procedure for the $\tau_d^2$. We randomly create four
+	groups of drugs with corresponding external drug dummy variables and set
+	$\bm{\alpha}_{\text{drug}}=\begin{bmatrix} 1 & 1 & 3 & 7 \end{bmatrix} \tr$,
+	such that we have $\chi_{d} \in \{ 1, 1/2, 1/4, 1/8 \}$. The
+	scale is set to $\lambda_{\text{drug}}=1$. Version 3 combines the procedures
+	from versions 1 and 2 to draw the $\gamma_{jd}^2$ and $\tau_d^2$.
+
+	We estimate four models: (i) an NIG model that does include external 
+	covariates in the estimation, (ii) the NIG model estimated as in Section 
+	\ref{sec:estimation}, and regular (iii) lasso and (iv) ridge models.
+	Exclusion of the external covariates as in model (i)
 	amounts to direct estimation of one common expected 
-	prior mean $\phi$, instead of regression estimates for the $\phi_{jd}$ as
-	in models (i) and (ii). All 
+	prior means $\phi$ and/or $\chi$, instead of regression estimates for the
+	$\phi_{jd}$ and/or $\chi_d$ as in model (ii). All 
 	models are fit on 241 randomly selected cell lines, while performance measures 
@@
 
-<<include=FALSE>>=
-load("results/simulations_gdsc_set1.Rdata")
-mapmse <- apply(res[, c(7:12)], 2, median)
-mapmset <- apply(res[, c(13:18)], 2, median)
-melbo <- apply(elbo, 2, median)
-phi <- 1/(set$alphaf %*% t(unname(model.matrix(~ factor(1:4)))))
-mphi <- apply(1/est[, c(11, 13)], 2, median)
-mlambdaf <- apply(est[, c(5, 12)], 2, median)
+	\subsection{Results}
+<<simulation_gdsc>>=
+res1 <- read.table("results/simulations_gdsc_res1.txt", row.names=NULL)
+res2 <- read.table("results/simulations_gdsc_res2.txt", row.names=NULL)
+res3 <- read.table("results/simulations_gdsc_res3.txt", row.names=NULL)
+temp1 <- res1[, 1]
+temp2 <- res2[, 1]
+temp3 <- res3[, 1]
+res1 <- as.matrix(res1[, -1])
+res2 <- as.matrix(res2[, -1])
+res3 <- as.matrix(res3[, -1])
+rownames(res1) <- temp1
+rownames(res2) <- temp2
+rownames(res3) <- temp3
+lambda <- apply(res1[rownames(res1)=="lambdaf", c(1, 2)], 2, median)
+apply(res2[rownames(res2)=="lambdad", c(1, 2)], 2, median)
+apply(res3[rownames(res3)=="lambdaf", c(1, 2)], 2, median)
+apply(res3[rownames(res3)=="lambdad", c(1, 2)], 2, median)
+est.phi <- median(1/res[rownames(res)=="alphaf0", 1])
+true.phi <- mean(1/(c(1, 1, 3, 7) %*% t(unname(model.matrix(~ factor(1:4))))))
+melbo <- apply(res[rownames(res)=="elbo", c(1, 2)], 2, median)
+melbot <- apply(res[rownames(res)=="elbot", c(1, 2)], 2, median)
+
+mphi <- median(1/res[rownames(res)=="alphaf0", 1])
+mlambdaf <- median(1/res[rownames(res)=="lambdaf", 1])
 @	
 	Figure \ref{fig:simulations_gdsc_est1} shows the estimated
-	$\bm{\alpha}_{\text{feat}}$ together with its true value for models (i) and 
-	(ii). The figure shows that $\bm{\alpha}_{\text{feat}}$ is consistently
-	underestimated 
-	in the full NIG model (both with estimated and fixed $\lambda_{\text{feat}}$),
-	while $\lambda_{\text{feat}}$ is overestimated as 
-	\Sexpr{round(mlambdaf[1], 2)} in model (i) (the true value is 1). 
-	This translates into
-	underestimated prior means $\phi_{jd}$ as presented in Figure 
-	\ref{fig:simulations_gdsc_phi1}, which in turn implies underestimated 
-	$\beta_{jd}$ variances. The overestimated $\lambda_{\text{feat}}$ translates 
+	$\bm{\alpha}_{\text{feat}}$ together with its true value for model
+	(ii) in version 1 of the simulation study. Figure 
+	\ref{fig:simulations_gdsc_est1}a shows that 
+	$\bm{\alpha}_{\text{feat}}$ is slightly but 
+	consistently underestimated. On the $\phi_{jd}$ scale this underestimation is
+	almost undetectable as depicted in Figure
+	\ref{fig:simulations_gdsc_est1}b. $\lambda_{\text{feat}}$ is slightly 
+	overestimated with median estimate \Sexpr{round(lambda[2], 2)} (the
+	true value is 1). The overestimated $\lambda_{\text{feat}}$ translates 
 	into underestimated $\gamma_{jd}^2$ variances. Underestimation of variances is 
 	a common phenomenon in variational Bayes approximations. In addition, 
-	$\bm{\alpha}_{\text{feat}}$ and $\lambda_{\text{feat}}$ are higher level 
-	parameters, which are, in general, hard to estimate. We note 
-	that even though $\bm{\alpha}_{\text{feat}}$ and thus $\phi_{jd}$ are 
-	estimated with bias, the orders of their elements are correct.
-<<simulations_gdsc_est1, fig.cap="Estimated hyperparameters $\\bm{\\alpha}_{\\text{feat}}$ from simulations with (a) estimated and (b) fixed $\\lambda_{\\text{feat}}$ (models (i) and (ii), respectively).", out.width="100%", fig.asp=1/2>>=
-@
-
-<<simulations_gdsc_phi1, fig.cap="Estimated prior means $\\bm{\\phi}_{jd}$ for the four groups of features from simulations with (a) estimated and (b) fixed $\\lambda_{\\text{feat}}$ (models (i) and (ii), respectively).", out.width="100%", fig.asp=1/2>>=
-@
-
-	The models (iii) and (iv) give median $\phi$ estimates of
-	\Sexpr{round(mphi[1], 2)} and \Sexpr{round(mphi[2], 2)}, 
-	respectively. The $\phi$ estimates are
-	underestimated, compared to true $\phi$ mean \Sexpr{round(mean(phi), 2)}. 
-	Model (iii) slightly underestimates the median
-	$\lambda_{\text{feat}}$ at \Sexpr{round(mlambdaf, 2)[2]} 
-	(the true value is one) and thus overestimates the 
+	$\lambda_{\text{feat}}$ is a higher level scale
+	parameter, which are, in general, hard to estimate. 
+<<simulations_gdsc_est1, fig.cap="Estimated (a) hyperparameters $\\bm{\\alpha}_{\\text{feat}}$ and (b) prior means $\\bm{\\phi}_{jd}$ from simulations with fixed $\\tau_d^2$.", out.width="100%", fig.asp=1/2>>=
+@
+
+	Model (i) gives a median $\phi$ estimate of \Sexpr{round(est.phi, 2)},  
+	equal to the true mean of the $\phi_{jd}$, \Sexpr{round(true.phi, 2)}. 
+	$\lambda_{\text{feat}}$ is slightly underestimated with median estimate
+	\Sexpr{round(lambda[1], 2)} (the true value is one) and thus overestimates the 
 	prior $\gamma_{jd}^2$ variance. This is likely to compensate for the common
-	prior mean 
-	$\phi$, that is not able to capture part of the variation in the true feature
-	specific $\phi_{jd}$. The median evidence 
+	prior mean $\phi$, that is not able to capture part of the variation in the 
+	true feature specific $\phi_{jd}$.
+
+	The median evidence 
 	lower bounds (ELBO), which approximate the marginal likelihoods of models
@@
 	data and gaps between training and test prediction MSE are displayed in Table
-	\ref{tab:pmse1}. Table \ref{tab:pmse1} confirms that the models that
+	\ref{tab:mse1}. Table \ref{tab:mse1} confirms that the models that
 	include the external data (models (i) and (ii)) learn the underlying structure 
@@
 	data.
-<<pmse1>>= 
+<<mse1>>= 
 library(knitr, quietly=TRUE, verbose=FALSE)
 suppressWarnings(library(kableExtra))
-load("results/simulations_gdsc_set1.Rdata")
-mapmse <- apply(res[, c(7:12)], 2, median)
-mapmset <- apply(res[, c(13:18)], 2, median)
-tab <- rbind(mapmse, mapmse - mapmset)[, -c(5, 6)]
-rownames(tab) <- c("$\\text{MSE}_{\\text{test}}$",
-                  "$\\text{MSE}_{\\text{test}} - \\text{MSE}_{\\text{train}}$")
+res <- read.table("results/simulations_gdsc_res1.txt", row.names=NULL)
+temp <- res[, 1]
+res <- as.matrix(res[, -1])
+rownames(res) <- temp
+memse <- apply(res[rownames(res)=="emse", ], 2, median)
+mpmse <- apply(res[rownames(res)=="pmse", ], 2, median)
+mpmset <- apply(res[rownames(res)=="pmset", ], 2, median)
+tab <- rbind(memse, mpmse, mpmse - mpmset)
+rownames(tab) <- c("$\\text{EMSE}_{\\text{test}}$",
+                   "$\\text{PMSE}_{\\text{test}}$",
+                   "$\\text{PMSE}_{\\text{test}} - \\text{PMSE}_{\\text{train}}$")
 options(knitr.kable.NA="", digits=3)
 kable(tab,
-      col.names=c("(i)", "(ii)", "(iii)", "(iv)"),
-      caption="Median prediction MSEs for the models (i)-(iv) in the fixed 
+      col.names=colnames(tab),
+      caption="Median estimation MSE, prediction MSEs, and difference between 
+      prediction MSEs on the test and training data in the fixed 
       $\\tau_d^2$ simulation setting.", 
@@
 
-  \subsection{Random drug component}
 	To investigate the interplay of the feature and drug variance components, we
@@
 
-<<include=FALSE>>=
-load("results/simulations_gdsc_set2.Rdata")
-mapmse <- apply(res[, c(7:12)], 2, median)
-mapmset <- apply(res[, c(13:18)], 2, median)
-melbo <- apply(elbo, 2, median)
-phi <- 1/(set$alphaf %*% t(unname(model.matrix(~ factor(1:4)))))
-chi <- 1/(set$alphad %*% t(unname(model.matrix(~ factor(1:4)))))
-mphi <- apply(1/est[, c(21, 25)], 2, median)
-mchi <- apply(1/est[, c(22, 26)], 2, median)
-mlambdaf <- apply(est[, c(9, 23)], 2, median)
-mlambdad <- apply(est[, c(10, 24)], 2, median)
-@	
-	Figures \ref{fig:simulations_gdsc_est2} and 
-	\ref{fig:simulations_gdsc_phi2} display the estimated $\bm{\alpha}$, 
+	Figures \ref{fig:simulations_gdsc_est2} displays the estimated $\bm{\alpha}$,
 	$\phi_{jd}$, and $\chi_d$ for models (i) and (ii). The scale parameters were
-	estimated as $\lambda_{\text{feat}}=\Sexpr{round(mlambdaf[1], 2)}$ and
-	$\lambda_{\text{drug}}=\Sexpr{round(mlambdad[1], 2)}$ in model (i). From 
+	estimated as $\lambda_{\text{feat}}=\Sexpr{round(rnorm(1), 2)}$ and
+	$\lambda_{\text{drug}}=\Sexpr{round(rnorm(1)[1], 2)}$ in model (i). From
 	Figure \ref{fig:simulations_gdsc_est2} we see that estimation of
@@
 	to a suboptimal EB estimation performance.
-
-<<simulations_gdsc_est2, fig.cap="Estimated hyperparameters $\\bm{\\alpha}_{\\text{feat}}$ and $\\bm{\\alpha}_{\\text{drug}}$ from simulations with (a) \\& (c) estimated and (b) \\& (d) fixed $\\lambda_{\\text{feat}}$ and $\\lambda_{\\text{drug}}$ (models (i) and (ii), respectively).", out.width="100%", fig.asp=1>>=
-@
-
-<<simulations_gdsc_phi2, fig.cap="Estimated prior means $\\bm{\\phi}_{jd}$ and $\\bm{\\chi}_{d}$ for the four groups of features and drugs from simulations with (a) \\& (c) estimated and (b) \\& (d) fixed $\\lambda_{\\text{feat}}$ and $\\lambda_{\\text{drug}}$ (models (i) and (ii), respectively).", out.width="100%", fig.asp=1>>=
+<<simulations_gdsc_est2, fig.cap="Estimated (a) hyperparameters $\\bm{\\alpha}_{\\text{feat}}$ and (b) prior means $\\bm{\\phi}_{jd}$ from simulations with fixed $\\tau_d^2$.", out.width="100%", fig.asp=1/2>>=
 @
@@
 	suboptimal, the external data do add something in predictive performance.
-<<pmse2>>= 
-library(knitr, quietly=TRUE, verbose=FALSE)
-suppressWarnings(library(kableExtra))
-load("results/simulations_gdsc_set2.Rdata")
-mapmse <- apply(res[, c(7:12)], 2, median)
-mapmset <- apply(res[, c(13:18)], 2, median)
-tab <- rbind(mapmse, mapmse - mapmset)[, -c(5, 6)]
-rownames(tab) <- c("$\\text{MSE}_{\\text{test}}$",
-                  "$\\text{MSE}_{\\text{test}} - \\text{MSE}_{\\text{train}}$")
-options(knitr.kable.NA="", digits=3)
-kable(tab,
-      col.names=c("(i)", "(ii)", "(iii)", "(iv)"),
-      caption="Median prediction MSEs for the models (i)-(iv) in the random 
-      $\\tau_d^2$ simulation setting.", 
-      format="latex", booktabs=TRUE, escape=FALSE) %>%
-  kable_styling(latex_options=c("HOLD_position"))
-@
 
@@
 suppressWarnings(library(kableExtra))
-load("results/data_gdsc_res1.Rdata")
-tab1 <- cbind(tab[, 1], c(NA, 0, NA), tab[, c(3, 2)], c(NA, 0, NA), 
-              tab[, c(4, 5)])
-options(knitr.kable.NA="", digits=3)
-kable(tab1,
-      col.names=c("intercept", "approved", "development", 
-                  "experimental", "targeted", "unkown", "cytotoxic"),
+# load("results/data_gdsc_res1.Rdata")
+# tab1 <- cbind(tab[, 1], c(NA, 0, NA), tab[, c(3, 2)], c(NA, 0, NA), 
+#               tab[, c(4, 5)])
+# options(knitr.kable.NA="", digits=3)
+# kable(tab1,
+#       col.names=c("intercept", "approved", "development", 
+#                   "experimental", "targeted", "unkown", "cytotoxic"),
+#       caption="$\\bm{\\alpha}_{\\text{drug}}$ estimates from GDSC data.", 
+#       format="latex", booktabs=TRUE, escape=FALSE) %>%
+#   kable_styling(latex_options=c("HOLD_position"))
+kable(matrix(1, nrow=1, ncol=2),
       caption="$\\bm{\\alpha}_{\\text{drug}}$ estimates from GDSC data.", 
@@
 suppressWarnings(library(kableExtra))
-load("results/data_gdsc_res1.Rdata")
-tab2 <- cbind(tab[, 6], c(NA, 0, 0), tab[, 7])
-options(knitr.kable.NA="", digits=3)
-kable(tab2,
-      col.names=c("intercept", "not in pathway", "in pathway"),
+# load("results/data_gdsc_res1.Rdata")
+# tab2 <- cbind(tab[, 6], c(NA, 0, 0), tab[, 7])
+# options(knitr.kable.NA="", digits=3)
+# kable(tab2,
+#       col.names=c("intercept", "not in pathway", "in pathway"),
+#       caption="$\\bm{\\alpha}_{\\text{feat}}$ estimates from GDSC data.", 
+#       format="latex", booktabs=TRUE, escape=FALSE) %>%
+#   kable_styling(latex_options=c("HOLD_position"))
+kable(matrix(1, nrow=1, ncol=2),
       caption="$\\bm{\\alpha}_{\\text{feat}}$ estimates from GDSC data.", 
@@
   is lowest in the NIG models 
-  (\Sexpr{round(mapmse[2], 2)} and \Sexpr{round(mapmse[3], 2)}, in the
+  (\Sexpr{round(rnorm(1))} and \Sexpr{rnorm(1)}, in the
   NIG1 and NIG2 models, respectively). The bSEM model is not able to 
   improve on the null model with an median prediction MSE of 
-  \Sexpr{round(mapmse[4], 2)} compared to the null model's MSE of 
-  \Sexpr{round(mapmse[1], 2)}.
+  \Sexpr{round(rnorm(1))} compared to the null model's MSE of 
+  \Sexpr{round(rnorm(1))}.
 <<results_data>>= 
-load("results/data_gdsc_res1.Rdata")
-mapmse <- apply(apmse, 2, median)
-melbo <- apply(elbo, 2, median)
-tab <- rbind(mapmse, c(melbo, NA, NA))
-rownames(tab) <- c("MSE", "ELBO")
+library(knitr, quietly=TRUE, verbose=FALSE)
+suppressWarnings(library(kableExtra))
+# load("results/data_gdsc_res1.Rdata")
+# mapmse <- apply(apmse, 2, median)
+# melbo <- apply(elbo, 2, median)
+# tab <- rbind(mapmse, c(melbo, NA, NA))
+# rownames(tab) <- c("MSE", "ELBO")
+# options(knitr.kable.NA="", digits=3)
+# kable(tab,
+#       col.names=c("NIG1", "NIG2", "bSEM", "lasso", "ridge"),
+#       caption="Median prediction MSE and median ELBO estimated on the test data", 
+#       format="latex", booktabs=TRUE, escape=FALSE) %>%
+#   kable_styling(latex_options=c("HOLD_position"))
 options(knitr.kable.NA="", digits=3)
-kable(tab,
-      col.names=c("NIG1", "NIG2", "bSEM", "lasso", "ridge"),
+kable(matrix(1, nrow=1, ncol=2),
       caption="Median prediction MSE and median ELBO estimated on the test data", 

### 2019-11-22 docs/supplement1.Rnw
fixed error in simulation_gdsc.R simulation 3

@@
   If we fill in the expectations and variances, we arrive at the estimating
-  equations in MD equations (\ref{md-eq:vbequations}).
+  % equations in MD equations (\ref{md-eq:vbequations}).
 
@@
   MCMC samples from the posterior corresponding to model 
-  (\ref{md-eq:prior}) may be generated for each equation independently. 
+  % (\ref{md-eq:prior}) may be generated for each equation independently. 
   MCMC samples may be generated by iteratively sampling the following 

### 2019-11-22 code/simulations_gdsc.R

@@ -10,337 +10,337 @@ sapply(packages, library, character.only=TRUE)
 ### load data
 load(file="data/data_gdsc_dat1.Rdata")
 
-################################# simulation 1 #################################
-### data preparation
-# select features
-D <- ncol(resp.prep)
-psel <- 100
-o <- order(-apply(expr.prep, 2, sd))
-idsel <- rep(list(o[c(1:psel)]), D)
-expr.sel <- lapply(idsel, function(s) {expr.prep[, s]})
-x <- lapply(expr.sel, function(s) {scale(s)})
-
-### data preparation
-p <- sapply(x, ncol)
-n <- nrow(expr.prep)
-ntrain <- floor(n/2)
-
-### simulation settings
-alphaf <- c(1, 1, 3, 7)
-lambdaf <- 1
-shape <- 3
-rate <- 2
-
-nreps <- 50
-methods <- c("NIG1", "NIG2", "lasso", "ridge")
-#
-# pmse <- pmset <- emse <- matrix(NA, nrow=nreps, ncol=length(methods) + 1)
-# est <- matrix(NA, nrow=5*nreps, ncol=length(methods) - 2)
-# elbo <- elbot <- matrix(NA, nrow=nreps, ncol=length(methods) - 2)
-
-# setup cluster
-cl <- makeCluster(ncores)
-if(ncores > 1) {
-  registerDoParallel(cl)
-} else {
-  registerDoSEQ()
-}
-res <- foreach(r=1:nreps, .packages=packages) %dopar% {
-# for(r in 1:nreps) {
-  cat("\r", "rep", r)
-  set.seed(2019 + r)
-
-  ### simulate parameters
-  C <- lapply(1:D, function(d) {
-    cbind(1, t(replicate(p[d], sample(c(1, 0, 0, 0), 4)[-1])))})
-  gamma <- lapply(1:D, function(d) {
-    sqrt(rinvgauss(p[d], 1/as.numeric(C[[d]] %*% alphaf), lambdaf))})
-  sigma <- 1/sqrt(rgamma(D, shape, rate))
-  beta <- lapply(1:D, function(d) {rnorm(p[d], 0, gamma[[d]]*sigma[d])})
-
-  # simulate data
-  idtrain <- sample(1:n, ntrain)
-  xtrain <- lapply(x, function(s) {scale(s[idtrain, ])})
[…]

### 2019-11-22 docs/manuscript1.Rnw
commit before rerunning analysis_gdsc.R

@@
 	We estimate four models: (i) an NIG model that does include external 
-	covariates in the estimation, (ii) the NIG model estimated as in Section 
-	\ref{sec:estimation}, and regular (iii) lasso and (iv) ridge models.
-	Exclusion of the external covariates as in model (i)
+	covariates in the estimation called NIG1 in the following, (ii) the NIG model
+	estimated as in Section \ref{sec:estimation} called NIG2, and regular (iii)
+	lasso and (iv) ridge models.
+	Exclusion of the external covariates as in the NIG1 model
 	amounts to direct estimation of one common expected 
 	prior means $\phi$ and/or $\chi$, instead of regression estimates for the
-	$\phi_{jd}$ and/or $\chi_d$ as in model (ii). All 
+	$\phi_{jd}$ and/or $\chi_d$ as in the NIG2 model. All 
 	models are fit on 241 randomly selected cell lines, while performance measures 
@@
 	\subsection{Results}
-<<simulation_gdsc>>=
-res1 <- read.table("results/simulations_gdsc_res1.txt", row.names=NULL)
-res2 <- read.table("results/simulations_gdsc_res2.txt", row.names=NULL)
-res3 <- read.table("results/simulations_gdsc_res3.txt", row.names=NULL)
-temp1 <- res1[, 1]
-temp2 <- res2[, 1]
-temp3 <- res3[, 1]
-res1 <- as.matrix(res1[, -1])
-res2 <- as.matrix(res2[, -1])
-res3 <- as.matrix(res3[, -1])
-rownames(res1) <- temp1
-rownames(res2) <- temp2
-rownames(res3) <- temp3
-lambda <- apply(res1[rownames(res1)=="lambdaf", c(1, 2)], 2, median)
-apply(res2[rownames(res2)=="lambdad", c(1, 2)], 2, median)
-apply(res3[rownames(res3)=="lambdaf", c(1, 2)], 2, median)
-apply(res3[rownames(res3)=="lambdad", c(1, 2)], 2, median)
+<<simulation_gdsc_res1>>=
+res <- read.table("results/simulations_gdsc_res1.txt", row.names=NULL)
+temp <- res[, 1]
+res <- as.matrix(res[, -1])
+rownames(res) <- temp
+true.phi <- mean(1/(c(1, 1, 3, 7) %*% t(unname(model.matrix(~ factor(1:4))))))
 est.phi <- median(1/res[rownames(res)=="alphaf0", 1])
-true.phi <- mean(1/(c(1, 1, 3, 7) %*% t(unname(model.matrix(~ factor(1:4))))))
-melbo <- apply(res[rownames(res)=="elbo", c(1, 2)], 2, median)
-melbot <- apply(res[rownames(res)=="elbot", c(1, 2)], 2, median)
-
-mphi <- median(1/res[rownames(res)=="alphaf0", 1])
-mlambdaf <- median(1/res[rownames(res)=="lambdaf", 1])
+est.lambdaf <- apply(res[rownames(res)=="lambdaf", c(1, 2)], 2, median)
+tab <- cbind(apply(res[rownames(res)=="elbo", ], 2, median),
+             apply(res[rownames(res)=="elbot", ], 2, median) -
+               apply(res[rownames(res)=="elbo", ], 2, median),
+             apply(res[rownames(res)=="emse", ], 2, median),
+             apply(res[rownames(res)=="pmse", ], 2, median),
+             apply(res[rownames(res)=="pmse", ], 2, median) - 
+               apply(res[rownames(res)=="pmset", ], 2, median))
+tab <- round(tab, 3)
+tab[which(tab[, 1]==max(tab[, 1], na.rm=TRUE)), 1] <- 
+  paste0("\\textbf{", tab[which(tab[, 1]==max(tab[, 1], na.rm=TRUE)), 1], "}")
+tab[which(tab[, 2]==min(tab[, 2], na.rm=TRUE)), 2] <- 
+  paste0("\\textbf{", tab[which(tab[, 2]==min(tab[, 2], na.rm=TRUE)), 2], "}")
+tab[which(tab[, 3]==min(tab[, 3], na.rm=TRUE)), 3] <- 
+  paste0("\\textbf{", tab[which(tab[, 3]==min(tab[, 3], na.rm=TRUE)), 3], "}")
+tab[which(tab[, 4]==min(tab[, 4], na.rm=TRUE)), 4] <- 
+  paste0("\\textbf{", tab[which(tab[, 4]==min(tab[, 4], na.rm=TRUE)), 4], "}")
+tab[which(tab[, 5]==min(tab[, 5], na.rm=TRUE)), 5] <- 
+  paste0("\\textbf{", tab[which(tab[, 5]==min(tab[, 5], na.rm=TRUE)), 5], "}")
+colnames(tab) <- c("$\\text{ELBO}_{\\text{test}}$",
+                   "$\\text{ELBO}_{\\text{train}} 
+                   - \\text{ELBO}_{\\text{test}}$",
+                   "$\\text{EMSE}_{\\text{test}}$",
+                   "$\\text{PMSE}_{\\text{test}}$",
+                   "$\\text{PMSE}_{\\text{test}} 
+                   - \\text{PMSE}_{\\text{train}}$")     
 @	
   Figure \ref{fig:simulations_gdsc_est1} shows the estimated
-	$\bm{\alpha}_{\text{feat}}$ together with its true value for model
-	(ii) in version 1 of the simulation study. Figure 
+	$\bm{\alpha}_{\text{feat}}$ together with its true value for NIG2 in 
+	version 1 of the simulation study. Figure 
 	\ref{fig:simulations_gdsc_est1}a shows that 
@@
 	\ref{fig:simulations_gdsc_est1}b. $\lambda_{\text{feat}}$ is slightly 
-	overestimated with median estimate \Sexpr{round(lambda[2], 2)} (the
+	overestimated with median estimate \Sexpr{round(est.lambdaf[2], 2)} (the
 	true value is 1). The overestimated $\lambda_{\text{feat}}$ translates 
@@
 	parameter, which are, in general, hard to estimate. 
-<<simulations_gdsc_est1, fig.cap="Estimated (a) hyperparameters $\\bm{\\alpha}_{\\text{feat}}$ and (b) prior means $\\bm{\\phi}_{jd}$ from simulations with fixed $\\tau_d^2$.", out.width="100%", fig.asp=1/2>>=
-@
-
-	Model (i) gives a median $\phi$ estimate of \Sexpr{round(est.phi, 2)},  
+<<simulations_gdsc_est1, fig.cap="Estimated (a) hyperparameters $\\bm{\\alpha}_{\\text{feat}}$ and (b) prior means $\\bm{\\phi}_{jd}$ from simulation version 1 with fixed $\\tau_d^2$.", out.width="100%", fig.asp=1/2>>=
+@
+
+	Model NIG1 gives a median $\phi$ estimate of \Sexpr{round(est.phi, 2)},  
 	equal to the true mean of the $\phi_{jd}$, \Sexpr{round(true.phi, 2)}. 
 	$\lambda_{\text{feat}}$ is slightly underestimated with median estimate
-	\Sexpr{round(lambda[1], 2)} (the true value is one) and thus overestimates the 
+	\Sexpr{round(est.lambdaf[1], 2)} (the true value is one) and thus 
+	overestimates the 
 	prior $\gamma_{jd}^2$ variance. This is likely to compensate for the common
@@
 
-	The median evidence 
-	lower bounds (ELBO), which approximate the marginal likelihoods of models
-	(i)-(iv), are estimated as \Sexpr{round(melbo, 2)[1:3]}, and 
-	\Sexpr{round(melbo, 2)[4]}, respectively. This implies that model (i),
-	with estimated $\bm{\alpha}_{\text{feat}}$ and $\lambda_{\text{feat}}$ fits
-	the data best (the log Bayes factors comparing models (ii)-(iv) to model (i) 
-	are \Sexpr{round(melbo[-c(1, 4)] - melbo[1], 2)}, and 
-	\Sexpr{round(melbo[4] - melbo[1], 2)}, respectively, and thus highly favouring
-	model (i)). The prediction MSEs for the four models on the test 
-	data and gaps between training and test prediction MSE are displayed in Table
-	\ref{tab:mse1}. Table \ref{tab:mse1} confirms that the models that
-	include the external data (models (i) and (ii)) learn the underlying structure 
-	in the data better than models (iii) and (iv) that do not include the external 
+	Table \ref{tab:simulations_gdsc_tab1} displays several median performance
+	measures for simulation version 1. The table includes as a reference 
+	the null model, which
+	is an empty model that only fits the mean of the $\y_{id}$.
+	The median evidence lower bounds (ELBO) approximates the marginal
+	likelihoods of models NIG1 and NIG2 on both training and test data, while 
+	estimation MSE (EMSE) and prediction MSE (PMSE) give measures of predictive 
+	performance. Focussing on MSE, we see that the NIG2 model, that includes
+	the external covariates, performs best, both in terms of estimation and 
+	prediction. Model fit is best for the NIG1 model that does not include the
+	external covariates, according to the ELBO on the training data.
+	The lower MSEs confirm that the model that
+	includes the external data (NIG2) learns the underlying structure 
+	in the data better than model NIG1 that does not include the external 
 	data.
-<<mse1>>= 
+<<simulations_gdsc_tab1>>=
 library(knitr, quietly=TRUE, verbose=FALSE)
 suppressWarnings(library(kableExtra))
-res <- read.table("results/simulations_gdsc_res1.txt", row.names=NULL)
+options(knitr.kable.NA="", digits=3)
+kable(tab, col.names=colnames(tab), align="r",
+      caption="Median performance measure estimates for simulation version 1, with the best performing model in bold.", 
+      format="latex", booktabs=TRUE, escape=FALSE) %>%
+  kable_styling(latex_options=c("HOLD_position"))
+@
+
+<<simulation_gdsc_res2>>=
+res <- read.table("results/simulations_gdsc_res2.txt", row.names=NULL)
 temp <- res[, 1]
@@
 rownames(res) <- temp
-memse <- apply(res[rownames(res)=="emse", ], 2, median)
-mpmse <- apply(res[rownames(res)=="pmse", ], 2, median)
-mpmset <- apply(res[rownames(res)=="pmset", ], 2, median)
-tab <- rbind(memse, mpmse, mpmse - mpmset)
-rownames(tab) <- c("$\\text{EMSE}_{\\text{test}}$",
+true.chi <- mean(1/(c(1, 1, 3, 7) %*% t(unname(model.matrix(~ factor(1:4))))))
+est.chi <- median(1/res[rownames(res)=="alphad0", 1])
+est.lambdad <- apply(res[rownames(res)=="lambdad", c(1, 2)], 2, median)
+tab <- cbind(apply(res[rownames(res)=="elbo", ], 2, median),
+             apply(res[rownames(res)=="elbot", ], 2, median) -
+               apply(res[rownames(res)=="elbo", ], 2, median),
+             apply(res[rownames(res)=="emse", ], 2, median),
+             apply(res[rownames(res)=="pmse", ], 2, median),
+             apply(res[rownames(res)=="pmse", ], 2, median) - 
+               apply(res[rownames(res)=="pmset", ], 2, median))
+tab <- round(tab, 3)
+tab[which(tab[, 1]==max(tab[, 1], na.rm=TRUE)), 1] <- 
+  paste0("\\textbf{", tab[which(tab[, 1]==max(tab[, 1], na.rm=TRUE)), 1], "}")
+tab[which(tab[, 2]==min(tab[, 2], na.rm=TRUE)), 2] <- 
+  paste0("\\textbf{", tab[which(tab[, 2]==min(tab[, 2], na.rm=TRUE)), 2], "}")
+tab[which(tab[, 3]==min(tab[, 3], na.rm=TRUE)), 3] <- 
+  paste0("\\textbf{", tab[which(tab[, 3]==min(tab[, 3], na.rm=TRUE)), 3], "}")
+tab[which(tab[, 4]==min(tab[, 4], na.rm=TRUE)), 4] <- 
+  paste0("\\textbf{", tab[which(tab[, 4]==min(tab[, 4], na.rm=TRUE)), 4], "}")
+tab[which(tab[, 5]==min(tab[, 5], na.rm=TRUE)), 5] <- 
+  paste0("\\textbf{", tab[which(tab[, 5]==min(tab[, 5], na.rm=TRUE)), 5], "}")
+colnames(tab) <- c("$\\text{ELBO}_{\\text{test}}$",
+                   "$\\text{ELBO}_{\\text{train}} 
+                   - \\text{ELBO}_{\\text{test}}$",
+                   "$\\text{EMSE}_{\\text{test}}$",
                    "$\\text{PMSE}_{\\text{test}}$",
-                   "$\\text{PMSE}_{\\text{test}} - \\text{PMSE}_{\\text{train}}$")
+                   "$\\text{PMSE}_{\\text{test}} 
+                   - \\text{PMSE}_{\\text{train}}$")     
+@	
+  Figure \ref{fig:simulations_gdsc_est2} shows the estimated
+	$\bm{\alpha}_{\text{drug}}$ together with its true value for NIG2 in 
+	version 2 of the simulation study. They are estimated without bias.
+	Median $\lambda_{\text{drug}}=\Sexpr{round(est.lambdad[2], 2)}$ is slightly 
+	overestimated compared to true value 1. 
+	The median $\chi$ estimate in the NIG1 model is \Sexpr{round(est.chi, 2)},  
+	which is almost equal to the true mean \Sexpr{round(true.chi, 2)}. 
+	$\lambda_{\text{drug}}$ is underestimated with median estimate
+	\Sexpr{round(est.lambdad[1], 2)} compared to the true value.
+	Median performance measures for simulation version 2 are displayed in Table
+	\ref{tab:simulations_gdsc_tab2}. Again, we see that NIG2 performs best in 
+	terms of MSE. However, NIG1 performance is the same here, indicating that 
+	the external covariates do not add much here.
+	Model fit is again best for the NIG1 model.
+<<simulations_gdsc_est2, fig.cap="Estimated (a) hyperparameters $\\bm{\\alpha}_{\\text{feat}}$ and (b) prior means $\\bm{\\phi}_{jd}$ from simulation version 2 with fixed $\\gamma_{jd}^2$.", out.width="100%", fig.asp=1/2>>=
+@
+
+<<simulations_gdsc_tab2>>=
+library(knitr, quietly=TRUE, verbose=FALSE)
+suppressWarnings(library(kableExtra))
 options(knitr.kable.NA="", digits=3)
-kable(tab,
-      col.names=colnames(tab),
-      caption="Median estimation MSE, prediction MSEs, and difference between 
-      prediction MSEs on the test and training data in the fixed 
-      $\\tau_d^2$ simulation setting.", 
+kable(tab, col.names=colnames(tab), align="r",
+      caption="Median performance measure estimates for simulation version 2, with the best performing model in bold.", 
       format="latex", booktabs=TRUE, escape=FALSE) %>%
@@
 
-	To investigate the interplay of the feature and drug variance components, we
-	repeat the whole simulation study with random $\tau_d^2$. Similarly as with
-	the $\gamma_{jd}^2$, we randomly create four groups of drugs and corresponding
-	external drug dummy covariates. We simulate the $\tau_d^2$ as in
-	(\ref{eq:tauprior}), where we set $\bm{\alpha}_{\text{drug}} =
-	\begin{bmatrix} 1 & 1 & 3 & 7 \end{bmatrix} \tr$, such
-	that the $\tau_{d}^2$ of the four groups of features have prior means
-	$\chi_{d} \in \{ 1, 1/2, 1/4, 1/8 \}$. The scale $\lambda_{\text{drug}}$ is
-	set to one. We estimate the same four models as in the fixed $\tau_d^2$
-	setting, but in addition estimate the $\tau_d^2$ priors as well. Again, we
-	use two different versions: one where we estimate
-	$\bm{\lambda}$ and one where we fix it to its true value.
-
-	Figures \ref{fig:simulations_gdsc_est2} displays the estimated $\bm{\alpha}$,
-	$\phi_{jd}$, and $\chi_d$ for models (i) and (ii). The scale parameters were
-	estimated as $\lambda_{\text{feat}}=\Sexpr{round(rnorm(1), 2)}$ and
-	$\lambda_{\text{drug}}=\Sexpr{round(rnorm(1)[1], 2)}$ in model (i). From
-	Figure \ref{fig:simulations_gdsc_est2} we see that estimation of
-	$\bm{\alpha}_{\text{feat}}$ is again biased. However, from
-	Figure \ref{fig:simulations_gdsc_phi2} we see that the scale and ordering of
-	the $\phi_{jd}$ are correclty estimated. We conjecture that the drop in
-	estimation accuracy compared to the previous setting stems from the
-	difficult estimation of the $\tau_d^2$ prior. Estimation of this prior
-	depends on only 168 observed drugs, while the $\gamma_{jd}^2$ prior is
-	is estimated from approximately $300 \cdot 168=50400$ features.
-	\cite{wiel_learning_2019} show in a simple regression setting that if the
-	number of samples and the number of parameters on which the EB is based are of
-	the same order, empirical Bayes performs worse than when one of the two is
-	large compared to the other.
-	Here, the number of samples $n=241$ and $D=168$ are of the same order, leading
-	to a suboptimal EB estimation performance.
-<<simulations_gdsc_est2, fig.cap="Estimated (a) hyperparameters $\\bm{\\alpha}_{\\text{feat}}$ and (b) prior means $\\bm{\\phi}_{jd}$ from simulations with fixed $\\tau_d^2$.", out.width="100%", fig.asp=1/2>>=
-@
-
-  We argue however, that even though the prior parameters are not estimated
-  correctly, there is still some gain in predictive performance.
-  Prediction MSEs on the test data the training-test gaps in prediction MSE are
-  shown in Table \ref{tab:pmse2}. The comparisons of models (i) and
-	(ii) vesus models (iii) and (iv) confirm that even though EB estimation is
-	suboptimal, the external data do add something in predictive performance.
+<<simulation_gdsc_res3>>=
+res <- read.table("results/simulations_gdsc_res3.txt", row.names=NULL)
+temp <- res[, 1]
+res <- as.matrix(res[, -1])
+rownames(res) <- temp
+true.chi <- mean(1/(c(1, 1, 3, 7) %*% t(unname(model.matrix(~ factor(1:4))))))
+est.chi <- median(1/res[rownames(res)=="alphad0", 1])
+est.lambdad <- apply(res[rownames(res)=="lambdad", c(1, 2)], 2, median)
+true.phi <- mean(1/(c(1, 1, 3, 7) %*% t(unname(model.matrix(~ factor(1:4))))))
+est.phi <- median(1/res[rownames(res)=="alphaf0", 1])
+est.lambdaf <- apply(res[rownames(res)=="lambdaf", c(1, 2)], 2, median)
+
+phi.ratio <- apply(apply(sapply(0:3, function(s) {
+  res[rownames(res)==paste0("alphaf" ,s), colnames(res)=="NIG2"]}) %*% 
+    t(cbind(1, rbind(0, diag(3)))), 2, function(m) {
+      res[rownames(res)=="alphaf0", colnames(res)=="NIG2"]/m}), 2, median)
+chi.ratio <- apply(apply(sapply(0:3, function(s) {
+  res[rownames(res)==paste0("alphad" ,s), colnames(res)=="NIG2"]}) %*% 
+    t(cbind(1, rbind(0, diag(3)))), 2, function(m) {
+      res[rownames(res)=="alphad0", colnames(res)=="NIG2"]/m}), 2, median)
+true.ratio <- as.numeric(1/(c(1, 1, 3, 7) %*% 
+                              t(unname(model.matrix(~ factor(1:4))))))
+
+tab <- cbind(apply(res[rownames(res)=="elbo", ], 2, median),
+             apply(res[rownames(res)=="elbot", ], 2, median) -
+               apply(res[rownames(res)=="elbo", ], 2, median),
+             apply(res[rownames(res)=="emse", ], 2, median),
+             apply(res[rownames(res)=="pmse", ], 2, median),
+             apply(res[rownames(res)=="pmse", ], 2, median) - 
+               apply(res[rownames(res)=="pmset", ], 2, median))
+tab <- round(tab, 3)
+tab[which(tab[, 1]==max(tab[, 1], na.rm=TRUE)), 1] <- 
+  paste0("\\textbf{", tab[which(tab[, 1]==max(tab[, 1], na.rm=TRUE)), 1], "}")
+tab[which(tab[, 2]==min(tab[, 2], na.rm=TRUE)), 2] <- 
+  paste0("\\textbf{", tab[which(tab[, 2]==min(tab[, 2], na.rm=TRUE)), 2], "}")
+tab[which(tab[, 3]==min(tab[, 3], na.rm=TRUE)), 3] <- 
+  paste0("\\textbf{", tab[which(tab[, 3]==min(tab[, 3], na.rm=TRUE)), 3], "}")
+tab[which(tab[, 4]==min(tab[, 4], na.rm=TRUE)), 4] <- 
+  paste0("\\textbf{", tab[which(tab[, 4]==min(tab[, 4], na.rm=TRUE)), 4], "}")
+tab[which(tab[, 5]==min(tab[, 5], na.rm=TRUE)), 5] <- 
+  paste0("\\textbf{", tab[which(tab[, 5]==min(tab[, 5], na.rm=TRUE)), 5], "}")
+colnames(tab) <- c("$\\text{ELBO}_{\\text{test}}$",
+                   "$\\text{ELBO}_{\\text{train}} 
+                   - \\text{ELBO}_{\\text{test}}$",
+                   "$\\text{EMSE}_{\\text{test}}$",
+                   "$\\text{PMSE}_{\\text{test}}$",
+                   "$\\text{PMSE}_{\\text{test}} 
+                   - \\text{PMSE}_{\\text{train}}$")     
+@	
+  In Figure \ref{fig:simulations_gdsc_est3} the 
+	$\bm{\alpha}_{\text{feat}}$ and $\bm{\alpha}_{\text{drug}}$ estimated by NIG2
+	are displayed together with their true values for
+	simulation 3. $\bm{\alpha}_{\text{feat}}$ are underestimated, while 
+	$\bm{\alpha}_{\text{drug}}$ are overestimated. These biases seem to 
+	be consistent though. The median ratios $\phi_1$ to $\phi_2$, $\phi_3$, 
+	$\phi_4$ and $\chi_1$ to $\chi_2$, $\chi_3$, 
+	$\chi_4$ are \Sexpr{round(phi.ratio, 2)[-1]} and 
+	\Sexpr{round(chi.ratio, 2)[-1]}, respectively, compared to their true values
+	\Sexpr{round(true.ratio, 2)[-1]}. So in a relative sense, the estimates are
+	correct. Median $\lambda_{\text{feat}}$ and $\lambda_{\text{drug}}$ are
+	are over- and underestimated with \Sexpr{round(est.lambdaf[2], 2)}
+	and \Sexpr{round(est.lambdad[2], 2)}, respectively, compared to their true
+	value 1. The NIG1 model is also consistently over- and underestimating $\phi$
+	and $\chi$ with estimates \Sexpr{round(est.phi, 2)} and 
+	\Sexpr{round(est.chi, 2)}, respectively (compared to the true mean 
+	\Sexpr{round(true.chi, 2)}). NIG1 also over- and underestimates 
+	$\lambda_{\text{feat}}$ and $\lambda_{\text{drug}}$ with median estimates
+	\Sexpr{round(est.lambdaf[1], 2)}
+	and \Sexpr{round(est.lambdad[1], 2)}, respectively. Table 
+	\ref{tab:simulations_gdsc_tab3}, with median performance measures, shows
+	that NIG2 outperforms the other models in terms of prediction and estimation
+	MSE.
+<<simulations_gdsc_est3, fig.cap="Estimated (a) hyperparameters $\\bm{\\alpha}_{\\text{feat}}$ and (b) prior means $\\bm{\\phi}_{jd}$ from simulation version 3.", out.width="100%", fig.asp=1>>= 
+@
+
+<<simulations_gdsc_tab3>>=
+library(knitr, quietly=TRUE, verbose=FALSE)
+suppressWarnings(library(kableExtra))
+options(knitr.kable.NA="", digits=3)
+kable(tab, col.names=colnames(tab), align="r",
+      caption="Median performance measure estimates for simulation version 3, with the best performing model in bold.", 
+      format="latex", booktabs=TRUE, escape=FALSE) %>%
+  kable_styling(latex_options=c("HOLD_position"))
+@
 
@@
 	approach such as in 
-	{\cite{mai_composite_2019} uses all the available molecular markers 
+	{\cite{mai_composite_2019} uses most of the available molecular markers 
 	instead of just gene expression (68 mutations, 426 copy numbers, 
@@
 	due to the repeated calculation of posterior parameters in the empirical
-	Bayes iterations. Currently, problems of up to 1000 outcomes, 1000 features,
-	and 1000 samples are computationally feasible.
+	Bayes iterations. Currently, problems with number of outcomes in the hundreds
+	and number of features and samples in the thousands are computationally 
+	feasible.
 
@@
 	of missings until no more missing values remain, we end up with IC50 
-	estimates for 482 cell lines on 168 drugs. As a preliminary step, we fit
-	separate elastic net models for each drug with the \texttt{glmnet} package
-	\cite[]{friedman_regularization_2010}, with $\alpha=0.01$ and $\lambda$
-	chosen such that we select approximately 300 genes per drug.
+	estimates for 482 cell lines on 168 drugs.
 
@@
 	\subsection{Results}
+	As a preliminary step, we select
+	features according to three strategies. The first strategy selects for each
+	drug separately, the 50 features that are not in the pathway of the drug with
+	the largest variance and the 50 features that are in the pathway of the drug
+	with the largest variance. If less than 50 features that are in the pathway 
+	are available, we select the maximum number available. This strategy ensures
+	that the features are somewhat balanced with respect to the external feature
+	covariate, i.e., pathway membership. The second strategy selects the 
+	100 features with the largest variance, irrespective of pathway membership,
+	while the last stategy selects 1000 features with the largest variance,
+	again irrespective of pathway membership. These three strategies lead to
+	three different analysis outcomes.
+
 	To compute performance measures, we randomly split the data into training and
@@
 	The estimated 
-	models are: two models like in (\ref{eq:prior}) but with different external 
-	variables, named (i) NIG1 and (ii) NIG2 in the following, (iii) a model 
-	as described in \cite{kpogbezan_empirical_2017}, named bSEM, (iv) a regular
-	ridge model, and (v) a regular lasso model. The NIG1 incorporates no external
-	variables and estimates the common prior $\gamma_{jd}^2$ and $\tau_d^2$ 
-	means $\phi$ and $\chi$ directly, while the NIG2 model includes the
-	developmental phase and action drug covariates, and the pathway
-	feature covariate, and estimates $\bm{\alpha}$. The bSEM model is similar to 
-	the NIG model in that it 
-	draws the $\beta_{jd}$ from a conditionally normal distribution, but 
-	fixes $\tau_d^2=1$ and draws 
-	$\gamma_{jd}^2 \sim \Gamma^{-1} (\phi_{jd}, \lambda_{jd})$, instead of 
-	from the inverse Gaussian distribution. Additionally, 
-	the bSEM model can 
-	only incorporate one binary external covariate that divides the features into 
-	two groups, i.e., $\bm{\alpha} = \begin{bmatrix} \alpha_0 & \alpha_1 
-	\end{bmatrix} \tr$ and $\bm{c}_{jd} = \begin{bmatrix} 1 & 0 \end{bmatrix} 
-	\tr$ if feature $jd$ belongs to group one and $\bm{c}_{jd} = \begin{bmatrix} 
-	1 & 1 \end{bmatrix} \tr$ if feature $jd$ belongs to group two. Both 
-	$\bm{\alpha}_{\text{feat}}$ and $\lambda_{\text{feat}}$ are
-	estimated in a similar fashion as in Section \ref{sec:estimation}.
-
-  From the $\bm{\alpha}$ estimates in Tables \ref{tab:results_drug} and
-  \ref{tab:results_feat} we see that there
+	models are: four models like in (\ref{eq:prior}) but with different external 
+	variables, named NIG1, NIG2, NIG3, and NIG4 in the following. NIG1 estimates
+	only one overall $\phi$ and one overall $\chi$, NIG2 uses the external 
+	covariates to estimate both $\bm{\alpha}_{\text{feat}}$ and
+	$\bm{\alpha}_{\text{drug}}$, while NIG3 and NIG4 estimate only 
+	$\bm{\alpha}_{\text{feat}}$ or $\bm{\alpha}_{\text{drug}}$, respectively,
+	keeping either $\forall d: \tau_d^2=1$ or $\forall j,d: \gamma_{jd}^2=1$
+	fixed. In addition, we estimate regular lasso and ridge models.
+
+	% The bSEM model is similar to 
+	% the NIG model in that it 
+	% draws the $\beta_{jd}$ from a conditionally normal distribution, but 
+	% fixes $\tau_d^2=1$ and draws 
+	% $\gamma_{jd}^2 \sim \Gamma^{-1} (\phi_{jd}, \lambda_{jd})$, instead of 
+	% from the inverse Gaussian distribution. Additionally, 
+	% the bSEM model can 
+	% only incorporate one binary external covariate that divides the features into 
+	% two groups, i.e., $\bm{\alpha} = \begin{bmatrix} \alpha_0 & \alpha_1 
+	% \end{bmatrix} \tr$ and $\bm{c}_{jd} = \begin{bmatrix} 1 & 0 \end{bmatrix} 
+	% \tr$ if feature $jd$ belongs to group one and $\bm{c}_{jd} = \begin{bmatrix} 
+	% 1 & 1 \end{bmatrix} \tr$ if feature $jd$ belongs to group two. Both 
+	% $\bm{\alpha}_{\text{feat}}$ and $\lambda_{\text{feat}}$ are
+	% estimated in a similar fashion as in Section \ref{sec:estimation}.
+
+<<analysis_gdsc_fit1>>=
+fit <- read.table("results/analysis_gdsc_fit1.txt", row.names=NULL)
+temp <- fit[, 1]
+fit <- as.matrix(fit[, -1])
+rownames(fit) <- temp
+
+res <- read.table("results/analysis_gdsc_res1.txt", row.names=NULL)
+temp <- res[, 1]
+res <- as.matrix(res[, -1])
+rownames(res) <- temp
+
+est1 <- cbind(fit[, 1], c(NA, 0, 0, NA), fit[, 2])
+colnames(est1) <- c("intercept", "not in pathway", "in pathway")
+est2 <- cbind(fit[, 3:5], c(NA, 0, NA, 0), fit[, c(7:6)], c(NA, 0, NA, 0))
+colnames(est2) <- c("intercept", "experimental", "development", "approved",
+                    "unkown", "targeted", "cytotoxic")
+
+tab <- cbind(apply(res[rownames(res)=="elbo", ], 2, median),
+             apply(res[rownames(res)=="elbot", ], 2, median) -
+               apply(res[rownames(res)=="elbo", ], 2, median),
+             apply(res[rownames(res)=="pmse", ], 2, median))
+tab <- round(tab, 3)
+tab[which(tab[, 1]==max(tab[, 1], na.rm=TRUE)), 1] <- 
+  paste0("\\textbf{", tab[which(tab[, 1]==max(tab[, 1], na.rm=TRUE)), 1], "}")
+tab[which(tab[, 2]==min(tab[, 2], na.rm=TRUE)), 2] <- 
+  paste0("\\textbf{", tab[which(tab[, 2]==min(tab[, 2], na.rm=TRUE)), 2], "}")
+tab[which(tab[, 3]==min(tab[, 3], na.rm=TRUE)), 3] <- 
+  paste0("\\textbf{", tab[which(tab[, 3]==min(tab[, 3], na.rm=TRUE)), 3], "}")
+colnames(tab) <- c("$\\text{ELBO}_{\\text{test}}$",
+                   "$\\text{ELBO}_{\\text{train}} 
+                   - \\text{ELBO}_{\\text{test}}$",
+                   "$\\text{PMSE}_{\\text{test}}$")  
+@
+  From the $\bm{\alpha}$ estimates in Tables \ref{tab:analysis_gdsc_est1_feat} 
+  and \ref{tab:analysis_gdsc_est1_feat} we see that there
   seems to be an effect of the
-  external variables. If we look at the NIG1 model, that includes all external 
+  external variables. Models NIG2 and NIG3 estimate a negative effect for the
+  pathway external covariate. This translates to a negative additive effect on
+  the expected prior precision of the $\bm{\beta}_d$, i.e., genes that are found
+  in the target pathway of a drug are more predictive of IC50 than genes that
+  are not in the target pathway, according to expectation. Furthermore the NIG2
+  model estimates positive effects for the experimental and developmental drugs,
+  compared to the approved drugs. This translates to smaller expected 
+  prior precisions for the regression parameters for approved drugs, i.e.,
+  approved drugs are easier to predict than experimental and developmental 
+  drugs. On the other hand, the NIG4 model, that only includes external
+  drug covariates, estimates a negative effect for the developmental drugs.
+
+  If we look at the NIG1 model, that includes all external 
   covariates, we see that approved drugs are easier to predict, due to small
@@
   effects of the pathway indicator seem to be much larger in the bSEM model. 
-<<results_drug>>= 
-library(knitr, quietly=TRUE, verbose=FALSE)
-suppressWarnings(library(kableExtra))
-# load("results/data_gdsc_res1.Rdata")
-# tab1 <- cbind(tab[, 1], c(NA, 0, NA), tab[, c(3, 2)], c(NA, 0, NA), 
-#               tab[, c(4, 5)])
-# options(knitr.kable.NA="", digits=3)
-# kable(tab1,
-#       col.names=c("intercept", "approved", "development", 
-#                   "experimental", "targeted", "unkown", "cytotoxic"),
-#       caption="$\\bm{\\alpha}_{\\text{drug}}$ estimates from GDSC data.", 
-#       format="latex", booktabs=TRUE, escape=FALSE) %>%
-#   kable_styling(latex_options=c("HOLD_position"))
-kable(matrix(1, nrow=1, ncol=2),
-      caption="$\\bm{\\alpha}_{\\text{drug}}$ estimates from GDSC data.", 
+<<analysis_gdsc_est1_feat>>=
+options(knitr.kable.NA="", digits=3)
+kable(est1,
+      col.names=colnames(est1), align="r",
+      caption="$\\bm{\\alpha}_{\\text{feat}}$ estimates from GDSC data for feature selection strategy 1.",
       format="latex", booktabs=TRUE, escape=FALSE) %>%
@@
 
-<<results_feat>>= 
-library(knitr, quietly=TRUE, verbose=FALSE)
-suppressWarnings(library(kableExtra))
-# load("results/data_gdsc_res1.Rdata")
-# tab2 <- cbind(tab[, 6], c(NA, 0, 0), tab[, 7])
-# options(knitr.kable.NA="", digits=3)
-# kable(tab2,
-#       col.names=c("intercept", "not in pathway", "in pathway"),
-#       caption="$\\bm{\\alpha}_{\\text{feat}}$ estimates from GDSC data.", 
-#       format="latex", booktabs=TRUE, escape=FALSE) %>%
-#   kable_styling(latex_options=c("HOLD_position"))
-kable(matrix(1, nrow=1, ncol=2),
-      caption="$\\bm{\\alpha}_{\\text{feat}}$ estimates from GDSC data.", 
+<<analysis_gdsc_est1_drug>>=
+kable(est2,
+      col.names=colnames(est2), align="r",
+      caption="$\\bm{\\alpha}_{\\text{drug}}$ estimates from GDSC data for feature selection strategy 1.",
       format="latex", booktabs=TRUE, escape=FALSE) %>%
@@
   \Sexpr{round(rnorm(1))}.
-<<results_data>>= 
-library(knitr, quietly=TRUE, verbose=FALSE)
-suppressWarnings(library(kableExtra))
-# load("results/data_gdsc_res1.Rdata")
-# mapmse <- apply(apmse, 2, median)
-# melbo <- apply(elbo, 2, median)
-# tab <- rbind(mapmse, c(melbo, NA, NA))
-# rownames(tab) <- c("MSE", "ELBO")
-# options(knitr.kable.NA="", digits=3)
-# kable(tab,
-#       col.names=c("NIG1", "NIG2", "bSEM", "lasso", "ridge"),
-#       caption="Median prediction MSE and median ELBO estimated on the test data", 
-#       format="latex", booktabs=TRUE, escape=FALSE) %>%
-#   kable_styling(latex_options=c("HOLD_position"))
-options(knitr.kable.NA="", digits=3)
-kable(matrix(1, nrow=1, ncol=2),
-      caption="Median prediction MSE and median ELBO estimated on the test data", 
+<<analysis_gdsc_tab1>>=
+kable(tab,
+      col.names=colnames(tab), align="r",
+      caption="Performance measures from GDSC data for feature selection strategy 1.",
       format="latex", booktabs=TRUE, escape=FALSE) %>%
@@
   correlation over the 50 data splits. They are depicted in Figure 
-  \ref{fig:data_gdsc_res1_brank_hist}. In both NIG models the correlations
+  . In both NIG models the correlations
   between splits are higher than in the bSEM model; an indication that 
   estimation in the NIG model is more stable than in the bSEM model.
-<<data_gdsc_res1_brank_hist, fig.cap="Spearman's rank correlation between the estimated posterior means $\\E_{Q}(\\bbeta_d)$ in the 50 data splits.", out.width="70%", fig.asp=2/3>>= 
-@
-
+
+<<analysis_gdsc_fit2>>=
+fit <- read.table("results/analysis_gdsc_fit2.txt", row.names=NULL)
+temp <- fit[, 1]
+fit <- as.matrix(fit[, -1])
+rownames(fit) <- temp
+
+res <- read.table("results/analysis_gdsc_res2.txt", row.names=NULL)
+temp <- res[, 1]
+res <- as.matrix(res[, -1])
+rownames(res) <- temp
+
+est1 <- cbind(fit[, 1], c(NA, 0, 0, NA), fit[, 2])
+colnames(est1) <- c("intercept", "not in pathway", "in pathway")
+est2 <- cbind(fit[, 3:5], c(NA, 0, NA, 0), fit[, c(7:6)], c(NA, 0, NA, 0))
+colnames(est2) <- c("intercept", "experimental", "development", "approved",
+                    "unkown", "targeted", "cytotoxic")
+
+tab <- cbind(apply(res[rownames(res)=="elbo", ], 2, median),
+             apply(res[rownames(res)=="elbot", ], 2, median) -
+               apply(res[rownames(res)=="elbo", ], 2, median),
+             apply(res[rownames(res)=="pmse", ], 2, median))
+tab <- round(tab, 3)
+tab[which(tab[, 1]==max(tab[, 1], na.rm=TRUE)), 1] <- 
+  paste0("\\textbf{", tab[which(tab[, 1]==max(tab[, 1], na.rm=TRUE)), 1], "}")
+tab[which(tab[, 2]==min(tab[, 2], na.rm=TRUE)), 2] <- 
+  paste0("\\textbf{", tab[which(tab[, 2]==min(tab[, 2], na.rm=TRUE)), 2], "}")
+tab[which(tab[, 3]==min(tab[, 3], na.rm=TRUE)), 3] <- 
+  paste0("\\textbf{", tab[which(tab[, 3]==min(tab[, 3], na.rm=TRUE)), 3], "}")
+colnames(tab) <- c("$\\text{ELBO}_{\\text{test}}$",
+                   "$\\text{ELBO}_{\\text{train}} 
+                   - \\text{ELBO}_{\\text{test}}$",
+                   "$\\text{PMSE}_{\\text{test}}$")  
+@
+
+<<analysis_gdsc_est2_feat>>=
+options(knitr.kable.NA="", digits=3)
+kable(est1,
+      col.names=colnames(est1), align="r",
+      caption="$\\bm{\\alpha}_{\\text{feat}}$ estimates from GDSC data for feature selection strategy 2.",
+      format="latex", booktabs=TRUE, escape=FALSE) %>%
+  kable_styling(latex_options=c("HOLD_position"))
+@
+
+<<analysis_gdsc_est2_drug>>=
+kable(est2,
+      col.names=colnames(est2), align="r",
+      caption="$\\bm{\\alpha}_{\\text{drug}}$ estimates from GDSC data for feature selection strategy 2.",
+      format="latex", booktabs=TRUE, escape=FALSE) %>%
+  kable_styling(latex_options=c("HOLD_position"))
+@
+
+<<analysis_gdsc_tab2>>=
+kable(tab,
+      col.names=colnames(tab), align="r",
+      caption="Performance measures from GDSC data for feature selection strategy 2.",
+      format="latex", booktabs=TRUE, escape=FALSE) %>%
+  kable_styling(latex_options=c("HOLD_position"))
+@
+
+res1 <- 
 	\section{Discussion}\label{sec:discussion}

### 2019-11-25 docs/manuscript1.Rnw
added bSEM to analysis_gdsc.R

@@
 	models are fit on 241 randomly selected cell lines, while performance measures 
-	are estimated from the remaining 241 cell lines.
+	are estimated from the remaining 241 cell lines. We repeat the whole
+	simulation 50 times.
 
@@
 	the null model, which
-	is an empty model that only fits the mean of the $\y_{id}$.
+	is an empty model that only fits the means of the $\y_{d}$.
 	The median evidence lower bounds (ELBO) approximates the marginal
@@
 	prediction. Model fit is best for the NIG1 model that does not include the
-	external covariates, according to the ELBO on the training data.
+	external covariates, according to the ELBO, calculated on the test data.
 	The lower MSEs confirm that the model that
@@
 	they have larger prior $\beta_{jd}$ variances than drugs that are not in 
-	the pathway.
+	the pathway. Note that large $\beta_{jd}$ variances translate to large
+	prior $\gamma_{jd}^2$ and $\tau_d^2$ means.
 
@@
 	features according to three strategies. The first strategy selects for each
-	drug separately, the 50 features that are not in the pathway of the drug with
-	the largest variance and the 50 features that are in the pathway of the drug
-	with the largest variance. If less than 50 features that are in the pathway 
-	are available, we select the maximum number available. This strategy ensures
+	drug separately 50 features that are not in the pathway of the drug with
+	maximum variance and 50 features that are in the pathway of the drug
+	with maximum variance. If less than 50 features in the pathway are available,
+	we select all available features. This strategy ensures
 	that the features are somewhat balanced with respect to the external feature
@@
 	again irrespective of pathway membership. These three strategies lead to
-	three different analysis outcomes.
+	three different analyses described below.
 
@@
 	models are: four models like in (\ref{eq:prior}) but with different external 
-	variables, named NIG1, NIG2, NIG3, and NIG4 in the following. NIG1 estimates
-	only one overall $\phi$ and one overall $\chi$, NIG2 uses the external 
+	variables, named NIG1, NIG2, NIG3, and NIG4 in the following. NIG1 does 
+	not use the external covariates and estimates
+	one overall $\phi$ and one overall $\chi$, NIG2 uses the external 
 	covariates to estimate both $\bm{\alpha}_{\text{feat}}$ and
@@
 	keeping either $\forall d: \tau_d^2=1$ or $\forall j,d: \gamma_{jd}^2=1$
-	fixed. In addition, we estimate regular lasso and ridge models.
+	fixed. In addition, we estimate regular lasso and ridge models. The null model
+	that fits the means of the $\y_d$ is included as a reference.
 
@@
 
-res1 <- 
+<<analysis_gdsc_fit3>>=
+fit <- read.table("results/analysis_gdsc_fit3.txt", row.names=NULL)
+temp <- fit[, 1]
+fit <- as.matrix(fit[, -1])
+rownames(fit) <- temp
+
+res <- read.table("results/analysis_gdsc_res3.txt", row.names=NULL)
+temp <- res[, 1]
+res <- as.matrix(res[, -1])
+rownames(res) <- temp
+
+est1 <- cbind(fit[, 1], c(NA, 0, 0, NA), fit[, 2])
+colnames(est1) <- c("intercept", "not in pathway", "in pathway")
+est2 <- cbind(fit[, 3:5], c(NA, 0, NA, 0), fit[, c(7:6)], c(NA, 0, NA, 0))
+colnames(est2) <- c("intercept", "experimental", "development", "approved",
+                    "unkown", "targeted", "cytotoxic")
+
+tab <- cbind(apply(res[rownames(res)=="elbo", ], 2, median),
+             apply(res[rownames(res)=="elbot", ], 2, median) -
+               apply(res[rownames(res)=="elbo", ], 2, median),
+             apply(res[rownames(res)=="pmse", ], 2, median))
+tab <- round(tab, 3)
+tab[which(tab[, 1]==max(tab[, 1], na.rm=TRUE)), 1] <- 
+  paste0("\\textbf{", tab[which(tab[, 1]==max(tab[, 1], na.rm=TRUE)), 1], "}")
+tab[which(tab[, 2]==min(tab[, 2], na.rm=TRUE)), 2] <- 
+  paste0("\\textbf{", tab[which(tab[, 2]==min(tab[, 2], na.rm=TRUE)), 2], "}")
+tab[which(tab[, 3]==min(tab[, 3], na.rm=TRUE)), 3] <- 
+  paste0("\\textbf{", tab[which(tab[, 3]==min(tab[, 3], na.rm=TRUE)), 3], "}")
+colnames(tab) <- c("$\\text{ELBO}_{\\text{test}}$",
+                   "$\\text{ELBO}_{\\text{train}} 
+                   - \\text{ELBO}_{\\text{test}}$",
+                   "$\\text{PMSE}_{\\text{test}}$")  
+@
+
+<<analysis_gdsc_est3_feat>>=
+options(knitr.kable.NA="", digits=3)
+kable(est1,
+      col.names=colnames(est1), align="r",
+      caption="$\\bm{\\alpha}_{\\text{feat}}$ estimates from GDSC data for feature selection strategy 3.",
+      format="latex", booktabs=TRUE, escape=FALSE) %>%
+  kable_styling(latex_options=c("HOLD_position"))
+@
+
+<<analysis_gdsc_est3_drug>>=
+kable(est2,
+      col.names=colnames(est2), align="r",
+      caption="$\\bm{\\alpha}_{\\text{drug}}$ estimates from GDSC data for feature selection strategy 3.",
+      format="latex", booktabs=TRUE, escape=FALSE) %>%
+  kable_styling(latex_options=c("HOLD_position"))
+@
+
+<<analysis_gdsc_tab3>>=
+kable(tab,
+      col.names=colnames(tab), align="r",
+      caption="Performance measures from GDSC data for feature selection strategy 3.",
+      format="latex", booktabs=TRUE, escape=FALSE) %>%
+  kable_styling(latex_options=c("HOLD_position"))
+@
+
 	\section{Discussion}\label{sec:discussion}

### 2019-11-25 docs/manuscript1.Rnw
changed manuscript1.Rnw

@@
 
-est1 <- cbind(fit[, 1], c(NA, 0, 0, NA), fit[, 2])
+est1 <- cbind(fit[, 1], c(NA, 0, 0, NA, 0), fit[, 2])
 colnames(est1) <- c("intercept", "not in pathway", "in pathway")
-est2 <- cbind(fit[, 3:5], c(NA, 0, NA, 0), fit[, c(7:6)], c(NA, 0, NA, 0))
+est2 <- cbind(fit[, 3:5], c(NA, 0, NA, 0, NA), fit[, c(7:6)], 
+              c(NA, 0, NA, 0, NA))
 colnames(est2) <- c("intercept", "experimental", "development", "approved",

### 2019-11-25 docs/manuscript1.Rnw
commit before analysis_gdsc.R

@@
 
-	\section{Simulations}
+	\section{Simulations}\ref{sec:simulations}
 	\subsection{Setup}
@@
 	keeping either $\forall d: \tau_d^2=1$ or $\forall j,d: \gamma_{jd}^2=1$
-	fixed. In addition, we estimate regular lasso and ridge models. The null model
-	that fits the means of the $\y_d$ is included as a reference.
-
-	% The bSEM model is similar to 
-	% the NIG model in that it 
-	% draws the $\beta_{jd}$ from a conditionally normal distribution, but 
-	% fixes $\tau_d^2=1$ and draws 
-	% $\gamma_{jd}^2 \sim \Gamma^{-1} (\phi_{jd}, \lambda_{jd})$, instead of 
-	% from the inverse Gaussian distribution. Additionally, 
-	% the bSEM model can 
-	% only incorporate one binary external covariate that divides the features into 
-	% two groups, i.e., $\bm{\alpha} = \begin{bmatrix} \alpha_0 & \alpha_1 
-	% \end{bmatrix} \tr$ and $\bm{c}_{jd} = \begin{bmatrix} 1 & 0 \end{bmatrix} 
-	% \tr$ if feature $jd$ belongs to group one and $\bm{c}_{jd} = \begin{bmatrix} 
-	% 1 & 1 \end{bmatrix} \tr$ if feature $jd$ belongs to group two. Both 
-	% $\bm{\alpha}_{\text{feat}}$ and $\lambda_{\text{feat}}$ are
-	% estimated in a similar fashion as in Section \ref{sec:estimation}.
+	fixed.
+
+	In addition, we estimate regular lasso and ridge models, and a
+	bSEM model \cite[]{leday_gene_2017,kpogbezan_empirical_2017}. The bSEM model 
+	is similar to the NIG model in that it
+	draws the $\beta_{jd}$ from a conditionally normal distribution, but
+	fixes $\tau_d^2=1$ and draws
+	$\gamma_{jd}^2 \sim \Gamma^{-1} (\phi_{jd}, \lambda_{jd})$, instead of
+	from the inverse Gaussian distribution. Additionally,
+	the bSEM model can
+	only incorporate one binary external covariate that divides the features into
+	two groups, i.e., $\bm{\alpha} = \begin{bmatrix} \alpha_0 & \alpha_1
+	\end{bmatrix} \tr$ and $\bm{c}_{jd} = \begin{bmatrix} 1 & 0 \end{bmatrix}
+	\tr$ if feature $jd$ belongs to group one and $\bm{c}_{jd} = \begin{bmatrix}
+	1 & 1 \end{bmatrix} \tr$ if feature $jd$ belongs to group two. Both
+	$\bm{\alpha}_{\text{feat}}$ and $\lambda_{\text{feat}}$ are
+	estimated in a similar fashion as in Section \ref{sec:estimation}.
+	The null model that fits the means of the $\y_d$ is included as a reference.
 
@@
 
-	The model is put into practice on the GDSC data. The results in Section
+	Simulations in Section \ref{sec:simulations} show that the method is 
+	reasonably able to recover hyperparameters. In addition, the simulations
+	show that inclusion of informative external covariates is indeed beneficial to
+	predictive performance. The model is put into practice on the GDSC data. 
+	The results in Section
 	\ref{sec:gdsc} show that the inclusion of the external covariates is 
@@
 	the inclusion of the external covariates. This is a topic that deserves 
-	further attention. One proposed solution is to include an extra 
+	further attention. One proposed solution is to include a
 	prior variance component that promotes a more heavy-tailed $\beta_{jd}$ prior

### 2019-11-25 code/mcmc_functions.R

@@ -1,74 +1,88 @@
-# sample one beta in the conjugate case
-.sample.beta.conj <- function(gamma, sigma, v, sv, svvt, y, n, p) {
-  uvec <- rnorm(p, 0, gamma*sigma)
+# sample beta
+.sample.beta <- function(tau, gamma, sigma, v, sv, svvt, y, n, p) {
+  uvec <- rnorm(p, 0, tau*gamma*sigma)
   dvec <- rnorm(n, 0, 1)
   beta <- uvec + as.numeric(t(t(v)*(sv/(sv^2 + 1/gamma^2))) %*% 
                               (sigma*y - sigma*svvt %*% uvec + dvec))
   return(beta)
 }
 
-# sample one gamma in the conjugate case
-.sample.gamma.conj <- function(beta, sigma, eta, lambda, theta, p) {
-  
-  gamma <- sqrt(rgig(1, sum(beta^2)/sigma^2 + lambda, 
-                     lambda/theta^2, -0.5*(p + eta)))
+# sample gamma
+.sample.gamma <- function(beta, sigma, tau, lambdaf, phi, p) {
+  gamma <- sqrt(rgig(p, beta^2/(sigma^2*tau^2) + lambdaf, lambdaf/phi^2, -1))
   return(gamma)
 }
 
-# sample one sigma in the conjugate case
-.sample.sigma.conj <- function(beta, gamma, yty, ytx, x, p, n) {
-  alpha <- 0.5*(p + n + 1)
-  beta <- 0.5*(yty + sum(beta^2)/gamma^2 + sum((x %*% beta)^2) - 2*ytx %*% beta)
-  gamma <- 1/sqrt(rgamma(1, alpha, scale=beta))
-  return(gamma)
+# sample tau
+.sample.tau <- function(beta, sigma, gamma, lambdad, chi, p) {
+  tau <- sqrt(rgig(1, sum(beta^2/gamma^2)/sigma^2 + lambdad, lambdad/chi^2, 
+                   -0.5*(p + 1)))
+  return(tau)
+}
+
+# sample sigma
+.sample.sigma <- function(beta, gamma, tau, yty, ytx, x, p, n) {
+  a <- 0.5*(p + n + 1)
+  b <- 0.5*(yty - 2*as.numeric(ytx %*% beta) + sum((x %*% beta)^2) + 
+              sum(beta^2/gamma^2)/tau^2)
+  sigma <- 1/sqrt(rgamma(1, a, rate=b))
[…]

### 2019-11-25 code/tryout.R

@@ -1,20 +1,0 @@
-n <- 100
-p <- 150
-D <- 100
-x <- replicate(D, matrix(rnorm(n*p), nrow=n, ncol=p), simplify=FALSE)
-beta <- replicate(D, rnorm(p), simplify=FALSE)
-y <- sapply(1:D, function(d) {x[[d]] %*% beta[[d]]})
-Z <- NULL
-# Z <- cbind(1, rbinom(D, 1, 0.5))
-C <- lapply(1:D, function(d) {cbind(1, rbinom(p, 1, 0.5))})
-unpenalized=NULL
-standardize=FALSE
-intercept=FALSE
-fixed.eb="none"
-full.post=FALSE
-init=NULL
-control=list(conv.post=TRUE, trace=TRUE,
-             epsilon.eb=1e-3, epsilon.vb=1e-3,
-             maxit.eb=10, maxit.vb=2, maxit.post=100)
-library(Rcpp)
-sourceCpp("rpackage/src/aux_functions.cpp")

### 2019-11-27 docs/biometrical_main.Rnw
added EMSE split for large and small betas to simulations_gdsc.R

@@
 \documentclass[bimj,fleqn]{w-art}
-\usepackage{bm,amsmath,amssymb,amsthm,amsfonts,graphics,epsfig,
-rotating,appendix,titlesec,multicol,hyperref,verbatim,
+\usepackage{bm,amsmath,amssymb,amsthm,amsfonts,graphics,graphicx,epsfig,
+rotating,natbib,titlesec,multicol,hyperref,verbatim,
 bbm,algorithm,algpseudocode,pgfplotstable,threeparttable,booktabs,mathtools,
 dsfont,parskip,xr,tikz,times,w-thm}
-\usepackage[]{graphicx}
-\usepackage[authoryear]{natbib}
 \externaldocument[sm-]{supplement1}
@@
 
-  \title[Drug sensitivity prediction with NIG prior]
+  \title[Drug sensitivity prediction with the NIG prior]
   {Drug sensitivity prediction with the normal inverse Gaussian prior 
@@
   several choices of $\beta_{jd}$ prior.
+
 <<dens_kappa, fig.cap="Implied prior densities $\\pi(\\kappa_{jd})$ for the (a) NIG, (b) Student's $t$, and (c) lasso priors.
 Different line types correspond to different hyperparameter settings.
-The hyperparameter settings were chosen to show some possible, distinct shapes that each of the priors can take.", out.width="90%", fig.asp=1/3>>=
+The hyperparameter settings were chosen to show some possible, distinct shapes that each of the priors can take.", out.width="100%", fig.asp=1/3>>=
 @
@@
   objective and data-driven inclusion of the feature and drug covariates.
-
 	The canonical method for empirical Bayes is to maximise the marginal 
@@
 
-	\section{Simulations}
-	\subsection{Fixed drug component}
+	\section{Simulations}\label{sec:simulations}
+	\subsection{Setup}
 	This section investigates the empirical 
@@
 	($\bm{\alpha}_{\text{feat}}=\begin{bmatrix} 1 & 1 & 3 & 7 \end{bmatrix} \tr$). 
-	The prior scale parameter is set to $\lambda_{\text{feat}}=1$. We estimate 
-	four models: (i) the NIG model as in Section \ref{sec:estimation}, (ii) an 
-	NIG model where we fix $\lambda_{\text{feat}}$ to its true value, and two NIG
-	models without the inclusion of the external covariates, again with 
-	$\lambda_{\text{feat}}$ either (iii) estimated or (iv) fixed to its true 
-	value. Exclusion of the external covariates as in models (iii) and (iv) 
+	The prior scale parameter is set to $\lambda_{\text{feat}}=1$. Version 2 
+	follows a similar procedure for the $\tau_d^2$. We randomly create four
+	groups of drugs with corresponding external drug dummy variables and set
+	$\bm{\alpha}_{\text{drug}}=\begin{bmatrix} 1 & 1 & 3 & 7 \end{bmatrix} \tr$,
+	such that we have $\chi_{d} \in \{ 1, 1/2, 1/4, 1/8 \}$. The
+	scale is set to $\lambda_{\text{drug}}=1$. Version 3 combines the procedures
+	from versions 1 and 2 to draw the $\gamma_{jd}^2$ and $\tau_d^2$.
+
+	We estimate four models: (i) an NIG model that does include external 
+	covariates in the estimation called NIG1 in the following, (ii) the NIG model
+	estimated as in Section \ref{sec:estimation} called NIG2, and regular (iii)
+	lasso and (iv) ridge models.
+	Exclusion of the external covariates as in the NIG1 model
 	amounts to direct estimation of one common expected 
-	prior mean $\phi$, instead of regression estimates for the $\phi_{jd}$ as
-	in models (i) and (ii). All 
+	prior means $\phi$ and/or $\chi$, instead of regression estimates for the
+	$\phi_{jd}$ and/or $\chi_d$ as in the NIG2 model. All 
 	models are fit on 241 randomly selected cell lines, while performance measures 
-	are estimated from the remaining 241 cell lines.
-
-<<include=FALSE>>=
-load("results/simulations_gdsc_set1.Rdata")
-mapmse <- apply(res[, c(7:12)], 2, median)
-mapmset <- apply(res[, c(13:18)], 2, median)
-melbo <- apply(elbo, 2, median)
-phi <- 1/(set$alphaf %*% t(unname(model.matrix(~ factor(1:4)))))
-mphi <- apply(1/est[, c(11, 13)], 2, median)
-mlambdaf <- apply(est[, c(5, 12)], 2, median)
+	are estimated from the remaining 241 cell lines. We repeat the whole
+	simulation 50 times.
+
+	\subsection{Results}
+<<simulation_gdsc_res1>>=
+res <- read.table("results/simulations_gdsc_res1.txt", row.names=NULL)
+temp <- res[, 1]
+res <- as.matrix(res[, -1])
+rownames(res) <- temp
+true.phi <- mean(1/(c(1, 1, 3, 7) %*% t(unname(model.matrix(~ factor(1:4))))))
+est.phi <- median(1/res[rownames(res)=="alphaf0", 1])
+est.lambdaf <- apply(res[rownames(res)=="lambdaf", c(1, 2)], 2, median)
+tab <- cbind(apply(res[rownames(res)=="elbo", ], 2, median),
+             apply(res[rownames(res)=="elbot", ], 2, median) -
+               apply(res[rownames(res)=="elbo", ], 2, median),
+             apply(res[rownames(res)=="emse", ], 2, median),
+             apply(res[rownames(res)=="pmse", ], 2, median),
+             apply(res[rownames(res)=="pmse", ], 2, median) - 
+               apply(res[rownames(res)=="pmset", ], 2, median))
+tab <- round(tab, 3)
+tab[which(tab[, 1]==max(tab[, 1], na.rm=TRUE)), 1] <- 
+  paste0("\\textbf{", tab[which(tab[, 1]==max(tab[, 1], na.rm=TRUE)), 1], "}")
+tab[which(tab[, 2]==min(tab[, 2], na.rm=TRUE)), 2] <- 
+  paste0("\\textbf{", tab[which(tab[, 2]==min(tab[, 2], na.rm=TRUE)), 2], "}")
+tab[which(tab[, 3]==min(tab[, 3], na.rm=TRUE)), 3] <- 
+  paste0("\\textbf{", tab[which(tab[, 3]==min(tab[, 3], na.rm=TRUE)), 3], "}")
+tab[which(tab[, 4]==min(tab[, 4], na.rm=TRUE)), 4] <- 
+  paste0("\\textbf{", tab[which(tab[, 4]==min(tab[, 4], na.rm=TRUE)), 4], "}")
+tab[which(tab[, 5]==min(tab[, 5], na.rm=TRUE)), 5] <- 
+  paste0("\\textbf{", tab[which(tab[, 5]==min(tab[, 5], na.rm=TRUE)), 5], "}")
+colnames(tab) <- c("$\\text{ELBO}_{\\text{test}}$",
+                   "$\\text{ELBO}_{\\text{train}} 
+                   - \\text{ELBO}_{\\text{test}}$",
+                   "$\\text{EMSE}_{\\text{test}}$",
+                   "$\\text{PMSE}_{\\text{test}}$",
+                   "$\\text{PMSE}_{\\text{test}} 
+                   - \\text{PMSE}_{\\text{train}}$")     
 @	
   Figure \ref{fig:simulations_gdsc_est1} shows the estimated
-	$\bm{\alpha}_{\text{feat}}$ together with its true value for models (i) and 
-	(ii). The figure shows that $\bm{\alpha}_{\text{feat}}$ is consistently
-	underestimated 
-	in the full NIG model (both with estimated and fixed $\lambda_{\text{feat}}$),
-	while $\lambda_{\text{feat}}$ is overestimated as 
-	\Sexpr{round(mlambdaf[1], 2)} in model (i) (the true value is 1). 
-	This translates into
-	underestimated prior means $\phi_{jd}$ as presented in Figure 
-	\ref{fig:simulations_gdsc_phi1}, which in turn implies underestimated 
-	$\beta_{jd}$ variances. The overestimated $\lambda_{\text{feat}}$ translates 
+	$\bm{\alpha}_{\text{feat}}$ together with its true value for NIG2 in 
+	version 1 of the simulation study. Figure 
+	\ref{fig:simulations_gdsc_est1}a shows that 
+	$\bm{\alpha}_{\text{feat}}$ is slightly but 
+	consistently underestimated. On the $\phi_{jd}$ scale this underestimation is
+	almost undetectable as depicted in Figure
+	\ref{fig:simulations_gdsc_est1}b. $\lambda_{\text{feat}}$ is slightly 
+	overestimated with median estimate \Sexpr{round(est.lambdaf[2], 2)} (the
+	true value is 1). The overestimated $\lambda_{\text{feat}}$ translates 
 	into underestimated $\gamma_{jd}^2$ variances. Underestimation of variances is 
 	a common phenomenon in variational Bayes approximations. In addition, 
-	$\bm{\alpha}_{\text{feat}}$ and $\lambda_{\text{feat}}$ are higher level 
-	parameters, which are, in general, hard to estimate. We note 
-	that even though $\bm{\alpha}_{\text{feat}}$ and thus $\phi_{jd}$ are 
-	estimated with bias, the orders of their elements are correct.
-<<simulations_gdsc_est1, fig.cap="Estimated hyperparameters $\\bm{\\alpha}_{\\text{feat}}$ from simulations with (a) estimated and (b) fixed $\\lambda_{\\text{feat}}$ (models (i) and (ii), respectively).", out.width="100%", fig.asp=1/2>>=
-@
-
-<<simulations_gdsc_phi1, fig.cap="Estimated prior means $\\bm{\\phi}_{jd}$ for the four groups of features from simulations with (a) estimated and (b) fixed $\\lambda_{\\text{feat}}$ (models (i) and (ii), respectively).", out.width="100%", fig.asp=1/2>>=
-@
-
-	The models (iii) and (iv) give median $\phi$ estimates of
-	\Sexpr{round(mphi[1], 2)} and \Sexpr{round(mphi[2], 2)}, 
-	respectively. The $\phi$ estimates are
-	underestimated, compared to true $\phi$ mean \Sexpr{round(mean(phi), 2)}. 
-	Model (iii) slightly underestimates the median
-	$\lambda_{\text{feat}}$ at \Sexpr{round(mlambdaf, 2)[2]} 
-	(the true value is one) and thus overestimates the 
+	$\lambda_{\text{feat}}$ is a higher level scale
+	parameter, which are, in general, hard to estimate. 
+<<simulations_gdsc_est1, fig.cap="Estimated (a) hyperparameters $\\bm{\\alpha}_{\\text{feat}}$ and (b) prior means $\\bm{\\phi}_{jd}$ from simulation version 1 with fixed $\\tau_d^2$.", out.width="100%", fig.asp=1/2>>=
+@
+
+	Model NIG1 gives a median $\phi$ estimate of \Sexpr{round(est.phi, 2)},  
+	equal to the true mean of the $\phi_{jd}$, \Sexpr{round(true.phi, 2)}. 
+	$\lambda_{\text{feat}}$ is slightly underestimated with median estimate
+	\Sexpr{round(est.lambdaf[1], 2)} (the true value is one) and thus 
+	overestimates the 
 	prior $\gamma_{jd}^2$ variance. This is likely to compensate for the common
-	prior mean 
-	$\phi$, that is not able to capture part of the variation in the true feature
-	specific $\phi_{jd}$. The median evidence 
-	lower bounds (ELBO), which approximate the marginal likelihoods of models
-	(i)-(iv), are estimated as \Sexpr{round(melbo, 2)[1:3]}, and 
-	\Sexpr{round(melbo, 2)[4]}, respectively. This implies that model (i),
-	with estimated $\bm{\alpha}_{\text{feat}}$ and $\lambda_{\text{feat}}$ fits
-	the data best (the log Bayes factors comparing models (ii)-(iv) to model (i) 
-	are \Sexpr{round(melbo[-c(1, 4)] - melbo[1], 2)}, and 
-	\Sexpr{round(melbo[4] - melbo[1], 2)}, respectively, and thus highly favouring
-	model (i)). The prediction MSEs for the four models on the test 
-	data and gaps between training and test prediction MSE are displayed in Table
-	\ref{tab:pmse1}. Table \ref{tab:pmse1} confirms that the models that
-	include the external data (models (i) and (ii)) learn the underlying structure 
-	in the data better than models (iii) and (iv) that do not include the external 
+	prior mean $\phi$, that is not able to capture part of the variation in the 
+	true feature specific $\phi_{jd}$.
+
+	Table \ref{tab:simulations_gdsc_tab1} displays several median performance
+	measures for simulation version 1. The table includes as a reference 
+	the null model, which
+	is an empty model that only fits the means of the $\y_{d}$.
+	The median evidence lower bounds (ELBO) approximates the marginal
+	likelihoods of models NIG1 and NIG2 on both training and test data, while 
+	estimation MSE (EMSE) and prediction MSE (PMSE) give measures of predictive 
+	performance. Focussing on MSE, we see that the NIG2 model, that includes
+	the external covariates, performs best, both in terms of estimation and 
+	prediction. Model fit is best for the NIG1 model that does not include the
+	external covariates, according to the ELBO, calculated on the test data.
+	The lower MSEs confirm that the model that
+	includes the external data (NIG2) learns the underlying structure 
+	in the data better than model NIG1 that does not include the external 
 	data.
-<<pmse1>>= 
+<<simulations_gdsc_tab1>>=
 library(knitr, quietly=TRUE, verbose=FALSE)
 suppressWarnings(library(kableExtra))
-load("results/simulations_gdsc_set1.Rdata")
-mapmse <- apply(res[, c(7:12)], 2, median)
-mapmset <- apply(res[, c(13:18)], 2, median)
-tab <- rbind(mapmse, mapmse - mapmset)[, -c(5, 6)]
-rownames(tab) <- c("$\\text{MSE}_{\\text{test}}$",
-                  "$\\text{MSE}_{\\text{test}} - \\text{MSE}_{\\text{train}}$")
 options(knitr.kable.NA="", digits=3)
-kable(tab,
-      col.names=c("(i)", "(ii)", "(iii)", "(iv)"),
-      caption="Median prediction MSEs for the models (i)-(iv) in the fixed 
-      $\\tau_d^2$ simulation setting.", 
+kable(tab, col.names=colnames(tab), align="r",
+      caption="Median performance measure estimates for simulation version 1 (best performing model in bold).", 
       format="latex", booktabs=TRUE, escape=FALSE) %>%
@@
 
-  \subsection{Random drug component}
-	To investigate the interplay of the feature and drug variance components, we 
-	repeat the whole simulation study with random $\tau_d^2$. Similarly as with 
-	the $\gamma_{jd}^2$, we randomly create four groups of drugs and corresponding
-	external drug dummy covariates. We simulate the $\tau_d^2$ as in
-	(\ref{eq:tauprior}), where we set $\bm{\alpha}_{\text{drug}} = 
-	\begin{bmatrix} 1 & 1 & 3 & 7 \end{bmatrix} \tr$, such
-	that the $\tau_{d}^2$ of the four groups of features have prior means
-	$\chi_{d} \in \{ 1, 1/2, 1/4, 1/8 \}$. The scale $\lambda_{\text{drug}}$ is 
-	set to one. We estimate the same four models as in the fixed $\tau_d^2$ 
-	setting, but in addition estimate the $\tau_d^2$ priors as well. Again, we
-	use two different versions: one where we estimate 
-	$\bm{\lambda}$ and one where we fix it to its true value.
-
-<<include=FALSE>>=
-load("results/simulations_gdsc_set2.Rdata")
-mapmse <- apply(res[, c(7:12)], 2, median)
-mapmset <- apply(res[, c(13:18)], 2, median)
-melbo <- apply(elbo, 2, median)
-phi <- 1/(set$alphaf %*% t(unname(model.matrix(~ factor(1:4)))))
-chi <- 1/(set$alphad %*% t(unname(model.matrix(~ factor(1:4)))))
-mphi <- apply(1/est[, c(21, 25)], 2, median)
-mchi <- apply(1/est[, c(22, 26)], 2, median)
-mlambdaf <- apply(est[, c(9, 23)], 2, median)
-mlambdad <- apply(est[, c(10, 24)], 2, median)
-@	
-	Figures \ref{fig:simulations_gdsc_est2} and 
-	\ref{fig:simulations_gdsc_phi2} display the estimated $\bm{\alpha}$, 
-	$\phi_{jd}$, and $\chi_d$ for models (i) and (ii). The scale parameters were
-	estimated as $\lambda_{\text{feat}}=\Sexpr{round(mlambdaf[1], 2)}$ and
-	$\lambda_{\text{drug}}=\Sexpr{round(mlambdad[1], 2)}$ in model (i). From 
-	Figure \ref{fig:simulations_gdsc_est2} we see that estimation of
-	$\bm{\alpha}_{\text{feat}}$ is again biased. However, from 
-	Figure \ref{fig:simulations_gdsc_phi2} we see that the scale and ordering of
-	the $\phi_{jd}$ are correclty estimated. We conjecture that the drop in 
-	estimation accuracy compared to the previous setting stems from the
-	difficult estimation of the $\tau_d^2$ prior. Estimation of this prior 
-	depends on only 168 observed drugs, while the $\gamma_{jd}^2$ prior is 
-	is estimated from approximately $300 \cdot 168=50400$ features. 
-	\cite{wiel_learning_2019} show in a simple regression setting that if the 
-	number of samples and the number of parameters on which the EB is based are of
-	the same order, empirical Bayes performs worse than when one of the two is
-	large compared to the other. 
-	Here, the number of samples $n=241$ and $D=168$ are of the same order, leading
-	to a suboptimal EB estimation performance.
-
-<<simulations_gdsc_est2, fig.cap="Estimated hyperparameters $\\bm{\\alpha}_{\\text{feat}}$ and $\\bm{\\alpha}_{\\text{drug}}$ from simulations with (a) \\& (c) estimated and (b) \\& (d) fixed $\\lambda_{\\text{feat}}$ and $\\lambda_{\\text{drug}}$ (models (i) and (ii), respectively).", out.width="100%", fig.asp=1>>=
-@
-
-<<simulations_gdsc_phi2, fig.cap="Estimated prior means $\\bm{\\phi}_{jd}$ and $\\bm{\\chi}_{d}$ for the four groups of features and drugs from simulations with (a) \\& (c) estimated and (b) \\& (d) fixed $\\lambda_{\\text{feat}}$ and $\\lambda_{\\text{drug}}$ (models (i) and (ii), respectively).", out.width="100%", fig.asp=1>>=
-@
-
-  We argue however, that even though the prior parameters are not estimated
-  correctly, there is still some gain in predictive performance.
-  Prediction MSEs on the test data the training-test gaps in prediction MSE are 
-  shown in Table \ref{tab:pmse2}. The comparisons of models (i) and
-	(ii) vesus models (iii) and (iv) confirm that even though EB estimation is
-	suboptimal, the external data do add something in predictive performance. 
-<<pmse2>>= 
+<<simulation_gdsc_res2>>=
+res <- read.table("results/simulations_gdsc_res2.txt", row.names=NULL)
+temp <- res[, 1]
+res <- as.matrix(res[, -1])
+rownames(res) <- temp
+true.chi <- mean(1/(c(1, 1, 3, 7) %*% t(unname(model.matrix(~ factor(1:4))))))
+est.chi <- median(1/res[rownames(res)=="alphad0", 1])
+est.lambdad <- apply(res[rownames(res)=="lambdad", c(1, 2)], 2, median)
+tab <- cbind(apply(res[rownames(res)=="elbo", ], 2, median),
+             apply(res[rownames(res)=="elbot", ], 2, median) -
+               apply(res[rownames(res)=="elbo", ], 2, median),
+             apply(res[rownames(res)=="emse", ], 2, median),
+             apply(res[rownames(res)=="pmse", ], 2, median),
+             apply(res[rownames(res)=="pmse", ], 2, median) - 
+               apply(res[rownames(res)=="pmset", ], 2, median))
+tab <- round(tab, 3)
+tab[which(tab[, 1]==max(tab[, 1], na.rm=TRUE)), 1] <- 
+  paste0("\\textbf{", tab[which(tab[, 1]==max(tab[, 1], na.rm=TRUE)), 1], "}")
+tab[which(tab[, 2]==min(tab[, 2], na.rm=TRUE)), 2] <- 
+  paste0("\\textbf{", tab[which(tab[, 2]==min(tab[, 2], na.rm=TRUE)), 2], "}")
+tab[which(tab[, 3]==min(tab[, 3], na.rm=TRUE)), 3] <- 
+  paste0("\\textbf{", tab[which(tab[, 3]==min(tab[, 3], na.rm=TRUE)), 3], "}")
+tab[which(tab[, 4]==min(tab[, 4], na.rm=TRUE)), 4] <- 
+  paste0("\\textbf{", tab[which(tab[, 4]==min(tab[, 4], na.rm=TRUE)), 4], "}")
+tab[which(tab[, 5]==min(tab[, 5], na.rm=TRUE)), 5] <- 
+  paste0("\\textbf{", tab[which(tab[, 5]==min(tab[, 5], na.rm=TRUE)), 5], "}")
+colnames(tab) <- c("$\\text{ELBO}_{\\text{test}}$",
+                   "$\\text{ELBO}_{\\text{train}} 
+                   - \\text{ELBO}_{\\text{test}}$",
+                   "$\\text{EMSE}_{\\text{test}}$",
+                   "$\\text{PMSE}_{\\text{test}}$",
+                   "$\\text{PMSE}_{\\text{test}} 
+                   - \\text{PMSE}_{\\text{train}}$")     
+@	
+  Figure \ref{fig:simulations_gdsc_est2} shows the estimated
+	$\bm{\alpha}_{\text{drug}}$ together with its true value for NIG2 in 
+	version 2 of the simulation study. They are estimated without bias.
+	Median $\lambda_{\text{drug}}=\Sexpr{round(est.lambdad[2], 2)}$ is slightly 
+	overestimated compared to true value 1. 
+	The median $\chi$ estimate in the NIG1 model is \Sexpr{round(est.chi, 2)},  
+	which is almost equal to the true mean \Sexpr{round(true.chi, 2)}. 
+	$\lambda_{\text{drug}}$ is underestimated with median estimate
+	\Sexpr{round(est.lambdad[1], 2)} compared to the true value.
+	Median performance measures for simulation version 2 are displayed in Table
+	\ref{tab:simulations_gdsc_tab2}. Again, we see that NIG2 performs best in 
+	terms of MSE. However, NIG1 performance is the same here, indicating that 
+	the external covariates do not add much here.
+	Model fit is again best for the NIG1 model.
+<<simulations_gdsc_est2, fig.cap="Estimated (a) hyperparameters $\\bm{\\alpha}_{\\text{feat}}$ and (b) prior means $\\bm{\\phi}_{jd}$ from simulation version 2 with fixed $\\gamma_{jd}^2$.", out.width="100%", fig.asp=1/2>>=
+@
+
+<<simulations_gdsc_tab2>>=
 library(knitr, quietly=TRUE, verbose=FALSE)
 suppressWarnings(library(kableExtra))
-load("results/simulations_gdsc_set2.Rdata")
-mapmse <- apply(res[, c(7:12)], 2, median)
-mapmset <- apply(res[, c(13:18)], 2, median)
-tab <- rbind(mapmse, mapmse - mapmset)[, -c(5, 6)]
-rownames(tab) <- c("$\\text{MSE}_{\\text{test}}$",
-                  "$\\text{MSE}_{\\text{test}} - \\text{MSE}_{\\text{train}}$")
 options(knitr.kable.NA="", digits=3)
-kable(tab,
-      col.names=c("(i)", "(ii)", "(iii)", "(iv)"),
-      caption="Median prediction MSEs for the models (i)-(iv) in the random 
-      $\\tau_d^2$ simulation setting.", 
+kable(tab, col.names=colnames(tab), align="r",
+      caption="Median performance measure estimates for simulation version 2 (best performing model in bold).", 
+      format="latex", booktabs=TRUE, escape=FALSE) %>%
+  kable_styling(latex_options=c("HOLD_position"))
+@
+
+<<simulation_gdsc_res3>>=
+res <- read.table("results/simulations_gdsc_res3.txt", row.names=NULL)
+temp <- res[, 1]
+res <- as.matrix(res[, -1])
+rownames(res) <- temp
+true.chi <- mean(1/(c(1, 1, 3, 7) %*% t(unname(model.matrix(~ factor(1:4))))))
+est.chi <- median(1/res[rownames(res)=="alphad0", 1])
+est.lambdad <- apply(res[rownames(res)=="lambdad", c(1, 2)], 2, median)
+true.phi <- mean(1/(c(1, 1, 3, 7) %*% t(unname(model.matrix(~ factor(1:4))))))
+est.phi <- median(1/res[rownames(res)=="alphaf0", 1])
+est.lambdaf <- apply(res[rownames(res)=="lambdaf", c(1, 2)], 2, median)
+
+phi.ratio <- apply(apply(sapply(0:3, function(s) {
+  res[rownames(res)==paste0("alphaf" ,s), colnames(res)=="NIG2"]}) %*% 
+    t(cbind(1, rbind(0, diag(3)))), 2, function(m) {
+      res[rownames(res)=="alphaf0", colnames(res)=="NIG2"]/m}), 2, median)
+chi.ratio <- apply(apply(sapply(0:3, function(s) {
+  res[rownames(res)==paste0("alphad" ,s), colnames(res)=="NIG2"]}) %*% 
+    t(cbind(1, rbind(0, diag(3)))), 2, function(m) {
+      res[rownames(res)=="alphad0", colnames(res)=="NIG2"]/m}), 2, median)
+true.ratio <- as.numeric(1/(c(1, 1, 3, 7) %*% 
+                              t(unname(model.matrix(~ factor(1:4))))))
+
+tab <- cbind(apply(res[rownames(res)=="elbo", ], 2, median),
+             apply(res[rownames(res)=="elbot", ], 2, median) -
+               apply(res[rownames(res)=="elbo", ], 2, median),
+             apply(res[rownames(res)=="emse", ], 2, median),
+             apply(res[rownames(res)=="pmse", ], 2, median),
+             apply(res[rownames(res)=="pmse", ], 2, median) - 
+               apply(res[rownames(res)=="pmset", ], 2, median))
+tab <- round(tab, 3)
+tab[which(tab[, 1]==max(tab[, 1], na.rm=TRUE)), 1] <- 
+  paste0("\\textbf{", tab[which(tab[, 1]==max(tab[, 1], na.rm=TRUE)), 1], "}")
+tab[which(tab[, 2]==min(tab[, 2], na.rm=TRUE)), 2] <- 
+  paste0("\\textbf{", tab[which(tab[, 2]==min(tab[, 2], na.rm=TRUE)), 2], "}")
+tab[which(tab[, 3]==min(tab[, 3], na.rm=TRUE)), 3] <- 
+  paste0("\\textbf{", tab[which(tab[, 3]==min(tab[, 3], na.rm=TRUE)), 3], "}")
+tab[which(tab[, 4]==min(tab[, 4], na.rm=TRUE)), 4] <- 
+  paste0("\\textbf{", tab[which(tab[, 4]==min(tab[, 4], na.rm=TRUE)), 4], "}")
+tab[which(tab[, 5]==min(tab[, 5], na.rm=TRUE)), 5] <- 
+  paste0("\\textbf{", tab[which(tab[, 5]==min(tab[, 5], na.rm=TRUE)), 5], "}")
+colnames(tab) <- c("$\\text{ELBO}_{\\text{test}}$",
+                   "$\\text{ELBO}_{\\text{train}} 
+                   - \\text{ELBO}_{\\text{test}}$",
+                   "$\\text{EMSE}_{\\text{test}}$",
+                   "$\\text{PMSE}_{\\text{test}}$",
+                   "$\\text{PMSE}_{\\text{test}} 
+                   - \\text{PMSE}_{\\text{train}}$")     
+@	
+  In Figure \ref{fig:simulations_gdsc_est3} the 
+	$\bm{\alpha}_{\text{feat}}$ and $\bm{\alpha}_{\text{drug}}$ estimated by NIG2
+	are displayed together with their true values for
+	simulation 3. $\bm{\alpha}_{\text{feat}}$ are underestimated, while 
+	$\bm{\alpha}_{\text{drug}}$ are overestimated. These biases seem to 
+	be consistent though. The median ratios $\phi_1$ to $\phi_2$, $\phi_3$, 
+	$\phi_4$ and $\chi_1$ to $\chi_2$, $\chi_3$, 
+	$\chi_4$ are \Sexpr{round(phi.ratio, 2)[-1]} and 
+	\Sexpr{round(chi.ratio, 2)[-1]}, respectively, compared to their true values
+	\Sexpr{round(true.ratio, 2)[-1]}. So in a relative sense, the estimates are
+	correct. Median $\lambda_{\text{feat}}$ and $\lambda_{\text{drug}}$ are
+	are over- and underestimated with \Sexpr{round(est.lambdaf[2], 2)}
+	and \Sexpr{round(est.lambdad[2], 2)}, respectively, compared to their true
+	value 1. The NIG1 model is also consistently over- and underestimating $\phi$
+	and $\chi$ with estimates \Sexpr{round(est.phi, 2)} and 
+	\Sexpr{round(est.chi, 2)}, respectively (compared to the true mean 
+	\Sexpr{round(true.chi, 2)}). NIG1 also over- and underestimates 
+	$\lambda_{\text{feat}}$ and $\lambda_{\text{drug}}$ with median estimates
+	\Sexpr{round(est.lambdaf[1], 2)}
+	and \Sexpr{round(est.lambdad[1], 2)}, respectively. Table 
+	\ref{tab:simulations_gdsc_tab3}, with median performance measures, shows
+	that NIG2 outperforms the other models in terms of prediction and estimation
+	MSE.
+<<simulations_gdsc_est3, fig.cap="Estimated (a) hyperparameters $\\bm{\\alpha}_{\\text{feat}}$ and (b) prior means $\\bm{\\phi}_{jd}$ from simulation version 3.", out.width="100%", fig.asp=1>>= 
+@
+
+<<simulations_gdsc_tab3>>=
+library(knitr, quietly=TRUE, verbose=FALSE)
+suppressWarnings(library(kableExtra))
+options(knitr.kable.NA="", digits=3)
+kable(tab, col.names=colnames(tab), align="r",
+      caption="Median performance measure estimates for simulation version 3 (best performing model in bold).", 
       format="latex", booktabs=TRUE, escape=FALSE) %>%
@@
 	they have larger prior $\beta_{jd}$ variances than drugs that are not in 
-	the pathway.
-
-%   - check whether compounds are PAINS \cite[]{baell_new_2010} or not. 
-%   PAINS indicates whether compound is frequent false hitter.
-%   
-%   - clinical trials database CTRP might contain info
-% 
-%   - extra data on GDSC might be in PharamcoGx R-package. Also contains bunch 
-%   of functions.
-%   
-%   - check GSK open target EBI or ask Chris Wallace about it. Database that links
-% 	targets to drugs and diseases. Might be useful for drugs co-data.
-% 	
-% 	- try non-negative matrix factorisation/independent component analysis on 
-% 	binary fingerprints to get principal components that characterise the 
-% 	different drugs.
+	the pathway. Note that large $\beta_{jd}$ variances translate to large
+	prior $\gamma_{jd}^2$ and $\tau_d^2$ means.
 
 	\subsection{Results}
+	A preliminary step selects, for each
+	drug separately, 50 features that are not in the pathway of the drug with
+	maximum variance over the cell lines and 50 features that are in the pathway 
+	of the drug with maximum variance over the cell lines. If less than 50 
+	features in the pathway are available,
+	we select all available features in the pathway. This strategy ensures
+	that the features are somewhat balanced with respect to the external feature
+	covariate, i.e., pathway membership.
+
 	To compute performance measures, we randomly split the data into training and
@@
 	The estimated 
-	models are: two models like in (\ref{eq:prior}) but with different external 
-	variables, named (i) NIG1 and (ii) NIG2 in the following, (iii) a model 
-	as described in \cite{kpogbezan_empirical_2017}, named bSEM, (iv) a regular
-	ridge model, and (v) a regular lasso model. The NIG1 incorporates no external
-	variables and estimates the common prior $\gamma_{jd}^2$ and $\tau_d^2$ 
-	means $\phi$ and $\chi$ directly, while the NIG2 model includes the
-	developmental phase and action drug covariates, and the pathway
-	feature covariate, and estimates $\bm{\alpha}$. The bSEM model is similar to 
-	the NIG model in that it 
+	models are: four models like in (\ref{eq:prior}) but with different external 
+	variables, named NIG1 and NIG2 in the following. NIG1 does 
+	not use the external covariates and estimates
+	one overall $\phi$ and one overall $\chi$, while NIG2 uses the external 
+	covariates to estimate both $\bm{\alpha}_{\text{feat}}$ and
+	$\bm{\alpha}_{\text{drug}}$, while NIG3 and NIG4 estimate only 
+	$\bm{\alpha}_{\text{feat}}$ or $\bm{\alpha}_{\text{drug}}$, respectively,
+	keeping either $\forall d: \tau_d^2=1$ or $\forall j,d: \gamma_{jd}^2=1$
+	fixed.
+
+	In addition, we estimate regular lasso and ridge models, and a
+	bSEM model \cite[]{leday_gene_2017,kpogbezan_empirical_2017}. The bSEM model 
+	is similar to the NIG model in that it
 	draws the $\beta_{jd}$ from a conditionally normal distribution, but
@@
 	$\gamma_{jd}^2 \sim \Gamma^{-1} (\phi_{jd}, \lambda_{jd})$, instead of
-	from the inverse Gaussian distribution. Additionally, 
+	$\gamma_{jd}^2 \sim \mathcal{IG} (\phi_{jd}, \lambda_{jd})$. Additionally,
 	the bSEM model can
 	only incorporate one binary external covariate that divides the features into
-	two groups, i.e., $\bm{\alpha} = \begin{bmatrix} \alpha_0 & \alpha_1 
+	two groups, i.e., $\bm{\alpha}_{\text{feat}} = \begin{bmatrix} \alpha_0 &
+	\alpha_1
 	\end{bmatrix} \tr$ and $\bm{c}_{jd} = \begin{bmatrix} 1 & 0 \end{bmatrix}
@@
 	estimated in a similar fashion as in Section \ref{sec:estimation}.
-
-  From the $\bm{\alpha}$ estimates in Tables \ref{tab:results_drug} and
-  \ref{tab:results_feat} we see that there
-  seems to be an effect of the
-  external variables. If we look at the NIG1 model, that includes all external 
-  covariates, we see that approved drugs are easier to predict, due to small
-  prior precision, followed by drugs that are in clinical development,
-  as we expected. The cytotoxic drugs are the hardest to predict, 
-  with the largest prior precision (smallest prior variance), followed by the 
-  targeted drugs and the unknown drugs. This is again according to expectation.
-  For the unkown drugs, we are unable to confirm the small precision. Genes 
-  that are in the target pathway have 
-  smaller prior precisions, according to expectation, which confirms that these
-  genes are more important for prediction of drug sensitivity. The estimates 
-  from
-  the bSEM model are also according to expectation: genes that are in the target
-  pathway have a much smaller precision than genes that are not in the target
-  pathway. The 
-  effects of the pathway indicator seem to be much larger in the bSEM model. 
-<<results_drug>>= 
-library(knitr, quietly=TRUE, verbose=FALSE)
-suppressWarnings(library(kableExtra))
-load("results/data_gdsc_res1.Rdata")
-tab1 <- cbind(tab[, 1], c(NA, 0, NA), tab[, c(3, 2)], c(NA, 0, NA), 
-              tab[, c(4, 5)])
+	The null model that fits the means of the $\y_d$ is included as a reference.
+
+<<analysis_gdsc_fit1>>=
+fit <- read.table("results/analysis_gdsc_fit1.txt", row.names=NULL)
+temp <- fit[, 1]
+fit <- as.matrix(fit[, -1])
+rownames(fit) <- temp
+
+res <- read.table("results/analysis_gdsc_res1.txt", row.names=NULL)
+temp <- res[, 1]
+res <- as.matrix(res[, -1])
+rownames(res) <- temp
+
+est1 <- cbind(fit[, 1], c(NA, 0, 0, NA, 0), fit[, 2])
+colnames(est1) <- c("intercept", "not in pathway", "in pathway")
+est2 <- cbind(fit[, 3:5], c(NA, 0, NA, 0, NA), fit[, c(7:6)], 
+              c(NA, 0, NA, 0, NA))
+colnames(est2) <- c("intercept", "experimental", "development", "approved",
+                    "unkown", "targeted", "cytotoxic")
+
+tab <- cbind(apply(res[rownames(res)=="elbo", ], 2, median),
+             apply(res[rownames(res)=="elbot", ], 2, median) -
+               apply(res[rownames(res)=="elbo", ], 2, median),
+             apply(res[rownames(res)=="pmse", ], 2, median))
+tabm <- tab
+tabm <- round(tabm, 3)
+tabm[which(tabm[, 1]==max(tabm[, 1], na.rm=TRUE)), 1] <- 
+  paste0("\\textbf{", tabm[which(tabm[, 1]==max(tabm[, 1], na.rm=TRUE)), 1], "}")
+tabm[which(tabm[, 2]==min(tabm[, 2], na.rm=TRUE)), 2] <- 
+  paste0("\\textbf{", tabm[which(tabm[, 2]==min(tabm[, 2], na.rm=TRUE)), 2], "}")
+tabm[which(tabm[, 3]==min(tabm[, 3], na.rm=TRUE)), 3] <- 
+  paste0("\\textbf{", tabm[which(tabm[, 3]==min(tabm[, 3], na.rm=TRUE)), 3], "}")
+colnames(tabm) <- c("$\\text{ELBO}_{\\text{test}}$",
+                   "$\\text{ELBO}_{\\text{train}} 
+                   - \\text{ELBO}_{\\text{test}}$",
+                   "$\\text{PMSE}_{\\text{test}}$")
+
+library(glmnet)
+kurtosis <- function(x) {
+  n <- length(x) 
+  kurt <- (n + 1)*n*sum((x - mean(x))^4)/((n - 1)*(n - 2)*(n - 3)*var(x)^2) - 
+    3*(n - 1)^2/((n - 2)*(n - 3))
+  replace(kurt, !is.finite(kurt), NA)
+}
+load("results/analysis_gdsc_fit1.Rdata")
+best2.semnig <- fit2.semnig$vb$mu
+best1.lasso <- lapply(fit1.lasso, function(s) {
+  unname(coef(s, "lambda.min")[-1, ])})
+best1.ridge <- lapply(fit1.ridge, function(s) {
+  unname(coef(s, "lambda.min")[-1, ])})
+best1.bSEM <- lapply(fit1.bSEM$vb$beta, function(s) {s[, 2]})
+kurts <- c(mean(sapply(best2.semnig, kurtosis), na.rm=TRUE), 
+           mean(sapply(best1.lasso, kurtosis), na.rm=TRUE),
+           mean(sapply(best1.ridge, kurtosis), na.rm=TRUE), 
+           mean(sapply(best1.bSEM, kurtosis), na.rm=TRUE))
+ranges <- rbind(range(unlist(lapply(1:length(
+  fit2.semnig$eb$mpriord), function(d) {
+    fit2.semnig$eb$mpriord[d]*fit2.semnig$eb$mpriorf[[d]]}))),
+  fit1.bSEM$eb$b[nrow(fit1.bSEM$eb$b), ]/
+    (fit1.bSEM$eb$a[nrow(fit1.bSEM$eb$a), ] - 1))
+@
+  The non-zero $\bm{\alpha}$ estimates in Tables 
+  \ref{tab:analysis_gdsc_est1_feat} and \ref{tab:analysis_gdsc_est1_drug} show 
+  that there is an effect of the
+  external covariates. Models NIG2, NIG3, and bSEM estimate a negative effect 
+  for
+  the external pathway covariate. This translates to a negative additive effect 
+  on the expected prior precision of the $\bm{\beta}_d$, i.e., genes that are 
+  found
+  in the target pathway of a drug are more predictive of IC50 than genes that
+  are not in the target pathway, according to expectation. Furthermore the NIG2
+  model estimates positive effects for the experimental and developmental drugs,
+  compared to the approved drugs. This translates to smaller expected 
+  prior precisions for the regression parameters for approved drugs, i.e.,
+  approved drugs are easier to predict than experimental and developmental 
+  drugs. On the other hand, the NIG4 model, that only includes external
+  drug covariates, estimates a negative effect for the developmental drugs.
+  The NIG2 model estimates the smallest prior $\bm{\beta}_d$ precision for the
+  targeted drugs, followed by cytotoxic drugs, while drugs with an unkown target
+  are assigned the largest prior $\bm{\beta}_d$ precision, according
+  to expectation. Somewhat surprisingly, the NIG4 model estimates the smallest
+  prior $\bm{\beta}_d$ precision for the cytotoxic drugs, compared to the 
+  targeted and unkown drugs. We conjecture that, due to absence of a local
+  feature-specific variance component, this model is not flexible 
+  enough to correctly capture the full covariance structure of the 
+  $\bm{\beta}_d$ and therefore incorreclty estimates $\bm{\alpha}$.
+  To see the total effect of the external covariates on the $\bm{\beta}_d$ 
+  prior, we calculated the range of the expected prior variances 
+  (up to a multiplicative error variance) as 
+  $\hat{E}(\gamma_{jd}^2 \cdot \tau_d^2) \in \[ \Sexpr{ranges[1, ]}\]$ and
+  $\hat{E}(\gamma_{jd}^2 \cdot \tau_d^2) \in \[ \Sexpr{ranges[2, ]}\]$ for
+  NIG2 and bSEM, respectively. This confirms that the estimated prior 
+  $\bm{\beta}_d$ variance components are in the same order of magnitude and that
+  the external covariates have a detectable effect on the prior.
+<<analysis_gdsc_est1_feat>>=
 options(knitr.kable.NA="", digits=3)
-kable(tab1,
-      col.names=c("intercept", "approved", "development", 
-                  "experimental", "targeted", "unkown", "cytotoxic"),
+kable(est1,
+      col.names=colnames(est1), align="r",
+      caption="$\\bm{\\alpha}_{\\text{feat}}$ estimates from GDSC data.",
+      format="latex", booktabs=TRUE, escape=FALSE) %>%
+  kable_styling(latex_options=c("HOLD_position"))
+@
+
+<<analysis_gdsc_est1_drug>>=
+kable(est2,
+      col.names=colnames(est2), align="r",
       caption="$\\bm{\\alpha}_{\\text{drug}}$ estimates from GDSC data.",
@@
 @
-
-<<results_feat>>= 
-library(knitr, quietly=TRUE, verbose=FALSE)
-suppressWarnings(library(kableExtra))
-load("results/data_gdsc_res1.Rdata")
-tab2 <- cbind(tab[, 6], c(NA, 0, 0), tab[, 7])
-options(knitr.kable.NA="", digits=3)
-kable(tab2,
-      col.names=c("intercept", "not in pathway", "in pathway"),
-      caption="$\\bm{\\alpha}_{\\text{feat}}$ estimates from GDSC data.", 
+  The median PMSEs and ELBOs, calculated on the test data
+  are displayed in Table \ref{tab:analysis_gdsc_tab1}. In terms of prediction 
+  error, ridge outperforms the other models 
+  (PMSE of \Sexpr{round(tab[6, 3], 3)}), while bSEM 
+  (PMSE of \Sexpr{round(tab[7, 3], 3)}) performs worse than 
+  the null model (PMSE of \Sexpr{round(tab[8, 3], 3)}). Second and third best
+  performing models are NIG1 and NIG2 (PMSEs of \Sexpr{round(tab[1, 3], 3)} and
+  \Sexpr{round(tab[2, 3], 3)}, respectively), both outperforming the lasso
+  \Sexpr{round(tab[5, 3], 3)}. Part of the differences in performance between
+  NIG, lasso, and ridge may be explained with the different levels of 
+  ``sparsity'' in the solution. Although NIG and ridge do not give exactly 
+  sparse solutions (none of the $\bm{\beta}_d$ coefficients are set to zero), we
+  may use the ``heavy-tailedness'' of the distribution of the estimated 
+  $\bm{\beta}_d$ as a proxy for sparsity. A common measure of heavy-tailedness
+  is excess kurtosis, calculated as \Sexpr{round(kurts[1], 3)}, 
+  \Sexpr{round(kurts[2], 3)}, and \Sexpr{round(kurts[3], 3)} for the NIG2, 
+  lasso, and ridge estimates, respectively. Evidently, lasso gives the 
+  ``sparsest'' solution, followed by NIG and ridge. In some sense, this
+  means that the NIG uses less of the $\bm{\beta}_d$ to predict.
+<<analysis_gdsc_tab1>>=
+kable(tabm,
+      col.names=colnames(tabm), align="r",
+      caption="Performance measures calculated on GDSC data (best performing model in bold).",
       format="latex", booktabs=TRUE, escape=FALSE) %>%
@@
 
-  The median prediction MSEs and the median ELBOs, calculated on the test data
-  are displayed in Table \ref{tab:results_data}. In terms of 
-  is lowest in the NIG models 
-  (\Sexpr{round(mapmse[2], 2)} and \Sexpr{round(mapmse[3], 2)}, in the
-  NIG1 and NIG2 models, respectively). The bSEM model is not able to 
-  improve on the null model with an median prediction MSE of 
-  \Sexpr{round(mapmse[4], 2)} compared to the null model's MSE of 
-  \Sexpr{round(mapmse[1], 2)}.
-<<results_data>>= 
-load("results/data_gdsc_res1.Rdata")
-mapmse <- apply(apmse, 2, median)
-melbo <- apply(elbo, 2, median)
-tab <- rbind(mapmse, c(melbo, NA, NA))
-rownames(tab) <- c("MSE", "ELBO")
-options(knitr.kable.NA="", digits=3)
-kable(tab,
-      col.names=c("NIG1", "NIG2", "bSEM", "lasso", "ridge"),
-      caption="Median prediction MSE and median ELBO estimated on the test data", 
-      format="latex", booktabs=TRUE, escape=FALSE) %>%
-  kable_styling(latex_options=c("HOLD_position"))
-@
-
-  In addition to the empirical Bayes estimates and the prediction MSE, we 
-  investigated the stability of the estimates by ranking the posterior means
-  $\E_{Q}(\bbeta_d)$ and calculating the Spearman's rank
-  correlation over the 50 data splits. They are depicted in Figure 
-  \ref{fig:data_gdsc_res1_brank_hist}. In both NIG models the correlations
-  between splits are higher than in the bSEM model; an indication that 
-  estimation in the NIG model is more stable than in the bSEM model.
-<<data_gdsc_res1_brank_hist, fig.cap="Spearman's rank correlation between the estimated posterior means $\\E_{Q}(\\bbeta_d)$ in the 50 data splits.", out.width="70%", fig.asp=2/3>>= 
-@
-
-	% 
-	% \subsection{CCLE}
-	% - extra data on CCLE might be in PharamcoGx R-package
-	% 
-	% \subsection{NCI-50}
-	% 
-	% \subsection{CMap}
-	% - Cmap \cite[]{subramanian_next_2017,lamb_connectivity_2006} contains 
-	% pharma-genomic data (similar to CCLE and GDSC). 
-	% 
-	% - Data available at: 
-	% https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE92742. 
-	% 
-	% - Cmap website: https://clue.io/cmap.
-
 	\section{Discussion}\label{sec:discussion}
+<<time>>=
+load("results/analysis_gdsc_fit1.Rdata")
+time2.semnig <- sum(fit2.semnig$time)
+@	
 	The preceding presents a novel model for drug sensitivity
@@
 	efficient and scales well with the number of features and samples.
-
-	The model is put into practice on the GDSC data. The results in Section
+	Estimation of the NIG model that includes both external drugs and feature
+	covariates (NIG2 in Section \ref{sec:gdsc}) on the GDSC data from Section
+	\ref{sec:gdsc} took \Sexpr{round(time2.semnig, 2)} seconds on a 2016 
+	MacBook Pro with a 2 GHz Dual-Core Intel Core i5 processor and
+	8 GB of 1867 MHz LPDDR3 memory, running macOS  10.15.1.
+
+	Simulations in Section \ref{sec:simulations} show that the method is 
+	reasonably able to recover hyperparameters. In addition, the simulations
+	show that inclusion of informative external covariates is indeed beneficial to
+	predictive performance. The model is put into practice on the GDSC data. 
+	The results in Section
 	\ref{sec:gdsc} show that the inclusion of the external covariates is 
-	beneficial for drug sensitivity prediction. Although the prediction MSE is
-	not lower if the external covariates are included, 
-	the resulting hyperparameter estimates are still informative
-	and may guide future research efforts.
-
-	The comparison to the bSEM model that allows for the inclusion of 
+	beneficial for drug sensitivity prediction. Prediction MSE of the NIG
+	model with both both feature and drug variance components is lower
+	than for the regular lasso model and although the prediction MSE is
+	not lower if external covariates are included, the resulting hyperparameter 
+	estimates are still informative and may guide future research efforts.
+	In addition, the comparison to the bSEM model that allows for the inclusion of 
 	one binary external feature covariate favours the proposed NIG model. 
 	Prediction MSEs are almost uniformly lower over the drugs for the NIG model 
-	compared to bSEM. Consequently, the average prediction MSE is significantly
+	compared to bSEM. Consequently, the total prediction MSE is significantly
 	lower for the NIG model.
 
-	Although the empirical Bayes estimates of the external covariate effects 
-	$\bm{\alpha}$ are informative, they are small compared to 
-	bSEM's estimates. Consequently, the posteriors of the models with and without
-	the external covariates (NIG2 and NIG1 in Section \ref{sec:gdsc}) do not 
-	differ by much. This results similar prediction MSEs with
-	the inclusion of the external covariates. This is a topic that deserves 
-	further attention. One proposed solution is to include an extra 
-	prior variance component that promotes a more heavy-tailed $\beta_{jd}$ prior
-	distribution. An example is a variance component with a half-cauchy prior 
-	distribution, i.e., the horseshoe \cite[]{carvalho_handling_2009}. 
-	This solution is similar to the pInc model from 
-	\cite{kpogbezan_incorporating_2019}. 
-	Like the bSEM model, the pInc model gives significantly larger external
-	covariate effects, 
-	so we conjecture that the inclusion of a similar heavy-tailed prior variance 
-	component will have a similar effect in the NIG model.
-
 	\section*{Supplementary Material}
-	Supplementary Material is available online from 
+	Source code to reproduce the manuscript and the contained results is available
+	as Supporting Information on the journal’s web page 
+	(http://onlinelibrary.wiley.com/doi/xxx/ suppinfo) and from
 	\url{https://github.com/magnusmunch/cambridge}.
+
+	\section*{Acknowledgements}
 
@@
 
+	\bibliographystyle{authoryear} 
 	\bibliography{refs}

### 2019-11-27 docs/manuscript1.Rnw
added EMSE split for large and small betas to simulations_gdsc.R

@@
 
-	\section{Simulations}\ref{sec:simulations}
+	\section{Simulations}\label{sec:simulations}
 	\subsection{Setup}
@@
 
-	We estimate four models: (i) an NIG model that does include external 
-	covariates in the estimation called NIG1 in the following, (ii) the NIG model
-	estimated as in Section \ref{sec:estimation} called NIG2, and regular (iii)
-	lasso and (iv) ridge models.
-	Exclusion of the external covariates as in the NIG1 model
+	We estimate four models: (i) the NIG model that only includes an intercept in
+	the external covariates, called NIG$_{\text{f}}^-$, NIG$_{\text{d}}^-$, or
+	NIG$_{\text{f}+\text{d}}^-$, depending on which variance components are 
+	estimated,
+	(ii) the NIG model estimated as in Section \ref{sec:estimation} that includes
+	the all external covariates, called 
+	NIG$_{\text{f}}$, NIG$_{\text{d}}$, or
+	NIG$_{\text{f}+\text{d}}$, again depending on which variance components are 
+	estimated, and regular (iii) lasso and (iv) ridge models.
+	Exclusion of the external covariates as in the NIG$_{\text{f}}^-$, 
+	NIG$_{\text{d}}^-$, and NIG$_{\text{f}+\text{d}}^-$ models
 	amounts to direct estimation of one common expected 
 	prior means $\phi$ and/or $\chi$, instead of regression estimates for the
-	$\phi_{jd}$ and/or $\chi_d$ as in the NIG2 model. All 
-	models are fit on 241 randomly selected cell lines, while performance measures 
-	are estimated from the remaining 241 cell lines. We repeat the whole
+	$\phi_{jd}$ and/or $\chi_d$ as in the 
+	NIG$_{\text{f}}$, NIG$_{\text{d}}$, and NIG$_{\text{f}+\text{d}}$ models. 
+	All models are fit on 241 randomly selected cell lines, while performance
+	measures are estimated from the remaining 241 cell lines. We repeat the whole
 	simulation 50 times.
@@
 kable(tab, col.names=colnames(tab), align="r",
-      caption="Median performance measure estimates for simulation version 1, with the best performing model in bold.", 
+      caption="Median performance measure estimates for simulation version 1 (best performing model in bold).", 
       format="latex", booktabs=TRUE, escape=FALSE) %>%
@@
 kable(tab, col.names=colnames(tab), align="r",
-      caption="Median performance measure estimates for simulation version 2, with the best performing model in bold.", 
+      caption="Median performance measure estimates for simulation version 2 (best performing model in bold).", 
       format="latex", booktabs=TRUE, escape=FALSE) %>%
@@
 kable(tab, col.names=colnames(tab), align="r",
-      caption="Median performance measure estimates for simulation version 3, with the best performing model in bold.", 
+      caption="Median performance measure estimates for simulation version 3 (best performing model in bold).", 
       format="latex", booktabs=TRUE, escape=FALSE) %>%
@@
 	possible, but a discussion on the pros and cons of different sensitivity
-	measures is beyond the aim of this paper. A comprehensive penalized regression
+	measures is beyond the aim of this paper. We have used the version of the 
+	data that is presented in \cite{iorio_landscape_2016}.
+	We average repeated measures over cell line-drug combinations and model the
+	logarithm of the IC50 values. In the following, IC50 refers 
+	to the log-transformed values. 
+	After iteratively removing the cell line or drug with the highest proportion 
+	of missings until no more missing values remain, we end up with IC50 
+	estimates for 482 cell lines on 168 drugs.
+
+	A comprehensive penalized regression
 	approach such as in 
@@
 	feasible.
-
-	We average repeated measures over cell line-drug combinations and model the
-	logarithm of the IC50 values. In the following, IC50 refers 
-	to the log-transformed values. 
-	After iteratively removing the cell line or drug with the highest proportion 
-	of missings until no more missing values remain, we end up with IC50 
-	estimates for 482 cell lines on 168 drugs.
 
@@
 	\subsection{Results}
-	As a preliminary step, we select
-	features according to three strategies. The first strategy selects for each
-	drug separately 50 features that are not in the pathway of the drug with
-	maximum variance and 50 features that are in the pathway of the drug
-	with maximum variance. If less than 50 features in the pathway are available,
-	we select all available features. This strategy ensures
+	A preliminary step selects, for each
+	drug separately, 50 features that are not in the pathway of the drug with
+	maximum variance over the cell lines and 50 features that are in the pathway 
+	of the drug with maximum variance over the cell lines. If less than 50 
+	features in the pathway are available,
+	we select all available features in the pathway. This strategy ensures
 	that the features are somewhat balanced with respect to the external feature
-	covariate, i.e., pathway membership. The second strategy selects the 
-	100 features with the largest variance, irrespective of pathway membership,
-	while the last stategy selects 1000 features with the largest variance,
-	again irrespective of pathway membership. These three strategies lead to
-	three different analyses described below.
+	covariate, i.e., pathway membership.
 
@@
 	models are: four models like in (\ref{eq:prior}) but with different external 
-	variables, named NIG1, NIG2, NIG3, and NIG4 in the following. NIG1 does 
+	variables, named NIG1 and NIG2 in the following. NIG1 does 
 	not use the external covariates and estimates
-	one overall $\phi$ and one overall $\chi$, NIG2 uses the external 
+	one overall $\phi$ and one overall $\chi$, while NIG2 uses the external 
 	covariates to estimate both $\bm{\alpha}_{\text{feat}}$ and
@@
 	$\gamma_{jd}^2 \sim \Gamma^{-1} (\phi_{jd}, \lambda_{jd})$, instead of
-	from the inverse Gaussian distribution. Additionally,
+	$\gamma_{jd}^2 \sim \mathcal{IG} (\phi_{jd}, \lambda_{jd})$. Additionally,
 	the bSEM model can
 	only incorporate one binary external covariate that divides the features into
-	two groups, i.e., $\bm{\alpha} = \begin{bmatrix} \alpha_0 & \alpha_1
+	two groups, i.e., $\bm{\alpha}_{\text{feat}} = \begin{bmatrix} \alpha_0 &
+	\alpha_1
 	\end{bmatrix} \tr$ and $\bm{c}_{jd} = \begin{bmatrix} 1 & 0 \end{bmatrix}
@@
              apply(res[rownames(res)=="pmse", ], 2, median))
-tab <- round(tab, 3)
-tab[which(tab[, 1]==max(tab[, 1], na.rm=TRUE)), 1] <- 
-  paste0("\\textbf{", tab[which(tab[, 1]==max(tab[, 1], na.rm=TRUE)), 1], "}")
-tab[which(tab[, 2]==min(tab[, 2], na.rm=TRUE)), 2] <- 
-  paste0("\\textbf{", tab[which(tab[, 2]==min(tab[, 2], na.rm=TRUE)), 2], "}")
-tab[which(tab[, 3]==min(tab[, 3], na.rm=TRUE)), 3] <- 
-  paste0("\\textbf{", tab[which(tab[, 3]==min(tab[, 3], na.rm=TRUE)), 3], "}")
-colnames(tab) <- c("$\\text{ELBO}_{\\text{test}}$",
+tabm <- tab
+tabm <- round(tabm, 3)
+tabm[which(tabm[, 1]==max(tabm[, 1], na.rm=TRUE)), 1] <- 
+  paste0("\\textbf{", tabm[which(tabm[, 1]==max(tabm[, 1], na.rm=TRUE)), 1], "}")
+tabm[which(tabm[, 2]==min(tabm[, 2], na.rm=TRUE)), 2] <- 
+  paste0("\\textbf{", tabm[which(tabm[, 2]==min(tabm[, 2], na.rm=TRUE)), 2], "}")
+tabm[which(tabm[, 3]==min(tabm[, 3], na.rm=TRUE)), 3] <- 
+  paste0("\\textbf{", tabm[which(tabm[, 3]==min(tabm[, 3], na.rm=TRUE)), 3], "}")
+colnames(tabm) <- c("$\\text{ELBO}_{\\text{test}}$",
                    "$\\text{ELBO}_{\\text{train}} 
@@
                    "$\\text{PMSE}_{\\text{test}}$")
-@
-  From the $\bm{\alpha}$ estimates in Tables \ref{tab:analysis_gdsc_est1_feat} 
-  and \ref{tab:analysis_gdsc_est1_feat} we see that there
-  seems to be an effect of the
-  external variables. Models NIG2 and NIG3 estimate a negative effect for the
-  pathway external covariate. This translates to a negative additive effect on
-  the expected prior precision of the $\bm{\beta}_d$, i.e., genes that are found
+
+library(glmnet)
+kurtosis <- function(x) {
+  n <- length(x) 
+  kurt <- (n + 1)*n*sum((x - mean(x))^4)/((n - 1)*(n - 2)*(n - 3)*var(x)^2) - 
+    3*(n - 1)^2/((n - 2)*(n - 3))
+  replace(kurt, !is.finite(kurt), NA)
+}
+load("results/analysis_gdsc_fit1.Rdata")
+best2.semnig <- fit2.semnig$vb$mu
+best1.lasso <- lapply(fit1.lasso, function(s) {
+  unname(coef(s, "lambda.min")[-1, ])})
+best1.ridge <- lapply(fit1.ridge, function(s) {
+  unname(coef(s, "lambda.min")[-1, ])})
+best1.bSEM <- lapply(fit1.bSEM$vb$beta, function(s) {s[, 2]})
+kurts <- c(mean(sapply(best2.semnig, kurtosis), na.rm=TRUE), 
+           mean(sapply(best1.lasso, kurtosis), na.rm=TRUE),
+           mean(sapply(best1.ridge, kurtosis), na.rm=TRUE), 
+           mean(sapply(best1.bSEM, kurtosis), na.rm=TRUE))
+ranges <- rbind(range(unlist(lapply(1:length(
+  fit2.semnig$eb$mpriord), function(d) {
+    fit2.semnig$eb$mpriord[d]*fit2.semnig$eb$mpriorf[[d]]}))),
+  fit1.bSEM$eb$b[nrow(fit1.bSEM$eb$b), ]/
+    (fit1.bSEM$eb$a[nrow(fit1.bSEM$eb$a), ] - 1))
+@
+  The non-zero $\bm{\alpha}$ estimates in Tables 
+  \ref{tab:analysis_gdsc_est1_feat} and \ref{tab:analysis_gdsc_est1_drug} show 
+  that there is an effect of the
+  external covariates. Models NIG2, NIG3, and bSEM estimate a negative effect 
+  for
+  the external pathway covariate. This translates to a negative additive effect 
+  on the expected prior precision of the $\bm{\beta}_d$, i.e., genes that are 
+  found
   in the target pathway of a drug are more predictive of IC50 than genes that
@@
   drug covariates, estimates a negative effect for the developmental drugs.
-
-  If we look at the NIG1 model, that includes all external 
-  covariates, we see that approved drugs are easier to predict, due to small
-  prior precision, followed by drugs that are in clinical development,
-  as we expected. The cytotoxic drugs are the hardest to predict, 
-  with the largest prior precision (smallest prior variance), followed by the 
-  targeted drugs and the unknown drugs. This is again according to expectation.
-  For the unkown drugs, we are unable to confirm the small precision. Genes 
-  that are in the target pathway have 
-  smaller prior precisions, according to expectation, which confirms that these
-  genes are more important for prediction of drug sensitivity. The estimates 
-  from
-  the bSEM model are also according to expectation: genes that are in the target
-  pathway have a much smaller precision than genes that are not in the target
-  pathway. The 
-  effects of the pathway indicator seem to be much larger in the bSEM model. 
+  The NIG2 model estimates the smallest prior $\bm{\beta}_d$ precision for the
+  targeted drugs, followed by cytotoxic drugs, while drugs with an unkown target
+  are assigned the largest prior $\bm{\beta}_d$ precision, according
+  to expectation. Somewhat surprisingly, the NIG4 model estimates the smallest
+  prior $\bm{\beta}_d$ precision for the cytotoxic drugs, compared to the 
+  targeted and unkown drugs. We conjecture that, due to absence of a local
+  feature-specific variance component, this model is not flexible 
+  enough to correctly capture the full covariance structure of the 
+  $\bm{\beta}_d$ and therefore incorreclty estimates $\bm{\alpha}$.
+  To see the total effect of the external covariates on the $\bm{\beta}_d$ 
+  prior, we calculated the range of the expected prior variances 
+  (up to a multiplicative error variance) as 
+  $\hat{E}(\gamma_{jd}^2 \cdot \tau_d^2) \in \[ \Sexpr{ranges[1, ]}\]$ and
+  $\hat{E}(\gamma_{jd}^2 \cdot \tau_d^2) \in \[ \Sexpr{ranges[2, ]}\]$ for
+  NIG2 and bSEM, respectively. This confirms that the estimated prior 
+  $\bm{\beta}_d$ variance components are in the same order of magnitude and that
+  the external covariates have a detectable effect on the prior.
 <<analysis_gdsc_est1_feat>>=
@@
       col.names=colnames(est1), align="r",
-      caption="$\\bm{\\alpha}_{\\text{feat}}$ estimates from GDSC data for feature selection strategy 1.",
+      caption="$\\bm{\\alpha}_{\\text{feat}}$ estimates from GDSC data.",
       format="latex", booktabs=TRUE, escape=FALSE) %>%
@@
       col.names=colnames(est2), align="r",
-      caption="$\\bm{\\alpha}_{\\text{drug}}$ estimates from GDSC data for feature selection strategy 1.",
+      caption="$\\bm{\\alpha}_{\\text{drug}}$ estimates from GDSC data.",
       format="latex", booktabs=TRUE, escape=FALSE) %>%
@@
 @
-
-  The median prediction MSEs and the median ELBOs, calculated on the test data
-  are displayed in Table \ref{tab:results_data}. In terms of 
-  is lowest in the NIG models 
-  (\Sexpr{round(rnorm(1))} and \Sexpr{rnorm(1)}, in the
-  NIG1 and NIG2 models, respectively). The bSEM model is not able to 
-  improve on the null model with an median prediction MSE of 
-  \Sexpr{round(rnorm(1))} compared to the null model's MSE of 
-  \Sexpr{round(rnorm(1))}.
+  The median PMSEs and ELBOs, calculated on the test data
+  are displayed in Table \ref{tab:analysis_gdsc_tab1}. In terms of prediction 
+  error, ridge outperforms the other models 
+  (PMSE of \Sexpr{round(tab[6, 3], 3)}), while bSEM 
+  (PMSE of \Sexpr{round(tab[7, 3], 3)}) performs worse than 
+  the null model (PMSE of \Sexpr{round(tab[8, 3], 3)}). Second and third best
+  performing models are NIG1 and NIG2 (PMSEs of \Sexpr{round(tab[1, 3], 3)} and
+  \Sexpr{round(tab[2, 3], 3)}, respectively), both outperforming the lasso
+  \Sexpr{round(tab[5, 3], 3)}. Part of the differences in performance between
+  NIG, lasso, and ridge may be explained with the different levels of 
+  ``sparsity'' in the solution. Although NIG and ridge do not give exactly 
+  sparse solutions (none of the $\bm{\beta}_d$ coefficients are set to zero), we
+  may use the ``heavy-tailedness'' of the distribution of the estimated 
+  $\bm{\beta}_d$ as a proxy for sparsity. A common measure of heavy-tailedness
+  is excess kurtosis, calculated as \Sexpr{round(kurts[1], 3)}, 
+  \Sexpr{round(kurts[2], 3)}, and \Sexpr{round(kurts[3], 3)} for the NIG2, 
+  lasso, and ridge estimates, respectively. Evidently, lasso gives the 
+  ``sparsest'' solution, followed by NIG and ridge. In some sense, this
+  means that the NIG uses less of the $\bm{\beta}_d$ to predict.
 <<analysis_gdsc_tab1>>=
-kable(tab,
-      col.names=colnames(tab), align="r",
-      caption="Performance measures from GDSC data for feature selection strategy 1.",
+kable(tabm,
+      col.names=colnames(tabm), align="r",
+      caption="Performance measures calculated on GDSC data (best performing model in bold).",
       format="latex", booktabs=TRUE, escape=FALSE) %>%
@@
 
-  In addition to the empirical Bayes estimates and the prediction MSE, we 
-  investigated the stability of the estimates by ranking the posterior means
-  $\E_{Q}(\bbeta_d)$ and calculating the Spearman's rank
-  correlation over the 50 data splits. They are depicted in Figure 
-  . In both NIG models the correlations
-  between splits are higher than in the bSEM model; an indication that 
-  estimation in the NIG model is more stable than in the bSEM model.
-
-<<analysis_gdsc_fit2>>=
-fit <- read.table("results/analysis_gdsc_fit2.txt", row.names=NULL)
-temp <- fit[, 1]
-fit <- as.matrix(fit[, -1])
-rownames(fit) <- temp
-
-res <- read.table("results/analysis_gdsc_res2.txt", row.names=NULL)
-temp <- res[, 1]
-res <- as.matrix(res[, -1])
-rownames(res) <- temp
-
-est1 <- cbind(fit[, 1], c(NA, 0, 0, NA), fit[, 2])
-colnames(est1) <- c("intercept", "not in pathway", "in pathway")
-est2 <- cbind(fit[, 3:5], c(NA, 0, NA, 0), fit[, c(7:6)], c(NA, 0, NA, 0))
-colnames(est2) <- c("intercept", "experimental", "development", "approved",
-                    "unkown", "targeted", "cytotoxic")
-
-tab <- cbind(apply(res[rownames(res)=="elbo", ], 2, median),
-             apply(res[rownames(res)=="elbot", ], 2, median) -
-               apply(res[rownames(res)=="elbo", ], 2, median),
-             apply(res[rownames(res)=="pmse", ], 2, median))
-tab <- round(tab, 3)
-tab[which(tab[, 1]==max(tab[, 1], na.rm=TRUE)), 1] <- 
-  paste0("\\textbf{", tab[which(tab[, 1]==max(tab[, 1], na.rm=TRUE)), 1], "}")
-tab[which(tab[, 2]==min(tab[, 2], na.rm=TRUE)), 2] <- 
-  paste0("\\textbf{", tab[which(tab[, 2]==min(tab[, 2], na.rm=TRUE)), 2], "}")
-tab[which(tab[, 3]==min(tab[, 3], na.rm=TRUE)), 3] <- 
-  paste0("\\textbf{", tab[which(tab[, 3]==min(tab[, 3], na.rm=TRUE)), 3], "}")
-colnames(tab) <- c("$\\text{ELBO}_{\\text{test}}$",
-                   "$\\text{ELBO}_{\\text{train}} 
-                   - \\text{ELBO}_{\\text{test}}$",
-                   "$\\text{PMSE}_{\\text{test}}$")  
-@
-
-<<analysis_gdsc_est2_feat>>=
-options(knitr.kable.NA="", digits=3)
-kable(est1,
-      col.names=colnames(est1), align="r",
-      caption="$\\bm{\\alpha}_{\\text{feat}}$ estimates from GDSC data for feature selection strategy 2.",
-      format="latex", booktabs=TRUE, escape=FALSE) %>%
-  kable_styling(latex_options=c("HOLD_position"))
-@
-
-<<analysis_gdsc_est2_drug>>=
-kable(est2,
-      col.names=colnames(est2), align="r",
-      caption="$\\bm{\\alpha}_{\\text{drug}}$ estimates from GDSC data for feature selection strategy 2.",
-      format="latex", booktabs=TRUE, escape=FALSE) %>%
-  kable_styling(latex_options=c("HOLD_position"))
-@
-
-<<analysis_gdsc_tab2>>=
-kable(tab,
-      col.names=colnames(tab), align="r",
-      caption="Performance measures from GDSC data for feature selection strategy 2.",
-      format="latex", booktabs=TRUE, escape=FALSE) %>%
-  kable_styling(latex_options=c("HOLD_position"))
-@
-
-<<analysis_gdsc_fit3>>=
-fit <- read.table("results/analysis_gdsc_fit3.txt", row.names=NULL)
-temp <- fit[, 1]
-fit <- as.matrix(fit[, -1])
-rownames(fit) <- temp
-
-res <- read.table("results/analysis_gdsc_res3.txt", row.names=NULL)
-temp <- res[, 1]
-res <- as.matrix(res[, -1])
-rownames(res) <- temp
-
-est1 <- cbind(fit[, 1], c(NA, 0, 0, NA), fit[, 2])
-colnames(est1) <- c("intercept", "not in pathway", "in pathway")
-est2 <- cbind(fit[, 3:5], c(NA, 0, NA, 0), fit[, c(7:6)], c(NA, 0, NA, 0))
-colnames(est2) <- c("intercept", "experimental", "development", "approved",
-                    "unkown", "targeted", "cytotoxic")
-
-tab <- cbind(apply(res[rownames(res)=="elbo", ], 2, median),
-             apply(res[rownames(res)=="elbot", ], 2, median) -
-               apply(res[rownames(res)=="elbo", ], 2, median),
-             apply(res[rownames(res)=="pmse", ], 2, median))
-tab <- round(tab, 3)
-tab[which(tab[, 1]==max(tab[, 1], na.rm=TRUE)), 1] <- 
-  paste0("\\textbf{", tab[which(tab[, 1]==max(tab[, 1], na.rm=TRUE)), 1], "}")
-tab[which(tab[, 2]==min(tab[, 2], na.rm=TRUE)), 2] <- 
-  paste0("\\textbf{", tab[which(tab[, 2]==min(tab[, 2], na.rm=TRUE)), 2], "}")
-tab[which(tab[, 3]==min(tab[, 3], na.rm=TRUE)), 3] <- 
-  paste0("\\textbf{", tab[which(tab[, 3]==min(tab[, 3], na.rm=TRUE)), 3], "}")
-colnames(tab) <- c("$\\text{ELBO}_{\\text{test}}$",
-                   "$\\text{ELBO}_{\\text{train}} 
-                   - \\text{ELBO}_{\\text{test}}$",
-                   "$\\text{PMSE}_{\\text{test}}$")  
-@
-
-<<analysis_gdsc_est3_feat>>=
-options(knitr.kable.NA="", digits=3)
-kable(est1,
-      col.names=colnames(est1), align="r",
-      caption="$\\bm{\\alpha}_{\\text{feat}}$ estimates from GDSC data for feature selection strategy 3.",
-      format="latex", booktabs=TRUE, escape=FALSE) %>%
-  kable_styling(latex_options=c("HOLD_position"))
-@
-
-<<analysis_gdsc_est3_drug>>=
-kable(est2,
-      col.names=colnames(est2), align="r",
-      caption="$\\bm{\\alpha}_{\\text{drug}}$ estimates from GDSC data for feature selection strategy 3.",
-      format="latex", booktabs=TRUE, escape=FALSE) %>%
-  kable_styling(latex_options=c("HOLD_position"))
-@
-
-<<analysis_gdsc_tab3>>=
-kable(tab,
-      col.names=colnames(tab), align="r",
-      caption="Performance measures from GDSC data for feature selection strategy 3.",
-      format="latex", booktabs=TRUE, escape=FALSE) %>%
-  kable_styling(latex_options=c("HOLD_position"))
-@
-
 	\section{Discussion}\label{sec:discussion}
+<<time>>=
+load("results/analysis_gdsc_fit1.Rdata")
+time2.semnig <- sum(fit2.semnig$time)
+@	
 	The preceding presents a novel model for drug sensitivity
@@
 	efficient and scales well with the number of features and samples.
+	Estimation of the NIG model that includes both external drugs and feature
+	covariates (NIG2 in Section \ref{sec:gdsc}) on the GDSC data from Section
+	\ref{sec:gdsc} took \Sexpr{round(time2.semnig, 2)} seconds on a 2016 
+	MacBook Pro with a 2 GHz Dual-Core Intel Core i5 processor and
+	8 GB of 1867 MHz LPDDR3 memory, running macOS  10.15.1.
 
@@
 	\ref{sec:gdsc} show that the inclusion of the external covariates is 
-	beneficial for drug sensitivity prediction. Although the prediction MSE is
-	not lower if the external covariates are included, 
-	the resulting hyperparameter estimates are still informative
-	and may guide future research efforts.
-
-	The comparison to the bSEM model that allows for the inclusion of 
+	beneficial for drug sensitivity prediction. Prediction MSE of the NIG
+	model with both both feature and drug variance components is lower
+	than for the regular lasso model and although the prediction MSE is
+	not lower if external covariates are included, the resulting hyperparameter 
+	estimates are still informative and may guide future research efforts.
+	In addition, the comparison to the bSEM model that allows for the inclusion of 
 	one binary external feature covariate favours the proposed NIG model. 
 	Prediction MSEs are almost uniformly lower over the drugs for the NIG model 
-	compared to bSEM. Consequently, the average prediction MSE is significantly
+	compared to bSEM. Consequently, the total prediction MSE is significantly
 	lower for the NIG model.
-
-	Although the empirical Bayes estimates of the external covariate effects 
-	$\bm{\alpha}$ are informative, they are small compared to 
-	bSEM's estimates. Consequently, the posteriors of the models with and without
-	the external covariates (NIG2 and NIG1 in Section \ref{sec:gdsc}) do not 
-	differ by much. This results similar prediction MSEs with
-	the inclusion of the external covariates. This is a topic that deserves 
-	further attention. One proposed solution is to include a
-	prior variance component that promotes a more heavy-tailed $\beta_{jd}$ prior
-	distribution. An example is a variance component with a half-cauchy prior 
-	distribution, i.e., the horseshoe \cite[]{carvalho_handling_2009}. 
-	This solution is similar to the pInc model from 
-	\cite{kpogbezan_incorporating_2019}. 
-	Like the bSEM model, the pInc model gives significantly larger external
-	covariate effects, 
-	so we conjecture that the inclusion of a similar heavy-tailed prior variance 
-	component will have a similar effect in the NIG model.
 

### 2019-11-27 code/figures.R

@@ -160,29 +160,6 @@ points(c(1:4), chi, pch=2, col=col[c(1:4)], cex=1.5,
        cex.lab=1.5, cex.axis=1.5)
 par(opar)
 
-# ---- analysis_gdsc_est1 ----
-res <- read.table("results/analysis_gdsc_fit1.txt", row.names=NULL)
-temp <- res[, 1]
-res <- as.matrix(res[, -1])
-rownames(res) <- temp
-boxplot(res[rownames(res)=="pmse", ])
-boxplot(res[rownames(res)=="elbo", ])
-boxplot(res[rownames(res)=="elbot", ])
-
-
-
-boxplot(res[rownames(res)=="brankdist", ])
-
-# ---- data_gdsc_res1_brank_hist ----
-# library(sp)
-# load("results/data_gdsc_cv1.Rdata")
-# col <- bpy.colors(ncol(brankcor), cutoff.tail=0.3)
-# opar <- par(no.readonly=TRUE)
-# par(mar=opar$mar*c(1, 1.1, 1, 1))
-# boxplot(brankcor, ylab=expression("Spearman's"~rho), col=col)
-# par(opar)
-hist(rnorm(10))
-
 ################################# presentation #################################
 # ---- dens_beta_prior1 ---- 
 library(GeneralizedHyperbolic)

### 2019-11-27 code/simulations_gdsc.R

@@ -32,11 +32,7 @@ shape <- 3
 rate <- 2
 
 nreps <- 50
-methods <- c("NIG1", "NIG2", "lasso", "ridge")
-#
-# pmse <- pmset <- emse <- matrix(NA, nrow=nreps, ncol=length(methods) + 1)
-# est <- matrix(NA, nrow=5*nreps, ncol=length(methods) - 2)
-# elbo <- elbot <- matrix(NA, nrow=nreps, ncol=length(methods) - 2)
+methods <- c("NIG$_{\text{f}}^-$", "NIG$_{\text{f}}$", "lasso", "ridge")
 
 # setup cluster
 cl <- makeCluster(ncores)
@@ -104,40 +100,6 @@ res <- foreach(r=1:nreps, .packages=packages) %dopar% {
   predt1.lasso <- sapply(1:D, function(d) {xtrain[[d]] %*% best1.lasso[[d]]})
   predt1.ridge <- sapply(1:D, function(d) {xtrain[[d]] %*% best1.ridge[[d]]})
 
-  # emse[r, 1] <- mean(sapply(1:D, function(d) {
-  #   mean((best1.semnig[[d]] - beta[[d]])^2)}))
-  # emse[r, 2] <- mean(sapply(1:D, function(d) {
-  #   mean((best2.semnig[[d]] - beta[[d]])^2)}))
-  # emse[r, 3] <- mean(sapply(1:D, function(d) {
-  #   mean((best1.lasso[[d]] - beta[[d]])^2)}))
-  # emse[r, 4] <- mean(sapply(1:D, function(d) {
-  #   mean((best1.ridge[[d]] - beta[[d]])^2)}))
-  # emse[r, 5] <- mean(unlist(beta)^2)
-  #
-  # pmse[r, 1] <- mean(colMeans((pred1.semnig - ytest)^2))
-  # pmse[r, 2] <- mean(colMeans((pred2.semnig - ytest)^2))
-  # pmse[r, 3] <- mean(colMeans((pred1.lasso - ytest)^2))
-  # pmse[r, 4] <- mean(colMeans((pred1.ridge - ytest)^2))
-  # pmse[r, 5] <- mean(apply(ytest, 1, "-", colMeans(ytrain))^2)
-  #
-  # pmset[r, 1] <- mean(colMeans((predt1.semnig - ytrain)^2))
[…]

### 2019-11-27 docs/biometrical_main.Rnw
git before old checkout

@@
   \footnote{Corresponding author: {\sf{e-mail: m.munch@amsterdamumc.nl}}, 
-  Phone: +31 71 527 4844}\inst{,1,2}} 
+  Phone: +31 71 527 4844}\inst{,1,2,3}} 
   \author[van de Wiel]{Mark A. van de Wiel\inst{1,3}}
@@
 	\section*{Acknowledgements}
+	A visit of MM to SR and GL was funded by the Amsterdam Public Health 
+	institute's Methodology travel grant.
 

### 2019-11-27 docs/manuscript1.Rnw
git before old checkout

@@
 \date{\today}
-\author{Magnus M. M\"unch$^{1,2}$\footnote{Correspondence to: 
+\author{Magnus M. M\"unch$^{1,2,3}$\footnote{Correspondence to: 
 \href{mailto:m.munch@amsterdamumc.nl}{m.munch@amsterdamumc.nl}}, Mark A. van de 
@@
 	\section*{Acknowledgements}
-	We thank Zhi Zhao and Leiv Tore Salte R{\o}nneberg from the Oslo Centre for 
-	Biostatistics and Epidemiology for their suggestions on the pre-processing of 
-	the GDSC data. \textit{Conflict of Interest}: None declared.
+	% We thank Zhi Zhao and Leiv Tore Salte R{\o}nneberg from the Oslo Centre for 
+	% Biostatistics and Epidemiology for their suggestions on the pre-processing of 
+	% the GDSC data. 
+	A visit of MM to SR and GL was funded by the Amsterdam Public Health 
+	institute's Methodology travel grant.
+	\textit{Conflict of Interest}: None declared.
 

### 2019-11-27 code/analysis_gdsc.R

@@ -157,7 +157,7 @@ res <- foreach(r=1:nreps, .packages=packages) %dopar% {
   cv1.bSEM <- bSEM(xtrain, split(ytrain, 1:ncol(ytrain)),
                    lapply(inpathway, "+", 1),
                    control=list(maxit=200, trace=FALSE, epsilon=1e-3))
-  
+
   # penalized regression models
   cv1.lasso <- lapply(1:D, function(d) {
     cv.glmnet(xtrain[[d]], ytrain[, d], intercept=FALSE, standardize=FALSE)})

### 2019-11-27 docs/manuscript1.Rnw
changed methods in simulations_gdsc.R

@@
 	(ii) the NIG model estimated as in Section \ref{sec:estimation} that includes
-	the all external covariates, called 
-	NIG$_{\text{f}}$, NIG$_{\text{d}}$, or
+	all external covariates, called NIG$_{\text{f}}$, NIG$_{\text{d}}$, or
 	NIG$_{\text{f}+\text{d}}$, again depending on which variance components are 
@@
   Figure \ref{fig:simulations_gdsc_est1} shows the estimated
-	$\bm{\alpha}_{\text{feat}}$ together with its true value for NIG2 in 
-	version 1 of the simulation study. Figure 
+	$\bm{\alpha}_{\text{feat}}$ together with its true value for NIG$_{\text{f}}$ 
+	in version 1 of the simulation study. Figure 
 	\ref{fig:simulations_gdsc_est1}a shows that 
@@
 
-	Model NIG1 gives a median $\phi$ estimate of \Sexpr{round(est.phi, 2)},  
+	Model NIG$_{\text{f}}^-$ gives a median $\phi$ estimate of 
+	\Sexpr{round(est.phi, 2)},  
 	equal to the true mean of the $\phi_{jd}$, \Sexpr{round(true.phi, 2)}. 

### 2019-11-27 docs/manuscript1.Rnw
fixed small mistake in simulations_gdsc.R

@@
 
-	\section{Introduction}
+	\section{Introduction}\label{sec:introduction}
 	Recently, promising results in precision medicine 
@@
 	NIG$_{\text{f}+\text{d}}^-$, depending on which variance components are 
-	estimated,
+	estimated (feature, drug, or both),
 	(ii) the NIG model estimated as in Section \ref{sec:estimation} that includes
@@
 est.lambdaf <- apply(res[rownames(res)=="lambdaf", c(1, 2)], 2, median)
-tab <- cbind(apply(res[rownames(res)=="elbo", ], 2, median),
-             apply(res[rownames(res)=="elbot", ], 2, median) -
-               apply(res[rownames(res)=="elbo", ], 2, median),
-             apply(res[rownames(res)=="emse", ], 2, median),
-             apply(res[rownames(res)=="pmse", ], 2, median),
-             apply(res[rownames(res)=="pmse", ], 2, median) - 
-               apply(res[rownames(res)=="pmset", ], 2, median))
+tab <- cbind(apply(res[rownames(res)=="emse", ], 2, median),
+             apply(res[rownames(res)=="pmse", ], 2, median))
 tab <- round(tab, 3)
-tab[which(tab[, 1]==max(tab[, 1], na.rm=TRUE)), 1] <- 
-  paste0("\\textbf{", tab[which(tab[, 1]==max(tab[, 1], na.rm=TRUE)), 1], "}")
+tab[which(tab[, 1]==min(tab[, 1], na.rm=TRUE)), 1] <- 
+  paste0("\\textbf{", tab[which(tab[, 1]==min(tab[, 1], na.rm=TRUE)), 1], "}")
 tab[which(tab[, 2]==min(tab[, 2], na.rm=TRUE)), 2] <- 
   paste0("\\textbf{", tab[which(tab[, 2]==min(tab[, 2], na.rm=TRUE)), 2], "}")
-tab[which(tab[, 3]==min(tab[, 3], na.rm=TRUE)), 3] <- 
-  paste0("\\textbf{", tab[which(tab[, 3]==min(tab[, 3], na.rm=TRUE)), 3], "}")
-tab[which(tab[, 4]==min(tab[, 4], na.rm=TRUE)), 4] <- 
-  paste0("\\textbf{", tab[which(tab[, 4]==min(tab[, 4], na.rm=TRUE)), 4], "}")
-tab[which(tab[, 5]==min(tab[, 5], na.rm=TRUE)), 5] <- 
-  paste0("\\textbf{", tab[which(tab[, 5]==min(tab[, 5], na.rm=TRUE)), 5], "}")
-colnames(tab) <- c("$\\text{ELBO}_{\\text{test}}$",
-                   "$\\text{ELBO}_{\\text{train}} 
-                   - \\text{ELBO}_{\\text{test}}$",
-                   "$\\text{EMSE}_{\\text{test}}$",
-                   "$\\text{PMSE}_{\\text{test}}$",
-                   "$\\text{PMSE}_{\\text{test}} 
-                   - \\text{PMSE}_{\\text{train}}$")     
+colnames(tab) <- c("$\\text{EMSE}_{\\text{test}}$",
+                   "$\\text{PMSE}_{\\text{test}}$")     
 @	
@@
 
-	Table \ref{tab:simulations_gdsc_tab1} displays several median performance
-	measures for simulation version 1. The table includes as a reference 
-	the null model, which
+	Table \ref{tab:simulations_gdsc_tab1} displays median estimation MSE (EMSE) 
+	and prediction MSE (PMSE) for simulation version 1. The table includes as a
+	reference the null model, which
 	is an empty model that only fits the means of the $\y_{d}$.
-	The median evidence lower bounds (ELBO) approximates the marginal
-	likelihoods of models NIG1 and NIG2 on both training and test data, while 
-	estimation MSE (EMSE) and prediction MSE (PMSE) give measures of predictive 
-	performance. Focussing on MSE, we see that the NIG2 model, that includes
-	the external covariates, performs best, both in terms of estimation and 
-	prediction. Model fit is best for the NIG1 model that does not include the
-	external covariates, according to the ELBO, calculated on the test data.
-	The lower MSEs confirm that the model that
-	includes the external data (NIG2) learns the underlying structure 
-	in the data better than model NIG1 that does not include the external 
-	data.
+	Estimation MSE (EMSE) and prediction MSE (PMSE) give measures of predictive 
+	performance. Focussing on MSE, we see that the NIG$_{\text{f}}$ model, 
+	that includes the external covariates, performs best, both in terms of 
+	estimation and prediction. The lower MSEs confirm that the model that
+	includes the external data (NIG$_{\text{f}}$) learns the underlying structure 
+	in the data better than model NIG$_{\text{f}}^-$ that does not include the
+	external data.
 <<simulations_gdsc_tab1>>=
@@
 kable(tab, col.names=colnames(tab), align="r",
-      caption="Median performance measure estimates for simulation version 1 (best performing model in bold).", 
+      caption="Median performance measure estimates for simulation version 1 (lowest in bold).", 
       format="latex", booktabs=TRUE, escape=FALSE) %>%
@@
 est.lambdad <- apply(res[rownames(res)=="lambdad", c(1, 2)], 2, median)
-tab <- cbind(apply(res[rownames(res)=="elbo", ], 2, median),
-             apply(res[rownames(res)=="elbot", ], 2, median) -
-               apply(res[rownames(res)=="elbo", ], 2, median),
-             apply(res[rownames(res)=="emse", ], 2, median),
-             apply(res[rownames(res)=="pmse", ], 2, median),
-             apply(res[rownames(res)=="pmse", ], 2, median) - 
-               apply(res[rownames(res)=="pmset", ], 2, median))
+tab <- cbind(apply(res[rownames(res)=="emse", ], 2, median),
+             apply(res[rownames(res)=="pmse", ], 2, median))
 tab <- round(tab, 3)
-tab[which(tab[, 1]==max(tab[, 1], na.rm=TRUE)), 1] <- 
-  paste0("\\textbf{", tab[which(tab[, 1]==max(tab[, 1], na.rm=TRUE)), 1], "}")
+tab[which(tab[, 1]==min(tab[, 1], na.rm=TRUE)), 1] <- 
+  paste0("\\textbf{", tab[which(tab[, 1]==min(tab[, 1], na.rm=TRUE)), 1], "}")
 tab[which(tab[, 2]==min(tab[, 2], na.rm=TRUE)), 2] <- 
   paste0("\\textbf{", tab[which(tab[, 2]==min(tab[, 2], na.rm=TRUE)), 2], "}")
-tab[which(tab[, 3]==min(tab[, 3], na.rm=TRUE)), 3] <- 
-  paste0("\\textbf{", tab[which(tab[, 3]==min(tab[, 3], na.rm=TRUE)), 3], "}")
-tab[which(tab[, 4]==min(tab[, 4], na.rm=TRUE)), 4] <- 
-  paste0("\\textbf{", tab[which(tab[, 4]==min(tab[, 4], na.rm=TRUE)), 4], "}")
-tab[which(tab[, 5]==min(tab[, 5], na.rm=TRUE)), 5] <- 
-  paste0("\\textbf{", tab[which(tab[, 5]==min(tab[, 5], na.rm=TRUE)), 5], "}")
-colnames(tab) <- c("$\\text{ELBO}_{\\text{test}}$",
-                   "$\\text{ELBO}_{\\text{train}} 
-                   - \\text{ELBO}_{\\text{test}}$",
-                   "$\\text{EMSE}_{\\text{test}}$",
-                   "$\\text{PMSE}_{\\text{test}}$",
-                   "$\\text{PMSE}_{\\text{test}} 
-                   - \\text{PMSE}_{\\text{train}}$")     
+colnames(tab) <- c("$\\text{EMSE}_{\\text{test}}$",
+                   "$\\text{PMSE}_{\\text{test}}$")     
 @	
   Figure \ref{fig:simulations_gdsc_est2} shows the estimated
-	$\bm{\alpha}_{\text{drug}}$ together with its true value for NIG2 in 
-	version 2 of the simulation study. They are estimated without bias.
+	$\bm{\alpha}_{\text{drug}}$ together with its true value for NIG$_{\text{d}}$ 
+	in version 2 of the simulation study. They are estimated without bias.
 	Median $\lambda_{\text{drug}}=\Sexpr{round(est.lambdad[2], 2)}$ is slightly 
 	overestimated compared to true value 1. 
-	The median $\chi$ estimate in the NIG1 model is \Sexpr{round(est.chi, 2)},  
+	The median $\chi$ estimate in the NIG$_{\text{d}}^-$ model is 
+	\Sexpr{round(est.chi, 2)},  
 	which is almost equal to the true mean \Sexpr{round(true.chi, 2)}. 
@@
 	Median performance measures for simulation version 2 are displayed in Table
-	\ref{tab:simulations_gdsc_tab2}. Again, we see that NIG2 performs best in 
-	terms of MSE. However, NIG1 performance is the same here, indicating that 
-	the external covariates do not add much here.
-	Model fit is again best for the NIG1 model.
+	\ref{tab:simulations_gdsc_tab2}. Again, we see that NIG$_{\text{d}}$ performs
+	best in terms of MSE. However, NIG$_{\text{d}}^-$ performance is the same 
+	here, indicating that the external covariates do not add much here.
 <<simulations_gdsc_est2, fig.cap="Estimated (a) hyperparameters $\\bm{\\alpha}_{\\text{feat}}$ and (b) prior means $\\bm{\\phi}_{jd}$ from simulation version 2 with fixed $\\gamma_{jd}^2$.", out.width="100%", fig.asp=1/2>>=
@@
 
-tab <- cbind(apply(res[rownames(res)=="elbo", ], 2, median),
-             apply(res[rownames(res)=="elbot", ], 2, median) -
-               apply(res[rownames(res)=="elbo", ], 2, median),
-             apply(res[rownames(res)=="emse", ], 2, median),
-             apply(res[rownames(res)=="pmse", ], 2, median),
-             apply(res[rownames(res)=="pmse", ], 2, median) - 
-               apply(res[rownames(res)=="pmset", ], 2, median))
+tab <- cbind(apply(res[rownames(res)=="emse", ], 2, median),
+             apply(res[rownames(res)=="pmse", ], 2, median))
 tab <- round(tab, 3)
-tab[which(tab[, 1]==max(tab[, 1], na.rm=TRUE)), 1] <- 
-  paste0("\\textbf{", tab[which(tab[, 1]==max(tab[, 1], na.rm=TRUE)), 1], "}")
+tab[which(tab[, 1]==min(tab[, 1], na.rm=TRUE)), 1] <- 
+  paste0("\\textbf{", tab[which(tab[, 1]==min(tab[, 1], na.rm=TRUE)), 1], "}")
 tab[which(tab[, 2]==min(tab[, 2], na.rm=TRUE)), 2] <- 
   paste0("\\textbf{", tab[which(tab[, 2]==min(tab[, 2], na.rm=TRUE)), 2], "}")
-tab[which(tab[, 3]==min(tab[, 3], na.rm=TRUE)), 3] <- 
-  paste0("\\textbf{", tab[which(tab[, 3]==min(tab[, 3], na.rm=TRUE)), 3], "}")
-tab[which(tab[, 4]==min(tab[, 4], na.rm=TRUE)), 4] <- 
-  paste0("\\textbf{", tab[which(tab[, 4]==min(tab[, 4], na.rm=TRUE)), 4], "}")
-tab[which(tab[, 5]==min(tab[, 5], na.rm=TRUE)), 5] <- 
-  paste0("\\textbf{", tab[which(tab[, 5]==min(tab[, 5], na.rm=TRUE)), 5], "}")
-colnames(tab) <- c("$\\text{ELBO}_{\\text{test}}$",
-                   "$\\text{ELBO}_{\\text{train}} 
-                   - \\text{ELBO}_{\\text{test}}$",
-                   "$\\text{EMSE}_{\\text{test}}$",
-                   "$\\text{PMSE}_{\\text{test}}$",
-                   "$\\text{PMSE}_{\\text{test}} 
-                   - \\text{PMSE}_{\\text{train}}$")     
+colnames(tab) <- c("$\\text{EMSE}_{\\text{test}}$",
+                   "$\\text{PMSE}_{\\text{test}}$")     
 @	
   In Figure \ref{fig:simulations_gdsc_est3} the 
-	$\bm{\alpha}_{\text{feat}}$ and $\bm{\alpha}_{\text{drug}}$ estimated by NIG2
+	$\bm{\alpha}_{\text{feat}}$ and $\bm{\alpha}_{\text{drug}}$ estimated by 
+	NIG$_{\text{f}+\text{d}}$
 	are displayed together with their true values for
@@
 	and \Sexpr{round(est.lambdad[2], 2)}, respectively, compared to their true
-	value 1. The NIG1 model is also consistently over- and underestimating $\phi$
+	value 1. The NIG$_{\text{f}+\text{d}}^-$ model is also consistently over- 
+	and underestimating $\phi$
 	and $\chi$ with estimates \Sexpr{round(est.phi, 2)} and 
 	\Sexpr{round(est.chi, 2)}, respectively (compared to the true mean 
-	\Sexpr{round(true.chi, 2)}). NIG1 also over- and underestimates 
+	\Sexpr{round(true.chi, 2)}). NIG$_{\text{f}+\text{d}}^-$ also over- and
+	underestimates 
 	$\lambda_{\text{feat}}$ and $\lambda_{\text{drug}}$ with median estimates
@@
 	\ref{tab:simulations_gdsc_tab3}, with median performance measures, shows
-	that NIG2 outperforms the other models in terms of prediction and estimation
-	MSE.
+	that NIG$_{\text{f}+\text{d}}$ outperforms the other models in terms of
+	prediction and estimation MSE.
 <<simulations_gdsc_est3, fig.cap="Estimated (a) hyperparameters $\\bm{\\alpha}_{\\text{feat}}$ and (b) prior means $\\bm{\\phi}_{jd}$ from simulation version 3.", out.width="100%", fig.asp=1>>= 
@@
 	models are: four models like in (\ref{eq:prior}) but with different external 
-	variables, named NIG1 and NIG2 in the following. NIG1 does 
-	not use the external covariates and estimates
-	one overall $\phi$ and one overall $\chi$, while NIG2 uses the external 
-	covariates to estimate both $\bm{\alpha}_{\text{feat}}$ and
-	$\bm{\alpha}_{\text{drug}}$, while NIG3 and NIG4 estimate only 
-	$\bm{\alpha}_{\text{feat}}$ or $\bm{\alpha}_{\text{drug}}$, respectively,
+	variables, named NIG$_{\text{f}+\text{d}}^-$, NIG$_{\text{f}+\text{d}}$,
+	NIG$_{\text{f}}$, and NIG$_{\text{d}}$ in the following. The nomenclature is
+	the same as in Section \ref{sec:simulations}, so NIG$_{\text{f}+\text{d}}^-$
+	does not use the external covariates and estimates
+	one overall $\phi$ and one overall $\chi$, while NIG$_{\text{f}+\text{d}}$ 
+	uses the external covariates to estimate both $\bm{\alpha}_{\text{feat}}$ and
+	$\bm{\alpha}_{\text{drug}}$. NIG$_{\text{f}}$ and NIG$_{\text{d}}$ estimate 
+	only $\bm{\alpha}_{\text{feat}}$ or $\bm{\alpha}_{\text{drug}}$, respectively,
 	keeping either $\forall d: \tau_d^2=1$ or $\forall j,d: \gamma_{jd}^2=1$
-	fixed.
+	fixed. In the language of \cite{bernardo_shrink_2011} as introduced in 
+	Section \ref{sec:introduction}, models NIG$_{\text{f}}$ and NIG$_{\text{d}}$
+	may be described as local and global shrinkage rules, respectively, as 
+	opposed to the global-local shrinkage models NIG$_{\text{f}+\text{d}}$ and
+	NIG$_{\text{f}+\text{d}}^-$.
 
@@
 est1 <- cbind(fit[, 1], c(NA, 0, 0, NA, 0), fit[, 2])
+est1 <- round(est1, 2)
+est1[is.na(est1)] <- "-"
 colnames(est1) <- c("intercept", "not in pathway", "in pathway")
@@
               c(NA, 0, NA, 0, NA))
+est2 <- round(est2, 2)
+est2[is.na(est2)] <- "-"
 colnames(est2) <- c("intercept", "experimental", "development", "approved",
@@
 
-tab <- cbind(apply(res[rownames(res)=="elbo", ], 2, median),
-             apply(res[rownames(res)=="elbot", ], 2, median) -
-               apply(res[rownames(res)=="elbo", ], 2, median),
-             apply(res[rownames(res)=="pmse", ], 2, median))
+tab <- cbind(apply(res[rownames(res)=="pmse", ], 2, median))
 tabm <- tab
 tabm <- round(tabm, 3)
-tabm[which(tabm[, 1]==max(tabm[, 1], na.rm=TRUE)), 1] <- 
-  paste0("\\textbf{", tabm[which(tabm[, 1]==max(tabm[, 1], na.rm=TRUE)), 1], "}")
-tabm[which(tabm[, 2]==min(tabm[, 2], na.rm=TRUE)), 2] <- 
-  paste0("\\textbf{", tabm[which(tabm[, 2]==min(tabm[, 2], na.rm=TRUE)), 2], "}")
-tabm[which(tabm[, 3]==min(tabm[, 3], na.rm=TRUE)), 3] <- 
-  paste0("\\textbf{", tabm[which(tabm[, 3]==min(tabm[, 3], na.rm=TRUE)), 3], "}")
-colnames(tabm) <- c("$\\text{ELBO}_{\\text{test}}$",
-                   "$\\text{ELBO}_{\\text{train}} 
-                   - \\text{ELBO}_{\\text{test}}$",
-                   "$\\text{PMSE}_{\\text{test}}$")
+tabm[which(tabm[, 1]==min(tabm[, 1], na.rm=TRUE)), 1] <- 
+  paste0("\\textbf{", tabm[which(tabm[, 1]==min(tabm[, 1], na.rm=TRUE)), 1], "}")
+colnames(tabm) <- c("$\\text{PMSE}_{\\text{test}}$")
 
@@
   that there is an effect of the
-  external covariates. Models NIG2, NIG3, and bSEM estimate a negative effect 
-  for
+  external covariates. Models NIG$_{\text{f}+\text{d}}$, NIG$_{\text{f}}$, and
+  bSEM estimate a negative effect for
   the external pathway covariate. This translates to a negative additive effect 
@@
   in the target pathway of a drug are more predictive of IC50 than genes that
-  are not in the target pathway, according to expectation. Furthermore the NIG2
+  are not in the target pathway, according to expectation. Furthermore the 
+  NIG$_{\text{f}+\text{d}}$
   model estimates positive effects for the experimental and developmental drugs,
@@
   approved drugs are easier to predict than experimental and developmental 
-  drugs. On the other hand, the NIG4 model, that only includes external
+  drugs. On the other hand, the NIG$_{\text{d}}$ model, that only includes
+  external
   drug covariates, estimates a negative effect for the developmental drugs.
-  The NIG2 model estimates the smallest prior $\bm{\beta}_d$ precision for the
+  The NIG$_{\text{f}+\text{d}}$ model estimates the smallest prior 
+  $\bm{\beta}_d$ precision for the
   targeted drugs, followed by cytotoxic drugs, while drugs with an unkown target
   are assigned the largest prior $\bm{\beta}_d$ precision, according
-  to expectation. Somewhat surprisingly, the NIG4 model estimates the smallest
+  to expectation. Somewhat surprisingly, the NIG$_{\text{d}}$ model
+  estimates the smallest
   prior $\bm{\beta}_d$ precision for the cytotoxic drugs, compared to the 
@@
   $\hat{E}(\gamma_{jd}^2 \cdot \tau_d^2) \in \[ \Sexpr{ranges[2, ]}\]$ for
-  NIG2 and bSEM, respectively. This confirms that the estimated prior 
+  NIG$_{\text{f}+\text{d}}$ and bSEM, respectively. This confirms that the
+  estimated prior 
   $\bm{\beta}_d$ variance components are in the same order of magnitude and that
@@
   error, ridge outperforms the other models 
-  (PMSE of \Sexpr{round(tab[6, 3], 3)}), while bSEM 
-  (PMSE of \Sexpr{round(tab[7, 3], 3)}) performs worse than 
-  the null model (PMSE of \Sexpr{round(tab[8, 3], 3)}). Second and third best
-  performing models are NIG1 and NIG2 (PMSEs of \Sexpr{round(tab[1, 3], 3)} and
-  \Sexpr{round(tab[2, 3], 3)}, respectively), both outperforming the lasso
-  \Sexpr{round(tab[5, 3], 3)}. Part of the differences in performance between
+  (PMSE of \Sexpr{round(tab[6, 1], 3)}), while bSEM 
+  (PMSE of \Sexpr{round(tab[7, 1], 3)}) performs worse than 
+  the null model (PMSE of \Sexpr{round(tab[8, 1], 3)}). Second and third best
+  performing models are NIG$_{\text{f}+\text{d}}^-$ and 
+  NIG$_{\text{f}+\text{d}}$ (PMSEs of \Sexpr{round(tab[1, 1], 3)} and
+  \Sexpr{round(tab[2, 1], 3)}, respectively), both outperforming the lasso
+  \Sexpr{round(tab[5, 1], 3)}. The underperformance of the 
+  NIG$_{\text{f}}$ and NIG$_{\text{d}}$ models shows the usefulness of a
+  global-local shrinkage prior as opposed to only global or local shrinkage as 
+  as in models NIG$_{\text{f}}$ and NIG$_{\text{d}}$.
+<<analysis_gdsc_tab1>>=
+kable(tabm,
+      col.names=colnames(tabm), align="r",
+      caption="Performance measures calculated on GDSC data (best performing model in bold).",
+      format="latex", booktabs=TRUE, escape=FALSE) %>%
+  kable_styling(latex_options=c("HOLD_position"))
+@
+
+  Part of the differences in performance between
   NIG, lasso, and ridge may be explained with the different levels of 
@@
   is excess kurtosis, calculated as \Sexpr{round(kurts[1], 3)}, 
-  \Sexpr{round(kurts[2], 3)}, and \Sexpr{round(kurts[3], 3)} for the NIG2, 
+  \Sexpr{round(kurts[2], 3)}, and \Sexpr{round(kurts[3], 3)} for the 
+  NIG$_{\text{f}+\text{d}}$, 
   lasso, and ridge estimates, respectively. Evidently, lasso gives the 
   ``sparsest'' solution, followed by NIG and ridge. In some sense, this
-  means that the NIG uses less of the $\bm{\beta}_d$ to predict.
-<<analysis_gdsc_tab1>>=
-kable(tabm,
-      col.names=colnames(tabm), align="r",
-      caption="Performance measures calculated on GDSC data (best performing model in bold).",
-      format="latex", booktabs=TRUE, escape=FALSE) %>%
-  kable_styling(latex_options=c("HOLD_position"))
-@
+  means that the NIG uses less of the $\bm{\beta}_d$ to predict than the ridge
+  model. In addition, the penalized regression methods estimate penalty 
+  parameters by cross-validation. Cross-validation directly minimises the 
+  (approximate) PMSE, as opposed to empirical Bayes in the NIG that maximises 
+  the (approximate) marginal likelihood, a measure of model fit. The superior 
+  PMSE of ridge regression is therefore not surprising. A caveat with penalized
+  regression methods is that they do not give measures of parameter uncertainty.
+  NIG, on the other hand, gives the full posterior of the parameters, either
+  through a variational Bayes approximation or with the Gibbs sampler from
+  SM Section \ref{sm-sec:gibbssampler}. The full posterior gives direct access
+  to the parameter uncertainties for a better interpretable model.
 
@@
 	Estimation of the NIG model that includes both external drugs and feature
-	covariates (NIG2 in Section \ref{sec:gdsc}) on the GDSC data from Section
+	covariates (NIG$_{\text{f}+\text{d}}$ in Section \ref{sec:gdsc}) on the GDSC
+	data from Section
 	\ref{sec:gdsc} took \Sexpr{round(time2.semnig, 2)} seconds on a 2016 
@@
 	\ref{sec:gdsc} show that the inclusion of the external covariates is 
-	beneficial for drug sensitivity prediction. Prediction MSE of the NIG
+	beneficial for drug sensitivity prediction. PMSE of the NIG
 	model with both both feature and drug variance components is lower
-	than for the regular lasso model and although the prediction MSE is
+	than for the regular lasso model and although the PMSE is
 	not lower if external covariates are included, the resulting hyperparameter 
@@
 	one binary external feature covariate favours the proposed NIG model. 
-	Prediction MSEs are almost uniformly lower over the drugs for the NIG model 
-	compared to bSEM. Consequently, the total prediction MSE is significantly
+	PMSEs are almost uniformly lower over the drugs for the NIG model 
+	compared to bSEM. Consequently, the total PMSE is significantly
 	lower for the NIG model.
+
+	A possible direction of future research that attempts 
+	to lower PMSE of the NIG model is the introduction of a cross-validated global
+	variance component in the prior of the $\bm{\beta}_d$. If we add the 
+	additional constraint that the drug and feature variance components have to 
+	average to one, this cross-validated variance component may be interpreted as
+	the overall level of shrinkage that is then tailored to lower PMSE.
 

### 2019-11-27 code/simulations_gdsc.R

@@ -34,6 +34,7 @@ rate <- 2
 nreps <- 50
 methods <- c("NIG$_{\text{f}}^-$", "NIG$_{\text{f}}$", "lasso", "ridge")
 
+nreps <- ncores <- 1
 # setup cluster
 cl <- makeCluster(ncores)
 if(ncores > 1) {
@@ -163,8 +164,8 @@ res <- foreach(r=1:nreps, .packages=packages) %dopar% {
   elbo <- c(mean(new.elbo(fit1.semnig, xtest, ytest)),
             mean(new.elbo(fit2.semnig, xtest, ytest)))
 
-  list(emse=emse, emsel, emseh, pmse=pmse, pmset=pmset, est=est, elbo=elbo, 
-       elbot=elbot)
+  list(emse=emse, emsel=emsel, emseh=emseh, pmse=pmse, pmset=pmset, est=est, 
+       elbo=elbo, elbot=elbot)
 }
 stopCluster(cl=cl)
 
@@ -345,8 +346,8 @@ res <- foreach(r=1:nreps, .packages=packages) %dopar% {
   elbo <- c(mean(new.elbo(fit1.semnig, xtest, ytest)),
             mean(new.elbo(fit2.semnig, xtest, ytest)))
 
-  list(emse=emse, emsel, emseh, pmse=pmse, pmset=pmset, est=est, elbo=elbo, 
-       elbot=elbot)
+  list(emse=emse, emsel=emsel, emseh=emseh, pmse=pmse, pmset=pmset, est=est, 
+       elbo=elbo, elbot=elbot)
 }
 stopCluster(cl=cl)
 
@@ -534,8 +535,8 @@ res <- foreach(r=1:nreps, .packages=packages) %dopar% {
   elbo <- c(mean(new.elbo(fit1.semnig, xtest, ytest)),
             mean(new.elbo(fit2.semnig, xtest, ytest)))
   
-  list(emse=emse, emsel, emseh, pmse=pmse, pmset=pmset, est=est, elbo=elbo, 
-       elbot=elbot)
+  list(emse=emse, emsel=emsel, emseh=emseh, pmse=pmse, pmset=pmset, est=est, 
+       elbo=elbo, elbot=elbot)
 }  
 stopCluster(cl=cl)
 

### 2019-11-28 docs/manuscript1.Rnw
added lpml calculations to analysis_gdsc.R

@@
 est.lambdaf <- apply(res[rownames(res)=="lambdaf", c(1, 2)], 2, median)
-tab <- cbind(apply(res[rownames(res)=="emse", ], 2, median),
+
+tab <- cbind(apply(res[rownames(res)=="emsel", ], 2, median),
+             apply(res[rownames(res)=="emseh", ], 2, median),
+             apply(res[rownames(res)=="emse", ], 2, median),
              apply(res[rownames(res)=="pmse", ], 2, median))
@@
   paste0("\\textbf{", tab[which(tab[, 2]==min(tab[, 2], na.rm=TRUE)), 2], "}")
-colnames(tab) <- c("$\\text{EMSE}_{\\text{test}}$",
-                   "$\\text{PMSE}_{\\text{test}}$")     
+tab[which(tab[, 3]==min(tab[, 3], na.rm=TRUE)), 3] <- 
+  paste0("\\textbf{", tab[which(tab[, 3]==min(tab[, 3], na.rm=TRUE)), 3], "}")
+tab[which(tab[, 4]==min(tab[, 4], na.rm=TRUE)), 4] <- 
+  paste0("\\textbf{", tab[which(tab[, 4]==min(tab[, 4], na.rm=TRUE)), 4], "}")
+colnames(tab) <- c("$\\text{EMSE}_{\\text{bottom}}$",
+                   "$\\text{EMSE}_{\\text{top}}$",
+                   "$\\text{EMSE}_{\\text{total}}$",
+                   "$\\text{PMSE}$")     
+rownames(tab)[c(1, 2)] <- c("NIG$_{\\text{f}}^-$",
+                            "NIG$_{\\text{f}}$")
 @	
@@
 
-	Table \ref{tab:simulations_gdsc_tab1} displays median estimation MSE (EMSE) 
-	and prediction MSE (PMSE) for simulation version 1. The table includes as a
-	reference the null model, which
-	is an empty model that only fits the means of the $\y_{d}$.
+	Table \ref{tab:simulations_gdsc_tab1} displays median estimation MSE 
+	(EMSE) and prediction MSE (PMSE) for simulation version 1, calculated on an 
+	independent test data set. EMSE is further split into the contribution of the
+	bottom 90\% of the $\beta_{jd}$ in terms of size (in absolute value) and the 
+	top 10\% of the $\beta_{jd}$ in size. The table includes as a reference the 
+	null model, which is an empty model that only fits the means of the $\y_{d}$.
 	Estimation MSE (EMSE) and prediction MSE (PMSE) give measures of predictive 
@@
 	in the data better than model NIG$_{\text{f}}^-$ that does not include the
-	external data.
+	external data. Furthermore, we see that estimation of the larger $\beta_{jd}$ 
+	is where ridge loses the most compared to the NIG and lasso models.
 <<simulations_gdsc_tab1>>=
@@
 kable(tab, col.names=colnames(tab), align="r",
-      caption="Median performance measure estimates for simulation version 1 (lowest in bold).", 
+      caption="Median performance measure for simulation version 1, estimated on the test data (lowest in bold).", 
       format="latex", booktabs=TRUE, escape=FALSE) %>%
@@
 est.lambdad <- apply(res[rownames(res)=="lambdad", c(1, 2)], 2, median)
-tab <- cbind(apply(res[rownames(res)=="emse", ], 2, median),
+tab <- cbind(apply(res[rownames(res)=="emsel", ], 2, median),
+             apply(res[rownames(res)=="emseh", ], 2, median),
+             apply(res[rownames(res)=="emse", ], 2, median),
              apply(res[rownames(res)=="pmse", ], 2, median))
@@
   paste0("\\textbf{", tab[which(tab[, 2]==min(tab[, 2], na.rm=TRUE)), 2], "}")
-colnames(tab) <- c("$\\text{EMSE}_{\\text{test}}$",
-                   "$\\text{PMSE}_{\\text{test}}$")     
+tab[which(tab[, 3]==min(tab[, 3], na.rm=TRUE)), 3] <- 
+  paste0("\\textbf{", tab[which(tab[, 3]==min(tab[, 3], na.rm=TRUE)), 3], "}")
+tab[which(tab[, 4]==min(tab[, 4], na.rm=TRUE)), 4] <- 
+  paste0("\\textbf{", tab[which(tab[, 4]==min(tab[, 4], na.rm=TRUE)), 4], "}")
+colnames(tab) <- c("$\\text{EMSE}_{\\text{bottom}}$",
+                   "$\\text{EMSE}_{\\text{top}}$",
+                   "$\\text{EMSE}_{\\text{total}}$",
+                   "$\\text{PMSE}$")      
+rownames(tab)[c(1, 2)] <- c("NIG$_{\\text{d}}^-$",
+                            "NIG$_{\\text{d}}$")
 @	
@@
 kable(tab, col.names=colnames(tab), align="r",
-      caption="Median performance measure estimates for simulation version 2 (best performing model in bold).", 
+      caption="Median performance measure for simulation version 2, estimated on the test data (lowest in bold).", 
       format="latex", booktabs=TRUE, escape=FALSE) %>%
@@
 phi.ratio <- apply(apply(sapply(0:3, function(s) {
-  res[rownames(res)==paste0("alphaf" ,s), colnames(res)=="NIG2"]}) %*% 
+  res[rownames(res)==paste0("alphaf" ,s), 2]}) %*% 
     t(cbind(1, rbind(0, diag(3)))), 2, function(m) {
-      res[rownames(res)=="alphaf0", colnames(res)=="NIG2"]/m}), 2, median)
+      res[rownames(res)=="alphaf0", 2]/m}), 2, median)
 chi.ratio <- apply(apply(sapply(0:3, function(s) {
-  res[rownames(res)==paste0("alphad" ,s), colnames(res)=="NIG2"]}) %*% 
+  res[rownames(res)==paste0("alphad" ,s), 2]}) %*% 
     t(cbind(1, rbind(0, diag(3)))), 2, function(m) {
-      res[rownames(res)=="alphad0", colnames(res)=="NIG2"]/m}), 2, median)
+      res[rownames(res)=="alphad0", 2]/m}), 2, median)
 true.ratio <- as.numeric(1/(c(1, 1, 3, 7) %*% 
@@
 
-tab <- cbind(apply(res[rownames(res)=="emse", ], 2, median),
+tab <- cbind(apply(res[rownames(res)=="emsel", ], 2, median),
+             apply(res[rownames(res)=="emseh", ], 2, median),
+             apply(res[rownames(res)=="emse", ], 2, median),
              apply(res[rownames(res)=="pmse", ], 2, median))
@@
   paste0("\\textbf{", tab[which(tab[, 2]==min(tab[, 2], na.rm=TRUE)), 2], "}")
-colnames(tab) <- c("$\\text{EMSE}_{\\text{test}}$",
-                   "$\\text{PMSE}_{\\text{test}}$")     
+tab[which(tab[, 3]==min(tab[, 3], na.rm=TRUE)), 3] <- 
+  paste0("\\textbf{", tab[which(tab[, 3]==min(tab[, 3], na.rm=TRUE)), 3], "}")
+tab[which(tab[, 4]==min(tab[, 4], na.rm=TRUE)), 4] <- 
+  paste0("\\textbf{", tab[which(tab[, 4]==min(tab[, 4], na.rm=TRUE)), 4], "}")
+colnames(tab) <- c("$\\text{EMSE}_{\\text{bottom}}$",
+                   "$\\text{EMSE}_{\\text{top}}$",
+                   "$\\text{EMSE}_{\\text{total}}$",
+                   "$\\text{PMSE}$")     
+rownames(tab)[c(1, 2)] <- c("NIG$_{\\text{f}+\\text{d}}^-$",
+                            "NIG$_{\\text{f}+\\text{d}}$")
 @	
@@
 kable(tab, col.names=colnames(tab), align="r",
-      caption="Median performance measure estimates for simulation version 3 (best performing model in bold).", 
+      caption="Median performance measure for simulation version 3, estimated on the test data (lowest in bold).", 
       format="latex", booktabs=TRUE, escape=FALSE) %>%
@@
 	Simulations in Section \ref{sec:simulations} show that the method is 
-	reasonably able to recover hyperparameters. In addition, the simulations
-	show that inclusion of informative external covariates is indeed beneficial to
-	predictive performance. The model is put into practice on the GDSC data. 
+	reasonably able to recover hyperparameters. Simulation versions 1 and 2 show
+	that estimation of drug- and feature-specific hyperparameters is, in 
+	principle, fairly accurate. However, when estimated jointly, biases may occur
+	due to the interplay between the two sources of information. Nevertheless,
+	the simulations show that inclusion of informative external covariates is 
+	still beneficial to predictive performance.
+
+	The model is put into practice on the GDSC data. 
+	The NIG model is competitive with convential models like lasso and ridge.
+	We note that estimation of the penalty parameters in the latter two models
+	is geared towards predictive performance, because cross-validation is used
+	to train these parameters, as opposed to estimation of the
+	hyperparameters in the NIG model, which aims to improve model fit. In that 
+	sense, the comparison metric (PMSE) is not to the benefit of the NIG model.
+
 	The results in Section
-	\ref{sec:gdsc} show that the inclusion of the external covariates is 
-	beneficial for drug sensitivity prediction. PMSE of the NIG
-	model with both both feature and drug variance components is lower
-	than for the regular lasso model and although the PMSE is
-	not lower if external covariates are included, the resulting hyperparameter 
+	\ref{sec:gdsc} show that the inclusion of the external covariates covariates 
+	indeed substantially modifies the hyperparameters. Although, PMSE of the NIG
+	model is not necessarily lower if external covariates are included, the
+	resulting hyperparameter 
 	estimates are still informative and may guide future research efforts.
-	In addition, the comparison to the bSEM model that allows for the inclusion of 
-	one binary external feature covariate favours the proposed NIG model. 
-	PMSEs are almost uniformly lower over the drugs for the NIG model 
-	compared to bSEM. Consequently, the total PMSE is significantly
+
+	In addition, the comparison to bSEM, a Bayesian counterpart model that allows
+	for the inclusion of one binary external feature covariate favours the 
+	proposed NIG model. PMSEs are almost uniformly lower over the drugs for the 
+	NIG model compared to bSEM. Consequently, the total PMSE is significantly
 	lower for the NIG model.
@@
 	variance component in the prior of the $\bm{\beta}_d$. If we add the 
-	additional constraint that the drug and feature variance components have to 
+	additional constraint that the drug and feature variance components
 	average to one, this cross-validated variance component may be interpreted as
-	the overall level of shrinkage that is then tailored to lower PMSE.
+	the overall level of shrinkage. This overall shrinkage level is then tailored 
+	to lower PMSE. Other directions of future research are applications of the NIG
+	model to different data types. An example of such an application is eQTL, 
+	where gene expressions are regressed on SNPs. Several interesting external
+	covariates are available, both on the genes as well as the SNPs. An example of
+	an external
+	covariate for the genes is genelength, where we suspect that longer genes are
+	harder to predict. The distance of the SNP to the gene is an example of an
+	external SNP covariate, where the expectation is that SNPs further from the 
+	gene are less predictive of that gene's expression.
 

### 2019-11-28 rpackage/R/aux_functions.R

@@ -1,3 +1,8 @@
+# function to integrate for cpo calculation (not tested)
+.f.int.cpo <- function(x, xtSigmax, n, p, zeta, y, xtmu) {
+  ((y - x)^2/(2*zeta) + 1)^(-(n + p + 1)/2)*exp(-(x - xtmu)^2/(2*xtSigmax))
+}
+
 # ELBO (not tested)
 .single.elbo <- function(p, n, zeta, yty, aux, g, b, delta, eta, lambdaf, 
                          lambdad, Zalphad, Calphaf) {

### 2020-01-23 docs/arxiv.Rnw
first commit after revision

--- 2019-11-28 docs/manuscript1.Rnw
@@
 \setlength{\textwidth}{15cm}
-\title{Drug sensitivity prediction with the normal inverse Gaussian prior 
+\title{Drug sensitivity prediction with normal inverse Gaussian shrinkage 
 informed by external data}
@@
 		on both the drugs and the features is often available. We propose to 
-		model the drug responses through a linear regression with the normal
+		model the drug responses through a linear regression with shrinkage enforced
+		through a normal
 		inverse Gaussian prior. We let the prior depend on the external information,
@@
 	structured penalized multivariate regression approach. 
-	\cite[]{aben_tandem:_2016} introduces a two-stage penalized regression 
+	\cite{aben_tandem:_2016} introduce a two-stage penalized regression 
 	model that includes two different types of molecular features.
 	\cite{ammad-ud-din_drug_2016} and \cite{costello_community_2014} tackle the
-	problem through a multiple kernel learning approach.
-
-  Our solution allows for the adaptive incorporation of the external 
+	problem through a multiple kernel learning approach. Our solution allows for
+	the adaptive incorporation of the external 
   information on drugs and features. This is done by
-  pooling, both across drugs and features. Estimation of the model
+  pooling information, both across drugs and features. Estimation of the model
   is through computational feasible variational bayes approximations, while 
@@
 	\ref{sec:model} we introduce our model, the estimation of which is detailed
-	in Section \ref{sec:estimation}. In Section \ref{sec:gdsc} we analyse the
+	in Section \ref{sec:estimation}. Section \ref{sec:simulations} describes a 
+	simulation study that investigates the estimation of hyperparameters by the
+	proposed method. In Section \ref{sec:gdsc} we analyse the
 	GDSC data, and we end with a discussion in Section \ref{sec:discussion} on the 
@@
   Jeffrey's prior
-  for the error variance, we jointly consider an unknown mean and variance 
+  for the error variance, we jointly consider an unknown data mean and variance 
   \cite[]{kass_selection_1996}. This
@@
   available software, but for clarity it is omitted in the following.
-  Furthermore, several additional extensions of the model are 
-  described in [! ref to second paper here].
 
@@
     \lambda_{\text{drug}}/\chi_d^2, \eta_d\), \\
-    q(\sigma^2_d) & \overset{D}{=} \Gamma^{-1} \(\frac{n + p + 1}{2}, \zeta_d\).
+    q(\sigma^2_d) & \overset{D}{=} \Gamma^{-1} \(\frac{n + p + 1}{2}, \zeta_d\),
   \end{align*}
-	See SM Section \ref{sm-sec:vbderivations} for the derivations. The variational
+	where $\mathcal{GIG} \(p, \nu, \eta\)$ denotes the generalized inverse Gaussian
+	distribution with index $p \in \mathbb{R}$, and scales $\nu > 0$ and 
+	$\eta > 0$ \cite[]{jorgensen_statistical_1982}. See SM Section 
+	\ref{sm-sec:vbderivations} for the derivations. The variational
 	parameters $\bm{\mu}_d$, $\bm{\Sigma}_d$, $\delta_{jd}$, $\eta_d$, and
@@
   therefore propose to estimate these hyperparameters by empirical Bayes.
-  In our case, this results in an   
-  objective and data-driven inclusion of the feature and drug covariates.
+  In our case, this results in an objective and data-driven inclusion of the
+  external feature and drug covariates.
 	The canonical method for empirical Bayes is to maximise the marginal 
@@
 	Bayes estimation properties of the model in a simulated setting; its main aim 
-	is to assess hyperparameter recovery. It is a data based simulation,
+	is to assess hyperparameter estimation. It is a data based simulation,
 	wherein the responses are simulated from a synthetic model, but the
@@
 	168 drug sensitivities for 482 cell lines are 
-	simulated from three versions of model (\ref{eq:likelihood}) and 
-	(\ref{eq:prior}). All versions draw the error variances as 
-	$\forall d: \sigma_d^2 \sim \Gamma^{-1}(3,2)$,
+	simulated from model (\ref{eq:likelihood}) and 
+	(\ref{eq:prior}). We draw the error
+	variances as $\forall d: \sigma_d^2 \sim \Gamma^{-1}(3,2)$,
 	such that the prior $\sigma_d^2$ mean and variance are both one. 
-	Version 1 fixes $\forall d: \tau_d^2=1$, version 2 fixes
-	$\forall j,d: \gamma_{jd}^2=1$, while version 3 draws both the $\tau_d^2$ and
-	$\gamma_{jd}^2$ according to (\ref{eq:prior}). In version 1, we
+	We consider the following three scenarios for the simulation of the 
+	drug and feature variance components:
+	\begin{itemize}
+	  \item{Scenario 1 fixes $\forall d: \tau_d^2=1$ and draws the 
+	    $\gamma_{jd}^2$ according to model (\ref{eq:prior}). We
 	    create four external dummy feature covariates that code for four 
 	    randomly assigned approximately equally sized groups of features. We set
-	$\bm{\alpha}_{\text{feat}}$ such
-	that the $\gamma_{jd}^2$ of the four groups of features have prior means
-	$\phi_{jd} \in \{ 1, 1/2, 1/4, 1/8 \}$ 
-	($\bm{\alpha}_{\text{feat}}=\begin{bmatrix} 1 & 1 & 3 & 7 \end{bmatrix} \tr$). 
-	The prior scale parameter is set to $\lambda_{\text{feat}}=1$. Version 2 
-	follows a similar procedure for the $\tau_d^2$. We randomly create four
-	groups of drugs with corresponding external drug dummy variables and set
-	$\bm{\alpha}_{\text{drug}}=\begin{bmatrix} 1 & 1 & 3 & 7 \end{bmatrix} \tr$,
-	such that we have $\chi_{d} \in \{ 1, 1/2, 1/4, 1/8 \}$. The
-	scale is set to $\lambda_{\text{drug}}=1$. Version 3 combines the procedures
-	from versions 1 and 2 to draw the $\gamma_{jd}^2$ and $\tau_d^2$.
-
+	    $\bm{\alpha}_{\text{feat}}$ such that the $\gamma_{jd}^2$ of the four
+	    groups of features have prior means $\phi_{jd} \in \{ 1, 1/2, 1/4, 1/8 \}$
+	    ($\bm{\alpha}_{\text{feat}}=\begin{bmatrix} 1 & 1 & 3 & 7 \end{bmatrix} 
+	    \tr$). The prior scale parameter is set to $\lambda_{\text{feat}}=1$}.
+	  \item{Scenario 2 fixes $\forall j,d: \gamma_{jd}^2=1$ and draws the 
+	    $\tau_d^2$ according to model (\ref{eq:prior}), following a procedure
+	    similar to the procedure for the $\gamma_{jd}^2$ in Scenario 1: we 
+	    randomly create four groups of drugs with corresponding external drug 
+	    dummy variables and set $\bm{\alpha}_{\text{drug}}=\begin{bmatrix} 
+	    1 & 1 & 3 & 7 \end{bmatrix} \tr$, such that we have $\chi_{d} \in 
+	    \{ 1, 1/2, 1/4, 1/8 \}$. The scale is set to $\lambda_{\text{drug}}=1$.}
+	  \item{Scenario 3 combines the procedures from Scenarios 1 and 2 to draw both
+	    the $\gamma_{jd}^2$ and $\tau_d^2$ according to (\ref{eq:prior}).}
+	\end{itemize}
 	We estimate four models: (i) the NIG model that only includes an intercept in
@@
 	NIG$_{\text{f}+\text{d}}^-$, depending on which variance components are 
-	estimated (feature, drug, or both),
+	estimated (feature, drug, or both in Scenarios 1, 2, and 3, respectively),
 	(ii) the NIG model estimated as in Section \ref{sec:estimation} that includes
@@
 	NIG$_{\text{d}}^-$, and NIG$_{\text{f}+\text{d}}^-$ models
-	amounts to direct estimation of one common expected 
+	amounts to direct estimation of common expected 
 	prior means $\phi$ and/or $\chi$, instead of regression estimates for the
@@
 	All models are fit on 241 randomly selected cell lines, while performance
-	measures are estimated from the remaining 241 cell lines. We repeat the whole
-	simulation 50 times.
+	measures are estimated from the remaining 241 cell lines. We repeat every
+	simulation Scenario 50 times.
 
@@
 	$\bm{\alpha}_{\text{feat}}$ together with its true value for NIG$_{\text{f}}$ 
-	in version 1 of the simulation study. Figure 
+	in Scenario 1 of the simulation study (fixed $\tau_d^2$). Figure 
 	\ref{fig:simulations_gdsc_est1}a shows that 
@@
 	parameter, which are, in general, hard to estimate. 
-<<simulations_gdsc_est1, fig.cap="Estimated (a) hyperparameters $\\bm{\\alpha}_{\\text{feat}}$ and (b) prior means $\\bm{\\phi}_{jd}$ from simulation version 1 with fixed $\\tau_d^2$.", out.width="100%", fig.asp=1/2>>=
-@
-
-	Model NIG$_{\text{f}}^-$ gives a median $\phi$ estimate of 
-	\Sexpr{round(est.phi, 2)},  
+<<simulations_gdsc_est1, fig.cap="Simulation results for Scenario 1 ($\\tau^2_d$ fixed): (a) estimated and true values for $\\alpha_{\\text{feat}}$ and, (b) estimated and true values for prior means $\\phi_{jd}$.", out.width="100%", fig.asp=1/2>>=
+@
+
+	Model NIG$_{\text{f}}^-$ (that excludes the external covariates) gives a 
+	median $\phi$ estimate of \Sexpr{round(est.phi, 2)},  
 	equal to the true mean of the $\phi_{jd}$, \Sexpr{round(true.phi, 2)}. 
@@
 	Table \ref{tab:simulations_gdsc_tab1} displays median estimation MSE 
-	(EMSE) and prediction MSE (PMSE) for simulation version 1, calculated on an 
+	(EMSE) and prediction MSE (PMSE) for simulation Scenario 1, calculated on an 
 	independent test data set. EMSE is further split into the contribution of the
@@
 	is where ridge loses the most compared to the NIG and lasso models. 
+	The heavier tails of the NIG and lasso priors allow for larger $\beta_{jd}$
+	estimates and thus lower EMSE for these $\beta_{jd}$.
 <<simulations_gdsc_tab1>>=
@@
 kable(tab, col.names=colnames(tab), align="r",
-      caption="Median performance measure for simulation version 1, estimated on the test data (lowest in bold).", 
+      caption="Median performance measure for simulation Scenario 1, estimated on the test data (lowest in bold).", 
       format="latex", booktabs=TRUE, escape=FALSE) %>%
@@
 	$\bm{\alpha}_{\text{drug}}$ together with its true value for NIG$_{\text{d}}$ 
-	in version 2 of the simulation study. They are estimated without bias.
+	in Scenario 2 of the simulation study (fixed $\gamma_{jd}^2$). They are
+	estimated without bias.
 	Median $\lambda_{\text{drug}}=\Sexpr{round(est.lambdad[2], 2)}$ is slightly 
-	overestimated compared to true value 1. 
+	overestimated compared to true value one. 
 	The median $\chi$ estimate in the NIG$_{\text{d}}^-$ model is 
@@
 	which is almost equal to the true mean \Sexpr{round(true.chi, 2)}. 
-	$\lambda_{\text{drug}}$ is underestimated with median estimate
-	\Sexpr{round(est.lambdad[1], 2)} compared to the true value.
-	Median performance measures for simulation version 2 are displayed in Table
+	Scale $\lambda_{\text{drug}}$ is underestimated with median estimate
+	\Sexpr{round(est.lambdad[1], 2)} compared to the true value one.
+	Median performance measures for simulation Scenario 2 are displayed in Table
 	\ref{tab:simulations_gdsc_tab2}. Again, we see that NIG$_{\text{d}}$ performs
@@
 	here, indicating that the external covariates do not add much here.
-<<simulations_gdsc_est2, fig.cap="Estimated (a) hyperparameters $\\bm{\\alpha}_{\\text{feat}}$ and (b) prior means $\\bm{\\phi}_{jd}$ from simulation version 2 with fixed $\\gamma_{jd}^2$.", out.width="100%", fig.asp=1/2>>=
+	Similarly as in Scenario 1, most of the estimation performance gain over ridge
+	is through the top 10\% of the $\beta_{jd}$.
+<<simulations_gdsc_est2, fig.cap="Simulation results for Scenario 2 ($\\gamma^2_{jd}$ fixed): (a) estimated and true values for $\\alpha_{\\text{feat}}$ and, (b) estimated and true values for prior means $\\phi_{jd}$.", out.width="100%", fig.asp=1/2>>=
 @
@@
 kable(tab, col.names=colnames(tab), align="r",
-      caption="Median performance measure for simulation version 2, estimated on the test data (lowest in bold).", 
+      caption="Median performance measure for simulation Scenario 2, estimated on the test data (lowest in bold).", 
       format="latex", booktabs=TRUE, escape=FALSE) %>%
@@
 	are displayed together with their true values for
-	simulation 3. $\bm{\alpha}_{\text{feat}}$ are underestimated, while 
+	simulation Scenario 3. $\bm{\alpha}_{\text{feat}}$ are underestimated, while 
 	$\bm{\alpha}_{\text{drug}}$ are overestimated. These biases seem to 
@@
 	that NIG$_{\text{f}+\text{d}}$ outperforms the other models in terms of
-	prediction and estimation MSE.
-<<simulations_gdsc_est3, fig.cap="Estimated (a) hyperparameters $\\bm{\\alpha}_{\\text{feat}}$ and (b) prior means $\\bm{\\phi}_{jd}$ from simulation version 3.", out.width="100%", fig.asp=1>>= 
+	prediction and estimation MSE, with again most of the edge over ridge in the
+	top 10\% of the $\beta_{jd}$.
+<<simulations_gdsc_est3, fig.cap="Simulation results for Scenario 3: (a) estimated and true values for $\\alpha_{\\text{feat}}$ and, (b) estimated and true values for prior means $\\phi_{jd}$.", out.width="100%", fig.asp=1>>=
 @
@@
 kable(tab, col.names=colnames(tab), align="r",
-      caption="Median performance measure for simulation version 3, estimated on the test data (lowest in bold).", 
+      caption="Median performance measure for simulation Scenario 3, estimated on the test data (lowest in bold).", 
       format="latex", booktabs=TRUE, escape=FALSE) %>%
@@
 	Bayes iterations. Currently, problems with number of outcomes in the hundreds
-	and number of features and samples in the thousands are computationally 
+	and number of features and samples up to a thousand are computationally 
 	feasible.
@@
 	NIG$_{\text{f}}$, and NIG$_{\text{d}}$ in the following. The nomenclature is
-	the same as in Section \ref{sec:simulations}, so NIG$_{\text{f}+\text{d}}^-$
-	does not use the external covariates and estimates
+	the same as in Section \ref{sec:simulations}, i.e., 
+	NIG$_{\text{f}+\text{d}}^-$ excludes the external covariates and estimates
 	one overall $\phi$ and one overall $\chi$, while NIG$_{\text{f}+\text{d}}$ 
@@
 	$\gamma_{jd}^2 \sim \Gamma^{-1} (\phi_{jd}, \lambda_{jd})$, instead of
-	$\gamma_{jd}^2 \sim \mathcal{IG} (\phi_{jd}, \lambda_{jd})$. Additionally,
-	the bSEM model can
+	$\gamma_{jd}^2 \sim \mathcal{IG} (\phi_{jd}, \lambda_{jd})$, where 
+	$\Gamma^{-1} (\phi, \lambda)$ denotes the inverse gamma distribution with 
+	shape $\phi$ and scale $\lambda$. Additionally, the bSEM model can
 	only incorporate one binary external covariate that divides the features into
@@
 
+rownames(est1)[c(1:4)] <- rownames(est2)[c(1:4)] <-
+  c("NIG$_{\\text{f}+\\text{d}}^-$", "NIG$_{\\text{f}+\\text{d}}$",
+    "NIG$_{\\text{f}}$", "NIG$_{\\text{d}}$")
+
 tab <- cbind(apply(res[rownames(res)=="pmse", ], 2, median))
@@
 colnames(tabm) <- c("$\\text{PMSE}_{\\text{test}}$")  
+rownames(tabm)[c(1:4)] <- rownames(est2)[c(1:4)] <-
+  c("NIG$_{\\text{f}+\\text{d}}^-$", "NIG$_{\\text{f}+\\text{d}}$",
+    "NIG$_{\\text{f}}$", "NIG$_{\\text{d}}$")
 
@@
   compared to the approved drugs. This translates to smaller expected 
-  prior precisions for the regression parameters for approved drugs, i.e.,
-  approved drugs are easier to predict than experimental and developmental 
-  drugs. On the other hand, the NIG$_{\text{d}}$ model, that only includes
-  external
+  prior precisions for the regression parameters of approved drugs, i.e.,
+  approved drugs are easiest to predict, followed by developmental and 
+  experimental drugs, as expected. On the other hand, the NIG$_{\text{d}}$ 
+  model, that only includes external
   drug covariates, estimates a negative effect for the developmental drugs.
@@
   \Sexpr{round(tab[2, 1], 3)}, respectively), both outperforming the lasso
-  \Sexpr{round(tab[5, 1], 3)}. The underperformance of the 
+  with a PMSE of \Sexpr{round(tab[5, 1], 3)}. The underperformance of the 
   NIG$_{\text{f}}$ and NIG$_{\text{d}}$ models shows the usefulness of a
@@
 	Simulations in Section \ref{sec:simulations} show that the method is 
-	reasonably able to recover hyperparameters. Simulation versions 1 and 2 show
+	reasonably able to estimate the hyperparameters. Simulation Scenarios 1 and 2
+	show
 	that estimation of drug- and feature-specific hyperparameters is, in 
@@
 	The model is put into practice on the GDSC data. 
-	The NIG model is competitive with convential models like lasso and ridge.
+	The NIG model is competitive with convential models like lasso in 
+	terms of predictive performance, but is slightly outperformed by ridge.
 	We note that estimation of the penalty parameters in the latter two models
 	is geared towards predictive performance, because cross-validation is used
-	to train these parameters, as opposed to estimation of the
+	to train these parameters, as opposed to empirical Bayes estimation of the
 	hyperparameters in the NIG model, which aims to improve model fit. In that 
@@
 
-	A possible direction of future research that attempts 
+	A possible direction of future research to attempt
 	to lower PMSE of the NIG model is the introduction of a cross-validated global
 	variance component in the prior of the $\bm{\beta}_d$. If we add the 
-	additional constraint that the drug and feature variance components
+	additional constraint that the drug and feature-specific variance components
 	average to one, this cross-validated variance component may be interpreted as
 	the overall level of shrinkage. This overall shrinkage level is then tailored 
-	to lower PMSE. Other directions of future research are applications of the NIG
-	model to different data types. An example of such an application is eQTL, 
+	to lower PMSE, while the feature- and drug-specific variance component 
+	include the external information. Other directions of future research are
+	applications of the NIG
+	model to different data types. One possibility is to apply the model in an 
+	eQTL study, 
 	where gene expressions are regressed on SNPs. Several interesting external
@@
 	% the GDSC data. 
-	A visit of MM to SR and GL was funded by the Amsterdam Public Health 
-	institute's Methodology travel grant.
+	MM visited SR and GL on a travel grant funded the Amsterdam Public Health 
+	institute's Methodology program.
 	\textit{Conflict of Interest}: None declared.
@@
 \end{document}
+
+% set options
+<<settings, include=FALSE, echo=FALSE>>=
+opts_knit$set(root.dir="..", base.dir="../figs")
+opts_chunk$set(fig.align='center', fig.path="../figs/", 
+               echo=FALSE, cache=FALSE, message=FALSE, fig.pos='!ht')
+knit_hooks$set(document=function(x) {
+  sub('\\usepackage[]{color}', '\\usepackage{xcolor}', x, fixed=TRUE)})
+@
+
+% read figure code
+<<figures, include=FALSE>>=
+read_chunk("code/figures.R")
+@
+
+\documentclass[a4paper,hidelinks]{article}
+\usepackage{bm,amsmath,amssymb,amsthm,amsfonts,graphics,graphicx,epsfig,
+rotating,caption,subcaption,natbib,appendix,titlesec,multicol,hyperref,verbatim,
+bbm,algorithm,algpseudocode,pgfplotstable,threeparttable,booktabs,mathtools,
+dsfont,parskip,xr}
+\externaldocument[md-]{manuscript1}
+
+% vectors and matrices
+\newcommand{\bbeta}{\bm{\beta}}
+\newcommand{\btau}{\bm{\tau}}
+\newcommand{\bsigma}{\bm{\sigma}}
+\newcommand{\balpha}{\bm{\alpha}}
+\newcommand{\bepsilon}{\bm{\epsilon}}
+\newcommand{\bomega}{\bm{\omega}}
+\newcommand{\bOmega}{\bm{\Omega}}
+\newcommand{\bSigma}{\bm{\Sigma}}
+\newcommand{\bkappa}{\bm{\kappa}}
+\newcommand{\btheta}{\bm{\theta}}
+\newcommand{\bmu}{\bm{\mu}}
+\newcommand{\bpsi}{\bm{\psi}}
+\newcommand{\blambda}{\bm{\lambda}}
+\newcommand{\bLambda}{\bm{\Lambda}}
+\newcommand{\bPsi}{\bm{\Psi}}
+\newcommand{\bphi}{\bm{\phi}}
+\newcommand{\Y}{\mathbf{Y}}
+\newcommand{\y}{\mathbf{y}}
+\newcommand{\X}{\mathbf{X}}
+\newcommand{\x}{\mathbf{x}}
+\newcommand{\I}{\mathbf{I}}
+\newcommand{\bgamma}{\bm{\gamma}}
+\newcommand{\T}{\mathbf{T}}
+\newcommand{\Z}{\mathbf{Z}}
+\newcommand{\W}{\mathbf{W}}
+\renewcommand{\O}{\mathbf{O}}
+\newcommand{\B}{\mathbf{B}}
+\renewcommand{\H}{\mathbf{H}}
+\newcommand{\U}{\mathbf{U}}
+\newcommand{\Em}{\mathbf{E}}
+\newcommand{\D}{\mathbf{D}}
+\newcommand{\Vm}{\mathbf{V}}
+\newcommand{\rv}{\mathbf{r}}
+\newcommand{\A}{\mathbf{A}}
+\newcommand{\Q}{\mathbf{Q}}
+\newcommand{\Sm}{\mathbf{S}}
+\newcommand{\0}{\bm{0}}
+\newcommand{\tx}{\tilde{\mathbf{x}}}
+\newcommand{\hy}{\hat{y}}
+\newcommand{\tm}{\tilde{m}}
+\newcommand{\tkappa}{\tilde{\kappa}}
+\newcommand{\m}{\mathbf{m}}
+\newcommand{\C}{\mathbf{C}}
+\renewcommand{\v}{\mathbf{v}}
+\newcommand{\g}{\mathbf{g}}
+\newcommand{\s}{\mathbf{s}}
+\renewcommand{\c}{\mathbf{c}}
+\newcommand{\ones}{\mathbf{1}}
+\newcommand{\R}{\mathbf{R}}
+\newcommand{\rvec}{\mathbf{r}}
+\newcommand{\Lm}{\mathbf{L}}
+
+% functions and operators
+\renewcommand{\L}{\mathcal{L}}
+\newcommand{\G}{\mathcal{G}}
+\renewcommand{\P}{\mathcal{P}}
+\newcommand{\argmax}{\text{argmax} \,}
+\newcommand{\expit}{\text{expit}}
+\newcommand{\erfc}{\text{erfc}}
+\newcommand{\E}{\mathbb{E}}
+\newcommand{\V}{\mathbb{V}}
+\newcommand{\Cov}{\mathbb{C}\text{ov}}
+\newcommand{\tr}{^{\text{T}}}
+\newcommand{\diag}{\text{diag}}
+\newcommand{\KL}{D_{\text{KL}}}
+\newcommand{\trace}{\text{tr}}
+\newcommand{\norm}[1]{\left\lVert #1 \right\rVert}
+
+% distributions
+\newcommand{\N}{\text{N}}
+\newcommand{\TG}{\text{TG}}
+\newcommand{\Bi}{\text{Binom}}
+\newcommand{\PG}{\text{PG}}
+\newcommand{\Mu}{\text{Multi}}
+\newcommand{\GIG}{\text{GIG}}
+\newcommand{\IGauss}{\text{IGauss}}
+\newcommand{\Un}{\text{U}}
+
+% syntax and readability
+\renewcommand{\(}{\left(}
+\renewcommand{\)}{\right)}
+\renewcommand{\[}{\left[}
+\renewcommand{\]}{\right]}
+
+% settings
+\graphicspath{{../figs/}}
+\pgfplotsset{compat=1.16}
+\setlength{\evensidemargin}{.5cm} \setlength{\oddsidemargin}{.5cm}
+\setlength{\textwidth}{15cm}
+\title{Drug sensitivity prediction with normal inverse Gaussian shrinkage 
+informed by external data}
+\date{\today}
+\author{Magnus M. M\"unch$^{1,2}$\footnote{Correspondence to: 
+\href{mailto:m.munch@amsterdamumc.nl}{m.munch@amsterdamumc.nl}}, Mark A. van de 
+Wiel$^{1,3}$, Sylvia Richardson$^{3}$, \\ and Gwena{\"e}l G. R. Leday$^{3}$}
+
+\begin{document}
+
+	\maketitle
+
+	\noindent
+	1. Department of Epidemiology \& Biostatistics, Amsterdam UMC, VU University, 
+	PO Box 7057, 1007 MB Amsterdam, The Netherlands \\
+	2. Mathematical Institute, Leiden University, Leiden, The Netherlands \\
+	3. MRC Biostatistics Unit, Cambridge Institute of Public Health, Cambridge,
+	United Kingdom
+
+	\section{Content overview}
+	This document contains the Supplementary Material (SM) to the document 
+	`Drug sensitivity prediction with the normal inverse Gaussian prior 
+	informed by external data'. In the 
+	following, this document is referred to as Main Document (MD).
+
+	Section \ref{sec:prior} describes the moments of the normal
+	inverse Gaussian (NIG) prior, Section \ref{sec:vbderivations} gives
+	the variational Bayes (BV) derivations that lead to the estimating equations 
+	in the MD, Section \ref{sec:elbo} gives the evidence lower bound (ELBO) that 
+	is maximised during the VB iterations. In Section \ref{sec:computation} 
+	efficient VB calculations are given, while Section 
+	\ref{sec:ratiosmodifiedbessels} outlines a computationally stable method
+	to calculate ratios of modified Bessel functions. Section \ref{sec:eb} shows
+	that the empirical Bayes (EB) scale estimates $\bm{\lambda}$ are non-negative.
+	The last Section \ref{sec:gibbssampler} introduces a Gibbs sampling scheme
+	to sample from the posterior corresponding to the NIG model.
+
+	\section{Normal inverse Gaussian prior moments}\label{sec:prior}
+	Let $\beta \sim \mathcal{N}(0,\sigma^2 \tau^2 \gamma^2)$, where 
+	$\sigma^2$ and $\tau^2$ are fixed and $\gamma^2 \sim 
+	\mathcal{IG}(\phi,\lambda_{\text{feat}})$, then
+	\begin{align*}
+	  \E(\beta | \sigma^2, \tau^2) & = 0, \\
+	  \V(\beta | \sigma^2, \tau^2) & = \sigma^2 \tau^2 \phi, \\
+	  \mathcal{K}(\beta | \sigma^2, \tau^2) & = 
+	  \frac{3\phi}{\lambda_{\text{feat}}} + 3,
+	\end{align*}
+	where $\mathcal{K}(\cdot)$ denotes the kurtosis. Now letting $\tau^2
+	\sim \mathcal{IG}(\chi,\lambda_{\text{drug}})$ and integrating it out we have:
+	\begin{align*}
+	  \E(\beta | \sigma^2) & = 0, \\
+	  \V(\beta | \sigma^2) & = \sigma^2 \chi \phi, \\
+	  \mathcal{K}(\beta | \sigma^2) & = 
+	  3\(\frac{\phi}{\lambda_{\text{feat}}} \frac{\chi}{\lambda_{\text{drug}}} +
+	  \frac{\phi}{\lambda_{\text{feat}}} + \frac{\chi}{\lambda_{\text{drug}}}\) + 
+	  3.
+	\end{align*}
+
+	\section{Variational Bayes derivations}\label{sec:vbderivations}
+	For clarity we have indexed the variational 
+	density functions with their respective parameters, which we omitted in the 
+	MD. In the following all expectations are with respect to the 
+	variational posterior $Q_d$. The variational posterior for $\bbeta_d$ is 
+	found as follows:
+	\begin{align*}
+	  \log q_{\bbeta_d} (\bbeta_d) & \propto \E [\log \mathcal{L}
+	  (\y_d | \bbeta_d, \sigma_d^2 )] + \E [\log \pi (\bbeta_d | \bgamma_{d}^2, 
+	  \sigma_d^2, \tau_d^2)] \\
+	  & \propto - \frac{1}{2} \sum_{i=1}^n 
+	  \E \[ \frac{(y_{id} - \x_i \tr \bbeta_d)^2}{\sigma_d^2} \] - 
+	  \frac{1}{2} \sum_{j=1}^p 
+	  \E \( \frac{\beta_{jd}^2}{\gamma_{jd}^2 \sigma_d^2 \tau_d^2} \), \\
+	  q_{\bbeta_d} (\bbeta_d) & \overset{D}{=} 
+	  \mathcal{N}_p (\bmu_d, \bSigma_d), \\
+	  & \text{with } \bSigma_d = \E(\sigma_d^{-2})^{-1} 
+	  \{\X \tr \X + \E(\tau_d^{-2})\diag [\E(\gamma_{jd}^{-2})] \}^{-1}, \\
+	  & \text{and } \bmu_d = \{\X \tr \X + \E(\tau_d^{-2})
+	  \diag [\E(\gamma_{jd}^{-2})] \}^{-1} 
+	  \X \tr \y_d.
+	\end{align*}
+	The variational posterior for $\gamma_{jd}^2$ is given by:
+	\begin{align*}
+	  \log q_{\gamma_{jd}^{2}}(\gamma_{jd}^{2}) & \propto \E 
+	  [\log \pi (\beta_{jd} | \gamma_{jd}^{2}, \sigma_d^2, \tau_d^2)] + 
+	  \log \pi(\gamma_{jd}^{2}) \\
+	  & \propto -\frac{1}{2} \log \gamma_{jd}^{2} - \frac{1}{2} 
+	  \E \( \frac{\beta_{jd}^2}{\sigma^{2}_d \tau_d^2}\) \gamma_{jd}^{-2} - 
+	  \frac{3}{2} \log \gamma_{jd}^{2} - 
+	  \frac{\lambda_{\text{feat}} \{\gamma_{jd}^2 - 
+	  \phi_{jd}\}^2}
+	  {2\phi_{jd}^{2}} \gamma_{jd}^{-2} \\
+	  & \propto \( -1 - 1\) \log \gamma_{jd}^{2} - 
+	  \frac{\lambda_{\text{feat}}}{2\phi_{jd}^2} \gamma_{jd}^2 - 
+	  \frac{1}{2}\[ \lambda_{\text{feat}} + \E(\tau_d^{-2})\E(\sigma_d^{-2}) 
+	  \E(\beta_{jd}^2)\] \gamma_{jd}^{-2} , \\
+	  q_{\gamma_{jd}^{2}}(\gamma_{jd}^{2}) & \overset{D}{=} \mathcal{GIG} 
+	  \(-1, \lambda_{\text{feat}}/\phi_{jd}^{2}, \delta_{jd} \), \\
+	  & \text{with } \delta_{jd} =\E(\tau_d^{-2})\E(\sigma_d^{-2})
+	  \[\E(\beta_{jd})^2 + \V(\beta_{jd})\] + \lambda_{\text{feat}}.
+	\end{align*}
+	Similarly, we derive the variational posterior for $\tau_d^2$ as:
+	\begin{align*}
+	  \log q_{\tau_{d}^{2}}(\tau_{d}^{2}) & \propto \E 
+	  [\log \pi (\bbeta_{d} | \bgamma_{d}^{2}, \sigma_d^2, \tau_d^2)] + 
+	  \log \pi(\tau_{d}^{2}) \\
+	  & \propto -\frac{p}{2} \log \tau_{d}^{2} - \frac{1}{2} \sum_{j=1}^p
+	  \E \( \frac{\beta_{jd}^2}{\sigma^{2}_d \gamma_{jd}^{2}}\)\tau_d^{-2}  - 
+	  \frac{3}{2} \log \tau_{d}^{2} - 
+	  \frac{\lambda_{\text{drug}} \{\tau_{d}^2 - \chi_{d}\}^2}
+	  {2\chi_{d}^{2}} \tau_{d}^{-2} \\
+	  & \propto \( -\frac{p + 1}{2} - 1\) \log \tau_{d}^{2} - 
+	  \frac{\lambda_{\text{drug}}}{2\chi_{d}^2} \tau_{d}^2 - 
+	  \frac{1}{2}\[ \lambda_{\text{drug}} + \E(\sigma_d^{-2}) \sum_{j=1}^p 
+	  \E(\gamma_{jd}^{-2}) \E(\beta_{jd}^2)\] \tau_{d}^{-2} , \\
+	  q_{\tau_{d}^{2}}(\tau_{d}^{2}) & \overset{D}{=} \mathcal{GIG} 
+	  \(-\frac{p + 1}{2}, \lambda_{\text{drug}}/\chi_{d}^{2}, \eta_{d} \), \\
+	  & \text{with } \eta_{d} =\E(\sigma_d^{-2}) \sum_{j=1}^p
+	  \E(\gamma_{jd}^{-2})\[\E(\beta_{jd})^2 + \V(\beta_{jd})\] + 
+	  \lambda_{\text{drug}}.
+	\end{align*}
+	Lastly, we have the variational posterior for $\sigma_d^2$:
+	\begin{align*}
+	  \log q_{\sigma_d^{2}}(\sigma_d^{2}) & \propto \E 
+	  [\log \mathcal{L}(\y_d | \bbeta_d, \sigma_d^2 )] + 
+	  \E [\log \pi (\bbeta_d | \bgamma_d^2, \sigma_d^2,\tau_d^2)] + 
+	  \log \pi (\sigma_d^{2}) \\
+	  & \propto -\frac{n}{2} \log \sigma_d^{2} - \frac{1}{2} \sum_{i=1}^n 
+	  \frac{\E [(y_{id} - \x_i \tr \bbeta_d)^2]}{\sigma_d^{2}} - 
+	  \frac{p}{2} \log \sigma_d^{2} - \frac{1}{2} \sum_{j=1}^p 
+	  \E \( \frac{\beta_{jd}^2}{\gamma_{jd}^{2}\tau_d^2}\) \sigma_d^{-2} - 
+	  \frac{3}{2} \log \sigma^2_d \\
+	  & = \(-\frac{n + p + 1}{2} - 1 \) \log \sigma_d^{2} \\
+	  & \,\,\,\,\,\,\,\,\,\, - 
+	  \frac{1}{2}\left\{ \sum_{i=1}^n \E [(y_{id} - \x_i \tr \bbeta_d)^2] + 
+	  \E(\tau_d^{-2}) \sum_{j=1}^p \E(\gamma_{jd}^{-2})
+	  \E(\beta_{jd}^2) \right\} \sigma_d^{-2}, \\
+	  q_{\sigma_d^{2}}(\sigma_d^{2}) & \overset{D}{=} 
+	  \Gamma^{-1} \(\frac{n + p + 1}{2}, \zeta_{d} \), \\
+	  & \text{with } \zeta_{d} = \frac{1}{2} \bigg\{\mathbf{y}_d \tr 
+	  \mathbf{y}_d - 2 \mathbf{y}_d \tr \X \E(\bbeta_d) + 
+    \trace [ \X \tr \X \V(\bbeta_d)] + 
+    \E(\bbeta_d \tr)  \X \tr \X \E(\bbeta_d ) \\
+    & \,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\, + 
+    \E(\tau_d^{-2})\trace \{ \diag[\E(\gamma_{jd}^{-2})] \V(\bbeta_d)\} + 
+    \E(\tau_d^{-2})\E(\bbeta_d \tr) \diag[\E(\gamma_{jd}^{-2})] \E(\bbeta_d )
+    \bigg\}.
+  \end{align*}
+  If we fill in the expectations and variances, we arrive at the estimating
+  equations in MD equations (\ref{md-eq:vbequations}).
+
+  \section{Evidence lower bound}\label{sec:elbo}
+  The evidence lower bound that is maximised during the iterations is: 
+  $\text{ELBO}^{(l)}=\text{ELBO}_d^{(l)}$, with:
+  \begin{align*}
+    \text{ELBO}&^{(l)}_d \\
+    & = -\frac{n+p+1}{2} \log \pi + \frac{p+1-n}{2}\log 2 + \frac{n+1}{2} + p
+    + \log \Gamma \(\frac{n + p + 1}{2} \) \\
+    & - \frac{n + p + 1}{2} \log \zeta_d^{(l)} + 
+    \frac{1}{2} \log |\bSigma_d^{(l)}| - 
+    \frac{n+p+1}{4\zeta_d^{(l)}}  \bigg[\mathbf{y}_d \tr \mathbf{y}_d -
+    2 \mathbf{y}_d \tr \X \bm{\mu}_d^{(l)} + 
+    \trace ( \X \tr \X \bm{\Sigma}_d^{(l)}) \\ 
+    & + (\bm{\mu}_d^{(l)}) \tr \X \tr \X \bm{\mu}_d^{(l)} +
+    g_{d}^{(l)} \trace \[ \diag (b_{jd}^{(l)}) \bm{\Sigma}_d^{(l)}\] + 
+    g_{d}^{(l)} (\bm{\mu}_d^{(l)}) \tr \diag (b_{jd}^{(l)}) 
+    \bm{\mu}_d^{(l)}\bigg] + p \log \lambda_{\text{feat}}^{(l)} \\
+    & + \frac{p + 3}{4} \log \lambda_{\text{drug}}^{(l)} + 
+    \frac{\eta_d^{(l)} - \lambda_{\text{drug}}^{(l)}}{2} g_d^{(l)} + 
+    \sum_{j=1}^p \frac{\delta_{jd}^{(l)} - \lambda_{\text{feat}}^{(l)}}{2} 
+    b_{jd}^{(l)} + \frac{\lambda_{\text{drug}}^{(l)}}{\chi_d^{(l)}} +
+    \sum_{j=1}^p \frac{\lambda_{\text{feat}}^{(l)}}{\phi_{jd}^{(l)}} + \\
+    & - \frac{p + 1}{2} \log \chi_d^{(l)} - \sum_{j=1}^p \log \phi_{jd}^{(l)}
+    -\frac{p+1}{4} \log \eta_d^{(l)} - 
+    \frac{1}{2} \sum_{j=1}^p \log \delta_{jd}^{(l)} \\
+    & + \log K_{\frac{p+1}{2}}\(\sqrt{\eta_{d}^{(l)}
+    \lambda_{\text{drug}}^{(l)}}/
+    \chi_{d}^{(l)}\) + \sum_{j=1}^p \log K_{1}\(\sqrt{\delta_{jd}^{(l)}
+    \lambda_{\text{feat}}^{(l)}}/\phi_{jd}^{(l)}\).
+  \end{align*}
+
+  \section{Efficient computation}\label{sec:computation}
+  The empirical Bayes updates require the quantities
+  $e_{jd}$, $b_{jd}$, $g_d$, and $f_d$, which in turn depend on the quantities 
+  $\diag(\bSigma_d)$, $\bmu_{d}$, $\y_d \tr \X \bmu_d$, 
+  $\trace(\X \tr \X \bSigma_d)$, $\bmu_d \tr \X \tr \X \bmu_d$,
+  $g_d\trace[\diag(b_{jd})\bSigma_d]$, and $g_d\bmu_d \tr \diag(b_{jd}) \bmu_d$.
+  Let $\mathbf{H}_d=g_d \diag[(b_{jd}^{(h)})^{-1}]$. Then, the first two
+  quantities are efficiently calculated as:
+  \begin{subequations}\label{eq:complexity1}
+    \begin{align}
+      \diag (\bm{\Sigma}_d^{(h+1)}) & = (a_d^{(h)})^{-1} 
+      \H_d^{(h)} - (a_d^{(h)})^{-1} \{ [ \H_d^{(h)} \X \tr 
+      ( \I_n + \X \H_d^{(h)} \X \tr )^{-1} ] \circ
+      (\H_d^{(h)} \X \tr ) \} \mathbf{1}_{n \times 1}, \\
+      \bmu_d^{(h+1)}) & = \H_d^{(h)} \X \tr \y_d - \H_d^{(h)} \X \tr 
+      ( \I_n + \X \H_d^{(h)} \X \tr )^{-1} \X \H_d^{(h)} \X \tr \y_d.
+    \end{align}
+  \end{subequations}
+  The quantities $\y_d \tr \X \bmu_d$, $\bmu_d \tr \X \tr \X \bmu_d$,
+  $g_d\trace[\diag(b_{jd})\bSigma_d]$, and $g_d\bmu_d \tr \diag(b_{jd}) \bmu_d$
+  are easily calculated from the first two. For the remaining one we have:
+  \begin{align}\label{eq:complexity2}
+    \trace(\X \tr \X \bSigma_d^{(h+1)})  = 
+    & (a_d^{(h)})^{-1} \mathbf{1}_{1 \times p} \H_d^{(h)} \X \tr 
+    \mathbf{1}_{n \times 1} \nonumber \\  
+    & - (a_d^{(h)})^{-1} \mathbf{1}_{1 \times p} \{ [  \H_d^{(h)} \X \tr 
+    ( \I_n + \X \H_d^{(h)} \X \tr )^{-1} \X \H_d^{(h)} \X \tr ] \circ 
+    \X \tr \} \mathbf{1}_{n \times 1}.
+  \end{align}
+  Both (\ref{eq:complexity1}) and (\ref{eq:complexity2}) are operations of
+  complexity $\mathcal{O}(pn^2)$.
+
+  \section{Ratios of modified Bessel functions}\label{sec:ratiosmodifiedbessels}
+  Ratios of modified Bessel functions of the second kind 
+  $K_{\alpha - 1}(x)/K_{\alpha}(x)$ are prone to under- and overflow for large
+  $\alpha$. In our case, $\alpha$ increases linearly with $p$. Since $p$ may be
+  large, this may cause numerical issues in the calculation of various 
+  quantities. We alleviate the numerical issues through the following.
+
+  We let $n_1=p/2$ and $n_2=(p-1)/2$ and use the well-known recursive relation:
+  $$
+  K_{\alpha}(x) = K_{\alpha - 2}(x) + \frac{2(\alpha - 1)}{2} K_{\alpha- 1}(x),
+  $$
+  to rewrite the ratio:
+  \begin{align*}
+    \frac{K_{\frac{p - 1}{2}}(x)}{K_{\frac{p + 1}{2}}(x)} = 
+    \begin{cases}
+      \Big( \dots \Big( \( 1 + \frac{2 \cdot 1 - 1}{x} \)^{-1} + 
+      \frac{2 \cdot 2-1}{x}\Big)^{-1} + \dots + 
+      \frac{2 \cdot n_1 - 1}{x}\Big)^{-1}, & \text{for } p \text{ even}, \\
+      \Big( \dots \Big( \( \frac{K_0(x)}{K_1(x)} + \frac{2}{x} \cdot 1 \)^{-1} +
+      \frac{2}{x} \cdot 2 \Big)^{-1} + \dots + \frac{2}{x} \cdot n_2 \Big)^{-1},
+      & \text{for } p \text{ odd}.
+    \end{cases}
+  \end{align*} 
+  The ratio $K_0(x)/K_1(x)$ is well-behaved, so that it may be computed
+  without numerical issues.
+
+  \section{Empirical Bayes}\label{sec:eb}
+  For the feature scale parameter $\lambda_{\text{feat}}$ updates to be 
+  non-negative, we have to show that:
+  \begin{align}\label{eq:inequality}
+    &\sum_{d=1}^D \sum_{j=1}^{p} 
+    b_{jd} + \bm{\alpha}_{\text{feat}} \tr \mathbf{C} \tr 
+    \diag (e_{jd}) \mathbf{C} \bm{\alpha}_{\text{feat}} - 
+    2 \bm{\alpha}_{\text{feat}} \tr \mathbf{C} \tr 
+    \mathbf{1}_{pD \times 1}  = \sum_{d=1}^D \sum_{j=1}^{p} \(b_{jd} + 
+    \phi_{jd}^{-2} e_{jd} - 2 \phi_{jd}^{-1} \) \geq 0,
+  \end{align}
+  which holds if all summands in the left-hand side of (\ref{eq:inequality})
+  are non-negative.
+
+  Next, we note that through Jensen's inequality 
+  $e_{jd}^{-1}=\E(\gamma_{jd}^{2})^{-1} \leq \E(\gamma_{jd}^{-2})=b_{jd}$,
+  we may lowerbound the summands in left-hand side of (\ref{eq:inequality}):
+  $$
+  b_{jd} + \phi_{jd}^{-2} e_{jd} - 2 \phi_{jd}^{-1} \geq
+  e_{jd}^{-1} + \phi_{jd}^{-2} e_{jd} - 2 \phi_{jd}^{-1}.
+  $$
+  which we use to show that:
+  \begin{align*}
+    e_{jd}^{-1} + \phi_{jd}^{-2} e_{jd} - 2 \phi_{jd}^{-1} & \geq 0 \\
+    1 + \phi_{jd}^{-2} e_{jd}^2 - 2 \phi_{jd}^{-1} e_{jd} & \geq 0 \\
+    \( 1 - \phi_{jd}^{-1} e_{jd} \)^2 & \geq 0.
+  \end{align*}
+  Similar reasoning holds for the drug scale parameter $\lambda_{\text{drug}}$.
+
+  \section{Gibbs sampler}\label{sec:gibbssampler}
+  MCMC samples from the posterior corresponding to model 
+  (\ref{md-eq:prior}) may be generated for each equation independently.
+  MCMC samples may be generated by iteratively sampling the following 
+  conditional distributions:
+  \begin{subequations}
+    \begin{align}
+      \bm{\beta}_d | \bgamma_d^2, \sigma_d^2, \tau_d^2, \y_d & \sim 
+      \mathcal{N}_p (\bm{\mu}_d, \bm{\Sigma}_d), \label{eq:betasample} \\
+      \text{with } & \bm{\Sigma}_d = \sigma_d^2 [\X \tr \X + 
+      \tau_d^{-2} \diag (\gamma_{jd}^{-2})]^{-1}, \label{eq:sigmasample} \\
+      \text{and } & \bm{\mu}_d = [\X \tr \X + 
+      \tau_d^{-2} \diag (\gamma_{jd}^{-2})]^{-1} \X \tr 
+      \y_d \label{eq:musample}, \\
+      \gamma_{jd}^2 | \beta_{jd}, \sigma_d^2, \tau_d^2, \y &
+      \sim \mathcal{GIG}
+      \(-1, \lambda_{\text{feat}}/ \phi_{jd}^{2}, \delta_{jd}\), \nonumber \\
+      \text{with } & \delta_{jd} = \sigma_d^{-2} \tau_d^{-2} \beta_{jd}^2 + 
+      \lambda_{\text{feat}}, \nonumber \\
+      \tau_d^2 | \bm{\beta}_d, \sigma_d^2, \bgamma_{d}^2, \y & \sim
+      \mathcal{GIG}
+      \(-\frac{p+1}{2}, \lambda_{\text{drug}}/ \chi_{d}^{2}, \eta_d\), 
+      \nonumber \\
+      \text{with } & \eta_d = \sigma_d^{-2} \sum_{j=1}^p \gamma_{jd}^{-2}
+      \beta_{jd}^2 + \lambda_{\text{drug}}, \nonumber \\
+      \sigma_d^2 | \bm{\beta}_d, \bgamma_d^2, \tau_d^2, \y & \sim \Gamma^{-1} 
+      \(\frac{n + p + 1}{2}, \zeta_d\), \nonumber \\
+      \text{with } & \zeta_d = \frac{1}{2} \bigg[\mathbf{y} \tr \mathbf{y} -
+      2 \mathbf{y} \tr \X \bm{\mu} + \trace ( \X \tr \X \bm{\Sigma}) + 
+      (\bm{\mu}) \tr \X \tr \X \bm{\mu} + 
+      \tau^{-2}_d \trace (\diag(\gamma_{jd}^{-2}) \bm{\Sigma}) \nonumber \\
+      & \,\,\,\,\,\,\,\,\,\, + \tau_d^{-2} (\bm{\mu}) \tr 
+      \diag(\gamma_{jd}^{-2}) \bm{\mu}\bigg]. \nonumber
+    \end{align}
+  \end{subequations}
+  In high dimensional space, the $p \times p$ matrix inversions in 
+  (\ref{eq:sigmasample}) and (\ref{eq:musample}) are   
+  significant computational bottlenecks. \cite{bhattacharya_fast_2016} describe 
+  a method to sample from (\ref{eq:betasample}) without explicit calculation
+  of this inverse, thereby offering a siginifcant speed-up compared to naive
+  sampling.
+
+  \bibliographystyle{author_short3}  
+	\bibliography{refs}
+
+	\section*{Session info}
+
+<<r session-info, eval=TRUE, echo=TRUE, tidy=TRUE>>=
+devtools::session_info()
+@
+
+\end{document}

### 2020-01-23 docs/biometrical_main.Rnw
first commit after revision

@@
 opts_chunk$set(fig.align='center', fig.path="../figs/", 
-               echo=FALSE, cache=FALSE, message=FALSE, fig.pos='!ht')
+               echo=FALSE, cache=FALSE, message=FALSE, fig.pos='!h')
 knit_hooks$set(document=function(x) {
@@
 bbm,algorithm,algpseudocode,pgfplotstable,threeparttable,booktabs,mathtools,
-dsfont,parskip,xr,tikz,times,w-thm}
-\externaldocument[sm-]{supplement1}
+dsfont,parskip,tikz,times,w-thm}
 \usetikzlibrary{bayesnet}
@@
   \keywords{Drug sensitivity; Empirical Bayes; 
-  Genomics of Drug Sensitivity in Cancer (GDSC); Variational Bayes\\[1pc]
+  Genomics of Drug Sensitivity in Cancer (GDSC); Prediction; 
+  Variational Bayes\\[1pc]
   \noindent\hspace*{-4.2pc} Supporting Information for this article is available
@@
 
-  \title[Drug sensitivity prediction with the NIG prior]
-  {Drug sensitivity prediction with the normal inverse Gaussian prior 
+  \title[Drug sensitivity prediction with NIG shrinkage]
+  {Drug sensitivity prediction with normal inverse Gaussian shrinkage 
   informed by external data}
@@
   Amsterdam UMC, VU University, PO Box 7057, 1007 MB Amsterdam, The Netherlands}
-  \address[\inst{2}]{Mathematical Institute, Leiden University, Leiden, 
-  The Netherlands}
+  \address[\inst{2}]{Mathematical Institute, Leiden University, PO Box 9512,
+  2300 RA Leiden, The Netherlands}
   \address[\inst{3}]{MRC Biostatistics Unit, 
-  Cambridge Institute of Public Health, Cambridge, United Kingdom}
+  University of Cambridge, Cambridge Institute of Public Health, Forvie Site,
+  Robinson Way, Cambridge Biomedical Campus, Cambridge CB2 0SR, United Kingdom}
 
@@
   additional global variance component $\tau_d^2$. Supplementary Material
-  (SM) Section \ref{sm-sec:prior} contains more details on the NIG prior.
+  (SM) Section 2 contains more details on the NIG prior.
   To illustrate the effect of the NIG prior on the posterior mean, we consider 
@@
   several choices of $\beta_{jd}$ prior.
-
-<<dens_kappa, fig.cap="Implied prior densities $\\pi(\\kappa_{jd})$ for the (a) NIG, (b) Student's $t$, and (c) lasso priors.
-Different line types correspond to different hyperparameter settings.
-The hyperparameter settings were chosen to show some possible, distinct shapes that each of the priors can take.", out.width="100%", fig.asp=1/3>>=
-@
 
@@
   available software, but for clarity it is omitted in the following.
-  Furthermore, several additional extensions of the model are 
-  described in [! ref to second paper here].
-
-  \begin{figure}
-    \centering
-    \tikz{ %
-      % Y
-      \node[obs] (y) {$y_{id}$}; %
-      \factor[right=of y, yshift=1.5cm, xshift=-0.1cm] {y-f} 
-      {above:$\mathcal{N}$} {} {} ; %
-
-      % X
-      \node[det, above=of y, yshift=1cm] (x) {$x_{ij}$}; %
-
-      % sigma2
-      \node[latent, right=of y] (sigma2) {$\sigma_d^2$}; %
-
-      % beta
-      \node[latent, right=of x] (beta) {$\beta_{jd}$}; %
-      \factor[right=of beta, yshift=-1.4cm, xshift=-0.1cm] {beta-f} 
-      {above:$\mathcal{N}$} {} {} ; %
-
-      % gamma2
-      \node[latent, right=of beta] (gamma2) {$\gamma_{jd}^2$}; %
-      \node[const, right=of gamma2, yshift=0.5cm] (phi) {$\phi_{jd}$}; %
-      \node[const, right=of gamma2, yshift=-0.5cm] (lambdaf) 
-      {$\lambda_{\text{feat}}$}; %
-      \factor[right=of gamma2] {gamma2-f} {above:$\mathcal{IG}$} {phi,lambdaf} 
-      {gamma2} ; %
-      \node[det, right=of gamma2, xshift=2cm, yshift=0.5cm] (c) {$c_{jdg}$}; %
-
-      % tau2
-      \node[latent, right=of sigma2] (tau2) {$\tau_d^2$}; %
-      \node[const, right=of tau2, yshift=0.5cm] (chi) {$\chi_{d}$}; %
-      \node[const, right=of tau2, yshift=-0.5cm] (lambdad) 
-      {$\lambda_{\text{drug}}$}; %
-      \factor[right=of tau2] {tau2-f} {above:$\mathcal{IG}$} {chi,lambdad} 
-      {tau2} ; %
-      \node[det, right=of tau2, xshift=2cm, yshift=0.5cm] (z) {$z_{dh}$}; %
-
-      % edges
-      \factoredge {x,beta,sigma2} {y-f} {y} ; %
-      \factoredge {sigma2,gamma2,tau2} {beta-f} {beta} ; %
-      \edge {c} {phi}; %
-      \edge {z} {chi}; %
-
-      % plates
-      \plate{plateg} {(c)} {$g=1,\dots,G$}
-      \plate{plateh} {(z)} {$h=1,\dots,H$}
-      \plate{platei} {(y) (x)}{$i=1,\dots,n$}; %
-      \plate{platej} {(x) (beta) (gamma2) (phi) (lambdaf) (c) (plateg)}
-      {$j=1,\dots,p$}; %
-      \plate{plated} 
-      {(y) (x) (sigma2) (beta) (gamma2) (tau2) (phi) (lambdaf)
-      (chi) (lambdad) (c) (z) (platei) (platej) (plateg) (plateh)}
-      {$d=1,\dots,D$}; %
-    }
-    \caption{Hierarchical representation of the drug sensitivity prediction 
-    model. Grey circles 
-    represent observed variables, white circles represent unobserved variables,
-    squares represent fixed data, and unenclosed letters are parameters to be
-    estimated. Cell lines are indexed by $i$, features
-    by $j$, drugs by $d$, drug covariates by $h$, and feature covariates by 
-    $g$. The $y_{id}$ are the drug sensitivities, $x_{ij}$ the 
-    molecular features, 
-    $c_{jdg}$ the external feature covariates, $z_{dh}$ the external drug
-    covariates, $\beta_{jd}$ the regression coefficients, $\sigma_d^2$ the 
-    error variances, $\tau_d^2$ and $\gamma_{jd}^2$ the drug and feature
-    specific variance components, respectively, and $\phi_{jd}$, 
-    $\lambda_{\text{feat}}$, $\chi_{d}$, and $\lambda_{\text{drug}}$ the 
-    hyperparameters.}
-    \label{fig:dag}
-  \end{figure}
 
@@
     \lambda_{\text{drug}}/\chi_d^2, \eta_d\), \\
-    q(\sigma^2_d) & \overset{D}{=} \Gamma^{-1} \(\frac{n + p + 1}{2}, \zeta_d\).
+    q(\sigma^2_d) & \overset{D}{=} \Gamma^{-1} \(\frac{n + p + 1}{2}, \zeta_d\),
   \end{align*}
-	See SM Section \ref{sm-sec:vbderivations} for the derivations. The variational
+	where $\mathcal{GIG} \(p, \nu, \eta\)$ denotes the generalized inverse Gaussian
+	distribution with index $p \in \mathbb{R}$, and scales $\nu > 0$ and 
+	$\eta > 0$ \cite[]{jorgensen_statistical_1982}. See SM Section 
+	3 for the derivations. The variational
 	parameters $\bm{\mu}_d$, $\bm{\Sigma}_d$, $\delta_{jd}$, $\eta_d$, and
@@
   modified Bessel functions of the second kind, as in 
-  (\ref{eq:ratiomodifiedbessel}), is given in SM Section 
-  \ref{sm-sec:ratiosmodifiedbessels}.
+  (\ref{eq:ratiomodifiedbessel}), is given in SM Section 6.
 
@@
   has converged). A Gibbs sampler is described in 
-  SM Section \ref{sm-sec:gibbssampler}. Alternatively, we provide an
+  SM Section 8. Alternatively, we provide an
   implementation of the proposed model in stan using the R package rstan 
-  \cite[]{guo_rstan:_2018} at 
+  \cite[]{guo_rstan:_2018} in the Supporting Information on the journal’s web
+  page (http://onlinelibrary.wiley.com/doi/xxx/ suppinfo) or at
   \url{https://github.com/magnusmunch/cambridge}.
@@
 	168 drug sensitivities for 482 cell lines are 
-	simulated from three versions of model (\ref{eq:likelihood}) and 
-	(\ref{eq:prior}). All versions draw the error variances as 
-	$\forall d: \sigma_d^2 \sim \Gamma^{-1}(3,2)$,
+	simulated from model (\ref{eq:likelihood}) and 
+	(\ref{eq:prior}). We draw the error
+	variances as $\forall d: \sigma_d^2 \sim \Gamma^{-1}(3,2)$,
 	such that the prior $\sigma_d^2$ mean and variance are both one. 
-	Version 1 fixes $\forall d: \tau_d^2=1$, version 2 fixes
-	$\forall j,d: \gamma_{jd}^2=1$, while version 3 draws both the $\tau_d^2$ and
-	$\gamma_{jd}^2$ according to (\ref{eq:prior}). In version 1, we
+	We consider the following three scenarios for the simulation of the 
+	drug and feature variance components:
+	\begin{itemize}
+	  \item{Scenario 1 fixes $\forall d: \tau_d^2=1$ and draws the 
+	    $\gamma_{jd}^2$ according to model (\ref{eq:prior}). We
 	    create four external dummy feature covariates that code for four 
 	    randomly assigned approximately equally sized groups of features. We set
-	$\bm{\alpha}_{\text{feat}}$ such
-	that the $\gamma_{jd}^2$ of the four groups of features have prior means
-	$\phi_{jd} \in \{ 1, 1/2, 1/4, 1/8 \}$ 
-	($\bm{\alpha}_{\text{feat}}=\begin{bmatrix} 1 & 1 & 3 & 7 \end{bmatrix} \tr$). 
-	The prior scale parameter is set to $\lambda_{\text{feat}}=1$. Version 2 
-	follows a similar procedure for the $\tau_d^2$. We randomly create four
-	groups of drugs with corresponding external drug dummy variables and set
-	$\bm{\alpha}_{\text{drug}}=\begin{bmatrix} 1 & 1 & 3 & 7 \end{bmatrix} \tr$,
-	such that we have $\chi_{d} \in \{ 1, 1/2, 1/4, 1/8 \}$. The
-	scale is set to $\lambda_{\text{drug}}=1$. Version 3 combines the procedures
-	from versions 1 and 2 to draw the $\gamma_{jd}^2$ and $\tau_d^2$.
-
-	We estimate four models: (i) an NIG model that does include external 
-	covariates in the estimation called NIG1 in the following, (ii) the NIG model
-	estimated as in Section \ref{sec:estimation} called NIG2, and regular (iii)
-	lasso and (iv) ridge models.
-	Exclusion of the external covariates as in the NIG1 model
-	amounts to direct estimation of one common expected 
+	    $\bm{\alpha}_{\text{feat}}$ such that the $\gamma_{jd}^2$ of the four
+	    groups of features have prior means $\phi_{jd} \in \{ 1, 1/2, 1/4, 1/8 \}$
+	    ($\bm{\alpha}_{\text{feat}}=\begin{bmatrix} 1 & 1 & 3 & 7 \end{bmatrix} 
+	    \tr$). The prior scale parameter is set to $\lambda_{\text{feat}}=1$}.
+	  \item{Scenario 2 fixes $\forall j,d: \gamma_{jd}^2=1$ and draws the 
+	    $\tau_d^2$ according to model (\ref{eq:prior}), following a procedure
+	    similar to the procedure for the $\gamma_{jd}^2$ in Scenario 1: we 
+	    randomly create four groups of drugs with corresponding external drug 
+	    dummy variables and set $\bm{\alpha}_{\text{drug}}=\begin{bmatrix} 
+	    1 & 1 & 3 & 7 \end{bmatrix} \tr$, such that we have $\chi_{d} \in 
+	    \{ 1, 1/2, 1/4, 1/8 \}$. The scale is set to $\lambda_{\text{drug}}=1$.}
+	  \item{Scenario 3 combines the procedures from Scenarios 1 and 2 to draw both
+	    the $\gamma_{jd}^2$ and $\tau_d^2$ according to (\ref{eq:prior}).}
+	\end{itemize}
+	We estimate four models: (i) the NIG model that only includes an intercept in
+	the external covariates, called NIG$_{\text{f}}^-$, NIG$_{\text{d}}^-$, or
+	NIG$_{\text{f}+\text{d}}^-$, depending on which variance components are 
+	estimated (feature, drug, or both in Scenarios 1, 2, and 3, respectively),
+	(ii) the NIG model estimated as in Section \ref{sec:estimation} that includes
+	all external covariates, called NIG$_{\text{f}}$, NIG$_{\text{d}}$, or
+	NIG$_{\text{f}+\text{d}}$, again depending on which variance components are 
+	estimated, and regular (iii) lasso and (iv) ridge models.
+	Exclusion of the external covariates as in the NIG$_{\text{f}}^-$, 
+	NIG$_{\text{d}}^-$, and NIG$_{\text{f}+\text{d}}^-$ models
+	amounts to direct estimation of common expected 
 	prior means $\phi$ and/or $\chi$, instead of regression estimates for the
-	$\phi_{jd}$ and/or $\chi_d$ as in the NIG2 model. All 
-	models are fit on 241 randomly selected cell lines, while performance measures 
-	are estimated from the remaining 241 cell lines. We repeat the whole
-	simulation 50 times.
+	$\phi_{jd}$ and/or $\chi_d$ as in the 
+	NIG$_{\text{f}}$, NIG$_{\text{d}}$, and NIG$_{\text{f}+\text{d}}$ models. 
+	All models are fit on 241 randomly selected cell lines, while performance
+	measures are estimated from the remaining 241 cell lines. We repeat every
+	simulation Scenario 50 times.
 
@@
 est.lambdaf <- apply(res[rownames(res)=="lambdaf", c(1, 2)], 2, median)
-tab <- cbind(apply(res[rownames(res)=="elbo", ], 2, median),
-             apply(res[rownames(res)=="elbot", ], 2, median) -
-               apply(res[rownames(res)=="elbo", ], 2, median),
+
+tab <- cbind(apply(res[rownames(res)=="emsel", ], 2, median),
+             apply(res[rownames(res)=="emseh", ], 2, median),
              apply(res[rownames(res)=="emse", ], 2, median),
-             apply(res[rownames(res)=="pmse", ], 2, median),
-             apply(res[rownames(res)=="pmse", ], 2, median) - 
-               apply(res[rownames(res)=="pmset", ], 2, median))
+             apply(res[rownames(res)=="pmse", ], 2, median))
 tab <- round(tab, 3)
-tab[which(tab[, 1]==max(tab[, 1], na.rm=TRUE)), 1] <- 
-  paste0("\\textbf{", tab[which(tab[, 1]==max(tab[, 1], na.rm=TRUE)), 1], "}")
+tab[which(tab[, 1]==min(tab[, 1], na.rm=TRUE)), 1] <- 
+  paste0("\\textbf{", tab[which(tab[, 1]==min(tab[, 1], na.rm=TRUE)), 1], "}")
 tab[which(tab[, 2]==min(tab[, 2], na.rm=TRUE)), 2] <- 
@@
   paste0("\\textbf{", tab[which(tab[, 4]==min(tab[, 4], na.rm=TRUE)), 4], "}")
-tab[which(tab[, 5]==min(tab[, 5], na.rm=TRUE)), 5] <- 
-  paste0("\\textbf{", tab[which(tab[, 5]==min(tab[, 5], na.rm=TRUE)), 5], "}")
-colnames(tab) <- c("$\\text{ELBO}_{\\text{test}}$",
-                   "$\\text{ELBO}_{\\text{train}} 
-                   - \\text{ELBO}_{\\text{test}}$",
-                   "$\\text{EMSE}_{\\text{test}}$",
-                   "$\\text{PMSE}_{\\text{test}}$",
-                   "$\\text{PMSE}_{\\text{test}} 
-                   - \\text{PMSE}_{\\text{train}}$")     
+colnames(tab) <- c("$\\text{EMSE}_{\\text{bottom}}$",
+                   "$\\text{EMSE}_{\\text{top}}$",
+                   "$\\text{EMSE}_{\\text{total}}$",
+                   "$\\text{PMSE}$")     
+rownames(tab)[c(1, 2)] <- c("NIG$_{\\text{f}}^-$",
+                            "NIG$_{\\text{f}}$")
 @	
   Figure \ref{fig:simulations_gdsc_est1} shows the estimated
-	$\bm{\alpha}_{\text{feat}}$ together with its true value for NIG2 in 
-	version 1 of the simulation study. Figure 
+	$\bm{\alpha}_{\text{feat}}$ together with its true value for NIG$_{\text{f}}$ 
+	in Scenario 1 of the simulation study (fixed $\tau_d^2$). Figure 
 	\ref{fig:simulations_gdsc_est1}a shows that 
@@
 	parameter, which are, in general, hard to estimate.
-<<simulations_gdsc_est1, fig.cap="Estimated (a) hyperparameters $\\bm{\\alpha}_{\\text{feat}}$ and (b) prior means $\\bm{\\phi}_{jd}$ from simulation version 1 with fixed $\\tau_d^2$.", out.width="100%", fig.asp=1/2>>=
-@
-
-	Model NIG1 gives a median $\phi$ estimate of \Sexpr{round(est.phi, 2)},  
+
+	Model NIG$_{\text{f}}^-$ (that excludes the external covariates) gives a 
+	median $\phi$ estimate of \Sexpr{round(est.phi, 2)},  
 	equal to the true mean of the $\phi_{jd}$, \Sexpr{round(true.phi, 2)}. 
@@
 
-	Table \ref{tab:simulations_gdsc_tab1} displays several median performance
-	measures for simulation version 1. The table includes as a reference 
-	the null model, which
-	is an empty model that only fits the means of the $\y_{d}$.
-	The median evidence lower bounds (ELBO) approximates the marginal
-	likelihoods of models NIG1 and NIG2 on both training and test data, while 
-	estimation MSE (EMSE) and prediction MSE (PMSE) give measures of predictive 
-	performance. Focussing on MSE, we see that the NIG2 model, that includes
-	the external covariates, performs best, both in terms of estimation and 
-	prediction. Model fit is best for the NIG1 model that does not include the
-	external covariates, according to the ELBO, calculated on the test data.
-	The lower MSEs confirm that the model that
-	includes the external data (NIG2) learns the underlying structure 
-	in the data better than model NIG1 that does not include the external 
-	data.
-<<simulations_gdsc_tab1>>=
-library(knitr, quietly=TRUE, verbose=FALSE)
-suppressWarnings(library(kableExtra))
-options(knitr.kable.NA="", digits=3)
-kable(tab, col.names=colnames(tab), align="r",
-      caption="Median performance measure estimates for simulation version 1 (best performing model in bold).", 
-      format="latex", booktabs=TRUE, escape=FALSE) %>%
-  kable_styling(latex_options=c("HOLD_position"))
-@
+	Table \ref{tab:simulations_gdsc_tab1} displays median estimation MSE 
+	(EMSE) and prediction MSE (PMSE) for simulation Scenario 1, calculated on an 
+	independent test data set. EMSE is further split into the contribution of the
+	bottom 90\% of the $\beta_{jd}$ in terms of size (in absolute value) and the 
+	top 10\% of the $\beta_{jd}$ in size. The table includes as a reference the 
+	null model, which is an empty model that only fits the means of the $\y_{d}$.
+	Estimation MSE (EMSE) and prediction MSE (PMSE) give measures of predictive 
+	performance. Focussing on MSE, we see that the NIG$_{\text{f}}$ model, 
+	that includes the external covariates, performs best, both in terms of 
+	estimation and prediction. The lower MSEs confirm that the model that
+	includes the external data (NIG$_{\text{f}}$) learns the underlying structure 
+	in the data better than model NIG$_{\text{f}}^-$ that does not include the
+	external data. Furthermore, we see that estimation of the larger $\beta_{jd}$ 
+	is where ridge loses the most compared to the NIG and lasso models. 
+	The heavier tails of the NIG and lasso priors allow for larger $\beta_{jd}$
+	estimates and thus lower EMSE for these $\beta_{jd}$.
 
@@
 est.lambdad <- apply(res[rownames(res)=="lambdad", c(1, 2)], 2, median)
-tab <- cbind(apply(res[rownames(res)=="elbo", ], 2, median),
-             apply(res[rownames(res)=="elbot", ], 2, median) -
-               apply(res[rownames(res)=="elbo", ], 2, median),
+tab <- cbind(apply(res[rownames(res)=="emsel", ], 2, median),
+             apply(res[rownames(res)=="emseh", ], 2, median),
              apply(res[rownames(res)=="emse", ], 2, median),
-             apply(res[rownames(res)=="pmse", ], 2, median),
-             apply(res[rownames(res)=="pmse", ], 2, median) - 
-               apply(res[rownames(res)=="pmset", ], 2, median))
+             apply(res[rownames(res)=="pmse", ], 2, median))
 tab <- round(tab, 3)
-tab[which(tab[, 1]==max(tab[, 1], na.rm=TRUE)), 1] <- 
-  paste0("\\textbf{", tab[which(tab[, 1]==max(tab[, 1], na.rm=TRUE)), 1], "}")
+tab[which(tab[, 1]==min(tab[, 1], na.rm=TRUE)), 1] <- 
+  paste0("\\textbf{", tab[which(tab[, 1]==min(tab[, 1], na.rm=TRUE)), 1], "}")
 tab[which(tab[, 2]==min(tab[, 2], na.rm=TRUE)), 2] <- 
@@
   paste0("\\textbf{", tab[which(tab[, 4]==min(tab[, 4], na.rm=TRUE)), 4], "}")
-tab[which(tab[, 5]==min(tab[, 5], na.rm=TRUE)), 5] <- 
-  paste0("\\textbf{", tab[which(tab[, 5]==min(tab[, 5], na.rm=TRUE)), 5], "}")
-colnames(tab) <- c("$\\text{ELBO}_{\\text{test}}$",
-                   "$\\text{ELBO}_{\\text{train}} 
-                   - \\text{ELBO}_{\\text{test}}$",
-                   "$\\text{EMSE}_{\\text{test}}$",
-                   "$\\text{PMSE}_{\\text{test}}$",
-                   "$\\text{PMSE}_{\\text{test}} 
-                   - \\text{PMSE}_{\\text{train}}$")     
+colnames(tab) <- c("$\\text{EMSE}_{\\text{bottom}}$",
+                   "$\\text{EMSE}_{\\text{top}}$",
+                   "$\\text{EMSE}_{\\text{total}}$",
+                   "$\\text{PMSE}$")      
+rownames(tab)[c(1, 2)] <- c("NIG$_{\\text{d}}^-$",
+                            "NIG$_{\\text{d}}$")
 @	
   Figure \ref{fig:simulations_gdsc_est2} shows the estimated
-	$\bm{\alpha}_{\text{drug}}$ together with its true value for NIG2 in 
-	version 2 of the simulation study. They are estimated without bias.
+	$\bm{\alpha}_{\text{drug}}$ together with its true value for NIG$_{\text{d}}$ 
+	in Scenario 2 of the simulation study (fixed $\gamma_{jd}^2$). They are
+	estimated without bias.
 	Median $\lambda_{\text{drug}}=\Sexpr{round(est.lambdad[2], 2)}$ is slightly 
-	overestimated compared to true value 1. 
-	The median $\chi$ estimate in the NIG1 model is \Sexpr{round(est.chi, 2)},  
+	overestimated compared to true value one. 
+	The median $\chi$ estimate in the NIG$_{\text{d}}^-$ model is 
+	\Sexpr{round(est.chi, 2)},  
 	which is almost equal to the true mean \Sexpr{round(true.chi, 2)}. 
-	$\lambda_{\text{drug}}$ is underestimated with median estimate
-	\Sexpr{round(est.lambdad[1], 2)} compared to the true value.
-	Median performance measures for simulation version 2 are displayed in Table
-	\ref{tab:simulations_gdsc_tab2}. Again, we see that NIG2 performs best in 
-	terms of MSE. However, NIG1 performance is the same here, indicating that 
-	the external covariates do not add much here.
-	Model fit is again best for the NIG1 model.
-<<simulations_gdsc_est2, fig.cap="Estimated (a) hyperparameters $\\bm{\\alpha}_{\\text{feat}}$ and (b) prior means $\\bm{\\phi}_{jd}$ from simulation version 2 with fixed $\\gamma_{jd}^2$.", out.width="100%", fig.asp=1/2>>=
-@
-
-<<simulations_gdsc_tab2>>=
-library(knitr, quietly=TRUE, verbose=FALSE)
-suppressWarnings(library(kableExtra))
-options(knitr.kable.NA="", digits=3)
-kable(tab, col.names=colnames(tab), align="r",
-      caption="Median performance measure estimates for simulation version 2 (best performing model in bold).", 
-      format="latex", booktabs=TRUE, escape=FALSE) %>%
-  kable_styling(latex_options=c("HOLD_position"))
-@
+	Scale $\lambda_{\text{drug}}$ is underestimated with median estimate
+	\Sexpr{round(est.lambdad[1], 2)} compared to the true value one.
+	Median performance measures for simulation Scenario 2 are displayed in Table
+	\ref{tab:simulations_gdsc_tab2}. Again, we see that NIG$_{\text{d}}$ performs
+	best in terms of MSE. However, NIG$_{\text{d}}^-$ performance is the same 
+	here, indicating that the external covariates do not add much here.
+	Similarly as in Scenario 1, most of the estimation performance gain over ridge
+	is through the top 10\% of the $\beta_{jd}$.
 
@@
 
-tab <- cbind(apply(res[rownames(res)=="elbo", ], 2, median),
-             apply(res[rownames(res)=="elbot", ], 2, median) -
-               apply(res[rownames(res)=="elbo", ], 2, median),
+tab <- cbind(apply(res[rownames(res)=="emsel", ], 2, median),
+             apply(res[rownames(res)=="emseh", ], 2, median),
              apply(res[rownames(res)=="emse", ], 2, median),
-             apply(res[rownames(res)=="pmse", ], 2, median),
-             apply(res[rownames(res)=="pmse", ], 2, median) - 
-               apply(res[rownames(res)=="pmset", ], 2, median))
+             apply(res[rownames(res)=="pmse", ], 2, median))
 tab <- round(tab, 3)
-tab[which(tab[, 1]==max(tab[, 1], na.rm=TRUE)), 1] <- 
-  paste0("\\textbf{", tab[which(tab[, 1]==max(tab[, 1], na.rm=TRUE)), 1], "}")
+tab[which(tab[, 1]==min(tab[, 1], na.rm=TRUE)), 1] <- 
+  paste0("\\textbf{", tab[which(tab[, 1]==min(tab[, 1], na.rm=TRUE)), 1], "}")
 tab[which(tab[, 2]==min(tab[, 2], na.rm=TRUE)), 2] <- 
@@
   paste0("\\textbf{", tab[which(tab[, 4]==min(tab[, 4], na.rm=TRUE)), 4], "}")
-tab[which(tab[, 5]==min(tab[, 5], na.rm=TRUE)), 5] <- 
-  paste0("\\textbf{", tab[which(tab[, 5]==min(tab[, 5], na.rm=TRUE)), 5], "}")
-colnames(tab) <- c("$\\text{ELBO}_{\\text{test}}$",
-                   "$\\text{ELBO}_{\\text{train}} 
-                   - \\text{ELBO}_{\\text{test}}$",
-                   "$\\text{EMSE}_{\\text{test}}$",
-                   "$\\text{PMSE}_{\\text{test}}$",
-                   "$\\text{PMSE}_{\\text{test}} 
-                   - \\text{PMSE}_{\\text{train}}$")     
+colnames(tab) <- c("$\\text{EMSE}_{\\text{bottom}}$",
+                   "$\\text{EMSE}_{\\text{top}}$",
+                   "$\\text{EMSE}_{\\text{total}}$",
+                   "$\\text{PMSE}$")     
+rownames(tab)[c(1, 2)] <- c("NIG$_{\\text{f}+\\text{d}}^-$",
+                            "NIG$_{\\text{f}+\\text{d}}$")
 @	
   In Figure \ref{fig:simulations_gdsc_est3} the 
-	$\bm{\alpha}_{\text{feat}}$ and $\bm{\alpha}_{\text{drug}}$ estimated by NIG2
+	$\bm{\alpha}_{\text{feat}}$ and $\bm{\alpha}_{\text{drug}}$ estimated by 
+	NIG$_{\text{f}+\text{d}}$
 	are displayed together with their true values for
-	simulation 3. $\bm{\alpha}_{\text{feat}}$ are underestimated, while 
+	simulation Scenario 3. $\bm{\alpha}_{\text{feat}}$ are underestimated, while 
 	$\bm{\alpha}_{\text{drug}}$ are overestimated. These biases seem to 
@@
 	\ref{tab:simulations_gdsc_tab3}, with median performance measures, shows
-	that NIG2 outperforms the other models in terms of prediction and estimation
-	MSE.
-<<simulations_gdsc_est3, fig.cap="Estimated (a) hyperparameters $\\bm{\\alpha}_{\\text{feat}}$ and (b) prior means $\\bm{\\phi}_{jd}$ from simulation version 3.", out.width="100%", fig.asp=1>>= 
-@
-
-<<simulations_gdsc_tab3>>=
-library(knitr, quietly=TRUE, verbose=FALSE)
-suppressWarnings(library(kableExtra))
-options(knitr.kable.NA="", digits=3)
-kable(tab, col.names=colnames(tab), align="r",
-      caption="Median performance measure estimates for simulation version 3 (best performing model in bold).", 
-      format="latex", booktabs=TRUE, escape=FALSE) %>%
-  kable_styling(latex_options=c("HOLD_position"))
-@
+	that NIG$_{\text{f}+\text{d}}$ outperforms the other models in terms of
+	prediction and estimation MSE, with again most of the edge over ridge in the
+	top 10\% of the $\beta_{jd}$.
 
@@
 	Bayes iterations. Currently, problems with number of outcomes in the hundreds
-	and number of features and samples in the thousands are computationally 
+	and number of features and samples up to a thousand are computationally 
 	feasible.
-
-	We average repeated measures over cell line-drug combinations and model the
-	logarithm of the IC50 values. In the following, IC50 refers 
-	to the log-transformed values. 
-	After iteratively removing the cell line or drug with the highest proportion 
-	of missings until no more missing values remain, we end up with IC50 
-	estimates for 482 cell lines on 168 drugs.
 
@@
 	models are: four models like in (\ref{eq:prior}) but with different external 
-	variables, named NIG1 and NIG2 in the following. NIG1 does 
-	not use the external covariates and estimates
-	one overall $\phi$ and one overall $\chi$, while NIG2 uses the external 
-	covariates to estimate both $\bm{\alpha}_{\text{feat}}$ and
-	$\bm{\alpha}_{\text{drug}}$, while NIG3 and NIG4 estimate only 
-	$\bm{\alpha}_{\text{feat}}$ or $\bm{\alpha}_{\text{drug}}$, respectively,
+	variables, named NIG$_{\text{f}+\text{d}}^-$, NIG$_{\text{f}+\text{d}}$,
+	NIG$_{\text{f}}$, and NIG$_{\text{d}}$ in the following. The nomenclature is
+	the same as in Section \ref{sec:simulations}, i.e., 
+	NIG$_{\text{f}+\text{d}}^-$ excludes the external covariates and estimates
+	one overall $\phi$ and one overall $\chi$, while NIG$_{\text{f}+\text{d}}$ 
+	uses the external covariates to estimate both $\bm{\alpha}_{\text{feat}}$ and
+	$\bm{\alpha}_{\text{drug}}$. NIG$_{\text{f}}$ and NIG$_{\text{d}}$ estimate 
+	only $\bm{\alpha}_{\text{feat}}$ or $\bm{\alpha}_{\text{drug}}$, respectively,
 	keeping either $\forall d: \tau_d^2=1$ or $\forall j,d: \gamma_{jd}^2=1$
-	fixed.
+	fixed. In the language of \cite{bernardo_shrink_2011} as introduced in 
+	Section \ref{sec:introduction}, models NIG$_{\text{f}}$ and NIG$_{\text{d}}$
+	may be described as local and global shrinkage rules, respectively, as 
+	opposed to the global-local shrinkage models NIG$_{\text{f}+\text{d}}$ and
+	NIG$_{\text{f}+\text{d}}^-$.
 
@@
 
-tab <- cbind(apply(res[rownames(res)=="elbo", ], 2, median),
-             apply(res[rownames(res)=="elbot", ], 2, median) -
-               apply(res[rownames(res)=="elbo", ], 2, median),
-             apply(res[rownames(res)=="pmse", ], 2, median))
+rownames(est1)[c(1:4)] <- rownames(est2)[c(1:4)] <-
+  c("NIG$_{\\text{f}+\\text{d}}^-$", "NIG$_{\\text{f}+\\text{d}}$",
+    "NIG$_{\\text{f}}$", "NIG$_{\\text{d}}$")
+
+tab <- cbind(apply(res[rownames(res)=="pmse", ], 2, median))
 tabm <- tab
 tabm <- round(tabm, 3)
-tabm[which(tabm[, 1]==max(tabm[, 1], na.rm=TRUE)), 1] <- 
-  paste0("\\textbf{", tabm[which(tabm[, 1]==max(tabm[, 1], na.rm=TRUE)), 1], "}")
-tabm[which(tabm[, 2]==min(tabm[, 2], na.rm=TRUE)), 2] <- 
-  paste0("\\textbf{", tabm[which(tabm[, 2]==min(tabm[, 2], na.rm=TRUE)), 2], "}")
-tabm[which(tabm[, 3]==min(tabm[, 3], na.rm=TRUE)), 3] <- 
-  paste0("\\textbf{", tabm[which(tabm[, 3]==min(tabm[, 3], na.rm=TRUE)), 3], "}")
-colnames(tabm) <- c("$\\text{ELBO}_{\\text{test}}$",
-                   "$\\text{ELBO}_{\\text{train}} 
-                   - \\text{ELBO}_{\\text{test}}$",
-                   "$\\text{PMSE}_{\\text{test}}$")
+tabm[which(tabm[, 1]==min(tabm[, 1], na.rm=TRUE)), 1] <- 
+  paste0("\\textbf{", tabm[which(tabm[, 1]==min(tabm[, 1], na.rm=TRUE)), 1], "}")
+colnames(tabm) <- c("$\\text{PMSE}_{\\text{test}}$")
+rownames(tabm)[c(1:4)] <- rownames(est2)[c(1:4)] <-
+  c("NIG$_{\\text{f}+\\text{d}}^-$", "NIG$_{\\text{f}+\\text{d}}$",
+    "NIG$_{\\text{f}}$", "NIG$_{\\text{d}}$")
 
@@
   in the target pathway of a drug are more predictive of IC50 than genes that
-  are not in the target pathway, according to expectation. Furthermore the NIG2
+  are not in the target pathway, according to expectation. Furthermore the 
+  NIG$_{\text{f}+\text{d}}$
   model estimates positive effects for the experimental and developmental drugs,
   compared to the approved drugs. This translates to smaller expected 
-  prior precisions for the regression parameters for approved drugs, i.e.,
-  approved drugs are easier to predict than experimental and developmental 
-  drugs. On the other hand, the NIG4 model, that only includes external
+  prior precisions for the regression parameters of approved drugs, i.e.,
+  approved drugs are easiest to predict, followed by developmental and 
+  experimental drugs, as expected. On the other hand, the NIG$_{\text{d}}$ 
+  model, that only includes external
   drug covariates, estimates a negative effect for the developmental drugs.
-  The NIG2 model estimates the smallest prior $\bm{\beta}_d$ precision for the
+  The NIG$_{\text{f}+\text{d}}$ model estimates the smallest prior 
+  $\bm{\beta}_d$ precision for the
   targeted drugs, followed by cytotoxic drugs, while drugs with an unkown target
   are assigned the largest prior $\bm{\beta}_d$ precision, according
-  to expectation. Somewhat surprisingly, the NIG4 model estimates the smallest
+  to expectation. Somewhat surprisingly, the NIG$_{\text{d}}$ model
+  estimates the smallest
   prior $\bm{\beta}_d$ precision for the cytotoxic drugs, compared to the 
@@
   $\hat{E}(\gamma_{jd}^2 \cdot \tau_d^2) \in \[ \Sexpr{ranges[2, ]}\]$ for
-  NIG2 and bSEM, respectively. This confirms that the estimated prior 
+  NIG$_{\text{f}+\text{d}}$ and bSEM, respectively. This confirms that the
+  estimated prior 
   $\bm{\beta}_d$ variance components are in the same order of magnitude and that
   the external covariates have a detectable effect on the prior.
-<<analysis_gdsc_est1_feat>>=
-options(knitr.kable.NA="", digits=3)
-kable(est1,
-      col.names=colnames(est1), align="r",
-      caption="$\\bm{\\alpha}_{\\text{feat}}$ estimates from GDSC data.",
-      format="latex", booktabs=TRUE, escape=FALSE) %>%
-  kable_styling(latex_options=c("HOLD_position"))
-@
-
-<<analysis_gdsc_est1_drug>>=
-kable(est2,
-      col.names=colnames(est2), align="r",
-      caption="$\\bm{\\alpha}_{\\text{drug}}$ estimates from GDSC data.",
-      format="latex", booktabs=TRUE, escape=FALSE) %>%
-  kable_styling(latex_options=c("HOLD_position"))
-@
+
   The median PMSEs and ELBOs, calculated on the test data
@@
   error, ridge outperforms the other models 
-  (PMSE of \Sexpr{round(tab[6, 3], 3)}), while bSEM 
-  (PMSE of \Sexpr{round(tab[7, 3], 3)}) performs worse than 
-  the null model (PMSE of \Sexpr{round(tab[8, 3], 3)}). Second and third best
-  performing models are NIG1 and NIG2 (PMSEs of \Sexpr{round(tab[1, 3], 3)} and
-  \Sexpr{round(tab[2, 3], 3)}, respectively), both outperforming the lasso
-  \Sexpr{round(tab[5, 3], 3)}. Part of the differences in performance between
+  (PMSE of \Sexpr{round(tab[6, 1], 3)}), while bSEM 
+  (PMSE of \Sexpr{round(tab[7, 1], 3)}) performs worse than 
+  the null model (PMSE of \Sexpr{round(tab[8, 1], 3)}). Second and third best
+  performing models are NIG$_{\text{f}+\text{d}}^-$ and 
+  NIG$_{\text{f}+\text{d}}$ (PMSEs of \Sexpr{round(tab[1, 1], 3)} and
+  \Sexpr{round(tab[2, 1], 3)}, respectively), both outperforming the lasso
+  with a PMSE of \Sexpr{round(tab[5, 1], 3)}. The underperformance of the 
+  NIG$_{\text{f}}$ and NIG$_{\text{d}}$ models shows the usefulness of a
+  global-local shrinkage prior as opposed to only global or local shrinkage as 
+  as in models NIG$_{\text{f}}$ and NIG$_{\text{d}}$.
+
+  Part of the differences in performance between
   NIG, lasso, and ridge may be explained with the different levels of 
@@
   is excess kurtosis, calculated as \Sexpr{round(kurts[1], 3)}, 
-  \Sexpr{round(kurts[2], 3)}, and \Sexpr{round(kurts[3], 3)} for the NIG2, 
+  \Sexpr{round(kurts[2], 3)}, and \Sexpr{round(kurts[3], 3)} for the 
+  NIG$_{\text{f}+\text{d}}$, 
   lasso, and ridge estimates, respectively. Evidently, lasso gives the 
   ``sparsest'' solution, followed by NIG and ridge. In some sense, this
-  means that the NIG uses less of the $\bm{\beta}_d$ to predict.
-<<analysis_gdsc_tab1>>=
-kable(tabm,
-      col.names=colnames(tabm), align="r",
-      caption="Performance measures calculated on GDSC data (best performing model in bold).",
-      format="latex", booktabs=TRUE, escape=FALSE) %>%
-  kable_styling(latex_options=c("HOLD_position"))
-@
+  means that the NIG uses less of the $\bm{\beta}_d$ to predict than the ridge
+  model. In addition, the penalized regression methods estimate penalty 
+  parameters by cross-validation. Cross-validation directly minimises the 
+  (approximate) PMSE, as opposed to empirical Bayes in the NIG that maximises 
+  the (approximate) marginal likelihood, a measure of model fit. The superior 
+  PMSE of ridge regression is therefore not surprising. A caveat with penalized
+  regression methods is that they do not give measures of parameter uncertainty.
+  NIG, on the other hand, gives the full posterior of the parameters, either
+  through a variational Bayes approximation or with the Gibbs sampler from
+  SM Section 8. The full posterior gives direct access
+  to the parameter uncertainties for a better interpretable model.
 
@@
 	Simulations in Section \ref{sec:simulations} show that the method is 
-	reasonably able to recover hyperparameters. In addition, the simulations
-	show that inclusion of informative external covariates is indeed beneficial to
-	predictive performance. The model is put into practice on the GDSC data. 
+	reasonably able to estimate the hyperparameters. Simulation Scenarios 1 and 2
+	show
+	that estimation of drug- and feature-specific hyperparameters is, in 
+	principle, fairly accurate. However, when estimated jointly, biases may occur
+	due to the interplay between the two sources of information. Nevertheless,
+	the simulations show that inclusion of informative external covariates is 
+	still beneficial to predictive performance.
+
+	The model is put into practice on the GDSC data. 
+	The NIG model is competitive with convential models like lasso in 
+	terms of predictive performance, but is slightly outperformed by ridge.
+	We note that estimation of the penalty parameters in the latter two models
+	is geared towards predictive performance, because cross-validation is used
+	to train these parameters, as opposed to empirical Bayes estimation of the
+	hyperparameters in the NIG model, which aims to improve model fit. In that 
+	sense, the comparison metric (PMSE) is not to the benefit of the NIG model.
+
 	The results in Section
-	\ref{sec:gdsc} show that the inclusion of the external covariates is 
-	beneficial for drug sensitivity prediction. Prediction MSE of the NIG
-	model with both both feature and drug variance components is lower
-	than for the regular lasso model and although the prediction MSE is
-	not lower if external covariates are included, the resulting hyperparameter 
+	\ref{sec:gdsc} show that the inclusion of the external covariates covariates 
+	indeed substantially modifies the hyperparameters. Although, PMSE of the NIG
+	model is not necessarily lower if external covariates are included, the
+	resulting hyperparameter 
 	estimates are still informative and may guide future research efforts.
-	In addition, the comparison to the bSEM model that allows for the inclusion of 
-	one binary external feature covariate favours the proposed NIG model. 
-	Prediction MSEs are almost uniformly lower over the drugs for the NIG model 
-	compared to bSEM. Consequently, the total prediction MSE is significantly
+
+	In addition, the comparison to bSEM, a Bayesian counterpart model that allows
+	for the inclusion of one binary external feature covariate favours the 
+	proposed NIG model. PMSEs are almost uniformly lower over the drugs for the 
+	NIG model compared to bSEM. Consequently, the total PMSE is significantly
 	lower for the NIG model.
+
+	A possible direction of future research to attempt
+	to lower PMSE of the NIG model is the introduction of a cross-validated global
+	variance component in the prior of the $\bm{\beta}_d$. If we add the 
+	additional constraint that the drug and feature-specific variance components
+	average to one, this cross-validated variance component may be interpreted as
+	the overall level of shrinkage. This overall shrinkage level is then tailored 
+	to lower PMSE, while the feature- and drug-specific variance component 
+	include the external information. Other directions of future research are
+	applications of the NIG
+	model to different data types. One possibility is to apply the model in an 
+	eQTL study, 
+	where gene expressions are regressed on SNPs. Several interesting external
+	covariates are available, both on the genes as well as the SNPs. An example of
+	an external
+	covariate for the genes is genelength, where we suspect that longer genes are
+	harder to predict. The distance of the SNP to the gene is an example of an
+	external SNP covariate, where the expectation is that SNPs further from the 
+	gene are less predictive of that gene's expression.
 
@@
 	\section*{Acknowledgements}
-	A visit of MM to SR and GL was funded by the Amsterdam Public Health 
-	institute's Methodology travel grant.
+	MM visited SR and GL on a travel grant funded the Amsterdam Public Health 
+	institute's Methodology program.
 
@@
 
+<<dens_kappa, fig.cap="Implied prior densities $\\pi(\\kappa_{jd})$ for the (a) NIG, (b) Student's $t$, and (c) lasso priors.
+Different line types correspond to different hyperparameter settings.
+The hyperparameter settings were chosen to show some possible, distinct shapes that each of the priors can take.", out.width="100%", fig.asp=1/3>>=
+@
+
+\begin{figure}
+    \centering
+    \tikz{ %
+      % Y
+      \node[obs] (y) {$y_{id}$}; %
+      \factor[right=of y, yshift=1.5cm, xshift=-0.1cm] {y-f} 
+      {above:$\mathcal{N}$} {} {} ; %
+
+      % X
+      \node[det, above=of y, yshift=1cm] (x) {$x_{ij}$}; %
+
+      % sigma2
+      \node[latent, right=of y] (sigma2) {$\sigma_d^2$}; %
+
+      % beta
+      \node[latent, right=of x] (beta) {$\beta_{jd}$}; %
+      \factor[right=of beta, yshift=-1.4cm, xshift=-0.1cm] {beta-f} 
+      {above:$\mathcal{N}$} {} {} ; %
+
+      % gamma2
+      \node[latent, right=of beta] (gamma2) {$\gamma_{jd}^2$}; %
+      \node[const, right=of gamma2, yshift=0.5cm] (phi) {$\phi_{jd}$}; %
+      \node[const, right=of gamma2, yshift=-0.5cm] (lambdaf) 
+      {$\lambda_{\text{feat}}$}; %
+      \factor[right=of gamma2] {gamma2-f} {above:$\mathcal{IG}$} {phi,lambdaf} 
+      {gamma2} ; %
+      \node[det, right=of gamma2, xshift=2cm, yshift=0.5cm] (c) {$c_{jdg}$}; %
+
+      % tau2
+      \node[latent, right=of sigma2] (tau2) {$\tau_d^2$}; %
+      \node[const, right=of tau2, yshift=0.5cm] (chi) {$\chi_{d}$}; %
+      \node[const, right=of tau2, yshift=-0.5cm] (lambdad) 
+      {$\lambda_{\text{drug}}$}; %
+      \factor[right=of tau2] {tau2-f} {above:$\mathcal{IG}$} {chi,lambdad} 
+      {tau2} ; %
+      \node[det, right=of tau2, xshift=2cm, yshift=0.5cm] (z) {$z_{dh}$}; %
+
+      % edges
+      \factoredge {x,beta,sigma2} {y-f} {y} ; %
+      \factoredge {sigma2,gamma2,tau2} {beta-f} {beta} ; %
+      \edge {c} {phi}; %
+      \edge {z} {chi}; %
+
+      % plates
+      \plate{plateg} {(c)} {$g=1,\dots,G$}
+      \plate{plateh} {(z)} {$h=1,\dots,H$}
+      \plate{platei} {(y) (x)}{$i=1,\dots,n$}; %
+      \plate{platej} {(x) (beta) (gamma2) (phi) (lambdaf) (c) (plateg)}
+      {$j=1,\dots,p$}; %
+      \plate{plated} 
+      {(y) (x) (sigma2) (beta) (gamma2) (tau2) (phi) (lambdaf)
+      (chi) (lambdad) (c) (z) (platei) (platej) (plateg) (plateh)}
+      {$d=1,\dots,D$}; %
+    }
+    \caption{Hierarchical representation of the drug sensitivity prediction 
+    model. Grey circles 
+    represent observed variables, white circles represent unobserved variables,
+    squares represent fixed data, and unenclosed letters are parameters to be
+    estimated. Cell lines are indexed by $i$, features
+    by $j$, drugs by $d$, drug covariates by $h$, and feature covariates by 
+    $g$. The $y_{id}$ are the drug sensitivities, $x_{ij}$ the 
+    molecular features, 
+    $c_{jdg}$ the external feature covariates, $z_{dh}$ the external drug
+    covariates, $\beta_{jd}$ the regression coefficients, $\sigma_d^2$ the 
+    error variances, $\tau_d^2$ and $\gamma_{jd}^2$ the drug and feature
+    specific variance components, respectively, and $\phi_{jd}$, 
+    $\lambda_{\text{feat}}$, $\chi_{d}$, and $\lambda_{\text{drug}}$ the 
+    hyperparameters.}
+    \label{fig:dag}
+  \end{figure}
+
+<<simulations_gdsc_est1, fig.cap="Simulation results for Scenario 1 ($\\tau^2_d$ fixed): (a) estimated and true values for $\\alpha_{\\text{feat}}$ and, (b) estimated and true values for prior means $\\phi_{jd}$.", out.width="100%", fig.asp=1/2>>=
+@
+
+<<simulations_gdsc_est2, fig.cap="Simulation results for Scenario 2 ($\\gamma^2_{jd}$ fixed): (a) estimated and true values for $\\alpha_{\\text{feat}}$ and, (b) estimated and true values for prior means $\\phi_{jd}$.", out.width="100%", fig.asp=1/2>>=
+@
+
+<<simulations_gdsc_est3, fig.cap="Simulation results for Scenario 3: (a) estimated and true values for $\\alpha_{\\text{feat}}$ and, (b) estimated and true values for prior means $\\phi_{jd}$.", out.width="100%", fig.asp=1>>=
+@
+
+<<simulations_gdsc_tab1>>=
+res <- read.table("results/simulations_gdsc_res1.txt", row.names=NULL)
+temp <- res[, 1]
+res <- as.matrix(res[, -1])
+rownames(res) <- temp
+true.phi <- mean(1/(c(1, 1, 3, 7) %*% t(unname(model.matrix(~ factor(1:4))))))
+est.phi <- median(1/res[rownames(res)=="alphaf0", 1])
+est.lambdaf <- apply(res[rownames(res)=="lambdaf", c(1, 2)], 2, median)
+
+tab <- cbind(apply(res[rownames(res)=="emsel", ], 2, median),
+             apply(res[rownames(res)=="emseh", ], 2, median),
+             apply(res[rownames(res)=="emse", ], 2, median),
+             apply(res[rownames(res)=="pmse", ], 2, median))
+tab <- round(tab, 3)
+tab[which(tab[, 1]==min(tab[, 1], na.rm=TRUE)), 1] <- 
+  paste0("\\textbf{", tab[which(tab[, 1]==min(tab[, 1], na.rm=TRUE)), 1], "}")
+tab[which(tab[, 2]==min(tab[, 2], na.rm=TRUE)), 2] <- 
+  paste0("\\textbf{", tab[which(tab[, 2]==min(tab[, 2], na.rm=TRUE)), 2], "}")
+tab[which(tab[, 3]==min(tab[, 3], na.rm=TRUE)), 3] <- 
+  paste0("\\textbf{", tab[which(tab[, 3]==min(tab[, 3], na.rm=TRUE)), 3], "}")
+tab[which(tab[, 4]==min(tab[, 4], na.rm=TRUE)), 4] <- 
+  paste0("\\textbf{", tab[which(tab[, 4]==min(tab[, 4], na.rm=TRUE)), 4], "}")
+colnames(tab) <- c("$\\text{EMSE}_{\\text{bottom}}$",
+                   "$\\text{EMSE}_{\\text{top}}$",
+                   "$\\text{EMSE}_{\\text{total}}$",
+                   "$\\text{PMSE}$")     
+rownames(tab)[c(1, 2)] <- c("NIG$_{\\text{f}}^-$",
+                            "NIG$_{\\text{f}}$")
+library(knitr, quietly=TRUE, verbose=FALSE)
+suppressWarnings(library(kableExtra))
+options(knitr.kable.NA="", digits=3)
+kable(tab, col.names=colnames(tab), align="r",
+      caption="Median performance measure for simulation Scenario 1, estimated on the test data (lowest in bold).", 
+      format="latex", booktabs=TRUE, escape=FALSE) %>%
+  kable_styling(latex_options=c("HOLD_position"))
+@
+
+<<simulations_gdsc_tab2>>=
+res <- read.table("results/simulations_gdsc_res2.txt", row.names=NULL)
+temp <- res[, 1]
+res <- as.matrix(res[, -1])
+rownames(res) <- temp
+true.chi <- mean(1/(c(1, 1, 3, 7) %*% t(unname(model.matrix(~ factor(1:4))))))
+est.chi <- median(1/res[rownames(res)=="alphad0", 1])
+est.lambdad <- apply(res[rownames(res)=="lambdad", c(1, 2)], 2, median)
+tab <- cbind(apply(res[rownames(res)=="emsel", ], 2, median),
+             apply(res[rownames(res)=="emseh", ], 2, median),
+             apply(res[rownames(res)=="emse", ], 2, median),
+             apply(res[rownames(res)=="pmse", ], 2, median))
+tab <- round(tab, 3)
+tab[which(tab[, 1]==min(tab[, 1], na.rm=TRUE)), 1] <- 
+  paste0("\\textbf{", tab[which(tab[, 1]==min(tab[, 1], na.rm=TRUE)), 1], "}")
+tab[which(tab[, 2]==min(tab[, 2], na.rm=TRUE)), 2] <- 
+  paste0("\\textbf{", tab[which(tab[, 2]==min(tab[, 2], na.rm=TRUE)), 2], "}")
+tab[which(tab[, 3]==min(tab[, 3], na.rm=TRUE)), 3] <- 
+  paste0("\\textbf{", tab[which(tab[, 3]==min(tab[, 3], na.rm=TRUE)), 3], "}")
+tab[which(tab[, 4]==min(tab[, 4], na.rm=TRUE)), 4] <- 
+  paste0("\\textbf{", tab[which(tab[, 4]==min(tab[, 4], na.rm=TRUE)), 4], "}")
+colnames(tab) <- c("$\\text{EMSE}_{\\text{bottom}}$",
+                   "$\\text{EMSE}_{\\text{top}}$",
+                   "$\\text{EMSE}_{\\text{total}}$",
+                   "$\\text{PMSE}$")      
+rownames(tab)[c(1, 2)] <- c("NIG$_{\\text{d}}^-$",
+                            "NIG$_{\\text{d}}$")
+library(knitr, quietly=TRUE, verbose=FALSE)
+suppressWarnings(library(kableExtra))
+options(knitr.kable.NA="", digits=3)
+kable(tab, col.names=colnames(tab), align="r",
+      caption="Median performance measure for simulation Scenario 2, estimated on the test data (lowest in bold).", 
+      format="latex", booktabs=TRUE, escape=FALSE) %>%
+  kable_styling(latex_options=c("HOLD_position"))
+@
+
+<<simulations_gdsc_tab3>>=
+res <- read.table("results/simulations_gdsc_res3.txt", row.names=NULL)
+temp <- res[, 1]
+res <- as.matrix(res[, -1])
+rownames(res) <- temp
+true.chi <- mean(1/(c(1, 1, 3, 7) %*% t(unname(model.matrix(~ factor(1:4))))))
+est.chi <- median(1/res[rownames(res)=="alphad0", 1])
+est.lambdad <- apply(res[rownames(res)=="lambdad", c(1, 2)], 2, median)
+true.phi <- mean(1/(c(1, 1, 3, 7) %*% t(unname(model.matrix(~ factor(1:4))))))
+est.phi <- median(1/res[rownames(res)=="alphaf0", 1])
+est.lambdaf <- apply(res[rownames(res)=="lambdaf", c(1, 2)], 2, median)
+
+phi.ratio <- apply(apply(sapply(0:3, function(s) {
+  res[rownames(res)==paste0("alphaf" ,s), 2]}) %*% 
+    t(cbind(1, rbind(0, diag(3)))), 2, function(m) {
+      res[rownames(res)=="alphaf0", 2]/m}), 2, median)
+chi.ratio <- apply(apply(sapply(0:3, function(s) {
+  res[rownames(res)==paste0("alphad" ,s), 2]}) %*% 
+    t(cbind(1, rbind(0, diag(3)))), 2, function(m) {
+      res[rownames(res)=="alphad0", 2]/m}), 2, median)
+true.ratio <- as.numeric(1/(c(1, 1, 3, 7) %*% 
+                              t(unname(model.matrix(~ factor(1:4))))))
+
+tab <- cbind(apply(res[rownames(res)=="emsel", ], 2, median),
+             apply(res[rownames(res)=="emseh", ], 2, median),
+             apply(res[rownames(res)=="emse", ], 2, median),
+             apply(res[rownames(res)=="pmse", ], 2, median))
+tab <- round(tab, 3)
+tab[which(tab[, 1]==min(tab[, 1], na.rm=TRUE)), 1] <- 
+  paste0("\\textbf{", tab[which(tab[, 1]==min(tab[, 1], na.rm=TRUE)), 1], "}")
+tab[which(tab[, 2]==min(tab[, 2], na.rm=TRUE)), 2] <- 
+  paste0("\\textbf{", tab[which(tab[, 2]==min(tab[, 2], na.rm=TRUE)), 2], "}")
+tab[which(tab[, 3]==min(tab[, 3], na.rm=TRUE)), 3] <- 
+  paste0("\\textbf{", tab[which(tab[, 3]==min(tab[, 3], na.rm=TRUE)), 3], "}")
+tab[which(tab[, 4]==min(tab[, 4], na.rm=TRUE)), 4] <- 
+  paste0("\\textbf{", tab[which(tab[, 4]==min(tab[, 4], na.rm=TRUE)), 4], "}")
+colnames(tab) <- c("$\\text{EMSE}_{\\text{bottom}}$",
+                   "$\\text{EMSE}_{\\text{top}}$",
+                   "$\\text{EMSE}_{\\text{total}}$",
+                   "$\\text{PMSE}$")     
+rownames(tab)[c(1, 2)] <- c("NIG$_{\\text{f}+\\text{d}}^-$",
+                            "NIG$_{\\text{f}+\\text{d}}$")
+library(knitr, quietly=TRUE, verbose=FALSE)
+suppressWarnings(library(kableExtra))
+options(knitr.kable.NA="", digits=3)
+kable(tab, col.names=colnames(tab), align="r",
+      caption="Median performance measure for simulation Scenario 3, estimated on the test data (lowest in bold).", 
+      format="latex", booktabs=TRUE, escape=FALSE) %>%
+  kable_styling(latex_options=c("HOLD_position"))
+@
+
+<<analysis_gdsc_est1_feat>>=
+fit <- read.table("results/analysis_gdsc_fit1.txt", row.names=NULL)
+temp <- fit[, 1]
+fit <- as.matrix(fit[, -1])
+rownames(fit) <- temp
+est1 <- cbind(fit[, 1], c(NA, 0, 0, NA, 0), fit[, 2])
+est1 <- round(est1, 2)
+est1[is.na(est1)] <- "-"
+colnames(est1) <- c("intercept", "not in pathway", "in pathway")
+est2 <- cbind(fit[, 3:5], c(NA, 0, NA, 0, NA), fit[, c(7:6)], 
+              c(NA, 0, NA, 0, NA))
+est2 <- round(est2, 2)
+est2[is.na(est2)] <- "-"
+colnames(est2) <- c("intercept", "experimental", "development", "approved",
+                    "unkown", "targeted", "cytotoxic")
+rownames(est1)[c(1:4)] <- rownames(est2)[c(1:4)] <-
+  c("NIG$_{\\text{f}+\\text{d}}^-$", "NIG$_{\\text{f}+\\text{d}}$",
+    "NIG$_{\\text{f}}$", "NIG$_{\\text{d}}$")
+options(knitr.kable.NA="", digits=3)
+kable(est1,
+      col.names=colnames(est1), align="r",
+      caption="$\\bm{\\alpha}_{\\text{feat}}$ estimates from GDSC data.",
+      format="latex", booktabs=TRUE, escape=FALSE) %>%
+  kable_styling(latex_options=c("HOLD_position"))
+@
+
+<<analysis_gdsc_est1_drug>>=
+kable(est2,
+      col.names=colnames(est2), align="r",
+      caption="$\\bm{\\alpha}_{\\text{drug}}$ estimates from GDSC data.",
+      format="latex", booktabs=TRUE, escape=FALSE) %>%
+  kable_styling(latex_options=c("HOLD_position"))
+@
+
+<<analysis_gdsc_tab1>>=
+kable(tabm,
+      col.names=colnames(tabm), align="r",
+      caption="Performance measures calculated on GDSC data (best performing model in bold).",
+      format="latex", booktabs=TRUE, escape=FALSE) %>%
+  kable_styling(latex_options=c("HOLD_position"))
+@
+
 \end{document}

### 2020-01-23 code/analysis_eqtl.R

@@ -1,0 +1,421 @@
+#!/usr/bin/env Rscript
+
+################################################################################
+# regression gene expression on SNP data
+################################################################################
+
+# number of cores to use
+ncores <- 50
+
+### libraries
+packages <- c("foreach", "doParallel", "cambridge", "glmnet", "pInc", "rstan")
+sapply(packages, library, character.only=TRUE)
+
+### load data
+load(file="data/data_eqtl_dat1.Rdata")
+
+################################## analysis 1 ##################################
+### data preparation
+y <- scale(expr.prep, scale=FALSE)
+x <- lapply(snps.prep, scale, scale=FALSE)
+G <- length(x)
+p <- sapply(x, ncol)
+n <- nrow(y)
+
+### model fitting
+# setting seed
+set.seed(2019)
+
+# estimation settings
+control <- list(conv.post=TRUE, trace=TRUE, epsilon.eb=1e-3, epsilon.vb=1e-3,
+                maxit.eb=200, maxit.vb=1, maxit.post=100)
+methods <- c("NIG1", "NIG2", "lasso", "ridge", "bSEM")
+
+# model with external covariates
+Z <- model.matrix(~ length, data=gene.prep)
+colnames(Z) <- c("intercept", "length")
+C <- sapply(1:G, function(g) {
+  mat <- cbind(model.matrix( ~ factor(feat.prep$exon[[g]], 
+                                      levels=c("inter", "intron", "exon"))), 
+               feat.prep$distance[[g]],
+               feat.prep$minoraf[[g]])
+  colnames(mat) <- c("intercept", "intron", "exon", "distance", "minoraf")
+  return(mat)})
+fit.semnig1 <- semnig(x=x, y=y, C=C, Z=Z, unpenalized=NULL,
+                      standardize=FALSE, intercept=FALSE, fixed.eb="none",
+                      full.post=TRUE, init=NULL, control=control)
+
+# model without external covariates
+Z <- matrix(1, nrow=G)
+colnames(Z) <- c("intercept")
+C <- lapply(p, function(s) {
+  mat <- matrix(1, nrow=s, ncol=1); colnames(mat) <- "intercept"; return(mat)})
[…]

### 2020-01-23 code/data_gdsc.R

@@ -89,6 +89,35 @@ if(file.exists("data/Ensembl2Reactome_All_Levels.txt")) {
                       header=TRUE, stringsAsFactors=FALSE)
 }
 
+# methylation data
+meth1 <- read.table(unz("data/METH_CELL_DATA.txt.zip", "F2_METH_CELL_DATA.txt"), 
+                    header=TRUE)
+meth2 <- read.xls("data/methSampleId_2_cosmicIds.xlsx", stringsAsFactors=FALSE)
+meth <- t(meth1)
+rownames(meth) <- meth2$cosmic_id[match(
+  paste0("X", meth2$Sentrix_ID, "_", meth2$Sentrix_Position), colnames(meth1))]
+meth <- meth[order(as.numeric(rownames(meth))), ]
+
+# library(biomaRt)
+# library(IlluminaHumanMethylation450kanno.ilmn12.hg19)
+# ann450k <- getAnnotation(IlluminaHumanMethylation450kanno.ilmn12.hg19)
+# annid <- sapply(1:nrow(meth1), function(s) {
+#   which(ann450k$Islands_Name==rownames(meth1)[s])})
+# geneid <- sapply(annid, function(s) {
+#   unique(unlist(strsplit(ann450k$UCSC_RefGene_Name[s], ";")))})
+# island <- sapply(annid, function(s) {
+#   unique(unlist(strsplit(ann450k$Relation_to_Island[s], ";")))})
+# 
+# # View(listEnsembl())
+# # View(listFilters(ensembl))
+# # View(listAttributes(ensembl))
+# ensembl <- useEnsembl(biomart="ensembl", dataset="hsapiens_gene_ensembl")
+# gb <- getBM(attributes=c("external_gene_name", "ensembl_gene_id", 
+#                          "hgnc_symbol"), 
+#             filters=c("external_gene_name"), 
+#             values=unique(unlist(geneid)), 
+#             mart=ensembl)
+
 # combining and removing double drugs and rescreens
 drug <- merge(merge(merge(drug1, drug2, by.x="DRUG_ID", by.y="drug_id",
                           all=TRUE), drug3, by.x="DRUG_ID", by.y="Identifier",
@@ -277,9 +306,10 @@ rownames(expr.temp) <- substr(rownames(expr.temp), 2, 10000L)
 expr <- expr.temp[order(as.numeric(rownames(expr.temp))), ]
 rm(expr.temp)
 
[…]

### 2020-01-30 docs/manuscript1.Rnw
added Bayesian ridge and lasso

@@
 
-	\subsection{Bayesian prior model}
+	\subsection{Bayesian prior model}\label{sec:prior}
 	We carry out inference by endowing the parameters with the following priors:
@@
 	where $\mathcal{IG}(\phi, \lambda)$ denotes an inverse Gaussian distribution 
-	with mean $\phi$ and shape $\lambda$. Prior distributions of the form
+	with mean $\phi$ and shape $\lambda$.
+
+	In model (\ref{eq:prior}), $\gamma_{jd}^2$ denotes a local variance 
+	component that is supposed to capture local, feature-specific variation in the
+	model parameters $\beta_{jd}$, while the global variance component $\tau$ 
+	captures the, drug-specific, general trend in $\bm{\beta}_{d}$. Each drug is
+	endowed with a random error variance $\sigma_d^2$.
+
+	Prior distributions of the form
 	(\ref{eq:prior}) are often referred to as global-local shrinkage rules 
 	\cite[]{bernardo_shrink_2011}, due to the multiplicative separation of the 
-	prior 
-	variance into a local component $\gamma_{jd}^2$ and a global component
-	$\tau_d^2$. The local components are supposed to capture local variation in
-	the model parameters $\beta_{jd}$, while the global component is supposed to
-	capture the general trend in $\bm{\beta}_{d}$. For appropriate 
-	local shrinkage in global-local 
+	prior variance into a local component $\gamma_{jd}^2$ and a global component
+	$\tau_d^2$. For appropriate local shrinkage in global-local 
 	shrinkage models it is important  to account for 

### 2020-01-30 code/figures.R

@@ -183,17 +183,48 @@ temp <- res[, 1]
 res <- as.matrix(res[, -1])
 rownames(res) <- temp
 boxplot(res[rownames(res)=="pmse", ])
-apply(res[rownames(res)=="pmse", ], 2, median)
+sort(apply(res[rownames(res)=="pmse", ], 2, median))
+sort(apply(res[rownames(res)=="elbo", ], 2, median), decreasing=TRUE)
+sort(apply(res[rownames(res)=="lpml", ], 2, median), decreasing=TRUE)
+sort(apply(res[rownames(res)=="brankdist", ], 2, median))
 boxplot(res[rownames(res)=="elbo", ])
 boxplot(res[rownames(res)=="lpml", ])
 boxplot(res[rownames(res)=="brankdist", ])
 
-load("results/analysis_gdsc_fit5.Rdata")
+load("results/analysis_gdsc_fit1.Rdata")
 fit1.semnig$eb$alphaf
 fit1.semnig$eb$alphad
 fit2.semnig$eb$alphaf
 fit2.semnig$eb$alphad
 fit3.semnig$eb$alphaf
-fit3.semnig$eb$alphad
+fit4.semnig$eb$alphad
+fit5.semnig$eb$alphaf
+fit5.semnig$eb$alphad
+
+
+test1 <- Reduce("cbind", fit2.semnig$vb$mu)
+test2 <- sapply(fit1.ridge, function(s) {as.numeric(coef(s, s="lambda.min"))[-1]})
+test3 <- sapply(1:D, function(d) {
+  fit1.optim[[d]]$par[names(fit1.optim[[d]]$par) %in% paste0("beta[", 1:p[d], "]")]})
+plot(test1, test3)
+abline(a=0, b=1, col=2)
+
+plot(colMeans((y - x[[1]] %*% test2)^2),
+      colMeans((y - x[[1]] %*% test3)^2), xlab="ridge", ylab="NIG")
+abline(a=0, b=1, col=2)
+
+mean(colMeans((y - x[[1]] %*% test2)^2))
+mean(colMeans((y - x[[1]] %*% test3)^2))
+
+plot(y - x[[1]] %*% test2, y - x[[1]] %*% test3)
+plot(y, x[[1]] %*% test2)
+plot(y, x[[1]] %*% test3)
+abline(a=0, b=1, col=2)
+
+dim(x[[1]])
+100*168
[…]

### 2020-02-06 docs/supplement1.Rnw
commit before analysis_gdsc.R
added EMSE split for large and small betas to simulations_gdsc.R
added mcmc functions

--- 2019-11-19 docs/supplement1.Rnw
@@
 \setlength{\textwidth}{15cm}
-\title{Supplementary material to: Drug sensitivity prediction with the normal 
-inverse Gaussian prior informed by external data}
+\title{Drug sensitivity prediction with normal inverse Gaussian shrinkage 
+informed by external data}
 \date{\today}
@@
 	$\sigma^2$ and $\tau^2$ are fixed and $\gamma^2 \sim 
-	\mathcal{IG}(\phi,\lambda)$, then
+	\mathcal{IG}(\phi,\lambda_{\text{feat}})$, then
 	\begin{align*}
@@
 	  \mathcal{K}(\beta | \sigma^2, \tau^2) & = 
-	  \frac{3\phi}{\lambda} + 3,
+	  \frac{3\phi}{\lambda_{\text{feat}}} + 3,
 	\end{align*}
-	where $\mathcal{K}(\cdot)$ denotes the kurtosis.
+	where $\mathcal{K}(\cdot)$ denotes the kurtosis. Now letting $\tau^2
+	\sim \mathcal{IG}(\chi,\lambda_{\text{drug}})$ and integrating it out we have:
+	\begin{align*}
+	  \E(\beta | \sigma^2) & = 0, \\
+	  \V(\beta | \sigma^2) & = \sigma^2 \chi \phi, \\
+	  \mathcal{K}(\beta | \sigma^2) & = 
+	  3\(\frac{\phi}{\lambda_{\text{feat}}} \frac{\chi}{\lambda_{\text{drug}}} +
+	  \frac{\phi}{\lambda_{\text{feat}}} + \frac{\chi}{\lambda_{\text{drug}}}\) + 
+	  3.
+	\end{align*}
 
@@
   conditional distributions:
-  \begin{subequations}
+  \begin{subequations}\label{eq:gibbssampler}
     \begin{align}
@@
       \(\frac{n + p + 1}{2}, \zeta_d\), \nonumber \\
-      \text{with } & \zeta_d = \frac{1}{2} \bigg[\mathbf{y} \tr \mathbf{y} -
-      2 \mathbf{y} \tr \X \bm{\mu} + \trace ( \X \tr \X \bm{\Sigma}) + 
-      (\bm{\mu}) \tr \X \tr \X \bm{\mu} + 
-      \tau^{-2}_d \trace (\diag(\gamma_{jd}^{-2}) \bm{\Sigma}) \nonumber \\
-      & \,\,\,\,\,\,\,\,\,\, + \tau_d^{-2} (\bm{\mu}) \tr 
-      \diag(\gamma_{jd}^{-2}) \bm{\mu}\bigg]. \nonumber
+      \text{with } & \zeta_d = \frac{1}{2} \[\mathbf{y}_d \tr \mathbf{y}_d -
+      2 \mathbf{y}_d \tr \X \bm{\beta}_d + \bm{\beta}_d \X \tr \X \bm{\beta}_d + 
+      \tau^{-2}_d \bm{\beta}_d \tr \diag(\gamma_{jd}^{-2}) \bm{\beta}_d) \]. 
+      \nonumber
     \end{align}
@@
   sampling.
+
+  Sampling from (\ref{eq:gibbssampler}) either requires specifying 
+  hyperparameters
+  $\bm{\alpha}_{\text{feat}}$, $\bm{\alpha}_{\text{drug}}$, 
+  $\lambda_{\text{feat}}$, and $\lambda_{\text{drug}}$, or to endow them with
+  and extra layer of hyperpriors. In general, specifying them requires 
+  rarely available, detailed subject knowledge, so we resort to endowing them
+  with vague (improper) hyperpriors:
+  \begin{align*}
+    \bm{\alpha}_{\text{feat}} & \sim c_1 \\
+    \bm{\alpha}_{\text{drug}} & \sim c_2, \\
+    \lambda_{\text{feat}} &\sim 1/\lambda_{\text{feat}}, \\
+    \lambda_{\text{drug}} &\sim 1/\lambda_{\text{drug}},
+  \end{align*}
+  where $c_1$ and $c_2$ are constants. The corresponding full conditionals are
+  given by:
+  \begin{align*}
+    \bm{\alpha}_{\text{feat}} | \lambda_{\text{feat}}, \bm{\Gamma} & \sim 
+    \mathcal{N} \([\mathbf{C} \tr \diag(\gamma_{jd}^2) \mathbf{C}]^{-1}
+    \mathbf{C} \tr \mathbf{1}_{pD \times 1}, \lambda_{\text{feat}}^{-1} 
+    [\mathbf{C} \tr \diag(\gamma_{jd}^2) \mathbf{C}]^{-1} \) \\
+    \bm{\alpha}_{\text{drug}} | \lambda_{\text{drug}}, \bm{\tau} & \sim 
+    \mathcal{N} \([\mathbf{Z} \tr \diag(\tau_{d}^2) \mathbf{Z}]^{-1}
+    \mathbf{Z} \tr \mathbf{1}_{D \times 1}, \lambda_{\text{drug}}^{-1} 
+    [\mathbf{Z} \tr \diag(\tau_{d}^2) \mathbf{Z}]^{-1} \), \\
+    \lambda_{\text{feat}} | \bm{\alpha}_{\text{feat}}, \bm{\Gamma} & 
+    \sim \Gamma \(\frac{pD}{2}, \frac{1}{2} 
+    \[\mathbf{C} \bm{\alpha}_{\text{feat}} - \diag(\gamma_{jd}^{-2}) 
+    \mathbf{1}_{pD \times 1}\] \tr \diag(\gamma_{jd}^2) 
+    \[\mathbf{C} \bm{\alpha}_{\text{feat}} - 
+    \diag(\gamma_{jd}^{-2}) \mathbf{1}_{pD \times 1}\] \), \\
+    \lambda_{\text{drug}} | \bm{\alpha}_{\text{drug}}, \bm{\tau} & 
+    \sim \Gamma \(\frac{D}{2}, \frac{1}{2} 
+    \[\mathbf{Z} \bm{\alpha}_{\text{drug}} - \diag(\tau_{d}^{-2}) 
+    \mathbf{1}_{D \times 1}\] \tr \diag(\tau_{d}^2) 
+    \[\mathbf{Z} \bm{\alpha}_{\text{drug}} - 
+    \diag(\tau_{d}^{-2}) \mathbf{1}_{D \times 1}\] \).
+  \end{align*}
 

### 2020-02-06 rpackage/R/mcmc_functions.R

@@ -1,0 +1,110 @@
+.draw_beta <- function(par, x, y, n, p) {
+  gamma <- par$gamma
+  sigma <- par$sigma
+  tau <- par$tau
+  u <- rnorm(p, 0, (tau*sigma*gamma)^2)
+  delta <- rnorm(n, 0, sigma^2)
+  beta <- u + (t(x)*gamma^2) %*% solve(x %*% (t(x)*gamma^2) + diag(n)/tau^2) %*%
+    (y - x %*% u - delta)  
+  return(beta)
+}
+
+.draw_gamma <- function(par, C, p) {
+  alphaf <- par$alphaf
+  lambdaf <- par$lambdaf
+  beta <- par$beta
+  sigma <- par$sigma
+  tau <- par$tau
+  gamma <- sqrt(rgig(p, -1, chi=beta^2/(sigma*tau)^2 + lambdaf, 
+                     psi=lambdaf*as.numeric(C %*% alphaf)^2))
+  return(gamma)
+}
+
+.draw_tau <- function(par, Z, p) {
+  alphad <- par$alphad
+  lambdad <- par$lambdad
+  beta <- par$beta
+  sigma <- par$sigma
+  gamma <- par$gamma
+  tau <- sqrt(rgig(1, -(p + 1)/2, chi=sum(beta^2/(gamma*sigma)^2) + lambdad, 
+                     psi=lambdad*as.numeric(Z %*% alphad)^2))
+  return(tau)
+}
+
+.draw_sigma <- function(par, x, y, n, p) {
+  beta <- par$beta
+  gamma <- par$gamma
+  tau <- par$tau
+  sigma <- 1/sqrt(rgamma(1, (n + p + 1)/2, 
+                         0.5*(sum(y^2) -2*sum(y*as.numeric(x %*% beta)) + 
+                                as.numeric(x %*% beta)^2 + 
+                                sum(beta^2/gamma^2)/tau^2)))
+  return(sigma)
+}
+
+.draw_alphaf <- function(par, Cmat) {
+  gamma <- unlist(par$gamma)
+  lambdaf <- par$lambdaf
+  mat <- solve(t(Cmat) %*% diag(gamma^2) %*% Cmat)
+  alphaf <- rmvnorm(1, rowSums(mat %*% t(Cmat)), mat/lambdaf)
+  return(alphaf)
+}
+
[…]

### 2020-02-06 docs/supplement1.Rnw
small changes

@@
 	\end{align*}
+
+	\section{Jeffrey's prior}
+	We parametrise the data with a Gaussian distribution with unknown mean and 
+	variance, i.e., $y_i \sim \mathcal{N} (\mu, \sigma^2)$. This gives us the
+	following Hessian (matrix of second derivatives) of the log density:
+	$$
+	\mathbf{H}(\log f(y_i | \mu, \sigma^2)) = \begin{bmatrix}
+	-\frac{1}{\sigma^2} & y_i - \mu \\
+	y_i - \mu & \frac{1}{2\sigma^4} - \frac{(y_i - \mu)^2}{\sigma^6}.
+	\end{bmatrix}
+	$$
+	Jeffrey's prior is now found as:
+	$$
+	\pi(\mu, \sigma^2) \propto |\mathcal{I}(\mu, \sigma^2)|^{1/2} = 
+	|-\E (\mathbf{H})|^{1/2} = \begin{vmatrix}
+	\frac{1}{\sigma^2} & 0 \\
+	0 & \frac{1}{2\sigma^4}
+	\end{vmatrix}^{1/2} \propto 1/\sigma^3,
+	$$
+	where $\mathcal{I}(\theta)$ denotes the Fisher information of $\theta$.
 

### 2020-02-06 code/analysis_gdsc.R

@@ -1138,181 +1138,181 @@ if(ncores > 1) {
 }
 res <- foreach(r=1:nfolds, .packages=packages, .errorhandling="pass", 
                .verbose=TRUE) %dopar% {
-                 cat("\r", "fold", r)
-                 set.seed(2020 + r)
-                 
-                 # splitting data
-                 ntrain <- sum(foldid!=r)
-                 xtrain <- lapply(x, function(s) {
-                   scale(as.matrix(s[foldid!=r, ], nrow=ntrain))})
-                 ytrain <- scale(as.matrix(y[foldid!=r, ], nrow=ntrain))
-                 xtest <- lapply(x, function(s) {
-                   scale(as.matrix(s[foldid==r, ], nrow=n - ntrain))})
-                 ytest <- scale(as.matrix(y[foldid==r, ], nrow=n - ntrain))
-                 
-                 # model without external covariates
-                 control <- list(conv.post=TRUE, trace=FALSE, epsilon.eb=1e-3, epsilon.vb=1e-3, 
-                                 maxit.eb=200, maxit.vb=1, maxit.post=100, maxit.block=0)
-                 Z <- matrix(1, nrow=D)
-                 colnames(Z) <- c("intercept")
-                 C <- lapply(inpathway, function(s) {
-                   s <- matrix(1, nrow=length(s)); colnames(s) <- c("intercept"); return(s)})
-                 cv1.semnig <- semnig(x=xtrain, y=ytrain, C=C, Z=Z, unpenalized=NULL, 
-                                      standardize=FALSE, intercept=FALSE, fixed.eb="none", 
-                                      full.post=TRUE, init=NULL, control=control)
-                 
-                 # models with external covariates
-                 Z <- model.matrix(~ replace(drug.prep$stage, is.na(drug.prep$stage),
-                                             "experimental") + replace(drug.prep$action, is.na(
-                                               drug.prep$action), "unknown"))
-                 colnames(Z) <- c("intercept", "experimental", "in clinical development",
-                                  "targeted", "unknown")
-                 C <- lapply(inpathway, function(s) {
-                   s <- cbind(1, s); colnames(s) <- c("intercept", "in pathway"); return(s)})
-                 cv2.semnig <- semnig(x=xtrain, y=ytrain, C=C, Z=Z, unpenalized=NULL, 
-                                      standardize=FALSE, intercept=FALSE, fixed.eb="none", 
[…]

### 2020-02-06 docs/supplement1.Rnw

@@
   rarely available, detailed subject knowledge, so we resort to endowing them
-  with vague (improper) hyperpriors:
+  with vague (improper) hyperpriors (as in \cite{upadhyay_bayesian_1996}:
   \begin{align*}

### 2020-02-13 docs/supplement1.Rnw
added EBridge

--- 2020-02-06 docs/supplement1.Rnw
@@
   rarely available, detailed subject knowledge, so we resort to endowing them
-  with vague (improper) hyperpriors:
+  with possibly flat, improper (as in \cite{upadhyay_bayesian_1996})
+  hyperpriors:
   \begin{align*}
-    \bm{\alpha}_{\text{feat}} & \sim c_1 \\
-    \bm{\alpha}_{\text{drug}} & \sim c_2, \\
-    \lambda_{\text{feat}} &\sim 1/\lambda_{\text{feat}}, \\
-    \lambda_{\text{drug}} &\sim 1/\lambda_{\text{drug}},
+    \bm{\alpha}_{\text{feat},g} | \lambda_{\text{feat}} & \sim 
+    \mathcal{N}(0, \nu_{\text{feat}}^2/\lambda_{\text{feat}}), \\
+    \bm{\alpha}_{\text{drug},h} | \lambda_{\text{drug}} & \sim 
+    \mathcal{N}(0, \nu_{\text{drug}}^2/\lambda_{\text{drug}}), \\
+    \lambda_{\text{feat}} & \sim \Gamma(\kappa_{\text{feat}}, 
+    \xi_{\text{feat}}), \\
+    \lambda_{\text{drug}} & \sim \Gamma(\kappa_{\text{drug}}, 
+    \xi_{\text{drug}}),
   \end{align*}
-  where $c_1$ and $c_2$ are constants. The corresponding full conditionals are
+  with hyperparameters $\nu_{\text{feat}}^2$, $\nu_{\text{drug}}^2$, 
+  $\kappa_{\text{feat}}$, $\xi_{\text{feat}}$, $\kappa_{\text{drug}}$, and
+  $\xi_{\text{drug}}$. Improper priors occur if we let 
+  $\nu_{\text{feat}}, \nu_{\text{drug}} \to \infty$ and 
+  $\kappa_{\text{feat}}, \kappa_{\text{drug}}, \xi_{\text{feat}}, 
+  \xi_{\text{drug}} \to 0$. The corresponding full conditionals are
   given by:
@@
     \bm{\alpha}_{\text{feat}} | \lambda_{\text{feat}}, \bm{\Gamma} & \sim 
-    \mathcal{N} \([\mathbf{C} \tr \diag(\gamma_{jd}^2) \mathbf{C}]^{-1}
-    \mathbf{C} \tr \mathbf{1}_{pD \times 1}, \lambda_{\text{feat}}^{-1} 
-    [\mathbf{C} \tr \diag(\gamma_{jd}^2) \mathbf{C}]^{-1} \) \\
+    \mathcal{N} \( \mathbf{m}_{\text{feat}}, \mathbf{S}_{\text{feat}} \) \\
+    \text{with } & \mathbf{S}_{\text{feat}} = \lambda_{\text{feat}}^{-1} 
+    [\mathbf{C} \tr \diag(\gamma_{jd}^2) \mathbf{C} +
+    \nu_{\text{feat}}^{-2} \I_G]^{-1}, \\
+    \text{and } & \mathbf{m}_{\text{feat}} = [\mathbf{C} \tr 
+    \diag(\gamma_{jd}^2) \mathbf{C} + \nu_{\text{feat}}^{-2} \I_G]^{-1}
+    \mathbf{C} \tr \mathbf{1}_{pD \times 1}, \\
     \bm{\alpha}_{\text{drug}} | \lambda_{\text{drug}}, \bm{\tau} & \sim 
-    \mathcal{N} \([\mathbf{Z} \tr \diag(\tau_{d}^2) \mathbf{Z}]^{-1}
-    \mathbf{Z} \tr \mathbf{1}_{D \times 1}, \lambda_{\text{drug}}^{-1} 
-    [\mathbf{Z} \tr \diag(\tau_{d}^2) \mathbf{Z}]^{-1} \), \\
+    \mathcal{N} \(\textbf{m}_{\text{drug}}, \mathbf{S}_{\text{drug}} \), \\
+    \text{with } & \mathbf{S}_{\text{drug}} = \lambda_{\text{drug}}^{-1} 
+    [\mathbf{Z} \tr \diag(\tau_{d}^2) \mathbf{Z} +
+    \nu_{\text{drug}}^{-2} \I_H]^{-1}, \\
+    \text{and } & \mathbf{m}_{\text{drug}} = [\mathbf{Z} \tr 
+    \diag(\tau_{d}^2) \mathbf{Z} + \nu_{\text{drug}}^{-2} \I_H]^{-1}
+    \mathbf{Z} \tr \mathbf{1}_{D \times 1}, \\
     \lambda_{\text{feat}} | \bm{\alpha}_{\text{feat}}, \bm{\Gamma} & 
-    \sim \Gamma \(\frac{pD}{2}, \frac{1}{2} 
+    \sim \Gamma \(\frac{pD}{2} + \kappa_{\text{feat}}, 
+    \upsilon_{\text{feat}}\), \\
+    \text{with } & \upsilon_{\text{feat}} =\frac{1}{2} 
     \[\mathbf{C} \bm{\alpha}_{\text{feat}} - \diag(\gamma_{jd}^{-2}) 
@@
     \[\mathbf{C} \bm{\alpha}_{\text{feat}} - 
-    \diag(\gamma_{jd}^{-2}) \mathbf{1}_{pD \times 1}\] \), \\
+    \diag(\gamma_{jd}^{-2}) \mathbf{1}_{pD \times 1}\] + \xi_{\text{feat}}, \\
     \lambda_{\text{drug}} | \bm{\alpha}_{\text{drug}}, \bm{\tau} & 
-    \sim \Gamma \(\frac{D}{2}, \frac{1}{2} 
+    \sim \Gamma \(\frac{D}{2} + \kappa_{\text{drug}}, 
+    \upsilon_{\text{drug}} \), \\
+    \text{with } & \upsilon_{\text{drug}} =\frac{1}{2} 
     \[\mathbf{Z} \bm{\alpha}_{\text{drug}} - \diag(\tau_{d}^{-2}) 
@@
     \[\mathbf{Z} \bm{\alpha}_{\text{drug}} - 
-    \diag(\tau_{d}^{-2}) \mathbf{1}_{D \times 1}\] \).
+    \diag(\tau_{d}^{-2}) \mathbf{1}_{D \times 1}\] + \xi_{\text{drug}}.
   \end{align*}

### 2020-02-13 rpackage/R/main_functions.R

@@ -421,3 +421,76 @@ cv.semnig <- function(x, y, nfolds=10, foldid=NULL, seed=NULL, phi=phi,
   
   return(fit)
 }
+
+# EBridge estimation
+ebridge <- function(x, y, C, Z, mult.lambda=FALSE,
+                    hyper=list(lambda=NULL, zeta=0, nu=0),
+                    control=list(epsilon=sqrt(.Machine$double.eps), 
+                                 maxit=500, trace=FALSE)) {
+  
+  yty <- colSums(y^2)
+  Cmat <- Reduce("rbind", C)
+  p <- ncol(x)
+  n <- nrow(y)
+  D <- ncol(y)
+  H <- ncol(Z)
+  G <- ncol(Cmat)
+  
+  if(is.null(hyper$lambda)) {
+    cv.fit1 <- lapply(1:D, function(d) {
+      cv.glmnet(x, y[, d], alpha=0, intercept=FALSE)})
+    if(mult.lambda) {
+      hyper$lambda <- 1/sqrt(n*sapply(cv.fit1, "[[", "lambda.min"))
+    } else {
+      hyper$lambda <- rep(1/sqrt(n*exp(mean(log(sapply(cv.fit1, "[[", 
+                                                       "lambda.min"))))), D)  
+    }
+  } else {
+    if(length(hyper$lambda)==1) {
+      hyper$lambda <- rep(hyper$lambda, D)  
+    }
+  }
+  cv.fit2 <- lapply(1:D, function(d) {
+    glmnet(x, y[, d], alpha=0, lambda=1/(exp(2*mean(log(hyper$lambda)))*n), 
+           intercept=FALSE)})  
+  
+  control$fnscale=-1
+  names(control)[1] <- "reltol"
+  opt <- optim(par=rep(0, H + G), fn=.f.optim, lambda=hyper$lambda, 
+               nu=hyper$nu, zeta=hyper$zeta, Cmat=Cmat, Z=Z, n=n, p=p, D=D, G=G, 
+               y=y, x=x, yty=yty, control=control)
+  conv <- opt$convergence
+  niter <- unname(opt$counts[1])
+  alphaf <- opt$par[1:G]
+  alphad <- opt$par[-c(1:G)]
+  tau <- exp(colSums(alphad*t(Z))/2)
+  gamma <- exp(colSums(alphaf*t(Cmat))/2)
+  
+  beta1 <- sapply(1:D, function(d) {
[…]

### 2020-03-04 docs/manuscript1.Rnw
commit bfeore push

@@
 		  \label{eq:betaprior}\\
-		  \gamma_{jd}^{2} & \sim \mathcal{IG}(\phi_{jd}, \lambda_{\text{feat}}), \\
+		  \gamma_{jd}^{2} & \sim \mathcal{IG}(\phi_{jd}, \lambda_{\text{feat}}), 
+		  \label{eq:gammaprior}\\
 		  \tau_d^{2} & \sim \mathcal{IG}(\chi_d, \lambda_{\text{drug}}), 
 		  \label{eq:tauprior}\\
-		  \sigma_d^{2} & \sim 1/\sigma_d^{3},
+		  \sigma_d^{2} & \sim 1/\sigma_d^{3}, \label{eq:sigmaprior}
 		\end{align}
@@
 
-	In model (\ref{eq:prior}), $\gamma_{jd}^2$ denotes a local variance 
+	In model (\ref{eq:prior}), $\gamma_{jd}^2$ in (\ref{eq:gammaprior}) denotes a 
+	local variance 
 	component that is supposed to capture local, feature-specific variation in the
-	model parameters $\beta_{jd}$, while the global variance component $\tau$ 
-	captures the, drug-specific, general trend in $\bm{\beta}_{d}$. Each drug is
-	endowed with a random error variance $\sigma_d^2$.
+	model parameters $\beta_{jd}$ in (\ref{eq:betaprior}), 
+	while the global variance components $\tau_d^2$ in (\ref{eq:tauprior})
+	capture the drug-specific, general trend in $\bm{\beta}_{d}$. Each drug 
+	response is endowed with a random error variance $\sigma_d^2$, distributed
+	according to (\ref{eq:sigmaprior}).
 
@@
 Different line types correspond to different hyperparameter settings.
-The hyperparameter settings were chosen to show some possible, distinct shapes that each of the priors can take.", out.width="100%", fig.asp=1/3>>=
+The hyperparameter settings (given in Section \\ref{sm-sec:hyperparameters} of the SM) were chosen to show some possible, distinct shapes that each of the priors can take.", out.width="100%", fig.asp=1/3>>=
 @
@@
   of hyperparameters, the NIG prior can behave similarly to the Student's $t$ 
-  prior (decreasing form zero, with substantial mass close to 0), 
-  but also rather differently. Our argumentation to model the 
-  $\gamma^2_{jd}$ by an inverse Gaussian 
+  prior (decreasing form zero, with substantial mass close to zero and 
+  little mass close to one, like the
+  solid lines in Figure \ref{fig:dens_kappa}a-\ref{fig:dens_kappa}b), 
+  but also rather differently (dashed and dotted lines in 
+  Figure \ref{fig:dens_kappa}a-\ref{fig:dens_kappa}b). Our argumentation to 
+  model the $\gamma^2_{jd}$ by an inverse Gaussian 
   distribution, as has been suggested in \cite{fabrizi_specification_2016} and 
@@
   more conveniently than the Student's $t$ prior, as explained in Section 
-  \ref{sec:empiricalbayes}, and (iii) like the horseshoe, the NIG shrinkage
-  weights prior can put mass both near 0 and 1, a desirable property of 
-  shrinkage priors.
+  \ref{sec:empiricalbayes}, and (iii) like the horseshoe 
+  \cite[]{carvalho_handling_2009}, the NIG shrinkage
+  weights prior can put mass both near zero and one, a desirable property of 
+  shrinkage priors \cite[]{bernardo_shrink_2011}.
 
@@
 	NIG$_{\text{f}+\text{d}}$, again depending on which variance components are 
-	estimated, and regular (iii) lasso and (iv) ridge models.
+	estimated, and frequentist (iii) lasso and (iv) ridge models.
 	Exclusion of the external covariates as in the NIG$_{\text{f}}^-$, 
@@
 	NIG$_{\text{f}}$, NIG$_{\text{d}}$, and NIG$_{\text{f}+\text{d}}$ models. 
+	The lasso and ridge models were fit using the \texttt{R} package 
+	\texttt{glmnet} \cite[]{friedman_regularization_2010} with cross validated
+	penalty parameters.
 	All models are fit on 241 randomly selected cell lines, while performance

### 2020-03-04 docs/supplement1.Rnw
commit bfeore push

@@
 
-	\section{Jeffrey's prior}
+	\section{Hyperparameter settings}\label{sec:hyperparameters}
+	For the NIG shrinkage weights prior in Figure \ref{md-fig:dens_kappa}a the 
+	following hyperparameter settings (in terms of the inverse Gaussian prior) 
+	were used: $\lambda=1$, $\phi=10$ (solid),
+	$\lambda=0.1$, $\phi=10$ (dashed), and $\lambda=1$, $\phi=1$ (dotted).
+	The Student`s $t$ prior on the regression parameters $\beta_{jd}$ arises if
+	$\gamma_{jd}^2 \sim \Gamma^{-1}(a,b)$, with $\Gamma^{-1}(a,b)$ an inverse
+	Gamma distribution with shape $a$ and scale $b$. Figure 
+	\ref{md-fig:dens_kappa}b is then obtained by setting: $a=0.9$, $b=1.3$ 
+	(solid), $a=1$, $b=4$ (dashed), and $a=1$, $b=0.2$ (dotted).
+	The lasso prior results from an exponential prior on $\gamma_{jd}^2$. To
+	create Figure \ref{md-fig:dens_kappa}c the following settings were
+	used for rate $\lambda$: $\lambda=1$ (solid), $\lambda=0.2$ (dashed), and 
+	$\lambda=0.1$ (dotted).
+
+	\section{Jeffrey's prior}\label{sec:jeffreysprior}
 	We parametrise the data with a Gaussian distribution with unknown mean and 

### 2020-03-04 code/figures.R

@@ -1,7 +1,7 @@
 ################################# main document ################################
 # ---- figures ----
 # ---- dens_kappa ----
-library(sp)
+suppressWarnings(suppressMessages(library(sp)))
 labels <- c("NIG", "Student's t", "lasso")
 col <- bpy.colors(length(labels), cutoff.tail=0.3)
 lty <- c(1:length(labels))
@@ -33,8 +33,8 @@ lines(x.seq, dknig(x.seq, 1, 1), col=col[1], lty=lty[3])
 
 plot(x.seq, dkstudentst(x.seq, a=0.9, b=1.3), type="l", ylim=c(0, 4), 
      xlab=expression(kappa), ylab=expression(pi(kappa)), main="(b)", col=col[2], 
-     lty=lty[1])
-lines(x.seq, dkstudentst(x.seq, a=1, b=4), col=col[2], lty=lty[2])
+     lty=lty[2])
+lines(x.seq, dkstudentst(x.seq, a=1, b=4), col=col[2], lty=lty[1])
 lines(x.seq, dkstudentst(x.seq, a=1, b=0.2), col=col[2], lty=lty[3])
 
 plot(x.seq, dklasso(x.seq, 1), type="l", ylim=c(0, 3.5), 
@@ -45,7 +45,7 @@ lines(x.seq, dklasso(x.seq, 0.1), col=col[3], lty=lty[3])
 par(opar)
 
 # ---- simulations_gdsc_est1 ----
-library(sp)
+suppressWarnings(suppressMessages(library(sp)))
 res <- read.table("results/simulations_gdsc_res1.txt", row.names=NULL)
 temp <- res[, 1]
 res <- as.matrix(res[, -1])
@@ -77,7 +77,7 @@ points(c(1:4), phi, pch=2, col=col[c(1:4)], cex=1.5,
 par(opar)
 
 # ---- simulations_gdsc_est2 ----
-library(sp)
+suppressWarnings(suppressMessages(library(sp)))
 res <- read.table("results/simulations_gdsc_res2.txt", row.names=NULL)
 temp <- res[, 1]
 res <- as.matrix(res[, -1])
@@ -109,7 +109,7 @@ points(c(1:4), phi, pch=2, col=col[c(1:4)], cex=1.5,
 par(opar)
 
 # ---- simulations_gdsc_est3 ----
-library(sp)
[…]

### 2020-03-04 code/simulations_gdsc.R

@@ -664,7 +664,7 @@ res <- foreach(r=1:nreps, .packages=packages, .errorhandling="pass") %dopar% {
   Z <- cbind(1, t(replicate(D, sample(c(1, 0, 0, 0), 4)[-1])))
   sigma <- rep(1, D)
   beta <- lapply(1:D, function(d) {
-    rnorm(p[d], 0, sigma[d]*sqrt(1/as.numeric(C[[d]] %*% alphaf))*
+    rnorm(p[d], 0, sqrt(1/as.numeric(C[[d]] %*% alphaf))*
             sqrt(1/as.numeric(Z %*% alphad)))})
   
   # simulate data

### 2020-03-30 docs/manuscript1.Rnw
changed simulations

@@
   \end{figure}
+
+  \subsection{Alternative prior}\label{sec:altprior}
+  Consider the following simpler hyperprior alternative to (\ref{eq:prior}):
+  \begin{subequations}\label{eq:altprior}
+		\begin{align}
+		  \beta_{jd} | \gamma^2_{jd}, \tau_d^2, \sigma^2_d & \sim \mathcal{N}_{p} 
+		  (0, \gamma_{jd}^2 \tau_d^2 \sigma_d^2), \\
+		  \log \gamma_{jd}^{2} & = \mathbf{c}_{jd} \tr \bm{\alpha}_{\text{feat}}, \\
+		  \log \tau_d^{2} & = \mathbf{z}_{d} \tr \bm{\alpha}_{\text{drug}}\\
+		  \sigma_d^{2} & \sim \Gamma^{-1} (\nu, \zeta).
+		\end{align}
+	\end{subequations}
+	This model is a Bayesian, multivariate extension of 
+	\cite{zeng_incorporating_2020}. The model does 
+	not propagate uncertainty in the variance components $\gamma_{jd}^{2}$
+	and $\tau_d^{2}$, but directly models the variance components as an
+	exponentiated linear function of the external covariates. The linear functions
+	are modelled on the log-scale to ensure non-negative variance components.
+	Additionally, it allows for more flexibility in the error variance
+	prior model. Note that in the limit $\zeta \to 0$ and $\nu=1/2$, the error
+	variance follows the Jeffrey`s prior from (\ref{eq:prior}). Throughout the
+	following, whenever $\zeta$ and $\nu$ are not specified, we use $\zeta \to 0$ 
+	and $\nu=1/2$.
+
+	Estimation of $\bm{\alpha}_{\text{feat}}$ and $\bm{\alpha}_{\text{feat}}$, and
+	ultimately $\bm{\beta}_d$ is explained in Section \ref{sm-sec:altprior} of the
+	SM. The results in Sections \ref{sec:simulations} and \ref{sec:gdsc}
+	include model (\ref{eq:altprior}) as an alternative to (\ref{eq:prior}).
 
@@
 
-<<analysis_gdsc_fit1>>=
-fit <- read.table("results/analysis_gdsc_fit1.txt", row.names=NULL)
-temp <- fit[, 1]
-fit <- as.matrix(fit[, -1])
-rownames(fit) <- temp
-
-res <- read.table("results/analysis_gdsc_res1.txt", row.names=NULL)
-temp <- res[, 1]
-res <- as.matrix(res[, -1])
-rownames(res) <- temp
-
-est1 <- cbind(fit[, 1], c(NA, 0, 0, NA, 0), fit[, 2])
-est1 <- round(est1, 2)
-est1[is.na(est1)] <- "-"
-colnames(est1) <- c("intercept", "not in pathway", "in pathway")
-est2 <- cbind(fit[, 3:5], c(NA, 0, NA, 0, NA), fit[, c(7:6)], 
-              c(NA, 0, NA, 0, NA))
-est2 <- round(est2, 2)
-est2[is.na(est2)] <- "-"
-colnames(est2) <- c("intercept", "experimental", "development", "approved",
-                    "unkown", "targeted", "cytotoxic")
-
-rownames(est1)[c(1:4)] <- rownames(est2)[c(1:4)] <-
-  c("NIG$_{\\text{f}+\\text{d}}^-$", "NIG$_{\\text{f}+\\text{d}}$",
-    "NIG$_{\\text{f}}$", "NIG$_{\\text{d}}$")
-
-tab <- cbind(apply(res[rownames(res)=="pmse", ], 2, median))
-tabm <- tab
-tabm <- round(tabm, 3)
-tabm[which(tabm[, 1]==min(tabm[, 1], na.rm=TRUE)), 1] <- 
-  paste0("\\textbf{", tabm[which(tabm[, 1]==min(tabm[, 1], na.rm=TRUE)), 1], "}")
-colnames(tabm) <- c("$\\text{PMSE}_{\\text{test}}$")  
-rownames(tabm)[c(1:4)] <- rownames(est2)[c(1:4)] <-
-  c("NIG$_{\\text{f}+\\text{d}}^-$", "NIG$_{\\text{f}+\\text{d}}$",
-    "NIG$_{\\text{f}}$", "NIG$_{\\text{d}}$")
-
-library(glmnet)
-kurtosis <- function(x) {
-  n <- length(x) 
-  kurt <- (n + 1)*n*sum((x - mean(x))^4)/((n - 1)*(n - 2)*(n - 3)*var(x)^2) - 
-    3*(n - 1)^2/((n - 2)*(n - 3))
-  replace(kurt, !is.finite(kurt), NA)
-}
-load("results/analysis_gdsc_fit1.Rdata")
-best2.semnig <- fit2.semnig$vb$mu
-best1.lasso <- lapply(fit1.lasso, function(s) {
-  unname(coef(s, "lambda.min")[-1, ])})
-best1.ridge <- lapply(fit1.ridge, function(s) {
-  unname(coef(s, "lambda.min")[-1, ])})
-best1.bSEM <- lapply(fit1.bSEM$vb$beta, function(s) {s[, 2]})
-kurts <- c(mean(sapply(best2.semnig, kurtosis), na.rm=TRUE), 
-           mean(sapply(best1.lasso, kurtosis), na.rm=TRUE),
-           mean(sapply(best1.ridge, kurtosis), na.rm=TRUE), 
-           mean(sapply(best1.bSEM, kurtosis), na.rm=TRUE))
-ranges <- rbind(range(unlist(lapply(1:length(
-  fit2.semnig$eb$mpriord), function(d) {
-    fit2.semnig$eb$mpriord[d]*fit2.semnig$eb$mpriorf[[d]]}))),
-  fit1.bSEM$eb$b[nrow(fit1.bSEM$eb$b), ]/
-    (fit1.bSEM$eb$a[nrow(fit1.bSEM$eb$a), ] - 1))
+<<analysis_gdsc>>==
+# fit <- read.table("results/analysis_gdsc_fit1.txt", row.names=NULL)
+# temp <- fit[, 1]
+# fit <- as.matrix(fit[, -1])
+# rownames(fit) <- temp
+# 
+# res <- read.table("results/analysis_gdsc_res1.txt", row.names=NULL)
+# temp <- res[, 1]
+# res <- as.matrix(res[, -1])
+# rownames(res) <- temp
+# 
+# est1 <- cbind(fit[, 1], c(NA, 0, 0, NA, 0), fit[, 2])
+# est1 <- round(est1, 2)
+# est1[is.na(est1)] <- "-"
+# colnames(est1) <- c("intercept", "not in pathway", "in pathway")
+# est2 <- cbind(fit[, 3:5], c(NA, 0, NA, 0, NA), fit[, c(7:6)],
+#               c(NA, 0, NA, 0, NA))
+# est2 <- round(est2, 2)
+# est2[is.na(est2)] <- "-"
+# colnames(est2) <- c("intercept", "experimental", "development", "approved",
+#                     "unkown", "targeted", "cytotoxic")
+# 
+# rownames(est1)[c(1:4)] <- rownames(est2)[c(1:4)] <-
+#   c("NIG$_{\\text{f}+\\text{d}}^-$", "NIG$_{\\text{f}+\\text{d}}$",
+#     "NIG$_{\\text{f}}$", "NIG$_{\\text{d}}$")
+# 
+# tab <- cbind(apply(res[rownames(res)=="pmse", ], 2, median))
+# tabm <- tab
+# tabm <- round(tabm, 3)
+# tabm[which(tabm[, 1]==min(tabm[, 1], na.rm=TRUE)), 1] <-
+#   paste0("\\textbf{", tabm[which(tabm[, 1]==min(tabm[, 1], na.rm=TRUE)), 1], "}")
+# colnames(tabm) <- c("$\\text{PMSE}_{\\text{test}}$")
+# rownames(tabm)[c(1:4)] <- rownames(est2)[c(1:4)] <-
+#   c("NIG$_{\\text{f}+\\text{d}}^-$", "NIG$_{\\text{f}+\\text{d}}$",
+#     "NIG$_{\\text{f}}$", "NIG$_{\\text{d}}$")
+# 
+# library(glmnet)
+# kurtosis <- function(x) {
+#   n <- length(x)
+#   kurt <- (n + 1)*n*sum((x - mean(x))^4)/((n - 1)*(n - 2)*(n - 3)*var(x)^2) -
+#     3*(n - 1)^2/((n - 2)*(n - 3))
+#   replace(kurt, !is.finite(kurt), NA)
+# }
+# load("results/analysis_gdsc_fit1.Rdata")
+# best2.semnig <- fit2.semnig$vb$mu
+# best1.lasso <- lapply(fit1.lasso, function(s) {
+#   unname(coef(s, "lambda.min")[-1, ])})
+# best1.ridge <- lapply(fit1.ridge, function(s) {
+#   unname(coef(s, "lambda.min")[-1, ])})
+# best1.bSEM <- lapply(fit1.bSEM$vb$beta, function(s) {s[, 2]})
+# kurts <- c(mean(sapply(best2.semnig, kurtosis), na.rm=TRUE),
+#            mean(sapply(best1.lasso, kurtosis), na.rm=TRUE),
+#            mean(sapply(best1.ridge, kurtosis), na.rm=TRUE),
+#            mean(sapply(best1.bSEM, kurtosis), na.rm=TRUE))
+# ranges <- rbind(range(unlist(lapply(1:length(
+#   fit2.semnig$eb$mpriord), function(d) {
+#     fit2.semnig$eb$mpriord[d]*fit2.semnig$eb$mpriorf[[d]]}))),
+#   fit1.bSEM$eb$b[nrow(fit1.bSEM$eb$b), ]/
+#     (fit1.bSEM$eb$a[nrow(fit1.bSEM$eb$a), ] - 1))
+ranges <- matrix(1, nrow=10, ncol=10)
+tab <- matrix(1, nrow=10, ncol=10)
+tabm <- matrix(1, nrow=10, ncol=10)
+kurts <- matrix(1, nrow=10, ncol=10)
 @
@@
 <<time>>=
-load("results/analysis_gdsc_fit1.Rdata")
-time2.semnig <- sum(fit2.semnig$time)
+# load("results/analysis_gdsc_fit1.Rdata")
+# time2.semnig <- sum(fit2.semnig$time)
+time2.semnig <- 1
 @

### 2020-03-30 docs/supplement1.Rnw
changed simulations

@@
 
+  \section{Estimation of alternative prior}\label{sec:altprior}
+  The hyperparameters $\bm{\alpha} = \begin{bmatrix} \bm{\alpha}_{\text{feat}}
+  \tr & \bm{\alpha}_{\text{drug}} \tr \end{bmatrix} \tr$ in model 
+  (\ref{md-eq:altprior}) are estimated by empirical Bayes maximisation of
+  the marginal log likelihood:
+  \begin{subequations}\label{eq:altmll}
+    \begin{align}
+      \log p (\bm{\alpha} | \mathbf{Y}) & = \sum_{d=1}^D
+      \log \[ \int_{\bm{\beta}_d} \int_{\sigma^2_d} \mathcal{L} 
+      (\mathbf{y}_d | \bm{\beta}_d, \sigma^2_d ) \pi(\bm{\beta}_d | \sigma^2_d) 
+      \pi(\sigma^2_d) d\bm{\beta}_d d\sigma_d^2 \] \\
+      & \propto - \sum_{d=1}^D \bigg[ \log \left|\I + \mathbf{X} 
+      \diag \[ \exp\( \tilde{\mathbf{C}}_d \bm{\alpha} \) \] \mathbf{X} \tr 
+      \right| \\
+      & \,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\, + (n + 1) \log \( \mathbf{y}_d \tr 
+      \left\{ \I + \mathbf{X} 
+      \diag \[ \exp\( \tilde{\mathbf{C}}_d \bm{\alpha} \) \] \mathbf{X} 
+      \tr\right\}^{-1} \mathbf{y}_d \) \bigg],
+    \end{align}
+  \end{subequations}
+  where $\tilde{\mathbf{C}}_d = \begin{bmatrix} \mathbf{C}_d & 
+  \mathbf{1}_{p \times 1} \cdot \mathbf{z}_d \tr \end{bmatrix}$, and
+  $\mathbf{C}_d = \begin{bmatrix} \mathbf{c}_{1d} & \cdots & 
+  \mathbf{c}_{pd} \end{bmatrix} \tr$.
+  The first term in the sum of (\ref{eq:altmll}) is concave in 
+  $\bm{\alpha}$, while the second term is convex in $\bm{\alpha}$. This renders
+  the marginal log likelihood a sum of non-concave (as well as non-convex) 
+  functions. General purpose optimisers, such as \texttt{optim} in \texttt{R} 
+  are generally not well suited to maximise non-concave functions. Maximisation
+  is therefore done through a minorise-maximisation (MM) algorithm, 
+  similar to \cite{zeng_incorporating_2020}.
+
+  The MM algorithm seeks to minorise the marginal log likelihood,
+  $\log p(\bm{\alpha} | \mathbf{Y})$, at current
+  parameter estimate $\bm{\alpha}^{(t)}$ with function 
+  $g(\bm{\alpha} | \bm{\alpha}^{(t)})$, such that
+  $\forall \bm{\alpha}: g(\bm{\alpha} | \bm{\alpha}^{(t)}) \leq 
+  \log p(\bm{\alpha} | \mathbf{Y})$ 
+  and $g(\bm{\alpha}^{(t)} | \bm{\alpha}^{(t)}) = 
+  \log p(\bm{\alpha} | \mathbf{Y})$. Then, the sequence
+  $$
+  \bm{\alpha}^{(t + 1)} = \underset{\bm{\alpha}}{\argmax} 
+  g(\bm{\alpha} | \bm{\alpha}^{(t+1)})
+  $$
+  converges to a local maximum of $f(\bm{\alpha})$. Generally, the minorisation
+  function is chosen such that it is concave and easy to maximise. In our case,
+  a suitable minorisation is:
+  \begin{subequations}\label{eq:minor}
+    \begin{align}
+      g(\bm{\alpha} | \bm{\alpha}^{(t+1)}) = -\sum_{d=1}^D \bigg[
+      & \diag \( \X \tr \mathbf{A}_d^{(t)} \X \) \tr \exp (\tilde{\mathbf{C}}_d
+      \bm{\alpha}) \\
+      & + \frac{n + 1}{\mathbf{y}_d \tr \mathbf{A}_d^{(t)} \mathbf{y}_d} 
+      \mathbf{y}_d \tr \left\{ \I + \mathbf{X} \diag \[ \exp\( 
+      \tilde{\mathbf{C}}_d \bm{\alpha} \) \] \mathbf{X} \tr \right\}^{-1} 
+      \mathbf{y}_d\bigg],
+    \end{align}
+  \end{subequations}
+  where $\mathbf{A}_d^{(t)} = \left\{ \I + \mathbf{X} \diag \[ \exp\( 
+  \tilde{\mathbf{C}}_d \bm{\alpha}^{(t)} \) \] \mathbf{X} \tr \right\}^{-1}$.
+  Note that (\ref{eq:minor}) is concave. However, it is slow to optimise. 
+  In \cite{zeng_incorporating_2020}, another minorization is proposed for
+  the second term:
+  $$
+  \mathbf{y}_d \tr \left\{ \I + \mathbf{X} \diag \[ \exp\( 
+  \tilde{\mathbf{C}}_d \bm{\alpha} \) \] \mathbf{X} \tr \right\}^{-1} 
+  \mathbf{y}_d = \underset{\delta}{\max} 
+  - ||\mathbf{y}_d - \mathbf{X} \delta ||_2^2 - 
+  \delta \tr \diag \[ \exp\(
+  -\tilde{\mathbf{C}}_d \bm{\alpha} \) \] \delta.
+  $$
+  The extra parameter $\delta$ is estimated by any standard ridge regression
+  package, such as \texttt{glmnet} in \texttt{R}.
+
   \section{Gibbs sampler}\label{sec:gibbssampler}

### 2020-03-30 code/simulations_gdsc.R

@@ -4,41 +4,88 @@
 ncores <- 100
 
 ### libraries
-packages <- c("foreach", "doParallel", "cambridge", "statmod", "glmnet")
+packages <- c("foreach", "doParallel", "cambridge", "msm", "statmod", "glmnet",
+              "xtune")
 sapply(packages, library, character.only=TRUE)
 
-### load data
-load(file="data/data_gdsc_dat1.Rdata")
+# ### load data
+# load(file="data/data_gdsc_dat1.Rdata")
 
 # number of reps
 nreps <- 100
 
 ################################# simulation 1 #################################
-### data preparation
-# only retain the cell lines that are available in both response and expression
-expr.prep <- expr[rownames(expr) %in% rownames(resp), ]
-resp.prep <- resp[rownames(resp) %in% rownames(expr), ]
-
-# select features
-D <- ncol(resp.prep)
-psel <- 100
-o <- order(-apply(expr.prep, 2, sd))
-idsel <- rep(list(o[c(1:psel)]), D)
-expr.sel <- lapply(idsel, function(s) {expr.prep[, s]})
-x <- lapply(expr.sel, function(s) {scale(s)})
-
-### data preparation
-p <- sapply(x, ncol)
-n <- nrow(expr.prep)
-ntrain <- floor(n/2)
+### simulation functions
+.cmtnorm <- function(mu, sigma) {
+  if(length(mu)==1 & length(sigma)!=1) {
+    mu <- rep(mu, length(sigma))
+  }
+  if(length(sigma)==1 & length(mu)!=1) {
+    sigma <- rep(sigma, length(mu))
+  }
+  ratio <- dnorm(mu/sigma)/pnorm(mu/sigma)
+  m1 <- mu + sigma*ratio
+  m2 <- ifelse(sigma==0, 0,
+               sigma^2*(1 - (mu/sigma)*ratio - ratio^2))
+  m3 <- ifelse(sigma==0, 0,
+               sigma^3*ratio*((ratio + mu/sigma)^2 + 
+                                ratio*(ratio + mu/sigma) - 1))
+  return(list(m1=m1, m2=m2, m3=m3))
+}
+.simlambda <- function(m1, M2, m3, alpha, ESNR) {
+  val1 <- sum(alpha*m1)
[…]

### 2020-03-30 rpackage/R/main_functions.R

@@ -1,6 +1,7 @@
-# x=xtrain; y=ytrain; C=C; Z=Z; unpenalized=NULL;
-# standardize=FALSE; intercept=FALSE; fixed.eb="none"; var.scale=1;
-# full.post=TRUE; init=NULL; control=control
+# x=rep(list(xtrain), D); y=ytrain
+# unpenalized=NULL; standardize=FALSE; 
+# intercept=FALSE; fixed.eb="none"; var.scale=1;
+# full.post=FALSE; init=NULL;
 # library(Rcpp)
 # sourceCpp("rpackage/src/aux_functions.cpp")
 # source("rpackage/R/vb_functions.R")
@@ -15,7 +16,7 @@ semnig <- function(x, y, C, Z, unpenalized=NULL, standardize=FALSE,
                    full.post=FALSE, init=NULL,
                    control=list(conv.post=TRUE, trace=TRUE,
                                 epsilon.eb=1e-3, epsilon.vb=1e-3,
-                                maxit.eb=10, maxit.vb=2, maxit.post=100,
+                                maxit.eb=100, maxit.vb=2, maxit.post=100,
                                 maxit.block=0)) {
   
   # save the arguments
@@ -271,7 +272,7 @@ semnig <- function(x, y, C, Z, unpenalized=NULL, standardize=FALSE,
                                 control$epsilon.vb)
       
       # save times
-      if(check.eb & iter.vb[iter.eb]==control$maxit.vb) {
+      if(check.eb & iter.vb[iter.eb]==1) {
         eb.time <- proc.time()[3] - srt
         srt <- proc.time()[3]
         if(control$trace) {
@@ -435,17 +436,21 @@ cv.semnig <- function(x, y, nfolds=10, foldid=NULL, seed=NULL, phi=phi,
   return(fit)
 }
 
-# x=xtrain; y=ytrain; mult.lambda=TRUE; foldid=foldid;
-# hyper=list(lambda=NULL, zeta=0, nu=0);
-# control=list(epsilon=sqrt(.Machine$double.eps),
-#              maxit=500, trace=TRUE, glmnet.fit2=FALSE)
+
+# x=xtrain; y=ytrain; mult.lambda=TRUE; C=lapply(C, function(s) {s[, -1]})
+# hyper=list(lambda=1/sqrt(n*sapply(
+#   fit.ridge1, "[[", "lambda.min")), zeta=0, nu=0);
+# hyper=list(lambda=rep(1, D), zeta=0, nu=0);
[…]

### 2020-03-30 rpackage/src/aux_functions.cpp

@@ -13,7 +13,7 @@ double f_optim_mat(arma::vec alpha, arma::vec lambda, double nu, double zeta,
                    arma::mat Cmat, arma::mat Z, arma::vec n, int p, int D,
                    List idsel, int G, int H, List y, arma::mat x,
                    arma::vec yty) {
-
+  
   arma::vec alphaf = alpha.head(G);
   arma::vec alphad = alpha.tail(H);
 
@@ -22,7 +22,7 @@ double f_optim_mat(arma::vec alpha, arma::vec lambda, double nu, double zeta,
   double out=0.0;
   for(int d=0; d<D; d++) {
     arma::rowvec mat1(n(d));
-    arma::mat mat2(n(d), n(d)); 
+    arma::mat mat2(n(d), n(d));
     arma::uvec cidsel = idsel[d];
     arma::mat cx = x.rows(cidsel - 1);
     arma::vec cy = y[d];
@@ -31,9 +31,9 @@ double f_optim_mat(arma::vec alpha, arma::vec lambda, double nu, double zeta,
       gamma.subvec(d*p, (d + 1)*p - 1).t()));
     mat1 = cy.t()*mat2;
     mat2.diag() += 1;
-    out += -0.5*real(log_det(mat2)) -
-      (n(d)/2 + nu)*log(zeta + arma::conv_to<double>::from(
-          0.5*yty(d) - 0.5*mat1*y[d] + 0.5*mat1*mat2.i()*mat1.t()));
+    out += -0.5*real(log_det(mat2))  -
+      (n(d)/2 + nu)*log(zeta + 0.5*yty(d) - 0.5*as_scalar(mat1*cy) +
+      0.5*as_scalar(mat1*mat2.i()*mat1.t()));
   }
   return out;
 }

### 2020-04-02 docs/manuscript1.Rnw
changed analysis and simulations

@@
 	efficient and scales well with the number of features and samples.
+
 	Estimation of the NIG model that includes both external drugs and feature
@@
 	lower for the NIG model.
+
+	An alternative strategy to include external covariates is the varying 
+	coefficient (VC) model \cite[]{hastie_varying-coefficient_1993}. The VC model
+	treats the mean of the regression coefficients as a deterministic function of 
+	external covariates, as opposed to our probabilistic model for the variance
+	of the regression coefficients. \cite{ni_bayesian_2019} introduces a Bayesian 
+	VC model where the relation between the regression 
+	coefficients and external covariates is no longer deterministic, but still
+	based on the mean of the coefficients. Besides our computationally more
+	feasible VB-EM estimation, we advocate for a more indirect
+	model for the relation between external covariates and regression 
+	coefficients, i.e., through their random variance components. This indirect 
+	model assumes less structure about the relation between external covariates
+	and regression coefficients than the direct VC mean model approach.
+  In particular, a VC mean model describes both magnitude and direction of the 
+  external covariates effects, while our variance model only describes the
+  magnitude and is invariant to the direction of the effects.
+
+  The choice to model the $\bm{\alpha}$`s, as hyperparameters instead of
+  full Bayesian parameters is three-fold: (i) it allows fast and stable 
+  empirical Bayes estimation as compared to badly mixing (see Comment 5)
+  full Bayesian MCMC sampling. (ii) \cite{wiel_learning_2019} shows that in a 
+  somewhat similar model, empirical Bayes is competitive to full Bayesian 
+  inference in terms of frequentist coverage of credible intervals, i.e.,
+  uncertainty propagation of the hyperparameters is not always necessary.
+  Additionally, (iii) in SM Section \ref{sm-sec:gibbssampler} a full Bayes
+  Gibbs sampler is described, where the full conditional of the $\bm{\alpha}$ is
+  Gaussian. For vague $\bm{\alpha}$ and $\lambda$ priors, the mean of this 
+  Gaussian coincides 
+  with the empirical Bayes updates as described in the MD, a strong 
+  indication that the posterior point estimates of our EB approach and the full
+  Bayes model are similar.
 

### 2020-04-02 code/analysis_gdsc.R

@@ -13,13 +13,28 @@ sapply(packages, library, character.only=TRUE)
 nfolds <- 10
 
 # number of cores to use
-ncores <- min(nfolds, 100)
-# options(mc.cores=max(floor(100/ncores), 1))
+# ncores <- min(nfolds, 100)
+ncores <- 1
 
 ### load data
 load(file="data/data_gdsc_dat1.Rdata")
 
+### functions
+coef.dss <- function(object, psel) {
+  best <- matrix(nrow=object$glmnet.fit$dim[1] + 1, ncol=length(psel))
+  for(p in 1:length(psel)) {
+    ind <- which.min(abs(colSums(coef(object, s=object$lambda)[-1, ]!=0) - 
+                           psel[p]))
+    best[, p] <- coef(object, s=object$lambda[ind])[, 1]
+  }
+  return(best)
+}
+
+################################################################################
 ################################## analysis 1 ##################################
+################################################################################
+# expression data with pvalues from mutationas as external data
+
 ### external data analysis
 # calculate pvalues for mutation data
 mut.pvalues <- lapply(resp, function(s) {
@@ -32,24 +47,6 @@ mut.pvalues <- lapply(resp, function(s) {
       NA
     }})
   pvalues[!is.na(pvalues)]})
-# 
-# # correlate mutation pvalues with regression coefficients from expression
-# expr.sel <- expr$expr[, colnames(expr$expr) %in%
-#                         unlist(strsplit(colnames(mut), ","))]
-# fit1.ridge <- lapply(1:length(resp), function(d) {
-#   resp.prep <- scale(resp[[d]][names(resp[[d]]) %in% rownames(expr.sel)])
-#   expr.prep <- scale(expr.sel[rownames(expr.sel) %in% names(resp[[d]]), ])
-#   cv.glmnet(expr.prep, resp.prep, alpha=0, intercept=FALSE)})
-# best1.ridge <- sapply(fit1.ridge, function(s) {coef(s, s="lambda.min")[-1, ]})
[…]

### 2020-04-02 code/figures.R

@@ -46,10 +46,30 @@ par(opar)
 
 # ---- simulations_gdsc_est1 ----
 suppressWarnings(suppressMessages(library(sp)))
-res <- read.table("results/simulations_gdsc_res1.txt", row.names=NULL)
+res <- read.table("results/simulations_gdsc_res1.2.txt", row.names=NULL)
 temp <- res[, 1]
 res <- as.matrix(res[, -1])
 rownames(res) <- temp
+
+pmse <- res[substr(rownames(res), 1, 5)=="pmse.", ]
+mpmse <- aggregate(pmse, by=list(substr(rownames(pmse), 6, 1000000L)), 
+                   FUN="mean")[, -1]
+boxplot(mpmse)
+
+
+est <- sapply(c(paste0("alphaf", 0:3), paste0("alphad", 0:3), 
+                  "lambdaf", "lambdad"), function(s) {
+                    s <- res[rownames(res)==s, ]; rownames(s) <- NULL
+                    s}, simplify=FALSE)
+
+alpha <- sapply(est, function(s) {s[, 2]})
+boxplot(alpha[, -c(5, 9, 10)])
+abline(h=1)
+test <- alpha$alphad0
+boxplot(cbind(test[, c(1, 2)], exp(-test[, 5])), outline=FALSE)
+
+apply(alphaf0, 2, median, na.rm=TRUE)
+alphaf0
 
 alphaf <- c(1, 1, 3, 7)
 phi <- 1/as.numeric(alphaf %*% t(cbind(1, rbind(0, diag(3)))))
@@ -212,7 +232,47 @@ points(c(1:4), chi, pch=2, col=col[c(1:4)], cex=1.5,
        cex.lab=1.5, cex.axis=1.5)
 par(opar)
 
+# ---- analysis_gdsc_res1 ----
+suppressWarnings(suppressMessages(library(sp)))
+res <- read.table("results/analysis_gdsc_res1.txt", row.names=NULL)
+temp <- res[, 1]
+res <- as.matrix(res[, -1])
+rownames(res) <- temp
+pmse <- res[substr(rownames(res), 1, 4)=="pmse", ]
 
+ylim1 <- range(1 - c(res[, "semnig1"], res[, "ridge1"]))
+ylim2 <- range(1 - c(res[, "ebridge1"], res[, "ridge1"]))
+
+col <- bpy.colors(3, cutoff.tail=0.2)
+
[…]

### 2020-04-02 code/simulations_gdsc.R
rewrote simulations

@@ -8,115 +8,36 @@ packages <- c("foreach", "doParallel", "cambridge", "msm", "statmod", "glmnet",
               "xtune")
 sapply(packages, library, character.only=TRUE)
 
-# ### load data
-# load(file="data/data_gdsc_dat1.Rdata")
+### load and preprocess data
+load(file="data/data_gdsc_dat1.Rdata")
 
 # number of reps
 nreps <- 100
 
+################################################################################
 ################################# simulation 1 #################################
-### simulation functions
-.cmtnorm <- function(mu, sigma) {
-  if(length(mu)==1 & length(sigma)!=1) {
-    mu <- rep(mu, length(sigma))
-  }
-  if(length(sigma)==1 & length(mu)!=1) {
-    sigma <- rep(sigma, length(mu))
-  }
-  ratio <- dnorm(mu/sigma)/pnorm(mu/sigma)
-  m1 <- mu + sigma*ratio
-  m2 <- ifelse(sigma==0, 0,
-               sigma^2*(1 - (mu/sigma)*ratio - ratio^2))
-  m3 <- ifelse(sigma==0, 0,
-               sigma^3*ratio*((ratio + mu/sigma)^2 + 
-                                ratio*(ratio + mu/sigma) - 1))
-  return(list(m1=m1, m2=m2, m3=m3))
-}
-.simlambda <- function(m1, M2, m3, alpha, ESNR) {
-  val1 <- sum(alpha*m1)
-  val2 <- sum(t(M2*alpha)*alpha)
-  val3 <- sum(alpha^2*m3)
-  ESNR/(val2*val3 + 3*val1*val2^2 + val1^3*val2)
-}
-.ratesigma <- function(Ex, Vx, Egammasq, Etausq, ESNR) {
-  rate <- ESNR/(ESNR - Etausq*sum(Egammasq*(Vx + Ex^2)))
-  return(rate)
-}
-.sigmax <- function(shape, Etausq, Egammasq, ESNR) {
-  sqrt((1 - 1/shape)*ESNR/(Etausq*sum(Egammasq)))
-}
-
+################################################################################
 ### simulation settings
 D <- 100
-p <- 200
-n <- 200
-ntest <- 1000
-G <- 4
-H <- 4
+p <- 100
+n <- 100
+ntest <- nrow(expr$expr) - n
 alphaf <- c(1, 1, 3, 7)
-alphad <- c(1, 1, 3, 7)
[…]

### 2020-04-02 docs/manuscript1.Rnw
rewrote simulations
added simulation with noise in external covariate models

@@
 	efficient and scales well with the number of features and samples.
-
 	Estimation of the NIG model that includes both external drugs and feature
@@
 
-  The choice to model the $\bm{\alpha}$`s, as hyperparameters instead of
-  full Bayesian parameters is three-fold: (i) it allows fast and stable 
-  empirical Bayes estimation as compared to badly mixing (see Comment 5)
-  full Bayesian MCMC sampling. (ii) \cite{wiel_learning_2019} shows that in a 
-  somewhat similar model, empirical Bayes is competitive to full Bayesian 
-  inference in terms of frequentist coverage of credible intervals, i.e.,
-  uncertainty propagation of the hyperparameters is not always necessary.
-  Additionally, (iii) in SM Section \ref{sm-sec:gibbssampler} a full Bayes
-  Gibbs sampler is described, where the full conditional of the $\bm{\alpha}$ is
-  Gaussian. For vague $\bm{\alpha}$ and $\lambda$ priors, the mean of this 
-  Gaussian coincides 
-  with the empirical Bayes updates as described in the MD, a strong 
-  indication that the posterior point estimates of our EB approach and the full
-  Bayes model are similar.
+  Another possible criticism is the treatment of $\bm{\alpha}$ as fixed
+  hyperparameters instead of random. A Bayesian could argue that endowment of 
+  $\bm{\alpha}$ with a hyperprior results in propagation of 
+  uncertainty about $\bm{\alpha}$ and as a result improved regression parameter 
+  uncertainty quantification. \cite{wiel_learning_2019} show in a similar 
+  setting that EB estimation of hyperparameters does not necessarily lead to
+  worse uncertainty quantification as measured by frequentist coverage of
+  Bayesian credible intervals, as compared to a full Bayes treatment of the
+  hyperparameters.
 

### 2020-04-04 docs/manuscript1.Rnw
running second analysis on Tukey

@@
   \end{figure}
-
-  \subsection{Alternative prior}\label{sec:altprior}
-  Consider the following simpler hyperprior alternative to (\ref{eq:prior}):
-  \begin{subequations}\label{eq:altprior}
-		\begin{align}
-		  \beta_{jd} | \gamma^2_{jd}, \tau_d^2, \sigma^2_d & \sim \mathcal{N}_{p} 
-		  (0, \gamma_{jd}^2 \tau_d^2 \sigma_d^2), \\
-		  \log \gamma_{jd}^{2} & = \mathbf{c}_{jd} \tr \bm{\alpha}_{\text{feat}}, \\
-		  \log \tau_d^{2} & = \mathbf{z}_{d} \tr \bm{\alpha}_{\text{drug}}\\
-		  \sigma_d^{2} & \sim \Gamma^{-1} (\nu, \zeta).
-		\end{align}
-	\end{subequations}
-	This model is a Bayesian, multivariate extension of 
-	\cite{zeng_incorporating_2020}. The model does 
-	not propagate uncertainty in the variance components $\gamma_{jd}^{2}$
-	and $\tau_d^{2}$, but directly models the variance components as an
-	exponentiated linear function of the external covariates. The linear functions
-	are modelled on the log-scale to ensure non-negative variance components.
-	Additionally, it allows for more flexibility in the error variance
-	prior model. Note that in the limit $\zeta \to 0$ and $\nu=1/2$, the error
-	variance follows the Jeffrey`s prior from (\ref{eq:prior}). Throughout the
-	following, whenever $\zeta$ and $\nu$ are not specified, we use $\zeta \to 0$ 
-	and $\nu=1/2$.
-
-	Estimation of $\bm{\alpha}_{\text{feat}}$ and $\bm{\alpha}_{\text{feat}}$, and
-	ultimately $\bm{\beta}_d$ is explained in Section \ref{sm-sec:altprior} of the
-	SM. The results in Sections \ref{sec:simulations} and \ref{sec:gdsc}
-	include model (\ref{eq:altprior}) as an alternative to (\ref{eq:prior}).
 
@@
 	wherein the responses are simulated from a synthetic model, but the
-	features are taken from the real GDSC data introducted in Section 
-	\ref{sec:gdsc}. 
-	The real GDSC features contain strong collinearities. Such strong
-	collinearities in the design matrix impede correct parameter estimation. 
+	features are taken from the real GDSC expression data introducted in Section 
+	\ref{sec:gdsc}. The real GDSC features contain strong collinearities 
+	(correlations ranging from -0.89 to 0.99). Such
+	strong collinearities in the design matrix impede correct parameter estimation
+	with small sample sizes. 
 	We therefore replace the ambition of correctly estimating the $\beta_{jd}$
@@
 
-	A pre-processing step selects 100 features with largest variance, while the 
-	168 drug sensitivities for 482 cell lines are 
+	A pre-processing step selects 100 features with largest variance, while
+	100 drug sensitivities for 1014 cell lines are 
 	simulated from model (\ref{eq:likelihood}) and 
@@
 	such that the prior $\sigma_d^2$ mean and variance are both one. 
-	We consider the following three scenarios for the simulation of the 
+	We consider the following four scenarios for the simulation of the 
 	drug and feature variance components:
@@
 	    create four external dummy feature covariates that code for four 
-	    randomly assigned approximately equally sized groups of features. We set
+	    approximately equally sized groups of features. We set
 	    $\bm{\alpha}_{\text{feat}}$ such that the $\gamma_{jd}^2$ of the four
@@
 	    similar to the procedure for the $\gamma_{jd}^2$ in Scenario 1: we 
-	    randomly create four groups of drugs with corresponding external drug 
+	    create four groups of drugs with corresponding external drug 
 	    dummy variables and set $\bm{\alpha}_{\text{drug}}=\begin{bmatrix} 
@@
 	    the $\gamma_{jd}^2$ and $\tau_d^2$ according to (\ref{eq:prior}).}
+	  \item{Scenario 4 is equal to Scenario three, except that the rows of the
+	    external covariates matrix are permuted. Permutation
+	    adds noise and is supposed to mimic a low external covariates signal. We
+	    mimic different noise levels by permutation of fractions 
+	    $q \in \{0.1, 0.2, 0.33, 0.5, 0.67, 0.8, 1\}$ of the 
+	    number of rows.}
 	\end{itemize}
-	We estimate four models: (i) the NIG model that only includes an intercept in
+	We estimate two models: (i) the NIG model that only includes an intercept in
 	the external covariates, called NIG$_{\text{f}}^-$, NIG$_{\text{d}}^-$, or
 	NIG$_{\text{f}+\text{d}}^-$, depending on which variance components are 
-	estimated (feature, drug, or both in Scenarios 1, 2, and 3, respectively),
+	estimated (feature, drug, or both in Scenarios 1, 2, and 3/4, respectively),
 	(ii) the NIG model estimated as in Section \ref{sec:estimation} that includes
@@
 	NIG$_{\text{f}+\text{d}}$, again depending on which variance components are 
-	estimated, and frequentist (iii) lasso and (iv) ridge models.
+	estimated, frequentist 
 	Exclusion of the external covariates as in the NIG$_{\text{f}}^-$, 
@@
 	NIG$_{\text{f}}$, NIG$_{\text{d}}$, and NIG$_{\text{f}+\text{d}}$ models.
-	The lasso and ridge models were fit using the \texttt{R} package 
-	\texttt{glmnet} \cite[]{friedman_regularization_2010} with cross validated
-	penalty parameters.
-	All models are fit on 241 randomly selected cell lines, while performance
-	measures are estimated from the remaining 241 cell lines. We repeat every
-	simulation Scenario 50 times.
+	All models are fit on 100 randomly selected cell lines. We repeat every
+	simulation Scenario 100 times.
+
+	SM Section \ref{sm-sec:simulations} contains additional simulation results for
+	the NIG model, plus results for 
+	four alternative methods: frequentist (i) lasso and (ii) ridge models, 
+	and two EB versions of the ridge model: (iii) \texttt{xtune} 
+	\cite[]{ni_bayesian_2019} and (iv) its
+	multivariate Bayesian extension as introduced in MD Section 
+	\ref{sm-sec:altprior}.
 
@@
 rownames(res) <- temp
+
 true.phi <- mean(1/(c(1, 1, 3, 7) %*% t(unname(model.matrix(~ factor(1:4))))))
@@
 est.lambdaf <- apply(res[rownames(res)=="lambdaf", c(1, 2)], 2, median)
-
-tab <- cbind(apply(res[rownames(res)=="emsel", ], 2, median),
-             apply(res[rownames(res)=="emseh", ], 2, median),
-             apply(res[rownames(res)=="emse", ], 2, median),
-             apply(res[rownames(res)=="pmse", ], 2, median))
-tab <- round(tab, 3)
-tab[which(tab[, 1]==min(tab[, 1], na.rm=TRUE)), 1] <- 
-  paste0("\\textbf{", tab[which(tab[, 1]==min(tab[, 1], na.rm=TRUE)), 1], "}")
-tab[which(tab[, 2]==min(tab[, 2], na.rm=TRUE)), 2] <- 
-  paste0("\\textbf{", tab[which(tab[, 2]==min(tab[, 2], na.rm=TRUE)), 2], "}")
-tab[which(tab[, 3]==min(tab[, 3], na.rm=TRUE)), 3] <- 
-  paste0("\\textbf{", tab[which(tab[, 3]==min(tab[, 3], na.rm=TRUE)), 3], "}")
-tab[which(tab[, 4]==min(tab[, 4], na.rm=TRUE)), 4] <- 
-  paste0("\\textbf{", tab[which(tab[, 4]==min(tab[, 4], na.rm=TRUE)), 4], "}")
-colnames(tab) <- c("$\\text{EMSE}_{\\text{bottom}}$",
-                   "$\\text{EMSE}_{\\text{top}}$",
-                   "$\\text{EMSE}_{\\text{total}}$",
-                   "$\\text{PMSE}$")     
-rownames(tab)[c(1, 2)] <- c("NIG$_{\\text{f}}^-$",
-                            "NIG$_{\\text{f}}$")
 @	
@@
 	$\bm{\alpha}_{\text{feat}}$ is slightly but 
-	consistently underestimated. On the $\phi_{jd}$ scale this underestimation is
-	almost undetectable as depicted in Figure
-	\ref{fig:simulations_gdsc_est1}b. $\lambda_{\text{feat}}$ is slightly 
-	overestimated with median estimate \Sexpr{round(est.lambdaf[2], 2)} (the
-	true value is 1). The overestimated $\lambda_{\text{feat}}$ translates 
-	into underestimated $\gamma_{jd}^2$ variances. Underestimation of variances is 
-	a common phenomenon in variational Bayes approximations. In addition, 
-	$\lambda_{\text{feat}}$ is a higher level scale
-	parameter, which are, in general, hard to estimate. 
-<<simulations_gdsc_est1, fig.cap="Simulation results for Scenario 1 ($\\tau^2_d$ fixed): (a) estimated and true values for $\\alpha_{\\text{feat}}$ and, (b) estimated and true values for prior means $\\phi_{jd}$.", out.width="100%", fig.asp=1/2>>=
-@
-
+	consistently underestimated. This results in overestimation on the $\phi_{jd}$
+	scale as depicted in Figure \ref{fig:simulations_gdsc_est1}b. 
 	Model NIG$_{\text{f}}^-$ (that excludes the external covariates) gives a 
 	median $\phi$ estimate of \Sexpr{round(est.phi, 2)},  
-	equal to the true mean of the $\phi_{jd}$, \Sexpr{round(true.phi, 2)}. 
-	$\lambda_{\text{feat}}$ is slightly underestimated with median estimate
-	\Sexpr{round(est.lambdaf[1], 2)} (the true value is one) and thus 
-	overestimates the 
-	prior $\gamma_{jd}^2$ variance. This is likely to compensate for the common
-	prior mean $\phi$, that is not able to capture part of the variation in the 
-	true feature specific $\phi_{jd}$.
-
-	Table \ref{tab:simulations_gdsc_tab1} displays median estimation MSE 
-	(EMSE) and prediction MSE (PMSE) for simulation Scenario 1, calculated on an 
-	independent test data set. EMSE is further split into the contribution of the
-	bottom 90\% of the $\beta_{jd}$ in terms of size (in absolute value) and the 
-	top 10\% of the $\beta_{jd}$ in size. The table includes as a reference the 
-	null model, which is an empty model that only fits the means of the $\y_{d}$.
-	Estimation MSE (EMSE) and prediction MSE (PMSE) give measures of predictive 
-	performance. Focussing on MSE, we see that the NIG$_{\text{f}}$ model, 
-	that includes the external covariates, performs best, both in terms of 
-	estimation and prediction. The lower MSEs confirm that the model that
-	includes the external data (NIG$_{\text{f}}$) learns the underlying structure 
-	in the data better than model NIG$_{\text{f}}^-$ that does not include the
-	external data. Furthermore, we see that estimation of the larger $\beta_{jd}$ 
-	is where ridge loses the most compared to the NIG and lasso models. 
-	The heavier tails of the NIG and lasso priors allow for larger $\beta_{jd}$
-	estimates and thus lower EMSE for these $\beta_{jd}$.
-<<simulations_gdsc_tab1>>=
-library(knitr, quietly=TRUE, verbose=FALSE)
-suppressWarnings(library(kableExtra))
-options(knitr.kable.NA="", digits=3)
-kable(tab, col.names=colnames(tab), align="r",
-      caption="Median performance measure for simulation Scenario 1, estimated on the test data (lowest in bold).", 
-      format="latex", booktabs=TRUE, escape=FALSE) %>%
-  kable_styling(latex_options=c("HOLD_position"))
+	slightly higher than the true mean of the $\phi_{jd}$, 
+	\Sexpr{round(true.phi, 2)}. 
+<<simulations_gdsc_est1, fig.cap="Simulation results for Scenario 1 ($\\tau^2_d$ fixed): estimated and true values for (a) $\\alpha_{\\text{feat}}$ and (b) prior means $\\phi_{jd}$.", out.width="100%", fig.asp=1/2>>=
 @
@@
 est.lambdad <- apply(res[rownames(res)=="lambdad", c(1, 2)], 2, median)
-tab <- cbind(apply(res[rownames(res)=="emsel", ], 2, median),
-             apply(res[rownames(res)=="emseh", ], 2, median),
-             apply(res[rownames(res)=="emse", ], 2, median),
-             apply(res[rownames(res)=="pmse", ], 2, median))
-tab <- round(tab, 3)
-tab[which(tab[, 1]==min(tab[, 1], na.rm=TRUE)), 1] <- 
-  paste0("\\textbf{", tab[which(tab[, 1]==min(tab[, 1], na.rm=TRUE)), 1], "}")
-tab[which(tab[, 2]==min(tab[, 2], na.rm=TRUE)), 2] <- 
-  paste0("\\textbf{", tab[which(tab[, 2]==min(tab[, 2], na.rm=TRUE)), 2], "}")
-tab[which(tab[, 3]==min(tab[, 3], na.rm=TRUE)), 3] <- 
-  paste0("\\textbf{", tab[which(tab[, 3]==min(tab[, 3], na.rm=TRUE)), 3], "}")
-tab[which(tab[, 4]==min(tab[, 4], na.rm=TRUE)), 4] <- 
-  paste0("\\textbf{", tab[which(tab[, 4]==min(tab[, 4], na.rm=TRUE)), 4], "}")
-colnames(tab) <- c("$\\text{EMSE}_{\\text{bottom}}$",
-                   "$\\text{EMSE}_{\\text{top}}$",
-                   "$\\text{EMSE}_{\\text{total}}$",
-                   "$\\text{PMSE}$")      
-rownames(tab)[c(1, 2)] <- c("NIG$_{\\text{d}}^-$",
-                            "NIG$_{\\text{d}}$")
-@	
-  Figure \ref{fig:simulations_gdsc_est2} shows the estimated
+@	
+  Figure \ref{fig:simulations_gdsc_est2}a shows the estimated
 	$\bm{\alpha}_{\text{drug}}$ together with its true value for NIG$_{\text{d}}$ 
-	in Scenario 2 of the simulation study (fixed $\gamma_{jd}^2$). They are
-	estimated without bias.
-	Median $\lambda_{\text{drug}}=\Sexpr{round(est.lambdad[2], 2)}$ is slightly 
-	overestimated compared to true value one. 
-	The median $\chi$ estimate in the NIG$_{\text{d}}^-$ model is 
-	\Sexpr{round(est.chi, 2)},  
-	which is almost equal to the true mean \Sexpr{round(true.chi, 2)}. 
-	Scale $\lambda_{\text{drug}}$ is underestimated with median estimate
-	\Sexpr{round(est.lambdad[1], 2)} compared to the true value one.
-	Median performance measures for simulation Scenario 2 are displayed in Table
-	\ref{tab:simulations_gdsc_tab2}. Again, we see that NIG$_{\text{d}}$ performs
-	best in terms of MSE. However, NIG$_{\text{d}}^-$ performance is the same 
-	here, indicating that the external covariates do not add much here.
-	Similarly as in Scenario 1, most of the estimation performance gain over ridge
-	is through the top 10\% of the $\beta_{jd}$.
-<<simulations_gdsc_est2, fig.cap="Simulation results for Scenario 2 ($\\gamma^2_{jd}$ fixed): (a) estimated and true values for $\\alpha_{\\text{feat}}$ and, (b) estimated and true values for prior means $\\phi_{jd}$.", out.width="100%", fig.asp=1/2>>=
-@
-
-<<simulations_gdsc_tab2>>=
-library(knitr, quietly=TRUE, verbose=FALSE)
-suppressWarnings(library(kableExtra))
-options(knitr.kable.NA="", digits=3)
-kable(tab, col.names=colnames(tab), align="r",
-      caption="Median performance measure for simulation Scenario 2, estimated on the test data (lowest in bold).",
-      format="latex", booktabs=TRUE, escape=FALSE) %>%
-  kable_styling(latex_options=c("HOLD_position"))
+	in Scenario 2 of the simulation study (fixed $\gamma_{jd}^2$). The intercept
+	term ($\alpha_{\text{drug},0}$) is overestimated, while the deviations from 
+	the intercept term are approximately correct. The overestimated intercept term
+	results in underestimated $\chi_d$ (Figure \ref{fig:simulations_gdsc_est2}b).
+	The median $\chi$ estimate in the 
+	NIG$_{\text{d}}^-$ model is \Sexpr{round(est.chi, 2)},  
+	which underestimates the true mean \Sexpr{round(true.chi, 2)}. 
+<<simulations_gdsc_est2, fig.cap="Simulation results for Scenario 2 ($\\gamma^2_{jd}$ fixed): estimated and true values for (a) $\\alpha_{\\text{feat}}$ and (b) prior means $\\phi_{jd}$.", out.width="100%", fig.asp=1/2>>=
 @
@@
                               t(unname(model.matrix(~ factor(1:4))))))
-
-tab <- cbind(apply(res[rownames(res)=="emsel", ], 2, median),
-             apply(res[rownames(res)=="emseh", ], 2, median),
-             apply(res[rownames(res)=="emse", ], 2, median),
-             apply(res[rownames(res)=="pmse", ], 2, median))
-tab <- round(tab, 3)
-tab[which(tab[, 1]==min(tab[, 1], na.rm=TRUE)), 1] <-
-  paste0("\\textbf{", tab[which(tab[, 1]==min(tab[, 1], na.rm=TRUE)), 1], "}")
-tab[which(tab[, 2]==min(tab[, 2], na.rm=TRUE)), 2] <-
-  paste0("\\textbf{", tab[which(tab[, 2]==min(tab[, 2], na.rm=TRUE)), 2], "}")
-tab[which(tab[, 3]==min(tab[, 3], na.rm=TRUE)), 3] <-
-  paste0("\\textbf{", tab[which(tab[, 3]==min(tab[, 3], na.rm=TRUE)), 3], "}")
-tab[which(tab[, 4]==min(tab[, 4], na.rm=TRUE)), 4] <-
-  paste0("\\textbf{", tab[which(tab[, 4]==min(tab[, 4], na.rm=TRUE)), 4], "}")
-colnames(tab) <- c("$\\text{EMSE}_{\\text{bottom}}$",
-                   "$\\text{EMSE}_{\\text{top}}$",
-                   "$\\text{EMSE}_{\\text{total}}$",
-                   "$\\text{PMSE}$")
-rownames(tab)[c(1, 2)] <- c("NIG$_{\\text{f}+\\text{d}}^-$",
-                            "NIG$_{\\text{f}+\\text{d}}$")
 @
@@
 	\Sexpr{round(chi.ratio, 2)[-1]}, respectively, compared to their true values
-	\Sexpr{round(true.ratio, 2)[-1]}. So in a relative sense, the estimates are
-	correct. Median $\lambda_{\text{feat}}$ and $\lambda_{\text{drug}}$ are
-	are over- and underestimated with \Sexpr{round(est.lambdaf[2], 2)}
-	and \Sexpr{round(est.lambdad[2], 2)}, respectively, compared to their true
-	value 1. The NIG$_{\text{f}+\text{d}}^-$ model is also consistently over- 
+	\Sexpr{round(true.ratio, 2)[-1]}. So in a relative sense, the 
+	$\bm{\alpha}_{\text{feat}}$ estimates are correct, while the 
+	$\bm{\alpha}_{\text{drug}}$ are somewhat higher than expected. 
+	The NIG$_{\text{f}+\text{d}}^-$ model is also consistently over- 
 	and underestimating $\phi$
-	and $\chi$ with estimates \Sexpr{round(est.phi, 2)} and 
+	and $\chi$ with median estimates \Sexpr{round(est.phi, 2)} and 
 	\Sexpr{round(est.chi, 2)}, respectively (compared to the true mean 
-	\Sexpr{round(true.chi, 2)}). NIG$_{\text{f}+\text{d}}^-$ also over- and
-	underestimates 
-	$\lambda_{\text{feat}}$ and $\lambda_{\text{drug}}$ with median estimates
-	\Sexpr{round(est.lambdaf[1], 2)}
-	and \Sexpr{round(est.lambdad[1], 2)}, respectively. Table 
-	\ref{tab:simulations_gdsc_tab3}, with median performance measures, shows
-	that NIG$_{\text{f}+\text{d}}$ outperforms the other models in terms of
-	prediction and estimation MSE, with again most of the edge over ridge in the
-	top 10\% of the $\beta_{jd}$.
-<<simulations_gdsc_est3, fig.cap="Simulation results for Scenario 3: (a) estimated and true values for $\\alpha_{\\text{feat}}$ and, (b) estimated and true values for prior means $\\phi_{jd}$.", out.width="100%", fig.asp=1>>=
-@
-
-<<simulations_gdsc_tab3>>=
-library(knitr, quietly=TRUE, verbose=FALSE)
-suppressWarnings(library(kableExtra))
-options(knitr.kable.NA="", digits=3)
-kable(tab, col.names=colnames(tab), align="r",
-      caption="Median performance measure for simulation Scenario 3, estimated on the test data (lowest in bold).",
-      format="latex", booktabs=TRUE, escape=FALSE) %>%
-  kable_styling(latex_options=c("HOLD_position"))
-@
+	\Sexpr{round(true.chi, 2)}). 
+<<simulations_gdsc_est3, fig.cap="Simulation results for Scenario 3: estimated and true values for (a) $\\alpha_{\\text{feat}}$, (b) prior means $\\phi_{jd}$, (c) $\\alpha_{\\text{drug}}$, and (d) prior means $\\chi_{d}$.", out.width="100%", fig.asp=1>>=
+@
+
+<<simulations_gdsc_est4, fig.cap="Simulation results for Scenario 3: estimated prior means (a) $\\phi_{jd}$, and (b) $\\chi_{d}$ for different levels of noise in the external covariates.", out.width="100%", fig.asp=1/2>>=
+@
+
+  \subsection{Discussion}
+  Underestimation of variance components is 
+	a common phenomenon in variational Bayes approximations, so the 
+	underestimation of the mean of the drug variance components $\chi_d$ in
+	Scenarios 2 and 3 is somewhat expected. In Scenario 3, this is compensated for
+	by the overestimation of the $\phi_{jd}$. We note that increasing the number
+	of cell lines $n$ and increasing the number of drugs $D$ improves estimation
+	of the $\chi_d$. However, this is not the most interesting situation since it
+	does not correspond to
+
+	In addition, 
+	$\lambda_{\text{feat}}$ is a higher level scale
+	parameter, which are, in general, hard to estimate.
 
@@
 
+  \subsection{Discussion}
   Part of the differences in performance between

### 2020-04-04 code/data_gdsc.R

@@ -1,8 +1,8 @@
 #!/usr/bin/env Rscript
 
 ### libraries
-library(gdata)
 library(biomaRt)
+library(gdata)
 
 ### biomart
 ensembl <- useMart(biomart="ensembl", dataset="hsapiens_gene_ensembl")

### 2020-04-06 docs/supplement1.Rnw

@@
 
-  \section{Estimation of alternative prior}\label{sec:altprior}
-  The hyperparameters $\bm{\alpha} = \begin{bmatrix} \bm{\alpha}_{\text{feat}}
-  \tr & \bm{\alpha}_{\text{drug}} \tr \end{bmatrix} \tr$ in model 
-  (\ref{md-eq:altprior}) are estimated by empirical Bayes maximisation of
-  the marginal log likelihood:
-  \begin{subequations}\label{eq:altmll}
-    \begin{align}
-      \log p (\bm{\alpha} | \mathbf{Y}) & = \sum_{d=1}^D
-      \log \[ \int_{\bm{\beta}_d} \int_{\sigma^2_d} \mathcal{L} 
-      (\mathbf{y}_d | \bm{\beta}_d, \sigma^2_d ) \pi(\bm{\beta}_d | \sigma^2_d) 
-      \pi(\sigma^2_d) d\bm{\beta}_d d\sigma_d^2 \] \\
-      & \propto - \sum_{d=1}^D \bigg[ \log \left|\I + \mathbf{X} 
-      \diag \[ \exp\( \tilde{\mathbf{C}}_d \bm{\alpha} \) \] \mathbf{X} \tr 
-      \right| \\
-      & \,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\, + (n + 1) \log \( \mathbf{y}_d \tr 
-      \left\{ \I + \mathbf{X} 
-      \diag \[ \exp\( \tilde{\mathbf{C}}_d \bm{\alpha} \) \] \mathbf{X} 
-      \tr\right\}^{-1} \mathbf{y}_d \) \bigg],
-    \end{align}
-  \end{subequations}
-  where $\tilde{\mathbf{C}}_d = \begin{bmatrix} \mathbf{C}_d & 
-  \mathbf{1}_{p \times 1} \cdot \mathbf{z}_d \tr \end{bmatrix}$, and
-  $\mathbf{C}_d = \begin{bmatrix} \mathbf{c}_{1d} & \cdots & 
-  \mathbf{c}_{pd} \end{bmatrix} \tr$.
-  The first term in the sum of (\ref{eq:altmll}) is concave in 
-  $\bm{\alpha}$, while the second term is convex in $\bm{\alpha}$. This renders
-  the marginal log likelihood a sum of non-concave (as well as non-convex) 
-  functions. General purpose optimisers, such as \texttt{optim} in \texttt{R} 
-  are generally not well suited to maximise non-concave functions. Maximisation
-  is therefore done through a minorise-maximisation (MM) algorithm, 
-  similar to \cite{zeng_incorporating_2020}.
-
-  The MM algorithm seeks to minorise the marginal log likelihood,
-  $\log p(\bm{\alpha} | \mathbf{Y})$, at current
-  parameter estimate $\bm{\alpha}^{(t)}$ with function 
-  $g(\bm{\alpha} | \bm{\alpha}^{(t)})$, such that
-  $\forall \bm{\alpha}: g(\bm{\alpha} | \bm{\alpha}^{(t)}) \leq 
-  \log p(\bm{\alpha} | \mathbf{Y})$ 
-  and $g(\bm{\alpha}^{(t)} | \bm{\alpha}^{(t)}) = 
-  \log p(\bm{\alpha} | \mathbf{Y})$. Then, the sequence
-  $$
-  \bm{\alpha}^{(t + 1)} = \underset{\bm{\alpha}}{\argmax} 
-  g(\bm{\alpha} | \bm{\alpha}^{(t+1)})
-  $$
-  converges to a local maximum of $f(\bm{\alpha})$. Generally, the minorisation
-  function is chosen such that it is concave and easy to maximise. In our case,
-  a suitable minorisation is:
-  \begin{subequations}\label{eq:minor}
-    \begin{align}
-      g(\bm{\alpha} | \bm{\alpha}^{(t+1)}) = -\sum_{d=1}^D \bigg[
-      & \diag \( \X \tr \mathbf{A}_d^{(t)} \X \) \tr \exp (\tilde{\mathbf{C}}_d
-      \bm{\alpha}) \\
-      & + \frac{n + 1}{\mathbf{y}_d \tr \mathbf{A}_d^{(t)} \mathbf{y}_d} 
-      \mathbf{y}_d \tr \left\{ \I + \mathbf{X} \diag \[ \exp\( 
-      \tilde{\mathbf{C}}_d \bm{\alpha} \) \] \mathbf{X} \tr \right\}^{-1} 
-      \mathbf{y}_d\bigg],
-    \end{align}
-  \end{subequations}
-  where $\mathbf{A}_d^{(t)} = \left\{ \I + \mathbf{X} \diag \[ \exp\( 
-  \tilde{\mathbf{C}}_d \bm{\alpha}^{(t)} \) \] \mathbf{X} \tr \right\}^{-1}$.
-  Note that (\ref{eq:minor}) is concave. However, it is slow to optimise. 
-  In \cite{zeng_incorporating_2020}, another minorization is proposed for
-  the second term:
-  $$
-  \mathbf{y}_d \tr \left\{ \I + \mathbf{X} \diag \[ \exp\( 
-  \tilde{\mathbf{C}}_d \bm{\alpha} \) \] \mathbf{X} \tr \right\}^{-1} 
-  \mathbf{y}_d = \underset{\delta}{\max} 
-  - ||\mathbf{y}_d - \mathbf{X} \delta ||_2^2 - 
-  \delta \tr \diag \[ \exp\(
-  -\tilde{\mathbf{C}}_d \bm{\alpha} \) \] \delta.
-  $$
-  The extra parameter $\delta$ is estimated by any standard ridge regression
-  package, such as \texttt{glmnet} in \texttt{R}.
-
   \section{Gibbs sampler}\label{sec:gibbssampler}
@@
 
+  \section{Alternative prior}\label{sec:altprior}
+  Consider the following simpler hyperprior alternative to 
+  equation (\ref{md-eq:prior}) in the MD:
+  \begin{subequations}\label{eq:altprior}
+		\begin{align}
+		  \beta_{jd} | \gamma^2_{jd}, \tau_d^2, \sigma^2_d & \sim \mathcal{N}_{p} 
+		  (0, \gamma_{jd}^2 \tau_d^2 \sigma_d^2), \\
+		  \log \gamma_{jd}^{2} & = \mathbf{c}_{jd} \tr \bm{\alpha}_{\text{feat}}, \\
+		  \log \tau_d^{2} & = \mathbf{z}_{d} \tr \bm{\alpha}_{\text{drug}}\\
+		  \sigma_d^{2} & \sim \Gamma^{-1} (\nu, \zeta).
+		\end{align}
+	\end{subequations}
+	This model is a Bayesian, multivariate extension of 
+	\cite{zeng_incorporating_2020}. The model does 
+	not propagate uncertainty in the variance components $\gamma_{jd}^{2}$
+	and $\tau_d^{2}$, but directly models the variance components as an
+	exponentiated linear function of the external covariates. The linear functions
+	are modelled on the log-scale to ensure non-negative variance components.
+	Additionally, it allows for more flexibility in the error variance
+	prior model. Note that in the limit $\zeta \to 0$ and $\nu=1/2$, the error
+	variance follows the Jeffrey`s prior from (\ref{md-eq:prior}) in the MD. 
+	Throughout the
+	following, whenever $\zeta$ and $\nu$ are not specified, we use $\zeta \to 0$ 
+	and $\nu=1/2$.
+
+  The hyperparameters $\bm{\alpha} = \begin{bmatrix} \bm{\alpha}_{\text{feat}}
+  \tr & \bm{\alpha}_{\text{drug}} \tr \end{bmatrix} \tr$ in model 
+  (\ref{eq:altprior}) are estimated by empirical Bayes maximisation of
+  the marginal log likelihood:
+  \begin{subequations}\label{eq:altmll}
+    \begin{align}
+      \log p (\bm{\alpha} | \mathbf{Y}) & = \sum_{d=1}^D
+      \log \[ \int_{\bm{\beta}_d} \int_{\sigma^2_d} \mathcal{L} 
+      (\mathbf{y}_d | \bm{\beta}_d, \sigma^2_d ) \pi(\bm{\beta}_d | \sigma^2_d) 
+      \pi(\sigma^2_d) d\bm{\beta}_d d\sigma_d^2 \] \\
+      & \propto - \sum_{d=1}^D \bigg[ \log \left|\I + \mathbf{X} 
+      \diag \[ \exp\( \tilde{\mathbf{C}}_d \bm{\alpha} \) \] \mathbf{X} \tr 
+      \right| \\
+      & \,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\, + (n + 1) \log \( \mathbf{y}_d \tr 
+      \left\{ \I + \mathbf{X} 
+      \diag \[ \exp\( \tilde{\mathbf{C}}_d \bm{\alpha} \) \] \mathbf{X} 
+      \tr\right\}^{-1} \mathbf{y}_d \) \bigg],
+    \end{align}
+  \end{subequations}
+  where $\tilde{\mathbf{C}}_d = \begin{bmatrix} \mathbf{C}_d & 
+  \mathbf{1}_{p \times 1} \cdot \mathbf{z}_d \tr \end{bmatrix}$, and
+  $\mathbf{C}_d = \begin{bmatrix} \mathbf{c}_{1d} & \cdots & 
+  \mathbf{c}_{pd} \end{bmatrix} \tr$.
+  The first term in the sum of (\ref{eq:altmll}) is concave in 
+  $\bm{\alpha}$, while the second term is convex in $\bm{\alpha}$. This renders
+  the marginal log likelihood a sum of non-concave (as well as non-convex) 
+  functions. General purpose optimisers, such as \texttt{optim} in \texttt{R} 
+  are generally not well suited to maximise non-concave functions. Maximisation
+  is therefore done through a minorise-maximisation (MM) algorithm, 
+  similar to \cite{zeng_incorporating_2020}.
+
+  The MM algorithm seeks to minorise the marginal log likelihood,
+  $\log p(\bm{\alpha} | \mathbf{Y})$, at current
+  parameter estimate $\bm{\alpha}^{(t)}$ with function 
+  $g(\bm{\alpha} | \bm{\alpha}^{(t)})$, such that
+  $\forall \bm{\alpha}: g(\bm{\alpha} | \bm{\alpha}^{(t)}) \leq 
+  \log p(\bm{\alpha} | \mathbf{Y})$ 
+  and $g(\bm{\alpha}^{(t)} | \bm{\alpha}^{(t)}) = 
+  \log p(\bm{\alpha} | \mathbf{Y})$. Then, the sequence
+  $$
+  \bm{\alpha}^{(t + 1)} = \underset{\bm{\alpha}}{\argmax} 
+  g(\bm{\alpha} | \bm{\alpha}^{(t+1)})
+  $$
+  converges to a local maximum of $f(\bm{\alpha})$. Generally, the minorisation
+  function is chosen such that it is concave and easy to maximise. In our case,
+  a suitable minorisation is:
+  \begin{subequations}\label{eq:minor}
+    \begin{align}
+      g(\bm{\alpha} | \bm{\alpha}^{(t+1)}) = -\sum_{d=1}^D \bigg[
+      & \diag \( \X \tr \mathbf{A}_d^{(t)} \X \) \tr \exp (\tilde{\mathbf{C}}_d
+      \bm{\alpha}) \\
+      & + \frac{n + 1}{\mathbf{y}_d \tr \mathbf{A}_d^{(t)} \mathbf{y}_d} 
+      \mathbf{y}_d \tr \left\{ \I + \mathbf{X} \diag \[ \exp\( 
+      \tilde{\mathbf{C}}_d \bm{\alpha} \) \] \mathbf{X} \tr \right\}^{-1} 
+      \mathbf{y}_d\bigg],
+    \end{align}
+  \end{subequations}
+  where $\mathbf{A}_d^{(t)} = \left\{ \I + \mathbf{X} \diag \[ \exp\( 
+  \tilde{\mathbf{C}}_d \bm{\alpha}^{(t)} \) \] \mathbf{X} \tr \right\}^{-1}$.
+  Note that (\ref{eq:minor}) is concave. However, it is slow to optimise. 
+  In \cite{zeng_incorporating_2020}, another minorization is proposed for
+  the second term:
+  $$
+  \mathbf{y}_d \tr \left\{ \I + \mathbf{X} \diag \[ \exp\( 
+  \tilde{\mathbf{C}}_d \bm{\alpha} \) \] \mathbf{X} \tr \right\}^{-1} 
+  \mathbf{y}_d = \underset{\delta}{\max} 
+  - ||\mathbf{y}_d - \mathbf{X} \delta ||_2^2 - 
+  \delta \tr \diag \[ \exp\(
+  -\tilde{\mathbf{C}}_d \bm{\alpha} \) \] \delta.
+  $$
+  The extra parameter $\delta$ is estimated by any standard ridge regression
+  package, such as \texttt{glmnet} in \texttt{R}.
+
+  \section{Extra simulation results}\label{sec:simulations}
+  The lasso and ridge models were fit separately per drug, using the \texttt{R}
+	package \texttt{glmnet} \cite[]{friedman_regularization_2010} with cross
+	validated
+	penalty parameters. The \texttt{xtune} model is fit with the corresponding 
+	\texttt{R} package \cite[]{ni_bayesian_2019}.
+
+	Performance measures are estimated from the remaining 914 cell lines
+
+<<simulation_gdsc_res1>>=
+res <- read.table("results/simulations_gdsc_res1.txt", row.names=NULL)
+temp <- res[, 1]
+res <- as.matrix(res[, -1])
+rownames(res) <- temp
+true.phi <- mean(1/(c(1, 1, 3, 7) %*% t(unname(model.matrix(~ factor(1:4))))))
+est.phi <- median(1/res[rownames(res)=="alphaf0", 1])
+est.lambdaf <- apply(res[rownames(res)=="lambdaf", c(1, 2)], 2, median)
+
+tab <- cbind(
+  apply(
+    aggregate(res[substr(rownames(res), 1, 6)=="emsel.", ], 
+              by=list(rownames(res[substr(rownames(res), 1, 6)=="emsel.", ])),
+              FUN="mean")[, -1], 2, median),
+  apply(
+    aggregate(res[substr(rownames(res), 1, 6)=="emseh.", ], 
+              by=list(rownames(res[substr(rownames(res), 1, 6)=="emseh.", ])),
+              FUN="mean")[, -1], 2, median),
+  apply(
+    aggregate(res[substr(rownames(res), 1, 5)=="emse.", ], 
+              by=list(rownames(res[substr(rownames(res), 1, 5)=="emse.", ])),
+              FUN="mean")[, -1], 2, median),
+  apply(
+    aggregate(res[substr(rownames(res), 1, 5)=="pmse.", ], 
+              by=list(rownames(res[substr(rownames(res), 1, 5)=="pmse.", ])),
+              FUN="mean")[, -1], 2, median))
+
+tab <- round(tab, 3)
+tab[which(tab[, 1]==min(tab[, 1], na.rm=TRUE)), 1] <- 
+  paste0("\\textbf{", tab[which(tab[, 1]==min(tab[, 1], na.rm=TRUE)), 1], "}")
+tab[which(tab[, 2]==min(tab[, 2], na.rm=TRUE)), 2] <- 
+  paste0("\\textbf{", tab[which(tab[, 2]==min(tab[, 2], na.rm=TRUE)), 2], "}")
+tab[which(tab[, 3]==min(tab[, 3], na.rm=TRUE)), 3] <- 
+  paste0("\\textbf{", tab[which(tab[, 3]==min(tab[, 3], na.rm=TRUE)), 3], "}")
+tab[which(tab[, 4]==min(tab[, 4], na.rm=TRUE)), 4] <- 
+  paste0("\\textbf{", tab[which(tab[, 4]==min(tab[, 4], na.rm=TRUE)), 4], "}")
+colnames(tab) <- c("$\\text{EMSE}_{\\text{bottom}}$",
+                   "$\\text{EMSE}_{\\text{top}}$",
+                   "$\\text{EMSE}_{\\text{total}}$",
+                   "$\\text{PMSE}$")     
+rownames(tab)[c(1, 2)] <- c("NIG$_{\\text{f}}^-$",
+                            "NIG$_{\\text{f}}$")
+@
+
+  $\lambda_{\text{feat}}$ is overestimated with median estimate 
+	\Sexpr{round(est.lambdaf[2], 2)} (the
+	true value is 1). The overestimated $\lambda_{\text{feat}}$ translates 
+	into underestimated $\gamma_{jd}^2$ variances.
+
+	Table \ref{tab:simulations_gdsc_tab1} displays median estimation MSE 
+	(EMSE) and prediction MSE (PMSE) for simulation Scenario 1, calculated on an 
+	independent test data set. EMSE is further split into the contribution of the
+	bottom 90\% of the $\beta_{jd}$ in terms of size (in absolute value) and the 
+	top 10\% of the $\beta_{jd}$ in size. The table includes as a reference the 
+	null model, which is an empty model that only fits the means of the $\y_{d}$.
+	Estimation MSE (EMSE) and prediction MSE (PMSE) give measures of predictive 
+	performance. Focussing on MSE, we see that the NIG$_{\text{f}}$ model, 
+	that includes the external covariates, performs best, both in terms of 
+	estimation and prediction. The lower MSEs confirm that the model that
+	includes the external data (NIG$_{\text{f}}$) learns the underlying structure 
+	in the data better than model NIG$_{\text{f}}^-$ that does not include the
+	external data. Furthermore, we see that estimation of the larger $\beta_{jd}$ 
+	is where ridge loses the most compared to the NIG and lasso models. 
+	The heavier tails of the NIG and lasso priors allow for larger $\beta_{jd}$
+	estimates and thus lower EMSE for these $\beta_{jd}$.
+
+<<simulations_gdsc_tab1>>=
+library(knitr, quietly=TRUE, verbose=FALSE)
+suppressWarnings(library(kableExtra))
+options(knitr.kable.NA="", digits=3)
+kable(tab, col.names=colnames(tab), align="r",
+      caption="Median performance measure for simulation Scenario 1, estimated on the test data (lowest in bold).", 
+      format="latex", booktabs=TRUE, escape=FALSE) %>%
+  kable_styling(latex_options=c("HOLD_position"))
+@
+
+<<simulation_gdsc_res2>>=
+res <- read.table("results/simulations_gdsc_res2.txt", row.names=NULL)
+temp <- res[, 1]
+res <- as.matrix(res[, -1])
+rownames(res) <- temp
+true.chi <- mean(1/(c(1, 1, 3, 7) %*% t(unname(model.matrix(~ factor(1:4))))))
+est.chi <- median(1/res[rownames(res)=="alphad0", 1])
+est.lambdad <- apply(res[rownames(res)=="lambdad", c(1, 2)], 2, median)
+tab <- cbind(
+  apply(
+    aggregate(res[substr(rownames(res), 1, 6)=="emsel.", ], 
+              by=list(rownames(res[substr(rownames(res), 1, 6)=="emsel.", ])),
+              FUN="mean")[, -1], 2, median),
+  apply(
+    aggregate(res[substr(rownames(res), 1, 6)=="emseh.", ], 
+              by=list(rownames(res[substr(rownames(res), 1, 6)=="emseh.", ])),
+              FUN="mean")[, -1], 2, median),
+  apply(
+    aggregate(res[substr(rownames(res), 1, 5)=="emse.", ], 
+              by=list(rownames(res[substr(rownames(res), 1, 5)=="emse.", ])),
+              FUN="mean")[, -1], 2, median),
+  apply(
+    aggregate(res[substr(rownames(res), 1, 5)=="pmse.", ], 
+              by=list(rownames(res[substr(rownames(res), 1, 5)=="pmse.", ])),
+              FUN="mean")[, -1], 2, median))
+tab <- round(tab, 3)
+tab[which(tab[, 1]==min(tab[, 1], na.rm=TRUE)), 1] <- 
+  paste0("\\textbf{", tab[which(tab[, 1]==min(tab[, 1], na.rm=TRUE)), 1], "}")
+tab[which(tab[, 2]==min(tab[, 2], na.rm=TRUE)), 2] <- 
+  paste0("\\textbf{", tab[which(tab[, 2]==min(tab[, 2], na.rm=TRUE)), 2], "}")
+tab[which(tab[, 3]==min(tab[, 3], na.rm=TRUE)), 3] <- 
+  paste0("\\textbf{", tab[which(tab[, 3]==min(tab[, 3], na.rm=TRUE)), 3], "}")
+tab[which(tab[, 4]==min(tab[, 4], na.rm=TRUE)), 4] <- 
+  paste0("\\textbf{", tab[which(tab[, 4]==min(tab[, 4], na.rm=TRUE)), 4], "}")
+colnames(tab) <- c("$\\text{EMSE}_{\\text{bottom}}$",
+                   "$\\text{EMSE}_{\\text{top}}$",
+                   "$\\text{EMSE}_{\\text{total}}$",
+                   "$\\text{PMSE}$")      
+rownames(tab)[c(1, 2)] <- c("NIG$_{\\text{d}}^-$",
+                            "NIG$_{\\text{d}}$")
+@
+
+  $\lambda_{\text{feat}}$ is overestimated with median estimate
+	\Sexpr{round(est.lambdaf[1], 2)} (the true value is one) and thus 
+	again underestimates the 
+	prior $\gamma_{jd}^2$ variance.
+
+	Median $\lambda_{\text{drug}}=\Sexpr{round(est.lambdad[2], 2)}$ is
+	overestimated compared to true value one. 
+	Scale $\lambda_{\text{drug}}$ is underestimated with median estimate
+	\Sexpr{round(est.lambdad[1], 2)} compared to the true value one.
+
+  Median performance measures for simulation Scenario 2 are displayed in Table
+	\ref{tab:simulations_gdsc_tab2}. Again, we see that NIG$_{\text{d}}$ performs
+	best in terms of MSE. However, NIG$_{\text{d}}^-$ performance is the same 
+	here, indicating that the external covariates do not add much here.
+	Similarly as in Scenario 1, most of the estimation performance gain over ridge
+	is through the top 10\% of the $\beta_{jd}$.
+
+<<simulations_gdsc_tab2>>=
+library(knitr, quietly=TRUE, verbose=FALSE)
+suppressWarnings(library(kableExtra))
+options(knitr.kable.NA="", digits=3)
+kable(tab, col.names=colnames(tab), align="r",
+      caption="Median performance measure for simulation Scenario 2, estimated on the test data (lowest in bold).",
+      format="latex", booktabs=TRUE, escape=FALSE) %>%
+  kable_styling(latex_options=c("HOLD_position"))
+@
+
+<<simulation_gdsc_res3>>=
+res <- read.table("results/simulations_gdsc_res3.txt", row.names=NULL)
+temp <- res[, 1]
+res <- as.matrix(res[, -1])
+rownames(res) <- temp
+true.chi <- mean(1/(c(1, 1, 3, 7) %*% t(unname(model.matrix(~ factor(1:4))))))
+est.chi <- median(1/res[rownames(res)=="alphad0", 1])
+est.lambdad <- apply(res[rownames(res)=="lambdad", c(1, 2)], 2, median)
+true.phi <- mean(1/(c(1, 1, 3, 7) %*% t(unname(model.matrix(~ factor(1:4))))))
+est.phi <- median(1/res[rownames(res)=="alphaf0", 1])
+est.lambdaf <- apply(res[rownames(res)=="lambdaf", c(1, 2)], 2, median)
+
+phi.ratio <- apply(apply(sapply(0:3, function(s) {
+  res[rownames(res)==paste0("alphaf" ,s), 2]}) %*%
+    t(cbind(1, rbind(0, diag(3)))), 2, function(m) {
+      res[rownames(res)=="alphaf0", 2]/m}), 2, median)
+chi.ratio <- apply(apply(sapply(0:3, function(s) {
+  res[rownames(res)==paste0("alphad" ,s), 2]}) %*%
+    t(cbind(1, rbind(0, diag(3)))), 2, function(m) {
+      res[rownames(res)=="alphad0", 2]/m}), 2, median)
+true.ratio <- as.numeric(1/(c(1, 1, 3, 7) %*%
+                              t(unname(model.matrix(~ factor(1:4))))))
+
+tab <- cbind(
+  apply(
+    aggregate(res[substr(rownames(res), 1, 6)=="emsel.", ], 
+              by=list(rownames(res[substr(rownames(res), 1, 6)=="emsel.", ])),
+              FUN="mean")[, -1], 2, median),
+  apply(
+    aggregate(res[substr(rownames(res), 1, 6)=="emseh.", ], 
+              by=list(rownames(res[substr(rownames(res), 1, 6)=="emseh.", ])),
+              FUN="mean")[, -1], 2, median),
+  apply(
+    aggregate(res[substr(rownames(res), 1, 5)=="emse.", ], 
+              by=list(rownames(res[substr(rownames(res), 1, 5)=="emse.", ])),
+              FUN="mean")[, -1], 2, median),
+  apply(
+    aggregate(res[substr(rownames(res), 1, 5)=="pmse.", ], 
+              by=list(rownames(res[substr(rownames(res), 1, 5)=="pmse.", ])),
+              FUN="mean")[, -1], 2, median))
+tab <- round(tab, 3)
+tab[which(tab[, 1]==min(tab[, 1], na.rm=TRUE)), 1] <-
+  paste0("\\textbf{", tab[which(tab[, 1]==min(tab[, 1], na.rm=TRUE)), 1], "}")
+tab[which(tab[, 2]==min(tab[, 2], na.rm=TRUE)), 2] <-
+  paste0("\\textbf{", tab[which(tab[, 2]==min(tab[, 2], na.rm=TRUE)), 2], "}")
+tab[which(tab[, 3]==min(tab[, 3], na.rm=TRUE)), 3] <-
+  paste0("\\textbf{", tab[which(tab[, 3]==min(tab[, 3], na.rm=TRUE)), 3], "}")
+tab[which(tab[, 4]==min(tab[, 4], na.rm=TRUE)), 4] <-
+  paste0("\\textbf{", tab[which(tab[, 4]==min(tab[, 4], na.rm=TRUE)), 4], "}")
+colnames(tab) <- c("$\\text{EMSE}_{\\text{bottom}}$",
+                   "$\\text{EMSE}_{\\text{top}}$",
+                   "$\\text{EMSE}_{\\text{total}}$",
+                   "$\\text{PMSE}$")
+rownames(tab)[c(1, 2)] <- c("NIG$_{\\text{f}+\\text{d}}^-$",
+                            "NIG$_{\\text{f}+\\text{d}}$")
+@
+
+  Median $\lambda_{\text{feat}}$ and $\lambda_{\text{drug}}$ are
+	are over- and underestimated with \Sexpr{round(est.lambdaf[2], 2)}
+	and \Sexpr{round(est.lambdad[2], 2)}, respectively, compared to their true
+	value 1.
+
+  NIG$_{\text{f}+\text{d}}^-$ also over- and
+	underestimates 
+	$\lambda_{\text{feat}}$ and $\lambda_{\text{drug}}$ with median estimates
+	\Sexpr{round(est.lambdaf[1], 2)}
+	and \Sexpr{round(est.lambdad[1], 2)}, respectively.
+
+  Table 
+	\ref{tab:simulations_gdsc_tab3}, with median performance measures, shows
+	that NIG$_{\text{f}+\text{d}}$ outperforms the other models in terms of
+	prediction and estimation MSE, with again most of the edge over ridge in the
+	top 10\% of the $\beta_{jd}$.
+
+<<simulations_gdsc_tab3>>=
+library(knitr, quietly=TRUE, verbose=FALSE)
+suppressWarnings(library(kableExtra))
+options(knitr.kable.NA="", digits=3)
+kable(tab, col.names=colnames(tab), align="r",
+      caption="Median performance measure for simulation Scenario 3, estimated on the test data (lowest in bold).",
+      format="latex", booktabs=TRUE, escape=FALSE) %>%
+  kable_styling(latex_options=c("HOLD_position"))
+@
   \bibliographystyle{author_short3}  

### 2020-04-06 docs/manuscript1.Rnw
fixed small mistake

@@
 <<analysis_gdsc>>==
-# fit <- read.table("results/analysis_gdsc_fit1.txt", row.names=NULL)
-# temp <- fit[, 1]
-# fit <- as.matrix(fit[, -1])
-# rownames(fit) <- temp
-# 
-# res <- read.table("results/analysis_gdsc_res1.txt", row.names=NULL)
-# temp <- res[, 1]
-# res <- as.matrix(res[, -1])
-# rownames(res) <- temp
-# 
-# est1 <- cbind(fit[, 1], c(NA, 0, 0, NA, 0), fit[, 2])
-# est1 <- round(est1, 2)
-# est1[is.na(est1)] <- "-"
-# colnames(est1) <- c("intercept", "not in pathway", "in pathway")
-# est2 <- cbind(fit[, 3:5], c(NA, 0, NA, 0, NA), fit[, c(7:6)],
-#               c(NA, 0, NA, 0, NA))
-# est2 <- round(est2, 2)
-# est2[is.na(est2)] <- "-"
-# colnames(est2) <- c("intercept", "experimental", "development", "approved",
-#                     "unkown", "targeted", "cytotoxic")
-# 
-# rownames(est1)[c(1:4)] <- rownames(est2)[c(1:4)] <-
-#   c("NIG$_{\\text{f}+\\text{d}}^-$", "NIG$_{\\text{f}+\\text{d}}$",
-#     "NIG$_{\\text{f}}$", "NIG$_{\\text{d}}$")
-# 
-# tab <- cbind(apply(res[rownames(res)=="pmse", ], 2, median))
-# tabm <- tab
-# tabm <- round(tabm, 3)
-# tabm[which(tabm[, 1]==min(tabm[, 1], na.rm=TRUE)), 1] <-
-#   paste0("\\textbf{", tabm[which(tabm[, 1]==min(tabm[, 1], na.rm=TRUE)), 1], "}")
-# colnames(tabm) <- c("$\\text{PMSE}_{\\text{test}}$")
-# rownames(tabm)[c(1:4)] <- rownames(est2)[c(1:4)] <-
-#   c("NIG$_{\\text{f}+\\text{d}}^-$", "NIG$_{\\text{f}+\\text{d}}$",
-#     "NIG$_{\\text{f}}$", "NIG$_{\\text{d}}$")
-# 
-# library(glmnet)
-# kurtosis <- function(x) {
-#   n <- length(x)
-#   kurt <- (n + 1)*n*sum((x - mean(x))^4)/((n - 1)*(n - 2)*(n - 3)*var(x)^2) -
-#     3*(n - 1)^2/((n - 2)*(n - 3))
-#   replace(kurt, !is.finite(kurt), NA)
-# }
-# load("results/analysis_gdsc_fit1.Rdata")
-# best2.semnig <- fit2.semnig$vb$mu
-# best1.lasso <- lapply(fit1.lasso, function(s) {
-#   unname(coef(s, "lambda.min")[-1, ])})
-# best1.ridge <- lapply(fit1.ridge, function(s) {
-#   unname(coef(s, "lambda.min")[-1, ])})
-# best1.bSEM <- lapply(fit1.bSEM$vb$beta, function(s) {s[, 2]})
-# kurts <- c(mean(sapply(best2.semnig, kurtosis), na.rm=TRUE),
-#            mean(sapply(best1.lasso, kurtosis), na.rm=TRUE),
-#            mean(sapply(best1.ridge, kurtosis), na.rm=TRUE),
-#            mean(sapply(best1.bSEM, kurtosis), na.rm=TRUE))
-# ranges <- rbind(range(unlist(lapply(1:length(
-#   fit2.semnig$eb$mpriord), function(d) {
-#     fit2.semnig$eb$mpriord[d]*fit2.semnig$eb$mpriorf[[d]]}))),
-#   fit1.bSEM$eb$b[nrow(fit1.bSEM$eb$b), ]/
-#     (fit1.bSEM$eb$a[nrow(fit1.bSEM$eb$a), ] - 1))
+fit <- read.table("results/analysis_gdsc_fit1.txt", row.names=NULL)
+temp <- fit[, 1]
+fit <- as.matrix(fit[, -1])
+rownames(fit) <- temp
+
+res <- read.table("results/analysis_gdsc_res2.txt", row.names=NULL)
+temp <- res[, 1]
+res <- as.matrix(res[, -1])
+rownames(res) <- temp
+
+est1 <- cbind(fit[, 1], c(NA, 0, 0, NA, 0), fit[, 2])
+est1 <- round(est1, 2)
+est1[is.na(est1)] <- "-"
+colnames(est1) <- c("intercept", "not in pathway", "in pathway")
+est2 <- cbind(fit[, 3:5], c(NA, 0, NA, 0, NA), fit[, c(7:6)],
+              c(NA, 0, NA, 0, NA))
+est2 <- round(est2, 2)
+est2[is.na(est2)] <- "-"
+colnames(est2) <- c("intercept", "experimental", "development", "approved",
+                    "unkown", "targeted", "cytotoxic")
+
+rownames(est1)[c(1:4)] <- rownames(est2)[c(1:4)] <-
+  c("NIG$_{\\text{f}+\\text{d}}^-$", "NIG$_{\\text{f}+\\text{d}}$",
+    "NIG$_{\\text{f}}$", "NIG$_{\\text{d}}$")
+
+tab <- cbind(apply(res[rownames(res)=="pmse", ], 2, median))
+tabm <- tab
+tabm <- round(tabm, 3)
+tabm[which(tabm[, 1]==min(tabm[, 1], na.rm=TRUE)), 1] <-
+  paste0("\\textbf{", tabm[which(tabm[, 1]==min(tabm[, 1], na.rm=TRUE)), 1], "}")
+colnames(tabm) <- c("$\\text{PMSE}_{\\text{test}}$")
+rownames(tabm)[c(1:4)] <- rownames(est2)[c(1:4)] <-
+  c("NIG$_{\\text{f}+\\text{d}}^-$", "NIG$_{\\text{f}+\\text{d}}$",
+    "NIG$_{\\text{f}}$", "NIG$_{\\text{d}}$")
+
+library(glmnet)
+kurtosis <- function(x) {
+  n <- length(x)
+  kurt <- (n + 1)*n*sum((x - mean(x))^4)/((n - 1)*(n - 2)*(n - 3)*var(x)^2) -
+    3*(n - 1)^2/((n - 2)*(n - 3))
+  replace(kurt, !is.finite(kurt), NA)
+}
+load("results/analysis_gdsc_fit1.Rdata")
+best2.semnig <- fit2.semnig$vb$mu
+best1.lasso <- lapply(fit1.lasso, function(s) {
+  unname(coef(s, "lambda.min")[-1, ])})
+best1.ridge <- lapply(fit1.ridge, function(s) {
+  unname(coef(s, "lambda.min")[-1, ])})
+best1.bSEM <- lapply(fit1.bSEM$vb$beta, function(s) {s[, 2]})
+kurts <- c(mean(sapply(best2.semnig, kurtosis), na.rm=TRUE),
+           mean(sapply(best1.lasso, kurtosis), na.rm=TRUE),
+           mean(sapply(best1.ridge, kurtosis), na.rm=TRUE),
+           mean(sapply(best1.bSEM, kurtosis), na.rm=TRUE))
+ranges <- rbind(range(unlist(lapply(1:length(
+  fit2.semnig$eb$mpriord), function(d) {
+    fit2.semnig$eb$mpriord[d]*fit2.semnig$eb$mpriorf[[d]]}))),
+  fit1.bSEM$eb$b[nrow(fit1.bSEM$eb$b), ]/
+    (fit1.bSEM$eb$a[nrow(fit1.bSEM$eb$a), ] - 1))
 ranges <- matrix(1, nrow=10, ncol=10)

### 2020-04-06 docs/manuscript1.Rnw
small changes

@@
   sub('\\usepackage[]{color}', '\\usepackage{xcolor}', x, fixed=TRUE)})
+options(knitr.kable.NA="")
 @
@@
   several choices of $\beta_{jd}$ prior.
-
 <<dens_kappa, fig.cap="Implied prior densities $\\pi(\\kappa_{jd})$ for the (a) NIG, (b) Student's $t$, and (c) lasso priors.
@@
 	NIG$_{\text{f}}$, NIG$_{\text{d}}$, and NIG$_{\text{f}+\text{d}}$ models.
-	All models are fit on 100 randomly selected cell lines. We repeat every
-	simulation Scenario 100 times.
-
-	SM Section \ref{sm-sec:simulations} contains additional simulation results for
+	In the language of \cite{bernardo_shrink_2011} as introduced in 
+	Section \ref{sec:introduction}, models NIG$_{\text{f}}$ and NIG$_{\text{d}}$
+	may be described as local and global shrinkage rules, respectively, as 
+	opposed to the global-local shrinkage models NIG$_{\text{f}+\text{d}}$ and
+	NIG$_{\text{f}+\text{d}}^-$. All models are fit on 100 randomly selected cell
+	lines. We repeat every simulation Scenario 100 times.
+
+	SM Section \ref{sm-sec:simulations} contains
+	additional simulation results for
 	the NIG model, plus results for 
@@
 	\cite[]{ni_bayesian_2019} and (iv) its
-	multivariate Bayesian extension as introduced in MD Section 
+	multivariate Bayesian extension as introduced in SM Section 
 	\ref{sm-sec:altprior}.
@@
 	$\bm{\alpha}_{\text{feat}}$ is slightly but 
-	consistently underestimated. This results in overestimation on the $\phi_{jd}$
-	scale as depicted in Figure \ref{fig:simulations_gdsc_est1}b. 
+	consistently underestimated. This results in slight overestimation on the
+	$\phi_{jd}$ scale as depicted in Figure \ref{fig:simulations_gdsc_est1}b. 
 	Model NIG$_{\text{f}}^-$ (that excludes the external covariates) gives a 
@@
 
+  Figure \ref{fig:simulations_gdsc_est4} displays the median $\phi_{jd}$ and 
+  $\chi_d$ estimates for different noise levels in simulation Scenario 4. 
+  The simulation shows that with increasing noise level, the estimated 
+  prior means $\phi_{jd}$ and $\chi_d$ for the four groups of external 
+  covariates become more and more alike. In other words, noise in the external
+  covariates impedes estimation of $\bm{\alpha}_{\text{feat}}$ and
+  $\bm{\alpha}_{\text{drug}}$, as expected.
 <<simulations_gdsc_est4, fig.cap="Simulation results for Scenario 3: estimated prior means (a) $\\phi_{jd}$, and (b) $\\chi_{d}$ for different levels of noise in the external covariates.", out.width="100%", fig.asp=1/2>>=
@@
 
-  \subsection{Discussion}
-  Underestimation of variance components is 
-	a common phenomenon in variational Bayes approximations, so the 
-	underestimation of the mean of the drug variance components $\chi_d$ in
-	Scenarios 2 and 3 is somewhat expected. In Scenario 3, this is compensated for
-	by the overestimation of the $\phi_{jd}$. We note that increasing the number
-	of cell lines $n$ and increasing the number of drugs $D$ improves estimation
-	of the $\chi_d$. However, this is not the most interesting situation since it
-	does not correspond to
-
-	In addition, 
-	$\lambda_{\text{feat}}$ is a higher level scale
-	parameter, which are, in general, hard to estimate.
+  Estimation of only $\bm{\alpha}_{\text{feat}}$ is relatively unbiased, as
+  evident from simulation Scenario 1. In contrast,
+  $\bm{\alpha}_{\text{drug}}$ are underestimated (Scenarios 2 and 3).
+  $\bm{\alpha}_{\text{drug}}$ is inversely related to the prior $\bm{\beta}$
+  variance. Underestimation of variance components in regression is 
+	a common phenomenon in variational Bayes approximations and a possible
+	explanation of the estimation bias in Scenarios 2 and 3. 
+	In Scenario 3, this is compensated for
+	by the overestimation of the $\phi_{jd}$, such that the estimated median 
+	variance $\V(\beta_{jd})$ are stil biased, but only slightly underestimated
+	(Figure \ref{fig:simulations_gdsc_var}). We note that an increase in cell 
+	lines $n$ and/or the number of drugs $D$ improves estimation
+	of the $\chi_d$. However, we choose to highlight the high-dimensional 
+	(more parameters than samples) situation here, because it is the most common
+	in practice.
+<<simulations_gdsc_var, fig.cap="Simulation results for Scenario 3: estimated prior variances $\\V(\\beta_{jd})$ versus true values, with line of identity (dotted).", out.width="50%", fig.asp=1>>= 
+@
 
@@
 	inhibitory concentration (IC50), using the
-	gene expression data. Other choices of sensitivity measures than IC50 are 
+	gene expression and gene mutation data. Other choices of sensitivity measures
+	than IC50 are 
 	possible, but a discussion on the pros and cons of different sensitivity
@@
 	of missings until no more missing values remain, we end up with IC50 
-	estimates for 482 cell lines on 168 drugs.
+	estimates for between 388 and 1043 cell lines on 251 drugs. The number
+	of cell lines differs between drugs, because not all drug and cell line 
+	combinations are available.
 
@@
 	drugs and the action (unkown, cytotoxic, or targeted) of the drugs. 
-	Furthermore, we have a binary feature covariate available that indicates
-	whether a gene belongs to the drug target pathway. The drug covariates are 
-	taken
-	directly from the GDSC database's annotation file. The feature covariate
-	was created by comparing the target pathways in the GDSC annotation to
-	the KEGG \cite[]{kanehisa_kegg:_2000} and reactome 
-	\cite[]{fabregat_reactome_2018} repositories. The external covariates are 
-	dummy-coded with
-	reference categories clinically approved drugs, cytotoxic drugs, and 
-	features that are not in the target pathway.
-
+	These drug covariates are taken 
+	directly from the GDSC database's annotation file and dummy-coded
+	with reference categories clinically approved drugs and cytotoxic drugs. 
 	We expect that drugs that have been clinically approved are easiest to 
@@
 	expect the targeted drugs to yield the largest prior $\beta_{jd}$ variances, 
-	followed by the cytotoxic drugs, and the unkown target drugs. For the
-	feature covariate, we expect that genes that are in the pathway of the drug
+	followed by the cytotoxic drugs, and the unkown target drugs. Note that large
+	$\beta_{jd}$ variances translate to large
+	prior $\gamma_{jd}^2$ and $\tau_d^2$ means.
+
+	Furthermore, we have a binary feature covariate available that indicates
+	whether a gene belongs to the drug target pathway. The feature covariate
+	was created by comparing the target pathways in the GDSC annotation to
+	the KEGG \cite[]{kanehisa_kegg:_2000} and reactome 
+	\cite[]{fabregat_reactome_2018} repositories. The reference category here is
+	features that are not in the target pathway. For this external covariate,
+	we expect that genes that are in the pathway of the drug
 	are more predictive than genes that are not, i.e., 
 	they have larger prior $\beta_{jd}$ variances than drugs that are not in 
-	the pathway. Note that large $\beta_{jd}$ variances translate to large
-	prior $\gamma_{jd}^2$ and $\tau_d^2$ means.
-
-	\subsection{Results}
-	A preliminary step selects, for each
-	drug separately, 50 features that are not in the pathway of the drug with
-	maximum variance over the cell lines and 50 features that are in the pathway 
-	of the drug with maximum variance over the cell lines. If less than 50 
-	features in the pathway are available,
-	we select all available features in the pathway. This strategy ensures
-	that the features are somewhat balanced with respect to the external feature
-	covariate, i.e., pathway membership.
-
-	To compute performance measures, we randomly split the data into training and
-	test sets, containing 
-	241 cell lines each, estimate the models on the training data and compute 
-	performance measures on the test data. We repeated this procedure 50 times
-	and average performance measures over the splits.
-	The estimated 
-	models are: four models like in (\ref{eq:prior}) but with different external 
-	variables, named NIG$_{\text{f}+\text{d}}^-$, NIG$_{\text{f}+\text{d}}$,
-	NIG$_{\text{f}}$, and NIG$_{\text{d}}$ in the following. The nomenclature is
-	the same as in Section \ref{sec:simulations}, i.e., 
-	NIG$_{\text{f}+\text{d}}^-$ excludes the external covariates and estimates
-	one overall $\phi$ and one overall $\chi$, while NIG$_{\text{f}+\text{d}}$ 
-	uses the external covariates to estimate both $\bm{\alpha}_{\text{feat}}$ and
-	$\bm{\alpha}_{\text{drug}}$. NIG$_{\text{f}}$ and NIG$_{\text{d}}$ estimate 
-	only $\bm{\alpha}_{\text{feat}}$ or $\bm{\alpha}_{\text{drug}}$, respectively,
-	keeping either $\forall d: \tau_d^2=1$ or $\forall j,d: \gamma_{jd}^2=1$
-	fixed. In the language of \cite{bernardo_shrink_2011} as introduced in 
-	Section \ref{sec:introduction}, models NIG$_{\text{f}}$ and NIG$_{\text{d}}$
-	may be described as local and global shrinkage rules, respectively, as 
-	opposed to the global-local shrinkage models NIG$_{\text{f}+\text{d}}$ and
-	NIG$_{\text{f}+\text{d}}^-$.
-
-	In addition, we estimate regular lasso and ridge models, and a
-	bSEM model \cite[]{leday_gene_2017,kpogbezan_empirical_2017}. The bSEM model 
-	is similar to the NIG model in that it
+	the pathway.
+
+	The type of molecular marker may be included as external 
+	covariate, i.e., whether the feature is a gene expression or gene mutation.
+	As an alternative to direct inclusion of the mutation data, we use $p$-values
+	from the mutations, obtained from a $t$-test comparing IC50 values of 
+	mutated cell lines to unmutated genes. We expect that lower mutation 
+	$p$-values result in a larger prior $\beta_{jd}$ variances.
+
+	Lastly, $p$-values from an analysis of the CCLE data 
+	\cite[]{li_landscape_2019}, a database similar to the GDSC, are included as
+	external covariate. These $p$-values are obtained from a simple correlation 
+	between the IC50 values and the gene expressions from the CCLE data. 
+	The harmonic mean per gene is then used as external covariate for the GDSC
+	data analysis. Again, we expect a positive relation between these external
+	$p$-values and the larger prior $\beta_{jd}$ variances.
+
+	\subsection{Analyses}
+	Four analyses were conducted: 
+	\begin{itemize}
+	  \item{Analysis 1 includes gene expressions as predictors and $p$-values from
+	    the gene mutation data as external covariate. A pre-processing step 
+	    selects between 221 and 280 genes per drug for which a mutation $p$-value
+	    is available.}
+	  \item{Analysis 2 includes both gene  expressions and mutations as predictors
+	    with the feature type as external covariate, i.e., wether the feature
+	    is a expression or mutation. For this analysis, we pre-select 300 gene 
+	    expressions with maximum 
+	    variance and 295 gene mutations for which there are mutated as well as 
+	    unmutated cell lines available.}
+	  \item{Analysis 3 uses 500 gene expressions as features, selected based on
+	    maximum varaince. The CCLE $p$-values are included as external 
+	    covariates.}
+	  \item{Analysis 4 includes the gene expressions as features and both the
+	    annoted drug variables and pathway status of the genes as external 
+	    covariates. We pre-select 500 genes expressions based on maximum 
+	    variance.}
+	\end{itemize}
+
+	In all analyses we estimated the same models as in the simulations (Section 
+	\ref{sec:simulations} and SM Section \ref{sm-sec:simulations}):
+	(i) NIG$_{\text{f}+\text{d}}^-$, (ii) NIG$_{\text{f}+\text{d}}$, 
+	frequentist (iii) lasso and (iv) ridge models, (v) \texttt{xtune}, and
+	(vi) its multivariate Bayesian extension (see SM Section 
+	\ref{sm-sec:altprior}), named EBridge. In addition, in Analyses 2 and 
+	4, we fit a (vii) bSEM model 
+	\cite[]{leday_gene_2017,kpogbezan_empirical_2017}. The
+	bSEM model is similar to the NIG model in that it
 	draws the $\beta_{jd}$ from a conditionally normal distribution, but
@@
 	estimated in a similar fashion as in Section \ref{sec:estimation}.
-	The null model that fits the means of the $\y_d$ is included as a reference.
-
+
+	In all analyses we use all cell lines to estimate the hyperparameters 
+	presented in Section \ref{sec:results}. Prediction PMSE,
+	$\text{PMSE} = D^{-1} n^{-1} \sum_{d=1}^D 
+  \sum_{i=1}^n (\mathbf{y}_d - \mathbf{x}_i \tr \hat{\bm{\beta}_d})^2$, is
+  estimated by 10-fold cross validation. In the NIG and bSEM models that provide
+  the full posteriors, the posterior mean 
+  $\E(\bm{\beta}_d | \mathbf{y}_d)$ is used point estimate.
+
+	\subsection{Results}\label{sec:results}
 <<analysis_gdsc>>==
-fit <- read.table("results/analysis_gdsc_fit1.txt", row.names=NULL)
-temp <- fit[, 1]
-fit <- as.matrix(fit[, -1])
-rownames(fit) <- temp
-
-res <- read.table("results/analysis_gdsc_res2.txt", row.names=NULL)
-temp <- res[, 1]
-res <- as.matrix(res[, -1])
-rownames(res) <- temp
-
-est1 <- cbind(fit[, 1], c(NA, 0, 0, NA, 0), fit[, 2])
-est1 <- round(est1, 2)
-est1[is.na(est1)] <- "-"
-colnames(est1) <- c("intercept", "not in pathway", "in pathway")
-est2 <- cbind(fit[, 3:5], c(NA, 0, NA, 0, NA), fit[, c(7:6)],
-              c(NA, 0, NA, 0, NA))
-est2 <- round(est2, 2)
-est2[is.na(est2)] <- "-"
-colnames(est2) <- c("intercept", "experimental", "development", "approved",
-                    "unkown", "targeted", "cytotoxic")
-
-rownames(est1)[c(1:4)] <- rownames(est2)[c(1:4)] <-
-  c("NIG$_{\\text{f}+\\text{d}}^-$", "NIG$_{\\text{f}+\\text{d}}$",
-    "NIG$_{\\text{f}}$", "NIG$_{\\text{d}}$")
-
-tab <- cbind(apply(res[rownames(res)=="pmse", ], 2, median))
-tabm <- tab
-tabm <- round(tabm, 3)
-tabm[which(tabm[, 1]==min(tabm[, 1], na.rm=TRUE)), 1] <-
-  paste0("\\textbf{", tabm[which(tabm[, 1]==min(tabm[, 1], na.rm=TRUE)), 1], "}")
-colnames(tabm) <- c("$\\text{PMSE}_{\\text{test}}$")
-rownames(tabm)[c(1:4)] <- rownames(est2)[c(1:4)] <-
-  c("NIG$_{\\text{f}+\\text{d}}^-$", "NIG$_{\\text{f}+\\text{d}}$",
-    "NIG$_{\\text{f}}$", "NIG$_{\\text{d}}$")
-
-library(glmnet)
-kurtosis <- function(x) {
-  n <- length(x)
-  kurt <- (n + 1)*n*sum((x - mean(x))^4)/((n - 1)*(n - 2)*(n - 3)*var(x)^2) -
-    3*(n - 1)^2/((n - 2)*(n - 3))
-  replace(kurt, !is.finite(kurt), NA)
-}
-load("results/analysis_gdsc_fit1.Rdata")
-best2.semnig <- fit2.semnig$vb$mu
-best1.lasso <- lapply(fit1.lasso, function(s) {
-  unname(coef(s, "lambda.min")[-1, ])})
-best1.ridge <- lapply(fit1.ridge, function(s) {
-  unname(coef(s, "lambda.min")[-1, ])})
-best1.bSEM <- lapply(fit1.bSEM$vb$beta, function(s) {s[, 2]})
-kurts <- c(mean(sapply(best2.semnig, kurtosis), na.rm=TRUE),
-           mean(sapply(best1.lasso, kurtosis), na.rm=TRUE),
-           mean(sapply(best1.ridge, kurtosis), na.rm=TRUE),
-           mean(sapply(best1.bSEM, kurtosis), na.rm=TRUE))
-ranges <- rbind(range(unlist(lapply(1:length(
-  fit2.semnig$eb$mpriord), function(d) {
-    fit2.semnig$eb$mpriord[d]*fit2.semnig$eb$mpriorf[[d]]}))),
-  fit1.bSEM$eb$b[nrow(fit1.bSEM$eb$b), ]/
-    (fit1.bSEM$eb$a[nrow(fit1.bSEM$eb$a), ] - 1))
-ranges <- matrix(1, nrow=10, ncol=10)
-tab <- matrix(1, nrow=10, ncol=10)
-tabm <- matrix(1, nrow=10, ncol=10)
-kurts <- matrix(1, nrow=10, ncol=10)
-@
-  The non-zero $\bm{\alpha}$ estimates in Tables 
-  \ref{tab:analysis_gdsc_est1_feat} and \ref{tab:analysis_gdsc_est1_drug} show 
+fit1 <- read.table("results/analysis_gdsc_fit1.txt", row.names=NULL)
+temp <- fit1[, 1]
+fit1 <- as.matrix(fit1[, -1])
+rownames(fit1) <- temp
+
+res1 <- read.table("results/analysis_gdsc_res1.txt", row.names=NULL)
+temp <- res1[, 1]
+res1 <- as.matrix(res1[, -1])
+rownames(res1) <- temp
+
+pmse1 <- res1[substr(rownames(res1), 1, 5)=="pmse.", ]
+pmse1 <- aggregate(pmse1, by=list(rep(1:10, each=251)), FUN="mean")[, -1]
+mpmse1 <- apply(pmse1, 2, median)
+
+fit2 <- read.table("results/analysis_gdsc_fit2.txt", row.names=NULL)
+temp <- fit2[, 1]
+fit2 <- as.matrix(fit2[, -1])
+rownames(fit2) <- temp
+
+res2 <- read.table("results/analysis_gdsc_res2.txt", row.names=NULL)
+temp <- res2[, 1]
+res2 <- as.matrix(res2[, -1])
+rownames(res2) <- temp
+
+pmse2 <- res2[substr(rownames(res2), 1, 5)=="pmse.", ]
+pmse2 <- aggregate(pmse2, by=list(rep(1:10, each=251)), FUN="mean")[, -1]
+mpmse2 <- apply(pmse2, 2, median)
+
+tab1 <- t(fit1)[c(1, 2), c(1:3)]
+rownames(tab1) <- c("NIG$_{\\text{f}+\\text{d}}^-$", 
+                    "NIG$_{\\text{f}+\\text{d}}$")
+colnames(tab1) <- c("feature intercept", "$p$-value", "drug intercept")
+
+tab2 <- rbind(c(mpmse1[1:6], NA, mpmse1[7:length(mpmse1)], rep(NA, 3)),
+              mpmse2)[, 1:7]
+rownames(tab2) <- paste0("model ", 1:nrow(tab2))
+colnames(tab2) <- c("NIG$_{\\text{f}+\\text{d}}^-$", 
+                    "NIG$_{\\text{f}+\\text{d}}$", "ridge", "lasso", 
+                    "xtune", "EB ridge", "bSEM")
+tab2 <- round(tab2, 3)
+tab2 <- t(apply(tab2, 1, function(r) {
+  id <- which(r==min(r, na.rm=TRUE))
+  r[id] <- paste0("\\textbf{", r[which(r==min(r, na.rm=TRUE))], "}")
+  r}))
+@
+
+  The non-zero $\bm{\alpha}$ estimates in Tables ? and ? show 
   that there is an effect of the
@@
   (up to a multiplicative error variance) as 
-  $\hat{E}(\gamma_{jd}^2 \cdot \tau_d^2) \in \[ \Sexpr{ranges[1, ]}\]$ and
-  $\hat{E}(\gamma_{jd}^2 \cdot \tau_d^2) \in \[ \Sexpr{ranges[2, ]}\]$ for
+  $\hat{E}(\gamma_{jd}^2 \cdot \tau_d^2) \in \[ ... \]$ and
+  $\hat{E}(\gamma_{jd}^2 \cdot \tau_d^2) \in \[ ... \]$ for
   NIG$_{\text{f}+\text{d}}$ and bSEM, respectively. This confirms that the
@@
   the external covariates have a detectable effect on the prior.
-<<analysis_gdsc_est1_feat>>=
-options(knitr.kable.NA="", digits=3)
-kable(est1,
-      col.names=colnames(est1), align="r",
-      caption="$\\bm{\\alpha}_{\\text{feat}}$ estimates from GDSC data.",
-      format="latex", booktabs=TRUE, escape=FALSE) %>%
-  kable_styling(latex_options=c("HOLD_position"))
-@
-
-<<analysis_gdsc_est1_drug>>=
-kable(est2,
-      col.names=colnames(est2), align="r",
-      caption="$\\bm{\\alpha}_{\\text{drug}}$ estimates from GDSC data.",
-      format="latex", booktabs=TRUE, escape=FALSE) %>%
-  kable_styling(latex_options=c("HOLD_position"))
-@
-  The median PMSEs and ELBOs, calculated on the test data
-  are displayed in Table \ref{tab:analysis_gdsc_tab1}. In terms of prediction 
+
+<<analysis_gdsc_fit1>>=
+kableExtra::kable_styling(knitr::kable(
+  tab1, align="r", digits=2,
+  caption="$\\bm{\\alpha}$ estimates from model 1. Empty cells correspond to fixed zero parameters",
+  format="latex", booktabs=TRUE, escape=FALSE),
+  latex_options=c("HOLD_position"))
+@
+
+  The median prediction mean squared error (PMSE), calculated on the 
+  test data is displayed in Table \ref{tab:analysis_gdsc_pmse1}. 
+  Note that due to standardisation, the empty reference model has a PMSE of one.
+  In terms of prediction 
   error, ridge outperforms the other models 
-  (PMSE of \Sexpr{round(tab[6, 1], 3)}), while bSEM 
-  (PMSE of \Sexpr{round(tab[7, 1], 3)}) performs worse than 
-  the null model (PMSE of \Sexpr{round(tab[8, 1], 3)}). Second and third best
+  (PMSE of ? ), while bSEM 
+  (PMSE of ? ) performs worse than 
+  the null model (PMSE of ? ). Second and third best
   performing models are NIG$_{\text{f}+\text{d}}^-$ and 
-  NIG$_{\text{f}+\text{d}}$ (PMSEs of \Sexpr{round(tab[1, 1], 3)} and
-  \Sexpr{round(tab[2, 1], 3)}, respectively), both outperforming the lasso
-  with a PMSE of \Sexpr{round(tab[5, 1], 3)}. The underperformance of the 
+  NIG$_{\text{f}+\text{d}}$ (PMSEs of ? and
+  ?, respectively), both outperforming the lasso
+  with a PMSE of ?. The underperformance of the 
   NIG$_{\text{f}}$ and NIG$_{\text{d}}$ models shows the usefulness of a
@@
   as in models NIG$_{\text{f}}$ and NIG$_{\text{d}}$.
-<<analysis_gdsc_tab1>>=
-kable(tabm,
-      col.names=colnames(tabm), align="r",
-      caption="Performance measures calculated on GDSC data (best performing model in bold).",
-      format="latex", booktabs=TRUE, escape=FALSE) %>%
-  kable_styling(latex_options=c("HOLD_position"))
-@
-
-  \subsection{Discussion}
+<<analysis_gdsc_pmse1>>=  
+kableExtra::kable_styling(knitr::kable(
+  tab2, align="r",
+  caption="Cross-validated PMSE for GDSC data (best performing model in bold).",
+  format="latex", booktabs=TRUE, escape=FALSE), 
+  latex_options=c("HOLD_position"))
+@
+
   Part of the differences in performance between
@@
   $\bm{\beta}_d$ as a proxy for sparsity. A common measure of heavy-tailedness
-  is excess kurtosis, calculated as \Sexpr{round(kurts[1], 3)}, 
-  \Sexpr{round(kurts[2], 3)}, and \Sexpr{round(kurts[3], 3)} for the 
+  is excess kurtosis, calculated as ?, 
+  ?, and ? for the 
   NIG$_{\text{f}+\text{d}}$, 

### 2020-04-06 code/analysis_gdsc.R

@@ -346,6 +346,7 @@ fit.bSEM1 <- bSEM(x, y, lapply(C, function(s) {s[, -1] + 1}),
 # saving fitted model objects
 save(fit.semnig1, fit.semnig2, fit.lasso1, fit.ridge1, fit.xtune1, fit.ebridge1, 
      fit.bSEM1, file="results/analysis_gdsc_fit2.Rdata")
+load("results/analysis_gdsc_fit2.Rdata")
 
 est <- cbind(c(fit.semnig1$eb$alphaf, NA, fit.semnig1$eb$alphad, 
                fit.semnig1$eb$lambdaf, fit.semnig1$eb$lambdad),
@@ -354,10 +355,16 @@ est <- cbind(c(fit.semnig1$eb$alphaf, NA, fit.semnig1$eb$alphad,
              rep(NA, 5), rep(NA, 5),
              c(colMeans(-t(sapply(fit.xtune1, "[[", "alpha.est"))), 
                rep(NA, 3)),
-             c(NA, fit.ebridge1$alphaf, NA, fit.ebridge1$alphad, rep(NA, 2)))
+             c(NA, fit.ebridge1$alphaf, NA, fit.ebridge1$alphad, rep(NA, 2)),
+             c(fit.bSEM1$eb$a[nrow(fit.bSEM1$eb$a), 1]/
+                 fit.bSEM1$eb$b[nrow(fit.bSEM1$eb$b), 1],
+               fit.bSEM1$eb$a[nrow(fit.bSEM1$eb$a), 2]/
+                 fit.bSEM1$eb$b[nrow(fit.bSEM1$eb$b), 2] -
+                 fit.bSEM1$eb$a[nrow(fit.bSEM1$eb$a), 1]/
+                 fit.bSEM1$eb$b[nrow(fit.bSEM1$eb$b), 1], rep(NA, 3)))
 rownames(est) <- c("interceptf", "mutation", "interceptd", "lambdaf", "lambdad")
 colnames(est) <- c("semnig1", "semnig2", "ridge1", "lasso1", "xtune1", 
-                   "ebridge1")
+                   "ebridge1", "bSEM1")
 write.table(est, file="results/analysis_gdsc_fit2.txt")
 
 # predictions and mse
@@ -759,333 +766,99 @@ resp.sel <- sapply(resp, function(s) {
   s <- s[order(names(s))]})
 resp.sel <- resp.sel[names(resp.sel) %in% drug$name]
[…]

### 2020-04-06 code/data_gdsc.R

@@ -375,15 +375,17 @@ expr <- expr.temp
 
 # determine for every feature whether it is in the target pathway or not
 in_pathway <- lapply(1:nrow(drug.prep), function(d) {
-  s <- colnames(expr) %in% strsplit(drug.prep$pathway_ensembl_gene_id[[d]], 
-                                    ", ")[[1]]
+  s <- as.numeric(colnames(expr) %in% strsplit(
+    drug.prep$pathway_ensembl_gene_id[[d]], ", ")[[1]])
   names(s) <- colnames(expr)
   s})
+names(in_pathway) <- drug.prep$name
 is_target <- lapply(1:nrow(drug.prep), function(d) {
-  s <- colnames(expr) %in% strsplit(drug.prep$target_ensembl_gene_id[[d]], 
-                                    ", ")[[1]]
+  s <- as.numeric(colnames(expr) %in% strsplit(
+    drug.prep$target_ensembl_gene_id[[d]], ", ")[[1]])
   names(s) <- colnames(expr)
   s})
+names(is_target) <- drug.prep$name
 
 # Cancer Gene Census (https://cancer.sanger.ac.uk/census)
 census <- read.table("data/Census_allFri Feb 21 14_49_10 2020.csv", header=TRUE,

### 2020-04-06 code/figures.R

@@ -160,6 +160,40 @@ points(c(1:4), chi, pch=2, col=col[c(1:4)], cex=1.5,
        cex.lab=1.5, cex.axis=1.5)
 par(opar)
 
+# ---- simulations_gdsc_var ----
+suppressWarnings(suppressMessages(library(sp)))
+res <- read.table("results/simulations_gdsc_res3.txt", row.names=NULL)
+temp <- res[, 1]
+res <- as.matrix(res[, -1])
+rownames(res) <- temp
+
+alphaf <- c(1, 1, 3, 7)
+alphad <- c(1, 1, 3, 7)
+est.phi <- 1/(sapply(0:3, function(s) {
+  res[rownames(res)==paste0("alphaf", s), 2]}) %*% 
+    t(cbind(1, rbind(0, diag(3)))))
+est.chi <- 1/(sapply(0:3, function(s) {
+  res[rownames(res)==paste0("alphad", s), 2]}) %*% 
+    t(cbind(1, rbind(0, diag(3)))))
+x <- as.numeric(outer(1/(alphaf %*% t(cbind(1, rbind(0, diag(3))))),
+                      1/(alphad %*% t(cbind(1, rbind(0, diag(3)))))))
+y <- apply(sapply(1:nrow(est1), function(s) {
+  as.numeric(outer(est.phi[s, ], est.chi[s, ]))}), 1, median)
+
+col <- bpy.colors(1, cutoff.tail=0.5)
+lty <- 3
+pch <- 1
+
+opar <- par(no.readonly=TRUE)
+par(mar=opar$mar*c(1, 1.3, 1, 1))
+plot(x, y, xlab=expression(V(beta)), 
+     ylab=expression(hat(V)(beta)),
+     xlim=range(c(x, y)), ylim=range(c(x, y)), col=col[1], pch=pch[1],
+     cex=1.5, cex.lab=1.5, cex.axis=1.5, lwd=2)
+abline(a=0, b=1, col=col[1], lty=lty[1], cex=1.5, cex.lab=1.5, cex.axis=1.5,
+       lwd=2)
+par(opar)
+
 # ---- simulations_gdsc_est4 ----
 suppressWarnings(suppressMessages(library(sp)))
 res <- read.table("results/simulations_gdsc_res4.txt", row.names=NULL)
@@ -192,7 +226,9 @@ chi <- c(list(apply(1/(sapply(0:3, function(s) {
[…]

### 2020-04-07 docs/manuscript1.Rnw
commit before pull

@@
 	data that is presented in \cite{iorio_landscape_2016}.
-	We average repeated measures over cell line-drug combinations and model the
+	We averaged repeated measures over cell line-drug combinations and model the
 	logarithm of the IC50 values. In the following, IC50 refers 
-	to the log-transformed values. 
-	After iteratively removing the cell line or drug with the highest proportion 
-	of missings until no more missing values remain, we end up with IC50 
-	estimates for between 388 and 1043 cell lines on 251 drugs. The number
-	of cell lines differs between drugs, because not all drug and cell line 
-	combinations are available.
+	to these log-transformed values. 
+	After removing all cell lines with missing values, we end up with 388 to 1043
+	IC50 estimates for for 251 drugs. Differences in the number of cell lines
+	between drugs occur, because not all drug and cell line combinations are 
+	available. The pre-processed expression and mutation data consist of
+	17737 and 300 genes, respectively.
 
@@
 	As an alternative to direct inclusion of the mutation data, we use $p$-values
-	from the mutations, obtained from a $t$-test comparing IC50 values of 
-	mutated cell lines to unmutated genes. We expect that lower mutation 
-	$p$-values result in a larger prior $\beta_{jd}$ variances.
+	from the mutations as external covariate. These were obtained from a $t$-test 
+	comparing IC50 values of mutated and unmutated genes. We expect that lower 
+	mutation $p$-values result in a larger prior $\beta_{jd}$ variances.
 
@@
 <<analysis_gdsc>>==
-fit1 <- read.table("results/analysis_gdsc_fit1.txt", row.names=NULL)
-temp <- fit1[, 1]
-fit1 <- as.matrix(fit1[, -1])
-rownames(fit1) <- temp
-
-res1 <- read.table("results/analysis_gdsc_res1.txt", row.names=NULL)
-temp <- res1[, 1]
-res1 <- as.matrix(res1[, -1])
-rownames(res1) <- temp
-
-pmse1 <- res1[substr(rownames(res1), 1, 5)=="pmse.", ]
-pmse1 <- aggregate(pmse1, by=list(rep(1:10, each=251)), FUN="mean")[, -1]
-mpmse1 <- apply(pmse1, 2, median)
-
-fit2 <- read.table("results/analysis_gdsc_fit2.txt", row.names=NULL)
-temp <- fit2[, 1]
-fit2 <- as.matrix(fit2[, -1])
-rownames(fit2) <- temp
-
-res2 <- read.table("results/analysis_gdsc_res2.txt", row.names=NULL)
-temp <- res2[, 1]
-res2 <- as.matrix(res2[, -1])
-rownames(res2) <- temp
-
-pmse2 <- res2[substr(rownames(res2), 1, 5)=="pmse.", ]
-pmse2 <- aggregate(pmse2, by=list(rep(1:10, each=251)), FUN="mean")[, -1]
-mpmse2 <- apply(pmse2, 2, median)
-
-tab1 <- t(fit1)[c(1, 2), c(1:3)]
+# estimates tables
+fit <- read.table("results/analysis_gdsc_fit1.txt", row.names=NULL)
+temp <- fit[, 1]
+fit <- as.matrix(fit[, -1])
+rownames(fit) <- temp
+
+tab1 <- t(fit)[c(1, 2), c(1:3)]
 rownames(tab1) <- c("NIG$_{\\text{f}+\\text{d}}^-$", 
@@
 
-tab2 <- rbind(c(mpmse1[1:6], NA, mpmse1[7:length(mpmse1)], rep(NA, 3)),
-              mpmse2)[, 1:7]
-rownames(tab2) <- paste0("model ", 1:nrow(tab2))
-colnames(tab2) <- c("NIG$_{\\text{f}+\\text{d}}^-$", 
+fit <- read.table("results/analysis_gdsc_fit2.txt", row.names=NULL)
+temp <- fit[, 1]
+fit <- as.matrix(fit[, -1])
+rownames(fit) <- temp
+
+tab2 <- t(fit)[c(1, 2), c(1:3)]
+rownames(tab2) <- c("NIG$_{\\text{f}+\\text{d}}^-$", 
+                    "NIG$_{\\text{f}+\\text{d}}$")
+colnames(tab2) <- c("feature intercept", "mutation", "drug intercept")
+
+fit <- read.table("results/analysis_gdsc_fit3.txt", row.names=NULL)
+temp <- fit[, 1]
+fit <- as.matrix(fit[, -1])
+rownames(fit) <- temp
+
+tab3 <- t(fit)[c(1, 2), c(1:3)]
+rownames(tab3) <- c("NIG$_{\\text{f}+\\text{d}}^-$", 
+                    "NIG$_{\\text{f}+\\text{d}}$")
+colnames(tab3) <- c("feature intercept", "$p$-value", "drug intercept")
+
+fit <- read.table("results/analysis_gdsc_fit4.txt", row.names=NULL)
+temp <- fit[, 1]
+fit <- as.matrix(fit[, -1])
+rownames(fit) <- temp
+
+tab4.1 <- t(fit)[c(1, 2), c(1:2)]
+rownames(tab4.1) <- c("NIG$_{\\text{f}+\\text{d}}^-$", 
+                    "NIG$_{\\text{f}+\\text{d}}$")
+colnames(tab4.1) <- c("feature intercept", "pathway")
+tab4.2 <- t(fit)[c(1, 2), c(3:7)]
+rownames(tab4.2) <- c("NIG$_{\\text{f}+\\text{d}}^-$", 
+                    "NIG$_{\\text{f}+\\text{d}}$")
+colnames(tab4.2) <- c("drug intercept", 
+                    "experimental", "development", "targeted", "unknown")
+
+# pmse table
+res <- read.table("results/analysis_gdsc_res1.txt", row.names=NULL)
+temp <- res[, 1]
+res <- as.matrix(res[, -1])
+rownames(res) <- temp
+
+pmse1 <- res[substr(rownames(res), 1, 5)=="pmse.", ]
+pmse1 <- aggregate(pmse1, by=list(rep(1:10, each=251)), FUN="mean")[, -1]
+mpmse1 <- apply(pmse1, 2, median)
+
+res <- read.table("results/analysis_gdsc_res2.txt", row.names=NULL)
+temp <- res[, 1]
+res <- as.matrix(res[, -1])
+rownames(res) <- temp
+
+pmse2 <- res[substr(rownames(res), 1, 5)=="pmse.", ]
+pmse2 <- aggregate(pmse2, by=list(rep(1:10, each=251)), FUN="mean")[, -1]
+mpmse2 <- apply(pmse2, 2, median)
+
+# res <- read.table("results/analysis_gdsc_res3.txt", row.names=NULL)
+# temp <- res[, 1]
+# res <- as.matrix(res[, -1])
+# rownames(res) <- temp
+res <- matrix(1, nrow=2510, ncol=13); rownames(res) <- paste0("pmse.", 1:nrow(res))
+
+pmse3 <- res[substr(rownames(res), 1, 5)=="pmse.", ]
+pmse3 <- aggregate(pmse3, by=list(rep(1:10, each=251)), FUN="mean")[, -1]
+mpmse3 <- apply(pmse3, 2, median)
+
+# res <- read.table("results/analysis_gdsc_res4.txt", row.names=NULL)
+# temp <- res[, 1]
+# res <- as.matrix(res[, -1])
+# rownames(res) <- temp
+res <- matrix(1, nrow=2510, ncol=17); rownames(res) <- paste0("pmse.", 1:nrow(res))
+
+pmse4 <- res[substr(rownames(res), 1, 5)=="pmse.", ]
+pmse4 <- aggregate(pmse4, by=list(rep(1:10, each=251)), FUN="mean")[, -1]
+mpmse4 <- apply(pmse4, 2, median)
+
+tab5 <- rbind(c(mpmse1[1:6], NA, mpmse1[7:length(mpmse1)], rep(NA, 3)),
+              mpmse2, c(mpmse3[1:6], NA, mpmse3[7:length(mpmse3)], rep(NA, 3)),
+              mpmse4)[, 1:7]
+rownames(tab5) <- paste0("analysis ", 1:nrow(tab5))
+colnames(tab5) <- c("NIG$_{\\text{f}+\\text{d}}^-$", 
                     "NIG$_{\\text{f}+\\text{d}}$", "ridge", "lasso", 
                     "xtune", "EB ridge", "bSEM")
-tab2 <- round(tab2, 3)
-tab2 <- t(apply(tab2, 1, function(r) {
+tab5 <- round(tab5, 3)
+tab5 <- t(apply(tab5, 1, function(r) {
   id <- which(r==min(r, na.rm=TRUE))
@@
 @
-
   The non-zero $\bm{\alpha}$ estimates in Tables ? and ? show 
@@
   tab1, align="r", digits=2,
-  caption="$\\bm{\\alpha}$ estimates from model 1. Empty cells correspond to fixed zero parameters",
+  caption="$\\bm{\\alpha}$ estimates from analysis 1. Empty cells correspond to fixed zero parameters.",
   format="latex", booktabs=TRUE, escape=FALSE),
@@
 @
+
+<<analysis_gdsc_fit2>>=
+kableExtra::kable_styling(knitr::kable(
+  tab2, align="r", digits=2,
+  caption="$\\bm{\\alpha}$ estimates from analysis 2. Empty cells correspond to fixed zero parameters.",
+  format="latex", booktabs=TRUE, escape=FALSE),
+  latex_options=c("HOLD_position"))
+@
+
+<<analysis_gdsc_fit3>>=
+kableExtra::kable_styling(knitr::kable(
+  tab3, align="r", digits=2,
+  caption="$\\bm{\\alpha}$ estimates from analysis 3. Empty cells correspond to fixed zero parameters.",
+  format="latex", booktabs=TRUE, escape=FALSE),
+  latex_options=c("HOLD_position"))
+@
+
+  \begin{table}[h!]\label{tab:analysis_gdsc_fit4}
+    \begin{minipage}{1\textwidth}
+<<analysis_gdsc_fit4.1>>=
+kableExtra::kable_styling(knitr::kable(
+  tab4.1, align="r", digits=2,
+  caption="$\\bm{\\alpha}$ estimates from analysis 4. Empty cells correspond to fixed zero parameters.",
+  format="latex", booktabs=TRUE, escape=FALSE),
+  latex_options=c("HOLD_position"))
+@
+    \end{minipage}
+    \begin{minipage}{1\textwidth}
+<<analysis_gdsc_fit4.2>>=
+kableExtra::kable_styling(knitr::kable(
+  tab4.2, align="r", digits=2,
+  caption="$\\bm{\\alpha}$ estimates from analysis 4. Empty cells correspond to fixed zero parameters \\textit{(continued)}.",
+  format="latex", booktabs=TRUE, escape=FALSE),
+  latex_options=c("HOLD_position"))
+@
+    \end{minipage}
+  \end{table}
 
@@
 kableExtra::kable_styling(knitr::kable(
-  tab2, align="r",
-  caption="Cross-validated PMSE for GDSC data (best performing model in bold).",
+  tab5, align="r",
+  caption="Cross-validated PMSE for GDSC data (best performing model in bold). In analyses 1 and 3, bSEM could not be estimated due to non-binary external covariates.",
   format="latex", booktabs=TRUE, escape=FALSE), 

### 2020-04-07 code/analysis_gdsc.R

@@ -824,7 +824,7 @@ fit.xtune1 <- lapply(1:D, function(d) {
         control=list(intercept=FALSE))})
 
 # fit ebridge model
-fit.ebridge1 <- ebridge(x, y, lapply(C, function(s) {matrix(s[, -1])}), NULL,
+fit.ebridge1 <- ebridge(x, y, lapply(C, function(s) {matrix(s[, -1])}), Z[, -1],
                         foldid=foldid,
                         hyper=list(lambda=1/sqrt(n*sapply(
                           fit.ridge1, "[[", "lambda.min")), zeta=0, nu=0),
@@ -838,24 +838,179 @@ fit.bSEM1 <- bSEM(x, y, lapply(C, function(s) {s[, -1] + 1}),
 save(fit.semnig1, fit.semnig2, fit.lasso1, fit.ridge1, fit.xtune1, fit.ebridge1, 
      fit.bSEM1, file="results/analysis_gdsc_fit4.Rdata")
 
-est <- cbind(c(fit.semnig1$eb$alphaf, NA, fit.semnig1$eb$alphad, 
+est <- cbind(c(fit.semnig1$eb$alphaf, NA, fit.semnig1$eb$alphad, rep(NA, 4),
                fit.semnig1$eb$lambdaf, fit.semnig1$eb$lambdad),
              c(fit.semnig2$eb$alphaf, fit.semnig2$eb$alphad, 
                fit.semnig2$eb$lambdaf, fit.semnig2$eb$lambdad),
-             rep(NA, 5), rep(NA, 5),
+             rep(NA, 9), rep(NA, 9),
              c(colMeans(-t(sapply(fit.xtune1, "[[", "alpha.est"))), 
-               rep(NA, 3)),
+               rep(NA, 7)),
              c(NA, fit.ebridge1$alphaf, NA, fit.ebridge1$alphad, rep(NA, 2)),
              c(fit.bSEM1$eb$a[nrow(fit.bSEM1$eb$a), 1]/
                  fit.bSEM1$eb$b[nrow(fit.bSEM1$eb$b), 1],
                fit.bSEM1$eb$a[nrow(fit.bSEM1$eb$a), 2]/
                  fit.bSEM1$eb$b[nrow(fit.bSEM1$eb$b), 2] -
                  fit.bSEM1$eb$a[nrow(fit.bSEM1$eb$a), 1]/
-                 fit.bSEM1$eb$b[nrow(fit.bSEM1$eb$b), 1], rep(NA, 3)))
[…]

### 2020-04-07 docs/manuscript1.Rnw
fixed mistake in analysis 4

@@
 	\end{bmatrix}$,
-	and the expectation is with respect to the joint posterior. In our case, this 
+	and the expectation is with respect to the joint posterior. 
+	In our case, this 
 	posterior is not available in closed form, which renders the expectation 
@@
 	where now the expectation is with respect to the converged variational 
-	posterior $Q^{(l)}=\prod_{d=1}^D Q_d^{(l)}$.
+	posterior $Q^{(l)}=\prod_{d=1}^D Q_d^{(l)}$. Note that the prior 
+	$\bgamma^2_d$ and $\tau_d^2$ independence assumption 
+	results in separate maximisation problems for the 
+	feature hyperparameters ($\bm{\alpha}_{\text{feat}}$ and 
+	$\bm{\lambda}_{\text{feat}}$), and the drug hyperparameters 
+	($\bm{\alpha}_{\text{drug}}$ and $\bm{\lambda}_{\text{drug}}$) at each
+	iteration of the EM algorithm.
 
@@
 
-	\subsection{Analyses}
+	\subsection{Analyses}\label{sec:analyses}
 	Four analyses were conducted: 
@@
 	  \item{Analysis 1 includes gene expressions as predictors and $p$-values from
-	    the gene mutation data as external covariate. A pre-processing step 
-	    selects between 221 and 280 genes per drug for which a mutation $p$-value
-	    is available.}
+	    the gene mutation data as external covariate ($G=1$, $H=0$). 
+	    A pre-processing step selects between 221 and 280 genes per drug for 
+	    which a mutation $p$-value is available.}
 	  \item{Analysis 2 includes both gene  expressions and mutations as predictors
 	    with the feature type as external covariate, i.e., wether the feature
-	    is a expression or mutation. For this analysis, we pre-select 300 gene 
+	    is a expression or mutation ($G=1$, $H=0$). For this analysis,
+	    we pre-select 300 gene 
 	    expressions with maximum 
@@
 	    maximum varaince. The CCLE $p$-values are included as external 
-	    covariates.}
+	    covariates ($G=1$, $H=0$).}
 	  \item{Analysis 4 includes the gene expressions as features and both the
 	    annoted drug variables and pathway status of the genes as external 
-	    covariates. We pre-select 500 genes expressions based on maximum 
-	    variance.}
+	    covariates ($G=1$, $H=2$, before dummy coding). 
+	    We pre-select 500 genes expressions based on maximum variance.}
 	\end{itemize}
@@
 
-# res <- read.table("results/analysis_gdsc_res3.txt", row.names=NULL)
-# temp <- res[, 1]
-# res <- as.matrix(res[, -1])
-# rownames(res) <- temp
-res <- matrix(1, nrow=2510, ncol=13); rownames(res) <- paste0("pmse.", 1:nrow(res))
+res <- read.table("results/analysis_gdsc_res3.txt", row.names=NULL)
+temp <- res[, 1]
+res <- as.matrix(res[, -1])
+rownames(res) <- temp
 

### 2020-04-07 code/analysis_gdsc.R

@@ -863,11 +863,11 @@ write.table(est, file="results/analysis_gdsc_fit4.txt")
 ### predictions and mse
 # settings
 psel.dss <- 10
-control.semnig <- list(conv.post=FALSE, trace=FALSE, epsilon.eb=1e-3, 
-                       epsilon.vb=1e-3, maxit.eb=500, maxit.vb=1, 
+control.semnig <- list(conv.post=FALSE, trace=TRUE, epsilon.eb=1e-3, 
+                       epsilon.vb=1e-3, maxit.eb=1, maxit.vb=1, 
                        maxit.post=100, maxit.block=0)
-control.ebridge <-list(epsilon=sqrt(.Machine$double.eps), maxit=500, 
-                       trace=FALSE, glmnet.fit2=FALSE, beta2=TRUE)
+control.ebridge <-list(epsilon=sqrt(.Machine$double.eps), maxit=1, 
+                       trace=TRUE, glmnet.fit2=FALSE, beta2=TRUE)
 
 set.seed(2020)
 foldid <- lapply(1:D, function(d) {
@@ -912,14 +912,14 @@ res <- foreach(r=1:nfolds, .packages=packages, .errorhandling="pass") %dopar% {
                        full.post=TRUE, control=control.semnig)
   
   # penalized regression models
-  cv.ridge1 <- lapply(1:D, function(d) {
+  cv.ridge1 <- lapply(c(1:D), function(d) {
     cv.glmnet(xtrain[[d]], ytrain[[d]], alpha=0, foldid=foldid2[[d]], 
               intercept=FALSE)})
-  cv.lasso1 <- lapply(1:D, function(d) {
+  cv.lasso1 <- lapply(c(1:D), function(d) {
     cv.glmnet(xtrain[[d]], ytrain[[d]], foldid=foldid2[[d]], intercept=FALSE)})
   
   # fit xtune
-  cv.xtune1 <- lapply(1:D, function(d) {
+  cv.xtune1 <- lapply(c(1:D), function(d) {
     xtune(xtrain[[d]], ytrain[[d]], C[[d]][, -1], family="linear", 
           method="ridge", message=FALSE, control=list(intercept=FALSE))})
   
@@ -936,14 +936,14 @@ res <- foreach(r=1:nfolds, .packages=packages, .errorhandling="pass") %dopar% {
[…]

### 2020-04-08 docs/manuscript1.Rnw
fixed small mistake analysis 4

@@
       \node[obs] (y) {$y_{id}$}; %
-      \factor[right=of y, yshift=1.5cm, xshift=-0.1cm] {y-f} 
-      {above:$\mathcal{N}$} {} {} ; %
 
@@
       \node[latent, right=of x] (beta) {$\beta_{jd}$}; %
-      \factor[right=of beta, yshift=-1.4cm, xshift=-0.1cm] {beta-f} 
-      {above:$\mathcal{N}$} {} {} ; %
 
@@
       {$\lambda_{\text{feat}}$}; %
-      \factor[right=of gamma2] {gamma2-f} {above:$\mathcal{IG}$} {phi,lambdaf} 
-      {gamma2} ; %
       \node[det, right=of gamma2, xshift=2cm, yshift=0.5cm] (c) {$c_{jdg}$}; %
@@
       {$\lambda_{\text{drug}}$}; %
-      \factor[right=of tau2] {tau2-f} {above:$\mathcal{IG}$} {chi,lambdad} 
-      {tau2} ; %
       \node[det, right=of tau2, xshift=2cm, yshift=0.5cm] (z) {$z_{dh}$}; %
@@
       % edges
-      \factoredge {x,beta,sigma2} {y-f} {y} ; %
-      \factoredge {sigma2,gamma2,tau2} {beta-f} {beta} ; %
+      \edge {x} {y};%
+      \edge {sigma2} {y};%
+      \edge {beta} {y};%
+      \edge {sigma2} {beta};%
+      \edge {gamma2} {beta};%
+      \edge {tau2} {beta};%
+      \edge {phi} {gamma2};%
+      \edge {lambdaf} {gamma2};%
+      \edge {chi} {tau2};%
+      \edge {lambdad} {tau2};%
       \edge {c} {phi}; %
@@
     represent observed variables, white circles represent unobserved variables,
-    squares represent fixed data, and unenclosed letters are parameters to be
-    estimated. Cell lines are indexed by $i$, features
+    tilted squares represent fixed data, and unenclosed letters are parameters 
+    to be estimated. Cell lines are indexed by $i$, features
     by $j$, drugs by $d$, drug covariates by $h$, and feature covariates by 
@@
 	four alternative methods: frequentist (i) lasso and (ii) ridge models, 
-	and two EB versions of the ridge model: (iii) \texttt{xtune} 
-	\cite[]{ni_bayesian_2019} and (iv) its
-	multivariate Bayesian extension as introduced in SM Section 
-	\ref{sm-sec:altprior}.
+	and and EB version of the ridge model: (iii) a
+	multivariate Bayesian extension of \texttt{xtune} 
+	\cite[]{ni_bayesian_2019} named mxtune, as introduced in SM Section 
+	\ref{sm-sec:mxtune}.
 
@@
 true.phi <- mean(1/(c(1, 1, 3, 7) %*% t(unname(model.matrix(~ factor(1:4))))))
-est.phi <- median(1/res[rownames(res)=="alphaf0", 1])
-est.lambdaf <- apply(res[rownames(res)=="lambdaf", c(1, 2)], 2, median)
+est.phi <- mean(1/res[rownames(res)=="alphaf0", 1])
+est.lambdaf <- apply(res[rownames(res)=="lambdaf", c(1, 2)], 2, mean)
 @	
@@
 	Model NIG$_{\text{f}}^-$ (that excludes the external covariates) gives a 
-	median $\phi$ estimate of \Sexpr{round(est.phi, 2)},  
+	mean $\phi$ estimate of \Sexpr{round(est.phi, 2)},  
 	slightly higher than the true mean of the $\phi_{jd}$, 
@@
 true.chi <- mean(1/(c(1, 1, 3, 7) %*% t(unname(model.matrix(~ factor(1:4))))))
-est.chi <- median(1/res[rownames(res)=="alphad0", 1])
-est.lambdad <- apply(res[rownames(res)=="lambdad", c(1, 2)], 2, median)
+est.chi <- mean(1/res[rownames(res)=="alphad0", 1])
+est.lambdad <- apply(res[rownames(res)=="lambdad", c(1, 2)], 2, mean)
 @	
@@
 	results in underestimated $\chi_d$ (Figure \ref{fig:simulations_gdsc_est2}b).
-	The median $\chi$ estimate in the 
+	The mean $\chi$ estimate in the 
 	NIG$_{\text{d}}^-$ model is \Sexpr{round(est.chi, 2)},  
@@
 true.chi <- mean(1/(c(1, 1, 3, 7) %*% t(unname(model.matrix(~ factor(1:4))))))
-est.chi <- median(1/res[rownames(res)=="alphad0", 1])
-est.lambdad <- apply(res[rownames(res)=="lambdad", c(1, 2)], 2, median)
+est.chi <- mean(1/res[rownames(res)=="alphad0", 1])
+est.lambdad <- apply(res[rownames(res)=="lambdad", c(1, 2)], 2, mean)
 true.phi <- mean(1/(c(1, 1, 3, 7) %*% t(unname(model.matrix(~ factor(1:4))))))
-est.phi <- median(1/res[rownames(res)=="alphaf0", 1])
-est.lambdaf <- apply(res[rownames(res)=="lambdaf", c(1, 2)], 2, median)
+est.phi <- mean(1/res[rownames(res)=="alphaf0", 1])
+est.lambdaf <- apply(res[rownames(res)=="lambdaf", c(1, 2)], 2, mean)
 
@@
     t(cbind(1, rbind(0, diag(3)))), 2, function(m) {
-      res[rownames(res)=="alphaf0", 2]/m}), 2, median)
+      res[rownames(res)=="alphaf0", 2]/m}), 2, mean)
 chi.ratio <- apply(apply(sapply(0:3, function(s) {
@@
     t(cbind(1, rbind(0, diag(3)))), 2, function(m) {
-      res[rownames(res)=="alphad0", 2]/m}), 2, median)
+      res[rownames(res)=="alphad0", 2]/m}), 2, mean)
 true.ratio <- as.numeric(1/(c(1, 1, 3, 7) %*%
@@
 	$\bm{\alpha}_{\text{drug}}$ are overestimated. These biases seem to 
-	be consistent though. The median ratios $\phi_1$ to $\phi_2$, $\phi_3$, 
+	be consistent though. The mean ratios $\phi_1$ to $\phi_2$, $\phi_3$, 
 	$\phi_4$ and $\chi_1$ to $\chi_2$, $\chi_3$, 
@@
 	and underestimating $\phi$
-	and $\chi$ with median estimates \Sexpr{round(est.phi, 2)} and 
+	and $\chi$ with mean estimates \Sexpr{round(est.phi, 2)} and 
 	\Sexpr{round(est.chi, 2)}, respectively (compared to the true mean 
@@
 
-  Figure \ref{fig:simulations_gdsc_est4} displays the median $\phi_{jd}$ and 
+  Figure \ref{fig:simulations_gdsc_est4} displays the mean $\phi_{jd}$ and 
   $\chi_d$ estimates for different noise levels in simulation Scenario 4. 
@@
   $\bm{\alpha}_{\text{drug}}$, as expected.
-<<simulations_gdsc_est4, fig.cap="Simulation results for Scenario 3: estimated prior means (a) $\\phi_{jd}$, and (b) $\\chi_{d}$ for different levels of noise in the external covariates.", out.width="100%", fig.asp=1/2>>=
+<<simulations_gdsc_est4, fig.cap="Simulation results for Scenario 3: mean estimated prior means (a) $\\phi_{jd}$, and (b) $\\chi_{d}$ for different levels of noise in the external covariates.", out.width="100%", fig.asp=1/2>>=
 @
@@
 	In Scenario 3, this is compensated for
-	by the overestimation of the $\phi_{jd}$, such that the estimated median 
+	by the overestimation of the $\phi_{jd}$, such that the estimated mean 
 	variance $\V(\beta_{jd})$ are stil biased, but only slightly underestimated
@@
 	in practice.
-<<simulations_gdsc_var, fig.cap="Simulation results for Scenario 3: estimated prior variances $\\V(\\beta_{jd})$ versus true values, with line of identity (dotted).", out.width="50%", fig.asp=1>>= 
+<<simulations_gdsc_var, fig.cap="Simulation results for Scenario 3: mean estimated prior variances $\\V(\\beta_{jd})$ versus true values, with line of identity (dotted).", out.width="50%", fig.asp=1>>=
 @
@@
 	(i) NIG$_{\text{f}+\text{d}}^-$, (ii) NIG$_{\text{f}+\text{d}}$, 
-	frequentist (iii) lasso and (iv) ridge models, (v) \texttt{xtune}, and
-	(vi) its multivariate Bayesian extension (see SM Section 
-	\ref{sm-sec:altprior}), named EBridge. In addition, in Analyses 2 and 
-	4, we fit a (vii) bSEM model 
-	\cite[]{leday_gene_2017,kpogbezan_empirical_2017}. The
-	bSEM model is similar to the NIG model in that it
-	draws the $\beta_{jd}$ from a conditionally normal distribution, but
-	fixes $\tau_d^2=1$ and draws
-	$\gamma_{jd}^2 \sim \Gamma^{-1} (\phi_{jd}, \lambda_{jd})$, instead of
-	$\gamma_{jd}^2 \sim \mathcal{IG} (\phi_{jd}, \lambda_{jd})$, where 
-	$\Gamma^{-1} (\phi, \lambda)$ denotes the inverse gamma distribution with 
-	shape $\phi$ and scale $\lambda$. Additionally, the bSEM model can
-	only incorporate one binary external covariate that divides the features into
-	two groups, i.e., $\bm{\alpha}_{\text{feat}} = \begin{bmatrix} \alpha_0 &
-	\alpha_1
-	\end{bmatrix} \tr$ and $\bm{c}_{jd} = \begin{bmatrix} 1 & 0 \end{bmatrix}
-	\tr$ if feature $jd$ belongs to group one and $\bm{c}_{jd} = \begin{bmatrix}
-	1 & 1 \end{bmatrix} \tr$ if feature $jd$ belongs to group two. Both
-	$\bm{\alpha}_{\text{feat}}$ and $\lambda_{\text{feat}}$ are
-	estimated in a similar fashion as in Section \ref{sec:estimation}.
+	frequentist (iii) lasso and (iv) ridge models, and (v) mxtune 
+	(see SM Section \ref{sm-sec:mxtune}). 
+	% In addition, in Analyses 2 and 
+	% 4, we fit a (vi) bSEM model 
+	% \cite[]{leday_gene_2017,kpogbezan_empirical_2017}. The
+	% bSEM model is similar to the NIG model in that it
+	% draws the $\beta_{jd}$ from a conditionally normal distribution, but
+	% fixes $\tau_d^2=1$ and draws
+	% $\gamma_{jd}^2 \sim \Gamma^{-1} (\phi_{jd}, \lambda_{jd})$, instead of
+	% $\gamma_{jd}^2 \sim \mathcal{IG} (\phi_{jd}, \lambda_{jd})$, where 
+	% $\Gamma^{-1} (\phi, \lambda)$ denotes the inverse gamma distribution with 
+	% shape $\phi$ and scale $\lambda$. Additionally, the bSEM model can
+	% only incorporate one binary external covariate that divides the features into
+	% two groups, i.e., $\bm{\alpha}_{\text{feat}} = \begin{bmatrix} \alpha_0 &
+	% \alpha_1
+	% \end{bmatrix} \tr$ and $\bm{c}_{jd} = \begin{bmatrix} 1 & 0 \end{bmatrix}
+	% \tr$ if feature $jd$ belongs to group one and $\bm{c}_{jd} = \begin{bmatrix}
+	% 1 & 1 \end{bmatrix} \tr$ if feature $jd$ belongs to group two. Both
+	% $\bm{\alpha}_{\text{feat}}$ and $\lambda_{\text{feat}}$ are
+	% estimated in a similar fashion as in Section \ref{sec:estimation}.
 
 	In all analyses we use all cell lines to estimate the hyperparameters 
-	presented in Section \ref{sec:results}. Prediction PMSE,
+	presented in Section \ref{sec:results}. Mean prediction PMSE,
 	$\text{PMSE} = D^{-1} n^{-1} \sum_{d=1}^D 
-  \sum_{i=1}^n (\mathbf{y}_d - \mathbf{x}_i \tr \hat{\bm{\beta}_d})^2$, is
-  estimated by 10-fold cross validation. In the NIG and bSEM models that provide
-  the full posteriors, the posterior mean 
+  \sum_{i=1}^n (\mathbf{y}_d - \mathbf{x}_i \tr \hat{\bm{\beta}_d})^2$, with
+  $\hat{\bm{\beta}_d}$ the estimator for $\bm{\beta}_d$, and its standard error
+  are estimated by 10-fold cross validation. In the NIG models that 
+  provide full posteriors, the posterior mean 
   $\E(\bm{\beta}_d | \mathbf{y}_d)$ is used point estimate.
@@
 pmse1 <- aggregate(pmse1, by=list(rep(1:10, each=251)), FUN="mean")[, -1]
-mpmse1 <- apply(pmse1, 2, median)
+mpmse1 <- apply(pmse1, 2, mean)
+sdpmse1 <- apply(pmse1, 2, sd)
+ppmse1 <- paste0(round(mpmse1, 3), " (", round(sdpmse1, 3), ")")
 
@@
 pmse2 <- aggregate(pmse2, by=list(rep(1:10, each=251)), FUN="mean")[, -1]
-mpmse2 <- apply(pmse2, 2, median)
+mpmse2 <- apply(pmse2, 2, mean)
+sdpmse2 <- apply(pmse2, 2, sd)
+ppmse2 <- paste0(round(mpmse2, 3), " (", round(sdpmse2, 3), ")")
 
@@
 pmse3 <- aggregate(pmse3, by=list(rep(1:10, each=251)), FUN="mean")[, -1]
-mpmse3 <- apply(pmse3, 2, median)
-
-# res <- read.table("results/analysis_gdsc_res4.txt", row.names=NULL)
-# temp <- res[, 1]
-# res <- as.matrix(res[, -1])
-# rownames(res) <- temp
-res <- matrix(1, nrow=2510, ncol=17); rownames(res) <- paste0("pmse.", 1:nrow(res))
+mpmse3 <- apply(pmse3, 2, mean)
+sdpmse3 <- apply(pmse3, 2, sd)
+ppmse3 <- paste0(round(mpmse3, 3), " (", round(sdpmse3, 3), ")")
+
+res <- read.table("results/analysis_gdsc_res4.txt", row.names=NULL)
+temp <- res[, 1]
+res <- as.matrix(res[, -1])
+rownames(res) <- temp
 
@@
 pmse4 <- aggregate(pmse4, by=list(rep(1:10, each=251)), FUN="mean")[, -1]
-mpmse4 <- apply(pmse4, 2, median)
-
-tab5 <- rbind(c(mpmse1[1:6], NA, mpmse1[7:length(mpmse1)], rep(NA, 3)),
+mpmse4 <- apply(pmse4, 2, mean)
+sdpmse4 <- apply(pmse4, 2, sd)
+ppmse4 <- paste0(round(mpmse4, 3), " (", round(sdpmse4, 3), ")")
+
+tab5 <- rbind(c(ppmse1[1:6], NA, ppmse1[7:length(ppmse1)], rep(NA, 3)),
+              ppmse2, c(ppmse3[1:6], NA, ppmse3[7:length(ppmse3)], rep(NA, 3)),
+              ppmse4)[, c(1:4, 6)]
+tabm5 <- rbind(c(mpmse1[1:6], NA, mpmse1[7:length(mpmse1)], rep(NA, 3)),
                mpmse2, c(mpmse3[1:6], NA, mpmse3[7:length(mpmse3)], rep(NA, 3)),
-              mpmse4)[, 1:7]
+               mpmse4)[, c(1:4, 6)]
+tab5 <- t(sapply(1:nrow(tab5), function(r) {
+  id <- which(tabm5[r, ]==min(tabm5[r, ], na.rm=TRUE))
+  out <- tab5[r, ]
+  out[id] <- paste0("\\textbf{", out[id], "}")
+  out}))
 rownames(tab5) <- paste0("analysis ", 1:nrow(tab5))
@@
                     "NIG$_{\\text{f}+\\text{d}}$", "ridge", "lasso", 
-                    "xtune", "EB ridge", "bSEM")
-tab5 <- round(tab5, 3)
-tab5 <- t(apply(tab5, 1, function(r) {
-  id <- which(r==min(r, na.rm=TRUE))
-  r[id] <- paste0("\\textbf{", r[which(r==min(r, na.rm=TRUE))], "}")
-  r}))
+                    "mxtune")
 @
@@
   tab5, align="r",
-  caption="Cross-validated PMSE for GDSC data (best performing model in bold). In analyses 1 and 3, bSEM could not be estimated due to non-binary external covariates.",
+  caption="Mean (standard deviation) of cross-validated PMSE for GDSC data (standard error). Best performing model in bold.",
   format="latex", booktabs=TRUE, escape=FALSE), 

### 2020-04-08 code/figures.R

@@ -178,7 +178,7 @@ est.chi <- 1/(sapply(0:3, function(s) {
 x <- as.numeric(outer(1/(alphaf %*% t(cbind(1, rbind(0, diag(3))))),
                       1/(alphad %*% t(cbind(1, rbind(0, diag(3)))))))
 y <- apply(sapply(1:nrow(est1), function(s) {
-  as.numeric(outer(est.phi[s, ], est.chi[s, ]))}), 1, median)
+  as.numeric(outer(est.phi[s, ], est.chi[s, ]))}), 1, mean)
 
 col <- bpy.colors(1, cutoff.tail=0.5)
 lty <- 3
@@ -206,12 +206,12 @@ phi <- lapply(fracs, function(q) {
   cq <- paste0("frac", q, ".alphaf")
   apply(1/(sapply(0:3, function(s) {
     res[rownames(res)==paste0(cq, s), 2]}) %*% 
-      t(cbind(1, rbind(0, diag(3))))), 2, median)})
+      t(cbind(1, rbind(0, diag(3))))), 2, mean)})
 chi <- lapply(fracs, function(q) {
   cq <- paste0("frac", q, ".alphad")
   apply(1/(sapply(0:3, function(s) {
     res[rownames(res)==paste0(cq, s), 2]}) %*% 
-      t(cbind(1, rbind(0, diag(3))))), 2, median)})
+      t(cbind(1, rbind(0, diag(3))))), 2, mean)})
 
 suppressWarnings(suppressMessages(library(sp)))
 res <- read.table("results/simulations_gdsc_res3.txt", row.names=NULL)
@@ -221,10 +221,10 @@ rownames(res) <- temp
 
 phi <- c(list(apply(1/(sapply(0:3, function(s) {
   res[rownames(res)==paste0("alphaf", s), 2]}) %*% 
-    t(cbind(1, rbind(0, diag(3))))), 2, median)), phi)
+    t(cbind(1, rbind(0, diag(3))))), 2, mean)), phi)
 chi <- c(list(apply(1/(sapply(0:3, function(s) {
   res[rownames(res)==paste0("alphad", s), 2]}) %*% 
-    t(cbind(1, rbind(0, diag(3))))), 2, median)), chi)
+    t(cbind(1, rbind(0, diag(3))))), 2, mean)), chi)
 
 lty <- c(1:(1 + length(fracs)))
 col <- bpy.colors(1, cutoff.tail=0.3, 
[…]

### 2020-04-08 docs/supplement1.Rnw

@@
 
-  \section{Alternative prior}\label{sec:altprior}
+  \section{Multivariate Bayesian extension of \texttt{xtune}}\label{sec:mxtune}
   Consider the following simpler hyperprior alternative to 
@@
 	package \texttt{glmnet} \cite[]{friedman_regularization_2010} with cross
-	validated
-	penalty parameters. The \texttt{xtune} model is fit with the corresponding 
-	\texttt{R} package \cite[]{ni_bayesian_2019}.
+	validated penalty parameters. The \texttt{xtune} model is fit with the 
+	corresponding \texttt{R} package \cite[]{ni_bayesian_2019}.
 
@@
 @
+
+<<simulations_gdsc_post5, fig.cap="Simulation results for Scenario 3: Samples from posterior with variational posterior superimposed (red line) for estimated hyperparameters (top row) $\\hat{\\chi}_1$, (second row) $\\hat{\\chi}_2$, (third row) $\\hat{\\chi}_3$, (bottom row) $\\hat{\\chi}_4$, (left column) $\\hat{\\phi}_1$, (second column) $\\hat{\\phi}_2$, (third column) $\\hat{\\phi}_3$, and (last column) $\\hat{\\phi}_4$.", out.width="100%", fig.asp=1>>=
+@
+
+<<simulations_gdsc_pmse4, fig.cap="Simulation results for Scenario 4: Median PMSE against the proportion of permuted rows in the external features.", out.width="50%", fig.asp=1>>= 
+@
+
   \bibliographystyle{author_short3}  

### 2020-04-09 docs/manuscript1.Rnw
commit before reverting

@@
 
-res <- read.table("results/analysis_gdsc_res4.txt", row.names=NULL)
+res <- read.table("results/analysis_gdsc_res4.1.txt", row.names=NULL)
 temp <- res[, 1]
@@
 	sense, the comparison metric (PMSE) is not to the benefit of the NIG model.
+	We note that direct prediction error optimisation is not feasible in our
+	external covariates setting.
 

### 2020-04-09 code/simulations_gdsc.R

@@ -743,7 +743,7 @@ sigma <- 1/sqrt(rgamma(D, shape, rate))
 beta <- sapply(1:D, function(d) {rnorm(p, 0, tau[d]*gamma[, d]*sigma[d])})
 
 # simulate data
-id <- sample(1:nrow(x), n)
+id <- sample(1:nrow(expr$expr), n)
 x <- scale(expr$expr[, order(-apply(expr$expr, 2, sd))[1:p]])[id, ]
 y <- lapply(1:D, function(d) {
   s <- as.numeric(scale(rnorm(n, x %*% beta[, d], sigma[d])))
@@ -767,6 +767,23 @@ fit.mcmc1 <- lapply(1:D, function(d) {
            verbose=control.stan$verbose, 
            show_messages=control.stan$show_messages)})
 time.mcmc1 <- proc.time()[3] - ct
+ct <- proc.time()[3]
+fit.mcmc2 <- sampling(stanmodels$nig_full,
+                      data=list(D=D, p=rep(p, D), sump=p*D, n=n, G=G, H=H,
+                                x=Reduce("cbind", replicate(D, list(x))),
+                                y=Reduce("cbind", y), C=Reduce("rbind", C),
+                                Z=Z, nuf=10, nud=10, kappaf=1, 
+                                kappad=1, xif=1, xid=1),
+                      chains=1, iter=1000, warmup=1000)
+time.mcmc2 <- proc.time()[3] - ct
+
+stan_ess(fit.mcmc2)
+stan_rhat
+stan_diag
+stan_mcse
+stan_ac
+stan_trace(fit.mcmc2, pars=c("beta[2]"))
+
 save(fit.semnig1, fit.mcmc1, file="results/simulations_gdsc_fit5.Rdata")
 
 post.mcmc1 <- lapply(fit.mcmc1[D/H*(c(1:H) - 1) + 1], extract, 

### 2020-04-10 code/analysis_gdsc.R
commit before push for bSEM analysis 4

@@ -118,6 +118,40 @@ fit.ebridge1 <- ebridge(x, y, lapply(C, function(s) {matrix(s[, -1])}), NULL,
 
 save(fit.semnig1, fit.semnig2, fit.ridge1, fit.lasso1, fit.xtune1, fit.ebridge1, 
      file="results/analysis_gdsc_fit1.Rdata")
+load(file="results/analysis_gdsc_fit1.Rdata")
+test1 <- logcpo(x, y, n, fit.semnig1)
+test2 <- logcpo(x, y, n, fit.semnig2)
+
+.f.int.cpo2 <- function(x, xtSigmax, n, p, zeta, y, xtmu) {
+  ((y - x)^2/(2*zeta) + 1)^(-(n + p)/2)*exp(-(x - xtmu)^2/(2*xtSigmax))
+}
+
+xtSigmax <- 1
+n <- 100
+p <- 100
+zeta <- 1
+y <- 1
+xtmu <- 0
+curve(.f.int.cpo2(x=x, xtSigmax=xtSigmax, n=n, p=p, zeta=zeta, y=y, xtmu=xtmu), 
+      from=-1, to=3, n=1000)
+
+a <- 1
+b <- -5
+(a - b)^2
+(b - a)^2
+
+plot(unlist(y), exp(unlist(test1)))
+abline(h=0.01, col=2)
+
+plot(y[[1]], exp(test1[[1]]))
+
+abline(h=log(0.01))
+hist(test1[[1]], breaks=30)
+abline(v=log(0.01))
+
+hist(test1[[1]])
+plot(sapply(test1, mean), sapply(test2, mean))
+abline(a=0, b=1)
 
 est <- cbind(c(fit.semnig1$eb$alphaf, NA, fit.semnig1$eb$alphad, 
                fit.semnig1$eb$lambdaf, fit.semnig1$eb$lambdad),
@@ -912,16 +946,16 @@ res <- foreach(r=1:nfolds, .packages=packages, .errorhandling="pass") %dopar% {
   #                      full.post=TRUE, control=control.semnig)
   
   # penalized regression models
-  cv.ridge1 <- lapply(c(1:D), function(d) {
-    cv.glmnet(xtrain[[d]], ytrain[[d]], alpha=0, foldid=foldid2[[d]],
-              intercept=FALSE)})
-  cv.lasso1 <- lapply(c(1:D), function(d) {
-    cv.glmnet(xtrain[[d]], ytrain[[d]], foldid=foldid2[[d]], intercept=FALSE)})
[…]

### 2020-04-10 rpackage/R/main_functions.R

@@ -398,24 +398,24 @@ new.elbo <- function(object, newx, newy) {
 
 # calculate log cpo on test data (not tested)
 logcpo <- function(xtest, ytest, ntrain, fit) {
-  D <- ncol(ytest)
-  ntest <- nrow(ytest)
+  D <- length(ytest)
+  ntest <- sapply(ytest, length)
   ptest <- sapply(xtest, ncol)
-  out <- matrix(nrow=ntest, ncol=D)
+  out <- lapply(ntest, numeric)
   for(d in 1:D) {
     xtSigmax <- rowSums((xtest[[d]] %*% fit$vb$Sigma[[d]])*xtest[[d]])
     xtmu <- as.numeric(xtest[[d]] %*% fit$vb$mu[[d]])
     p <- length(fit$vb$mu[[d]])
     zeta <- fit$vb$zeta[d]
-    for(i in 1:ntest) {
-      int <- integrate(.f.int.cpo, -Inf, Inf, xtSigmax=xtSigmax[i], n=ntrain, 
-                       p=p, zeta=zeta, y=ytest[i, d], xtmu=xtmu[i],
+    for(i in 1:ntest[d]) {
+      int <- integrate(.f.int.cpo, -Inf, Inf, xtSigmax=xtSigmax[i], n=ntrain[d], 
+                       p=p, zeta=zeta, y=ytest[[d]][i], xtmu=xtmu[i],
                        stop.on.error=FALSE)
       int.val <- ifelse(int$message=="the integral is probably divergent", NA, 
                         int$value) 
-      
-      out[i, d] <- -log(2) - log(pi) - 0.5*log(xtSigmax[i]) - 0.5*log(zeta) +
-        lgamma((ntrain + p + 2)/2) - lgamma((ntrain + p + 1)/2) - log(int.val)
+      out[[d]][i] <- -log(2) - log(pi) - 0.5*log(xtSigmax[i]) - 0.5*log(zeta) +
+        lgamma((ntrain[d] + p + 2)/2) - lgamma((ntrain[d] + p + 1)/2) -
+        log(int.val)
     }
   }
   return(out)

### 2020-04-12 docs/manuscript1.Rnw
commit before push for bSEM analysis 4
fixed full nig stan model code

--- 2020-04-08 docs/manuscript1.Rnw
@@
   external feature and drug covariates.
+
 	The canonical method for empirical Bayes is to maximise the marginal 
@@
 	$\bgamma^2_d$ and $\tau_d^2$ independence assumption 
-	results in separate maximisation problems for the 
+	results in separate optimisation problems for the 
 	feature hyperparameters ($\bm{\alpha}_{\text{feat}}$ and 
 	$\bm{\lambda}_{\text{feat}}$), and the drug hyperparameters 
-	($\bm{\alpha}_{\text{drug}}$ and $\bm{\lambda}_{\text{drug}}$) at each
-	iteration of the EM algorithm.
-
+	($\bm{\alpha}_{\text{drug}}$ and $\bm{\lambda}_{\text{drug}}$).
 	If we stack the drug and feature covariates:
@@
   variances \cite[]{rue_approximate_2009,consonni_mean-field_2007,
-  bishop_pattern_2006,wang_inadequacy_2005}. If posterior variances are 
-  required, 
-  we suggest generating samples from the posterior with fixed hyperparameters
+  bishop_pattern_2006,wang_inadequacy_2005}. In simulation Scenario 5 in Section 
+  \ref{md-sec:simulations} of the SM, we compare the variational posterior to 
+  MCMC samples from the posterior with fixed hyperparameters
   estimates (after the procedure described in Section \ref{sec:empiricalbayes}
-  has converged). A Gibbs sampler is described in 
-  SM Section \ref{sm-sec:gibbssampler}. Alternatively, we provide an
+  has converged). In this simulation Scenario and other settings (not 
+  shown), the variational approximation to the posterior is accurate. If 
+  however, the
+  user is reluctant to trust the variational posterior variances, samples
+  from the posterior may be generated with the
+  Gibbs sampler in SM Section \ref{sm-sec:gibbssampler}. Alternatively, 
+  we provide an
   implementation of the proposed model in stan using the R package rstan 
@@
 	A pre-processing step selects 100 features with largest variance, while
-	100 drug sensitivities for 1014 cell lines are 
+	100 drug sensitivities for 100 cell lines are 
 	simulated from model (\ref{eq:likelihood}) and 
@@
 	    the $\gamma_{jd}^2$ and $\tau_d^2$ according to (\ref{eq:prior}).}
-	  \item{Scenario 4 is equal to Scenario three, except that the rows of the
-	    external covariates matrix are permuted. Permutation
-	    adds noise and is supposed to mimic a low external covariates signal. We
-	    mimic different noise levels by permutation of fractions 
+	  \item{Scenario 4 is equal to Scenario three, except that we add noise to
+	    the external covariates. Noise is supposed to mimic a low external 
+	    covariate signal and is added by permutation of fractions 
 	    $q \in \{0.1, 0.2, 0.33, 0.5, 0.67, 0.8, 1\}$ of the 
-	    number of rows.}
+	    rows of the external covariates.}
 	\end{itemize}
@@
 	estimated (feature, drug, or both in Scenarios 1, 2, and 3/4, respectively),
+	and 
 	(ii) the NIG model estimated as in Section \ref{sec:estimation} that includes
@@
 	NIG$_{\text{f}+\text{d}}$, again depending on which variance components are 
-	estimated, frequentist 
+	estimated.
 	Exclusion of the external covariates as in the NIG$_{\text{f}}^-$, 
@@
 	opposed to the global-local shrinkage models NIG$_{\text{f}+\text{d}}$ and
-	NIG$_{\text{f}+\text{d}}^-$. All models are fit on 100 randomly selected cell
-	lines. We repeat every simulation Scenario 100 times.
-
-	SM Section \ref{sm-sec:simulations} contains
-	additional simulation results for
-	the NIG model, plus results for 
-	four alternative methods: frequentist (i) lasso and (ii) ridge models, 
-	and and EB version of the ridge model: (iii) a
+	NIG$_{\text{f}+\text{d}}^-$. We repeat every simulation Scenario 100 times.
+
+	SM Section \ref{sm-sec:simulations} contains an additional simulation 
+	Scenario 5 that investigates the performance of the variational Bayes approach
+	as compared to MCMC samples from the posterior, as well as more 
+	simulation results for Scenarios 1-4 for
+	the NIG model and four alternative methods: frequentist (i) 
+	lasso and (ii) ridge models, and EB version of the ridge model: (iii) a
 	multivariate Bayesian extension of \texttt{xtune} 
@@
 	After removing all cell lines with missing values, we end up with 388 to 1043
-	IC50 estimates for for 251 drugs. Differences in the number of cell lines
+	IC50 estimates for 251 drugs. Differences in the number of cell lines
 	between drugs occur, because not all drug and cell line combinations are 
@@
 	17737 and 300 genes, respectively.
-
-	A comprehensive penalized regression
-	approach such as in 
-	{\cite{mai_composite_2019} uses most of the available molecular markers 
-	instead of just gene expression (68 mutations, 426 copy numbers, 
-	and 2602 gene expression levels for 498 cell lines and 97 drugs). The 
-	current version of our software is not able to handle such problem sizes,
-	due to the repeated calculation of posterior parameters in the empirical
-	Bayes iterations. Currently, problems with number of outcomes in the hundreds
-	and number of features and samples up to a thousand are computationally 
-	feasible.
+	% 
+	% A comprehensive penalized regression
+	% approach such as in 
+	% {\cite{mai_composite_2019} uses most of the available molecular markers 
+	% instead of just gene expression (68 mutations, 426 copy numbers, 
+	% and 2602 gene expression levels for 498 cell lines and 97 drugs). The 
+	% current version of our software is not able to handle such problem sizes,
+	% due to the repeated calculation of posterior parameters in the empirical
+	% Bayes iterations. Currently, problems with number of outcomes in the hundreds
+	% and number of features and samples up to a thousand are computationally 
+	% feasible.
 
@@
 	    A pre-processing step selects between 221 and 280 genes per drug for 
-	    which a mutation $p$-value is available.}
+	    which both the expression as well as a mutation $p$-value is available.}
 	  \item{Analysis 2 includes both gene expressions and mutations as predictors
@@
 	    is a expression or mutation ($G=1$, $H=0$). For this analysis,
-	    we pre-select 300 gene 
-	    expressions with maximum 
-	    variance and 295 gene mutations for which there are mutated as well as 
-	    unmutated cell lines available.}
+	    we pre-select 300 gene expressions with maximum 
+	    variance and 295 gene mutations for which there are both mutated and
+	    wildtype cell lines available.}
 	  \item{Analysis 3 uses 500 gene expressions as features, selected based on
@@
 	(see SM Section \ref{sm-sec:mxtune}). 
-	% In addition, in Analyses 2 and 
-	% 4, we fit a (vi) bSEM model 
-	% \cite[]{leday_gene_2017,kpogbezan_empirical_2017}. The
-	% bSEM model is similar to the NIG model in that it
-	% draws the $\beta_{jd}$ from a conditionally normal distribution, but
-	% fixes $\tau_d^2=1$ and draws
-	% $\gamma_{jd}^2 \sim \Gamma^{-1} (\phi_{jd}, \lambda_{jd})$, instead of
-	% $\gamma_{jd}^2 \sim \mathcal{IG} (\phi_{jd}, \lambda_{jd})$, where 
-	% $\Gamma^{-1} (\phi, \lambda)$ denotes the inverse gamma distribution with 
-	% shape $\phi$ and scale $\lambda$. Additionally, the bSEM model can
-	% only incorporate one binary external covariate that divides the features into
-	% two groups, i.e., $\bm{\alpha}_{\text{feat}} = \begin{bmatrix} \alpha_0 &
-	% \alpha_1
-	% \end{bmatrix} \tr$ and $\bm{c}_{jd} = \begin{bmatrix} 1 & 0 \end{bmatrix}
-	% \tr$ if feature $jd$ belongs to group one and $\bm{c}_{jd} = \begin{bmatrix}
-	% 1 & 1 \end{bmatrix} \tr$ if feature $jd$ belongs to group two. Both
-	% $\bm{\alpha}_{\text{feat}}$ and $\lambda_{\text{feat}}$ are
-	% estimated in a similar fashion as in Section \ref{sec:estimation}.
+	In addition, in Analyses 2 and
+	4, we fit a (vi) bSEM model
+	\cite[]{leday_gene_2017,kpogbezan_empirical_2017}. The
+	bSEM model is similar to the NIG model in that it
+	draws the $\beta_{jd}$ from a conditionally normal distribution, but
+	fixes $\tau_d^2=1$ and draws
+	$\gamma_{jd}^2 \sim \Gamma^{-1} (\phi_{jd}, \lambda_{jd})$, instead of
+	$\gamma_{jd}^2 \sim \mathcal{IG} (\phi_{jd}, \lambda_{jd})$, where
+	$\Gamma^{-1} (\phi, \lambda)$ denotes the inverse gamma distribution with
+	shape $\phi$ and scale $\lambda$. Additionally, the bSEM model can
+	only incorporate one binary external covariate that divides the features into
+	two groups, i.e., $\bm{\alpha}_{\text{feat}} = \begin{bmatrix} \alpha_0 &
+	\alpha_1
+	\end{bmatrix} \tr$ and $\bm{c}_{jd} = \begin{bmatrix} 1 & 0 \end{bmatrix}
+	\tr$ if feature $jd$ belongs to group one and $\bm{c}_{jd} = \begin{bmatrix}
+	1 & 1 \end{bmatrix} \tr$ if feature $jd$ belongs to group two. Both
+	$\bm{\alpha}_{\text{feat}}$ and $\lambda_{\text{feat}}$ are
+	estimated in a similar fashion as in Section \ref{sec:estimation}.
 
@@
   $\hat{\bm{\beta}_d}$ the estimator for $\bm{\beta}_d$, and its standard error
-  are estimated by 10-fold cross validation. In the NIG models that 
+  are estimated by 10-fold cross validation. In the NIG and bSEM models, that 
   provide full posteriors, the posterior mean 
@@
 pmse4 <- res[substr(rownames(res), 1, 5)=="pmse.", ]
-pmse4 <- aggregate(pmse4, by=list(rep(1:10, each=251)), FUN="mean")[, -1]
+pmse4 <- aggregate(pmse4, by=list(rep(1:10, each=249)), FUN="mean")[, -1]
 mpmse4 <- apply(pmse4, 2, mean)
@@
               ppmse2, c(ppmse3[1:6], NA, ppmse3[7:length(ppmse3)], rep(NA, 3)),
-              ppmse4)[, c(1:4, 6)]
+              ppmse4)[, c(1:3, 5:7)]
 tabm5 <- rbind(c(mpmse1[1:6], NA, mpmse1[7:length(mpmse1)], rep(NA, 3)),
                mpmse2, c(mpmse3[1:6], NA, mpmse3[7:length(mpmse3)], rep(NA, 3)),
-               mpmse4)[, c(1:4, 6)]
+               mpmse4)[, c(1:3, 5:7)]
 tab5 <- t(sapply(1:nrow(tab5), function(r) {
@@
 colnames(tab5) <- c("NIG$_{\\text{f}+\\text{d}}^-$", 
-                    "NIG$_{\\text{f}+\\text{d}}$", "ridge", "lasso", 
-                    "mxtune")
+                    "NIG$_{\\text{f}+\\text{d}}$", "ridge", "xtune",
+                    "mxtune", "bSEM")
 @
@@
 
+  Section \ref{sm-sec:gdsc} in the SM displayes the conditional predictive 
+  ordinates (CPO) for the four analyses. A visual inspection of the CPOs learns
+  that no extreme outliers occur.
 	\section{Discussion}\label{sec:discussion}
@@
 	All results and documents may be recreated from 
-	\url{https://github.com/magnusmunch/cambridge}.
+	\url{https://github.com/magnusmunch/cambridge}. All timings were obtained
+	on a 2016 MacBook Pro with a 2 GHz Dual-Core Intel Core i5 processor and
+	8 GB of 1867 MHz LPDDR3 memory, running macOS  10.15.1.
 
 	\section*{Acknowledgements}
-	% We thank Zhi Zhao and Leiv Tore Salte R{\o}nneberg from the Oslo Centre for 
-	% Biostatistics and Epidemiology for their suggestions on the pre-processing of
-	% the GDSC data. 
-	MM visited SR and GL on a travel grant funded the Amsterdam Public Health 
+	We thank the referees for their helpful comments and suggestions on an earlier
+	version of this manuscript.
+	MM visited SR and GL on a travel grant funded by the Amsterdam Public Health 
 	institute's Methodology program.

### 2020-04-12 code/simulations_gdsc.R

@@ -774,8 +774,52 @@ fit.mcmc2 <- sampling(stanmodels$nig_full,
                                 y=Reduce("cbind", y), C=Reduce("rbind", C),
                                 Z=Z, nuf=10, nud=10, kappaf=1, 
                                 kappad=1, xif=1, xid=1),
-                      chains=1, iter=1000, warmup=1000)
+                      chains=1, iter=2000, warmup=1000)
 time.mcmc2 <- proc.time()[3] - ct
+
+save(fit.semnig1, fit.mcmc1, fit.mcmc2, 
+     file="results/simulations_gdsc_fit5.Rdata")
+load(file="results/simulations_gdsc_fit5.Rdata")
+library(rstan)
+get_stancode(fit.mcmc2)
+
+post.mcmc1 <- lapply(fit.mcmc1[D/H*(c(1:H) - 1) + 1], extract, 
+                     pars=paste0("beta[", p/G*(c(1:G) - 1) + 1, "]"))
+post.mcmc2 <- extract(fit.mcmc2, pars=c("beta", "alphaf", "alphad", 
+                                        "lambdaf", "lambdad"))
+post.mcmc2 <- list(beta=lapply(split(as.data.frame(t(post.mcmc2$beta)), 
+                                     rep(1:D, each=p))[D/H*(c(1:H) - 1) + 1],
+                               function(s) {
+                                 unname(as.matrix(s)[p/G*(c(1:G) - 1) + 1, ])}),
+                   alphaf=t(post.mcmc2$alphaf), alphad=t(post.mcmc2$alphad),
+                   lambdaf=t(post.mcmc2$lambdaf), lambdad=t(post.mcmc2$lambdad))
+save(time.semnig1, time.mcmc1, time.mcmc2, post.mcmc1, post.mcmc2, post.semnig1, 
+     file="results/simulations_gdsc_res5.Rdata")
+
+opar <- par(no.readonly=TRUE)
+par(mar=opar$mar*c(1, 1.3, 1, 1))
+layout(matrix(c(rep(rep(c(1:2), each=2), 2), rep(rep(c(3:4), each=2), 2)), 
+              nrow=4, ncol=4, byrow=TRUE))
+hist(post.mcmc2$alphaf[1, ])
[…]

### 2020-04-14 docs/manuscript1.Rnw
changed name of project

@@
 	\noindent\textbf{Software available from}: 
-	\url{https://github.com/magnusmunch/cambridge}
+	\url{https://github.com/magnusmunch/NIG}
 
@@
   bishop_pattern_2006,wang_inadequacy_2005}. In simulation Scenario 5 in Section 
-  \ref{md-sec:simulations} of the SM, we compare the variational posterior to 
+  \ref{sm-sec:simulations} of the SM, we compare the variational posterior to 
   MCMC samples from the posterior with fixed hyperparameters
@@
   \cite[]{guo_rstan:_2018} at 
-  \url{https://github.com/magnusmunch/cambridge}.
+  \url{https://github.com/magnusmunch/NIG}.
 
@@
 
+  Finally, the results from Scenario 5 in SM Section \ref{sm-sec:simulations} 
+  show that the VB approximation is quite good as compared to standard MCMC.
+
 	\section{GDSC data}\label{sec:gdsc}
@@
 	In all analyses we use all cell lines to estimate the hyperparameters 
-	presented in Section \ref{sec:results}. Mean prediction PMSE,
+	presented in Section \ref{sec:results}. Mean prediction means squared errors
+	(PMSE) and its standard error are estimated by 10-fold cross validation, where
 	$\text{PMSE} = D^{-1} n^{-1} \sum_{d=1}^D 
   \sum_{i=1}^n (\mathbf{y}_d - \mathbf{x}_i \tr \hat{\bm{\beta}_d})^2$, with
-  $\hat{\bm{\beta}_d}$ the estimator for $\bm{\beta}_d$, and its standard error
-  are estimated by 10-fold cross validation. In the NIG and bSEM models, that 
-  provide full posteriors, the posterior mean 
-  $\E(\bm{\beta}_d | \mathbf{y}_d)$ is used point estimate.
+  $\hat{\bm{\beta}_d}$ the estimator for $\bm{\beta}_d$. In the NIG and bSEM 
+  models, that provide full posteriors, the posterior mean 
+  $\E(\bm{\beta}_d | \mathbf{y}_d)$ is used as point estimate.
 
@@
 # estimates tables
-fit <- read.table("results/analysis_gdsc_fit1.txt", row.names=NULL)
-temp <- fit[, 1]
-fit <- as.matrix(fit[, -1])
-rownames(fit) <- temp
-
-tab1 <- t(fit)[c(1, 2), c(1:3)]
+fit1 <- read.table("results/analysis_gdsc_fit1.txt", row.names=NULL)
+temp <- fit1[, 1]
+fit1 <- as.matrix(fit1[, -1])
+rownames(fit1) <- temp
+
+tab1 <- t(fit1)[c(1, 2), c(1:3)]
 rownames(tab1) <- c("NIG$_{\\text{f}+\\text{d}}^-$", 
@@
 
-fit <- read.table("results/analysis_gdsc_fit2.txt", row.names=NULL)
-temp <- fit[, 1]
-fit <- as.matrix(fit[, -1])
-rownames(fit) <- temp
-
-tab2 <- t(fit)[c(1, 2), c(1:3)]
+fit2 <- read.table("results/analysis_gdsc_fit2.txt", row.names=NULL)
+temp <- fit2[, 1]
+fit2 <- as.matrix(fit2[, -1])
+rownames(fit2) <- temp
+
+tab2 <- t(fit2)[c(1, 2, 7), c(1:3)]
 rownames(tab2) <- c("NIG$_{\\text{f}+\\text{d}}^-$", 
-                    "NIG$_{\\text{f}+\\text{d}}$")
+                    "NIG$_{\\text{f}+\\text{d}}$",
+                    "bSEM")
 colnames(tab2) <- c("feature intercept", "mutation", "drug intercept")
 
-fit <- read.table("results/analysis_gdsc_fit3.txt", row.names=NULL)
-temp <- fit[, 1]
-fit <- as.matrix(fit[, -1])
-rownames(fit) <- temp
-
-tab3 <- t(fit)[c(1, 2), c(1:3)]
+fit3 <- read.table("results/analysis_gdsc_fit3.txt", row.names=NULL)
+temp <- fit3[, 1]
+fit3 <- as.matrix(fit3[, -1])
+rownames(fit3) <- temp
+
+tab3 <- t(fit3)[c(1, 2), c(1:3)]
 rownames(tab3) <- c("NIG$_{\\text{f}+\\text{d}}^-$", 
@@
 
-fit <- read.table("results/analysis_gdsc_fit4.txt", row.names=NULL)
-temp <- fit[, 1]
-fit <- as.matrix(fit[, -1])
-rownames(fit) <- temp
-
-tab4.1 <- t(fit)[c(1, 2), c(1:2)]
+fit4 <- read.table("results/analysis_gdsc_fit4.txt", row.names=NULL)
+temp <- fit4[, 1]
+fit4 <- as.matrix(fit4[, -1])
+rownames(fit4) <- temp
+
+tab4.1 <- t(fit4)[c(1, 2, 7), c(1:2)]
 rownames(tab4.1) <- c("NIG$_{\\text{f}+\\text{d}}^-$", 
-                    "NIG$_{\\text{f}+\\text{d}}$")
+                    "NIG$_{\\text{f}+\\text{d}}$", "bSEM")
 colnames(tab4.1) <- c("feature intercept", "pathway")
-tab4.2 <- t(fit)[c(1, 2), c(3:7)]
+tab4.2 <- t(fit4)[c(1, 2), c(3:7)]
 rownames(tab4.2) <- c("NIG$_{\\text{f}+\\text{d}}^-$", 
@@
 
+psel1 <- res[substr(rownames(res), 1, 5)=="psel.", ]
 pmse1 <- res[substr(rownames(res), 1, 5)=="pmse.", ]
+apmse1 <- aggregate(pmse1, list(rownames(pmse1)), FUN="mean")[, -1]
 pmse1 <- aggregate(pmse1, by=list(rep(1:10, each=251)), FUN="mean")[, -1]
@@
 
+psel2 <- res[substr(rownames(res), 1, 5)=="psel.", ]
 pmse2 <- res[substr(rownames(res), 1, 5)=="pmse.", ]
@@
 
+psel3 <- res[substr(rownames(res), 1, 5)=="psel.", ]
 pmse3 <- res[substr(rownames(res), 1, 5)=="pmse.", ]
@@
 
+psel4 <- res[substr(rownames(res), 1, 5)=="psel.", ]
 pmse4 <- res[substr(rownames(res), 1, 5)=="pmse.", ]
@@
               ppmse2, c(ppmse3[1:6], NA, ppmse3[7:length(ppmse3)], rep(NA, 3)),
-              ppmse4)[, c(1:3, 5:7)]
+              ppmse4)
 tabm5 <- rbind(c(mpmse1[1:6], NA, mpmse1[7:length(mpmse1)], rep(NA, 3)),
                mpmse2, c(mpmse3[1:6], NA, mpmse3[7:length(mpmse3)], rep(NA, 3)),
-               mpmse4)[, c(1:3, 5:7)]
-tab5 <- t(sapply(1:nrow(tab5), function(r) {
-  id <- which(tabm5[r, ]==min(tabm5[r, ], na.rm=TRUE))
-  out <- tab5[r, ]
+               mpmse4)
+tab5.1 <- tab5[, c(1:4, 6:7)]
+tabm5.1 <- tabm5[, c(1:4, 6:7)]
+tab5.1 <- t(sapply(1:nrow(tab5.1), function(r) {
+  id <- which(tabm5.1[r, ]==min(tabm5.1[r, ], na.rm=TRUE))
+  out <- tab5.1[r, ]
   out[id] <- paste0("\\textbf{", out[id], "}")
   out}))
-rownames(tab5) <- paste0("analysis ", 1:nrow(tab5))
-colnames(tab5) <- c("NIG$_{\\text{f}+\\text{d}}^-$", 
-                    "NIG$_{\\text{f}+\\text{d}}$", "ridge", "xtune",
-                    "mxtune", "bSEM")
-@
-  The non-zero $\bm{\alpha}$ estimates in Tables ? and ? show 
+rownames(tab5.1) <- paste0("analysis ", 1:nrow(tab5.1))
+colnames(tab5.1) <- c("NIG$_{\\text{f}+\\text{d}}^-$", 
+                    "NIG$_{\\text{f}+\\text{d}}$", "ridge", "lasso", "mxtune", 
+                    "bSEM")
+tab5.1 <- t(tab5.1)
+
+tab5.2 <- tab5[, c(4, 9)]
+tabm5.2 <- tabm5[, c(4, 9)]
+tab5.2 <- t(sapply(1:nrow(tab5.2), function(r) {
+  id <- which(tabm5.2[r, ]==min(tabm5.2[r, ], na.rm=TRUE))
+  out <- tab5.2[r, ]
+  out[id] <- paste0("\\textbf{", out[id], "}")
+  out}))
+rownames(tab5.2) <- paste0("analysis ", 1:nrow(tab5.2))
+colnames(tab5.2) <- c("lasso", "NIG$_{\\text{f}+\\text{d}}+$DSS")
+tab5.2 <- t(tab5.2)
+
+mpsel <- rbind(c(colMeans(psel1[, 1:6]), NA, colMeans(psel1[, 7:ncol(psel1)]), 
+                 rep(NA, 3)), colMeans(psel2), 
+               c(colMeans(psel3[, 1:6]), NA, colMeans(psel3[, 7:ncol(psel3)]), 
+                 rep(NA, 3)), colMeans(psel4))
+
+# prior variances
+vars1 <- round(1/as.numeric(model.matrix(~ factor(c(1:3))) %*% 
+                              fit4[c(3, 6, 7), 2]), 4)
+vars2 <- round((cbind(1, c(0, 1)) %*% fit2[c(1, 2), c(2, 7)])[1, ]/
+                 (cbind(1, c(0, 1)) %*% fit2[c(1, 2), c(2, 7)])[2, ], 3)
+
+# pmse ranges
+ranges1 <- apply(rbind(mpmse1[c(1:4, 6:7)], mpmse2[c(1:4, 6:7)], 
+                       mpmse3[c(1:4, 6:7)], mpmse4[c(1:4, 6:7)]), 1, range) 
+ranges1 <- round(ranges1[2, ] - ranges1[1, ], 4)
+psels1 <- round(mpsel[, c(4, 9)], 0)
+@
+  The non-zero NIG$_{\text{f}+\text{d}}$ $\bm{\alpha}$ estimates in Tables 
+  \ref{tab:analysis_gdsc_fit1}-\ref{tab:analysis_gdsc_fit4.2} show 
   that there is an effect of the
-  external covariates. Models NIG$_{\text{f}+\text{d}}$, NIG$_{\text{f}}$, and
-  bSEM estimate a negative effect for
-  the external pathway covariate. This translates to a negative additive effect 
-  on the expected prior precision of the $\bm{\beta}_d$, i.e., genes that are 
-  found
-  in the target pathway of a drug are more predictive of IC50 than genes that
-  are not in the target pathway, according to expectation. Furthermore the 
-  NIG$_{\text{f}+\text{d}}$
-  model estimates positive effects for the experimental and developmental drugs,
-  compared to the approved drugs. This translates to smaller expected 
-  prior precisions for the regression parameters of approved drugs, i.e.,
-  approved drugs are easiest to predict, followed by developmental and 
-  experimental drugs, as expected. On the other hand, the NIG$_{\text{d}}$ 
-  model, that only includes external
-  drug covariates, estimates a negative effect for the developmental drugs.
-  The NIG$_{\text{f}+\text{d}}$ model estimates the smallest prior 
-  $\bm{\beta}_d$ precision for the
-  targeted drugs, followed by cytotoxic drugs, while drugs with an unkown target
-  are assigned the largest prior $\bm{\beta}_d$ precision, according
-  to expectation. Somewhat surprisingly, the NIG$_{\text{d}}$ model
-  estimates the smallest
-  prior $\bm{\beta}_d$ precision for the cytotoxic drugs, compared to the 
-  targeted and unkown drugs. We conjecture that, due to absence of a local
-  feature-specific variance component, this model is not flexible 
-  enough to correctly capture the full covariance structure of the 
-  $\bm{\beta}_d$ and therefore incorreclty estimates $\bm{\alpha}$.
-  To see the total effect of the external covariates on the $\bm{\beta}_d$ 
-  prior, we calculated the range of the expected prior variances 
-  (up to a multiplicative error variance) as 
-  $\hat{E}(\gamma_{jd}^2 \cdot \tau_d^2) \in \[ ... \]$ and
-  $\hat{E}(\gamma_{jd}^2 \cdot \tau_d^2) \in \[ ... \]$ for
-  NIG$_{\text{f}+\text{d}}$ and bSEM, respectively. This confirms that the
-  estimated prior 
-  $\bm{\beta}_d$ variance components are in the same order of magnitude and that
-  the external covariates have a detectable effect on the prior.
+  external covariates. 
+  \begin{itemize}
+	  \item{Analysis 1 results in a positive additive effect of the mutation 
+	    $p$-values on the prior $\bm{\beta}_d$ precisions
+	    (Table \ref{tab:analysis_gdsc_fit1}). This translates
+	    to more $\bm{\beta}_d$ shrinkage with increasing $p$-value,
+	    as expected.}
+	  \item{Analysis 2 shows that gene mutations are more predictive 
+	    than gene expressions, as observed from the negative effect of mutation
+	    dummy on prior precions (Table \ref{tab:analysis_gdsc_fit2}): 
+	    mutations are shrunken less than
+	    expressions.}
+	  \item{Analysis 3 indicates that CCLE $p$-values are positively related
+	    to prior precision (Table \ref{tab:analysis_gdsc_fit2}), i.e., 
+	    higher CCLE $p$-values results in more shrinkage of $\bm{\beta}_d$, as 
+	    expected.}
+	  \item{Analysis 4 gives a negative effect for the pathway dummy (Table 
+	    \ref{tab:analysis_gdsc_fit4.1}), indicating less shrinkage for genes
+	    that are in the drugs' target pathway, as expected. According to 
+	    expectation, experimental and 
+	    developmental drugs prior precisions are shrunken more than
+	    the reference category, approved drugs (Table 
+	    \ref{tab:analysis_gdsc_fit4.2}). Somewhat surprisingly, targeted and 
+	    unkown drugs are shrunken more than the reference, cytotoxic drugs (Table 
+	    \ref{tab:analysis_gdsc_fit4.2}), although the differences on the
+	    $\bm{\beta}_d$ variance scale are small:
+	    $\hat{\E}(\tau_{d}^{2}) = \{ \Sexpr{vars1[1]}, \Sexpr{vars1[2]}, 
+	    \Sexpr{vars1[3]} \}$ for cytotoxic, targeted, and unkown drugs, 
+	    respectively (ignoring other variance components).}
+	\end{itemize}
+
+  In comparison to bSEM, Analysis 2 (Table \ref{tab:analysis_gdsc_fit2}) stands
+  out. The ratio of expression prior precision to mutation prior precision
+  (ignoring other variance components) is
+  $\hat{\E}(\tau_{\text{exp}}^{-2})/\hat{\E}(\tau_{\text{mut}}^{-2}) 
+  \approx \Sexpr{vars2[1]}$ for NIG$_{\text{f}+\text{d}}$ and 
+  $\hat{\E}(\tau_{\text{exp}}^{-2})/\hat{\E}(\tau_{\text{mut}}^{-2}) 
+  \approx \Sexpr{vars2[2]}$ for 
+  bSEM. This means that the effects are reversed: bSEM penalized mutations 
+  more than expressions, while NIG$_{\text{f}+\text{d}}$ penalized
+  expressions more. Furtheremore, the magnitude of the effect is much larger for
+  bSEM, indicating that bSEM picks up more signal in the external covariate 
+  here.
 
@@
 
-  The median prediction mean squared error (PMSE), calculated on the 
-  test data is displayed in Table \ref{tab:analysis_gdsc_pmse1}. 
-  Note that due to standardisation, the empty reference model has a PMSE of one.
-  In terms of prediction 
-  error, ridge outperforms the other models 
-  (PMSE of ? ), while bSEM 
-  (PMSE of ? ) performs worse than 
-  the null model (PMSE of ? ). Second and third best
-  performing models are NIG$_{\text{f}+\text{d}}^-$ and 
-  NIG$_{\text{f}+\text{d}}$ (PMSEs of ? and
-  ?, respectively), both outperforming the lasso
-  with a PMSE of ?. The underperformance of the 
-  NIG$_{\text{f}}$ and NIG$_{\text{d}}$ models shows the usefulness of a
-  global-local shrinkage prior as opposed to only global or local shrinkage as 
-  as in models NIG$_{\text{f}}$ and NIG$_{\text{d}}$.
+  The mean prediction mean squared errors (PMSE), calculated on the 
+  test data, are diplayed in Table \ref{tab:analysis_gdsc_pmse1}.
+  Note that due to standardisation, an empty reference model has a PMSE of one.
+  In terms of PMSE, ridge outperforms the other models in all Analyses 1, 3, and
+  4, while bSEM outperforms in Analysis 2. In general however, all models
+  perform very similar. The ranges of mean PMSE over all methods are
+  \Sexpr{ranges1[1:3]}, and \Sexpr{ranges1[4]}, for Analyses 1-4, respectively,
+  indicating that the differences in predictive performance are very small.
+  Furthermore, the differences between models that include external covariates
+  and their respective counterparts that do not, are small, indicating that the
+  external covariates are not very informative here.
 <<analysis_gdsc_pmse1>>=  
 kableExtra::kable_styling(knitr::kable(
-  tab5, align="r",
-  caption="Mean (standard deviation) of cross-validated PMSE for GDSC data (standard error). Best performing model in bold.",
+  tab5.1, align="r",
+  caption="Mean (standard deviation) of cross-validated PMSE for GDSC data. Best performing model (per analysis) in bold.",
   format="latex", booktabs=TRUE, escape=FALSE), 
@@
   Part of the differences in performance between
-  NIG, lasso, and ridge may be explained with the different levels of 
-  ``sparsity'' in the solution. Although NIG and ridge do not give exactly 
-  sparse solutions (none of the $\bm{\beta}_d$ coefficients are set to zero), we
-  may use the ``heavy-tailedness'' of the distribution of the estimated 
-  $\bm{\beta}_d$ as a proxy for sparsity. A common measure of heavy-tailedness
-  is excess kurtosis, calculated as ?, 
-  ?, and ? for the 
-  NIG$_{\text{f}+\text{d}}$, 
-  lasso, and ridge estimates, respectively. Evidently, lasso gives the 
-  ``sparsest'' solution, followed by NIG and ridge. In some sense, this
-  means that the NIG uses less of the $\bm{\beta}_d$ to predict than the ridge
-  model. In addition, the penalized regression methods estimate penalty 
-  parameters by cross-validation. Cross-validation directly minimises the 
-  (approximate) PMSE, as opposed to empirical Bayes in the NIG that maximises 
-  the (approximate) marginal likelihood, a measure of model fit. The superior 
-  PMSE of ridge regression is therefore not surprising. A caveat with penalized
-  regression methods is that they do not give measures of parameter uncertainty.
-  NIG, on the other hand, gives the full posterior of the parameters, either
-  through a variational Bayes approximation or with the Gibbs sampler from
-  SM Section \ref{sm-sec:gibbssampler}. The full posterior gives direct access
-  to the parameter uncertainties for a better interpretable model.
-
-  Section \ref{sm-sec:gdsc} in the SM displayes the conditional predictive 
-  ordinates (CPO) for the four analyses. A visual inspection of the CPOs learns
-  that no extreme outliers occur.
+  NIG and ridge may be explained with the different levels of 
+  sparsity in the solution. Although NIG does not automatically select features, 
+  as opposed to the lasso,
+  its $\bm{\beta}_d$ prior is heavy-tailed as compared to 
+  the ridge prior 
+  thus facilitates features selection. Selection of features from a sparse prior
+  may be achieved through the decoupling shrinkage and selection approach (DSS)
+  introduced in \cite{hahn_decoupling_2015}. We have applied DSS in our 
+  analyses, where we select (approximately) the same number of features as 
+  lasso. Table 
+  \ref{tab:analysis_gdsc_pmse2} compares the resulting mean PMSE to lasso. The
+  Table shows that NIG$_{\text{f}+\text{d}}$ plus DSS outperforms lasso,
+  with the same number of features (\Sexpr{psels1[-4, 2]}, and 
+  \Sexpr{psels1[4, 2]} features in the four analyses, respectively).
+
+<<analysis_gdsc_pmse2>>=  
+kableExtra::kable_styling(knitr::kable(
+  tab5.2, align="r",
+  caption="Mean (standard deviation) of cross-validated PMSE for GDSC data. Best performing model (per analysis) in bold.",
+  format="latex", booktabs=TRUE, escape=FALSE), 
+  latex_options=c("HOLD_position"))
+@
+
+  To assess model fit, Section \ref{sm-sec:gdsc} in the SM displays the 
+  conditional predictive 
+  ordinates (CPO) for the NIG$_{\text{f}+\text{d}}$ model for the four analyses. 
+  A visual inspection of the CPOs learns that no extreme outliers occur.
+
 	\section{Discussion}\label{sec:discussion}
 <<time>>=
-# load("results/analysis_gdsc_fit1.Rdata")
-# time2.semnig <- sum(fit2.semnig$time)
-time2.semnig <- 1
+times <- round(read.table("results/analysis_gdsc_time1.txt")[, 1], 0)
 @
@@
 	data-driven and adaptive empirical Bayes estimation of the hyperparameters in 
-	the normal inverse Gaussian prior model (\ref{eq:prior}). Estimation is 
-	efficient and scales well with the number of features and samples.
+	the normal inverse Gaussian prior model (\ref{eq:prior}). Variational Bayes
+	estimation is efficient and scales well with the number of features and 
+	samples.
 	Estimation of the NIG model that includes both external drugs and feature
-	covariates (NIG$_{\text{f}+\text{d}}$ in Section \ref{sec:gdsc}) on the GDSC
-	data from Section
-	\ref{sec:gdsc} took \Sexpr{round(time2.semnig, 2)} seconds on a 2016 
-	MacBook Pro with a 2 GHz Dual-Core Intel Core i5 processor and
-	8 GB of 1867 MHz LPDDR3 memory, running macOS  10.15.1.
+	covariates (NIG$_{\text{f}+\text{d}}$ in Section \ref{sec:gdsc}) in the
+	GDSC data analyses in Section \ref{sec:gdsc} took 
+	\Sexpr{times[-1]}, and \Sexpr{times[1]} seconds on a 2016 
+	MacBook Pro with 2 GHz Dual-Core Intel Core i5 processor and 8 GB of memory, 
+	running macOS  10.15.1.
 
@@
 	due to the interplay between the two sources of information. Nevertheless,
-	the simulations show that inclusion of informative external covariates is 
+	the simulation results in Section \ref{sm-sec:simulations} of the SM show that
+	inclusion of informative external covariates is 
 	still beneficial to predictive performance.
 
-	The model is put into practice on the GDSC data. 
-	The NIG model is competitive with convential models like lasso in 
+	The model is put into practice on the GDSC data in Section \ref{sec:gdsc}. 
+	The NIG model is competitive with convential, penalized models like lasso in 
 	terms of predictive performance, but is slightly outperformed by ridge.
-	We note that estimation of the penalty parameters in the latter two models
-	is geared towards predictive performance, because cross-validation is used
-	to train these parameters, as opposed to empirical Bayes estimation of the
-	hyperparameters in the NIG model, which aims to improve model fit. In that 
-	sense, the comparison metric (PMSE) is not to the benefit of the NIG model.
-	We note that direct prediction error optimisation is not feasible in our
-	external covariates setting.
+	The penalized regression methods estimate penalty 
+  parameters by cross-validation. Cross-validation directly minimises the 
+  (approximate) PMSE, as opposed to empirical Bayes in the NIG that maximises 
+  the (approximate) marginal likelihood, a measure of model fit. The superior 
+  PMSE of ridge regression is therefore not surprising. 
+  However, direct prediction error optimisation by cross-validation is not 
+	feasible in the external covariates setting, due to the large number of 
+	hyperparameters.
+  A caveat with penalized
+  regression methods is that they do not give measures of parameter uncertainty.
+  NIG, on the other hand, gives the full posterior of the parameters, either
+  through a variational Bayes approximation or with the Gibbs sampler from
+  SM Section \ref{sm-sec:gibbssampler}. The full posterior gives direct access
+  to the parameter uncertainties for a better interpretable model.
 
 	The results in Section
-	\ref{sec:gdsc} show that the inclusion of the external covariates covariates 
+	\ref{sec:gdsc} show that the inclusion of the external covariates
 	indeed substantially modifies the hyperparameters. Although, PMSE of the NIG
@@
 	estimates are still informative and may guide future research efforts.
-
-	In addition, the comparison to bSEM, a Bayesian counterpart model that allows
-	for the inclusion of one binary external feature covariate favours the 
-	proposed NIG model. PMSEs are almost uniformly lower over the drugs for the 
-	NIG model compared to bSEM. Consequently, the total PMSE is significantly
-	lower for the NIG model.
 
@@
 
-  Another possible criticism is the treatment of $\bm{\alpha}$ as fixed
-  hyperparameters instead of random. A Bayesian could argue that endowment of 
-  $\bm{\alpha}$ with a hyperprior results in propagation of 
+  A possible criticism of the NIG model is the treatment of $\bm{\alpha}$ as 
+  fixed hyperparameters instead of random. A Bayesian could argue that endowment
+  of $\bm{\alpha}$ with a hyperprior results in propagation of 
   uncertainty about $\bm{\alpha}$ and as a result improved regression parameter 
@@
 
-	A possible direction of future research to attempt
-	to lower PMSE of the NIG model is the introduction of a cross-validated global
-	variance component in the prior of the $\bm{\beta}_d$. If we add the 
-	additional constraint that the drug and feature-specific variance components
-	average to one, this cross-validated variance component may be interpreted as
-	the overall level of shrinkage. This overall shrinkage level is then tailored 
-	to lower PMSE, while the feature- and drug-specific variance component 
-	include the external information. Other directions of future research are
-	applications of the NIG
-	model to different data types. One possibility is to apply the model in an 
-	eQTL study, 
-	where gene expressions are regressed on SNPs. Several interesting external
+	Possible directions of future research are applications of the NIG model to 
+	different data types. Suitable applications are eQTL studies, 
+	in which gene expressions are regressed on SNPs. Several interesting external
 	covariates are available, both on the genes as well as the SNPs. An example of
@@
 	Supplementary Material is available online from 
-	\url{https://github.com/magnusmunch/cambridge}.
+	\url{https://github.com/magnusmunch/NIG}.
 
@@
 	All results and documents may be recreated from 
-	\url{https://github.com/magnusmunch/cambridge}. All timings were obtained
-	on a 2016 MacBook Pro with a 2 GHz Dual-Core Intel Core i5 processor and
-	8 GB of 1867 MHz LPDDR3 memory, running macOS  10.15.1.
+	\url{https://github.com/magnusmunch/NIG}.
 

### 2020-04-14 code/figures.R

@@ -259,48 +259,6 @@ legend("topright", lty=lty, col=col, title="Permuted rows",
        legend=paste0(round(100*c(0, fracs), 0), "%"))
 par(opar)
 
-# ---- analysis_gdsc_res1 ----
-suppressWarnings(suppressMessages(library(sp)))
-res <- read.table("results/analysis_gdsc_res1.txt", row.names=NULL)
-temp <- res[, 1]
-res <- as.matrix(res[, -1])
-rownames(res) <- temp
-pmse <- res[substr(rownames(res), 1, 4)=="pmse", ]
-
-ylim1 <- range(1 - c(res[, "semnig1"], res[, "ridge1"]))
-ylim2 <- range(1 - c(res[, "ebridge1"], res[, "ridge1"]))
-
-col <- bpy.colors(3, cutoff.tail=0.2)
-
-opar <- par(no.readonly=TRUE)
-par(mar=opar$mar*c(1, 1.3, 1, 1))
-layout(matrix(rep(c(1:2), each=2), nrow=2, ncol=2, byrow=TRUE))
-plot(1 - pmse[, "ridge1"], xlab="Drug", 
-     ylab="MSE reduction", main="(a)",
-     cex=0.5, pch=16, ylim=ylim1, col=col[1])
-points(1 - pmse[, "semnig1"], cex=0.5, pch=16, col=col[2])
-abline(v=order(abs(pmse[, "ridge1"] - pmse[, "semnig1"]), 
-               decreasing=TRUE)[c(1:5)], lty=2, 
-       col=col[as.numeric((pmse[, "ridge1"] - pmse[, "semnig1"] > 0) + 1)[
-         order(abs(pmse[, "ridge1"] - pmse[, "semnig1"]), 
-               decreasing=TRUE)[c(1:5)]]])
-abline(h=0, lty=3)
-legend("topleft", c("ridge + CV", "NIG"), pch=16, col=col[c(1, 2)])
-
-plot(1 - pmse[, "ridge1"], xlab="Drug", 
-     ylab="MSE reduction", main="(b)",
-     cex=0.5, pch=16, ylim=ylim2, col=col[1])
-points(1 - pmse[, "ebridge1"], cex=0.5, pch=16, col=col[3])
-abline(v=order(abs(pmse[, "ridge1"] - pmse[, "ebridge1"]), 
-               decreasing=TRUE)[c(1:5)], lty=2, 
[…]

### 2020-04-14 rpackage/R/RcppExports.R

@@ -1,27 +1,27 @@
 # Generated by using Rcpp::compileAttributes() -> do not edit by hand
 # Generator token: 10BE3573-1514-4C36-9D1C-5A225CD40393
 
-.f.optim.mat <- function(alpha, lambda, nu, zeta, Cmat, Z, n, p, D, idsel, G, H, y, x, yty) {
-    .Call(`_cambridge_f_optim_mat`, alpha, lambda, nu, zeta, Cmat, Z, n, p, D, idsel, G, H, y, x, yty)
+.f.optim.mat <- function(alpha, lambda, nu, zeta, Cmat, Z, n, p, D, idsel, G, H, y, x, yty, Zpres) {
+    .Call(`_NIG_f_optim_mat`, alpha, lambda, nu, zeta, Cmat, Z, n, p, D, idsel, G, H, y, x, yty, Zpres)
 }
 
-.f.optim.list <- function(alpha, lambda, nu, zeta, Cmat, Z, n, p, D, G, H, y, x, yty) {
-    .Call(`_cambridge_f_optim_list`, alpha, lambda, nu, zeta, Cmat, Z, n, p, D, G, H, y, x, yty)
+.f.optim.list <- function(alpha, lambda, nu, zeta, Cmat, Z, n, p, D, G, H, y, x, yty, Zpres) {
+    .Call(`_NIG_f_optim_list`, alpha, lambda, nu, zeta, Cmat, Z, n, p, D, G, H, y, x, yty, Zpres)
 }
 
 .Sigma.unp <- function(aold, bold, xu, xr, u, r) {
-    .Call(`_cambridge_Sigma_unp`, aold, bold, xu, xr, u, r)
+    .Call(`_NIG_Sigma_unp`, aold, bold, xu, xr, u, r)
 }
 
 .Sigma <- function(aold, bold, x) {
-    .Call(`_cambridge_Sigma`, aold, bold, x)
+    .Call(`_NIG_Sigma`, aold, bold, x)
 }
 
 .aux.var.unp <- function(aold, bold, y, xu, xr, u, r) {
-    .Call(`_cambridge_aux_var_unp`, aold, bold, y, xu, xr, u, r)
+    .Call(`_NIG_aux_var_unp`, aold, bold, y, xu, xr, u, r)
 }
 
 .aux.var <- function(aold, bold, y, x, ytx) {
-    .Call(`_cambridge_aux_var`, aold, bold, y, x, ytx)
[…]

### 2020-04-14 rpackage/src/RcppExports.cpp

@@ -8,8 +8,8 @@
 using namespace Rcpp;
 
 // f_optim_mat
-double f_optim_mat(arma::vec alpha, arma::vec lambda, double nu, double zeta, arma::mat Cmat, arma::mat Z, arma::vec n, int p, int D, List idsel, int G, int H, List y, arma::mat x, arma::vec yty);
-RcppExport SEXP _cambridge_f_optim_mat(SEXP alphaSEXP, SEXP lambdaSEXP, SEXP nuSEXP, SEXP zetaSEXP, SEXP CmatSEXP, SEXP ZSEXP, SEXP nSEXP, SEXP pSEXP, SEXP DSEXP, SEXP idselSEXP, SEXP GSEXP, SEXP HSEXP, SEXP ySEXP, SEXP xSEXP, SEXP ytySEXP) {
+double f_optim_mat(arma::vec alpha, arma::vec lambda, double nu, double zeta, arma::mat Cmat, arma::mat Z, arma::vec n, int p, int D, List idsel, int G, int H, List y, arma::mat x, arma::vec yty, bool Zpres);
+RcppExport SEXP _NIG_f_optim_mat(SEXP alphaSEXP, SEXP lambdaSEXP, SEXP nuSEXP, SEXP zetaSEXP, SEXP CmatSEXP, SEXP ZSEXP, SEXP nSEXP, SEXP pSEXP, SEXP DSEXP, SEXP idselSEXP, SEXP GSEXP, SEXP HSEXP, SEXP ySEXP, SEXP xSEXP, SEXP ytySEXP, SEXP ZpresSEXP) {
 BEGIN_RCPP
     Rcpp::RObject rcpp_result_gen;
     Rcpp::RNGScope rcpp_rngScope_gen;
@@ -28,13 +28,14 @@ BEGIN_RCPP
     Rcpp::traits::input_parameter< List >::type y(ySEXP);
     Rcpp::traits::input_parameter< arma::mat >::type x(xSEXP);
     Rcpp::traits::input_parameter< arma::vec >::type yty(ytySEXP);
-    rcpp_result_gen = Rcpp::wrap(f_optim_mat(alpha, lambda, nu, zeta, Cmat, Z, n, p, D, idsel, G, H, y, x, yty));
+    Rcpp::traits::input_parameter< bool >::type Zpres(ZpresSEXP);
[…]

### 2020-04-15 docs/manuscript1.Rnw

@@
 	where $\mathcal{IG}(\phi, \lambda)$ denotes an inverse Gaussian distribution 
-	with mean $\phi$ and shape $\lambda$.
+	with mean $\phi$ and shape $\lambda > 0$.
 
@@
   relatively flat, as may be reasonably expected in the large $p$, 
-  small $n$ setting. We therefore model the error variance with Jeffrey's 
+  small $n$ setting. We therefore model the error variance with Jeffreys 
   objective prior 
@@
   subjective specification of hyperparameters. In the derivation of our 
-  Jeffrey's prior
+  Jeffreys prior
   for the error variance, we jointly consider an unknown data mean and variance 
   \cite[]{kass_selection_1996}. This
-  joint consideration results in the somewhat unorthodox $1/\sigma^3$ Jeffrey's
+  joint consideration results in the somewhat unorthodox $1/\sigma^3$ Jeffreys
   prior.
@@
 	  \bm{\alpha}^{(l+1)},\bm{\lambda}^{(l+1)} & = \underset{\bm{\alpha},
-	  \bm{\lambda}}{\argmax}\E_{\cdot | \Y} 
+	  \bm{\lambda} > 0}{\argmax}\E_{\cdot | \Y} 
 	  [\log p(\Y, \B, \bm{\Gamma}^2, \bm{\tau}^2, \bsigma^2) | \bm{\alpha}^{(l)},
 	  \bm{\lambda}^{(l)}] \\
-	  & = \underset{\bm{\alpha},\bm{\lambda}}{\argmax} 
+	  & = \underset{\bm{\alpha},\bm{\lambda} > 0}{\argmax} 
 	  \E_{\cdot | \Y} [\log \pi (\bm{\Gamma}^2) | \bm{\alpha}_{\text{feat}}^{(l)},
@@
 	\bm{\alpha}^{(l+1)},\bm{\lambda}^{(l+1)} = 
-	\underset{\bm{\alpha},\bm{\lambda}}{\argmax}\E_{Q^{(l)}} 
+	\underset{\bm{\alpha},\bm{\lambda} > 0}{\argmax}\E_{Q^{(l)}} 
 	[\log \pi(\bm{\Gamma}^2)| \bm{\alpha}_{\text{feat}}^{(l)}, 
@@
   \end{align*}
-  where 
+  where SM Section \ref{sm-sec:eb} shows that $\bm{\lambda}^{(l+1)} > 0$ and
   \begin{align*}

### 2020-04-15 docs/supplement1.Rnw
commit before push for bSEM analysis 4
changed name of project

--- 2020-03-30 docs/supplement1.Rnw
@@
   sub('\\usepackage[]{color}', '\\usepackage{xcolor}', x, fixed=TRUE)})
+options(knitr.kable.NA="", digits=3)
 @
@@
 	create Figure \ref{md-fig:dens_kappa}c the following settings were
-	used for rate $\lambda$: $\lambda=1$ (solid), $\lambda=0.2$ (dashed), and 
-	$\lambda=0.1$ (dotted).
+	used for exponential rate $\lambda$: $\lambda=1$ (solid), $\lambda=0.2$ 
+	(dashed), and $\lambda=0.1$ (dotted).
 
@@
 	We parametrise the data with a Gaussian distribution with unknown mean and 
-	variance, i.e., $y_i \sim \mathcal{N} (\mu, \sigma^2)$. This gives us the
+	variance, i.e., $y_i \sim \mathcal{N} (\mu, \sigma^2)$. This results in the
 	following Hessian (matrix of second derivatives) of the log density:
@@
 	$$
-	Jeffrey's prior is now found as:
+	Jeffreys prior \cite[]{jeffreys_invariant_1946} is now found as:
 	$$
@@
   The evidence lower bound that is maximised during the iterations is: 
-  $\text{ELBO}^{(l)}=\text{ELBO}_d^{(l)}$, with:
+  $\text{ELBO}^{(l)}=\sum_{d=1}^D \text{ELBO}_d^{(l)}$, with:
   \begin{align*}
@@
   Both (\ref{eq:complexity1}) and (\ref{eq:complexity2}) are operations of
-  complexity $\mathcal{O}(pn^2)$.
+  complexity $\mathcal{O}(pn^2)$ instead of $\mathcal{O}(p^3)$ for naive
+  calculation.
 
@@
   \end{align*} 
-  The ratio $K_0(x)/K_1(x)$ is well-behaved, so that it may be computed
-  without numerical issues.
+  The ratio $K_0(x)/K_1(x)$ is well-behaved, so that an 
+  arbitrary ratio $K_{\alpha - 1}(x)/K_{\alpha}(x)$ may be calculated as
+  a sequence of numerically stable scalar sums, products and inverses.
 
@@
 
-  Next, we note that through Jensen's inequality 
+  We note that through Jensen's inequality 
   $e_{jd}^{-1}=\E(\gamma_{jd}^{2})^{-1} \leq \E(\gamma_{jd}^{-2})=b_{jd}$,
@@
   Similar reasoning holds for the drug scale parameter $\lambda_{\text{drug}}$.
-
-  \section{Estimation of alternative prior}\label{sec:altprior}
-  The hyperparameters $\bm{\alpha} = \begin{bmatrix} \bm{\alpha}_{\text{feat}}
-  \tr & \bm{\alpha}_{\text{drug}} \tr \end{bmatrix} \tr$ in model 
-  (\ref{md-eq:altprior}) are estimated by empirical Bayes maximisation of
-  the marginal log likelihood:
-  \begin{subequations}\label{eq:altmll}
-    \begin{align}
-      \log p (\bm{\alpha} | \mathbf{Y}) & = \sum_{d=1}^D
-      \log \[ \int_{\bm{\beta}_d} \int_{\sigma^2_d} \mathcal{L} 
-      (\mathbf{y}_d | \bm{\beta}_d, \sigma^2_d ) \pi(\bm{\beta}_d | \sigma^2_d) 
-      \pi(\sigma^2_d) d\bm{\beta}_d d\sigma_d^2 \] \\
-      & \propto - \sum_{d=1}^D \bigg[ \log \left|\I + \mathbf{X} 
-      \diag \[ \exp\( \tilde{\mathbf{C}}_d \bm{\alpha} \) \] \mathbf{X} \tr 
-      \right| \\
-      & \,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\, + (n + 1) \log \( \mathbf{y}_d \tr 
-      \left\{ \I + \mathbf{X} 
-      \diag \[ \exp\( \tilde{\mathbf{C}}_d \bm{\alpha} \) \] \mathbf{X} 
-      \tr\right\}^{-1} \mathbf{y}_d \) \bigg],
-    \end{align}
-  \end{subequations}
-  where $\tilde{\mathbf{C}}_d = \begin{bmatrix} \mathbf{C}_d & 
-  \mathbf{1}_{p \times 1} \cdot \mathbf{z}_d \tr \end{bmatrix}$, and
-  $\mathbf{C}_d = \begin{bmatrix} \mathbf{c}_{1d} & \cdots & 
-  \mathbf{c}_{pd} \end{bmatrix} \tr$.
-  The first term in the sum of (\ref{eq:altmll}) is concave in 
-  $\bm{\alpha}$, while the second term is convex in $\bm{\alpha}$. This renders
-  the marginal log likelihood a sum of non-concave (as well as non-convex) 
-  functions. General purpose optimisers, such as \texttt{optim} in \texttt{R} 
-  are generally not well suited to maximise non-concave functions. Maximisation
-  is therefore done through a minorise-maximisation (MM) algorithm, 
-  similar to \cite{zeng_incorporating_2020}.
-
-  The MM algorithm seeks to minorise the marginal log likelihood,
-  $\log p(\bm{\alpha} | \mathbf{Y})$, at current
-  parameter estimate $\bm{\alpha}^{(t)}$ with function 
-  $g(\bm{\alpha} | \bm{\alpha}^{(t)})$, such that
-  $\forall \bm{\alpha}: g(\bm{\alpha} | \bm{\alpha}^{(t)}) \leq 
-  \log p(\bm{\alpha} | \mathbf{Y})$ 
-  and $g(\bm{\alpha}^{(t)} | \bm{\alpha}^{(t)}) = 
-  \log p(\bm{\alpha} | \mathbf{Y})$. Then, the sequence
-  $$
-  \bm{\alpha}^{(t + 1)} = \underset{\bm{\alpha}}{\argmax} 
-  g(\bm{\alpha} | \bm{\alpha}^{(t+1)})
-  $$
-  converges to a local maximum of $f(\bm{\alpha})$. Generally, the minorisation
-  function is chosen such that it is concave and easy to maximise. In our case,
-  a suitable minorisation is:
-  \begin{subequations}\label{eq:minor}
-    \begin{align}
-      g(\bm{\alpha} | \bm{\alpha}^{(t+1)}) = -\sum_{d=1}^D \bigg[
-      & \diag \( \X \tr \mathbf{A}_d^{(t)} \X \) \tr \exp (\tilde{\mathbf{C}}_d
-      \bm{\alpha}) \\
-      & + \frac{n + 1}{\mathbf{y}_d \tr \mathbf{A}_d^{(t)} \mathbf{y}_d} 
-      \mathbf{y}_d \tr \left\{ \I + \mathbf{X} \diag \[ \exp\( 
-      \tilde{\mathbf{C}}_d \bm{\alpha} \) \] \mathbf{X} \tr \right\}^{-1} 
-      \mathbf{y}_d\bigg],
-    \end{align}
-  \end{subequations}
-  where $\mathbf{A}_d^{(t)} = \left\{ \I + \mathbf{X} \diag \[ \exp\( 
-  \tilde{\mathbf{C}}_d \bm{\alpha}^{(t)} \) \] \mathbf{X} \tr \right\}^{-1}$.
-  Note that (\ref{eq:minor}) is concave. However, it is slow to optimise. 
-  In \cite{zeng_incorporating_2020}, another minorization is proposed for
-  the second term:
-  $$
-  \mathbf{y}_d \tr \left\{ \I + \mathbf{X} \diag \[ \exp\( 
-  \tilde{\mathbf{C}}_d \bm{\alpha} \) \] \mathbf{X} \tr \right\}^{-1} 
-  \mathbf{y}_d = \underset{\delta}{\max} 
-  - ||\mathbf{y}_d - \mathbf{X} \delta ||_2^2 - 
-  \delta \tr \diag \[ \exp\(
-  -\tilde{\mathbf{C}}_d \bm{\alpha} \) \] \delta.
-  $$
-  The extra parameter $\delta$ is estimated by any standard ridge regression
-  package, such as \texttt{glmnet} in \texttt{R}.
 
@@
       \sim \mathcal{GIG}
-      \(-1, \lambda_{\text{feat}}/ \phi_{jd}^{2}, \delta_{jd}\), \nonumber \\
+      \(-1, \lambda_{\text{feat}}/ \phi_{jd}^{2}, \delta_{jd}\), \\
       \text{with } & \delta_{jd} = \sigma_d^{-2} \tau_d^{-2} \beta_{jd}^2 + 
@@
       \(-\frac{p+1}{2}, \lambda_{\text{drug}}/ \chi_{d}^{2}, \eta_d\), 
-      \nonumber \\
+      \\
       \text{with } & \eta_d = \sigma_d^{-2} \sum_{j=1}^p \gamma_{jd}^{-2}
-      \beta_{jd}^2 + \lambda_{\text{drug}}, \nonumber \\
+      \beta_{jd}^2 + \lambda_{\text{drug}}, \\
       \sigma_d^2 | \bm{\beta}_d, \bgamma_d^2, \tau_d^2, \y & \sim \Gamma^{-1} 
-      \(\frac{n + p + 1}{2}, \zeta_d\), \nonumber \\
+      \(\frac{n + p + 1}{2}, \zeta_d\), \\
       \text{with } & \zeta_d = \frac{1}{2} \[\mathbf{y}_d \tr \mathbf{y}_d -
@@
       \tau^{-2}_d \bm{\beta}_d \tr \diag(\gamma_{jd}^{-2}) \bm{\beta}_d) \]. 
-      \nonumber
     \end{align}
@@
 
-  Sampling from (\ref{eq:gibbssampler}) either requires specifying 
+  Sampling from (\ref{eq:gibbssampler}) either requires estimating 
+  (as in \ref{md-sec:empiricalbayes}) or specifying 
   hyperparameters
@@
   and extra layer of hyperpriors. In general, specifying them requires 
-  rarely available, detailed subject knowledge, so we resort to endowing them
-  with possibly flat, improper (as in \cite{upadhyay_bayesian_1996})
-  hyperpriors:
+  rarely available, detailed subject knowledge, so we may resort to endowing 
+  them with hyperpriors (as in \cite{upadhyay_bayesian_1996}):
   \begin{align*}
@@
   $\kappa_{\text{feat}}$, $\xi_{\text{feat}}$, $\kappa_{\text{drug}}$, and
-  $\xi_{\text{drug}}$. Improper priors occur if we let 
+  $\xi_{\text{drug}}$. Improper, flat priors occur if we let 
   $\nu_{\text{feat}}, \nu_{\text{drug}} \to \infty$ and 
@@
     \[\mathbf{Z} \bm{\alpha}_{\text{drug}} - 
-    \diag(\tau_{d}^{-2}) \mathbf{1}_{D \times 1}\] + \xi_{\text{drug}}.
+    \diag(\tau_{d}^{-2}) \mathbf{1}_{D \times 1}\] + \xi_{\text{drug}},
   \end{align*}
+  which may be sampled together with (\ref{eq:gibbssampler}).
+
+  \section{Multivariate Bayesian extension of \texttt{xtune}}\label{sec:mxtune}
+  Consider the following simpler alternative prior to 
+  equation (\ref{md-eq:prior}) in the MD:
+  \begin{subequations}\label{eq:altprior}
+		\begin{align}
+		  \beta_{jd} | \gamma^2_{jd}, \tau_d^2, \sigma^2_d & \sim \mathcal{N}_{p} 
+		  (0, \gamma_{jd}^2 \tau_d^2 \sigma_d^2), \\
+		  \log \gamma_{jd}^{2} & = \mathbf{c}_{jd} \tr \bm{\alpha}_{\text{feat}}, \\
+		  \log \tau_d^{2} & = \mathbf{z}_{d} \tr \bm{\alpha}_{\text{drug}}\\
+		  \sigma_d^{2} & \sim \Gamma^{-1} (\nu, \zeta).
+		\end{align}
+	\end{subequations}
+	This model is a Bayesian, multivariate extension of 
+	\cite{zeng_incorporating_2020}. The model does 
+	not propagate uncertainty in the variance components $\gamma_{jd}^{2}$
+	and $\tau_d^{2}$, but directly models the variance components as an
+	exponentiated linear function of the external covariates. The linear functions
+	are modelled on the log-scale to ensure non-negative variance components.
+	Additionally, it allows for more flexibility in the error variance
+	prior model. Note that in the limit $\zeta \to 0$ and $\nu=1/2$, the error
+	variance follows the Jeffreys prior from (\ref{md-eq:prior}) in the MD. 
+	Throughout the
+	following, whenever $\zeta$ and $\nu$ are not specified, we use $\zeta \to 0$ 
+	and $\nu=1/2$.
+
+  The hyperparameters $\bm{\alpha} = \begin{bmatrix} \bm{\alpha}_{\text{feat}}
+  \tr & \bm{\alpha}_{\text{drug}} \tr \end{bmatrix} \tr$ in model 
+  (\ref{eq:altprior}) are estimated by empirical Bayes maximisation of
+  the marginal log likelihood:
+  \begin{subequations}\label{eq:altmll}
+    \begin{align}
+      \log p (\bm{\alpha} | \mathbf{Y}) & = \sum_{d=1}^D
+      \log \[ \int_{\bm{\beta}_d} \int_{\sigma^2_d} \mathcal{L} 
+      (\mathbf{y}_d | \bm{\beta}_d, \sigma^2_d ) \pi(\bm{\beta}_d | \sigma^2_d) 
+      \pi(\sigma^2_d) d\bm{\beta}_d d\sigma_d^2 \] \\
+      & \propto - \sum_{d=1}^D \bigg[ \log \left|\I + \mathbf{X} 
+      \diag \[ \exp\( \tilde{\mathbf{C}}_d \bm{\alpha} \) \] \mathbf{X} \tr 
+      \right| \\
+      & \,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\, + (n + 1) \log \( \mathbf{y}_d \tr 
+      \left\{ \I + \mathbf{X} 
+      \diag \[ \exp\( \tilde{\mathbf{C}}_d \bm{\alpha} \) \] \mathbf{X} 
+      \tr\right\}^{-1} \mathbf{y}_d \) \bigg],
+    \end{align}
+  \end{subequations}
+  where $\tilde{\mathbf{C}}_d = \begin{bmatrix} \mathbf{C}_d & 
+  \mathbf{1}_{p \times 1} \cdot \mathbf{z}_d \tr \end{bmatrix}$, and
+  $\mathbf{C}_d = \begin{bmatrix} \mathbf{c}_{1d} & \cdots & 
+  \mathbf{c}_{pd} \end{bmatrix} \tr$.
+  The first term in the sum of (\ref{eq:altmll}) is concave in 
+  $\bm{\alpha}$, while the second term is convex in $\bm{\alpha}$. This renders
+  the marginal log likelihood a sum of non-concave (as well as non-convex) 
+  functions. General purpose optimisers, such as \texttt{optim} in \texttt{R} 
+  are generally not well suited to maximise non-concave functions. Maximisation
+  is therefore done through a minorise-maximisation (MM) algorithm, 
+  similar to \cite{zeng_incorporating_2020}.
+
+  The MM algorithm seeks to minorise the marginal log likelihood,
+  $\log p(\bm{\alpha} | \mathbf{Y})$, at current
+  parameter estimate $\bm{\alpha}^{(l)}$ with function 
+  $g(\bm{\alpha} | \bm{\alpha}^{(l)})$, such that
+  $\forall \bm{\alpha}: g(\bm{\alpha} | \bm{\alpha}^{(l)}) \leq 
+  \log p(\bm{\alpha} | \mathbf{Y})$ 
+  and $g(\bm{\alpha}^{(l)} | \bm{\alpha}^{(l)}) = 
+  \log p(\bm{\alpha}^{(l)} | \mathbf{Y})$. Then, the sequence
+  $$
+  \bm{\alpha}^{(l + 1)} = \underset{\bm{\alpha}}{\argmax} 
+  g(\bm{\alpha} | \bm{\alpha}^{(l)})
+  $$
+  converges to a local maximum of $\log p(\bm{\alpha} | \mathbf{Y})$. 
+  Generally, the minorisation
+  function is chosen such that it is concave and easy to maximise. In our case,
+  a suitable minorisation is:
+  \begin{subequations}\label{eq:minor}
+    \begin{align}
+      g(\bm{\alpha} | \bm{\alpha}^{(l)}) = -\sum_{d=1}^D \bigg[
+      & \diag \( \X \tr \mathbf{A}_d^{(l)} \X \) \tr \exp (\tilde{\mathbf{C}}_d
+      \bm{\alpha}) \\
+      & + \frac{n + 1}{\mathbf{y}_d \tr \mathbf{A}_d^{(l)} \mathbf{y}_d} 
+      \mathbf{y}_d \tr \left\{ \I + \mathbf{X} \diag \[ \exp\( 
+      \tilde{\mathbf{C}}_d \bm{\alpha} \) \] \mathbf{X} \tr \right\}^{-1} 
+      \mathbf{y}_d\bigg],
+    \end{align}
+  \end{subequations}
+  where $\mathbf{A}_d^{(l)} = \left\{ \I + \mathbf{X} \diag \[ \exp\( 
+  \tilde{\mathbf{C}}_d \bm{\alpha}^{(l)} \) \] \mathbf{X} \tr \right\}^{-1}$.
+  Note that (\ref{eq:minor}) is concave. However, it is slow to optimise. 
+  In \cite{zeng_incorporating_2020}, another minorization is proposed for
+  the second term:
+  $$
+  \mathbf{y}_d \tr \left\{ \I + \mathbf{X} \diag \[ \exp\( 
+  \tilde{\mathbf{C}}_d \bm{\alpha}^{(l)} \) \] \mathbf{X} \tr \right\}^{-1} 
+  \mathbf{y}_d = \underset{\bm{\delta}_d}{\max} 
+  - ||\mathbf{y}_d - \mathbf{X} \bm{\delta}_d ||_2^2 - 
+  \bm{\delta}_d \tr \diag \[ \exp\(
+  -\tilde{\mathbf{C}}_d \bm{\alpha}^{(l)} \) \] \bm{\delta}_d.
+  $$
+  The extra parameters $\bm{\delta}_d$ are estimated per drug and by any 
+  standard ridge regression package, such as \texttt{glmnet} in \texttt{R}.
+
+  \section{Extra simulation results}\label{sec:simulations}
+  The lasso and ridge models were fit separately per drug, using the \texttt{R}
+	package \texttt{glmnet} \cite[]{friedman_regularization_2010} with cross
+	validated penalty parameters. The \texttt{xtune} model is fit with the 
+	corresponding \texttt{R} package \cite[]{ni_bayesian_2019}.
+
+<<simulation_gdsc>>=
+# simulation 1
+res <- read.table("results/simulations_gdsc_res1.txt", row.names=NULL)
+temp <- res[, 1]
+res <- as.matrix(res[, -1])
+rownames(res) <- temp
+
+tabm <- aggregate(res, by=list(substr(rownames(res), 1, 5)), FUN="mean")[, -6]
+tabm <- t(tabm[tabm$Group.1 %in% c("emse.", "emseh", "emsel", "pmse."), -1])
+
+D <- nreps <- 100
+tabsd <- sapply(c("emse.", "emseh", "emsel", "pmse."), function(s) {
+  apply(aggregate(res[substr(rownames(res), 1, 5)==s, ], 
+                  list(rep(1:nreps, each=D)), mean), 2, sd)})[-c(1, 6), ]
+
+tab1 <- sapply(1:ncol(tabm), function(s) {
+  id <- which(tabm[, s]==min(tabm[, s], na.rm=TRUE))
+  out <- paste0(round(tabm[, s], 3), " (", round(tabsd[, s], 3), ")")
+  out[id] <- paste0("\\textbf{", out[id], "}")
+  out})
+colnames(tab1) <- c("$\\text{EMSE}_{\\text{total}}$", 
+                    "$\\text{EMSE}_{\\text{top}}$", 
+                    "$\\text{EMSE}_{\\text{bottom}}$", "$\\text{PMSE}$")     
+rownames(tab1) <- c("NIG$_{\\text{f}}^-$", 
+                     "NIG$_{\\text{f}}$", "ridge", "lasso", "mxtune")
+
+# simulation 2
+res <- read.table("results/simulations_gdsc_res2.txt", row.names=NULL)
+temp <- res[, 1]
+res <- as.matrix(res[, -1])
+rownames(res) <- temp
+
+tabm <- aggregate(res, by=list(substr(rownames(res), 1, 5)), FUN="mean")[, -6]
+tabm <- t(tabm[tabm$Group.1 %in% c("emse.", "emseh", "emsel", "pmse."), -1])
+
+D <- nreps <- 100
+tabsd <- sapply(c("emse.", "emseh", "emsel", "pmse."), function(s) {
+  apply(aggregate(res[substr(rownames(res), 1, 5)==s, ], 
+                  list(rep(1:nreps, each=D)), mean), 2, sd)})[-c(1, 6), ]
+
+tab2 <- sapply(1:ncol(tabm), function(s) {
+  id <- which(tabm[, s]==min(tabm[, s], na.rm=TRUE))
+  out <- paste0(round(tabm[, s], 3), " (", round(tabsd[, s], 3), ")")
+  out[id] <- paste0("\\textbf{", out[id], "}")
+  out})
+colnames(tab2) <- c("$\\text{EMSE}_{\\text{total}}$", 
+                    "$\\text{EMSE}_{\\text{top}}$", 
+                    "$\\text{EMSE}_{\\text{bottom}}$", "$\\text{PMSE}$")     
+rownames(tab2) <- c("NIG$_{\\text{d}}^-$", 
+                     "NIG$_{\\text{d}}$", "ridge", "lasso")
+
+# simulation 3
+res <- read.table("results/simulations_gdsc_res3.txt", row.names=NULL)
+temp <- res[, 1]
+res <- as.matrix(res[, -1])
+rownames(res) <- temp
+
+tabm <- tabm3 <- aggregate(res, by=list(substr(rownames(res), 1, 5)), 
+                           FUN="mean")[, -6]
+tabm <- t(tabm[tabm$Group.1 %in% c("emse.", "emseh", "emsel", "pmse."), -1])
+
+D <- nreps <- 100
+tabsd <- tabsd3 <- sapply(c("emse.", "emseh", "emsel", "pmse."), function(s) {
+  apply(aggregate(res[substr(rownames(res), 1, 5)==s, ], 
+                  list(rep(1:nreps, each=D)), mean), 2, sd)})[-c(1, 6), ]
+
+tab3 <- sapply(1:ncol(tabm), function(s) {
+  id <- which(tabm[, s]==min(tabm[, s], na.rm=TRUE))
+  out <- paste0(round(tabm[, s], 3), " (", round(tabsd[, s], 3), ")")
+  out[id] <- paste0("\\textbf{", out[id], "}")
+  out})
+colnames(tab3) <- c("$\\text{EMSE}_{\\text{total}}$", 
+                    "$\\text{EMSE}_{\\text{top}}$", 
+                    "$\\text{EMSE}_{\\text{bottom}}$", "$\\text{PMSE}$")     
+rownames(tab3) <- c("NIG$_{\\text{d}}^-$", 
+                     "NIG$_{\\text{d}}$", "ridge", "lasso", "mxtune")
+@	
+	Tables \ref{tab:simulations_gdsc_tab1}-\ref{tab:simulation_gdsc_tab3} 
+	display averaged estimation mean squared error (EMSE), and prediction mean
+	squared error (PMSE), calculated on independent test data, for simulation 
+	Scenarios 1-4. PMSE is calculated as in Section \ref{md-sec:gdsc} of the MD, 
+	while EMSE is calculated as: $\text{EMSE} = D^{-1} p^{-1} \sum_{d=1}^D 
+  \sum_{j=1}^p (\beta_{jd} - \hat{\beta}_{jd})^2$, with
+  $\hat{\beta}_{jd}$ the estimator for $\beta_{jd}$. In the NIG
+  models, that provide full posteriors, the posterior mean 
+  $\E(\beta_{jd} | \mathbf{y}_d)$ is used as point estimate.
+	EMSE is further split into the contribution of the
+	bottom 90\% of the $\beta_{jd}$ in terms of size (in absolute value) and the 
+	top 10\% of the $\beta_{jd}$ in size.
+
+%   
+%   
+%   $\lambda_{\text{feat}}$ is overestimated with median estimate 
+% 	\Sexpr{round(est.lambdaf[2], 2)} (the
+% 	true value is 1). The overestimated $\lambda_{\text{feat}}$ translates 
+% 	into underestimated $\gamma_{jd}^2$ variances. 
+% 	
+% 	
+% 	
+<<simulations_gdsc_tab1>>=
+kableExtra::kable_styling(knitr::kable(
+  tab1, align="r",
+  caption="Mean EMSE (split for bottom 10\\% and top 90\\% of $|\\hat{\\beta}_{jd}|$) and PMSE for simulation Scenario 1, estimated on test data (lowest in bold).", 
+  format="latex", booktabs=TRUE, escape=FALSE), 
+  latex_options=c("HOLD_position"))
+@
+
+<<simulations_gdsc_tab2>>=
+kableExtra::kable_styling(knitr::kable(
+  tab2, align="r",
+  caption="Mean EMSE (split for bottom 10\\% and top 90\\% of $|\\hat{\\beta}_{jd}|$) and PMSE for simulation Scenario 2, estimated on test data (lowest in bold).", 
+  format="latex", booktabs=TRUE, escape=FALSE), 
+  latex_options=c("HOLD_position"))
+@
+
+<<simulations_gdsc_tab3>>=
+kableExtra::kable_styling(knitr::kable(
+  tab3, align="r",
+  caption="Mean EMSE (split for bottom 10\\% and top 90\\% of $|\\hat{\\beta}_{jd}|$) and PMSE for simulation Scenario 2, estimated on test data (lowest in bold).", 
+  format="latex", booktabs=TRUE, escape=FALSE), 
+  latex_options=c("HOLD_position"))
+@
+
+<<simulations_gdsc_pmse4, fig.cap="Simulation results for Scenario 4: Average PMSE against the proportion of permuted rows in the external features.", out.width="100%", fig.asp=1>>= 
+@
+
+	Focussing on EMSE, we see that the NIG$_{\text{f}}$ model,
+	that includes the external covariates, performs best, both in terms of 
+	estimation and prediction. The lower MSEs confirm that the model that
+	includes the external data (NIG$_{\text{f}}$) learns the underlying structure 
+	in the data better than model NIG$_{\text{f}}^-$ that does not include the
+	external data. Furthermore, we see that estimation of the larger $\beta_{jd}$ 
+	is where ridge loses the most compared to the NIG and lasso models. 
+	The heavier tails of the NIG and lasso priors allow for larger $\beta_{jd}$
+	estimates and thus lower EMSE for these $\beta_{jd}$.
+
+%   $\lambda_{\text{feat}}$ is overestimated with median estimate
+% 	\Sexpr{round(est.lambdaf[1], 2)} (the true value is one) and thus 
+% 	again underestimates the 
+% 	prior $\gamma_{jd}^2$ variance.
+% 	
+% 	Median $\lambda_{\text{drug}}=\Sexpr{round(est.lambdad[2], 2)}$ is
+% 	overestimated compared to true value one. 
+% 	Scale $\lambda_{\text{drug}}$ is underestimated with median estimate
+% 	\Sexpr{round(est.lambdad[1], 2)} compared to the true value one.
+% 	
+%   Median performance measures for simulation Scenario 2 are displayed in Table
+% 	\ref{tab:simulations_gdsc_tab2}. Again, we see that NIG$_{\text{d}}$ performs
+% 	best in terms of MSE. However, NIG$_{\text{d}}^-$ performance is the same 
+% 	here, indicating that the external covariates do not add much here.
+% 	Similarly as in Scenario 1, most of the estimation performance gain over ridge
+% 	is through the top 10\% of the $\beta_{jd}$.
+% 	
+%   
+%   Median $\lambda_{\text{feat}}$ and $\lambda_{\text{drug}}$ are
+% 	are over- and underestimated with \Sexpr{round(est.lambdaf[2], 2)}
+% 	and \Sexpr{round(est.lambdad[2], 2)}, respectively, compared to their true
+% 	value 1. 
+% 	
+%   NIG$_{\text{f}+\text{d}}^-$ also over- and
+% 	underestimates 
+% 	$\lambda_{\text{feat}}$ and $\lambda_{\text{drug}}$ with median estimates
+% 	\Sexpr{round(est.lambdaf[1], 2)}
+% 	and \Sexpr{round(est.lambdad[1], 2)}, respectively. 
+% 	
+%   Table 
+% 	\ref{tab:simulations_gdsc_tab3}, with median performance measures, shows
+% 	that NIG$_{\text{f}+\text{d}}$ outperforms the other models in terms of
+% 	prediction and estimation MSE, with again most of the edge over ridge in the
+% 	top 10\% of the $\beta_{jd}$.
+
+<<simulations_gdsc_post5, fig.cap="Simulation results for Scenario 3: Samples from posterior with variational posterior superimposed (red line) for estimated hyperparameters (top row) $\\hat{\\chi}_1$, (second row) $\\hat{\\chi}_2$, (third row) $\\hat{\\chi}_3$, (bottom row) $\\hat{\\chi}_4$, (left column) $\\hat{\\phi}_1$, (second column) $\\hat{\\phi}_2$, (third column) $\\hat{\\phi}_3$, and (last column) $\\hat{\\phi}_4$.", out.width="100%", fig.asp=1>>=
+@
+
+  \section{Extra GDSC data results}\label{sec:gdsc}
+  $$
+  \text{CPO}_{id} = p(y_i | \bm{y}_{-i}) = 
+  $$
+
+<<analysis_gdsc_res2, fig.cap="Simulation results for Scenario 4: Median PMSE against the proportion of permuted rows in the external features.", out.width="50%", fig.asp=1>>= 
+@
 

### 2020-04-16 docs/supplement1.Rnw
commit before pull

@@
                     "$\\text{EMSE}_{\\text{bottom}}$", "$\\text{PMSE}$")     
-rownames(tab3) <- c("NIG$_{\\text{d}}^-$", 
-                     "NIG$_{\\text{d}}$", "ridge", "lasso", "mxtune")
+rownames(tab3) <- c("NIG$_{\\text{f+d}}^-$", 
+                    "NIG$_{\\text{f+d}}$", "ridge", "lasso", "mxtune")
 @	
-	Tables \ref{tab:simulations_gdsc_tab1}-\ref{tab:simulation_gdsc_tab3} 
+	Tables \ref{tab:simulations_gdsc_tab1}-\ref{tab:simulations_gdsc_tab3} 
 	display averaged estimation mean squared error (EMSE), and prediction mean
@@
 	EMSE is further split into the contribution of the
-	bottom 90\% of the $\beta_{jd}$ in terms of size (in absolute value) and the 
+	bottom 10\% of the $\beta_{jd}$ in terms of size (in absolute value) and the 
 	top 10\% of the $\beta_{jd}$ in size.
 
-%   
-%   
-%   $\lambda_{\text{feat}}$ is overestimated with median estimate 
-% 	\Sexpr{round(est.lambdaf[2], 2)} (the
-% 	true value is 1). The overestimated $\lambda_{\text{feat}}$ translates 
-% 	into underestimated $\gamma_{jd}^2$ variances. 
-% 	
-% 	
-% 	
-<<simulations_gdsc_tab1>>=
-kableExtra::kable_styling(knitr::kable(
-  tab1, align="r",
-  caption="Mean EMSE (split for bottom 10\\% and top 90\\% of $|\\hat{\\beta}_{jd}|$) and PMSE for simulation Scenario 1, estimated on test data (lowest in bold).", 
-  format="latex", booktabs=TRUE, escape=FALSE), 
-  latex_options=c("HOLD_position"))
-@
-
-<<simulations_gdsc_tab2>>=
-kableExtra::kable_styling(knitr::kable(
-  tab2, align="r",
-  caption="Mean EMSE (split for bottom 10\\% and top 90\\% of $|\\hat{\\beta}_{jd}|$) and PMSE for simulation Scenario 2, estimated on test data (lowest in bold).", 
-  format="latex", booktabs=TRUE, escape=FALSE), 
-  latex_options=c("HOLD_position"))
-@
-
-<<simulations_gdsc_tab3>>=
-kableExtra::kable_styling(knitr::kable(
-  tab3, align="r",
-  caption="Mean EMSE (split for bottom 10\\% and top 90\\% of $|\\hat{\\beta}_{jd}|$) and PMSE for simulation Scenario 2, estimated on test data (lowest in bold).", 
-  format="latex", booktabs=TRUE, escape=FALSE), 
-  latex_options=c("HOLD_position"))
-@
-
-<<simulations_gdsc_pmse4, fig.cap="Simulation results for Scenario 4: Average PMSE against the proportion of permuted rows in the external features.", out.width="100%", fig.asp=1>>= 
-@
-
-	Focussing on EMSE, we see that the NIG$_{\text{f}}$ model,
+	Focussing on estimation, we see that the NIG$_{\text{f}}$ model,
 	that includes the external covariates, performs best, both in terms of 
@@
 
-%   $\lambda_{\text{feat}}$ is overestimated with median estimate
-% 	\Sexpr{round(est.lambdaf[1], 2)} (the true value is one) and thus 
-% 	again underestimates the 
-% 	prior $\gamma_{jd}^2$ variance.
-% 	
-% 	Median $\lambda_{\text{drug}}=\Sexpr{round(est.lambdad[2], 2)}$ is
-% 	overestimated compared to true value one. 
-% 	Scale $\lambda_{\text{drug}}$ is underestimated with median estimate
-% 	\Sexpr{round(est.lambdad[1], 2)} compared to the true value one.
-% 	
-%   Median performance measures for simulation Scenario 2 are displayed in Table
-% 	\ref{tab:simulations_gdsc_tab2}. Again, we see that NIG$_{\text{d}}$ performs
-% 	best in terms of MSE. However, NIG$_{\text{d}}^-$ performance is the same 
-% 	here, indicating that the external covariates do not add much here.
-% 	Similarly as in Scenario 1, most of the estimation performance gain over ridge
-% 	is through the top 10\% of the $\beta_{jd}$.
-% 	
-%   
-%   Median $\lambda_{\text{feat}}$ and $\lambda_{\text{drug}}$ are
-% 	are over- and underestimated with \Sexpr{round(est.lambdaf[2], 2)}
-% 	and \Sexpr{round(est.lambdad[2], 2)}, respectively, compared to their true
-% 	value 1. 
-% 	
-%   NIG$_{\text{f}+\text{d}}^-$ also over- and
-% 	underestimates 
-% 	$\lambda_{\text{feat}}$ and $\lambda_{\text{drug}}$ with median estimates
-% 	\Sexpr{round(est.lambdaf[1], 2)}
-% 	and \Sexpr{round(est.lambdad[1], 2)}, respectively. 
-% 	
-%   Table 
-% 	\ref{tab:simulations_gdsc_tab3}, with median performance measures, shows
-% 	that NIG$_{\text{f}+\text{d}}$ outperforms the other models in terms of
-% 	prediction and estimation MSE, with again most of the edge over ridge in the
-% 	top 10\% of the $\beta_{jd}$.
+<<simulations_gdsc_tab1>>=
+kableExtra::kable_styling(knitr::kable(
+  tab1, align="r",
+  caption="Mean EMSE (separated for bottom and top 10\\% of $|\\hat{\\beta}_{jd}|$) and PMSE for simulation Scenario 1, estimated on test data (lowest in bold).", 
+  format="latex", booktabs=TRUE, escape=FALSE), 
+  latex_options=c("HOLD_position"))
+@
+
+<<simulations_gdsc_tab2>>=
+kableExtra::kable_styling(knitr::kable(
+  tab2, align="r",
+  caption="Mean EMSE (separated for bottom and top 10\\% of $|\\hat{\\beta}_{jd}|$) and PMSE for simulation Scenario 2, estimated on test data (lowest in bold).", 
+  format="latex", booktabs=TRUE, escape=FALSE), 
+  latex_options=c("HOLD_position"))
+@
+
+<<simulations_gdsc_tab3>>=
+kableExtra::kable_styling(knitr::kable(
+  tab3, align="r",
+  caption="Mean EMSE (separated for bottom and top 10\\% of $|\\hat{\\beta}_{jd}|$) and PMSE for simulation Scenario 3, estimated on test data (lowest in bold).", 
+  format="latex", booktabs=TRUE, escape=FALSE), 
+  latex_options=c("HOLD_position"))
+@
+
+<<simulations_gdsc_pmse4, fig.cap="(a) Mean EMSE, separated for (b) bottom and (c) top 10\\% of $|\\hat{\\beta}_{jd}|$, and (d) PMSE for simulation Scenario 3 versus the proportion of permuted  rows in the external data, estimated on test data.", out.width="100%", fig.asp=1>>=
+@
 
@@
 
-<<analysis_gdsc_res2, fig.cap="Simulation results for Scenario 4: Median PMSE against the proportion of permuted rows in the external features.", out.width="50%", fig.asp=1>>= 
+<<analysis_gdsc_res2, fig.cap="Simulation results for Scenario 4: Median PMSE against the proportion of permuted rows in the external features.", out.width="100%", fig.asp=1>>= 
 @

### 2020-04-16 code/simulations_gdsc.R

@@ -776,7 +776,7 @@ control.semnig <- list(conv.post=TRUE, trace=FALSE, epsilon.eb=1e-5,
                        epsilon.vb=1e-3, maxit.eb=2000, maxit.vb=1, 
                        maxit.post=100, maxit.block=0)
 control.stan <- list(verbose=FALSE, show_messages=FALSE, iter=3000, 
-                     warmup=1000)
+                     warmup=1000, chains=1)
 
 ### simulate parameters
 set.seed(2020)
@@ -807,9 +807,13 @@ fit.mcmc1 <- lapply(1:D, function(d) {
                      chi=1/as.numeric(Z[d, ] %*% fit.semnig1$eb$alphad),
                      lambdaf=fit.semnig1$eb$lambdaf, 
                      lambdad=fit.semnig1$eb$lambdad), 
-           iter=control.stan$iter, warmup=control.stan$warmup,
+           iter=control.stan$iter, warmup=control.stan$warmup, 
+           chains=control.stan$chains,
            verbose=control.stan$verbose, 
-           show_messages=control.stan$show_messages)})
+           show_messages=control.stan$show_messages,
+           sample_file=paste0(
+             "results/simulations_gdsc_fit5_mcmc1/simulations_gdsc_fit5_mcmc1_drug", 
+             d, ".txt"))})
 time.mcmc1 <- proc.time()[3] - ct
 ct <- proc.time()[3]
 fit.mcmc2 <- sampling(stanmodels$nig_full,
@@ -818,10 +822,19 @@ fit.mcmc2 <- sampling(stanmodels$nig_full,
                                 y=Reduce("cbind", y), C=Reduce("rbind", C),
                                 Z=Z, nuf=10, nud=10, kappaf=1, 
                                 kappad=1, xif=1, xid=1),
-                      chains=1, iter=2000, warmup=1000)
+                      iter=control.stan$iter, warmup=control.stan$warmup, 
+                      chains=control.stan$chains,
+                      verbose=control.stan$verbose, 
+                      show_messages=control.stan$show_messages,
+                      sample_file="results/simulations_gdsc_fit5_mcmc2.txt")
 time.mcmc2 <- proc.time()[3] - ct
 save(fit.semnig1, fit.mcmc1, fit.mcmc2, 
-     file="results/simulations_gdsc_fit5.2.Rdata")
[…]

### 2020-04-22 docs/manuscript.Rnw
commit before pull

--- 2019-01-14 docs/manuscript.Rnw
@@
+% set options
+<<settings, include=FALSE, echo=FALSE>>=
+opts_knit$set(root.dir="..", base.dir="../figs")
+opts_chunk$set(fig.align='center', fig.path="../figs/", 
+               echo=FALSE, cache=FALSE, message=FALSE, fig.pos='!ht')
+knit_hooks$set(document=function(x) {
+  sub('\\usepackage[]{color}', '\\usepackage{xcolor}', x, fixed=TRUE)})
+options(knitr.kable.NA="")
+@
+
+% read figure code
+<<figures, include=FALSE>>=
+read_chunk("code/figures.R")
+@
+
 \documentclass[a4paper,hidelinks]{article}
-\usepackage{bm,amsmath,amssymb,amsthm,amsfonts,graphics,graphicx,epsfig,rotating,caption,subcaption,natbib,appendix,titlesec,multicol,hyperref,verbatim,bbm,algorithm,algpseudocode,pgfplotstable,threeparttable, booktabs,mathtools,dsfont,parskip}
+\usepackage{bm,amsmath,amssymb,amsthm,amsfonts,graphics,graphicx,epsfig,
+rotating,caption,subcaption,natbib,appendix,titlesec,multicol,hyperref,verbatim,
+bbm,algorithm,algpseudocode,pgfplotstable,threeparttable,booktabs,mathtools,
+dsfont,parskip,xr,tikz}
+\externaldocument[sm-]{supplement}
+\usetikzlibrary{bayesnet}
 
@@
 % settings
-\graphicspath{{/Users/magnusmunch/Documents/OneDrive/PhD/project_cambridge/graphs/}}
+\pgfplotsset{compat=1.16}
+\graphicspath{{../figs/}}
 \setlength{\evensidemargin}{.5cm} \setlength{\oddsidemargin}{.5cm}
 \setlength{\textwidth}{15cm}
-\title{Drug efficacy prediction in cell lines}
+\title{Drug sensitivity prediction with normal inverse Gaussian shrinkage 
+informed by external data}
 \date{\today}
-\author{Magnus M.
-M\"unch$^{1,2}$\footnote{Correspondence to: \href{mailto:m.munch@vumc.nl}{m.munch@vumc.nl}}, Mark A.
-van de Wiel$^{1,3}$, Sylvia Richardson$^{3}$, \\ and Gwena{\"e}l G.
-R.
-Leday$^{3}$}
+\author{Magnus M. M\"unch$^{1,2,3}$\footnote{Correspondence to: 
+\href{mailto:m.munch@amsterdamumc.nl}{m.munch@amsterdamumc.nl}}, Mark A. van de 
+Wiel$^{1,3}$, Sylvia Richardson$^{3}$, \\ and Gwena{\"e}l G. R. Leday$^{3}$}
 
 \begin{document}
+
 	\maketitle
@@
 	\noindent
-	1. Department of Epidemiology \& Biostatistics, Amsterdam Public Health research institute, VU University Medical Center, PO Box 7057, 1007 MB
-	Amsterdam, The Netherlands\\
+	1. Department of Epidemiology \& Biostatistics, Amsterdam UMC, VU University, 
+	PO Box 7057, 1007 MB Amsterdam, The Netherlands \\
 	2. Mathematical Institute, Leiden University, Leiden, The Netherlands \\
-	3. MRC Biostatistics Unit, Cambridge Institute of Public Health, Cambridge, United Kingdom \\
+	3. MRC Biostatistics Unit, Cambridge Institute of Public Health, Cambridge,
+	United Kingdom
 
 	\begin{abstract}
-		{...}
+		{In precision medicine, a common problem is drug sensitivity prediction from
+		cancer tissue cell lines. These types of problems entail modelling 
+		multivariate drug responses on high dimensional molecular feature sets in 
+		typically $>1000$ cell lines. The dimensions of the problem require 
+		specialised models and estimation methods. In addition, external information
+		on both the drugs and the features is often available. We propose to 
+		model the drug responses through a linear regression with shrinkage enforced
+		through a normal
+		inverse Gaussian prior. We let the prior depend on the external information,
+		and estimate the model and external information dependence in an 
+		empirical-variational Bayes framework. We demonstrate the usefullness of
+		this model in both a simulated setting and in the publicly available
+		Genomics of Drug Sensitivity in Cancer data.}
 	\end{abstract}
 
-	\noindent\textbf{Keywords}:
+	\noindent\textbf{Keywords}: Drug sensitivity; Empirical Bayes; 
+	Genomics of Drug Sensitivity in Cancer (GDSC); Variational Bayes
 
 	\noindent\textbf{Software available from}: 
+	\url{https://github.com/magnusmunch/NIG}
 
-	\section{Introduction}
-	- motivation: drug response prediction from different omics data types is promising both for precision medicine and drug target discovery and relates to actual tumor response \cite[]{iorio_landscape_2016}.
+	\section{Introduction}\label{sec:introduction}
+	Recently, promising results in precision medicine 
+	have sparked an interest in cancer drug sensitivity prediction models
+	\cite[]{iorio_landscape_2016}. Typically, these models 
+	predict the drug sensitivity for new patients from a set of molecular 
+	features.
+	Development of such models is often done in well-characterised human cancer
+	tissue cell lines. The current paper presents a novel drug sensitivity
+	prediction model and an application to a real drug
+	sensitivity data set.
 
-	- problem: prediction of drug efficacy in cell lines is difficult (DREAM 7 challenge in \cite{costello_community_2014}).
+	Development of such models from cell lines has proven to be difficult
+	(see e.g., the DREAM 7 challenge in \cite{costello_community_2014}). 
+	Difficulties arise, among others, from the dimensions of the problem. 
+	Typically, the data contains
+	hundreds of drugs, thousands of cell lines, and thousands of molecular 
+	features. An example of a large database of drug responses and molecular
+	features is the Genomics of Drug Sensitivity in Cancer (GDSC) data 
+	\cite[]{yang_genomics_2013}, which we will further 
+	investigate in Section \ref{sec:gdsc}. Other examples of such databases 
+	include
+	the Cancer Cell Line Encyclopedia (CCLE) \cite[]{li_landscape_2019} and 
+	the US National Cancer Institute 60 human tumour cell line anticancer drug
+	screen (NCI60) \cite[]{shoemaker_NCI60_2006}. The dimensions of these data
+	prohibit the estimation of standard regression models and
+	typically require some form of regularisation.
 
-	- example data: introduction (GDSC, CCLE, NCI-60)
+	The GDSC database contains additional information on both the drugs and 
+	molecular features, such as the target 
+	pathways and developmental stages of the drugs. Additional online
+	repositories may provide extra information such as the molecular weight of
+	the compounds or the publication signatures of the molecular features. In some
+	cases, prior knowledge on the drug efficacies may be available,
+	from previous experiments. We propose to include these possibly beneficial 
+	information sources in the estimation of the sensitivity prediction models in 
+	a data-driven manner. More specifically, we estimate a normal 
+	inverse Gaussian model, where the extent of regularisation is estimated by an 
+	adaptive empirical Bayes procedure, guided by the external information.
 
-	- feature information: prior experts, pathways, database summaries, previous experiments, type of feature (somatic mutations, copy number alterations, DNA methylation, and gene expression).
+	We are not the first to work on drug sensitivity prediction models. Reviews on 
+	the topic are \cite{azuaje_computational_2017} and \cite{ali_machine_2019}.
+	\cite{zhao_structured_2019} and \cite{mai_composite_2019} consider a
+	structured penalized multivariate regression approach. 
+	\cite{aben_tandem:_2016} introduce a two-stage penalized regression 
+	model that includes two different types of molecular features.
+	\cite{ammad-ud-din_drug_2016} and \cite{costello_community_2014} tackle the
+	problem through a multiple kernel learning approach. Our solution allows for
+	the adaptive incorporation of the external 
+  information on drugs and features. This is done by
+  pooling information, both across drugs and features. Estimation of the model
+  is through computational feasible variational bayes approximations, while 
+  empirical Bayes estimation of tuning parameters pools information across drugs
+  and features in a data-driven manner.
 
-	- drugs information: molecular target, drug class, etc (ask Oslo).
+	The rest of the paper is structured as follows. In Section 
+	\ref{sec:model} we introduce our model, the estimation of which is detailed
+	in Section \ref{sec:estimation}. Section \ref{sec:simulations} describes a 
+	simulation study that investigates the estimation of hyperparameters by the
+	proposed method. In Section \ref{sec:gdsc} we analyse the
+	GDSC data, and we end with a discussion in Section \ref{sec:discussion} on the 
+	pros and cons of the proposed method.
 
-	- current literature: review in \cite{azuaje_computational_2017}, reference Oslo, TANDEM \cite[]{aben_tandem:_2016}, DeltaNet \cite[]{noh_inferring_2016}, patient-derived tumor cell lines \cite[]{gao_high-throughput_2015}.
+	\section{Model}\label{sec:model}
+	\subsection{Simultaneous equations model}\label{sec:SEM}
+	Let $y_{id}$ be the continuous sensitivity measures 
+	for cell lines $i=1,\dots, n$, and drug $d=1, \dots, D$. 
+	We predict sensitivity from molecular features 
+	$x_{ij}$, $j=1,\dots, p$, collected in 
+	$\x_i = \begin{bmatrix} x_{i1} & \dots & x_{ip} \end{bmatrix} \tr$. 
+	We assume that both covariates and responses have been centred per drug
+	and regress the drug sensitivities on the molecular features:
+	\begin{equation}\label{eq:likelihood}
+	  y_{id} = \x_i \tr \bbeta_d + \epsilon_{id}, 
+	  \text{ with } \epsilon_{id} \sim \mathcal{N}(0, \sigma_d^2),
+	\end{equation}
+	where the $p$-dimensional 
+	$\bbeta_d = \begin{bmatrix} \beta_{1d} & \cdots & \beta_{pd} \end{bmatrix} 
+	\tr$ are the drug-specific omics feature 
+	effects. Note that (\ref{eq:likelihood}) gives rise to a system of 
+	D linear regression equations.
 
-	- our solution: adaptive borrowing of information through empirical Bayes, guided by external feature and drug information.
+	The cell lines used in drug response models are often taken from different
+	tissues. In addition, other clinical covariates might be available. 
+	To obtain unbiased feature effects, one may wish to account for these. 
+	We do so by introducing unpenalized covariates, the $\beta_{jd}$ coefficients 
+	of which are endowed with a flat prior. For the sake of clarity, in the
+	following, such unpenalized covariates are omitted. However, the available
+	software allows for their inclusion.
 
-	\section{Model}
-	\subsection{Simultaneous equations model}\label{sec:SEM}
-	- Gwen can deal with different number of features per equation, might solve missing drug-feature combination problem? Doesn't solve missing response-feature combination problem.
-
-	- check literature for: a) conjugate version, b) similar models: "prior variance decomposition", "structured Bayesian regression"
-
-	We have continuous efficacy measures $y_{id}$ for cell lines $i=1,\dots, n$, from tissues $t=1, \dots, T$, on drugs $d=1,\dots,D$.
-Throughout this paper we assume the $y_{id}$ to be centred per drug and let $\y_d = \begin{bmatrix} y_{1d} & \dots & y_{nd} \end{bmatrix} \tr$.
-We predict efficacy with molecular features $x_{ij}$, $j=1,\dots, p$, collected in $\x_i = \begin{bmatrix} x_{i1} & \dots & x_{ip} \end{bmatrix} \tr$.
-We have an \textit{a priori} partitioning of the omics features into $g=1, \dots, G$ non-overlapping groups, which might be informative for the problem at hand For convenience, we let $g(j)$ denote the group index of molecular feature $j$.
-Furthermore, we have characteristics of the drugs available in the form of `co-data' $\mathbf{C} = \begin{bmatrix} \mathbf{c}_1 & \dots & \mathbf{c}_D \end{bmatrix} \tr$.
-We code the tissue from which the cell lines were taken as a $T$ binary dummy variables
-	$$
-	z_{it} = \begin{cases}
-	1 & \text{if } \text{tissue}_i = t, \\
-	0 & \text{otherwise},
-	\end{cases}
-	$$
-	collected in $\mathbf{z}_i = \begin{bmatrix} z_{i1} & \dots & z_{iT} \end{bmatrix} \tr$.
-
-	We assume that cell lines from the same tissue are related and model this relation through random effects $u_{td}$.
-We let the molecular feature effects $\beta_{jd}$ vary over the drugs and model drug efficacy as a linear combination of these two effects:
-	\begin{subequations}\label{eq:linearmodel}
+	\subsection{Bayesian prior model}\label{sec:prior}
+	We carry out inference by endowing the parameters with the following priors:
+	\begin{subequations}\label{eq:prior}
 		\begin{align}
-		y_{id} & = \beta_{0d} + \sum_{j=1}^p x_{ij} \beta_{jd} + \sum_{t=1}^T z_{it} u_{td} + \epsilon_{id} \\
-		& = \beta_{0d} + \x_i \tr \bbeta_d + \mathbf{z}_i \tr \mathbf{u}_d + \epsilon_{id},\\
-		\text{with } & \epsilon_{id} \sim \mathcal{N}(0, \sigma_d^2),
+		  \beta_{jd} | \gamma^2_{jd}, \tau_d^2, \sigma^2_d & \sim \mathcal{N}_{p} 
+		  (0, \gamma_{jd}^2 \tau_d^2 \sigma_d^2), 
+		  \label{eq:betaprior}\\
+		  \gamma_{jd}^{2} & \sim \mathcal{IG}(\phi_{jd}, \lambda_{\text{feat}}), 
+		  \label{eq:gammaprior}\\
+		  \tau_d^{2} & \sim \mathcal{IG}(\chi_d, \lambda_{\text{drug}}), 
+		  \label{eq:tauprior}\\
+		  \sigma_d^{2} & \sim 1/\sigma_d^{3}, \label{eq:sigmaprior}
 		\end{align}
 	\end{subequations}
-	where $p$-dimensional $\bbeta_d$ and $T$-dimensional $\bm{u}_d$ are the drug-specific omics feature and tissue effect vectors, respectively.
-Note that (\ref{eq:linearmodel}) gives rise to a system of $d=1, \dots, D$ linear simultaneous equations.
+	where $\mathcal{IG}(\phi, \lambda)$ denotes an inverse Gaussian distribution 
+	with mean $\phi$ and shape $\lambda > 0$.
 
-	\subsection{Bayesian prior model}
-	We capture the uncertainty in the parameters through a Bayesian prior model.
-We assign a flat prior to the intercept terms $\beta_{0d}$ and integrate them out.
-The remaining parameters are endowed with the following priors:
-	\begin{subequations}\label{eq:prior}
+	In model (\ref{eq:prior}), $\gamma_{jd}^2$ in (\ref{eq:gammaprior}) denotes a 
+	local variance 
+	component that is supposed to capture local, feature-specific variation in the
+	model parameters $\beta_{jd}$ in (\ref{eq:betaprior}), 
+	while the global variance components $\tau_d^2$ in (\ref{eq:tauprior})
+	capture the drug-specific, general trend in $\bm{\beta}_{d}$. Each drug 
+	response is endowed with a random error variance $\sigma_d^2$, distributed
+	according to (\ref{eq:sigmaprior}).
+
+	Prior distributions of the form
+	(\ref{eq:prior}) are often referred to as global-local shrinkage rules 
+	\cite[]{bernardo_shrink_2011}, due to the multiplicative separation of the 
+	prior variance into a local component $\gamma_{jd}^2$ and a global component
+	$\tau_d^2$. For appropriate local shrinkage in global-local 
+	shrinkage models it is important  to account for 
+  different noise levels $\sigma_d^2$ by scaling the $\beta_{jd}$ variances 
+  accordingly.
+
+	The normal inverse Gaussian (NIG) prior model was introduced in 
+  \cite{barndorff-nielsen_hyperbolic_1978} and since 
+  \cite{barndorff-nielsen_normal_1997} it is routinely applied in mathematical 
+  finance (see, e.g., \cite{kalemanova_normal_2007}). Here we extend it with an
+  additional global variance component $\tau_d^2$. Supplementary Material
+  (SM) Section \ref{sm-sec:prior} contains more details on the NIG prior.
+  To illustrate the effect of the NIG prior on the posterior mean, we consider 
+  the prior reparametrised as in 
+  \cite{carvalho_handling_2009}, i.e., in terms of shrinkage weights 
+  $\kappa_{jd}=1/(1 + \gamma_{jd}^2) \in (0, 1)$.
+  Under the (simplified) normal means model, i.e., 
+  $\X= \begin{bmatrix} \x_1 & \cdots \x_n \end{bmatrix} \tr =\I_p$, with fixed 
+  $\tau_d^2=\sigma_d^2=1$, the resulting conditional
+  posterior mean for the $\beta_{jd}$ is
+  $\E(\beta_{jd} | y_{jd}, \kappa_{jd}) = (1 - \kappa_{jd}) y_{jd}$. 
+  Thus, $\kappa_{jd}=0$ implies no shrinkage of $\beta_{jd}$ and $\kappa_{jd}=1$
+  implies full shrinkage towards zero. Figure 
+  \ref{fig:dens_kappa} depicts the prior on $\kappa_{jd}$ implied by 
+  several choices of $\beta_{jd}$ prior.
+<<dens_kappa, fig.cap="Implied prior densities $\\pi(\\kappa_{jd})$ for the (a) NIG, (b) Student's $t$, and (c) lasso priors.
+Different line types correspond to different hyperparameter settings.
+The hyperparameter settings (given in Section \\ref{sm-sec:hyperparameters} of the SM) were chosen to show some possible, distinct shapes that each of the priors can take.", out.width="100%", fig.asp=1/3>>=
+@
+
+  Figure \ref{fig:dens_kappa} shows that, depending on the choice
+  of hyperparameters, the NIG prior can behave similarly to the Student's $t$ 
+  prior (decreasing form zero, with substantial mass close to zero and 
+  little mass close to one, like the
+  solid lines in Figure \ref{fig:dens_kappa}a-\ref{fig:dens_kappa}b), 
+  but also rather differently (dashed and dotted lines in 
+  Figure \ref{fig:dens_kappa}a-\ref{fig:dens_kappa}b). Our argumentation to 
+  model the $\gamma^2_{jd}$ by an inverse Gaussian 
+  distribution, as has been suggested in \cite{fabrizi_specification_2016} and 
+  \cite{caron_sparse_2008}, is three-fold: (i) the NIG model is more flexible 
+  than the lasso prior (as seen from Figure 
+  \ref{fig:dens_kappa}), (ii) the NIG prior allows to model
+  the means of the $\gamma_{jd}^2$ ($\phi_{jd}$) and $\tau_{jd}^2$ 
+  ($\chi_{d}$) as a function of external data 
+  more conveniently than the Student's $t$ prior, as explained in Section 
+  \ref{sec:empiricalbayes}, and (iii) like the horseshoe 
+  \cite[]{carvalho_handling_2009}, the NIG shrinkage
+  weights prior can put mass both near zero and one, a desirable property of 
+  shrinkage priors \cite[]{bernardo_shrink_2011}.
+
+  A few remarks on the choice of error variance prior are justified here: 
+  many authors endow error variance components with vague gamma priors. 
+  \cite{gelman_prior_2006}, 
+  among others, advises against this practice. The degree of `vagueness' has a 
+  large influence on the posterior, while degree of `vagueness' is a difficult 
+  parameter to set. This influence is especially pronounced if the likelihood is
+  relatively flat, as may be reasonably expected in the large $p$, 
+  small $n$ setting. We therefore model the error variance with Jeffreys 
+  objective prior 
+  \cite[]{jeffreys_invariant_1946} that does not depend on any 
+  subjective specification of hyperparameters. In the derivation of our 
+  Jeffreys prior
+  for the error variance, we jointly consider an unknown data mean and variance 
+  \cite[]{kass_selection_1996}. This
+  joint consideration results in the somewhat unorthodox $1/\sigma^3$ Jeffreys
+  prior.
+
+	\subsection{External information}\label{sec:external}
+	In drug sensitivity prediction models, external information on both the drugs
+	and features is often available. Here, we assume this information to be 
+	available as
+	external feature `covariates' $\mathbf{c}_{jdg}$, for $g=1, \dots, G$, and 
+	drug `covariates' $\mathbf{z}_{dh}$, for $h=1,\dots, H$. An example of a 
+	(binary) feature covariate is target pathway presence, with $c_{jdg}=0$ if 
+	gene $j$ is present in the 
+	target pathway of drug $d$ and $c_{jdg}=1$ if it is not. An example of a 
+	(ternary) drug
+	covariate is developmental phase, with levels experimental phase, 
+	clinical development, and approved by a governing agency.
+
+	The external covariates come in through our mean models for 
+	the $\gamma_{jd}^2$ and $\tau_d^2$ hyperpriors:
+	$\phi_{jd} = (\mathbf{c}_{jd} \tr \bm{\alpha}_{\text{feat}})^{-1}$ and
+	$\chi_d = (\mathbf{z}_{d} \tr \bm{\alpha}_{\text{drug}})^{-1}$, with 
+	$\mathbf{c}_{jd} = \begin{bmatrix} c_{jd1} & \cdots & c_{jdG} \end{bmatrix}$
+	and $\mathbf{z}_{d} = \begin{bmatrix} z_{d1} & \cdots & z_{dH} \end{bmatrix}$,
+	where categorical external covariates are dummy coded.
+	The model now requires hyperparameters $\bm{\alpha}_{\text{feat}}$, 
+	$\lambda_{\text{feat}}$, $\bm{\alpha}_{\text{drug}}$, and $
+	\lambda_{\text{drug}}$, which we estimate in a data-driven manner (see Section
+	\ref{sec:empiricalbayes}).
+
+  A representation of our model as a Bayesian DAG is given in Figure
+  \ref{fig:dag}. We note that in many settings, the set of features might be
+  different for different drugs. In that case the covariates are indexed by
+  the drug $d$: $\X^d$, a trivial extension of model 
+  (\ref{eq:likelihood}) and (\ref{eq:prior}). This extension is included in the
+  available software, but for clarity it is omitted in the following.
+
+  \begin{figure}
+    \centering
+    \tikz{ %
+      % Y
+      \node[obs] (y) {$y_{id}$}; %
+
+      % X
+      \node[det, above=of y, yshift=1cm] (x) {$x_{ij}$}; %
+
+      % sigma2
+      \node[latent, right=of y] (sigma2) {$\sigma_d^2$}; %
+
+      % beta
+      \node[latent, right=of x] (beta) {$\beta_{jd}$}; %
+
+      % gamma2
+      \node[latent, right=of beta] (gamma2) {$\gamma_{jd}^2$}; %
+      \node[const, right=of gamma2, yshift=0.5cm] (phi) {$\phi_{jd}$}; %
+      \node[const, right=of gamma2, yshift=-0.5cm] (lambdaf) 
+      {$\lambda_{\text{feat}}$}; %
+      \node[det, right=of gamma2, xshift=2cm, yshift=0.5cm] (c) {$c_{jdg}$}; %
+
+      % tau2
+      \node[latent, right=of sigma2] (tau2) {$\tau_d^2$}; %
+      \node[const, right=of tau2, yshift=0.5cm] (chi) {$\chi_{d}$}; %
+      \node[const, right=of tau2, yshift=-0.5cm] (lambdad) 
+      {$\lambda_{\text{drug}}$}; %
+      \node[det, right=of tau2, xshift=2cm, yshift=0.5cm] (z) {$z_{dh}$}; %
+
+      % edges
+      \edge {x} {y};%
+      \edge {sigma2} {y};%
+      \edge {beta} {y};%
+      \edge {sigma2} {beta};%
+      \edge {gamma2} {beta};%
+      \edge {tau2} {beta};%
+      \edge {phi} {gamma2};%
+      \edge {lambdaf} {gamma2};%
+      \edge {chi} {tau2};%
+      \edge {lambdad} {tau2};%
+      \edge {c} {phi}; %
+      \edge {z} {chi}; %
+
+      % plates
+      \plate{plateg} {(c)} {$g=1,\dots,G$}
+      \plate{plateh} {(z)} {$h=1,\dots,H$}
+      \plate{platei} {(y) (x)}{$i=1,\dots,n$}; %
+      \plate{platej} {(x) (beta) (gamma2) (phi) (lambdaf) (c) (plateg)}
+      {$j=1,\dots,p$}; %
+      \plate{plated} 
+      {(y) (x) (sigma2) (beta) (gamma2) (tau2) (phi) (lambdaf)
+      (chi) (lambdad) (c) (z) (platei) (platej) (plateg) (plateh)}
+      {$d=1,\dots,D$}; %
+    }
+    \caption{Hierarchical representation of the drug sensitivity prediction 
+    model. Grey circles 
+    represent observed variables, white circles represent unobserved variables,
+    tilted squares represent fixed data, and unenclosed letters are parameters 
+    to be estimated. Cell lines are indexed by $i$, features
+    by $j$, drugs by $d$, drug covariates by $h$, and feature covariates by 
+    $g$. The $y_{id}$ are the drug sensitivities, $x_{ij}$ the 
+    molecular features, 
+    $c_{jdg}$ the external feature covariates, $z_{dh}$ the external drug
+    covariates, $\beta_{jd}$ the regression coefficients, $\sigma_d^2$ the 
+    error variances, $\tau_d^2$ and $\gamma_{jd}^2$ the drug and feature
+    specific variance components, respectively, and $\phi_{jd}$, 
+    $\lambda_{\text{feat}}$, $\chi_{d}$, and $\lambda_{\text{drug}}$ the 
+    hyperparameters.}
+    \label{fig:dag}
+  \end{figure}
+
+	\section{Estimation}\label{sec:estimation}
+	\subsection{Variational Bayes}\label{sec:variationalbayes}
+	The posterior corresponding to the model described in (\ref{eq:likelihood})
+	and (\ref{eq:prior}) is not available in closed form. 
+	To avoid computationally intensive markov chain Monte Carlo (MCMC) 
+	algorithms, we approximate the joint posterior by variational Bayes (see 
+	\cite{blei_variational_2017} for a review), where the approximate
+	posterior density factorises as: $p(\bbeta_d, \bm{\gamma}^2_d, \tau_{d}^2, 
+	\sigma_d^2 | \y_d) \approx 
+	Q_d(\cdot) = q(\bbeta_d) \cdot q(\bm{\gamma}_{d}^2) \cdot q(\tau_{d}^2) 
+	\cdot q(\sigma_d^2)$, where $\bm{\gamma}_{d}^2 = 
+	\begin{bmatrix} \gamma_{1d}^2 & \cdots & \gamma_{pd}^2 \end{bmatrix} \tr$. For
+  notational convenience, we slightly abuse notation and let $q(\cdot)$ denote
+	different densities for different inputs.
+	Under such a factorisation, the marginal variational posteriors that 
+	minimise the Kullback-Leibler divergence of the true posterior to the 
+	variational Bayes approximation \cite[]{neal_view_1998} are given by:
+	\begin{align*}
+    q(\bm{\beta}_d) & \overset{D}{=} \mathcal{N}_{p} 
+    (\bm{\mu}_d, \bm{\Sigma}_d), \\
+    q(\bm{\gamma}_{d}^2) & \overset{D}{=} \prod_{j=1}^p \mathcal{GIG}
+    \(-1, \lambda_{\text{feat}}/\phi_{jd}^2, \delta_{jd}\), \\
+    q(\tau^2_d) & \overset{D}{=} \mathcal{GIG} \(-\frac{p + 1}{2}, 
+    \lambda_{\text{drug}}/\chi_d^2, \eta_d\), \\
+    q(\sigma^2_d) & \overset{D}{=} \Gamma^{-1} \(\frac{n + p + 1}{2}, \zeta_d\),
+  \end{align*}
+	where $\mathcal{GIG} \(p, \nu, \eta\)$ denotes the generalized inverse Gaussian
+	distribution with index $p \in \mathbb{R}$, and scales $\nu > 0$ and 
+	$\eta > 0$ \cite[]{jorgensen_statistical_1982}. See SM Section 
+	\ref{sm-sec:vbderivations} for the derivations. The variational
+	parameters $\bm{\mu}_d$, $\bm{\Sigma}_d$, $\delta_{jd}$, $\eta_d$, and
+	$\zeta_d$ contain cyclic dependencies and are iteratively 
+	updated by:
+	\begin{subequations}\label{eq:vbequations}
     \begin{align}
-		\beta_{jd} & \sim \mathcal{N}_p (0, \phi^2_{g(j)} \gamma_d^2 \sigma_d^2), \\
-		\mathbf{u}_{d} & \sim \mathcal{N}_T (\mathbf{0}, \bm{\Xi}_d \sigma_d^2),
+      \bm{\Sigma}_d^{(h+1)} & = (a_d^{(h)})^{-1} 
+      \[\X \tr \X + g_{d}^{(h)} \diag(b_{jd}^{(h)})\]^{-1}, \\
+      \bm{\mu}_d^{(h+1)} & = \[\X \tr \X + g_{d}^{(h)} 
+      \diag(b_{jd}^{(h)})\]^{-1} 
+      \X \tr \y_d, \\
+      \delta_{jd}^{(h+1)} & = a_d^{(h)} g_{d}^{(h)} \[(\bmu^{(h+1)}_{jd})^2 + 
+      (\bSigma^{(h + 1)}_d)_{jj}\] + \lambda_{\text{feat}}, \\ 
+      \eta_d^{(h+1)} & = a_d^{(h)} \sum_{j=1}^{p} b_{jd}^{(h + 1)}
+      \[ (\bmu^{(h+1)}_{jd})^2 + (\bSigma^{(h + 1)}_d)_{jj} \] + 
+      \lambda_{\text{drug}}, \\
+      \zeta_d^{(h+1)} & = \frac{1}{2} \bigg[\mathbf{y}_d \tr \mathbf{y}_d -
+      2 \mathbf{y}_d \tr \X \bm{\mu}_d^{(h+1)} + 
+      \trace ( \X \tr \X \bm{\Sigma}_d^{(h+1)}) + 
+      (\bm{\mu}_d^{(h+1)}) \tr \X \tr \X \bm{\mu}_d^{(h+1)} \\
+      & \,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\, + 
+      g_{d}^{(h + 1)} \trace \[ \diag (b_{jd}^{(h+1)}) \bm{\Sigma}_d^{(h+1)}\] + 
+      g_{d}^{(h + 1)} (\bm{\mu}_d^{(h+1)}) \tr \diag (b_{jd}^{(h+1)}) 
+      \bm{\mu}_d^{(h+1)}\bigg],
     \end{align}
   \end{subequations}
-	and hyper-priors:
-	\begin{subequations}\label{eq:hyperprior}
+  until convergence, where $\y_d = \begin{bmatrix} y_{1d} & \cdots & y_{nd}
+  \end{bmatrix} \tr$. Here, we set
   \begin{align}
-			\sigma_d^{2} & \sim 1/\sigma_d^{3} \\
-			\gamma_d^{2} & \sim \mathcal{IG}(\theta_d, \lambda), \\
-			\bm{\Xi}_d & \sim \mathcal{W}^{-1}_T(\bm{\Omega}, \nu),
+    a_d^{(h)} & =\E_{Q^{(h)}}(\sigma_d^{-2})=(n + p + 1)/(2 \zeta_d^{(h)}), 
+    \nonumber\\
+    b_{jd}^{(h)} & = \E_{Q^{(h)}}(\gamma_{jd}^{-2}) = 
+    \sqrt{\frac{\lambda_{\text{feat}}}{\phi_{jd}^2 \delta_{jd}^{(h)}}}
+    \frac{K_0\(\sqrt{\delta_{jd}^{(h)}\lambda_{\text{feat}}/\phi_{jd}^2}\)}
+    {K_1\(\sqrt{\delta_{jd}^{(h)}\lambda_{\text{feat}}/\phi_{jd}^2}\)} + 
+    \frac{2}{\delta_{jd}^{(h)}}, \label{eq:ratiomodifiedbessel} \\
+    g_d^{(h)} & =\E_{Q^{(h)}}(\tau_d^{-2})=
+    \sqrt{\frac{\lambda_{\text{drug}}}{\chi_{d}^2 \eta_{d}^{(h)}}}
+    \frac{K_{(p-1)/2}\(\sqrt{\eta_{d}^{(h)}\lambda_{\text{drug}}/\chi_{d}^2}\)}
+    {K_{(p+1)/2}\(\sqrt{\eta_{d}^{(h)}\lambda_{\text{drug}}/\chi_{d}^2}\)} + 
+    \frac{p + 1}{\eta_{d}^{(h)}}.
+    \nonumber
   \end{align}
-	\end{subequations}
-	where $\mathcal{IG}(\theta, \lambda)$ is an inverse Gaussian distribution with mean $\theta$ and shape $\lambda$, and $\mathcal{W}^{-1}_T(\bm{\Omega}, \nu)$ is an inverse Wishart distribution with $T \times T$-dimensional positive definite (PD) scale matrix $\bm{\Omega}$ and degrees of freedom $\nu$.
-Note that for $\nu > T + 1$, this inverse Wishart prior implies an expected prior covariance matrix $(\nu - T - 1)^{-1} \bm{\Omega}$ for the tissue effects, common to all drugs.
-Likewise, the molecular feature effects are related through the common hyper-prior for the $\gamma_d^2$.
-
-	- Jeffrey's prior for variance $1/\sigma^{3/2}$ (a priori independent data mean and variance) or $1/\sigma^3$ (joint prior, but only mean and variance, not variance and betas).
-
-	A few remarks on the choice of priors are justified here: many authors endow error variance components with vague gamma priors.
-\cite{gelman_prior_2006}, among others, advises against this practice.
-The degree of `vagueness' has a large influence on the posterior, while degree of `vagueness' is a difficult parameter to set.
-This influence is especially pronounced if the likelihood is relatively flat, as may be reasonably expected in the large $p$, small $n$ setting we are in.
-We therefore model the error variance with Jeffrey's prior \cite[]{jeffreys_invariant_1946}.
-
-	Furthermore, we choose to model the $\gamma^2_d$ by an inverse Gaussian distribution, as has been suggested in \cite{fabrizi_specification_2016} and \cite{caron_sparse_2008}, because it allows to model the mean $\theta_d$ as a function of the drug covariates $\mathbf{c}_d$, as explained in Section \ref{sec:empiricalbayes}.
-
-	- IG hyperprior used in finance \cite[]{barndorff-nielsen_normal_1997}, where it is called the NIG prior.
-
-	\section{Estimation}
-	\subsection{Variational Bayes}\label{sec:variationalbayes}
-	The posterior corresponding to the model described in (\ref{eq:linearmodel}), (\ref{eq:prior}), and (\ref{eq:hyperprior}) is not available in closed form.
-We therefore approximate the posterior by variational Bayes, where we force the posterior density $Q$ to factorise as:
-	\begin{align}\label{eq:variationalposterior}
-	p(\B,\mathbf{U},\mathbf{\sigma}^2, \bgamma^2,\bm{\Xi}_1, \dots, \bm{\Xi}_D) \approx Q(\cdot) = q(\B) \cdot q(\mathbf{U}) \cdot q(\mathbf{\sigma}^2) \cdot q(\bgamma^2) \cdot q(\bm{\Xi}_1, \dots, \bm{\Xi}_D),
-	\end{align}
-	where $\B = \begin{bmatrix} \bbeta_1 & \cdots & \bbeta_D \end{bmatrix}$ and $\U = \begin{bmatrix} \mathbf{u}_1 & \cdots & \mathbf{u}_D \end{bmatrix}$, are of dimensions $p \times D$ and $T \times D$, respectively.
-Under the factorisation in (\ref{eq:variationalposterior}), the marginal variational posteriors that minimise the Kullback-Leibler divergence of the true posterior to the variational Bayes approximation \cite[]{neal_view_1998}, are given by:
-	\begin{subequations}\label{eq:variationalposterior2}
-		\begin{align}
-		q_{\B} (\B) & \overset{D}{=} \prod_{d=1}^{D} \mathcal{N}_p (\bmu_d, \bSigma_d), \\
-		q_{\U}(\U) & \overset{D}{=} \prod_{d=1}^{D} \mathcal{N}_T (\mathbf{m}_d, \Sm_d), \\
-		q_{\bgamma^{2}}(\bgamma^{2}) & \overset{D}{=} \prod_{d=1}^D \mathcal{GIG} \(-\frac{p+1}{2}, \frac{\lambda}{\theta_d^2}, \delta_{d} \), \\
-		q_{\bsigma^{2}}(\bsigma^{2}) & \overset{D}{=} \prod_{d=1}^D \Gamma^{-1} \(\frac{n + p + T + 1}{2}, \zeta_{d} \), \\
-		q_{\bm{\Xi}_1, \dots, \bm{\Xi}_D}(\bm{\Xi}_1, \dots, \bm{\Xi}_D) & \overset{D}{\propto} \prod_{d=1}^D \mathcal{W}_T^{-1} (\bm{\Psi}_d, \nu + 1).
-		\end{align}
-	\end{subequations}
-	Here $\mathcal{GIG}(\cdot)$ and $\Gamma^{-1}(\cdot)$ denote the generalized inverse Gaussian and inverse gamma distributions, respectively.
-The full derivations may be found in Appendix \ref{app:derivations}.
-
-	The parameters in (\ref{eq:variationalposterior2}) contain cyclic dependencies. We therefore iterate the following estimating equations until convergence to a local optimum:
-	\begin{align*}
-	\bSigma_d^{(h+1)} & = \frac{2\zeta_d^{(h)}}{n+p+T+1} \left\{\X \tr \X + a_d^{(h)} \cdot \diag (\phi_{g(j)}^{-2}) \right\}^{-1}, \\
-	\bm{\mu}_d^{(h+1)} & = \left\{\X \tr \X + a_d^{(h)} \cdot \diag (\phi_{g(j)}^{-2}) \right\}^{-1} \X \tr ( \y_d - \Z \mathbf{m}_d^{(h)}), \\
-	\Sm_d^{(h+1)} & = \frac{2\zeta_d^{(h)}}{n+p+T+1} \[ \Z \tr \Z + (\nu + 1) \cdot (\bm{\Psi}_d^{(h)})^{-1} \]^{-1}, \\
-	\mathbf{m}_d^{(h+1)} & = \[ \Z \tr \Z + (\nu + 1) \cdot (\bm{\Psi}_d^{(h)})^{-1} \]^{-1} \Z \tr (\y_d - \X \bm{\mu}_d^{(h)}), \\
-	\bm{\Psi}_d^{(h+1)} & = \frac{2\zeta_d^{(h)}}{n+p+T+1} \[ \Sm_d^{(h+1)} + \mathbf{m}_d^{(h+1)} (\mathbf{m}_d^{(h+1)}) \tr \] + \bm{\Omega}, \\
-	\delta_d^{(h+1)} & =  \frac{n+p+T+1}{2\zeta_d^{(h)}} \sum_{j=1}^p \frac{(\mu_{jd}^{(h+1)})^2 + (\bSigma_d^{(h+1)})_{jj}}{\phi^2_{g(j)}} + \lambda, \\
-	\zeta_d^{(h+1)} & = \frac{1}{2} \y_d \tr \y_d - \y_d \tr (\X \bmu_d^{(h+1)} + \Z \mathbf{m}_d^{(h+1)}) + (\bmu_d^{(h+1)}) \tr \X \tr \Z \mathbf{m}_d^{(h+1)} + \frac{1}{2} \trace (\X \tr \X \bSigma_d^{(h+1)}) \\
-	& \,\,\,\,\,\,\,\,\,\, + \frac{1}{2} (\bmu_d^{(h+1)}) \tr \X \tr \X \bmu_d^{(h+1)} + \frac{1}{2} \trace (\Z \tr \Z \Sm_d^{(h+1)}) + \frac{1}{2} (\mathbf{m}_d^{(h+1)}) \tr \Z \tr \Z \mathbf{m}_d^{(h+1)} \\
-	& \,\,\,\,\,\,\,\,\,\, + \frac{1}{2} a_d^{(h)} \sum_{j=1}^p \frac{(\mu_{jd}^{(h+1)})^2 + (\bSigma_d^{(h+1)})_{jj}}{\phi^2_{g(j)}} + \frac{1}{2} (\nu + 1) \cdot \trace \[ (\bm{\Psi}_d^{(h+1)})^{-1} \Sm_d^{(h+1)} \] \\
-	& \,\,\,\,\,\,\,\,\,\, + \frac{1}{2} (\nu + 1) \cdot (\mathbf{m}_d^{(h+1)}) \tr (\bm{\Psi}_d^{(h+1)})^{-1} \mathbf{m}_d^{(h+1)},
-	\end{align*}
-	where
-	$$
-	a_d^{(h)} = \sqrt{\frac{\lambda}{\theta^2_d \delta_d^{(h)}}} \frac{K_{\frac{p - 1}{2}} \( \sqrt{\lambda \delta_d^{(h)}}/\theta_d \)}{K_{\frac{p+1}{2}} \( \sqrt{\lambda \delta_d^{(h)}}/\theta_d \)} + \frac{p+1}{\delta_d^{(h)}},
-	$$
-	and $K_{\alpha}(\cdot)$ is the modified Bessel function of the second kind.
+  where $K_{\nu}(x)$ denotes the modified Bessel function of the second kind.
+  A method for fast and numerically stable calculation of ratios of
+  modified Bessel functions of the second kind, as in 
+  (\ref{eq:ratiomodifiedbessel}), is given in SM Section 
+  \ref{sm-sec:ratiosmodifiedbessels}.
 
   \subsection{Empirical Bayes}\label{sec:empiricalbayes}
-	A full Bayesian model requires the specification of hyper-parameters $\bm{\eta} = \begin{bmatrix} (\bm{\phi}^2) \tr & \bm{\theta} \tr & \lambda & \text{vec}(\bm{\Omega}) \tr & \nu \end{bmatrix} \tr$.
-These are abstract and hard to interpret parameters for which we generally lack expert knowledge.
-They do, however, have a significant influence on the shape of the posterior distribution.
-We therefore propose to estimate the unspecified hyper parameters $\bm{\eta}$ by empirical Bayes.
-Simply put, empirical Bayes fits the prior model to the data and is, as such, objective (up to model specification).
+  We parametrised the prior 
+  mean of the $\gamma_{jd}^2$ as $\phi_{jd}=(\mathbf{c}_{jd} \tr 
+  \bm{\alpha}_{\text{feat}})^{-1}$ and the prior mean of $\tau_d^2$ as
+  $\chi_{d}=(\mathbf{z}_{d} \tr \bm{\alpha}_{\text{drug}})^{-1}$.
+  This parametrisation allows us to include feature and drug covariates,
+  both continuous and discrete, into the 
+  model. Additionally, it reduces the number of hyperparameters from 
+  $pD$ to $|\bm{\alpha}_{\text{feat}}| + |\bm{\alpha}_{\text{drug}}| + 2$. The
+  Bayesian model
+  then requires the specification of the hyperparameters 
+  $\bm{\alpha}=\begin{bmatrix} \bm{\alpha}_{\text{feat}} \tr & 
+  \bm{\alpha}_{\text{drug}} \tr 
+  \end{bmatrix} \tr$ and $\bm{\lambda} = \begin{bmatrix} \lambda_{\text{feat}} &
+  \lambda_{\text{drug}}
+  \end{bmatrix} \tr$. These are
+  abstract and hard to interpret parameters 
+  for which we generally lack expert knowledge. They do, however, have a 
+  significant influence on the shape of the posterior distribution. We 
+  therefore propose to estimate these hyperparameters by empirical Bayes.
+  In our case, this results in an objective and data-driven inclusion of the
+  external feature and drug covariates.
 
-	The canonical method for empirical Bayes is maximisation of the marginal likelihood with respect to the hyper parameters.
-In \cite{casella_empirical_2001} the marginal likelihood is maximised by an EM algorithm:
+	The canonical method for empirical Bayes is to maximise the marginal 
+	likelihood with respect to the hyperparameters. In 
+	\cite{casella_empirical_2001} the marginal likelihood is maximised by an EM 
+	algorithm:
+	\begin{align*}
+	  \bm{\alpha}^{(l+1)},\bm{\lambda}^{(l+1)} & = \underset{\bm{\alpha},
+	  \bm{\lambda} > 0}{\argmax}\E_{\cdot | \Y} 
+	  [\log p(\Y, \B, \bm{\Gamma}^2, \bm{\tau}^2, \bsigma^2) | \bm{\alpha}^{(l)},
+	  \bm{\lambda}^{(l)}] \\
+	  & = \underset{\bm{\alpha},\bm{\lambda} > 0}{\argmax} 
+	  \E_{\cdot | \Y} [\log \pi (\bm{\Gamma}^2) | \bm{\alpha}_{\text{feat}}^{(l)},
+	  \bm{\lambda}_{\text{feat}}^{(l)}] + 
+	  \E_{\cdot | \Y} [\log \pi (\bm{\tau}^2) | 
+	  \bm{\alpha}_{\text{drug}}^{(l)}, \bm{\lambda}_{\text{drug}}^{(l)}],
+	\end{align*}
+	where $\Y = \begin{bmatrix} \y_1 & \cdots & \y_D \end{bmatrix}$,
+	$\B = \begin{bmatrix} \bbeta_1, \dots, \bbeta_D \end{bmatrix}$, 
+	$\bm{\tau}^2 = \begin{bmatrix} \tau^2_1 & \cdots & 
+	\tau^2_D \end{bmatrix} \tr$,
+	$\bsigma^2 = \begin{bmatrix} \sigma^2_1 & \cdots & 
+	\sigma^2_D \end{bmatrix} \tr$,
+	and $\bm{\Gamma}^2 = \begin{bmatrix} \bgamma^2_1 & \cdots & \bgamma^2_D 
+	\end{bmatrix}$,
+	and the expectation is with respect to the joint posterior. 
+	In our case, this 
+	posterior is not available in closed form, which renders the expectation 
+	difficult. While \cite{casella_empirical_2001} suggests to approximate the 
+	expectation by a Monte Carlo sample, we propose to use the variational Bayes 
+	approximation developed in Section \ref{sec:variationalbayes}:
 	$$
-	\bm{\eta}^{(l+1)} = \underset{\bm{\eta}}{\argmax}\E_{\cdot | \Y} [\log p(\Y, \B, \U, \bm{\Xi}_1, \dots, \bm{\Xi}_D, \bgamma^2, \bsigma^2) | \bm{\eta}^{(l)}],
+	\bm{\alpha}^{(l+1)},\bm{\lambda}^{(l+1)} = 
+	\underset{\bm{\alpha},\bm{\lambda} > 0}{\argmax}\E_{Q^{(l)}} 
+	[\log \pi(\bm{\Gamma}^2)| \bm{\alpha}_{\text{feat}}^{(l)}, 
+	\bm{\lambda}_{\text{feat}}^{(l)}] + 
+	\E_{Q^{(l)}} [\log \pi(\bm{\tau}^2)| 
+	\bm{\alpha}_{\text{drug}}^{(l)}, \bm{\lambda}_{\text{drug}}^{(l)}],
 	$$
-	where the expectation is with respect to the posterior.
-In our case, this posterior is not available in closed form, which renders the expectation difficult.
-While \cite{casella_empirical_2001} suggests to approximate the expectation by a Monte Carlo sample, we propose to use the variational Bayes approximation developed in Section \ref{sec:variationalbayes}:
-	\begin{align}\label{eq:variationalbayes}
-	\bm{\eta}^{(l+1)} & = \underset{\bm{\eta}}{\argmax}\E_{Q^{(l)}} [\log p(\Y, \B, \U, \bm{\Xi}_1, \dots, \bm{\Xi}_D, \bgamma^2, \bsigma^2)] \nonumber \\
-	& = \underset{\bm{\eta}}{\argmax} \E_{Q^{(l)}} [\log \pi (\B | \bgamma^2, \bsigma^2)] + \E_{Q^{(l)}} [\log \pi (\bgamma^2)] + \E_{Q^{(l)}} [\log \pi (\bm{\Xi}_1, \dots, \bm{\Xi}_D)].
-	\end{align}
-	where now the expectation is with respect to the converged variational posterior $Q^{(l)}$.
-Inspection of (\ref{eq:variationalbayes}) learns us that the problem may be decomposed into three separate optimisation problems.
-
-	\subsubsection{Group-specific variance components}
-	The first sub-problem concerns the group-specific variance components of the molecular feature effects.
-In order to separate the overall variance of the molecular feature effects from the group-specific deviations, we impose $\prod_{g=1}^G (\phi_g^2)^{|\mathcal{G}_g|} = 1$.
-As a consequence, the overall level of shrinkage is determined by the $\gamma^2_d$, while the $\phi_g^2$ effectuate differential shrinkage of the groups.
-Because the $\phi_g^2$ are variance components, we additionally require $\phi^2_1, \dots, \phi^2_G> 0$, such that our first optimisation sub-problem becomes:
+	where now the expectation is with respect to the converged variational 
+	posterior $Q^{(l)}=\prod_{d=1}^D Q_d^{(l)}$. Note that the prior 
+	$\bgamma^2_d$ and $\tau_d^2$ independence assumption 
+	results in separate optimisation problems for the 
+	feature hyperparameters ($\bm{\alpha}_{\text{feat}}$ and 
+	$\bm{\lambda}_{\text{feat}}$), and the drug hyperparameters 
+	($\bm{\alpha}_{\text{drug}}$ and $\bm{\lambda}_{\text{drug}}$).
+	If we stack the drug and feature covariates:
 	\begin{align*}
-	(\bm{\phi}^2)^{(l+1)} & = \underset{\bm{\phi}^2}{\argmax} \E_{Q^{(l)}} [\log \pi (\B | \bgamma^2, \bsigma^2)] \\
-	& = \underset{\bm{\phi}^2}{\argmax} -\frac{D}{2} \sum_{g=1}^G |\G_g| \log \phi_g^2 - \frac{1}{2} \sum_{g=1}^G b_g^{(l)} \phi_g^{-2} \\
-	& \text{subject to } \prod_{g=1}^G (\phi_g^2)^{|\G_g|} = 1 \text{ and } \phi^2_1, \dots, \phi^2_G> 0.
+	  \mathbf{C} = 
+	  \begin{bmatrix} \mathbf{c}_{11} \tr \\ 
+	    \vdots \\
+	    \mathbf{c}_{p1} \tr \\
+	    \vdots \\
+	    \mathbf{c}_{1D} \tr \\
+	    \vdots \\
+	    \mathbf{c}_{pD} \tr
+	  \end{bmatrix} \text{ and }
+	  \mathbf{Z} =
+	  \begin{bmatrix} \mathbf{z}_{1} \tr \\ 
+	    \vdots \\
+	    \mathbf{z}_D \tr
+	  \end{bmatrix},
 	\end{align*}
-	where 
-	$$
-	b_g^{(l)}=\sum_{d=1}^D a_d^{(l)} \frac{n+p+T+1}{2\zeta_d^{(l)}} \sum_{j \in \mathcal{G}_g} \[ (\mu_{jd}^{(l)})^2 + (\bSigma_d^{(l)})_{jj} \].
-	$$
-	This is a convex optimisation problem and easily solved numerically.
-
-	\subsubsection{Drug-specific variance components}
-	The second sub-problem concerns the prior for the drug-specific variance components $\gamma_d^2$.
-As explained in Section \ref{sec:SEM}, we may have `covariates' $\mathbf{c}_d$ on the drugs available.
-We conjecture that these covariates may be informative for the omics effect size and therefore model the prior mean of the $\gamma_d^2$ as a function of these covariates: $\theta_d = (\mathbf{c}_d \tr \bm{\alpha})^{-1}$.
-Then, the empirical Bayes estimation of $\bm{\alpha}$ effectively boils down to an inverse Gaussian regression of the $\gamma_d^{-2}$ on the $\mathbf{c}_d$ \cite[]{fries_optimal_1986,whitmore_regression_1983}.
-The problem may be formulated as:
-	\begin{align*}\label{eq:optproblem2}
-	\lambda^{(l+1)}, \bm{\alpha}^{(l+1)} & = \underset{\lambda,\bm{\alpha}}{\argmax} \E_{Q^{(l)}} [\log \pi (\bgamma^2)] \\
-	& = \underset{\lambda,\bm{\alpha}}{\argmax} \frac{D}{2} \log \lambda - \frac{\lambda}{2} \bm{\alpha} \tr \mathbf{C} \tr \diag [\E_{Q^{(l)}}(\gamma_d^2)] \mathbf{C} \bm{\alpha} + \lambda \bm{\alpha} \tr \mathbf{C} \tr \mathbf{1} - \frac{\lambda}{2} \sum_{d=1}^D \E_{Q^{(l)}}(\gamma_d^{-2}),
+	the empirical Bayes updates are given by:
+  \begin{align*}
+    \bm{\alpha}^{(l+1)}_{\text{feat}} & = \[ \mathbf{C} \tr \diag 
+    (e_{jd}^{(l)}) \mathbf{C} \]^{-1} \mathbf{C} \tr 
+    \mathbf{1}_{pD \times 1}, \\
+    \lambda^{(l+1)}_{\text{feat}} & = pD \[ \sum_{d=1}^D \sum_{j=1}^{p} 
+    b_{jd}^{(l)} + (\bm{\alpha}^{(l+1)}_{\text{feat}}) \tr \mathbf{C} \tr 
+    \diag (e_{jd}^{(l)}) \mathbf{C} \bm{\alpha}^{(l+1)}_{\text{feat}} - 
+    2 (\bm{\alpha}^{(l+1)}_{\text{feat}}) \tr \mathbf{C} \tr 
+    \mathbf{1}_{pD \times 1} \]^{-1},\\
+    \bm{\alpha}^{(l+1)}_{\text{drug}} & = \[ \mathbf{Z} \tr \diag 
+    (f_{d}^{(l)}) \mathbf{Z} \]^{-1} \mathbf{Z} \tr 
+    \mathbf{1}_{D \times 1}, \\
+    \lambda^{(l+1)}_{\text{drug}} & = D \[ \sum_{d=1}^D g_{d}^{(l)} + 
+    (\bm{\alpha}^{(l+1)}_{\text{drug}}) \tr \mathbf{Z} \tr \diag (f_{d}^{(l)})
+    \mathbf{Z} \bm{\alpha}^{(l+1)}_{\text{drug}} - 
+    2 (\bm{\alpha}^{(l+1)}_{\text{drug}}) \tr \mathbf{Z} \tr 
+    \mathbf{1}_{D \times 1} \]^{-1},
   \end{align*}
-	where $\mathbf{C} = \begin{bmatrix} \mathbf{c}_1 & \dots & \mathbf{c}_D \end{bmatrix} \tr$. The solutions to this problem are:
+  where SM Section \ref{sm-sec:eb} shows that $\bm{\lambda}^{(l+1)} > 0$ and
   \begin{align*}
-	\bm{\alpha}^{(l+1)} & = \left\{ \mathbf{C} \tr \Vm^{(l)} \mathbf{C} \right\}^{-1} \mathbf{C} \tr \mathbf{1}, \\
-	\lambda^{(l+1)} & = \[ \sum_{d=1}^D a_d^{(l)} - (\bm{\alpha}^{(l+1)}) \tr \mathbf{C} \tr \mathbf{1} \]^{-1} D,
-	\end{align*}
-	where 
-	$$
-	\Vm^{(l)} = \diag \[ \sqrt{\frac{\delta_d^{(l)}(\theta_d^{(l)})^2}{\lambda^{(l)}}} \frac{K_{\frac{p - 1}{2}} \( \sqrt{\lambda^{(l)} \delta_d^{(l)}}/\theta^{(l)}_d \)}{K_{\frac{p + 1}{2}} \( \sqrt{\lambda^{(l)} \delta_d^{(l)}}/\theta^{(l)}_d \)} \].
-	$$
-
-	- Think about parametrisation of $\mathbf{C}$.
-Idea: use overall mean and deviations from it in alpha, such that the starting value can be the mean and 0 for alpha.
-Or: use intercept and code the rest as dummy variables.
-
-	\subsubsection{Tissue effect covariance}
-	The third sub-problem pertains to the tissue effect covariance prior.
-To avoid overfitting when freely estimating the $\frac{T(T+1)}{2}$ elements in $\bm{\Omega}$ and $\nu$, we parametrise them as $\bm{\Omega}=  \tau [(\tau - \rho) \I_T + \rho \mathbf{1}_{T \times T}]$ and $\nu = \tau + T + 1$, respectively.
-The justification for this parametrisation is three-fold: (a) it implies an easy to interpret prior covariance matrix with $\tau$ on its diagonal and $\rho$ on its off-diagonals, (b) it reduces the number of parameters to  estimate to just two, and (c) simplifies the positive definite constraint on  $\bm{\Omega}$.
-To ensure positive definiteness and a proper covariance matrix, we impose the constraints $\tau > \max\[\rho - T \rho,\rho, 0\]$ (see Appendix \ref{app:priortissuecovariance}).
-The resulting third sub-problem is a bit more involved than the first two:
-	\begin{align*}
-	\tau^{(l + 1)}, \rho^{(l+1)} & = \underset{\tau, \rho}{\argmax} \frac{D (T^2 + T)}{2} \log \tau + \frac{DT}{2} \tau \log \tau + \frac{D(T+1)}{2} \log \[ \tau - (T - 1) \rho\] \\
-	& \,\,\,\,\,\,\,\,\,\, + \frac{D}{2} \tau \log \[ \tau - (T - 1) \rho\] + \frac{D(T^2 - 1)}{2} \log(\tau - \rho) +\frac{D(T - 1)}{2} \tau  \log(\tau - \rho) \\
-	& \,\,\,\,\,\,\,\,\,\, - D \sum_{t=1}^T \log \Gamma \( \frac{\tau + t + 1}{2} \) - w_1^{(l)} \tau^2 + w_2^{(l)} \tau \rho + w_3^{(l)} \tau, \\
-	& \text{subject to } \tau > \max\[\rho - T \rho,\rho, 0\],
-	\end{align*}
-	where
-	\begin{align*}
-	w_1^{(l)} & =  \frac{\tau^{(l)} + T + 2}{2} \sum_{d=1}^D \trace \[ (\bm{\Psi}_d^{(l)})^{-1}\] > 0,\\
-	w_2^{(l)} & = \frac{\tau^{(l)} + T + 2}{2} \sum_{d=1}^D \left\{\trace \[ (\bm{\Psi}_d^{(l)})^{-1}\] - \trace \[\mathbf{1}_{T \times T} \cdot (\bm{\Psi}_d^{(l)})^{-1}\] \right\} < w_1^{(l)},\\
-	w_3^{(l)} & = \frac{1}{2} \[D \sum_{t=1}^T \psi\(\frac{\tau^{(l)} + t + 2}{2}\) - \sum_{d=1}^D \log |\bm{\Psi}_d^{(l)}| \].
+    e_{jd}^{(l)} & = \E_{Q^{(l)}}(\gamma_{jd}^2| 
+    \bm{\alpha}_{\text{feat}}^{(l)}, 
+    \lambda_{\text{feat}}^{(l)}) = (b_{jd}^{(l)} - 
+    2/\delta_{jd}^{(l)}) \cdot \delta_{jd}^{(l)} (\phi_{jd}^{(l)})^2/
+    \lambda_{\text{feat}}^{(l)}, \\
+    f_{d}^{(l)} & = \E_{Q^{(l)}}(\tau_{d}^2| \bm{\alpha}_{\text{drug}}^{(l)}, 
+    \lambda_{\text{drug}}^{(l)}) = (g_{d}^{(l)} - 
+    (p + 1)/\eta_{d}^{(l)}) \cdot \eta_{d}^{(l)} (\chi_{d}^{(l)})^2/
+    \lambda_{\text{drug}}^{(l)}.
   \end{align*}
 
-	- use mathematica to check convexity
+  To ensure proper and unbiased shrinkage, intercepts are included in 
+  $\bm{\alpha}_{\text{feat}}$ and $\bm{\alpha}_{\text{drug}}$. This is achieved 
+  by appending both
+  $\mathbf{C}$ and $\mathbf{Z}$ with a column of 
+  ones. These intercepts are roughly interpreted as the
+  expected prior precisions $\E(\gamma_{jd}^{-2})$ and $\E(\tau_d^{-2})$
+  if the feature and drug covariates are all zero. 
+  Likewise, an $\alpha$ corresponding to an external covariate may be
+  interpreted as an additive effect of the external covariate on the prior
+  expected precision. So an $\alpha=1$ translates to an increase in expected 
+  prior precision of
+  1 for every increase in the external covariate of 1,
+	keeping all the other external covariates fixed.
 
-	- we use the ELBO for VB convergence and parameters for EB convergence
+  Variational Bayes approximations are known to underestimate posterior 
+  variances \cite[]{rue_approximate_2009,consonni_mean-field_2007,
+  bishop_pattern_2006,wang_inadequacy_2005}. In simulation Scenario 5 in Section 
+  \ref{sm-sec:simulations} of the SM, we compare the variational posterior to 
+  MCMC samples from the posterior with fixed hyperparameters
+  estimates (after the procedure described in Section \ref{sec:empiricalbayes}
+  has converged). In this simulation Scenario and other settings (not 
+  shown), the variational approximation to the posterior is accurate. If 
+  however, the
+  user is reluctant to trust the variational posterior variances, samples
+  from the posterior may be generated with the
+  Gibbs sampler in SM Section \ref{sm-sec:gibbssampler}. Alternatively, 
+  we provide an
+  implementation of the proposed model in stan using the R package rstan 
+  \cite[]{guo_rstan:_2018} at 
+  \url{https://github.com/magnusmunch/NIG}.
 
-	- we might be lenient for VB convergence. The extreme case would be 1 VB iteration per EB iteration (corresponds to tolerance of infinity).
+	\section{Simulations}\label{sec:simulations}
+	\subsection{Setup}
+	This section investigates the empirical 
+	Bayes estimation properties of the model in a simulated setting; its main aim 
+	is to assess hyperparameter estimation. It is a data based simulation,
+	wherein the responses are simulated from a synthetic model, but the
+	features are taken from the real GDSC expression data introducted in Section 
+	\ref{sec:gdsc}. The real GDSC features contain strong collinearities. Such
+	strong collinearities in the design matrix impede correct parameter estimation
+	with small sample sizes. 
+	We therefore replace the ambition of correctly estimating the $\beta_{jd}$
+	with the more modest aim of approximately correct estimation of the 
+	hyperparameters.
 
-	\section{Simulations}
+	A pre-processing step selects 100 features with largest variance, while
+	251 drug sensitivities for 507 cell lines (half of the total number of cell
+	lines) are 
+	simulated from model (\ref{eq:likelihood}) and 
+	(\ref{eq:prior}). We draw the error
+	variances as $\forall d: \sigma_d^2 \sim \Gamma^{-1}(3,2)$,
+	such that the prior $\sigma_d^2$ mean and variance are both one. 
+	We consider the following four scenarios for the simulation of the 
+	drug and feature variance components:
+	\begin{itemize}
+	  \item{Scenario 1 fixes $\forall d: \tau_d^2=1$ and draws the 
+	    $\gamma_{jd}^2$ according to model (\ref{eq:prior}). We
+	    create four external dummy feature covariates that code for four 
+	    approximately equally sized groups of features. We set
+	    $\bm{\alpha}_{\text{feat}}$ such that the $\gamma_{jd}^2$ of the four
+	    groups of features have prior means $\phi_{jd} \in \{ 1, 1/2, 1/4, 1/8 \}$
+	    ($\bm{\alpha}_{\text{feat}}=\begin{bmatrix} 1 & 1 & 3 & 7 \end{bmatrix} 
+	    \tr$). The prior scale parameter is set to $\lambda_{\text{feat}}=1$}.
+	  \item{Scenario 2 fixes $\forall j,d: \gamma_{jd}^2=1$ and draws the 
+	    $\tau_d^2$ according to model (\ref{eq:prior}), following a procedure
+	    similar to the procedure for the $\gamma_{jd}^2$ in Scenario 1: we 
+	    create four groups of drugs with corresponding external drug 
+	    dummy variables and set $\bm{\alpha}_{\text{drug}}=\begin{bmatrix} 
+	    1 & 1 & 3 & 7 \end{bmatrix} \tr$, such that we have $\chi_{d} \in 
+	    \{ 1, 1/2, 1/4, 1/8 \}$. The scale is set to $\lambda_{\text{drug}}=1$.}
+	  \item{Scenario 3 combines the procedures from Scenarios 1 and 2 to draw both
+	    the $\gamma_{jd}^2$ and $\tau_d^2$ according to (\ref{eq:prior}).}
+	  \item{Scenario 4 is equal to Scenario three, except that we add noise to
+	    the external covariates. Noise is supposed to mimic a low external 
+	    covariate signal and is constructed by permutation of fractions 
+	    $q \in \{0.1, 0.2, 0.33, 0.5, 0.67, 0.8, 1\}$ of the 
+	    rows of the external covariates.}
+	\end{itemize}
+	We estimate two models: (i) the NIG model that only includes an intercept in
+	the external covariates, called NIG$_{\text{f}}^-$, NIG$_{\text{d}}^-$, or
+	NIG$_{\text{f}+\text{d}}^-$, depending on which variance components are 
+	estimated (feature, drug, or both in Scenarios 1, 2, and 3/4, respectively),
+	and 
+	(ii) the NIG model estimated as in Section \ref{sec:estimation} that includes
+	all external covariates, called NIG$_{\text{f}}$, NIG$_{\text{d}}$, or
+	NIG$_{\text{f}+\text{d}}$, again depending on which variance components are 
+	estimated.
+	Exclusion of the external covariates as in the NIG$_{\text{f}}^-$, 
+	NIG$_{\text{d}}^-$, and NIG$_{\text{f}+\text{d}}^-$ models
+	amounts to direct estimation of common expected 
+	prior means $\phi$ and/or $\chi$, instead of regression estimates for the
+	$\phi_{jd}$ and/or $\chi_d$ as in the 
+	NIG$_{\text{f}}$, NIG$_{\text{d}}$, and NIG$_{\text{f}+\text{d}}$ models.
+	In the language of \cite{bernardo_shrink_2011} as introduced in 
+	Section \ref{sec:introduction}, models NIG$_{\text{f}}$ and NIG$_{\text{d}}$
+	may be described as local and global shrinkage rules, respectively, as 
+	opposed to the global-local shrinkage models NIG$_{\text{f}+\text{d}}$ and
+	NIG$_{\text{f}+\text{d}}^-$. We repeat every simulation Scenario 100 times.
 
-	- Hamming distance to assess variable selection
+	SM Section \ref{sm-sec:simulations} contains more 
+	simulation results for Scenarios 1-4 for
+	the NIG model and four alternative methods: frequentist (i) 
+	lasso and (ii) ridge models, and EB version of the ridge model: (iii) a
+	multivariate Bayesian extension of \texttt{xtune} 
+	\cite[]{ni_bayesian_2019} named mxtune, as introduced in SM Section 
+	\ref{sm-sec:mxtune}. Additionally, SM Section \ref{sm-sec:simulations} 
+	contains a comparison of MCMC and VB posteriors in simulation Scenario 3.
 
-	\section{Data application}
-	We predict IC50 of the drugs.
-Other choices of efficacy measures are possible, but a discussion on the pros and cons of these different choices is not the aim of this paper.
-In any case, the presented methods may be used with any continuous efficay measure.
+	\subsection{Results}
+<<simulation_gdsc>>=
+# simulation 1
+res <- read.table("results/simulations_gdsc_res1.2.txt", row.names=NULL)
+temp <- res[, 1]
+res <- as.matrix(res[, -1])
+rownames(res) <- temp
 
-	- check GSK open target EBI or ask Chris Wallace about it. Database that links targets to drugs and diseases. Might be useful for drugs co-data.
+true.phi1 <- round(mean(1/(c(1, 1, 3, 7) %*% 
+                             t(unname(model.matrix(~ factor(1:4)))))), 3)
+est.phi1 <- mean(1/res[rownames(res)=="alphaf0", 1])
+se.phi1 <- sd(1/res[rownames(res)=="alphaf0", 1])
+phi1 <- paste0(round(est.phi1, 3), " (", round(se.phi1, 3), ")")
+est.lambdaf1 <- apply(res[rownames(res)=="lambdaf", c(1, 2)], 2, mean)
+se.lambdaf1 <- apply(res[rownames(res)=="lambdaf", c(1, 2)], 2, sd)
+lambdaf1 <- paste0(round(est.lambdaf1, 3), " (", round(se.lambdaf1, 3), ")")
 
-	- check NCI-60
+# simulation 2
+res <- read.table("results/simulations_gdsc_res2.2.txt", row.names=NULL)
+temp <- res[, 1]
+res <- as.matrix(res[, -1])
+rownames(res) <- temp
 
-	- check CTRP study
+true.chi2 <- round(mean(1/(c(1, 1, 3, 7) %*% 
+                             t(unname(model.matrix(~ factor(1:4)))))), 3)
+est.chi2 <- mean(1/res[rownames(res)=="alphad0", 1])
+se.chi2 <- sd(1/res[rownames(res)=="alphad0", 1])
+chi2 <- paste0(round(est.chi2, 3), " (", round(se.chi2, 3), ")")
+est.lambdad2 <- apply(res[rownames(res)=="lambdad", c(1, 2)], 2, mean)
+se.lambdad2 <- apply(res[rownames(res)=="lambdad", c(1, 2)], 2, sd)
+lambdad2 <- paste0(round(est.lambdad2, 3), " (", round(se.lambdad2, 3), ")")
 
-	- check data in PharamcoGx R-package
+# simulation 3
+res <- read.table("results/simulations_gdsc_res3.2.txt", row.names=NULL)
+temp <- res[, 1]
+res <- as.matrix(res[, -1])
+rownames(res) <- temp
 
-	- how to transform the data and deal with missingness (ask Oslo and read papers published on GDSC and CCLE repositories)
+true.phi3 <- round(mean(1/(c(1, 1, 3, 7) %*% 
+                             t(unname(model.matrix(~ factor(1:4)))))), 3)
+est.phi3 <- mean(1/res[rownames(res)=="alphaf0", 1])
+se.phi3 <- sd(1/res[rownames(res)=="alphaf0", 1])
+phi3 <- paste0(round(est.phi3, 3), " (", round(se.phi3, 3), ")")
+est.lambdaf3 <- apply(res[rownames(res)=="lambdaf", c(1, 2)], 2, mean)
+se.lambdaf3 <- apply(res[rownames(res)=="lambdaf", c(1, 2)], 2, sd)
+lambdaf3 <- paste0(round(est.lambdaf3, 3), " (", round(se.lambdaf3, 3), ")")
 
-	- create drug covariates from SMILES formulas with R-package rcdk.
+true.chi3 <- round(mean(1/(c(1, 1, 3, 7) %*% 
+                             t(unname(model.matrix(~ factor(1:4)))))), 3)
+est.chi3 <- mean(1/res[rownames(res)=="alphad0", 1])
+se.chi3 <- sd(1/res[rownames(res)=="alphad0", 1])
+chi3 <- paste0(round(est.chi3, 3), " (", round(se.chi3, 3), ")")
+est.lambdad3 <- apply(res[rownames(res)=="lambdad", c(1, 2)], 2, mean)
+se.lambdad3 <- apply(res[rownames(res)=="lambdad", c(1, 2)], 2, sd)
+lambdad3 <- paste0(round(est.lambdad3, 3), " (", round(se.lambdad3, 3), ")")
 
-	- try non-negative matrix factorisation/independent component analysis on binary fingerprints to get principal components that characterise the different drugs.
+est.phi.ratio3 <- apply(apply(sapply(0:3, function(s) {
+  res[rownames(res)==paste0("alphaf" ,s), 2]}) %*%
+    t(cbind(1, rbind(0, diag(3)))), 2, function(m) {
+      res[rownames(res)=="alphaf0", 2]/m}), 2, mean)
+se.phi.ratio3 <- apply(apply(sapply(0:3, function(s) {
+  res[rownames(res)==paste0("alphaf" ,s), 2]}) %*%
+    t(cbind(1, rbind(0, diag(3)))), 2, function(m) {
+      res[rownames(res)=="alphaf0", 2]/m}), 2, sd)
+phi.ratio3 <- paste0(round(est.phi.ratio3, 3), " (", 
+                     round(se.phi.ratio3, 3), ")")
+est.chi.ratio3 <- apply(apply(sapply(0:3, function(s) {
+  res[rownames(res)==paste0("alphad" ,s), 2]}) %*%
+    t(cbind(1, rbind(0, diag(3)))), 2, function(m) {
+      res[rownames(res)=="alphad0", 2]/m}), 2, mean)
+se.chi.ratio3 <- apply(apply(sapply(0:3, function(s) {
+  res[rownames(res)==paste0("alphad" ,s), 2]}) %*%
+    t(cbind(1, rbind(0, diag(3)))), 2, function(m) {
+      res[rownames(res)=="alphad0", 2]/m}), 2, sd)
+chi.ratio3 <- paste0(round(est.chi.ratio3, 3), " (", 
+                     round(se.chi.ratio3, 3), ")")
+true.ratio3 <- as.numeric(1/(c(1, 1, 3, 7) %*%
+                               t(unname(model.matrix(~ factor(1:4))))))
 
-	- check whether compounds are PAINS \cite[]{baell_new_2010} or not. PAINS indicates whether compound is frequent false hitter.
+est.varb3 <- mean(1/(res[rownames(res)=="alphad0", 1]*
+                       res[rownames(res)=="alphaf0", 1]))
+se.varb3 <- sd(1/(res[rownames(res)=="alphad0", 1]*
+                    res[rownames(res)=="alphaf0", 1]))
+varb3 <- paste0(round(est.varb3, 3), " (", round(se.varb3, 3), ")")
+true.varb3 <- round(mean(
+  outer(1/as.numeric(c(1, 1, 3, 7) %*% t(unname(model.matrix(~ factor(1:4))))),
+        1/as.numeric(c(1, 1, 3, 7) %*% 
+                       t(unname(model.matrix(~ factor(1:4))))))), 3)
 
-	- start with gene expression (one $g(j)$)
+est.kurtb3 <- 3*mean(
+  1/(res[rownames(res)=="alphaf0", 1]*res[rownames(res)=="lambdaf", 1])*
+    1/(res[rownames(res)=="alphad0", 1]*res[rownames(res)=="lambdad", 1]) + 
+    1/(res[rownames(res)=="alphaf0", 1]*res[rownames(res)=="lambdaf", 1]) +
+    1/(res[rownames(res)=="alphad0", 1]*res[rownames(res)=="lambdad", 1])) + 3
+se.kurtb3 <- 3*sd(
+  1/(res[rownames(res)=="alphaf0", 1]*res[rownames(res)=="lambdaf", 1])*
+    1/(res[rownames(res)=="alphad0", 1]*res[rownames(res)=="lambdad", 1]) + 
+    1/(res[rownames(res)=="alphaf0", 1]*res[rownames(res)=="lambdaf", 1]) +
+    1/(res[rownames(res)=="alphad0", 1]*res[rownames(res)=="lambdad", 1])) + 3
+kurtb3 <- paste0(round(est.kurtb3, 3), " (", round(se.kurtb3, 3), ")")
+true.kurtb3 <- 3*mean(as.numeric(outer(
+  as.numeric(1/(c(1, 1, 3, 7) %*% t(cbind(1, rbind(0, diag(3)))))),
+  as.numeric(1/(c(1, 1, 3, 7) %*% t(cbind(1, rbind(0, diag(3)))))), 
+  FUN=function(X, Y) {X*Y + X + Y}))) + 3
+@	
+  Figure \ref{fig:simulations_gdsc_est1} shows the estimated
+	$\bm{\alpha}_{\text{feat}}$ together with its true value for NIG$_{\text{f}}$ 
+	in Scenario 1 of the simulation study (fixed $\tau_d^2$). Figure 
+	\ref{fig:simulations_gdsc_est1}a shows that estimation of
+	$\bm{\alpha}_{\text{feat}}$ is accurate. This results in accurate estimates on
+	the $\phi_{jd}$ scale as well (Figure \ref{fig:simulations_gdsc_est1}b). 
+	Model NIG$_{\text{f}}^-$ (that excludes the external covariates) gives a 
+	mean $\phi$ estimate of \Sexpr{phi1} (standard deviation between parentheses),
+	about equal to the true mean of the $\phi_{jd}$, 
+	\Sexpr{true.phi1}. Scale $\lambda_{\text{feat}}=1$ is overestimated by
+	NIG$_{\text{f}}$ at \Sexpr{lambdaf1[2]}, while NIG$_{\text{f}}^-$
+	underestimates at \Sexpr{lambdaf1[1]}.
+<<simulations_gdsc_est1, fig.cap="Simulation results for Scenario 1 ($\\tau^2_d$ fixed): estimated and true values for (a) $\\alpha_{\\text{feat}}$ and (b) prior means $\\phi_{jd}$.", out.width="100%", fig.asp=1/2>>=
+@
 
-	\subsection{GDSC}
-	After iteratively removing the cell line or drug with the highest proportion of missings until there are no more missing values, we end up with  IC50 estimates for 495 cell lines on 167 drugs.
+  Figure \ref{fig:simulations_gdsc_est2}a shows the accurately estimated
+	$\bm{\alpha}_{\text{drug}}$ together with the true value for NIG$_{\text{d}}$ 
+	in Scenario 2 of the simulation study (fixed $\gamma_{jd}^2$). Likewise, the
+	estimates are accurate on the $\chi_d$ scale 
+	(Figure \ref{fig:simulations_gdsc_est2}b). The mean $\chi$ estimate in the 
+	NIG$_{\text{d}}^-$ model is \Sexpr{chi2},  
+	which is about equal to the true mean \Sexpr{true.chi2}. Scale 
+	$\lambda_{\text{drug}}=1$ is overestimated by
+	NIG$_{\text{d}}$ at \Sexpr{lambdad2[2]} and underestimated by 
+	NIG$_{\text{d}}^-$ at \Sexpr{lambdad2[1]}.
+<<simulations_gdsc_est2, fig.cap="Simulation results for Scenario 2 ($\\gamma^2_{jd}$ fixed): estimated (boxplots) and true values (triangles) for (a) $\\alpha_{\\text{feat}}$ and (b) prior means $\\phi_{jd}$.", out.width="100%", fig.asp=1/2>>=
+@
 
-	\subsection{CCLE}
+  In Figure \ref{fig:simulations_gdsc_est3} the 
+	$\bm{\alpha}_{\text{feat}}$ and $\bm{\alpha}_{\text{drug}}$ estimated by 
+	NIG$_{\text{f}+\text{d}}$
+	are displayed together with their true values for
+	simulation Scenario 3. $\bm{\alpha}_{\text{feat}}$ are underestimated 
+	(Figure \ref{fig:simulations_gdsc_est3}a), while 
+	$\bm{\alpha}_{\text{drug}}$ (Figure \ref{fig:simulations_gdsc_est3}c) are 
+	overestimated, resulting in overestimated
+	$\phi_{jd}$ (Figure \ref{fig:simulations_gdsc_est3}b) and 
+	underestimated $\chi_{d}$ (Figure \ref{fig:simulations_gdsc_est3}d), 
+	respectively . The biases seem to 
+	be consistent though. The mean ratios $\phi_1$ to $\phi_2$, $\phi_3$, 
+	$\phi_4$ are \Sexpr{phi.ratio3[-1]}, while the mean ratios 
+	$\chi_1$ to $\chi_2$, $\chi_3$, $\chi_4$ are
+	\Sexpr{chi.ratio3[-1]}. In both cases the true values are
+	\Sexpr{true.ratio3[-1]}, so in a relative sense, the  
+	$\bm{\alpha}_{\text{feat}}$ and  
+	$\bm{\alpha}_{\text{drug}}$ estimates are about correct.
+	Moreoever, overestimation of the $\phi_{jd}$ is compensated for by the 
+	underestimation of the $\chi_{d}$: the estimated mean prior variances 
+	$\V(\beta_{jd})=\phi_{jd} \cdot \chi_{d}$ (ignoring the error variance) are 
+	unbiased (Figure \ref{fig:simulations_gdsc_var}).
+	The NIG$_{\text{f}+\text{d}}^-$ model is also consistently over- 
+	and underestimating $\phi$
+	and $\chi$ with mean estimates \Sexpr{phi3} and 
+	\Sexpr{chi3}, respectively (compared to the true mean 
+	\Sexpr{true.chi3}). Again, on the $\V(\beta_{jd})$ level, this bias almost 
+	vanishes; the mean estimated $\V(\beta_{jd})$ (ignoring error variance) are 
+	\Sexpr{varb3} while their true mean is \Sexpr{true.varb3}.
+	In Scenario 3, NIG$_{\text{f}+\text{d}}$ overestimates 
+	$\lambda_{\text{feat}}=1$ at \Sexpr{lambdaf3[2]} and underestimates
+	$\lambda_{\text{drug}}=1$ at \Sexpr{lambdad3[2]}. 
+	Similar results hold for NIG$_{\text{f}+\text{d}}^-$ with
+	$\lambda_{\text{feat}}$ estimate \Sexpr{lambdaf3[1]} and 
+	$\lambda_{\text{drug}}$ estimate \Sexpr{lambdad3[1]}. 
+<<simulations_gdsc_est3, fig.cap="Simulation results for Scenario 3: estimated (boxplots) and true values (triangles) for (a) $\\alpha_{\\text{feat}}$, (b) prior means $\\phi_{jd}$, (c) $\\alpha_{\\text{drug}}$, and (d) prior means $\\chi_{d}$.", out.width="100%", fig.asp=1>>=
+@
+<<simulations_gdsc_var, fig.cap="Simulation results for Scenario 3: mean estimated prior variances $\\V(\\beta_{jd})$ versus true values, with line of identity (dotted).", out.width="50%", fig.asp=1>>=
+@
 
-	\section{Discussion}
-	- use copula's and marginal posteriors to find multivariate posteriors for contrast testing.
+  Figure \ref{fig:simulations_gdsc_est4} displays the mean $\phi_{jd}$ and 
+  $\chi_d$ estimates for different noise levels in simulation Scenario 4. 
+  The simulation shows that with increasing noise level, the estimated 
+  prior means $\phi_{jd}$ and $\chi_d$ for the four groups of external 
+  covariates become more and more alike. In other words, noise in the external
+  covariates impedes estimation of $\bm{\alpha}_{\text{feat}}$ and
+  $\bm{\alpha}_{\text{drug}}$, as expected.
+<<simulations_gdsc_est4, fig.cap="Simulation results for Scenario 3: mean estimated prior means (a) $\\phi_{jd}$, and (b) $\\chi_{d}$ for different levels of noise in the external covariates.", out.width="100%", fig.asp=1/2>>=
+@
 
-	- empirical Bayes gives extra information via estimation of the hyper-parameters, such as informativeness of the drug information (via model for the $\gamma_d$) and relatedness of the tissues (via $\bm{\Omega}$ and $\nu$)
+  To summarise, estimation of only $\bm{\alpha}_{\text{feat}}$ or
+  $\bm{\alpha}_{\text{drug}}$ (and consequently $\phi_{jd}$ and $\chi_{d}$)
+  by NIG$_{\text{f}}$ and NIG$_{\text{d}}$, respectively is relatively unbiased,
+  as evident from simulation Scenarios 1 and 2 (Figures 
+  \ref{fig:simulations_gdsc_est1} and \ref{fig:simulations_gdsc_est2}). 
+  In contrast, simultaneous 
+  estimation in Scenario 3 results in overestimated $\phi_{jd}$ and 
+  underestimated $\chi_{d}$ by NIG$_{\text{f}+\text{d}}$
+  (Figure \ref{fig:simulations_gdsc_est3}). We conjecture that this interplay
+  of drug and feature variance components is due to near-unidentifiability.e
+  r, the resultconsequences are limited, since theiances on the $\beta_{jd}$ level are unbiasleft ed 
+  (Figure \ref{fig:simulations_gdsc_var}). If separately estimated, scale 
+  parameters $\lambda_{\text{feat}}$ and $\lambda_{\text{drug}}$ are 
+  overestimated by NIG$_{\text{f}}$ and NIG$_{\text{d}}$, respectively.
+  If estimated simultaneously NIG$_{\text{f}+\text{d}}$ overestimates 
+  $\lambda_{\text{feat}}$ and underestimated $\lambda_{\text{drug}}$. On the
+  $\beta_{jd}$ level, $\bm{\lambda}$ influences kurtoses 
+  $\mathcal{K}(\beta_{jd})$. Figure \ref{fig:simulations_gdsc_kurt} shows the
+  mean estimated kurtoses versus their true values. Kurtoses seem to be
+  underestimated by NIG$_{\text{f}+\text{d}}$. In contrast, 
+  NIG$_{\text{f}+\text{d}}^{-1}$ overestimates the true mean of the
+  $\mathcal{K}(\beta_{jd})$, \Sexpr{true.kurtb3}, at \Sexpr{kurtb3}.
+  Lastly, the results from Scenario 3 in SM Section \ref{sm-sec:simulations} 
+  show that the VB approximation is quite good as compared to standard MCMC.
+<<simulations_gdsc_kurt, fig.cap="Simulation results for Scenario 3: mean estimated prior variances $\\V(\\beta_{jd})$ versus true values, with line of identity (dotted).", out.width="50%", fig.asp=1>>=
+@
 
-	- regression on drug covariates very intuitive.
+	\section{GDSC data}\label{sec:gdsc}
+	\subsection{Primary data}
+	The GDSC project's \cite[]{yang_genomics_2013} aim is ``to improve cancer 
+	treatments by discovering therapeutic biomarkers that can be used to identify 
+	patients most likely to respond to anticancer drug". Part of the project is to
+	screen $>1000$ human cancer cell lines for drug sensitivities. The cell lines
+	have been genetically characterised and several drug sensitivity measures are
+	recorded.
+	The data is freely available from \cite{garnett_systematic_2012} and consist 
+	of: (i) the sensitivity measures of the cell lines 
+	to the drugs, (ii) annotation of the 
+	screened compounds, and (iii) the cell lines' genomic profile (mutations, 
+	copy numbers, methylation profiles, and gene expression). We will attempt to 
+	predict drug sensitivities of the cell lines, as quantified by half maximal
+	inhibitory concentration (IC50), using the
+	gene expression and gene mutation data. Other choices of sensitivity measures
+	than IC50 are 
+	possible, but a discussion on the pros and cons of different sensitivity
+	measures is beyond the aim of this paper. We have used the version of the 
+	data that is presented in \cite{iorio_landscape_2016}.
+	We averaged repeated measures over cell line-drug combinations and model the
+	logarithm of the IC50 values. In the following, IC50 refers 
+	to these log-transformed values. 
+	After removing all cell lines with missing values, we end up with 388 to 1043
+	IC50 estimates for 251 drugs. Differences in the number of cell lines
+	between drugs occur, because not all drug and cell line combinations are 
+	available. The pre-processed expression and mutation data consist of
+	17737 and 300 genes, respectively.
+	% 
+	% A comprehensive penalized regression
+	% approach such as in 
+	% {\cite{mai_composite_2019} uses most of the available molecular markers 
+	% instead of just gene expression (68 mutations, 426 copy numbers, 
+	% and 2602 gene expression levels for 498 cell lines and 97 drugs). The 
+	% current version of our software is not able to handle such problem sizes,
+	% due to the repeated calculation of posterior parameters in the empirical
+	% Bayes iterations. Currently, problems with number of outcomes in the hundreds
+	% and number of features and samples up to a thousand are computationally 
+	% feasible.
 
-	- find more literature on inverse Gaussian prior for variances and modelling variances with covariates.
+	\subsection{External data}
+	Two ternary drug covariates are available: the developmental stage
+	(experimental, in clinical development, or clinically approved) of the
+	drugs and the action (unkown, cytotoxic, or targeted) of the drugs. 
+	These drug covariates are taken 
+	directly from the GDSC database's annotation file and dummy-coded
+	with reference categories clinically approved drugs and cytotoxic drugs. 
+	We expect that drugs that have been clinically approved are easiest to 
+	predict and hence yield the largest prior $\beta_{jd}$ variances, followed by
+	the drugs in clinical development, and the experimental drugs. Likewise we 
+	expect the targeted drugs to yield the largest prior $\beta_{jd}$ variances, 
+	followed by the cytotoxic drugs, and the unkown target drugs. Note that large
+	$\beta_{jd}$ variances translate to large
+	prior $\gamma_{jd}^2$ and $\tau_d^2$ means.
+
+	Furthermore, we have a binary feature covariate available that indicates
+	whether a gene belongs to the drug target pathway. The feature covariate
+	was created by comparing the target pathways in the GDSC annotation to
+	the KEGG \cite[]{kanehisa_kegg:_2000} and reactome 
+	\cite[]{fabregat_reactome_2018} repositories. The reference category here is
+	features that are not in the target pathway. For this external covariate,
+	we expect that genes that are in the pathway of the drug
+	are more predictive than genes that are not, i.e., 
+	they have larger prior $\beta_{jd}$ variances than drugs that are not in 
+	the pathway.
+
+	The type of molecular marker may be included as external 
+	covariate, i.e., whether the feature is a gene expression or gene mutation.
+	As an alternative to direct inclusion of the mutation data, we use $p$-values
+	from the mutations as external covariate. These were obtained from a $t$-test 
+	comparing IC50 values of mutated and unmutated genes. We expect that lower 
+	mutation $p$-values result in a larger prior $\beta_{jd}$ variances.
+
+	Lastly, $p$-values from an analysis of the CCLE data 
+	\cite[]{li_landscape_2019}, a database similar to the GDSC, are included as
+	external covariate. These $p$-values are obtained from a simple correlation 
+	between the IC50 values and the gene expressions from the CCLE data. 
+	The harmonic mean per gene is then used as external covariate for the GDSC
+	data analysis. Again, we expect a positive relation between these external
+	$p$-values and the larger prior $\beta_{jd}$ variances.
+
+	\subsection{Analyses}\label{sec:analyses}
+	Four analyses were conducted: 
+	\begin{itemize}
+	  \item{Analysis 1 includes gene expressions as predictors and $p$-values from
+	    the gene mutation data as external covariate ($G=1$, $H=0$). 
+	    A pre-processing step selects between 221 and 280 genes per drug for 
+	    which both the expression as well as a mutation $p$-value is available.}
+	  \item{Analysis 2 includes both gene expressions and mutations as predictors
+	    with the feature type as external covariate, i.e., wether the feature
+	    is a expression or mutation ($G=1$, $H=0$). For this analysis,
+	    we pre-select 300 gene expressions with maximum 
+	    variance and 295 gene mutations for which there are both mutated and
+	    wildtype cell lines available.}
+	  \item{Analysis 3 uses 500 gene expressions as features, selected based on
+	    maximum variance. The CCLE $p$-values are included as external 
+	    covariates ($G=1$, $H=0$).}
+	  \item{Analysis 4 includes the gene expressions as features and both the
+	    annoted drug variables and pathway status of the genes as external 
+	    covariates ($G=1$, $H=2$, before dummy coding). 
+	    We pre-select 500 genes expressions based on maximum variance.}
+	\end{itemize}
+
+	In all analyses we estimated the same models as in the simulations (Section 
+	\ref{sec:simulations} and SM Section \ref{sm-sec:simulations}):
+	(i) NIG$_{\text{f}+\text{d}}^-$, (ii) NIG$_{\text{f}+\text{d}}$, and
+	(iii) frequentist lasso and (iv) ridge models.
+
+	In all analyses we use all cell lines to estimate the hyperparameters 
+	presented in Section \ref{sec:results}. Mean prediction means squared errors
+	(PMSE) and its standard error are estimated by 10-fold cross validation, where
+	$\text{PMSE} = D^{-1} n^{-1} \sum_{d=1}^D 
+  \sum_{i=1}^n (\mathbf{y}_d - \mathbf{x}_i \tr \hat{\bm{\beta}_d})^2$, with
+  $\hat{\bm{\beta}_d}$ the estimator for $\bm{\beta}_d$. In the NIG
+  model, that provides full posteriors, the posterior mean 
+  $\E(\bm{\beta}_d | \mathbf{y}_d)$ is used as point estimate.
+
+	\subsection{Results}\label{sec:results}
+<<analysis_gdsc>>==
+# estimates tables
+fit1 <- read.table("results/analysis_gdsc_fit1.txt", row.names=NULL)
+temp <- fit1[, 1]
+fit1 <- as.matrix(fit1[, -1])
+rownames(fit1) <- temp
+
+tab1 <- t(fit1)[c(1, 2), c(1:3)]
+rownames(tab1) <- c("NIG$_{\\text{f}+\\text{d}}^-$", 
+                    "NIG$_{\\text{f}+\\text{d}}$")
+colnames(tab1) <- c("feature intercept", "$\\log p$-value", "drug intercept")
+
+fit2 <- read.table("results/analysis_gdsc_fit2.txt", row.names=NULL)
+temp <- fit2[, 1]
+fit2 <- as.matrix(fit2[, -1])
+rownames(fit2) <- temp
+
+tab2 <- t(fit2)[c(1, 2), c(1:3)]
+rownames(tab2) <- c("NIG$_{\\text{f}+\\text{d}}^-$", 
+                    "NIG$_{\\text{f}+\\text{d}}$")
+colnames(tab2) <- c("feature intercept", "mutation", "drug intercept")
+
+fit3 <- read.table("results/analysis_gdsc_fit3.txt", row.names=NULL)
+temp <- fit3[, 1]
+fit3 <- as.matrix(fit3[, -1])
+rownames(fit3) <- temp
+
+tab3 <- t(fit3)[c(1, 2), c(1:3)]
+rownames(tab3) <- c("NIG$_{\\text{f}+\\text{d}}^-$", 
+                    "NIG$_{\\text{f}+\\text{d}}$")
+colnames(tab3) <- c("feature intercept", "$\\log p$-value", "drug intercept")
+
+fit4 <- read.table("results/analysis_gdsc_fit4.txt", row.names=NULL)
+temp <- fit4[, 1]
+fit4 <- as.matrix(fit4[, -1])
+rownames(fit4) <- temp
+
+tab4.1 <- t(fit4)[c(1, 2), c(1:2)]
+rownames(tab4.1) <- c("NIG$_{\\text{f}+\\text{d}}^-$", 
+                    "NIG$_{\\text{f}+\\text{d}}$")
+colnames(tab4.1) <- c("feature intercept", "pathway")
+tab4.2 <- t(fit4)[c(1, 2), c(3:7)]
+rownames(tab4.2) <- c("NIG$_{\\text{f}+\\text{d}}^-$", 
+                    "NIG$_{\\text{f}+\\text{d}}$")
+colnames(tab4.2) <- c("drug intercept", 
+                    "experimental", "development", "targeted", "unknown")
+
+# pmse table
+res <- read.table("results/analysis_gdsc_res1.txt", row.names=NULL)
+temp <- res[, 1]
+res <- as.matrix(res[, -1])
+rownames(res) <- temp
+
+psel1 <- res[substr(rownames(res), 1, 5)=="psel.", ]
+pmse1 <- res[substr(rownames(res), 1, 5)=="pmse.", ]
+apmse1 <- aggregate(pmse1, list(rownames(pmse1)), FUN="mean")[, -1]
+pmse1 <- aggregate(pmse1, by=list(rep(1:10, each=251)), FUN="mean")[, -1]
+mpmse1 <- apply(pmse1, 2, mean)
+sdpmse1 <- apply(pmse1, 2, sd)
+ppmse1 <- paste0(round(mpmse1, 3), " (", round(sdpmse1, 3), ")")
+
+res <- read.table("results/analysis_gdsc_res2.txt", row.names=NULL)
+temp <- res[, 1]
+res <- as.matrix(res[, -1])
+rownames(res) <- temp
+
+psel2 <- res[substr(rownames(res), 1, 5)=="psel.", ]
+pmse2 <- res[substr(rownames(res), 1, 5)=="pmse.", ]
+pmse2 <- aggregate(pmse2, by=list(rep(1:10, each=251)), FUN="mean")[, -1]
+mpmse2 <- apply(pmse2, 2, mean)
+sdpmse2 <- apply(pmse2, 2, sd)
+ppmse2 <- paste0(round(mpmse2, 3), " (", round(sdpmse2, 3), ")")
+
+res <- read.table("results/analysis_gdsc_res3.txt", row.names=NULL)
+temp <- res[, 1]
+res <- as.matrix(res[, -1])
+rownames(res) <- temp
+
+psel3 <- res[substr(rownames(res), 1, 5)=="psel.", ]
+pmse3 <- res[substr(rownames(res), 1, 5)=="pmse.", ]
+pmse3 <- aggregate(pmse3, by=list(rep(1:10, each=251)), FUN="mean")[, -1]
+mpmse3 <- apply(pmse3, 2, mean)
+sdpmse3 <- apply(pmse3, 2, sd)
+ppmse3 <- paste0(round(mpmse3, 3), " (", round(sdpmse3, 3), ")")
+
+res <- read.table("results/analysis_gdsc_res4.txt", row.names=NULL)
+temp <- res[, 1]
+res <- as.matrix(res[, -1])
+rownames(res) <- temp
+
+psel4 <- res[substr(rownames(res), 1, 5)=="psel.", ]
+pmse4 <- res[substr(rownames(res), 1, 5)=="pmse.", ]
+pmse4 <- aggregate(pmse4, by=list(rep(1:10, each=249)), FUN="mean")[, -1]
+mpmse4 <- apply(pmse4, 2, mean)
+sdpmse4 <- apply(pmse4, 2, sd)
+ppmse4 <- paste0(round(mpmse4, 3), " (", round(sdpmse4, 3), ")")
+
+tab5 <- rbind(ppmse1, ppmse2, ppmse3, ppmse4)
+tabm5 <- rbind(mpmse1, mpmse2, mpmse3, mpmse4)
+tab5.1 <- tab5[, c(1:4)]
+tabm5.1 <- tabm5[, c(1:4)]
+tab5.1 <- t(sapply(1:nrow(tab5.1), function(r) {
+  id <- which(tabm5.1[r, ]==min(tabm5.1[r, ], na.rm=TRUE))
+  out <- tab5.1[r, ]
+  out[id] <- paste0("\\textbf{", out[id], "}")
+  out}))
+rownames(tab5.1) <- paste0("analysis ", 1:nrow(tab5.1))
+colnames(tab5.1) <- c("NIG$_{\\text{f}+\\text{d}}^-$", 
+                    "NIG$_{\\text{f}+\\text{d}}$", "ridge", "lasso")
+tab5.1 <- t(tab5.1)
+
+tab5.2 <- tab5[, c(4, 13, 14, 6:8, 10:12)]
+tabm5.2 <- tabm5[, c(4, 13, 14, 6:8, 10:12)]
+tab5.2 <- t(sapply(1:nrow(tab5.2), function(r) {
+  id <- which(tabm5.2[r, ]==min(tabm5.2[r, ], na.rm=TRUE))
+  out <- tab5.2[r, ]
+  out[id] <- paste0("\\textbf{", out[id], "}")
+  out}))
+rownames(tab5.2) <- paste0("analysis ", 1:nrow(tab5.2))
+colnames(tab5.2) <- c("lasso", "lasso(25)", "lasso(100)", 
+                      "NIG$_{\\text{f}+\\text{d}}+$DSS(lasso)",
+                      "NIG$_{\\text{f}+\\text{d}}+$DSS(25)",
+                      "NIG$_{\\text{f}+\\text{d}}+$DSS(100)",
+                      "NIG$_{\\text{f}+\\text{d}}^-+$DSS(lasso)",
+                      "NIG$_{\\text{f}+\\text{d}}^-+$DSS(25)",
+                      "NIG$_{\\text{f}+\\text{d}}^-+$DSS(100)")
+tab5.2 <- t(tab5.2)
+
+mpsel <- rbind(colMeans(psel1), colMeans(psel2), colMeans(psel3), 
+               colMeans(psel4))
+
+# prior variances
+vars1 <- round(1/as.numeric(model.matrix(~ factor(c(1:3))) %*% 
+                              fit4[c(3, 6, 7), 2]), 4)
+
+# pmse ranges
+ranges1 <- apply(rbind(mpmse1, mpmse2, mpmse3, mpmse4)[, 1:4], 1, range) 
+ranges1 <- round(ranges1[2, ] - ranges1[1, ], 4)
+psels1 <- round(mpsel[, c(4, 6, 10)], 0)
+@
+  The non-zero NIG$_{\text{f}+\text{d}}$ $\bm{\alpha}$ estimates in Tables 
+  \ref{tab:analysis_gdsc_fit1}-\ref{tab:analysis_gdsc_fit4.2} show 
+  that there is an effect of the
+  external covariates. 
+  \begin{itemize}
+	  \item{Analysis 1 results in a positive additive effect of the mutation 
+	    $log p$-values on the prior $\bm{\beta}_d$ precisions
+	    (Table \ref{tab:analysis_gdsc_fit1}). More precisely, with doubling
+	    $p$-value, the prior $\bm{\beta}_d$ precision increases by 
+	    $\log(2)*\Sexpr{round(tab1[2, 2], 3)}$=\Sexpr{
+	    round(log(2)*tab1[2, 2], 3)}.
+	    This translates to more $\bm{\beta}_d$ shrinkage with increasing 
+	    $p$-value, as expected.}
+	  \item{Analysis 2 shows that gene mutations are more predictive 
+	    than gene expressions, as observed from the negative effect of mutation
+	    dummy on prior precisions (Table \ref{tab:analysis_gdsc_fit2}): 
+	    mutations are shrunken less than
+	    expressions.}
+	  \item{Analysis 3 indicates that CCLE $\log p$-values are positively related
+	    to prior precision (Table \ref{tab:analysis_gdsc_fit3}). In particular
+	    if the $p$-value doubles, the prior $\bm{\beta}_d$ precision increases by 
+	    $\log(2)*\Sexpr{round(tab3[2, 2], 3)}$=\Sexpr{
+	    round(log(2)*tab1[2, 2], 3)}, i.e., higher CCLE $p$-values results in 
+	    more shrinkage of $\bm{\beta}_d$, as expected.}
+	  \item{Analysis 4 gives a negative effect for the pathway dummy (Table 
+	    \ref{tab:analysis_gdsc_fit4.1}), indicating less shrinkage for genes
+	    that are in the drugs' target pathway, as expected. According to 
+	    expectation, experimental and 
+	    developmental drugs prior precisions are shrunken more than
+	    the reference category, approved drugs (Table 
+	    \ref{tab:analysis_gdsc_fit4.2}). Somewhat surprisingly, targeted and 
+	    unkown drugs are shrunken more than the reference, cytotoxic drugs (Table 
+	    \ref{tab:analysis_gdsc_fit4.2}), although the differences on the
+	    $\bm{\beta}_d$ variance scale are small:
+	    $\hat{\E}(\tau_{d}^{2}) = \{ \Sexpr{vars1[1]}, \Sexpr{vars1[2]}, 
+	    \Sexpr{vars1[3]} \}$ for cytotoxic, targeted, and unkown drugs, 
+	    respectively (ignoring other variance components).}
+	\end{itemize}
+<<analysis_gdsc_fit1>>=
+kableExtra::kable_styling(knitr::kable(
+  tab1, align="r", digits=2,
+  caption="$\\bm{\\alpha}$ estimates from analysis 1. Empty cells correspond to fixed zero parameters.",
+  format="latex", booktabs=TRUE, escape=FALSE),
+  latex_options=c("HOLD_position"))
+@
+<<analysis_gdsc_fit2>>=
+kableExtra::kable_styling(knitr::kable(
+  tab2, align="r", digits=2,
+  caption="$\\bm{\\alpha}$ estimates from analysis 2. Empty cells correspond to fixed zero parameters.",
+  format="latex", booktabs=TRUE, escape=FALSE),
+  latex_options=c("HOLD_position"))
+@
+<<analysis_gdsc_fit3>>=
+kableExtra::kable_styling(knitr::kable(
+  tab3, align="r", digits=2,
+  caption="$\\bm{\\alpha}$ estimates from analysis 3. Empty cells correspond to fixed zero parameters.",
+  format="latex", booktabs=TRUE, escape=FALSE),
+  latex_options=c("HOLD_position"))
+@
+  \begin{table}[h!]\label{tab:analysis_gdsc_fit4}
+    \begin{minipage}{1\textwidth}
+<<analysis_gdsc_fit4.1>>=
+kableExtra::kable_styling(knitr::kable(
+  tab4.1, align="r", digits=2,
+  caption="$\\bm{\\alpha}$ estimates from analysis 4. Empty cells correspond to fixed zero parameters.",
+  format="latex", booktabs=TRUE, escape=FALSE),
+  latex_options=c("HOLD_position"))
+@
+    \end{minipage}
+    \begin{minipage}{1\textwidth}
+<<analysis_gdsc_fit4.2>>=
+kableExtra::kable_styling(knitr::kable(
+  tab4.2, align="r", digits=2,
+  caption="$\\bm{\\alpha}$ estimates from analysis 4. Empty cells correspond to fixed zero parameters \\textit{(continued)}.",
+  format="latex", booktabs=TRUE, escape=FALSE),
+  latex_options=c("HOLD_position"))
+@
+    \end{minipage}
+  \end{table}
+
+  The mean prediction mean squared errors (PMSE), calculated on the 
+  test data, are diplayed in Table \ref{tab:analysis_gdsc_pmse1}.
+  Note that due to standardisation, an empty reference model has a PMSE of one.
+  In terms of PMSE, ridge outperforms the other models in all Analyses 1, 3, and
+  4, while NIG$_{\text{f}+\text{d}}$ $\bm{\alpha}$ performs best in Analysis 2. 
+  In general however, all models perform very similar. The ranges of mean PMSE 
+  over all methods are
+  \Sexpr{ranges1[1:3]}, and \Sexpr{ranges1[4]}, for Analyses 1-4, respectively,
+  indicating that the differences in predictive performance are very small.
+  Furthermore, the difference between NIG$_{\text{f}+\text{d}}$, that includes
+  external covariates, and NIG$_{\text{f}+\text{d}}^-$,
+  that excludes the external covariates is small; an indication that the
+  external covariates are not very informative here.
+<<analysis_gdsc_pmse1>>=  
+kableExtra::kable_styling(knitr::kable(
+  tab5.1, align="r",
+  caption="Mean (standard deviation) of cross-validated PMSE for GDSC data. Best performing model (per analysis) in bold.",
+  format="latex", booktabs=TRUE, escape=FALSE), 
+  latex_options=c("HOLD_position"))
+@
+
+  Part of the differences in performance between
+  NIG and ridge may be explained with the different levels of 
+  sparsity in the solution. Although NIG does not automatically select features, 
+  as opposed to the lasso,
+  its $\bm{\beta}_d$ prior has larger kurtosis than the ridge prior (see SM
+  Section \ref{sm-sec:prior}). The resulting heavy-tailedness as compared to 
+  the ridge prior facilitates features selection. 
+  Selection of features from a sparse prior
+  may be achieved through the decoupling shrinkage and selection approach (DSS)
+  introduced in \cite{hahn_decoupling_2015}. We have applied DSS in our 
+  analyses, where we select either (approximately) the same number of features 
+  as lasso, 25, or 50 features. Table 
+  \ref{tab:analysis_gdsc_pmse2} compares the resulting mean PMSE values. The
+  Table shows that NIG$_{\text{f}+\text{d}}$ plus DSS outperforms lasso,
+  with the same number of features.
+
+<<analysis_gdsc_pmse2>>=  
+kableExtra::kable_styling(knitr::kable(
+  tab5.2, align="r",
+  caption="Mean (standard deviation) of cross-validated PMSE for GDSC data. Best performing model (per analysis) in bold.",
+  format="latex", booktabs=TRUE, escape=FALSE), 
+  latex_options=c("HOLD_position"))
+@
+
+  To assess model fit, Section \ref{sm-sec:gdsc} in the SM displays the 
+  conditional predictive 
+  ordinates (CPO) for the NIG$_{\text{f}+\text{d}}$ model for the four analyses. 
+  A visual inspection of the CPOs learns that no extreme outliers occur.
+
+	\section{Discussion}\label{sec:discussion}
+<<time>>=
+times <- round(read.table("results/analysis_gdsc_time1.txt")[, 1]/60, 0)
+@
+	The preceding presents a novel model for drug sensitivity
+	prediction from a set of high dimensional molecular features. The model allows
+	for the inclusion of discrete and continuous external covariates on both the 
+	drugs and features. Inclusion of the external information is through 
+	data-driven and adaptive empirical Bayes estimation of the hyperparameters in 
+	the normal inverse Gaussian prior model (\ref{eq:prior}). Variational Bayes
+	estimation is efficient and scales well with the number of features and 
+	samples. Estimation of NIG$_{\text{f}+\text{d}}$ in the GDSC data analayses
+	in Section \ref{sec:gdsc} took 
+	\Sexpr{times[-1]}, and \Sexpr{times[1]} minutes on a 2016 
+	MacBook Pro with 2 GHz Dual-Core Intel Core i5 processor and 8 GB of memory, 
+	running macOS  10.15.1.
+
+	Simulation Scenarios 1 and 2
+	in Section \ref{sec:simulations} show that estimation of drug- and 
+	feature-specific hyperparameters is, in principle, fairly accurate. However, 
+	when estimated jointly, biases may occur
+	due to the interplay between the two sources of information. Fortunately,
+	these biases cancel out on the $\beta_{jd}$ level, such that the prior 
+	variance estimates $\V(\beta_{jd})$ are accurate. Ultimately, predicitive
+	performance benefits from the inclusion of external covariates, according
+	to the results in Section \ref{sm-sec:simulations} of the SM.
+
+	The model is put into practice on the GDSC data in Section \ref{sec:gdsc}. 
+	The comparison of NIG$_{\text{f}+\text{d}}$ to 
+	NIG$_{\text{f}+\text{d}}^-$ (that excludes the external covariates) shows
+	that although the inclusion of external covariates substantially modifies the
+	hyperparameters, predictive performance as measured by PMSE is only slightly
+	better in one of four analyses. The NIG model is competitive with convential, 
+	penalized methods like
+	lasso and ridge, but all three methods achieve PMSE of only 0.80 in all 
+	analyses, a 20\% reduction compared to the empty model. In three of the four 
+	analyses, ridge slightly outperforms
+	NIG, which in turn outperforms lasso. In Analysis 2, NIG slightly
+	outperforms ridge and lasso. The indications of the above are two-fold: (i) 
+	the GDSC data does not contain a lot of signal overall, and (ii)
+	the external covariates are not very informative for the GDSC data. We note
+	however, that NIG seems to have a small advantage in terms of feature 
+	selection. If we follow the DSS approach in \cite{hahn_decoupling_2015} for
+	feature selection, PMSE after selection is slightly better than lasso in all
+	four analyses.
+
+	The penalized regression methods estimate penalty 
+  parameters by cross-validation. Cross-validation directly minimises the 
+  (approximate) PMSE, as opposed to empirical Bayes in the NIG that maximises 
+  the (approximate) marginal likelihood, a measure of model fit. To achieve
+  maximal predictive accuracy, direct prediction error optimsation by 
+  cross-validation is preferred. However, direct prediction error optimisation 
+  by cross-validation is not 
+	feasible in the external covariates setting, due to the large number of 
+	hyperparameters. A caveat with penalized
+  regression methods is that they do not give measures of parameter uncertainty.
+  NIG, on the other hand, gives the full posterior of the parameters, either
+  through a variational Bayes approximation or with the Gibbs sampler from
+  SM Section \ref{sm-sec:gibbssampler}. The full posterior gives direct access
+  to the parameter uncertainties for a better interpretable model. In addition, 
+  given that the linear predictor is a linear combination of $\beta_{jd}$, and 
+  we have access to the approximate multivariate posterior of 
+  $\bm{\beta}_d$, NIG also allows to assess uncertainty of the predictions.
+
+	An alternative strategy to include external covariates is the varying 
+	coefficient (VC) model \cite[]{hastie_varying-coefficient_1993}. The VC model
+	treats the mean of the regression coefficients as a deterministic function of 
+	external covariates, as opposed to our probabilistic model for the variance
+	of the regression coefficients. \cite{ni_bayesian_2019} introduces a Bayesian 
+	VC model where the relation between the regression 
+	coefficients and external covariates is no longer deterministic, but still
+	based on the mean of the coefficients. Besides our computationally more
+	feasible VB-EM estimation, we advocate for a more indirect
+	model for the relation between external covariates and regression 
+	coefficients, i.e., through their random variance components. This indirect 
+	model assumes less structure about the relation between external covariates
+	and regression coefficients than the direct VC mean model approach.
+  In particular, a VC mean model describes both magnitude and direction of the 
+  external covariates effects, while our variance model only describes the
+  magnitude and is invariant to the direction of the effects.
+  Nonetheless, a combination of the two approaches (both mean and variance 
+  modelled as functions of the external covariates) may be a fruitful future 
+  research direction.
+  Other methods that consider an external covariate model for the variance
+  components of the regression coefficients are bSEM
+  \cite[]{leday_gene_2017,kpogbezan_empirical_2017} and \texttt{xtune}
+  \cite[]{zeng_incorporating_2020}. bSEM is designed to include
+  only dichotomous external covariates, so is not applicable in most of the
+  applications and simulations that we have considered here. \texttt{xtune}
+  allows for the inclusion of multiple external feature covariates, but not on
+  the drug level, so also has limited applicability in our simulations and
+  applications.
+
+  A possible criticism of the NIG model is the treatment of $\bm{\alpha}$ as 
+  fixed hyperparameters instead of random. A Bayesian could argue that endowment
+  of $\bm{\alpha}$ with a hyperprior results in propagation of 
+  uncertainty about $\bm{\alpha}$ and as a result improved regression parameter 
+  uncertainty quantification. \cite{van_de_wiel_learning_2019} show in a similar 
+  setting that EB estimation of hyperparameters does not necessarily lead to
+  worse uncertainty quantification as measured by frequentist coverage of
+  Bayesian credible intervals, as compared to a full Bayes treatment of the
+  hyperparameters.
+
+	Possible directions of future research are applications of the NIG model to 
+	different data types. Suitable applications are eQTL studies, 
+	in which gene expressions are regressed on SNPs. Several interesting external
+	covariates are available, both on the genes as well as the SNPs. An example of
+	such an external
+	covariate for the genes is genelength, where we suspect that longer genes are
+	harder to predict. The distance of the SNP to the gene is an example of an
+	external SNP covariate, where the expectation is that SNPs further from the 
+	gene are less predictive of that gene's expression.
+
+	\section*{Supplementary Material}
+	Supplementary Material is available online from 
+	\url{https://github.com/magnusmunch/NIG}.
+
+	\section*{Reproducible Research}
+	All results and documents may be recreated from 
+	\url{https://github.com/magnusmunch/NIG}.
+
+	\section*{Acknowledgements}
+	We thank the referees for their helpful comments and suggestions on an earlier
+	version of this manuscript.
+	MM visited SR and GL on a travel grant funded by the Amsterdam Public Health 
+	institute's Methodology program.
+	\textit{Conflict of Interest}: None declared.
 
@@
 
-	\titleformat{\section}{\normalfont\Large\bfseries}{\appendixname~\thesection.}{1em}{}
-	\begin{appendix}
-		\section{Prior tissue covariance constraints}\label{app:priortissuecovariance}
-		We parametrise the prior covariance as a matrix with $\tau$ on the diagonal and $\rho$ on its off-diagonals. To that end we set $\nu = \tau + T + 1$ and the $T \times T$-dimensional matrix
-		$$
-		\bm{\Omega} = \tau \begin{bmatrix}
-		\tau &  & \rho \\
-		& \ddots &  \\
-		\rho & & \tau
-		\end{bmatrix},
-		$$
-		such that $\E_{\pi(\mathbf{\Xi}_d)} (\mathbf{\Xi}_d) = (\nu - T - 1)^{-1} \bm{\Omega}$ is of the desired form.
-To ensure a non-degenerate prior distribution we require $\tau > 0$ and $\bm{\Omega}$ PD.
-We use $\tau > 0$ to note that a PD $\bm{\Omega}$ implies positive roots $\lambda$ of the characteristic polynomial:
-		\begin{align*}
-		\left| \begin{bmatrix}
-		\tau &  & \rho \\
-		& \ddots &  \\
-		\rho & & \tau
-		\end{bmatrix} -\lambda \I_T \right| = (\tau - \lambda - \rho)^{T-1}[\tau - \lambda - (1-T)\rho] = 0.
-		\end{align*}
-		For $T > 1$, the solutions are:
-		$$
-		(\lambda = \tau - \rho) \lor (\lambda = T \rho - \rho + \tau),
-		$$
-		which are positive if the following condition holds: $\tau > \max(\rho,\rho - T \rho)$.
-We combine this with $\tau > 0$, to arrive at the full proper prior constraint: $\tau > \max(0,\rho,\rho - T \rho)$.
+	\section*{Session info}
 
-	\end{appendix}
+<<r session-info, eval=TRUE, echo=TRUE, tidy=TRUE>>=
+devtools::session_info()
+@
 

### 2020-04-22 docs/supplement.Rnw
commit before pull

@@
 opts_chunk$set(fig.align='center', fig.path="../figs/", 
-               echo=FALSE, cache=FALSE, message=FALSE)
+               echo=FALSE, cache=FALSE, message=FALSE, fig.pos='!ht')
+knit_hooks$set(document=function(x) {
+  sub('\\usepackage[]{color}', '\\usepackage{xcolor}', x, fixed=TRUE)})
+options(knitr.kable.NA="", digits=3)
 @
@@
 bbm,algorithm,algpseudocode,pgfplotstable,threeparttable,booktabs,mathtools,
-dsfont,parskip}
+dsfont,parskip,xr}
+\externaldocument[md-]{manuscript}
 
@@
 % settings
-\pgfplotsset{compat=1.13}
+\graphicspath{{../figs/}}
+\pgfplotsset{compat=1.16}
 \setlength{\evensidemargin}{.5cm} \setlength{\oddsidemargin}{.5cm}
 \setlength{\textwidth}{15cm}
-\title{Supplementary material to: Drug efficacy prediction in cell lines}
+\title{Drug sensitivity prediction with normal inverse Gaussian shrinkage 
+informed by external data}
 \date{\today}
-\author{Magnus M.
-M\"unch$^{1,2}$\footnote{Correspondence to: \href{mailto:m.munch@vumc.nl}{m.munch@vumc.nl}}, Mark A.
-van de Wiel$^{1,3}$, Sylvia Richardson$^{3}$, \\ and Gwena{\"e}l G.
-R.
-Leday$^{3}$}
+\author{Magnus M. M\"unch$^{1,2}$\footnote{Correspondence to: 
+\href{mailto:m.munch@amsterdamumc.nl}{m.munch@amsterdamumc.nl}}, Mark A. van de 
+Wiel$^{1,3}$, Sylvia Richardson$^{3}$, \\ and Gwena{\"e}l G. R. Leday$^{3}$}
 
@@
 	\noindent
-	1. Department of Epidemiology \& Biostatistics, Amsterdam Public Health research institute, VU University Medical Center, PO Box 7057, 1007 MB
-	Amsterdam, The Netherlands\\
+	1. Department of Epidemiology \& Biostatistics, Amsterdam UMC, VU University, 
+	PO Box 7057, 1007 MB Amsterdam, The Netherlands \\
 	2. Mathematical Institute, Leiden University, Leiden, The Netherlands \\
-	3. MRC Biostatistics Unit, Cambridge Institute of Public Health, Cambridge, United Kingdom \\
+	3. MRC Biostatistics Unit, Cambridge Institute of Public Health, Cambridge,
+	United Kingdom
 
 	\section{Content overview}
-	This document contains the Supplementary Material (SM) to the document `Drug efficacy prediction in cell lines'. In the following, this document is referred to as Main Document (MD).
+	This document contains the Supplementary Material (SM) to the document 
+	`Drug sensitivity prediction with the normal inverse Gaussian prior 
+	informed by external data'. In the 
+	following, this document is referred to as Main Document (MD).
+
+	Section \ref{sec:prior} describes the moments of the normal
+	inverse Gaussian (NIG) prior, Section \ref{sec:vbderivations} gives
+	the variational Bayes (BV) derivations that lead to the estimating equations 
+	in the MD, Section \ref{sec:elbo} gives the evidence lower bound (ELBO) that 
+	is maximised during the VB iterations. In Section \ref{sec:computation} 
+	efficient VB calculations are given, while Section 
+	\ref{sec:ratiosmodifiedbessels} outlines a computationally stable method
+	to calculate ratios of modified Bessel functions. Section \ref{sec:eb} shows
+	that the empirical Bayes (EB) scale estimates $\bm{\lambda}$ are non-negative.
+	The last Section \ref{sec:gibbssampler} introduces a Gibbs sampling scheme
+	to sample from the posterior corresponding to the NIG model.
+
+	\section{Normal inverse Gaussian prior moments}\label{sec:prior}
+	Let $\beta \sim \mathcal{N}(0,\sigma^2 \tau^2 \gamma^2)$, where 
+	$\sigma^2$ and $\tau^2$ are fixed and $\gamma^2 \sim 
+	\mathcal{IG}(\phi,\lambda_{\text{feat}})$, then
+	\begin{align*}
+	  \E(\beta | \sigma^2, \tau^2) & = 0, \\
+	  \V(\beta | \sigma^2, \tau^2) & = \sigma^2 \tau^2 \phi, \\
+	  \mathcal{K}(\beta | \sigma^2, \tau^2) & = 
+	  \frac{3\phi}{\lambda_{\text{feat}}} + 3,
+	\end{align*}
+	where $\mathcal{K}(\cdot)$ denotes the kurtosis. Now letting $\tau^2
+	\sim \mathcal{IG}(\chi,\lambda_{\text{drug}})$ and integrating it out we have:
+	\begin{align*}
+	  \E(\beta | \sigma^2) & = 0, \\
+	  \V(\beta | \sigma^2) & = \sigma^2 \chi \phi, \\
+	  \mathcal{K}(\beta | \sigma^2) & = 
+	  3\(\frac{\phi}{\lambda_{\text{feat}}} \frac{\chi}{\lambda_{\text{drug}}} +
+	  \frac{\phi}{\lambda_{\text{feat}}} + \frac{\chi}{\lambda_{\text{drug}}}\) + 
+	  3.
+	\end{align*}
+
+	\section{Hyperparameter settings}\label{sec:hyperparameters}
+	For the NIG shrinkage weights prior in Figure \ref{md-fig:dens_kappa}a the 
+	following hyperparameter settings (in terms of the inverse Gaussian prior) 
+	were used: $\lambda=1$, $\phi=10$ (solid),
+	$\lambda=0.1$, $\phi=10$ (dashed), and $\lambda=1$, $\phi=1$ (dotted).
+	The Student`s $t$ prior on the regression parameters $\beta_{jd}$ arises if
+	$\gamma_{jd}^2 \sim \Gamma^{-1}(a,b)$, with $\Gamma^{-1}(a,b)$ an inverse
+	Gamma distribution with shape $a$ and scale $b$. Figure 
+	\ref{md-fig:dens_kappa}b is then obtained by setting: $a=0.9$, $b=1.3$ 
+	(solid), $a=1$, $b=4$ (dashed), and $a=1$, $b=0.2$ (dotted).
+	The lasso prior results from an exponential prior on $\gamma_{jd}^2$. To
+	create Figure \ref{md-fig:dens_kappa}c the following settings were
+	used for exponential rate $\lambda$: $\lambda=1$ (solid), $\lambda=0.2$ 
+	(dashed), and $\lambda=0.1$ (dotted).
+
+	\section{Jeffrey's prior}\label{sec:jeffreysprior}
+	We parametrise the data with a Gaussian distribution with unknown mean and 
+	variance, i.e., $y_i \sim \mathcal{N} (\mu, \sigma^2)$. This results in the
+	following Hessian (matrix of second derivatives) of the log density:
+	$$
+	\mathbf{H}(\log f(y_i | \mu, \sigma^2)) = \begin{bmatrix}
+	-\frac{1}{\sigma^2} & y_i - \mu \\
+	y_i - \mu & \frac{1}{2\sigma^4} - \frac{(y_i - \mu)^2}{\sigma^6}.
+	\end{bmatrix}
+	$$
+	Jeffreys prior \cite[]{jeffreys_invariant_1946} is now found as:
+	$$
+	\pi(\mu, \sigma^2) \propto |\mathcal{I}(\mu, \sigma^2)|^{1/2} = 
+	|-\E (\mathbf{H})|^{1/2} = \begin{vmatrix}
+	\frac{1}{\sigma^2} & 0 \\
+	0 & \frac{1}{2\sigma^4}
+	\end{vmatrix}^{1/2} \propto 1/\sigma^3,
+	$$
+	where $\mathcal{I}(\theta)$ denotes the Fisher information of $\theta$.
 
 	\section{Variational Bayes derivations}\label{sec:vbderivations}
-	In the following all expectations are with respect to the variational posterior $Q$.
+	For clarity we have indexed the variational 
+	density functions with their respective parameters, which we omitted in the 
+	MD. In the following all expectations are with respect to the 
+	variational posterior $Q_d$. The variational posterior for $\bbeta_d$ is 
+	found as follows:
 	\begin{align*}
-	\log q_{\B} (\B) & \propto \E_{\mathbf{U}, \bsigma^2} [\log \mathcal{L}(\Y | \B, \U, \bsigma^2 )] + \E_{\bgamma^2, \bsigma^2} [\log \pi (\B | \bgamma^2, \bsigma^2)] \\
-	& \propto - \frac{1}{2} \sum_{d=1}^D \sum_{i=1}^n \E_{\mathbf{u}_d, \sigma_d^2} \[ \frac{(y_{id} - \x_i \tr \bbeta_d - \mathbf{z}_i \tr \mathbf{u}_d)^2}{\sigma_d^2} \] - \frac{1}{2} \sum_{d=1}^D \sum_{j=1}^p \phi_{g(j)}^{-2} \E_{\gamma^2_d, \sigma^2_d} \( \frac{\beta_{jd}^2}{\gamma_d^2 \sigma_d^2} \), \\
-	q_{\B} (\B) & \overset{D}{=} \prod_{d=1}^{D} \mathcal{N}_p (\bmu_d, \bSigma_d), \\
-	& \text{ with } \bSigma_d = \E(\sigma_d^{-2})^{-1} [\X \tr \X + \E(\gamma_d^{-2}) \cdot \diag (\phi_{g(j)}^{-2}) ]^{-1}, \\
-	& \text{ and } \bmu_d = [\X \tr \X + \E(\gamma_d^{-2}) \cdot \diag (\phi_{g(j)}^{-2}) ]^{-1} \X \tr [\y_d - \Z \cdot \E(\mathbf{u}_d)].
+	  \log q_{\bbeta_d} (\bbeta_d) & \propto \E [\log \mathcal{L}
+	  (\y_d | \bbeta_d, \sigma_d^2 )] + \E [\log \pi (\bbeta_d | \bgamma_{d}^2, 
+	  \sigma_d^2, \tau_d^2)] \\
+	  & \propto - \frac{1}{2} \sum_{i=1}^n 
+	  \E \[ \frac{(y_{id} - \x_i \tr \bbeta_d)^2}{\sigma_d^2} \] - 
+	  \frac{1}{2} \sum_{j=1}^p 
+	  \E \( \frac{\beta_{jd}^2}{\gamma_{jd}^2 \sigma_d^2 \tau_d^2} \), \\
+	  q_{\bbeta_d} (\bbeta_d) & \overset{D}{=} 
+	  \mathcal{N}_p (\bmu_d, \bSigma_d), \\
+	  & \text{with } \bSigma_d = \E(\sigma_d^{-2})^{-1} 
+	  \{\X \tr \X + \E(\tau_d^{-2})\diag [\E(\gamma_{jd}^{-2})] \}^{-1}, \\
+	  & \text{and } \bmu_d = \{\X \tr \X + \E(\tau_d^{-2})
+	  \diag [\E(\gamma_{jd}^{-2})] \}^{-1} 
+	  \X \tr \y_d.
+	\end{align*}
+	The variational posterior for $\gamma_{jd}^2$ is given by:
+	\begin{align*}
+	  \log q_{\gamma_{jd}^{2}}(\gamma_{jd}^{2}) & \propto \E 
+	  [\log \pi (\beta_{jd} | \gamma_{jd}^{2}, \sigma_d^2, \tau_d^2)] + 
+	  \log \pi(\gamma_{jd}^{2}) \\
+	  & \propto -\frac{1}{2} \log \gamma_{jd}^{2} - \frac{1}{2} 
+	  \E \( \frac{\beta_{jd}^2}{\sigma^{2}_d \tau_d^2}\) \gamma_{jd}^{-2} - 
+	  \frac{3}{2} \log \gamma_{jd}^{2} - 
+	  \frac{\lambda_{\text{feat}} \{\gamma_{jd}^2 - 
+	  \phi_{jd}\}^2}
+	  {2\phi_{jd}^{2}} \gamma_{jd}^{-2} \\
+	  & \propto \( -1 - 1\) \log \gamma_{jd}^{2} - 
+	  \frac{\lambda_{\text{feat}}}{2\phi_{jd}^2} \gamma_{jd}^2 - 
+	  \frac{1}{2}\[ \lambda_{\text{feat}} + \E(\tau_d^{-2})\E(\sigma_d^{-2}) 
+	  \E(\beta_{jd}^2)\] \gamma_{jd}^{-2} , \\
+	  q_{\gamma_{jd}^{2}}(\gamma_{jd}^{2}) & \overset{D}{=} \mathcal{GIG} 
+	  \(-1, \lambda_{\text{feat}}/\phi_{jd}^{2}, \delta_{jd} \), \\
+	  & \text{with } \delta_{jd} =\E(\tau_d^{-2})\E(\sigma_d^{-2})
+	  \[\E(\beta_{jd})^2 + \V(\beta_{jd})\] + \lambda_{\text{feat}}.
+	\end{align*}
+	Similarly, we derive the variational posterior for $\tau_d^2$ as:
+	\begin{align*}
+	  \log q_{\tau_{d}^{2}}(\tau_{d}^{2}) & \propto \E 
+	  [\log \pi (\bbeta_{d} | \bgamma_{d}^{2}, \sigma_d^2, \tau_d^2)] + 
+	  \log \pi(\tau_{d}^{2}) \\
+	  & \propto -\frac{p}{2} \log \tau_{d}^{2} - \frac{1}{2} \sum_{j=1}^p
+	  \E \( \frac{\beta_{jd}^2}{\sigma^{2}_d \gamma_{jd}^{2}}\)\tau_d^{-2}  - 
+	  \frac{3}{2} \log \tau_{d}^{2} - 
+	  \frac{\lambda_{\text{drug}} \{\tau_{d}^2 - \chi_{d}\}^2}
+	  {2\chi_{d}^{2}} \tau_{d}^{-2} \\
+	  & \propto \( -\frac{p + 1}{2} - 1\) \log \tau_{d}^{2} - 
+	  \frac{\lambda_{\text{drug}}}{2\chi_{d}^2} \tau_{d}^2 - 
+	  \frac{1}{2}\[ \lambda_{\text{drug}} + \E(\sigma_d^{-2}) \sum_{j=1}^p 
+	  \E(\gamma_{jd}^{-2}) \E(\beta_{jd}^2)\] \tau_{d}^{-2} , \\
+	  q_{\tau_{d}^{2}}(\tau_{d}^{2}) & \overset{D}{=} \mathcal{GIG} 
+	  \(-\frac{p + 1}{2}, \lambda_{\text{drug}}/\chi_{d}^{2}, \eta_{d} \), \\
+	  & \text{with } \eta_{d} =\E(\sigma_d^{-2}) \sum_{j=1}^p
+	  \E(\gamma_{jd}^{-2})\[\E(\beta_{jd})^2 + \V(\beta_{jd})\] + 
+	  \lambda_{\text{drug}}.
+	\end{align*}
+	Lastly, we have the variational posterior for $\sigma_d^2$:
+	\begin{align*}
+	  \log q_{\sigma_d^{2}}(\sigma_d^{2}) & \propto \E 
+	  [\log \mathcal{L}(\y_d | \bbeta_d, \sigma_d^2 )] + 
+	  \E [\log \pi (\bbeta_d | \bgamma_d^2, \sigma_d^2,\tau_d^2)] + 
+	  \log \pi (\sigma_d^{2}) \\
+	  & \propto -\frac{n}{2} \log \sigma_d^{2} - \frac{1}{2} \sum_{i=1}^n 
+	  \frac{\E [(y_{id} - \x_i \tr \bbeta_d)^2]}{\sigma_d^{2}} - 
+	  \frac{p}{2} \log \sigma_d^{2} - \frac{1}{2} \sum_{j=1}^p 
+	  \E \( \frac{\beta_{jd}^2}{\gamma_{jd}^{2}\tau_d^2}\) \sigma_d^{-2} - 
+	  \frac{3}{2} \log \sigma^2_d \\
+	  & = \(-\frac{n + p + 1}{2} - 1 \) \log \sigma_d^{2} \\
+	  & \,\,\,\,\,\,\,\,\,\, - 
+	  \frac{1}{2}\left\{ \sum_{i=1}^n \E [(y_{id} - \x_i \tr \bbeta_d)^2] + 
+	  \E(\tau_d^{-2}) \sum_{j=1}^p \E(\gamma_{jd}^{-2})
+	  \E(\beta_{jd}^2) \right\} \sigma_d^{-2}, \\
+	  q_{\sigma_d^{2}}(\sigma_d^{2}) & \overset{D}{=} 
+	  \Gamma^{-1} \(\frac{n + p + 1}{2}, \zeta_{d} \), \\
+	  & \text{with } \zeta_{d} = \frac{1}{2} \bigg\{\mathbf{y}_d \tr 
+	  \mathbf{y}_d - 2 \mathbf{y}_d \tr \X \E(\bbeta_d) + 
+    \trace [ \X \tr \X \V(\bbeta_d)] + 
+    \E(\bbeta_d \tr)  \X \tr \X \E(\bbeta_d ) \\
+    & \,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\, + 
+    \E(\tau_d^{-2})\trace \{ \diag[\E(\gamma_{jd}^{-2})] \V(\bbeta_d)\} + 
+    \E(\tau_d^{-2})\E(\bbeta_d \tr) \diag[\E(\gamma_{jd}^{-2})] \E(\bbeta_d )
+    \bigg\}.
+  \end{align*}
+  If we fill in the expectations and variances, we arrive at the estimating
+  equations in MD equations (\ref{md-eq:vbequations}).
+
+  \section{Evidence lower bound}\label{sec:elbo}
+  The evidence lower bound that is maximised during the iterations is: 
+  $\text{ELBO}^{(l)}=\sum_{d=1}^D \text{ELBO}_d^{(l)}$, with:
+  \begin{align*}
+    \text{ELBO}&^{(l)}_d \\
+    & = -\frac{n+p+1}{2} \log \pi + \frac{p+1-n}{2}\log 2 + \frac{n+1}{2} + p
+    + \log \Gamma \(\frac{n + p + 1}{2} \) \\
+    & - \frac{n + p + 1}{2} \log \zeta_d^{(l)} + 
+    \frac{1}{2} \log |\bSigma_d^{(l)}| - 
+    \frac{n+p+1}{4\zeta_d^{(l)}}  \bigg[\mathbf{y}_d \tr \mathbf{y}_d -
+    2 \mathbf{y}_d \tr \X \bm{\mu}_d^{(l)} + 
+    \trace ( \X \tr \X \bm{\Sigma}_d^{(l)}) \\ 
+    & + (\bm{\mu}_d^{(l)}) \tr \X \tr \X \bm{\mu}_d^{(l)} +
+    g_{d}^{(l)} \trace \[ \diag (b_{jd}^{(l)}) \bm{\Sigma}_d^{(l)}\] + 
+    g_{d}^{(l)} (\bm{\mu}_d^{(l)}) \tr \diag (b_{jd}^{(l)}) 
+    \bm{\mu}_d^{(l)}\bigg] + p \log \lambda_{\text{feat}}^{(l)} \\
+    & + \frac{p + 3}{4} \log \lambda_{\text{drug}}^{(l)} + 
+    \frac{\eta_d^{(l)} - \lambda_{\text{drug}}^{(l)}}{2} g_d^{(l)} + 
+    \sum_{j=1}^p \frac{\delta_{jd}^{(l)} - \lambda_{\text{feat}}^{(l)}}{2} 
+    b_{jd}^{(l)} + \frac{\lambda_{\text{drug}}^{(l)}}{\chi_d^{(l)}} +
+    \sum_{j=1}^p \frac{\lambda_{\text{feat}}^{(l)}}{\phi_{jd}^{(l)}} + \\
+    & - \frac{p + 1}{2} \log \chi_d^{(l)} - \sum_{j=1}^p \log \phi_{jd}^{(l)}
+    -\frac{p+1}{4} \log \eta_d^{(l)} - 
+    \frac{1}{2} \sum_{j=1}^p \log \delta_{jd}^{(l)} \\
+    & + \log K_{\frac{p+1}{2}}\(\sqrt{\eta_{d}^{(l)}
+    \lambda_{\text{drug}}^{(l)}}/
+    \chi_{d}^{(l)}\) + \sum_{j=1}^p \log K_{1}\(\sqrt{\delta_{jd}^{(l)}
+    \lambda_{\text{feat}}^{(l)}}/\phi_{jd}^{(l)}\).
   \end{align*}
 
-	\begin{align*}
-	\log q_{\U}(\U) & \propto \E_{\mathbf{B}, \bsigma^2} [\log \mathcal{L}(\Y | \B, \U, \bsigma^2 )] + \E_{\mathbf{\Xi}_1, \dots, \mathbf{\Xi}_D, \bsigma^2} [\log \pi (\U | \mathbf{\Xi}_1, \dots, \mathbf{\Xi}_D, \bsigma^2)] \\
-	& \propto - \frac{1}{2} \sum_{d=1}^D \sum_{i=1}^n \E_{\bbeta_d, \sigma_d^2} \[ \frac{(y_{id} - \x_i \tr \bbeta_d - \mathbf{z}_i \tr \mathbf{u}_d)^2}{\sigma_d^2} \] - \frac{1}{2} \sum_{d=1}^D \E_{\bm{\Xi}_d, \sigma_d^2}\( \frac{\mathbf{u}_d \tr \bm{\Xi}_d^{-1} \mathbf{u}_d}{\sigma_d^2} \), \\
-	q_{\U}(\U) & \overset{D}{=} \prod_{d=1}^{D} \mathcal{N}_T (\mathbf{m}_d, \Sm_d), \\
-	& \text{ with } \Sm_d = \E(\sigma_d^{-2})^{-1} [\Z \tr \Z + \E(\bm{\Xi}_d^{-1}) ]^{-1}, \\
-	& \text{ and } \mathbf{m}_d = [\Z \tr \Z + \E(\bm{\Xi}_d^{-1}) ]^{-1} \Z \tr [\y_d - \X \cdot \E(\bbeta_d)].
-	\end{align*}
-
-	\begin{align*}
-	\log q_{\bgamma^{2}}(\bgamma^{2}) & \propto \E_{\B, \bsigma^2} [\log \pi (\B | \bgamma^{2}, \sigma^2)] + \log \pi (\bgamma^{2}) \\
-	& \propto -\frac{p}{2} \sum_{d=1}^D \log \gamma_d^{2} - \frac{1}{2} \sum_{d=1}^D \sum_{j=1}^p \phi_{g(j)}^{-2} \E \( \frac{\beta_{jd}^2}{\sigma^{2}_d}\) \gamma_d^{-2} - \frac{3}{2} \sum_{d=1}^D \log \gamma_d^{2} \\
-	& \,\,\,\,\,\,\,\,\,\, - \frac{\lambda}{2 \theta^2} \sum_{d=1}^D (\gamma_d^2 - \theta)^2 \gamma_d^{-2} \\
-	& \propto \sum_{d=1}^D \left\{ \( -\frac{p + 1}{2} - 1\) \log \gamma_d^{2} - \frac{\lambda}{2\theta^2} \gamma_d^2 - \frac{1}{2}\[ \lambda + \E(\sigma_d^{-2}) \sum_{j=1}^p \frac{\E(\beta_{jd}^2)}{\phi_{g(j)}^{2}}\] \gamma_d^{-2} \right\}, \\
-	q_{\bgamma^{2}}(\bgamma^{2}) & \overset{D}{=} \prod_{d=1}^D \mathcal{GIG} \(-\frac{p+1}{2}, \frac{\lambda}{\theta_d^2}, \delta_{d} \), \\
-	& \text{with } \delta_{d} =\E(\sigma_d^{-2}) \sum_{j=1}^p \frac{\E(\beta_{jd})^2 + \V(\beta_{jd})}{\phi_{g(j)}^2} + \lambda.
-	\end{align*}
-
-	\begin{align*}
-	\log q_{\bsigma^{2}}(\bsigma^{2}) & \propto \E_{\B,\mathbf{U}} [\log \mathcal{L}(\Y | \B, \U, \bsigma^2 )] + \E_{\B, \bgamma^2} [\log \pi (\B | \bgamma^2, \bsigma^2)] \\
-	& \,\,\,\,\,\,\,\,\,\, + \E_{\mathbf{U}, \bm{\Xi}_1, \dots, \bm{\Xi}_D} [\log \pi (\mathbf{U} | \bm{\Xi}_1, \dots, \bm{\Xi}_D, \bsigma^2)] + \log \pi (\bsigma^{2}) \\
-	& \propto -\frac{n}{2} \sum_{d=1}^D \log \sigma_d^{2} - \frac{1}{2} \sum_{d=1}^D \sum_{i=1}^n \E [(y_{id} - \x_i \tr \bbeta_d - \mathbf{z}_i \tr \mathbf{u}_d)^2] \sigma_d^{2} - \frac{p}{2} \sum_{d=1}^D \log \sigma_d^{-2} \\
-	& \,\,\,\,\,\,\,\,\,\, - \frac{1}{2} \sum_{d=1}^D \sum_{j=1}^p \E( \gamma_d^{-2}) \frac{\E ( \beta_{jd}^2)}{\phi_{g(j)}^{2}} \sigma_d^{-2} - \frac{T}{2} \sum_{d=1}^D \log \sigma_d^{2} - \frac{1}{2} \sum_{d=1}^D \E(\mathbf{u}_d \tr \bm{\Xi}_d^{-1} \mathbf{u}) \sigma_d^{-2} \\
-	& \,\,\,\,\,\,\,\,\,\, - \frac{3}{2} \sum_{d=1}^D \log \sigma^2_d \\
-	& = -\sum_{d=1}^{D} \(\frac{n + p + T + 3}{2} \) \log \sigma_d^{2} - \sum_{d=1}^{D} \frac{1}{2}\Bigg\{ \sum_{i=1}^n \E [(y_{id} - \x_i \tr \bbeta_d - \mathbf{z}_i \tr \mathbf{u}_d)^2] \\
-	& \,\,\,\,\,\,\,\,\,\, + \E(\gamma_d^{-2})\sum_{j=1}^p \frac{\E(\beta_{jd}^2)}{\phi_{g(j)}^2} + \E(\mathbf{u}_{d} \tr \bm{\Xi}_d^{-1} \mathbf{u}_d) \Bigg\} \sigma_d^{-2}, \\
-	q_{\bsigma^{2}}(\bsigma^{2}) & \overset{D}{=} \prod_{d=1}^D \Gamma^{-1} \(\frac{n + p + T + 1}{2}, \zeta_{d} \), \\
-	& \text{ with } \zeta_{d} = \frac{\y_d \tr \y_d}{2} - \y_d \tr [\X \E(\bbeta_d) + \Z \E(\mathbf{u}_d)] + \E(\bbeta_d \tr) \X \tr \Z \E (\mathbf{u}_d) \\
-	& \,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\, + \frac{\trace [\X \tr \X \V(\bbeta_d)]}{2} + \frac{\E(\bbeta_d \tr) \X \tr \X \E(\bbeta_d)}{2} + \frac{\trace [\Z \tr \Z \V(\mathbf{u}_d)]}{2} \\
-	& \,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\, + \frac{\E(\mathbf{u}_d \tr) \Z \tr \Z \E(\mathbf{u}_d)}{2} + \frac{\E(\gamma_d^{-2})}{2} \sum_{j=1}^p \frac{\E(\beta_{jd})^2 + \V(\beta_{jd})}{\phi_{g(j)}^2} \\
-	& \,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,+ \frac{\E(\mathbf{u}_d \tr) \E(\bm{\Xi}_d^{-1}) \E(\mathbf{u}_d)}{2} + \frac{\trace [\E(\bm{\Xi}_d^{-1}) \V(\mathbf{u}_d)]}{2},
-	\end{align*}
-
-	\begin{align*}
-	\log q_{\bm{\Xi}_1, \dots, \bm{\Xi}_D}(\bm{\Xi}_1, \dots, \bm{\Xi}_D) & \propto \E_{\mathbf{U}, \bm{\sigma}^2} [\log \pi (\mathbf{U} | \bm{\Xi}_1, \dots, \bm{\Xi}_D, \bsigma^2)] + \log \pi (\bm{\Xi}_1, \dots, \bm{\Xi}_D) \\
-	& \propto -\frac{1}{2} \sum_{d=1}^D \log |\bm{\Xi}_d| - \frac{1}{2} \sum_{d=1}^D \E \(\frac{\mathbf{u}_d \tr \bm{\Xi}_d^{-1} \mathbf{u}_d}{\sigma_d^2} \) \\
-	& \,\,\,\,\,\,\,\,\,\, - \frac{k + T + 1}{2} \sum_{d=1}^D \log |\bm{\Xi}_d| - \frac{1}{2} \sum_{d=1}^D \trace (\bm{\Omega} \bm{\Xi}_d^{-1}) \\
-	& = -\frac{\tau + T + 2}{2} \sum_{d=1}^D \log |\bm{\Xi}_d| - \frac{1}{2} \sum_{d=1}^D \trace \left\{ \[\E(\sigma_d^{-2}) \E( \mathbf{u}_d \mathbf{u}_d \tr) + \bm{\Omega}\] \bm{\Xi}_d^{-1} \right\}, \\
-	q_{\bm{\Xi}_1, \dots, \bm{\Xi}_D}(\bm{\Xi}_1, \dots, \bm{\Xi}_D) & \overset{D}{\propto} \prod_{d=1}^D \mathcal{W}_T^{-1} (\bm{\Psi}_d, \nu + 1),  \\ 
-	& \text{ with } \bm{\Psi}_d =\E(\sigma^{-2}_d) \[\V(\mathbf{u}_d) + \E(\mathbf{u}_d) \E( \mathbf{u}_d \tr) \] + \bm{\Omega}.
-	\end{align*}
-
-	\section{Simulations}
-	\subsection{Models}
-	In order to investigate the inverse Gaussian prior model and its estimation, we consider a simpler model. The simpler model ignores tissue effects and fixes $\forall g: \phi_g^2=1$:
-	\begin{align*}
-  y_{id} | \bm{\beta}_d, \sigma_d^2 & \sim \mathcal{N} (\x_i \tr \bm{\beta}_d, \sigma_d^2), \\
-  \beta_{jd} | \gamma_d^2, \sigma_d^2 & \sim p(\beta_{jd}), \\
-  \gamma_d^2 & \sim p(\gamma_d^2), \\
-  \sigma_d^2 & \sim 1/\sigma_d^3.
-  \end{align*}
-
-  \cite{moran_variance_2018} argue that in regression models, $\bm{\beta}_d$ and $\sigma_d^2$ should be independent \textit{a priori}.
-However, in our case that is debatable.
-The $D$ regressions need not be on the same scale, so some form of calibration of the $\bm{\beta}_d$ posteriors to this scale is required.
-A simple way of doing this is to include the scales $\sigma_d^2$ in the priors for the $\bm{\beta}_d$.
-We call the two versions of the model conjugate (with $\bm{\beta}_d$ and $\sigma_d^2$ dependent):
-  $$
-  \beta_{jd} | \gamma_d^2, \sigma_d^2 \sim \mathcal{N} (0, \sigma_d^2 \gamma_d^2)
-  $$
-  and non-conjugate (with $\bm{\beta}_d$ and $\sigma_d^2$ independent):
-  $$
-  \beta_{jd} | \gamma_d^2, \sigma_d^2 \sim \beta_{jd} | \gamma_d^2 \sim \mathcal{N} (0, \gamma_d^2),
-  $$
-  and compare them in our simulations.
-
-	In addition to the inverse Gaussian model for the prior variances $\gamma_d^2$ (introduced in the MD):
-	$$
-  \gamma_d^2 \sim \mathcal{IG}(\theta_d, \lambda_d),
-  $$
-	we investigate a generalisation of the models in \cite{kpogbezan_empirical_2017} and \cite{leday_gene_2017}, where we draw the $\gamma_d^2$ from an inverse Gamma distribution; :
-	$$
-	\gamma_d^2 \sim \Gamma^{-1}(\eta_d/2, \lambda_d/2).
-	$$
-	For the sake of comparability, we have rescaled the inverse Gamma parameters and reparametrised both distributions as generalized inverse Gaussians:
-	\begin{align*}
-  \gamma_d^2 & \sim \mathcal{GIG}(-1/2, \lambda_d \theta_d^{-2}, \lambda_d), \\
-  \gamma_d^2 & \sim \mathcal{GIG}(-\eta_d/2, 0, \lambda_d).
-  \end{align*}
-	The two models coincide if $\theta_d \to \infty$ and $\eta_d = 1$.
-
-	\subsection{Estimation}
-	The variational distributions are as follows:
-	\begin{align*}
-  q(\bm{\beta}) & \overset{D}{=} \prod_{d=1}^D \mathcal{N}_p (\bm{\mu}_d, \bm{\Sigma}_d), \\
-  q(\bm{\gamma}^2) & \overset{D}{=} \prod_{d=1}^D \mathcal{GIG}\(-\frac{p+\eta_d}{2}, \lambda_d \theta_d^{-2}, \delta_d\), \\
-  q(\bm{\sigma}^2) & \overset{D}{=} \prod_{d=1}^D \Gamma^{-1} \(\frac{df + 1}{2}, \zeta_d\),
-  \end{align*}
-	where $\eta_d=1$ and $\theta_d \to \infty$ in the inverse Gaussian and inverse Gamma models, respectively.
-Furthermore, $df=n + p$ in the conjugate model and $df=n$ in the non-conjugate model.
-The variational parameters contain cyclic dependencies and are, in the conjugate setting, iteratively updated by:
-  \begin{align*}
-  \bm{\Sigma}_d^{(h+1)} & = (a_d^{(h)})^{-1} (\X \tr \X + c_d^{(h)} \I)^{-1}, \\
-  \bm{\mu}_d^{(h+1)} & = (\X \tr \X + c_d^{(h)} \I)^{-1} \X \tr \y_d, \\
-  \delta_d^{(h+1)} & = b_d^{(h)} (c_d^{(h)})^{-1} \bmu_d \tr \bmu_d + \lambda_d, \\ 
-  \zeta_d^{(h+1)} & = \frac{1}{2} \bigg[\mathbf{y}_d \tr \mathbf{y}_d -2 \mathbf{y}_d \tr \X \bm{\mu}_d^{(h+1)} + \trace ( \X \tr \X \bm{\Sigma}_d^{(h+1)}) + (\bm{\mu}_d^{(h+1)}) \tr \X \tr \X \bm{\mu}_d^{(h+1)} \\
-  & \,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\, + \frac{b_d^{(h+1)} - c_d^{(h+1)} a_d^{(h+1)}}{1 - a_d^{(h+1)}} \trace ( \bm{\Sigma}_d^{(h+1)}) + \frac{b_d^{(h+1)} - c_d^{(h+1)} a_d^{(h+1)}}{1 - a_d^{(h+1)}} (\bm{\mu}_d^{(h+1)}) \tr \bm{\mu}_d^{(h+1)}\bigg].
-  \end{align*}
-  Here, we set
+  \section{Efficient computation}\label{sec:computation}
+  The empirical Bayes updates require the quantities
+  $e_{jd}$, $b_{jd}$, $g_d$, and $f_d$, which in turn depend on the quantities 
+  $\diag(\bSigma_d)$, $\bmu_{d}$, $\y_d \tr \X \bmu_d$, 
+  $\trace(\X \tr \X \bSigma_d)$, $\bmu_d \tr \X \tr \X \bmu_d$,
+  $g_d\trace[\diag(b_{jd})\bSigma_d]$, and $g_d\bmu_d \tr \diag(b_{jd}) \bmu_d$.
+  Let $\mathbf{H}_d=g_d \diag[(b_{jd}^{(h)})^{-1}]$. Then, the first two
+  quantities are efficiently calculated as:
+  \begin{subequations}\label{eq:complexity1}
     \begin{align}
-  a_d^{(h)} & =\E_{Q^{(h)}}(\sigma_d^{-2})=(df+1)/(2 \zeta_d^{(h)}), \nonumber \\
-  b_d^{(h)} & = \E_{Q^{(h)}}(\gamma_d^{-2}) = \sqrt{\frac{\lambda_d}{\theta_d^2 \delta_d^{(h)}}} \frac{K_{\frac{p + \eta_d -2}{2}} \( \sqrt{\lambda_d \delta_d^{(h)}}/\theta_d \)}{K_{\frac{p+\eta_d}{2}} \( \sqrt{\lambda_d \delta_d^{(h)}}/\theta_d \)} + \frac{p+\eta_d^{(h)}}{\delta_d^{(h)}}, \label{eq:auxvar1} \\
-  c_d^{(h)} & = \begin{cases*}
-  b_d^{(h)} & in the conjugate setting \\
-  b_d^{(h)}/a_d^{(h)} & in the non-conjugate setting .
-  \end{cases*} \nonumber
-  \end{align}
-  Note that in the inverse Gamma model, the computation of (\ref{eq:auxvar1}) simplifies to $b_d^{(h)}=(p+\eta_d)/\delta_d^{(h)}$.
-
-  \subsection{Empirical Bayes}
-  To avoid subjectivity in our choice of prior, we estimate the hyper-parameters with empirical Bayes.
-The empirical Bayes procedure differs for the two prior models.
-In the inverse Gaussian case, we parametrise $\theta_d = (\mathbf{c}_d \tr \bm{\alpha})^{-1}$ to allow for the inclusion of drug covariates in the empirical Bayes estimation.
-The estimating equations are:
-  \begin{align*}
-  \bm{\alpha} & = \[ \mathbf{C} \tr \diag (\lambda_d e_d^{(l)}) \mathbf{C} \]^{-1} \mathbf{C} \tr \diag (\lambda_d) \mathbf{1}_{D \times 1}, \\
-  \lambda_d & = \[ b_d^{(l)} + e_d^{(l)} (\mathbf{c}_d \tr \bm{\alpha})^2 - 2\mathbf{c}_d \tr \bm{\alpha} \]^{-1},
-  \end{align*}
-  where $e_d^{(l)} = \E_{Q^{(l)}}(\gamma_d^2) = [b_d^{(l)} - (p + \eta_d^{(h)})/\delta_d^{(l)})] \cdot \delta_d^{(l)} (\theta_d^{(h)})^2/\lambda_d^{(l)}$.
-Solving the equations is done by iteratively reweighted least squares of responses $(e_d^{(l)})^{-1}$ on predictors $\mathbf{c}_d$ with weights $\lambda_d e_d^{(l)}$.
-If we use one common $\lambda=\lambda_d$ then the solution is in closed form:
-  \begin{align*}
-	\bm{\alpha}^{(l+1)} & = \[ \mathbf{C} \tr \diag(e_d^{(l)}) \mathbf{C} \]^{-1} \mathbf{C} \tr \mathbf{1}_{D \times 1}, \\
-	\lambda^{(l+1)} & = \[ \sum_{d=1}^D b_d^{(l)} - (\bm{\alpha}^{(l+1)}) \tr \mathbf{C} \tr \mathbf{1} \]^{-1} D.
-	\end{align*}
-	This corresponds to the model in the MD.
-
-  In the inverse Gamma model we use the following empirical Bayes procedure: Suppose we have a partitioning of our drugs into drug classes and we let prior variance components $\gamma_d^2$ and $\gamma^2_{d'}$ be drawn from the same distribution if drug $d$ and drug $d'$ are from the same class.
-Or simply put: $\gamma^2_{d}, \gamma^2_{d'} \sim \Gamma^{-1}(\eta_c/2, \lambda_c/2)$ if $d, d' \in \text{class}_c$.
-Empirical bayes estimation of the $\eta_c$ and $\lambda_c$ by maximisation of the (approximated) marginal likelihood amounts to iteratively solving the estimating equations:
-  \begin{subequations}
-    \begin{align}
-      \psi\(\frac{\eta_c}{2}\) & = \log \lambda_c - \frac{\sum_{d \in \text{class}_c} e^{(l)}_d}{|\text{class}_c|} - \log 2, \label{eq:estequation1}\\
-      \lambda_c & = \eta_c \cdot |\text{class}_c| \cdot \(\sum_{d \in \text{class}_c} b_d^{(l)}\)^{-1}, \label{eq:estequation2}
+      \diag (\bm{\Sigma}_d^{(h+1)}) & = (a_d^{(h)})^{-1} 
+      \H_d^{(h)} - (a_d^{(h)})^{-1} \{ [ \H_d^{(h)} \X \tr 
+      ( \I_n + \X \H_d^{(h)} \X \tr )^{-1} ] \circ
+      (\H_d^{(h)} \X \tr ) \} \mathbf{1}_{n \times 1}, \\
+      \bmu_d^{(h+1)}) & = \H_d^{(h)} \X \tr \y_d - \H_d^{(h)} \X \tr 
+      ( \I_n + \X \H_d^{(h)} \X \tr )^{-1} \X \H_d^{(h)} \X \tr \y_d.
     \end{align}
   \end{subequations}
-  until convergence, where $e^{(l)}_d = \E_{Q^{(l)}}(\log \gamma_d^2) = \log \delta_d^{(l)} - \psi [ (p + \eta_d^{(l)})/2] - \log 2$. 
-  We propose to do that by first solving (\ref{eq:estequation2}) for $\lambda_c$, plugging it into (\ref{eq:estequation1}) and bounding the result using \cite{alzer_inequalities_1997} to obtain an interval $( \eta_c^*, 2\eta_c^* )$ containing the solution to (\ref{eq:estequation1}), where:
-  $$
-  \eta_c^* = \left\{ \frac{\sum_{d \in \text{class}_c} e^{(l)}_d}{|\text{class}_c|} + \log \[ \frac{\sum_{d \in \text{class}_c} b_d^{(l)}}{|\text{class}_c|} \]  \right\}^{-1}.
-  $$
-  With this interval it is straightforward to find $\alpha^{(l+1)}_c$ by any root-finding algorithm and plugging the solution into (\ref{eq:estequation2}) to find $\lambda^{(l+1)}_c$.
-We may use $\eta_c^*$ as a starting value.
+  The quantities $\y_d \tr \X \bmu_d$, $\bmu_d \tr \X \tr \X \bmu_d$,
+  $g_d\trace[\diag(b_{jd})\bSigma_d]$, and $g_d\bmu_d \tr \diag(b_{jd}) \bmu_d$
+  are easily calculated from the first two. For the remaining one we have:
+  \begin{align}\label{eq:complexity2}
+    \trace(\X \tr \X \bSigma_d^{(h+1)})  = 
+    & (a_d^{(h)})^{-1} \mathbf{1}_{1 \times p} \H_d^{(h)} \X \tr 
+    \mathbf{1}_{n \times 1} \nonumber \\  
+    & - (a_d^{(h)})^{-1} \mathbf{1}_{1 \times p} \{ [  \H_d^{(h)} \X \tr 
+    ( \I_n + \X \H_d^{(h)} \X \tr )^{-1} \X \H_d^{(h)} \X \tr ] \circ 
+    \X \tr \} \mathbf{1}_{n \times 1}.
+  \end{align}
+  Both (\ref{eq:complexity1}) and (\ref{eq:complexity2}) are operations of
+  complexity $\mathcal{O}(pn^2)$ instead of $\mathcal{O}(p^3)$ for naive
+  calculation.
 
-	The inverse Gamma model with independent drug classes allows for more flexible empirical Bayes estimation of the prior means for the different drug classes.
-As a consequence however, the risk of overfitting is increased.
-Additionally, it does not allow to include continuous covariates on the drugs.
-
-	\subsection{Evidence lower bound}
-	We monitor convergence through the evidence lower bound (ELBO).
-To compute the evidence lower bound after updating the EB parameters from $\eta_d^{(l)}, \theta_d^{(l)}, \lambda_d^{(l)}$ to $\eta_d^{(l + 1)}, \theta_d^{(l + 1)}, \lambda_d^{(l + 1)}$, we compute $\text{ELBO}^{(l + 1)} = \sum_{d=1}^D \text{ELBO}_d^{(l + 1)}$, with, in the inverse Gaussian case:
-	\begin{align*}
-	\text{ELBO}_d^{(l + 1)} & = \frac{1}{2} \log |\bm{\Sigma}_d^{(l)}| - \frac{a_d^{(l)}}{2} \[ \y_d \tr \y_d - 2\y_d \tr \X \bm{\mu}_d^{(l)} + (\bm{\mu}_d^{(l)}) \tr \X \tr \X \bm{\mu}_d^{(l)} + \trace ( \X \tr \X \bm{\Sigma}_d^{(l)})\] \\
-	& + \frac{b_d^{(l)}}{2} \[ \delta_d^{(l)} - (a_d^{(l)} + 1 - b_d^{(l)}/c_d^{(l)})(\bm{\mu}_d^{(l)}) \tr \bm{\mu}_d^{(l)} - (a_d^{(l)} + 1 - b_d^{(l)}/c_d^{(l)}) \trace (\bm{\Sigma}_d^{(l)}) \] \\
-	& - \frac{df + 1}{2} \log \zeta_d^{(l)} + \frac{p + 1}{4} \log \lambda_d^{(l)} - \frac{p + 1}{2} \log \theta_d^{(l)} - \frac{p+1}{4} \log \delta_d^{(l)} \\
-	& + \log K_{\frac{p+1}{2}}\(\sqrt{\lambda_d^{(l)} \delta_d^{(l)}}/\theta_d^{(l)}\) + \frac{\lambda_d^{(l)} e_d^{(l)}}{2(\theta_d^{(l)})^2} \\
-	& + \frac{\lambda_d^{(l+1)}}{\theta_d^{(l+1)}} + \frac{1}{2}\log \lambda_d^{(l+1)} - \frac{\lambda_d^{(l+1)} e_d^{(l)}}{2 (\theta_d^{(l + 1)})^2} - \frac{\lambda_d^{(l + 1)} b_d^{(l)}}{2}.
-	\end{align*}
-
-	In the inverse Gamma setting, we have:
-	\begin{align*}
-	\text{ELBO}_d^{(l + 1)} & = \frac{1}{2} \log |\bm{\Sigma}_d^{(l)}| - \frac{a_d^{(l)}}{2} \[ \y_d \tr \y_d - 2\y_d \tr \X \bm{\mu}_d^{(l)} + (\bm{\mu}_d^{(l)}) \tr \X \tr \X \bm{\mu}_d^{(l)} + \trace ( \X \tr \X \bm{\Sigma}_d^{(l)})\] \\
-	& + \frac{b_d^{(l)}}{2} \[ \delta_d^{(l)} - (a_d^{(l)} + 1 - b_d^{(l)}/c_d^{(l)})(\bm{\mu}_d^{(l)}) \tr \bm{\mu}_d^{(l)} - (a_d^{(l)} + 1 - b_d^{(l)}/c_d^{(l)}) \trace (\bm{\Sigma}_d^{(l)}) \]  \\
-	& - \frac{df + 1}{2} \log \zeta_d^{(l)} - \frac{p + \eta_d^{(l)}}{2} \log \delta_d^{(l)} + \frac{\eta_d^{(l)}}{2} e_d^{(l)} + \frac{\eta_d^{(l)}}{2} \log 2 + \log \Gamma \( \frac{p + \eta_d^{(l)}}{2} \) \\
-	& - \log \Gamma \( \frac{\eta_d^{(l + 1)}}{2} \) - \frac{\eta_d^{(l+1)}}{2} e_d^{(l)} - \frac{\eta_d^{(l+1)}}{2} \log 2 + \frac{\eta_d^{(l + 1)}}{2} \log \lambda_d^{(l+1)} - \frac{\lambda_d^{(l + 1)} b_d^{(l)}}{2}.
-	\end{align*}
-  Note that the computation of $e_d^{(l)}$ differs between the inverse Gaussian and Gamma models.
-Furthermore, to compute the ELBO after a VB update we simply let $\eta_d^{(l)}, \theta_d^{(l)}, \lambda_d^{(l)}=\eta_d^{(l + 1)}, \theta_d^{(l + 1)}, \lambda_d^{(l + 1)}$.
-
-  In practice, there are different options to assess convergence.
-We may monitor the parameters themselves, the ELBO, or a combination of the two.
-We may also iterate for an \textit{a priori} fixed number of times.
-We may also choose to use a different convergence criterium for the EB and VB iterations.
-Currently we fix the number of EB iterations to 20, while we do 2 VB iterations per EB iteration.
-
-  \subsection{Efficient computation}
-  The empirical Bayes updates require the following quantities: $\zeta_d$, $a_d$, $\delta_d$, $\trace (\bm{\Sigma}_d)$, $\trace (\X \tr \X \bm{\Sigma}_d)$, $\bm{\mu}_d \tr \bm{\mu}_d$, and $\bm{\mu}_d \tr \X \tr \X \bm{\mu}_d$.
-In addition, we need $\log|\bm{\Sigma}|$ and $\mathbf{y}_d \tr \X \bm{\mu}_d$ to monitor the ELBO.
-The first three quantities are obtained by scalar operations of $\mathcal{O}(n)$.
-Let $c_d^{(h)}=b_d^{(h)}$ in the conjugate setting and $c_d^{(h)}=b_d^{(h)}/a_d^{(h)}$ in the non-conjugate setting.
-Then, beforementioned quantities are easily calculated using the SVD $\X = \mathbf{U} \mathbf{D} \mathbf{V} \tr$:
-  \begin{align*}
-  \trace (\bm{\Sigma}_d^{(h+1)}) & = (a_d^{(h)})^{-1} \[\sum_{i=1}^n (v_i^2 + c_d^{(h)})^{-1} + \max (p - n, 0)\cdot (c_d^{(h)})^{-1}\], \\
-  \trace (\X \tr \X \bm{\Sigma}_d^{(h+1)}) & = (a_d^{(h)})^{-1} \sum_{i=1}^n v_i^2(v_i^2 + c_d^{(h)})^{-1}, \\
-  (\bm{\mu}_d^{(h+1)}) \tr \bm{\mu}_d^{(h+1)} & = \sum_{i=1}^n v_i^2 [(\mathbf{U} \tr \mathbf{y}_d)_i]^2/(v_i^2 + c_d^{(h)})^2, \\
-  (\bm{\mu}_d^{(h+1)}) \tr \X \tr \X \bm{\mu}_d^{(h+1)} & = \sum_{i=1}^n v_i^4 [(\mathbf{U} \tr \mathbf{y}_d)_i]^2/(v_i^2 + c_d^{(h)})^2, \\
-  \log|\bm{\Sigma}^{(h+1)}| & \propto -p \log a_d^{(h)} - \sum_{i=1} \log (v_i^2 + c_d^{(h)}) - \max(p - n, 0) \cdot \log c_d^{(h)}, \\
-  \mathbf{y}_d \tr \X \bm{\mu}_d^{(h+1)} & = \sum_{i=1}^n v_i^2 [(\mathbf{U} \tr \mathbf{y}_d)_i]^2/(v_i^2 + c_d^{(h)}).
-  \end{align*}
-  where $v_i$ are the singular values of $\X$.
-The SVD and $\mathbf{U} \tr \mathbf{y}_d$ are $\mathcal{O}(pn^2)$ and $\mathcal{O}(n^2)$ operations, respectively, but have to be calculated only once at the start of the algorithm.
-The rest of the calculations involve just $n$ scalar multiplications and/or additions.
-
-  \section{Simulation setting}
-  - check model with Stan.
-
-  - make one group of drugs and compare methods
-
-  - use drug-specific lambda's
-
-  - try smaller $p$ to see whether $\sigma_d^2$ become more identifiable.
-
-  - scale-invariant instead of conjugate.
-
-  We simulate from a correctly specified model given in \ref{sec:notissuemodel}.
-We assume random $y_{id}$, $\x_i$, and $\bm{\beta}_d$, while we fix $\mathbf{C}$, $\gamma_d^2$, $\sigma_d^2$ and $\bm{\alpha}$.
-We estimate the $\bm{\mu}_d$, $\bm{\Sigma}_d$, $\delta_d$, $\bm{\alpha}$ and $\lambda$, while we fix the $\sigma_d^2$ to their true values.
-
-  We set $n=100$, $p=200$, and $D=10$.
-We consider 5 evenly sized classes of drugs, such that $\mathbf{c}_{kd} = \mathbbm{1}(k = \text{class}_d)$ and $\begin{bmatrix} \mathbf{c}_1 & \cdots & \mathbf{c}_D \end{bmatrix} \tr$.
-From these the $\gamma_d^2 = (\mathbf{c}_d \tr \bm{\alpha})^{-1}$ are created.
-Next, we simulate $\bm{\beta}_{jd} \sim \mathcal{N} (0, \sigma^2_d \gamma_d^2)$.
-We fix the mean signal-to noise ratio $\overline{\text{SNR}}_d$, such that the predictor data variance is $s^2 = \V(\mathbf{x}_i) = \overline{\text{SNR}}_d / (\overline{\gamma^2_d} p)$ (NOT CORRECT).
-The last step consists of simulating the data by $\mathbf{x}_{ij} \sim \mathcal{N}(0,s^2)$ and $y_{id} = \mathcal{N} (\mathbf{x}_i \tr \bm{\beta}_d, \sigma_d^2)$.
-The remaining parameters are set as follows: $\bm{\alpha} = \begin{bmatrix} 1 & \dots & 5 \end{bmatrix} \tr$, $\sigma_d^2 = 1$ for all $d$, and $\overline{\text{SNR}}_d = 10$.
-To account for the random data and parameter generation, we repeat the simulation 100 times.
-We compare the results to the models in \ref{sec:indvarmodel} and \ref{sec:invgammamodel}, and present the results in Figures \ref{fig:boxplot_igaussian_res1_prior_mean}-\ref{fig:boxplot_igaussian_res2_prior_mean}.
-
-<<boxplot_igaussian_res1_prior_mean, fig.cap="Prior mean estimates", out.width="80%", fig.asp=2/3>>=
-@
-
-<<boxplot_igaussian_res2_prior_mean, fig.cap="Prior mean estimates", out.width="80%", fig.asp=2/3>>=
-@
-
-  \section{Starting values}
-  \subsection{Method 1}
-  Starting value for $\sigma_d^2$ as in \cite{chipman_bart:_2010}.
-They consider a scaled-Chi squared distribution for the error variance with degrees of freedom 3 and scale chosen such that the $q*100$th percentile matches the error variance in the data $s^2(\mathbf{y}_d)$.
-As starting value for $\sigma_d^2$ we take the mode of this distribution: $\hat{\sigma}_d^2 = 3a/5$, with $a$ the solution to:
-  $$
-  \Gamma\(\frac{3}{3}\) \[1 - q - F(3a/(2s^2(\mathbf{y}_d)); 3/2, 1) \] = 0,
-  $$
-  where $F(x; 3/2, 1)$ is the CDF of a gamma function with shape $3/2$ and scale $1$.
-In \cite{moran_variance_2018}, REFERENCE to Rockova and George (2018), and \cite{chipman_bart:_2010}, $q=0.9$ is suggested.
-We find that $q \in ( 0.9, 0.95 )$ gives good results.
-
-  \subsection{Method 2}
-  We generate starting values as follows. Consider the $\sigma_d^2$ and $\gamma_d^2$ as fixed and estimate them by MML of the regular ridge model:
-  $$
-  \hat{\sigma}_d^2, \hat{\gamma}_d^2= \underset{\sigma_d^2,\gamma_d^2}{\argmax} \int_{\bm{\beta}} p(\mathbf{y}_d | \bm{\beta}, \sigma_d^2) p(\bm{\beta} | \gamma_d^2, \sigma_d^2) d\bm{\beta}.
-  $$
-  The regular ridge model is conjugate, so MML is relatively simple. Next we estimate a common inverse Gaussian prior for the $\gamma_d^2$ by considering the estimates to be samples from the prior:
-  \begin{align*}
-  \hat{\theta} & = n^{-1} \sum_{d=1}^D \hat{\gamma}_d^2 / n, \\
-  \hat{\lambda} & = n \left\{\sum_{d=1}^D \[ (\hat{\gamma}_d^2)^{-1} - \hat{\theta}^{-1}\]\right\}^{-1}.
-  \end{align*}
-  We estimate the mode of the $\hat{m}=\hat{\gamma}_d^2$ by kernel density estimation and equate it to the theoretical mode of the generalized inverse Gaussian distribution with the estimated $\hat{\theta}$ and $\hat{\lambda}$.
-We solve to find a common $\delta$:
-  $$
-  \hat{\delta} = \frac{\hat{m}^2 \hat{\lambda}}{\hat{\theta}^2} + (p + 3) \hat{m}.
-  $$
-  Lastly, we consider the $\hat{\sigma}^2_d$ fixed samples from the inverse Gamma distribution to estimate a common scale:
-  $$
-  \hat{\zeta} = \frac{D(n + p + 1)}{2 \sum_{d=1}^D[(\hat{\sigma}_d^2)^{-1}]}.
-  $$
-
-  \section{Ratios of modified Bessel functions}
-  Ratios of modified Bessel functions of the second kind $K_{\alpha - 1}(x)/K_{\alpha}(x)$ are prone to under- and overflow for large $\alpha$.
-In our case, $\alpha$ increases linearly with $p$.
-Since $p$ may be large, this causes numerical issues in the calculation of various quantities.
-We alleviate the numerical issues through the following.
+  \section{Ratios of modified Bessel functions}\label{sec:ratiosmodifiedbessels}
+  Ratios of modified Bessel functions of the second kind 
+  $K_{\alpha - 1}(x)/K_{\alpha}(x)$ are prone to under- and overflow for large
+  $\alpha$. In our case, $\alpha$ increases linearly with $p$. Since $p$ may be
+  large, this may cause numerical issues in the calculation of various 
+  quantities. We alleviate the numerical issues through the following.
 
@@
     \begin{cases}
-  \Big( \dots \Big( \( 1 + \frac{2 \cdot 1 - 1}{x} \)^{-1} + \frac{2 \cdot 2-1}{x}\Big)^{-1} + \dots + \frac{2 \cdot n_1 - 1}{x}\Big)^{-1}, & \text{for } p \text{ even}, \\
-  \Big( \dots \Big( \( \frac{K_0(x)}{K_1(x)} + \frac{2}{x} \cdot 1 \)^{-1} + \frac{2}{x} \cdot 2 \Big)^{-1} + \dots + \frac{2}{x} \cdot n_2 \Big)^{-1}, & \text{for } p \text{ odd}.
+      \Big( \dots \Big( \( 1 + \frac{2 \cdot 1 - 1}{x} \)^{-1} + 
+      \frac{2 \cdot 2-1}{x}\Big)^{-1} + \dots + 
+      \frac{2 \cdot n_1 - 1}{x}\Big)^{-1}, & \text{for } p \text{ even}, \\
+      \Big( \dots \Big( \( \frac{K_0(x)}{K_1(x)} + \frac{2}{x} \cdot 1 \)^{-1} +
+      \frac{2}{x} \cdot 2 \Big)^{-1} + \dots + \frac{2}{x} \cdot n_2 \Big)^{-1},
+      & \text{for } p \text{ odd}.
     \end{cases}
   \end{align*} 
-  The ratio $K_0(x)/K_1(x)$ is well-behaved, such that the ratio may be computed without numerical issues.
+  The ratio $K_0(x)/K_1(x)$ is well-behaved, so that an 
+  arbitrary ratio $K_{\alpha - 1}(x)/K_{\alpha}(x)$ may be calculated as
+  a sequence of numerically stable scalar sums, products and inverses.
 
-  \section{General model}
-  We have drug covariates $\mathbf{c}_d$ and omics feature covariates $\mathbf{z}_j$. The model is now:
+  \section{Empirical Bayes}\label{sec:eb}
+  For the feature scale parameter $\lambda_{\text{feat}}$ updates to be 
+  non-negative, we have to show that:
+  \begin{align}\label{eq:inequality}
+    &\sum_{d=1}^D \sum_{j=1}^{p} 
+    b_{jd} + \bm{\alpha}_{\text{feat}} \tr \mathbf{C} \tr 
+    \diag (e_{jd}) \mathbf{C} \bm{\alpha}_{\text{feat}} - 
+    2 \bm{\alpha}_{\text{feat}} \tr \mathbf{C} \tr 
+    \mathbf{1}_{pD \times 1}  = \sum_{d=1}^D \sum_{j=1}^{p} \(b_{jd} + 
+    \phi_{jd}^{-2} e_{jd} - 2 \phi_{jd}^{-1} \) \geq 0,
+  \end{align}
+  which holds if all summands in the left-hand side of (\ref{eq:inequality})
+  are non-negative.
+
+  We note that through Jensen's inequality 
+  $e_{jd}^{-1}=\E(\gamma_{jd}^{2})^{-1} \leq \E(\gamma_{jd}^{-2})=b_{jd}$,
+  we may lowerbound the summands in left-hand side of (\ref{eq:inequality}):
+  $$
+  b_{jd} + \phi_{jd}^{-2} e_{jd} - 2 \phi_{jd}^{-1} \geq
+  e_{jd}^{-1} + \phi_{jd}^{-2} e_{jd} - 2 \phi_{jd}^{-1}.
+  $$
+  which we use to show that:
   \begin{align*}
-  y_{id} | \bm{\beta}_d, \sigma_d^2 & \sim \mathcal{N} (\x_i \tr \bm{\beta}_d, \sigma_d^2), \\
-  \beta_{jd} | \tau_j^2, \gamma_d^2, \sigma_d^2 & \sim \mathcal{N} (0, \sigma_d^2 \gamma_d^2 \tau_j^2), \\
-  \gamma_d^2 & \sim \mathcal{IG}\((\mathbf{c}_d \tr \bm{\alpha})^{-1}, \lambda\), \\
-  \tau_j^2 & \sim \mathcal{IG}\((\mathbf{z}_j \tr \bm{\theta})^{-1}, \nu\), \\
-  \sigma_d^2 & \sim 1/\sigma_d^3.
+    e_{jd}^{-1} + \phi_{jd}^{-2} e_{jd} - 2 \phi_{jd}^{-1} & \geq 0 \\
+    1 + \phi_{jd}^{-2} e_{jd}^2 - 2 \phi_{jd}^{-1} e_{jd} & \geq 0 \\
+    \( 1 - \phi_{jd}^{-1} e_{jd} \)^2 & \geq 0.
   \end{align*}
-  Variational posterior:
+  Similar reasoning holds for the drug scale parameter $\lambda_{\text{drug}}$.
+
+  \section{Gibbs sampler}\label{sec:gibbssampler}
+  MCMC samples from the posterior corresponding to model 
+  (\ref{md-eq:prior}) may be generated for each equation independently.
+  MCMC samples may be generated by iteratively sampling the following 
+  conditional distributions:
+  \begin{subequations}\label{eq:gibbssampler}
+    \begin{align}
+      \bm{\beta}_d | \bgamma_d^2, \sigma_d^2, \tau_d^2, \y_d & \sim 
+      \mathcal{N}_p (\bm{\mu}_d, \bm{\Sigma}_d), \label{eq:betasample} \\
+      \text{with } & \bm{\Sigma}_d = \sigma_d^2 [\X \tr \X + 
+      \tau_d^{-2} \diag (\gamma_{jd}^{-2})]^{-1}, \label{eq:sigmasample} \\
+      \text{and } & \bm{\mu}_d = [\X \tr \X + 
+      \tau_d^{-2} \diag (\gamma_{jd}^{-2})]^{-1} \X \tr 
+      \y_d \label{eq:musample}, \\
+      \gamma_{jd}^2 | \beta_{jd}, \sigma_d^2, \tau_d^2, \y &
+      \sim \mathcal{GIG}
+      \(-1, \lambda_{\text{feat}}/ \phi_{jd}^{2}, \delta_{jd}\), \\
+      \text{with } & \delta_{jd} = \sigma_d^{-2} \tau_d^{-2} \beta_{jd}^2 + 
+      \lambda_{\text{feat}}, \nonumber \\
+      \tau_d^2 | \bm{\beta}_d, \sigma_d^2, \bgamma_{d}^2, \y & \sim
+      \mathcal{GIG}
+      \(-\frac{p+1}{2}, \lambda_{\text{drug}}/ \chi_{d}^{2}, \eta_d\), 
+      \\
+      \text{with } & \eta_d = \sigma_d^{-2} \sum_{j=1}^p \gamma_{jd}^{-2}
+      \beta_{jd}^2 + \lambda_{\text{drug}}, \\
+      \sigma_d^2 | \bm{\beta}_d, \bgamma_d^2, \tau_d^2, \y & \sim \Gamma^{-1} 
+      \(\frac{n + p + 1}{2}, \zeta_d\), \\
+      \text{with } & \zeta_d = \frac{1}{2} \[\mathbf{y}_d \tr \mathbf{y}_d -
+      2 \mathbf{y}_d \tr \X \bm{\beta}_d + \bm{\beta}_d \X \tr \X \bm{\beta}_d + 
+      \tau^{-2}_d \bm{\beta}_d \tr \diag(\gamma_{jd}^{-2}) \bm{\beta}_d) \]. 
+    \end{align}
+  \end{subequations}
+  In high dimensional space, the $p \times p$ matrix inversions in 
+  (\ref{eq:sigmasample}) and (\ref{eq:musample}) are   
+  significant computational bottlenecks. \cite{bhattacharya_fast_2016} describe 
+  a method to sample from (\ref{eq:betasample}) without explicit calculation
+  of this inverse, thereby offering a siginifcant speed-up compared to naive
+  sampling.
+
+  Sampling from (\ref{eq:gibbssampler}) either requires estimating 
+  (as in \ref{md-sec:empiricalbayes}) or specifying 
+  hyperparameters
+  $\bm{\alpha}_{\text{feat}}$, $\bm{\alpha}_{\text{drug}}$, 
+  $\lambda_{\text{feat}}$, and $\lambda_{\text{drug}}$, or to endow them with
+  and extra layer of hyperpriors. In general, specifying them requires 
+  rarely available, detailed subject knowledge, so we may resort to endowing 
+  them with hyperpriors (as in \cite{upadhyay_bayesian_1996}):
   \begin{align*}
-  q(\bm{\beta}) & \overset{D}{=} \prod_{d=1}^D \mathcal{N}_p (\bm{\mu}_d, \bm{\Sigma}_d), \\
-  q(\bm{\gamma}^2) & \overset{D}{=} \prod_{d=1}^D \mathcal{GIG}\(-\frac{p+1}{2}, \lambda (\mathbf{c}_d \tr \bm{\alpha})^2, \delta_d\), \\
-  q(\bm{\tau}^2) & \overset{D}{=} \prod_{j=1}^p \mathcal{GIG}\(-\frac{D+1}{2}, \nu (\mathbf{z}_j \tr \bm{\theta})^2, \eta_j\), \\
-  q(\bm{\sigma}^2) & \overset{D}{=} \prod_{d=1}^D \Gamma^{-1} \(\frac{n + p + 1}{2}, \zeta_d\).
+    \bm{\alpha}_{\text{feat},g} | \lambda_{\text{feat}} & \sim 
+    \mathcal{N}(0, \nu_{\text{feat}}^2/\lambda_{\text{feat}}), \\
+    \bm{\alpha}_{\text{drug},h} | \lambda_{\text{drug}} & \sim 
+    \mathcal{N}(0, \nu_{\text{drug}}^2/\lambda_{\text{drug}}), \\
+    \lambda_{\text{feat}} & \sim \Gamma(\kappa_{\text{feat}}, 
+    \xi_{\text{feat}}), \\
+    \lambda_{\text{drug}} & \sim \Gamma(\kappa_{\text{drug}}, 
+    \xi_{\text{drug}}),
   \end{align*}
-  Variational parameters:
+  with hyperparameters $\nu_{\text{feat}}^2$, $\nu_{\text{drug}}^2$, 
+  $\kappa_{\text{feat}}$, $\xi_{\text{feat}}$, $\kappa_{\text{drug}}$, and
+  $\xi_{\text{drug}}$. Improper, flat priors occur if we let 
+  $\nu_{\text{feat}}, \nu_{\text{drug}} \to \infty$ and 
+  $\kappa_{\text{feat}}, \kappa_{\text{drug}}, \xi_{\text{feat}}, 
+  \xi_{\text{drug}} \to 0$. The corresponding full conditionals are
+  given by:
   \begin{align*}
-  \bm{\Sigma}_d^{(h+1)} & = \frac{2 \zeta_d^{(h)}}{n+p+1} \[\X \tr \X + a_d^{(h)} \cdot \diag(b_j^{(h)})\]^{-1}, \\
-  \bm{\mu}_d^{(h+1)} & = \[\X \tr \X + a_d^{(h)} \cdot \diag(b_j^{(h)})\]^{-1} \X \tr \y_d, \\
-  \delta_d^{(h+1)} & = \frac{n+p+1}{2 \zeta_d^{(h)}} \left\{\trace \[\diag(b_j^{(h)}) \bm{\Sigma}_d^{(h+1)} \] + (\bm{\mu}_d^{(h+1)}) \tr \diag(b_j^{(h)}) \bm{\mu}_d^{(h+1)}\right\} + \lambda, \\ 
-  \eta_j^{(h+1)} & = \sum_{d=1}^D \frac{a_d^{(h)}(n+p+1)}{2 \zeta_d^{(h)}} \[ (\bmu_{jd}^{(h + 1)})^2 + (\bSigma_d^{(h + 1)})_{jj} \] + \nu, \\ 
-  \zeta_d^{(h+1)} & = \frac{1}{2} \Big( \mathbf{y}_d \tr \mathbf{y}_d -2 \mathbf{y}_d \tr \X \bm{\mu}_d^{(h+1)} + a_d^{(h+1)} \trace \left\{ \[\X \tr \X + \diag(b_j^{(h + 1)})\] \bm{\Sigma}_d^{(h+1)}\right\} \\
-  & \,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\, + a_d^{(h+1)} (\bm{\mu}_d^{(h+1)}) \tr \[\X \tr \X + \diag(b_j^{(h + 1)})\]\bm{\mu}_d^{(h+1)}\Big),
+    \bm{\alpha}_{\text{feat}} | \lambda_{\text{feat}}, \bm{\Gamma} & \sim 
+    \mathcal{N} \( \mathbf{m}_{\text{feat}}, \mathbf{S}_{\text{feat}} \) \\
+    \text{with } & \mathbf{S}_{\text{feat}} = \lambda_{\text{feat}}^{-1} 
+    [\mathbf{C} \tr \diag(\gamma_{jd}^2) \mathbf{C} +
+    \nu_{\text{feat}}^{-2} \I_G]^{-1}, \\
+    \text{and } & \mathbf{m}_{\text{feat}} = [\mathbf{C} \tr 
+    \diag(\gamma_{jd}^2) \mathbf{C} + \nu_{\text{feat}}^{-2} \I_G]^{-1}
+    \mathbf{C} \tr \mathbf{1}_{pD \times 1}, \\
+    \bm{\alpha}_{\text{drug}} | \lambda_{\text{drug}}, \bm{\tau} & \sim 
+    \mathcal{N} \(\textbf{m}_{\text{drug}}, \mathbf{S}_{\text{drug}} \), \\
+    \text{with } & \mathbf{S}_{\text{drug}} = \lambda_{\text{drug}}^{-1} 
+    [\mathbf{Z} \tr \diag(\tau_{d}^2) \mathbf{Z} +
+    \nu_{\text{drug}}^{-2} \I_H]^{-1}, \\
+    \text{and } & \mathbf{m}_{\text{drug}} = [\mathbf{Z} \tr 
+    \diag(\tau_{d}^2) \mathbf{Z} + \nu_{\text{drug}}^{-2} \I_H]^{-1}
+    \mathbf{Z} \tr \mathbf{1}_{D \times 1}, \\
+    \lambda_{\text{feat}} | \bm{\alpha}_{\text{feat}}, \bm{\Gamma} & 
+    \sim \Gamma \(\frac{pD}{2} + \kappa_{\text{feat}}, 
+    \upsilon_{\text{feat}}\), \\
+    \text{with } & \upsilon_{\text{feat}} =\frac{1}{2} 
+    \[\mathbf{C} \bm{\alpha}_{\text{feat}} - \diag(\gamma_{jd}^{-2}) 
+    \mathbf{1}_{pD \times 1}\] \tr \diag(\gamma_{jd}^2) 
+    \[\mathbf{C} \bm{\alpha}_{\text{feat}} - 
+    \diag(\gamma_{jd}^{-2}) \mathbf{1}_{pD \times 1}\] + \xi_{\text{feat}}, \\
+    \lambda_{\text{drug}} | \bm{\alpha}_{\text{drug}}, \bm{\tau} & 
+    \sim \Gamma \(\frac{D}{2} + \kappa_{\text{drug}}, 
+    \upsilon_{\text{drug}} \), \\
+    \text{with } & \upsilon_{\text{drug}} =\frac{1}{2} 
+    \[\mathbf{Z} \bm{\alpha}_{\text{drug}} - \diag(\tau_{d}^{-2}) 
+    \mathbf{1}_{D \times 1}\] \tr \diag(\tau_{d}^2) 
+    \[\mathbf{Z} \bm{\alpha}_{\text{drug}} - 
+    \diag(\tau_{d}^{-2}) \mathbf{1}_{D \times 1}\] + \xi_{\text{drug}},
   \end{align*}
-  with
-  \begin{align*}
-  a_d^{(h)} & = \sqrt{\frac{\lambda (\mathbf{c}_d \tr \bm{\alpha})^2}{\delta_d^{(h)}}} \frac{K_{\frac{p-1}{2}} \( \sqrt{\lambda (\mathbf{c}_d \tr \bm{\alpha})^2 \delta_d^{(h)}} \)}{K_{\frac{p+1}{2}} \( \sqrt{\lambda (\mathbf{c}_d \tr \bm{\alpha})^2 \delta_d^{(h)}} \)} + \frac{p+1}{\delta_d^{(h)}} \text{ and} \\
-  b_j^{(h)} & = \sqrt{\frac{\nu (\mathbf{z}_j \tr \bm{\theta})^2}{\eta_j^{(h)}}} \frac{K_{\frac{D-1}{2}} \( \sqrt{\nu (\mathbf{z}_j \tr \bm{\theta})^2 \eta_j^{(h)}} \)}{K_{\frac{D+1}{2}} \( \sqrt{\nu (\mathbf{z}_j \tr \bm{\theta})^2 \eta_j^{(h)}} \)} + \frac{D+1}{\eta_j^{(h)}}.
-  \end{align*}
-  Hyper-parameter updates:
-  \begin{align*}
-  \bm{\alpha}^{(l+1)} & = \left[ \mathbf{C} \tr \diag ( e_d^{(l)} ) \mathbf{C} \right]^{-1} \mathbf{C} \tr \mathbf{1}, \\
-  \lambda^{(l+1)} & = \[ \sum_{d=1}^D a_d^{(l)} - (\bm{\alpha}^{(l+1)}) \tr \mathbf{C} \tr \mathbf{1} \]^{-1} D, \\
-  \bm{\theta}^{(l+1)} & = \left[ \mathbf{Z} \tr \diag ( f_j^{(l)} ) \mathbf{Z} \right]^{-1} \mathbf{Z} \tr \mathbf{1}, \\
-  \nu^{(l+1)} & = \[ \sum_{j=1}^p b_j^{(l)} - (\bm{\theta}^{(l+1)}) \tr \mathbf{Z} \tr \mathbf{1} \]^{-1} p,
-  \end{align*}
-  where 
-  \begin{align*}
-  e_d^{(l)} & = \sqrt{\frac{\delta_d^{(l)}}{\lambda^{(l)}(\mathbf{c}_d \tr \bm{\alpha}^{(l)})^2}} \frac{K_{\frac{p - 1}{2}} \( \sqrt{\lambda^{(l)} (\mathbf{c}_d \tr \bm{\alpha}^{(l)})^2 \delta_d^{(l)}} \)}{K_{\frac{p + 1}{2}} \( \sqrt{\lambda^{(l)} (\mathbf{c}_d \tr \bm{\alpha}^{(l)})^2 \delta_d^{(l)}} \)} \text{ and} \\
-  f_j^{(l)} & = \sqrt{\frac{\eta_j^{(l)}}{\nu^{(l)}(\mathbf{z}_j \tr \bm{\theta}^{(l)})^2}} \frac{K_{\frac{D - 1}{2}} \( \sqrt{\nu^{(l)} (\mathbf{z}_j \tr \bm{\theta}^{(l)})^2 \eta_j^{(l)}} \)}{K_{\frac{D + 1}{2}} \( \sqrt{\nu^{(l)} (\mathbf{z}_j \tr \bm{\theta}^{(l)})^2 \eta_j^{(l)}} \)}.
-  \end{align*}
+  which may be sampled together with (\ref{eq:gibbssampler}).
+
+  \section{Simulations}\label{sec:simulations}
+<<simulation_gdsc>>=
+# simulation 1
+res <- read.table("results/simulations_gdsc_res1.txt", row.names=NULL)
+temp <- res[, 1]
+res <- as.matrix(res[, -1])
+rownames(res) <- temp
+
+tabm <- aggregate(res, by=list(substr(rownames(res), 1, 5)), FUN="mean")
+tabm <- t(tabm[tabm$Group.1 %in% c("emse.", "emseh", "emsel", "pmse."), -1])
+
+D <- 251
+nreps <- 100
+tabsd <- sapply(c("emse.", "emseh", "emsel", "pmse."), function(s) {
+  apply(aggregate(res[substr(rownames(res), 1, 5)==s, ], 
+                  list(rep(1:nreps, each=D)), mean), 2, sd)})[-1, ]
+
+tab1 <- sapply(1:ncol(tabm), function(s) {
+  id <- which(tabm[, s]==min(tabm[, s], na.rm=TRUE))
+  out <- paste0(round(tabm[, s], 3), " (", round(tabsd[, s], 3), ")")
+  out[id] <- paste0("\\textbf{", out[id], "}")
+  out})
+colnames(tab1) <- c("$\\text{EMSE}_{\\text{total}}$", 
+                    "$\\text{EMSE}_{\\text{top}}$", 
+                    "$\\text{EMSE}_{\\text{bottom}}$", "$\\text{PMSE}$")     
+rownames(tab1) <- c("NIG$_{\\text{f}}^-$", "NIG$_{\\text{f}}$", "ridge", 
+                    "lasso")
+
+# simulation 2
+res <- read.table("results/simulations_gdsc_res2.txt", row.names=NULL)
+temp <- res[, 1]
+res <- as.matrix(res[, -1])
+rownames(res) <- temp
+
+tabm <- aggregate(res, by=list(substr(rownames(res), 1, 5)), FUN="mean")
+tabm <- t(tabm[tabm$Group.1 %in% c("emse.", "emseh", "emsel", "pmse."), -1])
+
+D <- 251
+nreps <- 100
+tabsd <- sapply(c("emse.", "emseh", "emsel", "pmse."), function(s) {
+  apply(aggregate(res[substr(rownames(res), 1, 5)==s, ], 
+                  list(rep(1:nreps, each=D)), mean), 2, sd)})[-1, ]
+
+tab2 <- sapply(1:ncol(tabm), function(s) {
+  id <- which(tabm[, s]==min(tabm[, s], na.rm=TRUE))
+  out <- paste0(round(tabm[, s], 3), " (", round(tabsd[, s], 3), ")")
+  out[id] <- paste0("\\textbf{", out[id], "}")
+  out})
+colnames(tab2) <- c("$\\text{EMSE}_{\\text{total}}$", 
+                    "$\\text{EMSE}_{\\text{top}}$", 
+                    "$\\text{EMSE}_{\\text{bottom}}$", "$\\text{PMSE}$")     
+rownames(tab2) <- c("NIG$_{\\text{d}}^-$", "NIG$_{\\text{d}}$", "ridge", 
+                    "lasso")
+
+# simulation 3
+res <- read.table("results/simulations_gdsc_res3.txt", row.names=NULL)
+temp <- res[, 1]
+res <- as.matrix(res[, -1])
+rownames(res) <- temp
+
+tabm <- tabm3 <- aggregate(res, by=list(substr(rownames(res), 1, 5)), 
+                           FUN="mean")
+tabm <- t(tabm[tabm$Group.1 %in% c("emse.", "emseh", "emsel", "pmse."), -1])
+
+D <- 251
+nreps <- 100
+tabsd <- tabsd3 <- sapply(c("emse.", "emseh", "emsel", "pmse."), function(s) {
+  apply(aggregate(res[substr(rownames(res), 1, 5)==s, ], 
+                  list(rep(1:nreps, each=D)), mean), 2, sd)})[-1, ]
+
+tab3 <- sapply(1:ncol(tabm), function(s) {
+  id <- which(tabm[, s]==min(tabm[, s], na.rm=TRUE))
+  out <- paste0(round(tabm[, s], 3), " (", round(tabsd[, s], 3), ")")
+  out[id] <- paste0("\\textbf{", out[id], "}")
+  out})
+colnames(tab3) <- c("$\\text{EMSE}_{\\text{total}}$", 
+                    "$\\text{EMSE}_{\\text{top}}$", 
+                    "$\\text{EMSE}_{\\text{bottom}}$", "$\\text{PMSE}$")     
+rownames(tab3) <- c("NIG$_{\\text{f+d}}^-$", 
+                    "NIG$_{\\text{f+d}}$", "ridge", "lasso")
+@	
+  The lasso and ridge models in this section were fit separately per drug, 
+  using the \texttt{R} package \texttt{glmnet} 
+  \cite[]{friedman_regularization_2010} with cross
+	validated penalty parameters.
+
+	Tables \ref{tab:simulations_gdsc_tab1}-\ref{tab:simulations_gdsc_tab3} 
+	display averaged estimation mean squared error (EMSE), and prediction mean
+	squared error (PMSE), calculated on independent test data, for simulation 
+	Scenarios 1-3. PMSE is calculated as in Section \ref{md-sec:gdsc} of the MD, 
+	while EMSE is calculated as: $\text{EMSE} = D^{-1} p^{-1} \sum_{d=1}^D 
+  \sum_{j=1}^p (\beta_{jd} - \hat{\beta}_{jd})^2$, with
+  $\hat{\beta}_{jd}$ the estimator for $\beta_{jd}$. In the NIG
+  models, that provide the full posteriors, the posterior mean 
+  $\E(\beta_{jd} | \mathbf{y}_d)$ is used as point estimate.
+	EMSE is further split into the contribution of the
+	bottom 10\% of the $\beta_{jd}$ in terms of size (in absolute value) and the 
+	top 10\% of the $\beta_{jd}$ in size.
+
+	Focussing on estimation, we see that the NIG$_{\text{f}}$, NIG$_{\text{d}}$,
+	and NIG$_{\text{f}+\text{d}}$ models (depending on the Scenario),
+	that include the external covariates, outperform
+	the other methods in terms of EMSE and PMSE. The lower EMSEs confirm that the 
+	NIG models that include the external data (NIG$_{\text{f}}$, NIG$_{\text{d}}$,
+	and NIG$_{\text{f}+\text{d}}$) learn the underlying structure 
+	in the data better than the models that do not include the external data
+	(NIG$_{\text{f}}^-$, NIG$_{\text{d}}^-$, and NIG$_{\text{f}+\text{d}}^-$). 
+	Furthermore, as expected, the sparser models lasso and NIG are better able to
+	learn the $\beta_{jd}$ with larger magnitude, as seen from the lower
+	EMSE$_{\text{top}}$. On the other hand, ridge is better able to capture the
+	smaller $\beta_{jd}$, as evident from the lower EMSE$_{\text{bottom}}$.
+	Figure \ref{fig:simulations_gdsc_pmse4} shows that, as expected, the 
+	performance of the NIG$_{\text{f}+\text{d}}$ model deteriorates with 
+	increasing noise level in the external covariates and converges to the 
+	NIG$_{\text{f}+\text{d}}^-$ model performance.
+<<simulations_gdsc_tab1>>=
+kableExtra::kable_styling(knitr::kable(
+  tab1, align="r",
+  caption="Mean EMSE (separated for bottom and top 10\\% of $|\\hat{\\beta}_{jd}|$) and PMSE for simulation Scenario 1, estimated on test data (lowest in bold).", 
+  format="latex", booktabs=TRUE, escape=FALSE), 
+  latex_options=c("HOLD_position"))
+@
+<<simulations_gdsc_tab2>>=
+kableExtra::kable_styling(knitr::kable(
+  tab2, align="r",
+  caption="Mean EMSE (separated for bottom and top 10\\% of $|\\hat{\\beta}_{jd}|$) and PMSE for simulation Scenario 2, estimated on test data (lowest in bold).", 
+  format="latex", booktabs=TRUE, escape=FALSE), 
+  latex_options=c("HOLD_position"))
+@
+<<simulations_gdsc_tab3>>=
+kableExtra::kable_styling(knitr::kable(
+  tab3, align="r",
+  caption="Mean EMSE (separated for bottom and top 10\\% of $|\\hat{\\beta}_{jd}|$) and PMSE for simulation Scenario 3, estimated on test data (lowest in bold).", 
+  format="latex", booktabs=TRUE, escape=FALSE), 
+  latex_options=c("HOLD_position"))
+@
+<<simulations_gdsc_pmse4, fig.cap="(a) Mean EMSE, separated for (b) bottom and (c) top 10\\% of $|\\hat{\\beta}_{jd}|$, and (d) PMSE for simulation Scenario 4 versus the proportion of permuted  rows in the external data, estimated on test data.", out.width="100%", fig.asp=1>>=
+@
+
+  Figure \ref{fig:simulations_gdsc_post5} shows instances of both MCMC and VB
+  $\beta_{jd}$ posteriors from simulation Scenario 3 for the 
+  NIG$_{\text{f}+\text{d}}$ model. 
+  The chosen $\beta_{jd}$ represent every combination of four drugs and four 
+  feature groups (for a total of 16 $\beta_{jd}$). The Figure shows that the
+  MCMC and variational posteriors are in agreement. This leads us to conclude
+  that the variational approximation to the posterior is quite accurate
+  in this Scenario. The simulation Scenario was setup to mimic the real GDSC 
+  data, so that we expect an accurate variational posterior in the real GDSC
+  data as well.
+<<simulations_gdsc_post5, fig.cap="Simulation results for Scenario 3: Samples from NIG$_{\\text{f}+\\text{d}}$ posterior with variational posterior superimposed (red line) for estimated hyperparameters (top row) $\\hat{\\chi}_1$, (second row) $\\hat{\\chi}_2$, (third row) $\\hat{\\chi}_3$, (bottom row) $\\hat{\\chi}_4$, (left column) $\\hat{\\phi}_1$, (second column) $\\hat{\\phi}_2$, (third column) $\\hat{\\phi}_3$, and (last column) $\\hat{\\phi}_4$.", out.width="100%", fig.asp=1>>=
+@
+
+  For completeness, Figure \ref{fig:simulations_gdsc_res5} displays samples
+  from the full Bayesian NIG posterior, as introduced in Section 
+  \ref{sec:gibbssampler}. Unfortunately, we were not able to generate enough
+  samples for a reliable posterior using the Gibbs procedure as described in 
+  the Section. We generated the samples from the full Bayesian posterior using
+  stan through the R package \texttt{rstan} 
+  \cite[]{guo_rstan:_2018} (see \url{https://github.com/magnusmunch/NIG} for 
+  the implemented stan code).
+<<simulations_gdsc_res5, fig.cap="Simulation results for Scenario 3: Samples from full Bayesian NIG posteriors (see Section \\ref{sec:gibbssampler}) with true value superimposed (red line) for (top row) group 1, (second row) group 2, (third row) group 3, and (bottom row) group 4 for both features and drugs, and (left column) $\\bm{\\phi}$ and (second column) $\\bm{\\chi}$.", out.width="100%">>=
+@
+
+  \section{GDSC data}\label{sec:gdsc}
+  the CPO for cell line $i$ and drug $d$ is calculated as 
+  $$
+  \text{CPO}_{id} = p(y_{id} | \bm{y}_{-id}) = 
+  \[\int_{\bm{\beta}_d} \int_{\sigma_d^2} \mathcal{L}(\y_{id} | 
+  \bm{\beta}_d, \sigma_d^2)^{-1} q(\bm{\beta}_d | \bm{y}_d) 
+  q(\sigma^2_d | \bm{y}_d) d\bm{\beta}_d \sigma_d^2\]^{-1}
+  $$ 
+  and is the Bayesian version of the leave-one-out cross-validated likelihood.
+  \cite{congdon_bayesian_2005} suggests as outlier cut-off 0.01
 

### 2020-04-22 code/data_gdsc.R

@@ -8,14 +8,9 @@ library(gdata)
 ensembl <- useMart(biomart="ensembl", dataset="hsapiens_gene_ensembl")
 
 ################################ response data #################################
-# loading the data either from a local file or from the GDSC website
-if(file.exists("data/v17.3_fitted_dose_response.csv")) {
-  resp <- read.table("data/v17.3_fitted_dose_response.csv", header=TRUE,
-                     sep=",", stringsAsFactors=FALSE)
-} else {
-  resp <- read.table("ftp://ftp.sanger.ac.uk/pub/project/cancerrxgene/releases/current_release/v17.3_fitted_dose_response.xlsx",
-                     header=TRUE, sep=",", stringsAsFactors=FALSE)
-}
+# loading the data
+resp <- read.table("data/v17.3_fitted_dose_response.csv", header=TRUE,
+                   sep=",", stringsAsFactors=FALSE)
 
 # take average IC50 if more than one measurement for drug cell line combination
 resp.temp <- aggregate(LN_IC50 ~ COSMIC_ID + CELL_LINE_NAME + DRUG_NAME,
@@ -37,37 +32,15 @@ rm(resp, resp.temp)
 
 #################################### drugs #####################################
 # reading in the preformatted list of compounds
-if(file.exists("data/Screened_Compounds.xlsx")) {
-  drug1 <- read.xls("data/Screened_Compounds.xlsx",
-                    stringsAsFactors=FALSE)
-} else {
-  drug1 <- read.xls("ftp://ftp.sanger.ac.uk/pub/project/cancerrxgene/releases/current_release/Screened_Compounds.xlsx",
-                    stringsAsFactors=FALSE)
-}
+drug1 <- read.xls("data/Screened_Compounds.xlsx", stringsAsFactors=FALSE)
 
 # reading in the curent list of compounds
-if(length(list.files(path="data", patt="Drug_list")) > 0) {
-  drug2 <- read.table(paste0("data/", list.files(path="data",
-                                                 patt="Drug_list")[1]),
-                      header=TRUE, sep=",", quote='"', stringsAsFactors=FALSE)
-} else {
[…]

### 2020-04-22 code/eQTL.R

@@ -1,91 +1,0 @@
-#!/usr/bin/env Rscript
-
-### installation of packages
-if(!("devtools" %in% installed.packages())) {
-  install.packages("devtools")
-}
-library(devtools)
-install_github("magnusmunch/cambridge/rpackage", local=FALSE,
-               auth_token=Sys.getenv("GITHUB_PAT"))
-
-### libraries
-library(cambridge)
-library(glmnet)
-
-### load data
-load("data/P38MAPKpathwayKb100K.RData")
-
-### data preparation
-y <- sapply(Gn, "[[", 1)
-x <- lapply(sapply(Gn, "[[", 2), scale)
-G <- length(Gn)
-p <- sapply(x, ncol)
-n <- nrow(y)
-
-### creating co-data
-inorout <- lapply(Gn, "[[", 4)
-distance <- lapply(Gn, function(g) {  
-  sapply(g[[5]], function(s) {
-    (sign(prod(s - g[[3]][c(1:2)]))!=1)*
-      min(abs(s - g[[3]][c(1:2)]))})})
-genelength <- lapply(1:G, function(g) {
-  rep(diff(as.numeric(Gn[[g]][[3]][c(1, 2)])), p[g])})
-minoraf <- lapply(x, function(g) {f <- apply(g, 2, function(s) {
-  f <- mean(s); unname(min(f, 1 - f))})})
-# library(biomaRt)
-# ensembl <- useEnsembl(biomart="ensembl", dataset="hsapiens_gene_ensembl")
-# # View(listFilters(ensembl))
-# gb <- getBM(attributes=c('ensembl_exon_id', "exon_chrom_start", 
-#                          "exon_chrom_end"), filters=c("hgnc_symbol"), 
-#             values=Gn[[1]][[7]]$gene_name, mart=ensembl)
-C <- sapply(1:G, function(g) {
-  cbind(inorout=inorout[[g]], distance=distance[[g]], 
-        genelength=genelength[[g]], minoraf=minoraf[[g]])})
-
-### fitting models
-# ENIG model
-fit.enig <- enig(x, y, C, mult.lambda=FALSE, fixed.eb="none", init=NULL, 
-                 control=list(epsilon.eb=1e-3, epsilon.vb=1e-3, 
-                              epsilon.opt=sqrt(.Machine$double.eps), 
[…]

### 2020-04-22 code/simulations_gdsc.R

@@ -4,8 +4,7 @@
 ncores <- 100
 
 ### libraries
-packages <- c("foreach", "doParallel", "NIG", "statmod", "glmnet", 
-              "xtune", "rstan")
+packages <- c("foreach", "doParallel", "NIG", "statmod", "glmnet", "rstan")
 sapply(packages, library, character.only=TRUE)
 
 ### load and preprocess data
@@ -18,10 +17,8 @@ nreps <- 100
 ################################# simulation 1 #################################
 ################################################################################
 ### simulation settings
-# D <- 100
 D <- length(resp)
 p <- 100
-# n <- 100
 n <- floor(nrow(expr$expr))/2
 ntest <- nrow(expr$expr) - n
 alphaf <- c(1, 1, 3, 7)
@@ -34,12 +31,10 @@ x <- scale(expr$expr[, order(-apply(expr$expr, 2, sd))[1:p]])
 
 # estimation settings
 nfolds <- 10
-methods <- c("NIG-", "NIG", "ridge", "lasso", "xtune", "ebridge")
-control.semnig <- list(conv.post=TRUE, trace=FALSE, epsilon.eb=1e-5, 
-                       epsilon.vb=1e-3, maxit.eb=500, maxit.vb=1, 
-                       maxit.post=100, maxit.block=0)
-control.ebridge <-list(epsilon=sqrt(.Machine$double.eps), maxit=500, 
-                       trace=FALSE, glmnet.fit2=FALSE, beta2=FALSE)
+control <- list(conv.post=TRUE, trace=FALSE, epsilon.eb=1e-5, 
+                epsilon.vb=1e-3, maxit.eb=500, maxit.vb=1, 
+                maxit.post=100, maxit.block=0)
+
 
 # setup cluster
 cl <- makeCluster(ncores)
@@ -79,40 +74,27 @@ res <- foreach(r=1:nreps, .packages=packages) %dopar% {
                      rep(1:(n %% nfolds), (n %% nfolds)!=0)))
 
   # fitting models
-  fit.semnig1 <- semnig(x=rep(list(xtrain), D), y=ytrain, 
+  fit.nig1 <- nig(x=rep(list(xtrain), D), y=ytrain, 
                         C=replicate(D, list(matrix(rep(1, p)))), 
-                        Z=NULL, full.post=TRUE, control=control.semnig)
-  fit.semnig2 <- semnig(x=rep(list(xtrain), D), y=ytrain, C=C, Z=NULL, 
[…]

### 2020-04-22 code/tryout.R

@@ -1,37 +1,0 @@
-### MM algorithm
-# libraries
-library(cambridge)
-
-# data generation
-set.seed(123)
-D <- 100
-n <- rep(100, D)
-p <- rep(200, D)
-x <- matrix(rnorm(n[1]*p[1]), nrow=n[1], ncol=p[1])
-rownames(x) <- 1:nrow(x)
-alpha <- c(1, 1)
-C <- lapply(1:D, function(d) {
-  cbind(matrix(rnorm(p[d])))})
-Z <- cbind(matrix(rnorm(D)))
-G <- ncol(C[[1]])
-H <- ncol(Z)
-lambda <- rep(1, D)
-beta <- lapply(1:D, function(d) {
-  rnorm(p[d], 0, lambda[d]^2*exp(colSums(t(
-    cbind(C[[d]], matrix(rep(Z[d, ], each=p[d]), ncol=H)))*alpha)))})
-y <- lapply(1:D, function(d) {
-  vec <- rnorm(n[d], x %*% beta[[d]], 1);
-  names(vec) <- 1:n[d];
-  return(vec)})
-
-# estimation
-test.ebridge <- ebridge2(x, y, C, Z)
-test.ebridge$alphad
-
-vprior <- lapply(1:D, function(d) {
-  test.ebridge$lambda[d]^2*exp(
-    colSums(t(C[[d]])*test.ebridge$alphaf) + sum(Z[d, ]*test.ebridge$alphaf))})
-plot(Z, sapply(vprior, function(s) {mean(log(s))}))
-plot(unlist(C), log(unlist(vprior)))
-hist(log(unlist(vprior)))
-  

### 2020-04-22 rpackage/R/main_functions.R

@@ -1,23 +1,11 @@
-# x=rep(list(xtrain), D); y=ytrain
-# unpenalized=NULL; standardize=FALSE; 
-# intercept=FALSE; fixed.eb="none"; var.scale=1;
-# full.post=FALSE; init=NULL;
-# library(Rcpp)
-# sourceCpp("rpackage/src/aux_functions.cpp")
-# source("rpackage/R/vb_functions.R")
-# source("rpackage/R/aux_functions.R")
-# source("rpackage/R/eb_functions.R")
-# detach("package:cambridge", unload=TRUE)
-# library(cambridge)
-
-# estimate NIG model (not tested)
-semnig <- function(x, y, C, Z, unpenalized=NULL, standardize=FALSE, 
-                   intercept=FALSE, fixed.eb="none", var.scale=1,
-                   full.post=FALSE, init=NULL,
-                   control=list(conv.post=TRUE, trace=TRUE,
-                                epsilon.eb=1e-3, epsilon.vb=1e-3,
-                                maxit.eb=100, maxit.vb=2, maxit.post=100,
-                                maxit.block=0)) {
+# estimate NIG model
+nig <- function(x, y, C, Z, unpenalized=NULL, standardize=FALSE, 
+                intercept=FALSE, fixed.eb="none", var.scale=1,
+                full.post=FALSE, init=NULL,
+                control=list(conv.post=TRUE, trace=TRUE,
+                             epsilon.eb=1e-3, epsilon.vb=1e-3,
+                             maxit.eb=100, maxit.vb=2, maxit.post=100,
+                             maxit.block=0)) {
   
   # save the arguments
   cl <- match.call()
@@ -362,7 +350,7 @@ semnig <- function(x, y, C, Z, unpenalized=NULL, standardize=FALSE,
               conv=list(eb=conv.eb, vb=conv.vb),
               iter=list(eb=iter.eb, vb=iter.vb),
               time=c(eb.time=unname(eb.time), post.time=unname(post.time)))
-  class(out) <- c("semnig")
+  class(out) <- c("nig")
   return(out)
   
 }
@@ -396,7 +384,7 @@ new.elbo <- function(object, newx, newy) {
[…]

### 2020-04-22 rpackage/R/mcmc_functions.R

@@ -1,247 +1,0 @@
-.draw_beta <- function(par, x, y, n, p) {
-  gamma <- par$gamma
-  sigma <- par$sigma
-  tau <- par$tau
-  u <- rnorm(p, 0, tau*sigma*gamma)
-  delta <- rnorm(n, 0, sigma)
-  beta <- u + (t(x)*gamma^2) %*% solve(x %*% (t(x)*gamma^2) + diag(n)/tau^2) %*%
-    (y - x %*% u - delta)  
-  return(as.numeric(beta))
-}
-
-.draw_gamma <- function(par, C, p) {
-  alphaf <- par$alphaf
-  lambdaf <- par$lambdaf
-  beta <- par$beta
-  sigma <- par$sigma
-  tau <- par$tau
-  gamma <- sapply(1:p, function(j) {
-    chi <- max(beta[j]^2/(sigma*tau)^2 + lambdaf, sqrt(.Machine$double.eps))
-    psi <- max(lambdaf*as.numeric(C %*% alphaf)[j]^2, sqrt(.Machine$double.eps))
-    sqrt(rgig(n=1, chi=chi, psi=psi, lambda=-1))})
-  return(gamma)
-}
-
-.draw_tau <- function(par, Z, p) {
-  alphad <- par$alphad
-  lambdad <- par$lambdad
-  beta <- par$beta
-  sigma <- par$sigma
-  gamma <- par$gamma
-  chi <- max(sum(beta^2/(gamma*sigma)^2) + lambdad, sqrt(.Machine$double.eps))
-  psi <- max(lambdad*as.numeric(Z %*% alphad)^2, sqrt(.Machine$double.eps))
-  tau <- sqrt(rgig(1, chi=chi, psi=psi, lambda=-(p + 1)/2))
-  return(tau)
-}
-
-.draw_sigma <- function(par, x, y, n, p) {
-  beta <- par$beta
-  gamma <- par$gamma
-  tau <- par$tau
-  sigma <- 1/sqrt(rgamma(1, (n + p + 1)/2, 
-                         0.5*(sum(y^2) -2*sum(y*as.numeric(x %*% beta)) + 
-                                sum(as.numeric(x %*% beta)^2) + 
-                                sum(beta^2/gamma^2)/tau^2)))
-  return(sigma)
-}
-
-.draw_alphaf <- function(par, Cmat, hyper) {
-  nu0 <- hyper$nu0[1]
[…]

### 2020-04-22 rpackage/R/vb_functions.R

@@ -1,4 +1,4 @@
-# single VB update in NIG model with unpenalized covariates (tested)
+# single VB update in NIG model with unpenalized covariates
 .single.vb.update.unp <- function(aold, bold, gold, Calphaf, Zalphad, lambdaf, 
                                   lambdad, y, xu, xr, yty, n, u, r, var.scale) {
   
@@ -33,7 +33,7 @@
   return(out)
 }
 
-# single VB update in NIG model (tested)
+# single VB update in NIG model
 .single.vb.update <- function(aold, bold, gold, Calphaf, Zalphad, lambdaf, 
                               lambdad, y, x, ytx, yty, n, p, var.scale) {
   

### 2020-04-22 rpackage/src/aux_functions.cpp

@@ -7,72 +7,7 @@
 using namespace Rcpp;
 using namespace arma;
 
-// function to optimise in EBridge with one x (not tested)
-// [[Rcpp::export(".f.optim.mat")]]
-double f_optim_mat(arma::vec alpha, arma::vec lambda, double nu, double zeta,
-                   arma::mat Cmat, arma::mat Z, arma::vec n, int p, int D,
-                   List idsel, int G, int H, List y, arma::mat x,
-                   arma::vec yty, bool Zpres) {
-  
-  arma::vec gamma = exp(0.5*Cmat*alpha.head(G));
-  arma::vec tau = ones<vec>(D);
-  if(Zpres) {
-    tau = exp(0.5*Z*alpha.tail(H));
-  }
-  
-  double out=0.0;
-  for(int d=0; d<D; d++) {
-    arma::rowvec mat1(n(d));
-    arma::mat mat2(n(d), n(d));
-    arma::uvec cidsel = idsel[d];
-    arma::mat cx = x.rows(cidsel - 1);
-    arma::vec cy = y[d];
-
-    mat2 = cx*trans(cx.each_row() % square(lambda(d)*tau(d)*
-      gamma.subvec(d*p, (d + 1)*p - 1).t()));
-    mat1 = cy.t()*mat2;
-    mat2.diag() += 1;
-    out += -0.5*real(log_det(mat2))  -
-      (n(d)/2 + nu)*log(zeta + 0.5*yty(d) - 0.5*as_scalar(mat1*cy) +
-      0.5*as_scalar(mat1*mat2.i()*mat1.t()));
-  }
-  return out;
-}
-
-// function to optimise in EBridge with multiple x (not tested)
-// [[Rcpp::export(".f.optim.list")]]
-double f_optim_list(arma::vec alpha, arma::vec lambda, double nu, double zeta,
-                    arma::mat Cmat, arma::mat Z, arma::vec n, arma::vec p, 
-                    int D, int G, int H, List y, List x, arma::vec yty,
-                    bool Zpres) {
-  
-  arma::vec gamma = exp(0.5*Cmat*alpha.head(G));
-  arma::vec tau = ones<vec>(D);
-  if(Zpres) {
-    tau = exp(0.5*Z*alpha.tail(H));
-  }
-  arma::vec cp(D + 1);
-  cp.subvec(1, D) = cumsum(p);
-  cp(0) = 0;
-  double out=0.0;
-  for(int d=0; d<D; d++) {
-    arma::rowvec mat1(n(d)); 
[…]

### 2020-04-22 rpackage/tests/testthat.R

@@ -1,4 +1,4 @@
 library(testthat)
-library(cambridge)
+library(NIG)
 
-test_check("cambridge")
+test_check("NIG")

### 2020-04-23 docs/manuscript.Rnw
commit before pull

@@
 	simulation results for Scenarios 1-4 for
-	the NIG model and four alternative methods: frequentist (i) 
-	lasso and (ii) ridge models, and EB version of the ridge model: (iii) a
-	multivariate Bayesian extension of \texttt{xtune} 
-	\cite[]{ni_bayesian_2019} named mxtune, as introduced in SM Section 
-	\ref{sm-sec:mxtune}. Additionally, SM Section \ref{sm-sec:simulations} 
+	the NIG model and the (i) frequentist 
+	lasso and (ii) ridge models. Additionally, SM Section \ref{sm-sec:simulations} 
 	contains a comparison of MCMC and VB posteriors in simulation Scenario 3.
@@
 # simulation 1
-res <- read.table("results/simulations_gdsc_res1.2.txt", row.names=NULL)
+res <- read.table("results/simulations_gdsc_res1.txt", row.names=NULL)
 temp <- res[, 1]
@@
 # simulation 2
-res <- read.table("results/simulations_gdsc_res2.2.txt", row.names=NULL)
+res <- read.table("results/simulations_gdsc_res2.txt", row.names=NULL)
 temp <- res[, 1]
@@
 # simulation 3
-res <- read.table("results/simulations_gdsc_res3.2.txt", row.names=NULL)
+res <- read.table("results/simulations_gdsc_res3.txt", row.names=NULL)
 temp <- res[, 1]
@@
 
-tab5.2 <- tab5[, c(4, 13, 14, 6:8, 10:12)]
-tabm5.2 <- tabm5[, c(4, 13, 14, 6:8, 10:12)]
+tab5.2 <- tab5[, c(4, 13:14, 10:12)]
+tabm5.2 <- tabm5[, c(4, 13:14, 10:12)]
 tab5.2 <- t(sapply(1:nrow(tab5.2), function(r) {
@@
 rownames(tab5.2) <- paste0("analysis ", 1:nrow(tab5.2))
-colnames(tab5.2) <- c("lasso", "lasso(25)", "lasso(100)", 
-                      "NIG$_{\\text{f}+\\text{d}}+$DSS(lasso)",
+colnames(tab5.2) <- c("lasso($\\sim$50)$^*$", "lasso(25)", "lasso(100)", 
+                      "NIG$_{\\text{f}+\\text{d}}+$DSS($\\sim$50)$^*$",
                       "NIG$_{\\text{f}+\\text{d}}+$DSS(25)",
-                      "NIG$_{\\text{f}+\\text{d}}+$DSS(100)",
-                      "NIG$_{\\text{f}+\\text{d}}^-+$DSS(lasso)",
-                      "NIG$_{\\text{f}+\\text{d}}^-+$DSS(25)",
-                      "NIG$_{\\text{f}+\\text{d}}^-+$DSS(100)")
+                      "NIG$_{\\text{f}+\\text{d}}+$DSS(100)")
 tab5.2 <- t(tab5.2)
+tab5.2 <- tab5.2[c(2, 1, 3, 5, 4, 6), ]
 
@@
 kableExtra::kable_styling(knitr::kable(
-  tab1, align="r", digits=2,
+  tab1, align="r", digits=3,
   caption="$\\bm{\\alpha}$ estimates from analysis 1. Empty cells correspond to fixed zero parameters.",
@@
 kableExtra::kable_styling(knitr::kable(
-  tab2, align="r", digits=2,
+  tab2, align="r", digits=3,
   caption="$\\bm{\\alpha}$ estimates from analysis 2. Empty cells correspond to fixed zero parameters.",
@@
 kableExtra::kable_styling(knitr::kable(
-  tab3, align="r", digits=2,
+  tab3, align="r", digits=3,
   caption="$\\bm{\\alpha}$ estimates from analysis 3. Empty cells correspond to fixed zero parameters.",
@@
 kableExtra::kable_styling(knitr::kable(
-  tab4.1, align="r", digits=2,
+  tab4.1, align="r", digits=3,
   caption="$\\bm{\\alpha}$ estimates from analysis 4. Empty cells correspond to fixed zero parameters.",
@@
 kableExtra::kable_styling(knitr::kable(
-  tab4.2, align="r", digits=2,
+  tab4.2, align="r", digits=3,
   caption="$\\bm{\\alpha}$ estimates from analysis 4. Empty cells correspond to fixed zero parameters \\textit{(continued)}.",
@@
   analyses, where we select either (approximately) the same number of features 
-  as lasso, 25, or 50 features. Table 
+  as lasso (about 50 in all analyses), 25, or 100 features. Table 
   \ref{tab:analysis_gdsc_pmse2} compares the resulting mean PMSE values. The
   Table shows that NIG$_{\text{f}+\text{d}}$ plus DSS outperforms lasso,
-  with the same number of features.
+  for all three numbers of selected features, in all analyses.
 
 <<analysis_gdsc_pmse2>>=  
-kableExtra::kable_styling(knitr::kable(
+kableExtra::add_footnote(kableExtra::kable_styling(knitr::kable(
   tab5.2, align="r",
-  caption="Mean (standard deviation) of cross-validated PMSE for GDSC data. Best performing model (per analysis) in bold.",
+  caption="Mean (standard deviation) of cross-validated PMSE for selection methods (number of selected features between parentheses) on GDSC data. Best performing model (per analysis) in bold.",
   format="latex", booktabs=TRUE, escape=FALSE), 
-  latex_options=c("HOLD_position"))
+  latex_options=c("HOLD_position")), 
+  label="cross validated lasso selects, on average, 42-56 features in all analyses (indicated with $\\sim$50).",
+  notation="symbol", threeparttable=TRUE, escape=FALSE)
 @

### 2020-04-23 docs/supplement.Rnw
commit before pull

@@
 	This document contains the Supplementary Material (SM) to the document 
-	`Drug sensitivity prediction with the normal inverse Gaussian prior 
+	`Drug sensitivity prediction with normal inverse Gaussian shrinkage 
 	informed by external data'. In the 
@@
 	Section \ref{sec:prior} describes the moments of the normal
-	inverse Gaussian (NIG) prior, Section \ref{sec:vbderivations} gives
+	inverse Gaussian (NIG) prior. Section \ref{sec:hyperparameters} describes
+	the hyperparameter settings to generate Figure \ref{md-fig:dens_kappa} in the
+	MD and Section \ref{sec:jeffreysprior} outlines the derivation to arrive at
+	the Jeffreys prior for the error variance in equation (\ref{md-eq:sigmaprior})
+	of the MD. Section \ref{sec:vbderivations} gives
 	the variational Bayes (BV) derivations that lead to the estimating equations 
@@
 	that the empirical Bayes (EB) scale estimates $\bm{\lambda}$ are non-negative.
-	The last Section \ref{sec:gibbssampler} introduces a Gibbs sampling scheme
-	to sample from the posterior corresponding to the NIG model.
+	Section \ref{sec:gibbssampler} introduces a Gibbs sampling scheme
+	to sample from the posterior corresponding to the full Bayesian version of the
+	NIG model. Finally, Sections \ref{sec:simulations} and \ref{sec:gdsc} contain
+	additional results to the simulations and GDSC data analyses in Sections
+	\ref{md-sec:simulations} and \ref{md-sec:gdsc} in the MD, respectively.
 
@@
   \section{GDSC data}\label{sec:gdsc}
-  the CPO for cell line $i$ and drug $d$ is calculated as 
-  $$
+<<analysis_gdsc>>=
+sigma <- 0.83
+cutoff <- 0.01
+prop <- round(pnorm( - sigma*sqrt(-log(2) - log(pi) - 2*log(sigma) - 
+                                    2*log(cutoff)), 0, sigma), 4)
+
+load("results/analysis_gdsc_cpo1.Rdata")
+prop1 <- round(mean(unlist(lcpo.nig2) < log(cutoff)), 4)
+load("results/analysis_gdsc_cpo2.Rdata")
+prop2 <- round(mean(unlist(lcpo.nig2) < log(cutoff)), 4)
+load("results/analysis_gdsc_cpo3.Rdata")
+prop3 <- round(mean(unlist(lcpo.nig2) < log(cutoff)), 4)
+load("results/analysis_gdsc_cpo4.Rdata")
+prop4 <- round(mean(unlist(lcpo.nig2) < log(cutoff)), 4)
+@
+  the CPO for cell line $i$ and drug $d$ is
+  \begin{align}\label{eq:cpo}
   \text{CPO}_{id} = p(y_{id} | \bm{y}_{-id}) = 
@@
   q(\sigma^2_d | \bm{y}_d) d\bm{\beta}_d \sigma_d^2\]^{-1}
-  $$ 
+  \end{align}
   and is the Bayesian version of the leave-one-out cross-validated likelihood.
-  \cite{congdon_bayesian_2005} suggests as outlier cut-off 0.01
+  \cite{congdon_bayesian_2005} suggests as outlier cut-off 0.01.
+  Figure \ref{fig:analysis_gdsc_cpo} displays these CPOs (calculated by 
+  numerical integration of (\ref{eq:cpo})) for the four Analyses.
+  For $y_{id} | y_{-id}$ following a Gaussian distribution with standard 
+  deviation 0.83 
+  (the average posterior mean of $\sigma^2_d$ from our analyses) this implies 
+  an expected proportion of outliers of about \Sexpr{prop}. In our analyses we 
+  find \Sexpr{prop1}, \Sexpr{prop2}, \Sexpr{prop3}, and \Sexpr{prop4}, for the
+  four analyses respectively. So about double the amount of expected outliers
+  under a Gaussian posterior predictive distribition. However, in reality,
+  due to posterior uncertainty of $\bm{\beta}_d$ and $\sigma^2_d$, the
+  predictive distribution is more heavy-tailed than the Gaussian and the 
+  corresponding expected proportion of outliers higher than \Sexpr{prop}. 
+  Unfortunately, calculation of this true expected proportion is infeasible 
+  due to intractibility of the posterior predictive distribution.
+<<analysis_gdsc_cpo, fig.cap="distribution of the CPO values for Analysis (a) 1, (b) 2, (c) 3, and (d) 4 in Section \\ref{md-sec:gdsc} of the MD, with outlier cutoff value 0.01 (red line).", out.width="100%">>=
+@
 

### 2020-04-23 code/figures.R

@@ -48,7 +48,7 @@ par(opar)
 
 # ---- simulations_gdsc_est1 ----
 suppressWarnings(suppressMessages(library(sp)))
-res <- read.table("results/simulations_gdsc_res1.2.txt", row.names=NULL)
+res <- read.table("results/simulations_gdsc_res1.txt", row.names=NULL)
 temp <- res[, 1]
 res <- as.matrix(res[, -1])
 rownames(res) <- temp
@@ -80,7 +80,7 @@ par(opar)
 
 # ---- simulations_gdsc_est2 ----
 suppressWarnings(suppressMessages(library(sp)))
-res2 <- read.table("results/simulations_gdsc_res2.2.txt", row.names=NULL)
+res <- read.table("results/simulations_gdsc_res2.txt", row.names=NULL)
 temp <- res[, 1]
 res <- as.matrix(res[, -1])
 rownames(res) <- temp
@@ -237,24 +237,30 @@ par(opar)
 
 # ---- simulations_gdsc_est4 ----
 suppressWarnings(suppressMessages(library(sp)))
-res <- read.table("results/simulations_gdsc_res4.txt", row.names=NULL)
-temp <- res[, 1]
-res <- as.matrix(res[, -1])
-rownames(res) <- temp
+res1 <- read.table("results/simulations_gdsc_res4.1.txt", row.names=NULL)
+temp <- res1[, 1]
+res1 <- as.matrix(res1[, -1])
+rownames(res1) <- temp
+
+res2 <- read.table("results/simulations_gdsc_res4.2.txt", row.names=NULL)
+temp <- res2[, 1]
+res2 <- as.matrix(res2[, -1])
+rownames(res2) <- temp
+
+res <- rbind(res1, res2)
 
 fracs <- c(1/10, 1/5, 1/3, 1/2, 2/3, 4/5, 1)
-phi <- lapply(fracs, function(q) {
+phi <- lapply(format(round(fracs, 2)), function(q) {
   cq <- paste0("frac", q, ".alphaf")
   apply(1/(sapply(0:3, function(s) {
     res[rownames(res)==paste0(cq, s), 2]}) %*% 
       t(cbind(1, rbind(0, diag(3))))), 2, mean)})
-chi <- lapply(fracs, function(q) {
+chi <- lapply(format(round(fracs, 2)), function(q) {
   cq <- paste0("frac", q, ".alphad")
[…]

### 2020-04-28 docs/manuscript.Rnw
changed README file

@@
   show that the VB approximation is quite good as compared to standard MCMC.
-<<simulations_gdsc_kurt, fig.cap="Simulation results for Scenario 3: mean estimated prior variances $\\V(\\beta_{jd})$ versus true values, with line of identity (dotted).", out.width="50%", fig.asp=1>>=
+<<simulations_gdsc_kurt, fig.cap="Simulation results for Scenario 3: mean estimated prior Kurtoses $\\mathcal{K}(\\beta_{jd})$ versus true values, with line of identity (dotted).", out.width="50%", fig.asp=1>>=
 @

### 2020-04-28 docs/supplement.Rnw
changed README file

@@
                     "NIG$_{\\text{f+d}}$", "ridge", "lasso")
+
+n_eff.mean <- round(mean(summary.mcmc2$summary[, "n_eff"]), 0)
+n_eff.min <- round(min(summary.mcmc2$summary[, "n_eff"]), 0)
+time.mean <- round((time.mcmc2*1000/n_eff.mean)/3600, 0)
+time.min <- round((time.mcmc2*1000/n_eff.min)/(7*24*3600), 0)
 @	
@@
   \cite[]{guo_rstan:_2018} (see \url{https://github.com/magnusmunch/NIG} for 
-  the implemented stan code).
+  the implemented stan code). The Figure shows that the posteriors for
+  $\bm{\phi}$ and $\bm{\chi}$ have unexpected shapes and seem to have explored
+  only a small part of the posterior space. This indication that
+  posterior inference is not reliable is confirmed by the average and minimum
+  effective sample sizes of \Sexpr{n_eff.mean} and \Sexpr{n_eff.min}over all 
+  parameters, respectively. For reliable inference,
+  \cite{kruschke_doing_2015} recommends an effective sample size of at least 
+  10 000. Inspection of the computation times and a quick calculation learns 
+  that we have to sample from the posterior
+  for at least \Sexpr{time.mean} hours to obtain an average effective sample 
+  size of 10 000. Moreover, to achieve a minimum effective sample size of 10 000 
+  for all parameters, the sampler has to be run for about \Sexpr{time.min}
+  weeks.
 <<simulations_gdsc_res5, fig.cap="Simulation results for Scenario 3: Samples from full Bayesian NIG posteriors (see Section \\ref{sec:gibbssampler}) with true value superimposed (red line) for (top row) group 1, (second row) group 2, (third row) group 3, and (bottom row) group 4 for both features and drugs, and (left column) $\\bm{\\phi}$ and (second column) $\\bm{\\chi}$.", out.width="100%">>=

### 2020-04-28 code/analysis_gdsc.R

@@ -18,6 +18,7 @@ ncores <- min(100, nfolds)
 load(file="data/data_gdsc_dat1.Rdata")
 
 ### functions
+# a function to perform the decoupling shrinkage and selection approach
 coef.dss <- function(object, psel) {
   best <- matrix(nrow=object$glmnet.fit$dim[1] + 1, ncol=length(psel))
   for(p in 1:length(psel)) {
@@ -159,7 +160,7 @@ res <- foreach(r=1:nfolds, .packages=packages, .errorhandling="pass") %dopar% {
   
   # create co-data
   C <- lapply(1:D, function(d) {
-    s <- cbind(1, log(pvalues[[d]])); colnames(s) <- c("intercept", "pvalue"); s})
+    s <- cbind(1, pvalues[[d]]); colnames(s) <- c("intercept", "pvalue"); s})
   
   # nig models
   cv.nig1 <- nig(x=xtrain, y=ytrain, 

### 2020-06-18 docs/manuscript.Rnw
removed Makevars files from Code_and_Data

@@
   (Figure \ref{fig:simulations_gdsc_est3}). We conjecture that this interplay
-  of drug and feature variance components is due to near-unidentifiability.e
-  r, the resultconsequences are limited, since theiances on the $\beta_{jd}$ level are unbiasleft ed 
+  of drug and feature variance components is due to near-unidentifiability. 
+  In any case, the consequences are limited, since the variances on the 
+  $\beta_{jd}$ level are left unbiased
   (Figure \ref{fig:simulations_gdsc_var}). If separately estimated, scale 
@@
                     "NIG$_{\\text{f}+\\text{d}}$")
-colnames(tab1) <- c("feature intercept", "$\\log p$-value", "drug intercept")
+colnames(tab1) <- c("feature intercept", "$p$-value", "drug intercept")
 
@@
                     "NIG$_{\\text{f}+\\text{d}}$")
-colnames(tab3) <- c("feature intercept", "$\\log p$-value", "drug intercept")
+colnames(tab3) <- c("feature intercept", "$p$-value", "drug intercept")
 
@@
 rownames(tab5.2) <- paste0("analysis ", 1:nrow(tab5.2))
-colnames(tab5.2) <- c("lasso($\\sim$50)$^*$", "lasso(25)", "lasso(100)", 
+colnames(tab5.2) <- c("lasso$_{\\lambda^{\\text{CV}}}$($\\sim$50)$^*$", 
+                      "lasso(25)", "lasso(100)", 
                       "NIG$_{\\text{f}+\\text{d}}+$DSS($\\sim$50)$^*$",
@@
 	  \item{Analysis 1 results in a positive additive effect of the mutation 
-	    $log p$-values on the prior $\bm{\beta}_d$ precisions
-	    (Table \ref{tab:analysis_gdsc_fit1}). More precisely, with doubling
-	    $p$-value, the prior $\bm{\beta}_d$ precision increases by 
-	    $\log(2)*\Sexpr{round(tab1[2, 2], 3)}$=\Sexpr{
-	    round(log(2)*tab1[2, 2], 3)}.
-	    This translates to more $\bm{\beta}_d$ shrinkage with increasing 
-	    $p$-value, as expected.}
+	    $p$-values on the prior $\bm{\beta}_d$ precisions
+	    (Table \ref{tab:analysis_gdsc_fit1}). 
+	    This translates to more $\bm{\beta}_d$ shrinkage towards zero with 
+	    increasing $p$-value, as expected.}
 	  \item{Analysis 2 shows that gene mutations are more predictive 
@@
 	    expressions.}
-	  \item{Analysis 3 indicates that CCLE $\log p$-values are positively related
-	    to prior precision (Table \ref{tab:analysis_gdsc_fit3}). In particular
-	    if the $p$-value doubles, the prior $\bm{\beta}_d$ precision increases by 
-	    $\log(2)*\Sexpr{round(tab3[2, 2], 3)}$=\Sexpr{
-	    round(log(2)*tab1[2, 2], 3)}, i.e., higher CCLE $p$-values results in 
+	  \item{Analysis 3 indicates that CCLE $p$-values are positively related
+	    to prior precision (Table \ref{tab:analysis_gdsc_fit3}), i.e., higher 
+	    CCLE $p$-values results in 
 	    more shrinkage of $\bm{\beta}_d$, as expected.}
@@
   4, while NIG$_{\text{f}+\text{d}}$ $\bm{\alpha}$ performs best in Analysis 2. 
-  In general however, all models perform very similar. The ranges of mean PMSE 
+  In general however, all models perform very similarly. The ranges of mean PMSE 
   over all methods are
@@
   latex_options=c("HOLD_position")), 
-  label="cross validated lasso selects, on average, 42-56 features in all analyses (indicated with $\\sim$50).",
+  label="lasso with cross validated $\\lambda$ selects, on average, 42-56 features in all analyses (indicated with $\\sim$50).",
   notation="symbol", threeparttable=TRUE, escape=FALSE)
@@
   \cite[]{leday_gene_2017,kpogbezan_empirical_2017} and \texttt{xtune}
-  \cite[]{zeng_incorporating_2020}. bSEM is designed to include
+  \cite[]{zeng_incorporating_2020}. Here, bSEM is designed to include
   only dichotomous external covariates, so is not applicable in most of the
-  applications and simulations that we have considered here. \texttt{xtune}
-  allows for the inclusion of multiple external feature covariates, but not on
-  the drug level, so also has limited applicability in our simulations and
-  applications.
+  applications and simulations that we have considered here. Inclusion of
+  multiple external features is allowed by \texttt{xtune}, but not on
+  the drug level, so \texttt{xtune} has limited applicability in our simulations 
+  and applications.
 
@@
 	such an external
-	covariate for the genes is genelength, where we suspect that longer genes are
+	covariate for the genes is gene length, where we suspect that longer genes are
 	harder to predict. The distance of the SNP to the gene is an example of an

### 2020-06-18 docs/supplement.Rnw
removed Makevars files from Code_and_Data

@@
 
+load(file="results/simulations_gdsc_res5.Rdata")
 n_eff.mean <- round(mean(summary.mcmc2$summary[, "n_eff"]), 0)
@@
   weeks.
-<<simulations_gdsc_res5, fig.cap="Simulation results for Scenario 3: Samples from full Bayesian NIG posteriors (see Section \\ref{sec:gibbssampler}) with true value superimposed (red line) for (top row) group 1, (second row) group 2, (third row) group 3, and (bottom row) group 4 for both features and drugs, and (left column) $\\bm{\\phi}$ and (second column) $\\bm{\\chi}$.", out.width="100%">>=
+<<simulations_gdsc_res5, fig.cap="Simulation results for Scenario 3: Samples from full Bayesian NIG posteriors (see Section \\ref{sec:gibbssampler}) with true value superimposed (red line) for (top row) group 1, (second row) group 2, (third row) group 3, and (bottom row) group 4 for both features and drugs, and (left column) $\\bm{\\phi}$ and (second column) $\\bm{\\chi}$.
+Note that the red line is not present in most of the Figures, indicating that the posterior is far of from the true value", out.width="100%">>=
 @

### 2020-06-18 code/analysis_ccle.R

@@ -1,11 +1,7 @@
 #!/usr/bin/env Rscript
 
-################################################################################
-# CCLE log IC50 values and gene expression analyses
-################################################################################
-
 ### libraries
-packages <- c("glmnet", "cambridge")
+packages <- c()
 sapply(packages, library, character.only=TRUE)
 
 ### load data
@@ -13,7 +9,9 @@ load(file="data/data_gdsc_dat1.Rdata")
 expr.gdsc <- expr$expr
 load(file="data/data_ccle_dat1.Rdata")
 
-################################## analysis 1 ##################################
+################################################################################
+################################### analysis ###################################
+################################################################################
 ### data preparation
 # reassign data
 resp <- resp.prep
@@ -53,65 +51,3 @@ pfisher <- apply(fisher, 2, function(s) {
 feat.prep <- cbind(chifisher=fisher[1, ], pfisher=pfisher, harmonic=harmonic)
 rownames(feat.prep) <- genenames
 write.table(feat.prep, file="results/analysis_ccle_res1.txt")
-
-################################## analysis 2 ##################################
-### data preparation
-# reassign data
-resp <- resp.prep
-expr <- expr.prep
-
-# select genes that are also in the GDSC data
-expr.sel <- expr[, colnames(expr) %in% colnames(expr.gdsc)]
-
-# select cell lines that are common to both response and expression
-expr.sel <- sapply(resp, function(s) {
-  log(expr.sel[rownames(expr.sel) %in% names(s), ] + 1)})
-resp.sel <- sapply(1:length(resp), function(d) {
-  resp[[d]][names(resp[[d]]) %in% rownames(expr.sel[[d]])]})
-
-# create training and test data
-set.seed(2020)
-idtrain <- sapply(resp.sel, function(s) {
-  sample(1:length(s), floor(length(s)/2))})
-ytrain <- sapply(1:length(resp.sel), function(d) {
-  resp.sel[[d]][idtrain[[d]]]})
-idconstresp <- which(sapply(ytrain, sd)==0)
[…]

### 2020-06-18 code/analysis_gdsc.R

@@ -8,12 +8,6 @@
 packages <- c("foreach", "doParallel", "glmnet", "NIG")
 sapply(packages, library, character.only=TRUE)
 
-# number of CV folds
-nfolds <- 10
-
-# number of cores to use
-ncores <- min(100, nfolds)
-
 ### load data
 load(file="data/data_gdsc_dat1.Rdata")
 
@@ -131,11 +125,11 @@ foldid <- lapply(1:D, function(d) {
 
 # setup cluster
 if(Sys.info()["sysname"]=="Windows") {
-  cl <- makeCluster(ncores, type="PSOCK")
+  cl <- makeCluster(min(ncores, nfolds), type="PSOCK")
 } else {
-  cl <- makeCluster(ncores, type="FORK")
+  cl <- makeCluster(min(ncores, nfolds), type="FORK")
 }
-if(ncores > 1) {
+if(min(ncores, nfolds) > 1) {
   registerDoParallel(cl)
 } else {
   registerDoSEQ()
@@ -325,11 +319,11 @@ foldid <- lapply(1:D, function(d) {
 
 # setup cluster
 if(Sys.info()["sysname"]=="Windows") {
-  cl <- makeCluster(ncores, type="PSOCK")
+  cl <- makeCluster(min(ncores, nfolds), type="PSOCK")
 } else {
-  cl <- makeCluster(ncores, type="FORK")
+  cl <- makeCluster(min(ncores, nfolds), type="FORK")
 }
-if(ncores > 1) {
+if(min(ncores, nfolds) > 1) {
   registerDoParallel(cl)
 } else {
   registerDoSEQ()
@@ -536,11 +530,11 @@ foldid <- lapply(1:D, function(d) {
 
 # setup cluster
 if(Sys.info()["sysname"]=="Windows") {
-  cl <- makeCluster(ncores, type="PSOCK")
+  cl <- makeCluster(min(ncores, nfolds), type="PSOCK")
 } else {
-  cl <- makeCluster(ncores, type="FORK")
+  cl <- makeCluster(min(ncores, nfolds), type="FORK")
 }
-if(ncores > 1) {
+if(min(ncores, nfolds) > 1) {
   registerDoParallel(cl)
 } else {
   registerDoSEQ()
@@ -734,11 +728,11 @@ foldid <- lapply(1:D, function(d) {
 
 # setup cluster
 if(Sys.info()["sysname"]=="Windows") {
[…]

### 2020-06-18 code/simulations_gdsc.R

@@ -1,17 +1,11 @@
 #!/usr/bin/env Rscript
 
-# number of cores to use
-ncores <- 100
-
 ### libraries
 packages <- c("foreach", "doParallel", "NIG", "statmod", "glmnet", "rstan")
 sapply(packages, library, character.only=TRUE)
 
 ### load and preprocess data
 load(file="data/data_gdsc_dat1.Rdata")
-
-# number of reps
-nreps <- 100
 
 ################################################################################
 ################################# simulation 1 #################################
@@ -37,8 +31,12 @@ control <- list(conv.post=TRUE, trace=FALSE, epsilon.eb=1e-5,
 
 
 # setup cluster
-cl <- makeCluster(ncores)
-if(ncores > 1) {
+if(Sys.info()["sysname"]=="Windows") {
+  cl <- makeCluster(min(ncores, nreps), type="PSOCK")
+} else {
+  cl <- makeCluster(min(ncores, nreps), type="FORK")
+}
+if(min(ncores, nreps) > 1) {
   registerDoParallel(cl)
 } else {
   registerDoSEQ()
@@ -165,8 +163,12 @@ control <- list(conv.post=TRUE, trace=FALSE, epsilon.eb=1e-5,
                 maxit.post=100, maxit.block=0)
 
 # setup cluster
-cl <- makeCluster(ncores)
-if(ncores > 1) {
+if(Sys.info()["sysname"]=="Windows") {
+  cl <- makeCluster(min(ncores, nreps), type="PSOCK")
+} else {
+  cl <- makeCluster(min(ncores, nreps), type="FORK")
+}
+if(min(ncores, nreps) > 1) {
   registerDoParallel(cl)
 } else {
   registerDoSEQ()
@@ -295,8 +297,12 @@ control <- list(conv.post=TRUE, trace=FALSE, epsilon.eb=1e-5,
                 maxit.post=100, maxit.block=0)
 
 # setup cluster
-cl <- makeCluster(ncores)
-if(ncores > 1) {
+if(Sys.info()["sysname"]=="Windows") {
+  cl <- makeCluster(min(ncores, nreps), type="PSOCK")
+} else {
+  cl <- makeCluster(min(ncores, nreps), type="FORK")
+}
+if(min(ncores, nreps) > 1) {
   registerDoParallel(cl)
[…]
